---
title: "Loss functions, and finding their bottom"
subtitle: "How 'pick the number that fits best' becomes 'make this curve as small as possible' — the idea papers write as argmin, and R writes as optimize()"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live cells below re-state their own data. Base R is all
# we need, so nothing to install.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```


## Where we're going

Almost every fitting method — every "find the best line," "estimate the best
parameter," "train the model" — is secretly the same task: **write down one number
that measures how wrong you are, then choose the settings that make that number as
small as possible.** That "how wrong" number is a **loss function**, and shrinking it
is **optimization**.

This lesson builds that whole loop from the ground up on tiny, hand-checkable numbers,
using base R. We'll see a loss as a *function* whose graph is a curve, learn the
one-line recipe "the bottom of the valley is where the slope is zero," get just enough
chain-rule to differentiate a squared term, and finish with the R functions that search
for the bottom when the algebra is too messy — `optimize()` and `optim()`. We stay in
**one dimension** (one knob to turn); tuning several knobs at once is a later module.

**Builds on:** [the derivative as a slope](../derivative-as-slope/lesson.qmd) — we lean
hard on "slope zero means momentarily flat," and we reuse its numeric `slope_at()`
tool; and [vectors & summation](../vectors-and-summation/lesson.qmd) — a loss is
usually a *sum* over your data points ($\sum$), one penalty per point; and
[mean, variance & covariance](../mean-variance-covariance/lesson.qmd) — the warm-up
uses squared deviations from the mean.

## Warm-up: do you remember?

Before the new stuff, three quick reps — answers from memory first, then check.

**1.** For `y <- c(4, 6, 8)`, what single number is $\sum_{i=1}^{n} y_i$, and what is
$n$? Say the summation in words, then give the value.

```{webr}
#| autorun: false
# warmup1-starter — edit and Run to check
y <- c(4, 6, 8)
```

::: {.callout-note collapse="true" title="Answer"}
"The sum from i equals 1 to n of y-sub-i" — add them all: $4 + 6 + 8 = 18$, with
$n = 3$. In R that's `sum(y)` and `length(y)`. A loss almost always has a $\sum$ in it,
so this is the move you'll repeat. Shaky? Revisit
[vectors & summation](../vectors-and-summation/lesson.qmd).
:::

**2.** For a U-shaped curve like $f(x) = x^2$, where along it is the slope exactly
**zero**, and what is special about that spot?

::: {.callout-note collapse="true" title="Answer"}
At the very **bottom** of the bowl (here $x = 0$). The curve is momentarily flat there
— neither uphill nor downhill. That "slope is zero at the bottom" fact is the entire
engine of this lesson. See
[the derivative as a slope](../derivative-as-slope/lesson.qmd).
:::

**3.** From [mean-variance-covariance](../mean-variance-covariance/lesson.qmd):
`y <- c(4, 6, 8)`. What does `sum((y - mean(y))^2)` measure — in words? Compute
it by hand.

::: {.callout-note collapse="true" title="Answer"}
It measures the **total squared deviation** from the mean — how spread out the data is
(unnormalized). The mean is $6$, so the squared deviations are $(4-6)^2 + (6-6)^2 +
(8-6)^2 = 4 + 0 + 4 = 8$. Divide by $n$ and you have the (population) variance; divide
by $n-1$ and you have R's `var()`. Shaky? Revisit
[mean-variance-covariance](../mean-variance-covariance/lesson.qmd).
:::

## A tiny example first

You measured the same thing **three times** and got $4$, $6$, and $8$ (say, three
readings of one plant's height in cm). You want to report **one** number `m` that best
stands in for all three. Which single number fits best?

To answer "fits best" you first need to say what *fitting badly* costs. A natural
choice: for a candidate `m`, look at how far each measurement sits from it, **square**
each gap (so overshooting and undershooting both count as positive cost), and add the
squares. Call that total the **loss** of `m`:

$$
L(m) = (4 - m)^2 + (6 - m)^2 + (8 - m)^2
$$

Let's just try a few candidates by hand:

- **$m = 4$:** $(4-4)^2 + (6-4)^2 + (8-4)^2 = 0 + 4 + 16 = 20$.
- **$m = 5$:** $(4-5)^2 + (6-5)^2 + (8-5)^2 = 1 + 1 + 9 = 11$.
- **$m = 6$:** $(4-6)^2 + (6-6)^2 + (8-6)^2 = 4 + 0 + 4 = \mathbf{8}$.
- **$m = 7$:** $(4-7)^2 + (6-7)^2 + (8-7)^2 = 9 + 1 + 1 = 11$.
- **$m = 8$:** $(4-8)^2 + (6-8)^2 + (8-8)^2 = 16 + 4 + 0 = 20$.

The loss falls, bottoms out at $m = 6$ (loss $8$), then climbs again — a **valley**.
The best single stand-in is $m = 6$. (No accident: that's the **mean** of $4, 6, 8$.
Squared-error loss is always smallest at the average.)

The same thing in R, so you can check every step:

```{r toy-example}
# The three measurements
y <- c(4, 6, 8)

# The loss of a candidate m: squared gaps, summed
L <- function(m) sum((y - m)^2)

# Try the same candidates we did by hand
sapply(c(4, 5, 6, 7, 8), L)   # 20 11 8 11 20 -> a valley, lowest at m = 6

L(6)          # 8  -> the smallest loss
mean(y)       # 6  -> the candidate that achieves it
```

## Getting a feel for it

**A loss function is a "how wrong am I?" meter with a dial.** Picture a single dial
labelled `m`. Turn it and a needle — the loss — swings: some settings read high (very
wrong), one setting reads lowest (least wrong). Optimization is just *turning the dial
to the reading you like least of all* — the smallest loss. The dial is the
**parameter** (the knob you get to choose); the needle is the **objective** (also
called loss or cost — three names, one idea: the number you're trying to shrink).
(Where the picture strains: a real dial you turn by feel; here we'll turn it with math,
because there are infinitely many settings and we want the exact best one.)

**Why square the gaps instead of just adding them?** Try plain gaps, $4 - m$ etc.: at
$m = 6$ they are $-2, 0, +2$, which **add to zero** — the overshoot and the undershoot
cancel, so raw gaps make every `m` near the middle look equally "free." Squaring kills
the cancelling (a negative squared is positive), so every miss adds real cost, and
bigger misses are punished disproportionately ($4$ costs $16$, not $4$). That's why the
loss has a genuine single low point instead of a flat useless line. (This is the same
"square so deviations stop cancelling" reason the variance squares its deviations.)

**What the loss's value feels like.** Our best loss is $8$ — not zero, because three
different readings can't *all* be matched by one number; $8$ is the irreducible
disagreement among them. Nudge `m` a little off $6$ and the loss barely rises at first
(the valley floor is nearly flat); nudge it a lot and the loss shoots up (the walls are
steep). "Flat at the bottom, steep on the sides" is exactly the slope story from the
derivative lesson — and it's what lets a slope *point us to* the bottom.

## What the notation is going to say

Two pieces of notation are coming. The first, **$L(m)$**, just names the curve we've
been computing: "the loss when the knob is set to $m$." The second, **$\arg\min$**,
names the *answer we want*: not the smallest loss itself, but the **setting of $m$ that
achieves it**. In our toy example the smallest loss is $8$, but the thing we report is
$m = 6$ — the argmin. And the recipe for finding it is the slope idea in words: *the
bottom of the valley is where the loss stops falling and hasn't yet started rising —
i.e. where its slope is zero* — so we'll take the derivative of $L$, set it to zero, and
solve for $m$.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $m$ | "em" | the **parameter** — the one knob/dial we get to choose | our candidate stand-in, e.g. $m = 6$ |
| $y_i$ | "why sub eye" | the $i$-th data point (in R: `y[i]`) | $y_1 = 4,\ y_2 = 6,\ y_3 = 8$ |
| $n$ | "en" | how many data points (in R: `length(y)`) | $3$ |
| $\sum_{i=1}^{n}$ | "the sum from i equals 1 to n" | add the term to its right, once per data point ([vectors & summation](../vectors-and-summation/lesson.qmd)) | $\sum_i (y_i-m)^2 = 8$ at $m=6$ |
| $L(m)$ | "ell of em" | the **loss** (objective / cost): how wrong the setting $m$ is, as one number | $L(6) = 8$ |
| $L'(m)$ | "ell prime of em" | the **slope** of the loss curve at $m$ — its derivative ([derivative as a slope](../derivative-as-slope/lesson.qmd)) | $L'(6) = 0$ |
| $\frac{d}{dm}$ | "dee by dee em" | "take the derivative with respect to $m$" — the same slope, written as an instruction | $\frac{d}{dm}L = 6m - 36$ |
| $\min_m L(m)$ | "min over em of ell of em" | the smallest **loss value** itself (NOT the setting that achieves it) | $8$ |
| $\arg\min_m L(m)$ | "arg min over em of ell of em" | the **value of $m$** that makes $L$ smallest (NOT the smallest loss itself) | $6$ |
| $\hat{m}$ | "em hat" | our estimate of the best $m$ from data — the hat always means "estimated from data" | $6$ |
| $\bar{y}$ | "why bar" | the mean (average) of the $y$ values (in R: `mean(y)`) | $6$ |

Note the split $\arg\min$ makes: $\min_m L(m)$ is the smallest **loss** ($8$);
$\arg\min_m L(m)$ is the **setting** that achieves it ($6$). Papers almost always want
the argmin — the best knob position.

## The formal version

**The loss, one term at a time.** For each data point $y_i$, the miss is $y_i - m$, and
its cost is that miss **squared**, $(y_i - m)^2$ (squared so undershoot and overshoot
both cost, and never cancel). Add one such cost per point — that's the $\sum$ — and you
have the whole loss:

$$
L(m) = \sum_{i=1}^{n} (y_i - m)^2
$$

Reading it aloud: "ell of em equals the sum, from i equals 1 to n, of (y-sub-i minus
em) squared" — or in everyday words: **"add up the squared gap between each measurement
and your guess $m$."** Mapping onto the toy at $m = 6$: the three squared gaps are
$(4-6)^2 = 4$, $(6-6)^2 = 0$, $(8-6)^2 = 4$, summing to $8$ — the same $8$ we got by
hand.

**The recipe: take the derivative, set it to zero, solve.** The bottom of the valley is
where the slope is zero (warm-up 2). "Slope of $L$" is its derivative $L'(m)$, so the
best $m$ is the one where $L'(m) = 0$. This three-word recipe — **differentiate, set to
zero, solve** — is the standard way to find a minimum by algebra, and it has a name we'll
reuse everywhere. To run it we need the derivative of $L$.

**Chain rule, lite (just enough to differentiate a square).** Two facts do the whole
job:

1. **The derivative of a sum is the sum of the derivatives** — differentiate each
   $(y_i - m)^2$ term on its own and add the results. (You already used "a sum splits
   into pieces" in [vectors & summation](../vectors-and-summation/lesson.qmd).)
2. **The derivative of (stuff)² is 2 · (stuff) · (derivative of stuff).** In symbols,
   the slope of $(\text{stuff})^2$ with respect to $m$ is $2\,(\text{stuff}) \cdot
   (\text{stuff})'$. Here "stuff" is $y_i - m$, whose slope with respect to $m$ is $-1$
   (as $m$ goes up by 1, $y_i - m$ goes *down* by 1). So each term's derivative is
   $2(y_i - m)(-1) = -2(y_i - m)$.

Putting the two facts together, one line at a time:

$$
\begin{aligned}
L'(m) &= \frac{d}{dm} \sum_{i=1}^{n} (y_i - m)^2 \\
      &= \sum_{i=1}^{n} \frac{d}{dm}(y_i - m)^2 \\
      &= \sum_{i=1}^{n} 2\,(y_i - m)\cdot(-1) \\
      &= -2 \sum_{i=1}^{n} (y_i - m)
\end{aligned}
$$

- Line 2: the derivative of a sum is the sum of the derivatives (fact 1) — pull the
  $\frac{d}{dm}$ inside the $\sum$.
- Line 3: the derivative of (stuff)² is $2\,(\text{stuff})\cdot(\text{stuff})'$ (fact 2),
  with stuff $= y_i - m$ and $(\text{stuff})' = -1$.
- Line 4: the constant $-2$ is the same in every term, so factor it out front of the
  sum (an ordinary arithmetic tidy-up).

**Set it to zero and read off the answer.** For our toy, $L'(m) = -2\big[(4-m) + (6-m)
+ (8-m)\big] = -2(18 - 3m) = 6m - 36$. Setting $6m - 36 = 0$ gives $m = 6$. That's the
argmin — and it equals $\bar y = \text{mean}(4,6,8) = 6$, matching the valley bottom we
found by trying candidates.

**Sanity-check the derivative numerically.** The slope should be zero at $m = 6$,
negative to the left (loss still falling), positive to the right (loss now rising). We
measure it with the exact `slope_at()` tool from the derivative lesson:

```{r sanity-check}
y <- c(4, 6, 8)
L <- function(m) sum((y - m)^2)

# The numeric slope tool from the derivative-as-slope lesson
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)

# Measured slope vs. our formula L'(m) = 6m - 36
rbind(
  measured = sapply(c(5, 6, 7), function(m) slope_at(L, m)),
  formula  = 6 * c(5, 6, 7) - 36
)
# both rows: -6, 0, 6  -> slope is 0 exactly at the bottom (m = 6)
```

## Explore it

```{webr}
#| autorun: true
#| fig-alt: "The loss curve L(m), a U-shaped parabola with its lowest point marked at m = 6 where the loss equals 8; a horizontal dashed tangent line touches the curve at the bottom to show the slope is zero there."
# explore-1 — LIVE: the loss as a CURVE, with its bottom marked.
# PREDICT: the flat dashed tangent sits at the lowest point. What m is that? Write it
# down, then run. TRY IT: change the data y and watch the bottom slide to the new mean.
y <- c(4, 6, 8)
L <- function(m) sum((y - m)^2)

m_best <- mean(y)                              # the argmin (bottom of the valley)
curve(sapply(x, L), from = 2, to = 10, lwd = 2, xlab = "m (the knob)",
      ylab = "L(m)  (how wrong)", main = sprintf("lowest at m = %.1f, loss = %.1f", m_best, L(m_best)))
points(m_best, L(m_best), pch = 16, cex = 1.5)
abline(h = L(m_best), col = "firebrick", lty = 2)   # flat line: slope 0 at the bottom
```

```{webr}
#| autorun: true
# explore-2 — LIVE: the recipe "derivative = 0" agrees with just searching the curve.
# By hand the derivative is L'(m) = 6m - 36, which is zero at m = 6. Let's confirm the
# slope is zero there, AND that R's search finds the same bottom.
# PREDICT: what will optimize() report as the minimizing m? Write it, then run.
y <- c(4, 6, 8)
L <- function(m) sum((y - m)^2)
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)

by_recipe <- 6           # solve 6m - 36 = 0  -> m = 6
by_search <- optimize(L, interval = c(0, 20))$minimum   # let R hunt for the bottom

c(slope_at_recipe = slope_at(L, by_recipe),  # ~0: the recipe's m sits at the flat bottom
  recipe_m        = by_recipe,
  search_m        = by_search)               # search lands on the same m
```

```{webr}
#| autorun: false
# explore-3 — LIVE: when the algebra is ugly, let the computer search downhill.
# optimize() needs only the loss function and a range to look in; it repeatedly tries
# values, feeling which way is downhill, until it settles at the bottom. You never write
# a derivative. TRY IT: swap in a loss with no tidy formula and it still works.
weird <- function(m) sum((c(4, 6, 8) - m)^2) + 0.3 * (m - 5)^4   # no clean "set to zero"
optimize(weird, interval = c(0, 20))$minimum    # R finds the bottom anyway

# optim() is the same idea for the general case; here in 1-D with the Brent method:
optim(par = 0, fn = weird, method = "Brent", lower = 0, upper = 20)$par
```

## Common traps

**Trap 1 — "argmin is the smallest loss."** Tempting, because you *find* the argmin by
minimizing the loss. But $\arg\min$ is the **input** that minimizes, not the minimum
**output**. In our toy the smallest loss is $\min_m L(m) = 8$, while
$\arg\min_m L(m) = 6$ — two different numbers, $8$ vs $6$. When a paper writes
$\hat m = \arg\min_m L(m)$, the answer it wants is the setting ($6$), not the leftover
cost ($8$). Mixing them up reports the wrong quantity entirely.

**Trap 2 — "slope zero always means the lowest point."** Slope zero means *momentarily
flat* — which is true at a valley bottom, but also at a hill **top** and on a flat
shelf. "Differentiate, set to zero, solve" finds the flat spots; you still must check
you landed in a valley, not on a peak. Concrete example: try $g(m) = -(m-6)^2$. Its
derivative is $-2(m-6)$, which is zero at $m = 6$ — but $g(6) = 0$ is the **largest**
value, not the smallest ($g(4) = -4$ and $g(8) = -4$ are both lower). The zero-slope
point is a hilltop, not a valley. Our squared-error loss is a single clean bowl (only one
flat spot, and it's the bottom), so we're safe here — but that's a property of this loss,
not a free guarantee. It's why real optimizers also *search* and compare, as `optimize()`
does, rather than trusting a single zero-slope point blindly.

```{r trap2-check}
# Quick sanity check: g(m) = -(m-6)^2 has slope zero at 6, but that's a MAXIMUM
g <- function(m) -(m - 6)^2
optimize(g, c(0, 12), maximum = TRUE)$maximum   # 6 — the peak, not a valley
```

## Check yourself

**1.** For the measurements below, compute the loss $L(m) = \sum (y_i - m)^2$ at
$m = 3$ **by hand**, then check in R.

```{webr}
#| autorun: false
# check1-starter — edit and Run to check your hand answer
y <- c(1, 3, 5)
```

::: {.callout-note collapse="true" title="Answer"}
The squared gaps at $m = 3$: $(1-3)^2 + (3-3)^2 + (5-3)^2 = 4 + 0 + 4 = 8$. In R,
`sum((c(1,3,5) - 3)^2)` returns `8`. (And $m = 3$ is the mean of $1, 3, 5$, so this is
already the argmin — the smallest possible loss for this data.)
:::

**2.** The derivative of the loss for `y <- c(4, 6, 8)` is $L'(m) = 6m - 36$. Using
"set the derivative to zero and solve," what is the best $m$? Is that value the argmin
or the minimum loss?

::: {.callout-note collapse="true" title="Answer"}
Set $6m - 36 = 0 \Rightarrow 6m = 36 \Rightarrow m = 6$. That $6$ is the **argmin** —
the *setting* of the knob that minimizes the loss. The **minimum loss** is the separate
number $L(6) = 8$. (Don't report $8$ when you're asked for the best $m$.)
:::

**3.** In your own words: explain to a colleague, in two sentences, what a loss function
is and why "find where the slope is zero" helps you minimize it.

::: {.callout-note collapse="true" title="A model answer"}
A loss function turns a candidate setting into a single "how wrong is this?" number, and
minimizing it means turning the knob to the setting with the smallest wrongness. The
bottom of that loss curve is momentarily flat, so its slope is zero there — which is why
setting the derivative to zero and solving pins down the best setting.
:::

## What you can now do

- **Read a loss (objective / cost) function as a curve** with a knob $m$ on the x-axis
  and "how wrong" on the y-axis, and evaluate it in R as `sum((y - m)^2)`.
- **Explain and use the recipe "differentiate, set to zero, solve"** to find the setting
  where a loss curve bottoms out, because the slope is zero at the bottom.
- **Differentiate a squared term** with chain-rule-lite: $(\text{stuff})^2 \to
  2\,(\text{stuff})\cdot(\text{stuff})'$, and use "the derivative of a sum is the sum of
  the derivatives."
- **Tell an argmin from a minimum** — the best *setting* versus the smallest *loss* —
  and report the one the question asks for.
- **Hand a loss to `optimize()` / `optim()`** and let R search for the bottom when the
  algebra is too messy to solve by hand.

If any of these feels shaky, revisit [A tiny example first](#a-tiny-example-first) and
[The formal version](#the-formal-version) before continuing — anything that *fits* a
model by minimizing an error assumes all of these.

## Where this goes next

Every "line of best fit," "maximum-likelihood estimate," and "trained model" is this
same loop — write a loss, then minimize it — scaled up. Two extensions come later: when
there are **several** knobs at once, the single slope becomes a whole vector of slopes
(a gradient, its own module); and when you can't solve "slope = zero" by algebra, a
computer takes small downhill steps to hunt the bottom (the gradient-descent course
builds that from the compass you just met). Ready for more reps? Do the
[practice problems](practice.qmd).
