---
title: "Five hypotheses"
subtitle: "H0 through H4 as concrete statements about the region — and why PP3 and PP4 are the two you watch"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live {webr} cells below re-state their own data. Base R only.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

## Where we're going

`coloc.abf` hands you five numbers — $\text{PP0}, \text{PP1}, \text{PP2}, \text{PP3},
\text{PP4}$ — that sum to 1. Before we compute them (that's Module 3), you need to know
*what each one is the probability of*. This module makes each of the five hypotheses a plain
sentence about the region, introduces the idea of a **configuration** (which SNP is causal
for which trait), and pins down the two numbers you'll actually report to a collaborator:
**PP3** (two distinct causal variants) and **PP4** (one shared causal variant). It also
introduces the three prior weights $p_1, p_2, p_{12}$ that tilt the five before the data
speaks.

**Builds on:** [Module 1](../01-the-colocalisation-question/lesson.qmd) (the two lABF
vectors $\ell_1, \ell_2$), and the foundation
[priors-posteriors-and-updating](../../../../foundations/priors-posteriors-and-updating/lesson.qmd).

## Warm-up: do you remember?

Before the new stuff, three quick reps — answers from memory first, then check.

**1.** *(Module 1)* For a region of $Q$ SNPs, what are $\ell_1$ and $\ell_2$, and where do
they come from?

::: {.callout-note collapse="true" title="Answer"}
$\ell_1$ and $\ell_2$ are the per-SNP log Bayes factor vectors for trait 1 and trait 2 —
each SNP's summary statistics run through `wakefield_abf`. See
[Module 1](../01-the-colocalisation-question/lesson.qmd).
:::

**2.** *(Module 1)* If trait 1's loud SNP and trait 2's loud SNP are *different* SNPs, does
that look more like a shared or a distinct signal?

::: {.callout-note collapse="true" title="Answer"}
**Distinct** — the two studies peak on different variants, so they're likely two separate
causal SNPs sitting close together. See [Module 1](../01-the-colocalisation-question/lesson.qmd).
:::

**3.** *(foundation)* In Bayesian updating, before you see any data, what does a **prior**
give each competing hypothesis?

::: {.callout-note collapse="true" title="Answer"}
A prior gives each hypothesis a starting weight — how plausible it is *before* the data —
which then gets multiplied by how well the hypothesis explains the data. See
[priors-posteriors-and-updating](../../../../foundations/priors-posteriors-and-updating/lesson.qmd).
:::

## A tiny example first

Keep the three-SNP region from Module 1. There are exactly five stories the region could be
telling. Let's list them concretely for **SNP 1, SNP 2, SNP 3**, and count how many specific
"which SNP" arrangements each story allows — call each arrangement a **configuration**.

- **$H_0$ — nothing here.** Neither trait has a causal SNP in this region. Just one
  configuration (no causal SNP at all): **1**.
- **$H_1$ — trait 1 only.** Trait 1 has one causal SNP; trait 2 has none. The causal SNP
  could be SNP 1, 2, or 3: **3** configurations.
- **$H_2$ — trait 2 only.** Mirror image: trait 2 has one causal SNP, trait 1 none: **3**.
- **$H_3$ — both, but distinct.** Trait 1 is causal at one SNP, trait 2 at a *different* one.
  Trait 1 picks one of 3, trait 2 picks one of the *other* 2: $3 \times 2 = $ **6**
  configurations (e.g. "trait 1 at SNP 1, trait 2 at SNP 3").
- **$H_4$ — both, shared.** One single SNP is causal for *both* traits. That SNP is 1, 2, or
  3: **3** configurations (e.g. "SNP 2 for both").

Notice $H_3$ and $H_4$ both mean "both traits have a signal here" — the difference is only
whether it's the *same* SNP. That single distinction is the entire scientific payload of
colocalisation.

The same counting in R (it's just combinatorics, no statistics yet):

```{r toy-example}
Q <- 3                        # three SNPs in the region
c(H0 = 1,                     # no causal SNP anywhere
  H1 = Q,                     # trait-1 causal at one of Q SNPs
  H2 = Q,                     # trait-2 causal at one of Q SNPs
  H3 = Q * (Q - 1),           # distinct: t1 at one SNP, t2 at a DIFFERENT one
  H4 = Q)                     # shared: one SNP causal for BOTH traits
```

## Getting a feel for it

**Analogy — two guest lists, one party.** Think of each trait as a party that may or may not
have invited exactly one special guest (its causal SNP) from a room of $Q$ people. $H_0$:
neither party invited anyone. $H_1$/$H_2$: only one party invited a guest. Then the
interesting case, both parties invited a guest, splits in two: **$H_4$** is "both parties
invited the *same* person" (they share a friend — colocalisation); **$H_3$** is "each party
invited a *different* person" (they just both happen to know someone in that crowded room).
The whole test is a careful way of asking: given who lit up at each party, did they invite
the same guest?

Where the analogy breaks: a real party either did or didn't invite someone, but we never get
to see the guest list — we only see noisy "brightness" (the lABFs). So we can't declare a
winner; we can only spread probability across the five stories.

**Why exactly these five, and why split $H_3$ from $H_4$?** Because those are the only
mutually exclusive, exhaustive stories under the working assumption that each trait has *at
most one* causal SNP in the region. Lumping $H_3$ and $H_4$ together (as an older "do they
both have a signal?" test would) throws away the one thing you care about. Splitting them is
what lets the method say "yes both have signal, but it's *not* the same variant" — the
false-colocalisation guard.

**What the numbers feel like.** For $Q = 3$ the configuration counts are $1, 3, 3, 6, 3$.
$H_3$ (distinct) has the *most* configurations — 6 ways to pick two different SNPs — so if
the data were totally uninformative and priors were equal *per configuration*, $H_3$ would be
favoured just by having more ways to happen. That's a hint of why the *priors* matter: coloc
deliberately makes sharing ($p_{12}$) rarer per SNP than a single-trait signal, to keep the
"more configurations" bookkeeping honest. We'll see the exact weights next.

## What the notation is going to say

The notation will attach a **prior weight** to each configuration, using three per-SNP
numbers:

- $p_1$ = the prior probability that any given SNP is causal for **trait 1 only**,
- $p_2$ = the same for **trait 2 only**,
- $p_{12}$ = the prior probability that a given SNP is causal for **both** traits.

So the prior for hypothesis $H_1$ (trait-1-only, one SNP out of $Q$) is built from $p_1$; for
$H_4$ (shared, one SNP) from $p_{12}$; and for $H_3$ (distinct) from $p_1 \times p_2$ (one SNP
causal for trait 1, a different one for trait 2). coloc's defaults are $p_1 = p_2 = 10^{-4}$
and $p_{12} = 10^{-5}$ — sharing is deliberately an order of magnitude rarer than a
single-trait hit.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $H_0 \dots H_4$ | "aitch zero … aitch four" | the five hypotheses about the region | the five stories above |
| $\text{PP}_k$ | "pee-pee sub kay" | posterior probability of hypothesis $H_k$ (they sum to 1) | computed in Module 3 |
| $Q$ | "cue" | number of SNPs in the region | 3 |
| $p_1$ | "pee one" | prior prob. a SNP is causal for trait 1 only | coloc default $10^{-4}$ |
| $p_2$ | "pee two" | prior prob. a SNP is causal for trait 2 only | coloc default $10^{-4}$ |
| $p_{12}$ | "pee one-two" | prior prob. a SNP is causal for both traits | coloc default $10^{-5}$ |
| configuration | — | one specific "which SNP is causal for which trait" arrangement | e.g. "SNP 2 causal for both" |

## The formal version

No algebra to derive yet — the "formal version" here is the exact statement of the five
hypotheses and the two you report. Under the one-causal-per-trait assumption:

$$
\begin{aligned}
H_0 &: \text{no causal SNP for either trait} \\
H_1 &: \text{a causal SNP for trait 1 only} \\
H_2 &: \text{a causal SNP for trait 2 only} \\
H_3 &: \text{causal SNPs for both traits, at \textbf{distinct} SNPs} \\
H_4 &: \text{one \textbf{shared} causal SNP for both traits}
\end{aligned}
$$

Reading it aloud: "$H_3$ says both traits have a causal SNP here but they're different SNPs;
$H_4$ says both traits are driven by the very same SNP." In everyday words: **PP4 large →
colocalisation (shared variant); PP3 large → two separate variants that only look joined.**

The two you report: after the data speaks, you get $\text{PP0}\dots\text{PP4}$ summing to 1.
A high **PP4** (coloc convention often uses $> 0.8$) is the evidence *for* a shared causal
variant; a high **PP3** is evidence *against* — two distinct variants. The other three
($\text{PP0}, \text{PP1}, \text{PP2}$) mostly tell you the region lacked the joint
signal to decide.

## Explore it

```{webr}
#| autorun: true
# TRY IT: change Q (the number of SNPs) and watch the configuration counts.
# PREDICT: which hypothesis's count grows FASTEST as Q gets big -- H1, H4, or H3?
Q <- 3
counts <- c(H0 = 1, H1 = Q, H2 = Q, H3 = Q * (Q - 1), H4 = Q)
counts
# H3 (distinct) grows like Q^2 -- there are far more ways to pick TWO different SNPs
# than one shared SNP. That is why coloc makes p12 smaller per SNP than p1, p2.
```

```{webr}
#| autorun: true
#| fig-alt: "A bar chart of the five hypotheses' configuration counts for a three-SNP region: H0=1, H1=3, H2=3, H3=6, H4=3. The H3 (distinct) bar is the tallest."
# LIVE: SEE why 'distinct' has the most ways to happen in a small region.
# PREDICT: for Q = 3, which bar is tallest -- H3 (distinct) or H4 (shared)?
Q <- 3
counts <- c(H0 = 1, H1 = Q, H2 = Q, H3 = Q * (Q - 1), H4 = Q)
barplot(counts, ylab = "number of configurations",
        main = paste0("How many ways can each story happen? (Q = ", Q, ")"),
        col = c("grey", "grey", "grey", "tomato", "steelblue"))
legend("topleft", fill = c("tomato", "steelblue"),
       legend = c("H3 = distinct (PP3)", "H4 = shared (PP4)"), bty = "n")
```

## Common traps

**Trap: "a high PP3 means the method failed / found nothing."** No — PP3 is a genuine,
informative result. It says: *both* traits really do have a causal variant in this region,
and the method is confident they are **different** variants. That is often the most important
finding (it stops you from wrongly claiming a shared mechanism). "Found nothing" is PP0
being high, not PP3.

**Trap: "PP4 near 0.5 is decent evidence for sharing."** Not really. Because there are five
hypotheses, the neutral baseline isn't 0.5 — the standard threshold people use is
**PP4 $> 0.8$** (and often they also want PP4/(PP3+PP4) high). A PP4 of 0.5 usually means the
data can't distinguish shared from distinct, not that sharing is "more likely than not" in
any useful sense. Report the number, not a coin-flip conclusion.

## Check yourself

**1.** A region has $Q = 4$ SNPs. How many **distinct** ($H_3$) configurations are there, and
how many **shared** ($H_4$) configurations?

```{webr}
#| autorun: false
# check1-starter — edit and Run
Q <- 4
```

::: {.callout-note collapse="true" title="Answer"}
$H_3$: $Q(Q-1) = 4 \times 3 = 12$ ways to put trait 1 at one SNP and trait 2 at a different
one. $H_4$: $Q = 4$ ways to pick one shared SNP. Distinct always outnumbers shared, and by
more as $Q$ grows.
:::

**2.** You run `coloc.abf` and get $\text{PP3} = 0.92$. In one sentence, what do you tell
your collaborator?

::: {.callout-note collapse="true" title="Answer"}
"Both traits have a real association signal in this region, but the method is 92% confident
they're driven by **two different** causal variants, not a shared one — so don't assume a
common mechanism." (High PP3 = distinct, not "no signal.")
:::

**3.** In your own words: explain to a colleague, in two sentences, the difference between
$H_3$ and $H_4$ and why coloc bothers to separate them.

::: {.callout-note collapse="true" title="A model answer"}
$H_4$ says both traits are driven by the *same* causal SNP (they colocalise), while $H_3$ says
each trait has its own *different* causal SNP that just sits nearby. coloc separates them
because "both have a signal here" is not the same as "they share a mechanism" — and only the
shared case (high PP4) supports a common causal variant.
:::

## What you can now do

- **State each of $H_0$–$H_4$** as a concrete sentence about a region, and count the
  configurations each allows for a small $Q$.
- **Explain why PP3 (distinct) and PP4 (shared) are the two numbers you report**, and why the
  usual bar for claiming colocalisation is PP4 $> 0.8$, not $0.5$.
- **Read the priors $p_1, p_2, p_{12}$** as per-SNP weights, and say why coloc makes sharing
  ($p_{12}$) rarer than a single-trait hit.

If any of these feels shaky, revisit the five-story list above before continuing — Module 3
turns each story into a number, and every formula there is named after one of these
hypotheses.

## Where this goes next

You can now say what the five probabilities *mean*. [Module 3](../03-combining-abfs-with-priors/lesson.qmd)
computes them: it combines $\ell_1$, $\ell_2$, and the priors $p_1, p_2, p_{12}$ into
$\ell_{H_0}\dots\ell_{H_4}$ (with the clever $\sum_{i\neq j}$ trick for the distinct case),
normalises to $\text{PP0}\dots\text{PP4}$, and builds `coloc_abf()`. Ready for reps? Do the
[practice problems](practice.qmd).
