---
title: "Practice: Five hypotheses"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

Work these in order — they get harder. Do them on paper or in an R session; only then open
the answer.

## Problem 1 — the toy example, new numbers

For a region with $Q = 5$ SNPs, count the configurations of each of the five hypotheses (as in
the lesson's $Q = 3$ example).

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve
Q <- 5
```

::: {.callout-note collapse="true" title="Answer"}
```{r}
Q <- 5
c(H0 = 1, H1 = Q, H2 = Q, H3 = Q*(Q-1), H4 = Q)
```
So $H_0 = 1$, $H_1 = 5$, $H_2 = 5$, $H_3 = 5\times4 = 20$, $H_4 = 5$. Distinct ($H_3$) has by
far the most ways to happen — 20 — because there are many ways to pick *two different* SNPs.
:::

## Problem 2 — fill in the missing step

You're describing the five hypotheses for a region. Fill in the blank statement.

- $H_0$: no causal SNP for either trait.
- $H_1$: a causal SNP for trait 1 only.
- $H_2$: a causal SNP for trait 2 only.
- $H_3$: **your step here** — state the "both, but…" hypothesis precisely.
- $H_4$: one shared causal SNP for both traits.

::: {.callout-note collapse="true" title="Answer"}
$H_3$: **both traits have a causal SNP in the region, but at two *distinct* (different)
SNPs.** The whole scientific point of coloc is separating this from $H_4$ (same SNP): "both
associated" is not the same as "shared mechanism." $H_3$ is what you report as PP3.
:::

## Problem 3 — interpret a result

`coloc.abf` returns these five posteriors for a region. In one or two sentences, what do you
conclude, and which single number did you base it on?

```{webr}
#| autorun: false
# p3-starter — read these off; no computation needed
pp <- c(PP0 = 0.01, PP1 = 0.02, PP2 = 0.01, PP3 = 0.05, PP4 = 0.91)
```

::: {.callout-note collapse="true" title="Answer"}
**PP4 = 0.91** is well above the usual 0.8 bar, so the data support a **shared causal
variant** — the two traits colocalise. PP3 (distinct) is only 0.05, and the "no/one-trait
signal" hypotheses are tiny, so this is a clean colocalisation call. (You'd still add the
"correlation, not proven causation" caveat when reporting it.)
:::

## Problem 4 — spot the error

A colleague summarises a run like this — it contains **exactly one mistake** in the
*reasoning*. Find it.

> "We got PP3 = 0.88. That means the method couldn't find any signal in the region, so there's
> nothing here for either trait."

::: {.callout-note collapse="true" title="Answer"}
The mistake: **PP3 = 0.88 is not "no signal."** A high PP3 says *both* traits have a real
causal variant in the region — the method is confident there are **two distinct** causal SNPs.
"No signal at all" would be a high **PP0**, not PP3. The correct summary: "both traits are
associated here, but by different variants, so don't assume a shared mechanism."
:::

## Problem 5 — mixing it with Module 1 *(uses Module 1's lABF fingerprint reading)*

Here are two lABF vectors for a three-SNP region (from Module 1's machine). Based on the loud
SNPs, which *hypothesis* ($H_3$ or $H_4$) do you expect the data to favour, and why?

```{webr}
#| autorun: false
# p5-starter — eyeball the loud SNPs, then name the hypothesis
l1 <- c(-1.1, -0.9, 6.4)   # trait 1
l2 <- c(6.0, -1.2, -1.0)   # trait 2
```

::: {.callout-note collapse="true" title="Answer"}
Trait 1 is loud on **SNP 3**; trait 2 is loud on **SNP 1**. The loud SNPs are *different*, so
the fingerprint (Module 1) points to **distinct** causal variants → expect the data to favour
**$H_3$** (high PP3). If both had been loud on the same SNP, you'd expect $H_4$ (PP4) instead.
Module 3 turns this eyeballed guess into the actual numbers.
:::

## Problem 6 — transfer / cliffhanger

coloc's default priors are $p_1 = p_2 = 10^{-4}$ and $p_{12} = 10^{-5}$. In one sentence, why
would raising $p_{12}$ (say to $10^{-4}$, equal to $p_1$) tend to *increase* PP4 for a given
region? You can reason it out even before Module 3's formulas.

::: {.callout-note collapse="true" title="Answer"}
$p_{12}$ is the prior weight on the *shared* hypothesis $H_4$. Since each hypothesis's posterior
is (prior $\times$ evidence), a bigger $p_{12}$ multiplies $H_4$'s support up before
normalising, so PP4 rises (and the others fall to compensate). Module 3 shows this exactly:
$\ell_{H_4} = \log p_{12} + s_{12}$, so raising $p_{12}$ adds a constant to $H_4$'s log
support. This is also why colocalisation results should be checked for prior sensitivity.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**, once the
material isn't fresh. A pass then is real evidence you've retained it. Five questions, graded on
the page; 4/5 or better means this module is solid. Score under 4? Revisit what the missed
questions point at, then retake in a **later session** — passing after another gap is the
evidence that counts. Results are saved in this browser only.

::: {.module-check module="courses/bayesian-colocalisation/modules/02-five-hypotheses" pass="4"}

::: {.quiz-q qid="fh-h3-vs-h4" qtype="mc" answer="c"}
What is the difference between $H_3$ and $H_4$?

a) $H_3$ means no signal; $H_4$ means a signal
b) $H_3$ means trait 1 only; $H_4$ means trait 2 only
c) $H_3$ means both traits are causal at distinct SNPs; $H_4$ means they share one SNP
d) $H_3$ uses a larger prior than $H_4$
:::

::: {.quiz-q qid="fh-h3-configs" qtype="numeric" answer="20" tolerance="0"}
For a region of $Q = 5$ SNPs, how many configurations does $H_3$ (distinct) have?
:::

::: {.quiz-q qid="fh-high-pp3" qtype="mc" answer="b"}
You get PP3 = 0.90. What do you conclude?

a) There is no causal variant in the region for either trait
b) Both traits have a causal variant here, but they are two distinct variants
c) The two traits share a single causal variant
d) Only trait 1 has a causal variant
:::

::: {.quiz-q qid="fh-p12-default" qtype="mc" answer="a"}
Why does coloc set $p_{12}$ smaller than $p_1$ and $p_2$?

a) A SNP being causal for both traits is rarer than being causal for just one
b) Sharing configurations vastly outnumber distinct ones
c) $p_{12}$ must sum to 1 with $p_1$ and $p_2$
d) It makes PP4 larger by default
:::

::: {.quiz-q qid="fh-labf-fingerprint" qtype="mc" answer="b"}
*(interleaved — from Module 1.)* Trait 1's lABF is loud on SNP 2 and trait 2's lABF is loud on
SNP 4. Which hypothesis does this fingerprint point toward?

a) $H_4$ (shared)
b) $H_3$ (distinct)
c) $H_0$ (no signal)
d) $H_1$ (trait 1 only)
:::

:::

::::
