---
title: "When fine-mapping breaks"
subtitle: "Why unequal statistical information across SNPs quietly biases fine-mapping — and gets worse as your study grows"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live cells below re-state their own data. Base R only.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

::: {.callout-tip title="This lesson runs in your browser" collapse="true"}
The code cells below with a **Run** button are live — powered by WebR (R compiled to
WebAssembly), they run right here on your phone or laptop, no install needed. Edit any of them
and re-run. The first run downloads a small R engine, so give it a few seconds. Cells without a
Run button are ordinary listings whose output was computed when the page was built.
:::

## Where we're going

Before we build a new colocalisation test, we need to understand the *disease* it cures. This
module is the first half of the paper's diagnosis. Fine-mapping — deciding which SNP in a region
is the causal one — works by giving each SNP a Bayes factor and normalising those across SNPs
into "probability this SNP is the causal one." That normalisation quietly assumes every SNP's
Bayes factor is measured on the same footing. When SNPs carry **different amounts of statistical
information** (different sample size, imputation quality, call rate, or even just rounding), that
assumption fails — and the failure gets *worse*, not better, as your study grows. By the end
you'll reproduce the paper's Figure 1 intuition with a few lines of base R.

**Builds on:** the [Approximate Bayes Factors course](../../../approximate-bayes-factors/syllabus.md)
(which built `wakefield_abf(beta, se, W)` — one SNP's summary stats → a log Bayes factor) and the
[Bayesian Colocalisation course](../../../bayesian-colocalisation/syllabus.md) (which normalised
per-SNP Bayes factors into posteriors); plus the foundation
[distributions & densities](../../../../foundations/distributions-and-densities/lesson.qmd).

## Warm-up: do you remember?

Three quick reps first — produce each from memory, then check.

**1.** *(ABF course)* A Wakefield Bayes factor for one SNP is built from that SNP's effect
estimate $\hat\beta$ and standard error. In one line, what does a **larger** Bayes factor mean?

::: {.callout-note collapse="true" title="Answer"}
A larger Bayes factor means **stronger evidence that this SNP is associated** with the trait
(more support for "there's a real effect here" over "no effect"). See the
[ABF course](../../../approximate-bayes-factors/syllabus.md).
:::

**2.** *(coloc course)* If SNP $i$ has Bayes factor $\text{BF}_i$, how do you turn a region's
Bayes factors into the **posterior probability** that SNP $i$ is the causal one?

::: {.callout-note collapse="true" title="Answer"}
Normalise: $\text{PP}_i = \text{BF}_i / \sum_j \text{BF}_j$ — each SNP's Bayes factor divided by
the total across all SNPs (so they sum to 1). See the
[Bayesian Colocalisation course](../../../bayesian-colocalisation/syllabus.md).
:::

**3.** *(foundation)* Two SNPs are in **perfect linkage disequilibrium (LD)** — perfectly
correlated. If they truly tag the same signal equally, what posterior probability of "being
causal" should each get?

::: {.callout-note collapse="true" title="Answer"}
**0.5 each** — with identical information they're indistinguishable, so the probability splits
evenly. This is the baseline the rest of the lesson watches break. LD (correlation between SNPs)
is the same correlation idea from
[distributions & densities](../../../../foundations/distributions-and-densities/lesson.qmd).
:::

## A tiny example first

Two SNPs, **perfectly correlated** (perfect LD), truly tagging the exact same causal signal. If
both were measured equally well, each would get a full-data z-score of $3$ and — being
identical — split the causal probability 50/50.

Now impose a little **information loss** on SNP 2 only: its genotype was imputed at $r^2 = 0.9$
(90% accurate) instead of perfectly. Imputation at $r^2 = 0.9$ shrinks SNP 2's *observed*
z-score by a factor $\sqrt{0.9} \approx 0.949$, so SNP 2's z drops from $3$ to
$3 \times 0.949 \approx 2.846$. SNP 1 is untouched.

Watch what that does. A Wakefield log Bayes factor from a z-score (with prior weight
$W = 0.04$ and standard error $0.03$, so the shrink factor $r = W/(W+\text{se}^2) = 0.04/0.0409
\approx 0.978$) is $\tfrac12(\log(1 - r) + r\,z^2)$:

- SNP 1: $\tfrac12(\log(0.022) + 0.978 \times 3^2) \approx 2.493$, so $\text{BF}_1 = e^{2.493}
  \approx 12.09$.
- SNP 2: $\tfrac12(\log(0.022) + 0.978 \times 2.846^2) \approx 2.053$, so $\text{BF}_2 =
  e^{2.053} \approx 7.79$.

Normalise: SNP 2's posterior is $7.79 / (12.09 + 7.79) \approx 0.392$. A *tiny* 10% imputation
shortfall pushed SNP 2 from the deserved **0.5** down to **0.39** — and SNP 1 up to 0.61, purely
because SNP 2's Bayes factor is now measured on weaker footing. The two SNPs are *identical in
truth*; only their information differs.

The same thing in R:

```{r toy-example}
# Wakefield log Bayes factor straight from a z-score (W = prior weight, se = standard error)
labf_z <- function(z, se, W = 0.04) {
  V <- se^2                # the estimate's variance
  r <- W / (W + V)         # shrinkage factor (how much the prior tempers the evidence)
  0.5 * (log(1 - r) + r * z^2)
}

se <- 0.03
z1 <- 3                    # SNP 1: full information
z2 <- 3 * sqrt(0.9)        # SNP 2: imputation r^2 = 0.9 shrinks the observed z

l1 <- labf_z(z1, se); l2 <- labf_z(z2, se)
BF1 <- exp(l1); BF2 <- exp(l2)
c(BF1 = BF1, BF2 = BF2)                 # 12.09 and 7.79

posterior_snp2 <- BF2 / (BF1 + BF2)
posterior_snp2                          # 0.392, not the deserved 0.5
```

## Getting a feel for it

**Analogy — grading two identical essays with two different graders.** Two students hand in
word-for-word identical essays (the two SNPs tag the same signal). A fair result gives them the
same grade — 50/50 of "best in class." But essay 2 is graded by a stricter, noisier grader (the
information-lossy SNP): same essay, systematically lower marks. The Bayes-factor normalisation
then declares essay 1 "more likely the best," not because it *is* better, but because it was
measured more generously. Fine-mapping compares Bayes factors *as if* every SNP were graded by
the same grader — and when they aren't, the comparison is rigged.

Where the analogy breaks: a grader's bias is a fixed offset, but here the distortion **grows
with the size of the numbers** — which is the twist that makes this dangerous in big studies
(next section).

**Why does information loss lower the Bayes factor?** Because a Bayes factor rewards a *large,
precise* signal. Information loss does both bad things: it can shrink the observed effect (a
noisily-imputed genotype attenuates the signal, like $\sqrt{r^2}$ above) *and* inflate the
standard error (fewer effective samples). A Wakefield Bayes factor climbs steeply with $z^2$, so
even a modest z-shrink from $3$ to $2.85$ costs real evidence. The normalisation step then
converts that unfair evidence gap into an unfair probability gap.

**What does the 0.39 feel like, and what makes it worse?** At a z of $3$, the drift is mild
(0.5 → 0.39). But the Bayes factor grows with $z^2$, so at a z of $10$ the *same* 10% shrink
opens a far larger BF gap — and the posterior of the lossy SNP collapses toward 0. Bigger
studies mean bigger z-scores mean bigger drift. That is the whole alarm of the paper: the error
**worsens as sample size grows**, exactly when you'd hope for more reliable answers.

## What the notation is going to say

The formula will say what you already did by hand: turn each SNP's z into a log Bayes factor
with $\tfrac12(\log(1-r) + r z^2)$, exponentiate, and normalise by the region's total. The one
new idea is that the *inputs* to that formula — each SNP's z and standard error — carry
different amounts of information, so the Bayes factors that come out are not comparable, and
normalising incomparable numbers gives a biased posterior.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $z$ | "zed" / "zee" | a SNP's z-score: effect estimate divided by its standard error, $\hat\beta/\text{se}$ | $3$ (SNP 1), $2.846$ (SNP 2) |
| $\text{se}$ | "ess-ee" | the standard error of the effect estimate — the estimate's typical wobble | $0.03$ |
| $W$ | "double-you" | the prior weight (prior variance of the effect); coloc's case-control default is $0.04$ | $0.04$ |
| $r$ | "little r" (here a **shrinkage factor**, not a correlation) | $W/(W + \text{se}^2)$ — how much of the raw evidence survives the prior | $\approx 0.978$ |
| $\text{BF}_i$ | "B-F sub eye" | SNP $i$'s Bayes factor: evidence it's associated | $12.09$, $7.79$ |
| $\ell_i$ | "ell sub eye" | the **log** Bayes factor, $\log \text{BF}_i$ (what `wakefield_abf` returns) | $2.493$, $2.053$ |
| $\text{PP}_i$ | "P-P sub eye" | posterior probability SNP $i$ is causal: $\text{BF}_i / \sum_j \text{BF}_j$ | $0.608$, $0.392$ |
| $r^2$ | "r squared" | imputation accuracy: how well a SNP's genotype is known (1 = perfect) | $0.9$ on SNP 2 |
| $N$ | "en" | study sample size | varies |

## The formal version

A SNP's posterior probability of being causal, built from its Wakefield log Bayes factor:

$$
\ell_i = \tfrac{1}{2}\big(\log(1 - r) + r\,z_i^2\big), \qquad r = \frac{W}{W + \text{se}_i^2},
\qquad \text{PP}_i = \frac{e^{\ell_i}}{\sum_j e^{\ell_j}} .
$$

Reading it aloud: "ell-sub-i equals one half times, log of one minus r, plus r times z-sub-i
squared; where r is W over W plus se-sub-i squared; and the posterior of SNP i is e-to-the-ell-i
divided by the sum over j of e-to-the-ell-j." Everyday words: "score each SNP by how big and
precise its signal is, then share out the causal probability in proportion to those scores."

**Mapping it onto the toy example.** With $W = 0.04$, $\text{se} = 0.03$ so $r \approx 0.978$:

- the "$r z_i^2$" term is the "big precise signal earns evidence" step → SNP 1: $0.978\times 9$;
  SNP 2: $0.978\times 8.10$;
- exponentiating gives $\text{BF}_1 = 12.09$, $\text{BF}_2 = 7.79$;
- the "$/\sum_j$" is the normalise-to-a-probability step → $\text{PP}_2 = 7.79/19.88 = 0.392$.

Every number matches the by-hand computation above. The bias lives entirely in the fact that
$z_2$ (and, in a fuller model, $\text{se}_2$) carries less information than $z_1$, even though
the two SNPs are truth-identical.

## Explore it

```{webr}
#| autorun: true
# TRY IT: change the imputation r^2 on SNP 2. PREDICT: at r^2 = 1 (no loss), what posterior
# should SNP 2 get? (0.5 — identical information.) Then try r^2 = 0.8.
labf_z <- function(z, se, W = 0.04) { V <- se^2; r <- W/(W + V); 0.5*(log(1 - r) + r*z^2) }
se <- 0.03
z_full <- 3
imp_r2 <- 0.9                          # <- change me

z2 <- z_full * sqrt(imp_r2)            # information loss shrinks SNP 2's observed z
l <- c(labf_z(z_full, se), labf_z(z2, se))
BF <- exp(l - max(l))                  # subtract max for numerical stability
post <- BF / sum(BF)
cat("posterior of the info-lossy SNP 2:", round(post[2], 4),
    " (deserved: 0.5)\n")
```

```{webr}
#| autorun: true
#| fig-alt: "A curve of the information-lossy SNP's posterior probability against the full-data z-score, starting near 0.5 at small z and sliding down toward 0 as z grows, with a dashed horizontal line at the deserved 0.5."
# TRY IT: this reproduces the paper's Figure-1 idea. Bigger studies -> bigger z-scores. Watch
# the info-lossy SNP's posterior slide AWAY from 0.5 as z grows. PREDICT: up or down?
labf_z <- function(z, se, W = 0.04) { V <- se^2; r <- W/(W + V); 0.5*(log(1 - r) + r*z^2) }
se <- 0.03; imp_r2 <- 0.95
z_grid <- seq(1, 25, length.out = 200)
post2 <- sapply(z_grid, function(zf) {
  l <- c(labf_z(zf, se), labf_z(zf*sqrt(imp_r2), se))
  BF <- exp(l - max(l)); (BF/sum(BF))[2]
})
plot(z_grid, post2, type = "l", lwd = 2, ylim = c(0, 0.6),
     xlab = "full-data z-score (grows with sample size)",
     ylab = "posterior of the info-lossy SNP",
     main = "Fine-mapping bias worsens as the study grows")
abline(h = 0.5, lty = 2, col = "tomato")
legend("bottomleft", legend = "deserved 0.5", lty = 2, col = "tomato", bty = "n")
```

```{webr}
#| autorun: true
# TRY IT: the sample-size story as a table. Fixed true effect; se shrinks with sqrt(N), so z
# grows with sqrt(N). PREDICT: which N gives the WORST (closest to 0) posterior for SNP 2?
labf_z <- function(z, se, W = 0.04) { V <- se^2; r <- W/(W + V); 0.5*(log(1 - r) + r*z^2) }
effect <- 0.06; imp_r2 <- 0.95
drift <- sapply(c(5000, 50000, 500000), function(N) {
  se1 <- 0.03 * sqrt(5000 / N)         # standard error shrinks with sqrt(N)
  z1  <- effect / se1                   # full-data z grows with sqrt(N)
  z2  <- z1 * sqrt(imp_r2)              # info-lossy SNP 2
  l <- c(labf_z(z1, se1), labf_z(z2, se1)); BF <- exp(l - max(l)); (BF/sum(BF))[2]
})
data.frame(N = c(5000, 50000, 500000), posterior_snp2 = round(drift, 4))
# ~0.48 at N=5,000 but ~0 at N=500,000 -- matching the paper's Figure 1
```

## Common traps

**Trap: "bigger sample size means more reliable fine-mapping."** Intuitive — more data is
usually better. But here it's the opposite: because the Bayes factor grows with $z^2$ and $z$
grows with $\sqrt{N}$, a fixed information gap between SNPs blows up as $N$ grows. Our table shows
the info-lossy SNP's posterior at $0.48$ for $N = 5{,}000$ but essentially $0$ for
$N = 500{,}000$ — the *same* 5% imputation shortfall, far more damage in the big study. More data
sharpens a *biased* answer, not a correct one.

**Trap: "the two SNPs really are different, since they got different posteriors."** No — in our
setup the two SNPs are in perfect LD and tag the *identical* true signal; they *deserve* 0.5
each. The only difference is measurement quality. The posterior gap is an **artifact of unequal
information**, not a real difference in causal status. That's exactly why a single-variant
credible set here would name the wrong SNP half the time.

## Check yourself

**1.** Two perfectly-correlated SNPs; SNP 2 loses information so its z drops from $4$ to $3.6$
while SNP 1 stays at $4$ (same se $= 0.03$). Is SNP 2's posterior above or below 0.5, and
roughly where?

```{webr}
#| autorun: false
# check1-starter — edit and Run to check
labf_z <- function(z, se, W = 0.04) { V <- se^2; r <- W/(W + V); 0.5*(log(1 - r) + r*z^2) }
se <- 0.03
```

::: {.callout-note collapse="true" title="Answer"}
Below 0.5. $\ell_1 = \tfrac12(\log(1-r) + r\cdot 16)$, $\ell_2 = \tfrac12(\log(1-r) + r\cdot
12.96)$ with $r \approx 0.978$; exponentiate and normalise to get SNP 2's posterior $\approx
0.18$. In R: `l <- c(labf_z(4, se), labf_z(3.6, se)); BF <- exp(l - max(l)); (BF/sum(BF))[2]`.
The larger z makes the gap much wider than the toy's 0.39.
:::

**2.** Name three real-world causes of "variable information across SNPs" the paper lists.

::: {.callout-note collapse="true" title="Answer"}
Any three of: **unequal sample size** across variants (common in meta-analyses), **variable
imputation accuracy** ($r^2$), **variable call rate / genotype missingness**, and even
**rounding summary statistics** to few significant figures (the rounding error varies across
SNPs). All change how much information a SNP carries, so its Bayes factor is no longer comparable
to its neighbours'.
:::

**3.** In your own words: explain to a colleague, in two sentences, why unequal information
across SNPs biases fine-mapping, and why it gets worse with sample size.

::: {.callout-note collapse="true" title="A model answer"}
Fine-mapping normalises each SNP's Bayes factor into a probability, which silently assumes every
SNP's Bayes factor is measured on the same footing — so when some SNPs carry less information
(smaller z, larger se), their Bayes factors are unfairly small and the posterior is biased toward
the better-measured SNPs. It worsens with sample size because the Bayes factor grows with the
square of the z-score, which itself grows with $\sqrt{N}$, so a fixed information gap between SNPs
opens into an ever-larger evidence gap as the study grows.
:::

## What you can now do

- **Compute a SNP's posterior of being causal** from its z and standard error via the Wakefield
  log Bayes factor and the normalise-across-SNPs step.
- **Explain why unequal statistical information** (sample size, imputation $r^2$, call rate,
  rounding) makes Bayes factors non-comparable and biases fine-mapping.
- **Reproduce the Figure-1 intuition** in base R: two perfectly-LD SNPs, information loss on one,
  its posterior drifting from the deserved 0.5.
- **State the sample-size twist:** the bias *worsens* as $N$ grows, because evidence scales with
  $z^2 \propto N$.

If any of these feels shaky, revisit the toy example and the two Common traps before continuing —
the next module shows this bias flowing straight into `coloc`.

## Where this goes next

You've seen fine-mapping break at the level of a single trait. The next module,
[How it corrupts coloc](../02-how-it-corrupts-coloc/lesson.qmd), shows this same bias propagating
into the `coloc` you built — turning a truly *shared* signal (H4) into a confident but wrong
*distinct* call (H3). That's what motivates a colocalisation test that avoids fine-mapping
entirely.

Ready for more reps? Do the [practice problems](practice.qmd).
