---
title: "Practice: When fine-mapping breaks"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

Work these in order — they get harder. Do them on paper or in an R session; only then open the
answer.

## Problem 1 — the toy example, new numbers

Two perfectly-LD SNPs; SNP 1 has full-data z $= 2.5$, SNP 2 loses information (imputation
$r^2 = 0.8$, so its z is $2.5\sqrt{0.8}$), both with se $= 0.03$. Compute SNP 2's posterior of
being causal.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
labf_z <- function(z, se, W = 0.04) { V <- se^2; r <- W/(W + V); 0.5*(log(1 - r) + r*z^2) }
se <- 0.03
z1 <- 2.5
z2 <- 2.5 * sqrt(0.8)
```

::: {.callout-note collapse="true" title="Answer"}
```{r p1-ans}
labf_z <- function(z, se, W = 0.04) { V <- se^2; r <- W/(W + V); 0.5*(log(1 - r) + r*z^2) }
se <- 0.03; z1 <- 2.5; z2 <- 2.5*sqrt(0.8)
l <- c(labf_z(z1, se), labf_z(z2, se)); BF <- exp(l - max(l))
round(BF/sum(BF), 4)          # posterior of SNP 1, SNP 2
```
SNP 2's posterior is about **0.35** — below the deserved 0.5. Even at this modest z, an
imputation $r^2$ of 0.8 opens a clear gap; bigger z's would push it lower still.
:::

## Problem 2 — fill in the missing step

SNP with z $= 3$, se $= 0.03$, $W = 0.04$. Compute its log Bayes factor:

- Step 1: $r = W/(W + \text{se}^2) = 0.04/(0.04 + 0.0009) \approx 0.978$.
- Step 2: **your step here** — plug into $\ell = \tfrac12(\log(1 - r) + r z^2)$.
- Step 3: exponentiate: $\text{BF} = e^{\ell}$.

Fill in step 2.

```{webr}
#| autorun: false
# p2-starter — edit and Run to check the blank
r <- 0.04/(0.04 + 0.03^2)
z <- 3
```

::: {.callout-note collapse="true" title="Answer"}
Step 2: $\ell = \tfrac12(\log(1 - 0.978) + 0.978\times 9) = \tfrac12(\log(0.022) + 8.802) =
\tfrac12(-3.816 + 8.802) = \tfrac12(4.986) \approx 2.493$.

```{r p2-ans}
r <- 0.04/(0.04 + 0.03^2); z <- 3
l <- 0.5*(log(1 - r) + r*z^2); l          # ~2.493
exp(l)                                     # BF ~ 12.09
```
:::

## Problem 3 — the sample-size drift

Fixed true effect $0.05$; imputation $r^2 = 0.9$ on SNP 2. Compute SNP 2's posterior at
$N = 5{,}000$, $50{,}000$, and $500{,}000$ (se shrinks with $\sqrt N$, so z grows with
$\sqrt N$), and say which is worst.

```{webr}
#| autorun: false
# p3-starter — your job: loop over N, compute SNP 2's posterior each time
labf_z <- function(z, se, W = 0.04) { V <- se^2; r <- W/(W + V); 0.5*(log(1 - r) + r*z^2) }
effect <- 0.05; imp_r2 <- 0.9
Ns <- c(5000, 50000, 500000)
```

::: {.callout-note collapse="true" title="Answer"}
```{r p3-ans}
labf_z <- function(z, se, W = 0.04) { V <- se^2; r <- W/(W + V); 0.5*(log(1 - r) + r*z^2) }
effect <- 0.05; imp_r2 <- 0.9
post2 <- sapply(c(5000, 50000, 500000), function(N) {
  se1 <- 0.03 * sqrt(5000/N); z1 <- effect/se1; z2 <- z1*sqrt(imp_r2)
  l <- c(labf_z(z1, se1), labf_z(z2, se1)); BF <- exp(l - max(l)); (BF/sum(BF))[2]
})
data.frame(N = c(5000, 50000, 500000), posterior_snp2 = round(post2, 4))
```
The **largest** study ($N = 500{,}000$) is worst — SNP 2's posterior collapses toward 0. More
data sharpens the biased answer; it does not fix it.
:::

## Problem 4 — spot the error

A colleague concludes their fine-mapping is trustworthy because they have a huge sample. They
wrote — it contains **exactly one mistake** in the reasoning. Find it.

> "We ran on 500,000 individuals, so any single-variant credible set is reliable — the huge
> sample size guarantees the fine-mapping bias from unequal imputation is negligible."

::: {.callout-note collapse="true" title="Answer"}
The mistake is "the huge sample size guarantees the bias is negligible." It's **backwards**: the
fine-mapping bias from unequal information *grows* with sample size, because the Bayes factor
scales with $z^2$ and $z \propto \sqrt N$. A 500,000-person study makes an imputation-driven
credible set *less* trustworthy, not more — it can confidently name the wrong SNP. The correct
statement: with variable information across SNPs, a large $N$ demands *more* caution about
single-variant credible sets, not less.
:::

## Problem 5 — mixing it with coloc *(uses the Bayesian Colocalisation course's normalisation)*

You have three SNPs in a region with log Bayes factors $\ell = (2.0, 2.0, 1.0)$ for one trait.
Using the coloc-course normalisation ($\text{PP}_i = e^{\ell_i}/\sum_j e^{\ell_j}$), compute each
SNP's posterior. Then say what would happen to SNP 3's posterior if information loss dropped its
$\ell$ to $0.5$.

```{webr}
#| autorun: false
# p5-starter — edit and Run to solve, right here
l <- c(2.0, 2.0, 1.0)
```

::: {.callout-note collapse="true" title="Answer"}
```{r p5-ans}
l <- c(2.0, 2.0, 1.0); BF <- exp(l - max(l))
round(BF/sum(BF), 4)                        # 0.4223 0.4223 0.1554
l2 <- c(2.0, 2.0, 0.5); BF2 <- exp(l2 - max(l2))
round(BF2/sum(BF2), 4)                       # SNP 3 drops further
```
Original: about $(0.42, 0.42, 0.16)$. Dropping SNP 3's $\ell$ to $0.5$ shrinks its posterior to
about $0.10$ and nudges the other two up — the same normalise-across-SNPs step from the coloc
course, now visibly biased by SNP 3's information loss.
:::

## Problem 6 — transfer: rounding as information loss

The paper notes that even **rounding summary statistics** to few decimal places causes variable
information across SNPs. Without computing, explain in one or two sentences why rounding acts like
information loss and why it varies across SNPs.

::: {.callout-note collapse="true" title="Answer"}
Rounding throws away precision — the reported number is the true value plus a small rounding
error — which is exactly a small dose of added noise, i.e. information loss. It varies across SNPs
because the *size* of the rounding error relative to each SNP's true statistic differs (rounding
0.123 to 0.1 loses proportionally more than rounding 0.789 to 0.8), so different SNPs end up with
different effective precision — and their Bayes factors become non-comparable, just like unequal
imputation.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**, once the
material isn't fresh. A pass then is real evidence you've retained it. Five questions, graded on
the page; 4/5 or better means this module is solid. Results are saved in this browser only.

::: {.module-check module="courses/proportional-colocalisation/modules/01-when-fine-mapping-breaks" pass="4"}

::: {.quiz-q qid="wfmb-perfect-ld-baseline" qtype="numeric" answer="0.5" tolerance="0"}
Two SNPs in perfect LD tag the identical true signal and are measured with equal information.
What posterior probability of being causal should each receive?
:::

::: {.quiz-q qid="wfmb-samplesize-direction" qtype="mc" answer="b"}
As sample size grows, the fine-mapping bias from unequal information across SNPs:

a) shrinks toward zero
b) gets worse, because the Bayes factor scales with $z^2$ and $z$ grows with $\sqrt{N}$
c) stays exactly the same
d) reverses sign
:::

::: {.quiz-q qid="wfmb-cause-of-variable-info" qtype="mc" answer="c"}
Which of these does the paper cite as a cause of variable statistical information across SNPs?

a) using a Bayesian instead of a frequentist test
b) choosing the prior weight $W$
c) variable imputation accuracy ($r^2$) across variants
d) the number of hypotheses in coloc
:::

::: {.quiz-q qid="wfmb-why-biased" qtype="mc" answer="a"}
Why does normalising Bayes factors across SNPs give a biased posterior when information varies?

a) It assumes every SNP's Bayes factor is measured on the same footing, but info-lossy SNPs get unfairly small ones
b) It divides by the wrong total
c) It uses the log scale instead of the raw scale
d) It forgets to include the prior weight $W$
:::

::: {.quiz-q qid="wfmb-normalise-interleave" qtype="numeric" answer="0.4223" tolerance="0.001"}
*(uses the Bayesian Colocalisation course's normalisation)* For three SNPs with log Bayes factors
$\ell = (2.0, 2.0, 1.0)$, what is SNP 1's posterior $e^{\ell_1}/\sum_j e^{\ell_j}$? (Enter to 4
decimals.)
:::

:::

::::
