---
title: "Many knobs, one model"
subtitle: "From one predictor to many — and the phrase that unlocks every coefficient"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — live `{webr}` cells below re-state their own data.
needed <- c()   # base R is all we need
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)

# The shared running example for this course: a simulated economic-indicator panel.
# Signal sits on ONE member of each correlated twin pair (rate_10y, cpi_infl, unemp),
# so later modules can show correlated predictors interfering. Base R only.
sim_indicators <- function(n = 200, seed = 1, rho_max = 0.95) {
  set.seed(seed)
  nm <- c("rate_10y", "rate_2y", "cpi_infl", "pce_infl", "unemp",
          "oil", "usd_index", "sentiment", "housing", "pmi",
          "credit_spread", "vix")
  p <- length(nm)
  R <- diag(p); dimnames(R) <- list(nm, nm)
  R["rate_10y", "rate_2y"]  <- R["rate_2y", "rate_10y"]  <- 0.90
  R["cpi_infl", "pce_infl"] <- R["pce_infl", "cpi_infl"] <- rho_max
  R["sentiment", "pmi"]     <- R["pmi", "sentiment"]     <- 0.70
  R["credit_spread", "vix"] <- R["vix", "credit_spread"] <- 0.65
  X <- matrix(rnorm(n * p), n, p) %*% chol(R)
  colnames(X) <- nm
  beta <- setNames(numeric(p), nm)
  beta[c("rate_10y", "cpi_infl", "unemp")] <- c(0.5, -0.4, -0.3)
  y <- drop(X %*% beta + rnorm(n, sd = 1))
  list(X = X, y = y, beta_true = beta)
}
```

## Where we're going

In the earlier course you fit a *line*: one predictor, one slope. Real questions
rarely come with one predictor. An analyst asking *"which indicators drive next
quarter's GDP growth?"* has a dozen candidates at once — interest rates, inflation,
unemployment, oil. **Multiple regression** fits them all in a single model, giving
each predictor its own coefficient. This module builds the picture and teaches the one
phrase you need to read any of those coefficients correctly: *"holding the others
fixed."* No new formula machinery yet — that's Module 2. Here we learn what the model
*says*.

**Builds on:** [Simple Linear Regression](../../../simple-linear-regression/syllabus.md)
(a slope is "change in y per unit change in x"),
[linear-combinations-and-data-matrices](../../../../foundations/linear-combinations-and-data-matrices/lesson.qmd)
(stacking predictors into columns; a weighted sum of columns).

## Warm-up: do you remember?

Three quick reps — answer from memory first, then check.

**1.** *(simple linear regression)* In the line $\hat{y} = 3 + 2x$, the slope is 2.
Say in one sentence what that 2 *means* in terms of $x$ and $\hat{y}$.

::: {.callout-note collapse="true" title="Answer"}
"When $x$ increases by 1 unit, the predicted $\hat{y}$ increases by 2 units." A slope
is a *change per unit change* — the single idea we're about to give many predictors.
See [Simple Linear Regression](../../../simple-linear-regression/syllabus.md).
:::

**2.** *(linear combinations)* You have columns $c_1 = (1, 2)$ and $c_2 = (10, 20)$.
What is the linear combination $3\,c_1 + 0.5\,c_2$?

::: {.callout-note collapse="true" title="Answer"}
$3\,(1,2) + 0.5\,(10,20) = (3, 6) + (5, 10) = (8, 16)$. A **linear combination** is a
weighted sum of columns — exactly how a multiple regression builds its predictions.
See [linear-combinations-and-data-matrices](../../../../foundations/linear-combinations-and-data-matrices/lesson.qmd).
:::

**3.** *(baseline)* If a recipe's cost is `2 + 1*flour + 3*sugar` (in some units), and
you use 2 flour and 1 sugar, what is the cost?

::: {.callout-note collapse="true" title="Answer"}
$2 + 1\times 2 + 3\times 1 = 2 + 2 + 3 = 7$. Plugging numbers into a
"intercept plus weighted ingredients" formula is exactly what prediction will be.
:::

## A tiny example first

Forget economics for a moment. Suppose the predicted value $\hat{y}$ is built from two
knobs $x_1$ and $x_2$ by this exact rule:

> $\hat{y} = 2 + 1\cdot x_1 + 3\cdot x_2.$

Here are four settings of the knobs and what the rule predicts:

| $x_1$ | $x_2$ | $\hat{y} = 2 + 1\cdot x_1 + 3\cdot x_2$ |
|---:|---:|---:|
| 0 | 0 | $2 + 0 + 0 = 2$ |
| 1 | 0 | $2 + 1 + 0 = 3$ |
| 0 | 1 | $2 + 0 + 3 = 5$ |
| 2 | 1 | $2 + 2 + 3 = 7$ |

Look at the first two rows. $x_2$ stays at 0; $x_1$ goes from 0 to 1; the prediction
goes from 2 to 3 — **up by 1**. That 1 is the number multiplying $x_1$. Now rows one
and three: $x_1$ stays at 0; $x_2$ goes from 0 to 1; the prediction goes from 2 to 5 —
**up by 3**, the number multiplying $x_2$. The "2" out front is what you predict when
both knobs are at zero.

The same thing in R:

```{r toy-example}
# The toy rule, applied to the four knob settings
x1 <- c(0, 1, 0, 2)
x2 <- c(0, 0, 1, 1)
y  <- c(2, 3, 5, 7)          # = 2 + 1*x1 + 3*x2, exactly

# Ask R to recover the rule from the data (Module 2 shows HOW; here just read it)
coef(lm(y ~ x1 + x2))        # (Intercept) 2, x1 1, x2 3 — the rule, read back out
```

## Getting a feel for it

**A studio mixing board.** Picture a sound engineer at a mixing board with a row of
sliders — one for vocals, one for bass, one for drums. The final sound is all of them
added together. Each slider has its own effect on the mix, and — this is the key —
you judge what the *bass* slider does by pushing **only** the bass slider and leaving
every other slider exactly where it is. Multiple regression is that board: each
predictor is a slider, its coefficient is how much the output moves when you nudge
*that* slider a notch **with all the others held still.**

**Why "holding the others fixed" is the whole game.** With one predictor there was
nothing to hold fixed, so a slope was just "change in $y$ per change in $x$." With many
predictors, if you change $x_1$ *and* let $x_2$ drift at the same time, you can't tell
which one moved $y$. So the coefficient on $x_1$ is defined as the effect of $x_1$
*when $x_2$ (and every other predictor) is pinned down.* In the toy table we engineered
exactly that: to read the $x_1$ effect we compared two rows where $x_2$ didn't change.

**What the coefficients feel like.** In the toy rule the $x_2$ knob (3) is "stronger"
than the $x_1$ knob (1) — a one-unit nudge of $x_2$ moves the prediction three times as
far. The intercept (2) is the baseline the prediction sits at before any knob is
touched. Where the analogy breaks: a real mixing board's sliders are independent, but
real predictors often move *together* (inflation measures rise as a group) — and that
is precisely the trouble Module 3 is about.

## What the notation is going to say

The formula will say: *the prediction is a baseline, plus each predictor multiplied by
its own coefficient, all added up.* In the toy example that baseline is 2, and the
"each predictor times its coefficient" part is $1\cdot x_1 + 3\cdot x_2$. Real data
won't land *exactly* on the rule, so the honest model also admits a leftover — the part
no straight-line-in-the-predictors rule can capture.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $y_i$ | "why sub eye" | the outcome for observation $i$ (in R: `y[i]`) | 2, 3, 5, 7 |
| $x_{ij}$ | "ex sub eye jay" | value of the $j$-th predictor for observation $i$ | e.g. $x_{4,1} = 2$ |
| $p$ | "pee" | how many predictors (not counting the intercept) | 2 |
| $n$ | "en" | how many observations (in R: `nrow`) | 4 |
| $\beta_0$ | "beta zero" | the intercept — the prediction when every predictor is 0 | 2 |
| $\beta_j$ | "beta sub jay" | the coefficient on predictor $j$: change in $y$ per unit of $x_j$, others held fixed | $\beta_1 = 1,\ \beta_2 = 3$ |
| $\hat{y}_i$ | "why hat sub eye" | the model's prediction for observation $i$ (in R: `predict`) | matches $y_i$ here |
| $\varepsilon_i$ | "epsilon sub eye" | the leftover for observation $i$ — what the predictors don't explain | 0 in this rigged toy |
| $\mathbf{X}$ | "the matrix ex" | the table of predictors, rows = observations, columns = predictors | the $x_1, x_2$ columns |
| $\sum_{j=1}^{p}$ | "the sum from jay equals 1 to pee" | add the term to its right once per predictor | $\beta_1 x_1 + \beta_2 x_2$ |

## The formal version

We build the model one piece at a time.

- Start with a **baseline** every prediction shares: $\beta_0$.
- Add each predictor scaled by its own coefficient: $\beta_1 x_{i1}$, then
  $\beta_2 x_{i2}$, and so on up to $\beta_p x_{ip}$. Each term is "one slider's
  contribution."
- Real data misses the rule slightly, so add a leftover $\varepsilon_i$ for each
  observation.

Put together, the multiple linear regression model for observation $i$ is:

$$
y_i = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \cdots + \beta_p x_{ip} + \varepsilon_i
    = \beta_0 + \sum_{j=1}^{p} \beta_j x_{ij} + \varepsilon_i .
$$

Reading it aloud: "why-sub-eye equals beta-zero, plus the sum from jay equals 1 to pee
of beta-sub-jay times ex-sub-eye-jay, plus epsilon-sub-eye" — or in everyday words:
"each outcome is a baseline, plus every predictor times its own weight, plus whatever's
left over."

The model's **prediction** drops the leftover (we don't know it in advance):

$$
\hat{y}_i = \beta_0 + \sum_{j=1}^{p} \beta_j x_{ij} .
$$

Mapping onto the toy example (row 4, where $x_1 = 2, x_2 = 1$):

- $\beta_0$ is the "2" baseline.
- $\sum_j \beta_j x_{ij}$ is $1\cdot 2 + 3\cdot 1 = 5$.
- So $\hat{y}_4 = 2 + 5 = 7$, matching the table — and $\varepsilon_4 = y_4 - \hat{y}_4 = 0$
  because we built the data to sit exactly on the rule.

**The prediction is a linear combination of the predictor columns** (plus the
intercept) — the warm-up's weighted-sum-of-columns idea, with the coefficients as the
weights.

## Explore it

```{webr}
#| autorun: true
# TRY IT: change the coefficient on x2 (the 3) and re-run. Watch how the gap between
# the x2=0 rows and the x2=1 rows changes — that gap IS the coefficient.
# PREDICT: if you change the 3 to a 10, what will the last prediction (x1=2, x2=1) be?
b0 <- 2; b1 <- 1; b2 <- 3
x1 <- c(0, 1, 0, 2)
x2 <- c(0, 0, 1, 1)

yhat <- b0 + b1 * x1 + b2 * x2      # the model's prediction, straight from the formula
data.frame(x1, x2, yhat)
```

```{webr}
#| autorun: false
# TRY IT: here the model has a NEGATIVE coefficient on x2. Before running:
# PREDICT: with b2 = -4 and (x1=1, x2=3), will yhat be above or below the intercept?
b0 <- 10; b1 <- 2; b2 <- -4
x1 <- c(0, 1, 1, 3)
x2 <- c(0, 0, 3, 2)

yhat <- b0 + b1 * x1 + b2 * x2
data.frame(x1, x2, yhat)
```

Now the real thing. Here is the course's running example: a simulated panel of GDP
growth against twelve economic indicators. We fit all twelve at once with `lm(y ~ .)`
("`.`" means "every other column as a predictor"). This is a listing to *read*, not
edit — it runs on the desktop/CI build:

```{r panel-fit}
# The simulated economic panel (defined in this page's hidden setup)
sim <- sim_indicators(n = 200, seed = 1)
panel <- data.frame(sim$X, gdp_growth = sim$y)

fit <- lm(gdp_growth ~ ., data = panel)
round(coef(fit), 3)          # one intercept + twelve coefficients, each "others held fixed"
```

```{r panel-predict}
# Predict GDP growth for one new quarter (all twelve indicators supplied at once).
# PREDICT: prediction = intercept + sum of (each indicator * its coefficient).
new_quarter <- panel[1, names(coef(fit))[-1]]   # reuse row 1's indicator values
predict(fit, newdata = new_quarter)              # a single number: the model's forecast
```

Every one of those twelve coefficients carries the "holding the *other eleven* fixed"
meaning. Reading them safely is what the rest of the course is about.

## Common traps

**Trap: "the coefficient on $x_1$ is the effect of $x_1$ on $y$, full stop."** It's
tempting to read $\beta_1$ as "$x_1$'s effect" with no strings attached. But the honest
reading always has strings: it is $x_1$'s effect *with every other predictor held
fixed*. Change which other predictors are in the model and $\beta_1$ can change. In the
toy table, "$x_1$'s effect is 1" was only true because we compared rows with $x_2$
frozen; if $x_2$ had drifted along with $x_1$, the apparent $x_1$ effect would have
been different. (Module 3 makes this concrete and a little alarming.)

**Trap: "the biggest coefficient is the most important predictor."** A coefficient's
size depends on the predictor's *units*. If $x_1$ were measured in thousands and $x_2$
in ones, their coefficients would scale accordingly even if the predictors mattered
equally. Comparing raw coefficient sizes across predictors on different scales is
comparing apples to kilometres. (Standardizing predictors first — the
[variance-structure](../../../../foundations/variance-structure-and-standardization/lesson.qmd)
module's job — is what makes such comparisons fair; Module 3 leans on it.)

## Check yourself

**1.** A fitted model is $\hat{y} = 5 + 2\,x_1 - 4\,x_2$. Two neighbourhoods have the
same $x_2$, but the second has $x_1$ larger by 3. How much higher is its predicted
$\hat{y}$?

::: {.callout-note collapse="true" title="Answer"}
Only $x_1$ changes (by 3), $x_2$ is held fixed, so $\hat{y}$ changes by
$2 \times 3 = 6$ — up 6. The $-4\,x_2$ term is identical for both and cancels.
:::

**2.** In that same model, what does the $-4$ tell you, stated with the crucial
qualifier?

::: {.callout-note collapse="true" title="Answer"}
Holding $x_1$ (and any other predictor) fixed, a one-unit increase in $x_2$ *lowers*
the prediction by 4. The qualifier "holding $x_1$ fixed" is not optional — it is what a
multiple-regression coefficient means.
:::

**3.** In your own words: explain to a colleague, in two sentences, what a coefficient
in a multiple regression measures and why the phrase "holding the others fixed" is part
of the definition.

::: {.callout-note collapse="true" title="A model answer"}
A coefficient is how much the prediction moves when you nudge that one predictor by a
unit while every other predictor stays put — like judging one slider on a mixing board
by moving only it. The "held fixed" part is essential because if the other predictors
were free to move too, you couldn't tell which predictor was responsible for the change
in $y$.
:::

## What you can now do

- **Read the multiple-regression model** $y_i = \beta_0 + \sum_j \beta_j x_{ij} + \varepsilon_i$
  and name every part (baseline, per-predictor weighted terms, leftover).
- **Interpret any coefficient** as "change in $y$ per unit of that predictor, holding
  all the others fixed" — and know why the qualifier can't be dropped.
- **Compute a prediction** as the intercept plus a linear combination of the predictor
  values, by hand or with `predict()`.
- **Resist two traps:** a coefficient is not an unconditional effect, and its raw size
  is not importance when predictors are on different scales.

If any of these feels shaky, revisit [Getting a feel for it](#getting-a-feel-for-it)
before continuing — Module 2 shows exactly how `lm()` finds those coefficients, and it
assumes you can read them.

## Where this goes next

You can read the model and its coefficients; you don't yet know where the numbers
*come from*. Next: [Least squares and the normal equations](../02-least-squares-and-the-normal-equations/lesson.qmd)
derives the coefficients by minimizing squared error over all the knobs at once, and
you'll rebuild `lm()`'s answer in one line of base R. Ready for more reps first? Do the
[practice problems](practice.qmd).
