---
title: "Practice: Many knobs, one model"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
set.seed(42)
```

Work these in order — they get harder. Do them on paper or in an R session; only then
open the answer.

## Problem 1 — the toy rule, new numbers

A model predicts by the rule $\hat{y} = 1 + 2\,x_1 + 5\,x_2$. Fill in the predictions
for these three knob settings, then say which number is the effect of $x_1$ (others
held fixed) and which is the effect of $x_2$.

| $x_1$ | $x_2$ | $\hat{y}$ |
|---:|---:|---:|
| 0 | 0 | ? |
| 1 | 0 | ? |
| 0 | 1 | ? |

```{webr}
#| autorun: false
# p1-starter — edit and Run to check your hand answers
b0 <- 1; b1 <- 2; b2 <- 5
x1 <- c(0, 1, 0); x2 <- c(0, 0, 1)
```

::: {.callout-note collapse="true" title="Answer"}
Predictions: $1+0+0 = 1$, $1+2+0 = 3$, $1+0+5 = 6$. Comparing rows 1→2 ($x_1$ up 1, $x_2$
fixed) the prediction rises by 2 — the effect of $x_1$. Comparing rows 1→3 ($x_2$ up 1,
$x_1$ fixed) it rises by 5 — the effect of $x_2$.

```{r}
b0 <- 1; b1 <- 2; b2 <- 5
x1 <- c(0, 1, 0); x2 <- c(0, 0, 1)
b0 + b1 * x1 + b2 * x2      # 1 3 6
```
:::

## Problem 2 — fill in the missing step

A colleague is interpreting the coefficient on `income` in the model
$\widehat{\text{spend}} = 200 + 0.3\,\text{income} - 15\,\text{kids}$. They write:

- Step 1: "The coefficient on `income` is 0.3."
- Step 2: **your step here** — state the full interpretation, with the qualifier.
- Step 3: "So two households differing by \$1,000 in income (but the same number of
  kids) are predicted to differ by \$300 in spending."

Supply Step 2.

::: {.callout-note collapse="true" title="Answer"}
Step 2: "Holding the number of kids fixed, each extra \$1 of income raises predicted
spending by \$0.30." The essential part is **holding kids fixed** — without it, "0.3 is
income's effect" is incomplete, because in a multiple regression a coefficient is always
a *ceteris paribus* (others-held-fixed) effect. Step 3 then just scales \$0.30 by 1,000.
:::

## Problem 3 — predict from a fitted model, in R

A fitted model of house price (in \$1000s) is
$\widehat{\text{price}} = 50 + 0.10\,\text{sqft} + 20\,\text{baths}$. Predict the price
of a 2000-sqft, 2-bath house, and of a 2000-sqft, 3-bath house. By how much does the
extra bathroom change the prediction?

```{webr}
#| autorun: false
# p3-starter — your job: compute both predictions and their difference
# PREDICT: the "+20 per bath" coefficient tells you the difference before you compute it.
price <- function(sqft, baths) 50 + 0.10 * sqft + 20 * baths
```

::: {.callout-note collapse="true" title="Answer"}
$50 + 0.10\times2000 + 20\times2 = 50 + 200 + 40 = 290$ (\$290k) for 2 baths, and
$50 + 200 + 60 = 310$ (\$310k) for 3 baths. The difference is \$20k — exactly the
`baths` coefficient, because `sqft` was held fixed at 2000.

```{r}
price <- function(sqft, baths) 50 + 0.10 * sqft + 20 * baths
c(two = price(2000, 2), three = price(2000, 3), diff = price(2000, 3) - price(2000, 2))
```
:::

## Problem 4 — spot the error

A colleague fits $\widehat{\text{gdp}} = 0.02 + 0.5\,\text{rate\_10y} + 0.001\,\text{oil\_price}$
(where `oil_price` is in dollars, roughly 20–120) and concludes: *"`rate_10y` matters
far more than `oil_price`, because 0.5 is 500 times bigger than 0.001."* This contains
**exactly one mistake** — find it.

::: {.callout-note collapse="true" title="Answer"}
The mistake is comparing raw coefficient sizes across predictors on **different scales**.
`rate_10y` moves over a small range (a few units) while `oil_price` ranges over ~100
dollars, so `oil_price`'s coefficient is small *per dollar* but its total contribution
($0.001 \times 100 = 0.1$) is not negligible. Coefficient size is not importance unless
the predictors are on a common scale (e.g. standardized). The correct statement compares
either standardized coefficients or each coefficient times its predictor's spread.
:::

## Problem 5 — mixing it with linear combinations *(uses linear-combinations' weighted-sum-of-columns)*

The prediction for a whole dataset is a **linear combination of the predictor columns**.
Given columns $x_1 = (2, 4, 6)$ and $x_2 = (1, 0, 1)$ and coefficients $\beta_0 = 3$,
$\beta_1 = 1$, $\beta_2 = 10$, compute the vector of fitted values
$\hat{y} = \beta_0 + \beta_1 x_1 + \beta_2 x_2$ (an intercept added to every entry).

```{webr}
#| autorun: false
# p5-starter — build the fitted-value vector as a linear combination
x1 <- c(2, 4, 6); x2 <- c(1, 0, 1)
b0 <- 3; b1 <- 1; b2 <- 10
```

::: {.callout-note collapse="true" title="Answer"}
$\hat{y} = 3 + 1\cdot(2,4,6) + 10\cdot(1,0,1) = 3 + (2,4,6) + (10,0,10) = (15, 7, 19)$.
The intercept adds 3 to every entry; the rest is the warm-up's weighted sum of columns.

```{r}
x1 <- c(2, 4, 6); x2 <- c(1, 0, 1); b0 <- 3; b1 <- 1; b2 <- 10
b0 + b1 * x1 + b2 * x2      # 15 7 19
```
:::

## Problem 6 — a gentle cliffhanger

You fit `gdp` on two inflation measures, `cpi_infl` and `pce_infl`, that are correlated
0.95. On its own, `pce_infl` has a solidly negative slope with `gdp`. You predict its
coefficient in the *joint* model (with `cpi_infl` also included) will also come out
negative. Will it be? *(You can't fully settle this yet — the next module is about
exactly this.)*

::: {.callout-note collapse="true" title="Answer"}
Not necessarily — and that's the whole point of Module 3. When two predictors are
0.95-correlated, the joint fit can shrink one twin's coefficient toward zero or even flip
its sign, because "holding the other fixed" leaves almost no independent variation to
work with. The marginal (alone) slope and the joint coefficient answer different
questions and can disagree sharply. Module 3, *Marginal vs joint*, shows this happen.
:::
