---
title: "LDA, QDA & naive Bayes"
subtitle: "Classify by modeling what each class looks like — then let Bayes' theorem do the scoring"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c("MASS")
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```


## Where we're going

Some classifiers learn "which class does this point belong to?" directly from the
data (like kNN voting on neighbors). The methods in this lesson take a different
route: first model what each class **looks like** — its shape as a data cloud — then
use Bayes' theorem to flip that into "given this new point, which class is most
likely?" These are called **generative classifiers** because they model the process
that *generated* the data within each class.
[@james2021introductionstatisticallearning, pp. 141--142]

We will build three variants — LDA, QDA, and naive Bayes — that differ only in how
much detail they use to describe each class's shape. All three follow the same recipe:
estimate a prior, estimate a class-conditional density, plug into Bayes' theorem.

**Builds on:**
[multivariate normal distribution](../multivariate-normal-distribution/lesson.qmd) —
the "data cloud described by a mean vector and covariance matrix" that LDA and QDA
assume each class follows;
[classification evaluation](../classification-evaluation/lesson.qmd) — the confusion
matrix and accuracy measures we use to judge the result;
[priors, posteriors & updating](../priors-posteriors-and-updating/lesson.qmd) —
the Bayes flip that turns "how likely is this data given the class?" into "how likely
is this class given the data?"


## Warm-up: do you remember?

Before the new stuff, a few quick reps — answers from memory first, then check.

**1.** A bivariate normal has mean vector $\boldsymbol{\mu} = (5, 10)$ and covariance
matrix with diagonal entries 4 and 9 and off-diagonal 3. What is the standard
deviation of $X_1$?

::: {.callout-note collapse="true" title="Answer"}
$\text{sd}(X_1) = \sqrt{4} = 2$. The diagonal of the covariance matrix gives the
variance of each variable — taught in
[multivariate normal distribution](../multivariate-normal-distribution/lesson.qmd);
revisit if shaky.
:::

**2.** In the weather/umbrella table (100 days: 25 rainy, 75 dry; 30 with umbrella, 70
without; 20 rainy-with-umbrella), what is $P(\text{rain} \mid \text{umbrella})$? Say
the Bayes flip in one sentence: what does it turn into what?

::: {.callout-note collapse="true" title="Answer"}
$P(\text{rain} \mid \text{umbrella}) = 20/30 \approx 0.667$. The Bayes flip turns
"how likely is this evidence given the hypothesis?" into "how likely is the hypothesis
given the evidence?" — taught in
[priors, posteriors & updating](../priors-posteriors-and-updating/lesson.qmd); revisit
if shaky.
:::

**3.** You have a confusion matrix with 45 true positives, 5 false negatives, 10 false
positives, and 40 true negatives. What is the sensitivity?

::: {.callout-note collapse="true" title="Answer"}
Sensitivity = TP / (TP + FN) = 45 / 50 = 0.90. Sensitivity measures how well the
classifier catches actual positives — taught in
[classification evaluation](../classification-evaluation/lesson.qmd); revisit if shaky.
:::


## A tiny example first

Imagine two species of fish — **trout** and **bass** — measured on two variables:
length (cm) and weight (kg). You have 6 fish total, 3 of each:

| Fish | Species | Length | Weight |
|------|---------|--------|--------|
| 1    | trout   | 2      | 1      |
| 2    | trout   | 4      | 2      |
| 3    | trout   | 3      | 1      |
| 4    | bass    | 6      | 5      |
| 5    | bass    | 8      | 6      |
| 6    | bass    | 7      | 5      |

A new fish arrives with length 5 and weight 3. Which species?

**Step 1 — prior:** each species has 3 of 6 fish, so the prior for each class is 3/6 =
0.5. If you had no measurements at all, you would guess 50/50.

**Step 2 — class means:** trout center is at (average of 2, 4, 3; average of 1, 2, 1)
= (3, 1.33). Bass center is at (average of 6, 8, 7; average of 5, 6, 5) = (7, 5.33).

**Step 3 — how close is the new fish to each center?** The new fish at (5, 3) is
closer to the trout center (3, 1.33) by about 2.6 units and to the bass center (7,
5.33) by about 3.1 units. But "closer" on raw distance ignores how spread out each
class is — that is what the covariance matrix fixes. For now, just note: trout center
is closer.

**Step 4 — Bayes' theorem:** once we compute each class's density at (5, 3) —
meaning "how plausible is (5, 3) if this fish is a trout?" and the same for bass —
we multiply by the prior and compare. The class with the larger product wins. We will
do this formally after the notation, but the recipe is: **prior times density,
compare**.

```{r toy-example}
# The toy data: 6 fish, 2 species
species <- c("trout","trout","trout","bass","bass","bass")
length  <- c(2, 4, 3, 6, 8, 7)
weight  <- c(1, 2, 1, 5, 6, 5)

# Step 1: priors (class frequencies)
n_trout <- 3; n_bass <- 3; n <- 6
pi_trout <- n_trout / n    # 0.5
pi_bass  <- n_bass / n     # 0.5
c(pi_trout = pi_trout, pi_bass = pi_bass)

# Step 2: class means
mu_trout <- c(mean(length[1:3]), mean(weight[1:3]))   # (3.0, 1.33)
mu_bass  <- c(mean(length[4:6]), mean(weight[4:6]))    # (7.0, 5.33)
rbind(trout = mu_trout, bass = mu_bass)

# Step 3: Euclidean distance to each center (ignoring covariance for now)
new_fish <- c(5, 3)
dist_trout <- sqrt(sum((new_fish - mu_trout)^2))   # ~2.6
dist_bass  <- sqrt(sum((new_fish - mu_bass)^2))     # ~3.1
c(dist_trout = round(dist_trout, 2), dist_bass = round(dist_bass, 2))
```


## Getting a feel for it

**The mental picture: every class is a data cloud, and you classify by asking "which
cloud does this point fit inside better?"** You already know from the
[multivariate normal module](../multivariate-normal-distribution/lesson.qmd) that a
mean vector and covariance matrix describe a cloud — its center and its shape (how
wide, how tilted). LDA says: fit one cloud per class, then score a new point by
asking each cloud "how likely are you to have produced this point?" The cloud that
answers louder wins.

**Why this construction?** Because raw distance to the center is not enough. A class
whose cloud is narrow and tall should "claim" nearby points more strongly than a class
whose cloud is wide and flat, even if the centers are the same distance away. The
density — the height of the bell at that point — accounts for both the distance and
the shape. Bayes' theorem then folds in the prior (if bass are 90% of the lake, a
tie goes to bass).

**Where the analogy breaks:** real data does not always look like a Gaussian cloud.
If the true class shapes are banana-curved or multi-modal, the Gaussian model is
wrong. LDA and QDA work best when the classes are roughly elliptical; for wild
shapes, other classifiers (like kNN) may win.

**What the scores feel like.** In our toy example, the trout center is closer to (5,
3), so you would expect trout's density to be higher — and since the priors are
equal, trout should win the classification. When we formalize this with the density
function, you will see trout win by a concrete margin.


## What the notation is going to say

The formula will say: for each class $k$, multiply the prior probability $\pi_k$
("how common is class $k$?") by the class-conditional density $f_k(x)$ ("how
plausible is this measurement $x$ if it came from class $k$?"). Divide by the total
across all classes (so the results add to 1). The class whose product is biggest wins.
In our toy, that means: multiply trout's prior (0.5) by the multivariate normal
density at (5, 3) using trout's mean and covariance, do the same for bass, and pick
the larger one.


## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $x$ | "ex" | the measurement vector for the point we want to classify | $(5, 3)$ — the new fish |
| $K$ | "K" | number of classes | 2 (trout, bass) |
| $k$, $l$ | "k", "ell" | a particular class index ($l$ is just a second letter for the same kind of index, used when summing over all classes) | 1 = trout, 2 = bass |
| $p$ | "p" | number of predictors (measurements per observation) | 2 (length, weight) |
| $\pi_k$ | "pi-k" | prior probability of class $k$ (estimated as $n_k / n$) | $\pi_\text{trout} = 0.5$, $\pi_\text{bass} = 0.5$ |
| $f_k(x)$ | "f-k of x" | class-conditional density: how plausible is measurement $x$ if it came from class $k$ | the multivariate normal density evaluated at (5, 3) using each class's mean and covariance |
| $p_k(x)$ | "p-k of x" | posterior probability of class $k$ given $x$ — the thing we want | what we compute via Bayes' theorem |
| $\boldsymbol{\mu}_k$ | "mu-k" | mean vector of class $k$ | $\boldsymbol{\mu}_\text{trout} = (3, 1.33)$ |
| $\boldsymbol{\Sigma}$ | "sigma" (the matrix) | shared covariance matrix (LDA); in QDA each class has its own $\boldsymbol{\Sigma}_k$ | the pooled covariance computed from all 6 fish |
| $\delta_k(x)$ | "delta-k of x" | discriminant score for class $k$ — a shortcut that avoids computing the full density | (derived below) |
| $\lvert\boldsymbol{\Sigma}_k\rvert$ | "the determinant of sigma-k" | a single number measuring how spread out (how large) the covariance cloud is; in R: `det(Sigma_k)` | (used in QDA below) |
| $j$, $x_j$ | "j", "x-sub-j" | predictor index and the $j$-th predictor value | $j = 1$ is length, $j = 2$ is weight; $x_1 = 5$, $x_2 = 3$ |
| $f_{kj}(x_j)$ | "f-k-j of x-sub-j" | the one-dimensional density of predictor $j$ within class $k$ (used in naive Bayes) | the univariate normal density of length within trout, evaluated at 5 |
| $\prod_{j=1}^{p}$ | "the product from j equals 1 to p" | multiply instead of add (in R: `prod(...)`) | multiply the length density by the weight density |


## The formal version

### The classification rule

Bayes' theorem for classification
[@james2021introductionstatisticallearning, pp. 141--142, Eq. 4.15]:

$$
p_k(x) = \frac{\pi_k \, f_k(x)}{\sum_{l=1}^{K} \pi_l \, f_l(x)}
$$

Reading it aloud: "p-k of x equals pi-k times f-k of x, divided by the sum over all
classes of pi-l times f-l of x" — or in everyday words: "the posterior probability of
class $k$ is that class's prior times its density, divided by the total density across
all classes."

Mapping it onto the toy example:

- $\pi_k$ is the 0.5 we computed from counting fish
- $f_k(x)$ is the multivariate normal density at (5, 3) using class $k$'s mean and
  covariance — the "how plausible" number
- the denominator just makes the posteriors add to 1

Classify to the class with the largest $p_k(x)$.

### LDA: shared covariance, linear boundary

LDA assumes every class is a multivariate normal with its own mean $\boldsymbol{\mu}_k$
but a **shared** covariance matrix $\boldsymbol{\Sigma}$
[@james2021introductionstatisticallearning, pp. 142--148]. Because the covariance
is the same for all classes, the quadratic terms in the Gaussian exponent cancel
when you compare two classes, and what remains is a **linear** function of $x$ — the
discriminant score:

$$
\delta_k(x) = x^\top \boldsymbol{\Sigma}^{-1} \boldsymbol{\mu}_k
  - \tfrac{1}{2}\,\boldsymbol{\mu}_k^\top \boldsymbol{\Sigma}^{-1} \boldsymbol{\mu}_k
  + \log \pi_k
$$

Reading it aloud: "delta-k of x equals x-transpose times sigma-inverse times mu-k,
minus one-half mu-k-transpose times sigma-inverse times mu-k, plus log of pi-k" — or
in everyday words: "a score that combines how close $x$ is to class $k$'s center
(accounting for the shape of the cloud) with a bonus for classes that are more common."

Classify to the class with the largest $\delta_k(x)$. The boundary between two classes
is the set of points where $\delta_k(x) = \delta_l(x)$ — and because $\delta$ is
linear in $x$, that boundary is a **straight line** (in 2D) or a flat plane (in
higher dimensions). That is why it is called *linear* discriminant analysis.

### Hand-scoring our toy fish

Now let's do the full LDA computation by hand, then confirm with R.

```{r lda-by-hand}
# The toy data again
trout_x <- cbind(c(2, 4, 3), c(1, 2, 1))   # 3 fish x 2 measurements
bass_x  <- cbind(c(6, 8, 7), c(5, 6, 5))

# Class means
mu_t <- colMeans(trout_x)   # (3.0, 1.333)
mu_b <- colMeans(bass_x)    # (7.0, 5.333)

# POOLED covariance: average the within-class covariance matrices
#  (each has n_k - 1 = 2 degrees of freedom; pool = sum of deviations / (n - K))
S_t <- cov(trout_x)   # within-trout covariance
S_b <- cov(bass_x)    # within-bass covariance
Sigma_pooled <- ((3 - 1) * S_t + (3 - 1) * S_b) / (6 - 2)
Sigma_pooled

# Inverse of pooled covariance
Sigma_inv <- solve(Sigma_pooled)

# Discriminant scores at x = (5, 3)
x_new <- c(5, 3)
pi_t <- 0.5; pi_b <- 0.5

delta_trout <- as.numeric(x_new %*% Sigma_inv %*% mu_t
  - 0.5 * mu_t %*% Sigma_inv %*% mu_t + log(pi_t))
delta_bass <- as.numeric(x_new %*% Sigma_inv %*% mu_b
  - 0.5 * mu_b %*% Sigma_inv %*% mu_b + log(pi_b))

c(delta_trout = round(delta_trout, 3), delta_bass = round(delta_bass, 3))
# The larger score wins
```

```{webr}
#| autorun: true
#| fig-alt: "Scatter plot of 6 fish (3 trout at lower-left, 3 bass at upper-right) with the new fish at (5,3) marked as a triangle. The LDA decision boundary is a straight line between the two classes."
# The toy data
trout_len <- c(2, 4, 3); trout_wt <- c(1, 2, 1)
bass_len  <- c(6, 8, 7); bass_wt  <- c(5, 6, 5)

# TRY IT: move the new fish and re-run — does the classification change?
new_fish <- c(5, 3)

# Pooled covariance and its inverse
trout_x <- cbind(trout_len, trout_wt)
bass_x  <- cbind(bass_len, bass_wt)
S_pool <- ((2) * cov(trout_x) + (2) * cov(bass_x)) / 4
S_inv  <- solve(S_pool)
mu_t <- colMeans(trout_x); mu_b <- colMeans(bass_x)

# Discriminant scores
d_t <- as.numeric(new_fish %*% S_inv %*% mu_t - 0.5 * mu_t %*% S_inv %*% mu_t)
d_b <- as.numeric(new_fish %*% S_inv %*% mu_b - 0.5 * mu_b %*% S_inv %*% mu_b)

# Plot
plot(c(trout_len, bass_len), c(trout_wt, bass_wt),
     col = c(rep("steelblue", 3), rep("coral", 3)), pch = 19, cex = 1.5,
     xlab = "Length", ylab = "Weight", xlim = c(0, 10), ylim = c(0, 8),
     main = paste("LDA classifies new fish as:", ifelse(d_t > d_b, "trout", "bass")))
points(new_fish[1], new_fish[2], pch = 17, cex = 2, col = "black")
text(new_fish[1] + 0.3, new_fish[2] + 0.4, "new fish", cex = 0.8)
legend("topleft", legend = c("trout", "bass", "new fish"),
       col = c("steelblue", "coral", "black"), pch = c(19, 19, 17))

# The decision boundary: where delta_trout = delta_bass
# delta_k(x) = x' S_inv mu_k - 0.5 mu_k' S_inv mu_k  (equal priors cancel)
# boundary: x' S_inv (mu_t - mu_b) = 0.5 (mu_t' S_inv mu_t - mu_b' S_inv mu_b)
w <- S_inv %*% (mu_t - mu_b)
c_val <- 0.5 * as.numeric(mu_t %*% S_inv %*% mu_t - mu_b %*% S_inv %*% mu_b)
# w[1]*x1 + w[2]*x2 = c_val  =>  x2 = (c_val - w[1]*x1) / w[2]
x1_seq <- seq(0, 10, length.out = 100)
x2_boundary <- (c_val - w[1] * x1_seq) / w[2]
lines(x1_seq, x2_boundary, lty = 2, lwd = 2, col = "gray40")
```


### QDA: per-class covariance, curved boundary

LDA forced both classes to share one covariance matrix. But look at the within-class
covariances for our fish:

```{r qda-toy}
trout_x <- cbind(c(2, 4, 3), c(1, 2, 1))
bass_x  <- cbind(c(6, 8, 7), c(5, 6, 5))
cat("Trout covariance:\n"); print(cov(trout_x))
cat("Bass covariance:\n");  print(cov(bass_x))
```

Trout's length-weight covariance is positive (longer trout tend to be heavier); bass
has a negative off-diagonal (in these 3 fish, the heaviest is not the longest). Pooling
hides that difference. QDA says: let each class keep its own
$\boldsymbol{\Sigma}_k$.
[@james2021introductionstatisticallearning, pp. 152--154]

Because each class now has a different covariance, the quadratic terms in the Gaussian
exponent **do not cancel** when you compare two classes. The discriminant score becomes
quadratic in $x$:

$$
\delta_k(x) = -\tfrac{1}{2}\,x^\top \boldsymbol{\Sigma}_k^{-1} x
  + x^\top \boldsymbol{\Sigma}_k^{-1} \boldsymbol{\mu}_k
  - \tfrac{1}{2}\,\boldsymbol{\mu}_k^\top \boldsymbol{\Sigma}_k^{-1} \boldsymbol{\mu}_k
  - \tfrac{1}{2}\,\log\lvert\boldsymbol{\Sigma}_k\rvert
  + \log \pi_k
$$

Reading it aloud: "delta-k of x equals negative one-half x-transpose sigma-k-inverse
x, plus x-transpose sigma-k-inverse mu-k, minus one-half mu-k-transpose
sigma-k-inverse mu-k, minus one-half the log-determinant of sigma-k, plus log pi-k"
— the same kind of distance-to-center score as LDA, but now each class uses its own
sigma-k, and there is an extra log-determinant term that penalizes classes with
larger (more spread-out) covariance. The boundary where $\delta_k = \delta_l$ is now a
**curve** — a quadratic function of $x$ — hence "quadratic" discriminant analysis.

Mapping it onto the toy: with only 3 fish per class and $p = 2$ predictors, each class
estimates 3 covariance parameters (variance of length, variance of weight, their
covariance). LDA estimates 3 total (shared); QDA estimates $2 \times 3 = 6$ from only
6 fish — that is why QDA can overfit with small samples.

**When to use QDA over LDA:** if you believe the classes truly have different spreads
or tilts (different covariance matrices), QDA can capture that. But QDA estimates many
more parameters ($K \times p(p+1)/2$ covariance entries instead of one shared set of
$p(p+1)/2$), so with small samples it can overfit. LDA is the safer default when $n$
is modest relative to $p$.
[@james2021introductionstatisticallearning, pp. 153--154]


### Naive Bayes: drop the covariance entirely

Here is a third option. Instead of estimating a full $p \times p$ covariance matrix (3
parameters for our 2 predictors), what if we pretend length and weight are independent
within each species? Then we only need each predictor's own variance — no covariance.
For trout: variance of length = 1, variance of weight = 0.33. For bass: variance of
length = 1, variance of weight = 0.33. That is 2 variances per class instead of 3
covariance parameters — and the savings grow fast with more predictors.

This is the **naive Bayes** assumption: the predictors are **independent within each
class**. [@james2021introductionstatisticallearning, pp. 154--158]

The class-conditional density factors into a product of one-dimensional densities:

$$
f_k(x) = \prod_{j=1}^{p} f_{kj}(x_j)
$$

Reading it aloud: "f-k of x equals the product from j equals 1 to p of f-k-j of
x-sub-j" — or in everyday words: "the class-conditional density is just each
predictor's density multiplied together, as if they don't talk to each other."

In our toy: the naive Bayes density for trout at (5, 3) is (density of length = 5
under a normal with mean 3, variance 1) $\times$ (density of weight = 3 under a
normal with mean 1.33, variance 0.33). We just multiply the two bell curves
independently.

```{r nb-toy}
# Naive Bayes for our toy fish — each predictor separately
trout_len <- c(2, 4, 3); trout_wt <- c(1, 2, 1)
bass_len  <- c(6, 8, 7); bass_wt  <- c(5, 6, 5)

# Density at new fish (5, 3) under each class, assuming independence
nb_trout <- dnorm(5, mean(trout_len), sd(trout_len)) *
            dnorm(3, mean(trout_wt),  sd(trout_wt))
nb_bass  <- dnorm(5, mean(bass_len),  sd(bass_len)) *
            dnorm(3, mean(bass_wt),   sd(bass_wt))
cat("NB density — trout:", round(nb_trout, 6), " bass:", round(nb_bass, 6), "\n")
cat("NB classifies as:", ifelse(nb_trout > nb_bass, "trout", "bass"), "\n")
```

This is a big assumption — it says length and weight carry no information about each
other within a species, which is probably wrong. But it dramatically reduces the
number of parameters: instead of a full $p \times p$ covariance matrix, you estimate
only $p$ variances per class. With many predictors and limited data, that variance
reduction can more than compensate for the bias of the wrong independence assumption.
[@james2021introductionstatisticallearning, pp. 156--158]


## Explore it

### LDA with `MASS::lda()`

Let's verify our hand computation with R's `MASS::lda()` and then try a bigger
dataset.

```{webr}
#| autorun: false
# Confirm our hand result matches MASS::lda()
library(MASS)
species <- factor(c("trout","trout","trout","bass","bass","bass"))
fish_data <- data.frame(length = c(2,4,3,6,8,7), weight = c(1,2,1,5,6,5))
fit <- lda(species ~ length + weight, data = fish_data)

# PREDICT: based on our hand calculation, which species should this return?
mass_result <- as.character(predict(fit, newdata = data.frame(length = 5, weight = 3))$class)

# Our hand computation (from the baked chunk above)
trout_x <- cbind(c(2,4,3), c(1,2,1)); bass_x <- cbind(c(6,8,7), c(5,6,5))
S_pool <- ((2)*cov(trout_x) + (2)*cov(bass_x)) / 4; S_inv <- solve(S_pool)
mu_t <- colMeans(trout_x); mu_b <- colMeans(bass_x); x <- c(5, 3)
d_t <- as.numeric(x %*% S_inv %*% mu_t - 0.5 * mu_t %*% S_inv %*% mu_t)
d_b <- as.numeric(x %*% S_inv %*% mu_b - 0.5 * mu_b %*% S_inv %*% mu_b)
hand_result <- ifelse(d_t > d_b, "trout", "bass")

cat("Hand computation:", hand_result, "\n")
cat("MASS::lda():    ", mass_result, "\n")
cat("Match:", hand_result == mass_result, "\n")
```

```{webr}
#| autorun: false
# TRY IT: change the new fish's measurements — at what length does the class flip?
# PREDICT: write down your guess, then run.
library(MASS)
species <- factor(c("trout","trout","trout","bass","bass","bass"))
fish_data <- data.frame(length = c(2,4,3,6,8,7), weight = c(1,2,1,5,6,5))
fit <- lda(species ~ length + weight, data = fish_data)

# Try a range of lengths at weight = 3
for (len in 3:7) {
  pred <- predict(fit, newdata = data.frame(length = len, weight = 3))
  cat("length =", len, "→", as.character(pred$class),
      " posterior:", round(pred$posterior, 3), "\n")
}
```

### LDA on the iris dataset

```{r lda-iris}
#| fig-alt: "LDA projection of the iris dataset onto the first two discriminant axes, showing three species as colored clusters: setosa clearly separated, versicolor and virginica partially overlapping."
library(MASS)
fit_iris <- lda(Species ~ ., data = iris)

# Project onto the discriminant axes
proj <- as.matrix(iris[, 1:4]) %*% fit_iris$scaling
cols <- c(setosa = "steelblue", versicolor = "coral", virginica = "forestgreen")
plot(proj[, 1], proj[, 2], col = cols[iris$Species], pch = 19,
     xlab = "LD1", ylab = "LD2", main = "LDA projection of iris")
legend("topright", legend = names(cols), col = cols, pch = 19, cex = 0.8)

# Accuracy on training data
preds <- predict(fit_iris)$class
cat("Training accuracy:", mean(preds == iris$Species), "\n")
```


## Common traps

**Trap: "LDA needs the features to be normally distributed — if they're not, it
fails."** LDA *assumes* Gaussian class-conditional densities, but it does not
"fail" on non-normal data — it still produces a linear boundary, and that boundary
may work fine in practice. Our toy fish data has only 3 points per class (nowhere
near a Gaussian bell), yet LDA still correctly classified the new fish as trout. The
real danger is not non-normality but when the classes have very different covariance
shapes: then the shared-covariance assumption hurts, and QDA is the better choice.
[@james2021introductionstatisticallearning, pp. 163--165]

**Trap: "QDA is always better than LDA because it is more flexible."** More
flexibility means more parameters. In our toy, QDA estimates $2 \times 3 = 6$
covariance parameters from 6 fish — one parameter per data point, so every
covariance estimate is pure noise. LDA estimates only 3, pooling across classes, so
each estimate has twice the data behind it. With small $n$, the extra parameters in
QDA add variance without reducing bias.
[@james2021introductionstatisticallearning, pp. 153--154]


## Check yourself

**1.** In our toy fish example, both priors are 0.5. If the lake were 80% bass and
20% trout, what would $\pi_\text{bass}$ be, and how would that change the LDA
discriminant score for bass (higher or lower)?

::: {.callout-note collapse="true" title="Answer"}
$\pi_\text{bass} = 0.8$. The score $\delta_\text{bass}$ includes a $+\log \pi_k$
term, and $\log(0.8) > \log(0.5)$, so the bass score goes **up**. A more common
class gets a bonus in the discriminant score — ties go to the more common species.
:::

**2.** LDA uses one shared covariance matrix. QDA uses one per class. Which estimates
more parameters, and what is the practical risk of estimating more?

::: {.callout-note collapse="true" title="Answer"}
QDA estimates more parameters ($K$ covariance matrices instead of one). The risk is
overfitting: with small $n$ relative to $p$, those extra parameters are estimated
noisily and the curved boundary chases noise rather than signal.
:::

**3.** In your own words: explain to a colleague, in two sentences, what LDA does and
why it uses Bayes' theorem.

::: {.callout-note collapse="true" title="A model answer"}
LDA models each class as a Gaussian cloud (same shape, different centers) and asks
"how plausible is this new point under each cloud?" Bayes' theorem then flips those
plausibilities into posterior probabilities — "given this point, which class is most
likely?" — folding in how common each class is.
:::


## What you can now do

- Compute a prior from class frequencies and say what role it plays in Bayes' theorem
  for classification
- Hand-score a new point under LDA: compute the pooled covariance, the discriminant
  scores, and say which class wins
- Explain why LDA's boundary is linear (shared covariance cancels the quadratic
  terms) and QDA's boundary is curved (per-class covariance does not cancel)
- State the naive Bayes independence assumption and explain the bias-variance
  trade-off it creates

If any of these feels shaky, revisit the relevant section above before continuing —
the next module that uses classifiers assumes all of them.


## Where this goes next

These three methods are part of a larger toolbox of classifiers — you already know
[kNN](../k-nearest-neighbors/lesson.qmd) (a discriminative, distance-based approach),
and later modules will add tree-based classifiers that partition the predictor space
in a completely different way. The choice among methods depends on the true decision
boundary shape, the sample size, and the number of predictors — no single classifier
dominates. [@james2021introductionstatisticallearning, pp. 166--170]

Ready for more reps? Do the [practice problems](practice.qmd).
