---
title: "Practice: LDA, QDA & naive Bayes"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c("MASS")
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Do them on paper or in an R session; only then
open the answer.


## Problem 1 — confidence rep with new numbers

Two classes (A and B), two predictors ($x_1$, $x_2$), three observations each:

| Class | $x_1$ | $x_2$ |
|-------|--------|--------|
| A     | 1      | 2      |
| A     | 3      | 4      |
| A     | 2      | 2      |
| B     | 7      | 8      |
| B     | 9      | 6      |
| B     | 8      | 7      |

A new point arrives at (5, 5). Compute the prior for each class and the class mean
vectors. Which class center is closer (Euclidean distance)?

```{webr}
#| autorun: false
# p1-starter
A_x1 <- c(1, 3, 2); A_x2 <- c(2, 4, 2)
B_x1 <- c(7, 9, 8); B_x2 <- c(8, 6, 7)
new_pt <- c(5, 5)
```

::: {.callout-note collapse="true" title="Answer"}
Priors: $\pi_A = \pi_B = 3/6 = 0.5$.
Mean vectors: $\boldsymbol{\mu}_A = (2, 2.67)$, $\boldsymbol{\mu}_B = (8, 7)$.
Distance from (5, 5) to $\boldsymbol{\mu}_A$: $\sqrt{(5-2)^2 + (5-2.67)^2} = \sqrt{9 + 5.44} \approx 3.80$.
Distance to $\boldsymbol{\mu}_B$: $\sqrt{(5-8)^2 + (5-7)^2} = \sqrt{9 + 4} \approx 3.61$.
Class B's center is closer by Euclidean distance.
:::


## Problem 2 — faded worked example

Below is an LDA computation for the lesson's toy fish data (new fish at (5, 3)).
One step is blanked out — fill in the missing piece.

- Step 1: priors are $\pi_\text{trout} = 0.5$, $\pi_\text{bass} = 0.5$.
- Step 2: class means are $\boldsymbol{\mu}_\text{trout} = (3, 1.33)$,
  $\boldsymbol{\mu}_\text{bass} = (7, 5.33)$.
- Step 3: the pooled covariance and its inverse are given:

  $\boldsymbol{\Sigma}_\text{pooled} = \begin{pmatrix} 1 & 0.5 \\ 0.5 & 0.333 \end{pmatrix}$, $\quad \boldsymbol{\Sigma}^{-1} = \begin{pmatrix} 4 & -6 \\ -6 & 12 \end{pmatrix}$

- Step 4: **your step here** — compute the discriminant score for trout:
  $\delta_\text{trout}(x) = x^\top \boldsymbol{\Sigma}^{-1} \boldsymbol{\mu}_\text{trout} - \tfrac{1}{2}\,\boldsymbol{\mu}_\text{trout}^\top \boldsymbol{\Sigma}^{-1} \boldsymbol{\mu}_\text{trout} + \log(0.5)$.
  Plug in $x = (5, 3)$ and $\boldsymbol{\mu}_\text{trout} = (3, 1.33)$.
- Step 5: compare $\delta_\text{trout}$ to $\delta_\text{bass}$ — the larger wins.

::: {.callout-note collapse="true" title="Answer"}
Step 4: first compute $\boldsymbol{\Sigma}^{-1} \boldsymbol{\mu}_\text{trout}$:
$(4 \times 3 + (-6) \times 1.33,\; -6 \times 3 + 12 \times 1.33) = (4, -2)$.
Then $x^\top$ times that: $5 \times 4 + 3 \times (-2) = 14$.
The constant: $\tfrac{1}{2} \times (3 \times 4 + 1.33 \times (-2)) = \tfrac{1}{2}(12 - 2.67) = 4.67$.
So $\delta_\text{trout} = 14 - 4.67 + \log(0.5) = 14 - 4.67 - 0.69 = 8.64$.
(The lesson's R code gives the same value; trout wins because $\delta_\text{trout} > \delta_\text{bass}$.)
:::


## Problem 3 — full problem, by hand then in R

Using the Problem 1 data (classes A and B), compute the pooled covariance matrix.
Then compute the LDA discriminant scores $\delta_A(x)$ and $\delta_B(x)$ at the new
point (5, 5). Which class does LDA assign?

```{webr}
#| autorun: false
# p3-starter
# PREDICT: class B's center is closer — does LDA agree or does the covariance change things?
A <- cbind(c(1, 3, 2), c(2, 4, 2))
B <- cbind(c(7, 9, 8), c(8, 6, 7))
S_A <- cov(A); S_B <- cov(B)
S_pool <- ((3 - 1) * S_A + (3 - 1) * S_B) / (6 - 2)
```

::: {.callout-note collapse="true" title="Answer"}
```{r}
A <- cbind(c(1, 3, 2), c(2, 4, 2))
B <- cbind(c(7, 9, 8), c(8, 6, 7))
S_pool <- ((2) * cov(A) + (2) * cov(B)) / 4
S_inv <- solve(S_pool)
mu_A <- colMeans(A); mu_B <- colMeans(B)
x <- c(5, 5)
delta_A <- as.numeric(x %*% S_inv %*% mu_A - 0.5 * mu_A %*% S_inv %*% mu_A)
delta_B <- as.numeric(x %*% S_inv %*% mu_B - 0.5 * mu_B %*% S_inv %*% mu_B)
c(delta_A = round(delta_A, 3), delta_B = round(delta_B, 3))
```

LDA assigns to class B ($\delta_B > \delta_A$), agreeing with the Euclidean distance
result. The pooled covariance is not the identity here (A and B have different
covariance structures), but in this case the covariance adjustment reinforces rather
than overturns the raw-distance ranking.
:::


## Problem 4 — spot the error

A colleague says: *"I ran LDA on my 500 patients (20 features) and got 80% accuracy.
Then I ran QDA on the same data and got 72% — QDA must be worse than LDA."* This
contains **exactly one mistake** — find it.

::: {.callout-note collapse="true" title="Answer"}
The mistake: concluding QDA is worse in general from one dataset. QDA estimates far
more parameters ($20 \times 21 / 2 = 210$ covariance entries per class vs. one shared set), so with only
500 observations and 20 features, QDA likely **overfitted**. The lower accuracy
reflects high variance from too many parameters, not that QDA is a worse method. With
more data or fewer features, QDA could outperform LDA — the right conclusion is "QDA
needs more data than LDA to work well here," not "QDA is worse."
:::


## Problem 5 — interleaved *(uses classification evaluation's confusion matrix)*

You fit LDA to a test set and get this confusion matrix:

|              | Predicted A | Predicted B |
|--------------|:-----------:|:-----------:|
| **Actual A** | 35          | 15          |
| **Actual B** | 5           | 45          |

(a) What is the sensitivity for class A (treating A as the "positive" class)?
(b) The LDA model assumed a shared covariance. If you suspect class B is much more
spread out than class A, which method should you try instead, and why?

::: {.callout-note collapse="true" title="Answer"}
(a) Sensitivity = TP / (TP + FN) = 35 / (35 + 15) = 35/50 = 0.70.

(b) Try **QDA**, because it gives each class its own covariance matrix. If B truly
has a larger or differently shaped covariance than A, the shared-covariance
assumption in LDA is wrong, and QDA can capture the difference — at the cost of
estimating more parameters.
:::


## Problem 6 — transfer: when does naive Bayes beat LDA?

You have 50 observations and 40 predictors. Both LDA and QDA perform poorly. A
colleague suggests naive Bayes. Explain, in terms of the number of parameters each
method estimates, why naive Bayes might work better here.

::: {.callout-note collapse="true" title="Answer"}
With $p = 40$ predictors, LDA estimates one $40 \times 40$ covariance matrix (820
unique entries). QDA estimates one per class — even worse. With only 50 observations,
those covariance estimates are extremely noisy. Naive Bayes assumes independence, so
it estimates only $p = 40$ variances per class (one per predictor) instead of a full
covariance matrix. That is 40 parameters per class instead of 820. The independence
assumption is probably wrong, but the massive reduction in parameters to estimate
(the variance side of the bias-variance trade-off) can more than compensate —
especially when $n$ is small relative to $p$.
:::


## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/lda-qda-and-naive-bayes" pass="4"}

::: {.quiz-q qid="lqnb-bayes-recipe" qtype="mc" answer="b"}
What do LDA, QDA, and naive Bayes all have in common?

a) They use the same shared covariance matrix for every class in the dataset
b) They estimate a class-conditional density and plug it into Bayes' theorem
c) They assume all predictors are conditionally independent within each class
:::

::: {.quiz-q qid="lqnb-lda-boundary" qtype="mc" answer="a"}
Why is LDA's decision boundary a straight line (in 2D)?

a) The shared covariance cancels the quadratic terms when comparing classes
b) LDA ignores the covariance matrix and uses raw Euclidean distance only
c) LDA restricts coefficients to be equal, which forces a linear function
:::

::: {.quiz-q qid="lqnb-qda-risk" qtype="mc" answer="c"}
What is the main practical risk of using QDA instead of LDA?

a) QDA cannot handle more than two classes at a time in its model
b) QDA ignores the prior class probabilities when scoring new points
c) QDA estimates more parameters and can overfit with small samples
:::

::: {.quiz-q qid="lqnb-nb-assumption" qtype="mc" answer="b"}
Naive Bayes assumes that within each class, the predictors are:

a) All equal in variance
b) Independent of each other
c) Drawn from a uniform distribution
:::

::: {.quiz-q qid="lqnb-prior-role" qtype="mc" answer="a"}
(From priors, posteriors & updating.) In LDA's discriminant score $\delta_k(x)$, the $\log \pi_k$ term means that:

a) A more common class gets a higher score, all else equal
b) The class with the smallest variance always wins
c) The prior has no effect on the classification
:::

:::

::::
