---
title: "Randomness, samples & sampling distributions"
subtitle: "Why a number computed from data is itself random — and what its wobble (the standard error) means"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live cells below re-state their own data. This module
# USES randomness, so the seed is what makes every histogram reproducible. Base R only.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

::: {.callout-tip title="This lesson runs in your browser" collapse="true"}
The code cells below with a **Run** button are live — powered by WebR (R compiled to
WebAssembly), they run right here on your phone or laptop, no install needed. Edit any
of them and re-run. The first run downloads a small R engine, so give it a few seconds.
Cells without a Run button are ordinary listings whose output was computed when the
page was built.
:::

## Where we're going

Every number you ever compute from data — an average, a proportion, a slope — is
computed from *one* sample, and if you had drawn a *different* sample you'd have gotten
a *different* number. So the number itself is random: it has its own spread, its own
"typical wobble." That wobble has a name, the **standard error**, and it is the single
idea that separates "here is a number" from "here is how much I trust this number."
This lesson builds the idea by *simulating* the re-draws in R and watching the number
jump around.

We stay with neutral, hand-checkable examples (dice, a bag of numbered chips, a
machine that spits out measurements) and base R. We will **not** touch p-values, null
hypotheses, or how to *read* a confidence interval's endpoints — that machinery belongs
to a later hypothesis-testing module. Here we build only the foundation those tools
stand on: a statistic is random, its spread is the standard error, and for averages
that spread turns out bell-shaped.

**Builds on:** [the mean, variance & covariance](../mean-variance-covariance/lesson.qmd) —
we lean on the mean $\bar{x}$ and the standard deviation $s$ throughout; and
[probability, odds & log-odds](../probability-and-odds/lesson.qmd) (soft — for the
baseline idea that a random draw has probabilities attached).

## Warm-up: do you remember?

Before the new stuff, three quick reps — produce each answer from memory, then check.

**1.** *(mean-variance-covariance)* For `x <- c(2, 4, 6)`, what is the mean $\bar{x}$,
and what is the standard deviation `sd(x)` measuring in one plain phrase?

```{webr}
#| autorun: false
# warmup1-starter — edit and Run to check
x <- c(2, 4, 6)
```

::: {.callout-note collapse="true" title="Answer"}
$\bar{x} = (2 + 4 + 6)/3 = 4$. The standard deviation measures **how far a typical data
point sits from that mean** — the size of the spread of the numbers *in the data*. In R:
`mean(x)` is 4; `sd(x)` is 2. Keep the phrase "spread of the data" in mind — halfway
through this lesson we meet a *different* spread that is easy to confuse with it. Shaky?
Revisit [mean, variance & covariance](../mean-variance-covariance/lesson.qmd).
:::

**2.** *(probability-and-odds / baseline)* You roll one fair six-sided die. What is the
probability of each face, and what is the *average* face value you'd expect over many,
many rolls?

::: {.callout-note collapse="true" title="Answer"}
Each face has probability $1/6 \approx 0.167$. The long-run average face is
$(1+2+3+4+5+6)/6 = 3.5$. That "what you'd get on average over many repeats" idea is the
engine of this whole lesson — except we'll repeat *whole samples*, not single rolls. See
[probability, odds & log-odds](../probability-and-odds/lesson.qmd).
:::

**3.** In R, `sample(1:6, 4, replace = TRUE)` does what, in plain words? And what does
`replicate(3, mean(rnorm(5)))` hand you back?

::: {.callout-note collapse="true" title="Answer"}
`sample(1:6, 4, replace = TRUE)` rolls a fair die **4 times** (with replacement, so the
same face can repeat) and returns those 4 values. `replicate(3, mean(rnorm(5)))` runs
"draw 5 random normal values, take their mean" **3 separate times** and returns the 3
means as a vector. `replicate()` is the "do this whole thing over and over" tool at the
heart of this lesson.
:::

## A tiny example first

You have a bag of chips numbered **1, 2, 3, 4, 5, 6** — like a fair die. The *whole
bag* is the **population**: everything you could possibly draw. You don't get to
average the whole bag; instead you reach in and pull out a **sample** of 4 chips (with
replacement — you note the number and toss it back). From that sample you compute one
summary — say, the **mean** of the 4 numbers. That single summary computed from a
sample is called a **statistic**.

Here is the key move. Draw one sample of 4 and average it. Now *redraw* — a fresh grab
of 4 chips — and average again. The two averages will almost never be the same number.

Let's do exactly that in R, with a fixed seed so you get these same draws:

```{r toy-example}
set.seed(42)

# The POPULATION: a bag of chips numbered 1..6 (a fair die).
# One SAMPLE of 4 chips, drawn with replacement:
s1 <- sample(1:6, 4, replace = TRUE)
s1                       # 1 5 1 1
mean(s1)                 # the STATISTIC (sample mean): 2

# REDRAW: a fresh sample of 4 from the same bag
s2 <- sample(1:6, 4, replace = TRUE)
s2                       # 2 4 2 2
mean(s2)                 # a DIFFERENT statistic: 2.5
```

Same bag, same recipe ("grab 4, take the mean"), two different answers: **2** and
**2.5**. Nothing was done wrong — the *sample* changed, so the *statistic* changed.
That is the entire seed of this lesson: **a statistic is random.** The number you
report is one roll of a die you didn't know you were rolling.

## Getting a feel for it

**The mental picture: a lottery machine that spits out numbers.** Think of "compute the
mean of a sample of 4" as a single lottery machine. You press the button (draw a sample,
average it) and out pops a number. Press it again: a different number. The population —
the bag of chips — never changes; the machine's *rule* never changes; but each press
gives a fresh output because the draw is fresh. The population has a mean (for the fair
bag it's $3.5$) sitting quietly behind the curtain, and the machine's outputs scatter
*around* that hidden value. Where the analogy strains: a real lottery is designed to be
unpredictable on purpose, whereas here the randomness is just the unavoidable luck of
*which* few chips you happened to grab — it shrinks if you grab more chips per press.

**Why we even care that the statistic is random.** Because you almost always get to
press the button *once*. You have one dataset, you compute one mean, and you have to
decide how much to trust it. If pressing the button again would give you something wildly
different, your one number is shaky; if repeated presses cluster tightly, your one number
is trustworthy. So the useful question is never "what is my number?" — it's "how much
would my number *jump around* if I could redraw?" That jump-around-size is the whole
prize, and it has a name: the **standard error**.

**What the wobble feels like, and the one trap baked into it.** With only 4 chips per
draw, the two means above (2 and 2.5) landed a fair bit below the true 3.5 and differed
from each other — small samples wobble a lot. Grab 100 chips per draw instead of 4 and
the means would huddle tightly around 3.5. So the standard error *shrinks as the sample
grows* — more data, steadier number. Here is the trap to plant a flag on now: the
standard error is **not** the same as the spread of the data (the `sd(x)` from the
warm-up). The data's spread barely changes with sample size; the statistic's spread
collapses as the sample grows. They are two different spreads, and confusing them is the
most common mistake in this whole area. We'll pin the difference down with numbers below.

## What the notation is going to say

There is only one small formula in this lesson, and it says something you can already
picture. We are going to build the histogram of "the statistic, over many redraws" —
that histogram is called the **sampling distribution** of the statistic. Its *center*
sits at the true population value, and its *standard deviation* is the standard error.
So the formula will just say: **the standard error is the standard deviation of the
sampling distribution** — take all those redrawn statistics and compute their `sd()`,
exactly the `sd()` you already know from the last module, only applied to the *list of
redrawn numbers* instead of to one sample's raw data.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $\bar{x}$ | "ex bar" | the mean of one sample — here, the **statistic** we track (in R: `mean(x)`) | 2 for the first sample of 4 |
| $s$ | "ess" | the standard deviation of the data *in one sample* (in R: `sd(x)`) | ≈ 19.6 for the big sample below |
| $\mu$ | "mew" | the true population mean — the hidden value the statistic scatters around | 3.5 for the fair bag; 100 below |
| $\sigma$ | "sigma (lowercase)" | the true population standard deviation (the spread of the whole population) | 15 for the machine below |
| $n$ | "en" | the sample size — how many items in one draw (in R: `length(x)`) | 4 chips, or 25 below |
| $\mathrm{SE}$ | "the standard error" | the standard deviation of the statistic across redraws — the wobble of the number itself | ≈ 3 below |
| $\bar{x}_1, \bar{x}_2, \dots, \bar{x}_R$ | "ex bar sub one, … up to ex bar sub R" | the statistic from redraw 1, redraw 2, …, redraw $R$ — the list we histogram | 2, 2.5, … |
| $R$ | "big R" | how many times we redraw the sample in the simulation (in R: `replicate(R, ...)`) | 2000 below |

Greek letters ($\mu$, $\sigma$) are the *true, hidden* population quantities; the Latin
$\bar{x}$ and $s$ are what we *compute from a sample*. The standard error, $\mathrm{SE}$,
is a spread — but a spread *of statistics*, not of raw data, which is why it gets its own
name instead of just being called a standard deviation.

## The formal version

Simulate the redraws, collect the statistic each time, and the **standard error is the
`sd()` of that collection**:

$$
\mathrm{SE} \;=\; \mathrm{sd}\big(\bar{x}_1,\; \bar{x}_2,\; \dots,\; \bar{x}_R\big)
$$

Reading it aloud: "the standard error equals the standard deviation of x-bar-sub-one,
x-bar-sub-two, up through x-bar-sub-R" — or in everyday words: **"redraw the sample $R$
times, write down the statistic each time, and take the standard deviation of that whole
list."** Here $R$ is the number of redraws (make it big — a few thousand — so the
histogram is smooth).

Let's run it. The population is now a measurement machine whose output is bell-shaped
with true mean $\mu = 100$ and true spread $\sigma = 15$ (`rnorm(n, 100, 15)` draws $n$
values from it). We take samples of $n = 25$, and our statistic is the sample mean:

```{r se-simulation}
set.seed(42)
pop_mean <- 100   # mu:    the true (hidden) population mean
pop_sd   <- 15    # sigma: the true population spread
n        <- 25    # sample size per draw

# Redraw the whole sample 2000 times; each time record the sample MEAN (x-bar).
# replicate() is the "do this whole thing R times" tool from the warm-up.
sample_means <- replicate(2000, mean(rnorm(n, mean = pop_mean, sd = pop_sd)))

mean(sample_means)   # ~100: the sampling distribution is CENTERED on the truth
sd(sample_means)     # ~3.0: THIS is the standard error — the wobble of x-bar
```

Mapping it onto the numbers:

- Each `rnorm(25, 100, 15)` is one press of the lottery machine — one sample of 25.
- `mean(...)` inside is the statistic $\bar{x}$ for that press.
- `replicate(2000, ...)` presses the button 2000 times → the list
  $\bar{x}_1, \dots, \bar{x}_{2000}$.
- `sd(sample_means)` is the formula above → $\mathrm{SE} \approx 3.0$.

**The standard error is NOT the spread of the data.** Look at one single sample and its
own `sd()`:

```{r se-vs-sd}
set.seed(42)
one_sample <- rnorm(25, mean = 100, sd = 15)
mean(one_sample)   # ~102.8 — one sample's mean
sd(one_sample)     # ~19.6 — the spread of the DATA (near the true sigma = 15)
```

The data's spread is about **19.6** (it estimates $\sigma = 15$, the population spread).
The standard error is about **3.0**. These are wildly different numbers describing two
different things: 19.6 is "how spread out the 25 measurements are"; 3.0 is "how much the
*average of 25* would jump if I redrew." Same data, two spreads — never swap them.

::: {.callout-tip collapse="true" title="A shortcut for the mean's SE (optional but handy)"}
For the sample *mean* specifically, you don't always have to simulate: the standard
error is well approximated by $\mathrm{SE} \approx s / \sqrt{n}$ — the data's standard
deviation divided by the square root of the sample size. Here $15 / \sqrt{25} = 3$,
matching the simulation. This is why "more data → steadier number": $\sqrt{n}$ grows, so
$\mathrm{SE}$ shrinks. It's a shortcut *for the mean only*; the simulate-and-take-`sd()`
recipe above works for **any** statistic (median, proportion, slope) where no tidy
formula exists — which is exactly why we lead with it.
:::

## Explore it

```{webr}
#| autorun: true
#| fig-alt: "Histogram of 2000 sample means clustered in a tight bell shape around 100, only a few units wide"
# explore-1 — LIVE: watch the sampling distribution appear, and see SE is not sd(data).
# TRY IT: change n (try 4, then 100) and re-run. PREDICT: as n grows, does the histogram
# get WIDER or NARROWER? Commit to an answer, then run.
set.seed(42)
n <- 25
sample_means <- replicate(2000, mean(rnorm(n, mean = 100, sd = 15)))

hist(sample_means, breaks = 30, col = "steelblue",
     main = sprintf("SE (sd of these means) = %.2f", sd(sample_means)),
     xlab = "sample mean (x-bar) over 2000 redraws")
abline(v = 100, col = "red", lwd = 2)   # the true population mean the means cluster on
```

```{webr}
#| autorun: true
# explore-2 — LIVE: the standard error is the sd of the STATISTIC, not of the data.
# PREDICT: which will be bigger — the sd of ONE sample, or the sd of the 2000 means?
# Write down your guess, then run.
set.seed(42)
one_sample   <- rnorm(25, mean = 100, sd = 15)
sample_means <- replicate(2000, mean(rnorm(25, mean = 100, sd = 15)))

c(sd_of_one_sample = sd(one_sample),      # ~19.6 — spread of the DATA
  standard_error   = sd(sample_means))    # ~3.0  — spread of the STATISTIC
```

```{webr}
#| autorun: true
#| fig-alt: "Two histograms: lumpy raw population values, and bell-shaped sample means from that same lumpy population"
# explore-3 — LIVE: the bell shows up for MEANS even when the population is NOT a bell.
# The population here is lumpy: mostly 1s with an occasional 10. Yet the sample means
# pile into a bell. TRY IT: change sample size from 30 to 2, re-run, watch the bell break.
set.seed(42)
pop <- c(1, 1, 1, 1, 1, 10)     # a lumpy, non-bell population
means_30 <- replicate(3000, mean(sample(pop, 30, replace = TRUE)))

op <- par(mfrow = c(1, 2))
barplot(table(pop), col = "grey", main = "the population (lumpy)")
hist(means_30, breaks = 30, col = "seagreen",
     main = "means of 30 (a bell!)", xlab = "sample mean")
par(op)   # restore the layout so later cells don't inherit mfrow = c(1, 2)
```

The bell appearing in explore-3 is the famous **Central Limit Theorem**, and we are
treating it exactly as what it looks like: an **observation**. Redraw a *mean* enough
times, from almost any population, and the histogram of those means is bell-shaped and
centered on the true mean — even when the population itself is lumpy, skewed, or spiky.
We are not proving it here; we are watching it happen. (The bigger the sample size per
draw, the better the bell — try dropping it to 2 and the bell falls apart.)

## What a confidence interval is trying to do

Here is the payoff the standard error unlocks — **conceptually only**; the recipe for
computing and *reading* interval endpoints belongs to the hypothesis-testing module.

You press the button once and get, say, $\bar{x} = 102.8$. You know from the sampling
distribution that your number wobbles by about one $\mathrm{SE} \approx 3$ around the
hidden truth. So instead of reporting the bare point $102.8$ — which is almost certainly
a little off — you report a **range around it wide enough that it usually captures the
true value**. That range is a **confidence interval**: an honest "the truth is somewhere
around here, give or take the wobble," built by stretching out from your statistic by a
few standard errors.

The subtle part — and the reason it gets its own module — is what "usually captures"
means. The *truth* is fixed; the *interval* is what's random (it moves every time you
redraw). A "95%" interval is one built by a recipe such that, across many redraws, about
95% of the intervals it produces would cover the true value. We can literally watch that
happen: build the interval $\bar{x} \pm 2\,\mathrm{SE}$ on each of many redraws and count
how often it covers the true $\mu = 100$:

```{r ci-coverage}
set.seed(42)
covers <- replicate(1000, {
  s  <- rnorm(25, mean = 100, sd = 15)   # one sample
  se <- sd(s) / sqrt(25)                 # this sample's estimated SE
  lo <- mean(s) - 2 * se                 # stretch 2 SEs below the statistic
  hi <- mean(s) + 2 * se                 # ...and 2 SEs above
  lo <= 100 && 100 <= hi                 # did THIS interval catch the truth?
})
mean(covers)   # ~0.94: about 94% of these intervals covered mu = 100
```

About **94%** of the intervals caught the truth — close to the 95% the "$\pm 2$ SE"
recipe aims for. That is *all* a confidence interval is trying to do: catch the hidden
value most of the time. **How to pick the multiplier, read the endpoints, and phrase the
claim carefully is the hypothesis-testing module's job** — here you only need the idea
that the interval is a wobble-sized net cast around your one statistic.

## Common traps

**Trap 1 — "the standard error is the standard deviation of my data."** Tempting, because
both are spreads computed with `sd()`, and both are around when you have a sample. But
they measure different things. In the simulation, one sample's `sd()` was about **19.6**
(the spread of the 25 raw measurements, estimating $\sigma = 15$) while the standard
error was about **3.0** (the spread of the *mean* across redraws). Swapping them
overstates your uncertainty by more than sixfold here. Tell them apart by asking "the
spread of *what*?": `sd(data)` is the spread of the raw numbers; the standard error is
the spread of the *statistic*.

**Trap 2 — "more data makes the standard deviation shrink."** People see the standard
error fall as $n$ grows and conclude the data itself gets tighter. It doesn't — the
population spread $\sigma$ is fixed, so `sd(one_sample)` hovers near it no matter how big
the sample. What shrinks with $n$ is the **standard error**, the wobble of the
*average*. Watch both at once:

```{r trap2-numbers}
set.seed(42)
sd_small  <- sd(rnorm(10,   100, 15))                       # data spread, n = 10
sd_big    <- sd(rnorm(1000, 100, 15))                       # data spread, n = 1000
se_small  <- sd(replicate(2000, mean(rnorm(10,   100, 15))))# SE, n = 10
se_big    <- sd(replicate(2000, mean(rnorm(1000, 100, 15))))# SE, n = 1000
c(sd_small = sd_small, sd_big = sd_big)   # both near 15 — the DATA spread barely moves
c(se_small = se_small, se_big = se_big)   # ~4.7 -> ~0.47 — the SE collapses with n
```

The data's spread stays near 15 both times; the standard error drops roughly tenfold
(from ~4.7 to ~0.47) as $n$ goes from 10 to 1000. Only the *statistic's* spread cares
about sample size.

## Check yourself

**1.** You draw one sample and get `sd(x) = 8` from `n = 16` observations. Using the
mean's shortcut $\mathrm{SE} \approx s / \sqrt{n}$, what is the approximate standard
error of the mean — and is that bigger or smaller than 8?

::: {.callout-note collapse="true" title="Answer"}
$\mathrm{SE} \approx 8 / \sqrt{16} = 8 / 4 = 2$. It is **smaller** than 8. That is the
whole point: the average of 16 numbers wobbles far less (SE ≈ 2) than the individual
numbers spread (sd = 8). Dividing by $\sqrt{n}$ is what makes the mean steadier than any
single observation.
:::

**2.** A statistic other than the mean — say the **median** — is computed from a random
sample. Does it *also* have a sampling distribution and a standard error? Sketch how
you'd find its SE in R.

```{webr}
#| autorun: false
# check2-starter — edit and Run: histogram the median over many redraws, take its sd
set.seed(42)
# your job: replicate( ... median(rnorm(25, 100, 15)) ... ), then sd() of the result
```

::: {.callout-note collapse="true" title="Answer"}
Yes — **any** statistic computed from a random sample is random, so it has a sampling
distribution and a standard error. The simulate-and-take-`sd()` recipe doesn't care
which statistic you use: `sample_medians <- replicate(2000, median(rnorm(25, 100, 15)))`,
then `sd(sample_medians)` (about **3.7** here). The mean's tidy $s/\sqrt{n}$ shortcut is
special to the mean; the simulation works for everything.
:::

**3.** In your own words: explain to a colleague, in two sentences, what the standard
error is and why it is *not* the same as the standard deviation of the data.

::: {.callout-note collapse="true" title="A model answer"}
The standard error is how much a number you compute from data (like an average) would
jump around if you kept redrawing fresh samples — it's the spread of the *statistic*, not
of the raw data. The data's standard deviation says how scattered the individual
measurements are and barely changes with sample size, while the standard error shrinks as
the sample grows, because averaging more numbers steadies the result.
:::

## What you can now do

- **Name the three players** — population, one sample, and a statistic computed from that
  sample — and explain why the statistic is random (redraw the sample, get a different
  number).
- **Simulate re-runs in R** with `replicate()` + `sample()`/`rnorm()`, histogram the
  statistic, and read that histogram as the **sampling distribution**.
- **Compute a standard error** as the `sd()` of the redrawn statistics, and explain — with
  numbers — why it is *not* the standard deviation of the data and why it shrinks as $n$
  grows.
- **Recognize the bell** that emerges for sample means (the Central Limit Theorem, as an
  observation) even from a non-bell population.
- **State what a confidence interval is trying to do** — cast a wobble-sized net around
  your statistic that usually captures the hidden true value — without yet reading its
  endpoints.

If any of these feels shaky, revisit [Getting a feel for it](#getting-a-feel-for-it) and
[The formal version](#the-formal-version) before continuing — anything that reports
uncertainty on a number assumes all of these.

## Where this goes next

The standard error is the raw material for every statement of the form "this estimate,
give or take." A hypothesis-testing module turns it into p-values and the recipe for
*reading* confidence-interval endpoints; a resampling module reuses the same
simulate-and-take-`sd()` engine — the **bootstrap** — to get a standard error when you
can't even name the population (its sibling, the permutation test, reuses resampling to
build *null* distributions instead). Ready for more reps? Do the
[practice problems](practice.qmd).
