---
title: "Practice: Randomness, samples & sampling distributions"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only (a {webr} starter does not see it — each starter re-states its
# own data). This module uses randomness, so set.seed(42) fixes every baked answer.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

Work these in order — they get harder. Do them in an R session (this module lives on
simulation); only then open the answer. Use `set.seed(42)` so your numbers match ours.

## Problem 1 — the toy example, new numbers

The population is now a measurement machine with true mean $\mu = 50$ and true spread
$\sigma = 10$ (`rnorm(n, 50, 10)`). Take samples of $n = 16$, and let the statistic be
the sample mean. Simulate 2000 redraws, histogram the means, and report the **standard
error** (the `sd()` of those means). **PREDICT** first: will the SE be near 10, or much
smaller? Write it down, then run.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
set.seed(42)
sample_means <- replicate(2000, mean(rnorm(16, mean = 50, sd = 10)))
# your job: sd(sample_means), and hist(sample_means)
```

::: {.callout-note collapse="true" title="Answer"}
The standard error is the standard deviation of the redrawn means. It is **much smaller
than 10** — the average of 16 numbers wobbles far less than the numbers themselves. The
mean's shortcut predicts $\mathrm{SE} \approx \sigma/\sqrt{n} = 10/\sqrt{16} = 2.5$, and
the simulation lands right there.

```{r p1-answer}
set.seed(42)
sample_means <- replicate(2000, mean(rnorm(16, mean = 50, sd = 10)))
sd(sample_means)          # ~2.53 — the standard error, close to the 2.5 shortcut
mean(sample_means)        # ~50 — the sampling distribution centers on the truth
```
:::

## Problem 2 — fill in the missing step

Here is the simulate-and-take-`sd()` recipe for a new setup, with the **core step
blanked**. Population: $\mu = 8$, $\sigma = 2$, sample size $n = 4$; statistic = the
mean. Fill in ONLY the blanked line.

> Step 1 — one press of the machine is `mean(rnorm(4, 8, 2))`: draw 4 values, average
> them.
> Step 2 — **`<your step here>`**: press the button 2000 times, collecting the mean each
> time into a vector called `sample_means`.
> Step 3 — the standard error is `sd(sample_means)`, and the center is
> `mean(sample_means)`.

```{webr}
#| autorun: false
# p2-starter — edit and Run to check your blanked step
set.seed(42)
# sample_means <- <your step here>
# sd(sample_means); mean(sample_means)
```

::: {.callout-note collapse="true" title="Answer"}
The blanked step is the `replicate()` call — the "do this whole thing over and over"
move that turns one press into a sampling distribution:

```{r p2-answer}
set.seed(42)
sample_means <- replicate(2000, mean(rnorm(4, mean = 8, sd = 2)))  # the missing step
sd(sample_means)          # ~1.03 — the standard error (shortcut: 2/sqrt(4) = 1)
mean(sample_means)        # ~7.97 — centered on the true mu = 8
```

Without `replicate()` you'd have a single mean, not a distribution — and no way to see
the wobble. `replicate(2000, ...)` is what manufactures the redraws.
:::

## Problem 3 — a proportion is a statistic too

A bag of chips is **30% red, 70% blue**. You draw $n = 40$ chips (with replacement) and
compute the **proportion red** — code red as 1, blue as 0, and the proportion is just the
`mean()` of those 0/1s. Simulate 2000 redraws and find the sampling distribution's center
and standard error. **PREDICT:** roughly what value will the proportions center on?

```{webr}
#| autorun: false
# p3-starter — your job: build `props` with replicate(), then its center and SE
set.seed(42)
# one draw of 40 chips (red = 1, blue = 0), then the proportion red:
#   mean(sample(c(1, 0), 40, replace = TRUE, prob = c(0.3, 0.7)))
# wrap that in replicate(2000, ...) to get `props`, then mean(props) and sd(props)
```

::: {.callout-note collapse="true" title="Answer"}
A proportion is the mean of 0/1 data, so it is a statistic like any other — random, with
its own sampling distribution and standard error. The center sits at the true proportion
0.3; the standard error is the `sd()` of the redrawn proportions.

```{r p3-answer}
set.seed(42)
props <- replicate(2000, mean(sample(c(1, 0), 40, replace = TRUE, prob = c(0.3, 0.7))))
mean(props)   # ~0.303 — centered on the true 30%
sd(props)     # ~0.071 — the standard error of the proportion
```

The exact same simulate-and-`sd()` recipe you used for a mean works unchanged for a
proportion — that's the power of the simulation approach.
:::

## Problem 4 — spot the error

A colleague reports the uncertainty of a sample mean. Their reasoning contains **exactly
one mistake** — find it, explain why it's wrong, and give the fix.

> "I drew a sample of $n = 9$ values and computed `sd(x)`, which came out to about 2.57.
> That `sd` **is** the standard error of my mean, so I'll report the mean give-or-take
> 2.57."

```{webr}
#| autorun: false
# p4-starter — copy this to see the two DIFFERENT spreads
set.seed(42)
x <- rnorm(9, mean = 0, sd = 3)
# sd(x)            # the spread of the DATA
# sd(x) / sqrt(9)  # the standard error of the MEAN
```

::: {.callout-note collapse="true" title="Answer"}
The mistake is calling `sd(x)` the standard error. `sd(x) ≈ 2.57` is the spread of the
**data** — how scattered the 9 raw values are. The standard error is the spread of the
**mean**, which for the mean is about $s/\sqrt{n}$, far smaller:

```{r p4-answer}
set.seed(42)
x <- rnorm(9, mean = 0, sd = 3)
sd(x)             # ~2.57 — spread of the DATA (what the colleague wrongly reported)
sd(x) / sqrt(9)   # ~0.86 — the correct standard error of the mean
```

Reporting 2.57 overstates the mean's wobble by roughly threefold. The fix: divide by
$\sqrt{n}$ (here $\sqrt{9} = 3$) to get $\mathrm{SE} \approx 0.86$, or simulate redraws
and take the `sd()` of the means. The tell is always "the spread of *what*?" — the data,
or the statistic.
:::

## Problem 5 — mixing it with the mean & standard deviation *(uses [mean-variance-covariance](../mean-variance-covariance/lesson.qmd)'s `sd`)*

You have **one** sample of $n = 100$ measurements from a machine with $\sigma = 4$. You
cannot redraw — you only have this one dataset. Using the mean's shortcut, turn the
data's standard deviation (the `sd()` from the mean-variance-covariance module) into the
standard error of the mean. **PREDICT:** which is bigger, `sd(x)` or the SE?

```{webr}
#| autorun: false
# p5-starter — edit and Run: get sd(x), then convert it to the SE of the mean
set.seed(42)
x <- rnorm(100, mean = 20, sd = 4)
```

::: {.callout-note collapse="true" title="Answer"}
`sd(x)` is the data spread you learned to compute in
[mean-variance-covariance](../mean-variance-covariance/lesson.qmd); the mean's standard
error is that divided by $\sqrt{n}$. The SE is much smaller.

```{r p5-answer}
set.seed(42)
x <- rnorm(100, mean = 20, sd = 4)
sd(x)              # ~4.17 — the spread of the DATA (estimates sigma = 4)
sd(x) / sqrt(100)  # ~0.42 — the standard error of the MEAN
```

The data spread (~4.17) is about tenfold larger than the SE (~0.42), because $\sqrt{100}
= 10$. This is the bridge: the standard deviation you already know, divided by $\sqrt{n}$,
becomes the standard error of the mean — no re-simulation needed when the statistic is a
mean.
:::

## Problem 6 — transfer: does the bell always show up?

The Central Limit Theorem says means go bell-shaped — but *how big a sample* it takes
depends on how weird the population is. Here is a nasty population: **four 0s and one
100** (extremely lopsided). Simulate the sampling distribution of the mean for a *tiny*
sample ($n = 2$) and a *larger* one ($n = 50$), and compare their shapes and standard
errors. **PREDICT:** which one looks more like a bell?

```{webr}
#| autorun: false
# p6-starter — your job: histogram the means at n = 2 and n = 50, and compare their sd()
set.seed(42)
pop <- c(0, 0, 0, 0, 100)      # a very lopsided population; true mean = 20
```

::: {.callout-note collapse="true" title="Answer"}
The bell shows up for the *larger* sample. At $n = 2$ the means are lumpy and jagged (few
possible averages, huge spread); at $n = 50$ they smooth into a bell centered on the true
mean of 20, with a much smaller standard error.

```{r p6-answer}
set.seed(42)
pop <- c(0, 0, 0, 0, 100)          # true mean = 20
m2  <- replicate(3000, mean(sample(pop,  2, replace = TRUE)))
m50 <- replicate(3000, mean(sample(pop, 50, replace = TRUE)))
c(SE_n2 = sd(m2), SE_n50 = sd(m50)) # ~28.7 vs ~5.5 — SE shrinks with n
c(center_n2 = mean(m2), center_n50 = mean(m50))  # both ~20 — centered on the truth
```

Two lessons in one: the Central Limit Theorem is an *observation with fine print* — the
bell is a large-sample tendency, and a lopsided population needs a bigger $n$ before it
kicks in; and no matter the shape, the standard error keeps shrinking as $n$ grows
(~28.7 at $n=2$ down to ~5.5 at $n=50$). A later module turns this simulate-the-redraws
engine into the bootstrap. For now: means steady and bell-ify as samples grow.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/randomness-samples-and-sampling-distributions" pass="4"}

::: {.quiz-q qid="rss-se-vs-sd-data" qtype="mc" answer="b"}
The standard error and the data's standard deviation both use `sd()`. How do they
differ?

a) They are the same thing
b) `sd(data)` is the spread of the raw numbers; the standard error is the spread of the *statistic* across redraws
c) The standard error is always the larger of the two
:::

::: {.quiz-q qid="rss-n-shrinks-se" qtype="mc" answer="c"}
As the sample size $n$ grows, what happens?

a) Both the data's sd and the standard error shrink
b) The data's sd shrinks while the standard error stays fixed
c) The data's sd stays near $\sigma$ while the standard error shrinks
:::

::: {.quiz-q qid="rss-statistic-random" qtype="mc" answer="a"}
Why is a statistic such as the sample mean random?

a) Because redrawing the sample yields a different value each time
b) Because the population itself changes over time
c) Because `sd()` introduces noise
:::

::: {.quiz-q qid="rss-clt" qtype="mc" answer="b"}
Even when the population is not bell-shaped, the histogram of the sample *mean* across
many redraws tends toward what shape?

a) The same shape as the population
b) A bell (normal) curve — the Central Limit Theorem
c) A flat, uniform spread
:::

::: {.quiz-q qid="rss-se-from-variance" qtype="numeric" answer="4" tolerance="0"}
(From mean, variance & covariance.) The standard error is the standard deviation of the
redrawn statistics. If those redrawn means have variance $16$, what is the standard
error?
:::

:::

::::
