---
title: "Practice: The Approximate Bayes Factor"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

Work these in order — they get harder. Do them on paper or in an R session; only then
open the answer. A helper to paste as needed:
`wakefield_abf <- function(beta, se, W) { z <- beta/se; V <- se^2; r <- W/(W+V); 0.5*(log(1-r) + r*z^2) }`.

## Problem 1 — the toy example, new numbers

A variant has $\hat\beta = 0.24$ and $\text{se} = 0.06$, with prior variance $W = 0.04$.
Compute $Z$, $r$, and the lABF from the closed form. Confirm with `wakefield_abf`.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
wakefield_abf <- function(beta, se, W) { z <- beta/se; V <- se^2; r <- W/(W+V); 0.5*(log(1-r) + r*z^2) }
beta_hat <- 0.24; se <- 0.06; W <- 0.04
```

::: {.callout-note collapse="true" title="Answer"}
$Z = 0.24/0.06 = 4$; $V = 0.06^2 = 0.0036$; $r = 0.04/(0.04 + 0.0036) = 0.9174$. lABF
$= \tfrac12(\log(1 - 0.9174) + 0.9174 \times 16) = \tfrac12(-2.494 + 14.679) = 6.092$.

```{r p1-answer}
wakefield_abf <- function(beta, se, W) { z <- beta/se; V <- se^2; r <- W/(W+V); 0.5*(log(1-r) + r*z^2) }
c(Z = 0.24/0.06, lABF = wakefield_abf(0.24, 0.06, 0.04))    # 4, 6.0924
```
(Slightly below the lesson's $6.11$: same $Z = 4$, but a bigger $V$ shrinks $r$, so the
signal is discounted a touch more.)
:::

## Problem 2 — fill in the missing step

A variant has $Z = 2$, $V = 0.0025$, and prior $W = 0.04$. Here is the worked solution with
the core step blanked — supply only that step.

- Step 1 (shown): the shrinkage factor is $r = W/(W + V) = 0.04/0.0425 = 0.9412$.
- Step 2 (**your step here**): assemble the lABF from $r$ and $Z$. Write the exact
  arithmetic.
- Step 3 (shown): the lABF is $\approx 0.466$ (so BF $\approx 1.59$ — only mild evidence).

```{webr}
#| autorun: false
# p2-starter — edit and Run to check the blank
Z <- 2; r <- 0.9412
```

::: {.callout-note collapse="true" title="Answer"}
The missing move is plugging $r$ and $Z$ into the closed form: lABF $= \tfrac12(\log(1 - r)
+ rZ^2) = \tfrac12(\log(0.0588) + 0.9412 \times 4) = \tfrac12(-2.833 + 3.765) = \tfrac12(0.932)
= 0.466$.

```{r p2-answer}
Z <- 2; r <- 0.04 / 0.0425
0.5 * (log(1 - r) + r * Z^2)      # 0.466 -- half of (width penalty + signal reward)
```
Note how close the signal ($3.765$) is to the penalty ($2.833$) here: at $Z = 2$ the ABF is
barely positive, unlike the crushing $Z = 4$ case.
:::

## Problem 3 — a full computation, both ways

A variant has $\hat\beta = 0.30$, $\text{se} = 0.10$, prior $W = 0.04$. Compute the lABF two
ways and confirm they match: (a) as the log of the ratio of the two `dnorm` curve heights
(Module 2's curves), and (b) from the closed form. Report the Bayes factor $e^{\text{lABF}}$
too.

```{webr}
#| autorun: false
# p3-starter — your job: both routes to the lABF, and the BF
beta_hat <- 0.30; se <- 0.10; W <- 0.04
```

::: {.callout-note collapse="true" title="Answer"}
$V = 0.01$, $Z = 3$, $r = 0.04/0.05 = 0.8$. Closed form: lABF $= \tfrac12(\log(0.2) + 0.8
\times 9) = \tfrac12(-1.609 + 7.2) = 2.795$; density-ratio route gives the same $2.795$. BF
$= e^{2.795} \approx 16.4$.

```{r p3-answer}
beta_hat <- 0.30; se <- 0.10; W <- 0.04
V <- se^2; Z <- beta_hat / se; r <- W / (W + V)

ratio  <- log(dnorm(beta_hat, 0, sqrt(V + W)) / dnorm(beta_hat, 0, sqrt(V)))   # curve route
closed <- 0.5 * (log(1 - r) + r * Z^2)                                          # formula
c(density_ratio = ratio, closed_form = closed, BF = exp(closed))    # 2.795, 2.795, 16.4
```
:::

## Problem 4 — spot the error

A colleague computes the lABF for a variant with $Z = 4$, $V = 0.0025$, $W = 0.04$ — the
computation contains **exactly one mistake** — find it and fix it.

> "The shrinkage factor is $r = W/(W + V) = 0.04/0.0425 = 0.9412$. So lABF $= \tfrac12(\log(1
> + r) + rZ^2) = \tfrac12(\log(1.9412) + 0.9412 \times 16) = \tfrac12(0.663 + 15.06) =
> 7.86$."

::: {.callout-note collapse="true" title="Answer"}
The mistake is $\log(1 + r)$ instead of $\log(1 - r)$. The width penalty is $\log(1 - r) =
\log(0.0588) = -2.833$ (a *penalty*, so negative), not $\log(1 + r) = +0.663$. With the sign
right, lABF $= \tfrac12(-2.833 + 15.06) = 6.11$, not $7.86$. Using $1 + r$ turns the
alternative's width *penalty* into a spurious *reward* — it would make a wider curve look
like more evidence, which is backwards.

```{r p4-answer}
r <- 0.04 / 0.0425; Z <- 4
c(wrong = 0.5 * (log(1 + r) + r * Z^2),          # 7.86 -- sign error
  correct = 0.5 * (log(1 - r) + r * Z^2))         # 6.11
```
:::

## Problem 5 — mixing it with Module 2 *(uses Module 2's two curves)*

A variant has $\hat\beta = 0.10$, $\text{se} = 0.05$, prior $W = 0.04$. (a) Write down the
two Module-2 curves $N(0, V)$ and $N(0, V+W)$ and their heights at $\hat\beta = 0.10$. (b)
Take their log-ratio to get the lABF, and confirm it equals the closed form. (c) Is this
strong evidence?

```{webr}
#| autorun: false
# p5-starter — edit and Run to solve, right here
beta_hat <- 0.10; se <- 0.05; W <- 0.04
```

::: {.callout-note collapse="true" title="Answer"}
$V = 0.0025$, $Z = 2$. (a) Null $N(0, 0.0025)$ height $\approx 1.080$; alternative $N(0,
0.0425)$ height $\approx 1.720$ (from Module 2's practice). (b) lABF $= \log(1.720/1.080) =
\log(1.593) = 0.466$, matching the closed form $\tfrac12(\log(0.0588) + 0.9412 \times 4) =
0.466$. (c) Not strong: BF $= e^{0.466} \approx 1.59$ — the alternative fits only about
$1.6\times$ better, mild evidence at best.

```{r p5-answer}
beta_hat <- 0.10; se <- 0.05; W <- 0.04
V <- se^2; Z <- beta_hat / se; r <- W / (W + V)
curve_lABF <- log(dnorm(beta_hat, 0, sqrt(V + W)) / dnorm(beta_hat, 0, sqrt(V)))
c(curve = curve_lABF, closed = 0.5 * (log(1 - r) + r * Z^2), BF = exp(curve_lABF))
```
:::

## Problem 6 — transfer (interpretation)

Keep $Z = 4$ and $V = 0.0025$ fixed, but sweep the prior variance $W$ over
$\{0.001, 0.04, 1, 100\}$. Compute the lABF at each. What happens as $W$ grows very large,
and why does that make sense given the two terms in the formula?

```{webr}
#| autorun: false
# p6-starter — edit and Run to solve, right here
Z <- 4; V <- 0.0025
W_grid <- c(0.001, 0.04, 1, 100)
```

::: {.callout-note collapse="true" title="Answer"}
The lABF rises then *falls*: about $2.12$ at $W = 0.001$, $6.11$ at $W = 0.04$, $4.98$ at
$W = 1$, and $2.70$ at $W = 100$. Very large $W$ **hurts** the evidence.

```{r p6-answer}
Z <- 4; V <- 0.0025
labf <- function(W) { r <- W / (W + V); 0.5 * (log(1 - r) + r * Z^2) }
round(sapply(c(0.001, 0.04, 1, 100), labf), 3)     # 2.117, 6.113, 4.983, 2.701
```
Why: $W$ pushes the signal reward $rZ^2$ up (good) but also deepens the width penalty
$\log(1 - r)$ (bad, since $r \to 1$ makes $1 - r \to 0$ and $\log(1 - r) \to -\infty$). For
small $W$ the reward is throttled ($r$ near 0); for huge $W$ the penalty dominates. The best
evidence sits at a moderate $W$ matched to plausible effect sizes — which is exactly why
choosing `sd.prior` sensibly matters.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="courses/approximate-bayes-factors/modules/03-the-approximate-bayes-factor" pass="4"}

::: {.quiz-q qid="abf-shrinkage-numeric" qtype="numeric" answer="0.9412" tolerance="0.001"}
For a variant with $V = 0.0025$ and prior $W = 0.04$, compute the shrinkage factor $r =
W/(W + V)$. Round to four decimals.
:::

::: {.quiz-q qid="abf-labf-numeric" qtype="numeric" answer="6.11" tolerance="0.02"}
For a variant with $Z = 4$, $V = 0.0025$, and $W = 0.04$ (so $r = 0.9412$), compute the log
ABF $= \tfrac12(\log(1 - r) + rZ^2)$. Round to two decimals.
:::

::: {.quiz-q qid="abf-two-terms" qtype="mc" answer="b"}
In $\text{lABF} = \tfrac12(\log(1 - r) + rZ^2)$, what are the signs and roles of the two
terms?

a) Both terms are positive: a width reward and a signal reward
b) $\log(1 - r)$ is a negative width penalty; $rZ^2$ is a positive signal reward
c) $\log(1 - r)$ is a positive signal reward; $rZ^2$ is a negative width penalty
:::

::: {.quiz-q qid="abf-what-is-r" qtype="mc" answer="c"}
What is $r$ in the ABF formula?

a) The correlation between the estimate and the true effect, between $-1$ and $1$
b) The p-value re-expressed on the $[0, 1]$ scale
c) The shrinkage factor $W/(W + V)$, a variance ratio between $0$ and $1$
:::

::: {.quiz-q qid="abf-curve-ratio" qtype="mc" answer="a"}
*(Interleaved — uses Module 2's two curves.)* The Bayes factor for association equals which
ratio, evaluated at the observed $\hat\beta$?

a) The alternative curve $N(0, V+W)$ height over the null curve $N(0, V)$ height
b) The null curve height over the alternative curve height
c) The prior $N(0, W)$ height over the estimate's noise $N(0, V)$ height
:::

:::

::::
