---
title: "The derivative, as a slope you can measure"
subtitle: "How steep is a curve right here? Measure it with arithmetic — no calculus rules needed"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live cells below re-state their own data. Base R only.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```


## Where we're going

You already have the *idea* of a derivative: it's the **slope** of a curve — how fast
the output changes as you nudge the input. This module turns that idea into something
you can actually **compute**, using nothing but subtraction and division. No
differentiation rules, no limits-in-symbols: we'll measure the slope the way you'd
measure the steepness of a hill — take a tiny step and look at the rise over the run.

That one skill — "how steep is this curve right here?" — is the compass behind a huge
amount of applied math: finding where a curve bottoms out (optimization), rates of
change in physics and finance, and the sensitivity of any model to its inputs.

**Builds on:** just the baseline (you can read a line's slope as rise over run, and
plug a number into a function).

## Warm-up: do you remember?

This is a foundation module, so there's nothing earlier to recall — just three
baseline reflexes the idea leans on. Produce each answer before opening it.

**1.** A straight line passes through the points $(1, 2)$ and $(3, 8)$. What is its
slope?

::: {.callout-note collapse="true" title="Answer"}
Slope is *rise over run*: $(8 - 2) / (3 - 1) = 6 / 2 = 3$. The line climbs 3 units of
output for every 1 unit of input. Everything in this lesson is this one calculation,
done over a tiny run.
:::

**2.** For the function $f(x) = x^2$, what are $f(3)$ and $f(3.1)$?

::: {.callout-note collapse="true" title="Answer"}
$f(3) = 3^2 = 9$ and $f(3.1) = 3.1^2 = 9.61$. "Plug the number in and square it" — that's
all $f(x) = x^2$ asks.
:::

**3.** Look at a U-shaped curve (like $x^2$). Where along it is the curve *flat* —
slope zero — and where is it steepest: near the bottom, or far out on the sides?

::: {.callout-note collapse="true" title="Answer"}
It's flat at the very bottom (slope 0 there) and gets steeper the farther out you go.
Slope changes from point to point — which is exactly why we need a way to measure it
*at a chosen point*.
:::

## A tiny example first

Take the curve $f(x) = x^2$, and ask: **how steep is it right at $x = 3$?**

Here's the snag. Slope is rise over run — but that needs *two* points, and we only
have one. So we borrow a second point a tiny step away and measure the slope of the
little line between them. Watch what happens as the step shrinks:

- **Step of 0.5:** go from $x = 3$ to $x = 3.5$. Rise $= f(3.5) - f(3) = 12.25 - 9 =
  3.25$. Run $= 0.5$. Slope $= 3.25 / 0.5 = 6.5$.
- **Step of 0.1:** from $3$ to $3.1$. Rise $= 9.61 - 9 = 0.61$. Run $= 0.1$. Slope $=
  6.1$.
- **Step of 0.01:** from $3$ to $3.01$. Rise $= 9.0601 - 9 = 0.0601$. Run $= 0.01$.
  Slope $= 6.01$.

The numbers are marching toward **6**. That settled-on value — the slope the little
lines approach as the step shrinks to nothing — is the derivative of $f$ at $x = 3$.

The same thing in R, so you can check every step:

```{r toy-example}
# The curve, as an R function
f <- function(x) x^2

x0 <- 3
for (h in c(0.5, 0.1, 0.01)) {
  rise  <- f(x0 + h) - f(x0)   # change in output over the little step
  run   <- h                   # the step itself
  slope <- rise / run          # rise over run
  cat("step", h, "-> slope", slope, "\n")
}
# 6.5, 6.1, 6.01 -> heading for 6
```

## Getting a feel for it

**A speedometer is a derivative.** Your car's speed is how fast distance changes with
time. You can't see "speed at this instant" directly — but measure how far you travel
in the next 1 second (that's an average speed), then the next 0.1 second, then the next
0.01 second, and the averages home in on the number the speedometer shows *right now*.
Instantaneous speed is the derivative of distance. Same move as our shrinking step.
(Where the picture bends: a real speedometer measures forward in time only; in a moment
we'll get a sharper estimate by looking a step *each way*.)

**Why rise-over-a-shrinking-run?** Over a wide step, "rise / run" is the *average*
slope across the whole interval — it blurs together the gentle part and the steep part.
Shrinking the run zooms in until the curve, over that tiny stretch, is essentially
straight — and the slope of that straight little piece is the steepness *right at the
point*. It's the mathematical version of zooming into a map until the curved coastline
looks like a straight edge.

**What the slope's value feels like.** A bigger number means steeper. A **positive**
slope means the curve is heading uphill as $x$ increases; **negative** means downhill;
**zero** means momentarily flat — the top of a hill or the bottom of a valley. At
$x = 3$ our slope is $+6$: steep, and climbing. At $x = 0$ (the bottom of the $x^2$
bowl) it would be $0$. That "slope is zero at the bottom" fact is the whole reason
slopes help us *find* bottoms later.

## What the notation is going to say

The formula coming up says exactly what we just did: "pick a point $x$; take a tiny
step of size $h$; the slope is the change in the function, $f(x+h) - f(x)$, divided by
the step $h$; and as $h$ shrinks toward $0$, that ratio settles onto *the* slope at
$x$, which we name $f'(x)$." In our toy example that's the "$(9.61 - 9)/0.1 = 6.1$,
and shrinking" we just watched.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $f$ | "eff" | a function — a rule that turns an input into an output (an R function!) | $f(x) = x^2$ |
| $f(x)$ | "eff of ex" | the output of $f$ at input $x$ | $f(3) = 9$ |
| $x$ | "ex" | the input, and the point we measure the slope at | $3$ |
| $h$ | "aitch" | the tiny step we take away from $x$ | $0.5$, then $0.1$, then $0.01$ |
| $f(x+h)$ | "eff of ex plus aitch" | the output one small step to the right | $f(3.1) = 9.61$ |
| $f(x-h)$ | "eff of ex minus aitch" | the output one small step to the left | $f(2.9) = 8.41$ |
| $2h$ | "two aitch" | the total run from $x-h$ to $x+h$ — twice the step | $0.2$ |
| $f'(x)$ | "eff prime of ex" | the slope of $f$ at the point $x$ — the derivative | $f'(3) = 6$ |
| $\approx$ | "approximately equals" | almost equal (our estimate, before $h$ reaches $0$) | $6.1 \approx 6$ |
| $\to$ | "goes to" or "approaches" | tends toward a value (as the step shrinks, the slope tends toward its true value) | $h \to 0$ |

## The formal version

The slope over one small step — called the **difference quotient** — is:

$$
\frac{f(x + h) - f(x)}{h}
$$

Reading it aloud: "eff of ex plus aitch, minus eff of ex, all over aitch" — or in
everyday words: "the rise (how much the output changed) divided by the run (the little
step you took)." For our toy at $x = 3$, $h = 0.1$: $(9.61 - 9)/0.1 = 6.1$ — exactly
the middle row of the table above.

The derivative $f'(x)$ is the value this settles on as the step $h$ shrinks toward
zero:

$$
f'(x) \;\approx\; \frac{f(x+h) - f(x)}{h} \qquad \text{(sharper and sharper as } h \to 0\text{)}
$$

Reading it aloud: "eff prime of ex is approximately the difference quotient, and it
gets exact as aitch goes to zero." For $x^2$ the numbers marched $6.5 \to 6.1 \to 6.01
\to 6$, so $f'(3) = 6$.

**A sharper way to measure it (the one we'll use in code).** Instead of stepping only
*forward*, step the same distance each way — back to $x - h$ and forward to $x + h$ —
and take the rise across both:

$$
f'(x) \;\approx\; \frac{f(x + h) - f(x - h)}{2h}
$$

Reading it aloud: "eff of ex-plus-aitch minus eff of ex-minus-aitch, all over two
aitch" — or in everyday words: "output a step to the right minus output a step to
the left, divided by the total run across both steps." Looking symmetrically each
way cancels most of the curve's bending, so this
**central difference** is far more accurate for the same $h$. Check it on our toy at
$x = 3$ with $h = 0.1$: $(f(3.1) - f(2.9)) / 0.2 = (9.61 - 8.41)/0.2 = 1.2/0.2 = 6$ —
bang on, no shrinking needed.

```{r sanity-check}
# Sanity-check the central difference against the value we approached by hand
f <- function(x) x^2
x0 <- 3; h <- 0.1
(f(x0 + h) - f(x0 - h)) / (2 * h)   # 6 exactly — matches f'(3) = 6
```

(For this particular curve the exact slope happens to be $2x$, which is $6$ at
$x = 3$. We won't *derive* that rule — the point of this module is that you can always
**measure** the slope numerically, whatever the function, without knowing a rule for
it.)

## Explore it

```{webr}
#| autorun: true
#| fig-alt: "The parabola f(x)=x squared with the point at x=3 marked and a straight secant line through x=3 and x=3+h; the secant's slope is shown in the title."
# explore-1 — LIVE: this draws the curve, the point at x = 3, and the little line
# through (3, f(3)) and (3+h, f(3+h)). Shrink h and watch that line become the tangent.
# PREDICT: as you make h smaller (try 0.5, then 0.1, then 0.01), what number does the
# slope in the title head toward? Write it down, then edit h and re-run.
f  <- function(x) x^2
x0 <- 3
h  <- 0.5                                  # <-- shrink me

slope <- (f(x0 + h) - f(x0)) / h           # rise over run, stepping forward by h

curve(f, from = 0, to = 5, lwd = 2, ylab = "f(x) = x^2",
      main = sprintf("step h = %.2f   ->   measured slope = %.3f", h, slope))
points(x0, f(x0), pch = 16, cex = 1.5)
# the little "rise over run" line, extended so you can see its steepness
abline(a = f(x0) - slope * x0, b = slope, col = "firebrick", lwd = 2, lty = 2)
```

```{webr}
#| autorun: true
# explore-2 — LIVE: a reusable tool. `slope_at(f, x)` measures the slope of ANY
# function f at any point x, using the symmetric central difference. This little
# function is the workhorse the rest of the optimization story is built on.
# TRY IT: change the function, or the point. PREDICT each slope before running.
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)

f <- function(x) x^2
c(at_x0 = slope_at(f, 0),    # bottom of the bowl -> should be 0
  at_x3 = slope_at(f, 3),    # -> should be 6
  at_xm2 = slope_at(f, -2))  # downhill on the left -> should be -4
```

```{webr}
#| autorun: true
# explore-3 — LIVE: explicit check — does our numerical tool agree with the
# known analytical slope of x^2 (which is 2x)?
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
f <- function(x) x^2

# Side by side: numerical measurement vs. the exact formula 2x
data.frame(
  x         = c(0, 3, -2),
  numerical = c(slope_at(f, 0), slope_at(f, 3), slope_at(f, -2)),
  analytic  = c(2 * 0,          2 * 3,           2 * (-2))
)
```

## Common traps

**Trap: "smaller $h$ is always better — use $h = 10^{-16}$."** It's tempting: the
formula is exact only as $h \to 0$, so tinier looks more accurate. But computers store
numbers with limited precision, and when $h$ is *too* small, $f(x+h)$ and $f(x-h)$ round
to nearly the same stored value — their difference is then mostly rounding noise, and
dividing by a tiny $h$ blows that noise up. There's a sweet spot (around $h = 10^{-5}$
for the central difference). See it fail:

```{r trap-tiny-h}
f <- function(x) x^2
central <- function(h) (f(3 + h) - f(3 - h)) / (2 * h)
c(good_h = central(1e-5),   # ~6, correct
  tiny_h = central(1e-14))  # noticeably off — floating-point cancellation
```

**Trap: "the slope at a point comes from that one point alone."** A single point has a
*height*, not a *slope* — steepness is inherently about how the height *changes*
between nearby points. Try it: $f(3) = 9$. That $9$ alone tells you *nothing* about
steepness — there is no rise and no run to divide. Only once you bring in a second
point ($f(3.1) = 9.61$) can you form rise/run: $(9.61 - 9)/0.1 = 6.1$. That's why
every method here nudges to a second point (or two); "the slope at $x$" is really
shorthand for "what the nearby rise-over-runs settle on."

**Trap: "the central difference divides by $h$."** When you step from $x - h$ to
$x + h$, the total run is $2h$, not $h$. Dividing by $h$ instead of $2h$ doubles the
answer: $(f(3.1) - f(2.9))/0.1 = 1.2/0.1 = 12$ — twice the true slope of $6$. The
fix: the denominator must match the full span, $2h$.

## Check yourself

**1.** Use the numerical method to estimate the slope of $f(x) = x^2$ at $x = 5$. What
value should it approach?

```{webr}
#| autorun: false
# check1-starter — edit and Run. Define f, pick x = 5, measure the slope.
f <- function(x) x^2
```

::: {.callout-note collapse="true" title="Answer"}
Using the central difference: `slope_at <- function(f, x, h = 1e-5) (f(x+h) - f(x-h))/(2*h)`
then `slope_at(f, 5)` gives essentially **10**. (The exact slope of $x^2$ is $2x = 10$
at $x = 5$ — but you found it by measuring, not by any rule.)
:::

**2.** At a point where a curve is heading *downhill* as $x$ increases, is the measured
slope positive or negative?

::: {.callout-note collapse="true" title="Answer"}
Negative. Going downhill means the output *falls* as $x$ rises, so the rise is
negative, and rise/run comes out below zero. (Uphill → positive; momentarily flat →
zero.)
:::

**3.** In your own words: explain to a colleague, in two sentences, what the derivative
of a function at a point measures, and why we compute it with a tiny step instead of a
single point.

::: {.callout-note collapse="true" title="A model answer"}
The derivative at a point is the steepness of the curve right there — how fast the
output changes for a small nudge in the input. We use a tiny step because steepness is
about *change* between points, so we look at the rise over a run and shrink the run
until it reflects the slope at that single spot.
:::

## What you can now do

- **Measure the slope of any function at any point** numerically — rise over a small
  run — using `(f(x+h) - f(x-h)) / (2*h)`, without knowing a single differentiation
  rule.
- **Read what the slope's value means**: sign tells you uphill / downhill / flat, and
  magnitude tells you how steep.
- **Explain the derivative as a rate of change** (the speedometer picture) and why a
  slope needs two nearby points, not one.
- **Pick a sensible step size** — small, but not so small that rounding noise takes
  over.

If any of these feels shaky, revisit [Getting a feel for it](#getting-a-feel-for-it)
and [Explore it](#explore-it) before moving on — anything that *searches* for the
bottom of a curve leans on all of them.

## Where this goes next

A slope tells you which way is uphill and how steep — so if you're trying to find the
*lowest* point of a curve, the slope is a compass: step against it and you head
downhill. That's the engine of gradient-based optimization, and it's where this tool
goes to work next. Ready for more reps first? Do the [practice problems](practice.qmd).
