---
title: "Practice: The derivative, as a slope you can measure"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

Work these in order — they get harder. Do them on paper or in the live cells; only
then open the answer.

## Problem 1 — the toy example, new numbers

In the lesson we measured the slope of $f(x) = x^2$ at $x = 3$ and got $6$. Now measure
it at $x = 2$. Take a forward step of $h = 0.1$ by hand (rise over run), then check in R.
What value should the slope approach as $h$ shrinks?

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
f <- function(x) x^2
x0 <- 2
```

::: {.callout-note collapse="true" title="Answer"}
By hand with $h = 0.1$: rise $= f(2.1) - f(2) = 4.41 - 4 = 0.41$; run $= 0.1$; slope
$= 0.41 / 0.1 = 4.1$. Shrinking $h$ marches this toward **4**.

```{r p1-answer}
f <- function(x) x^2
x0 <- 2
(f(x0 + 0.1) - f(x0)) / 0.1                 # 4.1, heading for 4
(f(x0 + 1e-5) - f(x0 - 1e-5)) / (2 * 1e-5)  # ~4, the settled-on slope
```

The exact slope of $x^2$ is $2x = 4$ at $x = 2$ — found here by measuring, not by a rule.
:::

## Problem 2 — fill in the missing step

Estimate the slope of $f(x) = x^3$ at $x = 1$ with the central difference, $h = 0.01$.
The work is laid out below with the **core step blanked** — supply just that line.

> Step 1 — the two nearby outputs: $f(1.01) = 1.01^3 = 1.030301$ and
> $f(0.99) = 0.99^3 = 0.970299$.
> Step 2 — **`your step here`**: form the central difference (rise across both steps,
> over the total run $2h = 0.02$).
> Step 3 — read the result: it should be close to $3$.

```{webr}
#| autorun: false
# p2-starter — edit and Run to check the blank
f <- function(x) x^3
x0 <- 1; h <- 0.01
```

::: {.callout-note collapse="true" title="Answer"}
The missing step is the central-difference ratio:
$$
\frac{f(1.01) - f(0.99)}{2h} = \frac{1.030301 - 0.970299}{0.02} = \frac{0.060002}{0.02}
= 3.0001 \;\approx\; 3.
$$
The move is: subtract the two outputs to get the rise, divide by the *total* run $2h$.

```{r p2-answer}
f <- function(x) x^3
x0 <- 1; h <- 0.01
(f(x0 + h) - f(x0 - h)) / (2 * h)   # 3.0001 -> the slope of x^3 at x = 1 is 3
```
:::

## Problem 3 — a curve that falls

Measure the slope of $f(x) = 1/x$ at $x = 2$ using the numerical method. Then say, in
words, what the *sign* of your answer tells you about the curve there.

```{webr}
#| autorun: false
# p3-starter — your job: measure the slope of 1/x at x = 2, then interpret its sign
f <- function(x) 1 / x
x0 <- 2
```

::: {.callout-note collapse="true" title="Answer"}
```{r p3-answer}
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
f <- function(x) 1 / x
slope_at(f, 2)   # about -0.25
```
The slope is about $-0.25$ — **negative**, so the curve $1/x$ is heading *downhill* at
$x = 2$: as $x$ increases, $1/x$ decreases. (It's also fairly gentle — a magnitude of
only a quarter — because $1/x$ has flattened out by $x = 2$.)
:::

## Problem 4 — spot the error

A colleague wants the slope of $f(x) = x^2$ at $x = 3$ (which we know is $6$). They
write the central difference like this — it contains **exactly one mistake**. Find it
and fix it.

```{webr}
#| autorun: false
# p4-starter — this is the buggy code; run it, see the wrong answer, then fix it
f <- function(x) x^2
x0 <- 3; h <- 1e-5
(f(x0 + h) - f(x0 - h)) / h      # they got about 12 — but the slope is 6
```

::: {.callout-note collapse="true" title="Answer"}
The mistake is the **run**: stepping from $x - h$ to $x + h$ covers a distance of
$2h$, not $h$. Dividing by $h$ instead of $2h$ doubles the answer (that's why $6$ came
out as $12$). Fix the denominator:

```{r p4-answer}
f <- function(x) x^2
x0 <- 3; h <- 1e-5
(f(x0 + h) - f(x0 - h)) / (2 * h)   # 6 — correct once the run is the full 2h
```
:::

## Problem 5 — a gentle cliffhanger

Here is the curve $f(x) = (x - 4)^2 + 1$ — a bowl whose lowest point is somewhere near
$x = 4$. You can't solve for the bottom with a formula yet, but you *can* hunt for it
with slopes: the bottom is where the slope is **zero**. Scan a few $x$ values, measure
the slope at each, and find where it flips from negative to positive.

```{webr}
#| autorun: false
# p5-starter — measure the slope at several x; where does it cross zero?
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
f <- function(x) (x - 4)^2 + 1
xs <- 1:7
```

::: {.callout-note collapse="true" title="Answer"}
```{r p5-answer}
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
f <- function(x) (x - 4)^2 + 1
xs <- 1:7
sapply(xs, function(x) slope_at(f, x))   # -6 -4 -2 0 2 4 6
```
The slopes go $-6, -4, -2, 0, +2, +4, +6$ across $x = 1 \dots 7$: negative (downhill)
to the left of $x = 4$, positive (uphill) to the right, and **zero exactly at
$x = 4$** — the bottom of the bowl. That "slope is zero at the lowest point, negative
before it, positive after" pattern is the whole idea behind *rolling downhill to the
minimum* — the tool that turns this scan into an automatic search is next.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/derivative-as-slope" pass="4"}

::: {.quiz-q qid="das-central-difference" qtype="numeric" answer="8" tolerance="0"}
Using the central-difference formula $(f(x+h) - f(x-h))/(2h)$ with a small $h$, estimate
the slope of $f(x) = x^2$ at $x = 4$.
:::

::: {.quiz-q qid="das-flat-slope-zero" qtype="mc" answer="a"}
At a point where a curve is momentarily flat (the top of a hill), what does the measured
slope equal?

a) Zero — the output isn't changing there, so rise over run is $0$
b) Undefined — a flat point has no slope
c) A large positive number — the peak is the steepest part of the curve
:::

::: {.quiz-q qid="das-tiny-h-noise" qtype="mc" answer="b"}
A student sets $h = 10^{-16}$ in the central-difference formula, reasoning that smaller
is always more accurate. What actually happens?

a) The estimate keeps getting more accurate, with no lower limit on $h$
b) $f(x+h)$ and $f(x-h)$ round to nearly the same stored value, so the result is mostly rounding noise
c) The formula returns the exact slope, since $h \to 0$ is the definition
:::

::: {.quiz-q qid="das-two-points-needed" qtype="mc" answer="c"}
Why can't the slope at a point be read from that single point's height alone?

a) Because the height must first be converted into a rate
b) Because a single point's height is always zero
c) Because steepness is about how the height *changes* between nearby points, so it needs at least two
:::

::: {.quiz-q qid="das-negative-slope" qtype="numeric" answer="-6" tolerance="0"}
Using the central difference on $f(x) = x^2$, estimate the slope at $x = -3$. (The sign
tells you which way the curve is heading.)
:::

:::

::::
