---
title: "The colocalisation question"
subtitle: "Two studies, one region, one yes/no question — and the two per-SNP vectors it all runs on"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live {webr} cells below re-state their own data. Base R only.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

## Where we're going

Two research groups each study a stretch of the genome. One ran a GWAS on, say, LDL
cholesterol; the other measured how strongly the same genetic variants change the
*expression* of a nearby gene (an eQTL study). Both find a signal in the same region. The
question this course answers is deceptively simple: **is it the same underlying variant
driving both, or two different variants that just happen to sit close together?** That
question is *colocalisation*. This module builds the picture and, crucially, sets up the
two per-SNP number-vectors — $\ell_1$ and $\ell_2$ — that every formula in the rest of the
course consumes. We do **not** re-derive where those numbers come from; you already built
that machine.

**Builds on:** the [Approximate Bayes Factors course](../../../approximate-bayes-factors/syllabus.md)
(its `wakefield_abf(beta, se, W)` produces the per-SNP log Bayes factors we start from), and
the foundation [distributions-and-densities](../../../../foundations/distributions-and-densities/lesson.qmd).

## Warm-up: do you remember?

Before the new stuff, three quick reps — answers from memory first, then check.

**1.** *(ABF course, Module 1)* A variant reports $\hat\beta = 0.15$ and $\text{se} = 0.05$.
What is its z-statistic $Z$?

::: {.callout-note collapse="true" title="Answer"}
$Z = \hat\beta / \text{se} = 0.15 / 0.05 = 3$. The z-statistic is just the effect estimate
divided by its standard error — taught in
[ABF Module 1](../../../approximate-bayes-factors/modules/01-the-summary-of-an-association/lesson.qmd).
:::

**2.** *(ABF course, Module 3)* In the coloc sign convention, does a **large** log Bayes
factor (lABF) mean strong or weak evidence of association?

::: {.callout-note collapse="true" title="Answer"}
**Strong.** coloc flips Wakefield's original sign so that large lABF = strong evidence *for*
association (alternative over null). See
[ABF Module 3](../../../approximate-bayes-factors/modules/03-the-approximate-bayes-factor/lesson.qmd)
and its capstone.
:::

**3.** *(foundation)* A normal density $N(0, \sigma^2)$ is a bell curve. If you widen it (bigger
$\sigma^2$), does the height at the centre go up or down?

::: {.callout-note collapse="true" title="Answer"}
**Down** — a density always integrates to 1, so spreading it wider pushes the peak lower. See
[distributions-and-densities](../../../../foundations/distributions-and-densities/lesson.qmd).
:::

## A tiny example first

Forget formulas. Picture a short stretch of genome with just **three** genetic variants
(SNPs) — call them SNP 1, SNP 2, SNP 3. We have two studies over exactly these three SNPs:

- **Trait 1** is a GWAS (say, an LDL cholesterol study). For each SNP it reports an effect
  estimate $\hat\beta$ and a standard error.
- **Trait 2** is an eQTL study (how much each SNP changes a nearby gene's expression). It too
  reports a $\hat\beta$ and standard error per SNP.

Here are the summary statistics. Notice SNP 2 has the biggest effect in **both** studies —
that is the pattern that will eventually say "shared."

| SNP | Trait 1 $\hat\beta$ | Trait 1 se | Trait 2 $\hat\beta$ | Trait 2 se |
|---|---|---|---|---|
| 1 | 0.01 | 0.05 | −0.01 | 0.05 |
| 2 | 0.20 | 0.05 | 0.18 | 0.05 |
| 3 | −0.02 | 0.05 | 0.02 | 0.05 |

Now the key move: run each column through the ABF machine you already built. With prior
variance $W = 0.04$ (coloc's case-control default $\sqrt{W} = 0.2$), each SNP in each trait
becomes **one number** — its log Bayes factor for association. By hand for SNP 2, trait 1:
$Z = 0.20/0.05 = 4$, and `wakefield_abf` turns that into an lABF of about $6.11$ (a huge
Bayes factor: $e^{6.11} \approx 452$). The two weak SNPs give small, slightly negative lABFs.

The same thing in R (this *is* `wakefield_abf` from the ABF course — no new math):

```{r toy-example}
# The ABF machine you built in the previous course (base R, coloc sign convention).
wakefield_abf <- function(beta, se, W) {
  z <- beta / se          # z-statistic  (ABF Module 1)
  V <- se^2               # variance of the estimate
  r <- W / (W + V)        # shrinkage factor r  (ABF Module 3)
  0.5 * (log(1 - r) + r * z^2)   # the closed-form log Bayes factor
}

W <- 0.04                                  # coloc case-control default: sqrt(W)=0.2
beta1 <- c(0.01, 0.20, -0.02); se1 <- rep(0.05, 3)   # trait 1 (GWAS)
beta2 <- c(-0.01, 0.18, 0.02); se2 <- rep(0.05, 3)   # trait 2 (eQTL)

l1 <- wakefield_abf(beta1, se1, W)   # per-SNP lABF, trait 1
l2 <- wakefield_abf(beta2, se2, W)   # per-SNP lABF, trait 2
round(rbind(l1 = l1, l2 = l2), 3)
```

Those two rows — `l1` and `l2` — are the entire input to colocalisation. From here on, the
raw $\hat\beta$s and standard errors never appear again.

## Getting a feel for it

**Analogy — two spotlights on the same stage.** Imagine a dark theatre stage lined with
three actors (the SNPs). Two spotlight operators (the two studies) each sweep a beam along
the row and mark where their beam is brightest. Colocalisation asks: **are both spotlights
lighting up the same actor, or are they lighting up two different actors who happen to stand
close together?** If both beams peak on actor 2, the traits plausibly share a causal
variant. If one peaks on actor 1 and the other on actor 3, they are two separate stories
that only *look* joined because the actors are crowded together.

Where the analogy breaks: real spotlights point at exactly one spot, but each study's
"beam" is fuzzy — the lABF vector is a *profile of brightness* across all three SNPs, not a
single pointer. That fuzziness (from finite sample size and from SNPs being correlated) is
exactly why we need a probabilistic test instead of just eyeballing which SNP is biggest.

**Why per-SNP log Bayes factors, and not raw effect sizes?** Two reasons. First, the lABF
already folds in *how sure* we are (a big effect with a huge standard error is not strong
evidence — the ABF course built exactly that shrinkage). Second — and this is the payoff we
cash in Module 3 — Bayes factors **multiply** when you combine independent pieces of
evidence, and adding their logs is the numerically safe way to multiply. Starting from lABFs
sets us up to combine two studies with additions instead of dangerous products of tiny
numbers.

**What the numbers feel like.** Look at the toy `l1`, `l2`: SNP 2's entries are around $+6$
and $+4.7$ (loud), the other four are around $-1.4$ (quiet, slightly *against* association).
Both loud entries land on the **same** SNP — the shared-signal fingerprint. If instead
trait 1 were loud on SNP 1 and trait 2 loud on SNP 3, the loud entries would sit in
different columns: the distinct-signal fingerprint. The whole course is machinery for
turning "which columns are loud, and do they line up?" into five clean probabilities.

## What the notation is going to say

There is no new formula in this module — only new *names*. The notation will say: for a
region with $Q$ SNPs and two traits, collect trait 1's per-SNP log Bayes factors into a
vector $\ell_1 = (\ell_{1,1}, \ell_{1,2}, \dots, \ell_{1,Q})$, and trait 2's into
$\ell_2$. In our toy example $Q = 3$, and $\ell_1$ is exactly the `l1` row we just computed:
$\ell_{1,2} \approx 6.11$ is trait 1's log Bayes factor at SNP 2. That's it — these two
vectors are the alphabet of the whole course.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $Q$ | "cue" | how many SNPs are in the region (in R: `length(l1)`) | 3 |
| $i$ | "eye" | which SNP we're pointing at (a row index) | 1, 2, or 3 |
| $\ell_{1,i}$ | "ell one sub eye" | trait 1's log Bayes factor for association at SNP $i$ (`l1[i]`) | $\ell_{1,2}\approx 6.11$ |
| $\ell_{2,i}$ | "ell two sub eye" | trait 2's log Bayes factor for association at SNP $i$ (`l2[i]`) | $\ell_{2,2}\approx 4.68$ |
| $\ell_1, \ell_2$ | "ell one, ell two" | the whole per-SNP lABF vector for each trait | the two rows above |
| $\hat\beta$ | "beta hat" | an effect estimate from a fitted study (a log odds ratio for a case-control GWAS) | $0.20$ at SNP 2, trait 1 |
| $\text{se}$ | "ess ee" | the standard error of an effect estimate | $0.05$ |
| $Z$ | "zee" | the z-statistic $\hat\beta/\text{se}$ | $4$ at SNP 2, trait 1 |
| $W$ | "double-you" | the prior variance on the true effect (coloc: $\sqrt{W}=0.2$) | $0.04$ |

## The formal version

No derivation here — the "formal version" of this module is just the definition of the two
vectors, tied to the toy numbers. For a region of $Q$ SNPs:

$$
\ell_1 = \big(\ell_{1,1}, \dots, \ell_{1,Q}\big), \qquad
\ell_{1,i} = \text{wakefield\_abf}(\hat\beta_{1,i}, \text{se}_{1,i}, W),
$$

and the same for trait 2 with $\ell_2$.

Reading it aloud: "ell-one is the list, over SNPs 1 through Q, of trait 1's Wakefield log
Bayes factor at each SNP" — or in everyday words: "run every SNP's summary statistics for
trait 1 through the ABF machine and stack the answers into a vector; do it again for trait
2."

Mapping it onto the toy example:

- $\ell_{1,2} \approx 6.11$ is the "$Z = 4$ at SNP 2 → lABF" step for trait 1.
- $\ell_{2,2} \approx 4.68$ is the same for trait 2 (its $Z = 3.6$).
- The other four entries ($\approx -1.4$) are the quiet SNPs.

## Explore it

```{webr}
#| autorun: true
# TRY IT: move the big signal. Right now SNP 2 is loud in BOTH traits (shared-looking).
#         Change beta2 so trait 2 is loud on SNP 3 instead -> a distinct-looking pattern.
# PREDICT: before running, which SNP's l1 will be largest? which SNP's l2?
wakefield_abf <- function(beta, se, W) {
  z <- beta / se; V <- se^2; r <- W / (W + V)
  0.5 * (log(1 - r) + r * z^2)
}
W <- 0.04
beta1 <- c(0.01, 0.20, -0.02); se1 <- rep(0.05, 3)   # trait 1 (GWAS)
beta2 <- c(-0.01, 0.18, 0.02); se2 <- rep(0.05, 3)   # trait 2 (eQTL)

l1 <- wakefield_abf(beta1, se1, W)
l2 <- wakefield_abf(beta2, se2, W)
data.frame(snp = 1:3, Z1 = beta1/se1, Z2 = beta2/se2,
           l1 = round(l1, 3), l2 = round(l2, 3))
```

```{webr}
#| autorun: true
#| fig-alt: "Two side-by-side bar charts over three SNPs. In both charts SNP 2's bar towers above SNPs 1 and 3, showing both traits' log Bayes factors peak on the same SNP — the shared-signal fingerprint."
# LIVE: SEE the two brightness profiles. Do the loud bars line up on the same SNP?
# PREDICT: before running, will BOTH charts' tallest bar sit over the same SNP number?
wakefield_abf <- function(beta, se, W) { z<-beta/se; V<-se^2; r<-W/(W+V); 0.5*(log(1-r)+r*z^2) }
W <- 0.04
l1 <- wakefield_abf(c(0.01,0.20,-0.02), rep(0.05,3), W)
l2 <- wakefield_abf(c(-0.01,0.18,0.02), rep(0.05,3), W)

op <- par(mfrow = c(1, 2))
barplot(l1, names.arg = paste0("SNP", 1:3), main = "Trait 1 (GWAS): l1", ylab = "lABF")
barplot(l2, names.arg = paste0("SNP", 1:3), main = "Trait 2 (eQTL): l2", ylab = "lABF")
par(op)
```

## Common traps

**Trap: "the study with the bigger effect size wins."** No — colocalisation never compares
the two studies' effect *sizes* against each other, and it doesn't even use raw $\hat\beta$s
after this step. It compares *where each study's evidence peaks*. In the toy data trait 1's
big effect ($0.20$) is larger than trait 2's ($0.18$), but that's irrelevant; what matters
is that **both** peak on SNP 2. A tiny-effect trait and a huge-effect trait can perfectly
colocalise, and two huge-effect traits can fail to.

**Trap: "if both studies are significant in the region, they colocalise."** Also no. That is
exactly the mistake the method exists to prevent. Two studies can both have a genuine,
strong signal in the same region and yet be driven by *different* causal SNPs sitting close
together (linkage). "Both significant here" is the *setup* for the question, not the answer —
distinguishing shared (PP4) from distinct (PP3) is the whole point, and we build it in
Module 2.

## Check yourself

**1.** A region has $Q = 5$ SNPs. How many entries does the vector $\ell_2$ have, and what
does its 4th entry $\ell_{2,4}$ mean?

::: {.callout-note collapse="true" title="Answer"}
$\ell_2$ has 5 entries (one per SNP). $\ell_{2,4}$ is trait 2's log Bayes factor for
association at SNP 4 — how strong the eQTL evidence is that SNP 4 specifically is a causal
variant for trait 2.
:::

**2.** For a SNP with $\hat\beta = 0.10$, $\text{se} = 0.05$, $W = 0.04$: is its lABF for
that trait positive or negative? (Try it, then predict what happens if $\hat\beta$ doubles.)

```{webr}
#| autorun: false
# check2-starter — edit and Run
wakefield_abf <- function(beta, se, W) { z<-beta/se; V<-se^2; r<-W/(W+V); 0.5*(log(1-r)+r*z^2) }
beta <- 0.10; se <- 0.05; W <- 0.04
```

::: {.callout-note collapse="true" title="Answer"}
$Z = 0.10/0.05 = 2$, giving lABF $\approx 0.47$ — slightly **positive** (mild evidence for
association). Double $\hat\beta$ to $0.20$ and $Z = 4$; the lABF jumps to $\approx 6.1$
because the $r Z^2$ term grows with $Z$ squared. Evidence climbs fast as the z-statistic
grows.
:::

**3.** In your own words: explain to a colleague, in two sentences, what colocalisation asks
and why we start from per-SNP log Bayes factors instead of raw effect sizes.

::: {.callout-note collapse="true" title="A model answer"}
Colocalisation asks whether two traits' association signals in one region are driven by the
same causal variant (shared) or by two different variants that happen to sit close together
(distinct). We start from per-SNP log Bayes factors because they already weigh each SNP's
evidence by how certain it is, and because working in log space lets us combine the two
studies' evidence by adding rather than by multiplying tiny numbers.
:::

## What you can now do

- **Describe the colocalisation question** in plain words for two studies over one region,
  and say why "both significant here" is not the same as "shared causal variant."
- **Turn two studies' per-SNP summary statistics into the two lABF vectors** $\ell_1$ and
  $\ell_2$ using the `wakefield_abf` machine from the previous course.
- **Read the shared-vs-distinct fingerprint** off the two lABF profiles: do the loud SNPs
  line up on the same SNP, or not?

If any of these feels shaky, revisit the toy example above — or the
[ABF course capstone](../../../approximate-bayes-factors/modules/05-capstone-reading-wakefields-abf/lesson.qmd)
if `wakefield_abf` itself is hazy — before continuing; the next module assumes you can
produce $\ell_1$ and $\ell_2$ on demand.

## Where this goes next

Now that every SNP in every trait is a single log Bayes factor, we can state the actual
question precisely. [Module 2](../02-five-hypotheses/lesson.qmd) turns "shared or distinct?"
into **five** competing hypotheses about the region — $H_0$ through $H_4$ — with PP3
(distinct) and PP4 (shared) as the two answers you'll care about. Ready for reps? Do the
[practice problems](practice.qmd).
