---
title: "Practice: The colocalisation question"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

Work these in order — they get harder. Do them on paper or in an R session; only then open
the answer.

## Problem 1 — the toy example, new numbers

A three-SNP region has these summary statistics ($W = 0.04$). Build the two lABF vectors
$\ell_1$ and $\ell_2$, and say which SNP looks loud in each trait.

| SNP | Trait 1 $\hat\beta$ | Trait 1 se | Trait 2 $\hat\beta$ | Trait 2 se |
|---|---|---|---|---|
| 1 | 0.03 | 0.04 | 0.22 | 0.04 |
| 2 | −0.01 | 0.04 | 0.01 | 0.04 |
| 3 | 0.24 | 0.04 | −0.02 | 0.04 |

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
wakefield_abf <- function(beta, se, W) { z<-beta/se; V<-se^2; r<-W/(W+V); 0.5*(log(1-r)+r*z^2) }
beta1 <- c(0.03, -0.01, 0.24); se1 <- rep(0.04, 3)
beta2 <- c(0.22, 0.01, -0.02); se2 <- rep(0.04, 3)
W <- 0.04
```

::: {.callout-note collapse="true" title="Answer"}
Run each column through `wakefield_abf`:

```{r}
wakefield_abf <- function(beta, se, W) { z<-beta/se; V<-se^2; r<-W/(W+V); 0.5*(log(1-r)+r*z^2) }
l1 <- wakefield_abf(c(0.03, -0.01, 0.24), rep(0.04, 3), 0.04)   # trait 1
l2 <- wakefield_abf(c(0.22, 0.01, -0.02), rep(0.04, 3), 0.04)   # trait 2
round(rbind(l1 = l1, l2 = l2), 3)
```

Trait 1 is loud on **SNP 3** ($Z = 0.24/0.04 = 6$, a big lABF); trait 2 is loud on **SNP 1**
($Z = 5.5$). The loud SNPs are *different* — this looks like a **distinct** pattern (heading
toward PP3, once we can compute it).
:::

## Problem 2 — fill in the missing step

For a single SNP with $\hat\beta = 0.18$, $\text{se} = 0.06$, $W = 0.04$, here is the lABF
computation with one step blanked. Supply the missing step.

- Step 1: $Z = \hat\beta / \text{se} = 0.18 / 0.06 = 3$, and $V = \text{se}^2 = 0.0036$.
- Step 2: **your step here** — compute the shrinkage factor $r$.
- Step 3: $\text{lABF} = \tfrac12\big(\log(1 - r) + r Z^2\big)$.

```{webr}
#| autorun: false
# p2-starter — edit and Run to check the blank
beta <- 0.18; se <- 0.06; W <- 0.04
```

::: {.callout-note collapse="true" title="Answer"}
The missing move is $r = W/(W + V)$, the shrinkage factor:
$$ r = \frac{0.04}{0.04 + 0.0036} = \frac{0.04}{0.0436} \approx 0.9174. $$
That's the one number that turns the raw $Z$ into a Bayes factor — it says what fraction of the
total variance is "real prior effect" versus noise. Then
$\text{lABF} = \tfrac12(\log(1 - 0.9174) + 0.9174 \times 9) \approx \tfrac12(-2.494 + 8.257)
\approx 2.88$.

```{r}
beta <- 0.18; se <- 0.06; W <- 0.04
V <- se^2; r <- W/(W+V); Z <- beta/se
c(r = r, lABF = 0.5*(log(1-r) + r*Z^2))
```
:::

## Problem 3 — read the fingerprint

Here are two lABF vectors already computed for a four-SNP region. Without any more math, say
whether this region looks **shared** or **distinct**, and name the SNP(s) involved.

```{webr}
#| autorun: false
# p3-starter — your job: eyeball the loud SNPs in each trait
l1 <- c(-1.2, 5.8, -0.9, -1.1)   # trait 1 lABFs
l2 <- c(-1.0, 6.3, -1.3, -0.8)   # trait 2 lABFs
```

::: {.callout-note collapse="true" title="Answer"}
Both traits are loud on **SNP 2** (lABF $\approx 5.8$ and $6.3$; every other SNP is small and
negative). The loud SNPs line up on the *same* SNP → this looks **shared** (heading toward a
high PP4). Recall from the lesson: colocalisation is about whether the loud SNPs coincide, not
about which trait's number is bigger.
:::

## Problem 4 — spot the error

A colleague wants trait 1's lABF at a SNP with $\hat\beta = 0.20$, $\text{se} = 0.05$,
$W = 0.04$. They wrote this — it contains **exactly one mistake**. Find and fix it.

```{webr}
#| autorun: false
# p4-starter — there is one bug in this computation
beta <- 0.20; se <- 0.05; W <- 0.04
Z <- beta / se          # 4
V <- se                 # <- they used se, not se^2
r <- W / (W + V)
lABF <- 0.5 * (log(1 - r) + r * Z^2)
```

::: {.callout-note collapse="true" title="Answer"}
The bug is `V <- se`: the variance is $V = \text{se}^2$, **not** the standard error itself.
With `V <- se` you get $V = 0.05$ (should be $0.0025$), so $r = 0.04/0.09 \approx 0.444$ and a
badly wrong lABF. Fixing it:

```{r}
beta <- 0.20; se <- 0.05; W <- 0.04
Z <- beta / se
V <- se^2                # the fix: variance is se squared
r <- W / (W + V)
c(V = V, r = r, lABF = 0.5*(log(1-r) + r*Z^2))   # lABF ~ 6.11, the correct value
```

This is the same "variance vs standard error" trap you'll meet again in `coloc.abf`, where the
argument is `varbeta` ($= \text{se}^2$), not `se`.
:::

## Problem 5 — mixing it with the ABF course *(uses the ABF course's posterior-of-association step)*

A SNP has trait-1 lABF $= 6.11$. Two different questions:
(a) Within a region assumed to hold exactly one causal SNP, its *regional* posterior weight is
$e^{\text{lABF}}$ relative to the others. (b) Its *genome-wide* posterior probability of
association uses the tiny prior $\pi = 10^{-4}$: posterior odds $= \tfrac{\pi}{1-\pi}
e^{\text{lABF}}$, then PPA $=$ odds$/(1+$odds$)$. Compute the genome-wide PPA.

```{webr}
#| autorun: false
# p5-starter — edit and Run to solve
labf <- 6.11; prior <- 1e-4
```

::: {.callout-note collapse="true" title="Answer"}
This is the posterior-odds step from the
[ABF course Module 4](../../../approximate-bayes-factors/modules/04-posterior-probability-of-association/lesson.qmd):

```{r}
labf <- 6.11; prior <- 1e-4
po <- (prior/(1-prior)) * exp(labf)   # posterior odds
c(post_odds = po, PPA = po/(1+po))     # PPA ~ 0.043
```

The genome-wide PPA is only about **0.043** — even a Bayes factor of $e^{6.11} \approx 450$
barely moves a $10^{-4}$ prior. That's the key contrast: colocalisation works with the
*regional* evidence ($e^{\text{lABF}}$ and the sums of it), not the tiny genome-wide prior, so
its numbers can be decisive where the genome-wide PPA is not.
:::

## Problem 6 — transfer / cliffhanger

You have $\ell_1 = \log(1, 2, 5)$ and $\ell_2 = \log(2, 1, 4)$ for a three-SNP region. You want
the total evidence that trait 1 has *some* causal SNP here — i.e. $\sum_i \text{BF}_{1,i}$ — but
computed from the **logs** without leaving log space. You can't fully do this yet; sketch what
you'd need.

::: {.callout-note collapse="true" title="Answer"}
You need a "log of a sum" helper: $\text{logsum}(\ell_1) = \log\big(e^{\ell_{1,1}} +
e^{\ell_{1,2}} + e^{\ell_{1,3}}\big) = \log(1 + 2 + 5) = \log 8 \approx 2.079$. Doing this
*stably* (subtracting the max first, so huge real lABFs don't overflow) is `logsum`, which
[Module 3](../03-combining-abfs-with-priors/lesson.qmd) builds — along with its partner
`logdiff` for the distinct-evidence subtraction. That module turns these evidence sums, plus
the priors, into the five posterior probabilities.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**, once the
material isn't fresh. A pass then is real evidence you've retained it. Five questions, graded on
the page; 4/5 or better means this module is solid. Score under 4? Revisit what the missed
questions point at, then retake in a **later session** — passing after another gap is the
evidence that counts. Results are saved in this browser only.

::: {.module-check module="courses/bayesian-colocalisation/modules/01-the-colocalisation-question" pass="4"}

::: {.quiz-q qid="cq-what-coloc-asks" qtype="mc" answer="c"}
What does colocalisation ask about two studies' signals in one genomic region?

a) Which study has the larger effect size
b) Whether both studies reach genome-wide significance
c) Whether the same causal variant drives both, or two different ones do
d) How many SNPs are in the region
:::

::: {.quiz-q qid="cq-length-l2" qtype="numeric" answer="7" tolerance="0"}
A region has 7 SNPs. How many entries does the vector $\ell_2$ have?
:::

::: {.quiz-q qid="cq-both-sig" qtype="mc" answer="b"}
Both studies have a strong, real signal in the same region. What does that tell you about
colocalisation?

a) They definitely share a causal variant
b) Nothing yet — they could share one variant or have two distinct nearby variants
c) They definitely have distinct variants
d) The region has no causal variant
:::

::: {.quiz-q qid="cq-labf-from-z" qtype="numeric" answer="6.113" tolerance="0.01"}
For a SNP with $\hat\beta = 0.20$, $\text{se} = 0.05$, $W = 0.04$: what is its lABF (coloc
convention)? Enter to 3 decimals.
:::

::: {.quiz-q qid="cq-abf-large-strong" qtype="mc" answer="a"}
*(interleaved — from the Approximate Bayes Factors course, Module 3.)* In coloc's sign
convention, a **large** log Bayes factor means:

a) Strong evidence for association
b) Strong evidence against association
c) The prior variance $W$ is large
d) The standard error is large
:::

:::

::::
