---
title: "Likelihood and log-likelihood"
subtitle: "Flip the question: with the data fixed, which parameter value makes what you saw least surprising?"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live cells below re-state their own data. Base R is all
# we need, so nothing to install.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

::: {.callout-tip title="This lesson runs in your browser" collapse="true"}
The code cells below with a **Run** button are live — powered by WebR (R compiled to
WebAssembly), they run right here on your phone or laptop, no install needed. Edit any
of them and re-run. The first run downloads a small R engine, so give it a few seconds.
Cells without a Run button are ordinary listings whose output was computed when the
page was built.
:::

## Where we're going

You already know how to ask "given a coin with probability $p$ of heads, how likely is
it to see 2 heads in 3 flips?" — that's a **probability** question, with $p$ fixed and
the data varying. This lesson **flips the question around**: you've *already flipped the
coin and seen the data* — now which value of $p$ makes that data least surprising? That
flipped quantity is the **likelihood**, and the value of $p$ that maximizes it is the
**maximum-likelihood estimate** — the single most important recipe in applied
statistics for turning data into a parameter estimate.

We'll build it with a plain coin: multiply the per-flip pieces into a likelihood, take
logs to turn that product into a tame sum, and find the peak. We will *not* touch any
particular model's likelihood formula, nor the machinery of optimization — just the
idea, on numbers you can check by hand.

**Builds on:** [probability, odds & log-odds](../probability-and-odds/lesson.qmd) — a
probability is a number in $[0,1]$ and independent events multiply; and
[exponentials & logs](../exponentials-and-logs/lesson.qmd) — the log rule that turns a
product into a sum, $\log(ab) = \log a + \log b$, is the whole reason log-likelihood
exists.

## Warm-up: do you remember?

Before the new stuff, three quick reps — answers from memory first, then check.

**1.** A bent coin lands heads with probability $p = 0.6$. What is its probability of
*tails*? State the one-line rule you used.

::: {.callout-note collapse="true" title="Answer"}
Tails probability is $1 - p = 1 - 0.6 = 0.4$. The rule: "probability of not-A is one
minus probability of A" — the two outcomes must together account for everything, so
they sum to 1. Shaky? Revisit
[probability, odds & log-odds](../probability-and-odds/lesson.qmd).
:::

**2.** Two independent events have probabilities $0.6$ and $0.4$. What is the
probability that **both** happen? (Independent = one outcome tells you nothing about the
other.)

::: {.callout-note collapse="true" title="Answer"}
Multiply: $0.6 \times 0.4 = 0.24$. For **independent** events, the probability that all
of them happen is the *product* of their individual probabilities. That multiply is the
engine of the whole lesson. See
[probability, odds & log-odds](../probability-and-odds/lesson.qmd).
:::

**3.** Read this aloud and give its value: $\log(0.6 \times 0.4)$ written as a *sum* of
two logs. Which log rule are you using?

```{webr}
#| autorun: false
# warmup3-starter — edit and Run to check (log() with no base is the NATURAL log)
a <- 0.6
b <- 0.4
```

::: {.callout-note collapse="true" title="Answer"}
"The log of a product equals the sum of the logs": $\log(0.6 \times 0.4) =
\log(0.6) + \log(0.4)$. In R, `log(0.6 * 0.4)` and `log(0.6) + log(0.4)` both give
about $-1.427$. This **product-becomes-sum** rule is exactly what we'll use to tame the
likelihood. See [exponentials & logs](../exponentials-and-logs/lesson.qmd).
:::

## A tiny example first

You flip a coin **three times** and see **heads, heads, tails**. That's your data — it
is now *fixed*, it already happened. The coin has some unknown probability of heads;
call it $p$. You want to reason about $p$.

Here's the key move. Ask: *if the true probability of heads were some particular value
$p$, how probable was the exact sequence I saw?* Because the flips are independent, you
**multiply** the per-flip probabilities: heads contributes $p$, heads again contributes
$p$, tails contributes $1 - p$. So the probability of "heads, heads, tails" under that
$p$ is

$$
p \times p \times (1 - p).
$$

Now try a few guesses for $p$ and see which makes the observed data most probable:

- **$p = 0.5$** (a fair coin): $0.5 \times 0.5 \times 0.5 = 0.125$.
- **$p = 0.6$**: $0.6 \times 0.6 \times 0.4 = 0.144$.
- **$p = 0.8$**: $0.8 \times 0.8 \times 0.2 = 0.128$.

So a coin with $p = 0.6$ makes "heads, heads, tails" *more* probable ($0.144$) than a
fair coin does ($0.125$), and even more than a strongly-biased $p = 0.8$ coin ($0.128$).
That number — "how probable was my fixed data, as a function of the guessed $p$" — is
the **likelihood** of $p$. The value of $p$ that makes it biggest is the **best
estimate** of the coin's bias from this data.

The same thing in R, so you can check every step:

```{r toy-example}
# The likelihood of a value p, given our FIXED data H, H, T:
# two heads contribute p each, one tail contributes (1 - p). Multiply (independence).
lik <- function(p) p * p * (1 - p)

lik(0.5)   # 0.125  -- a fair coin
lik(0.6)   # 0.144  -- makes H,H,T MORE probable
lik(0.8)   # 0.128  -- too biased; worse than 0.6

# Which of these three guesses wins?
c(p0.5 = lik(0.5), p0.6 = lik(0.6), p0.8 = lik(0.8))
```

## Getting a feel for it

**The mental picture: a dial, not a die.** In a probability question the coin's bias
$p$ is *bolted down* and you imagine the data wobbling — "how often would 2-heads-in-3
show up?" Likelihood turns the whole setup inside out: the data is bolted down (you
already saw heads, heads, tails), and now *you* hold a dial labelled $p$ and turn it.
At each dial setting you read off one number: "how probable was my data at this
setting?" Sweep the dial from 0 to 1 and you trace out a curve; its highest point is
the setting the data likes best. Same formula $p\,p\,(1-p)$ as a probability — but you
read it as a function of the dial, not of the data. (Where the picture strains: a
likelihood is *not* a probability distribution over $p$ — the area under the curve
doesn't have to be 1, and you can't ask "the probability that $p = 0.6$." It's only a
*score* for comparing dial settings.)

**Why multiply the per-flip pieces?** Because the flips are independent — one flip's
result says nothing about the next — and independent events combine by *multiplication*
(warm-up 2). Adding them would be nonsense: three certain events ($p = 1$ each) would
"add" to 3, but a probability can never exceed 1. Multiplying keeps everything a proper
probability: $1 \times 1 \times 1 = 1$, and any doubt on any flip drags the product
*down*, which is exactly what "all of these had to happen" should feel like.

**What the number feels like — and why the peak sits where it does.** Our data was 2
heads out of 3 flips, a heads-rate of $2/3 \approx 0.667$. The likelihood peaks right
there, at $p = 2/3$ — the dial setting that says "heads is a two-thirds coin" matches
what you saw and so makes it least surprising. Turn the dial too low (say $0.2$) and
two heads become a fluke, so the score collapses (to $0.032$); turn it too high (say
$0.9$) and the lone tail becomes the fluke, so the score falls again. The likelihood
numbers themselves are small ($0.125$, $0.144$…) and get *tiny* fast as you add flips —
hold that thought, it's the whole reason for the next section.

## What the notation is going to say

The formula will just spell out the recipe you already did by hand: **for each
observation, write down its probability under the guessed $p$, then multiply them all
together** — heads gives $p$, tails gives $1 - p$. For our three flips that product is
$p \times p \times (1-p)$, the exact "$0.6 \times 0.6 \times 0.4 = 0.144$" from above.
Then the **log-likelihood** applies one rule from [exponentials &
logs](../exponentials-and-logs/lesson.qmd) — "a log of a product is a sum of logs" — to
turn that long multiplication into a plain *addition*, which is both easier to
differentiate and safe from the tiny-number problem.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $p$ | "pee" | the parameter we're estimating: the coin's probability of heads, a number in $[0,1]$ | the dial we turn; best value $2/3$ |
| $1 - p$ | "one minus pee" | the probability of the *opposite* outcome (tails) | $0.4$ when $p = 0.6$ |
| $x_i$ | "ex sub eye" | the $i$-th observation (here: 1 for heads, 0 for tails) (in R: `x[i]`) | $x_1=1, x_2=1, x_3=0$ |
| $n$ | "en" | the number of observations (in R: `length(x)`) | 3 |
| $k$ | "kay" | the number of "successes" (heads) | 2 |
| $L(p)$ | "the likelihood of pee" | how probable the FIXED data is, as a function of the guessed $p$ | $L(0.6) = 0.144$ |
| $\prod_{i=1}^{n}$ | "the product from i equals 1 to n" | multiply the term to its right, once per observation (in R: `prod(...)`) | multiplies the 3 per-flip pieces |
| $\sum_{i=1}^{n}$ | "the sum from i equals 1 to n" | add the term to its right, once per observation (in R: `sum(...)`) | adds the 3 per-flip log-pieces |
| $\log$ | "log" (natural log) | the natural logarithm, base $e$ — turns a product into a sum (in R: `log(x)`) | $\log(0.144) \approx -1.938$ |
| $\ell(p)$ | "ell of pee" (script L) | the log-likelihood: $\log L(p)$ | $\ell(0.6) \approx -1.938$ |
| $\arg\max$ | "arg max" | the VALUE of the parameter that makes the expression biggest (not the biggest value itself) | $\arg\max_p L(p) = 2/3$ |
| $\hat{p}$ | "pee hat" | our *estimate* of $p$ from data — the hat means "estimated from data" | $\hat{p} = 2/3$ |

The hat on $\hat{p}$ ("pee hat") is the standard flag for "this is estimated from data,
not the unknown truth" — see
[reading math notation](../reading-math-notation/lesson.qmd) for the accent
conventions.

## The formal version

**The likelihood.** Give each observation its probability under $p$ and multiply. Using
$\prod$ ("the product") the way $\sum$ meant "the sum" in
[vectors & summation](../vectors-and-summation/lesson.qmd):

$$
L(p) = \prod_{i=1}^{n} \big(\text{probability of observation } i \text{ under } p\big)
$$

For a coin, "heads" has probability $p$ and "tails" has probability $1 - p$. If we
write each flip as $x_i$ (1 for heads, 0 for tails), a compact way to say "$p$ for
heads, $1-p$ for tails" is $p^{x_i}(1-p)^{1-x_i}$ — check it: heads ($x_i = 1$) gives
$p^1(1-p)^0 = p$, tails ($x_i = 0$) gives $p^0(1-p)^1 = 1-p$. So

$$
L(p) = \prod_{i=1}^{n} p^{x_i}(1-p)^{1-x_i} = p^{k}(1-p)^{n-k},
$$

where $k$ is the number of heads. The last step just counts: every head contributes one
factor of $p$ (there are $k$ of them) and every tail one factor of $1-p$ (there are
$n - k$).

Reading it aloud: "the likelihood of $p$ equals the product, from $i$ equals 1 to $n$,
of $p$ to the $x_i$ times one-minus-$p$ to the one-minus-$x_i$ — which collapses to $p$
to the $k$ times one-minus-$p$ to the $n$-minus-$k$." Everyday words: **"multiply a
factor of $p$ for each head and a factor of $1-p$ for each tail."**

Mapping it onto the toy example ($n = 3$, $k = 2$): $L(p) = p^{2}(1-p)^{1}$, so
$L(0.6) = 0.6^2 \times 0.4 = 0.144$ — exactly the hand number.

**The log-likelihood.** Now take the log of $L(p)$ and use the log rules from
[exponentials & logs](../exponentials-and-logs/lesson.qmd) — one algebra step per line:

$$
\begin{aligned}
\ell(p) = \log L(p) &= \log\!\big(p^{k}(1-p)^{n-k}\big) \\
                    &= \log(p^{k}) + \log\!\big((1-p)^{n-k}\big) \\
                    &= k\log p + (n-k)\log(1-p)
\end{aligned}
$$

- Line 2: **a log of a product is a sum of logs** — $\log(ab) = \log a + \log b$
  (taught in [exponentials & logs](../exponentials-and-logs/lesson.qmd)).
- Line 3: **a log of a power pulls the exponent out front** — $\log(a^b) = b\log a$
  (same module).

Reading it aloud: "ell of $p$ equals $k$ times log $p$, plus $n$-minus-$k$ times log of
one-minus-$p$." Everyday words: **"the multiplication became a plain addition."** The
peak didn't move — $\log$ is an increasing function, so whatever $p$ makes $L$ biggest
makes $\ell$ biggest too — but the arithmetic is now a sum, which is both easier to work
with and immune to the tiny-number trap.

Sanity-check against the hand number ($n=3$, $k=2$, $p=0.6$):
$\ell(0.6) = 2\log(0.6) + 1\log(0.4) \approx -1.938$, and $\log(0.144) \approx -1.938$ —
they agree.

**The maximum-likelihood idea.** The best estimate is the $p$ that makes the data least
surprising — the top of the curve:

$$
\hat{p} = \arg\max_{p}\; L(p) = \arg\max_{p}\; \ell(p).
$$

Reading it aloud: "$p$-hat is the arg-max over $p$ of the likelihood" — everyday words:
**"the value of $p$ where the likelihood is highest."** That $\hat{p}$ is called the
**maximum-likelihood estimate (MLE)**. For the coin, the peak sits at the observed
heads-rate, $\hat{p} = k/n = 2/3$ — which matches the intuition above.

**Optional — how the peak is actually found** (one line of what the depth buys: it names
the tool the next courses use, so the phrase "set the derivative to zero" isn't a
mystery later). Two routes, both owned by other modules: **(a)** solve
"slope of $\ell$ equals 0" with calculus — the top of a smooth hill is where the slope
flattens ([derivative as slope](../derivative-as-slope/lesson.qmd) is the slope idea);
for this coin that equation gives $\hat p = k/n$ exactly. **(b)** *climb* $\ell$
numerically, taking uphill steps until the slope flattens — that's gradient
ascent/descent, taught elsewhere. This lesson stops at "the MLE is the top of the
log-likelihood"; *finding* the top is the optimization module's job.

## Explore it

```{webr}
#| autorun: true
# explore-1 — LIVE: likelihood is "how probable is my FIXED data, as p varies."
# Data is H, H, T (2 heads, 1 tail). We turn the p dial and read the score.
# TRY IT: change the guesses in `ps`. PREDICT: which p gives the biggest likelihood
# -- and roughly what number is the heads-rate 2/3? Commit, then run.
lik <- function(p) p^2 * (1 - p)          # p for each head, (1-p) for the tail

ps <- c(0.2, 0.5, 0.6, 2/3, 0.8)
round(setNames(lik(ps), ps), 4)           # peak sits at p = 2/3, the heads-rate
```

```{webr}
#| autorun: true
#| fig-alt: "The likelihood curve for the coin data H,H,T: probability-of-data on the vertical axis against the parameter p from 0 to 1, a hump that rises from zero, peaks at p equals two-thirds, and falls back to zero at p equals 1. A dashed vertical line marks the peak at two-thirds."
# explore-2 — LIVE: draw the whole likelihood curve and mark its peak.
# TRY IT: change the data by editing k (heads) and n (flips) -- watch the peak slide
# to k/n. PREDICT: if you set k <- 1, n <- 4, where will the peak land? Guess, run.
k <- 2; n <- 3
lik <- function(p) p^k * (1 - p)^(n - k)

p_grid <- seq(0, 1, length.out = 400)
plot(p_grid, lik(p_grid), type = "l", lwd = 2,
     xlab = "p (the dial)", ylab = "likelihood L(p)",
     main = "How probable is H,H,T, as p varies")
peak <- k / n                              # the maximum-likelihood estimate
abline(v = peak, lty = 2)
text(peak, max(lik(p_grid)), paste0("  peak at k/n = ", round(peak, 3)), pos = 4)
```

```{webr}
#| autorun: true
# explore-3 — LIVE: logs turn the product into a sum, and the peak does NOT move.
# The two ways below must agree, and both must argmax at the same p.
# TRY IT: change k and n. PREDICT: will log(L) be positive or negative here? (Hint:
# L is between 0 and 1, and log of a number below 1 is negative.) Guess, then run.
k <- 2; n <- 3
p_grid <- seq(0.001, 0.999, length.out = 999)

L  <- p_grid^k * (1 - p_grid)^(n - k)          # likelihood  (a product)
l  <- k * log(p_grid) + (n - k) * log(1 - p_grid)  # log-likelihood (a sum)

# Same argmax from both -- taking logs never moves the peak:
c(argmax_L = p_grid[which.max(L)],
  argmax_loglik = p_grid[which.max(l)])         # both about 0.667 = 2/3
```

```{webr}
#| autorun: false
# explore-4 — LIVE: why we bother with logs. Pile up 50 flips (30 heads).
# PREDICT: is the raw likelihood a normal-sized number, or absurdly tiny? Guess, run.
k <- 30; n <- 50; p <- 0.6

raw <- p^k * (1 - p)^(n - k)     # the product -- watch how tiny it gets
log_version <- k * log(p) + (n - k) * log(1 - p)  # the sum -- a tame number

c(raw_likelihood = raw, log_likelihood = log_version)
# raw is ~2.4e-15 (heading for underflow as n grows); the log is a comfortable -33.65.
```

## Common traps

**Trap 1 — "likelihood is the probability that $p$ takes this value."** Tempting,
because $L(0.6) = 0.144$ looks like a probability and $p$ is the thing we're unsure
about. But it is *not* a probability *of $p$*. It's the probability of the **data** read
as a function of $p$. Two tells: the likelihood values across different $p$ do **not**
add up to 1 (sum $L$ over a grid of $p$ and you get whatever you get, not 1), and the
peak value $0.148$ isn't "a 15% chance that $p = 2/3$." Likelihood *ranks* dial
settings; it does not assign probabilities to them.

**Trap 2 — "combine the flips by adding their probabilities."** With
$p = 0.6$ someone might write $0.6 + 0.6 + 0.4 = 1.6$ for "heads, heads, tails." That
can't be a probability — it exceeds 1. Independent events combine by **multiplying**:
$0.6 \times 0.6 \times 0.4 = 0.144$. (Addition only enters *after* the log, where the
log rule legitimately turns the product $0.144$ into the sum
$\log 0.6 + \log 0.6 + \log 0.4 \approx -1.938$ — same quantity, different scale.)

## Check yourself

**1.** A coin is flipped 3 times and comes up **tails, tails, heads** (so $k = 1$ head,
$n = 3$). Write $L(p)$, then compute $L(0.3)$ by hand and check in R. Where will the
peak be?

```{webr}
#| autorun: false
# check1-starter — edit and Run to check your hand answer
lik <- function(p) p^1 * (1 - p)^2   # 1 head, 2 tails
```

::: {.callout-note collapse="true" title="Answer"}
One head gives a factor $p$; two tails give $(1-p)$ each, so
$L(p) = p\,(1-p)^2$. By hand, $L(0.3) = 0.3 \times 0.7^2 = 0.3 \times 0.49 = 0.147$. In
R, `lik(0.3)` returns `0.147`. The peak sits at the heads-rate $\hat{p} = k/n = 1/3
\approx 0.333$.
:::

**2.** True or false: taking the log of the likelihood can move the location of its peak
(the best $p$). Explain in one line.

::: {.callout-note collapse="true" title="Answer"}
**False.** $\log$ is an increasing function — bigger input, bigger output — so the $p$
that makes $L$ largest is the very same $p$ that makes $\log L$ largest. Logs change the
*scale* (product $\to$ sum, tiny $\to$ tame), never the *location* of the peak. That's
exactly why it's safe to maximize the log-likelihood instead of the likelihood.
:::

**3.** In your own words: explain to a colleague, in two sentences, what the likelihood
of a parameter measures and why we usually work with its log.

::: {.callout-note collapse="true" title="A model answer"}
The likelihood takes your already-observed data as fixed and scores each candidate
parameter value by how probable it would have made that data — the winner is the value
that leaves you least surprised by what you saw. We take the log because the likelihood
is a product of many per-observation pieces (which gets vanishingly small and awkward to
handle), and the log rule turns that product into a sum without moving the peak, so the
best parameter is unchanged but the arithmetic is tame.
:::

## What you can now do

- **Flip a probability question into a likelihood** — hold the data fixed, treat the
  parameter as the dial, and read $L(p)$ as "how probable was my data at this setting?"
- **Build a likelihood from independent observations** by multiplying the per-observation
  pieces ($p$ for a head, $1-p$ for a tail), and collapse it to $p^{k}(1-p)^{n-k}$.
- **Turn a likelihood into a log-likelihood** with the product-to-sum log rule, and say
  why that keeps the numbers tame without moving the peak.
- **State the maximum-likelihood idea** — the estimate $\hat{p}$ is the parameter value
  that maximizes $L$ (equivalently $\ell$), i.e. the top of the curve — and name it the
  **MLE**.

If any of these feels shaky, revisit [A tiny example first](#a-tiny-example-first) and
[The formal version](#the-formal-version) before continuing — every method that
*estimates* a parameter from data leans on all of these.

## Where this goes next

Maximum likelihood is the recipe most statistical methods use to turn data into
parameter estimates: write down the likelihood of the parameters given the data, take
its log, and climb to the top. *Finding* that top — solving "slope $= 0$" or stepping
uphill numerically — is the job of the optimization material (the
[derivative as slope](../derivative-as-slope/lesson.qmd) is the first piece of that
tool). For observations measured on a continuous scale, the per-observation "pieces"
become probability *densities* rather than probabilities — the subject of a separate
distributions-and-densities module — but the multiply-then-log recipe is identical.
Ready for more reps? Do the [practice problems](practice.qmd).
