---
title: "Practice: Likelihood and log-likelihood"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only (a {webr} starter does not see it — each starter re-states its
# own data). Base R is all we need, so nothing to install.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Do them on paper or in an R session; only
then open the answer.

## Problem 1 — the toy example, new numbers

You flip a coin **four times** and see **heads, heads, heads, tails** (so $k = 3$
heads, $n = 4$). Write the likelihood $L(p)$, compute $L(0.5)$ and $L(0.75)$ by hand,
and say which is bigger. Then check with R. Where do you expect the peak?

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
lik <- function(p) p^3 * (1 - p)^1   # 3 heads, 1 tail
```

::: {.callout-note collapse="true" title="Answer"}
Three heads give a factor $p$ each, one tail gives $(1-p)$, so
$L(p) = p^3(1-p)$.

- $L(0.5) = 0.5^3 \times 0.5 = 0.125 \times 0.5 = 0.0625$.
- $L(0.75) = 0.75^3 \times 0.25 = 0.421875 \times 0.25 \approx 0.1055$.

So $p = 0.75$ makes the data more probable than a fair coin does — and $0.75$ is exactly
the heads-rate $k/n = 3/4$, which is where the peak sits.

```{r p1-answer}
lik <- function(p) p^3 * (1 - p)^1
c(L_0.5 = lik(0.5), L_0.75 = lik(0.75))   # 0.0625 and ~0.1055
# confirm the peak is at k/n = 3/4:
g <- seq(0.001, 0.999, length.out = 999)
g[which.max(lik(g))]                        # ~0.75
```
:::

## Problem 2 — fill in the missing step (product → log-sum)

Here is the log-likelihood for the P1 data ($k = 3$, $n = 4$) worked out — but the
**core step is blanked**. Fill in ONLY the blanked line: turn the product into a sum of
logs using the two log rules.

> $L(p) = p^3(1-p)^1$. Take the log of both sides:
> $\ell(p) = \log\!\big(p^3(1-p)\big)$.
> **`<your step here>`** — apply "log of a product is a sum of logs," then "log of a
> power pulls the exponent out front," to get
> $\ell(p) = \underline{\phantom{000000}} \cdot \log p + \underline{\phantom{000000}}
> \cdot \log(1-p)$.

```{webr}
#| autorun: false
# p2-starter — edit and Run to check your log-likelihood against the product's log
p <- 0.75
```

::: {.callout-note collapse="true" title="Answer"}
The blanked step is the heart of the log-likelihood:

$$
\begin{aligned}
\ell(p) = \log\!\big(p^3(1-p)\big) &= \log(p^3) + \log(1-p) \\
                                   &= 3\log p + 1\cdot\log(1-p)
\end{aligned}
$$

- Line 1: **log of a product is a sum of logs**, $\log(ab) = \log a + \log b$.
- Line 2: **log of a power pulls the exponent out front**, $\log(a^b) = b\log a$
  (here the $(1-p)$ term has exponent 1, so it's just $\log(1-p)$).

So $\ell(p) = 3\log p + \log(1-p)$. Check at $p = 0.75$: it must equal $\log L(0.75)$.

```{r p2-answer}
p <- 0.75
via_sum     <- 3 * log(p) + log(1 - p)   # the log-likelihood, as a sum
via_product <- log(p^3 * (1 - p))        # log of the raw likelihood
c(via_sum = via_sum, via_product = via_product)   # both ~ -2.2493 -- identical
```
:::

## Problem 3 — a bigger sample, by log-likelihood

You flip a coin **10 times** and see **7 heads, 3 tails** ($k = 7$, $n = 10$). The raw
likelihood is a tiny number, so work on the **log** scale. Write $\ell(p)$, then find
the $p$ that maximizes it on a grid. **PREDICT** first: where should the peak be? Write
it down, then run.

```{webr}
#| autorun: false
# p3-starter — edit and Run to solve, right here
k <- 7; n <- 10
loglik <- function(p) k * log(p) + (n - k) * log(1 - p)
```

::: {.callout-note collapse="true" title="Answer"}
With $k = 7$ heads and $n - k = 3$ tails, $\ell(p) = 7\log p + 3\log(1-p)$. The peak
sits at the heads-rate $\hat{p} = k/n = 7/10 = 0.7$ — the maximum-likelihood estimate.

```{r p3-answer}
k <- 7; n <- 10
loglik <- function(p) k * log(p) + (n - k) * log(1 - p)

g <- seq(0.001, 0.999, length.out = 999)
g[which.max(loglik(g))]        # ~0.7 = k/n, the MLE

# The raw likelihood is minuscule -- that's WHY we used the log:
c(raw = 0.7^7 * 0.3^3, loglik = loglik(0.7))   # ~0.00222 vs a tame -6.109
```

The raw likelihood at the peak is only about $0.0022$; on the log scale it's a
comfortable $-6.11$. With hundreds of flips the raw number would underflow to 0, but the
log stays well-behaved.
:::

## Problem 4 — spot the error (multiply vs add)

A colleague computes the likelihood of the P1 data (**H, H, H, T**, so $k=3$, $n=4$) at
$p = 0.6$ and writes the following. The reasoning has **exactly one** core mistake —
find it, explain why it's wrong, and give the correct value.

> "Each head has probability $0.6$ and the tail has probability $0.4$. To get the
> likelihood of all four flips, I add them up: $0.6 + 0.6 + 0.6 + 0.4 = 2.2$. So
> $L(0.6) = 2.2$."

```{webr}
#| autorun: false
# p4-starter — copy this to check the correct value
p <- 0.6
```

::: {.callout-note collapse="true" title="Answer"}
The core error is **adding the per-flip probabilities instead of multiplying them**. The
flips are independent, so "all four happened" combines by *multiplication*, not
addition. The giveaway: $2.2$ is greater than 1, and a likelihood built from
probabilities of a single fixed outcome can never exceed 1 (it's a product of numbers
each in $[0,1]$).

The correct value multiplies: $L(0.6) = 0.6 \times 0.6 \times 0.6 \times 0.4 =
0.6^3 \times 0.4 = 0.0864$.

```{r p4-answer}
p <- 0.6
p^3 * (1 - p)      # 0.0864 -- the correct likelihood (a product, not a sum)
```
:::

## Problem 5 — mixing it with exponentials & logs *(uses [exponentials-and-logs](../exponentials-and-logs/lesson.qmd)'s product→sum rule)*

For the P1 data at $p = 0.6$ you have three "heads" pieces and one "tails" piece:
$0.6, 0.6, 0.6, 0.4$. Compute the log-likelihood two ways — (a) as **the log of the
product** of the four pieces, and (b) as **the sum of the four logs** — and confirm they
agree, exactly the product-to-sum rule from exponentials & logs.

```{webr}
#| autorun: false
# p5-starter — edit and Run to solve, right here
pieces <- c(0.6, 0.6, 0.6, 0.4)   # per-flip probabilities: 3 heads, 1 tail at p = 0.6
```

**PREDICT** first: will the two results be equal? And will they be positive or negative?
Write it down, then run.

::: {.callout-note collapse="true" title="Answer"}
The log-likelihood is the log of the product of the per-flip pieces. The log rule
$\log(ab\cdots) = \log a + \log b + \cdots$ lets you compute it either as one log of the
whole product, or as the sum of the individual logs — they are the same number:

- (a) $\log(0.6 \times 0.6 \times 0.6 \times 0.4) = \log(0.0864) \approx -2.4488$.
- (b) $\log 0.6 + \log 0.6 + \log 0.6 + \log 0.4 \approx -2.4488$.

Both are **negative**, because every piece is below 1 and the log of a number below 1 is
negative. This equality is *why* log-likelihood is a sum — it's the product-to-sum rule
from [exponentials & logs](../exponentials-and-logs/lesson.qmd) applied to the
likelihood.

```{r p5-answer}
pieces <- c(0.6, 0.6, 0.6, 0.4)
c(log_of_product = log(prod(pieces)),   # (a)
  sum_of_logs    = sum(log(pieces)))    # (b) -- both ~ -2.4488, identical
```
:::

## Problem 6 — a gentle cliffhanger (finding the peak)

You've been *reading off* the peak of the log-likelihood as $k/n$. But how would you
*find* it if you couldn't guess? For the coin, the log-likelihood is
$\ell(p) = k\log p + (n-k)\log(1-p)$. The top of a smooth hill is where its **slope is
zero**. You'll get the calculus tool for this in a later module — but you can *see* it
now. For the P3 data ($k = 7$, $n = 10$), the slope of $\ell$ works out to
$\frac{k}{p} - \frac{n-k}{1-p}$. **Predict:** at which $p$ does that slope hit zero — and
does it match the peak $k/n$ you found before? Then run to check.

```{webr}
#| autorun: false
# p6-starter — the slope of the log-likelihood, evaluated at several p values
k <- 7; n <- 10
slope <- function(p) k / p - (n - k) / (1 - p)
```

::: {.callout-note collapse="true" title="Answer"}
Setting the slope to zero, $\frac{k}{p} = \frac{n-k}{1-p}$, cross-multiplying gives
$k(1-p) = (n-k)p$, i.e. $k = np$, so $p = k/n$. For P3 that's $7/10 = 0.7$ — the exact
same peak you found on the grid. The slope is **positive** below $0.7$ (the curve is
still climbing) and **negative** above it (past the top, heading down).

```{r p6-answer}
k <- 7; n <- 10
slope <- function(p) k / p - (n - k) / (1 - p)
round(c(at_0.5 = slope(0.5), at_0.7 = slope(0.7), at_0.9 = slope(0.9)), 4)
# +8 at 0.5 (climbing), 0 at 0.7 (the top), -22.22 at 0.9 (past the peak, descending)
```

"Set the slope to zero and solve" is the calculus route to the MLE — the
[derivative as slope](../derivative-as-slope/lesson.qmd) module gives you that tool, and
the optimization module shows how to *climb* to the top when no clean formula exists.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/likelihood-and-log-likelihood" pass="4"}

::: {.quiz-q qid="lll-log-likelihood-value" qtype="numeric" answer="-2.079" tolerance="0.001"}
Three independent flips at $p = 0.5$ give a likelihood of $0.125$. Its log-likelihood is
$\ln(0.125)$. Compute it, rounded to three decimals.
:::

::: {.quiz-q qid="lll-combine-multiply" qtype="mc" answer="b"}
For $p = 0.6$, how do you combine three independent flips (heads, heads, tails) into one
likelihood?

a) Add them: $0.6 + 0.6 + 0.4 = 1.6$
b) Multiply them: $0.6 \times 0.6 \times 0.4 = 0.144$
c) Average them: $(0.6 + 0.6 + 0.4)/3$
:::

::: {.quiz-q qid="lll-not-prob-of-p" qtype="mc" answer="a"}
Why is the likelihood $L(p)$ *not* a probability distribution over $p$?

a) $L(p)$ gives the probability of the *data* as $p$ varies, and these values do not sum to $1$ across $p$
b) Because likelihood values are always negative
c) Because $p$ is known rather than random
:::

::: {.quiz-q qid="lll-mle-idea" qtype="mc" answer="c"}
What is the maximum-likelihood estimate $\hat{p}$?

a) The average of all the likelihood values
b) The probability that the parameter is correct
c) The parameter value that maximizes the likelihood — the top of the curve
:::

::: {.quiz-q qid="lll-log-preserves-peak" qtype="mc" answer="a"}
(From exponentials & logs.) Taking the log turns the likelihood's product into a sum.
Does it move where the peak sits?

a) No — the log is increasing, so it preserves the peak's location while taming the numbers
b) Yes — the log shifts the maximizing $p$
c) Yes — the peak moves to $p = 0.5$
:::

:::

::::
