---
title: "Least squares and the normal equations"
subtitle: "Where the coefficients come from — and how to rebuild lm() in one line"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

## Where we're going

Module 1 told you what the coefficients *mean*. This module shows where they *come
from*. In the earlier course, one predictor gave the tidy formula
$\hat{\beta}_1 = s_{xy}/s_x^2$ (the x–y covariance $s_{xy}$ over $x$'s variance $s_x^2$).
With many predictors that single formula falls apart —
the predictors interfere, and you need to solve for all the coefficients *together*.
The tool is the **normal equations**, written compactly as $X^\top X\,\hat\beta = X^\top y$,
and by the end you'll reproduce `lm()`'s entire coefficient vector with one line of base
R: `solve(t(X) %*% X, t(X) %*% y)`.

**Builds on:** [Module 1](../01-many-knobs-one-model/lesson.qmd) (the model and its
coefficients),
[partial-derivatives-and-the-gradient](../../../../foundations/partial-derivatives-and-the-gradient/lesson.qmd)
(set every partial slope to zero at the bottom),
[matrices-and-linear-transforms](../../../../foundations/matrices-and-linear-transforms/lesson.qmd)
(transpose $X^\top$, the product $X^\top X$, and `solve()` as matrix inverse).

## Warm-up: do you remember?

**1.** *(Module 1)* In $\hat{y}_i = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2}$, what is
the **residual** for observation $i$ (observed minus predicted)?

::: {.callout-note collapse="true" title="Answer"}
$e_i = y_i - \hat{y}_i = y_i - \beta_0 - \beta_1 x_{i1} - \beta_2 x_{i2}$. The residual
is the leftover the fitted model misses. See
[Module 1](../01-many-knobs-one-model/lesson.qmd).
:::

**2.** *(partial derivatives)* At the bottom of a valley $f(b_1, b_2)$ with two knobs,
what is true of **both** partial derivatives $\partial f/\partial b_1$ and
$\partial f/\partial b_2$?

::: {.callout-note collapse="true" title="Answer"}
Both are **zero** — the ground is level in every direction at the bottom. Finding the
minimum means setting each partial slope to 0 and solving the resulting equations. See
[partial-derivatives-and-the-gradient](../../../../foundations/partial-derivatives-and-the-gradient/lesson.qmd).
:::

**3.** *(matrices)* For $X = \begin{pmatrix} 1 & 1 \\ 2 & 0 \\ 3 & 2 \end{pmatrix}$
(columns $x_1 = (1,2,3)$, $x_2 = (1,0,2)$), what is the top-left entry of $X^\top X$?

::: {.callout-note collapse="true" title="Answer"}
The top-left entry is $x_1 \cdot x_1 = 1^2 + 2^2 + 3^2 = 14$. Each entry of $X^\top X$
is a dot product of two predictor columns (in R: `t(X) %*% X`). See
[matrices-and-linear-transforms](../../../../foundations/matrices-and-linear-transforms/lesson.qmd).
:::

## A tiny example first

Take two predictors and three observations, and — to keep the arithmetic clean — pick
numbers that sit *exactly* on a plane through the origin (no intercept, no noise):

$$
x_1 = (1, 2, 3), \qquad x_2 = (1, 0, 2), \qquad y = (5, 4, 12).
$$

We'll look for the two coefficients $b_1, b_2$ that best predict $y$ from
$b_1 x_1 + b_2 x_2$. "Best" means the smallest sum of squared residuals, just like the
earlier course — only now there are two knobs to turn at once.

Here is the trick that finds them, worked in plain arithmetic. The best coefficients
make the residual "use up" all the information each predictor column carries about
$y$. Concretely, they satisfy these two equations (we derive them below):

$$
\begin{aligned}
(\textstyle\sum x_{i1}^2)\,b_1 + (\textstyle\sum x_{i1}x_{i2})\,b_2 &= \textstyle\sum x_{i1} y_i \\
(\textstyle\sum x_{i1}x_{i2})\,b_1 + (\textstyle\sum x_{i2}^2)\,b_2 &= \textstyle\sum x_{i2} y_i .
\end{aligned}
$$

Plug in the sums from our data (each is a dot product):

- $\sum x_{i1}^2 = 1 + 4 + 9 = 14$
- $\sum x_{i1} x_{i2} = 1\cdot1 + 2\cdot0 + 3\cdot2 = 7$
- $\sum x_{i2}^2 = 1 + 0 + 4 = 5$
- $\sum x_{i1} y_i = 1\cdot5 + 2\cdot4 + 3\cdot12 = 49$
- $\sum x_{i2} y_i = 1\cdot5 + 0\cdot4 + 2\cdot12 = 29$

So the two equations are:

$$
14\,b_1 + 7\,b_2 = 49, \qquad 7\,b_1 + 5\,b_2 = 29 .
$$

Solve them like any two-equations-two-unknowns pair. Multiply the first by 5 and the
second by 7 to line up $b_2$... or just use the shortcut for a 2×2 system: the
determinant is $14\cdot5 - 7\cdot7 = 70 - 49 = 21$, and

$$
b_1 = \frac{5\cdot49 - 7\cdot29}{21} = \frac{245 - 203}{21} = \frac{42}{21} = 2,
\qquad
b_2 = \frac{14\cdot29 - 7\cdot49}{21} = \frac{406 - 343}{21} = \frac{63}{21} = 3 .
$$

So $b_1 = 2$, $b_2 = 3$. The same thing in R:

```{r toy-example}
x1 <- c(1, 2, 3); x2 <- c(1, 0, 2); y <- c(5, 4, 12)
X <- cbind(x1, x2)                 # design matrix: two predictor columns, no intercept

t(X) %*% X                          # the 2x2 grid of column dot products: [[14,7],[7,5]]
t(X) %*% y                          # the right-hand side: (49, 29)
solve(t(X) %*% X, t(X) %*% y)       # solve the pair: b1 = 2, b2 = 3
```

(These three points were rigged to lie exactly on the plane $2x_1 + 3x_2$, so the fit
is perfect and the residuals are all zero. Real data won't be exact — but the solving
step is *identical*.)

## Getting a feel for it

**Casting a shadow onto a plane.** Picture the response $y$ as a single point floating
in space, and the predictor columns as directions spanning a flat sheet (a plane). The
fitted values $\hat{y}$ are forced to live *on* that sheet — they can only be
combinations of the predictor columns. The best fit is the point on the sheet
**closest** to $y$, and the closest point is the *shadow* $y$ casts straight down onto
it. The line from $y$ to its shadow — the residual — is **perpendicular** to the sheet.

**Why "perpendicular" gives exactly our two equations.** Perpendicular to the sheet
means perpendicular to every direction *in* it — to every predictor column. Two vectors
are perpendicular when their dot product is zero. So "residual perpendicular to column
$x_1$" is $\sum x_{i1}\,e_i = 0$, and "residual perpendicular to column $x_2$" is
$\sum x_{i2}\,e_i = 0$. Those two "zero dot product" statements, written out, are
exactly the two equations above. This is why they're called the **normal** equations —
"normal" is the geometry word for "perpendicular."

**What the numbers feel like.** $X^\top X$ is just the grid of dot products *among the
predictor columns* — how much each column overlaps with itself and with the others; big
off-diagonal entries mean the predictors point in similar directions (they're
correlated — the seed of Module 3's trouble). $X^\top y$ is how much each column
overlaps with the response. The normal equations balance the two: they pick the
coefficients so the predictor columns, combined, line up with $y$ as well as a flat
sheet can.

## What the notation is going to say

The formula will say: *stack the predictors into a matrix $X$; the best coefficients
make the residual perpendicular to every column of $X$; written compactly that is
"$X^\top X$ times the coefficients equals $X^\top y$."* Solving it is exactly the two
lines of R you just ran, and $X^\top X$ is exactly the little dot-product grid you
computed as $\left(\begin{smallmatrix}14&7\\7&5\end{smallmatrix}\right)$.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $y_i$ | "why sub eye" | the outcome for observation $i$ (in R: `y[i]`) | 5, 4, 12 |
| $x_{ij}$ | "ex sub eye jay" | the $j$-th predictor value for observation $i$ | $x_{1,2} = 0$ (row 2, column $x_2$) |
| $n$ | "en" | number of observations (in R: `nrow`) | 3 |
| $\sum_{i=1}^{n}$ | "the sum from eye equals 1 to en" | add the thing to its right once per observation | $\sum x_{i1}^2 = 1+4+9$ |
| $\mathbf{X}$ | "the matrix ex" | design matrix: rows = observations, columns = predictors (in R: `cbind(...)`) | the $x_1, x_2$ columns |
| $\mathbf{y}$ | "the vector why" | the response column | $(5, 4, 12)$ |
| $\boldsymbol{\beta}$, $b$ | "beta" / "bee" | the coefficient vector we solve for | $(2, 3)$ |
| $\hat{\beta}$ | "beta hat" | the *estimated* coefficients (the solution) | $(2, 3)$ |
| $X^\top$ | "ex transpose" | $X$ flipped so rows become columns (in R: `t(X)`) | a 2×3 matrix |
| $X^\top X$ | "ex transpose ex" | the grid of dot products among the columns of $X$ (the "Gram matrix") | $\left(\begin{smallmatrix}14&7\\7&5\end{smallmatrix}\right)$ |
| $X^\top y$ | "ex transpose why" | each column's dot product with $y$ | $(49, 29)$ |
| $A^{-1}$ | "A inverse" | the matrix that undoes $A$ (in R: `solve(A)`) | — |
| $e_i$ | "ee sub eye" | residual $y_i - \hat{y}_i$ | 0 (rigged toy) |
| $\mathrm{RSS}$ | "residual sum of squares" | $\sum_i e_i^2$, the total squared miss (in R: `sum(resid^2)`) | 0 (rigged toy) |
| $\nabla$ | "del" / "the gradient" | the vector of all partial derivatives at once | see Optional box |

## The formal version

The thing we minimize is the **residual sum of squares** — the same badness score as
the earlier course, now over all the knobs at once. For the two-predictor case:

$$
\mathrm{RSS}(b_1, b_2) = \sum_{i=1}^{n}\big(y_i - b_1 x_{i1} - b_2 x_{i2}\big)^2 .
$$

Find the bottom by setting each partial derivative to zero.

### Step 1 — partial derivative with respect to $b_1$

$$
\begin{aligned}
\frac{\partial\,\mathrm{RSS}}{\partial b_1}
&= \sum_{i=1}^{n} 2\big(y_i - b_1 x_{i1} - b_2 x_{i2}\big)\cdot(-x_{i1}) \\
&= -2\sum_{i=1}^{n} x_{i1}\big(y_i - b_1 x_{i1} - b_2 x_{i2}\big) .
\end{aligned}
$$

- Line 1: chain rule, lite — derivative of (stuff)² is $2\cdot$(stuff)$\cdot$(derivative
  of the stuff); treating $b_2$ as constant, the inside's derivative with respect to
  $b_1$ is $-x_{i1}$. The derivative of a sum is the sum of the derivatives.
- Line 2: pull the constant $-2$ out of the sum (a constant multiple factors out — the
  sum rule from [vectors-and-summation](../../../../foundations/vectors-and-summation/lesson.qmd)).

Set it to zero and divide by $-2$, then split the sum:

$$
\begin{aligned}
\sum_i x_{i1}\big(y_i - b_1 x_{i1} - b_2 x_{i2}\big) &= 0 \\
\Rightarrow\quad \sum_i x_{i1} y_i &= b_1 \sum_i x_{i1}^2 + b_2 \sum_i x_{i1}x_{i2} .
\end{aligned}
$$

- Line 2: multiply through and split into three sums; pull each constant coefficient
  ($b_1$, $b_2$) out of its sum, then move the two knob-terms to the right.

That is the **first normal equation** — the top row of the pair we used above.

### Step 2 — partial derivative with respect to $b_2$

The identical move, now differentiating the inside by $b_2$ (its derivative is
$-x_{i2}$):

$$
\sum_i x_{i2} y_i = b_1 \sum_i x_{i1}x_{i2} + b_2 \sum_i x_{i2}^2 .
$$

That is the **second normal equation** — the bottom row.

### Packaging the pair as one matrix equation

Line the two equations up. On the left, the coefficient grid multiplying $(b_1, b_2)$
is exactly the dot-product grid $X^\top X$; on the right is $X^\top y$:

$$
\underbrace{\begin{pmatrix} \sum x_{i1}^2 & \sum x_{i1}x_{i2} \\ \sum x_{i1}x_{i2} & \sum x_{i2}^2 \end{pmatrix}}_{X^\top X}
\underbrace{\begin{pmatrix} b_1 \\ b_2 \end{pmatrix}}_{\hat\beta}
=
\underbrace{\begin{pmatrix} \sum x_{i1} y_i \\ \sum x_{i2} y_i \end{pmatrix}}_{X^\top y} .
$$

Nothing about this is special to two predictors: with $p$ predictors (and an intercept
column of 1s folded into $X$) you get $p{+}1$ normal equations, and they package into
the very same shape:

$$
\boxed{\,X^\top X\,\hat\beta = X^\top y\,} \qquad\Longrightarrow\qquad \hat\beta = (X^\top X)^{-1} X^\top y .
$$

Reading it aloud: "ex-transpose-ex times beta-hat equals ex-transpose-why" — or in
everyday words: "the column dot-product grid, times the coefficients, equals the
columns' overlap with $y$; solve for the coefficients." In R you never form the inverse
by hand — `solve(A, b)` solves $A\hat\beta = b$ directly and more stably:

$$
\texttt{beta\_hat <- solve(t(X) \%*\% X, t(X) \%*\% y)} .
$$

**Sanity check (toy numbers).** $X^\top X = \left(\begin{smallmatrix}14&7\\7&5\end{smallmatrix}\right)$,
$X^\top y = (49, 29)$, and solving gave $(b_1, b_2) = (2, 3)$ — matching the hand
arithmetic and `lm()` (below).

::: {.callout-note title="Optional: the whole derivation in one vector-calculus line"}
If you're comfortable with the gradient $\nabla$ (all partial derivatives stacked into
a vector), the two steps above collapse to one. Writing $\mathrm{RSS} = \lVert y - Xb \rVert^2$,
its gradient is $\nabla_b\,\mathrm{RSS} = -2X^\top(y - Xb)$; setting it to the zero
vector gives $X^\top(y - X\hat\beta) = 0$, i.e. $X^\top X\hat\beta = X^\top y$. This
buys brevity, not new truth — it's the same "residual perpendicular to every column"
statement. Skip it freely; the scalar derivation above is the whole story.
:::

## Explore it

```{webr}
#| autorun: true
# TRY IT: change y (say, make the last value 13 instead of 12) and re-run. The fit is
# no longer perfect, but solve() still returns the least-squares coefficients.
# PREDICT: will solve()'s answer still match lm()'s exactly? Write your guess, then run.
x1 <- c(1, 2, 3); x2 <- c(1, 0, 2); y <- c(5, 4, 12)
X  <- cbind(x1, x2)

beta_hat <- solve(t(X) %*% X, t(X) %*% y)   # normal equations, from scratch
lm_coef  <- coef(lm(y ~ x1 + x2 - 1))        # lm(), no intercept (-1), as the referee
cbind(from_scratch = as.numeric(beta_hat), lm = as.numeric(lm_coef))
```

### Rebuild `lm()` on the real panel, and check it against the frozen proof

Here is the payoff. The `equivalence/` harness in this repo froze `lm()`'s answer on a
simulated 30-row indicator panel (validation **tier: full** — the reference is R's own
`stats::lm`, which also runs right here in your browser). The cell below regenerates
that exact panel, fits it two ways, and checks both against the frozen numbers:

```{webr}
#| autorun: true
# The SAME generator the equivalence fixture used (base R; reproduced by seed).
sim_indicators <- function(n = 30, seed = 1, rho_max = 0.95) {
  set.seed(seed)
  nm <- c("rate_10y","rate_2y","cpi_infl","pce_infl","unemp","oil","usd_index",
          "sentiment","housing","pmi","credit_spread","vix")
  p <- length(nm); R <- diag(p); dimnames(R) <- list(nm, nm)
  R["rate_10y","rate_2y"]  <- R["rate_2y","rate_10y"]  <- 0.90
  R["cpi_infl","pce_infl"] <- R["pce_infl","cpi_infl"] <- rho_max
  R["sentiment","pmi"]     <- R["pmi","sentiment"]     <- 0.70
  R["credit_spread","vix"] <- R["vix","credit_spread"] <- 0.65
  X <- matrix(rnorm(n * p), n, p) %*% chol(R); colnames(X) <- nm
  beta <- setNames(numeric(p), nm); beta[c("rate_10y","cpi_infl","unemp")] <- c(0.5,-0.4,-0.3)
  list(X = X, y = drop(X %*% beta + rnorm(n, sd = 1)))
}
sim <- sim_indicators()
Xd  <- cbind(`(Intercept)` = 1, sim$X)          # design matrix WITH the intercept column
y   <- sim$y

beta_scratch <- as.numeric(solve(t(Xd) %*% Xd, t(Xd) %*% y))   # from scratch
beta_lm      <- as.numeric(coef(lm(y ~ sim$X)))                # lm(), live

# The frozen reference from equivalence/fixtures/multiple-regression-full-lm.json:
beta_frozen <- c(-0.12945157, 0.38793328, 0.11812846, -1.69810596, 0.73806047,
                 -0.08225275, 0.05353800, 0.04667165, -0.07781305, 0.22603827,
                  0.10805845, -0.32544980, 0.63931982)

cat("scratch vs lm  max diff:", max(abs(beta_scratch - beta_lm)),    "\n")
cat("scratch vs frozen ref  :", max(abs(beta_scratch - beta_frozen)),"\n")
# Both differences are ~1e-8 or smaller: your one-line rebuild IS what lm() computes.
```

## Common traps

**Trap: "`solve()` always returns the coefficients."** `solve(t(X) %*% X, ...)` needs
$X^\top X$ to be invertible, which fails when a predictor column is a copy of another
(or an exact linear combination of others). Then there is *no unique* answer — infinitely
many coefficient vectors fit equally well — and R throws a "system is computationally
singular" error. Try it: give the toy a third column equal to `x1 + x2` and `solve()`
refuses.

```{webr}
#| autorun: false
# p-trap-starter: add a redundant column and watch solve() fail.
# PREDICT: what happens when x3 is exactly x1 + x2?
x1 <- c(1, 2, 3); x2 <- c(1, 0, 2); y <- c(5, 4, 12)
X  <- cbind(x1, x2, x3 = x1 + x2)     # x3 carries no new information
solve(t(X) %*% X, t(X) %*% y)          # error: X'X is singular
```

**Trap: "the normal equations are the *only* way, and always the best."** They're the
cleanest to *understand*, but forming $X^\top X$ squares how spread-out the predictors'
scales are (technically, it squares the condition number), which can lose precision when
predictors are nearly collinear. That's why `lm()` actually fits with a QR
decomposition instead — same answer, more numerically stable. On well-behaved data (our
first panel) the two agree to ~15 digits; on a near-collinear panel they still agree, but
to fewer digits (the `equivalence/` harness records exactly this by widening the
tolerance for its collinear fixture). Knowing the normal equations is knowing *what*
`lm()` computes; QR is a *how*.

## Check yourself

**1.** For predictors $x_1 = (2, 0, 1)$ and $x_2 = (1, 1, 0)$, what are the two
diagonal entries of $X^\top X$ (i.e. $\sum x_{i1}^2$ and $\sum x_{i2}^2$)?

```{webr}
#| autorun: false
# check1-starter
x1 <- c(2, 0, 1); x2 <- c(1, 1, 0)
```

::: {.callout-note collapse="true" title="Answer"}
$\sum x_{i1}^2 = 4 + 0 + 1 = 5$ and $\sum x_{i2}^2 = 1 + 1 + 0 = 2$. In R,
`diag(t(cbind(x1,x2)) %*% cbind(x1,x2))` gives `5 2`.
:::

**2.** The normal equations say the residual is perpendicular to every predictor
column. What single number does that make $\sum_i x_{i1}\, e_i$ equal to at the best
fit?

::: {.callout-note collapse="true" title="Answer"}
Zero. "Perpendicular to column $x_1$" *is* "the dot product of $x_1$ with the residuals
is 0," i.e. $\sum_i x_{i1} e_i = 0$. One such equation per column — that's the whole set
of normal equations.
:::

**3.** In your own words: explain to a colleague, in two sentences, what
$X^\top X\,\hat\beta = X^\top y$ says and why solving it gives the best-fitting
coefficients.

::: {.callout-note collapse="true" title="A model answer"}
It says: pick the coefficients that make the leftover residual perpendicular to every
predictor column — equivalently, make the predictors' combined pull line up with $y$ as
closely as a flat combination of them can. Solving it lands $\hat{y}$ on the point of the
predictor-plane closest to $y$, which is exactly the least-squares (smallest total
squared miss) fit.
:::

## What you can now do

- **Derive the normal equations** for two predictors by setting each partial derivative
  of RSS to zero, naming the trick behind each step.
- **Read $X^\top X\,\hat\beta = X^\top y$** as "residual perpendicular to every column,"
  and recognize $X^\top X$ as the grid of column dot products.
- **Fit a multiple regression from scratch** with `solve(t(X) %*% X, t(X) %*% y)` and
  confirm it reproduces `lm()` — the validation this course's `equivalence/` harness
  freezes.
- **Explain when the solve fails** (a redundant/collinear column makes $X^\top X$
  singular) and why `lm()` prefers QR for stability.

If any of these feels shaky, revisit [The formal version](#the-formal-version) before
continuing — Module 3 takes the collinearity crack we just saw and shows the damage it
does to the coefficients themselves.

## Where this goes next

You can now compute the coefficients. Next: [Marginal vs joint](../03-marginal-vs-joint/lesson.qmd)
asks what they *say* when predictors travel together — and shows a predictor that looks
important on its own vanishing (or flipping sign) in the joint fit. That "singular when
collinear" crack becomes the main event. Ready for more reps first? Do the
[practice problems](practice.qmd).
