---
title: "Practice: Least squares and the normal equations"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
set.seed(42)
```

Work these in order — they get harder. Do them on paper or in an R session; only then
open the answer.

## Problem 1 — solve the normal equations by hand, new numbers

Two predictors, three observations (no intercept), rigged to fit exactly:
$x_1 = (1, 1, 2)$, $x_2 = (0, 1, 1)$, $y = (3, 5, 8)$. Build $X^\top X$ and $X^\top y$
from the column dot products, then solve $X^\top X\,\hat\beta = X^\top y$ for
$(b_1, b_2)$.

```{webr}
#| autorun: false
# p1-starter — edit and Run to check your hand solve
x1 <- c(1, 1, 2); x2 <- c(0, 1, 1); y <- c(3, 5, 8)
X <- cbind(x1, x2)
```

::: {.callout-note collapse="true" title="Answer"}
Dot products: $\sum x_1^2 = 1+1+4 = 6$, $\sum x_1 x_2 = 0+1+2 = 3$, $\sum x_2^2 = 0+1+1 = 2$,
$\sum x_1 y = 3+5+16 = 24$, $\sum x_2 y = 0+5+8 = 13$. So
$X^\top X = \left(\begin{smallmatrix}6&3\\3&2\end{smallmatrix}\right)$, $X^\top y = (24, 13)$,
determinant $6\cdot2 - 3\cdot3 = 3$. Then $b_1 = (2\cdot24 - 3\cdot13)/3 = 9/3 = 3$ and
$b_2 = (6\cdot13 - 3\cdot24)/3 = 6/3 = 2$.

```{r}
x1 <- c(1, 1, 2); x2 <- c(0, 1, 1); y <- c(3, 5, 8); X <- cbind(x1, x2)
solve(t(X) %*% X, t(X) %*% y)      # 3, 2
```
:::

## Problem 2 — fill in the missing step

Here is the start of the derivation of the first normal equation. Supply the blanked
step.

- Step 1: $\dfrac{\partial}{\partial b_1}\sum_i (y_i - b_1 x_{i1} - b_2 x_{i2})^2
  = \sum_i 2(y_i - b_1 x_{i1} - b_2 x_{i2})(-x_{i1})$.
- Step 2: **your step here** — set the derivative to zero and simplify by dividing out
  the constant.
- Step 3: $\sum_i x_{i1} y_i = b_1 \sum_i x_{i1}^2 + b_2 \sum_i x_{i1} x_{i2}$.

::: {.callout-note collapse="true" title="Answer"}
Step 2: setting the derivative to 0 and dividing both sides by $-2$ gives
$\sum_i x_{i1}(y_i - b_1 x_{i1} - b_2 x_{i2}) = 0$. Multiplying out and splitting the sum
(then moving the knob-terms to the right) yields Step 3. The move is: constant multiple
$-2$ factors out of the sum, and a sum of three terms splits into three sums.
:::

## Problem 3 — from scratch vs `lm()`, in R

For $x_1 = (1, 2, 3, 4)$, $x_2 = (2, 1, 0, 1)$, $y = (5, 4, 6, 9)$ (with an intercept),
fit the model two ways: the normal equations with an explicit intercept column, and
`lm(y ~ x1 + x2)`. Confirm they match.

```{webr}
#| autorun: false
# p3-starter — build a design matrix WITH an intercept column of 1s, then solve
# PREDICT: will your solve() coefficients match lm()'s to full precision?
x1 <- c(1, 2, 3, 4); x2 <- c(2, 1, 0, 1); y <- c(5, 4, 6, 9)
X <- cbind(1, x1, x2)     # the leading 1s column is the intercept
```

::: {.callout-note collapse="true" title="Answer"}
Include the intercept as a column of 1s in $X$, then `solve(t(X) %*% X, t(X) %*% y)`
gives the three coefficients (intercept, $x_1$, $x_2$), matching `coef(lm(y ~ x1 + x2))`
to machine precision.

```{r}
x1 <- c(1, 2, 3, 4); x2 <- c(2, 1, 0, 1); y <- c(5, 4, 6, 9)
X <- cbind(1, x1, x2)
rbind(from_scratch = as.numeric(solve(t(X) %*% X, t(X) %*% y)),
      lm           = as.numeric(coef(lm(y ~ x1 + x2))))
```
(The two rows are identical.)
:::

## Problem 4 — spot the error

A colleague wants the coefficients for `y ~ x1 + x2` *with an intercept*, and writes:

```r
X <- cbind(x1, x2)                 # two predictor columns
beta <- solve(t(X) %*% X, t(X) %*% y)
```

They're surprised the intercept is missing and the slopes don't match `lm()`. This code
contains **exactly one mistake** — find and fix it.

::: {.callout-note collapse="true" title="Answer"}
The mistake: no **intercept column**. `lm(y ~ x1 + x2)` fits an intercept by default, but
this `X` has only the two predictor columns, so `solve()` fits a no-intercept model — a
different model, hence different slopes. Fix: `X <- cbind(1, x1, x2)` so the design
matrix has a leading column of 1s. Then the first coefficient is the intercept and the
slopes match `lm()`.
:::

## Problem 5 — mixing it with Module 1 *(uses Module 1's "holding others fixed")*

You solve the normal equations and get $\hat\beta = (2, 3, -1)$ for the model
$\hat{y} = \beta_0 + \beta_1 x_1 + \beta_2 x_2$. Two observations have the same $x_2$,
but the second has $x_1$ larger by 2. Using the coefficient's Module-1 meaning, how much
larger is the second observation's fitted value?

```{webr}
#| autorun: false
# p5-starter — you can reason it out, or check with the numbers
beta <- c(b0 = 2, b1 = 3, b2 = -1)
```

::: {.callout-note collapse="true" title="Answer"}
Only $x_1$ changes (by 2), $x_2$ held fixed, so the fitted value changes by
$\beta_1 \times 2 = 3 \times 2 = 6$ — larger by 6. The intercept and the $-1\cdot x_2$
term are identical for both observations and cancel. (Solving the normal equations gave
$\beta_1$; Module 1 told you how to read it.)
:::

## Problem 6 — transfer: what happens when a column is (nearly) redundant

Take $x_1 = (1, 2, 3, 4)$ and set $x_2 = 2 x_1$ exactly (so $x_2 = (2, 4, 6, 8)$). Try to
fit `y ~ x1 + x2` from scratch with the normal equations. Predict what `solve()` does,
then say in one sentence what near-collinearity (say $x_2 \approx 2x_1$ plus a whisker of
noise) would do instead. *(Module 3 is about this whisker.)*

```{webr}
#| autorun: false
# p6-starter — PREDICT before running: does solve() return coefficients, or error?
x1 <- c(1, 2, 3, 4); x2 <- 2 * x1; y <- c(3, 5, 8, 9)
X <- cbind(1, x1, x2)
```

::: {.callout-note collapse="true" title="Answer"}
With $x_2$ an exact multiple of $x_1$, the column carries no new information, $X^\top X$
is singular, and `solve()` errors ("system is computationally singular") — there is no
unique coefficient vector. With a whisker of noise ($x_2 \approx 2x_1$), $X^\top X$
becomes *barely* invertible: `solve()` returns coefficients, but huge and unstable ones
that swing wildly with tiny data changes. That instability — real, near-collinear
predictors — is exactly what Module 3 dissects.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="courses/multiple-regression/modules/02-least-squares-and-the-normal-equations" pass="4"}

::: {.quiz-q qid="lsn-normal-equations" qtype="mc" answer="b"}
The normal equations that `lm()` solves for $\hat\beta$ are:

a) $X\hat\beta = y$
b) $X^\top X\,\hat\beta = X^\top y$
c) $\hat\beta = X^\top y$
:::

::: {.quiz-q qid="lsn-solve-fails" qtype="mc" answer="c"}
`solve(t(X) %*% X, t(X) %*% y)` errors with "system is singular." What causes it?

a) The sample size is too small
b) The response $y$ has negative values
c) A predictor column is an exact linear combination of others, so $X^\top X$ is not invertible
:::

::: {.quiz-q qid="lsn-why-qr" qtype="mc" answer="a"}
Why does `lm()` fit with a QR decomposition rather than forming $X^\top X$ directly?

a) Forming $X^\top X$ squares the predictors' scale spread, losing precision on near-collinear data; QR is more stable for the same answer
b) QR gives different, better coefficients
c) QR is the only method that handles more than two predictors
:::

::: {.quiz-q qid="lsn-xtx-dotproduct" qtype="numeric" answer="7" tolerance="0"}
$X^\top X$ is the grid of dot products between the predictor columns. For columns
$x_1 = (1, 2, 3)$ and $x_2 = (1, 0, 2)$, what is their dot product (the $x_1, x_2$ entry)?
:::

::: {.quiz-q qid="lsn-coef-interpret" qtype="mc" answer="b"}
(From module 1.) The solved coefficient is $\hat\beta_1 = 2$. How do you interpret it?

a) $x_1$ is twice as important as the other predictors
b) $y$ changes by $2$ per unit of $x_1$, holding the other predictors fixed
c) The prediction is always $2$
:::

:::

::::
