---
title: "Practice: Hypothesis tests, p-values & confidence intervals"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only (a {webr} starter does not see it — each starter re-states its
# own data). Base R is all we need. The seed matters: some answers SIMULATE a p-value.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Do them on paper or in an R session; only
then open the answer.

## Problem 1 — the toy example, new numbers

A new batch of 4 plants; extra growth (treated minus untreated), in cm:

$$
d = (5,\; 2,\; 6,\; 3)
$$

The null is "no effect" (true average extra growth $= 0$). By hand: find the estimate,
the SE (use $\mathrm{SE} = \mathrm{sd}(d)/\sqrt{n}$), and the test statistic
$t = \hat{\theta}/\mathrm{SE}$. Is $t$ past the rough "±2" line? Then check in R.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
d <- c(5, 2, 6, 3)
```

::: {.callout-note collapse="true" title="Answer"}
Estimate: $(5+2+6+3)/4 = 16/4 = 4$ cm. The spread is $\mathrm{sd}(d) \approx 1.826$,
so $\mathrm{SE} = 1.826 / \sqrt{4} = 1.826 / 2 \approx 0.913$. The test statistic is
$t = 4 / 0.913 \approx 4.38$ — well past 2, so the estimate is over four noise-widths
from 0; you'd expect to reject the null.

```{r p1-answer}
d <- c(5, 2, 6, 3)
n <- length(d)
est <- mean(d)                 # 4
se  <- sd(d) / sqrt(n)         # 0.9129
tstat <- est / se              # 4.3818
c(estimate = est, SE = se, statistic = tstat)
t.test(d)$p.value              # 0.0220 — below 0.05, reject the null
```
:::

## Problem 2 — fill in the missing step

Here is a worked test — but the **core step is blanked**. Fill in ONLY the blanked line:
turning the estimate and SE into the test statistic.

Five measurements give estimate $\hat{\theta} = 3$ and (already computed for you)
$\mathrm{SE} = 0.71$. Testing against the null value 0:

> Step 1 (given): the estimate is $\hat{\theta} = 3$ and its standard error is
> $\mathrm{SE} = 0.71$.
> Step 2: **`<your step here>`** — combine the estimate and its SE into the test
> statistic $t$.
> Step 3 (given): that $t$ is far past 2, and `t.test` returns $p \approx 0.013 < 0.05$,
> so we **reject the null**.

```{webr}
#| autorun: false
# p2-starter — edit and Run to check your statistic
d <- c(4, 2, 5, 1, 3)   # this data has mean 3 and SE ~0.71
```

::: {.callout-note collapse="true" title="Answer"}
The blanked step is the heart of the test — **distance from the null in noise units**:

$$
t = \frac{\hat{\theta}}{\mathrm{SE}} = \frac{3}{0.7071} \approx 4.24
$$

That's the move: divide the estimate by its own SE to measure it in standard-error
units. It comes out about 4.24 — over four noise-widths from 0, which is why the p-value
is tiny.

```{r p2-answer}
d <- c(4, 2, 5, 1, 3)
n <- length(d)
est <- mean(d)                 # 3
se  <- sd(d) / sqrt(n)         # 0.7071
est / se                       # 4.2426 — the blanked statistic
t.test(d)$p.value              # 0.0132 — reject at 0.05
```
:::

## Problem 3 — a full test, by hand then in R

Six measurements of an effect:

$$
g = (2,\; -1,\; 3,\; 1,\; 0,\; 4)
$$

Test the null "true effect $= 0$." Compute the estimate, SE, and statistic by hand,
then in R get the p-value and the 95% confidence interval. **PREDICT first:** with a
statistic you're about to compute, will the CI contain 0 or not — and will $p$ be under
0.05? Commit, then run.

```{webr}
#| autorun: false
# p3-starter — your job: estimate, SE, statistic, then t.test for p and CI
g <- c(2, -1, 3, 1, 0, 4)
```

::: {.callout-note collapse="true" title="Answer"}
Estimate: $(2 - 1 + 3 + 1 + 0 + 4)/6 = 9/6 = 1.5$. Spread $\mathrm{sd}(g) \approx 1.871$,
so $\mathrm{SE} = 1.871/\sqrt{6} \approx 0.764$. Statistic $t = 1.5/0.764 \approx 1.96$
— *just under* 2, so it's borderline and we'd expect **not** to reject.

The p-value is about **0.107** ($> 0.05$: fail to reject), and the 95% CI is about
$(-0.46,\; 3.46)$ — which **contains 0**. The two agree exactly, as they must: the CI
includes the null value $\iff$ $p \geq 0.05$.

```{r p3-answer}
g <- c(2, -1, 3, 1, 0, 4)
n <- length(g)
est <- mean(g)                 # 1.5
se  <- sd(g) / sqrt(n)         # 0.7638
c(estimate = est, SE = se, statistic = est / se)   # t ~ 1.96
tt <- t.test(g)
c(p = tt$p.value, lower = tt$conf.int[1], upper = tt$conf.int[2])
# p ~ 0.107 (fail to reject); CI (-0.46, 3.46) contains 0 — same verdict
```
:::

## Problem 4 — spot the error

A colleague reports a test with $p = 0.02$ and writes the following. It contains
**exactly one mistake** — find it, say why it's wrong, and give the correct reading.

> "We got $p = 0.02$. That means there is a **2% chance the null hypothesis is true**,
> so we're 98% sure the effect is real."

::: {.callout-note collapse="true" title="Answer"}
The mistake is "$p$ is the probability the null is true" (this lesson's **Trap 1**). A
p-value is computed *assuming the null is already true*: $p = 0.02$ means *if the null
were true, data at least this extreme would occur only 2% of the time* — a conditional
probability given $H_0$, written $P(\text{data this extreme} \mid H_0)$. It says nothing
about $P(H_0 \text{ is true})$, which is the flipped condition and a different number
entirely (see [probability & odds](../probability-and-odds/lesson.qmd) on conditional
probability). The correct reading: "data this extreme are unlikely under the null, so we
reject the null at 0.05" — *not* "98% sure the effect is real."
:::

## Problem 5 — mixing it with probability *(uses [probability & odds](../probability-and-odds/lesson.qmd)'s complement rule)*

A test reports $p = 0.03$. Interpreting $p$ as "the probability, **if the null were
true**, of a result at least this extreme":

(a) In that same null world, what is the probability of a result that is **less**
extreme than the one observed? (b) Explain in one sentence why this is *not* "a 97%
chance the effect is real."

::: {.callout-note collapse="true" title="Answer"}
(a) Using the complement rule from [probability & odds](../probability-and-odds/lesson.qmd)
("probability of the opposite is one minus"): the probability of a *less* extreme result
in the null world is $1 - 0.03 = 0.97$.

(b) That 0.97 is still a probability computed **inside the null world**
($P(\text{less extreme} \mid H_0)$) — it's about what data the null tends to produce,
not about whether the null is true. "The chance the effect is real" would be
$P(H_0 \text{ false} \mid \text{data})$, the flipped condition, which this machinery
never computes. Same Trap-1 confusion, dressed up with a complement.
:::

## Problem 6 — transfer to a new context

You run a *paired* study — the same 6 subjects measured before and after — and record
each subject's change (after minus before):

$$
c = (2,\; -1,\; 0,\; 1,\; -2,\; 3)
$$

The null is "no change on average" (true mean change $= 0$). Compute the estimate, SE,
statistic, the $\pm 2\,\mathrm{SE}$ interval, and decide. This is a new *context* (a
before/after design) but the *machine is identical* — one column of numbers, one null of
0. **PREDICT:** will the interval contain 0? Then run.

```{webr}
#| autorun: false
# p6-starter — same machine, new context: estimate, SE, statistic, +-2SE interval
c_change <- c(2, -1, 0, 1, -2, 3)
```

::: {.callout-note collapse="true" title="Answer"}
Estimate: $(2 - 1 + 0 + 1 - 2 + 3)/6 = 3/6 = 0.5$. Spread $\approx 1.871$, so
$\mathrm{SE} = 1.871/\sqrt{6} \approx 0.764$. Statistic $t = 0.5/0.764 \approx 0.65$ —
well within the noise. The $\pm 2\,\mathrm{SE}$ interval is
$0.5 \pm 2(0.764) = 0.5 \pm 1.53$, about $(-1.03,\; 2.03)$ — it **contains 0**.

Verdict: fail to reject; the before/after change is easily explained by noise
($p \approx 0.54$). The design changed (paired before/after instead of treated/control),
but the reasoning — estimate over its SE, compare to 0, check whether the interval
straddles the null — is exactly the toy example's. That transferability is the whole
point: these four numbers read the same way whatever produced them.

```{r p6-answer}
c_change <- c(2, -1, 0, 1, -2, 3)
n <- length(c_change)
est <- mean(c_change)          # 0.5
se  <- sd(c_change) / sqrt(n)  # 0.7638
c(estimate = est, SE = se, statistic = est / se)          # t ~ 0.65
c(lower = est - 2 * se, upper = est + 2 * se)             # (-1.03, 2.03): contains 0
t.test(c_change)$p.value       # 0.5416 — fail to reject
```
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/hypothesis-tests-p-values-and-confidence-intervals" pass="4"}

::: {.quiz-q qid="ht-pvalue-meaning" qtype="mc" answer="b"}
A study reports $p = 0.045$. What does that mean?

a) The null hypothesis has only a 4.5% chance of being true
b) Assuming the null, data this extreme would arise about 4.5% of the time
c) The probability the effect is real is 95.5%
:::

::: {.quiz-q qid="ht-fail-to-reject" qtype="mc" answer="a"}
With $p = 0.10$ you fail to reject the null. What can you conclude?

a) Not enough evidence; the effect might be real but too small to catch here
b) The null is proven true — the effect is definitely zero, not just undetected
c) The alternative is proven true — failing to reject confirms the effect exists
:::

::: {.quiz-q qid="ht-test-statistic" qtype="mc" answer="c"}
A test statistic is built as:

a) The estimate minus the null value, scaled by the sample size
b) The p-value multiplied by the sample size to get a count of extreme outcomes
c) The estimate divided by its standard error
:::

::: {.quiz-q qid="ht-type-i-error" qtype="mc" answer="a"}
The cutoff $\alpha = 0.05$ sets the rate of which error?

a) Type I — a false alarm: rejecting the null when it is actually true
b) Type II — a missed signal: failing to reject when there is a real effect
c) Neither — $\alpha$ is the probability the null is true
:::

::: {.quiz-q qid="ht-ci-upper" qtype="numeric" answer="14" tolerance="0"}
(From randomness & sampling distributions.) A 95% confidence interval is roughly estimate
$\pm\ 2$ standard errors. If the estimate is $10$ with SE $= 2$, what is the upper end
(estimate $+ 2 \times$ SE)?
:::

:::

::::
