---
title: "Odds, log-odds, and the logit link"
subtitle: "Logistic regression fits a line to the log-odds — here's why, and how to read the slope"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live cells below re-state their own data. Base R only.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

::: {.callout-tip title="This lesson runs in your browser" collapse="true"}
The code cells below with a **Run** button are live — powered by WebR (R compiled to
WebAssembly), they run right here on your phone or laptop, no install needed. Edit any
of them and re-run. The first run downloads a small R engine, so give it a few seconds.
Cells without a Run button are ordinary listings whose output was computed when the
page was built.
:::

## Where we're going

Last module we bent a line into a probability with the sigmoid. Now we answer the
question it leaves hanging: *what is the line actually a line for?* The surprising,
tidy answer is that logistic regression models the **log-odds** as a straight line in
$x$ — a relationship called the **logit link**. That single choice is what makes the
coefficients readable: this module teaches you to read the slope $\beta_1$ two ways — as
a "change in log-odds per unit of $x$," and, after one exponentiation, as an **odds
ratio**, the form you'll actually see quoted in papers.

**Builds on:** [the sigmoid](../01-from-line-to-probability/lesson.qmd) (module 1), and
the [probability-and-odds](../../../../foundations/probability-and-odds/lesson.qmd)
foundation (odds, log-odds, and their conversions), which itself leans on
[exponentials-and-logs](../../../../foundations/exponentials-and-logs/lesson.qmd).

## Warm-up: do you remember?

Three recalls spanning module 1 and the foundations. Produce each answer before opening
it.

**1.** *(module 1)* What is the sigmoid $\sigma(z)$, and what is $\sigma(0)$?

::: {.callout-note collapse="true" title="Answer"}
$\sigma(z) = 1/(1 + e^{-z})$, the S-shaped squasher, and $\sigma(0) = 0.5$. See
[module 1](../01-from-line-to-probability/lesson.qmd).
:::

**2.** *(probability-and-odds)* A probability is $p = 0.8$. What are its odds, and its
log-odds?

```{webr}
#| autorun: false
# warmup2-starter — edit and Run to check
p <- 0.8
```

::: {.callout-note collapse="true" title="Answer"}
Odds $= p/(1-p) = 0.8/0.2 = 4$; log-odds $= \log(4) \approx 1.386$. Shaky? Revisit
[probability-and-odds](../../../../foundations/probability-and-odds/lesson.qmd).
:::

**3.** *(exponentials-and-logs)* What is $e^{\log(4)}$? And what does the log rule say
$\log(a) - \log(b)$ equals as a single log?

::: {.callout-note collapse="true" title="Answer"}
$e^{\log(4)} = 4$ (exp undoes log). And $\log(a) - \log(b) = \log(a/b)$ — a difference of
logs is the log of the ratio. See
[exponentials-and-logs](../../../../foundations/exponentials-and-logs/lesson.qmd). Both
facts do heavy lifting in this lesson.
:::

## A tiny example first

We're modelling whether a loan is repaid, from the applicant's income $x$ (in tens of
thousands). Suppose the true relationship, on the **log-odds scale**, is a straight line:

$$
\text{log-odds of repayment} = -2 + 1 \cdot x.
$$

So the log-odds is $-2$ at $x = 0$, and climbs by $1$ for every extra unit of income.
Let's tabulate three incomes and walk each all the way to a probability, by hand:

- **$x = 1$:** log-odds $= -2 + 1(1) = -1$. Odds $= e^{-1} \approx 0.368$. Probability
  $= \tfrac{0.368}{1 + 0.368} \approx 0.269$.
- **$x = 2$:** log-odds $= -2 + 1(2) = 0$. Odds $= e^{0} = 1$. Probability $=
  \tfrac{1}{2} = 0.5$.
- **$x = 3$:** log-odds $= -2 + 1(3) = +1$. Odds $= e^{1} \approx 2.718$. Probability
  $= \tfrac{2.718}{3.718} \approx 0.731$.

Two things to notice. First, the **log-odds** moved in equal steps ($-1, 0, +1$) — it's
the thing that's linear in $x$. Second, going from log-odds to probability is exactly the
[probability-and-odds](../../../../foundations/probability-and-odds/lesson.qmd) round trip
(exponentiate to get odds, then odds/(1+odds)), and — this is the punchline — it lands on
the very same numbers the **sigmoid** gives: $\sigma(-1) \approx 0.269$, $\sigma(0) = 0.5$,
$\sigma(1) \approx 0.731$.

The same thing in R:

```{r toy-example}
b0 <- -2; b1 <- 1
x  <- c(1, 2, 3)

log_odds <- b0 + b1 * x                 # the LINE: -1, 0, 1 (equal steps)
odds     <- exp(log_odds)               # exp undoes log -> odds
prob     <- odds / (1 + odds)           # odds -> probability

sigmoid  <- function(z) 1 / (1 + exp(-z))
rbind(log_odds = log_odds,
      odds     = odds,
      prob     = prob,
      sigmoid  = sigmoid(log_odds))     # bottom two rows are IDENTICAL
```

## Getting a feel for it

**A mental picture: a volume knob wired to a dimmer.** In module 1 the sigmoid was a
dimmer switch — score in, brightness (probability) out. Now picture *turning a volume
knob* labelled $x$. Each equal click of the knob adds the *same amount* to the log-odds
(that's the line: constant slope). But because the log-odds feeds a dimmer, those equal
clicks produce *unequal* changes in brightness: near the middle (probability around
$0.5$) a click swings the light a lot; out near "off" or "full," the same click barely
changes it. So the story is "**even steps on the log-odds scale, uneven steps on the
probability scale**" — which is exactly the shape of the S-curve.

Where the picture bends: a volume knob has a floor and a ceiling you can slam into, but
the log-odds knob has neither — you can keep clicking $x$ forever and the log-odds keeps
climbing, even as the probability creeps toward (but never reaches) 1.

**Why model the log-odds with a line, and not the probability directly?** Two reasons,
both of which you can now feel:

1. **Range.** A line runs over all real numbers, and so does the log-odds (from the
   [probability-and-odds](../../../../foundations/probability-and-odds/lesson.qmd)
   module: probability is boxed in $[0,1]$, but log-odds is unbounded both ways). A line
   fits the log-odds scale perfectly — no poking outside a legal range, the exact problem
   that killed the plain-line-on-probability idea in module 1.
2. **A constant, readable effect.** On the log-odds scale the effect of $x$ is *constant*
   — one extra unit of income always adds $\beta_1$ to the log-odds, whether you start
   rich or poor. "Every extra \$10k adds the same amount to the log-odds" is a clean
   sentence; "every extra \$10k adds *some* amount to the probability that depends on
   where you started" is not.

**Why exponentiating the slope gives an odds ratio — and what the number feels like.**
Here's the move that makes $\beta_1$ quotable. Increasing $x$ by 1 adds $\beta_1$ to the
log-odds. On the *odds* scale, adding $\beta_1$ to a log means **multiplying** the odds
by $e^{\beta_1}$ (from the undo pair — a difference of logs is a log of a ratio, run
backwards). So $e^{\beta_1}$ is the factor the odds get multiplied by per unit of $x$ —
the **odds ratio**. In our example $\beta_1 = 1$, so $e^{1} \approx 2.718$: each extra
unit of income makes repayment about **2.7 times** the odds it was. An odds ratio of $1$
means "no effect" (multiplying by 1 changes nothing); above 1 means $x$ raises the odds;
below 1 means $x$ lowers them.

## What the notation is going to say

The link equation will say: "the log-odds of the outcome is a straight line in $x$" —
$\log\!\big(\tfrac{p}{1-p}\big) = \beta_0 + \beta_1 x$. In our toy example that's the
"$-2 + 1 \cdot x$" we tabulated. Read $\beta_1$ as "how much the log-odds rises per unit
of $x$," and $e^{\beta_1}$ as "the factor the *odds* get multiplied by per unit of $x$."

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $p$ | "pee" | the probability of the outcome (here, repayment) | $0.269, 0.5, 0.731$ |
| $\dfrac{p}{1-p}$ | "pee over one-minus-pee" | the **odds** of the outcome | $0.368, 1, 2.718$ |
| $\log$ | "log" (natural log) | natural log, base $e$ (in R: `log`); from [exponentials-and-logs](../../../../foundations/exponentials-and-logs/lesson.qmd) | $\log(3) \approx 1.099$ |
| $e^{(\cdot)}$ | "ee to the …" | the exponential (in R: `exp`); undoes $\log$, so $e^{\log u} = u$ | $e^{1} \approx 2.718$ |
| $\operatorname{logit}(p)$ | "the LOH-jit of pee" | the **log-odds**: $\log\!\big(\tfrac{p}{1-p}\big)$ | $-1, 0, 1$ |
| $\sigma(z)$ | "SIG-muh of zee" | the **sigmoid** from [module 1](../01-from-line-to-probability/lesson.qmd): score → probability (undoes the logit) | $\sigma(1) \approx 0.731$ |
| $\beta_0$ | "beta zero" | intercept: the log-odds when $x = 0$ | $-2$ |
| $\beta_1$ | "beta one" | slope: the change in log-odds per one-unit rise in $x$ | $1$ |
| $e^{\beta_1}$ | "ee to the beta one" | the **odds ratio**: the factor the odds multiply by per unit of $x$ | $e^{1} \approx 2.718$ |
| $x$ | "ex" | the predictor (here, income) | $1, 2, 3$ |

## The formal version

The **logit link** — the defining equation of logistic regression:

$$
\underbrace{\log\!\left(\frac{p}{1-p}\right)}_{\text{log-odds of the outcome}}
= \beta_0 + \beta_1 x.
$$

Reading it aloud: "log of (pee over one-minus-pee) equals beta-zero plus beta-one ex" —
or in everyday words: *the log-odds of the outcome is a straight line in $x$.*

It is the exact inverse of module 1's sigmoid. Solving the link for $p$ un-does the
logit and lands you back on the sigmoid — one algebra chain, each step a move you've
already met:

$$
\begin{aligned}
\log\!\left(\frac{p}{1-p}\right) &= \beta_0 + \beta_1 x \\
\frac{p}{1-p} &= e^{\,\beta_0 + \beta_1 x} \\
p &= e^{\,\beta_0 + \beta_1 x}\,(1 - p) \\
p\left(1 + e^{\,\beta_0 + \beta_1 x}\right) &= e^{\,\beta_0 + \beta_1 x} \\
p &= \frac{e^{\,\beta_0 + \beta_1 x}}{1 + e^{\,\beta_0 + \beta_1 x}}
   = \frac{1}{1 + e^{-(\beta_0 + \beta_1 x)}} = \sigma(\beta_0 + \beta_1 x)
\end{aligned}
$$

- Line 2: apply $e^{(\cdot)}$ to both sides; $e^{\log(u)} = u$ cancels the log on the left
  (the undo pair, from
  [exponentials-and-logs](../../../../foundations/exponentials-and-logs/lesson.qmd)).
- Line 3: multiply both sides by $(1-p)$ — clearing a fraction, ordinary algebra.
- Line 4: expand the right side and gather every $p$ term on the left (distribute, then
  collect like terms — the "solve for the unknown" move from
  [simple linear regression](../../../simple-linear-regression/syllabus.md)).
- Line 5: divide both sides by $\left(1 + e^{\beta_0 + \beta_1 x}\right)$; the final tidy
  form comes from dividing top and bottom by $e^{\beta_0 + \beta_1 x}$ (multiplying by
  $\tfrac{e^{-(\cdot)}}{e^{-(\cdot)}} = 1$), which is exactly the sigmoid of the score.

So **logit and sigmoid are two views of one model**: the logit takes probability → score
(a line), the sigmoid takes score → probability. Sanity-check that the link really
reproduces module 1's sigmoid on the toy numbers:

```{r formal-check}
b0 <- -2; b1 <- 1
x  <- c(1, 2, 3)

# Route A: the link -> solve for p by the derived formula
score <- b0 + b1 * x
p_via_link <- exp(score) / (1 + exp(score))

# Route B: module 1's sigmoid of the same score
sigmoid <- function(z) 1 / (1 + exp(-z))
p_via_sigmoid <- sigmoid(score)

rbind(p_via_link = p_via_link, p_via_sigmoid = p_via_sigmoid)  # identical rows

# And the odds ratio: exp(beta1) is the per-unit multiplier on the ODDS
odds <- exp(score)                       # odds at x = 1, 2, 3
c(odds_ratio_step = odds[2] / odds[1],   # odds(x=2)/odds(x=1)
  exp_beta1        = exp(b1))            # both = e^1 ~ 2.718
```

The bottom check makes the odds-ratio claim concrete: the odds at $x = 2$ divided by the
odds at $x = 1$ equals $e^{\beta_1}$ — the odds get multiplied by that same factor for
*every* one-unit step.

## Explore it

```{webr}
#| autorun: true
#| fig-alt: "Two side-by-side plots: on the left, log-odds as a straight line rising with x; on the right, the same model as probability, an S-shaped curve rising from near 0 to near 1."
# explore-1 — LIVE: the SAME model, two scales. Straight line on the log-odds scale (left)
# becomes the S-curve on the probability scale (right). PREDICT: where the left line
# crosses 0, what is the probability on the right? Guess, then run.
b0 <- -2; b1 <- 1
sigmoid <- function(z) 1 / (1 + exp(-z))
xs <- seq(-2, 6, length.out = 200)

op <- par(mfrow = c(1, 2))
plot(xs, b0 + b1 * xs, type = "l", lwd = 2, col = "steelblue",
     xlab = "x", ylab = "log-odds", main = "log-odds: a straight line")
abline(h = 0, lty = 3)
plot(xs, sigmoid(b0 + b1 * xs), type = "l", lwd = 2, col = "firebrick",
     xlab = "x", ylab = "probability p", main = "probability: an S-curve")
abline(h = 0.5, lty = 3)
par(op)
```

```{webr}
#| autorun: false
# explore-2 — LIVE: read a coefficient as an ODDS RATIO. With slope b1, one extra unit of
# x multiplies the odds by exp(b1). PREDICT: with b1 = 0.7, is the odds ratio above or
# below 1, and what does that mean? Guess, then run. (Try b1 = -0.5 too.)
b1 <- 0.7                          # <-- change me
odds_ratio <- exp(b1)
c(beta1 = b1, odds_ratio = odds_ratio,
  meaning_above_1 = odds_ratio > 1)   # >1 means x raises the odds
```

```{webr}
#| autorun: false
# explore-3 — LIVE: equal steps in x add equal amounts to the LOG-ODDS, but UNEQUAL
# amounts to the PROBABILITY. PREDICT: as x goes 0,1,2,3,4, will the jumps in probability
# be constant? Guess, then run and look at diff().
b0 <- -2; b1 <- 1
sigmoid <- function(z) 1 / (1 + exp(-z))
x <- 0:4
lo <- b0 + b1 * x
p  <- sigmoid(lo)
rbind(x = x, log_odds = lo, prob = round(p, 3))
list(step_in_log_odds = diff(lo),      # constant (= b1)
     step_in_prob      = round(diff(p), 3))   # NOT constant
```

## Common traps

**Trap: "the slope $\beta_1$ is the change in *probability* per unit of $x$."** Very
tempting — that's what slope means in ordinary regression. But here $\beta_1$ is the
change in **log-odds**, and the probability change is *not* constant. In our model
($\beta_0 = -2, \beta_1 = 1$), going from $x = 1$ to $x = 2$ raises the probability from
$0.269$ to $0.5$ (a jump of $0.231$), but from $x = 4$ to $x = 5$ it goes $0.881 \to
0.953$ (a jump of only $0.072$). Same one-unit step in $x$, same $+1$ in log-odds — but
very different probability jumps, because of the S-curve's flattening ends.

**Trap: "an odds ratio of 2 means the probability doubles."** No — it means the *odds*
double, and doubling the odds does *not* double the probability. Start at $p = 0.4$ (odds
$0.667$); double the odds to $1.333$; the new probability is $1.333/2.333 \approx 0.571$,
not $0.8$. Odds ratios multiply odds, and only near very small probabilities does that
approximately match "multiplying the probability." Always convert if you want the
probability.

## Check yourself

**1.** A logistic model has log-odds $= -1 + 0.5x$. What is the log-odds at $x = 4$, and
the corresponding probability?

```{webr}
#| autorun: false
# check1-starter — edit and Run to solve, right here
b0 <- -1; b1 <- 0.5
x <- 4
```

::: {.callout-note collapse="true" title="Answer"}
Log-odds $= -1 + 0.5(4) = 1$. Odds $= e^{1} \approx 2.718$; probability $= 2.718/3.718
\approx 0.731$ (equivalently $\sigma(1)$).

```{r check1-answer}
b0 <- -1; b1 <- 0.5; x <- 4
lo <- b0 + b1 * x                    # 1
c(log_odds = lo, prob = exp(lo) / (1 + exp(lo)))   # 1, 0.731
```
:::

**2.** A coefficient is $\beta_1 = 0.69$. What is the odds ratio, and — in one sentence —
what does it tell you about the effect of a one-unit rise in $x$?

::: {.callout-note collapse="true" title="Answer"}
Odds ratio $= e^{0.69} \approx 1.99 \approx 2$. So each one-unit rise in $x$ roughly
**doubles the odds** of the outcome (multiplies them by about 2). (It does *not* double
the probability — see the trap above.)
:::

**3.** In your own words: explain to a colleague, in two sentences, why logistic
regression puts a line on the log-odds instead of on the probability, and what
$e^{\beta_1}$ means.

::: {.callout-note collapse="true" title="A model answer"}
It puts the line on the log-odds because the log-odds ranges over all real numbers (like a
line does), so the model can't predict an impossible probability, and because the effect
of $x$ is then a clean constant — the same log-odds change per unit no matter where you
start. $e^{\beta_1}$ is the odds ratio: the factor by which the odds of the outcome get
multiplied for each one-unit increase in $x$.
:::

## What you can now do

- **Write the logit link** $\log\!\big(\tfrac{p}{1-p}\big) = \beta_0 + \beta_1 x$ and
  explain why the log-odds (not the probability) is the thing modelled as a line.
- **Travel between scales**: log-odds → odds ($e^{(\cdot)}$) → probability
  ($\text{odds}/(1+\text{odds})$), and recognize this is the sigmoid run backwards.
- **Interpret a coefficient two ways**: $\beta_1$ as the change in log-odds per unit of
  $x$, and $e^{\beta_1}$ as the odds ratio (the per-unit multiplier on the odds).
- **Avoid the two classic misreadings**: $\beta_1$ is not a change in probability, and an
  odds ratio is not a probability ratio.

If any of these feels shaky, revisit [Getting a feel for it](#getting-a-feel-for-it) and
[The formal version](#the-formal-version) — the next module needs the log-odds ↔
probability round trip cold.

## Where this goes next

We now have the *shape* of the model — a line on the log-odds — but not yet the
$\beta$'s themselves. Where do the fitted coefficients come from? Unlike least squares
there's no neat formula; instead we ask "which $\beta$'s make the observed 0/1 data most
probable?" and climb to the top of that **likelihood** by gradient ascent — reusing your
[gradient-descent](../../../gradient-descent/syllabus.md) course with the sign flipped:
[likelihood and fitting by gradient ascent](../03-likelihood-and-gradient-ascent/lesson.qmd).
Ready for more reps first? Do the [practice problems](practice.qmd).
