---
title: "How wrong is the model, really?"
subtitle: "Residuals, RSS/MSE/RMSE, R², over- vs under-fitting, and grading on data you didn't study — train/test and cross-validation"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live cells below re-state their own data. Base R is all
# we need, so nothing to install.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

::: {.callout-tip title="This lesson runs in your browser" collapse="true"}
The code cells below with a **Run** button are live — powered by WebR (R compiled to
WebAssembly), they run right here on your phone or laptop, no install needed. Edit any
of them and re-run. The first run downloads a small R engine, so give it a few seconds.
Cells without a Run button are ordinary listings whose output was computed when the
page was built.
:::

## Where we're going

Once a method has produced predictions, one question decides everything: **how wrong
are they, and will they *stay* wrong on data the model has never seen?** This lesson
builds the whole toolkit of answers, in plain base R: the **residual** (one prediction's
miss), the summary numbers that roll all the misses into one score (**RSS, MSE, RMSE**),
a "better than just guessing the mean?" score (**R²**), the twin failures of
**overfitting and underfitting**, and the honesty devices that catch them —
**train/test split** and **k-fold cross-validation**. Almost every fitting method,
anywhere, is graded with exactly these.

We stay with tiny, hand-checkable numbers and neutral examples (predicted vs observed
values, a cloud of points). We will *not* touch hypothesis tests, standard errors, or
any method-specific score (those live in the modules and courses that need them).

**Builds on:** [the mean, variance & covariance](../mean-variance-covariance/lesson.qmd)
— we reuse its **why-we-square** story (squaring stops positive and negative misses from
cancelling) and the mean as a balance point; and
[loss functions & optimization](../loss-functions-and-optimization/lesson.qmd) — the
error score here *is* a loss, the one a fitting method already minimized on its
training data.

## Warm-up: do you remember?

Before the new stuff, three quick reps — answers from memory first, then check.

**1.** For `d <- c(-2, 0, -1, 1, 2)` (a set of "misses"), why can't you just *add* them
up to measure total error? What single trick fixes it, and what does `sum(d^2)` give?

```{webr}
#| autorun: false
# warmup1-starter — edit and Run to check
d <- c(-2, 0, -1, 1, 2)
```

::: {.callout-note collapse="true" title="Answer"}
Adding them gives $-2 + 0 - 1 + 1 + 2 = 0$: the over- and under-shoots cancel, exactly
like deviations from the mean do. The fix is to **square first** — a squared number is
never negative, so nothing cancels — which is the same move variance uses. `sum(d^2)` =
$4 + 0 + 1 + 1 + 4 = 10$. Shaky on why-we-square? Revisit
[the mean, variance & covariance](../mean-variance-covariance/lesson.qmd).
:::

**2.** For `y <- c(3, 5, 6, 9, 12)`, what is the mean $\bar{y}$ (say it "why-bar"), and
what R call gives it?

```{webr}
#| autorun: false
# warmup2-starter — edit and Run to check
y <- c(3, 5, 6, 9, 12)
```

::: {.callout-note collapse="true" title="Answer"}
$\bar{y} = (3 + 5 + 6 + 9 + 12)/5 = 35/5 = 7$. In R: `mean(y)`. The mean is the balance
point of the data — hold onto it, because "just guess the mean for everyone" is the
baseline we'll grade every model against. See
[the mean, variance & covariance](../mean-variance-covariance/lesson.qmd).
:::

**3.** In one sentence: in [loss functions & optimization](../loss-functions-and-optimization/lesson.qmd),
what does a fitting method actually *do* with an error score?

::: {.callout-note collapse="true" title="Answer"}
It **minimizes** it — it turns the knobs (the parameters) to make that "how wrong am I"
number as small as possible on the data it was given. The scores in this lesson are those
very error numbers; the new idea is checking them on data the method did *not* get to
minimize over. See
[loss functions & optimization](../loss-functions-and-optimization/lesson.qmd).
:::

## A tiny example first

A model made predictions for 5 items. Here are the **observed** values it was trying to
hit and the values it **predicted**:

| item | observed $y$ | predicted $\hat{y}$ |
|---|---|---|
| 1 | 3  | 4  |
| 2 | 5  | 5  |
| 3 | 6  | 7  |
| 4 | 9  | 8  |
| 5 | 12 | 11 |

**Residual = observed − predicted.** For each item, subtract the prediction from what
actually happened:
$3-4 = -1$, then $5-5 = 0$, then $6-7 = -1$, then $9-8 = +1$, then $12-11 = +1$.
So the residuals are $(-1,\ 0,\ -1,\ 1,\ 1)$. A negative residual means the model
guessed *too high*; a positive one, too low; zero is a bullseye.

**Roll them into one score.** Add them up and you get $-1+0-1+1+1 = 0$ — useless, they
cancelled (the warm-up trap!). So **square each residual first**, then add:
$(-1)^2 + 0^2 + (-1)^2 + 1^2 + 1^2 = 1+0+1+1+1 = 4$. That total, **4**, is the
**RSS — the residual sum of squares.** Divide by how many items (5) to get the *average*
squared miss, the **MSE** (mean squared error): $4/5 = 0.8$. Take the square root to get
back to the original units, the **RMSE** (root mean squared error):
$\sqrt{0.8} \approx 0.894$.

**Compare to just guessing the mean.** How well would a lazy "predict $\bar{y}=7$ for
everyone" model do? Its total squared miss is $\sum (y_i - 7)^2 =
(-4)^2 + (-2)^2 + (-1)^2 + 2^2 + 5^2 = 16+4+1+4+25 = 50$. Call that **TSS** (total sum of
squares — the spread of $y$ around its own mean). Our model's leftover error is only 4 out
of that 50, so it explained $50-4 = 46$ of it. The fraction explained is
$46/50 = 0.92$ — that is **R², "R-squared."**

The same thing in R, so you can check every step:

```{r toy-example}
y    <- c(3, 5, 6, 9, 12)     # observed
yhat <- c(4, 5, 7, 8, 11)     # the model's predictions

resid <- y - yhat             # residual = observed - predicted
resid                         # -1  0 -1  1  1

RSS  <- sum(resid^2)          # square each miss, then add: 1+0+1+1+1
RSS                           # 4
MSE  <- RSS / length(y)       # average squared miss: 4 / 5
MSE                           # 0.8
RMSE <- sqrt(MSE)             # back to the original units
RMSE                          # 0.8944272

ybar <- mean(y)               # the "guess the mean" baseline: 7
TSS  <- sum((y - ybar)^2)     # total spread of y around its mean
TSS                           # 50
R2   <- 1 - RSS / TSS         # fraction of that spread the model removed
R2                            # 0.92
```

## Getting a feel for it

**The mental picture: a report card, not a single grade.** Think of the model as a
student who sat an exam. Each **residual** is one question's error — how far the
student's answer sat from the right one, with a sign for "too high" or "too low."
**RSS/MSE/RMSE** are three ways of averaging the whole report card into one number, and
**R²** is the report card *graded on a curve*: not "how many points did you drop?" but
"how much better than the class dunce — the student who writes the class average on every
question — did you do?"

**Why square the residuals (again)?** Same reason variance squares deviations: raw
misses cancel (our five summed to 0), and squaring is the cleanest fix that never
cancels — it also punishes one big miss more than several small ones, which is usually
what you want. Because squaring inflates the units (squared points), RMSE takes the
square root at the end to land back in the *original* units, so an RMSE of 0.894 reads
as "typical miss of about nine-tenths of a point" — a number you can actually picture.
This square-then-average-then-un-square shape is exactly the mean → variance →
standard-deviation move from
[the mean, variance & covariance](../mean-variance-covariance/lesson.qmd); MSE is the
variance of the misses, RMSE their standard deviation.

**Why divide by TSS for R²?** A raw RSS of 4 means nothing on its own — is 4 good? It
depends entirely on how spread out $y$ was to begin with. Dividing the *explained* part
($50-4$) by the *total* spread (50) rescales the score to a fraction, so it reads the
same whether $y$ is measured in millimetres or miles. **R² = 1 is a perfect fit** (zero
leftover error), **R² = 0 means "no better than guessing the mean"** (all the spread is
still there), and R² can even go **negative** for a model worse than that lazy baseline.

**What our numbers feel like.** RMSE ≈ 0.894 on values that run 3 to 12 is small — the
model is usually within a point. R² = 0.92 says it wiped out 92% of the raw spread. Both
say "good fit." But a warning we'll spend the rest of the lesson on: these are graded on
the *same* data the model studied. A good grade on your own practice exam is not the same
as a good grade on the real one.

**Cousin, one breath:** for models fit by likelihood rather than squared error (you'll
meet those in a GLM course), the analogue of RSS is called the **deviance** — same job,
"total leftover badness," measured on the likelihood scale instead of the squared-error
scale. The detail stays in those courses; just know the word points at the same idea.

## What the notation is going to say

The formulas below will say, in symbols, exactly the arithmetic we just did. The
**residual** formula will say "take each observed value and subtract the model's
prediction for it" — our $3-4=-1$. **RSS** will say "square every residual and add them
up" — our $1+0+1+1+1=4$. **MSE** divides that by $n$ (how many points), **RMSE** square-
roots it. And **R²** will say "one minus the ratio of leftover error to total spread" —
our $1 - 4/50 = 0.92$. No new arithmetic, just compact names.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $y_i$ | "why sub eye" | the $i$-th observed value (in R: `y[i]`) | $y_1=3,\dots,y_5=12$ |
| $\hat{y}_i$ | "why hat sub eye" | the model's predicted value for item $i$ (the hat = "estimated") | $\hat{y}_1=4,\dots,\hat{y}_5=11$ |
| $e_i$ or $r_i$ | "ee sub eye" / "residual" | observed minus predicted: $y_i - \hat{y}_i$ | $(-1,0,-1,1,1)$ |
| $n$ | "en" | the number of data points (in R: `length(y)`) | 5 |
| $\bar{y}$ | "why bar" | the mean of the observed values (in R: `mean(y)`) | 7 |
| $\sum_{i=1}^{n}$ | "the sum from i equals 1 to n" | add the term to its right, once per item (in R: `sum(...)`) | over the 5 items |
| $\text{RSS}$ | "R-S-S" | residual sum of squares: $\sum (y_i-\hat{y}_i)^2$ | 4 |
| $\text{TSS}$ | "T-S-S" | total sum of squares: $\sum (y_i-\bar{y})^2$ (spread of $y$) | 50 |
| $\text{MSE}$ | "M-S-E" | mean squared error: $\text{RSS}/n$ | 0.8 |
| $\text{RMSE}$ | "R-M-S-E" | root mean squared error: $\sqrt{\text{MSE}}$ | 0.894 |
| $R^2$ | "R-squared" | fraction of $y$'s spread the model explains: $1-\text{RSS}/\text{TSS}$ | 0.92 |

## The formal version

**Residual — one prediction's miss.** For each item $i$:

$$
e_i = y_i - \hat{y}_i
$$

Reading it aloud: "ee-sub-eye equals why-sub-eye minus why-hat-sub-eye" — in everyday
words: **"the residual is what actually happened minus what the model said would
happen."** Item 3: $e_3 = 6 - 7 = -1$, matching our hand computation.

**RSS, MSE, RMSE — rolling the misses into one score.** Build them one piece at a time.
Start with the residuals; they cancel if you just add them, so square each first (the
warm-up trick), then add — that is RSS:

$$
\text{RSS} = \sum_{i=1}^{n} (y_i - \hat{y}_i)^2
$$

Reading it aloud: "the sum, from i equals 1 to n, of the quantity why-sub-eye minus
why-hat-sub-eye, squared" — everyday words: **"square every miss and add them up."**
Our value: $(-1)^2+0^2+(-1)^2+1^2+1^2 = 4$. RSS grows with the number of points, so
divide by $n$ to get the *average* squared miss (MSE), then square-root it to return to
the original units (RMSE):

$$
\text{MSE} = \frac{\text{RSS}}{n} = \frac{1}{n}\sum_{i=1}^{n}(y_i-\hat{y}_i)^2,
\qquad
\text{RMSE} = \sqrt{\text{MSE}}
$$

Everyday words: **"MSE is the average squared miss; RMSE puts it back in the units you
care about."** Ours: $\text{MSE}=4/5=0.8$ and $\text{RMSE}=\sqrt{0.8}\approx 0.894$ —
the numbers from the toy chunk.

**R² — better than guessing the mean?** First the baseline's error, TSS, which is just
RSS if the "model" predicts $\bar{y}$ for everyone:

$$
\text{TSS} = \sum_{i=1}^{n} (y_i - \bar{y})^2
$$

Then R² is one minus how much error is *left over* as a fraction of that baseline:

$$
R^2 = 1 - \frac{\text{RSS}}{\text{TSS}}
$$

Reading it aloud: "R-squared equals one minus RSS over TSS" — everyday words: **"of all
the spread there was to explain, what fraction did the model remove?"** Ours:
$1 - 4/50 = 1 - 0.08 = 0.92$. If the model were no better than the mean-guesser,
$\text{RSS}=\text{TSS}$ and $R^2 = 0$; a perfect model has $\text{RSS}=0$ and $R^2=1$.

## Explore it

```{webr}
#| autorun: true
# explore-1 — LIVE: the error scores ARE the by-hand arithmetic.
# TRY IT: change a prediction in yhat and re-run — watch every score move.
# PREDICT: if you make yhat exactly equal to y (a perfect model), what will RSS,
# RMSE, and R2 become? Commit to three numbers, then run.
y    <- c(3, 5, 6, 9, 12)
yhat <- c(4, 5, 7, 8, 11)

resid <- y - yhat
RSS   <- sum(resid^2)                 # by hand: square the misses, add
MSE   <- RSS / length(y)
RMSE  <- sqrt(MSE)
TSS   <- sum((y - mean(y))^2)
R2    <- 1 - RSS / TSS

c(RSS = RSS, MSE = MSE, RMSE = RMSE, R2 = R2)   # 4, 0.8, 0.894, 0.92
```

```{webr}
#| autorun: true
# explore-2 — LIVE: R2 = 0 is EXACTLY the mean-guesser.
# PREDICT: this model predicts the mean of y for every point. What is its RSS
# compared to TSS, and therefore its R2? Write it down, then run.
y    <- c(2, 4, 6, 8, 10)
yhat <- rep(mean(y), length(y))       # "guess the mean for everyone"

RSS <- sum((y - yhat)^2)
TSS <- sum((y - mean(y))^2)
c(RSS = RSS, TSS = TSS, R2 = 1 - RSS / TSS)   # RSS == TSS, so R2 = 0
```

Now the picture at the heart of the second half of the lesson — **overfitting vs
underfitting.** The dots are noisy data whose *true* pattern is a straight line. A
too-simple model (a flat/near-flat guess) **underfits** — it misses the real trend. A
too-flexible model (a wiggly curve) **overfits** — it bends to chase the noise, so it
fits *these* dots beautifully but would miss new dots badly. The right model captures the
trend without chasing the wiggle.

```{webr}
#| autorun: true
#| fig-alt: "A scatter of points rising left to right. A straight line passes through the trend (a good fit); a nearly flat line sits below the trend (underfitting); a wiggly curve threads through every point (overfitting)."
# explore-3 — LIVE: underfit vs good fit vs overfit, on the same dots.
# TRY IT: change the wiggly degree from 10 to 2 and re-run — watch the curve calm down.
set.seed(42)
n <- 30
x <- sort(runif(n, 0, 10))
y <- 2 + 1.5 * x + rnorm(n, 0, 2)        # true pattern = a straight line + noise

fit_line   <- lm(y ~ poly(x, 1))          # good fit: matches the true shape
fit_wiggly <- lm(y ~ poly(x, 10))         # overfit: bends to chase the noise
grid <- data.frame(x = seq(0, 10, length.out = 200))

plot(x, y, pch = 19, col = "grey40", main = "Under-, good, and over-fitting")
abline(h = mean(y), lty = 3, lwd = 2, col = "orange")                 # underfit: flat
lines(grid$x, predict(fit_line,   grid), col = "blue", lwd = 2)       # good fit
lines(grid$x, predict(fit_wiggly, grid), col = "red",  lwd = 2)       # overfit
legend("topleft", c("underfit (mean)", "good (line)", "overfit (wiggly)"),
       col = c("orange", "blue", "red"), lty = c(3, 1, 1), lwd = 2, bty = "n")
```

**Train/test split — grade on data you didn't study.** The fix for overfitting is to
hide some data from the model while it fits, then grade it only on that hidden slice.
Watch the wiggly model *win* on training data but *lose* on held-out data:

```{webr}
#| autorun: false
# explore-4 — LIVE: the same wiggly model, judged two ways.
# PREDICT: the wiggly (degree-9) curve will have the SMALLER training RMSE. Will it
# also have the smaller TEST RMSE? Commit to yes/no, then run.
set.seed(42)
n <- 30
x <- sort(runif(n, 0, 10))
y <- 2 + 1.5 * x + rnorm(n, 0, 2)
dat <- data.frame(x = x, y = y)

set.seed(42)
test_idx <- sample(seq_len(n), size = 10)   # hide 10 points; train on the other 20
train <- dat[-test_idx, ]
test  <- dat[ test_idx, ]

rmse <- function(obs, pred) sqrt(mean((obs - pred)^2))
fit_line   <- lm(y ~ poly(x, 1), data = train)
fit_wiggly <- lm(y ~ poly(x, 9), data = train)

rbind(
  train = c(line = rmse(train$y, predict(fit_line)),
            wiggly = rmse(train$y, predict(fit_wiggly))),
  test  = c(line = rmse(test$y,  predict(fit_line,   test)),
            wiggly = rmse(test$y,  predict(fit_wiggly, test)))
)
# train: line 1.985, wiggly 1.528  (wiggly looks better...)
# test : line 2.886, wiggly 7.367  (...but is far worse on unseen data)
```

**k-fold cross-validation — the systematic version.** One split throws away a chunk of
data and depends on *which* points landed in the test slice. k-fold cross-validation
rotates the test slice: chop the data into $k$ equal **folds**, then $k$ times, hold out
one fold, train on the other $k-1$, and grade on the held-out fold. Average the $k$
scores. Every point gets used for testing exactly once. Here is the whole loop in R:

```{webr}
#| autorun: false
# explore-5 — LIVE: 5-fold cross-validation, spelled out as a loop.
# PREDICT: cross-validation grades on held-out data every time. Will the LINE or the
# WIGGLY model get the smaller average CV RMSE? Commit, then run.
set.seed(42)
n <- 30
x <- sort(runif(n, 0, 10))
y <- 2 + 1.5 * x + rnorm(n, 0, 2)
dat <- data.frame(x = x, y = y)
rmse <- function(obs, pred) sqrt(mean((obs - pred)^2))

set.seed(42)
k <- 5
fold <- sample(rep(1:k, length.out = n))   # give each row a fold label 1..5

cv_rmse <- function(degree) {
  errs <- numeric(k)
  for (f in 1:k) {                         # each fold takes a turn as the test set
    tr <- dat[fold != f, ]                 # train on the other k-1 folds
    te <- dat[fold == f, ]                 # grade on THIS fold only
    m  <- lm(y ~ poly(x, degree), data = tr)
    errs[f] <- rmse(te$y, predict(m, te))
  }
  mean(errs)                               # average the k held-out scores
}

c(line = cv_rmse(1), wiggly = cv_rmse(9))  # 2.433 vs 6.457: the simple line wins
```

## Common traps

**Trap 1 — "higher R² (or lower training error) always means a better model."** It is
tempting because on the *training* data more flexibility always drives error down: our
wiggly model had the smaller training RMSE (1.528 vs the line's 1.985), and a curve
flexible enough will hit every point for a training R² of 1. But that same wiggly model
scored a **test** RMSE of 7.367 against the line's 2.886 — more than twice as wrong on
data it hadn't seen. A training score you keep pushing up by adding flexibility measures
*memorization*, not skill. Always judge on held-out data.

**Trap 2 — "R² = 0.5 means the model got half the predictions right."** R² is not an
accuracy percentage and has nothing to do with counting correct answers. It is one minus
the ratio of *leftover squared error* to *total squared spread*: $R^2 = 1-\text{RSS}/
\text{TSS}$. In the toy data, $R^2 = 0.92$ means the model removed 92% of the *variation*
in $y$ (dropped RSS from 50 down to 4), not that 92% of items were "correct." And because
a bad model can leave *more* error than the mean-guesser, R² can go **below zero** —
impossible for any "percent correct."

## Check yourself

**1.** A model predicted `yhat <- c(2, 4, 5)` for observed `y <- c(1, 4, 7)`. Compute
the residuals, then RSS and RMSE by hand, and check in R.

```{webr}
#| autorun: false
# check1-starter — edit and Run to check your hand answer
y    <- c(1, 4, 7)
yhat <- c(2, 4, 5)
```

::: {.callout-note collapse="true" title="Answer"}
Residuals = observed − predicted: $1-2=-1$, $4-4=0$, $7-5=2$, so $(-1, 0, 2)$.
$\text{RSS} = (-1)^2 + 0^2 + 2^2 = 1 + 0 + 4 = 5$. $\text{MSE} = 5/3 \approx 1.667$, so
$\text{RMSE} = \sqrt{5/3} \approx 1.291$. In R, `sum((y-yhat)^2)` returns 5 and
`sqrt(mean((y-yhat)^2))` returns about 1.291.
:::

**2.** A model scores training R² = 0.99 but test R² = 0.20. In one line: what is
happening, and which number do you trust for how it will behave on new data?

::: {.callout-note collapse="true" title="Answer"}
The model is **overfitting** — it memorized the training data (R² near 1 there) but
generalizes poorly (R² only 0.20 on unseen data). Trust the **test** R², 0.20: that is
the honest estimate of how it behaves on data it did not study.
:::

**3.** In your own words: explain to a colleague, in two sentences, what R² measures and
why a great R² on the training data can still be bad news.

::: {.callout-note collapse="true" title="A model answer"}
R² measures the fraction of the data's total spread that the model explains, compared to
the do-nothing baseline of just predicting the mean for everyone (1 = perfect, 0 = no
better than the mean). A great *training* R² can be bad news because a flexible enough
model can drive training error to nearly zero by memorizing the noise, so it looks perfect
on the data it studied yet falls apart on new data — which is why you always check the
score on a held-out set.
:::

## The bias–variance trade-off (in words) {#bias-variance}

One last idea ties the whole lesson together, and for this module we keep it entirely in
words. Every model's expected error on new data splits into two competing pieces:

- **Bias** is error from the model being *too rigid* to capture the real pattern — the
  underfitting side. Our flat mean-line has high bias: it is systematically wrong because
  a horizontal line simply cannot follow a rising trend, no matter what data you feed it.
- **Variance** here means error from the model being *too sensitive* to the particular
  data it was trained on — the overfitting side. Our wiggly curve has high variance:
  reshuffle the points and it would draw a wildly different curve, because it is chasing
  noise that changes from sample to sample.

The **trade-off**: making a model more flexible lowers bias but raises variance, and
making it simpler does the reverse. You cannot drive both to zero at once, so the sweet
spot — the model that generalizes best — is a *balance*, not an extreme. That balance is
exactly what a low **cross-validation** error points you toward: in our run the flexible
degree-9 curve had low bias but its variance blew up (CV RMSE 6.457), while the straight
line balanced the two and won (CV RMSE 2.433). (Depth note: putting a *number* on bias and
variance separately is beyond this module — here the trade-off is a mental model for
*why* the middle wins.)

## What you can now do

- **Compute a residual, RSS, MSE, and RMSE by hand** from observed and predicted values,
  and reproduce them with `sum((y-yhat)^2)`, `mean(...)`, and `sqrt(...)`.
- **Compute and interpret R²** as "fraction of spread explained versus guessing the
  mean," knowing that 1 is perfect, 0 is no better than the mean, and negative is worse.
- **Tell overfitting from underfitting** — memorizing noise vs missing the trend — and
  picture each against a scatter of points.
- **Run a train/test split** (`set.seed(42)`, `sample()`) and a **k-fold
  cross-validation loop** in base R, and explain why a held-out score beats a training
  score for judging a model.
- **State the bias–variance trade-off in plain words** and say why the best-generalizing
  model is a balance, not the most flexible one.

If any of these feels shaky, revisit [A tiny example first](#a-tiny-example-first) and
the [Explore it](#explore-it) split/CV cells before continuing — anything that fits a
model and reports how good it is assumes all of these.

## Where this goes next

Every method that produces predictions is graded with these scores, and every method
flexible enough to overfit is defended with a held-out check like train/test or
cross-validation — so these tools reappear the moment a course fits anything to data.
Ready for more reps? Do the [practice problems](practice.qmd).
