---
title: "Practice: How wrong is the model, really?"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only (a {webr} starter does not see it — each starter re-states its
# own data). Base R is all we need, so nothing to install.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Do them on paper or in an R session; only
then open the answer.

## Problem 1 — the toy example, new numbers

A model produced these observed and predicted values:

| observed $y$ | 10 | 12 | 15 | 14 | 19 |
|---|---|---|---|---|---|
| predicted $\hat{y}$ | 11 | 12 | 14 | 16 | 17 |

By hand, compute the residuals, then RSS, MSE, and RMSE. Then check with R.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
y    <- c(10, 12, 15, 14, 19)
yhat <- c(11, 12, 14, 16, 17)
```

::: {.callout-note collapse="true" title="Answer"}
Residuals = observed − predicted: $10-11=-1$, $12-12=0$, $15-14=1$, $14-16=-2$,
$19-17=2$, so $(-1, 0, 1, -2, 2)$.
$\text{RSS} = (-1)^2 + 0^2 + 1^2 + (-2)^2 + 2^2 = 1+0+1+4+4 = 10$.
$\text{MSE} = 10/5 = 2$, and $\text{RMSE} = \sqrt{2} \approx 1.414$.

```{r p1-answer}
y    <- c(10, 12, 15, 14, 19)
yhat <- c(11, 12, 14, 16, 17)

resid <- y - yhat
resid                       # -1  0  1 -2  2
RSS  <- sum(resid^2)        # 10
c(RSS = RSS, MSE = RSS/length(y), RMSE = sqrt(RSS/length(y)))   # 10, 2, 1.414
```
:::

## Problem 2 — fill in the missing step (R²)

Here is R² worked out for the Problem-1 data — but the **core step is blanked.** Fill in
ONLY the blanked line: computing TSS, the baseline "guess the mean" error.

> The mean is $\bar{y} = (10+12+15+14+19)/5 = 70/5 = 14$.
> **`<your step here>`** — TSS is the sum of squared distances of each $y$ from $\bar{y}$:
> $\text{TSS} = \underline{\phantom{0000}}$.
> Then, using $\text{RSS} = 10$ from Problem 1, $R^2 = 1 - \text{RSS}/\text{TSS}$.

```{webr}
#| autorun: false
# p2-starter — edit and Run to check your TSS and the resulting R2
y <- c(10, 12, 15, 14, 19)
```

::: {.callout-note collapse="true" title="Answer"}
The blanked step is TSS — the spread of $y$ around its own mean of 14:

$$
\text{TSS} = (10-14)^2 + (12-14)^2 + (15-14)^2 + (14-14)^2 + (19-14)^2
= 16 + 4 + 1 + 0 + 25 = 46.
$$

Then $R^2 = 1 - \text{RSS}/\text{TSS} = 1 - 10/46 \approx 0.783$ — the model removed about
78% of the raw spread. TSS is the move because R² is always "leftover error *relative to*
the mean-guesser's error," and TSS is that baseline error.

```{r p2-answer}
y   <- c(10, 12, 15, 14, 19)
TSS <- sum((y - mean(y))^2)     # 46
c(TSS = TSS, R2 = 1 - 10 / TSS) # 46, 0.7826...
```
:::

## Problem 3 — interpretation, not computation

Two models are fit to the same data. Model A has training R² = 0.95 and test R² = 0.55.
Model B has training R² = 0.80 and test R² = 0.78. Which model would you deploy on new
data, and why? (No arithmetic — reason it out.)

::: {.callout-note collapse="true" title="Answer"}
Deploy **Model B.** What matters for new data is the **test** score, and B's is higher
(0.78 vs 0.55). Model A's huge gap between training (0.95) and test (0.55) is the
signature of **overfitting** — it memorized the training data but generalizes poorly.
Model B's training and test scores are close (0.80 vs 0.78), which says it learned the
real pattern rather than the noise, so it should hold up on data it hasn't seen. A high
training R² is not the goal; a high *held-out* R² is.
:::

## Problem 4 — spot the error

A colleague measured a model's error like this — it contains **exactly one mistake.**
Find it and fix it.

> "Observed `y <- c(4, 7, 9, 10)`, predicted `yhat <- c(5, 6, 9, 12)`. The residuals are
> $-1, 1, 0, -2$. To get the total error I add the residuals: $-1 + 1 + 0 - 2 = -2$. So
> the total error is $-2$, and the model is doing pretty well since that's close to
> zero."

```{webr}
#| autorun: false
# p4-starter — copy this to check the right way
y    <- c(4, 7, 9, 10)
yhat <- c(5, 6, 9, 12)
```

::: {.callout-note collapse="true" title="Answer"}
The mistake is **adding the raw residuals** instead of squaring them first. Raw residuals
cancel — here $-1+1+0-2 = -2$, which is near zero *not because the model is accurate but
because the misses offset each other* (a model that was off by $-1000$ and $+998$ would
also sum near zero). The fix is RSS: **square each residual, then add.**
$\text{RSS} = (-1)^2 + 1^2 + 0^2 + (-2)^2 = 1+1+0+4 = 6$, so $\text{RMSE} =
\sqrt{6/4} \approx 1.225$ — the honest "typical miss." A negative "total error" should
have been the giveaway: an error size can't be negative. This is the cancellation trap
the lesson opens with (the warm-up), and the whole reason RSS squares before adding.

```{r p4-answer}
y    <- c(4, 7, 9, 10)
yhat <- c(5, 6, 9, 12)
resid <- y - yhat
sum(resid)                       # -2  <- the misleading (wrong) number
sum(resid^2)                     # 6   <- RSS: square first, THEN add
sqrt(mean(resid^2))              # 1.225  <- RMSE
```
:::

## Problem 5 — mixing it with variance *(uses [mean-variance-covariance](../mean-variance-covariance/lesson.qmd)'s variance)*

TSS is secretly the sample variance in disguise. For `y <- c(10, 12, 15, 14, 19)` from
Problem 1, show that $\text{TSS} = (n-1)\times \texttt{var(y)}$, where `var()` is R's
sample variance (the one that divides the summed squared deviations by $n-1$).

**PREDICT** first: `var(y)` will be TSS divided by what number? Write it down, then run.

```{webr}
#| autorun: false
# p5-starter — edit and Run to solve, right here
y <- c(10, 12, 15, 14, 19)
```

::: {.callout-note collapse="true" title="Answer"}
From [the mean, variance & covariance](../mean-variance-covariance/lesson.qmd), the sample
variance is the sum of squared deviations from the mean, divided by $n-1$:
$\texttt{var(y)} = \frac{1}{n-1}\sum (y_i - \bar{y})^2$. But that sum *is* TSS! So
$\text{TSS} = (n-1)\times\texttt{var(y)}$. Here $n=5$, so `var(y)` = TSS / 4 = 46 / 4 =
11.5, and $(n-1)\times\texttt{var(y)} = 4 \times 11.5 = 46 = \text{TSS}$. The fit score
TSS and the spread measure variance are the same summed-squared-deviations idea, just
divided by different things.

```{r p5-answer}
y   <- c(10, 12, 15, 14, 19)
TSS <- sum((y - mean(y))^2)      # 46
n   <- length(y)
c(TSS = TSS, var_y = var(y), reconstructed = (n - 1) * var(y))  # 46, 11.5, 46
```
:::

## Problem 6 — transfer: write a cross-validation loop

New context. You have a small dataset and want to estimate how a straight-line model will
do on unseen data using **3-fold cross-validation.** Fill in the loop: for each fold,
train on the other folds and grade on the held-out fold, then average the three RMSEs.
Use `set.seed(42)` exactly as shown so your numbers match.

```{webr}
#| autorun: false
# p6-starter — complete the loop body (the two data lines and the model)
set.seed(42)
d <- data.frame(x = 1:9, y = c(2, 3, 5, 4, 6, 7, 6, 9, 10))
rmse <- function(obs, pred) sqrt(mean((obs - pred)^2))

set.seed(42)
k <- 3
fold <- sample(rep(1:k, length.out = nrow(d)))   # fold label for each row

errs <- numeric(k)
for (f in 1:k) {
  # your job — the three lines the loop needs:
  #   tr <- ...   # train: the rows where fold != f
  #   te <- ...   # test:  the rows where fold == f
  #   m  <- ...   # fit lm(y ~ x) on tr
  errs[f] <- rmse(te$y, predict(m, te))   # scores fold f once tr/te/m are defined
}
mean(errs)
```

::: {.callout-note collapse="true" title="Answer"}
The core of cross-validation is the rotation: each fold takes a turn as the test set
(`te <- d[fold == f, ]`) while everything else trains the model (`tr <- d[fold != f, ]`).
You fit `lm()` on `tr`, predict onto `te`, and record that fold's held-out RMSE; averaging
the three gives one honest estimate. With this seed the fold labels come out
`1 2 3 1 2 1 2 3 3`, the three fold RMSEs are about `0.679, 1.133, 1.350`, and their
average is about **1.054**.

```{r p6-answer}
set.seed(42)
d <- data.frame(x = 1:9, y = c(2, 3, 5, 4, 6, 7, 6, 9, 10))
rmse <- function(obs, pred) sqrt(mean((obs - pred)^2))

set.seed(42)
k <- 3
fold <- sample(rep(1:k, length.out = nrow(d)))

errs <- numeric(k)
for (f in 1:k) {
  tr <- d[fold != f, ]
  te <- d[fold == f, ]
  m  <- lm(y ~ x, data = tr)
  errs[f] <- rmse(te$y, predict(m, te))
}
round(errs, 3)      # 0.679 1.133 1.350
round(mean(errs), 3)  # 1.054
```

Every point was used for testing exactly once, and no point was ever graded by a model
that had already seen it — that is what makes 1.054 an honest guess at new-data error.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/model-fit-error-and-validation" pass="4"}

::: {.quiz-q qid="mfev-r-squared" qtype="numeric" answer="0.92" tolerance="0.01"}
A model has RSS $= 4$ and TSS $= 50$. What is $R^2 = 1 - \text{RSS}/\text{TSS}$? Round to
two decimals.
:::

::: {.quiz-q qid="mfev-training-error-trap" qtype="mc" answer="b"}
A wiggly model has lower training error than a straight line but much higher *test* error.
Which is the better model?

a) The wiggly model — lower training error means a better fit
b) The straight line — a held-out (test) score measures real skill, not training fit
c) They are equally good
:::

::: {.quiz-q qid="mfev-r-squared-meaning" qtype="mc" answer="a"}
$R^2 = 0.5$ means:

a) The model removed half the *variation* in $y$ compared with guessing the mean
b) The model got half the predictions exactly right
c) The model is correct 50% of the time
:::

::: {.quiz-q qid="mfev-overfitting" qtype="mc" answer="c"}
A model that fits every wiggle in the training data — memorizing its noise — is:

a) Underfitting — it missed the trend
b) Perfectly calibrated
c) Overfitting — it memorized noise, so it generalizes poorly
:::

::: {.quiz-q qid="mfev-loss-minimize" qtype="mc" answer="b"}
(From loss functions & optimization.) A fitting method adjusts parameters to do what to
its loss function on the training data?

a) Maximize it
b) Minimize it
c) Set it to 1
:::

:::

::::
