---
title: "How it corrupts coloc"
subtitle: "The fine-mapping bias flows straight into coloc — turning a truly shared signal into a confident, wrong 'distinct' call"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live cells below re-state their own data. Base R only.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

::: {.callout-tip title="This lesson runs in your browser" collapse="true"}
The code cells below with a **Run** button are live — powered by WebR (R compiled to
WebAssembly), they run right here on your phone or laptop, no install needed. Edit any of them
and re-run. The first run downloads a small R engine, so give it a few seconds. Cells without a
Run button are ordinary listings whose output was computed when the page was built.
:::

## Where we're going

Module 1 showed fine-mapping breaking for a single trait: unequal information across SNPs biases
the "which SNP is causal?" posterior. This module is the second half of the paper's diagnosis —
the bad news that makes the whole paper necessary. **`coloc` is built on fine-mapping.** It
fine-maps each trait, then compares the two traits' per-SNP posteriors to decide whether they
share a causal variant (H4) or have distinct ones (H3). So the fine-mapping bias flows straight
in — and it can flip a truly *shared* signal into a **confident but wrong** *distinct* call
(PP3 near 1). By the end you'll make `coloc` (the one you built) do exactly that, on purpose.

**Builds on:** [Module 1 (when fine-mapping breaks)](../01-when-fine-mapping-breaks/lesson.qmd)
and the [Bayesian Colocalisation course](../../../bayesian-colocalisation/syllabus.md) (which
built `coloc_abf(...)` and the PP3-vs-PP4 split).

## Warm-up: do you remember?

Three quick reps first — produce each from memory, then check.

**1.** *(Module 1)* Two SNPs in perfect LD tag the same signal, but SNP 2 has information loss.
Which way does SNP 2's fine-mapping posterior move from 0.5, and does it get better or worse with
sample size?

::: {.callout-note collapse="true" title="Answer"}
SNP 2's posterior drifts **below** 0.5 (toward the better-measured SNP), and it gets **worse** as
$N$ grows because the Bayes factor scales with $z^2 \propto N$. See
[Module 1](../01-when-fine-mapping-breaks/lesson.qmd).
:::

**2.** *(coloc course)* In `coloc`, which posterior probability means "**shared** causal variant"
and which means "**distinct** causal variants"?

::: {.callout-note collapse="true" title="Answer"}
**PP4 = shared** (colocalisation); **PP3 = distinct** (two separate causal variants). See the
[Bayesian Colocalisation course](../../../bayesian-colocalisation/syllabus.md).
:::

**3.** *(coloc course)* In one line, what is coloc's H4 support built from — i.e. what makes PP4
large?

::: {.callout-note collapse="true" title="Answer"}
H4 support is large when the *same SNP* carries strong evidence in **both** traits — the diagonal
of the "which SNP for trait 1 × which SNP for trait 2" grid, where the two traits' Bayes factors
line up on one variant. See the
[Bayesian Colocalisation course](../../../bayesian-colocalisation/syllabus.md).
:::

## A tiny example first

Two SNPs in perfect LD, and a signal that is **genuinely shared** — both traits are driven by the
same underlying causal variant. Because the two SNPs are perfectly correlated, each trait's
signal shows up equally on both, so with clean data each trait fine-maps to a 50/50 split and
`coloc` correctly declares **PP4 ≈ 0.999** (shared). Truth: shared. Verdict: shared.

Now impose the paper's mischief. Trait 1 suffers information loss on **SNP 2** (e.g. 95%
genotype missingness there), so trait 1's evidence now piles onto **SNP 1**: its fine-mapping
posterior becomes $(1.0,\ 0.0)$. Independently, trait 2 suffers information loss on **SNP 1**, so
trait 2's evidence piles onto **SNP 2**: posterior $(0.0,\ 1.0)$. Each trait, fine-mapped alone,
now confidently names a *different* SNP.

`coloc` compares those two fine-maps. Trait 1 says "SNP 1," trait 2 says "SNP 2" — different
SNPs! — so coloc concludes the causal variants are **distinct** and reports **PP3 = 1.0**,
supremely confident. But the truth is unchanged: the signal is *shared*. The missingness didn't
change the biology; it changed which SNP each trait *looked* strongest at, and coloc — trusting
the fine-maps — was fooled completely.

```{r toy-example}
# The coloc_abf you built (base R), condensed.
wakefield_abf <- function(beta, se, W) { z<-beta/se; V<-se^2; r<-W/(W+V); 0.5*(log(1-r)+r*z^2) }
.logsum  <- function(x) { m <- max(x); m + log(sum(exp(x - m))) }
.logdiff <- function(x, y) { m <- max(x, y); m + log(exp(x - m) - exp(y - m)) }
coloc_abf <- function(b1, se1, b2, se2, W1, W2, p1 = 1e-4, p2 = 1e-4, p12 = 1e-5) {
  l1 <- wakefield_abf(b1, se1, W1); l2 <- wakefield_abf(b2, se2, W2)
  s1 <- .logsum(l1); s2 <- .logsum(l2); s12 <- .logsum(l1 + l2)
  all <- c(0, log(p1)+s1, log(p2)+s2, log(p1)+log(p2)+.logdiff(s1+s2, s12), log(p12)+s12)
  exp(all - .logsum(all))
}
z <- 8                                    # a strong, big-study signal
# CLEAN data: two SNPs in perfect LD, signal genuinely shared -> both traits split 0.5/0.5
b1 <- c(z*0.03, z*0.03);   se1 <- c(0.03, 0.03)
b2 <- c(z*0.035, z*0.035); se2 <- c(0.035, 0.035)
clean <- coloc_abf(b1, se1, b2, se2, 0.04, 0.04)
round(clean[4:5], 4)                       # PP3, PP4 -> 0.001, 0.999  (correctly SHARED)

# INFO LOSS: trait 1 loses SNP 2, trait 2 loses SNP 1 (95% missingness each)
b1b <- c(z*0.03,  z*0.03*sqrt(0.05));  se1b <- c(0.03,  0.03/sqrt(0.05))
b2b <- c(z*0.035*sqrt(0.05), z*0.035); se2b <- c(0.035/sqrt(0.05), 0.035)
lossy <- coloc_abf(b1b, se1b, b2b, se2b, 0.04, 0.04)
round(lossy[4:5], 4)                       # PP3, PP4 -> 1, 0  (confidently, wrongly DISTINCT)
```

## Getting a feel for it

**Analogy — two witnesses pointing at a line-up.** A crime had one culprit (the shared causal
variant). Two witnesses each point at the line-up. With good lighting, both point at the same
person — case closed, "same culprit" (PP4). Now dim the lights *differently* for each witness:
witness 1 can't see the left half of the line-up, witness 2 can't see the right half. Witness 1
points right, witness 2 points left — at *different* people — and the detective (coloc) concludes
"two different culprits" (PP3), with total confidence. The witnesses didn't disagree because
there were two culprits; they disagreed because each was handicapped differently. coloc trusts
the pointing, so it inherits the handicap.

Where the analogy breaks: coloc doesn't just take the single most-likely SNP — it weighs the
whole posterior across SNPs. But when information loss makes each trait's posterior *confident*
(a single-variant credible set), the effect is exactly as stark as two witnesses each pointing
decisively at the wrong person.

**Why does the bias flip H4 into H3 specifically?** Because H4 (shared) support comes from the
two traits' evidence *lining up on the same SNP*. Information loss that pushes trait 1's peak to
SNP 1 and trait 2's peak to SNP 2 *destroys* that alignment — the traits' evidence now sits on
different SNPs, which is the literal definition of H3. The loss doesn't just add noise; it
manufactures a fake disagreement in the exact place coloc looks for agreement.

**What do these numbers feel like?** Look at the jump: clean PP4 = 0.999, lossy PP3 = 1.0. This
is not a mild wobble — it's the *maximum possible* error, delivered with the *maximum possible*
confidence. That is the paper's alarm: coloc doesn't fail loudly (a shrug, a 50/50); it fails
*silently and confidently*, which is far more dangerous for someone about to design a follow-up
experiment on the "distinct" call.

## What the notation is going to say

There's no new formula here — it's the exact `coloc_abf` you already built. The new content is
the *causal chain*: biased per-SNP posteriors (Module 1) → misaligned peaks across the two traits
→ coloc's diagonal (shared) support collapses while its off-diagonal (distinct) support wins → PP3
near 1. The lesson traces that chain through numbers you can run.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $\text{PP}_i$ | "P-P sub eye" | posterior probability SNP $i$ is causal, for one trait | $(1, 0)$ or $(0, 1)$ after loss |
| PP3 | "P-P three" | coloc posterior of **distinct** causal variants (H3) | $1.0$ (lossy) |
| PP4 | "P-P four" | coloc posterior of a **shared** causal variant (H4) | $0.999$ (clean) |
| $z$ | "zed" / "zee" | a SNP's z-score, $\hat\beta/\text{se}$; here the same big signal on both SNPs | $8$ |
| $r^2$ | "r squared" | data completeness / imputation accuracy (here $0.05$ = 95% missing) | $0.05$ on the lossy SNP |
| $W$ | "double-you" | coloc's prior weight | $0.04$ |

## The formal version

The mechanism, stated plainly (no new equation — the equation is `coloc_abf`):

$$
\underbrace{\text{unequal info across SNPs}}_{\text{Module 1}}
\;\Rightarrow\;
\underbrace{\text{biased per-SNP } \text{PP}_i \text{ (peaks move)}}_{\text{each trait alone}}
\;\Rightarrow\;
\underbrace{\text{traits' peaks misalign}}_{\text{H4 support collapses}}
\;\Rightarrow\;
\underbrace{\text{PP3} \to 1}_{\text{confident, wrong}} .
$$

Reading it aloud: "unequal information across SNPs leads to biased per-SNP posteriors whose peaks
move, which makes the two traits' peaks misalign, which collapses coloc's shared-variant support
and drives the distinct-variant posterior PP3 toward 1." Everyday words: "break each trait's
fine-map differently, and coloc reads the resulting disagreement as two separate causes."

**Mapping it onto the toy example.** Let's expose the middle of the chain — the per-SNP
posteriors — to see the peaks move:

```{r formal-check}
wakefield_abf <- function(beta, se, W) { z<-beta/se; V<-se^2; r<-W/(W+V); 0.5*(log(1-r)+r*z^2) }
pp <- function(l) { b <- exp(l - max(l)); round(b/sum(b), 3) }
z <- 8

# CLEAN: each trait splits 0.5/0.5 across the two perfect-LD SNPs
l1c <- wakefield_abf(c(z*0.03, z*0.03), c(0.03, 0.03), 0.04)
l2c <- wakefield_abf(c(z*0.035, z*0.035), c(0.035, 0.035), 0.04)
rbind(trait1 = pp(l1c), trait2 = pp(l2c))     # (0.5, 0.5) and (0.5, 0.5) -> aligned -> H4

# LOSSY: trait1 peaks on SNP1, trait2 peaks on SNP2 -> misaligned -> H3
l1l <- wakefield_abf(c(z*0.03, z*0.03*sqrt(0.05)), c(0.03, 0.03/sqrt(0.05)), 0.04)
l2l <- wakefield_abf(c(z*0.035*sqrt(0.05), z*0.035), c(0.035/sqrt(0.05), 0.035), 0.04)
rbind(trait1 = pp(l1l), trait2 = pp(l2l))     # (1, 0) and (0, 1) -> misaligned -> H3
```

The clean per-SNP posteriors are aligned (both 0.5/0.5) → coloc's diagonal support wins → PP4.
The lossy ones are misaligned ($(1,0)$ vs $(0,1)$) → the diagonal is empty, the off-diagonal wins
→ PP3. That misalignment is the entire bug.

## Explore it

```{webr}
#| autorun: true
# TRY IT: dial the missingness. imp = 1 is clean; smaller imp = more info loss. PREDICT: at what
# imp does coloc flip from PP4 (shared) to PP3 (distinct)?
wakefield_abf <- function(beta, se, W) { z<-beta/se; V<-se^2; r<-W/(W+V); 0.5*(log(1-r)+r*z^2) }
.logsum  <- function(x) { m <- max(x); m + log(sum(exp(x - m))) }
.logdiff <- function(x, y) { m <- max(x, y); m + log(exp(x - m) - exp(y - m)) }
coloc_abf <- function(b1, se1, b2, se2, W1, W2, p1=1e-4, p2=1e-4, p12=1e-5) {
  l1<-wakefield_abf(b1,se1,W1); l2<-wakefield_abf(b2,se2,W2)
  s1<-.logsum(l1); s2<-.logsum(l2); s12<-.logsum(l1+l2)
  all<-c(0, log(p1)+s1, log(p2)+s2, log(p1)+log(p2)+.logdiff(s1+s2,s12), log(p12)+s12)
  exp(all-.logsum(all))
}
z <- 8; imp <- 0.05                       # <- change me: 1 = clean, 0.05 = severe loss
b1 <- c(z*0.03,  z*0.03*sqrt(imp));  se1 <- c(0.03,  0.03/sqrt(imp))
b2 <- c(z*0.035*sqrt(imp), z*0.035); se2 <- c(0.035/sqrt(imp), 0.035)
pp <- coloc_abf(b1, se1, b2, se2, 0.04, 0.04)
cat("imp =", imp, " -> PP3(distinct) =", round(pp[4], 4),
    " PP4(shared) =", round(pp[5], 4), "\n")
```

```{webr}
#| autorun: true
#| fig-alt: "Two curves against the data-completeness level: PP4 (shared) starts near 1 at full completeness and falls to 0 as completeness drops, while PP3 (distinct) rises from 0 to 1 — the confident wrong call taking over as information is lost."
# TRY IT: watch the crossover. As data completeness falls, PP4 (correct) collapses and PP3
# (wrong) takes over -- both confidently. PREDICT: do the two curves cross near the middle?
wakefield_abf <- function(beta, se, W) { z<-beta/se; V<-se^2; r<-W/(W+V); 0.5*(log(1-r)+r*z^2) }
.logsum  <- function(x) { m <- max(x); m + log(sum(exp(x - m))) }
.logdiff <- function(x, y) { m <- max(x, y); m + log(exp(x - m) - exp(y - m)) }
coloc_abf <- function(b1, se1, b2, se2, W1, W2, p1=1e-4, p2=1e-4, p12=1e-5) {
  l1<-wakefield_abf(b1,se1,W1); l2<-wakefield_abf(b2,se2,W2)
  s1<-.logsum(l1); s2<-.logsum(l2); s12<-.logsum(l1+l2)
  all<-c(0, log(p1)+s1, log(p2)+s2, log(p1)+log(p2)+.logdiff(s1+s2,s12), log(p12)+s12)
  exp(all-.logsum(all))
}
z <- 8; imps <- seq(1, 0.02, length.out = 60)
res <- sapply(imps, function(imp) {
  b1 <- c(z*0.03, z*0.03*sqrt(imp)); se1 <- c(0.03, 0.03/sqrt(imp))
  b2 <- c(z*0.035*sqrt(imp), z*0.035); se2 <- c(0.035/sqrt(imp), 0.035)
  coloc_abf(b1, se1, b2, se2, 0.04, 0.04)[4:5]
})
plot(imps, res[2, ], type = "l", lwd = 2, col = "steelblue", ylim = c(0, 1),
     xlim = c(1, 0.02), xlab = "data completeness (1 = clean)", ylab = "posterior",
     main = "A shared signal, called distinct as information is lost")
lines(imps, res[1, ], lwd = 2, col = "tomato")
legend("right", legend = c("PP4 shared (correct)", "PP3 distinct (WRONG)"),
       col = c("steelblue", "tomato"), lwd = 2, bty = "n")
```

## Common traps

**Trap: "a confident coloc result (PP3 or PP4 near 1) must be correct."** Tempting — confidence
usually tracks reliability. But our lossy case reports **PP3 = 1.0**, maximally confident, and is
completely wrong (the truth is shared). Information-driven bias produces *confident* errors, not
hesitant ones, precisely because it sharpens each trait's fine-map onto a single (wrong-for-the-
comparison) SNP. A posterior near 1 is not a safety guarantee.

**Trap: "the problem is just noise, so more data fixes it."** No — as Module 1 showed, the bias
*grows* with sample size. Our toy uses a strong signal ($z = 8$, a big study) *because* that's
when the misalignment is sharpest. Adding data makes each trait's biased fine-map more confident,
which makes coloc's wrong PP3 more confident, not less. This is a *systematic* error from unequal
information, not random noise.

## Check yourself

**1.** In the lossy toy, trait 1's per-SNP posterior is $(1, 0)$ and trait 2's is $(0, 1)$. In
one line, why does that combination drive PP3 (distinct) rather than PP4 (shared)?

::: {.callout-note collapse="true" title="Answer"}
Because H4 (shared) needs both traits' evidence on the **same** SNP; here trait 1 points at SNP 1
and trait 2 at SNP 2 — different SNPs — which is exactly the H3 (distinct) configuration, so
coloc's off-diagonal support wins and PP3 → 1.
:::

**2.** True or false: the information loss in the lossy toy changed the underlying biology (there
really are two distinct causal variants now). Explain.

::: {.callout-note collapse="true" title="Answer"}
**False.** The two SNPs are in perfect LD and the signal is genuinely shared throughout — the
biology never changed. Only the *measurement* changed (differential missingness), which moved
where each trait *appeared* strongest. coloc's PP3 = 1 is an artifact of the measurement, not a
real second causal variant.
:::

**3.** In your own words: explain to a colleague, in two sentences, how a fine-mapping bias in a
single trait ends up producing a confident, wrong `coloc` call.

::: {.callout-note collapse="true" title="A model answer"}
coloc decides "shared vs distinct" by checking whether the two traits' fine-mapped evidence lands
on the same SNP, so if unequal information pushes each trait's peak onto a *different* SNP, coloc
sees a disagreement and calls the variants distinct (PP3). Because the biased fine-maps are
themselves confident (single-variant credible sets), coloc inherits that confidence and reports a
near-certain but wrong answer.
:::

## What you can now do

- **Explain why coloc inherits the fine-mapping bias** — it fine-maps each trait and compares the
  results, so biased per-SNP posteriors flow straight into the H3-vs-H4 decision.
- **Trace the causal chain** from unequal information → moved peaks → misaligned traits → PP3
  near 1.
- **Reproduce the failure** in base R: a genuinely shared signal that your `coloc_abf` calls
  confidently distinct once you impose differential information loss.
- **State the danger** — the error is *confident* (posterior near 1) and *worsens with sample
  size*, which is why a new, fine-mapping-free test is needed.

If any of these feels shaky, revisit the toy example and the misalignment picture before
continuing — the next module introduces the fix.

## Where this goes next

You've now seen the full disease: fine-mapping bias (Module 1) corrupting coloc into confident,
wrong calls (this module). The cure is to test colocalisation **without fine-mapping at all**.
The next module,
[Proportionality is colocalisation](../03-proportionality-is-colocalisation/lesson.qmd), reframes
"do these traits share a signal?" as "are their effect estimates *proportional*?" — a question
you can answer straight from the summary statistics, with no per-SNP posterior to bias.

Ready for more reps? Do the [practice problems](practice.qmd).
