---
title: "Practice: How it corrupts coloc"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

Work these in order — they get harder. Do them on paper or in an R session; only then open the
answer. Several problems reuse the condensed `coloc_abf` you built; the starters include it.

## Problem 1 — the toy example, new numbers

Repeat the clean-vs-lossy comparison with a signal of $z = 6$ (a smaller study). Compute the
clean PP3/PP4 and the lossy PP3/PP4 (trait 1 loses SNP 2, trait 2 loses SNP 1, imputation
$r^2 = 0.05$).

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
wakefield_abf <- function(beta, se, W) { z<-beta/se; V<-se^2; r<-W/(W+V); 0.5*(log(1-r)+r*z^2) }
.logsum  <- function(x) { m <- max(x); m + log(sum(exp(x - m))) }
.logdiff <- function(x, y) { m <- max(x, y); m + log(exp(x - m) - exp(y - m)) }
coloc_abf <- function(b1, se1, b2, se2, W1, W2, p1=1e-4, p2=1e-4, p12=1e-5) {
  l1<-wakefield_abf(b1,se1,W1); l2<-wakefield_abf(b2,se2,W2)
  s1<-.logsum(l1); s2<-.logsum(l2); s12<-.logsum(l1+l2)
  all<-c(0, log(p1)+s1, log(p2)+s2, log(p1)+log(p2)+.logdiff(s1+s2,s12), log(p12)+s12)
  exp(all-.logsum(all))
}
z <- 6; imp <- 0.05
```

::: {.callout-note collapse="true" title="Answer"}
```{r p1-ans}
wakefield_abf <- function(beta, se, W) { z<-beta/se; V<-se^2; r<-W/(W+V); 0.5*(log(1-r)+r*z^2) }
.logsum  <- function(x) { m <- max(x); m + log(sum(exp(x - m))) }
.logdiff <- function(x, y) { m <- max(x, y); m + log(exp(x - m) - exp(y - m)) }
coloc_abf <- function(b1, se1, b2, se2, W1, W2, p1=1e-4, p2=1e-4, p12=1e-5) {
  l1<-wakefield_abf(b1,se1,W1); l2<-wakefield_abf(b2,se2,W2)
  s1<-.logsum(l1); s2<-.logsum(l2); s12<-.logsum(l1+l2)
  all<-c(0, log(p1)+s1, log(p2)+s2, log(p1)+log(p2)+.logdiff(s1+s2,s12), log(p12)+s12)
  exp(all-.logsum(all))
}
z <- 6; imp <- 0.05
clean <- coloc_abf(c(z*0.03,z*0.03), c(0.03,0.03), c(z*0.035,z*0.035), c(0.035,0.035), 0.04, 0.04)
lossy <- coloc_abf(c(z*0.03, z*0.03*sqrt(imp)), c(0.03, 0.03/sqrt(imp)),
                   c(z*0.035*sqrt(imp), z*0.035), c(0.035/sqrt(imp), 0.035), 0.04, 0.04)
rbind(clean = round(clean[4:5], 4), lossy = round(lossy[4:5], 4))  # cols: PP3, PP4
```
Clean → PP4 near 1 (correctly shared); lossy → PP3 near 1 (confidently, wrongly distinct). Even
at $z = 6$ the flip is decisive.
:::

## Problem 2 — fill in the missing step

The causal chain that flips H4 to H3:

- Step 1: unequal information across SNPs biases each trait's per-SNP posterior (Module 1).
- Step 2: **your step here** — what does that do to *where each trait's peak sits*, and how does
  that affect coloc's shared (H4) support?
- Step 3: with the diagonal (shared) support gone, the off-diagonal wins → PP3 → 1.

Fill in step 2.

::: {.callout-note collapse="true" title="Answer"}
Step 2: the biased posteriors move each trait's **peak onto a different SNP** (trait 1 → SNP 1,
trait 2 → SNP 2). H4 (shared) support comes from both traits' evidence lining up on the *same*
SNP — the diagonal of the SNP × SNP grid — so moving the peaks apart **empties the diagonal** and
collapses H4 support. That's the move that hands the win to H3.
:::

## Problem 3 — expose the per-SNP peaks

For the lossy toy at $z = 8$, $\text{imp} = 0.05$, compute each trait's per-SNP posterior (not
coloc's PP3/PP4 — the individual fine-maps) and confirm they are misaligned.

```{webr}
#| autorun: false
# p3-starter — your job: compute pp() for each trait's log Bayes factors
wakefield_abf <- function(beta, se, W) { z<-beta/se; V<-se^2; r<-W/(W+V); 0.5*(log(1-r)+r*z^2) }
pp <- function(l) { b <- exp(l - max(l)); round(b/sum(b), 3) }
z <- 8; imp <- 0.05
```

::: {.callout-note collapse="true" title="Answer"}
```{r p3-ans}
wakefield_abf <- function(beta, se, W) { z<-beta/se; V<-se^2; r<-W/(W+V); 0.5*(log(1-r)+r*z^2) }
pp <- function(l) { b <- exp(l - max(l)); round(b/sum(b), 3) }
z <- 8; imp <- 0.05
l1 <- wakefield_abf(c(z*0.03, z*0.03*sqrt(imp)), c(0.03, 0.03/sqrt(imp)), 0.04)
l2 <- wakefield_abf(c(z*0.035*sqrt(imp), z*0.035), c(0.035/sqrt(imp), 0.035), 0.04)
rbind(trait1 = pp(l1), trait2 = pp(l2))
```
Trait 1 → $(1, 0)$, trait 2 → $(0, 1)$: each confidently names a *different* SNP. That
misalignment is what coloc reads as "distinct causal variants."
:::

## Problem 4 — spot the error

A colleague reports a colocalisation analysis — it contains **exactly one mistake** in the
reasoning. Find it.

> "coloc gave PP3 = 0.98, so we're 98% sure the two traits have distinct causal variants. The
> posterior is so high, and our sample is large, so there's no way this is an artifact of data
> quality."

::: {.callout-note collapse="true" title="Answer"}
The mistake is treating the high posterior + large sample as *proof against* an artifact — it's
the opposite. Information-driven bias produces **confident** wrong calls, and it **worsens with
sample size**, so a PP3 of 0.98 in a large study is *more* susceptible to being a differential-
information artifact, not less. The high posterior is not evidence the call is real; you'd want to
check data quality (imputation, missingness, rounding) and, ideally, a fine-mapping-free test.
:::

## Problem 5 — mixing it with Module 1 *(uses Module 1's posterior drift)*

Explain, using Module 1's result, *why* imposing information loss on trait 2's SNP 1 makes trait
2's fine-mapping peak move to SNP 2 — and connect that to this module's PP3 flip.

::: {.callout-note collapse="true" title="Answer"}
Module 1 showed that information loss on a SNP shrinks its Bayes factor, dragging its
fine-mapping posterior below its fair share — so information loss on trait 2's SNP 1 shrinks SNP
1's Bayes factor for trait 2, and the normalisation shifts trait 2's posterior mass onto SNP 2
(its now-better-measured neighbour in perfect LD). This module then adds the coloc step: with
trait 1's peak still on SNP 1 and trait 2's peak now on SNP 2, the two traits' peaks are
misaligned, coloc's shared support empties, and PP3 → 1.
:::

## Problem 6 — transfer: the paper's headline number

The paper reports that when the causal variant is truly *shared* but genotype missingness and LD
are severe, coloc can falsely call H3 (distinct) up to **100%** of the time. In one or two
sentences, interpret what "100% false H3" means and why it is more alarming than, say, "coloc is
noisy."

::: {.callout-note collapse="true" title="Answer"}
"100% false H3" means that across repeated simulations of a genuinely shared signal, coloc called
it *distinct* every single time — a systematic, reproducible error, not occasional bad luck. It's
more alarming than noise because noise averages out and looks uncertain (posteriors near 0.5),
whereas this failure is *confident and consistent* (PP3 near 1 every time), so it would reliably
send an investigator to design the wrong follow-up experiment.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**, once the
material isn't fresh. A pass then is real evidence you've retained it. Five questions, graded on
the page; 4/5 or better means this module is solid. Results are saved in this browser only.

::: {.module-check module="courses/proportional-colocalisation/modules/02-how-it-corrupts-coloc" pass="4"}

::: {.quiz-q qid="hicc-why-inherits" qtype="mc" answer="b"}
Why does coloc inherit the fine-mapping bias?

a) coloc uses a different prior than fine-mapping
b) coloc fine-maps each trait and compares the results, so biased per-SNP posteriors flow straight into the H3-vs-H4 decision
c) coloc ignores standard errors entirely
d) coloc only works on one trait at a time
:::

::: {.quiz-q qid="hicc-flip-direction" qtype="mc" answer="c"}
In the lossy toy, a truly **shared** signal is called:

a) H0 (no association)
b) H4 (shared), correctly
c) H3 (distinct), with high confidence
d) undetermined, with all posteriors near 0.2
:::

::: {.quiz-q qid="hicc-misalign" qtype="mc" answer="a"}
Trait 1's per-SNP posterior is $(1, 0)$ and trait 2's is $(0, 1)$. This drives PP3 (distinct)
because:

a) H4 needs both traits' evidence on the same SNP, and here they sit on different SNPs
b) the posteriors don't sum to 1
c) coloc always prefers PP3 when posteriors are extreme
d) the prior weight $W$ was set too low
:::

::: {.quiz-q qid="hicc-confidence" qtype="mc" answer="c"}
A coloc posterior near 1 (e.g. PP3 = 0.99) from a large study:

a) is guaranteed correct because of the large sample
b) is guaranteed correct because the posterior is so high
c) can still be a confident artifact of differential information, which worsens with sample size
d) means the two traits have no association
:::

::: {.quiz-q qid="hicc-drift-interleave" qtype="mc" answer="b"}
*(uses Module 1)* Imposing information loss on trait 2's SNP 1 moves trait 2's fine-mapping peak
to SNP 2 because:

a) SNP 2 has a larger prior weight
b) the loss shrinks SNP 1's Bayes factor, so normalisation shifts posterior mass to SNP 2
c) SNP 1 is removed from the region
d) the two SNPs are no longer in LD
:::

:::

::::
