---
title: "Bayes factors and posterior odds"
subtitle: "Bayes' rule as a multiplication on odds: prior odds times a Bayes factor gives posterior odds"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live cells below re-state their own data. Base R only.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

::: {.callout-tip title="This lesson runs in your browser" collapse="true"}
The code cells below with a **Run** button are live — powered by WebR (R compiled to
WebAssembly), they run right here on your phone or laptop, no install needed. Edit any
of them and re-run. The first run downloads a small R engine, so give it a few seconds.
Cells without a Run button are ordinary listings whose output was computed when the
page was built.
:::

## Where we're going

You've met Bayes' rule in *probability* form: posterior $\propto$ prior $\times$
likelihood. This lesson rewrites the same rule in a cleaner, more usable currency —
**odds** — where it becomes a plain multiplication: **posterior odds = prior odds ×
Bayes factor**. The **Bayes factor** is the star: it is exactly the ratio of the two
models' evidences (the marginal likelihoods from the previous foundation), and it
measures *how much the data alone shifts your belief* from one model toward the other,
with no prior mixed in. This "belief in, evidence multiplier, belief out" shape is the
backbone of Bayesian model comparison — and of the applied Bayes-factor methods this
foundation feeds.

We build it on two coin models, then read off the general recipe: turn a probability into
odds, multiply by the Bayes factor, turn the answer back into a probability.

**Builds on:**
[marginal likelihood & model evidence](../marginal-likelihood-and-model-evidence/lesson.qmd)
— a model's evidence $P(D)$ is its likelihood averaged over its prior;
[probability, odds & log-odds](../probability-and-odds/lesson.qmd) — odds are $p/(1-p)$,
and you can convert freely between probability and odds; and
[priors, posteriors & updating](../priors-posteriors-and-updating/lesson.qmd) — Bayes'
rule in probability form, posterior $\propto$ prior $\times$ likelihood.

## Warm-up: do you remember?

Before the new stuff, three quick reps — answers from memory first, then check.

**1.** A probability is $p = 0.75$. What are the corresponding **odds**? State the rule.

::: {.callout-note collapse="true" title="Answer"}
Odds $= p/(1-p) = 0.75/0.25 = 3$ — "3 to 1 on." The rule: odds are the probability of
the event divided by the probability of its complement. Shaky? Revisit
[probability, odds & log-odds](../probability-and-odds/lesson.qmd).
:::

**2.** In one line: what is a model's **evidence** (marginal likelihood), and how do you
get it from a likelihood and a prior?

::: {.callout-note collapse="true" title="Answer"}
A model's evidence is how probable the data is under that model, *averaged over its
prior* — multiply each parameter value's likelihood by its prior weight and sum (or
integrate). See
[marginal likelihood & model evidence](../marginal-likelihood-and-model-evidence/lesson.qmd).
:::

**3.** Recall Bayes' rule in probability form (posterior $\propto$ prior $\times$
likelihood). If the prior on a hypothesis is small but the likelihood of the data under
it is large, does the posterior go up or down relative to the prior?

::: {.callout-note collapse="true" title="Answer"}
**Up** — a large likelihood pulls the posterior above the prior; the data supports the
hypothesis, so belief rises. (How *far* up depends on how the rival hypotheses fit too —
which is exactly what the Bayes factor packages.) See
[priors, posteriors & updating](../priors-posteriors-and-updating/lesson.qmd).
:::

## A tiny example first

You flip a coin **three times** and see **heads, heads, tails** — your fixed data. You
are weighing two simple models against each other:

- **Model 1 ("fair"):** the coin is fair, $p = 0.5$.
- **Model 2 ("biased"):** the coin favors heads, $p = 0.6$.

Each model here is a single parameter value, so its **evidence** is just the likelihood
of the data at that value, $L(p) = p^2(1-p)$:

- Model 1's evidence: $L(0.5) = 0.5^2 \times 0.5 = 0.125$.
- Model 2's evidence: $L(0.6) = 0.6^2 \times 0.4 = 0.144$.

The **Bayes factor** for Model 2 over Model 1 is the ratio of their evidences:

$$
\text{BF} = \frac{0.144}{0.125} = 1.152.
$$

Read it as: "the data made Model 2 about $1.15$ times more probable than Model 1 —
a mild nudge toward the biased coin."

Now bring in what you believed *before* flipping. Suppose you thought the two models
equally likely — **prior odds of $1$** (1-to-1). Then:

$$
\text{posterior odds} = \text{prior odds} \times \text{BF} = 1 \times 1.152 = 1.152.
$$

Convert that back to a probability for Model 2: an odds of $1.152$ means a probability of
$1.152/(1 + 1.152) = 0.535$. So after seeing the data your belief in the biased coin rose
from $50\%$ to about $53.5\%$ — a small shift, because the evidence was weak.

The same thing in R, so you can check every step:

```{r toy-example}
lik <- function(p) p^2 * (1 - p)          # likelihood of H, H, T at heads-probability p

evidence_1 <- lik(0.5)                     # Model 1 (fair):   0.125
evidence_2 <- lik(0.6)                     # Model 2 (biased): 0.144

BF <- evidence_2 / evidence_1              # Bayes factor, Model 2 over Model 1
BF                                          # 1.152

prior_odds     <- 1                         # believed the two models equally likely
posterior_odds <- prior_odds * BF           # 1.152
post_prob_2    <- posterior_odds / (1 + posterior_odds)   # back to a probability
c(BF = BF, posterior_odds = posterior_odds, post_prob_M2 = post_prob_2)  # 0.535
```

## Getting a feel for it

**The mental picture: a medical test.** Think of a rare disease. Before any test, a
patient's **prior odds** of having it are tiny (say 1 in 1000). A test result is
**evidence** — and a good test comes with a number, the *likelihood ratio*: "a positive
result is 20 times more likely in a sick person than a healthy one." That likelihood
ratio is *exactly* a Bayes factor. Bayes' rule on odds says: multiply the prior odds by
that factor to get the **posterior odds**. A single positive test multiplies the odds by
20 — but $1/1000$ times $20$ is still only about $1$ in $50$, so the patient is *probably*
still healthy. The Bayes factor tells you *how strong the evidence is*; the prior odds
tell you *where you started*; the posterior odds combine them. (Where the analogy breaks:
a medical likelihood ratio compares "disease vs no disease," two fixed states, whereas a
general Bayes factor can compare two whole models each averaged over its own prior — but
the multiply-the-odds mechanics are identical.)

**Why odds, and why does the rule become a multiplication?** In probability form, Bayes'
rule carries an awkward normalizing denominator (the total evidence) so the posterior
probabilities sum to $1$. Write it as *odds* — one model's belief divided by the other's —
and that shared denominator **cancels**. What's left is startlingly clean: your
belief-ratio going in, times the evidence-ratio (the Bayes factor), equals your
belief-ratio coming out. Multiplication, not a messy fraction. This is why practitioners
reason in odds and log-odds: updating is just "multiply by the evidence" (or "add the
log-evidence").

**What the number feels like.** Our Bayes factor was $1.152$ — barely above $1$, which is
the "no evidence either way" mark. A Bayes factor of $1$ leaves your odds unchanged; above
$1$ favors Model 2, below $1$ favors Model 1. Rough reading rules of thumb: a factor near
$3$ is "worth a mention," near $10$ is "strong," near $100$ is "decisive." Our $1.15$ is
almost nothing — three coin flips just don't carry much evidence — which is why belief
moved only from $50\%$ to $53.5\%$.

## What the notation is going to say

The formula will say three plain things in a row. First, a **Bayes factor** is one
model's evidence divided by the other's — the ratio of the two marginal likelihoods you
learned to compute last time. Second, **posterior odds equal prior odds times that Bayes
factor** — the belief you carried in, scaled up or down by how the data landed. Third, to
report a probability rather than odds, convert the posterior odds back with "probability =
odds over one-plus-odds." Everything traces to the two-coin arithmetic:
$0.144/0.125 = 1.152$, then $1 \times 1.152$, then $1.152/(1 + 1.152) = 0.535$.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $M_1, M_2$ | "model one, model two" | the two competing models/hypotheses | fair coin ($p{=}0.5$) vs biased ($p{=}0.6$) |
| $D$ | "dee" | the observed data, held fixed | heads, heads, tails |
| $P(D \mid M_1)$ | "the probability of the data given model one" | model $M_1$'s **evidence** (marginal likelihood) | $0.125$ |
| $P(D)$ | "the probability of the data" | the total evidence, $\sum_M P(D\mid M)P(M)$ — the shared denominator that cancels in the odds form | cancels out |
| $\text{BF}_{21}$ | "the Bayes factor of two over one" | ratio of the two evidences: $P(D\mid M_2)/P(D\mid M_1)$ | $1.152$ |
| $\dfrac{P(M_2)}{P(M_1)}$ | "prior odds of two over one" | how much you favored $M_2$ over $M_1$ *before* the data | $1$ (equal belief) |
| $\dfrac{P(M_2\mid D)}{P(M_1\mid D)}$ | "posterior odds of two over one" | how much you favor $M_2$ over $M_1$ *after* the data | $1.152$ |
| $\pi$ | "pi" (prior probability) | the prior probability of the favored model | $0.5$ here |
| $\text{PP}$ | "posterior probability" | the posterior probability, recovered from posterior odds | $0.535$ |

The subscript order on $\text{BF}_{21}$ says "how many times more probable the data made
$M_2$ *relative to* $M_1$" — flip the subscripts and you invert the number
($\text{BF}_{12} = 1/\text{BF}_{21}$). See
[reading math notation](../reading-math-notation/lesson.qmd) for subscript conventions.

## The formal version

**The Bayes factor.** One model's evidence over the other's:

$$
\text{BF}_{21} = \frac{P(D \mid M_2)}{P(D \mid M_1)}.
$$

Reading it aloud: "the Bayes factor of two over one equals the probability of the data
given model two, divided by the probability of the data given model one." Everyday words:
**"how many times better the data fits model 2 than model 1, each model judged by its own
averaged-over-the-prior evidence."**

Mapping it onto the toy example: $P(D\mid M_2) = 0.144$, $P(D\mid M_1) = 0.125$, so
$\text{BF}_{21} = 0.144/0.125 = 1.152$.

**Bayes' rule on odds.** Start from Bayes' rule in probability form for each model,
$P(M \mid D) \propto P(D \mid M)\,P(M)$, and take the ratio of the two models — one algebra
step per line:

$$
\begin{aligned}
\frac{P(M_2 \mid D)}{P(M_1 \mid D)}
  &= \frac{P(D \mid M_2)\,P(M_2)\,/\,P(D)}{P(D \mid M_1)\,P(M_1)\,/\,P(D)} \\
  &= \frac{P(D \mid M_2)\,P(M_2)}{P(D \mid M_1)\,P(M_1)} \\
  &= \underbrace{\frac{P(D \mid M_2)}{P(D \mid M_1)}}_{\text{Bayes factor}}
     \times \underbrace{\frac{P(M_2)}{P(M_1)}}_{\text{prior odds}}
\end{aligned}
$$

- Line 1: **Bayes' rule for each model** — posterior $\propto$ likelihood $\times$ prior,
  with the *same* denominator $P(D)$ under both (from
  [priors, posteriors & updating](../priors-posteriors-and-updating/lesson.qmd)).
- Line 2: **cancel the common factor** $1/P(D)$ — it appears in numerator and
  denominator, so it divides out (a cancellation move from
  [probability, odds & log-odds](../probability-and-odds/lesson.qmd)).
- Line 3: **regroup the product into two ratios** — separate the evidence-ratio from the
  prior-belief-ratio.

So, in words: **posterior odds = Bayes factor × prior odds.** Reading it aloud:
"posterior odds of two over one equals the Bayes factor times the prior odds." Everyday
words: **"your belief-ratio after the data is your belief-ratio before, scaled by how much
better the data fit one model than the other."**

Sanity-check against the hand numbers: $\text{BF}_{21} = 1.152$ and prior odds $= 1$, so
posterior odds $= 1.152 \times 1 = 1.152$ — matching the toy computation above.

**Back to a probability.** Posterior odds convert to a posterior probability the same way
any odds do (from [probability, odds & log-odds](../probability-and-odds/lesson.qmd)):

$$
\text{PP} = \frac{\text{posterior odds}}{1 + \text{posterior odds}}.
$$

Reading it aloud: "the posterior probability equals the posterior odds over one plus the
posterior odds." Everyday words: **"turn the odds back into a probability."** For the toy:
$1.152/(1 + 1.152) = 0.535$ — the biased coin's probability rose from $0.5$ to $0.535$.

## Explore it

```{webr}
#| autorun: true
# explore-1 — LIVE: the whole update, from evidences to a posterior probability.
# TRY IT: change prior_odds. PREDICT: with prior_odds = 3 (you favored the biased coin
# 3-to-1), what will the posterior probability be -- bigger or smaller than 0.535? Guess, run.
lik <- function(p) p^2 * (1 - p)
BF  <- lik(0.6) / lik(0.5)                    # 1.152 -- evidence ratio, biased over fair

prior_odds     <- 1                            # try 3 (favor biased 3-to-1), or 1/3
posterior_odds <- prior_odds * BF              # Bayes' rule on odds: just a multiply
PP             <- posterior_odds / (1 + posterior_odds)
c(BF = round(BF, 3), posterior_odds = round(posterior_odds, 3), PP = round(PP, 3))
```

```{webr}
#| autorun: true
# explore-2 — LIVE: the medical-test intuition. A strong Bayes factor still can't
# overcome tiny prior odds in one step.
# PREDICT: rare disease (1 in 1000), a test with Bayes factor 20 comes back positive.
# Is the patient now probably sick? Commit to yes/no, then run.
prior_prob <- 1 / 1000
prior_odds <- prior_prob / (1 - prior_prob)    # about 0.001
BF <- 20                                        # a strong single piece of evidence

posterior_odds <- prior_odds * BF               # multiply the odds
PP <- posterior_odds / (1 + posterior_odds)
c(prior_prob = prior_prob, posterior_prob = round(PP, 4))   # ~0.0196 -- still ~2%!
```

```{webr}
#| autorun: true
#| fig-alt: "A curve of posterior probability against the Bayes factor on a logarithmic x-axis, starting from prior odds of 1: it passes through 0.5 at a Bayes factor of 1, rises toward 1 for large factors, and falls toward 0 for small factors. Dashed reference lines mark Bayes factor 1 (no evidence) and the resulting 0.5 probability."
# explore-3 — LIVE: SEE how the Bayes factor moves belief, starting from even odds.
# TRY IT: change prior_odds from 1 to 0.1 and watch the whole curve shift down.
prior_odds <- 1
BF_grid <- 10^seq(-2, 3, length.out = 300)       # Bayes factors from 0.01 to 1000
PP <- (prior_odds * BF_grid) / (1 + prior_odds * BF_grid)

plot(BF_grid, PP, type = "l", lwd = 2, log = "x",
     xlab = "Bayes factor (log scale)", ylab = "posterior probability",
     main = "Evidence moves belief (starting from even odds)")
abline(v = 1, lty = 2, col = "grey50"); abline(h = 0.5, lty = 2, col = "grey50")
```

## Common traps

**Trap 1 — "the Bayes factor is the posterior probability that Model 2 is true."** No —
the Bayes factor is only the *evidence ratio*; it ignores your prior. Our $\text{BF} =
1.152$ is **not** "a $115\%$ chance" (probabilities can't exceed 1) and not even a
probability at all. To get a probability you must combine it with prior odds and convert:
$1 \times 1.152 \to 0.535$. Skip the prior and you've answered a different question.

**Trap 2 — "a big Bayes factor means the hypothesis is probably true."** Tempting, but a
strong factor can still leave a hypothesis unlikely if it started implausible. The medical
example proves it: a Bayes factor of $20$ (strong!) applied to prior odds of $1/1000$
gives posterior odds of only $0.02$ — about a $2\%$ chance, still very unlikely. Evidence
*multiplies* your starting odds; it doesn't reset them.

## Check yourself

**1.** Two point models for some data have evidences $P(D\mid M_2) = 0.30$ and
$P(D\mid M_1) = 0.10$. Compute $\text{BF}_{21}$. If your prior odds for $M_2$ over $M_1$
were $2$, what are the posterior odds, and the posterior probability of $M_2$?

```{webr}
#| autorun: false
# check1-starter — edit and Run to check your hand answer
evidence_2 <- 0.30
evidence_1 <- 0.10
prior_odds <- 2
```

::: {.callout-note collapse="true" title="Answer"}
$\text{BF}_{21} = 0.30/0.10 = 3$. Posterior odds $= \text{prior odds} \times \text{BF} =
2 \times 3 = 6$. Posterior probability $= 6/(1+6) = 6/7 \approx 0.857$. In R:
`(2 * 3) / (1 + 2 * 3)` gives `0.857`.
:::

**2.** True or false: if the Bayes factor $\text{BF}_{21}$ equals exactly $1$, the data
have left your beliefs about $M_2$ vs $M_1$ unchanged. Explain in one line.

::: {.callout-note collapse="true" title="Answer"}
**True.** Posterior odds $= 1 \times \text{prior odds} = \text{prior odds}$ — multiplying
by $1$ changes nothing, so the data provided no evidence either way. A Bayes factor of $1$
is the "no information" mark.
:::

**3.** In your own words: explain to a colleague, in two sentences, what a Bayes factor is
and why posterior odds are just prior odds times that factor.

::: {.callout-note collapse="true" title="A model answer"}
A Bayes factor is the ratio of how well the data fit two competing models — each model
judged by its own evidence — so it measures the strength of the data as a multiplier on
your beliefs, with no prior baked in. Posterior odds are prior odds times the Bayes factor
because, when you write Bayes' rule as a ratio of the two models, the shared normalizing
constant cancels and updating collapses to "start with your belief-ratio, multiply by the
evidence-ratio."
:::

## What you can now do

- **Compute a Bayes factor** as the ratio of two models' evidences (marginal likelihoods).
- **Update beliefs on the odds scale** — multiply prior odds by the Bayes factor to get
  posterior odds — and explain why the update is a multiplication.
- **Convert between probability, odds, and posterior probability** so you can start from a
  prior probability and report a posterior probability.
- **Read a Bayes factor's strength** and avoid mistaking it for a posterior probability or
  ignoring the prior it must be combined with.

If any of these feels shaky, revisit [A tiny example first](#a-tiny-example-first) and
[The formal version](#the-formal-version) before continuing — every method that turns
evidence into a probability of a hypothesis leans on all of these.

## Where this goes next

This "prior odds × Bayes factor = posterior odds" machine is the engine of Bayesian model
comparison. A closed-form Bayes-factor method takes an estimated effect and its standard
error, computes the Bayes factor for "there is an effect" versus "there is none" in closed
form (using the normal-marginal result from
[marginal likelihood & model evidence](../marginal-likelihood-and-model-evidence/lesson.qmd)),
then runs it through exactly this machine to get a posterior probability of an effect —
which is what the applied courses that build on this foundation do. Ready for more reps?
Do the [practice problems](practice.qmd).
