---
title: "Practice: Bayes factors and posterior odds"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Do them on paper or in an R session; only then
open the answer.

## Problem 1 — the toy example, new numbers

Same coin data (heads, heads, tails), so a model at heads-probability $p$ has evidence
$L(p) = p^2(1-p)$. Weigh **Model 2 (biased, $p = 0.7$)** against **Model 1 (fair, $p =
0.5$)**, with **prior odds $1$** (equal belief). Compute the Bayes factor $\text{BF}_{21}$,
the posterior odds, and the posterior probability of the biased model.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
lik <- function(p) p^2 * (1 - p)
prior_odds <- 1
```

::: {.callout-note collapse="true" title="Answer"}
Evidences: $L(0.7) = 0.7^2 \times 0.3 = 0.147$ and $L(0.5) = 0.125$. Bayes factor:
$\text{BF}_{21} = 0.147/0.125 = 1.176$. Posterior odds $= 1 \times 1.176 = 1.176$.
Posterior probability $= 1.176/(1 + 1.176) = 0.540$.

```{r p1-answer}
lik <- function(p) p^2 * (1 - p)
BF  <- lik(0.7) / lik(0.5)                    # 1.176
po  <- 1 * BF                                  # posterior odds (prior odds 1)
c(BF = round(BF, 3), post_odds = round(po, 3), PP = round(po / (1 + po), 3))
```
:::

## Problem 2 — fill in the missing step

Two point models have evidences $P(D\mid M_2) = 0.20$ and $P(D\mid M_1) = 0.05$, and your
prior odds are $1$. Here is the worked solution with the core step blanked — supply only
that step.

- Step 1 (shown): the two evidences are $0.20$ and $0.05$.
- Step 2 (**your step here**): compute the Bayes factor $\text{BF}_{21}$ from the two
  evidences. Write the exact arithmetic.
- Step 3 (shown): with prior odds $1$, the posterior odds are $4$ and the posterior
  probability of $M_2$ is $4/(1+4) = 0.8$.

```{webr}
#| autorun: false
# p2-starter — edit and Run to check the blank
evidence_2 <- 0.20
evidence_1 <- 0.05
```

::: {.callout-note collapse="true" title="Answer"}
The missing move is the **evidence ratio**: $\text{BF}_{21} = P(D\mid M_2)/P(D\mid M_1) =
0.20/0.05 = 4$. The data fit $M_2$ four times better than $M_1$. Then posterior odds $= 1
\times 4 = 4$, and the posterior probability is $4/5 = 0.8$.

```{r p2-answer}
BF <- 0.20 / 0.05                # 4 -- the evidence ratio
po <- 1 * BF                      # 4
c(BF = BF, post_odds = po, PP = po / (1 + po))    # 0.8
```
:::

## Problem 3 — start from a prior probability

You believe a hypothesis $M_2$ with **prior probability $0.2$** (so $M_1$ has $0.8$). New
data give a Bayes factor $\text{BF}_{21} = 5$. Convert the prior probability to odds,
update on the odds scale, and report the posterior probability of $M_2$.

```{webr}
#| autorun: false
# p3-starter — your job: prior prob -> prior odds -> posterior odds -> posterior prob
prior_prob <- 0.2
BF <- 5
```

::: {.callout-note collapse="true" title="Answer"}
Prior odds $= 0.2/0.8 = 0.25$. Posterior odds $= 0.25 \times 5 = 1.25$. Posterior
probability $= 1.25/(1 + 1.25) = 0.556$. So the data lifted belief in $M_2$ from $0.2$ to
about $0.556$ — it crossed $0.5$, becoming the more likely of the two.

```{r p3-answer}
prior_prob <- 0.2
prior_odds <- prior_prob / (1 - prior_prob)     # 0.25
post_odds  <- prior_odds * 5                      # 1.25
c(prior_odds = prior_odds, post_odds = post_odds,
  PP = round(post_odds / (1 + post_odds), 3))     # 0.556
```
:::

## Problem 4 — spot the error

A colleague reports — it contains **exactly one mistake** — find it and give the correct
reading.

> "The Bayes factor for the treatment working versus not working is $\text{BF} = 8$, so
> there is an $8\%$ chance the treatment works."

::: {.callout-note collapse="true" title="Answer"}
The mistake is treating the **Bayes factor as a probability**. A Bayes factor is an
evidence *ratio*, not a probability — it isn't even between $0$ and $1$ here ($8 > 1$), and
"an $8\%$ chance" mangles the value on top of that. A Bayes factor of $8$ means the data
are $8$ times more consistent with "works" than "doesn't"; to get a probability you must
combine it with prior odds and convert. For example, with even prior odds ($1$): posterior
odds $= 1 \times 8 = 8$, posterior probability $= 8/9 \approx 0.889$ — an $89\%$ chance,
nowhere near $8\%$.

```{r p4-answer}
BF <- 8
po <- 1 * BF                      # even prior odds
po / (1 + po)                      # 0.889 -- the actual probability, not 0.08
```
:::

## Problem 5 — mixing it with marginal likelihoods *(uses marginal-likelihood-and-model-evidence's averaging)*

The two models needn't be single points. Compare **Model B** (a point model, fair coin
$p = 0.5$) against **Model A** (which spreads equal prior weight over $p = 0.5$ and
$p = 0.7$) for the data heads, heads, tails. Compute each model's evidence by averaging its
likelihood over its own prior, then the Bayes factor of A over B, and the posterior
probability of A at prior odds $1$.

```{webr}
#| autorun: false
# p5-starter — edit and Run to solve, right here
lik <- function(p) p^2 * (1 - p)
```

::: {.callout-note collapse="true" title="Answer"}
Model A's evidence is the *average* likelihood over its prior:
$\tfrac12\big(L(0.5) + L(0.7)\big) = \tfrac12(0.125 + 0.147) = 0.136$ (the marginal
likelihood move). Model B's evidence is the point value $L(0.5) = 0.125$. So
$\text{BF}_{AB} = 0.136/0.125 = 1.088$, posterior odds $= 1 \times 1.088$, and the
posterior probability of A is $1.088/(1 + 1.088) = 0.521$.

```{r p5-answer}
lik <- function(p) p^2 * (1 - p)
evidence_A <- mean(lik(c(0.5, 0.7)))          # average over A's prior -> 0.136
evidence_B <- lik(0.5)                          # point model -> 0.125
BF <- evidence_A / evidence_B                    # 1.088
po <- 1 * BF
c(BF = round(BF, 3), PP_A = round(po / (1 + po), 3))     # 0.521
```
The Bayes factor still works the same way — only how you *get* each model's evidence
changes (averaging over a prior vs a single point).
:::

## Problem 6 — transfer (interpretation)

A screening test for a rare condition has a Bayes factor of $8$ for a positive result. In
Population X the condition's prior probability is $1/500$; in Population Y it is $1/5$. A
patient from each population tests positive. Compute both posterior probabilities, and
explain in one sentence why the *same* test result means very different things in the two
populations.

```{webr}
#| autorun: false
# p6-starter — edit and Run to solve, right here
BF <- 8
prior_X <- 1 / 500
prior_Y <- 1 / 5
```

::: {.callout-note collapse="true" title="Answer"}
Population X: prior odds $= (1/500)/(499/500) \approx 0.002$; posterior odds $= 0.002
\times 8 \approx 0.016$; posterior probability $\approx 0.0158$ (about $1.6\%$).
Population Y: prior odds $= (1/5)/(4/5) = 0.25$; posterior odds $= 0.25 \times 8 = 2$;
posterior probability $= 2/3 \approx 0.667$.

```{r p6-answer}
BF <- 8
odds <- function(p) p / (1 - p)
pp <- function(prior) { po <- odds(prior) * BF; po / (1 + po) }
c(pop_X = round(pp(1/500), 4), pop_Y = round(pp(1/5), 4))    # 0.0158 vs 0.6667
```
The evidence multiplier is identical ($8\times$), but it acts on very different starting
odds — a strong factor can't rescue a hypothesis that started nearly impossible (X), yet
easily tips one that started plausible (Y). Where you begin matters as much as the
evidence.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/bayes-factors-and-posterior-odds" pass="4"}

::: {.quiz-q qid="bfpo-what-is-bf" qtype="mc" answer="b"}
What is a Bayes factor $\text{BF}_{21}$?

a) The posterior probability that model 2 is the true model
b) The ratio of the two models' evidences, $P(D\mid M_2)/P(D\mid M_1)$
c) The prior odds of model 2 over model 1, before any data
d) The product of the prior odds and the posterior probability
:::

::: {.quiz-q qid="bfpo-odds-update" qtype="mc" answer="a"}
On the odds scale, how does Bayes' rule combine what you knew before with the data?

a) posterior odds = prior odds × Bayes factor
b) posterior odds = prior odds + Bayes factor
c) posterior odds = prior odds ÷ Bayes factor
:::

::: {.quiz-q qid="bfpo-bf-numeric" qtype="numeric" answer="4" tolerance="0"}
Two point models have evidences $P(D\mid M_2) = 0.24$ and $P(D\mid M_1) = 0.06$. Compute
the Bayes factor $\text{BF}_{21}$.
:::

::: {.quiz-q qid="bfpo-post-prob" qtype="numeric" answer="0.667" tolerance="0.005"}
With a Bayes factor $\text{BF}_{21} = 4$ and prior odds of $0.5$, compute the posterior
probability of $M_2$. Round to three decimals.
:::

::: {.quiz-q qid="bfpo-evidence-averaged" qtype="mc" answer="c"}
*(Interleaved — uses marginal likelihood & model evidence.)* When one of the two models
spreads its prior over several parameter values, what plays the role of that model's
evidence in the Bayes factor?

a) The single largest likelihood value among the parameter values covered by that model's prior
b) The sum of the likelihoods at each parameter value, without dividing by the number of values
c) The likelihood averaged over that model's own prior (its marginal likelihood)
:::

:::

::::
