---
title: "Practice: Priors, posteriors & updating"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only (a {webr} starter does not see it — each starter re-states its
# own data). Base R is all we need, so nothing to install.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Do them on paper or in an R session; only
then open the answer.

## Problem 1 — the toy example, new numbers

Here are 200 emails sorted by whether each is **spam** and whether it contains the word
**"free"**:

|              | free: yes | free: no | row total |
|--------------|:---------:|:--------:|:---------:|
| **spam**     | 40        | 10       | 50        |
| **ham**      | 20        | 130      | 150       |
| **col total**| 60        | 140      | 200       |

By hand, compute (a) $P(\text{spam})$ over all emails, and (b) $P(\text{spam} \mid
\text{free})$ — the chance an email is spam *given* it contains "free." Then check in R.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
tab <- matrix(c(40, 10,
                20, 130), nrow = 2, byrow = TRUE,
              dimnames = list(class = c("spam", "ham"),
                              free = c("yes", "no")))
```

::: {.callout-note collapse="true" title="Answer"}
(a) $P(\text{spam})$ = spam emails over all emails = $50 / 200 = 0.25$ — the belief
**before** you look for the word.
(b) $P(\text{spam} \mid \text{free})$: narrow to the 60 "free" emails; 40 are spam →
$40 / 60 = 0.667$ — the belief **after** you see "free." The word shoved belief from
0.25 up to 0.667.

```{r p1-answer}
tab <- matrix(c(40, 10,
                20, 130), nrow = 2, byrow = TRUE,
              dimnames = list(class = c("spam", "ham"),
                              free = c("yes", "no")))
sum(tab["spam", ]) / sum(tab)              # 0.25   (before)
tab["spam", "yes"] / sum(tab[, "yes"])     # 0.667  (after learning "free")
```
:::

## Problem 2 — fill in the missing step (prior × likelihood → posterior)

A die is either **fair** (chance of a six $= 1/6$) or **loaded** (chance of a six
$= 1/2$). Your prior is 50/50. You roll it **twice and get two sixes**. Here is the
update worked out — but the **core step is blanked**. Fill in ONLY the blanked line: the
prior × likelihood for each die.

> Likelihood of two sixes: fair $= (1/6)^2 = 0.0278$, loaded $= (1/2)^2 = 0.25$.
> **`<your step here>`** — multiply each prior by its likelihood (the unnormalized
> posterior): fair $= \underline{\phantom{0000}}$, loaded $= \underline{\phantom{0000}}$.
> Then rescale so they sum to 1.

```{webr}
#| autorun: false
# p2-starter — edit and Run to check your two unnormalized numbers
p     <- c(fair = 1/6, loaded = 1/2)
prior <- c(fair = 0.5, loaded = 0.5)
```

::: {.callout-note collapse="true" title="Answer"}
The blanked step is the heart of Bayes: **prior × likelihood.**

- fair: $0.5 \times 0.0278 = 0.0139$
- loaded: $0.5 \times 0.25 = 0.125$

Rescale by their total ($0.0139 + 0.125 = 0.1389$): fair $= 0.0139 / 0.1389 = 0.10$,
loaded $= 0.125 / 0.1389 = 0.90$. After two sixes, belief moved from 50/50 to **10%
fair, 90% loaded.**

```{r p2-answer}
p     <- c(fair = 1/6, loaded = 1/2)
prior <- c(fair = 0.5, loaded = 0.5)
lik    <- p^2                       # 0.0278, 0.25
unnorm <- prior * lik              # 0.0139, 0.125  <- the blanked step
post   <- unnorm / sum(unnorm)     # 0.10, 0.90
round(rbind(lik, unnorm, post), 4)
```
:::

## Problem 3 — a full update with a non-flat prior

A coin is either **fair** ($p = 0.5$) or **biased** toward heads ($p = 0.8$). This time
you lean toward fair beforehand: **prior 70% fair, 30% biased.** You flip **twice and get
2 heads.** Compute the posterior probability the coin is biased, by hand, then in R.

```{webr}
#| autorun: false
# p3-starter — edit and Run to solve
p     <- c(fair = 0.5, biased = 0.8)
prior <- c(fair = 0.7, biased = 0.3)
```

::: {.callout-note collapse="true" title="Answer"}
Likelihood of 2 heads: fair $= 0.5^2 = 0.25$, biased $= 0.8^2 = 0.64$.
Prior × likelihood: fair $= 0.7 \times 0.25 = 0.175$, biased $= 0.3 \times 0.64 = 0.192$.
Rescale (total $0.175 + 0.192 = 0.367$): biased $= 0.192 / 0.367 = 0.523$.

So the posterior is about **48% fair, 52% biased** — the data leans biased, but the
skeptical prior keeps it a near-tie. Notice how much the prior matters here.

```{r p3-answer}
p     <- c(fair = 0.5, biased = 0.8)
prior <- c(fair = 0.7, biased = 0.3)
lik    <- p^2                        # 0.25, 0.64
post   <- (prior * lik) / sum(prior * lik)
round(post, 4)                       # fair 0.4768, biased 0.5232
```
:::

## Problem 4 — spot the error (the prior got dropped)

A colleague updates on **3 heads** from a coin that is either fair ($p = 0.5$) or biased
($p = 0.8$), with a prior of **80% fair, 20% biased.** Their work contains **exactly one
mistake** — find it, explain why it's wrong, and give the correct posterior.

> "Likelihood of 3 heads: fair $= 0.5^3 = 0.125$, biased $= 0.8^3 = 0.512$. To get the
> posterior I just rescale the likelihoods so they sum to 1: biased
> $= 0.512 / (0.125 + 0.512) = 0.804$. So there's an 80% chance the coin is biased."

```{webr}
#| autorun: false
# p4-starter — copy this to check the correct update
p     <- c(fair = 0.5, biased = 0.8)
prior <- c(fair = 0.8, biased = 0.2)   # <- the 80/20 prior the colleague ignored
```

::: {.callout-note collapse="true" title="Answer"}
The mistake: **the prior was dropped.** The colleague normalized the *likelihoods* alone,
but posterior ∝ likelihood **× prior**. Their "0.804" is what you'd get only with a flat
50/50 prior — but the prior here is 80/20 toward fair.

Doing it correctly:

- prior × likelihood: fair $= 0.8 \times 0.125 = 0.100$, biased $= 0.2 \times 0.512 =
  0.1024$.
- rescale (total $0.100 + 0.1024 = 0.2024$): biased $= 0.1024 / 0.2024 = 0.506$.

So the correct posterior is about **49% fair, 51% biased** — a coin toss, *not* 80%
biased. The skeptical prior nearly cancels the data's lean. Dropping the prior isn't a
rounding slip; it changes the conclusion.

```{r p4-answer}
p     <- c(fair = 0.5, biased = 0.8)
prior <- c(fair = 0.8, biased = 0.2)
lik    <- p^3
wrong  <- lik / sum(lik)                       # 0.196, 0.804 — prior IGNORED
right  <- (prior * lik) / sum(prior * lik)     # 0.494, 0.506 — the correct update
round(rbind(wrong, right), 4)
```
:::

## Problem 5 — mixing it with sampling *(uses [randomness, samples & sampling distributions](../randomness-samples-and-sampling-distributions/lesson.qmd))*

Suppose you already have a posterior over a coin's bias: a `Beta(8, 4)` curve (from a
flat prior and 7 heads in 10 flips). Instead of reading its interval off a formula, do it
the **MCMC way** — draw a big pile of samples from the posterior and summarize them, the
way you learned to summarize a sample. Draw 20,000 values with `rbeta`, then report the
mean and the 2.5% / 97.5% quantiles.

**PREDICT** first: roughly where will the mean of the draws land, and will the interval
be wide or narrow with only 10 flips of data? Write it down, then run.

```{webr}
#| autorun: false
# p5-starter — edit and Run to solve
set.seed(42)
a <- 8; b <- 4
```

::: {.callout-note collapse="true" title="Answer"}
Summarizing draws from a posterior is *exactly* the sample-summary move from
[randomness & sampling](../randomness-samples-and-sampling-distributions/lesson.qmd): the
posterior mean is the average of the draws, and the 95% credible interval is the middle
95% of them (the 2.5% and 97.5% quantiles).

```{r p5-answer}
set.seed(42)
a <- 8; b <- 4
draws <- rbeta(20000, a, b)                       # 20,000 samples FROM the posterior
round(c(mean  = mean(draws),
        lower = quantile(draws, 0.025),
        upper = quantile(draws, 0.975)), 3)       # ~0.666, ~0.386, ~0.893
```

The mean lands near 0.67 and the 95% credible interval is roughly $[0.39, 0.89]$ — wide,
because 10 flips is little data. This is the whole payoff of MCMC: once you have the
pile of draws, the answers are ordinary `mean()` and `quantile()`.
:::

## Problem 6 — a gentle cliffhanger (more data, sharper belief)

You keep flipping the same coin and now have **15 heads in 20 flips.** With a flat prior,
the posterior over the bias is a `Beta(16, 6)` curve. **Predict:** compared with the
`Beta(8, 4)` from 10 flips (interval ≈ $[0.39, 0.89]$), will the new 95% credible interval
be *wider* or *narrower*? Then run the starter to read off the mean and the interval.

```{webr}
#| autorun: false
# p6-starter — solve for the Beta(16, 6) posterior
a <- 16; b <- 6
```

::: {.callout-note collapse="true" title="Answer"}
More data sharpens the belief, so the interval should get **narrower** — and it does.
For a flat prior with $h$ heads in $n$ flips, the posterior is `Beta(h+1, n-h+1)`, and
`qbeta` reads its quantiles straight off.

```{r p6-answer}
a <- 16; b <- 6
post_mean <- a / (a + b)                    # 0.727
ci95      <- qbeta(c(0.025, 0.975), a, b)   # 0.528, 0.887
round(c(mean = post_mean, lower = ci95[1], upper = ci95[2]), 3)
```

The interval $[0.53, 0.89]$ is noticeably narrower than the 10-flip $[0.39, 0.89]$: more
evidence, tighter posterior. Keep flipping and it would shrink further toward the true
bias — which is the sense in which "the data eventually wins" over the prior. *The next
step — where the posterior has no tidy formula and MCMC does the drawing for you — is a
course-level topic; here you've built every piece it stands on.*
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/priors-posteriors-and-updating" pass="4"}

::: {.quiz-q qid="ppu-bayes-flip" qtype="mc" answer="b"}
Bayes' rule says the posterior is proportional to what?

a) The prior alone
b) Likelihood $\times$ prior
c) Likelihood $\div$ prior
:::

::: {.quiz-q qid="ppu-prior-matters" qtype="mc" answer="a"}
With little data, is it safe to ignore the prior and treat the posterior as just the
likelihood?

a) No — with little data the prior pulls hard and can flip the conclusion
b) Yes — the prior washes out and the likelihood determines the conclusion
c) Yes — the prior only rescales and changes nothing
:::

::: {.quiz-q qid="ppu-credible-vs-confidence" qtype="mc" answer="c"}
How does a 95% credible interval differ from a 95% confidence interval?

a) They are identical in meaning
b) The credible interval is always the wider of the two
c) The credible interval is a statement about the parameter given your data; the confidence interval is about the procedure across re-runs
:::

::: {.quiz-q qid="ppu-conditional-prob" qtype="numeric" answer="0.25" tolerance="0.01"}
Of $100$ people, $40$ own a dog; of those $40$, $10$ also own a cat. What is
$P(\text{cat} \mid \text{dog})$? Round to two decimals.
:::

::: {.quiz-q qid="ppu-likelihood-ratio" qtype="mc" answer="a"}
(From likelihood & log-likelihood.) If the data are $4\times$ more likely under "biased"
than under "fair," what is that ratio of likelihoods called?

a) Likelihood ratio — how strongly the data favor one hypothesis over the other
b) Prior odds — the initial belief ratio before any data arrive
c) Posterior probability — the updated belief after seeing the data
:::

:::

::::
