---
title: "Practice: Variance structure & standardization"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only (a {webr} starter does not see it — each starter re-states its
# own data). Base R is all we need, so nothing to install.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

Work these in order — they get harder. Do them on paper or in an R session; only
then open the answer.

## Problem 1 — standardize, new numbers

The column `x <- c(10, 20, 30)`. By hand, compute its mean, its standard deviation, and
then its three **z-scores** (center, then scale). Check with `scale()`.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
x <- c(10, 20, 30)
```

::: {.callout-note collapse="true" title="Answer"}
Mean $= (10+20+30)/3 = 20$. Deviations $-10, 0, +10$; squared $100, 0, 100$; sum $200$;
divide by $n - 1 = 2$ → variance $100$ → standard deviation $\sqrt{100} = 10$.
Standardize each: $-10/10 = -1$, $0/10 = 0$, $10/10 = +1$. So the z-scores are
$-1, 0, +1$.

```{r p1-answer}
x <- c(10, 20, 30)
(x - mean(x)) / sd(x)   # -1  0  1
as.vector(scale(x))     # -1  0  1 — scale() does center-then-scale for us
```
:::

## Problem 2 — fill in the missing step (the covariance matrix)

Here is the covariance matrix of a two-column data frame, worked out — but the
**diagonal entries are blanked**. Fill in ONLY the two blanks, using the rule "the
diagonal of a covariance matrix is each column's variance."

$$
\text{df} = \begin{bmatrix} x & y \\ 1 & 2 \\ 2 & 4 \\ 3 & 6 \end{bmatrix}
$$

> The off-diagonal is $\mathrm{Cov}(x, y) = 2$ (given). The full grid is
> $\begin{bmatrix} \underline{\phantom{00}} & 2 \\ 2 & \underline{\phantom{00}} \end{bmatrix}$
> — the top-left blank is $\mathrm{Var}(x)$ and the bottom-right
> blank is $\mathrm{Var}(y)$. Compute those two variances.

```{webr}
#| autorun: false
# p2-starter — edit and Run to check your two diagonal entries
df <- data.frame(x = c(1, 2, 3), y = c(2, 4, 6))
```

::: {.callout-note collapse="true" title="Answer"}
The blanked step is "the diagonal holds each column's own variance."

- $\mathrm{Var}(x)$ for $x = (1,2,3)$: mean 2, deviations $-1, 0, +1$, squared $1,0,1$,
  sum 2, divide by $n-1 = 2$ → **1**.
- $\mathrm{Var}(y)$ for $y = (2,4,6)$: mean 4, deviations $-2, 0, +2$, squared $4,0,4$,
  sum 8, divide by $2$ → **4**.

So the covariance matrix is $\begin{bmatrix} 1 & 2 \\ 2 & 4 \end{bmatrix}$ — symmetric,
with variances on the diagonal.

```{r p2-answer}
df <- data.frame(x = c(1, 2, 3), y = c(2, 4, 6))
cov(df)              # diagonal 1 and 4, off-diagonal 2 and 2
c(var(df$x), var(df$y))   # 1 4 — exactly the diagonal
```
:::

## Problem 3 — split the spread (within vs. between)

Four measurements in two groups:

- **Group X:** `4, 6` (mean 5)
- **Group Y:** `14, 16` (mean 15)

By hand, compute the grand mean, the **total** sum of squared deviations (around the
grand mean), the **within-group** sum, and the **between-group** sum. Confirm within +
between = total. Then check in R.

```{webr}
#| autorun: false
# p3-starter — your job: total_ss, within_ss, between_ss
prices <- c(4, 6, 14, 16)
group  <- factor(c("X", "X", "Y", "Y"))
```

::: {.callout-note collapse="true" title="Answer"}
Grand mean $= (4+6+14+16)/4 = 10$.

- **Total:** $(4-10)^2 + (6-10)^2 + (14-10)^2 + (16-10)^2 = 36 + 16 + 16 + 36 = 104$.
- **Within** (around each group's own mean, 5 and 15): $(4-5)^2 + (6-5)^2 = 1 + 1 = 2$
  for X, and $(14-15)^2 + (16-15)^2 = 1 + 1 = 2$ for Y → **4** total.
- **Between** (each group mean vs. grand, times group size 2):
  $2(5-10)^2 + 2(15-10)^2 = 2(25) + 2(25) = 100$.

Check: $4 + 100 = 104$ ✓ — almost all the spread is *between* the groups.

```{r p3-answer}
prices <- c(4, 6, 14, 16)
group  <- factor(c("X", "X", "Y", "Y"))

grand <- mean(prices)                               # 10
total_ss <- sum((prices - grand)^2)                 # 104

gm <- tapply(prices, group, mean)                   # X = 5, Y = 15
within_ss <- sum((prices - gm[group])^2)            # 4
between_ss <- sum(table(group) * (gm - grand)^2)    # 100

c(total = total_ss, within = within_ss, between = between_ss)
within_ss + between_ss                              # 104
```
:::

## Problem 4 — spot the error (standardizing)

A colleague standardized the area column below — the computation contains **exactly one
mistake**. Find it, explain why it's wrong, and give the corrected z-scores.

> "z-score = value minus mean, divided by the **variance**. For
> `area <- c(80, 100, 120, 140, 160)`: mean is 120, variance is 1000, so the z-scores are
> $-40/1000, -20/1000, 0, 20/1000, 40/1000 = -0.04, -0.02, 0, 0.02, 0.04$."

```{webr}
#| autorun: false
# p4-starter — edit and Run to see the correct z-scores
area <- c(80, 100, 120, 140, 160)
```

::: {.callout-note collapse="true" title="Answer"}
The one mistake: standardizing divides by the **standard deviation**, not the variance.
The variance is in *squared* units, so dividing by it shrinks the z-scores to a
meaningless tiny scale ($\pm 0.04$). The standard deviation is $\sqrt{1000} \approx
31.62$, so the correct z-scores are $-40/31.62 \approx -1.26$, $-20/31.62 \approx -0.63$,
$0$, $+0.63$, $+1.26$ — the recognizable "roughly $\pm 1$" spread of standardized data.

```{r p4-answer}
area <- c(80, 100, 120, 140, 160)
round((area - mean(area)) / sd(area), 2)   # -1.26 -0.63 0.00 0.63 1.26 — divide by sd
```
:::

## Problem 5 — mixing it with matrices *(uses [matrices & linear transforms](../matrices-and-linear-transforms/lesson.qmd)'s diagonal & symmetry)*

The house data again:

```{webr}
#| autorun: false
# p5-starter — edit and Run to solve, right here
houses <- data.frame(area = c(80, 100, 120, 140, 160),
                     beds = c(2, 2, 3, 4, 4))
```

(a) Build the covariance matrix with `cov(houses)`. (b) Using `diag()` from
[matrices & linear transforms](../matrices-and-linear-transforms/lesson.qmd), pull out
its diagonal — **PREDICT** first what those two numbers should be and why. (c) Confirm
the matrix is symmetric.

::: {.callout-note collapse="true" title="Answer"}
(a) `cov(houses)` is $\begin{bmatrix} 1000 & 30 \\ 30 & 1 \end{bmatrix}$.

(b) The **diagonal** of a covariance matrix is each column's own variance — the
covariance-with-itself identity. So `diag(cov(houses))` should be
`var(area) = 1000` and `var(beds) = 1`.

(c) Symmetric means entry $(1,2)$ equals entry $(2,1)$: both are 30 — the same
$\mathrm{Cov}(\text{area}, \text{beds})$ read in either order. Every covariance matrix is
symmetric for exactly this reason.

```{r p5-answer}
houses <- data.frame(area = c(80, 100, 120, 140, 160),
                     beds = c(2, 2, 3, 4, 4))
S <- cov(houses)
S
diag(S)                    # 1000 1 — each column's variance
c(var(houses$area), var(houses$beds))   # 1000 1 — matches the diagonal
S[1, 2] == S[2, 1]         # TRUE — symmetric
```
:::

## Problem 6 — pooling / borrowing strength (a gentle cliffhanger)

A new neighborhood has a single recorded sale at **300** (thousands). The overall grand
mean across neighborhoods is **160**. A pooled ("shrinkage") estimate blends the group's
own mean toward the grand mean, weighted by how much you trust the group.

**Predict:** with trust-weight $0.5$, will the pooled estimate land closer to 300 or to
160? What if you had many sales in the new neighborhood instead of one — should the weight
go up or down, and which way would the estimate move? (You can't yet compute the "right"
weight — that formula is course-level — but you can reason about the blend.)

```{webr}
#| autorun: false
# p6-starter — try a few weights and watch the blend move
grand      <- 160
group_mean <- 300
weight     <- 0.5
```

::: {.callout-note collapse="true" title="Answer"}
With weight $0.5$ the pooled estimate is a straight average of the two:
$0.5 \times 300 + 0.5 \times 160 = 230$ — halfway between, so exactly between 300 and 160.

With **many** sales you'd trust the group more, so the weight goes **up** (toward 1), and
the estimate moves back toward the group's own mean of 300. With a single flimsy sale you
trust it less, the weight drops toward 0, and the estimate is pulled hard toward the
stable grand mean of 160. That "trust more data more" dial is the heart of pooling — the
actual formula for the weight (built from within- and between-group spread) is what
course-level mixed-model material derives next.

```{r p6-answer}
grand      <- 160
group_mean <- 300

for (w in c(0.25, 0.5, 0.9)) {
  cat("weight", w, "-> pooled estimate", w * group_mean + (1 - w) * grand, "\n")
}
# weight 0.25 -> 195   (trust group little: near the grand mean)
# weight 0.5  -> 230   (halfway)
# weight 0.9  -> 286   (trust group lots: near the group's own mean)
```
:::
