---
title: "The joint fit from summaries"
subtitle: "Rebuild a whole multiple regression from a correlation matrix and marginal coefficients — no raw data"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)

sim_panel <- function(n = 200, seed = 2, rho_max = 0.95) {
  set.seed(seed); p <- 12
  R <- diag(p)
  R[1, 2] <- R[2, 1] <- 0.90; R[3, 4] <- R[4, 3] <- rho_max
  R[8, 10] <- R[10, 8] <- 0.70; R[11, 12] <- R[12, 11] <- 0.65
  X <- matrix(rnorm(n * p), n, p) %*% chol(R)
  colnames(X) <- c("rate_10y","rate_2y","cpi_infl","pce_infl","unemp","oil",
                   "usd_index","sentiment","housing","pmi","credit_spread","vix")
  beta <- numeric(p); beta[c(1, 3, 5)] <- c(0.5, -0.4, -0.3)
  list(X = X, y = drop(X %*% beta + rnorm(n, sd = 1)))
}
```

## Where we're going

Here is the meta-analysis problem that motivates this whole course. Five published studies
each report, for the same economic indicators, a *marginal* (one-at-a-time) regression
coefficient — the effect of each indicator on GDP fitted alone. You never see their raw
data. You also have a **correlation matrix** among the indicators from a reference panel.
Can you reconstruct the *joint* multiple regression — the effect of each indicator holding
the others fixed — from just those two ingredients? Remarkably, **yes**: one matrix
formula, $\hat\beta_{\text{joint}} = R^{-1}b$, rebuilds the entire joint fit, its $R^2$,
and even its BIC — no raw rows required. This module builds and proves that identity.

**Builds on:** [Multiple Regression](../../../multiple-regression/syllabus.md) (the joint
fit via the normal equations $X^\top X\hat\beta = X^\top y$; joint ≠ marginal),
[variance-structure-and-standardization](../../../../foundations/variance-structure-and-standardization/lesson.qmd)
(standardizing; the correlation matrix),
[matrices-and-linear-transforms](../../../../foundations/matrices-and-linear-transforms/lesson.qmd)
(the inverse, `solve()`).

## Warm-up: do you remember?

**1.** *(Multiple Regression)* What matrix equation do the least-squares joint coefficients
$\hat\beta$ satisfy?

::: {.callout-note collapse="true" title="Answer"}
The normal equations $X^\top X\,\hat\beta = X^\top y$, so $\hat\beta = (X^\top X)^{-1}X^\top y$.
See [Multiple Regression, Module 2](../../../multiple-regression/modules/02-least-squares-and-the-normal-equations/lesson.qmd).
:::

**2.** *(standardization)* After you standardize every column (subtract mean, divide by SD),
what does the matrix $X^\top X / n$ become — and what are its diagonal entries?

::: {.callout-note collapse="true" title="Answer"}
It becomes the **correlation matrix** $R$ among the predictors: each entry is a
correlation, and the diagonal is all 1s (a standardized variable correlates perfectly with
itself). See [variance-structure-and-standardization](../../../../foundations/variance-structure-and-standardization/lesson.qmd).
:::

**3.** *(Multiple Regression, Module 3)* If a predictor's *marginal* slope is large but its
*joint* coefficient is near zero, what's the usual cause?

::: {.callout-note collapse="true" title="Answer"}
It's correlated with another predictor that carries the real effect; alone it borrows that
association, but held-fixed against the real driver it adds little. Correlation between
predictors is exactly what $R$ encodes — and what $R^{-1}$ will undo. See
[Multiple Regression, Module 3](../../../multiple-regression/modules/03-marginal-vs-joint/lesson.qmd).
:::

## A tiny example first

Two standardized indicators. Their correlation matrix and the two marginal coefficients
(each indicator's one-at-a-time slope on $y$) are:

$$
R = \begin{pmatrix} 1 & 0.5 \\ 0.5 & 1 \end{pmatrix}, \qquad b = (0.7,\ 0.5).
$$

Marginally, *both* indicators look positive (0.7 and 0.5). What are the **joint**
coefficients? The claim is $\hat\beta = R^{-1}b$. For a 2×2, $R^{-1} = \frac{1}{1 - 0.5^2}
\begin{pmatrix} 1 & -0.5 \\ -0.5 & 1 \end{pmatrix} = \frac{1}{0.75}\begin{pmatrix} 1 & -0.5 \\ -0.5 & 1 \end{pmatrix}$.
So

$$
\hat\beta = \frac{1}{0.75}\begin{pmatrix} 1 & -0.5 \\ -0.5 & 1 \end{pmatrix}\begin{pmatrix} 0.7 \\ 0.5 \end{pmatrix}
= \frac{1}{0.75}\begin{pmatrix} 0.7 - 0.25 \\ -0.35 + 0.5 \end{pmatrix}
= \frac{1}{0.75}\begin{pmatrix} 0.45 \\ 0.15 \end{pmatrix} = \begin{pmatrix} 0.6 \\ 0.2 \end{pmatrix}.
$$

The joint coefficient on the second indicator shrank from a marginal 0.5 to **0.2** — held
against its correlated partner, it adds much less. And the fit's $R^2$ comes from the same
two ingredients: $R^2 = b^\top R^{-1} b = b^\top \hat\beta = 0.7(0.6) + 0.5(0.2) = 0.42 + 0.10 = 0.52$.

```{r toy-example}
R <- matrix(c(1, 0.5, 0.5, 1), 2); b <- c(0.7, 0.5)
beta <- solve(R, b)                          # R^{-1} b, without forming the inverse
r2   <- as.numeric(t(b) %*% beta)            # b' R^{-1} b
c(beta1 = beta[1], beta2 = beta[2], R2 = r2) # 0.6, 0.2, 0.52
```

We rebuilt the joint fit *and* its $R^2$ from a correlation matrix and two marginal numbers
— no data rows anywhere.

## Getting a feel for it

**$R^{-1}$ is the "discount for overlap" operator.** The marginal coefficient $b_j$ credits
indicator $j$ with everything it appears to explain about $y$ — including the part that's
really its correlated neighbours' doing. The correlation matrix $R$ records exactly how
much the indicators overlap. Multiplying $b$ by $R^{-1}$ **subtracts out the shared credit**,
leaving each indicator only what it uniquely adds. When $R = I$ (no correlation),
$R^{-1} = I$ and joint = marginal — nothing to discount. The more correlated the
predictors, the more $R^{-1}$ reshuffles the credit — which is the marginal-vs-joint story
from the last course, now as a single matrix multiply.

**Why $R^{-1}b$ is exactly the joint fit.** In the standardized world the normal equations
$X^\top X\hat\beta = X^\top y$ become $R\hat\beta = b$ once you divide both sides by $n$
(because $X^\top X/n = R$ and $X^\top y/n = b$). Solving that is $\hat\beta = R^{-1}b$ — the
*same* least-squares fit as Course A, but expressed in ingredients a meta-analysis actually
publishes. Nothing new is assumed; we just renamed the pieces.

**What the numbers feel like.** $R^2 = b^\top R^{-1}b = 0.52$ says the two indicators
together explain about half the variation in (standardized) $y$; the leftover residual
variance is $1 - 0.52 = 0.48$. That single number is the doorway to the likelihood and BIC
— all still from summaries.

## What the notation is going to say

The formula will say: *to get the joint coefficients, take the marginal ones and hit them
with the inverse correlation matrix — that discounts each for its overlap with the others.*
The fit quality is the marginal vector dotted with the joint vector, $b^\top R^{-1}b$; one
minus that is the residual variance, and from residual variance the likelihood and BIC
follow — every quantity a full regression reports, from two summary ingredients.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $R$ | "the correlation matrix R" | correlations among the predictors ($X^\top X/n$ standardized); diagonal all 1 | $\left(\begin{smallmatrix}1&.5\\.5&1\end{smallmatrix}\right)$ |
| $b$ | "the marginal vector b" | each predictor's one-at-a-time (marginal) coefficient ($X^\top y/n$ standardized) | $(0.7, 0.5)$ |
| $R^{-1}$ | "R inverse" | the matrix that undoes $R$ (in R: `solve(R)`) | $\frac{1}{0.75}\left(\begin{smallmatrix}1&-.5\\-.5&1\end{smallmatrix}\right)$ |
| $\hat\beta$ | "beta hat" | the reconstructed **joint** coefficients, $R^{-1}b$ | $(0.6, 0.2)$ |
| $b^\top R^{-1} b$ | "b-transpose R-inverse b" | the fit's $R^2$ (standardized $y$) | $0.52$ |
| $1 - b^\top R^{-1}b$ | — | residual variance of standardized $y$ | $0.48$ |
| $S$ | "ess" | the number of studies being combined (meta-analysis) | 5 |

## The formal version

**The identity.** Standardize every predictor and $y$ (mean 0, SD 1). Then Course A's normal
equations, divided through by $n$, read

$$
\underbrace{\tfrac{1}{n}X^\top X}_{=\,R}\,\hat\beta = \underbrace{\tfrac{1}{n}X^\top y}_{=\,b}
\qquad\Longrightarrow\qquad \boxed{\;\hat\beta_{\text{joint}} = R^{-1} b\;}.
$$

- Line 1: in standardized coordinates $X^\top X/n$ *is* the predictor correlation matrix
  $R$, and $X^\top y/n$ *is* the vector of marginal (one-at-a-time) coefficients $b$ —
  because for standardized columns a simple-regression slope equals the correlation. (Both
  facts from the [variance-structure](../../../../foundations/variance-structure-and-standardization/lesson.qmd) module.)
- Box: solve $R\hat\beta = b$ for $\hat\beta$. That's the whole reconstruction.

Reading it aloud: "beta-hat-joint equals R-inverse b" — or in everyday words: "discount the
marginal coefficients for how much the predictors overlap, and you get the joint ones."

**Fit quality, likelihood, and BIC — all from summaries.** With standardized $y$ (variance
1), the fraction of variance the model explains is

$$
R^2 = b^\top R^{-1} b, \qquad \text{residual variance} = 1 - b^\top R^{-1} b.
$$

Because the Gaussian log-likelihood depends on the data only through that residual variance,
the BIC of a subset $S$ is (up to an additive constant that is the *same* for every model on
the same data, so it cancels in every comparison):

$$
\mathrm{BIC}(S) = n\,\log\!\big(1 - b_S^\top R_S^{-1} b_S\big) + |S|\,\log n,
$$

where $R_S, b_S$ are the sub-matrix and sub-vector for the indicators in $S$. So the
[model-selection](../../../../foundations/model-selection-and-information-criteria/lesson.qmd)
score we'll search over is computable from summaries alone — the engine of Module 2.

**Sanity check (toy):** $R^2 = 0.7(0.6) + 0.5(0.2) = 0.52$, matching the R output above.

## Explore it

### The reconstruction matches a full-data `lm()`

```{r reconstruct}
# PREDICT: will the summary-only reconstruction (solve(R, b)) match lm() on the full raw
# data to full precision, or only roughly? Write your guess, then read the two rows.
sim <- sim_panel(n = 200, seed = 2)          # a full 12-indicator panel (hidden setup)
Xs <- scale(sim$X); ys <- as.numeric(scale(sim$y))
R  <- cor(sim$X)                              # correlation matrix (a "summary")
b  <- as.numeric(cor(sim$X, sim$y))           # marginal standardized coefs (a "summary")

beta_summaries <- solve(R, b)                 # R^{-1} b — from summaries only
beta_fulldata  <- coef(lm(ys ~ Xs - 1))       # lm() on the standardized raw data
round(rbind(from_summaries = beta_summaries, from_lm = unname(beta_fulldata)), 4)
```

The two rows are identical: the summary-only reconstruction reproduces the full-data joint
fit. This is the course's central "wow" — and it's what the equivalence harness freezes.

### Self-check against the frozen proof (validation tier: full)

The `equivalence/` harness froze `lm()`'s standardized coefficients for this exact seed-2
panel (**tier: full** — the reference is R's own `lm`, which also runs in your browser). The
cell below rebuilds them from summaries and checks against the frozen numbers:

```{webr}
#| autorun: true
# Regenerate the exact seed-2 panel, reconstruct the joint fit from summaries, and compare
# to the frozen reference from equivalence/fixtures/switchstep-joint-from-summaries.json.
sim_panel <- function(n = 200, seed = 2) {
  set.seed(seed); p <- 12; R <- diag(p)
  R[1,2]<-R[2,1]<-0.90; R[3,4]<-R[4,3]<-0.95; R[8,10]<-R[10,8]<-0.70; R[11,12]<-R[12,11]<-0.65
  X <- matrix(rnorm(n*p), n, p) %*% chol(R)
  beta <- numeric(p); beta[c(1,3,5)] <- c(0.5,-0.4,-0.3)
  list(X = X, y = drop(X %*% beta + rnorm(n, sd = 1)))
}
sim <- sim_panel(); R <- cor(sim$X); b <- as.numeric(cor(sim$X, sim$y))
beta_summaries <- solve(R, b)                       # from summaries only
r2 <- as.numeric(t(b) %*% beta_summaries)

beta_frozen <- c(0.508104, -0.035850, -0.705554, 0.337197, -0.139818, 0.067403,
                 -0.073332, 0.013897, 0.040463, -0.111275, 0.037216, -0.014408)
cat("beta max diff vs frozen lm reference:", max(abs(beta_summaries - beta_frozen)), "\n")
cat("R2 from summaries:", round(r2, 6), " (frozen: 0.402275)\n")
# Note the inflation twins: cpi_infl (index 3) joint = -0.71 while its 0.95-correlated
# partner pce_infl (index 4) = +0.34 — R^{-1} split the shared effect, from summaries alone.
```

### The meta-analysis hook

In practice the marginal vector $b$ comes from *combining* several studies. Each study
reports its own marginal coefficients (with standard errors); you **inverse-variance-average**
them — weight each by $1/\text{SE}^2$, so more precise studies count more — into one $b$,
then reconstruct once. Module 5 assembles this; here is the averaging step:

```{r meta-average}
# Three studies' marginal coefficients for the same 2 indicators, with standard errors.
est <- rbind(study1 = c(0.68, 0.52), study2 = c(0.73, 0.47), study3 = c(0.70, 0.51))
se  <- rbind(study1 = c(0.05, 0.06), study2 = c(0.04, 0.05), study3 = c(0.06, 0.05))
w   <- 1 / se^2                                     # inverse-variance weights
b_combined <- colSums(w * est) / colSums(w)         # precision-weighted average
round(b_combined, 4)                                 # one b vector, ready for R^{-1} b
```

## Common traps

**Trap: "you can't fit a joint regression without the raw data."** You can, *if* you have
the predictor correlation matrix and the marginal coefficients — $\hat\beta = R^{-1}b$
reconstructs it exactly (the `from_summaries` and `from_lm` rows matched to every digit).
This is precisely what makes the meta-analysis setting workable: published summaries are
enough.

**Trap: "the marginal coefficients *are* the joint effects — just read them off."** Only when
predictors are uncorrelated ($R = I$). Otherwise $R^{-1}$ reshuffles them: in the toy, the
second indicator went from a marginal 0.5 to a joint 0.2; in the panel, `cpi_infl` and its
0.95-correlated twin `pce_infl` had their shared effect split (−0.71 vs +0.34). Reading
marginals as joint effects re-commits the last course's headline error.

## Check yourself

**1.** With $R = \begin{pmatrix}1 & 0 \\ 0 & 1\end{pmatrix}$ (uncorrelated) and $b = (0.4, 0.9)$,
what is $\hat\beta = R^{-1}b$, and why?

::: {.callout-note collapse="true" title="Answer"}
$R^{-1} = I$, so $\hat\beta = b = (0.4, 0.9)$ — with uncorrelated predictors there is no
overlap to discount, so joint = marginal. $R^{-1}$ only does work when $R$ has off-diagonal
correlation.
:::

**2.** A reconstruction gives $R^2 = b^\top R^{-1} b = 0.65$ on standardized $y$. What is the
residual variance, and would a smaller residual variance make the log-likelihood larger or
smaller?

::: {.callout-note collapse="true" title="Answer"}
Residual variance $= 1 - 0.65 = 0.35$. A smaller residual variance means the model fits
better, which makes the log-likelihood **larger** (and $-2\log L$, the misfit term in BIC,
smaller). This is the link from the reconstruction to model selection.
:::

**3.** In your own words: explain to a colleague, in two sentences, how you can rebuild a
joint multiple regression from a correlation matrix and marginal coefficients, and what
$R^{-1}$ is doing.

::: {.callout-note collapse="true" title="A model answer"}
In the standardized world the predictor correlation matrix $R$ plays the role of $X^\top X$
and the marginal coefficients $b$ play the role of $X^\top y$, so the joint fit is just
$\hat\beta = R^{-1}b$ — no raw data needed. $R^{-1}$ discounts each marginal coefficient for
how much its predictor overlaps with the others, leaving the unique (joint) effect.
:::

## What you can now do

- **Reconstruct a joint regression from summaries** with $\hat\beta = R^{-1}b$, and explain
  why it equals the full-data least-squares fit (standardized normal equations).
- **Recover $R^2$ and residual variance** as $b^\top R^{-1}b$ and $1 - b^\top R^{-1}b$ — and
  hence the log-likelihood and BIC — from summaries alone.
- **Interpret $R^{-1}$** as the operator that discounts marginal coefficients for predictor
  overlap (joint = marginal only when $R = I$).
- **Combine studies** into one $b$ by inverse-variance averaging.

If any of these feels shaky, revisit [The formal version](#the-formal-version) — the
switching search scores every candidate subset with exactly this summary-stat BIC.

## Where this goes next

You can score any subset's BIC from summaries. Next: [The switching search](../02-the-switching-search/lesson.qmd)
uses that score to *search* for the best subset — adding, removing, and swapping indicators.
Ready for reps? Do the [practice problems](practice.qmd). External readings:
[resources](../../resources.md).
