---
title: "Practice: The joint fit from summaries"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
set.seed(42)
```

Work these in order — they get harder. Do them on paper or in an R session; only then
open the answer.

## Problem 1 — reconstruct a joint fit, new numbers

Two standardized indicators with correlation matrix and marginal coefficients

$$
R = \begin{pmatrix} 1 & 0.6 \\ 0.6 & 1 \end{pmatrix}, \qquad b = (0.8,\ 0.4).
$$

Compute the joint coefficients $\hat\beta = R^{-1}b$ and the fit's $R^2 = b^\top R^{-1}b$.

```{webr}
#| autorun: false
# p1-starter
R <- matrix(c(1, 0.6, 0.6, 1), 2); b <- c(0.8, 0.4)
```

::: {.callout-note collapse="true" title="Answer"}
$R^{-1} = \frac{1}{1-0.36}\begin{pmatrix}1&-0.6\\-0.6&1\end{pmatrix} = \frac{1}{0.64}\begin{pmatrix}1&-0.6\\-0.6&1\end{pmatrix}$.
So $\hat\beta_1 = (0.8 - 0.6\cdot0.4)/0.64 = 0.56/0.64 = 0.875$ and
$\hat\beta_2 = (0.4 - 0.6\cdot0.8)/0.64 = -0.08/0.64 = -0.125$. The second indicator's
coefficient **flipped negative** jointly (marginal $+0.4$)! $R^2 = 0.8(0.875) + 0.4(-0.125) = 0.7 - 0.05 = 0.65$.

```{r}
R <- matrix(c(1, 0.6, 0.6, 1), 2); b <- c(0.8, 0.4); beta <- solve(R, b)
c(beta = beta, R2 = sum(b * beta))     # 0.875, -0.125, 0.65
```
:::

## Problem 2 — fill in the missing step

You're recovering the residual variance of standardized $y$ for a subset with
$b = (0.5, 0.5)$ and $R = \begin{pmatrix}1 & 0.5\\0.5 & 1\end{pmatrix}$.

- Step 1: $\hat\beta = R^{-1}b = \frac{1}{0.75}(0.5 - 0.25,\ 0.5 - 0.25) = \frac{1}{0.75}(0.25, 0.25) = (0.333, 0.333)$.
- Step 2: **your step here** — compute $R^2 = b^\top \hat\beta$.
- Step 3: residual variance $= 1 - R^2$.

::: {.callout-note collapse="true" title="Answer"}
Step 2: $R^2 = b^\top\hat\beta = 0.5(0.333) + 0.5(0.333) = 0.333$. Step 3: residual
variance $= 1 - 0.333 = 0.667$. (Two correlated indicators, each marginally 0.5, jointly
explain only a third of the variance — the overlap is discounted.)
:::

## Problem 3 — summaries vs full data, in R

Generate a small correlated panel, then reconstruct its joint fit from *summaries*
($R = $ `cor(X)`, $b = $ `cor(X, y)`) and confirm it matches `lm()` on the standardized raw
data.

```{webr}
#| autorun: false
# p3-starter — build R and b, reconstruct, and compare to lm on standardized data
# PREDICT: will solve(R, b) match coef(lm(scale(y) ~ scale(X) - 1)) to full precision?
set.seed(7); n <- 150
z <- rnorm(n); X <- cbind(z + rnorm(n, sd = .4), z + rnorm(n, sd = .4), rnorm(n))
y <- X %*% c(1, 0, 0.5) + rnorm(n)
```

::: {.callout-note collapse="true" title="Answer"}
The summary reconstruction reproduces `lm()` exactly (to ~1e-14).

```{r}
set.seed(7); n <- 150
z <- rnorm(n); X <- cbind(z + rnorm(n, sd = .4), z + rnorm(n, sd = .4), rnorm(n))
y <- X %*% c(1, 0, 0.5) + rnorm(n)
R <- cor(X); b <- as.numeric(cor(X, y))
rbind(from_summaries = solve(R, b),
      from_lm        = unname(coef(lm(scale(y) ~ scale(X) - 1))))
```
:::

## Problem 4 — spot the error

A colleague writes: *"Each study only published marginal coefficients, so we can't possibly
recover the joint effects — we'd need the raw data to run a multiple regression."* This
contains **exactly one mistake** — find it.

::: {.callout-note collapse="true" title="Answer"}
The mistake: believing joint effects require raw data. Given the marginal coefficients $b$
**and** a correlation matrix $R$ among the predictors, the joint fit is exactly
$\hat\beta = R^{-1}b$ — reconstructed from those two summaries, no raw rows. That identity is
the whole basis of the meta-analysis setting; the correlation matrix supplies the "how the
predictors overlap" information that raw data would otherwise carry.
:::

## Problem 5 — mixing it with Multiple Regression *(uses the normal equations)*

In the last course the joint fit solved $X^\top X\,\hat\beta = X^\top y$. Explain, in a
sentence or two, why $\hat\beta = R^{-1}b$ is the *same* fit once variables are standardized —
i.e. what $R$ and $b$ correspond to.

::: {.callout-note collapse="true" title="Answer"}
When every column is standardized (mean 0, SD 1), $X^\top X/n$ becomes the correlation matrix
$R$ and $X^\top y/n$ becomes the marginal-coefficient vector $b$ (a standardized simple-
regression slope equals the correlation). Dividing the normal equations $X^\top X\hat\beta = X^\top y$
by $n$ gives $R\hat\beta = b$, so $\hat\beta = R^{-1}b$ is the identical least-squares fit,
just written in the summary ingredients.
:::

## Problem 6 — transfer: combine two studies, then reconstruct

Two studies report marginal coefficients for the same two indicators, with standard errors:
Study A gives $b = (0.60, 0.30)$ with SE $(0.10, 0.10)$; Study B gives $b = (0.70, 0.20)$ with
SE $(0.05, 0.05)$. Inverse-variance-average them into one $b$, then (with
$R = \begin{pmatrix}1 & 0.5\\0.5 & 1\end{pmatrix}$) reconstruct the joint fit. *(This is the
meta-analysis pipeline the capstone assembles.)*

```{webr}
#| autorun: false
# p6-starter — weight each study by 1/SE^2, average, then reconstruct
est <- rbind(A = c(0.60, 0.30), B = c(0.70, 0.20))
se  <- rbind(A = c(0.10, 0.10), B = c(0.05, 0.05))
R   <- matrix(c(1, 0.5, 0.5, 1), 2)
```

::: {.callout-note collapse="true" title="Answer"}
Study B is 4× as precise (SE half ⇒ weight $1/\text{SE}^2$ four times larger), so the average
leans toward B: $b \approx (0.68, 0.22)$. Reconstructing gives the joint fit for the combined
evidence.

```{r}
est <- rbind(A = c(0.60, 0.30), B = c(0.70, 0.20)); se <- rbind(A = c(0.10, 0.10), B = c(0.05, 0.05))
w <- 1 / se^2; b_comb <- colSums(w * est) / colSums(w)
R <- matrix(c(1, 0.5, 0.5, 1), 2)
list(b_combined = round(b_comb, 3), joint = round(solve(R, b_comb), 3))
```
The precision-weighted $b$ leans toward the tighter Study B; $R^{-1}$ then turns it into the
joint effects.
:::
