---
title: "PCA and SVD: finding the directions that matter"
subtitle: "How the covariance matrix's hidden directions compress a many-column dataset into a few numbers that capture most of the information"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

## Where we're going

You know that a covariance matrix packs all pairwise co-movements into one grid, and
you know how to project a vector onto a direction. Now: that covariance grid has hidden
directions built into it -- directions along which the data spreads the most. **PCA**
(principal component analysis) finds those directions. **SVD** (singular value
decomposition) is the machinery that does the finding. The payoff: take a dataset with
50 columns and compress it to 3 or 4 numbers per row that capture most of the
information.

**Builds on:**
[distance, similarity & geometry](../distance-similarity-and-geometry/lesson.qmd)
(projection, orthogonality),
[matrices & linear transforms](../matrices-and-linear-transforms/lesson.qmd)
(matrix multiply, transpose, $X^\top X$),
[variance structure & standardization](../variance-structure-and-standardization/lesson.qmd)
(covariance matrix, standardization, `cov()`, `cor()`)

## Warm-up: do you remember?

Three quick reps first -- answers from memory, then check.

**1.** Project vector `a <- c(3, 4)` onto direction `b <- c(1, 0)`. What is the
projection length? (The formula: dot product of a and b, divided by the length of b.)

```{webr}
#| autorun: false
# warmup1-starter -- edit and Run to check
a <- c(3, 4)
b <- c(1, 0)
```

::: {.callout-note collapse="true" title="Answer"}
The dot product is $3 \times 1 + 4 \times 0 = 3$, and the length of b is
$\sqrt{1^2 + 0^2} = 1$, so the projection length is $3 / 1 = 3$. In R:
`sum(a * b) / sqrt(sum(b^2))` returns 3. Taught in
[distance, similarity & geometry](../distance-similarity-and-geometry/lesson.qmd);
revisit if shaky.
:::

**2.** If two vectors are orthogonal (their dot product is 0), what does that mean in
plain words?

::: {.callout-note collapse="true" title="Answer"}
They are at right angles to each other -- neither has any shadow on the other, so they
carry separate, non-overlapping information. In R: `sum(u * w) == 0` is the test.
Taught in
[distance, similarity & geometry](../distance-similarity-and-geometry/lesson.qmd).
:::

**3.** What does the diagonal of a covariance matrix tell you? And the off-diagonal?

::: {.callout-note collapse="true" title="Answer"}
The diagonal holds each column's own variance. The off-diagonal holds the pairwise
covariances -- how each pair of columns moves together. And the matrix is symmetric:
entry $(j, k)$ equals entry $(k, j)$. Taught in
[variance structure & standardization](../variance-structure-and-standardization/lesson.qmd);
revisit if shaky.
:::


## Sub-spine A: The covariance matrix as a matrix product {#sub-spine-a}

### A tiny example first

We measured 4 people's **height** (cm) and **weight** (kg):

```{r toy-data}
height <- c(160, 170, 175, 180)
weight <- c(55, 65, 68, 75)
X <- cbind(height, weight)   # 4-by-2 data matrix: one ROW per person, one COLUMN per variable
X
```

You already know `cov(X)` gives the covariance matrix -- the grid whose diagonal is
each column's variance and whose off-diagonal is how the columns move together. Let's
see it:

```{r cov-direct}
S <- cov(X)
S
```

The covariance matrix is roughly $\begin{bmatrix} 72.9 & 70.4 \\ 70.4 & 68.9
\end{bmatrix}$. Both columns spread a lot (diagonal), and they move together strongly
(off-diagonal nearly as big as the diagonal).

Now here is the connection to matrices. First, **center** the data -- subtract each
column's mean:

```{r center}
# Column means: height = 171.25, weight = 65.75
col_means <- colMeans(X)
col_means

# Subtract them: centered data (deviations from the mean)
X_c <- scale(X, center = TRUE, scale = FALSE)
X_c
```

Now watch: the matrix product $X_c^\top X_c$ divided by $(n - 1)$ gives the exact same
covariance matrix. Reading it aloud: "ex-sub-c transpose, ex-sub-c, over n minus one" --
in plain words, multiply the centered data table by its own transpose (flip its rows and
columns, then matrix-multiply), and divide by one less than the number of observations.
$X_c$ is the centered data matrix (4-by-2, here), the raised $\top$ means transpose, and
$n$ is the number of observations (4, here).

```{r cov-as-matrix-product}
# t(X_c) is 2-by-4; X_c is 4-by-2; their product is 2-by-2
t(X_c) %*% X_c / (4 - 1)   # (n-1) = 3

# Compare to cov(X):
S
```

They match. The covariance matrix you already know from
[variance structure](../variance-structure-and-standardization/lesson.qmd) **is** a
matrix product you already know from
[matrices](../matrices-and-linear-transforms/lesson.qmd).

### Getting a feel for it

**Why does $X_c^\top X_c$ give covariances?** Each entry of $X_c^\top X_c$ is a dot
product between two centered columns. The dot product of centered-height with
centered-weight is $(-11.25)(-10.75) + (-1.25)(-0.75) + (3.75)(2.25) + (8.75)(9.25)$.
That sum of products of paired deviations is exactly the numerator of the covariance
formula from
[mean, variance & covariance](../mean-variance-covariance/lesson.qmd) -- all we need
to finish it off is dividing by $(n - 1)$. When a column is dotted with *itself*, the
sum of products of deviations is the numerator of that column's variance. So the
diagonal of $X_c^\top X_c / (n - 1)$ is each column's variance, and the off-diagonal
is each pair's covariance -- the covariance matrix, assembled by the matrix machinery
you already own.

This means: every time a paper writes "the sample covariance matrix," it's talking
about a matrix product of centered data with itself (transposed). One recipe, two
views.


## Sub-spine B: Eigenvalues and eigenvectors {#sub-spine-b}

### A tiny example first

Take the covariance matrix $S$ we just computed. What happens when you multiply $S$ by
a vector? Normally, matrix multiplication both **stretches** and **rotates** a vector
-- it changes the vector's length and its direction. But some special vectors are only
stretched, never rotated. Those are **eigenvectors**, and the stretch factor for each
one is its **eigenvalue**.

Let's find them. R's `eigen()` does the work:

```{r eigen-decomp}
e <- eigen(S)

# Two eigenvalues: the stretch factors
e$values

# Two eigenvectors: the special directions (each is a column of e$vectors)
e$vectors
```

The first eigenvalue is about **141.4** and the second is about **0.47**. The first
eigenvector is roughly $(-0.717, -0.697)$ and the second is roughly $(0.697, -0.717)$.

Now verify what "eigenvector" means -- by hand. Take the first eigenvector and multiply
$S$ by it. Then separately multiply the eigenvalue by the same vector. They should give
the same result:

```{r eigen-verify}
v1 <- e$vectors[, 1]         # first eigenvector
lambda1 <- e$values[1]       # first eigenvalue (about 141.4)

# Multiply S by v1 (matrix times vector)
S %*% v1

# Multiply lambda1 by v1 (scalar times vector)
lambda1 * v1
```

They match. When $S$ acts on its eigenvector, all it does is scale the vector by the
eigenvalue -- no rotation. The matrix pushed this vector straight along its own
direction, just faster (or slower).

One more check: the two eigenvectors should be **perpendicular** to each other
(orthogonal -- dot product 0), just like the orthogonal vectors in the
[distance & geometry](../distance-similarity-and-geometry/lesson.qmd) lesson:

```{r ortho-check}
v2 <- e$vectors[, 2]     # second eigenvector
sum(v1 * v2)              # dot product: 0 = orthogonal
```

Zero. They are at right angles.

### Getting a feel for it

**The funnel picture.** Imagine pushing a ball through a funnel. Most directions
deflect the ball -- it enters heading one way and exits heading another. But one
direction -- straight through the center -- just speeds the ball up or slows it down
without deflecting it at all. That direction is an eigenvector. The speed-up factor is
the eigenvalue. A large eigenvalue means the matrix stretches a lot along that
direction; a small eigenvalue means it barely stretches at all. (Where the analogy
breaks: the funnel has one special direction; a matrix has as many eigenvectors as it
has rows/columns.)

**Why symmetric matrices are special.** The covariance matrix is symmetric -- entry
$(1, 2)$ equals entry $(2, 1)$ -- you learned this in
[variance structure](../variance-structure-and-standardization/lesson.qmd). Symmetric
matrices always have **real** eigenvalues (not complex numbers)
[@austin2022understandinglinearalgebra, p. 421] and **perpendicular**
eigenvectors [@austin2022understandinglinearalgebra, pp. 421--422]. This is why PCA works: the "special directions" of the covariance matrix
are guaranteed to be at right angles to each other, so they carry separate,
non-overlapping information -- exactly the orthogonality idea from the
[distance & geometry](../distance-similarity-and-geometry/lesson.qmd) lesson.

**What the eigenvalues feel like.** Our first eigenvalue (141.4) is enormous compared
to the second (0.47). That tells you the covariance matrix stretches almost entirely
along one direction and barely stretches at all along the perpendicular direction. As
we'll see next, that "big-stretch direction" is where the data spreads the most -- and
that is PC1.

### What the notation is going to say

The formula below will say: when you multiply the covariance matrix $S$ by a special
vector $\mathbf{v}$, the result is just a number $\lambda$ times that same vector --
the matrix stretched it without rotating it. In our example, that is exactly the
"multiply $S$ by $(-0.717, -0.697)$ and get $141.4 \times (-0.717, -0.697)$ back" step
we just verified.

We use $S$ for the sample covariance matrix (not the capital sigma $\Sigma$ you saw in
[variance structure](../variance-structure-and-standardization/lesson.qmd)) to avoid
confusion with the $\Sigma$ in the SVD decomposition $X = U\Sigma V^\top$ that comes
later. Same concept, different letter, chosen to keep the two apart.

### How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $S$ | "ess" | sample covariance matrix (in R: `cov(X)`) | the 2-by-2 grid $\approx \begin{bmatrix} 72.9 & 70.4 \\ 70.4 & 68.9 \end{bmatrix}$ |
| $X$ | "ex" | the raw (uncentered) data matrix -- one row per observation, one column per variable (in R: `cbind(height, weight)`) | our 4-by-2 raw height/weight matrix |
| $X_c$ | "ex sub see" or "centered ex" | data matrix with column means subtracted (in R: `scale(X, center=TRUE, scale=FALSE)`) | our 4-by-2 centered height and weight |
| $n$ | "en" | the number of observations (in R: `nrow(X)`) | 4 |
| $\lambda$ | "lambda" | eigenvalue -- how much the matrix stretches along one eigenvector direction (in R: `eigen(S)$values`) | $\lambda_1 \approx 141.4$, $\lambda_2 \approx 0.47$ |
| $\mathbf{v}$ | "vee" (bold) | eigenvector -- a direction the matrix only stretches, never rotates (in R: `eigen(S)$vectors`) | $\mathbf{v}_1 \approx (-0.717, -0.697)$ |
| $S\mathbf{v} = \lambda\mathbf{v}$ | "ess times vee equals lambda times vee" | the eigenvalue equation: applying $S$ to $\mathbf{v}$ just scales $\mathbf{v}$ by $\lambda$ | verified by multiplying $S$ by $\mathbf{v}_1$ |
| $U$ | "yoo" | left singular vectors (in SVD; in R: `svd(X_c)$u`) | a 4-by-2 matrix from `svd(X_c)$u` |
| $\Sigma$ (in SVD) | "sigma" (capital) | diagonal matrix of singular values (in SVD; in R: `svd(X_c)$d`) | singular values $\approx 20.6$ and $1.19$ |
| $d_k$ | "dee sub kay" | the $k$-th singular value -- the $k$-th number on $\Sigma$'s diagonal ("sub k" means: the k-th one; in R: `svd(X_c)$d[k]`) | $d_1 \approx 20.6$, $d_2 \approx 1.19$ |
| $V$ | "vee" (capital) | right singular vectors = the principal component directions (in R: `svd(X_c)$v` or `prcomp(X)$rotation`) | matches `prcomp(X)$rotation` |
| $\mathbf{u}^\top S \mathbf{u}$ | "yoo transpose ess yoo" | variance of data projected onto direction $\mathbf{u}$ | equals $\lambda$ when $\mathbf{u}$ is an eigenvector of $S$ |

### The formal version

The eigenvalue equation:

$$
S\mathbf{v} = \lambda\mathbf{v}
$$

Reading it aloud: "ess times vee equals lambda times vee" -- or in everyday words:
**"multiplying the covariance matrix by this special direction just scales it by a
number, without rotating it."** [@austin2022understandinglinearalgebra, p. 230]

Mapping onto the toy example:

- $S$ is our 2-by-2 covariance matrix
- $\mathbf{v}_1 = (-0.717, -0.697)$ is the first eigenvector
- $\lambda_1 = 141.4$ is the first eigenvalue
- $S \mathbf{v}_1$ gave us $(-101.4, -98.5)$, and $141.4 \times (-0.717, -0.697)$
  gave us the same $(-101.4, -98.5)$ -- verified.

**Why the eigenvalue is the variance along that direction (the punchline).** If you
project all your centered data points onto an eigenvector direction and compute the
variance of those projections, you get exactly the eigenvalue. In formula:

$$
\mathbf{u}^\top S \mathbf{u} = \lambda \quad \text{when } \mathbf{u} \text{ is a unit eigenvector of } S
$$

Reading it aloud: "yoo transpose ess yoo equals lambda" -- the variance of the data's
shadow on direction $\mathbf{u}$ equals the eigenvalue for that direction, provided
$\mathbf{u}$ has unit length ($\lVert \mathbf{u} \rVert = 1$). R's `eigen()` returns
unit eigenvectors by default [@austin2022understandinglinearalgebra, p. 431]. We will
verify this numerically in sub-spine C.


## Sub-spine C: PCA = eigenvectors of S, ordered by eigenvalue {#sub-spine-c}

### A tiny example first

Using the same centered data $X_c$ (4 people, height and weight), project every point
onto the first eigenvector. That gives one number per person -- their "score" along
that direction:

```{r project-pc1}
# Scores = centered data projected onto eigenvector 1
scores_1 <- X_c %*% v1
scores_1

# The variance of these scores equals the first eigenvalue
var(as.vector(scores_1))
lambda1
```

The variance of the scores on the first eigenvector is **141.4** -- exactly $\lambda_1$.
Now do the same for the second eigenvector:

```{r project-pc2}
scores_2 <- X_c %*% v2
scores_2
var(as.vector(scores_2))
e$values[2]   # lambda2
```

The variance of the scores on the second eigenvector is **0.47** -- exactly
$\lambda_2$. The first eigenvector "captures" far more of the data's spread than the
second. That is why it is **PC1** (principal component 1): it is the direction of
maximum variance.

**This is PCA.** Line up the eigenvectors from largest eigenvalue to smallest. Call
them PC1, PC2, and so on. Each one is an orthogonal direction in the data, and each
one's eigenvalue tells you how much variance it captures
[@austin2022understandinglinearalgebra, p. 447]. The first few PCs often
capture the lion's share of the total variance, so you can throw the rest away and
still keep most of the information.

### Getting a feel for it

**PCA is finding the "best camera angle."** Imagine you have a 3D sculpture and one
photograph to take. The best angle is the one that shows the most variation in the
sculpture's shape -- the one where the shadow of the sculpture is widest. PC1 is that
best angle. PC2 is the next-best angle at right angles to it. Each PC is a camera angle
that captures as much of the remaining variation as possible, and no two cameras point
in overlapping directions.

**Why build it from the covariance matrix?** Because the covariance matrix already
holds all the information about how the data spreads and co-moves. Its eigenvectors are
the directions of maximum spread, and its eigenvalues are the amounts of spread along
those directions. PCA just reads those directions out of the matrix you already know
how to build.

### Checking with prcomp()

R's `prcomp()` is the standard PCA function. Let's compare its output to our `eigen()`
results:

```{r prcomp-compare}
p <- prcomp(X)

# The "rotation" matrix = the eigenvectors (the PC directions)
p$rotation

# Compare to eigen:
e$vectors
```

The numbers match up to a **sign flip** -- `prcomp()` gives $(0.717, 0.697)$ where
`eigen()` gave $(-0.717, -0.697)$. This is fine: an eigenvector and its negative point
in opposite directions along the same line, like "north" and "south" on the same axis.
The direction matters, not which end you call positive.

```{r prcomp-eigenvalues}
# prcomp's sdev^2 = the eigenvalues
p$sdev^2

# Compare to eigen:
e$values
```

`prcomp(X)$sdev` gives the standard deviations of each PC; squaring them gives the
eigenvalues. Everything matches.

### The SVD connection

There is a second way to get PCA: the **singular value decomposition** (SVD). Instead
of computing the covariance matrix and then finding its eigenvectors, SVD works
directly on the centered data. Any matrix can be decomposed as $X = U \Sigma V^\top$
[@austin2022understandinglinearalgebra, p. 464]
-- three pieces that R's `svd()` hands you:

```{r svd-connection}
s <- svd(X_c)

# V from SVD = the rotation from prcomp (the PC directions)
s$v
p$rotation

# The singular values (diagonal of Sigma)
s$d

# Singular values squared, divided by (n-1), = the eigenvalues
s$d^2 / (4 - 1)
e$values
```

The match is exact. Here is the relationship in words:

- $V$ from SVD = the eigenvectors of the covariance matrix = the PC directions
  (`prcomp(X)$rotation`)
- The singular values $d_k$, squared and divided by $(n - 1)$, give the eigenvalues =
  the variance along each PC [@austin2022understandinglinearalgebra, p. 478]

**Why does `prcomp()` use SVD internally?** Computing `eigen(cov(X))` requires building
the covariance matrix first, which can lose numerical precision when the numbers are
large or the columns are nearly collinear. SVD works on the raw centered data and is
more numerically stable. The result is mathematically identical.

This is **not** a full SVD tutorial -- just the connection you need: PCA directions
come from SVD's $V$, and the variances come from the singular values. When a paper
writes $X = U \Sigma V^\top$, you know where the PCA pieces live.

### Scree plot: how many PCs to keep?

With only 2 columns, our toy data has only 2 PCs, so a scree plot would be just two
bars. To see a meaningful scree plot, here is the classic `iris` dataset (4 numeric
columns -- sepal length, sepal width, petal length, petal width):

```{r scree-iris}
#| fig-alt: "Bar chart showing variance explained by each of four principal components of the iris dataset, with PC1 capturing about 73% and PC2 about 23%"
ip <- prcomp(iris[, 1:4], scale. = TRUE)

# Proportion of variance explained by each PC
prop_var <- ip$sdev^2 / sum(ip$sdev^2)
round(prop_var * 100, 1)

# Cumulative proportion
round(cumsum(prop_var) * 100, 1)

# The scree plot: a bar for each PC's share of the total variance
barplot(prop_var * 100, names.arg = paste0("PC", 1:4),
        ylab = "% of variance explained",
        main = "Scree plot: iris (4 columns, scaled)",
        col = "steelblue", border = NA)
```

PC1 captures about 73% of the total variance, and PC1 + PC2 together capture about
96%. Two PCs summarize 4 columns with only a 4% loss of information -- each bar is
that PC's **proportion of variance explained**, its eigenvalue divided by the sum of
all the eigenvalues [@james2021introductionstatisticallearning, pp. 503--504]. The
**scree plot** (the bar chart above) is the standard tool for making this call: look
for the "elbow" where the bars drop off sharply, and keep the PCs above that elbow
[@james2021introductionstatisticallearning, pp. 507--508]. One hard ceiling on the
count: the number of nonzero eigenvalues can never exceed the **rank** of the data
matrix -- if the columns are not all independent, some PCs are guaranteed to carry
zero variance [@austin2022understandinglinearalgebra, pp. 466--467]; see
[rank & linear independence](../rank-and-linear-independence/lesson.qmd).

## Explore it

```{webr}
#| autorun: true
#| fig-alt: "Scatter plot of two correlated variables with arrows showing PC1 and PC2 directions overlaid on the data"
# explore-1 -- LIVE: PCA on correlated 2D data, with PC directions as arrows.
# PREDICT: if I INCREASE the correlation (change 0.8 to 0.95), will PC1 explain MORE
# or LESS of the total variance? Commit to an answer, then run.
set.seed(42)
n <- 100
r <- 0.8   # TRY IT: change this to 0.3, then 0.95, then re-run
x1 <- rnorm(n)
x2 <- r * x1 + sqrt(1 - r^2) * rnorm(n)   # x2 correlated with x1

dat <- cbind(x1, x2)
pc <- prcomp(dat)
prop <- pc$sdev^2 / sum(pc$sdev^2)

plot(dat, pch = 16, col = "grey60", asp = 1,
     xlab = "x1", ylab = "x2",
     main = sprintf("PC1 explains %.0f%% of variance (r = %.1f)", prop[1]*100, r))
# Draw PC directions as arrows from the center of the cloud
ctr <- colMeans(dat)
for (k in 1:2) {
  arrows(ctr[1], ctr[2],
         ctr[1] + pc$rotation[1, k] * pc$sdev[k],
         ctr[2] + pc$rotation[2, k] * pc$sdev[k],
         col = c("firebrick", "steelblue")[k], lwd = 3, length = 0.12)
}
legend("topleft", c("PC1", "PC2"), col = c("firebrick", "steelblue"), lwd = 3)
```

```{webr}
#| autorun: false
# explore-2 -- LIVE: change the correlation and watch PC1 rotate.
# TRY IT: set r to 0.0. What happens to the PC1 arrow? Why?
# Then set r to -0.9 and see what changes.
set.seed(42)
n <- 100
r <- 0.3   # CHANGE ME: try 0.0, 0.5, -0.9
x1 <- rnorm(n)
x2 <- r * x1 + sqrt(1 - r^2) * rnorm(n)

pc <- prcomp(cbind(x1, x2))
cat("PC1 direction:", round(pc$rotation[, 1], 3), "\n")
cat("PC1 explains:", round(pc$sdev[1]^2 / sum(pc$sdev^2) * 100, 1), "% of variance\n")
cat("PC2 explains:", round(pc$sdev[2]^2 / sum(pc$sdev^2) * 100, 1), "% of variance\n")
```

```{webr}
#| autorun: false
#| fig-alt: "Two side-by-side scree plots for the iris dataset: unscaled PCA where PC1 towers over the rest, and scaled PCA where the variance is spread more evenly across the first two components"
# explore-3 -- LIVE: iris PCA, unscaled vs scaled, side by side.
# PREDICT: will scaling (scale. = TRUE) change which PCs matter? Commit, then run.
ip_raw    <- prcomp(iris[, 1:4], scale. = FALSE)
ip_scaled <- prcomp(iris[, 1:4], scale. = TRUE)

prop_raw    <- ip_raw$sdev^2    / sum(ip_raw$sdev^2)
prop_scaled <- ip_scaled$sdev^2 / sum(ip_scaled$sdev^2)

par(mfrow = c(1, 2))
barplot(prop_raw * 100, names.arg = paste0("PC", 1:4),
        ylab = "% of variance explained",
        main = "Unscaled", col = "steelblue", border = NA)
barplot(prop_scaled * 100, names.arg = paste0("PC", 1:4),
        ylab = "% of variance explained",
        main = "Scaled", col = "steelblue", border = NA)
par(mfrow = c(1, 1))

cat("UNSCALED proportion:", round(prop_raw * 100, 1), "\n")
cat("UNSCALED cumulative:", round(cumsum(prop_raw) * 100, 1), "\n")
cat("SCALED   proportion:", round(prop_scaled * 100, 1), "\n")
cat("SCALED   cumulative:", round(cumsum(prop_scaled) * 100, 1), "\n")
cat("PCs needed for 90% -- unscaled:", which(cumsum(prop_raw) >= 0.90)[1],
    "| scaled:", which(cumsum(prop_scaled) >= 0.90)[1], "\n")
```

Did your prediction hold? Unscaled, PC1 alone carries about 92% of the variance
(petal length's centimeter spread dominates the raw covariance matrix), so one PC
clears 90%. Scaled, the shares rebalance to about 73% and 23%, and you need two PCs
to clear 90% -- scaling really does change which PCs matter.

## Common traps

**Trap 1 -- "PCA gives you the most important VARIABLES."** Tempting, because we talk
about "reducing columns." But PC1 is not one of the original columns -- it is a
**[linear combination](../linear-combinations-and-data-matrices/lesson.qmd)** of all
of them. In our toy data, the first PC direction is
roughly $(0.72, 0.70)$: that is $0.72 \times \text{height} + 0.70 \times \text{weight}$.
Both variables contribute almost equally. A principal component is a *direction through
variable space*, not a column you can point at.

```{r trap-1-demo}
# The rotation matrix shows HOW MUCH each variable contributes to each PC
p <- prcomp(cbind(height = c(160, 170, 175, 180),
                  weight = c(55, 65, 68, 75)))
round(p$rotation, 2)
# PC1 = 0.72*height + 0.70*weight -- both variables matter
```

**Trap 2 -- "Always run PCA on raw (unscaled) data."** If height is in millimeters
(range 1,600--1,800) and weight is in kilograms (range 55--75), the millimeter
column's variance (7,292) is about a hundred times larger than the kilogram column's
(68.9), and PC1 will just point along millimeters -- not because millimeters are more
informative, but because their numbers are bigger. This is exactly the scale-swamping
trap from
[variance structure](../variance-structure-and-standardization/lesson.qmd). The fix:
`prcomp(X, scale. = TRUE)` standardizes each column first, so all columns start on
equal footing [@james2021introductionstatisticallearning, pp. 505--506].

```{r trap-2-demo}
# height in mm (large variance), weight in kg (small variance)
height_mm <- c(1600, 1700, 1750, 1800)
weight_kg <- c(55, 65, 68, 75)

# The variances: about 7292 vs about 69 -- a factor of ~106
var(height_mm)
var(weight_kg)

p_raw <- prcomp(cbind(height_mm, weight_kg))
p_scaled <- prcomp(cbind(height_mm, weight_kg), scale. = TRUE)

# Unscaled: PC1 is almost entirely height_mm (loading 0.9954)
round(p_raw$rotation, 4)

# Scaled: both variables contribute equally to PC1 (loadings 0.7071)
round(p_scaled$rotation, 4)
```

**Trap 3 -- "PC1 is the most important predictor (or risk factor)."** Tempting,
because "principal" sounds like "most important." But PCA never looks at any outcome
-- it finds the direction of maximum **variance**, not maximum predictive power
[@chan2021probabilitydatascience, p. 311]. Trap 2's own numbers refute it: unscaled,
PC1 loads 0.9954 on `height_mm` -- not because millimeter-height predicts anything,
but because its variance (7,292) is about a hundred times the other column's (68.9).
The "leading" component can be nothing more than a scale artifact. Whether a
direction *predicts* an outcome is a separate, supervised question -- you would need
a regression against that outcome to answer it.

**Trap 4 (Optional) -- "PCA can compress anything."** One sentence on what this
buys: knowing PCA's structural limit tells you when to reach for a different tool
instead of blaming your data. PCA is **linear** -- each PC is a straight-line
direction through variable space. Data that lives on a *curved* shape has no good
low-dimensional straight-line summary [@chan2021probabilitydatascience, p. 311].
Watch it fail on a ring:

```{r trap-4-demo}
#| fig-alt: "Scatter plot of 200 points arranged in a ring (donut shape), with the caption reporting that PC1 explains 52.1% and PC2 explains 47.9% of the variance -- no dominant direction"
# 200 points on a circle plus a little noise -- a donut of data
set.seed(42)
n <- 200
theta <- runif(n, 0, 2 * pi)
ring <- cbind(x = cos(theta) + rnorm(n, sd = 0.05),
              y = sin(theta) + rnorm(n, sd = 0.05))

pr <- prcomp(ring)
prop_ring <- pr$sdev^2 / sum(pr$sdev^2)
round(prop_ring * 100, 1)   # PC1: 52.1%, PC2: 47.9% -- a near 50/50 split

plot(ring, pch = 16, col = "grey60", asp = 1,
     main = sprintf("Ring data: PC1 %.1f%%, PC2 %.1f%% -- no dominant direction",
                    prop_ring[1] * 100, prop_ring[2] * 100))
```

PC1 explains 52.1% and PC2 explains 47.9% -- an almost even split, so dropping
either PC throws away half the information. The ring really has just ONE underlying
dimension (the angle around the circle), but no *straight line* captures it. The
escape hatch is **kernel PCA**: first stretch the data through a nonlinear
transformation that unbends the curve, then run ordinary PCA on the result
[@chan2021probabilitydatascience, p. 311].

## Check yourself

**1.** A 3-by-3 covariance matrix $S$ has eigenvalues 50, 8, and 2. What percentage of
the total variance does PC1 capture?

::: {.callout-note collapse="true" title="Answer"}
Total variance = sum of eigenvalues = $50 + 8 + 2 = 60$. PC1 captures
$50 / 60 \approx 83.3\%$. The eigenvalue of a PC *is* the variance it captures, so you
just divide by the total.
:::

**2.** You run `eigen(cov(X))` and get eigenvectors $\mathbf{v}_1 = (0.6, 0.8)$ and
$\mathbf{v}_2 = (0.8, -0.6)$. Verify these are orthogonal by computing their dot
product.

```{webr}
#| autorun: false
# check2-starter -- edit and Run to check
v1 <- c(0.6, 0.8)
v2 <- c(0.8, -0.6)
```

::: {.callout-note collapse="true" title="Answer"}
Dot product: $0.6 \times 0.8 + 0.8 \times (-0.6) = 0.48 - 0.48 = 0$. Zero means
orthogonal -- the two PCs are at right angles, carrying separate information. In R:
`sum(v1 * v2)` returns 0.
:::

**3.** Explain to a colleague, in two sentences, what an eigenvalue tells you about its
eigenvector, and why PCA uses the covariance matrix.

::: {.callout-note collapse="true" title="A model answer"}
An eigenvalue tells you how much the covariance matrix stretches along its eigenvector's
direction, and that stretch equals the variance of the data when projected onto that
direction -- so a big eigenvalue means the data spreads a lot that way. PCA uses the
covariance matrix because its eigenvectors are guaranteed to be perpendicular directions
of maximum spread, which means each principal component captures as much of the data's
remaining variation as possible without overlapping with the others.
:::

## What you can now do

- **Write the covariance matrix as a matrix product**: $S = X_c^\top X_c / (n - 1)$,
  and explain why each entry is a dot product of centered columns.
- **Find eigenvalues and eigenvectors** with `eigen()` and verify the eigenvalue
  equation $S\mathbf{v} = \lambda\mathbf{v}$ by multiplying both sides.
- **Run PCA with `prcomp()`** and read its output: `$rotation` is the PC directions
  (eigenvectors), `$sdev^2` is the eigenvalues (variances along each PC).
- **Connect SVD to PCA**: `svd(X_c)$v` gives the same PC directions, and the singular
  values squared divided by $(n - 1)$ give the eigenvalues.
- **Read a scree plot** to decide how many PCs to keep, and explain why `scale. = TRUE`
  matters when columns have different units.

If any of these feels shaky, revisit [Sub-spine B](#sub-spine-b) (eigenvalues) and
[Sub-spine C](#sub-spine-c) (PCA) before continuing -- anything that uses PCA assumes
all of these.

## Where this goes next

PCA shows up wherever high-dimensional data needs to be compressed or visualized. In
genetics, the first few PCs of a genotype matrix reveal **population structure** --
groups of individuals with shared ancestry -- and those PCs are routinely included as
covariates in GWAS to prevent confounding. With the covariance matrix, eigenvalues,
eigenvectors, `prcomp()`, and scree plots in hand, you are ready for any course that
uses PCA as a building block.

Ready for more reps? Do the [practice problems](practice.qmd).
