---
title: "Practice: PCA and SVD"
subtitle: "Answers are hidden -- try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order -- they get harder. Do them on paper or in an R session; only
then open the answer.

## Problem 1 -- the toy example, new numbers

Four students took two tests. Center the data, compute the covariance matrix, find the
eigenvectors with `eigen()`, and project the centered data onto PC1.

```{webr}
#| autorun: false
# p1-starter -- edit and Run to solve, right here
math    <- c(70, 80, 90, 100)
english <- c(60, 75, 85, 80)
X <- cbind(math, english)
```

::: {.callout-note collapse="true" title="Answer"}

```{r p1-answer}
math    <- c(70, 80, 90, 100)
english <- c(60, 75, 85, 80)
X <- cbind(math, english)

# Step 1: center the data (subtract column means)
col_means <- colMeans(X)
col_means                          # math = 85, english = 75

X_c <- scale(X, center = TRUE, scale = FALSE)
X_c

# Step 2: covariance matrix
S <- cov(X)
S

# Step 3: eigendecomposition
e <- eigen(S)
e$values          # two eigenvalues
e$vectors         # two eigenvectors (columns)

# Step 4: verify Sv = lambda*v for the first eigenvector
v1 <- e$vectors[, 1]
S %*% v1
e$values[1] * v1   # should match

# Step 5: project centered data onto PC1
scores_pc1 <- X_c %*% v1
scores_pc1

# Step 6: the variance of these scores should equal eigenvalue 1
var(as.vector(scores_pc1))
e$values[1]   # matches
```

The covariance matrix is `cov(X)`. The first eigenvalue is the larger one, and its
eigenvector is PC1. Projecting centered data onto PC1 gives each student a single
score, and the variance of those scores equals the first eigenvalue -- confirming the
core PCA relationship.
:::

## Problem 2 -- fill in the missing step

You have centered data `X_c` and its covariance matrix `S`. The worked solution for
finding how much variance PC1 captures is:

1. Compute the covariance matrix: `S <- cov(X)`
2. **Your step here:** use R to get the eigenvalues and eigenvectors of `S`.
3. Proportion of variance explained by PC1: `e$values[1] / sum(e$values)`

Fill in step 2: what R function call produces `e$values` and `e$vectors`?

```{webr}
#| autorun: false
# p2-starter -- fill in the blank, then Run to check
math    <- c(70, 80, 90, 100)
english <- c(60, 75, 85, 80)
X <- cbind(math, english)
S <- cov(X)

# YOUR STEP HERE: compute eigenvalues and eigenvectors of S
# e <- ???

# Then check:
# e$values[1] / sum(e$values)   # proportion explained by PC1
```

::: {.callout-note collapse="true" title="Answer"}

```{r p2-answer}
math    <- c(70, 80, 90, 100)
english <- c(60, 75, 85, 80)
X <- cbind(math, english)
S <- cov(X)

# The missing step: eigendecompose the covariance matrix
e <- eigen(S)

# Now the proportion:
e$values[1] / sum(e$values)
```

The missing step is `e <- eigen(S)`. The `eigen()` function takes a square matrix and
returns `$values` (eigenvalues) and `$vectors` (eigenvectors as columns). Dividing the
first eigenvalue by the sum of all eigenvalues gives the proportion of total variance
captured by PC1.
:::


## Problem 3 -- full PCA with prcomp()

A dataset has 6 observations and 3 variables. Run `prcomp()`, interpret the rotation
matrix and `sdev`, and make a scree plot.

```{webr}
#| autorun: false
# p3-starter -- your job: run PCA, interpret the output, make a scree plot
X <- matrix(c(
  10, 20, 5,
  12, 22, 6,
  14, 18, 7,
  16, 24, 4,
  18, 26, 8,
  20, 28, 3
), nrow = 6, byrow = TRUE)
colnames(X) <- c("A", "B", "C")
```

::: {.callout-note collapse="true" title="Answer"}

```{r p3-answer}
#| fig-alt: "Scree plot with three bars showing percent variance explained by each of three principal components"
X <- matrix(c(
  10, 20, 5,
  12, 22, 6,
  14, 18, 7,
  16, 24, 4,
  18, 26, 8,
  20, 28, 3
), nrow = 6, byrow = TRUE)
colnames(X) <- c("A", "B", "C")

# Run PCA
pc <- prcomp(X)

# The rotation matrix: each column shows how much each variable contributes to that PC
round(pc$rotation, 3)

# Eigenvalues (variance along each PC) = sdev^2
pc$sdev^2

# Proportion of variance explained
prop <- pc$sdev^2 / sum(pc$sdev^2)
round(prop * 100, 1)   # percentages

# Cumulative proportion
round(cumsum(prop) * 100, 1)

# Scree plot
barplot(prop * 100, names.arg = paste0("PC", 1:3),
        ylab = "% of variance explained",
        main = "Scree plot", col = "steelblue", border = NA)
```

Reading the output: `pc$rotation` shows the PC directions. Each column is an
eigenvector telling you the linear combination of original variables that defines that
PC. `pc$sdev^2` gives the eigenvalue (variance) for each PC. The scree plot shows how
much variance each PC captures -- if the first two capture most of it, you could
summarize these 3 columns with just 2 numbers per row.
:::


## Problem 4 -- spot the error

A colleague ran PCA on data where column A is height in millimeters (range
1,500--1,900) and column B is a test score (range 0--100). They report: "PC1
explains 98.6% of the variance, so column B is unimportant and can be dropped."

This conclusion contains **exactly one mistake**. Find it and explain the fix.

```{webr}
#| autorun: false
# p4-starter -- the colleague's data, to check your reasoning in R
height_mm <- c(1520, 1600, 1650, 1700, 1780, 1850, 1900)
score     <- c(35, 60, 72, 55, 88, 45, 66)
X <- cbind(height_mm, score)
```

::: {.callout-note collapse="true" title="Answer"}
The mistake is the conclusion "column B is unimportant." PC1 explains 98.6% of the
variance because column A's raw variance (about 18,729) is roughly *sixty times
larger* than column B's (about 306) -- the millimeter column's spread swamps
everything else. This is the scale-domination trap, not evidence that B is
uninformative.

```{r p4-answer}
height_mm <- c(1520, 1600, 1650, 1700, 1780, 1850, 1900)
score     <- c(35, 60, 72, 55, 88, 45, 66)
X <- cbind(height_mm, score)

# The raw variances: about 18729 vs about 306 -- a ~61x gap
var(height_mm)
var(score)

# Raw PCA: PC1 explains 98.6% -- a pure scale artifact
p_raw <- prcomp(X)
round(p_raw$sdev^2 / sum(p_raw$sdev^2) * 100, 1)

# Scaled PCA: the split becomes 67.7% / 32.3% -- column B carries real information
p_scaled <- prcomp(X, scale. = TRUE)
round(p_scaled$sdev^2 / sum(p_scaled$sdev^2) * 100, 1)
```

The fix: run `prcomp(X, scale. = TRUE)` to standardize both columns to mean 0 and
standard deviation 1 before computing PCs. After scaling, both columns start on equal
footing and PC1's share drops from 98.6% to 67.7% -- the "column B is unimportant"
story evaporates. This is exactly the lesson's Trap 2.
:::


## Problem 5 -- mixing it with variance-structure-and-standardization *(uses variance-structure's covariance vs. correlation matrix)*

Compute the covariance matrix and the correlation matrix of the data below. Run PCA on
both the raw data (`prcomp(X)`) and the scaled data (`prcomp(X, scale. = TRUE)`). Why
do the PC directions differ?

```{webr}
#| autorun: false
# p5-starter -- edit and Run to solve, right here
temp   <- c(15, 20, 25, 30, 35)       # temperature in Celsius
sales  <- c(200, 350, 400, 600, 750)  # ice cream sales in dollars
X <- cbind(temp, sales)
```

::: {.callout-note collapse="true" title="Answer"}

```{r p5-answer}
temp   <- c(15, 20, 25, 30, 35)
sales  <- c(200, 350, 400, 600, 750)
X <- cbind(temp, sales)

# Covariance matrix (raw units)
cov(X)

# Correlation matrix (standardized units, all on [-1, 1])
cor(X)

# PCA on raw data (uses covariance matrix)
pc_raw <- prcomp(X)
round(pc_raw$rotation, 3)
round(pc_raw$sdev^2 / sum(pc_raw$sdev^2) * 100, 1)

# PCA on scaled data (uses correlation matrix)
pc_scaled <- prcomp(X, scale. = TRUE)
round(pc_scaled$rotation, 3)
round(pc_scaled$sdev^2 / sum(pc_scaled$sdev^2) * 100, 1)
```

The PC directions differ because the raw covariance matrix is dominated by `sales`
(variance in the tens of thousands) while `temp` (variance = 62.5) is tiny by
comparison. Raw PCA points PC1 almost entirely along `sales`. After scaling, both
columns have variance 1, so the PCA reflects their *correlation pattern* rather than
their raw scales. `prcomp(X, scale. = TRUE)` is PCA on the correlation matrix, which
is the covariance matrix of the standardized columns --
this connects the standardizing idea from
[variance structure](../variance-structure-and-standardization/lesson.qmd)
to PCA.
:::


## Problem 6 -- transfer: PCA in genetics

A genetics paper says: "We included the first 5 PCs as covariates to correct for
population stratification." Explain in your own words: (a) what those PCs are, (b)
what information they capture, and (c) why including them helps.

::: {.callout-note collapse="true" title="Answer"}
(a) The 5 PCs are the first five principal components from running PCA on the genotype
data -- a matrix where each row is a person and each column is a genetic variant. Each
PC is a direction (a linear combination of all the genetic variants) that captures a
large share of the overall genetic variation across people.

(b) The top PCs of genotype data tend to capture **population structure** -- systematic
differences in genetic ancestry between groups of people (e.g., people with mostly
European ancestry cluster apart from people with mostly East Asian ancestry). These
ancestry-driven patterns usually account for the most variance in the genotype matrix,
so they show up as the leading PCs.

(c) If ancestry is correlated with both the trait you are studying *and* the genetic
variants you are testing, it confounds the analysis -- you might find a variant that
looks associated with the trait when it is really just associated with ancestry.
Including the first few PCs as covariates absorbs the ancestry-driven variation, so the
remaining signal reflects genuine variant-trait associations rather than population
structure.
:::


## Problem 7 -- transfer: reading a real PCA (USArrests)

The `USArrests` dataset ships with base R: 50 states, 4 variables -- `Murder`,
`Assault`, and `Rape` (arrests per 100,000 residents) and `UrbanPop` (percent of the
population living in urban areas). This is the textbook example from
[@james2021introductionstatisticallearning, pp. 498--501].

Run `prcomp(USArrests, scale. = TRUE)` and look at the `rotation` matrix (why
`scale. = TRUE` matters here: the variables are on wildly different scales --
Problem 4 showed what happens otherwise).

(a) Which variables load together on PC1, and which variable dominates PC2? Name, in
plain words, what direction each of PC1 and PC2 measures.

(b) PREDICT before you look at any scores: where would a **high-crime, mostly rural**
state sit in the PC1--PC2 score plot (which sign on each axis)? Then check your
prediction against `pu$x["Mississippi", ]` (Mississippi: high crime, only 44% urban).

```{webr}
#| autorun: false
# p7-starter -- USArrests ships with base R; nothing to retype
head(USArrests)
pu <- prcomp(USArrests, scale. = TRUE)
```

::: {.callout-note collapse="true" title="Answer"}

```{r p7-answer}
pu <- prcomp(USArrests, scale. = TRUE)

# (a) The loadings
round(pu$rotation, 3)

# Share of variance: PC1 62%, PC2 24.7% -- together 86.8%
round(pu$sdev^2 / sum(pu$sdev^2) * 100, 1)

# (b) Check the prediction for a high-crime, rural state
round(pu$x["Mississippi", 1:2], 2)
```

(a) On PC1, the three crime variables load together and with the same sign --
`Murder` $-0.536$, `Assault` $-0.583$, `Rape` $-0.543$ -- while `UrbanPop`
contributes much less ($-0.278$). On PC2, `UrbanPop` dominates ($0.873$) and the
crime variables are secondary. In words: **PC1 is an overall-crime direction** (one
number summarizing how much serious crime a state has) and **PC2 is an urbanization
direction** (how urban the state is, mostly separate from its crime level)
[@james2021introductionstatisticallearning, pp. 498--501].

(b) The crime variables load *negatively* on PC1 here, so high crime means a
**negative PC1 score**; `UrbanPop` loads *positively* on PC2, so rural means a
**negative PC2 score**. A high-crime rural state should land in the
negative-PC1, negative-PC2 corner -- and Mississippi does: its scores are
PC1 $= -0.99$, PC2 $= -2.37$. (Remember the lesson's sign-flip note: another
machine may flip the signs of a whole column, which flips the scores with it --
the *corner* changes name, but the interpretation is identical.)
:::


## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice -- come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** --
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/pca-and-svd" pass="4"}

::: {.quiz-q qid="pca-eigenvalue-meaning" qtype="mc" answer="c"}
What does the eigenvalue of a principal component tell you?

a) The number of original variables that contribute to that particular PC
b) The correlation between that PC and the original data matrix
c) The variance of the data projected onto that PC's direction
d) The percentage of observations captured by that PC
:::

::: {.quiz-q qid="pca-ortho-check" qtype="numeric" answer="0" tolerance="0"}
Two eigenvectors of a covariance matrix are $\mathbf{v}_1 = (0.6, 0.8)$ and $\mathbf{v}_2 = (-0.8, 0.6)$. What is their dot product?
:::

::: {.quiz-q qid="pca-prop-var" qtype="numeric" answer="80" tolerance="1"}
A PCA gives eigenvalues 40, 8, and 2. What percentage of the total variance does PC1 explain? (Enter a whole number.)
:::

::: {.quiz-q qid="pca-scale-trap" qtype="mc" answer="b"}
You run PCA on a dataset where column A is in meters and column B is in millimeters. PC1 explains 99.8% of the variance. What is the most likely explanation?

a) Column A carries nearly all the real information in the dataset, so column B adds nothing
b) Column B's larger raw numbers dominate the variance, so PCA aligns along B
c) The two columns are nearly uncorrelated with each other, so one PC is enough to summarize both
d) The data genuinely varies along only one direction in variable space, so one PC suffices
:::

::: {.quiz-q qid="pca-interleave-covmat" qtype="mc" answer="d"}
*(From [variance structure & standardization](../variance-structure-and-standardization/lesson.qmd).)* A 3-by-3 covariance matrix has diagonal entries 100, 25, and 4. What does the entry 100 represent?

a) The correlation between column 1 and column 2
b) The mean of column 1
c) The covariance between columns 1 and 3
d) The variance of column 1
:::

:::

::::
