---
title: "Practice: The multivariate normal distribution"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only (a {webr} starter does not see it -- each starter re-states
# its own data). Pure base R, so nothing to install.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order -- they get harder. Do them on paper or in an R session; only
then open the answer.


## Problem 1 -- the toy example, new numbers

A bivariate normal distribution has mean vector $\boldsymbol{\mu} = (3, 7)$ and
covariance matrix

$$
\boldsymbol{\Sigma} = \begin{pmatrix} 1 & -1 \\ -1 & 4 \end{pmatrix}
$$

1. What are $\text{Var}(X_1)$, $\text{Var}(X_2)$, and $\text{Cov}(X_1, X_2)$?
2. What is the correlation between $X_1$ and $X_2$?
3. Generate 500 samples from this distribution using the Cholesky approach. Verify
   that `colMeans()` and `cov()` of the samples are close to the inputs.

```{webr}
#| autorun: false
# p1-starter -- edit and Run to solve, right here
mu    <- c(3, 7)
Sigma <- matrix(c(1, -1,
                  -1, 4), nrow = 2)
```

::: {.callout-note collapse="true" title="Answer"}
1. $\text{Var}(X_1) = 1$, $\text{Var}(X_2) = 4$, $\text{Cov}(X_1, X_2) = -1$.
   The variances are the diagonal entries; the covariance is the off-diagonal.
2. Correlation $= \text{Cov}/(sd_1 \times sd_2) = -1 / (1 \times 2) = -0.5$.
   Negative: when $X_1$ is above its mean, $X_2$ tends to be below.
3. Code below.

```{r p1-answer}
mu    <- c(3, 7)
Sigma <- matrix(c(1, -1, -1, 4), nrow = 2)

# Read off the parameters
c(var_X1 = Sigma[1, 1], var_X2 = Sigma[2, 2], cov = Sigma[1, 2])  # 1, 4, -1
Sigma[1, 2] / (sqrt(Sigma[1, 1]) * sqrt(Sigma[2, 2]))              # -0.5

# Generate samples
rmvn <- function(n, mu, Sigma) {
  p <- length(mu)
  L <- chol(Sigma)
  Z <- matrix(rnorm(n * p), n, p)
  sweep(Z %*% L, 2, mu, "+")
}

set.seed(99)
pts <- rmvn(500, mu, Sigma)
colMeans(pts)   # close to (3, 7)
cov(pts)        # close to our Sigma
```
:::


## Problem 2 -- fill in the missing step

Here is a worked computation of the Mahalanobis distance from $(4, 5)$ to the center
$(3, 7)$ under the covariance matrix above -- but **the core step is blanked**.

> Step 1: deviation = $(4 - 3,\; 5 - 7) = (1, -2)$.
>
> Step 2: **`<your step here>`** -- compute $\boldsymbol{\Sigma}^{-1}$ (the inverse of
> the covariance matrix).
>
> Step 3: quadratic form = $(1, -2)^\top \boldsymbol{\Sigma}^{-1} (1, -2)$.
>
> Step 4: $d_M = \sqrt{\text{quadratic form}}$.

Fill in step 2: what is $\boldsymbol{\Sigma}^{-1}$ for
$\boldsymbol{\Sigma} = \begin{pmatrix} 1 & -1 \\ -1 & 4 \end{pmatrix}$?
Then compute the full Mahalanobis distance.

```{webr}
#| autorun: false
# p2-starter -- compute the inverse and the Mahalanobis distance
Sigma <- matrix(c(1, -1, -1, 4), nrow = 2)
dev   <- c(1, -2)
```

::: {.callout-note collapse="true" title="Answer"}
Step 2: The inverse of a $2 \times 2$ matrix $\begin{pmatrix} a & b \\ c & d \end{pmatrix}$
is $\frac{1}{ad - bc}\begin{pmatrix} d & -b \\ -c & a \end{pmatrix}$. Here
$\det = 1 \times 4 - (-1)(-1) = 3$, so

$$
\boldsymbol{\Sigma}^{-1} = \frac{1}{3}\begin{pmatrix} 4 & 1 \\ 1 & 1 \end{pmatrix}
= \begin{pmatrix} 4/3 & 1/3 \\ 1/3 & 1/3 \end{pmatrix}
$$

Then the quadratic form: $(1, -2) \begin{pmatrix} 4/3 & 1/3 \\ 1/3 & 1/3 \end{pmatrix} (1, -2)^\top$.

First, $\boldsymbol{\Sigma}^{-1} \mathbf{d} = (4/3 \times 1 + 1/3 \times (-2),\;
1/3 \times 1 + 1/3 \times (-2)) = (2/3, -1/3)$.

Then $\mathbf{d}^\top (\boldsymbol{\Sigma}^{-1} \mathbf{d}) = 1 \times 2/3 + (-2) \times (-1/3) = 2/3 + 2/3 = 4/3 \approx 1.333$.

$d_M = \sqrt{4/3} \approx 1.155$.

```{r p2-answer}
Sigma <- matrix(c(1, -1, -1, 4), nrow = 2)
dev   <- c(1, -2)

# Step 2: the inverse
Sigma_inv <- solve(Sigma)
round(Sigma_inv, 4)                               # 4/3, 1/3, 1/3, 1/3

# Step 3: quadratic form
quad <- t(dev) %*% Sigma_inv %*% dev
quad                                               # 1.333

# Step 4: Mahalanobis distance
sqrt(as.numeric(quad))                             # 1.155

# Check with R's built-in
sqrt(mahalanobis(c(4, 5), center = c(3, 7), cov = Sigma))  # 1.155
```
:::


## Problem 3 -- 3D marginals

A 3-dimensional multivariate normal has mean vector $\boldsymbol{\mu} = (0, 5, -2)$
and covariance matrix

$$
\boldsymbol{\Sigma} = \begin{pmatrix} 2 & 1 & 0 \\ 1 & 5 & -1 \\ 0 & -1 & 3 \end{pmatrix}
$$

1. Generate 500 samples from this distribution.
2. Extract just the third column ($X_3$). What should its theoretical distribution be?
3. Verify: is the sample mean close to its theoretical mean? Is the sample variance
   close to its theoretical variance?

```{webr}
#| autorun: false
# p3-starter -- generate 3D samples and check the marginal of X3
mu    <- c(0, 5, -2)
Sigma <- matrix(c(2,  1,  0,
                  1,  5, -1,
                  0, -1,  3), nrow = 3)
```

::: {.callout-note collapse="true" title="Answer"}
The marginal of $X_3$ is $N(\mu_3, \Sigma_{33}) = N(-2, 3)$. Just read off the
third mean and the (3,3) entry of the covariance matrix.

```{r p3-answer}
mu    <- c(0, 5, -2)
Sigma <- matrix(c(2,  1,  0,
                  1,  5, -1,
                  0, -1,  3), nrow = 3)

rmvn <- function(n, mu, Sigma) {
  p <- length(mu)
  L <- chol(Sigma)
  Z <- matrix(rnorm(n * p), n, p)
  sweep(Z %*% L, 2, mu, "+")
}

set.seed(101)
pts <- rmvn(500, mu, Sigma)

# Marginal of X3 should be N(-2, 3)
cat("Theory: N(-2, 3)\n")
cat("Sample mean of X3:", round(mean(pts[, 3]), 2), " (should be near -2)\n")
cat("Sample var  of X3:", round(var(pts[, 3]),  2), " (should be near 3)\n")
```

The sample statistics are close to the theoretical values, confirming the marginal is
$N(-2, 3)$.
:::


## Problem 4 -- spot the error

A colleague built this covariance matrix for a bivariate normal and tries to generate
samples:

```
Sigma <- matrix(c(4, 5, 5, 1), nrow = 2)
L <- chol(Sigma)
```

They get an error. The code contains **exactly one mistake** -- find it and explain why
it fails.

```{webr}
#| autorun: false
# p4-starter -- try running this and diagnosing the error
Sigma <- matrix(c(4, 5, 5, 1), nrow = 2)
```

::: {.callout-note collapse="true" title="Answer"}
The matrix is **not positive definite**. The determinant is $4 \times 1 - 5 \times 5
= -21 < 0$, so `chol()` will fail with an error about a non-positive-definite matrix.

Why? The off-diagonal covariance of 5 implies a correlation of $5 / (\sqrt{4} \times
\sqrt{1}) = 5 / 2 = 2.5$, which is impossible -- correlations must live in $[-1, 1]$.
The covariance between two variables can never exceed the product of their standard
deviations. Here, $sd_1 = 2$ and $sd_2 = 1$, so the covariance must be between $-2$
and $+2$.

**Fix:** use a covariance within the valid range, e.g. `matrix(c(4, 1.5, 1.5, 1), 2, 2)`
-- this has $\det = 4 - 2.25 = 1.75 > 0$ and correlation $1.5 / 2 = 0.75$.

```{r p4-answer}
# The broken matrix
Sigma_bad <- matrix(c(4, 5, 5, 1), nrow = 2)
det(Sigma_bad)                                      # -21 -- negative!
Sigma_bad[1, 2] / (sqrt(Sigma_bad[1, 1]) * sqrt(Sigma_bad[2, 2]))  # 2.5 -- impossible

# A fixed version
Sigma_ok <- matrix(c(4, 1.5, 1.5, 1), nrow = 2)
det(Sigma_ok)                                       # 1.75 -- positive
chol(Sigma_ok)                                      # works
```
:::


## Problem 5 -- mixing it with variance-structure-and-standardization *(uses variance-structure's covariance matrix and correlation skills)*

Given a 3-variable covariance matrix:

$$
\boldsymbol{\Sigma} = \begin{pmatrix} 16 & 6 & -4 \\ 6 & 9 & 3 \\ -4 & 3 & 4 \end{pmatrix}
$$

1. What are the standard deviations of $X_1$, $X_2$, and $X_3$?
2. Compute the full $3 \times 3$ correlation matrix from $\boldsymbol{\Sigma}$.
3. If $\mathbf{X} \sim N(\boldsymbol{\mu}, \boldsymbol{\Sigma})$, what is the
   correlation between $X_1$ and $X_3$? Is it positive or negative, and what does that
   tell you about how $X_1$ and $X_3$ move?

```{webr}
#| autorun: false
# p5-starter -- compute the correlation matrix
Sigma <- matrix(c(16,  6, -4,
                   6,  9,  3,
                  -4,  3,  4), nrow = 3)
```

::: {.callout-note collapse="true" title="Answer"}
1. Standard deviations are the square roots of the diagonal entries:
   $sd_1 = \sqrt{16} = 4$, $sd_2 = \sqrt{9} = 3$, $sd_3 = \sqrt{4} = 2$.

2. Correlation matrix: divide each off-diagonal entry by the product of the
   corresponding standard deviations.
   - $r_{12} = 6 / (4 \times 3) = 0.5$
   - $r_{13} = -4 / (4 \times 2) = -0.5$
   - $r_{23} = 3 / (3 \times 2) = 0.5$

3. The correlation between $X_1$ and $X_3$ is $-0.5$ -- negative. When $X_1$ is above
   its mean, $X_3$ tends to be below its mean. They move in opposite directions.

```{r p5-answer}
Sigma <- matrix(c(16,  6, -4,
                   6,  9,  3,
                  -4,  3,  4), nrow = 3)

# Standard deviations
sds <- sqrt(diag(Sigma))                            # 4, 3, 2
sds

# Correlation matrix: cov2cor() does this, or manual
D_inv <- diag(1 / sds)                              # diagonal of 1/sd
R <- D_inv %*% Sigma %*% D_inv                      # the correlation matrix
round(R, 2)

# Check with R's shortcut
cov2cor(Sigma)

# r(X1, X3) = -0.5 -- negative, they move in opposite directions
```
:::


## Problem 6 -- transfer: interpreting a genetics paper

A genetics paper says: "We model the effect sizes as
$\boldsymbol{\beta} \sim N(\mathbf{0}, \boldsymbol{\Sigma}_\beta)$ where
$\boldsymbol{\Sigma}_\beta$ encodes LD between SNPs."

In one or two sentences, what does this assumption mean in plain language? (Hint: think
about what each piece of the multivariate normal -- the mean vector and the covariance
matrix -- controls.)

::: {.callout-note collapse="true" title="Answer"}
The effect sizes are random, centered at zero on average (the mean vector is
$\mathbf{0}$, so no SNP is expected to have a positive or negative effect *a priori*),
and **correlated with each other** according to the LD (linkage disequilibrium)
structure. Nearby SNPs that are in high LD will have similar entries in
$\boldsymbol{\Sigma}_\beta$, meaning their effect sizes tend to be similar in magnitude
and sign -- the covariance matrix encodes how tightly each pair of SNPs' effects track
each other, just like the covariance matrix in our toy example encoded how $X_1$ and
$X_2$ moved together.
:::


## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice -- come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** --
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/multivariate-normal-distribution" pass="4"}

::: {.quiz-q qid="mvn-marginal" qtype="mc" answer="c"}
If $\mathbf{X} \sim N(\boldsymbol{\mu}, \boldsymbol{\Sigma})$ in 3 dimensions, and
$\boldsymbol{\mu} = (1, 4, 9)$ with $\boldsymbol{\Sigma}$ having diagonal entries
$(2, 5, 7)$, what is the marginal distribution of $X_2$?

a) $N(4, 2)$
b) $N(5, 4)$
c) $N(4, 5)$
d) $N(4, 7)$
:::

::: {.quiz-q qid="mvn-det-toy" qtype="numeric" answer="27" tolerance="0"}
For the lesson's toy covariance matrix $\boldsymbol{\Sigma} = \begin{pmatrix} 4 & 3 \\ 3 & 9 \end{pmatrix}$, what is the determinant $\lvert\boldsymbol{\Sigma}\rvert$?
:::

::: {.quiz-q qid="mvn-posdef" qtype="mc" answer="a"}
A colleague proposes $\boldsymbol{\Sigma} = \begin{pmatrix} 1 & 2 \\ 2 & 1 \end{pmatrix}$ as a covariance matrix. What happens when they try `chol(Sigma)`?

a) It errors because the matrix is not positive definite
b) It succeeds because the matrix is symmetric and has positive entries
c) It errors because the diagonal entries must be larger than 1
d) It succeeds because all entries are positive real numbers
:::

::: {.quiz-q qid="mvn-mahal-vs-euclid" qtype="mc" answer="d"}
Why is Mahalanobis distance more informative than Euclidean distance for multivariate normal data?

a) Mahalanobis distance is always smaller than Euclidean distance for the same pair of points
b) Euclidean distance requires the covariance matrix, which may not be available in practice
c) Mahalanobis distance only uses the diagonal entries of the covariance matrix, not off-diagonals
d) It accounts for different variances and correlations between variables
:::

::: {.quiz-q qid="mvn-cov-interleave" qtype="mc" answer="b"}
*(Recall from [variance structure & standardization](../variance-structure-and-standardization/lesson.qmd).)* If a covariance matrix has $\Sigma_{11} = 9$ and $\Sigma_{22} = 16$, and $\text{Cov}(X_1, X_2) = 6$, what is the correlation $r_{12}$?

a) $6 / (9 \times 16) = 0.042$
b) $6 / (3 \times 4) = 0.5$
c) $6 / (81 \times 256) = 0.0003$
d) $6 / (9 + 16) = 0.24$
:::

:::

::::
