---
title: "Posterior probability of association"
subtitle: "Run the Bayes factor through prior odds to a probability тАФ and see why a p-value threshold must shrink with n"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only тАФ the live cells below re-state their own data. Base R only.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

::: {.callout-tip title="This lesson runs in your browser" collapse="true"}
The code cells below with a **Run** button are live тАФ powered by WebR (R compiled to
WebAssembly), they run right here on your phone or laptop, no install needed. Edit any
of them and re-run. The first run downloads a small R engine, so give it a few seconds.
Cells without a Run button are ordinary listings whose output was computed when the
page was built.
:::

## Where we're going

You have a Bayes factor for association. But a Bayes factor is only *evidence* тАФ it isn't yet
a probability that a variant is associated. This module runs the ABF through the machine you
built in the foundations: **posterior odds = prior odds ├Ч Bayes factor**, then convert to the
**posterior probability of association (PPA)**. The prior odds come from $\pi$, the small
prior probability that any given variant is truly associated (in a GWAS, maybe $1$ in
$10{,}000$). Along the way we reach the paper's headline: because a p-value's meaning depends
on sample size while an honest Bayesian threshold does not, **a fixed p-value becomes weaker
and weaker evidence as $n$ grows** тАФ so a sensible significance threshold must *shrink* with
sample size. And we'll see the specific prior that makes ABF and p-value *rankings* line up.

**Builds on:** the [previous module](../03-the-approximate-bayes-factor/lesson.qmd) тАФ the
lABF;
[bayes factors & posterior odds](../../../../foundations/bayes-factors-and-posterior-odds/lesson.qmd)
тАФ posterior odds = prior odds ├Ч Bayes factor, and probability from odds;
[probability & odds](../../../../foundations/probability-and-odds/lesson.qmd) тАФ converting
between $\pi$ and odds; and
[hypothesis tests & p-values](../../../../foundations/hypothesis-tests-p-values-and-confidence-intervals/lesson.qmd)
тАФ what a p-value is.

## Warm-up: do you remember?

Before the new stuff, three quick reps тАФ answers from memory first, then check.

**1.** *(Foundation)* Prior odds of $0.5$ and a Bayes factor of $10$. What are the posterior
odds, and the posterior probability?

::: {.callout-note collapse="true" title="Answer"}
Posterior odds $= 0.5 \times 10 = 5$. Posterior probability $= 5/(1 + 5) = 0.833$. See
[bayes factors & posterior odds](../../../../foundations/bayes-factors-and-posterior-odds/lesson.qmd).
:::

**2.** *(Module 3)* Our running variant has $Z = 4$, $V = 0.0025$, $W = 0.04$, giving lABF
$= 6.11$. What is the Bayes factor $e^{\text{lABF}}$?

```{webr}
#| autorun: false
# warmup2-starter тАФ edit and Run to check
lABF <- 6.1128
```

::: {.callout-note collapse="true" title="Answer"}
$e^{6.11} \approx 451.6$. See
[the Approximate Bayes Factor](../03-the-approximate-bayes-factor/lesson.qmd).
:::

**3.** *(Foundation)* A prior probability is $\pi = 10^{-4}$. What are the prior *odds*
$\pi/(1 - \pi)$, roughly?

::: {.callout-note collapse="true" title="Answer"}
$\pi/(1 - \pi) = 10^{-4}/(1 - 10^{-4}) \approx 1.0001 \times 10^{-4}$ тАФ for tiny $\pi$ the
odds are almost exactly $\pi$ itself. See
[probability & odds](../../../../foundations/probability-and-odds/lesson.qmd).
:::

## A tiny example first

Our running variant has lABF $= 6.11$, so its Bayes factor for association is $e^{6.11}
\approx 451.6$ тАФ the data are about $452$ times more consistent with "associated" than "not."
Is the variant *probably* associated? That depends on how rare true associations are.

In a GWAS, most variants are *not* associated with the trait, so the prior probability that
any one is takes a small value like $\pi = 10^{-4}$ (1 in 10,000). Turn that into **prior
odds**:
$$
\text{prior odds} = \frac{\pi}{1 - \pi} = \frac{10^{-4}}{1 - 10^{-4}} \approx 10^{-4}.
$$
Now multiply by the Bayes factor to get **posterior odds** (the foundation's rule):
$$
\text{posterior odds} = \text{prior odds} \times \text{BF} = 10^{-4} \times 451.6 \approx
0.0452.
$$
Finally convert odds back to a probability тАФ the **posterior probability of association**:
$$
\text{PPA} = \frac{\text{posterior odds}}{1 + \text{posterior odds}} = \frac{0.0452}{1.0452}
\approx 0.0432.
$$

So even with a Bayes factor of $452$, our variant is only about **$4.3\%$** likely to be a
true association тАФ because it started so implausible ($1$ in $10{,}000$). Strong evidence,
applied to long odds, still leaves the posterior modest. The same thing in R:

```{r toy-example}
lABF  <- 6.11
BF    <- exp(lABF)                       # 451.6

pi_prior   <- 1e-4                        # prior prob any variant is associated
prior_odds <- pi_prior / (1 - pi_prior)   # ~1e-4
post_odds  <- prior_odds * BF             # ~0.0452
PPA        <- post_odds / (1 + post_odds) # ~0.0432
c(BF = round(BF, 1), prior_odds = signif(prior_odds, 3),
  post_odds = round(post_odds, 4), PPA = round(PPA, 4))
```

## Getting a feel for it

**The mental picture: the rare-disease test, again.** This is exactly the medical-test anchor
from the [Bayes-factors
foundation](../../../../foundations/bayes-factors-and-posterior-odds/lesson.qmd), now wearing
GWAS clothes. A GWAS variant is like a patient screened for a rare disease: the prior
probability of association ($\pi = 10^{-4}$) is the disease's rarity, and the ABF is the
test's likelihood ratio. A strong "positive" (BF $= 452$) multiplies the tiny prior odds up тАФ
but $10^{-4} \times 452$ is still only about $1$ in $22$, so the variant is *probably not*
associated despite impressive-looking evidence. This is why a genome-wide scan of a million
variants needs *overwhelming* per-variant evidence before believing any single hit тАФ the same
reason a rare-disease screen needs confirmation. (Where the anchor holds perfectly: both are
"multiply long prior odds by an evidence ratio"; the only difference is the ABF comes from a
closed form rather than a test's sensitivity/specificity.)

**Why a p-value can't do this job тАФ and why its threshold must shrink with $n$.** A p-value
answers "if there were no effect, how surprising is this data?" It never asks "how rare are
real effects?" тАФ so it can't produce a posterior probability. Worse, a p-value's *evidential
meaning* depends on sample size, while the ABF's does not. Here's the mechanism: fix the
p-value (fix $Z$) and grow the sample. Bigger $n$ shrinks the standard error, so $V \to 0$ and
the shrinkage $r = W/(W + V) \to 1$. In the lABF $= \tfrac12(\log(1 - r) + rZ^2)$, the signal
term $rZ^2$ stays bounded (it's about $Z^2$), but the width penalty $\log(1 - r) \to -\infty$.
So the **same p-value gives less and less Bayesian evidence as $n$ grows тАФ eventually
favoring the null.** A p-value of $0.05$ that meant "mild evidence for an effect" in a small
study means "evidence *against* an effect" in a huge one. To keep a fixed level of real
evidence, the p-value threshold has to get *stricter* as $n$ rises тАФ Wakefield's headline.

**What the numbers feel like.** With $\pi = 10^{-4}$, our BF of $452$ gave PPA $\approx 0.043$
тАФ unconvincing. Bump the prior to $\pi = 10^{-3}$ and PPA jumps to $0.31$; to $\pi = 10^{-2}$
and it's $0.82$. And on the shrinking-threshold point: a fixed $Z = 2$ ($p \approx 0.05$) has
lABF $\approx 0.65$ (BF $\approx 1.9$) when $V = 0.04$, but lABF $\approx -3.3$ (BF $\approx
0.04$) when $V = 10^{-6}$ тАФ the *identical* p-value flipped from weak-for to strong-against as
the study grew.

## What the notation is going to say

The formulas chain three steps you've already met. Prior odds is $\pi/(1 - \pi)$ тАФ the rare
prior probability turned into odds. Posterior odds is prior odds times the Bayes factor
$e^{\text{lABF}}$ тАФ Bayes' rule on odds. PPA is posterior odds over one-plus-posterior-odds тАФ
odds back to a probability. And the threshold argument is read straight off the lABF formula:
as $V \to 0$, $r \to 1$, and $\log(1 - r) \to -\infty$ drags the lABF down for any fixed $Z$.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $\pi$ | "pi" | the **prior probability** that a given variant is truly associated | $10^{-4}$ |
| $\pi/(1-\pi)$ | "pi over one minus pi" | the prior **odds** of association | $\approx 10^{-4}$ |
| $\text{BF}$ | "bee-eff" | the Bayes factor for association, $e^{\text{lABF}}$ | $451.6$ |
| $\text{lABF}$ | "log-A-B-F" | the log Bayes factor (Module 3) | $6.11$ |
| posterior odds | тАФ | prior odds $\times$ BF | $0.0452$ |
| $\text{PPA}$ | "pee-pee-ay" | the **posterior probability of association** | $0.0432$ |
| $r$ | "little r" | shrinkage factor $W/(W+V)$ (Module 3) | $0.9412$ |
| $Z$ | "zed" | z-statistic (Module 1) | $4$ |
| $V$ | "vee" | variance of the estimate, $\text{se}^2$ (Module 1); shrinks as $n$ grows | $0.0025$ |
| $W$ | "double-you" | prior variance on the true effect (Module 2) | $0.04$ |
| $\log$ | "log" (natural, base $e$) | natural logarithm (in R: `log`) | $\log(1-r) = -2.833$ |
| $\to$ | "goes to" / "tends to" | approaches, in a limit (e.g. as the sample grows) | $V \to 0$ as $n$ grows |

## The formal version

**From Bayes factor to posterior probability.** Three steps, all from the
[foundation](../../../../foundations/bayes-factors-and-posterior-odds/lesson.qmd):

$$
\text{prior odds} = \frac{\pi}{1 - \pi}, \qquad
\text{posterior odds} = \frac{\pi}{1 - \pi}\times \text{BF}, \qquad
\text{PPA} = \frac{\text{posterior odds}}{1 + \text{posterior odds}}.
$$

Reading it aloud: "prior odds is pi over one-minus-pi; posterior odds is that times the Bayes
factor; the posterior probability of association is the posterior odds over one-plus-posterior
odds." Everyday words: **"turn the rarity of real hits into odds, scale by the evidence, turn
back into a probability."**

Mapping it onto the toy example ($\pi = 10^{-4}$, BF $= 451.6$): prior odds $\approx 10^{-4}$,
posterior odds $\approx 0.0452$, PPA $\approx 0.0432$ тАФ the hand numbers.

**Why the p-value threshold must shrink with $n$.** Recall lABF $= \tfrac12(\log(1 - r) +
rZ^2)$ with $r = W/(W + V)$. Hold the p-value fixed тАФ equivalently, hold $Z$ fixed. As the
sample grows, the standard error shrinks, so $V \to 0$ and therefore

$$
r = \frac{W}{W + V} \longrightarrow 1, \qquad
\log(1 - r) = \log\!\frac{V}{V + W} \longrightarrow -\infty,
$$

while $rZ^2 \to Z^2$ stays finite. So $\text{lABF} \to -\infty$: **a fixed $Z$ (fixed
p-value) delivers ever-weaker тАФ eventually negative тАФ Bayesian evidence as $n$ grows.** To
report the *same* strength of real evidence, you must demand a larger $|Z|$ (a smaller
p-value) in a bigger study. There is no single p-value threshold that means the same thing at
every sample size тАФ the honest threshold shrinks.

Reading it aloud: "as vee goes to zero, r goes to one, log of one-minus-r goes to minus
infinity, so the log ABF goes to minus infinity." Everyday words: **"the same p-value is worth
less and less evidence the more data you have."**

**The prior that makes ABF and p-value rankings identical (Optional).** Order matters in a
GWAS: you rank variants and follow up the top. When does the ABF rank variants the *same* way
the p-value does? A p-value ranks purely by $Z^2$ (bigger $Z^2$ = smaller p). The lABF ranks
by $\tfrac12(\log(1 - r) + rZ^2)$. If you choose the prior so that **$W$ is proportional to
$V$** тАФ that is, $W = cV$ for a constant $c$, which makes $r = W/(W + V) = c/(1 + c)$ the
*same* for every variant тАФ then the lABF becomes $\tfrac12(\log(1 - r) + rZ^2)$ with $r$ fixed,
a strictly increasing function of $Z^2$. Rank by that and you rank by $Z^2$ тАФ **exactly the
p-value's ranking.** So a prior whose variance tracks each variant's own variance links the
two approaches: same order, even though the ABF still gives you a probability the p-value
can't. (This is optional depth тАФ it explains *why* the paper can say the methods sometimes
agree on ranking while disagreeing on interpretation.)

Sanity-check the PPA chain and the threshold claim numerically in the Explore cells below.

## Explore it

```{webr}
#| autorun: true
# explore-1 тАФ LIVE: Bayes factor to posterior probability of association.
# This is the from-scratch wakefield_ppa the course validates (base R).
# TRY IT: change pi_prior. PREDICT: raising pi from 1e-4 to 1e-3 (ten times more likely
# a priori) -- will PPA roughly ten-fold? Commit, then run.
wakefield_ppa <- function(labf, prior = 1e-4) {
  prior_odds <- prior / (1 - prior)
  po <- prior_odds * exp(labf)             # posterior odds = prior odds x BF
  po / (1 + po)                             # PPA
}
lABF <- 6.1128
c(pi_1e4 = wakefield_ppa(lABF, 1e-4),
  pi_1e3 = wakefield_ppa(lABF, 1e-3),
  pi_1e2 = wakefield_ppa(lABF, 1e-2))       # 0.0432, 0.311, 0.820
```

```{webr}
#| autorun: true
# explore-2 тАФ LIVE: the headline. Fix the p-value (fix Z), grow the sample (shrink V),
# and watch the SAME p-value lose all its Bayesian evidence.
# PREDICT: for Z = 2 (p ~ 0.05), as V shrinks toward 1e-6, does lABF stay positive? Guess, run.
labf <- function(z, V, W) { r <- W / (W + V); 0.5 * (log(1 - r) + r * z^2) }
Z <- 2; W <- 0.04                            # fixed p-value ~ 0.0455
V_seq <- c(0.04, 0.01, 0.0025, 1e-4, 1e-6)   # shrinking variance = growing sample

data.frame(V = V_seq,
           lABF = round(sapply(V_seq, function(V) labf(Z, V, W)), 3),
           BF   = round(exp(sapply(V_seq, function(V) labf(Z, V, W))), 3))
# lABF falls from +0.65 to -3.30: p = 0.05 goes from weak-FOR to strong-AGAINST an effect
```

```{webr}
#| autorun: true
#| fig-alt: "Posterior probability of association plotted against the log Bayes factor, one curve per prior probability pi (1e-4, 1e-3, 1e-2). All curves are S-shaped rising from 0 to 1, but the curve for a larger pi is shifted left, reaching high PPA at a smaller lABF. A dashed vertical line marks the running variant's lABF of 6.11."
# explore-3 тАФ LIVE: SEE how the prior pi sets how much evidence you need for a given PPA.
# TRY IT: add pi = 0.5 to the list and watch its curve sit far to the left.
labf_grid <- seq(-2, 14, length.out = 300)
ppa <- function(labf, pi) { po <- pi / (1 - pi) * exp(labf); po / (1 + po) }

plot(labf_grid, ppa(labf_grid, 1e-4), type = "l", lwd = 2, ylim = c(0, 1),
     xlab = "log ABF", ylab = "posterior prob. of association",
     main = "Rarer prior -> more evidence needed for the same PPA")
lines(labf_grid, ppa(labf_grid, 1e-3), lwd = 2, lty = 2)
lines(labf_grid, ppa(labf_grid, 1e-2), lwd = 2, lty = 3)
abline(v = 6.11, col = "grey50")
legend("bottomright", c("pi = 1e-4", "pi = 1e-3", "pi = 1e-2"), lwd = 2, lty = 1:3, bty = "n")
```

## Common traps

**Trap 1 тАФ "a big Bayes factor means the variant is probably associated."** Not on its own.
Our BF of $452$ sounds decisive, but with $\pi = 10^{-4}$ the PPA is only $0.043$ тАФ the
variant is still *probably not* a true hit. Evidence multiplies the prior odds; it doesn't
reset them. This is the same trap as the medical-test foundation, and it's why GWAS demands
enormous per-variant evidence.

**Trap 2 тАФ "a p-value of $0.05$ always means the same strength of evidence."** No тАФ its
evidential meaning depends on sample size. For $Z = 2$ (i.e. $p \approx 0.05$) with $W =
0.04$: at $V = 0.04$ the lABF is $+0.65$ (BF $\approx 1.9$, mild evidence *for*), but at $V =
10^{-6}$ the lABF is $-3.3$ (BF $\approx 0.04$, strong evidence *against*). Same p-value,
opposite conclusions тАФ because a huge study reaching only $p = 0.05$ has found a *tiny* effect,
which the ABF reads as unimpressive. That's precisely why the threshold must shrink with $n$.

## Check yourself

**1.** A variant has lABF $= 4$ (so BF $= e^4 \approx 54.6$). Using $\pi = 10^{-4}$, compute
the prior odds, posterior odds, and PPA.

```{webr}
#| autorun: false
# check1-starter тАФ edit and Run to check your hand answer
lABF <- 4
pi_prior <- 1e-4
```

::: {.callout-note collapse="true" title="Answer"}
BF $= e^4 \approx 54.6$. Prior odds $= 10^{-4}/(1 - 10^{-4}) \approx 10^{-4}$. Posterior odds
$= 10^{-4} \times 54.6 \approx 0.00546$. PPA $= 0.00546/(1.00546) \approx 0.00543$ тАФ about
half a percent. (A BF of $\sim$55 barely dents a $1$-in-$10{,}000$ prior.)
:::

**2.** True or false: increasing the sample size makes a fixed p-value of $0.01$ *stronger*
evidence for an effect. Explain in one line.

::: {.callout-note collapse="true" title="Answer"}
**False** тАФ it makes it *weaker*. A fixed p-value means a fixed $Z$; as $n$ grows, $V \to 0$,
$r \to 1$, and the width penalty $\log(1 - r) \to -\infty$ drags the lABF down, so the same
p-value carries less (eventually negative) Bayesian evidence тАФ the reason the threshold must
shrink with $n$.
:::

**3.** In your own words: explain to a colleague, in two sentences, why a variant with strong
ABF evidence can still be unlikely to be truly associated, and why a p-value threshold should
tighten as studies grow.

::: {.callout-note collapse="true" title="A model answer"}
The posterior probability of association multiplies the prior odds of a hit тАФ which are tiny
in a genome-wide scan тАФ by the Bayes factor, so even a large ABF applied to $1$-in-$10{,}000$
odds can leave the variant probably-not-associated. A p-value threshold should tighten with
sample size because a fixed p-value corresponds to a shrinking true effect as $n$ grows, and
the ABF reads that shrinking effect as ever-weaker evidence тАФ so holding the threshold fixed
would let progressively flimsier findings through.
:::

## What you can now do

- **Convert a Bayes factor into a posterior probability of association** via prior odds
  $\pi/(1-\pi)$, posterior odds, and PPA тАФ and build `wakefield_ppa()` in base R.
- **Explain why a strong ABF can still give a small PPA** when the prior $\pi$ is tiny (the
  rare-hit / rare-disease anchor).
- **Explain why a fixed p-value loses evidence as $n$ grows** (as $V \to 0$, $r \to 1$, the
  width penalty diverges), so the significance threshold must shrink.
- **State the prior that makes ABF and p-value rankings identical** тАФ $W$ proportional to $V$,
  which fixes $r$ and makes the lABF increase with $Z^2$.

If any of these feels shaky, revisit [A tiny example first](#a-tiny-example-first) and the
threshold argument in [The formal version](#the-formal-version) before the capstone тАФ you'll
decode Wakefield's own statement of these results next.

## Where this goes next

You've built the whole pipeline: $(\hat\beta, \text{se}) \to Z \to \text{lABF} \to \text{PPA}$.
The [capstone](../05-capstone-reading-wakefields-abf/lesson.qmd) returns to the source тАФ it
decodes Wakefield's own equations (including his reciprocal sign convention) and the
`coloc::approx.bf.estimates` / `coloc.abf` interface line by line, assembles the full
`wakefield_abf()`, runs it on a simulated multi-SNP GWAS region, and validates it once more
against the frozen `coloc` reference. Ready for more reps? Do the
[practice problems](practice.qmd).
