---
title: "Classification evaluation"
subtitle: "When your model predicts categories, a single error number hides the mistakes that matter"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

## Where we're going

You already know how to check whether a model overfits (train/test split,
cross-validation). But when the model predicts *categories* --- sick or healthy,
spam or not --- a single error number hides critical information. A model that calls
everyone "healthy" is 95% accurate if only 5% are sick, but it misses every sick
person. This module teaches you to read a confusion matrix, understand the trade-offs
between different error types, and use an ROC curve to compare classifiers.

**Builds on:** [probability-and-odds](../probability-and-odds/lesson.qmd)
(probabilities as rates and proportions) and
[model-fit-error-and-validation](../model-fit-error-and-validation/lesson.qmd)
(soft --- the idea of grading predictions against truth).


## Warm-up: do you remember?

Before the new stuff, a few quick reps --- answers from memory first, then check.

**1.** *(probability-and-odds)* If $P(\text{disease}) = 0.05$, what is
$P(\text{no disease})$?

::: {.callout-note collapse="true" title="Answer"}
$P(\text{no disease}) = 1 - 0.05 = 0.95$. The complement rule: if only two outcomes
exist, their probabilities add to 1. Taught in
[probability-and-odds](../probability-and-odds/lesson.qmd); revisit if shaky.
:::

**2.** *(model-fit-error-and-validation)* In one sentence, why do we grade a model on
held-out data, not training data?

::: {.callout-note collapse="true" title="Answer"}
A model can memorize its training data (overfit), scoring perfectly on data it already
saw while performing badly on new data. Held-out data reveals how well the model
*generalizes*. See
[model-fit-error-and-validation](../model-fit-error-and-validation/lesson.qmd);
revisit if shaky.
:::


## A tiny example first

Eight patients visit a clinic. A model predicts whether each one is **sick** or
**healthy**. We also know the truth (say, from a follow-up test). Here is every
patient, side by side:

| Patient | Model says | Truth   |
|---------|-----------|---------|
| 1       | sick      | sick    |
| 2       | sick      | sick    |
| 3       | sick      | healthy |
| 4       | healthy   | sick    |
| 5       | healthy   | healthy |
| 6       | healthy   | healthy |
| 7       | healthy   | healthy |
| 8       | sick      | sick    |

Now count the four kinds of outcome:

- **True Positive (TP):** the model said "sick" AND the patient really is sick.
  Patients 1, 2, 8. That is **3** TPs.
- **False Positive (FP):** the model said "sick" BUT the patient is actually healthy.
  Patient 3. That is **1** FP.
- **False Negative (FN):** the model said "healthy" BUT the patient is actually sick.
  Patient 4. That is **1** FN.
- **True Negative (TN):** the model said "healthy" AND the patient really is healthy.
  Patients 5, 6, 7. That is **3** TNs.

We can arrange these four counts in a 2-by-2 table called the **confusion matrix**
[@james2021introductionstatisticallearning, pp. 148--149]:

|                    | Actually sick | Actually healthy |
|--------------------|:------------:|:----------------:|
| **Predicted sick**     | TP = 3       | FP = 1           |
| **Predicted healthy**  | FN = 1       | TN = 3           |

From these four numbers we can compute three rates, all by hand:

**Accuracy** --- "of all 8 predictions, how many were correct?" Add the correct ones:
3 (TP) + 3 (TN) = 6. Divide by the total number of patients: 6 / 8 = **0.75**. Six
out of eight correct. Sounds decent, but watch what happens when we zoom in on each
class separately.

**Sensitivity** --- "of the patients who are truly sick, how many did the model
catch?" There are 4 truly sick patients (3 TP + 1 FN). The model caught 3 of them:
3 / 4 = **0.75**.

**Specificity** --- "of the patients who are truly healthy, how many did the model
correctly call healthy?" There are 4 truly healthy patients (3 TN + 1 FP). The model
got 3 of those right: 3 / 4 = **0.75**.

The same thing in R, so you can check every step:

```{r toy-example}
# The raw predictions and truth for 8 patients
predicted <- c("sick", "sick", "sick", "healthy",
               "healthy", "healthy", "healthy", "sick")
actual    <- c("sick", "sick", "healthy", "sick",
               "healthy", "healthy", "healthy", "sick")

# Count the four cells of the confusion matrix
TP <- sum(predicted == "sick"    & actual == "sick")      # 3
FP <- sum(predicted == "sick"    & actual == "healthy")   # 1
FN <- sum(predicted == "healthy" & actual == "sick")      # 1
TN <- sum(predicted == "healthy" & actual == "healthy")   # 3

# Accuracy: correct / total
accuracy <- (TP + TN) / (TP + FP + TN + FN)              # 0.75

# Sensitivity: caught sick / truly sick
sensitivity <- TP / (TP + FN)                             # 0.75

# Specificity: correctly cleared / truly healthy
specificity <- TN / (TN + FP)                             # 0.75

c(TP = TP, FP = FP, FN = FN, TN = TN)
c(accuracy = accuracy, sensitivity = sensitivity, specificity = specificity)
```


## Getting a feel for it

**A fire alarm.** Think of a fire alarm: its job is to detect fires. **Sensitivity**
is how often it goes off when there really is a fire --- you want this high, because
missing a fire is dangerous. **Specificity** is how often it stays quiet when there is
no fire --- you want this high too, because false alarms at 3 a.m. are miserable.
The trade-off: a hair-trigger alarm that screams at burnt toast has *perfect*
sensitivity (it never misses a fire) but terrible specificity (it fires on everything
else). A broken alarm that never rings has perfect specificity (no false alarms) but
zero sensitivity (it misses every fire). Every real alarm sits somewhere between those
extremes, and where you set the trigger is a *choice* about which kind of mistake
hurts more.

Where the analogy breaks: a fire alarm has one threshold (how much smoke trips it),
while a classification model usually outputs a probability that you *then* threshold.
But the core trade-off is the same.

**Why accuracy alone misleads.** Accuracy lumps all errors together. In our toy
example accuracy is 0.75, which sounds reasonable. But suppose only 5% of patients are
sick and the model lazily predicts "healthy" for everyone. That model gets 95%
accuracy --- and catches zero sick people. The confusion matrix forces you to see
*which* errors you are making: is the model failing the sick patients (low
sensitivity) or overwhelming healthy people with false alarms (low specificity)?

**What these numbers feel like.** In our toy example, sensitivity and specificity are
both 0.75. That means one in four sick patients slips through (the model misses them),
and one in four healthy patients gets a false alarm. Making the model more aggressive
(calling more people "sick") would catch that missed patient (sensitivity goes up) but
create more false alarms (specificity goes down). That is the *threshold trade-off* in
a nutshell.

**Precision and recall.** There is one more pair worth knowing. **Precision** asks:
"of the ones I *flagged* as sick, how many really were?" In our toy data the model
flagged 4 patients as sick (TP + FP = 3 + 1 = 4), and 3 of them really were, so
precision is 3/4 = 0.75. **Recall** is just another name for sensitivity --- "of the
ones who really were sick, how many did I catch?" When the positive class is rare,
precision drops because even a small FP rate produces many false alarms relative to
the few true positives. Imagine 1,000 patients with only 10 actually sick. Even if the
model catches all 10 (perfect recall/sensitivity), it might also flag 50 healthy
patients, giving precision of just 10/60 $\approx$ 0.17. That is a lot of unnecessary
biopsies for every real catch.


## What the notation is going to say

The formulas will give names to the four counts (TP, FP, TN, FN) and then write each
metric as a fraction of those counts. Accuracy is the total correct divided by the
total count. Sensitivity divides the correctly-caught positives by all actual
positives. Specificity divides the correctly-identified negatives by all actual
negatives. Precision divides the correctly-caught positives by everything the model
*called* positive. The **false positive rate** (FPR) is one minus specificity --- the
fraction of healthy people who got a false alarm. The ROC curve will plot sensitivity
(y-axis) against FPR (x-axis) as we slide the decision threshold from strict to
lenient, and AUC will measure the total area under that curve.


## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $\text{TP}$ | "true positive" | model said positive AND it really is positive | 3 |
| $\text{FP}$ | "false positive" | model said positive BUT it is really negative | 1 |
| $\text{TN}$ | "true negative" | model said negative AND it really is negative | 3 |
| $\text{FN}$ | "false negative" | model said negative BUT it is really positive | 1 |
| $\text{Accuracy}$ | "accuracy" | fraction of all predictions that are correct | 6/8 = 0.75 |
| $\text{Sensitivity}$ | "sensitivity" (also called recall or true positive rate) | fraction of actual positives the model catches (in R: `TP / (TP + FN)`) | 3/4 = 0.75 |
| $\text{TPR}$ | "T-P-R" (true positive rate) | just another name for sensitivity --- same fraction, same formula | 3/4 = 0.75 |
| $\text{Specificity}$ | "specificity" (also called true negative rate) | fraction of actual negatives the model correctly clears (in R: `TN / (TN + FP)`) | 3/4 = 0.75 |
| $\text{TNR}$ | "T-N-R" (true negative rate) | just another name for specificity --- same fraction, same formula | 3/4 = 0.75 |
| $\text{Precision}$ | "precision" (also called positive predictive value, PPV) | fraction of predicted positives that are truly positive (in R: `TP / (TP + FP)`) | 3/4 = 0.75 |
| $\text{PPV}$ | "P-P-V" (positive predictive value) | just another name for precision --- same fraction, same formula | 3/4 = 0.75 |
| $\text{Recall}$ | "recall" | same as sensitivity --- another name for the same thing | 3/4 = 0.75 |
| $\text{FPR}$ | "false positive rate" | fraction of actual negatives that get a false alarm: $1 - \text{Specificity}$ | 1/4 = 0.25 |
| $t$ | "tee" (the threshold) | the cutoff probability above which the model predicts positive | depends on context |
| $\text{ROC}$ | "R-O-C" (receiver operating characteristic) | a curve: TPR (y) vs FPR (x) as $t$ varies | see the Explore section |
| $\text{AUC}$ | "A-U-C" (area under the ROC curve) | one number summarizing the whole ROC curve: 0.5 = random guessing, 1.0 = perfect | computed below |


## The formal version

$$
\text{Accuracy} = \frac{TP + TN}{TP + FP + TN + FN}
$$

Reading it aloud: "accuracy equals TP plus TN, divided by TP plus FP plus TN plus
FN" --- or in everyday words: "the fraction of all predictions that turned out
correct." [@james2021introductionstatisticallearning, pp. 148--149]

$$
\text{Sensitivity} = \text{Recall} = \text{TPR} = \frac{TP}{TP + FN}
$$

Reading it aloud: "sensitivity equals TP divided by TP plus FN" --- or in everyday
words: "of all the people who truly are positive, the fraction the model caught."
[@james2021introductionstatisticallearning, p. 149]

$$
\text{Specificity} = \text{TNR} = \frac{TN}{TN + FP}
$$

Reading it aloud: "specificity equals TN divided by TN plus FP" --- or in everyday
words: "of all the people who truly are negative, the fraction the model correctly
cleared." [@james2021introductionstatisticallearning, pp. 149, 152--153]

$$
\text{Precision} = \text{PPV} = \frac{TP}{TP + FP}
$$

Reading it aloud: "precision equals TP divided by TP plus FP" --- or in everyday
words: "of everyone the model flagged as positive, the fraction that actually were."
[@james2021introductionstatisticallearning, pp. 152--153]

$$
\text{FPR} = 1 - \text{Specificity} = \frac{FP}{FP + TN}
$$

Reading it aloud: "the false positive rate equals one minus specificity, which is FP
divided by FP plus TN" --- or in everyday words: "the fraction of truly negative
people who got a false alarm."
[@james2021introductionstatisticallearning, pp. 149, 152--153]

**The ROC curve:** as you slide the threshold $t$ from very strict (almost nobody
predicted positive) to very lenient (almost everybody predicted positive), you get a
different confusion matrix at each threshold. At each one, compute the TPR and FPR.
Plot TPR on the y-axis and FPR on the x-axis. Connect the dots --- that is the ROC
curve. [@james2021introductionstatisticallearning, pp. 150--151]

**AUC (area under the ROC curve):** the total area under the ROC curve, always between
0 and 1. AUC = 0.5 means the model is no better than flipping a coin. AUC = 1.0
means the model separates the two classes perfectly. AUC gives you a single number
that summarizes performance across *all possible thresholds* rather than at one
particular cutoff. [@james2021introductionstatisticallearning, p. 151]

Mapping it onto the toy example:

- $TP + TN = 3 + 3 = 6$ is the "six out of eight correct" we computed by hand.
- $TP + FN = 3 + 1 = 4$ is the total number of truly sick patients --- the
  denominator for sensitivity.
- $TN + FP = 3 + 1 = 4$ is the total number of truly healthy patients --- the
  denominator for specificity.
- $TP + FP = 3 + 1 = 4$ is everything the model *called* sick --- the denominator
  for precision.


## Explore it

**PREDICT:** this cell rebuilds the confusion matrix with R's `table()` function
instead of counting by hand. Before running it, predict: will sensitivity,
specificity, and precision come out the same as the by-hand values above (0.75,
0.75, 0.75)? Run the cell to check.

```{webr}
#| autorun: true
# The 8-patient data from the toy example
predicted <- c("sick", "sick", "sick", "healthy",
               "healthy", "healthy", "healthy", "sick")
actual    <- c("sick", "sick", "healthy", "sick",
               "healthy", "healthy", "healthy", "sick")

# Build the confusion matrix with table()
# Rows = predicted, columns = actual
conf <- table(Predicted = predicted, Actual = actual)
conf

# Compute the four counts from the table
TP <- conf["sick", "sick"]        # 3
FP <- conf["sick", "healthy"]     # 1
FN <- conf["healthy", "sick"]     # 1
TN <- conf["healthy", "healthy"]  # 3

# Compute the rates
sensitivity <- TP / (TP + FN)     # caught sick / all truly sick
specificity <- TN / (TN + FP)     # cleared healthy / all truly healthy
precision   <- TP / (TP + FP)     # truly sick / everyone called sick

cat("Sensitivity (recall):", sensitivity, "\n")
cat("Specificity:         ", specificity, "\n")
cat("Precision:           ", precision, "\n")
```

The numbers match --- 0.75, 0.75, 0.75 --- because `table()` does exactly the same
counting the manual `sum()` calls did earlier; it is just a faster way to build the
2-by-2 grid. Now the real question: what happens to those numbers if the model gets
MORE aggressive about calling patients "sick"? Catching more of the truly sick people
(sensitivity up) always costs some false alarms (specificity down) --- that is the
threshold trade-off, and the ROC curve below traces it out completely.

Now let's see the ROC curve built from scratch. Ten patients, each with a model-output
probability of being sick. We will sweep the threshold and compute TPR and FPR at each
cutoff.

```{webr}
#| autorun: true
#| fig-alt: "An ROC curve plotting true positive rate versus false positive rate for 10 patients, with a diagonal dashed line showing random chance. The curve bows toward the upper-left corner, indicating the model is better than random."
# 10 patients: model outputs a probability of being sick
prob  <- c(0.95, 0.85, 0.70, 0.65, 0.55, 0.45, 0.35, 0.30, 0.15, 0.05)
truth <- c(   1,    1,    0,    1,    1,    0,    0,    1,    0,    0)
# 1 = truly sick, 0 = truly healthy

# Sweep thresholds from 1.0 down to 0.0
thresholds <- seq(1.0, 0.0, by = -0.05)

# At each threshold, compute TPR and FPR
tpr <- numeric(length(thresholds))
fpr <- numeric(length(thresholds))

for (i in seq_along(thresholds)) {
  t <- thresholds[i]
  pred_pos <- prob >= t            # predict "sick" if prob >= threshold
  TP <- sum(pred_pos & truth == 1)
  FP <- sum(pred_pos & truth == 0)
  FN <- sum(!pred_pos & truth == 1)
  TN <- sum(!pred_pos & truth == 0)
  tpr[i] <- TP / (TP + FN)        # sensitivity at this threshold
  fpr[i] <- FP / (FP + TN)        # false positive rate at this threshold
}

# Plot the ROC curve
plot(fpr, tpr, type = "l", lwd = 2, col = "steelblue",
     xlab = "False Positive Rate (1 - Specificity)",
     ylab = "True Positive Rate (Sensitivity)",
     main = "ROC curve (10-patient example)",
     xlim = c(0, 1), ylim = c(0, 1))
abline(a = 0, b = 1, lty = 2, col = "grey50")  # diagonal = random chance
points(fpr, tpr, pch = 19, cex = 0.6, col = "steelblue")

# Compute AUC by the trapezoidal rule
# Sort by FPR so the integral goes left to right
ord <- order(fpr)
auc <- sum(diff(fpr[ord]) * (tpr[ord][-1] + tpr[ord][-length(tpr[ord])]) / 2)
cat("AUC:", round(auc, 3), "\n")
cat("(0.5 = random, 1.0 = perfect)\n")
```

Now try changing the threshold yourself and watch the confusion matrix change.

```{webr}
#| autorun: false
# TRY IT: change the threshold below and re-run to see how the confusion
# matrix and rates shift.
# PREDICT: what happens to sensitivity and specificity if you set t = 0.30?
prob  <- c(0.95, 0.85, 0.70, 0.65, 0.55, 0.45, 0.35, 0.30, 0.15, 0.05)
truth <- c(   1,    1,    0,    1,    1,    0,    0,    1,    0,    0)

t <- 0.50  # <-- change this threshold

pred_label <- ifelse(prob >= t, "sick", "healthy")
actual_label <- ifelse(truth == 1, "sick", "healthy")

conf <- table(Predicted = pred_label, Actual = actual_label)
conf

TP <- sum(pred_label == "sick"    & actual_label == "sick")
FP <- sum(pred_label == "sick"    & actual_label == "healthy")
FN <- sum(pred_label == "healthy" & actual_label == "sick")
TN <- sum(pred_label == "healthy" & actual_label == "healthy")

cat("Sensitivity:", TP / (TP + FN), "\n")
cat("Specificity:", TN / (TN + FP), "\n")
cat("Precision:  ", TP / (TP + FP), "\n")
cat("Accuracy:   ", (TP + TN) / (TP + FP + TN + FN), "\n")
```


## Common traps

**Trap: "High accuracy = good model."** This is the most dangerous misconception in
classification. Go back to our 8-patient example: 4 patients are healthy, 4 are sick.
A model that always predicts "healthy" --- never examines anything --- gets every
healthy patient right and every sick patient wrong, for accuracy $= 4/8 = 0.5$. Its
sensitivity is $0/4 = 0\%$: it misses every sick person. Now scale it up to an
imbalanced population: if 90 out of 100 patients are healthy and only 10 are sick,
that same lazy model scores accuracy $= 90/100 = 0.9$ --- 90% sounds impressive ---
while its sensitivity is still $0/10 = 0\%$. The lazier the model, the better accuracy
looks as the healthy class comes to dominate the count, even though it still catches
zero sick people. The confusion matrix catches this instantly: TP = 0 means the model
is useless for the job it was built for. Whenever the positive class is rare (called
**class imbalance**), accuracy is misleading.
[@james2021introductionstatisticallearning, pp. 148--149]

**Trap: "Sensitivity and specificity are properties of the disease."** They are
properties of the *test* or *model*, not the condition. The same disease, tested with
a better model, produces different sensitivity and specificity. Saying "this disease
has 90% sensitivity" is like saying "this road has 90% braking power" --- it depends
on the car (the model), not the road (the disease).

**Trap: "The 50% threshold is always the right one."** The 50% cutoff (predict
"sick" whenever the model's probability exceeds 0.5) minimizes the *overall* error
rate when the model outputs well-calibrated posterior probabilities and
misclassification costs are equal, but "overall" quietly averages the two kinds of mistakes together as if they
cost the same --- and they usually don't. Look back at the 10-patient ROC data above:
at $t = 0.50$, sensitivity and specificity are both 0.8. Lower the threshold to
$t = 0.30$ and sensitivity rises to 1.0 (every truly sick patient is now caught) while
specificity falls to 0.4 (more healthy patients get a false alarm). If missing a sick
patient is far more costly than one extra follow-up test --- true for most medical
screening --- the 0.30 threshold is the better choice even though it makes the raw
error rate worse. There is no universally "correct" threshold; the right one depends
on how much each kind of mistake costs in your specific problem.
[@james2021introductionstatisticallearning, pp. 149--150]

::: {.callout-note title="If you've done hypothesis testing, this vocabulary is not new"}
The confusion matrix maps directly onto the Type I/Type II error language from
[hypothesis tests, p-values, and confidence intervals](../hypothesis-tests-p-values-and-confidence-intervals/lesson.qmd):
a **false positive (FP)** is a **Type I error** (a false alarm --- rejecting a true
null when there was nothing there), a **false negative (FN)** is a **Type II error**
(a missed signal --- failing to reject when there really was something there), and
**sensitivity** is the same idea as **power** (the fraction of real positives you
correctly detect). Same fire alarm, same trade-off, different vocabulary.
[@james2021introductionstatisticallearning, p. 152]
:::


## Check yourself

**1.** In our 8-patient confusion matrix (TP = 3, FP = 1, FN = 1, TN = 3), what is
the false positive rate (FPR)?

::: {.callout-note collapse="true" title="Answer"}
$\text{FPR} = \frac{FP}{FP + TN} = \frac{1}{1 + 3} = 0.25$. One in four healthy
patients got a false alarm. Equivalently, $\text{FPR} = 1 - \text{Specificity} =
1 - 0.75 = 0.25$.
:::

**2.** If a classifier has AUC = 0.5, what does that tell you?

::: {.callout-note collapse="true" title="Answer"}
AUC = 0.5 means the model is no better than flipping a coin. Its ROC curve sits right
on the diagonal line from (0, 0) to (1, 1). The model cannot distinguish between the
two classes at any threshold.
:::

**3.** In your own words: explain to a colleague, in two sentences, why accuracy can
be misleading for a classification model and what to look at instead.

::: {.callout-note collapse="true" title="A model answer"}
Accuracy counts correct predictions without caring which class the errors land in, so
a model that guesses the majority class every time can score high while missing
everything you care about. The confusion matrix shows you exactly how many positives
the model caught (sensitivity) versus how many false alarms it generated
(specificity), so you can see whether the errors are acceptable for your problem.
:::


## What you can now do

- Build a confusion matrix from predictions and truth labels, counting TP, FP, TN,
  FN by hand.
- Compute accuracy, sensitivity, specificity, and precision from those four counts.
- Explain the threshold trade-off: making the model more aggressive raises
  sensitivity but lowers specificity.
- Sketch what an ROC curve shows (TPR vs FPR across thresholds) and why the diagonal
  means "no better than random."
- Interpret AUC as a single-number summary of classifier quality (0.5 = random,
  1.0 = perfect).

If any of these feels shaky, revisit
[A tiny example first](#a-tiny-example-first) (confusion matrix, accuracy,
sensitivity, specificity), [Getting a feel for it](#getting-a-feel-for-it) (the
threshold trade-off), or [Explore it](#explore-it) (ROC curve and AUC) before
continuing --- the next module that uses classification evaluation assumes all of
them.


## Where this goes next

These tools apply every time you build or evaluate a classifier. In the
[k-nearest-neighbors](../k-nearest-neighbors/lesson.qmd) module you will train your
first classifier, and in
[LDA, QDA & naive Bayes](../lda-qda-and-naive-bayes/lesson.qmd) you will see
generative classifiers that model each class as a data cloud and use Bayes' theorem
to score. The confusion matrix, sensitivity/specificity, and ROC curve you just
learned are exactly how you would grade any of these classifiers.

Ready for more reps? Do the [practice problems](practice.qmd).
