---
title: "Practice: Classification evaluation"
subtitle: "Answers are hidden --- try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order --- they get harder. Do them on paper or in the live cells; only
then open the answer. Reminders: Sensitivity $= TP/(TP + FN)$, Specificity
$= TN/(TN + FP)$, Precision $= TP/(TP + FP)$, FPR $= 1 -$ Specificity.


## Problem 1 --- the toy example, new numbers

Ten patients. A model predicts "sick" or "healthy" for each. Here are the results:

| Patient | Model says | Truth   |
|---------|-----------|---------|
| 1       | sick      | sick    |
| 2       | sick      | healthy |
| 3       | healthy   | healthy |
| 4       | sick      | sick    |
| 5       | healthy   | sick    |
| 6       | healthy   | healthy |
| 7       | sick      | healthy |
| 8       | sick      | sick    |
| 9       | healthy   | healthy |
| 10      | healthy   | sick    |

Build the confusion matrix by hand. Then compute accuracy, sensitivity, specificity,
and precision.

```{webr}
#| autorun: false
# p1-starter --- count TP/FP/FN/TN, then compute the four rates
predicted <- c("sick", "sick", "healthy", "sick", "healthy",
               "healthy", "sick", "sick", "healthy", "healthy")
actual    <- c("sick", "healthy", "healthy", "sick", "sick",
               "healthy", "healthy", "sick", "healthy", "sick")
```

::: {.callout-note collapse="true" title="Answer"}
Count: TP = 3 (patients 1, 4, 8), FP = 2 (patients 2, 7), FN = 2 (patients 5, 10),
TN = 3 (patients 3, 6, 9).

|                    | Actually sick | Actually healthy |
|--------------------|:------------:|:----------------:|
| **Predicted sick**     | TP = 3       | FP = 2           |
| **Predicted healthy**  | FN = 2       | TN = 3           |

- Accuracy $= (3 + 3) / (3 + 2 + 2 + 3) = 6/10 = 0.60$
- Sensitivity $= 3 / (3 + 2) = 3/5 = 0.60$
- Specificity $= 3 / (3 + 2) = 3/5 = 0.60$
- Precision $= 3 / (3 + 2) = 3/5 = 0.60$

```{r p1-answer}
predicted <- c("sick", "sick", "healthy", "sick", "healthy",
               "healthy", "sick", "sick", "healthy", "healthy")
actual    <- c("sick", "healthy", "healthy", "sick", "sick",
               "healthy", "healthy", "sick", "healthy", "sick")

TP <- sum(predicted == "sick"    & actual == "sick")      # 3
FP <- sum(predicted == "sick"    & actual == "healthy")   # 2
FN <- sum(predicted == "healthy" & actual == "sick")      # 2
TN <- sum(predicted == "healthy" & actual == "healthy")   # 3

c(TP = TP, FP = FP, FN = FN, TN = TN)
c(accuracy    = (TP + TN) / (TP + FP + TN + FN),   # 0.60
  sensitivity = TP / (TP + FN),                     # 0.60
  specificity = TN / (TN + FP),                     # 0.60
  precision   = TP / (TP + FP))                     # 0.60
```
:::


## Problem 2 --- fill in the missing step

We are building an ROC curve. Six patients have model probabilities and true labels:

| Patient | Prob  | Truth |
|---------|-------|-------|
| A       | 0.90  | 1     |
| B       | 0.75  | 1     |
| C       | 0.60  | 0     |
| D       | 0.40  | 1     |
| E       | 0.25  | 0     |
| F       | 0.10  | 0     |

Here is a partial threshold sweep:

> **Threshold = 0.80:** predict sick if prob $\ge$ 0.80.
> Predicted positive: A only. TP = 1, FP = 0, FN = 2, TN = 3.
> TPR = 1/3 $\approx$ 0.333. FPR = 0/3 = 0.

> **Threshold = 0.50:** predict sick if prob $\ge$ 0.50.
> **Your step here** --- list who is predicted positive, count TP/FP/FN/TN, and
> compute TPR and FPR.

> **Threshold = 0.20:** predict sick if prob $\ge$ 0.20.
> Predicted positive: A, B, C, D, E. TP = 3, FP = 2, FN = 0, TN = 1.
> TPR = 3/3 = 1.0. FPR = 2/3 $\approx$ 0.667.

Supply the missing step at threshold = 0.50.

```{webr}
#| autorun: false
# p2-starter --- work out threshold = 0.50
prob  <- c(0.90, 0.75, 0.60, 0.40, 0.25, 0.10)
truth <- c(   1,    1,    0,    1,    0,    0)
t <- 0.50
```

::: {.callout-note collapse="true" title="Answer"}
At threshold = 0.50, predict sick if prob $\ge$ 0.50. That flags patients A (0.90),
B (0.75), and C (0.60).

- TP = 2 (A and B are truly sick)
- FP = 1 (C is truly healthy)
- FN = 1 (D, prob 0.40, is truly sick but falls below the threshold)
- TN = 2 (E and F are truly healthy and predicted healthy)
- TPR = 2/3 $\approx$ 0.667
- FPR = 1/3 $\approx$ 0.333

```{r p2-answer}
prob  <- c(0.90, 0.75, 0.60, 0.40, 0.25, 0.10)
truth <- c(   1,    1,    0,    1,    0,    0)
t <- 0.50
pred_pos <- prob >= t
TP <- sum(pred_pos & truth == 1)   # 2
FP <- sum(pred_pos & truth == 0)   # 1
FN <- sum(!pred_pos & truth == 1)  # 1
TN <- sum(!pred_pos & truth == 0)  # 2
c(TP = TP, FP = FP, FN = FN, TN = TN)
c(TPR = TP / (TP + FN), FPR = FP / (FP + TN))  # 0.667, 0.333
```
:::


## Problem 3 --- full ROC curve in R

Twelve patients with model probabilities and true labels:

| Patient | Prob  | Truth |
|---------|-------|-------|
| 1       | 0.98  | 1     |
| 2       | 0.89  | 1     |
| 3       | 0.82  | 0     |
| 4       | 0.74  | 1     |
| 5       | 0.66  | 1     |
| 6       | 0.58  | 0     |
| 7       | 0.49  | 0     |
| 8       | 0.41  | 1     |
| 9       | 0.33  | 0     |
| 10      | 0.22  | 0     |
| 11      | 0.14  | 1     |
| 12      | 0.07  | 0     |

Write R code that sweeps thresholds from 1.0 to 0.0 (by 0.01), computes TPR and FPR
at each, plots the ROC curve, and computes the AUC. **PREDICT:** will this model's
AUC be closer to 0.5 or to 1.0? (Hint: look at how well the probabilities separate
the 1s from the 0s.)

```{webr}
#| autorun: false
# p3-starter --- build the full ROC curve and compute AUC
prob  <- c(0.98, 0.89, 0.82, 0.74, 0.66, 0.58,
           0.49, 0.41, 0.33, 0.22, 0.14, 0.07)
truth <- c(   1,    1,    0,    1,    1,    0,
              0,    1,    0,    0,    1,    0)
```

::: {.callout-note collapse="true" title="Answer"}
The 1s and 0s are partially interleaved (patient 3 at 0.82 is a 0, patient 11 at 0.14
is a 1), so the model is decent but not perfect. AUC should be noticeably above 0.5
but not near 1.0.

```{r p3-answer}
#| fig-alt: "ROC curve plotting true positive rate against false positive rate, with AUC annotated."
prob  <- c(0.98, 0.89, 0.82, 0.74, 0.66, 0.58,
           0.49, 0.41, 0.33, 0.22, 0.14, 0.07)
truth <- c(   1,    1,    0,    1,    1,    0,
              0,    1,    0,    0,    1,    0)

thresholds <- seq(1.0, 0.0, by = -0.01)
tpr <- numeric(length(thresholds))
fpr <- numeric(length(thresholds))

for (i in seq_along(thresholds)) {
  pred_pos <- prob >= thresholds[i]
  TP <- sum(pred_pos & truth == 1)
  FP <- sum(pred_pos & truth == 0)
  FN <- sum(!pred_pos & truth == 1)
  TN <- sum(!pred_pos & truth == 0)
  tpr[i] <- TP / (TP + FN)       # sensitivity at this threshold
  fpr[i] <- FP / (FP + TN)       # false positive rate
}

# Plot the ROC curve
plot(fpr, tpr, type = "l", lwd = 2, col = "steelblue",
     xlab = "False Positive Rate", ylab = "True Positive Rate",
     main = "ROC curve (12-patient data)")
abline(a = 0, b = 1, lty = 2, col = "grey50")

# AUC via trapezoidal rule
ord <- order(fpr)
auc <- sum(diff(fpr[ord]) * (tpr[ord][-1] + tpr[ord][-length(tpr)]) / 2)
cat("AUC:", round(auc, 3), "\n")
```

AUC $\approx$ 0.722 --- above 0.5 (better than random) but well short of perfect,
which matches the partial overlap between the two groups' probabilities (patient 3 is
a healthy person with a high probability, and patient 11 is a sick person with a low
probability --- those crossings keep the AUC down).
:::


## Problem 4 --- spot the error

A colleague reports the following about their disease-screening model:

> "We tested 200 patients. 190 are healthy, 10 are sick. Our model predicts 'healthy'
> for everyone. Accuracy is 190/200 = 0.95. The model is great!"

The colleague's confusion matrix is:

|                    | Actually sick | Actually healthy |
|--------------------|:------------:|:----------------:|
| **Predicted sick**     | 0            | 0                |
| **Predicted healthy**  | 10           | 190              |

This argument contains **exactly one mistake** in reasoning. Find it and explain why
the model is not great.

::: {.callout-note collapse="true" title="Answer"}
The arithmetic is correct --- accuracy really is 190/200 = 0.95. The mistake is in
the *reasoning*: claiming high accuracy means the model is great. Look at the
confusion matrix: TP = 0. The model catches zero sick patients. Sensitivity
$= 0 / (0 + 10) = 0$. It has *zero* ability to detect the condition it was built to
detect. The 95% accuracy is entirely driven by the overwhelming number of healthy
patients (class imbalance). This is the "high accuracy = good model" trap: accuracy
is misleading when one class dominates.
:::


## Problem 5 --- mixing it with probability-and-odds *(uses probability-and-odds' rates and proportions --- Bayes' formula is given below, just plug in)*

A diagnostic test has sensitivity 0.90 and specificity 0.95. The prevalence of the
disease is $P(\text{disease}) = 0.01$ (1 in 100 people). A person tests positive.
What is $P(\text{disease} \mid \text{positive test})$?

Use Bayes' theorem:

$$
P(\text{disease} \mid +) = \frac{P(+ \mid \text{disease}) \cdot P(\text{disease})}{P(+ \mid \text{disease}) \cdot P(\text{disease}) + P(+ \mid \text{no disease}) \cdot P(\text{no disease})}
$$

where $P(+ \mid \text{disease})$ is the sensitivity and $P(+ \mid \text{no disease})$
is $1 - \text{specificity} = \text{FPR}$.

```{webr}
#| autorun: false
# p5-starter --- plug in the three numbers and compute
sensitivity <- 0.90
specificity <- 0.95
prevalence  <- 0.01
```

::: {.callout-note collapse="true" title="Answer"}
$P(+ \mid \text{disease}) = 0.90$ (sensitivity).
$P(+ \mid \text{no disease}) = 1 - 0.95 = 0.05$ (FPR).
$P(\text{disease}) = 0.01$, $P(\text{no disease}) = 0.99$.

$$
P(\text{disease} \mid +) = \frac{0.90 \times 0.01}{0.90 \times 0.01 + 0.05 \times 0.99}
= \frac{0.009}{0.009 + 0.0495} = \frac{0.009}{0.0585} \approx 0.154
$$

Even with a good test (90% sensitivity, 95% specificity), a positive result gives
only about a 15% chance of actually having the disease when the disease is rare. The
large pool of healthy people (99%) generates enough false positives to swamp the true
positives. This connects back to the rates-and-proportions ideas from
[probability-and-odds](../probability-and-odds/lesson.qmd): the *base rate* matters
enormously.

```{r p5-answer}
sens <- 0.90
spec <- 0.95
prev <- 0.01

# Bayes' theorem
numerator   <- sens * prev                         # P(+ | disease) * P(disease)
denominator <- sens * prev + (1 - spec) * (1 - prev)  # + P(+ | no disease) * P(no disease)
prob_disease_given_pos <- numerator / denominator   # ~ 0.154
prob_disease_given_pos
```
:::


## Problem 6 --- the real-world trade-off

Your company uses a fraud detection model. The CEO wants to minimize false alarms
(maximize precision --- "when we flag a transaction, it had better be real fraud").
The compliance officer wants to catch every fraudulent transaction (maximize
recall/sensitivity --- "we cannot let even one slip through"). These are
contradictory goals.

In two to three sentences, explain why maximizing precision and maximizing recall push
in opposite directions. Then explain how an ROC curve (or a precision-recall curve)
helps the CEO and the compliance officer find a compromise.

::: {.callout-note collapse="true" title="Answer"}
Maximizing recall means lowering the threshold so the model flags more transactions as
fraud --- this catches more true fraud (TP goes up), but it also flags more legitimate
transactions (FP goes up), which drags precision down. Maximizing precision means
raising the threshold so the model only flags transactions it is very confident about
--- FP drops, but some real fraud slips through (FN goes up), dragging recall down.

The ROC curve shows every possible threshold at once. The CEO and compliance officer
can walk along the curve together: at the upper-left corner, recall is high but FPR
(and thus false alarms) is also high. Moving down and left lowers both. They find the
point where the trade-off is acceptable for their business --- maybe catching 95% of
fraud (recall = 0.95) while accepting a certain false alarm rate. The curve makes the
trade-off *visible* instead of hidden behind a single number.
:::


## Problem 7 --- the credit-card default threshold trade-off (real data)

A bank's classifier predicts whether each of 10,000 credit-card customers will
default. Of the 10,000 customers, 333 actually default and 9,667 do not. Here are the
confusion matrices the bank measured at two different classification thresholds:

**Threshold = 50%** (the default cutoff --- predict "default" if predicted
probability $\ge 0.50$):

|                          | Actually default | Actually no default |
|--------------------------|:-----------------:|:---------------------:|
| **Predicted default**    | TP = 81           | FP = 23                |
| **Predicted no default** | FN = 252          | TN = 9644              |

**Threshold = 20%** (predict "default" if predicted probability $\ge 0.20$):

|                          | Actually default | Actually no default |
|--------------------------|:-----------------:|:---------------------:|
| **Predicted default**    | TP = 195          | FP = 235               |
| **Predicted no default** | FN = 138          | TN = 9432              |

Compute sensitivity, specificity, and precision at both thresholds. Then answer: if
you ran this bank's risk team, would you prefer the 50% threshold or the 20%
threshold? Justify your answer in terms of the cost of missing a defaulter versus the
cost of a false alarm (an unnecessary credit hold on a customer who would have paid).

```{webr}
#| autorun: false
# p7-starter --- the two confusion matrices as named vectors
threshold_50 <- c(TP = 81,  FP = 23,  FN = 252, TN = 9644)
threshold_20 <- c(TP = 195, FP = 235, FN = 138, TN = 9432)
```

::: {.callout-note collapse="true" title="Answer"}
At the 50% threshold: sensitivity $= 81/(81+252) = 81/333 \approx 0.243$, specificity
$= 9644/(9644+23) = 9644/9667 \approx 0.998$, precision $= 81/(81+23) = 81/104
\approx 0.779$.

At the 20% threshold: sensitivity $= 195/(195+138) = 195/333 \approx 0.586$,
specificity $= 9432/(9432+235) = 9432/9667 \approx 0.976$, precision $=
195/(195+235) = 195/430 \approx 0.453$.

```{r p7-answer}
threshold_50 <- c(TP = 81,  FP = 23,  FN = 252, TN = 9644)
threshold_20 <- c(TP = 195, FP = 235, FN = 138, TN = 9432)

rates <- function(cm) {
  c(sensitivity = cm["TP"] / (cm["TP"] + cm["FN"]),
    specificity = cm["TN"] / (cm["TN"] + cm["FP"]),
    precision   = cm["TP"] / (cm["TP"] + cm["FP"]))
}
rbind("50% threshold" = rates(threshold_50),
      "20% threshold" = rates(threshold_20))
```

Lowering the threshold from 50% to 20% more than doubles sensitivity (24.3% up to
58.6%) --- the bank now catches over half of the defaulters instead of barely a
quarter. The cost: false positives rise from 23 to 235 (specificity drops from 99.8%
to 97.6%), and precision falls from 77.9% to 45.3% (fewer than half the flagged
customers actually default). For a bank, missing a defaulter usually costs far more
than one unnecessary credit hold, so the 20% threshold --- despite its worse
precision --- is very likely the better business choice. This is the same
asymmetric-cost reasoning as the lesson's "50% threshold is always right" trap: the
"optimal" threshold depends on what a mistake costs, not on minimizing errors in the
abstract. [@james2021introductionstatisticallearning, pp. 148--150]
:::


## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice --- come back to it **at least a day from
now**, once the material is not fresh. A pass then is real evidence you have retained
it. Five questions, graded on the page; 4/5 or better means this module is solid.
Score under 4? Revisit what the missed questions point at, then retake in a **later
session** --- passing after another gap is the evidence that counts. Results are saved
in this browser only.

::: {.module-check module="foundations/classification-evaluation" pass="4"}

::: {.quiz-q qid="ce-sensitivity-calc" qtype="numeric" answer="0.75" tolerance="0.01"}
A confusion matrix has TP = 6, FP = 2, FN = 2, TN = 10. What is the sensitivity?
Round to two decimals.
:::

::: {.quiz-q qid="ce-accuracy-trap" qtype="mc" answer="c"}
A model predicts "negative" for every observation. The dataset has 950 negatives and
50 positives. What is the model's sensitivity?

a) 0.95
b) 0.50
c) 0.00
d) 1.00
:::

::: {.quiz-q qid="ce-fpr-meaning" qtype="mc" answer="b"}
The false positive rate (FPR) measures what?

a) The fraction of predicted positives that are actually wrong
b) The fraction of actual negatives that get a false alarm
c) The fraction of actual positives the model fails to catch
:::

::: {.quiz-q qid="ce-auc-random" qtype="numeric" answer="0.5" tolerance="0"}
A classifier that performs no better than random chance has an AUC of what value?
:::

::: {.quiz-q qid="ce-complement-interleaved" qtype="mc" answer="a"}
*(From probability-and-odds.)* If $P(\text{event}) = 0.15$, what is
$P(\text{not event})$?

a) 0.85
b) 0.15
c) 1.15
d) 0.75
:::

:::

::::
