---
title: "Integration and area under a curve"
subtitle: "What's the total accumulated amount? Measure it with rectangles, then with R's integrate() — and discover why probability IS area"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live cells below re-state their own data. Base R only.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```


## Where we're going

The derivative asked: *how fast is this changing right now?* This module asks the
mirror question: *how much has accumulated over this whole stretch?* That accumulated
total is the **integral** — the area under a curve between two points.

The same "area under a curve" picture appears in two places that matter deeply in
statistics: computing the probability that a quantity falls in some range (probability
IS area under a density), and measuring total exposure in pharmacokinetics, risk, or
signal processing. Once you can measure area numerically and connect it to
antiderivatives, those ideas become genuinely readable.

**Builds on:**
[The derivative as a slope](../derivative-as-slope/lesson.qmd) — the slope idea and
the `slope_at()` measuring tool;
[Differentiation rules](../differentiation-rules/lesson.qmd) — the power rule,
constant-multiple rule, and sum rule whose reverses this lesson teaches.

## Warm-up: do you remember?

Before the new stuff, three quick reps — answers from memory first, then check.

**1.** What is the derivative of $f(x) = x^3$? (State the power rule in words, then
apply it.)

::: {.callout-note collapse="true" title="Answer"}
The power rule: "bring the exponent down as a multiplier, then reduce the exponent by
one." For $x^3$: bring 3 down to get $3x^2$, reduce the exponent to get $f'(x) = 3x^2$.
Shaky? Revisit [Differentiation rules](../differentiation-rules/lesson.qmd).
:::

**2.** For the function $f(x) = x^2$, the `slope_at()` tool from the derivative lesson
gives `slope_at(f, 0)` = 0. What does that zero tell you about the curve at $x = 0$?

::: {.callout-note collapse="true" title="Answer"}
The curve is momentarily flat at $x = 0$ — the slope is zero, meaning the parabola
is neither rising nor falling right there. That's the very bottom of the bowl. See
[The derivative as a slope](../derivative-as-slope/lesson.qmd) §Getting a feel for it.
:::

**3.** If a function has a negative derivative at $x = 5$, what does that tell you
about the curve at that point?

::: {.callout-note collapse="true" title="Answer"}
A negative derivative means the curve is falling (sloping downward) at $x = 5$ — as
you move to the right, the function's value is decreasing. The steeper the negative
slope, the faster it's dropping. See
[The derivative as a slope](../derivative-as-slope/lesson.qmd) §Getting a feel for it.
:::

## Concept 1: Area as accumulated rectangles

### A tiny example first

Imagine you're driving, and your speedometer reads a steady **2 metres per second**
for the first 3 seconds, then **4 metres per second** for the next 3 seconds. How far
have you gone?

That's easy: $2 \times 3 + 4 \times 3 = 6 + 12 = 18$ metres. You just computed an
**area** — the area of two rectangles on a speed-time graph.

Now a harder version: your speed varies continuously. Speed is $f(t) = t$ at time $t$
(so you start slow and accelerate steadily). How far do you travel from $t = 0$ to
$t = 3$?

The speed-time graph is a straight line rising from $0$ to $3$. The area underneath
it is a triangle: base $3$, height $3$, area $= \frac{1}{2} \times 3 \times 3 = 4.5$
metres.

But what if the curve weren't a tidy triangle? The rectangle trick still works — we
just need more, thinner rectangles.

Let's use $f(x) = x^2$ from $x = 0$ to $x = 3$ as our running example. Chop the
interval into 3 equal strips of width 1. For each strip, use the function's value at
its left edge as the rectangle's height:

- Strip 1: $x = 0$, height $= f(0) = 0$, width $= 1$, area $= 0$.
- Strip 2: $x = 1$, height $= f(1) = 1$, width $= 1$, area $= 1$.
- Strip 3: $x = 2$, height $= f(2) = 4$, width $= 1$, area $= 4$.

Total estimate: $0 + 1 + 4 = 5$. The actual area (we'll verify this shortly) is
exactly $9$. Fat rectangles miss a lot — but watch what happens as we use more of them:

```{r toy-rectangles}
# f(x) = x^2, area from 0 to 3
f <- function(x) x^2

# Left-endpoint Riemann sum with n rectangles
riemann_left <- function(f, a, b, n) {
  width  <- (b - a) / n                  # each rectangle's width
  lefts  <- a + (0:(n-1)) * width        # left edges of the n strips
  sum(f(lefts) * width)                  # sum of height * width for each strip
}

# Watch the estimate settle as n grows
sapply(c(3, 10, 100, 1000, 10000), function(n) riemann_left(f, 0, 3, n))
# 5, 7.695, 8.865, ..., marching toward 9
```

### Getting a feel for it

**A bathtub filling up.** The curve $f(x)$ is how fast the water level is rising at
each moment $x$. The area under the curve from $a$ to $b$ is the total water added
between times $a$ and $b$. A big spike in the curve (fast flow) contributes a lot of
area; a flat stretch near zero contributes almost nothing. (Where this picture bends:
real bathtubs can't be drained mid-fill; here the "flow rate" can be negative, which
would mean draining — and the area below the x-axis counts as negative accumulated
total.)

**Why chop into rectangles?** A rectangle's area is "height times width" — the
simplest area formula there is. We can compute it for any function, on any interval,
no matter how complicated the curve. The trick is that as the rectangles get thinner,
each one hugs the curve more tightly, and the accumulated error shrinks. In the
shrinking-step spirit of the derivative lesson: "zoom in until the curved piece looks
flat, then measure."

**What the area's value feels like.** Our area of $9$ for $f(x) = x^2$ from $0$ to
$3$ is "the total amount accumulated under a curve that grows from $0$ to $9$." Change
the upper limit from $3$ to $4$ and the area jumps (you're adding a taller, wider
strip). Change the function to something smaller — say $f(x) = x$ — and the area
shrinks (the curve is lower everywhere). The area grows when the curve is taller or
when the interval is wider.

### What the notation is going to say

The formula coming up says: "add up infinitely many infinitely thin rectangles, each
with height $f(x)$ and width $dx$ (an infinitely thin slice of $x$), running from
$a$ to $b$." That infinite sum is written with a stylized S — the integral sign — and
the $dx$ at the end is the "thin slice" notation. In our toy example the formula
says: "add up $f(x) \cdot dx$ from $0$ to $3$", which is the limit our rectangle sums
were approaching.

### How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $\int_a^b$ | "the integral from a to b" | the total area under the curve between $a$ and $b$ (in R: `integrate(f, a, b)`) | $a = 0$, $b = 3$ |
| $f(x)$ | "eff of ex" | the height of the curve at position $x$ | $f(x) = x^2$ |
| $dx$ | "dee-ex" | an infinitely thin slice of width in the $x$ direction — the limit of a rectangle's width shrinking to zero | the width of one rectangle, shrunk past any finite number |
| $\int_a^b f(x)\,dx$ | "the integral from a to b of f of x dee-ex" | the accumulated area under $f$ between $a$ and $b$ | $\int_0^3 x^2\,dx = 9$ |
| $a$, $b$ | "lower limit, upper limit" | the two endpoints of the interval being measured | $0$ and $3$ |

### The formal version

$$
\int_a^b f(x)\,dx
$$

Reading it aloud: "the integral from $a$ to $b$ of $f$ of $x$ dee-ex" — or in
everyday words: **"the total accumulated amount under the curve $f$ between $x = a$
and $x = b$."**

The key mapping onto our toy example:

- $f(x) = x^2$ is the curve whose area we're measuring.
- $a = 0$, $b = 3$ are the endpoints.
- The integral equals $9$: the total area of infinitely many infinitely thin rectangles
  stacked under the parabola from $x = 0$ to $x = 3$.

R's `integrate()` does this numerically — the same shrinking-rectangles logic,
done very cleverly and very precisely:

```{r formal-integrate}
f <- function(x) x^2

# R measures the area numerically — analogous to slope_at() for derivatives
result <- integrate(f, lower = 0, upper = 3)
result          # 9 with a small error estimate (essentially 0 here)
result$value    # extract the area number: 9
```

### Explore it

```{webr}
#| autorun: true
#| fig-alt: "The parabola f(x) = x squared from x = 0 to x = 3, with filled rectangles showing a left-Riemann sum approximation; the title shows the current rectangle count and area estimate."
# explore-1 — LIVE: watch the rectangle estimate improve as you add more strips.
# PREDICT: with n = 3 rectangles the estimate is 5. What do you think n = 30 gives?
# Write a number, then change n and re-run.
f <- function(x) x^2
n <- 3                                     # <-- increase me (try 10, 30, 100)

a <- 0; b <- 3
width  <- (b - a) / n
lefts  <- a + (0:(n-1)) * width
area   <- sum(f(lefts) * width)

# Draw the rectangles
x_seq <- seq(a, b, length.out = 300)
plot(x_seq, f(x_seq), type = "l", lwd = 2, xlab = "x", ylab = "f(x) = x^2",
     main = sprintf("n = %d rectangles  ->  area estimate = %.4f  (true = 9)", n, area))
rect(lefts, 0, lefts + width, f(lefts),
     col = adjustcolor("steelblue", 0.4), border = "steelblue")
```

```{webr}
#| autorun: true
# explore-2 — LIVE: R's integrate() as the area-measuring tool.
# Compare it with riemann_left() — both compute area, one is exact.
# PREDICT: how close is the rectangle estimate with n = 50?
f_sq <- function(x) x^2

riemann_left <- function(f, a, b, n) {
  width  <- (b - a) / n
  lefts  <- a + (0:(n-1)) * width
  sum(f(lefts) * width)
}

# Area under x^2 from 0 to 3: true answer is 9
c(riemann_n50  = riemann_left(f_sq, 0, 3, 50),
  integrate    = integrate(f_sq, lower = 0, upper = 3)$value)
```

---

## Concept 2: Antiderivatives — undoing the derivative

### A tiny example first

From the [differentiation rules](../differentiation-rules/lesson.qmd) lesson you know
that the derivative of $x^3$ is $3x^2$. Run that connection backwards: if the
derivative of $x^3$ is $3x^2$, then $x^3$ is an **antiderivative** of $3x^2$.

Let's check with numbers. If $F(x) = x^3$, is $F'(x) = 3x^2$?

- At $x = 2$: the slope of $x^3$ should be $3 \times 2^2 = 12$. Measure it:
  $(F(2.01) - F(1.99))/(0.02) = (8.120601 - 7.880599)/0.02 \approx 12$. Check.

So the antiderivative of $3x^2$ is $x^3$ — and the remarkable thing is that the
definite integral of $3x^2$ from $0$ to $3$ equals exactly $F(3) - F(0) = 27 - 0 = 27$.

Meanwhile, the antiderivative of $x^2$ is $x^3/3$ (since $(x^3/3)' = x^2$). So the
area under $x^2$ from $0$ to $3$ is $(3^3/3) - (0^3/3) = 9 - 0 = 9$ — exactly what
our rectangles were marching toward.

```{r antideriv-check}
f <- function(x) x^2
F <- function(x) x^3 / 3   # the antiderivative: (x^3/3)' = x^2

# The area equals F(3) - F(0)
F(3) - F(0)   # 9

# Confirm with integrate()
integrate(f, 0, 3)$value   # also 9 — they match
```

### Getting a feel for it

**The reverse gear.** Differentiation has a reverse gear: just as subtraction undoes
addition, the antiderivative undoes the derivative. If you start with a curve and
differentiate it, you get its slope function. If you start with a slope function and
antidifferentiate, you get back the original curve. The derivative lesson went "from
$F$ to $f$"; this lesson goes "from $f$ back to $F$."

**Why does the antiderivative give the area?** This is the fundamental surprise of
calculus, and here's the intuition: think of $F(x)$ as "area accumulated under $f$
from some starting point up to $x$." As you push the right edge $x$ a tiny bit
further — by $dx$ — you add a thin rectangle of height $f(x)$ and width $dx$. So
$F$ is growing at a rate of $f(x)$. That's exactly what "the derivative of $F$ is
$f$" means. Running the logic backwards: find any $F$ whose derivative is $f$, and
$F(b) - F(a)$ gives you the accumulated area from $a$ to $b$. (Where this picture
bends: it only works when $f$ is continuous — no jumps, no gaps. For the functions in
this repo, that's always true.)

**What an antiderivative feels like.** The antiderivative is always a *family* of
curves — not a single function. $x^3/3$, $x^3/3 + 5$, and $x^3/3 - \pi$ all
differentiate back to $x^2$. They differ only in their starting height — the "$+ C$"
you'll see in the notation. For a *definite* area (between two fixed points), that
constant cancels out in $F(b) - F(a)$, so it doesn't matter which family member you
use.

### What the notation is going to say

Two kinds of integral are coming. The **definite integral** $\int_a^b f(x)\,dx$ has
two specific endpoints and gives a number (the area). The **indefinite integral**
$\int f(x)\,dx$ (no limits) asks for the whole family of antiderivatives — the answer
is a function, not a number, and it always includes a "$+ C$" to acknowledge that
the constant is free.

### How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $F(x)$ | "big eff of ex" | an antiderivative of $f$: the function whose derivative is $f$ | $F(x) = x^3/3$ is an antiderivative of $f(x) = x^2$ |
| $F'(x)$ | "big eff prime of ex" | the derivative of $F$ — should equal $f(x)$ | $F'(x) = x^2 = f(x)$ ✓ |
| $\int f(x)\,dx$ | "the indefinite integral of f of x dee-ex" | the family of all antiderivatives of $f$ (a function + C) | $\int x^2\,dx = x^3/3 + C$ |
| $C$ | "the constant of integration" | an unknown constant added to any antiderivative — represents "where you started counting" | any real number; cancels in $F(b) - F(a)$ |
| $\Big[F(x)\Big]_a^b$ | "big eff of ex, evaluated from a to b" | shorthand for $F(b) - F(a)$ | $\Big[x^3/3\Big]_0^3 = 9 - 0 = 9$ |

### The formal version

The **Fundamental Theorem of Calculus** in one line (the useful half):

$$
\int_a^b f(x)\,dx = F(b) - F(a)
$$

where $F$ is any antiderivative of $f$ (any function whose derivative is $f$).

Reading it aloud: "the integral from $a$ to $b$ of $f$ of $x$ dee-ex equals $F$ of
$b$ minus $F$ of $a$" — or in everyday words: **"the area under $f$ from $a$ to $b$
is found by evaluating the antiderivative at the two endpoints and subtracting."**

Mapping onto the toy:

- $f(x) = x^2$, so we need $F$ with $F'(x) = x^2$.
- The reverse power rule (next concept) gives $F(x) = x^3/3$.
- $F(3) - F(0) = 27/3 - 0 = 9$ — matches our rectangle sums and `integrate()`.

**Verification:** plug the numbers in both ways:

```{r fundamental-theorem}
f <- function(x) x^2
F <- function(x) x^3 / 3      # antiderivative: (x^3/3)' = x^2

# The Fundamental Theorem: definite integral = F(b) - F(a)
F(3) - F(0)                   # 9

# Cross-check with R's numerical integrator
integrate(f, lower = 0, upper = 3)$value   # 9 — both methods agree
```

### Explore it

```{webr}
#| autorun: true
# explore-3 — LIVE: verify that differentiating the antiderivative gives back f.
# The derivative of F(x) = x^3/3 should equal f(x) = x^2 at every point.
# PREDICT: what should slope_at(F, 2) return? (Hint: f(2) = ?)  Write your answer first.
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)

F <- function(x) x^3 / 3    # antiderivative of x^2
f <- function(x) x^2        # the original function

# slope of F at several points -- should equal f(x) at each
xs <- c(0, 1, 2, 3)
rbind(
  slope_of_F = sapply(xs, function(x) slope_at(F, x)),   # F'(x)
  f_value    = f(xs)                                      # f(x)
)
# Both rows should match -- F'(x) = f(x), that's what "antiderivative" means
```

---

## Concept 3: The reverse power rule

### A tiny example first

Every derivative rule runs backwards. The power rule says: $\frac{d}{dx} x^n = n
x^{n-1}$. So if the derivative gives $x^n$, the antiderivative of $x^n$ must be the
function whose derivative *produces* $x^n$.

Working backwards from $\frac{d}{dx} x^{n+1} = (n+1)\,x^n$: the antiderivative of
$x^n$ is $\frac{x^{n+1}}{n+1}$ (we divided both sides by $n+1$ to "undo" the
multiplication). Let's verify three small cases:

- $n = 1$: antiderivative of $x$ is $x^2/2$. Check: $\frac{d}{dx}(x^2/2) = 2x/2 = x$ ✓
- $n = 2$: antiderivative of $x^2$ is $x^3/3$. Check: $\frac{d}{dx}(x^3/3) = 3x^2/3 = x^2$ ✓
- $n = 3$: antiderivative of $x^3$ is $x^4/4$. Check: $\frac{d}{dx}(x^4/4) = 4x^3/4 = x^3$ ✓

```{r power-rule-check}
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)

# Antiderivative of x^2 is x^3/3: verify its slope equals x^2 everywhere
F2 <- function(x) x^3 / 3
c(at_x1 = slope_at(F2, 1),   # should be 1^2 = 1
  at_x2 = slope_at(F2, 2),   # should be 2^2 = 4
  at_x3 = slope_at(F2, 3))   # should be 3^2 = 9
```

### Getting a feel for it

**"Undo the power rule" with one arithmetic step.** The derivative rule says "bring
the exponent down, reduce the exponent by 1." The antiderivative rule says "raise the
exponent by 1, divide by the new exponent." Both rules are mirror operations —
one adds to the exponent while the other subtracts, one multiplies while the other
divides.

**Why divide by $(n+1)$?** Because differentiating $x^{n+1}$ brings $(n+1)$ down as a
multiplier. To end up back at plain $x^n$ — no extra coefficient — we pre-divide by
$(n+1)$ so the multiplication cancels out. It's the same reason you divide by $n$
when undoing a multiplication by $n$.

**What the $+C$ feels like.** The derivative of any constant is zero, so constants
are "invisible" to differentiation. When you antidifferentiate, you can't know what
constant was there originally — it could be $0$, $5$, $-3.7$, anything. The $+C$
says: "I know the shape of the antiderivative, but not where it sits vertically." For
a definite integral the two endpoints pin down the answer and the $C$ cancels, so it
disappears. For the indefinite integral you keep the $C$ as a reminder.

### What the notation is going to say

The reverse power rule says: the indefinite integral of $x^n$ is $x^{n+1}/(n+1)$,
plus a constant $C$. Two helper rules mirror the derivative rules: the constant
multiple rule (a multiplier in front of $f$ stays in front after integration), and the
sum rule (an integral of a sum is the sum of integrals). Together they let you
integrate any polynomial term by term.

### How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $x^n$ | "ex to the n" | the input raised to the power $n$ | $x^2$ (so $n = 2$) |
| $n+1$ | "n plus one" | the new exponent after raising by 1 | $3$ (when $n = 2$) |
| $\frac{x^{n+1}}{n+1}$ | "ex to the n-plus-one, over n-plus-one" | the antiderivative of $x^n$ | $x^3/3$ |
| $+ C$ | "plus C" | the constant of integration — any real number | disappears in $F(b) - F(a)$ |
| $k$ | "kay" | a constant multiplier in front of $f$ | e.g. $3$ in $3x^2$ |
| $\frac{d}{dx}$ | "dee by dee ex" | the derivative with respect to x — gives the slope of what follows | $\frac{d}{dx}(x^3/3) = x^2$ |
| $k\,F(x)$ | "kay times big eff of ex" | antiderivative of $k\,f(x)$ (constant multiple rule) | antiderivative of $3x^2$ is $3 \cdot (x^3/3) = x^3$ |
| $g(x)$ | "gee of ex" | a second function (any function besides $f$) | e.g., $g(x) = 2x$ |

### The formal version

**Reverse power rule:**

$$
\int x^n\,dx = \frac{x^{n+1}}{n+1} + C \qquad (n \neq -1)
$$

Reading it aloud: "the indefinite integral of $x$ to the $n$ dee-ex equals $x$ to
the $(n+1)$, divided by $(n+1)$, plus $C$" — or in everyday words: **"raise the
exponent by one, then divide by that new exponent, and add a free constant."**

**Constant multiple rule:**

$$
\int k\,f(x)\,dx = k \int f(x)\,dx
$$

A constant in front of $f$ passes through the integral unchanged — same as with
derivatives. So $\int 3x^2\,dx = 3 \cdot \int x^2\,dx = 3 \cdot (x^3/3 + C) = x^3 + C$.

**Sum rule:**

$$
\int \bigl(f(x) + g(x)\bigr)\,dx = \int f(x)\,dx + \int g(x)\,dx
$$

An integral of a sum splits into the sum of integrals — again, the same pattern as
derivatives. So you can integrate a polynomial term by term.

**Toy verification.** For $f(x) = x^2$: $n = 2$, so the antiderivative is
$x^{2+1}/(2+1) = x^3/3$. Then $\int_0^3 x^2\,dx = [x^3/3]_0^3 = 27/3 - 0 = 9$.

```{r reverse-power-rule}
# Antiderivative of x^2: apply the reverse power rule by hand -> x^3/3
F <- function(x) x^3 / 3

# Definite integral 0 to 3
F(3) - F(0)   # 9

# Sum-rule example: antiderivative of 2x + 3x^2 is x^2 + x^3
G <- function(x) x^2 + x^3     # antiderivative of 2x + 3x^2
g <- function(x) 2*x + 3*x^2   # the function

G(2) - G(0)                     # 4 + 8 = 12
integrate(g, 0, 2)$value        # 12 -- matches
```

### Explore it

```{webr}
#| autorun: false
# explore-4 — LIVE: apply the reverse power rule and check with integrate().
# For f(x) = x^4, what is the antiderivative? Fill in the blank, then verify.
# PREDICT: what should integrate(f, 0, 2)$value return once you have the right F?
f <- function(x) x^4

# Write the antiderivative using the reverse power rule: x^(n+1)/(n+1)
# n = 4, so n+1 = 5 -> F(x) = ???
F <- function(x) stop("apply the reverse power rule here")   # <-- replace stop() with your answer

# Fundamental theorem check
by_antideriv <- F(2) - F(0)
by_integrate <- integrate(f, 0, 2)$value
c(by_antideriv = by_antideriv, by_integrate = by_integrate)   # should match
```

---

## Concept 4: Definite vs. indefinite integrals, and the density/area connection

### A tiny example first

**Definite integral:** $\int_0^3 x^2\,dx = 9$. A number. No $C$. The two limits pin
down one specific area.

**Indefinite integral:** $\int x^2\,dx = x^3/3 + C$. A family of functions. The
$+ C$ is essential — without it you're claiming there's only one antiderivative, but
there are infinitely many.

Now here's the payoff: `dnorm(x)` in R gives the height of the normal density curve
at $x$ — the bell curve. And the probability that a standard-normal value falls
between $a$ and $b$ is precisely the area under `dnorm` between $a$ and $b$:

$$
P(a \leq X \leq b) = \int_a^b \texttt{dnorm}(x)\,dx
$$

Let's check this for the range $0$ to $1$:

```{r density-area}
# pnorm gives the "area to the left": P(X <= b) for a standard normal
pnorm(1) - pnorm(0)              # about 0.3413

# integrate() with dnorm gives the same area directly
integrate(dnorm, lower = 0, upper = 1)$value   # also about 0.3413
```

They match — probability *is* area under the density.

### Getting a feel for it

**Definite integral = a number.** You're asking "how much area is packed under the
curve between these two fences?" The answer is one real number, pinned down by the
two endpoints. Move a fence and the area changes. This is `pnorm(b) - pnorm(a)` — a
probability.

**Indefinite integral = a family of curves.** You're asking "what family of functions
has $f$ as its derivative?" The answer is a template — "$F(x)$ plus any constant" —
because the constant shifts the antiderivative up or down without changing its slope.
Think of it as a silhouette: many curves have the same silhouette (same shape, different
elevations). The $+C$ acknowledges that we only know the shape.

**Why probability = area under the density.** The density curve $\texttt{dnorm}(x)$
tells you how tightly probability is packed at each value of $x$ — it's a probability
per unit of $x$. Multiply "probability per unit length" by "a thin slice of length
$dx$" and you get a tiny probability. Integrate (add up all the slices between $a$ and
$b$) and you get the total probability in that range. This is exactly what
`pnorm(b) - pnorm(a)` computes — it's the same sum, just done analytically.

**The antiderivative of dnorm is pnorm.** That's what `pnorm` literally is: the
cumulative area under the normal density from $-\infty$ up to $x$. So
`pnorm(b) - pnorm(a)` is exactly $F(b) - F(a)$ from the Fundamental Theorem, with
$F = \texttt{pnorm}$ and $f = \texttt{dnorm}$.

### What the notation is going to say

For a continuous random variable $X$ with density $f_X(x)$, the probability of
landing in the interval $[a, b]$ is $\int_a^b f_X(x)\,dx$ — the integral of the
density over that range. The cumulative distribution function (CDF) $F_X(x)$, which
gives $P(X \leq x)$, is the antiderivative of the density. So `pnorm` and `dnorm` are
just one particular $F$ and $f$ in this notation.

### How to read the symbols

| Symbol | How to say it | What it means | In our density example |
|---|---|---|---|
| $f_X(x)$ | "eff sub ex of ex" | the probability density function (pdf) of random variable $X$ | `dnorm(x)` for a standard normal |
| $F_X(x)$ | "big eff sub ex of ex" | the cumulative distribution function (CDF): $P(X \leq x)$ | `pnorm(x)` |
| $P(a \leq X \leq b)$ | "the probability that X is between a and b" | the probability the random variable falls in that range | `pnorm(1) - pnorm(0)` ≈ 0.341 |
| $X \sim N(0, 1)$ | "X is distributed standard normal" | $X$ follows a standard-normal distribution: mean 0, standard deviation 1 | our test variable |

### The formal version

For a continuous random variable $X$ with density $f_X$:

$$
P(a \leq X \leq b) = \int_a^b f_X(x)\,dx = F_X(b) - F_X(a)
$$

Reading it aloud: "the probability that $X$ falls between $a$ and $b$ equals the
integral from $a$ to $b$ of the density function dee-ex, which equals the CDF at $b$
minus the CDF at $a$" — or in everyday words: **"probability is the area under the
density between the two boundary values."**

For the standard normal, $F_X = \texttt{pnorm}$ and $f_X = \texttt{dnorm}$:

```{r density-formal}
# All three ways to get P(0 <= X <= 1) for a standard normal:

# 1. Using pnorm (the CDF, i.e. the antiderivative)
pnorm(1) - pnorm(0)

# 2. Numerical integration of the density
integrate(dnorm, lower = 0, upper = 1)$value

# 3. The Fundamental Theorem view: F(b) - F(a) with F = pnorm
pnorm(1) - pnorm(0)   # same as #1 -- pnorm IS the antiderivative of dnorm
```

### Explore it

```{webr}
#| autorun: true
#| fig-alt: "The standard normal density curve with the region between x = 0 and x = 1 shaded in blue; the shaded area equals approximately 0.341, the probability that a standard normal falls in that range."
# explore-5 — LIVE: see probability as a shaded area under the bell curve.
# PREDICT: what fraction of the total bell curve area do you think lies between 0 and 1?
# (Hint: the whole bell curve has total area 1, and its mean is at 0.)
# TRY IT: change a and b to shade a different region. What does pnorm give for it?
a <- 0
b <- 1

x_seq <- seq(-3.5, 3.5, length.out = 400)
y_seq <- dnorm(x_seq)

plot(x_seq, y_seq, type = "l", lwd = 2, xlab = "x", ylab = "dnorm(x)",
     main = sprintf("P(%.1f ≤ X ≤ %.1f) = area = %.4f", a, b,
                    pnorm(b) - pnorm(a)))
# Shade the region between a and b
x_shade <- x_seq[x_seq >= a & x_seq <= b]
y_shade <- dnorm(x_shade)
polygon(c(a, x_shade, b), c(0, y_shade, 0),
        col = adjustcolor("steelblue", 0.5), border = NA)
abline(v = c(a, b), lty = 2, col = "steelblue")
```

```{webr}
#| autorun: true
# explore-6 — LIVE: pnorm IS the antiderivative of dnorm — the Fundamental Theorem live.
# The slope of pnorm at any x should equal dnorm(x): F'(x) = f(x).
# PREDICT: what should slope_at(pnorm, 0) return? (Check: dnorm(0) = ?)
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)

xs <- c(-1, 0, 1, 2)
rbind(
  slope_of_pnorm = sapply(xs, function(x) slope_at(pnorm, x)),   # F'(x)
  dnorm_value    = dnorm(xs)                                       # f(x)
)
# Both rows match: pnorm' = dnorm, confirming pnorm is an antiderivative of dnorm
```

---

## Common traps

**Trap 1: "The indefinite integral is the same as the definite integral."** These are
two different things. The definite integral $\int_0^3 x^2\,dx = 9$ is a single number
— area under the curve on a specific interval. The indefinite integral $\int x^2\,dx =
x^3/3 + C$ is an entire family of functions. The confusion is tempting because you
use the antiderivative *to compute* the definite integral, so they look related — and
they are, but the results are different types. Numbers: $\int_0^3 x^2\,dx = 9$ (a
number). Function: $\int x^2\,dx = x^3/3 + C$ (a function family). Omitting the $+C$
from an indefinite integral loses information — the answer is incomplete.

**Trap 2: "Raise the exponent — done!"** The reverse power rule has TWO moves: raise
the exponent by one AND divide by the new exponent. Forgetting the division is the
most common antiderivative mistake. For $x^2$: the antiderivative is $x^3/3$, not
$x^3$. Check: if $F(x) = x^3/3$ then $F'(x) = 3x^2/3 = x^2$ ✓. If you wrote $x^3$
then $F'(x) = 3x^2$ — off by a factor of 3.

```{r trap2-check-power}
# Quick sanity: does the antiderivative's derivative give back the original?
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
F_wrong   <- function(x) x^3       # forgot to divide
F_correct <- function(x) x^3 / 3   # divided by 3
c(wrong_deriv = slope_at(F_wrong, 2), correct_deriv = slope_at(F_correct, 2), target = 2^2)
```

**Trap 3: "The height of the density at $x$ is the probability of $X = x$."** The
density `dnorm(0)` = 0.399 is not a probability — it can be greater than 1 for narrow
distributions, and a single point has probability zero. The probability lives in the
*area*: a thin slice of width $dx$ at $x$ has probability $\approx \texttt{dnorm}(x)
\times dx$, and you add up slices with an integral. Verify with numbers: `dnorm(0)`
$\approx 0.399$, but `pnorm(0.001) - pnorm(0)` $\approx 0.000399$ — the probability
of the tiny range $[0, 0.001]$ is 0.399 times the width $0.001$, not $0.399$ itself.

```{r trap2-check}
# dnorm at 0: this is a density height, NOT a probability
dnorm(0)   # ~0.399 -- larger than many probabilities, but valid for a density

# Probability of a TINY range: density * width
dnorm(0) * 0.001            # ~0.000399 -- a genuine probability (very small range)
pnorm(0.001) - pnorm(0)     # same thing via the CDF -- agrees
```

---

## Check yourself

**1.** Using the reverse power rule, find the antiderivative of $f(x) = x^4$. Then
use it to compute $\int_0^2 x^4\,dx$ by hand (just plug in the endpoints and
subtract), and verify with R.

```{webr}
#| autorun: false
# check1-starter — edit and Run to verify your hand answer
f <- function(x) x^4
# Write the antiderivative F here, then compute F(2) - F(0)
F <- function(x) stop("apply the reverse power rule here")   # <-- replace stop() with your answer
```

::: {.callout-note collapse="true" title="Answer"}
Reverse power rule: $n = 4$, so antiderivative is $x^{4+1}/(4+1) = x^5/5$.

$$
\int_0^2 x^4\,dx = \left[\frac{x^5}{5}\right]_0^2 = \frac{32}{5} - 0 = 6.4
$$

```{r check1-answer}
F <- function(x) x^5 / 5
F(2) - F(0)                              # 6.4

integrate(function(x) x^4, 0, 2)$value  # also 6.4 -- matches
```
:::

**2.** A standard normal variable $X$ has `pnorm(b) - pnorm(a)` = `pnorm(-1) - pnorm(-2)`.
What does that number mean in plain English, and is it bigger or smaller than
`pnorm(0) - pnorm(-1)`?

::: {.callout-note collapse="true" title="Answer"}
`pnorm(-1) - pnorm(-2)` is the probability that a standard normal value falls between
$-2$ and $-1$ — that is, between two and one standard deviation below the mean. The
bell curve is lower there (narrower density) than between $-1$ and $0$, so this area
is smaller. Numerically: `pnorm(-1) - pnorm(-2)` $\approx 0.136$ while
`pnorm(0) - pnorm(-1)` $\approx 0.341$ — yes, smaller.
:::

**3.** In your own words: explain to a colleague, in two sentences, why `pnorm(b) -
pnorm(a)` gives the probability that a standard normal variable falls between $a$ and
$b$, and what the integral has to do with it.

::: {.callout-note collapse="true" title="A model answer"}
`pnorm` is the antiderivative of `dnorm`, so `pnorm(b) - pnorm(a)` is exactly the
Fundamental Theorem's $F(b) - F(a)$, which equals $\int_a^b \texttt{dnorm}(x)\,dx$.
That integral is the area under the density curve between $a$ and $b$, and for a
continuous distribution the probability of landing in a range is defined as that area.
:::

---

## What you can now do

- **Estimate the area under any curve numerically** using left-endpoint rectangles,
  and confirm with R's `integrate()` — the area-measuring tool that mirrors
  `slope_at()` for derivatives.
- **Compute a definite integral by hand** using the Fundamental Theorem: find the
  antiderivative $F$, evaluate $F(b) - F(a)$, and verify with `integrate()`.
- **Apply the reverse power rule**: $\int x^n\,dx = x^{n+1}/(n+1) + C$, and use the
  constant-multiple and sum rules to integrate any polynomial term by term.
- **Distinguish definite from indefinite integrals**: a definite integral is a number
  (area on a specific interval); an indefinite integral is a function family (the
  antiderivative $+ C$).
- **Connect probability to area**: read `pnorm(b) - pnorm(a)` as a definite integral
  of `dnorm`, and explain why density height is not a probability but density times
  width is.

If any of these feels shaky, revisit the corresponding concept section before
continuing — the density/area connection is foundational for anything that computes
probabilities from a distribution.

## Where this goes next

The density/area connection appears everywhere in applied statistics: computing
p-values, credible intervals, and likelihood contributions all reduce to "find this
area under a curve." The next natural step is
[Distributions and densities](../distributions-and-densities/lesson.qmd), which builds
the full toolkit for working with the normal distribution and its family. You'll also
use integration later when building intuition for likelihood functions in
[Likelihood and log-likelihood](../likelihood-and-log-likelihood/lesson.qmd).

Ready for more reps? Do the [practice problems](practice.qmd).
