---
title: "Practice: Integration and area under a curve"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they ramp from imitation to transfer. Do them on paper or in
the live cells; only then open the answer.

## Problem 1 — the toy example, new numbers

In the lesson we computed $\int_0^3 x^2\,dx = 9$ using both rectangle sums and the
antiderivative. Now compute $\int_0^2 x^2\,dx$ by hand using the reverse power rule
(find the antiderivative $F(x) = x^3/3$, then evaluate $F(2) - F(0)$). Then verify
with R's `integrate()`.

```{webr}
#| autorun: false
# p1-starter — edit and Run to check your hand answer
f <- function(x) x^2
```

::: {.callout-note collapse="true" title="Answer"}
Antiderivative: $F(x) = x^3/3$ (reverse power rule: $n=2$, so $n+1=3$, divide by
$3$).

$$
\int_0^2 x^2\,dx = \left[\frac{x^3}{3}\right]_0^2 = \frac{8}{3} - 0 = \frac{8}{3} \approx 2.667
$$

```{r p1-answer}
f <- function(x) x^2
F <- function(x) x^3 / 3

F(2) - F(0)                        # 2.666... = 8/3
integrate(f, lower = 0, upper = 2)$value   # same -- 8/3
```
:::

## Problem 2 — fill in the missing step

The work below computes $\int_0^4 x^3\,dx$ by the Fundamental Theorem with the
**core step blanked**. Supply just that step.

> **Step 1 — apply the reverse power rule:** $f(x) = x^3$, so $n = 3$. The
> antiderivative is $F(x) = \ldots$
>
> **Step 2 — blank:** evaluate $F(4) - F(0)$. (Do this by hand: what is $4^4/4$?)
>
> **Step 3 — verify in R:** `integrate(function(x) x^3, 0, 4)$value` should match.

```{webr}
#| autorun: false
# p2-starter — fill in F below, then compute F(4) - F(0)
f <- function(x) x^3
F <- function(x) stop("apply the reverse power rule here")
```

::: {.callout-note collapse="true" title="Answer"}
The missing step: the antiderivative of $x^3$ is $F(x) = x^4/4$ (reverse power rule:
$n=3$, so $n+1=4$, divide by $4$).

$$
\int_0^4 x^3\,dx = \left[\frac{x^4}{4}\right]_0^4 = \frac{256}{4} - 0 = 64
$$

```{r p2-answer}
F <- function(x) x^4 / 4
F(4) - F(0)                              # 64
integrate(function(x) x^3, 0, 4)$value  # 64 -- matches
```
:::

## Problem 3 — by hand, then in R

Use the reverse power rule and sum rule to evaluate
$\int_0^2 (3x^2 + 2x)\,dx$ by hand, then verify in R.

Work it term by term:

- Antiderivative of $3x^2$: use the constant-multiple rule ($k = 3$) and the reverse
  power rule ($n = 2$).
- Antiderivative of $2x$: same approach with $n = 1$.
- Then evaluate $F(2) - F(0)$.

```{webr}
#| autorun: false
# p3-starter — define the function and antiderivative, then check
g <- function(x) 3*x^2 + 2*x
```

::: {.callout-note collapse="true" title="Answer"}
Term by term:

- $\int 3x^2\,dx = 3 \cdot (x^3/3) = x^3$ (constant-multiple + reverse power rule)
- $\int 2x\,dx = 2 \cdot (x^2/2) = x^2$ (constant-multiple + reverse power rule with $n=1$)

So $F(x) = x^3 + x^2$.

$$
\int_0^2 (3x^2 + 2x)\,dx = F(2) - F(0) = (8 + 4) - 0 = 12
$$

```{r p3-answer}
g <- function(x) 3*x^2 + 2*x
G <- function(x) x^3 + x^2    # antiderivative: sum of the two terms

G(2) - G(0)                             # 12
integrate(g, lower = 0, upper = 2)$value   # 12 -- matches
```
:::

## Problem 4 — spot the error

A colleague wants $\int_1^3 x^2\,dx$. They write:

> "Antiderivative of $x^2$ is $x^3$. So the integral is $3^3 - 1^3 = 27 - 1 = 26$."

The answer contains **exactly one mistake**. Find it and fix it.

```{webr}
#| autorun: false
# p4-starter — run the buggy version to see the wrong answer, then fix it
f <- function(x) x^2
# Buggy antiderivative (their claim): F(x) = x^3
F_wrong <- function(x) x^3
F_wrong(3) - F_wrong(1)   # 26 -- wrong; what should it be?
```

::: {.callout-note collapse="true" title="Answer"}
The mistake: they forgot to divide by $(n+1)$. The reverse power rule for $x^2$ ($n =
2$) gives $F(x) = x^{2+1}/(2+1) = x^3/3$, not $x^3$. Using just $x^3$ is like
"undoing" only half of the power rule — they raised the exponent but didn't divide by
the new exponent.

$$
\int_1^3 x^2\,dx = \left[\frac{x^3}{3}\right]_1^3 = \frac{27}{3} - \frac{1}{3} = 9 - \frac{1}{3} = \frac{26}{3} \approx 8.667
$$

```{r p4-answer}
F_correct <- function(x) x^3 / 3   # divide by 3 -- the step they skipped
F_correct(3) - F_correct(1)        # 26/3 ≈ 8.667

integrate(function(x) x^2, 1, 3)$value   # 8.666... -- matches the corrected version
```
:::

## Problem 5 — interleaved *(uses derivative-as-slope and differentiation-rules)*

*(This problem reaches back to [derivative-as-slope](../derivative-as-slope/lesson.qmd)
and [differentiation-rules](../differentiation-rules/lesson.qmd).)*

You have $F(x) = x^4/4 - x^2$. This is claimed to be the antiderivative of some
function $f(x)$.

**(a)** Differentiate $F$ by hand (using the power rule and sum rule from
differentiation-rules) to find $f(x) = F'(x)$.

**(b)** Verify numerically: use `slope_at()` to measure the slope of $F$ at $x = 2$,
and confirm it matches $f(2)$ computed directly.

**(c)** Compute the definite integral $\int_0^2 f(x)\,dx$ using the Fundamental
Theorem: $F(2) - F(0)$.

```{webr}
#| autorun: false
# p5-starter — define F and slope_at, then work parts (a), (b), (c)
F <- function(x) x^4/4 - x^2
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
```

::: {.callout-note collapse="true" title="Answer"}
**(a)** Differentiating term by term (power rule on each):

- $(x^4/4)' = 4x^3/4 = x^3$
- $(-x^2)' = -2x$

So $f(x) = F'(x) = x^3 - 2x$.

**(b)** Numerical check at $x = 2$: $f(2) = 2^3 - 2 \times 2 = 8 - 4 = 4$.
`slope_at(F, 2)` should return approximately $4$.

**(c)** Fundamental Theorem: $F(2) - F(0) = (16/4 - 4) - (0 - 0) = (4 - 4) - 0 = 0$.

```{r p5-answer}
F <- function(x) x^4/4 - x^2
f <- function(x) x^3 - 2*x          # (a): the derivative of F
slope_at <- function(fn, x, h = 1e-5) (fn(x+h) - fn(x-h)) / (2*h)

# (b): slope of F at x = 2 should equal f(2) = 4
c(slope_of_F = slope_at(F, 2), f_at_2 = f(2))   # both ~4

# (c): definite integral via the Fundamental Theorem
F(2) - F(0)                              # 0
integrate(f, lower = 0, upper = 2)$value # also 0 -- F rises then falls back to 0
```

Note: the integral is $0$ because $f(x) = x^3 - 2x$ is negative on part of
$[0, 2]$ (it dips below the x-axis between $x = 0$ and $x \approx 1.41$) and
positive on the rest — the positive and negative areas cancel exactly. This is a
feature, not a bug: a definite integral can be zero or negative even when the function
isn't. Area "below the x-axis" counts as negative.
:::

## Problem 6 — transfer: probability as area *(cliffhanger)*

*(Connects to [distributions-and-densities](../distributions-and-densities/lesson.qmd).)*

The standard normal density is `dnorm(x)` in R. You know from the lesson that
`pnorm(b) - pnorm(a)` gives the probability $P(a \leq X \leq b)$ because `pnorm` is
the antiderivative of `dnorm`.

**(a)** PREDICT: without running any code, is `integrate(dnorm, -Inf, Inf)$value` equal
to 1, greater than 1, or less than 1? Write your prediction and your reasoning, then
check.

**(b)** The 95% probability region for a standard normal is approximately $[-1.96,
1.96]$. Verify this with `integrate()` (not `pnorm`) — what area does `integrate(dnorm,
-1.96, 1.96)$value` return?

**(c)** A friend says: "The height of the normal density at $x = 0$ is 0.399, so there's
a 39.9% chance of getting exactly 0." What's wrong with their reasoning?

```{webr}
#| autorun: false
# p6-starter — work all three parts
```

::: {.callout-note collapse="true" title="Answer"}
**(a)** The total area under any density must equal $1$ — all probabilities sum to
100%. So `integrate(dnorm, -Inf, Inf)$value` $= 1$ exactly (R returns very close to
$1$).

**(b)** `integrate(dnorm, -1.96, 1.96)$value` $\approx 0.95$ — confirming the
standard 95% interval.

**(c)** The density height is not a probability. `dnorm(0)` $\approx 0.399$ is the
probability *per unit length* — a density. The probability of landing in any single
point is exactly $0$ (a point has zero width, so zero area). The 39.9% comes from
multiplying by a width: $0.399 \times \Delta x$ gives the probability of a *tiny
range* of width $\Delta x$ around $0$, not the probability of exactly $0$.

```{r p6-answer}
# (a) Total area under the density = 1 (all probabilities sum to 1)
integrate(dnorm, lower = -Inf, upper = Inf)$value   # ~1

# (b) 95% interval: the 1.96-sigma rule
integrate(dnorm, lower = -1.96, upper = 1.96)$value   # ~0.9500

# (c) Height vs. probability -- a tiny range width makes this concrete
dnorm(0)              # 0.399 -- density height (NOT a probability)
pnorm(0.001) - pnorm(-0.001)   # ~0.000799 -- probability of a 0.002-wide interval
# 0.399 * 0.002 = 0.000798 -- density * width approximates the probability
```
:::

---

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/integration-and-area" pass="4"}

::: {.quiz-q qid="ia-reverse-power-rule" qtype="numeric" answer="9" tolerance="0"}
Using the reverse power rule, find the antiderivative of $f(x) = x^2$ and evaluate
$\int_0^3 x^2\,dx$. Enter the exact integer value.
:::

::: {.quiz-q qid="ia-definite-vs-indefinite" qtype="mc" answer="c"}
What is the correct form of the indefinite integral $\int x^3\,dx$?

a) $3x^2 + C$
b) $x^4/4$
c) $x^4/4 + C$
d) $4x^4 + C$
:::

::: {.quiz-q qid="ia-density-height-not-prob" qtype="mc" answer="b"}
For a standard normal variable, `dnorm(0)` returns approximately 0.399. What does
this number represent?

a) The probability that the variable equals 0
b) The density curve's height at 0, not a probability
c) The probability that the variable is less than 0
d) The area under the curve from -1 to 1
:::

::: {.quiz-q qid="ia-pnorm-antideriv" qtype="mc" answer="a"}
*(Interleaved from [derivative-as-slope](../derivative-as-slope/lesson.qmd).)*
`pnorm` is the antiderivative of `dnorm`. What does that mean in slope terms?

a) The slope of `pnorm` at any point $x$ equals `dnorm(x)`
b) The slope of `dnorm` at any point $x$ equals `pnorm(x)`
c) `pnorm` and `dnorm` have the same slope everywhere
d) The slope of `pnorm` is always zero
:::

::: {.quiz-q qid="ia-fundamental-theorem" qtype="numeric" answer="6.4" tolerance="0.01"}
Using the Fundamental Theorem of Calculus with antiderivative $F(x) = x^5/5$,
evaluate $\int_0^2 x^4\,dx$. Enter your answer to one decimal place.
:::

:::

::::
