---
title: "Practice: Capstone — reading a multiple regression"
subtitle: "Cumulative. No labels saying which tool — recalling that is part of the test."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
set.seed(42)
```

These draw on the whole course, in mixed order — deciding *which* idea each problem needs
is part of the workout. Try each honestly before opening the answer.

## Problem 1 — warm-up prediction

A fitted model is $\widehat{\text{gdp}} = 0.5 + 0.3\,x_1 - 0.2\,x_2$. Predict `gdp` when
$x_1 = 4$ and $x_2 = 5$.

```{webr}
#| autorun: false
# p1-starter
x1 <- 4; x2 <- 5
```

::: {.callout-note collapse="true" title="Answer"}
$0.5 + 0.3\times4 - 0.2\times5 = 0.5 + 1.2 - 1.0 = 0.7$. A prediction is the intercept
plus each predictor times its coefficient.
:::

## Problem 2 — fit it from scratch

For a centred two-predictor problem you're given
$X^\top X = \left(\begin{smallmatrix}3&1\\1&2\end{smallmatrix}\right)$ and
$X^\top y = (9, 8)$. Solve the normal equations for $(b_1, b_2)$ by hand, then check in R.

```{webr}
#| autorun: false
# p2-starter
XtX <- matrix(c(3, 1, 1, 2), 2); Xty <- c(9, 8)
```

::: {.callout-note collapse="true" title="Answer"}
Determinant $3\cdot2 - 1\cdot1 = 5$; $b_1 = (2\cdot9 - 1\cdot8)/5 = 10/5 = 2$,
$b_2 = (3\cdot8 - 1\cdot9)/5 = 15/5 = 3$. So $(b_1, b_2) = (2, 3)$.

```{r}
solve(matrix(c(3, 1, 1, 2), 2), c(9, 8))    # 2, 3
```
:::

## Problem 3 — read one coefficient row

A regression table row reads:

```
              Estimate Std. Error t value Pr(>|t|)
unemployment   -0.240      0.080   -3.00    0.003 **
```

State (a) the `t value` from the estimate and standard error, (b) what the coefficient
means in words with the necessary qualifier, and (c) what the `**` says.

::: {.callout-note collapse="true" title="Answer"}
(a) $t = -0.240 / 0.080 = -3.00$ — the estimate sits 3 standard errors below 0. (b)
Holding every other predictor fixed, a one-unit rise in unemployment is associated with a
0.24 *decrease* in predicted GDP. (c) `**` marks $p < 0.01$ (here $p = 0.003$): strong
evidence the coefficient isn't zero.
:::

## Problem 4 — spot the error

A colleague reports: *"Our model's $R^2$ is 0.85 and the F-test p-value is 2e-9, so every
coefficient in the table is trustworthy and can be interpreted on its own."* This
contains **exactly one mistake** — find it.

::: {.callout-note collapse="true" title="Answer"}
The mistake: a high $R^2$ and a tiny F-test p-value certify the model **as a whole**
explains real signal — they say nothing about whether *individual* coefficients are
trustworthy. Collinear predictors can have unreliable, even sign-flipped coefficients
while the model predicts well. Trustworthiness of a single coefficient is a separate
question, checked with its standard error and VIF, not with $R^2$ or the overall F-test.
:::

## Problem 5 — interpretation

A model of GDP includes two interest-rate measures correlated 0.97. In the joint fit,
one has a large positive coefficient and the other a moderate negative one, and both have
big standard errors. Your manager asks: "Does this mean one rate raises GDP and the other
lowers it?" Answer in two or three sentences.

::: {.callout-note collapse="true" title="Answer"}
No — with the two rates correlated 0.97 (VIFs will be very high), the joint fit can't
separate their shared effect, so it splits it into a large-positive / moderate-negative
pair with wide uncertainty; the opposite signs are an artifact of collinearity, not a
real opposing mechanism. The honest read is that the *pair* carries interest-rate signal
but the individual coefficients can't be trusted — you'd report the group effect, drop or
combine one measure, or flag the VIFs rather than tell a story about one raising and one
lowering GDP.
:::

## Problem 6 — decode a fresh regression

Here is a `summary(lm())` you haven't seen, predicting `gdp` from a lagged GDP term and
two near-identical interest-rate measures (`rate`, `rate2`), on 80 quarters:

```
Coefficients:
            Estimate Std. Error t value Pr(>|t|)
(Intercept)   0.0049     0.1147   0.043   0.9661
gdp_lag       0.5811     0.1302   4.464   1.4e-05 ***
rate          1.1015     0.4978   2.213   0.0299 *
rate2        -0.5982     0.5215  -1.147   0.2549

Residual standard error: 0.98 on 76 degrees of freedom
Multiple R-squared:  0.3661,  Adjusted R-squared:  0.3411
F-statistic: 14.63 on 3 and 76 DF,  p-value: 1.3e-07
```

You're told `rate` and `rate2` are correlated 0.97 (VIFs ≈ 20 each), and that the data
were generated with `gdp_lag` and `rate` truly mattering while `rate2`'s true effect is
zero. Decode it: (a) Is at least one predictor useful? (b) Which coefficient is most
the most trustworthy? (c) Explain the odd-looking `rate` / `rate2` pair.

::: {.callout-note collapse="true" title="Answer"}
(a) **Yes** — the F-statistic (14.63, p ≈ 1.3e-07) says the model as a whole explains
real signal, well beyond noise; the degrees of freedom are $80 - 4 = 76$. (b) **`gdp_lag`**:
$t = 0.5811/0.1302 = 4.46$, $p \approx 1.4\text{e-}5$ (`***`), and it isn't tangled in the
collinear pair — a trustworthy, well-determined coefficient. (c) `rate` and `rate2` are
0.97-correlated (VIF ≈ 20 ≫ 10), so the fit can't attribute the shared interest-rate
effect: it inflated `rate` to 1.10 (true value 0.4) and handed `rate2` a spurious −0.60
(true 0), both with large standard errors. The pair carries signal together, but neither
individual coefficient should be read as that predictor's real effect — the course's
headline, on a brand-new table.
:::
