---
title: "Capstone: reading a real logistic regression"
subtitle: "Open a full summary(glm(..., family = binomial)) and decode every number"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

::: {.callout-tip title="This lesson runs in your browser" collapse="true"}
The R cell below with a **Run** button is live (WebR) — you rebuild the predicted
probabilities from the fitted coefficients right here. The `glm()` output itself is baked
in at build time (it uses R's `stats` package, which isn't in the browser build), so it's
shown for reading, with a note beside it.
:::

## Where we're going

No new ideas here — this capstone proves you reached the goal. When you started, a
logistic-regression printout was a wall of jargon: *coefficients on the log-odds scale*,
*std. error*, *z value*, *deviance*, *odds ratio*. Now you've **built** every one of those
things. So we'll open a real `summary(glm(..., family = binomial))` — R fitting a genuine
model to the built-in `mtcars` data — and decode every field, tracing each back to the
module where you built it. Then we'll read the same fit the way a paper's methods section
would report it. If you can read this, you can read the logistic regression in almost any
methods section.

**Builds on:** every module — the [sigmoid (1)](../01-from-line-to-probability/lesson.qmd),
[odds, log-odds & the logit (2)](../02-odds-log-odds-and-the-logit/lesson.qmd), and
[likelihood & gradient ascent (3)](../03-likelihood-and-gradient-ascent/lesson.qmd), plus
the foundations they rest on.

## Warm-up: do you remember?

Three recalls spanning the course. Produce each before opening it.

**1.** *(Module 1)* A car has linear score $z = 1.98$ from a logistic model. What is its
predicted probability, and which function turns the score into it?

::: {.callout-note collapse="true" title="Answer"}
The sigmoid: $\sigma(1.98) = 1/(1 + e^{-1.98}) \approx 0.879$. See
[Module 1](../01-from-line-to-probability/lesson.qmd).
:::

**2.** *(Module 2)* A coefficient is $\beta_1 = -4.02$. What is the odds ratio, and does
this predictor raise or lower the odds of the outcome?

::: {.callout-note collapse="true" title="Answer"}
Odds ratio $= e^{-4.02} \approx 0.018$. Since it's **below 1**, each one-unit rise in the
predictor *lowers* the odds (multiplies them by about $0.018$ — a steep drop). See
[Module 2](../02-odds-log-odds-and-the-logit/lesson.qmd).
:::

**3.** *(Module 3)* How are logistic-regression coefficients found, given there's no
closed-form formula like least squares?

::: {.callout-note collapse="true" title="Answer"}
By **maximizing the log-likelihood** — climbing the single-summit surface with gradient
ascent until the gradient is zero. See
[Module 3](../03-likelihood-and-gradient-ascent/lesson.qmd).
:::

## The artifact

Here is a real logistic regression. We predict whether a car has a **manual transmission**
(`am = 1`) from its **weight** in thousands of pounds (`wt`), using R's built-in `mtcars`
data (32 cars). This is exactly the kind of call and output you'll meet in real code and
papers:

::: {.callout-note title="Runs on your machine, not here"}
`glm()` lives in R's `stats` package, which isn't in the browser's WebR build, so this
cell is baked at build time — copy it into your own R session to reproduce it. (In
[Module 3](../03-likelihood-and-gradient-ascent/lesson.qmd) you confirmed a from-scratch
gradient ascent lands on these very same coefficients.)
:::

```{r artifact}
# THE ARTIFACT — a real logistic regression, fit with base R's glm
fit <- glm(am ~ wt, data = mtcars, family = binomial)
summary(fit)
```

## The decode, field by field

Read each piece aloud, then trace it home to the module that built it.

**The model being fit — `am ~ wt`, `family = binomial`.** This says "model the outcome
`am` (manual = 1, automatic = 0) as a logistic regression on `wt`." Under the hood it is
the [logit link from Module 2](../02-odds-log-odds-and-the-logit/lesson.qmd):
$$
\log\!\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 \cdot \texttt{wt},
$$
the log-odds of a manual transmission as a straight line in weight. `family = binomial` is
what tells `glm` to use the sigmoid and the Bernoulli log-likelihood rather than ordinary
least squares.

**`Coefficients — Estimate`: `(Intercept) 12.040`, `wt -4.024`.** These are the fitted
$\hat\beta_0$ and $\hat\beta_1$ — the [Module 3](../03-likelihood-and-gradient-ascent/lesson.qmd)
summit of the log-likelihood, the exact numbers your from-scratch ascent reached. They
live on the **log-odds scale** ([Module 2](../02-odds-log-odds-and-the-logit/lesson.qmd)):

- $\hat\beta_0 = 12.040$ is the log-odds of a manual transmission at $\texttt{wt} = 0$ (an
  extrapolation — no car weighs zero — but that's what the intercept means).
- $\hat\beta_1 = -4.024$ is the change in log-odds per extra 1000 lb. It's **negative**:
  heavier cars are less likely to be manual. As an odds ratio ([Module 2](../02-odds-log-odds-and-the-logit/lesson.qmd)),
  $e^{-4.024} \approx 0.018$ — each extra 1000 lb multiplies the odds of manual by about
  $0.018$, roughly a **98% drop** in the odds.

**`Std. Error`: `4.510`, `1.436`.** The uncertainty in each estimate — how much the
coefficient would wobble if you refit on fresh data from the same process. With only 32
cars these are sizeable. (It plays the same role as the standard errors you met reading
`lm()` output in the
[simple-linear-regression course](../../../simple-linear-regression/syllabus.md); the
machinery differs, the meaning is the same.)

**`z value`: `2.670`, `-2.801`.** Each is `Estimate / Std. Error` — how many
standard errors the coefficient sits from zero. For `wt`: $-4.024 / 1.436 \approx -2.80$.
Big in magnitude → the coefficient is far from zero relative to its noise → strong
evidence of a real effect. (The logistic-regression analogue of the *t value* from
ordinary regression; here the reference curve is the normal, so it's called `z`.)

**`Pr(>|z|)`: `0.00759`, `0.00509`.** The **p-value** for each coefficient: the chance of
seeing a `z` value this far from zero if the coefficient were truly zero (no effect). Both
are well under $0.05$, and the `**` codes flag them as significant at the $0.01$ level.
Weight is a statistically clear predictor of transmission type. (Same reading as the
p-value stars in a
[regression table](../../../simple-linear-regression/syllabus.md).)

**`Null deviance: 43.230 on 31 df` and `Residual deviance: 19.176 on 30 df`.** Deviance is
$-2 \times$ log-likelihood — so a **smaller** deviance means a **larger** log-likelihood,
i.e. a better fit (the sign-flip of "less negative log-likelihood is better" from
[Module 3](../03-likelihood-and-gradient-ascent/lesson.qmd)). *Null* deviance is the fit
using only an intercept (no `wt`); *residual* deviance is the fit with `wt` added. The drop
from $43.230$ to $19.176$ — about $24$ — is how much better the model got by letting weight
matter. The `df` are bookkeeping: 32 cars minus 1 parameter (null) or 2 parameters
(residual).

**`AIC: 23.176`.** A model-comparison score that rewards fit (low deviance) but penalizes
using more parameters — handy for choosing between competing models. *(Optional: you don't
need AIC to read this fit; it matters when comparing several models. Lower is better.)*

**`Number of Fisher Scoring iterations: 6`.** How many steps `glm`'s optimizer took to
reach the summit — the same *idea* as the iteration count you watched in gradient ascent
([Module 3](../03-likelihood-and-gradient-ascent/lesson.qmd)), though `glm` uses a smarter,
curvature-aware climber (Fisher scoring) that needs far fewer steps than plain ascent, just
as `optim`'s L-BFGS-B did in the
[gradient-descent capstone](../../../gradient-descent/modules/04-capstone-decode-the-optimizer/lesson.qmd).

### Rebuild the predictions from the coefficients

Everything the model predicts flows from the two coefficients through the machinery you
built. Take the coefficients, form the score, squash with the sigmoid — and you get the
same probabilities `glm`'s own `predict()` returns. Run it:

```{webr}
#| autorun: true
# LIVE: rebuild glm's predicted probabilities by hand, from the fitted coefficients.
# PREDICT: a light 2.5k-lb car vs a heavy 4k-lb car -- which has the higher probability
# of being manual? Guess, then run.
sigmoid <- function(z) 1 / (1 + exp(-z))
b0 <- 12.040; b1 <- -4.024               # the fitted coefficients from the summary

wt    <- c(2.5, 4.0)
score <- b0 + b1 * wt                     # the log-odds (Module 2): 1.98 and -4.06
prob  <- sigmoid(score)                   # squash to probability (Module 1)
rbind(wt = wt, log_odds = round(score, 3), prob = round(prob, 3))
# 2.5k-lb car: ~0.879 chance manual ; 4.0k-lb car: ~0.017 -- matches glm's predict()
```

A light car ($2.5$k lb) is predicted about $88\%$ likely to be manual; a heavy one ($4$k
lb) only about $2\%$ — the negative weight coefficient at work, run all the way through the
sigmoid to a probability.

## How a paper would report the same fit

A methods section rarely prints the raw `glm` block. It writes something like:

> We modelled transmission type (manual vs. automatic) with logistic regression on vehicle
> weight. Heavier vehicles were significantly less likely to be manual (log-odds
> $\hat\beta = -4.02$ per 1000 lb; odds ratio $0.018$, 95% CI $[0.001, 0.30]$; $z =
> -2.80$, $p = 0.005$).

Decode that sentence, all of it now familiar:

- "**logistic regression on vehicle weight**" → the [logit link](../02-odds-log-odds-and-the-logit/lesson.qmd),
  log-odds modelled as a line in weight.
- "**log-odds $\hat\beta = -4.02$ per 1000 lb**" → the `wt` coefficient, on the log-odds
  scale ([Module 2](../02-odds-log-odds-and-the-logit/lesson.qmd)).
- "**odds ratio $0.018$**" → $e^{\hat\beta} = e^{-4.02}$ ([Module 2](../02-odds-log-odds-and-the-logit/lesson.qmd));
  each extra 1000 lb multiplies the odds of manual by $0.018$.
- "**95% CI $[0.001, 0.30]$**" → a plausible range for that odds ratio, built from the
  standard error as $e^{\hat\beta \pm 1.96\,\text{SE}}$. The whole interval sits below $1$
  — "no effect" — which is the CI-form of "the p-value is below $0.05$".
- "**$z = -2.80$, $p = 0.005$**" → the coefficient's `z value` and p-value straight from
  the table.

Same fit, same numbers — just prose instead of a printout.

## Common traps

**Trap: "the coefficient $-4.02$ means weight lowers the *probability* of manual by
$4.02$ (or by $402\%$)."** No — a coefficient is a change in **log-odds** per unit, never a
change in probability ([Module 2's trap](../02-odds-log-odds-and-the-logit/lesson.qmd#common-traps)).
Its clean multiplicative reading is the odds ratio $e^{-4.02} \approx 0.018$ (odds
multiplied by $0.018$ per 1000 lb). The *probability* change depends on the car's weight —
big near the middle of the S-curve, tiny out at the ends.

**Trap: "the null deviance being bigger than the residual deviance means the model got
worse."** Backwards. Deviance is $-2 \times$ log-likelihood, so **smaller is better**. The
*null* model (intercept only) has the bigger deviance ($43.230$); adding `wt` *reduced* it
to $19.176$. A drop in deviance is an *improvement* in fit — the mirror of "less negative
log-likelihood is better" from [Module 3](../03-likelihood-and-gradient-ascent/lesson.qmd).

## What you can now do

- **Read a `summary(glm(..., family = binomial))` field by field** — the model formula and
  family, the coefficients (on the log-odds scale), their standard errors, z values, and
  p-values, the null and residual deviance, the iteration count.
- **Interpret each coefficient two ways** — as a change in log-odds per unit, and as an
  odds ratio $e^{\hat\beta}$ (above 1 raises the odds, below 1 lowers them).
- **Rebuild any predicted probability** from the coefficients: form the score
  $\beta_0 + \beta_1 x$, squash with the sigmoid.
- **Decode the same model as a paper reports it** — coefficient, odds ratio with a
  confidence interval, z, and p — and know which module every piece came from.

If any field still reads as jargon, revisit the module it points to above — reading that
output cold is the entire goal of this course, and you just did it on real data.

## What you still can't read (yet)

This course took you through logistic regression with a **single** predictor. A few honest
edges to know about, each a natural next step:

- **Several predictors at once** — `glm(y ~ x1 + x2 + x3, family = binomial)`. Nothing new
  in spirit: the score becomes $\beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots$ (more knobs on
  the same gradient-ascent climb), and each coefficient's odds ratio is read "holding the
  others fixed." The linear-algebra bookkeeping for many predictors at once is its own
  topic.
- **Other outcomes** — counts (Poisson regression), 3+ categories (multinomial), ordered
  categories (ordinal). These swap the sigmoid and the likelihood for relatives of them;
  the "line on a transformed scale, fit by maximum likelihood" skeleton is identical.
- **How the standard errors and that 95% CI are actually computed** — this comes from the
  *curvature* of the log-likelihood at its summit (how sharply the hill falls away from the
  top). It leans on more calculus and a little linear algebra than this course assumed.

None of these change what you built; they extend it. You can read the core of a logistic
regression now — the rest is more of the same shape.

## Where this goes next

You've built and decoded logistic regression end to end. The natural continuations are
above: multiple logistic regression (many predictors), the wider family of *generalized
linear models* (Poisson, multinomial — same skeleton, different link and likelihood), and
the curvature-based standard-error machinery. Each reuses exactly what you now own — a line
on a transformed scale, squashed to a valid range, fit by maximizing a likelihood. Ready
for the final cumulative set? Do the [practice problems](practice.qmd).
