---
title: "Practice: Capstone — the whole model, mixed"
subtitle: "No hints about which idea each problem needs — recalling that is the test."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

These draw on every module, in mixed order — figuring out *which* idea each one needs is
part of the test, so there are no "uses Module N" labels. A helper to paste as needed:
`sigmoid <- function(z) 1 / (1 + exp(-z))`.

## Problem 1

A logistic model gives a subject the linear score $z = -0.5$. What is the predicted
probability, and is the subject more or less likely than not to have the outcome?

```{webr}
#| autorun: false
# p1-starter
sigmoid <- function(z) 1 / (1 + exp(-z))
z <- -0.5
```

::: {.callout-note collapse="true" title="Answer"}
```{r p1-answer}
sigmoid <- function(z) 1 / (1 + exp(-z))
sigmoid(-0.5)   # ~0.378
```
$\sigma(-0.5) \approx 0.378$ — **less** likely than not, because the score is negative
(any $z < 0$ gives a probability below $0.5$).
:::

## Problem 2

A published logistic model reports two coefficients: $\beta_{\text{age}} = 1.10$ and
$\beta_{\text{dose}} = -0.30$. Convert each to an odds ratio, and say in one sentence what
each tells you about the effect of that predictor on the odds of the outcome.

```{webr}
#| autorun: false
# p2-starter
b_age  <- 1.10
b_dose <- -0.30
```

::: {.callout-note collapse="true" title="Answer"}
```{r p2-answer}
c(OR_age = exp(1.10), OR_dose = exp(-0.30))   # 3.004, 0.741
```
Age: odds ratio $e^{1.10} \approx 3.0$ — each extra unit of age **triples** the odds of the
outcome (above 1 → raises the odds). Dose: odds ratio $e^{-0.30} \approx 0.74$ — each extra
unit of dose multiplies the odds by about $0.74$, a roughly **26% drop** (below 1 → lowers
the odds).
:::

## Problem 3

For a tiny dataset $x = (0, 1, 2)$, $y = (0, 1, 1)$, compute the log-likelihood at the
coefficients $\beta = (-0.5, 1)$. Show the three predicted probabilities as part of your
answer.

```{webr}
#| autorun: false
# p3-starter
sigmoid <- function(z) 1 / (1 + exp(-z))
x <- c(0, 1, 2); y <- c(0, 1, 1)
b <- c(-0.5, 1)
```

::: {.callout-note collapse="true" title="Answer"}
```{r p3-answer}
sigmoid <- function(z) 1 / (1 + exp(-z))
x <- c(0, 1, 2); y <- c(0, 1, 1); b <- c(-0.5, 1)
p <- sigmoid(b[1] + b[2] * x)
round(p, 4)                                       # 0.3775, 0.6225, 0.8176
sum(y * log(p) + (1 - y) * log(1 - p))            # log-likelihood ~ -1.150
```
Predicted probabilities $0.378, 0.622, 0.818$. Each point contributes the log of the
probability it assigned to the outcome that happened: the failure ($y=0$, $x=0$) contributes
$\log(1 - 0.378)$, the two successes contribute $\log(0.622)$ and $\log(0.818)$. The sum is
about $-1.150$.
:::

## Problem 4 — spot the error

A colleague summarizes a logistic fit: "Adding the predictor `dose` **raised** the residual
deviance from a null deviance of $30$ to a residual deviance of $42$, so `dose` clearly <!-- lint-ok -->
**improves** the model." The statement contains **exactly one conceptual mistake** — find
it and explain the correct reading.

::: {.callout-note collapse="true" title="Answer"}
Two things are tangled, but the single mistake is the claim that a **higher** residual
deviance means a **better** model. Deviance is $-2 \times$ log-likelihood, so **lower**
deviance is better (higher log-likelihood, more probable data). Adding a useful predictor
should *reduce* the residual deviance below the null deviance, not raise it. If the residual
deviance ($42$) were genuinely *above* the null deviance ($30$), that would signal the model
got *worse* — the opposite of what the colleague concluded. (In a correct fit like the
lesson's, the null deviance $43.2$ dropped to a residual deviance $19.2$ — a *drop* is the
improvement.)
:::

## Problem 5

A logistic model of germination has fitted log-odds $= -2 + 0.8x$, where $x$ is temperature
in some units. Predict the germination probability at $x = 3$, and show the two-step route
(score, then probability) you used.

```{webr}
#| autorun: false
# p5-starter
sigmoid <- function(z) 1 / (1 + exp(-z))
b0 <- -2; b1 <- 0.8
x <- 3
```

::: {.callout-note collapse="true" title="Answer"}
```{r p5-answer}
sigmoid <- function(z) 1 / (1 + exp(-z))
b0 <- -2; b1 <- 0.8; x <- 3
score <- b0 + b1 * x        # log-odds: -2 + 0.8(3) = 0.4
sigmoid(score)              # probability ~ 0.599
```
Step 1 — the score (log-odds): $-2 + 0.8(3) = 0.4$. Step 2 — squash with the sigmoid:
$\sigma(0.4) \approx 0.599$. So about a **60%** germination probability at $x = 3$. (A
positive score, so above $0.5$, as expected.)
:::

## Problem 6 — interpretation

Two logistic models are fit to the same 0/1 outcome on the same 32 subjects. Model A uses
predictor $x_1$ and has residual deviance $27$; Model B uses predictor $x_2$ and has
residual deviance $21$. Both start from the same null deviance of $43$. Which predictor
explains the outcome better, and roughly how confident should you be that either helps at
all? Answer in two or three sentences.

::: {.callout-note collapse="true" title="Answer"}
Model B fits better: its residual deviance ($21$) is lower than Model A's ($27$), and lower
deviance means higher log-likelihood — the observed outcomes are more probable under Model
B. Both predictors help *something*, since each cut the deviance below the null value of
$43$ (Model A by $16$, Model B by $22$) — a drop in deviance is the improvement from adding
the predictor. Model B's larger drop makes it the stronger single predictor of the two.
:::

## Problem 7 — decode a fresh artifact

Here is a real logistic regression you haven't seen — predicting whether a car has a
**straight engine** (`vs = 1`) from its fuel economy (`mpg`), on the built-in `mtcars`
data. Read the output and answer: (a) the fitted coefficient for `mpg` and its odds ratio;
(b) is higher `mpg` associated with higher or lower odds of a straight engine, and is it
statistically significant; (c) did adding `mpg` improve the fit over the null model?

```{r p7-artifact}
# A fresh glm fit to decode (you have every tool for this now)
summary(glm(vs ~ mpg, data = mtcars, family = binomial))
```

::: {.callout-note collapse="true" title="Answer"}
```{r p7-answer}
fit <- glm(vs ~ mpg, data = mtcars, family = binomial)
c(coef_mpg = coef(fit)["mpg"], odds_ratio_mpg = exp(coef(fit)["mpg"]))
c(null_deviance = fit$null.deviance, residual_deviance = fit$deviance)
```
(a) The `mpg` coefficient is $\hat\beta \approx 0.430$ (on the log-odds scale); its odds
ratio is $e^{0.430} \approx 1.54$.

(b) **Higher** odds — the coefficient is positive, so each extra mpg multiplies the odds of
a straight engine by about $1.54$ (a ~54% rise in the odds). It's statistically significant:
$z \approx 2.72$, $p \approx 0.007$ (well under $0.05$, flagged `**`).

(c) **Yes** — adding `mpg` dropped the deviance from a null $43.86$ to a residual $25.53$
(a fall of about $18$). Lower deviance means a better fit, so fuel economy is a useful
predictor of engine type.

Reading a fresh `glm` summary and pulling out the coefficient, the odds ratio, the
significance, and the deviance improvement — that's the whole point of the course, and you
just did it on data you'd never seen.
:::
