---
title: "Linear combinations & the data matrix"
subtitle: "How a table of data plus a list of weights becomes one number per row — the columns papers pack into X"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live cells below re-state their own data. Base R is all
# we need, so nothing to install.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

::: {.callout-tip title="This lesson runs in your browser" collapse="true"}
The code cells below with a **Run** button are live — powered by WebR (R compiled to
WebAssembly), they run right here on your phone or laptop, no install needed. Edit any
of them and re-run. The first run downloads a small R engine, so give it a few seconds.
Cells without a Run button are ordinary listings whose output was computed when the
page was built.
:::

## Where we're going

Almost every quantitative method eventually does the same humble thing: it takes a
**data table** — one row per thing you measured, one column per variable — and
combines each row's numbers into a single number, using a list of **weights**. That
"multiply each column by its weight and add" move is called a **linear combination**,
and it's just a dot product wearing work clothes. This lesson shows how the raw data
table becomes the matrix that combining acts on, and how the tricky-looking columns —
an all-1s column, 0/1 category columns, a product-of-two-columns — get built. After
this, when you see a paper stack its variables into a matrix and hit it with a weight
vector, you'll read it as ordinary R.

We stay with tiny, hand-checkable tables and base R. We will *not* teach the matrix
*operations* themselves — multiplying, transposing, inverting a matrix — those live in
[matrices & linear transforms](../matrices-and-linear-transforms/lesson.qmd); we only
*use* `%*%` here and lean on that lesson for what it does. And we make **no** claim
about how the weights get *chosen* — that's a job for specific methods, not this
building block.

**Builds on:** [matrices & linear transforms](../matrices-and-linear-transforms/lesson.qmd)
— we reuse `%*%` (matrix × vector = one dot product per row) without re-teaching it;
[vectors & summation](../vectors-and-summation/lesson.qmd) — a dot product is a sum of
products, $\sum_i a_i b_i$; and
[mean, variance & covariance](../mean-variance-covariance/lesson.qmd) — for the idea of
a column of data as a variable measured across rows.

## Warm-up: do you remember?

Three quick reps first — produce each answer from memory, then check.

**1.** For `a <- c(2, 5, 1)` and `b <- c(3, 0, 4)`, what single number is the **dot
product** of `a` and `b`? Say the formula in words, then give the value.

```{webr}
#| autorun: false
# warmup1-starter — edit and Run to check
a <- c(2, 5, 1)
b <- c(3, 0, 4)
```

::: {.callout-note collapse="true" title="Answer"}
"Multiply the paired entries and add them up": $2\times3 + 5\times0 + 1\times4 =
6 + 0 + 4 = 10$. In R that's `sum(a * b)`. This one operation — a **sum of products**
— is the whole engine of the linear combinations below. Shaky? Revisit
[vectors & summation](../vectors-and-summation/lesson.qmd).
:::

**2.** If `A` is a matrix and `v` is a vector, what does `A %*% v` compute, in one
sentence? What has to be true about the length of `v`?

::: {.callout-note collapse="true" title="Answer"}
`A %*% v` runs **one dot product per row** of `A`: each row is paired entry-by-entry
with `v`, multiplied, and summed, giving one output number per row. So `v` must have
one entry for every **column** of `A`, or the pairing wouldn't line up. That's the
whole content of [matrices & linear transforms](../matrices-and-linear-transforms/lesson.qmd).
:::

**3.** In R, a column of a data frame — like `df$height` — is just a vector of one
variable's values across all the rows. For `height <- c(2, 4, 6)`, what is
`mean(height)`, and what does that number describe?

```{webr}
#| autorun: false
# warmup3-starter — edit and Run to check
height <- c(2, 4, 6)
```

::: {.callout-note collapse="true" title="Answer"}
`mean(height)` is $(2 + 4 + 6)/3 = 4$ — the balance point of that one column's values.
Thinking of each column as a variable measured across rows is the mental model from
[mean, variance & covariance](../mean-variance-covariance/lesson.qmd); a data matrix is
just several such columns side by side.
:::

## A tiny example first

You track **4 short tasks**. For each you recorded how many **hours** it took and which
**tier** of difficulty it was (A, B, or C):

| task | hours | tier |
|---|---|---|
| 1 | 4 | A |
| 2 | 2 | B |
| 3 | 6 | C |
| 4 | 3 | B |

Now suppose someone hands you a **scoring rule** — a list of weights — and asks for one
**score** per task. The rule: start at a base of 10, add 2 per hour, add 5 if it's
tier B, add −3 if it's tier C (tier A adds nothing — it's the baseline).

**Score task 1** (4 hours, tier A): $10 + 2\times4 = 10 + 8 = \mathbf{18}$. Tier A adds
nothing, so we stop there.

**Score task 2** (2 hours, tier B): $10 + 2\times2 + 5 = 10 + 4 + 5 = \mathbf{19}$.

**Score task 3** (6 hours, tier C): $10 + 2\times6 - 3 = 10 + 12 - 3 = \mathbf{19}$.

**Score task 4** (3 hours, tier B): $10 + 2\times3 + 5 = 10 + 6 + 5 = \mathbf{21}$.

Every score is "multiply each thing by its weight, then add" — a **linear combination**
of the task's numbers. The catch: "base of 10," "tier B," and "tier C" aren't numbers
you measured; they're *columns you have to build* so the same multiply-and-add works on
every row. That's the whole job of this lesson: turn the table into a grid of numbers —
a **data matrix** — where combining is one dot product per row.

Here's the grid, built column by column, and the scores in R so you can check every
step:

```{r toy-example}
# What we measured
hours <- c(4, 2, 6, 3)
tier  <- c("A", "B", "C", "B")

# Build the columns the scoring rule needs, one at a time:
one   <- rep(1, 4)                 # an all-1s column, so the base of 10 rides along
tierB <- as.integer(tier == "B")  # 1 where tier is B, else 0
tierC <- as.integer(tier == "C")  # 1 where tier is C, else 0 (tier A is the baseline)

# Stack them into a data matrix: one row per task, one column per variable
X <- cbind(one = one, hours = hours, tierB = tierB, tierC = tierC)
X

# The weights, in the SAME order as the columns
beta <- c(10, 2, 5, -3)   # base, per-hour, tier-B bump, tier-C bump

# One score per row = one dot product per row (matrices & linear transforms)
X %*% beta                # 18, 19, 19, 21

# Prove it's just multiply-and-add for row 1:
sum(X[1, ] * beta)        # 18 -> 10*1 + 2*4 + 5*0 + (-3)*0
```

## Getting a feel for it

**A linear combination is a shopping cart total.** Picture the columns as products on a
shelf — hours, "is it tier B", "is it tier C" — and the weights as their prices. One
row is one shopping trip: you take some amount of each product (4 hours, 0 of tier-B,
0 of tier-C), multiply each amount by its price, and the register adds them into a
single total. The **all-1s column is the "cover charge"**: a fixed item every trip buys
exactly one of, so its weight (the base of 10) always lands in the total. Change the
prices (the weights) and every total shifts; change what's in *this* cart (the row) and
only this total shifts. The register never does anything but multiply-and-add — that's
what makes it *linear*. (Where the analogy strains: prices in a real store can't be
negative, but a weight can — tier C's −3 is a discount, pulling the total down.)

**Why an all-1s column, instead of just adding 10 at the end?** Because it lets the
*same* dot-product machine handle the base with no special case. If the base were bolted
on outside the multiply-and-add, every method would need a separate "and then add the
base" rule. Give the base its own column of 1s and its weight, and it's just one more
term in the same sum — $10 \times 1$ for every row. One uniform operation, no
exceptions. That uniformity is precisely why papers write the whole thing as a single
matrix times a single weight vector.

**Why does a 3-level category (A/B/C) become only *two* 0/1 columns, not three?** Here's
the trap it dodges. Suppose you made three columns — is-A, is-B, is-C. Every row has
exactly one of them equal to 1, so is-A + is-B + is-C is a column of all 1s — the same
as the base column you already have. That redundancy makes the combination ambiguous
(you could shuffle weight between the base and is-A and get identical scores), which is
poison for anything that later tries to pin the weights down. Dropping one level — here
tier A — fixes it: tier A becomes the **reference**, its "effect" folded into the base,
and B and C are read as *bumps relative to A*. A $k$-level category always needs
$k-1$ columns. Our tier-B weight of 5 means "a tier-B task scores 5 more than an
otherwise-identical tier-A task," and tier-C's −3 means "3 fewer than tier A."

**What the scores feel like.** They came out 18, 19, 19, 21 — all clustered near the
base of 10 plus a couple times the hours, which feels right: hours (weight 2) is the
main driver, and the tier bumps (+5, −3) nudge from there. Tasks 2 and 3 tie at 19
from opposite directions: task 3 has far more hours (+12) but the tier-C penalty (−3)
and no tier-B bonus drag it back to where task 2's fewer hours plus the tier-B bonus
land. Double every weight and every score's gap from zero doubles; add a task with 0
hours and tier A and it scores exactly the base, 10.

## What the notation is going to say

The formula below will say: *to score one row, walk across its columns, multiply each
column's value by that column's weight, and add the results.* For task 1 that's exactly
the "$10\times1 + 2\times4 + 5\times0 + (-3)\times0 = 18$" we did by hand. Stack all the
rows and it's "do that for every row" — which, since each row's score is a dot product,
is precisely the matrix-times-vector `X %*% beta` from
[matrices & linear transforms](../matrices-and-linear-transforms/lesson.qmd).

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $\mathbf{X}$ (bold capital) | "the matrix X" | the data matrix: rows = observations, columns = variables (in R: a `matrix`) | the 4×4 grid `X` |
| $\mathbf{x}_i$ | "ex sub eye" | row $i$ of the data matrix — one observation's values across all columns | row 1 is $(1, 4, 0, 0)$ |
| $X_{ij}$ | "X-sub-i-j" | the entry in row $i$, column $j$ (in R: `X[i, j]`) | $X_{12} = 4$ (task 1's hours) |
| $x_j$ | "ex sub jay" | the value in column $j$ of the row being scored | column 2 (hours) of a row |
| $\beta_j$ | "beta sub jay" | the weight multiplying column $j$ (in R: an entry of `beta`); column 1 is the all-1s column, so $\beta_1$ is the base | $\beta_1 = 10$ (base), $\beta_2 = 2$ (per hour) |
| $\boldsymbol{\beta}$ (bold) | "beta" / "the weight vector" | the whole column of weights, one per matrix column (in R: `beta`) | $(10, 2, 5, -3)$ |
| $\beta_0$ | "beta zero" | the name papers give the base term when they pull it out of the matrix — the same number as $\beta_1$ here | $10$ |
| $p$ | "pee" | the number of columns (variables) in the data matrix | 4 |
| $\sum_{j=1}^{p}$ | "the sum from j equals 1 to p" | add the term to its right once per column $j$ — a dot product ([vectors & summation](../vectors-and-summation/lesson.qmd)) | $\sum_j X_{1j}\beta_j = 18$ |
| $\mathbf{X}\boldsymbol{\beta}$ | "X beta" | the whole matrix times the whole weight vector: one linear combination per row (in R: `X %*% beta`) | $(18, 19, 19, 21)$ |

Bold lowercase ($\mathbf{x}_i$) is one row read as a vector; bold capital
($\mathbf{X}$) is the whole matrix; a plain letter with a subscript ($\beta_j$, $x_j$)
is a single number. When $\mathbf{X}$ sits right next to $\boldsymbol{\beta}$ with
nothing between them, that juxtaposition *is* the multiply — the same implied
multiplication as $2x$ meaning $2\times x$.

## The formal version

Assemble the score for one row, one term at a time. Row $i$ has values
$x_1, x_2, \dots, x_p$ across its $p$ columns, and there's a weight $\beta_j$ for each
column. The score is every value times its weight, added up:

$$
\text{score}_i \;=\; \beta_1 x_1 + \beta_2 x_2 + \dots + \beta_p x_p
\;=\; \sum_{j=1}^{p} \beta_j\, x_j
$$

Reading it aloud: "the score for row i equals the sum, from j equals 1 to p, of
beta-sub-j times ex-sub-j" — or in everyday words: **"multiply each column by its
weight and add."** That sum-of-products is a dot product, so it's the same engine as
the warm-up.

The all-1s column is not a special case — it's just column 1 with every value equal to
1, so its term is $\beta_1 \times 1 = \beta_1$ for every row: a constant $\beta_1$ added
to every score. Papers usually lift that constant out of the sum and rename it $\beta_0$
(the *intercept*) — the same number ($10$ here), just written outside the Σ instead of
as the weight on a column of ones.

Now stack all the rows. Each row's score is a dot product, and "one dot product per
row" is exactly what matrix-times-vector does, so the whole batch of scores is:

$$
\mathbf{X}\boldsymbol{\beta}
$$

Reading it aloud: "X beta" — everyday words: **"the data matrix times the weight
vector gives one linear combination (one score) per row."** Mapping onto the toy
example, row $i = 1$ is $\mathbf{x}_1 = (1, 4, 0, 0)$ and
$\boldsymbol{\beta} = (10, 2, 5, -3)$, so
$\sum_j X_{1j}\beta_j = 10\cdot1 + 2\cdot4 + 5\cdot0 + (-3)\cdot0 = 18$ — the "18" we got
by hand. Let's re-do the whole thing through the formula and confirm it matches:

```{r formal-check}
X <- cbind(one   = rep(1, 4),
           hours = c(4, 2, 6, 3),
           tierB = as.integer(c("A","B","C","B") == "B"),
           tierC = as.integer(c("A","B","C","B") == "C"))
beta <- c(10, 2, 5, -3)

# The whole batch of scores, straight from X %*% beta:
as.vector(X %*% beta)                       # 18 19 19 21

# The same scores, spelled out as one dot product (sum of products) per row:
apply(X, 1, function(row) sum(row * beta))  # 18 19 19 21 — identical
```

### Reading a weight: "change per unit, others held fixed"

Because each column contributes its own term and nothing else, a weight has a clean
reading: **$\beta_j$ is how much the score changes when column $j$ goes up by one and
every other column stays put.** Bump task 1's hours from 4 to 5 and only the
$\beta_2 x_2 = 2\times\text{hours}$ term moves — up by exactly $\beta_2 = 2$. A 0/1
column reads the same way: flipping tier-B from 0 to 1 (holding hours fixed) adds
exactly $\beta_3 = 5$, which is why we call it "5 more than tier A."

```{r held-fixed-check}
beta <- c(10, 2, 5, -3)
row1  <- c(1, 4, 0, 0)   # task 1 as it is
row1b <- c(1, 5, 0, 0)   # same task, one more hour
sum(row1b * beta) - sum(row1 * beta)   # 2 — exactly beta for the hours column
```

### The R bridge: `model.matrix()`

Building the all-1s and 0/1 columns by hand gets tedious, so base R has a function that
does it from a formula: `model.matrix()`. Shown here **neutrally** — it's just a
column-builder; it makes *no* decision about what the columns are for. Notice it
produces exactly the grid we built by hand: an intercept column of 1s, the numeric
column as-is, and $k-1$ indicator columns for the $k$-level category (tier A absorbed as
the reference).

```{r model-matrix}
hours <- c(4, 2, 6, 3)
tier  <- factor(c("A", "B", "C", "B"), levels = c("A", "B", "C"))

model.matrix(~ hours + tier)   # (Intercept), hours, tierB, tierC — our grid exactly
```

## Explore it

```{webr}
#| autorun: true
# explore-1 — LIVE: a linear combination IS one dot product per row.
# TRY IT: change a weight in beta, or a value in the matrix, then Run.
# PREDICT: set beta <- c(0, 0, 0, 0). What will every score be? Commit, then run.
X <- cbind(one   = rep(1, 4),
           hours = c(4, 2, 6, 3),
           tierB = as.integer(c("A","B","C","B") == "B"),
           tierC = as.integer(c("A","B","C","B") == "C"))
beta <- c(10, 2, 5, -3)

built_in <- as.vector(X %*% beta)                    # matrix times vector
by_hand  <- apply(X, 1, function(r) sum(r * beta))   # dot product per row, by hand

cbind(built_in = built_in, by_hand = by_hand)        # identical columns
```

```{webr}
#| autorun: true
# explore-2 — LIVE: a k-level category becomes k-1 columns; the reference is absorbed.
# TRY IT: change the levels order to c("C","A","B") and Run — watch which level becomes
# the reference (the one that gets NO column of its own).
# PREDICT: with levels A,B,C, how many indicator columns for tier? Commit, then run.
tier <- factor(c("A", "B", "C", "B"), levels = c("A", "B", "C"))

model.matrix(~ tier)   # (Intercept) + tierB + tierC : 3 levels -> 2 columns, A absorbed
```

```{webr}
#| autorun: false
# explore-3 — LIVE: an interaction column is literally the product of two columns.
# PREDICT: for hours = (4,2,6,3) and tierB = (0,1,0,1), what is hours * tierB,
# entry by entry? Write the four numbers, then Run.
hours <- c(4, 2, 6, 3)
tierB <- c(0, 1, 0, 1)

hours * tierB                        # the interaction column, built by hand
# Compare with what model.matrix builds for the "hours:tierB" term:
tier <- factor(c("A","B","C","B"), levels = c("A","B","C"))
# model.matrix(~ hours * tier)       # uncomment to see hours:tierB and hours:tierC
```

## Common traps

**Trap 1 — "a 3-level category needs 3 indicator columns, one per level."** It feels
natural: one column per category. But with an all-1s column already present, three
indicator columns are redundant — is-A + is-B + is-C equals the all-1s column exactly,
so the weights can't be pinned down (shift 4 from the base to is-A and every score is
unchanged). A $k$-level category takes $k-1$ columns; the dropped level (here tier A)
becomes the reference, and the remaining weights read as differences *from* it. In our
grid, `model.matrix(~ tier)` gives just `tierB` and `tierC` — two columns for three
levels — precisely to avoid this.

**Trap 2 — "the interaction column is the sum of the two columns."** Tempting, because
"combine two columns" sounds like adding. But an interaction is the **product**,
entry-by-entry, not the sum. For `hours = (4, 2, 6, 3)` and `tierB = (0, 1, 0, 1)`, the
interaction is `hours * tierB = (0, 2, 0, 3)` — it's `hours` only where tier is B, and
zero elsewhere. The *sum* `hours + tierB = (4, 3, 6, 4)` is a completely different
column with no such meaning. An interaction column lets one variable's weight *differ by
group*; a sum can't do that.

## Check yourself

**1.** Using the scoring weights `beta <- c(10, 2, 5, -3)` (base, per-hour, tier-B,
tier-C), score a new task: **5 hours, tier C**. Build its row first, then take the dot
product by hand, and check in R.

```{webr}
#| autorun: false
# check1-starter — edit and Run to check your hand answer
beta <- c(10, 2, 5, -3)
# build the row: one, hours, tierB, tierC
```

::: {.callout-note collapse="true" title="Answer"}
The row is $(1, 5, 0, 1)$: the all-1s entry, 5 hours, tier-B is 0, tier-C is 1. Dot it
with the weights: $10\times1 + 2\times5 + 5\times0 + (-3)\times1 = 10 + 10 + 0 - 3 =
\mathbf{17}$. In R, `sum(c(1, 5, 0, 1) * beta)` returns 17.
:::

**2.** A category has **4 levels** (say sizes S, M, L, XL). If your data matrix already
has an all-1s column, how many 0/1 indicator columns does this category contribute, and
what is the "missing" level called?

::: {.callout-note collapse="true" title="Answer"}
It contributes $4 - 1 = 3$ indicator columns. The level with no column of its own (say
S, if it's first) is the **reference**: its effect is folded into the base weight, and
the other three weights read as bumps relative to S. Using all 4 columns alongside the
all-1s column would be redundant — they'd sum to the all-1s column and the weights
couldn't be pinned down.
:::

**3.** In your own words: explain to a colleague, in two sentences, what a linear
combination of a data row is and why we bother packing the data into a matrix first.

::: {.callout-note collapse="true" title="A model answer"}
A linear combination scores one row by multiplying each of its column values by a
matching weight and adding the results — a shopping-cart total where the columns are
items and the weights are prices. We pack the data into a matrix so the *same*
multiply-and-add runs on every row at once as one matrix-times-vector, with the base and
the categories built in as ordinary columns rather than special cases.
:::

## What you can now do

- **Compute a linear combination of a data row by hand** — multiply each column value
  by its weight and add — and get the same answer as `X %*% beta`.
- **Read a data table as a matrix**: one row per observation, one column per variable,
  and pull entry $(i, j)$ with `X[i, j]`.
- **Build the special columns**: an all-1s column so a base term rides along, $k-1$
  indicator (0/1) columns for a $k$-level category with one level as the reference, and
  an interaction column as the entry-by-entry product of two columns.
- **Read a weight as "change per unit of this column, with the others held fixed,"** and
  a category weight as a difference from the reference level.
- **Recognize `model.matrix()`** as the R bridge that builds exactly these columns from
  a formula.

If any of these feels shaky, revisit [A tiny example first](#a-tiny-example-first) and
[The formal version](#the-formal-version) before continuing — anything that scores rows
from a table of variables assumes all of these.

## Where this goes next

A data matrix plus a weight vector is the shape almost every multi-variable method is
written in: the method's job is to *choose* the weights, but the scoring itself is
always this one linear combination per row. You now have the columns (base, categories,
interactions) and the reading of each weight. Ready for more reps? Do the
[practice problems](practice.qmd).
