---
title: "Practice: Linear combinations & the data matrix"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only (a {webr} starter does not see it — each starter re-states its
# own data). Base R is all we need, so nothing to install.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Do them on paper or in an R session; only
then open the answer.

## Problem 1 — the toy example, new numbers

Three items, each with a `units` count and a 0/1 `flag`. The scoring rule: base of 4,
add 3 per unit, add 2 if flagged. Build the data matrix (remember the all-1s column for
the base), then score every row as one dot product per row. Check with `%*%`.

$$
\text{weights} = (\underbrace{4}_{\text{base}},\ \underbrace{3}_{\text{per unit}},\ \underbrace{2}_{\text{flag}}), \qquad
\begin{array}{c|cc}
\text{item} & \text{units} & \text{flag} \\ \hline
1 & 2 & 1 \\
2 & 5 & 0 \\
3 & 1 & 1
\end{array}
$$

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
units <- c(2, 5, 1)
flag  <- c(1, 0, 1)
beta  <- c(4, 3, 2)   # base, per-unit, flag
```

::: {.callout-note collapse="true" title="Answer"}
Build the matrix with an all-1s column so the base of 4 rides along, then dot each row
with the weights:

- Item 1: $4\times1 + 3\times2 + 2\times1 = 4 + 6 + 2 = 12$
- Item 2: $4\times1 + 3\times5 + 2\times0 = 4 + 15 + 0 = 19$
- Item 3: $4\times1 + 3\times1 + 2\times1 = 4 + 3 + 2 = 9$

So the scores are $(12, 19, 9)$.

```{r p1-answer}
units <- c(2, 5, 1)
flag  <- c(1, 0, 1)
beta  <- c(4, 3, 2)

X <- cbind(one = 1, units = units, flag = flag)   # all-1s column carries the base
X %*% beta                                         # 12, 19, 9
c(sum(X[1, ] * beta), sum(X[2, ] * beta), sum(X[3, ] * beta))  # 12 19 9 — same, by hand
```
:::

## Problem 2 — fill in the missing step (indicator columns)

A 3-level category `grade` takes values `("lo", "hi", "mid", "hi")` across 4 rows, with
**`hi` chosen as the reference** level. Here is the build of its indicator columns — but
the **core step is blanked**. Fill in ONLY the blanked line: the two 0/1 columns.

> With an all-1s column already in the matrix, a 3-level category needs $3 - 1 = 2$
> indicator columns, and `hi` is the reference (it gets no column of its own).
> **`<your step here>`** — build `gradeLo` (1 where grade is "lo", else 0) and
> `gradeMid` (1 where grade is "mid", else 0):
> $\texttt{gradeLo} = (\underline{\phantom{0}},\underline{\phantom{0}},\underline{\phantom{0}},\underline{\phantom{0}})$,
> $\texttt{gradeMid} = (\underline{\phantom{0}},\underline{\phantom{0}},\underline{\phantom{0}},\underline{\phantom{0}})$.
> Every row that is `hi` shows 0 in both columns — that's how the reference is absorbed.

```{webr}
#| autorun: false
# p2-starter — edit and Run to check your two columns
grade <- c("lo", "hi", "mid", "hi")
```

::: {.callout-note collapse="true" title="Answer"}
The blanked step is the heart of turning a category into numbers — one 0/1 column per
non-reference level:

- `gradeLo` = 1 where `grade == "lo"`, else 0 $= (1, 0, 0, 0)$
- `gradeMid` = 1 where `grade == "mid"`, else 0 $= (0, 0, 1, 0)$

Rows 2 and 4 (both `hi`) are 0 in both columns — `hi` is the reference, folded into the
base. R's `model.matrix()` builds the identical pair when `hi` is the first factor level:

```{r p2-answer}
grade <- c("lo", "hi", "mid", "hi")
gradeLo  <- as.integer(grade == "lo")    # 1 0 0 0
gradeMid <- as.integer(grade == "mid")   # 0 0 1 0
rbind(gradeLo, gradeMid)

# model.matrix agrees (hi = reference because it's the first level here):
g <- factor(grade, levels = c("hi", "lo", "mid"))
model.matrix(~ g)                         # (Intercept), glo, gmid
```
:::

## Problem 3 — score a table with a category

A data matrix `X` holds an all-1s column, a numeric column `x1`, and a 0/1 column `catB`
(1 for category B, 0 for the reference category A). The weights are base 1, `x1` weight
2, `catB` weight 4. Score all three rows **by hand** as one dot product per row, then
confirm in R.

$$
X = \begin{bmatrix} 1 & 3 & 0 \\ 1 & 1 & 1 \\ 1 & 2 & 1 \end{bmatrix}, \qquad
\boldsymbol{\beta} = (1,\ 2,\ 4)
$$

```{webr}
#| autorun: false
# p3-starter — your job: score each row (dot product), then check with %*%
X <- matrix(c(1, 3, 0,
              1, 1, 1,
              1, 2, 1), nrow = 3, byrow = TRUE)
beta <- c(1, 2, 4)
```

::: {.callout-note collapse="true" title="Answer"}
Each score is the row dotted with the weights — base, plus 2 per unit of `x1`, plus 4 if
`catB` is 1:

- Row 1 (cat A): $1\times1 + 2\times3 + 4\times0 = 1 + 6 + 0 = 7$
- Row 2 (cat B): $1\times1 + 2\times1 + 4\times1 = 1 + 2 + 4 = 7$
- Row 3 (cat B): $1\times1 + 2\times2 + 4\times1 = 1 + 4 + 4 = 9$

So the scores are $(7, 7, 9)$. Rows 1 and 2 tie at 7 from opposite directions — row 1
has more `x1`, row 2 gets the category-B bump.

```{r p3-answer}
X <- matrix(c(1, 3, 0,
              1, 1, 1,
              1, 2, 1), nrow = 3, byrow = TRUE)
beta <- c(1, 2, 4)
X %*% beta                                   # 7, 7, 9
apply(X, 1, function(r) sum(r * beta))       # 7 7 9 — one dot product per row
```
:::

## Problem 4 — spot the error

A colleague turns a 3-level category `size` = ("S", "M", "L") into columns for a matrix
that already has an all-1s column. Their reasoning contains **exactly one mistake** —
find it, explain why it's wrong, and fix it.

> "There are 3 levels, so I'll make 3 indicator columns: `isS`, `isM`, `isL`, one per
> level. Each row gets a 1 in exactly one of them. Then I stack them next to the all-1s
> column and I'm done."

```{webr}
#| autorun: false
# p4-starter — copy this to see the redundancy for yourself
size <- c("S", "M", "L", "M", "S")
isS <- as.integer(size == "S")
isM <- as.integer(size == "M")
isL <- as.integer(size == "L")
one <- rep(1, length(size))
# Compare the all-1s column to isS + isM + isL:
# cbind(one, isS + isM + isL)   # <- uncomment and Run
```

::: {.callout-note collapse="true" title="Answer"}
The mistake is making **3** indicator columns for a 3-level category *when an all-1s
column is already present*. A $k$-level category needs $k - 1$ columns, not $k$. With all
three, `isS + isM + isL` equals 1 in every row — exactly the all-1s column — so the
columns are redundant and the weights can't be pinned down (you could move weight
between the base and `isS` and get identical scores).

The fix: drop one level to be the **reference** (say S), keeping only `isM` and `isL`.
Then S is absorbed into the base, and the `isM`, `isL` weights read as differences from
S.

```{r p4-answer}
size <- c("S", "M", "L", "M", "S")
isS <- as.integer(size == "S"); isM <- as.integer(size == "M"); isL <- as.integer(size == "L")
one <- rep(1, length(size))
cbind(one, redundant_sum = isS + isM + isL)   # the sum IS the all-1s column: redundant
# Correct: keep only the k-1 = 2 non-reference columns
cbind(isM, isL)                                # S is the reference, absorbed into the base
```
:::

## Problem 5 — mixing it with vectors & summation *(uses [vectors-and-summation](../vectors-and-summation/lesson.qmd)'s dot product)*

A data matrix `X` has an all-1s column, a numeric `size` column, and a 0/1 `pro` column;
the weights are $\boldsymbol{\beta} = (5, 3, 7)$. Compute the scores $\mathbf{X}\boldsymbol{\beta}$
two ways — (a) directly with `%*%`, and (b) as an explicit **dot product** per row using
`sum(...)`, the sum-of-products from vectors & summation — and confirm they agree.

$$
X = \begin{bmatrix} 1 & 4 & 1 \\ 1 & 2 & 0 \end{bmatrix}, \qquad \boldsymbol{\beta} = (5, 3, 7)
$$

**PREDICT** first: what will the two scores be? Write them down, then run.

```{webr}
#| autorun: false
# p5-starter — edit and Run to solve, right here
X <- matrix(c(1, 4, 1,
              1, 2, 0), nrow = 2, byrow = TRUE)
beta <- c(5, 3, 7)
```

::: {.callout-note collapse="true" title="Answer"}
A linear combination per row is *literally* the dot product from vectors & summation,
$\sum_j X_{ij}\beta_j$:

- Row 1 (pro = 1): $5\times1 + 3\times4 + 7\times1 = 5 + 12 + 7 = 24$
- Row 2 (pro = 0): $5\times1 + 3\times2 + 7\times0 = 5 + 6 + 0 = 11$

So $\mathbf{X}\boldsymbol{\beta} = (24, 11)$.

```{r p5-answer}
X <- matrix(c(1, 4, 1,
              1, 2, 0), nrow = 2, byrow = TRUE)
beta <- c(5, 3, 7)

X %*% beta                                        # 24, 11
c(sum(X[1, ] * beta), sum(X[2, ] * beta))         # 24, 11 — dot product per row, spelled out
```

The `sum(X[i, ] * beta)` line is exactly the dot product from
[vectors & summation](../vectors-and-summation/lesson.qmd); `%*%` just does it for every
row at once.
:::

## Problem 6 — a gentle cliffhanger (an interaction column)

So far every weight has meant the same thing for every row. But suppose you want the
per-hour rate to be **different for tier B than for tier A**. The tool is an
**interaction column**: the entry-by-entry product of `hours` and the `tierB` indicator.

For `hours = (4, 2, 6, 3)` and `tierB = (0, 1, 0, 1)`, first build the interaction
column `hours * tierB`. Then score the four rows with weights base 10, per-hour 2,
tier-B 5, **and interaction 1**. **Predict:** which rows will the interaction term change
the score of, and which will it leave untouched? Then run to find out.

```{webr}
#| autorun: false
# p6-starter — build the interaction column, then score with the 4 weights
hours <- c(4, 2, 6, 3)
tierB <- c(0, 1, 0, 1)
beta  <- c(10, 2, 5, 1)   # base, per-hour, tier-B, interaction
```

::: {.callout-note collapse="true" title="Answer"}
The interaction column is `hours * tierB` = $(0, 2, 0, 3)$ — it equals `hours` where the
tier is B, and 0 everywhere else. So the interaction term only moves the tier-B rows
(rows 2 and 4); the tier-A rows (1 and 3) are untouched, because their interaction entry
is 0.

Stack `(one, hours, tierB, interaction)` and dot with $(10, 2, 5, 1)$:

- Row 1 (A): $10 + 2\times4 + 5\times0 + 1\times0 = 18$
- Row 2 (B): $10 + 2\times2 + 5\times1 + 1\times2 = 10 + 4 + 5 + 2 = 21$
- Row 3 (A): $10 + 2\times6 + 5\times0 + 1\times0 = 22$
- Row 4 (B): $10 + 2\times3 + 5\times1 + 1\times3 = 10 + 6 + 5 + 3 = 24$

```{r p6-answer}
hours <- c(4, 2, 6, 3)
tierB <- c(0, 1, 0, 1)
inter <- hours * tierB               # (0, 2, 0, 3) — the interaction column
X     <- cbind(one = 1, hours = hours, tierB = tierB, inter = inter)
beta  <- c(10, 2, 5, 1)
X %*% beta                           # 18, 21, 22, 24
```

The cliffhanger: with the interaction, a tier-B task now effectively earns $2 + 1 = 3$
per hour instead of 2 — the per-hour rate *bends by group*. **Choosing** good weights
for columns like these is exactly the job a specific method takes on; here you've built
the columns it would act on.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/linear-combinations-and-data-matrices" pass="4"}

::: {.quiz-q qid="lcdm-linear-combination" qtype="numeric" answer="16" tolerance="0"}
A data row is $(3, 2)$ with weights $\beta = (2, 5)$. Its score is the linear combination
$\sum_j x_j \beta_j$. What is it?
:::

::: {.quiz-q qid="lcdm-k-minus-1-indicators" qtype="mc" answer="b"}
A category has $3$ levels (A, B, C) and the model already carries an all-1s column. How
many indicator columns should you add?

a) $3$ — one per level
b) $2$ — a $k$-level category takes $k-1$ columns, with the dropped level as the reference
c) $1$ — a single column coding A/B/C as $1/2/3$
:::

::: {.quiz-q qid="lcdm-interaction-product" qtype="mc" answer="c"}
An interaction column between `hours` and `tierB` is built how?

a) By adding the two columns entry by entry
b) By concatenating them into one longer column
c) By multiplying the two columns entry by entry
:::

::: {.quiz-q qid="lcdm-weight-meaning" qtype="mc" answer="a"}
In a linear model, what does a variable's weight (coefficient) mean?

a) The change in the score per one-unit change in that column, with the others held fixed
b) The total contribution of that column summed over all rows
c) The correlation between that column and the outcome
:::

::: {.quiz-q qid="lcdm-matrix-product-shape" qtype="mc" answer="b"}
(From matrices & linear transforms.) Scoring every row at once is the matrix product
$X\beta$. For $X$ of shape $4 \times 2$ and $\beta$ a length-$2$ vector, what is $X\beta$?

a) A $4 \times 2$ matrix
b) A length-$4$ vector — one score per row
c) A length-$2$ vector
:::

:::

::::
