---
title: "Practice: Capstone — reading colocPropTest"
subtitle: "The cumulative set — every module, mixed order, no labels. Recalling which tool applies is the point."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

These draw on the **whole course**, in mixed order, with no labels telling you which module each
uses — figuring that out is part of the test. Try each honestly before opening the answer.

## Problem 1 — a fresh pairwise test

Two SNPs, $b_1 = (0.18, 0.09)$, $b_2 = (0.20, 0.10)$, all variances $0.0009$, LD $\rho = 0.6$.
Eyeball proportionality, then run the pairwise test. Colocalised?

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
prop_V <- function(vbeta, rho) { V <- diag(vbeta); V[1,2] <- V[2,1] <- sqrt(vbeta[1]*vbeta[2])*rho; V }
prop_test_pair <- function(beta1, vbeta1, beta2, vbeta2, rho1, rho2 = rho1) {
  V1 <- prop_V(vbeta1, rho1); V2 <- prop_V(vbeta2, rho2)
  chisq <- function(theta) { d <- sin(theta)*beta1 - cos(theta)*beta2
    Vs <- sin(theta)^2*V1 + cos(theta)^2*V2; as.numeric(t(d) %*% solve(Vs, d)) }
  oL <- optimize(chisq, c(0, pi/2)); oR <- optimize(chisq, c(pi/2, pi))
  o <- if (oL$objective < oR$objective) oL else oR
  list(chisquare = o$objective, p = pchisq(o$objective, df = length(beta1)-1L, lower.tail = FALSE))
}
```

::: {.callout-note collapse="true" title="Answer"}
Ratios $b_1/b_2 = (0.9, 0.9)$ — equal, so proportional → expect a **large** p (colocalised).

```{r p1-ans}
prop_V <- function(vbeta, rho) { V <- diag(vbeta); V[1,2] <- V[2,1] <- sqrt(vbeta[1]*vbeta[2])*rho; V }
prop_test_pair <- function(beta1, vbeta1, beta2, vbeta2, rho1, rho2 = rho1) {
  V1 <- prop_V(vbeta1, rho1); V2 <- prop_V(vbeta2, rho2)
  chisq <- function(theta) { d <- sin(theta)*beta1 - cos(theta)*beta2
    Vs <- sin(theta)^2*V1 + cos(theta)^2*V2; as.numeric(t(d) %*% solve(Vs, d)) }
  oL <- optimize(chisq, c(0, pi/2)); oR <- optimize(chisq, c(pi/2, pi))
  o <- if (oL$objective < oR$objective) oL else oR
  list(chisquare = o$objective, p = pchisq(o$objective, df = length(beta1)-1L, lower.tail = FALSE))
}
prop_test_pair(c(0.18, 0.09), c(0.0009, 0.0009), c(0.20, 0.10), c(0.0009, 0.0009), 0.6)
```
Chi-square $\approx 0$, p $\approx 1$ → **fail to reject** proportionality → colocalised. (Recall
the flip: large p is the friendly outcome.)
:::

## Problem 2 — interpretation, two methods

A region gives coloc PP4 = 0.05, PP3 = 0.94, and the proportional test `min_adjusted` = 0.80.
Do the two methods agree? What should you suspect?

::: {.callout-note collapse="true" title="Answer"}
They **disagree**. coloc says *distinct* (PP3 = 0.94). The proportional test's *large*
`min_adjusted` (0.80) means *fail to reject* colocalisation — i.e. it *supports* colocalisation.
When coloc confidently calls "distinct" but the proportional test says "consistent with shared,"
you should suspect coloc's call is a **differential-information artifact**: unequal statistical
information across SNPs biased coloc's fine-maps (the whole point of the course). Investigate data
quality (imputation, missingness, rounding) before trusting the "distinct" call.
:::

## Problem 3 — spot the error

A colleague reports — it contains **exactly one mistake**. Find it.

> "The proportional test gave `min_adjusted` = 0.001. That's a small p-value, so we conclude the
> two traits colocalise. We used `p.adjust`'s default, which gives a proper Benjamini–Hochberg
> false discovery rate."

::: {.callout-note collapse="true" title="Answer"}
There are actually two things off, but the load-bearing mistake is the **direction**: the null is
colocalisation, so `min_adjusted` = 0.001 (small) *rejects* it → the traits are **NOT** colocalised
(distinct). (Secondary: `p.adjust`'s default is **Holm**, a family-wise correction, not
Benjamini–Hochberg — the package names the column `fdr` but applies Holm.) The corrected
conclusion: "small adjusted p → reject colocalisation → distinct signals," and "the correction is
Holm, not BH."
:::

## Problem 4 — why coloc broke

Two SNPs in perfect LD carry a genuinely shared signal, but coloc returns PP3 = 0.99 (distinct).
In two sentences, explain the chain of reasoning that produced this wrong, confident answer.

::: {.callout-note collapse="true" title="Answer"}
Unequal statistical information across the two SNPs (e.g. differential imputation or missingness)
biased each trait's per-SNP Bayes factors, pushing trait 1's fine-mapping peak onto one SNP and
trait 2's onto the other. coloc decides shared-vs-distinct by checking whether the traits' peaks
align, so the misaligned peaks made it report distinct (PP3) — confidently, because each biased
fine-map was itself a confident single-variant credible set.
:::

## Problem 5 — build the covariance and the statistic

For a pair with $\text{se} = (0.03, 0.04)$, LD $\rho = 0.5$: build $V$ for one trait, then, at
angle $\theta = \pi/3$ with gap $d = (0.02, -0.01)$, compute the quadratic form $d^\top V^{-1} d$.

```{webr}
#| autorun: false
# p5-starter — edit and Run to solve, right here
se <- c(0.03, 0.04); rho <- 0.5
d <- c(0.02, -0.01)
```

::: {.callout-note collapse="true" title="Answer"}
$V$ diagonal $0.0009, 0.0016$; off-diagonal $0.03\times 0.04\times 0.5 = 0.0006$.

```{r p5-ans}
prop_V <- function(vbeta, rho) { V <- diag(vbeta); V[1,2] <- V[2,1] <- sqrt(vbeta[1]*vbeta[2])*rho; V }
V <- prop_V(c(0.03^2, 0.04^2), 0.5); V
d <- c(0.02, -0.01)
as.numeric(t(d) %*% solve(V, d))          # the quadratic form d' V^{-1} d
```
The quadratic form is about $0.9$ — a small standardised squared distance (the gaps are modest
relative to their standard errors, and the two gaps partly offset through the covariance). This is
the same "sandwich the inverse covariance" computation the test statistic uses at each angle; a
value near 1 is unremarkable for pure noise.
:::

## Problem 6 — decode a fresh interface line

You've never seen this exact line, but you built its meaning. Decode it in plain words, naming
which of your functions it corresponds to:

```r
V1 <- marg_with_V(beta = S1$beta, vbeta = S1$varbeta, rho = LD[i, j])
```

::: {.callout-note collapse="true" title="Answer"}
It builds trait 1's covariance matrix for the pair of SNPs $(i, j)$ from that trait's effect
estimates' *variances* (`varbeta` = $\text{se}^2$) and the LD between the two SNPs (`LD[i, j]`),
using $V = (\sigma\sigma^\top)\odot\Sigma$. It is exactly your **`prop_V(vbeta, rho)`** — diagonal
$\text{se}^2$, off-diagonal $\text{se}_1\text{se}_2\rho$. `marg_with_V` is `colocPropTest`'s name
for it.
:::

## Problem 7 — the full pipeline, fresh region

Run `prop_run` on this 4-SNP region and decide colocalised or not. Trait 1 peaks on SNP 1, trait 2
peaks on SNP 3 (distinct).

```{webr}
#| autorun: false
# p7-starter — edit and Run to solve, right here
prop_V <- function(vbeta, rho) { V <- diag(vbeta); V[1,2] <- V[2,1] <- sqrt(vbeta[1]*vbeta[2])*rho; V }
prop_test_pair <- function(beta1, vbeta1, beta2, vbeta2, rho1, rho2 = rho1) {
  V1 <- prop_V(vbeta1, rho1); V2 <- prop_V(vbeta2, rho2)
  chisq <- function(theta) { d <- sin(theta)*beta1 - cos(theta)*beta2
    Vs <- sin(theta)^2*V1 + cos(theta)^2*V2; as.numeric(t(d) %*% solve(Vs, d)) }
  oL <- optimize(chisq, c(0, pi/2)); oR <- optimize(chisq, c(pi/2, pi))
  o <- if (oL$objective < oR$objective) oL else oR
  list(p = pchisq(o$objective, df = length(beta1)-1L, lower.tail = FALSE))
}
prop_run <- function(b1, v1, b2, v2, LD, pairs, ntests) {
  p <- numeric(ncol(pairs))
  for (m in seq_len(ncol(pairs))) { i <- pairs[1,m]; j <- pairs[2,m]
    p[m] <- prop_test_pair(b1[c(i,j)], v1[c(i,j)], b2[c(i,j)], v2[c(i,j)], LD[i,j])$p }
  list(adjusted = p.adjust(p, "holm", n = ntests), min_adjusted = min(p.adjust(p, "holm", n = ntests)))
}
b1 <- c(0.30, 0.02, 0.01, 0.00); b2 <- c(0.01, 0.02, 0.32, 0.01)
LD <- matrix(0.7, 4, 4); diag(LD) <- 1; pairs <- combn(4, 2)
```

::: {.callout-note collapse="true" title="Answer"}
```{r p7-ans}
prop_V <- function(vbeta, rho) { V <- diag(vbeta); V[1,2] <- V[2,1] <- sqrt(vbeta[1]*vbeta[2])*rho; V }
prop_test_pair <- function(beta1, vbeta1, beta2, vbeta2, rho1, rho2 = rho1) {
  V1 <- prop_V(vbeta1, rho1); V2 <- prop_V(vbeta2, rho2)
  chisq <- function(theta) { d <- sin(theta)*beta1 - cos(theta)*beta2
    Vs <- sin(theta)^2*V1 + cos(theta)^2*V2; as.numeric(t(d) %*% solve(Vs, d)) }
  oL <- optimize(chisq, c(0, pi/2)); oR <- optimize(chisq, c(pi/2, pi))
  o <- if (oL$objective < oR$objective) oL else oR
  list(p = pchisq(o$objective, df = length(beta1)-1L, lower.tail = FALSE))
}
prop_run <- function(b1, v1, b2, v2, LD, pairs, ntests) {
  p <- numeric(ncol(pairs))
  for (m in seq_len(ncol(pairs))) { i <- pairs[1,m]; j <- pairs[2,m]
    p[m] <- prop_test_pair(b1[c(i,j)], v1[c(i,j)], b2[c(i,j)], v2[c(i,j)], LD[i,j])$p }
  min(p.adjust(p, "holm", n = ntests))
}
b1 <- c(0.30, 0.02, 0.01, 0.00); b2 <- c(0.01, 0.02, 0.32, 0.01)
LD <- matrix(0.7, 4, 4); diag(LD) <- 1; pairs <- combn(4, 2)
prop_run(b1, rep(0.0009,4), b2, rep(0.0009,4), LD, pairs, ntests = 6)
```
`min_adjusted` is tiny (far below 0.05) → **reject colocalisation** → distinct. The pair (SNP 1,
SNP 3) breaks proportionality hard (trait 1 huge on SNP 1, trait 2 huge on SNP 3).
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**, once the
material isn't fresh. A pass then is real evidence you've retained it. Five questions, graded on
the page; 4/5 or better means the whole course is solid. Results are saved in this browser only.

::: {.module-check module="courses/proportional-colocalisation/modules/06-capstone-reading-colocproptest" pass="4"}

::: {.quiz-q qid="cap-null-flip" qtype="mc" answer="c"}
`colocPropTest` returns `min_adjusted` = 0.002 for a region. The correct conclusion is:

a) the traits colocalise, with strong evidence
b) the test failed
c) the traits do NOT colocalise — the small adjusted p rejects the null of proportionality
d) the traits colocalise only at SNP 1
:::

::: {.quiz-q qid="cap-marg-with-V" qtype="mc" answer="b"}
`marg_with_V(beta, vbeta, rho)` in `colocPropTest` computes:

a) the five coloc posteriors
b) the LD covariance $V = (\sigma\sigma^\top)\odot\Sigma$ for a pair
c) the profiled chi-square statistic
d) the Holm-adjusted p-values
:::

::: {.quiz-q qid="cap-df" qtype="numeric" answer="1" tolerance="0"}
The profiled Fieller statistic for a **pair** of SNPs follows a chi-square with how many degrees of
freedom?
:::

::: {.quiz-q qid="cap-when-to-use" qtype="mc" answer="a"}
Which is true about the proportional test versus coloc?

a) the proportional test requires equal LD between studies but does not require the causal variant to be typed
b) the proportional test requires the causal variant to be typed but not equal LD
c) both methods require the causal variant to be typed
d) neither method makes any assumption about LD
:::

::: {.quiz-q qid="cap-corruption-cause" qtype="mc" answer="c"}
The paper's whole motivation is that coloc can give confident wrong calls because:

a) coloc uses the wrong prior
b) coloc runs too few tests
c) variable statistical information across SNPs biases the fine-mapping coloc is built on
d) coloc ignores linkage disequilibrium
:::

:::

::::
