---
title: "Choosing a model: information criteria"
subtitle: "Why 'lowest error' picks the wrong model, and how AIC/BIC fix it"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

## Where we're going

You have a pool of candidate predictors and want the *best* model. The obvious idea —
"pick the model with the smallest error" — is a trap: adding *any* predictor, even pure
noise, can only lower the training error, so "smallest error" always votes for the
biggest model. This module builds the fix: score a model by **fit *plus* a penalty for
complexity**. That's what **AIC** and **BIC** do, and it's how methods choose *how many*
predictors to keep. You'll also meet the search problem — there are $2^p$ possible
subsets — and the greedy shortcuts (forward/backward stepwise) that don't always find the
best one.

**Builds on:** [model-fit-error-and-validation](../model-fit-error-and-validation/lesson.qmd)
(RSS, $R^2$, overfitting), [likelihood-and-log-likelihood](../likelihood-and-log-likelihood/lesson.qmd)
(the log-likelihood, which the criteria are built from).

## Warm-up: do you remember?

**1.** *(model fit)* What is the residual sum of squares $\mathrm{RSS}$, in one line, and
does adding a predictor to a model ever *increase* it?

::: {.callout-note collapse="true" title="Answer"}
$\mathrm{RSS} = \sum_i (y_i - \hat{y}_i)^2$, the total squared miss. Adding a predictor
**never increases** it — the fit can always ignore the new column (coefficient 0) or use
it to shave off a little more, so RSS only stays equal or drops. See
[model-fit-error-and-validation](../model-fit-error-and-validation/lesson.qmd).
:::

**2.** *(likelihood)* What does the **log-likelihood** measure, and do we want it large or
small for a good fit?

::: {.callout-note collapse="true" title="Answer"}
The log-likelihood scores how plausible the observed data is under the model; a
better-fitting model makes the data more plausible, so we want it **large** (equivalently,
$-2\times$ log-likelihood **small**). See
[likelihood-and-log-likelihood](../likelihood-and-log-likelihood/lesson.qmd).
:::

## A tiny example first

Five points, one predictor $x_1 = (1,2,3,4,5)$ and $y = (1,3,2,5,4)$.

**Model 0 — just the mean.** With no predictor, the best guess is $\bar{y} = 3$. Its
error is $\mathrm{RSS}_0 = (1-3)^2 + (3-3)^2 + (2-3)^2 + (5-3)^2 + (4-3)^2 = 4+0+1+4+1 = 10$.

**Model 1 — add $x_1$.** Fit a line; its $\mathrm{RSS}_1 = 3.6$ — much lower. Good, $x_1$
helps.

**Model 2 — add a second predictor $x_2$ that is pure noise** (random numbers unrelated
to $y$). Its $\mathrm{RSS}_2 = 3.351$. **Lower still** — even though $x_2$ is meaningless!
That's the whole problem: "smallest RSS" would pick Model 2, chasing noise.

```{r toy-example}
x1 <- c(1, 2, 3, 4, 5); y <- c(1, 3, 2, 5, 4)
set.seed(7); x2 <- rnorm(5)                 # a predictor with NO real relationship to y

rss <- function(fit) sum(resid(fit)^2)
c(mean_only = rss(lm(y ~ 1)),               # 10.0
  with_x1   = rss(lm(y ~ x1)),              # 3.6
  add_noise = rss(lm(y ~ x1 + x2)))         # 3.351 — lower, chasing noise
```

The cure: charge a **penalty** for each predictor, so a tiny RSS improvement isn't worth
it unless it beats the penalty.

## Getting a feel for it

**Paying rent for each predictor.** Think of each predictor as an employee you hire to
explain $y$. Every hire reduces the "unexplained work" (RSS) at least a little — even a
lazy one who just shuffles noise. If you judged your team only by "work remaining," you'd
hire everyone. So you charge **rent**: each employee must reduce the unexplained work by
more than their salary, or they're not worth keeping. Information criteria are that
salary rule — they add a per-predictor cost to the fit score, so only predictors that
*earn their keep* lower the total.

**Why a penalty and not just "use a test set."** Holding out data (cross-validation, the
[model-fit](../model-fit-error-and-validation/lesson.qmd) module's tool) also fights
overfitting, and it's excellent. Information criteria are a *cheaper, all-at-once*
alternative: one formula on the full-data fit, no re-splitting, with a penalty derived
from theory. They shine when you're comparing many models quickly — like a search over
predictor subsets.

**What the penalty feels like.** In the toy, going from Model 1 to Model 2 dropped RSS by
only $3.6 - 3.351 = 0.249$ — a rounding error's worth of "improvement" bought with a whole
extra predictor. A good penalty is bigger than that, so it *rejects* Model 2. AIC and BIC
both do; BIC (the heavier penalty) rejects extra predictors more aggressively.

## What the notation is going to say

The criteria say: *take a "badness" score built from the log-likelihood (small = good
fit), then add a penalty that grows with the number of parameters $k$.* AIC's penalty is
$2k$; BIC's is $k\log n$ — and because $\log n > 2$ once $n > 7$, BIC charges more per
parameter on any real dataset, so it prefers **smaller** models. Lowest score wins.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $\mathrm{RSS}$ | "residual sum of squares" | total squared miss $\sum_i (y_i-\hat{y}_i)^2$ | 10, 3.6, 3.351 |
| $\mathrm{TSS}$ | "total sum of squares" | total spread of $y$ around its mean $\sum_i (y_i-\bar{y})^2$ (the mean-only RSS) | 10 |
| $n$ | "en" | number of observations | 5 |
| $k$ | "kay" | number of free parameters (predictors + intercept + noise variance) | 3 for Model 1 |
| $p$ | "pee" | number of candidate predictors in the pool | 10 later |
| $\log$ | "log" (natural log) | natural logarithm (in R: `log`) | — |
| $\ell$, $\log L$ | "log-likelihood" | how plausible the data is under the fit (bigger = better) | −6.27 for Model 1 |
| $\mathrm{AIC}$ | "A-I-C" | $-2\log L + 2k$ (in R: `AIC`) | 18.55 for Model 1 |
| $\mathrm{BIC}$ | "B-I-C" | $-2\log L + k\log n$ (in R: `BIC`) | 17.38 for Model 1 |
| $2^p$ | "two to the pee" | how many subsets of $p$ predictors exist | $2^{10}=1024$ |

## The formal version

Both criteria have the same shape — **fit term + complexity penalty** — and both are
scored so that **lower is better**:

$$
\mathrm{AIC} = \underbrace{-2\log L}_{\text{misfit}} + \underbrace{2k}_{\text{penalty}},
\qquad
\mathrm{BIC} = \underbrace{-2\log L}_{\text{misfit}} + \underbrace{k\log n}_{\text{penalty}}.
$$

- The **misfit term** $-2\log L$ is small when the model makes the data plausible (from
  the [likelihood](../likelihood-and-log-likelihood/lesson.qmd) module). For a
  least-squares fit with Gaussian noise it is a rising function of $\mathrm{RSS}$, so
  "smaller RSS" *does* lower this term — it's the penalty that stops the runaway.
- The **penalty** counts parameters $k$. AIC charges 2 per parameter; BIC charges
  $\log n$. Since $\log n$ exceeds 2 for any $n > 7$, **BIC penalizes complexity harder
  and selects smaller models**; its penalty also *grows with the sample size*, so with
  lots of data BIC demands a predictor prove itself more.

Reading it aloud: "B-I-C equals minus-two log-likelihood, plus k times log n" — or in
everyday words: "badness of fit, plus a fine of $\log n$ for every knob you added."

Mapping onto the toy (R computes the log-likelihood; you read the ranking):

```{r toy-ic}
# PREDICT: which model will BIC crown — mean-only, x1, or x1+x2? (x2 is pure noise.)
x1 <- c(1, 2, 3, 4, 5); y <- c(1, 3, 2, 5, 4); set.seed(7); x2 <- rnorm(5)
m0 <- lm(y ~ 1); m1 <- lm(y ~ x1); m2 <- lm(y ~ x1 + x2)

rbind(AIC = c(mean = AIC(m0), x1 = AIC(m1), x1_x2 = AIC(m2)),
      BIC = c(BIC(m0), BIC(m1), BIC(m2)))

# BIC from the raw formula (-2 logL + k log n) equals the built-in BIC() — the formula IS
# the function. logLik's `df` is k (slopes + intercept + noise variance).
k1 <- attr(logLik(m1), "df")
c(from_formula = -2 * as.numeric(logLik(m1)) + k1 * log(5), builtin = BIC(m1))  # identical
```

Both criteria bottom out at **Model 1** (`x1` alone): BIC is 20.87 → **17.38** → 18.63.
The noise predictor `x2` *lowered RSS* but *raised BIC* — its tiny fit gain didn't cover
the penalty. The criterion made the call RSS couldn't.

## Explore it

### Best-subset: the right size, on a bigger pool

```{webr}
#| autorun: true
#| fig-alt: "A curve of BIC against model size from 0 to 5 predictors: it falls to a minimum at size 3 and then rises. A printed table shows RSS falling monotonically while BIC turns around at 3."
# TRY IT: change the true model (e.g. make only x1, x2 matter, or add x4) and re-run.
# PREDICT: the BIC curve's minimum should land on the number of predictors you kept real.
# Simulate y from just 3 of 10 predictors; find the best-RSS subset of each size,
# then score each by BIC. Watch RSS keep falling while BIC bottoms at the true size 3.
set.seed(1); n <- 120; p <- 10
X <- matrix(rnorm(n * p), n, p); colnames(X) <- paste0("x", 1:p)
y <- 2 * X[, 1] - 1.5 * X[, 2] + 1 * X[, 3] + rnorm(n)   # only x1, x2, x3 matter
dat <- data.frame(X, y)

best_of_size <- function(k) {                       # smallest-RSS model with k predictors
  if (k == 0) { fit <- lm(y ~ 1, dat); return(c(rss = sum(resid(fit)^2), bic = BIC(fit))) }
  combs <- combn(p, k)
  fits  <- apply(combs, 2, function(cols) lm(y ~ ., dat[, c(cols, p + 1)]))
  rss   <- sapply(fits, function(f) sum(resid(f)^2))
  best  <- fits[[which.min(rss)]]
  c(rss = min(rss), bic = BIC(best))
}
tab <- sapply(0:5, best_of_size); colnames(tab) <- 0:5
round(tab, 1)

plot(0:5, tab["bic", ], type = "b", lwd = 2, col = "firebrick",
     xlab = "number of predictors", ylab = "BIC (lower = better)",
     main = "BIC bottoms at the true size (3); RSS would never stop")
abline(v = 3, lty = 3)
```

RSS falls all the way to size 5; BIC turns around and bottoms at **3** — the true number
of real predictors.

### The search is huge, so we use greedy shortcuts

There are $2^p$ possible subsets — $2^{10} = 1024$ here, but $2^{30} \approx 10^9$ for 30
predictors. **Best-subset** (check them all) is infeasible past ~30 predictors, so we use
**forward stepwise** (start empty, add the single best predictor each step) or
**backward** (start full, drop the worst). R's `step()` does this:

```{r stepwise}
set.seed(1); n <- 120; p <- 10
X <- matrix(rnorm(n * p), n, p); colnames(X) <- paste0("x", 1:p)
y <- 2 * X[, 1] - 1.5 * X[, 2] + X[, 3] + rnorm(n)
dat <- data.frame(X, y)

fwd <- step(lm(y ~ 1, dat), scope = ~ x1+x2+x3+x4+x5+x6+x7+x8+x9+x10,
            direction = "forward", trace = 0)      # greedy: add best predictor each step
sort(names(coef(fwd))[-1])
```

## Common traps

**Trap: "the model with the lowest RSS (or highest $R^2$) is the best model."** False on
training data — RSS can only fall as you add predictors, so it always crowns the biggest
model (Model 2 beat Model 1 on RSS with a *noise* predictor). $R^2$ has the same flaw
(it's $1 - \mathrm{RSS}/\mathrm{TSS}$). Information criteria and adjusted $R^2$ exist
precisely because raw fit can't choose model size.

**Trap: "stepwise found it, so it's the best subset."** Greedy is not exhaustive. Forward
stepwise commits to its first pick and can never remove it, so an early mistake sticks —
its size-$k$ model can differ from the true best-subset-of-$k$ (a point ISLR's Table 6.1
makes). Above, the greedy search overshot the true `{x1, x2, x3}`. Greedy is a practical
necessity for large $p$, not a guarantee of the optimum.

## Check yourself

**1.** A model has log-likelihood $\log L = -50$ and $k = 4$ parameters, fit on $n = 100$
observations. Compute its AIC and BIC.

::: {.callout-note collapse="true" title="Answer"}
$\mathrm{AIC} = -2(-50) + 2\times4 = 100 + 8 = 108$.
$\mathrm{BIC} = -2(-50) + 4\log(100) = 100 + 4(4.605) = 100 + 18.42 = 118.42$. BIC is
larger because $\log(100) = 4.6 > 2$ — it fines each parameter more.
:::

**2.** You add a predictor and RSS drops from 3.6 to 3.55. Under what condition does BIC
still get *worse* (increase)?

::: {.callout-note collapse="true" title="Answer"}
When the drop in the misfit term ($-2\log L$) from that tiny RSS improvement is **smaller
than the added penalty** $\log n$. A 0.05 RSS drop barely moves the fit term, while the
penalty adds a full $\log n$, so BIC rises and rejects the predictor — exactly what
happened to the noise predictor in the toy.
:::

**3.** In your own words: explain to a colleague, in two sentences, why we don't just pick
the model with the smallest training error, and what AIC/BIC add.

::: {.callout-note collapse="true" title="A model answer"}
Training error only ever drops as you add predictors, so "smallest error" always picks the
biggest, most overfit model — including predictors that are just fitting noise. AIC and
BIC add a per-parameter penalty to the fit score, so a predictor is kept only if it
improves the fit by more than its complexity costs.
:::

## What you can now do

- **Explain why RSS/$R^2$ can't choose model size** — they only improve as predictors are
  added, even useless ones.
- **Write down AIC $= -2\log L + 2k$ and BIC $= -2\log L + k\log n$**, and say why BIC
  prefers smaller models (heavier, sample-size-growing penalty).
- **Read a BIC-vs-size curve** and locate the chosen model at its minimum.
- **Describe the search problem** ($2^p$ subsets) and why forward/backward stepwise are
  greedy approximations that can miss the best subset.

If any of these feels shaky, revisit [Getting a feel for it](#getting-a-feel-for-it)
before continuing — methods that select variables lean on all of it.

## Where this goes next

Information criteria are the *stopping rule* for any variable-selection search. A
switching search that adds, drops, and swaps predictors uses BIC to decide when to stop —
built in the `swap-stepwise-selection` course. Ready for reps? Do the
[practice problems](practice.qmd). External videos/readings: [resources](resources.md).
