---
title: "Practice: Choosing a model with information criteria"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Do them on paper or in an R session; only then
open the answer.

## Problem 1 — AIC and BIC by hand, new numbers

A model has log-likelihood $\log L = -80$ and $k = 3$ parameters, fit on $n = 50$
observations. Compute its AIC and BIC.

::: {.callout-note collapse="true" title="Answer"}
$\mathrm{AIC} = -2(-80) + 2\times3 = 160 + 6 = 166$.
$\mathrm{BIC} = -2(-80) + 3\log(50) = 160 + 3(3.912) = 160 + 11.74 = 171.74$. BIC exceeds
AIC because $\log(50) = 3.91 > 2$.

```{r}
logL <- -80; k <- 3; n <- 50
c(AIC = -2*logL + 2*k, BIC = -2*logL + k*log(n))   # 166, 171.74
```
:::

## Problem 2 — fill in the missing step

Two models fit on $n = 100$ points: Model A has $\log L = -120$, $k = 2$; Model B has
$\log L = -118$, $k = 5$. Which does BIC prefer?

- Step 1: $\mathrm{BIC}_A = -2(-120) + 2\log(100) = 240 + 2(4.605) = 249.21$.
- Step 2: **your step here** — compute $\mathrm{BIC}_B$.
- Step 3: compare and state the winner.

::: {.callout-note collapse="true" title="Answer"}
Step 2: $\mathrm{BIC}_B = -2(-118) + 5\log(100) = 236 + 5(4.605) = 236 + 23.03 = 259.03$.
Step 3: $\mathrm{BIC}_A = 249.21 < \mathrm{BIC}_B = 259.03$, so **BIC prefers Model A**.
Model B's better fit (log-likelihood up by 2, so misfit down by 4) didn't cover its 3
extra parameters' penalty ($3\log 100 \approx 13.8$).
:::

## Problem 3 — RSS falls but BIC picks smaller (in R)

Fit `y` on `x1` alone, then on `x1` plus two pure-noise predictors. Confirm RSS drops but
BIC rises, and state which model BIC selects.

```{webr}
#| autorun: false
# p3-starter — compare RSS and BIC across the three models
# PREDICT: RSS will fall each time; will BIC?
set.seed(3); n <- 40
x1 <- rnorm(n); y <- 1.5 * x1 + rnorm(n)
z1 <- rnorm(n); z2 <- rnorm(n)     # pure noise, unrelated to y
```

::: {.callout-note collapse="true" title="Answer"}
RSS falls as the noise predictors are added (they always can), but BIC is lowest for the
`x1`-only model — the penalty rejects the noise.

```{r}
set.seed(3); n <- 40
x1 <- rnorm(n); y <- 1.5 * x1 + rnorm(n); z1 <- rnorm(n); z2 <- rnorm(n)
m1 <- lm(y ~ x1); m2 <- lm(y ~ x1 + z1); m3 <- lm(y ~ x1 + z1 + z2)
rbind(RSS = c(sum(resid(m1)^2), sum(resid(m2)^2), sum(resid(m3)^2)),
      BIC = c(BIC(m1), BIC(m2), BIC(m3)))
```
RSS decreases left to right; BIC is smallest for `m1`. BIC selects `x1` alone.
:::

## Problem 4 — spot the error

A colleague compares five nested models and reports: *"Model 5 has the highest $R^2$
(0.94) of all five, so it's the best model — we should report it."* This contains
**exactly one mistake** — find it.

::: {.callout-note collapse="true" title="Answer"}
The mistake: using $R^2$ (equivalently, training RSS) to *choose among models of
different sizes*. $R^2$ can only rise as predictors are added, so the largest model
(Model 5) will essentially always have the highest $R^2$ — even if its extra predictors
are noise. To choose model size you need a criterion that penalizes complexity (AIC, BIC,
or adjusted $R^2$), or an out-of-sample estimate (cross-validation).
:::

## Problem 5 — mixing it with likelihood *(uses likelihood-and-log-likelihood's $\log L$)*

R reports `logLik(fit)` is $-45.2$ with `df = 4` (that `df` is $k$, the parameter count
including the noise variance), on $n = 60$ points. Reconstruct the BIC R would print, and
confirm with `BIC()`-style arithmetic.

```{webr}
#| autorun: false
# p5-starter
logL <- -45.2; k <- 4; n <- 60
```

::: {.callout-note collapse="true" title="Answer"}
$\mathrm{BIC} = -2\log L + k\log n = -2(-45.2) + 4\log(60) = 90.4 + 4(4.094) = 90.4 +
16.38 = 106.78$. R's `BIC()` uses exactly this: `-2*logLik + k*log(n)` with `k` the
`df` from `logLik` (parameters *including* the estimated variance).

```{r}
logL <- -45.2; k <- 4; n <- 60
-2 * logL + k * log(n)      # 106.78
```
:::

## Problem 6 — transfer: greedy can miss

You run forward stepwise on 8 predictors and it selects `{x1, x4, x6}`. A colleague runs
best-subset (all $2^8 = 256$ subsets) and the best size-3 model is `{x2, x4, x6}`. Is one
of them wrong? What does the disagreement tell you? *(The switching search in the next
course is built to reduce exactly this gap.)*

::: {.callout-note collapse="true" title="Answer"}
Neither is a bug. Forward stepwise is **greedy**: it added whichever single predictor
helped most first (`x1`), then could never remove it, so it can get stuck on a subset that
best-subset beats. The disagreement means the greedy path made an early commitment it
couldn't undo. A search that can also *swap* predictors in and out (the switching search)
can escape such traps and get closer to the best-subset answer without checking all $2^p$.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/model-selection-and-information-criteria" pass="4"}

::: {.quiz-q qid="msic-rss-cant-choose" qtype="mc" answer="a"}
Why can't the model with the lowest training RSS be crowned the best?

a) RSS always falls as predictors are added, even useless ones — so it always picks the biggest model
b) RSS is impossible to compute for large models
c) RSS ignores the training data
:::

::: {.quiz-q qid="msic-aic" qtype="numeric" answer="38" tolerance="0"}
AIC $= -2\log L + 2k$. For $-2\log L = 30$ and $k = 4$ parameters, what is the AIC?
:::

::: {.quiz-q qid="msic-bic-smaller" qtype="mc" answer="b"}
Compared with AIC, why does BIC ($-2\log L + k\log n$) tend to prefer smaller models?

a) BIC ignores the likelihood entirely
b) Its per-parameter penalty $\log n$ grows with sample size, so it penalizes extra parameters harder
c) BIC always picks the model with the most predictors
:::

::: {.quiz-q qid="msic-stepwise-greedy" qtype="mc" answer="c"}
Forward stepwise selection is greedy. What does that risk?

a) Nothing — a greedy search always finds the best subset
b) It must evaluate all $2^p$ subsets first
c) It commits to early picks it cannot undo, so it can miss the true best subset of a given size
:::

::: {.quiz-q qid="msic-loglik" qtype="mc" answer="a"}
(From likelihood & log-likelihood.) AIC and BIC both contain $\log L$. What is $L$?

a) The likelihood — how probable the observed data is under the fitted model
b) The number of predictors
c) The residual sum of squares
:::

:::

::::
