---
title: "Practice: Likelihood and gradient ascent"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

Work these in order — they get harder. Use paper or the live cells; only then open the
answer. Two helpers to paste as needed:

```r
sigmoid  <- function(z) 1 / (1 + exp(-z))
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) {
    up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
  })
```

## Problem 1 — the toy example, new numbers

Five students, hours $x = (1,2,3,4,5)$ and outcomes $y = (0,1,0,1,1)$. Compute the
log-likelihood at the starting guess $\beta = (0,0)$ **by hand** (every $p_i = 0.5$), then
fit by gradient ascent (lr $= 0.1$, 8000 steps) in R and report the fitted coefficients.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
sigmoid  <- function(z) 1 / (1 + exp(-z))
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) { up<-v; dn<-v; up[i]<-up[i]+h; dn[i]<-dn[i]-h; (f(up)-f(dn))/(2*h) })
x <- c(1, 2, 3, 4, 5); y <- c(0, 1, 0, 1, 1)
loglik <- function(b) { p <- sigmoid(b[1] + b[2]*x); sum(y*log(p) + (1-y)*log(1-p)) }
```

::: {.callout-note collapse="true" title="Answer"}
By hand at $\beta = (0,0)$: every $p_i = 0.5$, so each of the 5 terms is $\log(0.5)$, and
the log-likelihood is $5 \times \log(0.5) \approx -3.466$.

```{r p1-answer}
sigmoid  <- function(z) 1 / (1 + exp(-z))
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) { up<-v; dn<-v; up[i]<-up[i]+h; dn[i]<-dn[i]-h; (f(up)-f(dn))/(2*h) })
x <- c(1, 2, 3, 4, 5); y <- c(0, 1, 0, 1, 1)
loglik <- function(b) { p <- sigmoid(b[1] + b[2]*x); sum(y*log(p) + (1-y)*log(1-p)) }

loglik(c(0, 0))                        # -3.466 = 5 * log(0.5)
b <- c(0, 0); for (t in 1:8000) b <- b + 0.1 * gradient(loglik, b)
c(intercept = b[1], slope = b[2], loglik = loglik(b))   # ~ -2.649, 1.090, -2.422
```
Ascent climbs from $-3.466$ up to about $-2.422$ (less negative = more probable data), at
coefficients $\hat\beta_0 \approx -2.649$, $\hat\beta_1 \approx 1.090$.
:::

## Problem 2 — fill in the missing step

We're building the log-likelihood contribution of a single **failure** ($y_i = 0$) with
predicted probability $p_i$. Here's the work with the **core step blanked**:

> Step 1: the on/off switch $p_i^{\,y_i}(1 - p_i)^{1 - y_i}$ with $y_i = 0$ becomes
> $p_i^{0}(1 - p_i)^{1}$
> Step 2: **your step here** — simplify that expression
> Step 3: take the log → the contribution is $\log(1 - p_i)$

Supply Step 2.

```{webr}
#| autorun: false
# p2-starter — check the failure contribution for p = 0.3
p <- 0.3   # a failure: y = 0
```

::: {.callout-note collapse="true" title="Answer"}
Step 2: $p_i^{0} = 1$ (anything to the power 0 is 1) and $(1-p_i)^{1} = 1 - p_i$, so the
switch simplifies to $1 \times (1 - p_i) = 1 - p_i$. Taking the log then gives
$\log(1 - p_i)$.

```{r p2-answer}
p <- 0.3
c(switch = p^0 * (1 - p)^1, contribution_log = log(1 - p))   # 0.7, -0.357
```
A failure contributes $1 - p = 0.7$ to the likelihood (the model gave the non-event a
$0.7$ chance), i.e. $\log(0.7) \approx -0.357$ to the log-likelihood.
:::

## Problem 3 — which coefficients fit better?

For the lesson's data ($x = 1{:}6$, $y = 0,0,1,0,1,1$), a colleague can't decide between
two candidate coefficient pairs: $\beta_A = (-1, 0.5)$ and $\beta_B = (-4, 1)$. Compute
the log-likelihood of each and say which fits the data better, and why.

```{webr}
#| autorun: false
# p3-starter — compare the log-likelihood at two candidate coefficient pairs
sigmoid <- function(z) 1 / (1 + exp(-z))
x <- c(1, 2, 3, 4, 5, 6); y <- c(0, 0, 1, 0, 1, 1)
loglik <- function(b) { p <- sigmoid(b[1] + b[2]*x); sum(y*log(p) + (1-y)*log(1-p)) }
```

::: {.callout-note collapse="true" title="Answer"}
```{r p3-answer}
sigmoid <- function(z) 1 / (1 + exp(-z))
x <- c(1, 2, 3, 4, 5, 6); y <- c(0, 0, 1, 0, 1, 1)
loglik <- function(b) { p <- sigmoid(b[1] + b[2]*x); sum(y*log(p) + (1-y)*log(1-p)) }
c(beta_A = loglik(c(-1, 0.5)), beta_B = loglik(c(-4, 1)))   # -3.283 vs -2.622
```
$\beta_B = (-4, 1)$ fits **better**: its log-likelihood is $-2.622$, less negative than
$\beta_A$'s $-3.283$. "Less negative" means the observed 0/1 outcomes are *more probable*
under $\beta_B$. (Neither is the true maximum — that's about $-2.478$ at $(-4.249, 1.214)$
— but $\beta_B$ is closer to the summit.)
:::

## Problem 4 — spot the error

A colleague adapts their gradient-descent code to fit logistic regression, but the
log-likelihood *plummets* instead of climbing. Their code has **exactly one mistake** —
find it and fix it.

```{webr}
#| autorun: false
# p4-starter — run it, watch the log-likelihood crash, then fix the one mistake
sigmoid  <- function(z) 1 / (1 + exp(-z))
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) { up<-v; dn<-v; up[i]<-up[i]+h; dn[i]<-dn[i]-h; (f(up)-f(dn))/(2*h) })
x <- c(1, 2, 3, 4, 5, 6); y <- c(0, 0, 1, 0, 1, 1)
loglik <- function(b) { p <- sigmoid(b[1] + b[2]*x); sum(y*log(p) + (1-y)*log(1-p)) }

b <- c(0, 0)
for (t in 1:30) b <- b - 0.1 * gradient(loglik, b)   # log-likelihood goes the wrong way
c(b = b, loglik = loglik(b))
```

::: {.callout-note collapse="true" title="Answer"}
The mistake is the **sign**: they wrote `b - 0.1 * gradient(...)`, which is gradient
*descent* — it steps *downhill*, minimizing the log-likelihood (driving it toward
$-\infty$). We want to *maximize* it, so we step *uphill* with a **plus**:

```{r p4-answer}
sigmoid  <- function(z) 1 / (1 + exp(-z))
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) { up<-v; dn<-v; up[i]<-up[i]+h; dn[i]<-dn[i]-h; (f(up)-f(dn))/(2*h) })
x <- c(1, 2, 3, 4, 5, 6); y <- c(0, 0, 1, 0, 1, 1)
loglik <- function(b) { p <- sigmoid(b[1] + b[2]*x); sum(y*log(p) + (1-y)*log(1-p)) }

b <- c(0, 0)
for (t in 1:30) b <- b + 0.1 * gradient(loglik, b)   # PLUS: climb uphill
c(intercept = b[1], slope = b[2], loglik = loglik(b))   # climbing: loglik ~ -2.912 at 30 steps
```
With the minus sign, after 30 steps the log-likelihood was down around $-593$ and falling;
with the plus sign it *rises* toward the summit (about $-2.912$ at 30 steps, heading for
$-2.478$). Ascent maximizes, so it adds the gradient.
:::

## Problem 5 — mixing it with gradient descent *(uses the gradient-descent course's update rule)*

Do **one** ascent step by hand-and-check. On the lesson data, starting at $\beta = (0,0)$,
the gradient of the log-likelihood is $(0, 3.5)$ (worked out in the lesson). Using the
**ascent update rule** with learning rate $\alpha = 0.1$, what is $\beta$ after one step?
Confirm by computing the gradient and stepping in R.

```{webr}
#| autorun: false
# p5-starter — take exactly ONE ascent step from (0,0)
sigmoid  <- function(z) 1 / (1 + exp(-z))
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) { up<-v; dn<-v; up[i]<-up[i]+h; dn[i]<-dn[i]-h; (f(up)-f(dn))/(2*h) })
x <- c(1, 2, 3, 4, 5, 6); y <- c(0, 0, 1, 0, 1, 1)
loglik <- function(b) { p <- sigmoid(b[1] + b[2]*x); sum(y*log(p) + (1-y)*log(1-p)) }
```

::: {.callout-note collapse="true" title="Answer"}
The ascent rule is $\beta \leftarrow \beta + \alpha \nabla\ell$. With $\nabla\ell = (0,
3.5)$ and $\alpha = 0.1$: $\beta = (0,0) + 0.1(0, 3.5) = (0,\ 0.35)$.

```{r p5-answer}
sigmoid  <- function(z) 1 / (1 + exp(-z))
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) { up<-v; dn<-v; up[i]<-up[i]+h; dn[i]<-dn[i]-h; (f(up)-f(dn))/(2*h) })
x <- c(1, 2, 3, 4, 5, 6); y <- c(0, 0, 1, 0, 1, 1)
loglik <- function(b) { p <- sigmoid(b[1] + b[2]*x); sum(y*log(p) + (1-y)*log(1-p)) }

g <- gradient(loglik, c(0, 0))       # (0, 3.5)
c(0, 0) + 0.1 * g                    # (0, 0.35)
```
The intercept doesn't move (its slope was 0 — three passes and three fails balanced), and
the slope rises to $0.35$. This is the **same update rule** from the
[gradient-descent course](../../../gradient-descent/syllabus.md), with the sign flipped to
$+$ so we climb the log-likelihood instead of descending an error bowl.
:::

## Problem 6 — a gentle cliffhanger

You've fit the coefficients from scratch. In practice you'd call `glm(y ~ x, family =
binomial)` — and its `summary()` prints not just the coefficients but a whole table:
**standard errors**, **z values**, **p-values**, and a quantity called **deviance**. Here
is the coefficient table for the lesson's data. Read off the two fitted coefficients (you
already know these!), and guess: what do you think the "Std. Error" column is telling you
about each coefficient?

```{r p6-artifact}
# A real glm fit -- the coefficient table you'll fully decode in the capstone
x <- c(1, 2, 3, 4, 5, 6); y <- c(0, 0, 1, 0, 1, 1)
summary(glm(y ~ x, family = binomial))$coefficients
```

::: {.callout-note collapse="true" title="Answer"}
The `Estimate` column holds the fitted coefficients you found by ascent: intercept
$\approx -4.249$ and slope $\approx 1.214$ — exactly your from-scratch answer. The
`Std. Error` column is a measure of **uncertainty** in each estimate: roughly, how much
the coefficient would wobble if you refit on fresh data from the same process. A large
standard error (as here, with only 6 noisy points) means the estimate is imprecise. You
can't fully unpack the z value, p-value, and deviance yet — the
[capstone](../04-capstone-reading-a-real-logistic-regression/lesson.qmd) decodes every one
of them and ties each back to what you built.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="courses/logistic-regression/modules/03-likelihood-and-gradient-ascent" pass="4"}

::: {.quiz-q qid="mle-loglik-zero" qtype="numeric" answer="-4.159" tolerance="0.0005"}
For the lesson's six-student data ($x = 1{:}6$, $y = 0,0,1,0,1,1$) at the starting
guess $\beta = (0,0)$, every predicted probability is $0.5$. What is the
log-likelihood? Round to three decimals.
:::

::: {.quiz-q qid="mle-why-log" qtype="mc" answer="c"}
Why does fitting logistic regression use the **log**-likelihood rather than the
likelihood directly?

a) The log converts probabilities to log-odds, reconnecting the optimizer to the logit link
b) The log-likelihood can be positive, which makes it easier to tell when it is maximized
c) Taking log turns a product of sub-1 probabilities into a sum, preventing underflow
:::

::: {.quiz-q qid="mle-ascent-step" qtype="mc" answer="b"}
The gradient of the log-likelihood at $\beta = (0,0)$ is $(0,\ 3.5)$. Using gradient
ascent with learning rate $\alpha = 0.1$, what is $\beta$ after one step?

a) $(0,\ -0.35)$ — stepping downhill by subtracting the scaled gradient
b) $(0,\ 0.35)$ — stepping uphill by adding the scaled gradient
c) $(0,\ 3.5)$ — using the gradient vector itself as the new parameter
:::

::: {.quiz-q qid="mle-sign-convention" qtype="mc" answer="a"}
After gradient ascent on the lesson's data, the log-likelihood rises to about
$-2.478$ from its starting value at $\beta = (0,0)$. What does this change show?

a) The fitted coefficients make the observed outcomes more probable
b) The model worsened — more negative log-likelihood values are preferred
c) The likelihood itself doubled, from roughly $-4$ to $-2$
:::

::: {.quiz-q qid="mle-sigmoid-recall" qtype="numeric" answer="0.269" tolerance="0.0005"}
(From module 2, the logit link.) For the lesson's loan model ($\beta_0 = -2$,
$\beta_1 = 1$), the log-odds at $x = 1$ is $-1$. What is the predicted repayment
probability? Round to three decimals.
:::

:::

::::
