---
title: "k-nearest neighbors"
subtitle: "Predict by asking the closest data points — no model required"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

## Where we're going

You know how to measure how far apart two data points are. Now: use that distance
to make predictions. Find the $k$ closest labeled data points to a new unlabeled
point, and let them vote. That is the entire algorithm --- no model to fit, no
parameters to estimate. The data IS the model.

**Builds on:**
[distance, similarity & geometry](../distance-similarity-and-geometry/lesson.qmd)
(Euclidean distance, scaling effects),
[classification evaluation](../classification-evaluation/lesson.qmd)
(soft --- for confusion matrix, sensitivity/specificity),
[model fit, error & validation](../model-fit-error-and-validation/lesson.qmd)
(soft --- for overfitting, train/test split)

## Warm-up: do you remember?

A few quick reps --- answers from memory first, then check.

**1.** Compute the Euclidean distance between the points $(1, 3)$ and $(4, 7)$.

```{webr}
#| autorun: false
# warmup1-starter — edit and Run to check
p <- c(1, 3)
q <- c(4, 7)
```

::: {.callout-note collapse="true" title="Answer"}
Gaps: $4-1=3$ and $7-3=4$. Square: $9$ and $16$. Add: $25$. Square-root:
$\sqrt{25} = 5$. In R: `sqrt(sum((p - q)^2))` gives 5. --- taught in
[distance, similarity & geometry](../distance-similarity-and-geometry/lesson.qmd);
revisit if shaky.
:::

**2.** What happens to distances when one column is measured in thousands and
another in single digits?

::: {.callout-note collapse="true" title="Answer"}
The large-scale column dominates every distance computation. A 1-unit change in
the big column moves the distance far more than a 1-unit change in the small
column, regardless of which variable actually matters more. --- taught in
[distance, similarity & geometry](../distance-similarity-and-geometry/lesson.qmd),
"Closer = more similar" section; revisit if shaky.
:::

**3.** In one sentence: why is a perfect training score suspicious?

::: {.callout-note collapse="true" title="Answer"}
It usually means the model has memorized the training data (overfitting) and will
perform poorly on new data it has never seen. --- taught in
[model fit, error & validation](../model-fit-error-and-validation/lesson.qmd);
revisit if shaky.
:::

## A tiny example first

Six labeled points in 2D, each belonging to class A or class B. One new point ---
the **query** --- has no label yet. Our job: classify it.

```{r toy-data}
# The labeled training data: 6 points, each with an (x, y) position and a class label
train_x <- c(1, 2, 3, 6, 7, 8)
train_y <- c(2, 3, 1, 7, 8, 6)
label   <- c("A", "A", "A", "B", "B", "B")

# The new point we want to classify
query <- c(4, 4)
```

**Step 1 --- compute the distance from the query to every training point.**

Point (1, 2): gaps are $4-1=3$ and $4-2=2$, so $\sqrt{3^2+2^2} = \sqrt{9+4} = \sqrt{13} \approx 3.61$.

Point (2, 3): gaps are $4-2=2$ and $4-3=1$, so $\sqrt{4+1} = \sqrt{5} \approx 2.24$.

Point (3, 1): gaps are $4-3=1$ and $4-1=3$, so $\sqrt{1+9} = \sqrt{10} \approx 3.16$.

Point (6, 7): gaps are $4-6=-2$ and $4-7=-3$, so $\sqrt{4+9} = \sqrt{13} \approx 3.61$.

Point (7, 8): gaps are $4-7=-3$ and $4-8=-4$, so $\sqrt{9+16} = \sqrt{25} = 5.00$.

Point (8, 6): gaps are $4-8=-4$ and $4-6=-2$, so $\sqrt{16+4} = \sqrt{20} \approx 4.47$.

**Step 2 --- sort by distance (nearest first).**

| Rank | Point | Distance | Label |
|------|-------|----------|-------|
| 1st  | (2, 3) | $\sqrt{5} \approx 2.24$  | A |
| 2nd  | (3, 1) | $\sqrt{10} \approx 3.16$ | A |
| 3rd  | (1, 2) | $\sqrt{13} \approx 3.61$ | A |
| 4th  | (6, 7) | $\sqrt{13} \approx 3.61$ | B |
| 5th  | (8, 6) | $\sqrt{20} \approx 4.47$ | B |
| 6th  | (7, 8) | $\sqrt{25} = 5.00$       | B |

(Notice that points (1, 2) and (6, 7) are tied at $\sqrt{13}$. Tie-breaking rules
vary by implementation, but here it does not matter --- the classification comes
out the same either way.)

**Step 3 --- for k=3, take the top 3 and vote.**

The 3 nearest neighbors are (2, 3), (3, 1), and (1, 2) --- all labeled **A**. The
majority vote is A (3 votes to 0), so we classify the query as **A**.

Even if the tie between the 3rd and 4th closest had broken the other way ---
giving us (6, 7) as the 3rd neighbor instead of (1, 2) --- the vote would be 2
A's and 1 B: still **A**.

The same computation in R:

```{r toy-classify}
# Compute distance from the query to each training point
dists <- sqrt((train_x - query[1])^2 + (train_y - query[2])^2)
round(dists, 2)    # 3.61  2.24  3.16  3.61  5.00  4.47

# Sort by distance: order() returns the indices from nearest to farthest
ranked <- order(dists)
ranked              # 2 3 1 4 6 5  (index of nearest first)

# The k=3 nearest neighbors and their labels
k <- 3
neighbors <- ranked[1:k]
label[neighbors]    # "A" "A" "A"

# The majority vote: whichever label appears most
vote <- names(which.max(table(label[neighbors])))
vote                # "A"
```

R also ships this exact procedure as a function. `class::knn()` is a
base-recommended package (it comes with every R installation, no `install.packages()`
needed) that does the same distance-sort-vote we just did by hand:

```{r toy-classify-builtin}
train_mat <- cbind(train_x, train_y)   # knn() wants training features as a matrix
query_mat <- matrix(query, nrow = 1)   # the query as a 1-row matrix

class::knn(train_mat, query_mat, cl = label, k = 3)   # "A"
```

Same answer as the hand vote: A. `class::knn()` is not a different algorithm ---
it is exactly the sort-by-distance-and-vote you just did, vectorized and packaged.

## Getting a feel for it

**The restaurant-recommendation picture.** Imagine you move to a new city and want
a restaurant recommendation. You do not survey everyone in the city --- you ask
your 3 nearest neighbors. If 2 out of 3 say Italian, you go Italian. That is kNN:
trust the people closest to you, ignore the rest. (Where the analogy breaks:
real neighbors do not all live at measured distances from you, and their opinions
might be swayed by things distance cannot capture.)

**Why this construction --- the lazy learner.** Most models do work up front: they
fit a line, estimate coefficients, search for a split. kNN does nothing at
training time --- it stores the labeled data and waits. All the work happens at
prediction time, when someone hands it a new point and it searches for the
nearest neighbors. This is called a **lazy learner**: it does no work until
someone asks a question. The training data is not a stepping stone to a model ---
it IS the model.

**Why k matters --- flexibility vs. smoothness.** As $k$ shrinks, the decision
boundary gets more flexible and irregular (low bias, high variance); as $k$ grows,
it gets smoother and closer to a straight line (high bias, low variance)
[@james2021introductionstatisticallearning, pp. 41--42].

- **k = 1** trusts a single neighbor. It is maximally flexible: the prediction
  changes every time the nearest point changes. But it is noisy --- one outlier
  can flip the answer. On our toy data, k=1 for the query at (4, 4) gives point
  (2, 3) with label A. If that one training point were mislabeled, the
  prediction would be wrong.
- **k = 6** (all training points) averages everyone. It always predicts the
  overall majority class --- here, a tie between 3 A's and 3 B's. With 5
  labeled points (say 3 A, 2 B), k=5 would always predict A regardless of where
  the query sits. That is too smooth: it ignores local structure entirely.
- **k = 3** is a middle ground: local enough to capture the structure (the A
  cluster is in the lower-left, the B cluster is in the upper-right), but
  averaging over enough neighbors to smooth out noise.

**What the number feels like.** Our k=3 vote was 3-vs-0 for class A --- a
unanimous decision because the query at (4, 4) sits closer to the A cluster
than to the B cluster. What if the query were at (5, 5) instead? The nearest
neighbors change: the 3 closest become (6, 7), (8, 6), and either (2, 3) or
(7, 8) --- now the majority is B. Moving the query two units shifts the
prediction entirely, which is exactly the local sensitivity that makes kNN
useful for data with well-separated groups.

## The curse of dimensionality

In 2D, "nearest" means something --- some points genuinely sit close and others
sit far. But in high dimensions, a strange thing happens: **all points become
roughly the same distance from each other.** When that happens, the "nearest"
neighbor is barely nearer than the farthest one, and the whole concept of
neighborhood breaks down.

Here is a quick simulation. Generate 50 random points, and compute the ratio of
the maximum pairwise distance to the minimum pairwise distance. In low dimensions,
the farthest pair is many times farther than the closest pair; in high dimensions,
that ratio collapses toward 1.

```{r curse-sim}
# Generate 50 random points in 2D, 50D, and 500D.
# For each, compute max-distance / min-distance.
set.seed(42)
n_points <- 50

ratio_for_dim <- function(d) {
  pts <- matrix(rnorm(n_points * d), nrow = n_points, ncol = d)
  all_dists <- as.numeric(dist(pts))  # all pairwise Euclidean distances
  max(all_dists) / min(all_dists)     # how spread out are the distances?
}

dims <- c(2, 10, 50, 200, 500)
ratios <- sapply(dims, ratio_for_dim)
data.frame(dimensions = dims, max_over_min = round(ratios, 2))
```

In 2D, the farthest pair is many times farther than the closest --- distances are
spread out, and neighbors mean something. By 500D, the ratio is close to small
single digits: every point is about the same distance from every other point. When
all points are equally far away, the "nearest" neighbor is not really near --- the
concept breaks down. This is the **curse of dimensionality**.

## What the notation is going to say

The formula will say: given a new point, find the $k$ training points closest to
it (the neighborhood), then take the most common label among those $k$ points as
your prediction. For regression instead of classification, take the average of
those $k$ neighbors' values. In our toy example, the neighborhood of size 3
around (4, 4) contained three A's, so the prediction is A --- exactly the
"sort-by-distance, take the top $k$, vote" procedure we just did by hand.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $k$ | "kay" | how many nearest neighbors to consult (a choice you make) | $3$ |
| $\mathbf{x}$ (bold) | "the vector ex" | a data point --- its coordinates as a vector (in R: a numeric vector) | the query $(4, 4)$ |
| $\mathbf{x}_i$ | "ex sub eye" | the $i$-th training point (in R: the $i$-th row of the training data) | $\mathbf{x}_2 = (2, 3)$ is the nearest |
| $y_i$ | "why sub eye" | the label (or value) of training point $i$ | $y_2 = \text{A}$ |
| $d(\mathbf{x}, \mathbf{x}_i)$ | "d of ex and ex-sub-eye" | the distance from point $\mathbf{x}$ to training point $\mathbf{x}_i$ (in R: `sqrt(sum((x - x_i)^2))`) | $d(\text{query}, \mathbf{x}_2) = \sqrt{5} \approx 2.24$ |
| $N_k(\mathbf{x})$ | "the k-neighborhood of ex" / "en-sub-kay of ex" | the set of indices of the $k$ nearest training points to $\mathbf{x}$ | $N_3(\text{query}) = \{2, 3, 1\}$ (the three closest) |
| $\hat{y}$ | "why hat" | the predicted label or value --- the hat means "estimated from data" | $\hat{y} = \text{A}$ |
| $n$ | "en" | the number of data points (in R: `length(x)`) | $6$ training points |
| $\in$ | "is in" / "is an element of" | belongs to a set: $i \in N_k(\mathbf{x})$ means $i$ is one of the neighbor indices | $2 \in \{2, 3, 1\}$ |
| $\sum_{i \in N_k(\mathbf{x})}$ | "the sum over i in the k-neighborhood of x" | add up one term for each neighbor index (in R: `sum(...)`) | add the 3 nearest neighbors' values |

## The formal version

**Classification --- majority vote.**

$$
\hat{y} = \text{mode of } \{y_i : i \in N_k(\mathbf{x})\}
$$

Reading it aloud: "y-hat equals the mode of the set of y-sub-i for all i in
the k-neighborhood of x" --- or in everyday words: **"the prediction is
whichever label appears most often among the k nearest neighbors."**

Mapping onto the toy example: $N_3(\text{query}) = \{2, 3, 1\}$, so we
collect $\{y_2, y_3, y_1\} = \{\text{A}, \text{A}, \text{A}\}$. The mode
(the value that appears most) is A. So $\hat{y} = \text{A}$ --- exactly what
we got by hand. This is exactly the kNN classifier as defined in ISL: find the
$k$ closest training points and estimate the class probability as the fraction of
neighbors voting for it, then pick the class with the largest share
[@james2021introductionstatisticallearning, p. 39].

**Regression --- average the neighbors.**

$$
\hat{y} = \frac{1}{k}\sum_{i \in N_k(\mathbf{x})} y_i
$$

Reading it aloud: "y-hat equals one over k times the sum, for each i in the
k-neighborhood of x, of y-sub-i" --- or in everyday words: **"the prediction
is the average of the k nearest neighbors' values."**

This version applies when the target is a number (like predicting someone's
test score) rather than a class label. The arithmetic is the same --- find the
k closest, then average their values instead of voting.

$N_k(\mathbf{x})$ in both formulas is the set of indices of the $k$ nearest
training points to $\mathbf{x}$, measured by $d(\mathbf{x}, \mathbf{x}_i)$.

## Explore it

```{webr}
#| autorun: true
#| fig-alt: "Six training points in 2D colored by label (A in blue, B in orange) with a query point shown as a red triangle. A dashed circle centered on the query encloses the k=3 nearest neighbors."
# explore-1 — LIVE: plot the training data, the query, and the k-nearest circle.
# TRY IT: change query_x and query_y to move the query point — watch which
#   neighbors fall inside the circle.
# PREDICT: if you move the query to (7, 7), right in the B cluster, what class
#   will k=3 predict? Commit your answer, then change query_x and query_y and run.

train_x <- c(1, 2, 3, 6, 7, 8)
train_y <- c(2, 3, 1, 7, 8, 6)
label   <- c("A", "A", "A", "B", "B", "B")
query_x <- 4
query_y <- 4
k <- 3

# Compute distances and find the k nearest
dists  <- sqrt((train_x - query_x)^2 + (train_y - query_y)^2)
ranked <- order(dists)
nbrs   <- ranked[1:k]
radius <- max(dists[nbrs])  # distance to the k-th nearest neighbor

# Majority vote
vote <- names(which.max(table(label[nbrs])))

# Plot
cols <- ifelse(label == "A", "steelblue", "darkorange")
plot(train_x, train_y, pch = 16, cex = 2, col = cols,
     xlim = c(-1, 10), ylim = c(-1, 10),
     xlab = "x", ylab = "y",
     main = sprintf("k=%d vote: %s (neighbors: %s)",
                    k, vote,
                    paste(label[nbrs], collapse = ", ")))
points(query_x, query_y, pch = 17, cex = 2.5, col = "red")
text(query_x, query_y, "query", pos = 1, col = "red")
text(train_x, train_y, label, pos = 3, cex = 0.9)

# Draw a circle around the k nearest neighbors
theta <- seq(0, 2 * pi, length.out = 100)
lines(query_x + radius * cos(theta),
      query_y + radius * sin(theta),
      lty = 2, col = "grey40")
```

```{webr}
#| autorun: false
# explore-2 — LIVE: what happens when you change k?
# PREDICT: Set k=1. Which single point is nearest to (4,4)? What label does it have?
# Then set k=6. With ALL 6 points voting (3 A's and 3 B's), what happens?
# Commit to your answers, then change k below and run.

train_x <- c(1, 2, 3, 6, 7, 8)
train_y <- c(2, 3, 1, 7, 8, 6)
label   <- c("A", "A", "A", "B", "B", "B")
query   <- c(4, 4)
k <- 3   # TRY IT: change to 1, then to 6

dists  <- sqrt((train_x - query[1])^2 + (train_y - query[2])^2)
ranked <- order(dists)
nbrs   <- ranked[1:k]

cat("k =", k, "\n")
cat("Neighbors:", paste(sprintf("(%d,%d) %s dist=%.2f",
    train_x[nbrs], train_y[nbrs], label[nbrs], dists[nbrs]),
    collapse = "\n           "), "\n")
cat("Vote:", names(which.max(table(label[nbrs]))), "\n")
```

```{webr}
#| autorun: true
#| fig-alt: "Line plot showing the ratio of maximum to minimum pairwise distance on the y-axis against the number of dimensions on the x-axis. The ratio is large in low dimensions and shrinks toward small values in high dimensions, illustrating the curse of dimensionality."
# explore-3 — LIVE: the curse of dimensionality.
# The baked simulation above already showed the max/min distance ratio at
# 2, 10, 50, 200, and 500 dimensions.
# PREDICT: at 25 dimensions — a value NOT shown above — will the ratio be
# closer to the 10-dimension value or the 50-dimension value? Write down a
# specific number, then run and check.

set.seed(42)
n_points <- 50

ratio_for_dim <- function(d) {
  pts <- matrix(rnorm(n_points * d), nrow = n_points, ncol = d)
  all_dists <- as.numeric(dist(pts))
  max(all_dists) / min(all_dists)
}

dims   <- c(2, 5, 10, 25, 50, 100, 200, 500)
ratios <- sapply(dims, ratio_for_dim)

cat("Ratio at 25 dimensions:", round(ratios[dims == 25], 2), "\n")

plot(dims, ratios, type = "b", pch = 16, col = "steelblue",
     xlab = "number of dimensions",
     ylab = "max distance / min distance",
     main = "the curse: distances converge in high dimensions")
abline(h = 1, lty = 2, col = "grey50")
```

## Common traps

**Trap 1 --- "More neighbors is always better."** It is tempting to set $k$ high
for a smoother, more "stable" prediction. But k=6 on our toy data asks all 6
training points to vote: 3 A's and 3 B's, a perfect tie. With an odd number of
training points --- say 5, with 3 A's and 2 B's --- k=5 would always predict A,
no matter where the query sits. That is too smooth: it ignores local structure
and just predicts the overall majority class. In effect, a large $k$ throws away
the whole point of looking at neighbors --- it pushes the model toward the
high-bias, low-variance end of the flexibility trade-off
[@james2021introductionstatisticallearning, p. 42]. On our data:

```{r trap1-demo}
train_x <- c(1, 2, 3, 6, 7, 8)
train_y <- c(2, 3, 1, 7, 8, 6)
label   <- c("A", "A", "A", "B", "B", "B")
query   <- c(4, 4)

dists  <- sqrt((train_x - query[1])^2 + (train_y - query[2])^2)
ranked <- order(dists)

# k=3: local vote — the 3 nearest are all A
k3_vote <- names(which.max(table(label[ranked[1:3]])))
k3_vote   # "A"

# k=6: global vote — everyone votes, 3 A's and 3 B's = tie
table(label[ranked[1:6]])   # A: 3, B: 3 — a tie, not a useful prediction
```

**Trap 2 --- "kNN doesn't need feature scaling."** It absolutely does. Recording
the very same real-world difference in a finer unit inflates its number, and
distance takes your columns at face value --- a column with big numbers
dominates, exactly as in the
[distance module](../distance-similarity-and-geometry/lesson.qmd). Scale your
features before running kNN.

```{r trap2-demo}
# Two people who differ by 20 kg in weight and 5 cm in height
person_a <- c(weight_kg = 60, height_cm = 170)
person_b <- c(weight_kg = 80, height_cm = 175)

# The SAME two people, the SAME real difference — just recording height in
# millimeters instead of centimeters
person_a_mm <- c(weight_kg = 60, height_mm = 1700)
person_b_mm <- c(weight_kg = 80, height_mm = 1750)

dist_cm <- sqrt(sum((person_a - person_b)^2))        # height measured in cm
dist_mm <- sqrt(sum((person_a_mm - person_b_mm)^2))  # height measured in mm

c(distance_height_in_cm = round(dist_cm, 2), distance_height_in_mm = round(dist_mm, 2))
# 20.62 vs 53.85 — switching height's unit from cm to mm changed NOTHING about
# these two people, but it more than doubled the total distance. The height
# term alone grew from 5 (cm) to 50 (mm) — now bigger than the 20 kg weight
# difference it should never have outweighed. Without scaling, kNN would let
# an arbitrary choice of unit decide which feature matters most.
```

## Check yourself

**1.** Using our toy data, what are the 2 nearest neighbors (k=2) of the query
point (4, 4)? What class would k=2 predict?

```{webr}
#| autorun: false
# check1-starter — edit and Run to check your hand answer
train_x <- c(1, 2, 3, 6, 7, 8)
train_y <- c(2, 3, 1, 7, 8, 6)
label   <- c("A", "A", "A", "B", "B", "B")
query   <- c(4, 4)
```

::: {.callout-note collapse="true" title="Answer"}
The 2 nearest are (2, 3) at distance $\sqrt{5} \approx 2.24$ (label A) and
(3, 1) at distance $\sqrt{10} \approx 3.16$ (label A). Both are A, so the
majority vote is **A** --- unanimous. In R: `label[order(dists)[1:2]]` gives
`"A" "A"`.
:::

**2.** If a dataset has 1000 features and 50 observations, is kNN likely to work
well? Why or why not?

::: {.callout-note collapse="true" title="Answer"}
Probably not. With 1000 dimensions and only 50 points, the curse of
dimensionality kicks in: all 50 points will be roughly the same distance from
each other, so the "nearest" neighbor is barely nearer than the farthest. The
concept of neighborhood breaks down, and the majority vote becomes nearly
random with respect to local structure.
:::

**3.** Explain to a colleague, in two sentences, what kNN does and when it is a
poor choice.

::: {.callout-note collapse="true" title="A model answer"}
kNN classifies a new point by finding the k closest labeled training points and
letting them vote --- whichever label wins the majority is the prediction. It
struggles when the data has many features relative to the number of observations
(the curse of dimensionality makes all distances similar) or when features are
on wildly different scales and have not been standardized.
:::

## What you can now do

- **Classify a new point with kNN by hand** --- compute distances, sort, take the
  top $k$, and hold a majority vote.
- **Explain what $k$ controls** --- small $k$ is flexible but noisy (sensitive to
  outliers), large $k$ is smooth but can ignore local structure entirely.
- **Recognize the curse of dimensionality** --- in high dimensions, all distances
  converge, and the "nearest" neighbor stops being meaningfully near.
- **Know that kNN requires feature scaling** --- unscaled columns with big numbers
  drown out columns with small numbers, exactly as in the distance module.

If any of these feels shaky, revisit [A tiny example first](#a-tiny-example-first)
and [Getting a feel for it](#getting-a-feel-for-it) before continuing --- the
clustering module assumes all of them.

## Where this goes next

kNN finds the neighbors of one specific query point. Clustering (e.g.,
[k-means](../clustering-k-means-and-hierarchical/lesson.qmd)) takes the next
step: group ALL points into clusters without any labels at all --- unsupervised
rather than supervised. The distance, scaling, and dimensionality intuitions you
built here carry directly into that world.

Ready for more reps? Do the [practice problems](practice.qmd).
