---
title: "Practice: k-nearest neighbors"
subtitle: "Answers are hidden --- try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order --- they get harder. Do them on paper or in an R session; only
then open the answer.

## Problem 1 --- the toy example, new numbers

Six points in 2D with class labels X and Y. Classify the query point using k=3.

```{webr}
#| autorun: false
# p1-starter — compute distances, sort, take top 3, vote
train_x <- c(1, 3, 4, 7, 9, 10)
train_y <- c(2, 1, 3, 8, 9, 7)
label   <- c("X", "X", "X", "Y", "Y", "Y")
query   <- c(5, 5)
k <- 3
```

**By hand:** compute the distance from (5, 5) to each training point, sort by
distance, take the 3 nearest, and hold the majority vote. What class does k=3
predict?

::: {.callout-note collapse="true" title="Answer"}
Distances from (5, 5):

- (1, 2): $\sqrt{16+9} = \sqrt{25} = 5.00$ --- X
- (3, 1): $\sqrt{4+16} = \sqrt{20} \approx 4.47$ --- X
- (4, 3): $\sqrt{1+4} = \sqrt{5} \approx 2.24$ --- X
- (7, 8): $\sqrt{4+9} = \sqrt{13} \approx 3.61$ --- Y
- (9, 9): $\sqrt{16+16} = \sqrt{32} \approx 5.66$ --- Y
- (10, 7): $\sqrt{25+4} = \sqrt{29} \approx 5.39$ --- Y

Sorted: (4, 3) at $\sqrt{5}$, (7, 8) at $\sqrt{13}$, (3, 1) at $\sqrt{20}$,
(1, 2) at 5.00, (10, 7) at $\sqrt{29}$, (9, 9) at $\sqrt{32}$.

The 3 nearest are (4, 3) X, (7, 8) Y, (3, 1) X. Vote: 2 X vs 1 Y = **X**.

```{r p1-answer}
train_x <- c(1, 3, 4, 7, 9, 10)
train_y <- c(2, 1, 3, 8, 9, 7)
label   <- c("X", "X", "X", "Y", "Y", "Y")
query   <- c(5, 5)

dists <- sqrt((train_x - query[1])^2 + (train_y - query[2])^2)
round(dists, 2)   # 5.00  4.47  2.24  3.61  5.66  5.39

ranked <- order(dists)
k <- 3
nbrs <- ranked[1:k]
label[nbrs]   # "X" "Y" "X"

names(which.max(table(label[nbrs])))   # "X"
```
:::

## Problem 2 --- fill in the missing step

Here is a worked kNN classification with two steps blanked out. Fill them in.

Using the same data as Problem 1 with query = (5, 5) and k = 3:

**Step 1 (shown):** Compute the Euclidean distance from the query to each
training point.

| Point   | Distance |
|---------|----------|
| (1, 2)  | 5.00     |
| (3, 1)  | 4.47     |
| (4, 3)  | 2.24     |
| (7, 8)  | 3.61     |
| (9, 9)  | 5.66     |
| (10, 7) | 5.39     |

**Step 2: your step here** --- What are the indices (or points) of the k=3
nearest neighbors, listed from closest to farthest?

**Step 3: your step here** --- Given those 3 neighbors and their labels, what
is the majority vote? What class does kNN predict?

```{webr}
#| autorun: false
# p2-starter — use this to check your blanked steps
train_x <- c(1, 3, 4, 7, 9, 10)
train_y <- c(2, 1, 3, 8, 9, 7)
label   <- c("X", "X", "X", "Y", "Y", "Y")
query   <- c(5, 5)
```

::: {.callout-note collapse="true" title="Answer"}
**Step 2:** Sort by distance. The 3 nearest are: (4, 3) at distance 2.24,
(7, 8) at 3.61, (3, 1) at 4.47.

**Step 3:** Their labels are X, Y, X. The majority vote is 2 X vs 1 Y, so
kNN predicts **X**.

The core move is the **sort-then-take-top-k** step --- this is where you turn a
list of distances into a neighborhood, and it is the step that makes $k$ matter.
:::

## Problem 3 --- implement kNN from scratch

Write a function `my_knn(train_x, train_y, train_label, query, k)` that:

1. Computes the Euclidean distance from `query` to each training point
2. Finds the `k` nearest neighbors
3. Returns the majority-vote label

Test it on the lesson's original toy data with query = (4, 4) and k = 3.

```{webr}
#| autorun: false
# p3-starter — implement your own knn function and test it
train_x <- c(1, 2, 3, 6, 7, 8)
train_y <- c(2, 3, 1, 7, 8, 6)
train_label <- c("A", "A", "A", "B", "B", "B")
query   <- c(4, 4)
k <- 3

# Your function here:
# my_knn <- function(train_x, train_y, train_label, query, k) {
#   ...
# }
```

::: {.callout-note collapse="true" title="Answer"}
```{r p3-answer}
# From-scratch kNN
my_knn <- function(train_x, train_y, train_label, query, k) {
  dists  <- sqrt((train_x - query[1])^2 + (train_y - query[2])^2)  # Euclidean distances
  ranked <- order(dists)                  # sort indices by distance (nearest first)
  nbrs   <- ranked[1:k]                  # take the k closest indices
  names(which.max(table(train_label[nbrs])))  # majority vote
}

# Test on the lesson's toy data
train_x <- c(1, 2, 3, 6, 7, 8)
train_y <- c(2, 3, 1, 7, 8, 6)
train_label <- c("A", "A", "A", "B", "B", "B")
query   <- c(4, 4)

my_result <- my_knn(train_x, train_y, train_label, query, k = 3)
my_result   # "A"

# Compare with class::knn() — class is a base-recommended package
train_mat <- cbind(train_x, train_y)   # knn() wants a matrix of training features
query_mat <- matrix(query, nrow = 1)   # query as a 1-row matrix

library_result <- class::knn(train_mat, query_mat, cl = train_label, k = 3)
library_result   # "A"

# They agree
c(from_scratch = my_result, class_knn = as.character(library_result))
```
:::

## Problem 4 --- spot the error

A colleague classifies loan applicants as high-risk or low-risk using kNN with
two features: age in years and income in dollars. Here is their distance
computation, and their report: "The predictions seem to depend only on income ---
age makes no difference at all."

```{r p4-code}
# The colleague's code — three applicants, distances fed straight to kNN
age    <- c(25, 40, 60)                 # years
income <- c(30000, 45000, 120000)       # dollars

applicants <- cbind(age, income)
dist(applicants)    # the pairwise distances their kNN uses
```

The code contains **exactly one mistake**. Find it and explain what to fix.

::: {.callout-note collapse="true" title="Answer"}
The mistake is **feeding unscaled features to `dist()`.** Income spans tens of
thousands of dollars while age spans a few decades, so the income column swallows
the distance computation whole. Look at the printed distances: applicants 1 and 2
differ by 15 years of age and \$15,000 of income, and their distance is 15000.01
--- the age gap contributes about 0.01 of that. Age is effectively invisible,
exactly what the colleague observed.

The fix: standardize both columns (subtract the mean, divide by the standard
deviation) before computing distances, so each feature contributes on an equal
footing.

```{r p4-answer}
age    <- c(25, 40, 60)
income <- c(30000, 45000, 120000)

# Scaled: each column centered and divided by its sd
applicants_scaled <- cbind(scale(age), scale(income))
dist(applicants_scaled)   # now age and income both matter
```
:::

## Problem 5 --- mixing it with distance *(uses distance-similarity-and-geometry's Manhattan distance)*

Using the lesson's original toy data (query at (4, 4), k=3), compute the
**Manhattan distance** ($\sum |a_i - b_i|$) from the query to each of the 6
training points. Sort them. Does the k=3 classification change compared to
Euclidean distance? Why or why not?

```{webr}
#| autorun: false
# p5-starter — compute Manhattan distances, sort, vote
train_x <- c(1, 2, 3, 6, 7, 8)
train_y <- c(2, 3, 1, 7, 8, 6)
label   <- c("A", "A", "A", "B", "B", "B")
query   <- c(4, 4)
k <- 3
```

::: {.callout-note collapse="true" title="Answer"}
Manhattan distances from (4, 4): sum the absolute coordinate gaps.

- (1, 2): $|3| + |2| = 5$ --- A
- (2, 3): $|2| + |1| = 3$ --- A
- (3, 1): $|1| + |3| = 4$ --- A
- (6, 7): $|2| + |3| = 5$ --- B
- (7, 8): $|3| + |4| = 7$ --- B
- (8, 6): $|4| + |2| = 6$ --- B

Sorted (Manhattan): (2, 3) = 3, (3, 1) = 4, (1, 2) = 5 and (6, 7) = 5 (tied),
(8, 6) = 6, (7, 8) = 7.

k=3: (2, 3) A, (3, 1) A, then the 3rd slot is a tie between (1, 2) A and
(6, 7) B. Either way the majority is A.

**The classification does not change** because the query is well inside the A
cluster regardless of which distance metric we use. The ordering of points
shifted slightly (Euclidean had (2, 3) then (3, 1) then a tie; Manhattan has the
same first two, then a different tie), but the 3-nearest neighborhood is the same
group. Distance choice matters more when the query sits near the decision boundary
between clusters, where a different metric can rearrange which points qualify as
neighbors.

```{r p5-answer}
train_x <- c(1, 2, 3, 6, 7, 8)
train_y <- c(2, 3, 1, 7, 8, 6)
label   <- c("A", "A", "A", "B", "B", "B")
query   <- c(4, 4)

# Manhattan distances: sum of absolute gaps
man_dists <- abs(train_x - query[1]) + abs(train_y - query[2])
man_dists   # 5 3 4 5 7 6

# Euclidean for comparison
euc_dists <- sqrt((train_x - query[1])^2 + (train_y - query[2])^2)
round(euc_dists, 2)   # 3.61 2.24 3.16 3.61 5.00 4.47

# k=3 vote under each metric
k <- 3
man_nbrs <- order(man_dists)[1:k]
euc_nbrs <- order(euc_dists)[1:k]

cat("Manhattan k=3 labels:", label[man_nbrs], "-> vote:",
    names(which.max(table(label[man_nbrs]))), "\n")
cat("Euclidean k=3 labels:", label[euc_nbrs], "-> vote:",
    names(which.max(table(label[euc_nbrs]))), "\n")
# Both predict A
```
:::

## Problem 6 --- transfer: too many dimensions

A dataset has 500 columns and 50 rows. A colleague wants to use kNN to classify
new observations. What concern would you raise, and why?

::: {.callout-note collapse="true" title="Answer"}
The primary concern is the **curse of dimensionality**. With 500 dimensions and
only 50 observations, all pairwise distances converge to roughly the same value
--- the ratio of the farthest distance to the nearest distance shrinks toward 1.
When every point is about equally far from every other point, the "nearest"
neighbor is not meaningfully nearer than the rest, and the majority vote becomes
essentially arbitrary.

The lesson's simulation showed this directly: in 500 dimensions, the max/min
distance ratio collapsed to near 1, while in 2 dimensions it was large (meaning
neighbors were genuinely nearby). With 10 times more features than observations,
kNN is almost certainly a poor choice unless the features are heavily reduced
first (e.g., by PCA or feature selection to bring the effective dimensionality
down).
:::


## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice --- come back to it **at least a day from
now**, once the material isn't fresh. A pass then is real evidence you've retained
it. Five questions, graded on the page; 4/5 or better means this module is solid.
Score under 4? Revisit what the missed questions point at, then retake in a
**later session** --- passing after another gap is the evidence that counts.
Results are saved in this browser only.

::: {.module-check module="foundations/k-nearest-neighbors" pass="4"}

::: {.quiz-q qid="knn-vote-result" qtype="mc" answer="c"}
For the lesson's toy data (6 points, query at (4, 4), k = 3), what does kNN predict?

a) B, because the query is equidistant from both clusters
b) A tie --- 3 A's and 3 B's in the training data cancel out
c) A, because all three nearest neighbors are labeled A
d) B, because the nearest single point is labeled B
:::

::: {.quiz-q qid="knn-lazy-meaning" qtype="mc" answer="a"}
kNN is called a "lazy learner" because:

a) It stores training data and defers computation to prediction time
b) It uses a simpler distance formula than other classification methods do
c) It requires fewer training observations than other classification methods
:::

::: {.quiz-q qid="knn-large-k" qtype="mc" answer="b"}
What happens when you set k equal to the total number of training points?

a) The model becomes maximally flexible and captures every local pattern
b) The model always predicts the overall majority class, ignoring local structure
c) The model produces the most accurate predictions because it uses all available information
d) The model refuses to make a prediction because there are no "nearest" neighbors
:::

::: {.quiz-q qid="knn-scaling-interleaved" qtype="mc" answer="d"}
*(Interleaved from [distance, similarity & geometry](../distance-similarity-and-geometry/lesson.qmd).)* A kNN model uses height (cm, range 150--190) and income (dollars, range 30000--80000) as features, without scaling. Which feature dominates the distance computation?

a) Height, because it is measured first
b) Neither --- kNN automatically weights features equally
c) Height, because centimeters are a more precise unit than dollars
d) Income, because its numeric range is thousands of times larger
:::

::: {.quiz-q qid="knn-dist-query-to-86" qtype="numeric" answer="4.47" tolerance="0.01"}
Using the lesson's toy data, what is the Euclidean distance from the query point (4, 4) to the training point (8, 6)? Round to two decimal places.
:::

:::

::::
