---
title: "A prior on the effect"
subtitle: "Two competing bell curves for β̂: N(0, V) if there's no effect, N(0, V+W) if there is"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live cells below re-state their own data. Base R only.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

::: {.callout-tip title="This lesson runs in your browser" collapse="true"}
The code cells below with a **Run** button are live — powered by WebR (R compiled to
WebAssembly), they run right here on your phone or laptop, no install needed. Edit any
of them and re-run. The first run downloads a small R engine, so give it a few seconds.
Cells without a Run button are ordinary listings whose output was computed when the
page was built.
:::

## Where we're going

Last module you reduced an association to the pair $(\hat\beta, V)$. Now we add the one
Bayesian ingredient the Approximate Bayes Factor needs: a **prior on the true effect** — a
normal curve $N(0, W)$ that says, before seeing any data, "the true effect is probably
small, near zero, but could plausibly be as large as $W$'s spread allows." With that prior
in hand, the estimate $\hat\beta$ has **two** competing predicted distributions: a narrow
one, $N(0, V)$, if there is truly *no* effect (the null), and a wider one, $N(0, V + W)$,
if there *is* an effect drawn from the prior (the alternative). The ratio of those two
curves — next module — is the Bayes factor. This module builds the two curves and explains
what the single knob $W$ controls.

**Builds on:** the [previous module](../01-the-summary-of-an-association/lesson.qmd) — the
pair $(\hat\beta, V)$;
[distributions & densities](../../../../foundations/distributions-and-densities/lesson.qmd)
— the normal curve $N(\mu, \sigma^2)$ and `dnorm`; and
[marginal likelihood & model evidence](../../../../foundations/marginal-likelihood-and-model-evidence/lesson.qmd)
— averaging an estimate's density over a prior gives the marginal $N(0, V + W)$, the exact
result the alternative curve uses.

## Warm-up: do you remember?

Before the new stuff, three quick reps — answers from memory first, then check.

**1.** *(Module 1)* A variant has $\hat\beta = 0.20$ and $\text{se} = 0.05$. What is its
variance $V$?

```{webr}
#| autorun: false
# warmup1-starter — edit and Run to check
se <- 0.05
```

::: {.callout-note collapse="true" title="Answer"}
$V = \text{se}^2 = 0.05^2 = 0.0025$. This is our running variant for the rest of the
course. Shaky? Revisit
[the summary of an association](../01-the-summary-of-an-association/lesson.qmd).
:::

**2.** *(Foundation)* You average a normal estimate with variance $V$ over a normal prior
$N(0, W)$ on the true value. What is the resulting marginal distribution of the estimate?

::: {.callout-note collapse="true" title="Answer"}
$N(0, V + W)$ — a normal curve centered at zero whose variance is the two variances
**added**. See
[marginal likelihood & model evidence](../../../../foundations/marginal-likelihood-and-model-evidence/lesson.qmd).
:::

**3.** *(Foundation)* In `dnorm(x, mean, sd)`, what does the returned number represent — a
probability, or something else?

::: {.callout-note collapse="true" title="Answer"}
A **density** — the *height* of the bell curve at $x$, not a probability. Only *area* under
the curve is a probability. See
[distributions & densities](../../../../foundations/distributions-and-densities/lesson.qmd).
:::

## A tiny example first

Take our running variant: the estimate is $\hat\beta = 0.20$, with variance $V = 0.0025$
(so $\text{se} = 0.05$). We are weighing two stories about the *true* effect $\theta$ (the
value $\hat\beta$ is trying to estimate):

- **The null story ($H_0$):** there is no effect, $\theta = 0$. Then the estimate just
  wobbles around zero with its own noise, so $\hat\beta \sim N(0, V) = N(0, 0.0025)$ — a
  narrow bell curve, standard deviation $\sqrt{0.0025} = 0.05$.
- **The alternative story ($H_1$):** there *is* an effect, and before the data we believe
  it's drawn from a prior $\theta \sim N(0, W)$ with $W = 0.04$. Averaging the estimate's
  noise over that prior (the foundation result) widens the curve to $\hat\beta \sim N(0, V
  + W) = N(0, 0.0425)$ — standard deviation $\sqrt{0.0425} \approx 0.206$.

Now ask: **at our observed value $\hat\beta = 0.20$, which curve is taller?** The height of
each curve at $0.20$ is a `dnorm`:

- Null curve height: `dnorm(0.20, 0, sqrt(0.0025))` $= 0.00268$.
- Alternative curve height: `dnorm(0.20, 0, sqrt(0.0425))` $= 1.209$.

The alternative curve is *far* taller at $0.20$ — about $450$ times taller. In plain terms:
an estimate as big as $0.20$ is a shock under the narrow null curve (it's four standard
errors out) but perfectly ordinary under the wider alternative curve. That height ratio is
the seed of the Bayes factor.

The same thing in R, so you can check every step:

```{r toy-example}
beta_hat <- 0.20
V <- 0.0025          # variance of the estimate (from Module 1: se^2)
W <- 0.04            # prior variance on the true effect

# Two competing predicted curves for the estimate:
height_null <- dnorm(beta_hat, mean = 0, sd = sqrt(V))       # N(0, V)   -- no effect
height_alt  <- dnorm(beta_hat, mean = 0, sd = sqrt(V + W))   # N(0, V+W) -- effect

c(null = height_null, alt = height_alt)          # 0.00268 vs 1.209
height_alt / height_null                          # ~451: alt curve is far taller here
```

## Getting a feel for it

**The mental picture: a wide net vs a tight net.** Think of each hypothesis as a fisher
casting a net for where the estimate will land. The null caster believes the true effect is
exactly zero, so they cast a **tight net** right around zero (spread $V$) — they're betting
the estimate lands very close to zero. The alternative caster admits the effect *could* be
anywhere the prior allows, so they cast a **wide net** (spread $V + W$) that still centers
on zero but reaches much further out. When the estimate actually lands at $0.20$, far from
zero, it falls *outside* the tight net but comfortably *inside* the wide one — so the wide
net "predicted" it far better. A hypothesis is rewarded for having cast a net where the
data actually landed. (Where the picture strains: both nets are centered at zero even though
the alternative allows non-zero effects — that's because the prior itself is centered at
zero, spreading the *possible* effects symmetrically around no-effect; the alternative bets
on "an effect of some size," not on a specific direction.)

**Why is the alternative curve wider — why $V + W$ and not something else?** The estimate
carries two independent sources of wobble under the alternative: the measurement noise of
the estimate (variance $V$) *and* the genuine spread of the true effect the prior allows
(variance $W$). Two independent wobbles stacked on top of each other combine by **adding
variances** (the foundation result), so the total spread is $V + W$. This is also why we
carried $V = \text{se}^2$ as a variance back in Module 1 — so it could add cleanly to $W$
here. The null curve has only the first wobble ($V$), because under "no effect" the true
value isn't spread at all — it's pinned at zero.

**What the knob $W$ feels like.** $W$ is the *only* free choice in the whole method, and it
encodes "how big could a real effect plausibly be." Its square root is the prior standard
deviation of the log-odds-ratio: with $W = 0.04$, the prior sd is $\sqrt{0.04} = 0.2$, so
the prior thinks true log-ORs of roughly $\pm 0.4$ (two sds) are plausible — odds ratios
from about $e^{-0.4} \approx 0.67$ to $e^{0.4} \approx 1.49$. Turn $W$ up and the
alternative net widens, expecting bigger effects; turn it toward zero and the alternative
net shrinks onto the null net, and the two hypotheses become hard to tell apart. The
case-control default in `coloc`, `sd.prior = 0.2`, is exactly this $W = 0.2^2 = 0.04$.

## What the notation is going to say

The formulas just name the two nets. The **prior** on the true effect is $\theta \sim N(0,
W)$: a bell curve for the unknown true effect, centered at zero (no effect is the default
expectation), with spread $W$ (how big effects could plausibly get). The **null predicted
distribution** of the estimate is $\hat\beta \sim N(0, V)$ — only the estimate's own noise.
The **alternative predicted distribution** is $\hat\beta \sim N(0, V + W)$ — the estimate's
noise plus the prior's spread, variances added. Every piece is a `dnorm` you can evaluate,
exactly as in the toy computation.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $\theta$ | "theta" | the unknown **true** effect the estimate is chasing (a true log odds ratio) | the true log-OR (unknown) |
| $\hat\beta$ | "beta hat" | the effect **estimate** from the data | $0.20$ |
| $V$ | "vee" | variance of the estimate ($V = \text{se}^2$, from Module 1) | $0.0025$ |
| $W$ | "double-you" | the **prior variance** on the true effect — the one knob | $0.04$ |
| $\sqrt{W}$ | "root double-you" | the prior standard deviation, `sd.prior` — the plausible log-OR scale | $0.2$ |
| $N(0, V)$ | "normal, mean zero, variance vee" | the null predicted curve for $\hat\beta$ (no effect) | $N(0, 0.0025)$ |
| $N(0, V+W)$ | "normal, mean zero, variance vee plus double-you" | the alternative predicted curve for $\hat\beta$ | $N(0, 0.0425)$ |
| $H_0, H_1$ | "aitch-nought, aitch-one" | the null (no effect) and alternative (effect) hypotheses | the two stories |
| $\sim$ | "is distributed as" | "follows the distribution" (in R: `rnorm`/`dnorm` with these parameters) | $\hat\beta \sim N(0, V)$ under $H_0$ |

## The formal version

**The prior.** Before the data, the true effect is modelled as a normal centered at zero:

$$
\theta \sim N(0, W).
$$

Reading it aloud: "theta is distributed normal with mean zero and variance double-you."
Everyday words: **"we expect the true effect to be small (near zero), but allow it to be as
large as a spread of $W$ makes plausible."** Centering at zero is the honest default — no
effect is the baseline expectation — and $W$ is the dial for how far real effects might
reach.

**The two predicted distributions of the estimate.** Under each hypothesis, the estimate
$\hat\beta$ has its own predicted (prior-predictive) curve:

$$
\text{Under } H_0:\quad \hat\beta \sim N(0, V),
\qquad
\text{Under } H_1:\quad \hat\beta \sim N(0, V + W).
$$

Reading it aloud: "under aitch-nought, beta-hat is normal with variance vee; under
aitch-one, beta-hat is normal with variance vee plus double-you." Everyday words: **"if
there's no effect the estimate just carries its own noise; if there is an effect the
estimate carries its noise plus the prior's spread — so the variances add."**

The $H_1$ curve is the marginal-likelihood result from the foundation, applied here:
averaging the estimate's density $N(\theta, V)$ over the prior $\theta \sim N(0, W)$ gives
$N(0, V + W)$ — the variances $V$ and $W$ add. The $H_0$ curve is the special case $W = 0$
(the prior collapses to a spike at zero), leaving just $N(0, V)$.

Mapping it onto the toy example: with $V = 0.0025$ and $W = 0.04$, the two curves are
$N(0, 0.0025)$ and $N(0, 0.0425)$, whose heights at $\hat\beta = 0.20$ are $0.00268$ and
$1.209$ — exactly the hand numbers. Sanity-check them in the Explore cells below.

## Explore it

```{webr}
#| autorun: true
# explore-1 — LIVE: the two competing curve HEIGHTS at the observed estimate.
# TRY IT: change W (the prior variance / how big effects could be) and watch the heights.
# PREDICT: if you shrink W to 0.0001 (barely any effect allowed), will the two heights get
# CLOSER together or further apart? Commit, then run.
beta_hat <- 0.20
V <- 0.0025          # from Module 1: se^2
W <- 0.04            # prior variance on the true effect

height_null <- dnorm(beta_hat, 0, sqrt(V))         # N(0, V)
height_alt  <- dnorm(beta_hat, 0, sqrt(V + W))     # N(0, V + W)
c(null = round(height_null, 5), alt = round(height_alt, 5),
  ratio_alt_over_null = round(height_alt / height_null, 1))     # ~451
```

```{webr}
#| autorun: true
# explore-2 — LIVE: confirm the alternative curve IS the marginal-likelihood result.
# Averaging the estimate's noise N(theta, V) over the prior N(0, W) should give N(0, V+W).
# PREDICT: will brute-force integration match the closed-form dnorm below? Guess, run.
beta_hat <- 0.20; V <- 0.0025; W <- 0.04

integrand <- function(theta) dnorm(beta_hat, theta, sqrt(V)) * dnorm(theta, 0, sqrt(W))
brute  <- integrate(integrand, -Inf, Inf)$value    # average over the prior
closed <- dnorm(beta_hat, 0, sqrt(V + W))           # N(0, V+W): the alternative curve

c(brute_force = brute, closed_form = closed)        # equal -- the alt curve is the marginal
```

```{webr}
#| autorun: true
#| fig-alt: "Two normal density curves centered at zero on the same axes: a tall narrow curve labelled 'null N(0, V)' and a much shorter wider curve labelled 'alternative N(0, V+W)'. A dashed vertical line at beta-hat = 0.20 sits far out in the tail of the narrow null curve but well within the wide alternative curve, where the alternative curve is clearly higher."
# explore-3 — LIVE: SEE the two nets and where the estimate landed.
# TRY IT: drag the estimate by changing beta_hat -- near 0 the null curve is taller;
# far out the alternative curve is taller.
beta_hat <- 0.20; V <- 0.0025; W <- 0.04
grid <- seq(-0.6, 0.6, length.out = 500)

plot(grid, dnorm(grid, 0, sqrt(V)), type = "l", lwd = 2,
     xlab = expression(hat(beta)), ylab = "density",
     main = "Two predicted curves for the estimate")
lines(grid, dnorm(grid, 0, sqrt(V + W)), lwd = 2, lty = 2)
abline(v = beta_hat, col = "grey50")
legend("topright", c("null  N(0, V)", "alternative  N(0, V+W)"),
       lwd = 2, lty = c(1, 2), bty = "n")
```

::: {.callout-note title="Optional — priors that vary with allele frequency"}
Wakefield also considers priors whose variance $W$ depends on the variant's **minor allele
frequency (MAF)**. The idea: a fixed prior on the log-OR isn't always the most natural — one
can instead put the prior on a *standardized* effect, which makes the implied $W$ larger for
rare variants (small MAF) and smaller for common ones, because a rare allele carries less
information per copy. A common device makes $W$ scale like $1/[2\,\text{MAF}(1-\text{MAF})]$
— roughly $10.5$ at MAF $0.05$ versus $2$ at MAF $0.5$. You don't need this for the core
method (the case-control default uses a fixed $W = 0.04$), which is why it's optional; it
buys you a prior that adapts to how much each variant could plausibly reveal.
:::

## Common traps

**Trap 1 — "the two curves differ in where they're centered."** No — *both* the null curve
$N(0, V)$ and the alternative curve $N(0, V + W)$ are centered at **zero**. They differ only
in **spread**: the alternative is wider ($V + W = 0.0425$ vs $V = 0.0025$). The alternative
doesn't bet on a specific non-zero effect; it bets on "an effect of some size, either
direction," which spreads its curve symmetrically around zero rather than shifting it.

**Trap 2 — "to widen the alternative curve, add the standard deviations."** No — you add the
*variances*. Here $V = 0.0025$ (sd $0.05$) and $W = 0.04$ (sd $0.2$), so the alternative
variance is $V + W = 0.0425$, whose sd is $\sqrt{0.0425} \approx 0.206$ — **not** $0.05 +
0.2 = 0.25$. Adding sds would overstate the width. Variances (spreads²) add; standard
deviations don't.

## Check yourself

**1.** A variant has $\hat\beta = 0.15$ and $V = 0.0025$, and you use a prior variance $W =
0.04$. Compute the height of the null curve $N(0, V)$ and the alternative curve $N(0, V+W)$
at $\hat\beta = 0.15$. Which is taller?

```{webr}
#| autorun: false
# check1-starter — edit and Run to check your hand answer
beta_hat <- 0.15
V <- 0.0025
W <- 0.04
```

::: {.callout-note collapse="true" title="Answer"}
Null height: `dnorm(0.15, 0, sqrt(0.0025))` $= 0.0886$. Alternative height: `dnorm(0.15, 0,
sqrt(0.0425))` $= 1.485$. The **alternative** is taller — $0.15$ is three standard errors
out under the narrow null but ordinary under the wide alternative. (Note the alternative
height is even a touch higher than at $\hat\beta = 0.20$, where it was $1.209$, because
$0.15$ is closer to the wide curve's center.)
:::

**2.** You change the prior from $W = 0.04$ to $W = 0.0001$ (you now believe any real effect
is minuscule). What happens to the alternative curve $N(0, V + W)$, and to its ability to
out-predict the null curve?

::: {.callout-note collapse="true" title="Answer"}
The alternative variance drops from $0.0425$ to $0.0026$, nearly equal to the null's
$0.0025$ — the wide net shrinks almost onto the tight net. The two curves become nearly
identical, so the alternative can barely out-predict the null anywhere: with such a tiny
prior, "there's an effect" and "there's no effect" make almost the same prediction, and the
data can't distinguish them.
:::

**3.** In your own words: explain to a colleague, in two sentences, what the prior $N(0, W)$
does and why the alternative predicted curve for the estimate is wider than the null one.

::: {.callout-note collapse="true" title="A model answer"}
The prior $N(0, W)$ encodes our before-the-data belief about the true effect — probably
small (centered at zero) but plausibly as large as a spread of $W$ allows. The alternative
curve for the estimate is wider because, under "there is an effect," the estimate carries
*two* independent sources of spread — its own measurement noise ($V$) plus the genuine
spread of the true effect the prior permits ($W$) — and independent variances add, giving
$V + W$; the null curve has only the noise, so its variance is just $V$.
:::

## What you can now do

- **Place a normal prior $N(0, W)$ on the true effect** and state what $W$ (and its root,
  the `sd.prior`) encodes about the plausible odds-ratio range.
- **Write down the two predicted curves for the estimate** — $N(0, V)$ under the null and
  $N(0, V + W)$ under the alternative — and say why the variances add.
- **Evaluate and compare the two curves' heights** at an observed $\hat\beta$ with `dnorm`,
  and read the taller curve as "the hypothesis that predicted the data better."
- **Avoid the centering and standard-deviation traps** — both curves center at zero, and
  widths combine by adding variances, not standard deviations.

If any of these feels shaky, revisit [A tiny example first](#a-tiny-example-first) and
[The formal version](#the-formal-version) before continuing — the next module divides these
two curves' heights to build the Approximate Bayes Factor.

## Where this goes next

You have the two competing curves for the estimate. The
[next module](../03-the-approximate-bayes-factor/lesson.qmd) takes their ratio at the
observed $\hat\beta$ — the taller-curve-wins comparison you just did numerically — and shows
it collapses to a clean closed-form log Bayes factor you can build in a few lines of base R,
$\text{lABF} = \tfrac12\big(\log(1 - r) + r Z^2\big)$. Ready for more reps? Do the
[practice problems](practice.qmd).
