---
title: "Rank and linear independence"
subtitle: "When columns carry their own information — and when they don't"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

## Where we're going

You have already seen that a matrix can be **singular** — its inverse does not
exist, and `solve()` errors out. You have also seen that **collinear** columns
make regression estimates blow up. This module explains *why* both of those
happen: some columns are just weighted sums of others, so they carry no new
information. The number of genuinely independent columns is the **rank**, and
rank controls what a matrix can and cannot do — including how many directions
PCA can find.

**Builds on:**
[matrices & linear transforms](../matrices-and-linear-transforms/lesson.qmd)
(matrix multiply, transpose, inverse, singular),
[vectors, subscripts & summation](../vectors-and-summation/lesson.qmd)
(vectors, dot product, weighted sums)

## Warm-up: do you remember?

Before the new stuff, a few quick reps — answers from memory first, then check.

**1.** *(matrices)* What happens when you call `solve()` on the matrix
$S = \begin{bmatrix} 1 & 2 \\ 2 & 4 \end{bmatrix}$? Why?

::: {.callout-note collapse="true" title="Answer"}
R errors with "system is exactly singular" — row 2 is exactly twice row 1, so
the matrix squashes different inputs onto the same output and cannot be undone.
Taught in [matrices & linear transforms](../matrices-and-linear-transforms/lesson.qmd);
revisit the inverse section if shaky.
:::

**2.** *(vectors)* If $\mathbf{a} = (3, 1)$ and $\mathbf{b} = (0, 2)$, what is the
weighted sum $2\mathbf{a} + 1\mathbf{b}$?

::: {.callout-note collapse="true" title="Answer"}
$2(3, 1) + 1(0, 2) = (6, 2) + (0, 2) = (6, 4)$. In R: `2 * c(3,1) + 1 * c(0,2)`.
Taught in [vectors, subscripts & summation](../vectors-and-summation/lesson.qmd);
revisit if shaky.
:::

## A tiny example first

You measured three things about each of four houses: square metres, number of
bedrooms, and "double bedrooms" (which happens to be exactly 2 × bedrooms).

```{r toy-data}
sqm      <- c(80, 100, 120, 140)
bedrooms <- c(2,   3,   3,   4)
dbl_bed  <- 2 * bedrooms              # col 3 is exactly 2 * col 2

X <- cbind(sqm, bedrooms, dbl_bed)
X
```

Look at the third column: every entry is exactly twice the second column.
$\text{dbl\_bed} = 2 \times \text{bedrooms} + 0 \times \text{sqm}$ — it is a
weighted sum of the other two columns. It carries no information that was not
already there.

Can you invert $X^\top X$?

```{r xtx-singular}
XtX <- t(X) %*% X
XtX

# Try to invert it:
tryCatch(solve(XtX), error = function(e) message(e$message))
```

R refuses: "system is exactly singular: U[3,3] = 0." The third column being a
weighted sum of the others made $X^\top X$ singular — exactly the problem from
the [matrices lesson](../matrices-and-linear-transforms/lesson.qmd).

Now drop the redundant column:

```{r xtx-good}
X_good <- X[, 1:2]   # just sqm and bedrooms
solve(t(X_good) %*% X_good)   # works fine
```

Two genuinely independent columns → $X^\top X$ is invertible. Three columns,
one redundant → singular. That difference is what **linear independence** and
**rank** capture.

## Getting a feel for it

**Directions of information.** Think of each column as an arrow pointing in
some direction in "data space." If every arrow points in its own direction, each
column adds something new — you can reach places you could not reach before. But
if one arrow is just a stretched or rotated combination of the others, it is not
a new direction; it is a repeat. Adding it to your team does not let you go
anywhere new.

The dbl_bed column pointed in exactly the same direction as bedrooms, just
twice as long. Stretching an arrow does not create a new direction — it is still
the same line.

Here is that picture with three small arrows:

```{r independence-picture}
#| fig-alt: "Two solid arrows from the origin pointing in different directions, labelled (2, 1) and (-1, 2), and a dashed arrow to (4, 2) lying exactly along the (2, 1) arrow's line, labelled (4, 2) = 2 times (2, 1). The dashed arrow adds no new direction."
plot(NULL, xlim = c(-2, 5), ylim = c(-0.5, 3), asp = 1,
     xlab = "", ylab = "",
     main = "A dependent arrow adds no new direction")
arrows(0, 0, 2, 1, lwd = 2, length = 0.12)                      # (2, 1) — its own direction
arrows(0, 0, -1, 2, lwd = 2, length = 0.12)                     # (-1, 2) — its own direction
arrows(0, 0, 4, 2, lwd = 2, lty = 2, col = "grey40",
       length = 0.12)                                            # (4, 2) = 2 x (2, 1) — same line
text(1.1, 0.55, "(2, 1)", pos = 1)
text(-1, 2, "(-1, 2)", pos = 2)
text(4, 2, "(4, 2) = 2 × (2, 1)", pos = 3)
```

The two solid arrows, $(2, 1)$ and $(-1, 2)$, each point somewhere the other
cannot reach — two genuine directions. The dashed arrow $(4, 2)$ is exactly
$2 \times (2, 1)$: it sits on top of the first arrow's line, just extended. It
adds no new direction — that IS dependence, drawn.

**Why this matters for regression.** When you ask R to split credit among
columns using `lm()`, each column needs its own "direction" to get a unique
share. If two columns point in the same direction, their shares are not
determined — you can give all the credit to one or all to the other, and the
fit is the same. That is why R hands back `NA` for the redundant column, and
why the [collinearity discussion in variance structure](../variance-structure-and-standardization/lesson.qmd)
calls it "two knobs wired to the same wheel."

**What the numbers feel like.** Our three-column matrix has two genuine
directions (sqm, bedrooms); the third is fake. So this matrix has "rank 2" —
it carries two columns' worth of real information, even though it has three
columns.

## What the notation is going to say

The formal definition will say: a set of columns is **linearly independent** if
no column can be written as a weighted sum of the others. Equivalently, the
*only* way to combine them all to get zero is with every weight equal to zero.
In our toy example, dbl_bed = 2 × bedrooms + 0 × sqm, so the set {sqm,
bedrooms, dbl_bed} is linearly **dependent** — that weighted sum exists with
nonzero weights. Drop the redundant column and {sqm, bedrooms} is independent:
neither is a multiple of the other.

The **rank** is then the largest number of columns you can pick that are still
independent. Our 4×3 matrix $X$ has rank 2; the good 4×2 matrix has rank 2 too.

The **span** is every vector you could build by taking weighted sums of the
columns. Adding a dependent column does not enlarge the span — the span of
{sqm, bedrooms, dbl_bed} is the same as the span of {sqm, bedrooms}, because
dbl_bed was already reachable.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $c_1, c_2, \ldots$ | "see one, see two, ..." | scalar weights in a weighted sum | the 2 and 0 in $\text{dbl\_bed} = 2 \cdot \text{bedrooms} + 0 \cdot \text{sqm}$ |
| $c_1 \mathbf{v}_1 + c_2 \mathbf{v}_2 + \cdots$ | "see-one times vee-one plus see-two times vee-two plus ..." | a **linear combination** — a weighted sum of vectors | $0 \cdot \text{sqm} + 2 \cdot \text{bedrooms}$ |
| $\text{Span}\{\mathbf{v}_1, \ldots, \mathbf{v}_k\}$ | "the span of vee-one through vee-k" | every vector reachable by weighted sums of these vectors | every 4-long vector reachable by mixing sqm and bedrooms |
| $\text{rank}(A)$ | "the rank of A" | the number of genuinely independent columns | $\text{rank}(X) = 2$ |
| $\mathbf{0}$ | "the zero vector" | a vector of all zeros | $(0, 0, 0, 0)$ |
| $X^\top X$ | "ex-transpose ex" | the square matrix whose invertibility rank controls | our 3×3 singular matrix |
| $\min(n, p)$ | "the minimum of n and p" | whichever is smaller: the number of rows or columns | $\min(4, 3) = 3$ |
| $\Longrightarrow$ | "implies" | if the left side is true, the right side must be true | the only weights that produce zero are all-zero weights |
| $\in$ | "is in" | belongs to a set | $c_1 \in \mathbb{R}$ means "see-one is a real number" |
| $\mathbb{R}$ | "the real numbers" | the set of all real numbers | the weights can be any real number |
| $\{\, \cdot \,\}$ (braces) | "the set of" | everything between the braces is a description of the items that belong to the set | $\{c_1\mathbf{v}_1 + \cdots\}$ = "the set of all such weighted sums" |
| $\mid$ (the vertical bar inside braces) | "such that" | separates *what's in the set* (left of the bar) from *the condition it must satisfy* (right of the bar) | "weighted sums $c_1\mathbf{v}_1 + \cdots$, **such that** $c_1,\ldots,c_k$ are any real numbers" |
| $\mathbf{u}\mathbf{v}^\top$ | "u v-transpose" | a column times a row — a rank-1 matrix | a 4×4 matrix with every row a multiple of $\mathbf{v}^\top$ |

## The formal version

**Linear independence.** A set of vectors $\mathbf{v}_1, \mathbf{v}_2, \ldots, \mathbf{v}_k$ is **linearly independent** if:

$$
c_1 \mathbf{v}_1 + c_2 \mathbf{v}_2 + \cdots + c_k \mathbf{v}_k = \mathbf{0}
\quad \Longrightarrow \quad
c_1 = c_2 = \cdots = c_k = 0
$$

Reading it aloud: "see-one times vee-one plus see-two times vee-two plus dot-dot-dot
equals the zero vector *implies* every see equals zero" — or in everyday words:
"the only way to combine them to get nothing is to use no contribution from any
of them." [@austin2022understandinglinearalgebra, p. 98]

If there exist weights that are NOT all zero and still produce the zero vector,
the set is **linearly dependent** — at least one vector is a weighted sum of the
others. Why are those two statements the same? Take the combination that equals
zero, move any term whose weight is nonzero across the equals sign, and divide
both sides by that (nonzero) weight — that vector is now written as a weighted
sum of the rest.

Mapping it onto the toy example: $0 \cdot \text{sqm} + 2 \cdot \text{bedrooms} + (-1) \cdot \text{dbl\_bed} = \mathbf{0}$,
so {sqm, bedrooms, dbl_bed} is dependent (the weights 0, 2, −1 are not all zero).

**Span.** The **span** of a set of vectors is the set of all vectors reachable
by linear combinations of those vectors. [@austin2022understandinglinearalgebra, p. 81]

$$
\text{Span}\{\mathbf{v}_1, \ldots, \mathbf{v}_k\}
= \{ c_1 \mathbf{v}_1 + \cdots + c_k \mathbf{v}_k \mid c_1, \ldots, c_k \in \mathbb{R} \}
$$

Reading it aloud: "the span of vee-one through vee-k equals the braces — **the set
of** — see-one times vee-one plus dot-dot-dot, the vertical bar — **such that** —
see-one through see-k are in the real numbers" — or in everyday words: "everything
you can reach by mixing these vectors, where the weights can be any real numbers."

Removing a dependent column does not shrink the span — the dependent vector was
already reachable, so it was not contributing anything new.
[@austin2022understandinglinearalgebra, p. 97]

**Rank.** The **rank** of a matrix is the number of linearly independent columns
(equivalently, the number of pivots in its row-reduced form).
[@austin2022understandinglinearalgebra, p. 100]

$$
\text{rank}(A) = \text{number of independent columns of } A
$$

In R: `qr(A)$rank`.

## Explore it

```{webr}
#| autorun: true
# The toy data: three columns, one redundant
sqm      <- c(80, 100, 120, 140)
bedrooms <- c(2,   3,   3,   4)
dbl_bed  <- 2 * bedrooms

X <- cbind(sqm, bedrooms, dbl_bed)

# Rank: how many genuinely independent columns?
qr(X)$rank     # 2 — dbl_bed adds nothing

# Drop the redundant column:
qr(X[, 1:2])$rank   # still 2
```

```{webr}
#| autorun: false
# TRY IT: make col3 genuinely different and re-check the rank.
# PREDICT: what rank will the new X have? Write your guess, then run.
sqm      <- c(80, 100, 120, 140)
bedrooms <- c(2,   3,   3,   4)
col3     <- c(1,   0,   1,   0)    # change these — anything NOT a combo of the first two

X_new <- cbind(sqm, bedrooms, col3)
qr(X_new)$rank
```

```{webr}
#| autorun: false
# Rank deficiency makes X'X singular -- solve() fails, but only when a column truly IS
# a weighted sum of the others. Two candidate third columns below: one repeats old
# information, one adds a genuinely new direction.
# PREDICT, for EACH candidate: will solve(t(X) %*% X) work or error? Write both
# predictions first, then run and check.
sqm      <- c(80, 100, 120, 140)
bedrooms <- c(2,   3,   3,   4)

col_dep   <- 2 * bedrooms          # a weighted sum of bedrooms -> dependent
col_indep <- c(1,   0,   1,   0)   # not a combo of sqm or bedrooms -> independent

X_dep   <- cbind(sqm, bedrooms, col_dep)
X_indep <- cbind(sqm, bedrooms, col_indep)

cat("rank with col_dep:  ", qr(X_dep)$rank, "\n")
cat("rank with col_indep:", qr(X_indep)$rank, "\n")

tryCatch(
  { solve(t(X_dep) %*% X_dep); cat("X_dep:   solve() worked\n") },
  error = function(e) cat("X_dep:   ERROR -", e$message, "\n")
)
tryCatch(
  { solve(t(X_indep) %*% X_indep); cat("X_indep: solve() worked\n") },
  error = function(e) cat("X_indep: ERROR -", e$message, "\n")
)
```

**Optional (preview — uses eigenvalues from [PCA & SVD](../pca-and-svd/lesson.qmd),
fine to skip on first read).** Rank also caps how many components PCA can find:

```{webr}
#| autorun: false
# Rank caps PCA: you can't get more components than the rank.
# PREDICT: how many nonzero eigenvalues will a rank-2 matrix have?
sqm      <- c(80, 100, 120, 140)
bedrooms <- c(2,   3,   3,   4)
dbl_bed  <- 2 * bedrooms
X <- cbind(sqm, bedrooms, dbl_bed)

S <- cov(X)
eigenvalues <- eigen(S)$values
cat("rank:", qr(X)$rank, "\n")
cat("eigenvalues:", round(eigenvalues, 4), "\n")
cat("nonzero eigenvalues:", sum(abs(eigenvalues) > 1e-10), "\n")
```

### Rank-1: a column times a row

One special case shows up often: a **rank-1 matrix** is a single column times a
single row. [@austin2022understandinglinearalgebra, pp. 337--340]

```{webr}
#| autorun: false
# A rank-1 matrix: one column times one row
u <- c(1, 2, 3)
v <- c(4, 5)
R1 <- u %*% t(v)     # 3x2: every row is a multiple of v
R1
qr(R1)$rank           # 1 — only one independent column

# Every row is a scaled copy of v = (4, 5):
R1[1, ] / v    # all 1's
R1[2, ] / v    # all 2's
R1[3, ] / v    # all 3's
```

Every row of $\mathbf{u}\mathbf{v}^\top$ is a scaled copy of $\mathbf{v}^\top$
— there is only one "direction" in the matrix, hence rank 1. This is exactly
the kind of change the
[Sherman–Morrison formula](../sherman-morrison-rank-one-updates/lesson.qmd)
handles: a rank-1 update to a matrix.

## Common traps

**Trap 1 — "Correlated columns are linearly dependent."** Two columns with a
high correlation (say 0.95) point in *nearly* the same direction, but not
exactly. They are linearly independent — neither is an *exact* weighted sum of
the others. The matrix is technically invertible, but it is **ill-conditioned**:
small data changes cause large swings in the inverse. This is the collinearity
problem from [variance structure](../variance-structure-and-standardization/lesson.qmd) —
it makes regression wobbly, but the matrix is not singular. Dependence is an
exact, all-or-nothing property; collinearity is a matter of degree.

```{r trap1-illconditioned}
n <- 6
set.seed(1)
x1 <- scale(rnorm(n))[, 1]
z  <- scale(rnorm(n))[, 1]
z  <- z - sum(z * x1) / sum(x1^2) * x1              # make z orthogonal to x1
x2 <- 0.95 * x1 + sqrt(1 - 0.95^2) * scale(z)[, 1]   # cor(x1, x2) = 0.95 exactly

X <- cbind(x1, x2)
cat("correlation:", round(cor(x1, x2), 2), "  rank:", qr(X)$rank, "\n")
round(solve(t(X) %*% X), 2)
```

Rank 2 — full rank, so `solve()` succeeds — but the diagonal entries of the
inverse come out around **2.05**, versus about **0.2** for two *uncorrelated*
standardized columns of the same size. That is roughly ten times larger from a
correlation of 0.95 alone, with no dependence anywhere in sight. Push the
correlation even closer to 1 and the entries keep growing — the matrix stays
invertible right up to the edge, but the inverse gets less and less trustworthy
as it gets there.

**Trap 2 — "A matrix with more columns than rows always has full rank."**
Actually, a matrix can have at most $\min(n, p)$ independent columns (there are
only $n$ rows, so at most $n$ pivot positions). A 3×5 matrix has rank at most 3,
no matter how many columns it has. And if some columns are redundant, the rank
can be even lower. [@austin2022understandinglinearalgebra, p. 101]

## Check yourself

**1.** The matrix below has 3 columns. Without computing, is it rank 3, 2, or 1?

$$
M = \begin{bmatrix} 1 & 0 & 2 \\ 0 & 1 & -1 \\ 3 & 1 & 5 \end{bmatrix}
$$

*Hint:* check whether column 3 is a combination of columns 1 and 2.

```{webr}
#| autorun: false
# check1-starter
M <- matrix(c(1, 0, 3,
              0, 1, 1,
              2, -1, 5), nrow = 3)
# Check: does 2*col1 + (-1)*col2 equal col3?
2 * M[, 1] + (-1) * M[, 2]
M[, 3]
qr(M)$rank
```

::: {.callout-note collapse="true" title="Answer"}
Column 3 = 2 × column 1 + (−1) × column 2: check entry by entry —
$2(1) + (-1)(0) = 2$ ✓, $2(0) + (-1)(1) = -1$ ✓, $2(3) + (-1)(1) = 5$ ✓.
So the set is dependent and the rank is **2**, which `qr(M)$rank` confirms.
:::

**2.** **Optional (preview — uses eigenvalues from [PCA & SVD](../pca-and-svd/lesson.qmd),
fine to skip on first read).** If a 100×4 data matrix $X$ has rank 3, how many nonzero
eigenvalues does its covariance matrix have? How many principal components could PCA
extract?

::: {.callout-note collapse="true" title="Answer"}
3 nonzero eigenvalues and at most 3 principal components. One of the four
columns is a linear combination of the other three, so there are only three
genuine directions of variation. The covariance matrix inherits that rank, and
PCA can find at most as many components as the rank.
:::

**3.** In your own words: explain to a colleague, in two sentences, what the rank
of a matrix tells you and why it matters.

::: {.callout-note collapse="true" title="A model answer"}
The rank is the number of columns that carry genuinely independent information
— columns that point in their own direction rather than repeating what's already
there. It matters because rank controls whether the matrix is invertible, how
many regression coefficients can be estimated, and how many meaningful PCA
components exist.
:::

## What you can now do

- **Spot a dependent column** by checking whether it is a weighted sum of the
  others, and explain why that makes $X^\top X$ singular.
- **Find the rank** of a matrix using `qr(X)$rank` in R.
- **Connect rank deficiency to collinearity** ("two knobs, one wheel") and to
  PCA (rank caps the number of components).
- **Recognise a rank-1 matrix** as a column times a row — one direction of
  information.

If any of these feels shaky, revisit the
[tiny example](#a-tiny-example-first) (for independence) or the
[Explore section](#explore-it) (for rank and rank-1) before continuing —
later modules assume all of them.

## Where this goes next

You have now seen the *why* behind two ideas that came before: the "singular"
error in [matrices & linear transforms](../matrices-and-linear-transforms/lesson.qmd)
and the collinearity warning in
[variance structure](../variance-structure-and-standardization/lesson.qmd). You
have also met the "rank-1" label that
[Sherman–Morrison](../sherman-morrison-rank-one-updates/lesson.qmd) builds on.
In [PCA & SVD](../pca-and-svd/lesson.qmd), you will see rank control how many
principal components are available — rank is the ceiling on PCA's output.

Ready for more reps? Do the [practice problems](practice.qmd).
