---
title: "Practice: Rank and linear independence"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Do them on paper or in an R session; only
then open the answer.

## Problem 1 — the toy example, new numbers

Three employees have their weekly hours recorded in three columns: regular hours,
overtime hours, and total hours (which is regular + overtime).

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
regular  <- c(35, 40, 38)
overtime <- c(5, 10, 2)
total    <- regular + overtime

X <- cbind(regular, overtime, total)
X

# 1. Is {regular, overtime, total} linearly independent or dependent? Why?
# 2. What is the rank of X?
# 3. Can you invert X'X?
```

::: {.callout-note collapse="true" title="Answer"}
**1.** Dependent — total = 1 × regular + 1 × overtime, so column 3 is a weighted
sum of the other two (weights 1, 1, not all zero).

**2.** Rank 2, because only two columns carry independent information.

**3.** No — $X^\top X$ is singular because $X$ is rank-deficient.

```{r p1-answer}
regular  <- c(35, 40, 38)
overtime <- c(5, 10, 2)
total    <- regular + overtime
X <- cbind(regular, overtime, total)

qr(X)$rank   # 2

tryCatch(solve(t(X) %*% X),
         error = function(e) cat("ERROR:", e$message, "\n"))
```
:::

## Problem 2 — fill in the missing step

You want to check whether three columns $\mathbf{a}$, $\mathbf{b}$, $\mathbf{c}$
are linearly dependent. Here is the approach:

**Step 1:** Set up the test — if $c_1 \mathbf{a} + c_2 \mathbf{b} + c_3 \mathbf{c} = \mathbf{0}$ has a nonzero solution, the set is dependent.

**Step 2:** **Your step here** — what do you actually check in R to determine
whether a nonzero solution exists?

**Step 3:** The rank is 2, so the set is dependent.

```{webr}
#| autorun: false
# p2-starter — fill in Step 2
a <- c(1, 0, 2)
b <- c(0, 1, 1)
cc <- c(2, 1, 5)   # "cc" because "c" is a function in R

M <- cbind(a, b, cc)

# Step 2: what do you compute here?
```

::: {.callout-note collapse="true" title="Answer"}
**Step 2:** Check the rank of the matrix formed by the three columns. If the rank
is less than the number of columns (3), a nonzero solution exists and the set is
dependent.

```{r p2-answer}
a <- c(1, 0, 2)
b <- c(0, 1, 1)
cc <- c(2, 1, 5)
M <- cbind(a, b, cc)

qr(M)$rank   # 2 < 3 columns → dependent

# Verify: cc = 2*a + 1*b
2 * a + 1 * b    # (2, 1, 5) = cc ✓
```

The rank is the key diagnostic: `qr(M)$rank < ncol(M)` means dependent.
:::

## Problem 3 — rank of a data matrix

A 5×3 data matrix records height, weight, and BMI for five people. BMI is
weight / height² (not a linear combination of the other two).

```{webr}
#| autorun: false
# p3-starter — your job: find the rank and explain it
height <- c(1.65, 1.70, 1.75, 1.80, 1.85)   # metres
weight <- c(60, 70, 75, 80, 90)              # kg
bmi    <- weight / height^2

X <- cbind(height, weight, bmi)

# PREDICT: is the rank 1, 2, or 3? Write your guess, then run.
qr(X)$rank
```

::: {.callout-note collapse="true" title="Answer"}
The rank is **3**. BMI is a *nonlinear* function of height and weight
(weight / height²), not a linear combination ($c_1 \cdot \text{height} + c_2 \cdot \text{weight}$).
Linear independence only cares about *weighted sums* — nonlinear functions create
genuinely new columns.

```{r p3-answer}
height <- c(1.65, 1.70, 1.75, 1.80, 1.85)
weight <- c(60, 70, 75, 80, 90)
bmi    <- weight / height^2
X <- cbind(height, weight, bmi)
qr(X)$rank   # 3 — all independent
```
:::

## Problem 4 — spot the error

A colleague says: "I have a 50×4 data matrix where two columns have correlation
0.98. The rank must be 3, because those two columns are basically the same."

This reasoning contains **exactly one mistake**. Find it and fix it.

::: {.callout-note collapse="true" title="Answer"}
The mistake: **high correlation does not make columns linearly dependent.** Two
columns with correlation 0.98 are *nearly* but not *exactly* collinear. Linear
dependence is all-or-nothing: one column must be an *exact* weighted sum of the
others for the rank to drop. With correlation 0.98, the matrix almost certainly
still has rank 4 — it is just **ill-conditioned** (close to singular but not
actually singular).

```{r p4-answer}
set.seed(42)
x1 <- rnorm(50)
x2 <- 0.98 * x1 + sqrt(1 - 0.98^2) * rnorm(50)   # cor ≈ 0.98
x3 <- rnorm(50)
x4 <- rnorm(50)
X <- cbind(x1, x2, x3, x4)

cat("correlation:", round(cor(x1, x2), 3), "\n")
cat("rank:", qr(X)$rank, "\n")   # still 4!
```

The fix: "The columns are highly correlated, so the matrix is **ill-conditioned**
(nearly singular), but the rank is still 4."
:::

## Problem 5 — mixing it with matrices *(uses matrices & linear transforms' `solve()`)*

You have a 3×3 matrix $A$. First check whether it is invertible. If it is,
compute $A^{-1}$ using `solve()`. If it is not, explain why.

```{webr}
#| autorun: false
# p5-starter — edit and Run to solve, right here
A <- matrix(c(1, 2, 3,
              4, 5, 6,
              5, 7, 9), nrow = 3, byrow = TRUE)

# Step 1: check the rank
qr(A)$rank

# Step 2: try to invert (will it work?)
```

::: {.callout-note collapse="true" title="Answer"}
The rank is 2 (less than 3), so $A$ is singular — `solve(A)` will error.

Why: row 3 = row 1 + row 2 ($5 = 1+4$, $7 = 2+5$, $9 = 3+6$), so the
third row is a linear combination of the other two.

```{r p5-answer}
A <- matrix(c(1, 2, 3,
              4, 5, 6,
              5, 7, 9), nrow = 3, byrow = TRUE)

cat("rank:", qr(A)$rank, "\n")  # 2

# Row 3 = Row 1 + Row 2:
A[1, ] + A[2, ]    # (5, 7, 9) = A[3,] ✓

tryCatch(solve(A),
         error = function(e) cat("ERROR:", e$message, "\n"))
```

The rank check (`qr(A)$rank < nrow(A)`) predicts the failure — this is the
connection between linear independence and invertibility from the
[matrices lesson](../matrices-and-linear-transforms/lesson.qmd).
:::

## Problem 6 — rank and PCA *(a gentle cliffhanger)*

A researcher runs PCA on a 200×10 data matrix and finds that the first 3
principal components explain 100% of the variance. The other 7 components
have eigenvalue zero.

What does this tell you about the rank of the data matrix? If you had to
guess, what happened in the data collection?

::: {.callout-note collapse="true" title="Answer"}
The rank is **3**. Only 3 eigenvalues of the covariance matrix are nonzero, so
only 3 columns carry independent information — the other 7 are linear
combinations of those 3.

What probably happened: the 10 variables were not all measured independently.
Perhaps 3 are genuine measurements and the other 7 are derived quantities
(e.g., sums, differences, or rescalings of the original 3). This is the
"rank caps PCA" idea in action — PCA cannot find more directions of variance
than the rank allows, no matter how many columns the data frame has.

The next step would be to inspect the correlation matrix for columns with
$|r| = 1$ and drop the redundant ones before modelling.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/rank-and-linear-independence" pass="4"}

::: {.quiz-q qid="rli-dep-def" qtype="mc" answer="c"}
A set of vectors is **linearly dependent** if:

a) Every vector in the set is a scalar multiple of one other vector
b) The vectors all have the same length
c) One vector can be written as a weighted sum of the others
d) The vectors are uncorrelated
:::

::: {.quiz-q qid="rli-rank-value" qtype="numeric" answer="2" tolerance="0"}
A 4×3 matrix has columns: $\mathbf{a}$, $\mathbf{b}$, and $\mathbf{c} = 3\mathbf{a} - \mathbf{b}$. What is its rank?
:::

::: {.quiz-q qid="rli-max-rank" qtype="numeric" answer="5" tolerance="0"}
What is the maximum possible rank of a 5×8 matrix?
:::

::: {.quiz-q qid="rli-trap" qtype="mc" answer="b"}
Two columns have correlation 0.99. Which statement is correct?

a) The columns are linearly dependent, which means the matrix is singular
b) The columns are linearly independent but the matrix is ill-conditioned
c) The rank of the matrix drops by 1 because the columns are nearly equal
d) PCA will find one fewer component than the number of columns
:::

::: {.quiz-q qid="rli-mm-conformability" qtype="mc" answer="a"}
*(interleaved — uses matrices & linear transforms)* A matrix $A$ is $4 \times 6$ and a
matrix $B$ is $6 \times 2$. What shape is $AB$, and is the product defined?

a) 4×2; defined, because the inner dimensions (6 and 6) match
b) 6×6; defined, because both matrices have a 6 in their shape
c) 4×2; undefined, because the outer dimensions (4 and 2) don't match
d) 2×4; defined, by swapping the order of multiplication
:::

:::

::::
