---
title: "The false discovery rate"
subtitle: "When you run many tests at once, plain p-values lie — how Bonferroni, Holm, and Benjamini–Hochberg keep you honest"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live cells below re-state their own data. Base R only.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```


## Where we're going

A single hypothesis test comes with a clean promise: if nothing is really going on, you'll only
be fooled into shouting "found something!" 5% of the time. But the moment you run *many* tests
at once — one per gene, one per variant, one per pair — that 5% promise quietly breaks. Run 100
dead-boring tests and you'll get about 5 "significant" results *purely by chance*. This lesson
shows why that happens, and the two families of fixes: control the chance of **even one** false
alarm (family-wise: Bonferroni, Holm), or control the **fraction** of your alarms that are
false (false discovery rate: Benjamini–Hochberg). You'll do all of it with R's one-line
`p.adjust`.

**Builds on:**
[hypothesis tests, p-values & confidence intervals](../hypothesis-tests-p-values-and-confidence-intervals/lesson.qmd)
— what a p-value is, the 0.05 convention, and the idea of a false positive (Type I error).

## Warm-up: do you remember?

Two quick reps first — produce each from memory, then check.

**1.** In one sentence: what does a p-value of $0.03$ mean, and would you "reject the null" at
the usual 0.05 threshold?

::: {.callout-note collapse="true" title="Answer"}
It's the chance of seeing a result at least this extreme *if the null (nothing going on) were
true*. Since $0.03 < 0.05$, you **reject** the null at the 0.05 threshold. See
[hypothesis tests, p-values & confidence intervals](../hypothesis-tests-p-values-and-confidence-intervals/lesson.qmd).
:::

**2.** What is a **Type I error** (false positive), and what is the chance of one on a single
test run at the 0.05 threshold when the null is true?

::: {.callout-note collapse="true" title="Answer"}
A Type I error is a **false alarm** — rejecting the null when it's actually true. On one test at
the 0.05 threshold, that chance is exactly $0.05$. See
[hypothesis tests, p-values & confidence intervals](../hypothesis-tests-p-values-and-confidence-intervals/lesson.qmd).
:::

## A tiny example first

Imagine you test **4** completely inert things — four sugar pills, none of which does anything.
Each test independently has a 5% chance of a false alarm. What's the chance that *at least one*
of the four falsely lights up?

Work it in prose. The chance a single test does **not** false-alarm is $0.95$. For all four to
stay quiet, that must happen four times over: $0.95 \times 0.95 \times 0.95 \times 0.95 =
0.95^4 \approx 0.8145$. So the chance that **at least one** shouts "found it!" is
$1 - 0.8145 = 0.1855$ — about **19%**, not 5%. With just four tests your false-alarm rate has
nearly quadrupled.

The same thing in R:

```{r toy-example}
# 4 inert tests, each with a 5% false-alarm chance
m <- 4
per_test_ok <- 0.95          # chance one test stays quiet

# All four stay quiet
all_quiet <- per_test_ok^m
all_quiet                    # ~0.8145

# At least one false alarm = 1 - (all quiet)
at_least_one <- 1 - all_quiet
at_least_one                 # ~0.1855, i.e. ~19%
```

Push $m$ to 100 and the chance of at least one false alarm is $1 - 0.95^{100} \approx 0.994$ —
virtually certain. That is the multiple-testing problem in one number.

## Getting a feel for it

**Analogy — a metal detector at the beach.** One sweep of a cheap detector beeps by mistake
now and then; annoying but rare. Now sweep the *whole beach*, thousands of sweeps: you're
guaranteed a pile of false beeps even if there's no treasure anywhere. You have two honest ways
to respond. (1) Make the detector far less twitchy so that across the *entire day* you'd expect
essentially zero false beeps — you'll miss some real treasure, but every beep is trustworthy.
That's **family-wise error control** (Bonferroni/Holm). (2) Accept that some beeps will be
false, but insist that *among the spots you actually dig*, only a small fraction are duds —
you'll dig more holes and find more treasure, tolerating a controlled share of wasted digs.
That's the **false discovery rate** (Benjamini–Hochberg).

Where the analogy breaks: real tests aren't always independent (nearby variants are
correlated), and the exact guarantees of each method depend on that — but the *spirit* (few
false beeps ever, vs. few false beeps among your digs) is exactly right.

**Why multiply p-values (or divide the threshold)?** Because "at least one false alarm among
$m$ tries" grows roughly like $m$ times the per-test rate when that rate is small
($1 - 0.95^m \approx 0.05\,m$ for small $m$: $0.05\times 4 = 0.20$, close to our $0.19$). So to
pull the *whole-family* false-alarm rate back down to 0.05, the crudest fix is to demand each
individual test clear a threshold $m$ times stricter — $0.05/m$. That's Bonferroni. Everything
else is a cleverer version of the same instinct.

**What does 19% feel like?** It's the difference between "I'm probably fine" and "I'll likely be
wrong at least once." Drop to $m = 2$ tests and it's only $1 - 0.95^2 \approx 9.75\%$; climb to
$m = 20$ and it's $1 - 0.95^{20} \approx 64\%$. The number races toward certainty as tests pile
up.

## What the notation is going to say

The corrections all *transform your raw p-values into adjusted ones you can still compare to
0.05*. Bonferroni's adjusted p-value will say: "multiply each raw p by the number of tests
$m$ (capped at 1)" — so our smallest inert p of, say, $0.03$ becomes $0.03 \times 4 = 0.12$,
no longer significant. Benjamini–Hochberg's will say: "rank the p-values, and scale each by
$m$ divided by its rank, so the smallest p is judged most strictly and later ones more
leniently." In R, every one of these is a `method=` inside `p.adjust`.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $m$ | "em" | the number of tests you ran | $4$ |
| $p_i$ | "p sub eye" | the raw p-value of the $i$-th test | one per test |
| $p_{(i)}$ | "p paren eye" | the $i$-th **smallest** p-value (p-values sorted low to high) | — |
| $\alpha$ | "alpha" | the false-alarm threshold you're willing to accept | $0.05$ |
| FWER | "eff-double-you-ee-arr" | family-wise error rate: chance of **at least one** false alarm across all tests | ~$0.19$ uncorrected |
| FDR | "eff-dee-arr" | false discovery rate: expected **fraction** of your rejections that are false | — |
| $\tilde{p}_i$ | "p sub eye tilde" | the *adjusted* p-value, comparable to $\alpha$ after correction (in R: `p.adjust(p, ...)`) | — |
| $\tilde{p}_{(i)}$ | "p paren eye tilde" | the BH-adjusted p-value of the $i$-th smallest raw p-value | — |

## The formal version

Two questions, two families of answers. Start from $m$ raw p-values $p_1, \dots, p_m$.

**Family-wise error control (few false alarms EVER).**

*Bonferroni* multiplies each raw p-value by $m$ (and caps at 1):

$$
\tilde{p}_i = \min(1,\ m \cdot p_i).
$$

Read aloud: "adjusted-p equals the smaller of 1 and m-times-raw-p." Everyday words: "make every
test $m$ times harder to call significant." Equivalently, reject only when $p_i < \alpha/m$.

*Holm* is a uniformly better version: sort the p-values, and multiply the smallest by $m$, the
next by $m-1$, the next by $m-2$, … (with a running-maximum so adjusted values never decrease).
It controls the same family-wise error but rejects at least as much as Bonferroni — strictly
more powerful, never worse. **This is R's `p.adjust` default.**

**False discovery rate control (few false alarms among your DIGS).**

*Benjamini–Hochberg* sorts the p-values smallest to largest and compares the $i$-th smallest,
$p_{(i)}$, to a *sloped* threshold $\frac{i}{m}\alpha$:

$$
\text{reject the } i\text{-th if } \quad p_{(i)} \le \frac{i}{m}\,\alpha .
$$

Read aloud: "reject the i-th smallest p-value if it's at most i-over-m times alpha." Everyday
words: "the very smallest p-value must clear the strictest bar ($\alpha/m$, same as
Bonferroni), but each successive one gets a proportionally easier bar." Its adjusted p-value is
$\tilde{p}_{(i)} = \min(1,\ \frac{m}{i}\,p_{(i)})$ (with a running-minimum so they stay
sorted). Because most bars are looser than Bonferroni's single strict bar, BH rejects *more* —
it finds more real effects, at the cost of a controlled fraction of false ones.

**Mapping it onto the toy example.** Suppose the four inert tests returned raw p-values
$(0.03, 0.20, 0.51, 0.78)$ — the $0.03$ is a chance false alarm.

- Bonferroni: $0.03 \times 4 = 0.12 > 0.05$ → **not** significant. The false alarm is caught.
- Holm: smallest is $0.03 \times 4 = 0.12 > 0.05$ → stop; nothing significant. Also caught.
- BH: smallest bar is $\frac{1}{4}(0.05) = 0.0125$; since $0.03 > 0.0125$, the $0.03$ doesn't
  clear even its own bar → nothing significant. Caught too.

All three refuse to be fooled by the lone $0.03$. Let's confirm with `p.adjust`:

```{r formal-check}
p <- c(0.03, 0.20, 0.51, 0.78)   # four inert tests; the 0.03 is a chance false alarm

# Bonferroni: multiply by m = 4, cap at 1
p.adjust(p, method = "bonferroni")   # 0.12 0.80 1.00 1.00

# Holm (p.adjust's DEFAULT): stepwise family-wise control
p.adjust(p)                          # same verdict here: smallest is 0.12

# Benjamini-Hochberg: false discovery rate
p.adjust(p, method = "BH")           # smallest adjusted is 0.12 as well

# None drop below 0.05 -> none declared significant. Good.
any(p.adjust(p) < 0.05)              # FALSE
```

## Explore it

```{webr}
#| autorun: true
# TRY IT: change m and re-run. PREDICT: at m = 50, is the chance of >=1 false alarm closer to
# 0.5 or to 0.9?
m <- c(1, 2, 4, 10, 20, 50, 100)
fwer <- 1 - 0.95^m                        # chance of at least one false alarm among m inert tests
data.frame(m = m, chance_of_a_false_alarm = round(fwer, 3))
# Watch it climb from 0.05 toward near-certainty as m grows
```

```{webr}
#| autorun: true
#| fig-alt: "A line rising from about 0.05 at one test toward nearly 1.0 by 100 tests, showing the family-wise false-alarm rate climbing as the number of inert tests grows; a dashed horizontal line marks the intended 0.05."
# TRY IT: change the 0.05 per-test rate to 0.01 and re-run — the whole curve drops.
m <- 1:100
plot(m, 1 - 0.95^m, type = "l", lwd = 2, ylim = c(0, 1),
     xlab = "number of inert tests (m)",
     ylab = "chance of at least one false alarm",
     main = "Why plain p-values fail with many tests")
abline(h = 0.05, lty = 2, col = "tomato")
legend("bottomright", legend = "intended 0.05", lty = 2, col = "tomato", bty = "n")
```

```{webr}
#| autorun: true
# TRY IT: a mix of real effects (tiny p's) and noise. PREDICT: which method flags MORE tests,
# Holm or BH? (BH is more lenient, so it flags more.)
set.seed(8)
# 5 real effects (small p) buried among 45 nulls (uniform p)
p <- c(runif(5, 0, 0.001), runif(45, 0, 1))
cat("Uncorrected significant (p<0.05):", sum(p < 0.05), "\n")
cat("Holm significant:                ", sum(p.adjust(p, "holm") < 0.05), "\n")
cat("BH significant:                  ", sum(p.adjust(p, "BH")   < 0.05), "\n")
# Uncorrected over-calls (6); Holm is strictest (5); BH recovers one more (6)
```

## Common traps

**Trap: "each test is at 0.05, so I'm safe."** Tempting, because 0.05 *is* safe for one test.
But with $m = 4$ inert tests the chance of at least one false alarm is $1 - 0.95^4 \approx 0.19$,
not 0.05 — the per-test guarantee says nothing about the *family*. The whole point of the
correction is that "0.05 each" and "0.05 overall" are different promises.

**Trap: "FDR and family-wise control are the same thing."** No — they control different
quantities. Family-wise (Bonferroni/Holm) caps the chance of **even one** false alarm across
all tests; FDR (BH) caps the **fraction** of your *rejections* that are false. Suppose you make
20 rejections and 2 are false: your false *discovery* fraction is $2/20 = 10\%$ (BH would allow
this at $\alpha = 0.10$), but you *did* make false alarms, so the family-wise error occurred.
BH deliberately tolerates some false alarms to find more true effects; Holm refuses almost all
false alarms at the cost of missing weaker true ones. Pick the one that matches your cost of
being wrong.

## Check yourself

**1.** You run $10$ inert tests, each at the 0.05 threshold. Roughly what's the chance of at
least one false alarm?

```{webr}
#| autorun: false
# check1-starter — edit and Run to check
m <- 10
```

::: {.callout-note collapse="true" title="Answer"}
$1 - 0.95^{10} \approx 0.40$ — about a **40%** chance of at least one false alarm. In R:
`1 - 0.95^10`. Ten independent shots at a 5% mistake add up fast.
:::

**2.** Raw p-values $(0.001, 0.30, 0.60)$ from $m = 3$ tests. What are the Bonferroni-adjusted
values, and which are significant at 0.05?

```{webr}
#| autorun: false
# check2-starter — edit and Run to check
p <- c(0.001, 0.30, 0.60)
```

::: {.callout-note collapse="true" title="Answer"}
Multiply each by $3$ (cap at 1): $(0.003, 0.90, 1.00)$. Only the first, $0.003$, is below 0.05,
so **one** test survives correction. In R: `p.adjust(c(0.001, 0.30, 0.60), "bonferroni")`.
:::

**3.** In your own words: explain to a colleague, in two sentences, the difference between
controlling the family-wise error rate and controlling the false discovery rate.

::: {.callout-note collapse="true" title="A model answer"}
Family-wise control (Bonferroni, Holm) keeps the chance of making *even a single* false alarm
across all your tests small — great when any false positive is costly. False-discovery-rate
control (Benjamini–Hochberg) instead keeps the *fraction* of your flagged results that are false
small, which lets you flag more real effects when you can tolerate a known share of duds among
them.
:::

## What you can now do

- **Explain why plain 0.05 p-values fail** under many tests, and compute the family-wise
  false-alarm chance $1 - (1-\alpha)^m$.
- **Apply Bonferroni and Holm** (multiply/step-down the p-values) to control the chance of any
  false alarm, and know Holm is `p.adjust`'s default.
- **Apply Benjamini–Hochberg** to control the *fraction* of false rejections, and say when you'd
  prefer it over family-wise control.
- **Use `p.adjust(p, method = ...)`** in R and read its output back against a 0.05 threshold.

If any of these feels shaky, revisit the toy example and the "what the notation says" section —
methods that run many tests and correct over the total number of possible tests assume all of them.

## Where this goes next

Any method that runs a *pile* of tests and then asks "is any of them real?" has to correct for
multiplicity, or it drowns in false alarms. A common pattern in applied courses is to run *all
pairwise* tests over a set of items and correct over the total number of possible pairs with
exactly this `p.adjust` machinery — and to watch for a subtlety worth knowing: some released
software names an output column "fdr" while actually calling `p.adjust`'s Holm default, which is
why knowing both the FWER and FDR families matters.

Ready for more reps? Do the [practice problems](practice.qmd).
