---
title: "Practice: The false discovery rate"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

Work these in order — they get harder. Do them on paper or in an R session; only then open the
answer.

## Problem 1 — the toy example, new numbers

You run **5** completely inert tests, each at the 0.05 threshold. By hand, what's the chance
that at least one false-alarms? Then check in R.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
m <- 5
```

::: {.callout-note collapse="true" title="Answer"}
Chance all five stay quiet: $0.95^5 \approx 0.7738$. Chance of at least one false alarm:
$1 - 0.7738 = 0.2262$, about **23%**.

```{r p1-ans}
1 - 0.95^5   # ~0.226
```
:::

## Problem 2 — fill in the missing step

Raw p-values $(0.004, 0.02, 0.40)$ from $m = 3$ tests. Bonferroni adjustment, step by step:

- Step 1: the multiplier is $m = 3$.
- Step 2: **your step here** — form the adjusted values (multiply each raw p by 3, cap at 1).
- Step 3: compare to 0.05; the significant ones are those below 0.05.

Fill in step 2 and give the final verdict.

```{webr}
#| autorun: false
# p2-starter — edit and Run to check the blank
p <- c(0.004, 0.02, 0.40)
```

::: {.callout-note collapse="true" title="Answer"}
Step 2: $(0.004\times 3,\ 0.02\times 3,\ 0.40\times 3) = (0.012,\ 0.06,\ 1.00)$ (the last
capped from $1.2$ to $1$). Step 3: only $0.012 < 0.05$, so **only the first test** is
significant after Bonferroni.

```{r p2-ans}
p.adjust(c(0.004, 0.02, 0.40), "bonferroni")   # 0.012 0.060 1.000
```
Note $0.02$ was significant on its own but does **not** survive correcting for 3 tests.
:::

## Problem 3 — Holm vs Bonferroni

For raw p-values $(0.01, 0.02, 0.04)$ from $m = 3$ tests, compute both the Bonferroni and Holm
adjusted values in R, and say which method flags more tests at 0.05.

```{webr}
#| autorun: false
# p3-starter — your job: compare p.adjust with "bonferroni" and "holm"
p <- c(0.01, 0.02, 0.04)
```

::: {.callout-note collapse="true" title="Answer"}
```{r p3-ans}
p <- c(0.01, 0.02, 0.04)
rbind(bonferroni = p.adjust(p, "bonferroni"),
      holm       = p.adjust(p, "holm"))
```
Bonferroni gives $(0.03, 0.06, 0.12)$ → only the first is significant. Holm gives
$(0.03, 0.04, 0.04)$ → the first **two** are significant. Holm multiplies the smallest by 3,
the next by 2, the next by 1 (with a running max), so it is uniformly at least as powerful as
Bonferroni while controlling the same family-wise error.
:::

## Problem 4 — spot the error

A colleague ran $m = 20$ tests and wants to control the chance of *any* false alarm. They wrote
— it contains **exactly one mistake**. Find it and fix it.

```{webr}
#| autorun: false
# p4-starter — this contains exactly one mistake
p <- runif(20)              # 20 raw p-values
# "Each test is already at 0.05, so I'm safe to just use the raw values."
significant <- p < 0.05
sum(significant)
```

::: {.callout-note collapse="true" title="Answer"}
The mistake is **not correcting at all**. Comparing 20 raw p-values to 0.05 gives a family-wise
false-alarm chance of $1 - 0.95^{20} \approx 0.64$ — about a 64% chance of at least one false
positive. To control the chance of *any* false alarm you must adjust first:

```{r p4-ans}
set.seed(1); p <- runif(20)
sum(p.adjust(p, "holm") < 0.05)   # correct: family-wise controlled
```
:::

## Problem 5 — mixing it with hypothesis tests *(uses hypothesis-tests-p-values-and-confidence-intervals' Type I error)*

You test 100 variants; suppose **none** are truly associated (all nulls true). Using what a
Type I error rate means, predict how many raw p-values you'd expect below 0.05 by chance, then
confirm by simulation.

```{webr}
#| autorun: false
# p5-starter — edit and Run to solve, right here
set.seed(3)
p <- runif(100)     # 100 nulls: p-values are uniform on [0,1] under the null
```

::: {.callout-note collapse="true" title="Answer"}
Under a true null, a p-value is uniform on $[0,1]$, so the chance any one falls below 0.05 is
exactly 0.05 — that's the Type I error rate. Across 100 nulls you expect about
$100 \times 0.05 = 5$ false positives.

```{r p5-ans}
set.seed(3); p <- runif(100)
sum(p < 0.05)                    # about 5 by chance
sum(p.adjust(p, "holm") < 0.05)  # 0 after correction: the false alarms are removed
```
Roughly 5 raw p-values dip below 0.05 with nothing real going on — exactly the false alarms
correction exists to catch.
:::

## Problem 6 — transfer: read an FDR back as a sentence

An analysis flags 40 genes as significant and reports an FDR (BH) of 0.10. In plain words, what
does that 0.10 promise, and roughly how many of the 40 might be false?

::: {.callout-note collapse="true" title="Answer"}
An FDR of 0.10 means that, *among the genes you flagged*, an expected 10% are false discoveries
— so of the 40 flagged, roughly $40 \times 0.10 = 4$ may be duds. It does **not** say the chance
of any false alarm is 10% (that's family-wise); it says the *fraction* of your flags that are
wrong is controlled at 10% on average. You accept ~4 likely-wrong flags in exchange for
catching more of the real signals.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**, once the
material isn't fresh. A pass then is real evidence you've retained it. Five questions, graded on
the page; 4/5 or better means this module is solid. Results are saved in this browser only.

::: {.module-check module="foundations/false-discovery-rate" pass="4"}

::: {.quiz-q qid="fdr-fwer-toy" qtype="numeric" answer="0.19" tolerance="0.01"}
For $m = 4$ inert tests each at the 0.05 threshold, what is the chance of at least one false
alarm, $1 - 0.95^4$? (Enter to 2 decimals.)
:::

::: {.quiz-q qid="fdr-bonferroni-move" qtype="mc" answer="b"}
Bonferroni's adjusted p-value for the $i$-th test is:

a) the raw p-value divided by $m$
b) the smaller of 1 and $m$ times the raw p-value
c) the raw p-value minus $0.05$
d) always $0.05/m$
:::

::: {.quiz-q qid="fdr-vs-fwer" qtype="mc" answer="c"}
The false discovery rate (Benjamini–Hochberg) controls:

a) the chance of making even a single false alarm across all tests
b) the total number of tests you may run
c) the expected fraction of your rejections that are false
d) the size of each individual p-value
:::

::: {.quiz-q qid="fdr-padjust-default" qtype="mc" answer="a"}
R's `p.adjust(p)` with no `method=` argument uses which correction by default?

a) Holm
b) Benjamini–Hochberg
c) Bonferroni
d) none — it returns the raw p-values
:::

::: {.quiz-q qid="fdr-typeI-interleave" qtype="numeric" answer="5" tolerance="0"}
*(uses hypothesis-tests-p-values-and-confidence-intervals)* If you run 100 tests where every
null is true, about how many raw p-values do you expect below 0.05 purely by chance
($100 \times 0.05$)?
:::

:::

::::
