---
title: "Reading lm() output — and the same model in papers"
subtitle: "Decode every number summary() prints, then recognize it in a journal table"
---

```{r setup}
#| message: false
# Packages this lesson uses. Missing ones are installed automatically.
needed <- c()   # base R is all we need
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

## Where we're going

You can already produce the `Estimate` column of `summary(lm())` by hand. This
module decodes the rest — `Std. Error`, `t value`, `Pr(>|t|)`, `R-squared`,
`Residual standard error` — and then shows how the identical numbers dress up in an
academic paper: a model equation with $\varepsilon_i$, and a coefficient table with
parentheses and stars. After this, a regression table in a paper is something you
*read*, not skip.

**Builds on:** [module 02](../02-deriving-the-ols-formulas/lesson.qmd) — plus the
Greek/Latin/hat letter code from
[mean, variance & covariance](../../../../foundations/mean-variance-covariance/lesson.qmd).

## Warm-up: do you remember?

Pull these back before the new columns — you'll lean on all three. Produce each
answer before opening it.

**1.** *(module 2)* The `Estimate` column you can already produce. Say
$\hat{\beta}_1 = s_{xy}/s_x^2$ in plain words — what is the slope, in one sentence?

::: {.callout-note collapse="true" title="Answer"}
"Beta-one hat equals the covariance of x and y over the variance of x" — the slope is
how much x and y move together, per unit of how much x moves on its own. In R:
`cov(x, y) / var(x)`, which matches `coef(lm(y ~ x))[2]`. See
[module 2](../02-deriving-the-ols-formulas/lesson.qmd).
:::

**2.** *(mean, variance & covariance)* The letter code: of $\mu$ and $\bar{x}$, which
is the *true but unknown* quantity and which is the *estimate from data*? Why does it
matter here?

::: {.callout-note collapse="true" title="Answer"}
$\mu$ (Greek) is the true, unknown population value; $\bar{x}$ (Latin) is what we
compute from the sample to estimate it. This module lives on that gap: $\beta_1$ is
the unknown truth, $\hat{\beta}_1$ is our estimate, and the standard error measures
how far the estimate might bounce from the truth. See
[mean, variance & covariance](../../../../foundations/mean-variance-covariance/lesson.qmd).
:::

**3.** *(module 1)* What is a residual $e_i$, in symbols and in words?

::: {.callout-note collapse="true" title="Answer"}
$e_i = y_i - \hat{y}_i$ — actual minus predicted, the line's miss on point $i$. The
"residual standard error" this module decodes is just the typical size of these
misses. See [module 1](../01-least-squares-idea/lesson.qmd).
:::

## A tiny example first

The new idea behind ALL the remaining columns is: **your estimate would have come
out differently with different data.** Let's *see* that.

Pretend, just for this game, that we secretly know the truth: tomatoes really
follow $y = 1 + 2x + \text{noise}$ — intercept 1, slope 2, plus random luck each
season. We grow 5 plants at $x = (1, 2, 3, 4, 5)$ and run three seasons:

- Season 1 luck: $(+1, -1, 0, +1, -1)$ → harvest $y = (4, 4, 7, 10, 10)$.
  Fit the line (module 2 formulas): slope $\hat{\beta}_1 = 1.8$.
- Season 2 luck: $(-1, 0, +1, -1, +1)$ → $y = (2, 5, 8, 8, 12)$. Slope: $2.3$.
- Season 3 luck: $(0, +1, -1, +1, 0)$ → $y = (3, 6, 6, 10, 11)$. Slope: $2.0$.

Same true slope (2) every time — but the *estimates* came out 1.8, 2.3, 2.0. The
estimate is a lottery ticket: mostly near the truth, never exactly it. The
**standard error** is the typical size of that bounce, estimated from the
one sample you actually have.

The same thing in R, so you can check every step:

```{r toy-example}
# The secret truth (only simulations get to know it): y = 1 + 2x + noise
x <- c(1, 2, 3, 4, 5)

season1 <- c(4, 4, 7, 10, 10)    # truth + luck (+1,-1,0,+1,-1)
season2 <- c(2, 5, 8, 8, 12)     # truth + luck (-1,0,+1,-1,+1)
season3 <- c(3, 6, 6, 10, 11)    # truth + luck (0,+1,-1,+1,0)

# The slope estimate from each season — covariance over variance, as derived
c(cov(x, season1) / var(x),      # 1.8
  cov(x, season2) / var(x),      # 2.3
  cov(x, season3) / var(x))      # 2.0  — bouncing around the true 2
```

## Getting a feel for it

**Your estimate is a lottery ticket.** Each season of data is one draw from the same
machine, and the slope you compute from it is the ticket that draw hands you: 1.8,
then 2.3, then 2.0, all circling the true 2 you never get to see directly. Run the
season once more and you'd get yet another ticket. The **standard error** is just the
typical spread of those tickets — how far a single ticket usually lands from the
middle of the pile. (Where the picture bends: a real study buys only *one* ticket,
so the standard error is itself *estimated* from that lone draw, not measured by
actually re-running.)

**Why some estimates bounce more than others.** Two things widen the spread of
tickets, and both make plain sense. More *noise* in the harvest — bigger random luck
each season — jostles every estimate harder, so the tickets scatter wider. And fewer
data points, or fertilizer doses bunched close together, give the line less to brace
against, so it wobbles more; spreading the $x$'s out and adding plants both pin the
line down and pull the tickets into a tighter pile. That's the whole content of the
standard-error formula before we write a single symbol: *noise pushes the spread up,
and well-anchored, plentiful data pushes it down.*

**What the number feels like.** A small standard error means "if I'd run a different
season, I'd have landed in nearly the same place" — the estimate is trustworthy. A
large one means "easily could have come out quite different" — hold it loosely. The
histogram in "Explore it" below draws the whole invisible pile of tickets you could
have drawn, so you can *see* the spread the standard error is trying to summarize
from your single draw.

## What the notation is going to say

Papers describe this exact game in one line: "each y is the line plus a random
error; the errors are independent draws centered on zero." Then every column of the
output answers a question about the lottery: *how big is the bounce?* (standard
error), *how many bounce-lengths from zero is my estimate?* (t value), *if the true
slope were zero, how often would luck alone produce an estimate this far out?*
(p-value), and *how much of y's variation does the line soak up?* ($R^2$).

## How to read the symbols

Reused from earlier modules (they appear in this lesson's formulas, so they get a row
here too — season 1's data $x = (1,2,3,4,5)$, $y = (4,4,7,10,10)$ as the anchor):

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $x_i$ | "ex sub eye" | the i-th predictor value — fertilizer scoops (in R: `x[i]`) | 1, 2, 3, 4, 5 |
| $y_i$ | "why sub eye" | the i-th observed outcome — a season's harvest (in R: `y[i]`) | 4, 4, 7, 10, 10 |
| $\bar{y}$ | "why bar" | the mean of the y values (in R: `mean(y)`) | 7 |
| $\beta_0, \beta_1$ | "beta zero, beta one" | the TRUE (unknown) intercept and slope — no hats | 1 and 2 in our game |
| $\hat{\beta}_0, \hat{\beta}_1$ | "beta-zero hat, beta-one hat" | the fitted estimates from one sample ($s_{xy}/s_x^2$, module 2) | season 1: 1.6 and 1.8 |
| $\hat{y}_i$ | "why hat sub eye" | the fitted line's prediction for point i | season 1: 3.4, 5.2, 7, 8.8, 10.6 |
| $e_i$ | "ee sub eye" / "residual i" | the leftover miss $y_i - \hat{y}_i$ (module 1) | season 1: 0.6, −1.2, 0, 1.2, −0.6 |
| $\mathrm{SSE}$ | "sum of squared errors" | the fitted line's total squared miss, $\sum e_i^2$ (module 1) | season 1: 3.6 |

New in this lesson:

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $\varepsilon_i$ | "epsilon sub eye" | the random error: the part of $y_i$ the line doesn't explain | season 1's luck: +1, −1, 0, +1, −1 |
| $\sim$ | "is distributed as" | the left side follows the distribution on the right | $\varepsilon_i \sim N(0, \sigma^2)$ |
| $N(0, \sigma^2)$ | "normal with mean zero and variance sigma-squared" | a bell curve centered at 0 with spread $\sigma$ | luck averaging out to 0 |
| $\sigma$ | "sigma" | the true typical error size (unknown) | 1-ish in our game |
| $\hat{\sigma}$ | "sigma hat" (R: `Residual standard error`) | the estimate of $\sigma$ from the residuals | ≈ 0.79 below |
| $\mathrm{SE}(\hat{\beta}_1)$ | "the standard error of beta-one hat" | the typical bounce of the slope estimate across hypothetical re-runs | ≈ 0.25 below |
| $t$ | "the t value" | estimate ÷ its standard error: how many bounce-lengths from zero | ≈ 7.7 below |
| $p$ or $\Pr(>\|t\|)$ | "the p-value" | if the true coefficient were 0, the chance of an estimate ≥ this many bounce-lengths out | ≈ 0.005 below |
| $R^2$ | "R squared" | the fraction of y's variance the line explains, from 0 to 1 | ≈ 0.95 below |
| $H_0$ | "the null hypothesis" | the boring default being tested — here "true slope is 0" | "fertilizer does nothing" |

## The formal version

**The model, as papers write it:**

$$
y_i = \beta_0 + \beta_1 x_i + \varepsilon_i, \qquad \varepsilon_i \sim N(0, \sigma^2)
$$

Reading it aloud: "why-sub-eye equals beta-zero plus beta-one ex-sub-eye plus
epsilon-sub-eye, where the epsilons are normal with mean zero and variance
sigma-squared." In everyday words: *each harvest is the line's value plus that
season's luck, and luck is bell-curve noise centered on zero.* This is exactly the
game we played — $\beta_0 = 1$, $\beta_1 = 2$, luck as $\varepsilon_i$. When a
paper writes this equation, it is declaring which lottery it believes generated the
data. (Note it uses hatless Greek $\beta$s: the *true* line, not the estimate.)

**The columns of `summary(lm())`, one question each:**

$$
t = \frac{\hat{\beta}_1 - 0}{\mathrm{SE}(\hat{\beta}_1)}
\qquad\qquad
R^2 = 1 - \frac{\mathrm{SSE}}{\sum_{i=1}^n (y_i - \bar{y})^2}
$$

- $\mathrm{SE}(\hat{\beta}_1)$ — the estimated bounce. (Its own formula is
  $\hat{\sigma}/\sqrt{\sum(x_i-\bar{x})^2}$ — spread of the errors over spread of
  the x's: noisier data → bouncier estimates; wider-spread x → steadier ones.)
- $t$: "the t value equals the estimate minus zero, over its standard error" —
  distance from zero measured in bounce-lengths. Rule of thumb: beyond about ±2
  bounce-lengths is hard to explain by luck.
- $p$: the probability, *assuming the true coefficient is zero* ($H_0$), of luck
  producing an estimate at least this many bounce-lengths out. Small p → "luck
  alone rarely does this" → the stars in papers.
- $R^2$: "one minus SSE over the total sum of squares" — compare the line's leftover
  error to the error of the do-nothing flat line at $\bar{y}$ (module 1's broken
  candidate!). Soaking up all the error → $R^2 = 1$; doing no better than the flat
  line → $R^2 = 0$.

**The same model in a paper's table** (this is the payoff — the format to
recognize):

> | | (1) Tomatoes |
> |---|---|
> | Fertilizer | 1.926** |
> | | (0.249) |
> | Constant | 1.662 |
> | | (0.827) |
> | Observations | 5 |
> | R² | 0.952 |
>
> *Notes:* Standard errors in parentheses. \*p < 0.05, \*\*p < 0.01, \*\*\*p < 0.001.

(These are exactly the numbers `summary()` prints in the Explore section below —
same fit, journal costume.) Same numbers, new costume: each cell is $\hat{\beta}$ with $\mathrm{SE}(\hat{\beta})$
in parentheses *below* it (some journals use brackets, or t values instead — the
table notes always say). Stars encode the p-value thresholds. Divide any estimate by
its parenthesis and you've reconstructed the t value yourself.

## Explore it

```{r explore-1}
# TRY IT: the full decode, on data where WE set the truth. Change the true slope
# or sigma and re-run — watch every column respond.
# PREDICT: the true slope is 2. Will the Estimate row read exactly 2.000? Guess,
# then find the slope line in the summary() output below.
set.seed(42)
n <- 5
x <- c(1, 2, 3, 4, 5)
y <- 1 + 2 * x + rnorm(n, sd = 1)     # beta0 = 1, beta1 = 2, sigma = 1

fit <- lm(y ~ x)
summary(fit)
```

```{r explore-2}
# Rebuild each printed number from raw ingredients — nothing is magic:
est   <- coef(fit)[2]                          # beta-one hat: cov/var
resid <- y - fitted(fit)                       # the leftovers e_i
sigma_hat <- sqrt(sum(resid^2) / (n - 2))      # residual standard error*
se    <- sigma_hat / sqrt(sum((x - mean(x))^2))# SE: error spread / x spread
tval  <- est / se                              # bounce-lengths from zero
r2    <- 1 - sum(resid^2) / sum((y - mean(y))^2)

round(c(estimate = unname(est), SE = se, t = unname(tval), R2 = r2), 4)
# compare with the summary() printout above — identical.
# (*n - 2 because two parameters were estimated; same story as var()'s n - 1.)
```

```{r explore-3}
#| fig-alt: "Histogram of slope estimates across 2000 simulated re-runs, with width matching the standard error"
# TRY IT: is SE really "the typical bounce"? Run the lottery 2000 times and
# check. Increase sd or decrease n — the histogram widens and SE tracks it.
# PREDICT: will the spread (SD) of the 2000 re-run slopes come out close to the SE
# computed from the SINGLE sample above? Guess, then compare the two numbers in
# the plot title.
set.seed(42)
slopes <- replicate(2000, {
  y_new <- 1 + 2 * x + rnorm(n, sd = 1)     # a fresh season each time
  cov(x, y_new) / var(x)                    # its slope estimate
})

hist(slopes, breaks = 40, col = "steelblue",
     main = sprintf("SD of 2000 re-run slopes: %.3f  |  SE from ONE sample: %.3f",
                    sd(slopes), se))
abline(v = 2, col = "firebrick", lwd = 2)   # the truth
```

That histogram is the deepest idea in the module: the standard error printed from
your *single* dataset estimates the spread of a whole histogram you could never
actually observe. (With only n = 5 the two numbers agree roughly, not perfectly —
the SE is itself an estimate, and small samples make everything bouncy. Re-run with
`x <- 1:50` and watch them converge.)

## Common traps

**Trap 1 — reading the p-value as "the chance the effect is nothing."** Our toy fit
has $p \approx 0.005$. It is tempting to say "so there's a 0.5% chance fertilizer does
nothing." That's backwards. The p-value is computed *assuming the true slope is
already zero* ($H_0$) — it's the chance that **luck alone** would produce an estimate
this many bounce-lengths out, IF the null were true. It says nothing directly about
how probable the null itself is. Correct reading: "if fertilizer truly did nothing,
data this extreme would show up only about 0.5% of the time — so 'nothing' is a poor
explanation."

**Trap 2 — "more stars = bigger effect."** Since $t = \hat{\beta}/\mathrm{SE}$, the t
value (and its stars) grows when the estimate is large *or* when the standard error
is small — and SE shrinks with more data and less noise. So a microscopically small
slope, measured very precisely in a huge sample, can earn three stars. Stars measure
*confidence the effect isn't zero*, not the *size* or importance of the effect. To
judge size, read the estimate itself (in its real units); to judge how much the
predictor explains, read $R^2$.

## Check yourself

**1.** A paper's table shows `Fertilizer  1.84 (0.23)`. Reconstruct the t value and
say, in plain English, what it means.

::: {.callout-note collapse="true" title="Answer"}
$t = 1.84 / 0.23 = 8$. The slope estimate is 8 bounce-lengths away from zero — if
fertilizer truly did nothing, luck essentially never produces an estimate that far
out, so the paper will report it as strongly significant (and star it accordingly).
:::

**2.** Read aloud, then unpack in everyday words:
$y_i = \beta_0 + \beta_1 x_i + \varepsilon_i,\; \varepsilon_i \sim N(0, \sigma^2)$.
Which symbols would get hats in the fitted version?

::: {.callout-note collapse="true" title="Answer"}
"Why-sub-eye equals beta-zero plus beta-one ex-sub-eye plus epsilon-sub-eye, where
epsilon-sub-eye is distributed normal with mean zero and variance sigma-squared."
Everyday words: each observed y is a straight-line part plus a random error, and
the errors are bell-curve noise centered at zero with typical size sigma.

The fitted version hats everything estimated from data: $\hat{\beta}_0$,
$\hat{\beta}_1$, $\hat{\sigma}$ — and the errors are replaced by their observable
stand-ins, the residuals $e_i = y_i - \hat{y}_i$. ($x_i$ and $y_i$ are data;
they never wear hats.)
:::

## What you can now do

- **Read every column** of `summary(lm())` aloud and say what question each answers:
  Estimate, Std. Error, t value, Pr(>|t|), residual standard error, $R^2$.
- **Reconstruct** a t value from a journal table's estimate and parenthesized
  standard error, and decode the stars.
- **Read the model equation** $y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$ aloud, and
  say which symbols get hats once it's fitted from data.
- **Keep the ideas separate** — significance (p, stars) answers "is the effect
  nonzero?", $R^2$ answers "how much does it explain?", and the intercept is just the
  line's height at $x = 0$.

**Mastery rule:** if any bullet feels shaky, revisit the matching section above (the
three-seasons game for standard error, the columns list for t/p/$R^2$) and its
practice before moving to the capstone — which assumes you can read all of these
cold.

## Where this goes next

You can now score lines, derive the estimator, and decode both R output and journal
tables — every skill the course set out to teach. One module remains: the
[capstone](../04-capstone-reading-a-real-regression/lesson.qmd) points all of it at a
*real* regression end-to-end and finishes with a cumulative problem set spanning the
whole course. First, do this module's [practice problems](practice.qmd).
