---
title: "Practice: Reading lm() output and papers"
subtitle: "Answers are hidden — try honestly first, then click to check."
---

```{r setup}
#| message: false
# Packages this practice set uses. Missing ones are installed automatically.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

Work these in order — they get harder. Do them on paper or in an R session; only
then open the answer.

## Problem 1 — the toy example, new numbers

A fourth season arrives: luck $(+1, 0, -1, 0, +1)$ on top of the true line
$y = 1 + 2x$ at $x = (1, 2, 3, 4, 5)$. Write down the harvest $y$, then compute the
season's slope estimate by hand (covariance over variance — the deviations of $x$
are $(-2, -1, 0, 1, 2)$ and $\mathrm{Var}(x) = 2.5$). Check in R.

```{r p1-starter}
# Starter — copy this to begin (build y from the true line plus the luck)
x <- c(1, 2, 3, 4, 5); luck <- c(1, 0, -1, 0, 1)
```

::: {.callout-note collapse="true" title="Answer"}
True line values: $3, 5, 7, 9, 11$. Plus luck: $y = (4, 5, 6, 9, 12)$.

$\bar{y} = 36/5 = 7.2$, so y's deviations are $(-3.2, -2.2, -1.2, 1.8, 4.8)$.
Products with x's deviations: $6.4, 2.2, 0, 1.8, 9.6$ — sum $20$. Covariance:
$20/4 = 5$. Slope: $5 / 2.5 = \mathbf{2.0}$. This season's luck happened to cancel
almost perfectly.

```{r p1-answer}
x <- c(1, 2, 3, 4, 5)
y <- c(4, 5, 6, 9, 12)
cov(x, y) / var(x)   # 2 — another draw from the bouncing-estimates lottery
```
:::

## Problem 2 — faded worked example (standard error → t)

A fit on $x = (1, 2, 3, 4, 5)$ gives slope $\hat{\beta}_1 = 3$ with residual standard
error $\hat{\sigma} = 2$. The standard error and t value are worked out below, but the
**core step is blanked** — the one that turns $\hat{\sigma}$ into a slope bounce. Fill
in ONLY that line, then finish the t value.

> $\bar{x} = 3$; deviations $(-2, -1, 0, 1, 2)$; so $\sum (x_i - \bar{x})^2 = 4 + 1 + 0 + 1 + 4 = 10$.
> **`<your step here>`** — plug into $\mathrm{SE}(\hat{\beta}_1) = \hat{\sigma}/\sqrt{\sum(x_i-\bar{x})^2}$:
> $\mathrm{SE}(\hat{\beta}_1) = \underline{\phantom{000}}$.
> Then the t value $= \hat{\beta}_1 / \mathrm{SE}(\hat{\beta}_1) = \underline{\phantom{000}}$.

```{r p2-starter}
# Starter — copy this to check the two blanks
x <- c(1, 2, 3, 4, 5); sigma_hat <- 2; b1 <- 3
```

::: {.callout-note collapse="true" title="Answer"}
The blanked step is the module's signature move — error size over x-spread:
$\mathrm{SE}(\hat{\beta}_1) = 2 / \sqrt{10} = 2 / 3.162 \approx 0.632$.

Then $t = 3 / 0.632 \approx 4.74$ — the estimate sits almost five bounce-lengths from
zero, comfortably past the ±2 rule of thumb.

```{r p2-answer}
x <- c(1, 2, 3, 4, 5); sigma_hat <- 2; b1 <- 3
se <- sigma_hat / sqrt(sum((x - mean(x))^2))   # 0.632
c(SE = se, t = b1 / se)                          # 0.632 and 4.74
```
:::

## Problem 3 — decode a real table

From a (fictional but realistic) paper:

| | Model 1: Yield |
|---|---|
| Rainfall | 0.42* |
| | (0.18) |
| Constant | 12.10*** |
| | (1.95) |
| Observations | 214 |
| R² | 0.11 |

*Notes:* Standard errors in parentheses. \*p < 0.05, \*\*\*p < 0.001.

(a) Reconstruct the t value for Rainfall. (b) Say in plain English what the 0.42
means. (c) The R² is only 0.11 — does that contradict the star on Rainfall?

::: {.callout-note collapse="true" title="Answer"}
(a) $t = 0.42 / 0.18 \approx 2.33$ — a bit beyond 2 bounce-lengths from zero,
matching the single star (p < 0.05 but not < 0.01).

(b) Each extra unit of rainfall is associated with 0.42 more units of yield, on
average — the estimated slope $\hat{\beta}_1$.

(c) No contradiction. The star answers "is the slope distinguishable from zero?"
(yes, barely). R² answers "how much of yield's variation does rainfall explain?"
(11% — rainfall matters, but most variation comes from everything else). A
coefficient can be reliably nonzero and still explain little; with n = 214 even
modest effects earn stars. Reading both numbers together is exactly the skill
papers assume.
:::

## Problem 4 — the full decode, on mtcars (in R)

Starter below. From `summary(fit)`: (a) write the fitted line as an equation with
numbers in it; (b) rebuild the slope's t value from Estimate and Std. Error;
(c) state what `Pr(>|t|)` for `wt` claims, in one sentence with no jargon.
**Predict first:** before running, will the slope be positive or negative?

```{r p4-starter}
# Starter — the model: fuel efficiency as a function of car weight
fit <- lm(mpg ~ wt, data = mtcars)
summary(fit)
```

::: {.callout-note collapse="true" title="Answer"}
Predicted sign: negative — heavier cars get worse mileage.

(a) $\widehat{\text{mpg}} = 37.29 - 5.34 \times \text{wt}$ — a 1000-lb-heavier car
is predicted to get about 5.3 fewer miles per gallon.

(b) $t = -5.3445 / 0.5591 \approx -9.56$ — the estimate sits almost ten
bounce-lengths below zero.

```{r p4-answer}
est <- coef(summary(fit))["wt", ]
unname(est["Estimate"] / est["Std. Error"])   # -9.559, matches the t value column
```

(c) If car weight truly had no relationship with fuel efficiency, the chance of a
sample of 32 cars producing a slope estimate this far from zero is roughly
1-in-10-billion ($1.3 \times 10^{-10}$) — so "no relationship" is not a believable
explanation for what we see.
:::

## Problem 5 — interleaved *(uses Modules 1–2)*

Tiny data: $x = (1, 2, 3)$, $y = (2, 4, 3)$. Do the whole chain by hand:
(a) fit the line with the Module-2 formula ($\hat{\beta}_1 = s_{xy}/s_x^2$,
$\hat{\beta}_0 = \bar{y} - \hat{\beta}_1\bar{x}$); (b) compute its SSE (Module 1);
(c) form $R^2 = 1 - \mathrm{SSE}/\sum(y_i-\bar{y})^2$ (this module). **Predict first:**
with only three not-very-linear points, do you expect $R^2$ near 1 or well below it?

```{r p5-starter}
# Starter — copy this to begin
x <- c(1, 2, 3); y <- c(2, 4, 3)
```

::: {.callout-note collapse="true" title="Answer"}
Prediction: well below 1 — the points don't sit near any single line.

(a) $\bar{x} = 2$, $\bar{y} = 3$. Deviations of x: $(-1, 0, 1)$; of y: $(-1, 1, 0)$.
Products: $1, 0, 0$; sum 1, so $s_{xy} = 1/2$. Squared x-deviations sum to 2, so
$s_x^2 = 1$. Slope $\hat{\beta}_1 = 0.5$; intercept $\hat{\beta}_0 = 3 - 0.5\times2 = 2$.

(b) Predictions $2 + 0.5x$: $2.5, 3, 3.5$. Residuals: $-0.5, +1, -0.5$. Squared and
summed: $0.25 + 1 + 0.25 = 1.5$, so SSE $= 1.5$.

(c) Total sum of squares $\sum(y_i-\bar{y})^2 = 1 + 1 + 0 = 2$. So
$R^2 = 1 - 1.5/2 = \mathbf{0.25}$ — the line explains only a quarter of y's variation,
as predicted.

```{r p5-answer}
x <- c(1, 2, 3); y <- c(2, 4, 3)
fit <- lm(y ~ x)
coef(fit)                       # intercept 2, slope 0.5
c(SSE = sum(resid(fit)^2),      # 1.5
  TSS = sum((y - mean(y))^2),   # 2
  R2  = summary(fit)$r.squared) # 0.25
```
:::

## Problem 6 — spot the error

A colleague reports the `wt` result from `lm(mpg ~ wt, data = mtcars)` like this. It
contains **exactly one mistake** — a misreading of one number. Find it and give the
correct interpretation.

> "The p-value for weight is about $1.3\times10^{-10}$. So there's only a
> 0.00000001% chance that weight has no effect on mpg — we're 99.99999999% sure the
> effect is real."

```{r p6-starter}
# Starter — copy this to inspect the actual output
fit <- lm(mpg ~ wt, data = mtcars)
summary(fit)$coefficients
```

::: {.callout-note collapse="true" title="Answer"}
The mistake is treating the p-value as "the probability that weight has no effect."
A p-value is computed *assuming the null is true* (that weight has no effect): it is
the chance that luck alone, under that assumption, would produce a slope estimate
this far from zero. It is **not** the probability the null itself is true, so you
can't turn it into "99.99999999% sure the effect is real." Correct reading: "*if*
weight truly had no effect on mpg, a sample of 32 cars would almost never (about
1-in-10-billion) show a slope this extreme — so 'no effect' is a very poor
explanation for what we observe."

```{r p6-answer}
fit <- lm(mpg ~ wt, data = mtcars)
summary(fit)$coefficients["wt", ]   # the tiny p is P(data this extreme | null true)
```
:::

## Problem 7 — a gentle cliffhanger

Papers rarely stop at one predictor: you'll see
$y_i = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \dots + \varepsilon_i$ with a
whole *column* of coefficients. Fit `lm(mpg ~ wt + hp, data = mtcars)` and look at
the `wt` row. Its estimate changed from −5.34 to about −3.88 — yet nothing about
the cars changed. Why might adding `hp` change `wt`'s coefficient? (A full answer
needs the multiple-regression course; a good guess is enough here.)

::: {.callout-note collapse="true" title="Answer"}
```{r p7-answer}
coef(lm(mpg ~ wt + hp, data = mtcars))
```

Heavy cars also tend to have powerful engines (wt and hp are correlated). In the
one-predictor model, `wt`'s coefficient absorbed credit for everything traveling
with weight — including horsepower's effect. With `hp` in the model, each
coefficient now means "the effect of this variable *holding the other fixed*", so
`wt` keeps only the part that isn't explainable by `hp`. Making "holding fixed"
precise — and the matrix notation $(\mathbf{X}^\top\mathbf{X})^{-1}\mathbf{X}^\top\mathbf{y}$
papers use for it — is the natural next course to ask for.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="courses/simple-linear-regression/modules/03-reading-lm-output-and-papers" pass="4"}

::: {.quiz-q qid="lmout-t-recon" qtype="numeric" answer="7.7" tolerance="0.05"}
The lesson's paper table shows the Fertilizer estimate as 1.926 with SE 0.249.
Reconstruct the t value (estimate ÷ SE). Round to one decimal.
:::

::: {.quiz-q qid="lmout-pvalue" qtype="mc" answer="c"}
A regression reports $p = 0.002$ for a slope. Which interpretation is correct?

a) There is a 0.2% chance the true slope is zero — so we have 99.8% confidence the effect is genuine
b) The effect size must be large — a p-value this small only occurs when slopes are far from zero
c) Under a true slope of zero, data this extreme would arise only 0.2% of the time by chance
:::

::: {.quiz-q qid="lmout-stars-vs-effect" qtype="mc" answer="a"}
A predictor earns three stars ($p < 0.001$) but $R^2 = 0.08$. What does this mean?

a) The slope is reliably nonzero but explains very little of y's variation
b) The p-value and R² contradict each other — one of them must be misreported
c) The model fits well, as three significance stars indicate a strong relationship
:::

::: {.quiz-q qid="lmout-r2" qtype="numeric" answer="0.9" tolerance="0"}
For the season-1 data ($x = (1,2,3,4,5)$, $y = (4,4,7,10,10)$): SSE $= 3.6$ and
$\sum(y_i - \bar{y})^2 = 36$. Compute $R^2 = 1 - \mathrm{SSE}/\sum(y_i-\bar{y})^2$.
Enter one decimal.
:::

::: {.quiz-q qid="lmout-sse-interleaved" qtype="numeric" answer="2" tolerance="0"}
(From Module 1.) For $x = (1,2,3,4)$, $y = (2,3,5,6)$: candidate line "intercept 1,
slope 1" has residuals $(0, 0, +1, +1)$. What is its SSE? Enter a whole number.
:::

:::

::::
