---
title: "Ridge regularization: a leash on the coefficients"
subtitle: "Trading a little fit for a lot of stability, with one dial λ"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

## Where we're going

Least squares chases one goal: make the residuals small. When predictors are correlated
or few data points are available, that single-minded fit can hand you wild, unstable
coefficients — huge positive and negative values that cancel. **Ridge regularization**
adds a *second* goal: keep the coefficients themselves small. One dial, $\lambda$
("lambda"), trades between the two. Turn it up and the coefficients shrink toward zero;
turn it to zero and you're back to ordinary least squares. This module builds that idea
with one and two coefficients; the matrix form $(X^\top X + \lambda I)^{-1}$ is taught
where a method needs it.

**Builds on:** [loss-functions-and-optimization](../loss-functions-and-optimization/lesson.qmd)
(minimizing a cost), [model-fit-error-and-validation](../model-fit-error-and-validation/lesson.qmd)
(RSS, overfitting, bias–variance),
[variance-structure-and-standardization](../variance-structure-and-standardization/lesson.qmd)
(standardizing predictors — why it must come first).

## Warm-up: do you remember?

**1.** *(loss functions)* Least squares picks coefficients to make one quantity as small
as possible. Which quantity?

::: {.callout-note collapse="true" title="Answer"}
The residual sum of squares, $\mathrm{RSS} = \sum_i (y_i - \hat{y}_i)^2$ (each $\hat{y}_i$
is the model's predicted value for observation $i$) — the total squared miss. See
[loss-functions-and-optimization](../loss-functions-and-optimization/lesson.qmd).
:::

**2.** *(standardization)* To *standardize* a predictor you subtract its mean and divide
by its standard deviation. After standardizing, what are every predictor's mean and
spread?

::: {.callout-note collapse="true" title="Answer"}
Mean 0 and standard deviation 1 — all predictors put on the *same scale*, so none looks
"bigger" just because of its units. See
[variance-structure-and-standardization](../variance-structure-and-standardization/lesson.qmd).
:::

## A tiny example first

One predictor, three points on an exact line: $x = (1, 2, 3)$, $y = (2, 4, 6)$, so the
true slope is 2. Fit a slope $b$ through the origin, but now pay a penalty for making $b$
large. The thing we minimize is **penalized RSS**:

$$
\text{cost}(b) = \underbrace{\sum_i (y_i - b\,x_i)^2}_{\text{fit}} \; + \; \underbrace{\lambda\, b^2}_{\text{penalty}} .
$$

For one predictor we can solve it with the same move as ordinary least squares —
differentiate the cost with respect to $b$ and set the slope to zero — only now the
penalty adds an extra term:

$$
\frac{d}{db}\,\text{cost}(b) = \underbrace{-2\sum_i x_i(y_i - b\,x_i)}_{\text{from the fit}} + \underbrace{2\lambda b}_{\text{from the penalty}} = 0
\;\;\Longrightarrow\;\; b\Big(\textstyle\sum_i x_i^2 + \lambda\Big) = \sum_i x_i y_i .
$$

- Term 1: the derivative of $(y_i - b x_i)^2$ is $2(y_i - b x_i)(-x_i)$ (chain rule), summed
  over the data — the ordinary-least-squares part.
- Term 2: the derivative of $\lambda b^2$ is $2\lambda b$ — the *only* new piece the penalty adds.
- Divide by 2 and collect the $b$ terms: the $\lambda$ lands right next to $\sum x_i^2$.

So the clean answer is $b = \dfrac{\sum x_i y_i}{\sum x_i^2 + \lambda}$ — ordinary least
squares' $\sum x_i y_i / \sum x_i^2$, with $\lambda$ added to the denominator.
Here $\sum x_i y_i = 1\cdot2 + 2\cdot4 + 3\cdot6 = 2 + 8 + 18 = 28$ and
$\sum x_i^2 = 1 + 4 + 9 = 14$. So:

- $\lambda = 0$: $b = 28/14 = 2$ — ordinary least squares, the true slope.
- $\lambda = 14$: $b = 28/(14+14) = 28/28 = 1$ — shrunk to half.
- $\lambda = 56$: $b = 28/70 = 0.4$ — shrunk further.
- $\lambda \to \infty$: $b \to 0$ — the penalty wins entirely.

```{r toy-example}
x <- c(1, 2, 3); y <- c(2, 4, 6)
ridge_1d <- function(lambda) sum(x * y) / (sum(x^2) + lambda)   # closed form
sapply(c(0, 14, 56, 1e6), ridge_1d)                              # 2, 1, 0.4, ~0
```

The penalty puts $\lambda$ in the denominator: bigger $\lambda$, smaller $b$. That's the
whole mechanism — a **leash** on the coefficient.

## Getting a feel for it

**Two goals on one leash.** Ordinary least squares is a dog chasing one ball: "fit the
data, whatever it takes." Ridge puts the dog on a leash held from the origin: it still
lunges toward the fit, but the leash pulls its coefficients back toward zero. Lambda is
the leash's tension. Slack leash (lambda near 0): the dog reaches the OLS fit. Tight
leash (lambda huge): the dog can barely move from zero, and predictions collapse toward
a flat "predict the average" model. The useful settings live in between.

**Why penalize the coefficients' size at all.** When predictors are correlated, OLS can
fit the data well with *enormous* opposing coefficients (one huge positive, its correlated
twin huge negative) that nearly cancel — a fit that's accurate on this sample but
wildly unstable, flipping with tiny data changes. Squaring the coefficients in the penalty
makes those blow-ups expensive, so ridge trades a *tiny* bit of fit for a *lot* of
stability. That's the **bias–variance trade-off** from the
[model-fit](../model-fit-error-and-validation/lesson.qmd) module: ridge accepts a little
bias (coefficients pulled off the exact OLS values) to slash variance (their sample-to-
sample wobble).

**Why standardize first.** The penalty adds up the square of every coefficient and treats
them all equally. But a predictor measured in dollars (range ~1000) gets a tiny per-unit
coefficient, while one measured in fractions (range ~1) gets a large one — so the penalty
would clobber the second and ignore the first, purely because of units. Standardizing
puts all predictors on one scale first, so the penalty is *fair*. (Ridge on unstandardized
predictors penalizes the wrong things.)

## What the notation is going to say

The formula will say: *minimize the usual squared error, plus $\lambda$ times the sum of
the squared coefficients.* The first part wants a good fit; the second wants small
coefficients; $\lambda$ sets the exchange rate. At $\lambda = 0$ the second part vanishes
and it's plain least squares; as $\lambda$ grows, the second part dominates and drags
every coefficient toward 0.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $b$, $\beta_j$ | "bee", "beta sub jay" | a coefficient we're fitting | slope $b$ |
| $\lambda$ | "lambda" | the penalty dial (≥ 0): how hard to shrink coefficients | swept 0 → ∞ |
| $n$ | "en" | number of observations | 3 |
| $y_i$ | "why sub eye" | the outcome / response value for observation $i$ | 2, 4, 6 |
| $x_i$ | "ex sub eye" | the predictor value for observation $i$ (one predictor; $j$ dropped) | 1, 2, 3 |
| $x_{ij}$ | "ex sub eye jay" | predictor $j$'s value for observation $i$ | the $x$ values |
| $\hat\beta$ | "beta-hat" | the estimated coefficient vector (the ridge solution) | — (matrix form) |
| $X$ | "the data matrix" | predictors stacked in rows (used in the matrix form) | — (matrix form) |
| $y$ | "the response vector" | all $n$ outcome values stacked (used in the matrix form) | — (matrix form) |
| $\arg\min_{\beta}$ | "arg min over beta" | the coefficient vector $\beta$ that makes the expression smallest | the fitted $\hat\beta$ |
| $\sum_i$ | "the sum over i" | add over the data points | over the 3 points |
| $\sum_j \beta_j^2$ | "the sum of the squared betas" | the total coefficient size the penalty charges for | $b^2$ (one predictor) |
| $\lVert \beta \rVert^2$ | "the squared norm of beta" | same thing in vector shorthand: $\sum_j \beta_j^2$ | $b^2$ |
| $I$ | "the identity matrix" | the matrix version of "1" (used in the matrix form) | — |

## The formal version

Ridge regression chooses the coefficient vector $\beta$ that minimizes **penalized RSS**:

$$
\hat\beta_{\text{ridge}} = \arg\min_{\beta} \; \underbrace{\sum_{i=1}^{n}\Big(y_i - \textstyle\sum_j \beta_j x_{ij}\Big)^2}_{\text{fit the data}} \; + \; \underbrace{\lambda \sum_{j} \beta_j^2}_{\text{keep coefficients small}} .
$$

Reading it aloud: "beta-hat-ridge is the beta minimizing the sum of squared residuals,
plus lambda times the sum of the squared betas" — or in everyday words: "fit the data
*and* keep the knobs small, with lambda deciding how much the second goal matters."

- $\lambda = 0$ removes the penalty → ordinary least squares.
- $\lambda \to \infty$ makes any nonzero coefficient infinitely expensive → all
  coefficients driven to 0 (the model predicts the mean).
- Every $\lambda$ in between gives a different, shrunk fit. Crucially, the penalty is on
  the *slopes only* — the intercept is left unpenalized, which is why we standardize
  predictors (and often centre $y$) first.

**The matrix form (stated, used later).** With predictors stacked in $X$, the ridge
solution is $\hat\beta = (X^\top X + \lambda I)^{-1} X^\top y$ — ordinary least squares
with $\lambda$ added down the diagonal of $X^\top X$. The payoff, proved in the
a later course: $X^\top X + \lambda I$ is invertible for **any** $\lambda > 0$, even
when $X^\top X$ alone is singular (near-collinear predictors). That's ridge's other job —
making a fit *computable* when plain least squares can't invert.

## Explore it

::: {.callout-tip title="PREDICT before you run"}
At lambda = 0 you get OLS. What do both coefficients approach as lambda gets big?
Commit your guess, then run the cell.
:::

```{webr}
#| autorun: false
#| fig-alt: "Two coefficient paths plotted against increasing lambda on a log scale: both coefficients start at their OLS values and shrink smoothly toward zero as lambda grows."
# TWO correlated predictors; trace each coefficient as lambda grows (the 'ridge path').
# TRY IT: change the shared-noise sd (the 0.3 below — smaller sd -> more correlated x1, x2)
# or the lambda range, and re-run.
set.seed(1); n <- 60
z <- rnorm(n)
x1 <- scale(z + rnorm(n, sd = 0.3))          # standardized, correlated predictors
x2 <- scale(z + rnorm(n, sd = 0.3))
y  <- as.numeric(scale(2 * x1 - 1 * x2 + rnorm(n)))
X  <- cbind(x1, x2)

lambdas <- 10^seq(-2, 3, length.out = 40)
paths <- sapply(lambdas, function(l)
  solve(t(X) %*% X + l * diag(2), t(X) %*% y))  # ridge matrix form
matplot(log10(lambdas), t(paths), type = "l", lwd = 2, lty = 1,
        xlab = "log10(lambda)", ylab = "coefficient", main = "Ridge paths: shrink to 0")
abline(h = 0, lty = 3); legend("topright", c("b1", "b2"), col = 1:2, lwd = 2, bty = "n")
```

::: {.callout-tip title="PREDICT before you run"}
At lambda = 14, the closed form gave b = 1 for x = (1,2,3), y = (2,4,6). Will the
brute-force optimizer agree? Commit your guess, then run the cell.
:::

```{webr}
#| autorun: false
# TRY IT: the 1-D closed form vs a brute-force minimizer of penalized RSS — they agree.
x <- c(1, 2, 3); y <- c(2, 4, 6); lambda <- 14
closed  <- sum(x * y) / (sum(x^2) + lambda)
bruteforce <- optimize(function(b) sum((y - b * x)^2) + lambda * b^2, c(-5, 5))$minimum
c(closed_form = closed, optimize = bruteforce)   # both 1
```

## Common traps

**Trap: "a larger $\lambda$ throws away real signal."** Not quite — it *shrinks* toward
zero, it doesn't discard. Tiny $\lambda$ is essentially OLS (keeps all the signal); huge
$\lambda$ collapses to predicting the mean (keeps none); the useful $\lambda$ is the sweet
spot that trades a little fit for much less variance. In the toy, $\lambda = 14$ shrank the
true slope from 2 to 1 — not "wrong," just biased toward zero in exchange for stability;
you'd tune $\lambda$ (often by cross-validation) to balance the two.

**Trap: "ridge zeroes out useless predictors, so it does variable selection."** No — ridge
shrinks coefficients *toward* zero but almost never *to* exactly zero, so every predictor
stays in the model. In the toy, $\lambda = 56$ shrank the slope to 0.4 — small, but still
not zero. Even at $\lambda = 1{,}000{,}000$ the slope is $28/1{,}000{,}014 \approx 0.00003$,
not zero. Its cousin the **lasso** (which penalizes $\sum_j |\beta_j|$ instead of
$\sum_j \beta_j^2$) *does* drive some coefficients to exactly zero and so selects
variables. Ridge stabilizes; lasso selects.

## Check yourself

**1.** For one predictor with $\sum x_i y_i = 40$ and $\sum x_i^2 = 10$, what is the ridge
slope at $\lambda = 0$, at $\lambda = 10$, and at $\lambda = 30$?

::: {.callout-note collapse="true" title="Answer"}
$b = 40/(10 + \lambda)$: at $\lambda = 0$, $b = 4$ (OLS); at $\lambda = 10$, $b = 40/20 = 2$;
at $\lambda = 30$, $b = 40/40 = 1$. Each increase in $\lambda$ shrinks $b$ toward 0.
:::

**2.** Why must predictors be standardized before ridge, but not before ordinary least
squares?

::: {.callout-note collapse="true" title="Answer"}
OLS is unaffected by rescaling a predictor — its coefficient just rescales to compensate,
and the fit is identical. Ridge's penalty $\lambda\sum_j \beta_j^2$ is *not* scale-free: a
predictor on a large scale has a small coefficient and is barely penalized, one on a small
scale has a large coefficient and is heavily penalized. Standardizing puts all predictors
on one scale so the penalty treats them fairly.
:::

**3.** In your own words: explain to a colleague, in two sentences, what $\lambda$ does and
why you'd ever accept a *worse* fit to the training data.

::: {.callout-note collapse="true" title="A model answer"}
$\lambda$ sets how hard the model is pulled toward small coefficients: zero means plain
least squares, large means everything shrinks toward the mean. You accept a slightly worse
training fit because the shrunk coefficients are far more stable — less likely to be wild,
noise-chasing values that predict new data badly (trading a bit of bias for a lot less
variance).
:::

## What you can now do

- **Write penalized RSS** (fit + $\lambda \sum_j \beta_j^2$) and explain each part.
- **Predict how coefficients move** as $\lambda$ goes from 0 (OLS) to $\infty$ (all shrink
  to 0), and read a ridge path plot.
- **Say why standardization comes first**, and why the intercept is left unpenalized.
- **Distinguish ridge from lasso** (shrinks toward 0 vs sets exactly to 0 / selects), and
  state ridge's second job: making $(X^\top X + \lambda I)$ invertible for any $\lambda>0$.

If any of these feels shaky, revisit [Getting a feel for it](#getting-a-feel-for-it) —
methods that fit near-collinear predictors depend on the leash idea.

## Where this goes next

The matrix form $(X^\top X + \lambda I)^{-1}$ becomes essential when a correlation matrix
is near-singular: a later course uses a ridge $\lambda$ to
invert a near-collinear predictor-correlation matrix stably. Ready for reps? Do the
[practice problems](practice.qmd). External videos/readings: [resources](resources.md).
