---
title: "Practice: Ridge regularization"
subtitle: "Answers are hidden ŌĆö try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
set.seed(42)
```

Work these in order ŌĆö they get harder. Do them on paper or in an R session; only then
open the answer.

## Problem 1 ŌĆö 1-D ridge by hand, new numbers

One predictor with $\sum x_i y_i = 60$ and $\sum x_i^2 = 20$. Using
$b(\lambda) = \dfrac{\sum x_i y_i}{\sum x_i^2 + \lambda}$, compute the ridge slope at
$\lambda = 0$, $\lambda = 20$, and $\lambda = 40$.

```{webr}
#| autorun: false
# p1-starter
Sxy <- 60; Sxx <- 20
```

::: {.callout-note collapse="true" title="Answer"}
$b(0) = 60/20 = 3$ (OLS); $b(20) = 60/40 = 1.5$; $b(40) = 60/60 = 1$. Each rise in
$\lambda$ shrinks the slope toward 0.
:::

## Problem 2 ŌĆö fill in the missing step

Derive the 1-D ridge slope: minimize $\text{cost}(b) = \sum_i (y_i - b x_i)^2 + \lambda b^2$.

- Step 1: $\dfrac{d\,\text{cost}}{db} = \sum_i 2(y_i - b x_i)(-x_i) + 2\lambda b$.
- Step 2: **your step here** ŌĆö set the derivative to 0 and solve for $b$.
- Step 3: $b = \dfrac{\sum_i x_i y_i}{\sum_i x_i^2 + \lambda}$.

::: {.callout-note collapse="true" title="Answer"}
Step 2: set to 0 and divide by 2: $-\sum_i x_i y_i + b\sum_i x_i^2 + \lambda b = 0$, so
$b(\sum_i x_i^2 + \lambda) = \sum_i x_i y_i$, giving Step 3. The $\lambda b$ from the
penalty is what adds $\lambda$ to the denominator ŌĆö the shrinkage.
:::

## Problem 3 ŌĆö watch the shrinkage in R

For the two-predictor standardized data below, compute the ridge coefficients at
$\lambda = 0$, $\lambda = 5$, and $\lambda = 100$ using the matrix form
$(X^\top X + \lambda I)^{-1} X^\top y$. Confirm they shrink toward 0.

```{webr}
#| autorun: false
# p3-starter ŌĆö fill in the ridge solve for each lambda
# PREDICT: lambda = 0 gives OLS; lambda = 100 gives coefficients near 0.
set.seed(5); n <- 50
x1 <- scale(rnorm(n)); x2 <- scale(rnorm(n))
y  <- as.numeric(scale(1.5 * x1 - 0.8 * x2 + rnorm(n)))
X  <- cbind(x1, x2)
```

::: {.callout-note collapse="true" title="Answer"}
Each larger $\lambda$ pulls both coefficients toward 0; $\lambda = 0$ reproduces OLS.

```{r}
set.seed(5); n <- 50
x1 <- scale(rnorm(n)); x2 <- scale(rnorm(n))
y  <- as.numeric(scale(1.5 * x1 - 0.8 * x2 + rnorm(n))); X <- cbind(x1, x2)
ridge <- function(l) as.numeric(solve(t(X) %*% X + l * diag(2), t(X) %*% y))
sapply(c(0, 5, 100), ridge)      # columns: lambda 0, 5, 100 ŌĆö magnitudes shrink
```
:::

## Problem 4 ŌĆö spot the error

A colleague says: *"I used ridge with a big $\lambda$ and some coefficients are still not
zero ŌĆö ridge must be broken as a variable selector, since it's supposed to drop useless
predictors."* This contains **exactly one mistake** ŌĆö find it.

::: {.callout-note collapse="true" title="Answer"}
The mistake: expecting **ridge** to zero out predictors. Ridge shrinks coefficients
*toward* zero but essentially never *to* exactly zero, so it does **not** select variables
ŌĆö every predictor stays. The tool that drives coefficients to exactly zero (and so
selects) is the **lasso** ($\ell_1$ penalty, $\sum_j|\beta_j|$). Ridge is behaving
correctly; it was the wrong tool for the colleague's goal.
:::

## Problem 5 ŌĆö mixing it with loss functions *(uses loss-functions-and-optimization's "minimize a cost")*

Ridge minimizes fit + penalty. For $x = (2, 4)$, $y = (2, 4)$ (true slope 1), and
$\lambda = 4$, evaluate the penalized cost $\text{cost}(b) = \sum_i (y_i - b x_i)^2 + \lambda b^2$
at $b = 1$ and at $b = 0.5$, and say which the minimizer prefers between them.

```{webr}
#| autorun: false
# p5-starter
x <- c(2, 4); y <- c(2, 4); lambda <- 4
cost <- function(b) sum((y - b * x)^2) + lambda * b^2
```

::: {.callout-note collapse="true" title="Answer"}
At $b = 1$: fit $= (2-2)^2 + (4-4)^2 = 0$, penalty $= 4\cdot1 = 4$, cost $= 4$. At
$b = 0.5$: fit $= (2-1)^2 + (4-2)^2 = 1 + 4 = 5$, penalty $= 4\cdot0.25 = 1$, cost $= 6$.
So between these two, the minimizer prefers $b = 1$ (cost 4 < 6). (The true ridge optimum
here is $b = \sum xy/(\sum x^2 + \lambda) = 20/(20+4) \approx 0.83$ ŌĆö between OLS's 1 and 0.)
:::

## Problem 6 ŌĆö transfer: ridge rescues a singular fit

Two predictors are *identical*: $x_2 = x_1$. Then $X^\top X$ is singular and OLS
(`solve(t(X) %*% X, ...)`) fails. Predict what happens if you instead solve
$(X^\top X + \lambda I)^{-1}X^\top y$ with $\lambda = 0.1$. *(This is exactly the trick
the next course uses on a near-collinear correlation matrix.)*

```{webr}
#| autorun: false
# p6-starter ŌĆö PREDICT: OLS errors; does the ridge version return an answer?
set.seed(1); n <- 30
x1 <- rnorm(n); x2 <- x1                 # identical columns -> X'X singular
y  <- x1 + rnorm(n); X <- cbind(x1, x2)
```

::: {.callout-note collapse="true" title="Answer"}
Plain OLS errors ("system is computationally singular") because $X^\top X$ can't be
inverted. Adding $\lambda I$ (here $0.1$ down the diagonal) makes $X^\top X + \lambda I$
invertible, so the ridge solve **returns coefficients** ŌĆö it splits the shared effect
between the two identical columns instead of failing. Adding a positive $\lambda$ to the
diagonal guarantees invertibility for any $\lambda > 0$, which is why ridge stabilizes
near-collinear (and even exactly-collinear) problems.

```{r}
set.seed(1); n <- 30; x1 <- rnorm(n); x2 <- x1; y <- x1 + rnorm(n); X <- cbind(x1, x2)
solve(t(X) %*% X + 0.1 * diag(2), t(X) %*% y)   # returns an answer; plain solve() would error
```
:::
