---
title: "Practice: Distance, similarity & the geometry of data"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only (a {webr} starter does not see it — each starter re-states its
# own data). Base R is all we need, so nothing to install.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Do them on paper or in an R session; only
then open the answer.

## Problem 1 — the toy example, new numbers

Two new towns: $\mathbf{C} = (2, 1)$ and $\mathbf{D} = (5, 5)$. By hand, compute the
Euclidean distance between them (square the coordinate gaps, add, take the square root).
Then check with `dist()`.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
C <- c(2, 1)
D <- c(5, 5)
```

::: {.callout-note collapse="true" title="Answer"}
Gaps: $5 - 2 = 3$ and $5 - 1 = 4$. Square them: $3^2 = 9$ and $4^2 = 16$. Add:
$9 + 16 = 25$. Square-root: $\sqrt{25} = 5$. So the distance is **5** — the same 3-4-5
triangle as the lesson, just shifted on the map.

```{r p1-answer}
C <- c(2, 1)
D <- c(5, 5)
sqrt(sum((C - D)^2))              # 5 — by hand
as.numeric(dist(rbind(C, D)))    # 5 — built-in, same answer
```
:::

## Problem 2 — fill in the missing step (the norm of a vector)

Here is the **length (norm)** of the vector $\mathbf{v} = (5, 12)$ worked out — but the
**core step is blanked**. Fill in ONLY the blanked line.

> The norm is the vector's distance from the origin $(0, 0)$, so it's Pythagoras with
> the origin as the other point.
> **`<your step here>`** — square each coordinate and add them:
> $5^2 + 12^2 = \underline{\phantom{000}}$.
> Then take the square root of that sum to get $\lVert \mathbf{v} \rVert$.

```{webr}
#| autorun: false
# p2-starter — edit and Run to check your blanked step
v <- c(5, 12)
```

::: {.callout-note collapse="true" title="Answer"}
The blanked step is "square each coordinate and add": $5^2 + 12^2 = 25 + 144 = 169$.
Then $\lVert \mathbf{v} \rVert = \sqrt{169} = 13$. That sum-of-squares-then-root is
exactly the distance recipe with the origin as the second point.

```{r p2-answer}
v <- c(5, 12)
sum(v^2)                              # 169 — the blanked step (square, add)
sqrt(sum(v^2))                        # 13 — the norm (length) of v
as.numeric(dist(rbind(c(0, 0), v)))   # 13 — same, as distance from the origin
```
:::

## Problem 3 — Euclidean vs Manhattan in 3-D

Two points now have **three** coordinates: $\mathbf{p} = (1, 2, 2)$ and
$\mathbf{q} = (3, 5, 8)$. Compute BOTH their Euclidean distance (square the gaps, add,
root) and their Manhattan distance (add the *absolute* gaps). You can't draw the
triangle any more — but the recipe doesn't care.

**PREDICT** first: which will be larger, Euclidean or Manhattan? Write it down, then run.

```{webr}
#| autorun: false
# p3-starter — your job: Euclidean AND Manhattan distance
p <- c(1, 2, 2)
q <- c(3, 5, 8)
```

::: {.callout-note collapse="true" title="Answer"}
Gaps: $3-1 = 2$, $5-2 = 3$, $8-2 = 6$.

- **Euclidean:** square and add — $2^2 + 3^2 + 6^2 = 4 + 9 + 36 = 49$; root:
  $\sqrt{49} = 7$.
- **Manhattan:** add the absolute gaps — $|2| + |3| + |6| = 2 + 3 + 6 = 11$.

Manhattan (11) is larger than Euclidean (7) — always true, because the straight-line
path never exceeds the grid-walking path (the shortcut across the hypotenuse beats
walking the legs).

```{r p3-answer}
p <- c(1, 2, 2)
q <- c(3, 5, 8)
sqrt(sum((p - q)^2))                                  # 7  — Euclidean
sum(abs(p - q))                                       # 11 — Manhattan
as.numeric(dist(rbind(p, q)))                         # 7  — built-in Euclidean
as.numeric(dist(rbind(p, q), method = "manhattan"))  # 11 — built-in Manhattan
```
:::

## Problem 4 — spot the error (adding the gaps)

A colleague computed the straight-line distance between $\mathbf{A} = (0, 0)$ and
$\mathbf{B} = (3, 4)$ like this — it contains **exactly one mistake**. Find it, explain
why it's wrong, and give the corrected distance.

> "The east gap is 3 and the north gap is 4, so the distance is just $3 + 4 = 7$."

```{webr}
#| autorun: false
# p4-starter — copy this to see the two competing distances
A <- c(0, 0)
B <- c(3, 4)
```

::: {.callout-note collapse="true" title="Answer"}
The mistake is **adding the gaps instead of squaring-adding-rooting them**. Adding the
gaps ($3 + 4 = 7$) gives the *Manhattan* distance — the path that walks east then north,
along the two legs. The straight-line (Euclidean) distance cuts across the hypotenuse
and is shorter: $\sqrt{3^2 + 4^2} = \sqrt{9 + 16} = \sqrt{25} = \mathbf{5}$, not 7.

```{r p4-answer}
A <- c(0, 0)
B <- c(3, 4)
sum(abs(A - B))          # 7 — the colleague's number: Manhattan, not straight-line
sqrt(sum((A - B)^2))     # 5 — the correct Euclidean distance
```
:::

## Problem 5 — mixing it with vectors & summation *(uses [vectors-and-summation](../vectors-and-summation/lesson.qmd)'s dot product)*

Find the **shadow length** (projection) of $\mathbf{a} = (3, 3)$ onto the direction
$\mathbf{b} = (4, 0)$ — how far $\mathbf{a}$ reaches along $\mathbf{b}$. Use the
formula $(\mathbf{a} \cdot \mathbf{b}) / \lVert \mathbf{b} \rVert$, building the dot
product $\mathbf{a} \cdot \mathbf{b}$ as the sum-of-products you learned in
vectors & summation.

**PREDICT** first: $\mathbf{b}$ points straight east, so the shadow should equal one of
$\mathbf{a}$'s coordinates — which one? Then compute and check.

```{webr}
#| autorun: false
# p5-starter — edit and Run to solve, right here
a <- c(3, 3)
b <- c(4, 0)
```

::: {.callout-note collapse="true" title="Answer"}
The dot product is the sum-of-products from vectors & summation:
$\mathbf{a} \cdot \mathbf{b} = 3\times4 + 3\times0 = 12 + 0 = 12$. The norm of
$\mathbf{b}$ is $\lVert \mathbf{b} \rVert = \sqrt{4^2 + 0^2} = \sqrt{16} = 4$. So the
shadow length is $12 / 4 = \mathbf{3}$ — which is exactly $\mathbf{a}$'s east coordinate,
as predicted (a shadow cast onto the east-axis is just the east component).

```{r p5-answer}
a <- c(3, 3)
b <- c(4, 0)
dp <- sum(a * b)          # 12 — the dot product (vectors & summation)
len_b <- sqrt(sum(b^2))   # 4  — the norm of b
dp / len_b                # 3  — the shadow length of a along b
```

The `sum(a * b)` line is precisely the dot product from
[vectors & summation](../vectors-and-summation/lesson.qmd); the projection just rescales
it by $\mathbf{b}$'s length.
:::

## Problem 6 — a transfer: correlation distance (interpretation)

Three sensors record a reading at three times:

- Sensor **X**: $(1, 2, 3)$ — rising.
- Sensor **Y**: $(2, 4, 6)$ — also rising, but on a bigger scale.
- Sensor **Z**: $(3, 2, 1)$ — falling.

**Euclidean** distance would call X and Z fairly close (their numbers overlap) and X and
Y far (Y's numbers are much bigger). But sometimes you care about the *pattern* — does
the reading rise and fall together? — not the raw size. That's what **correlation
distance**, $1 - r$, measures ($r$ is the correlation from
[mean, variance & covariance](../mean-variance-covariance/lesson.qmd)). Without
computing exact numbers: rank the pairs (X&Y, X&Z) from *most* similar to *least*
similar by **correlation distance**, and say why. Then run the starter to confirm.

```{webr}
#| autorun: false
# p6-starter — edit and Run to check your ranking
X <- c(1, 2, 3)
Y <- c(2, 4, 6)
Z <- c(3, 2, 1)
```

::: {.callout-note collapse="true" title="Answer"}
X and Y move in **perfect lockstep** (Y is just X doubled — both rise by the same
proportion each step), so their correlation is $r = 1$ and the correlation distance is
$1 - 1 = 0$: as similar as it gets, *despite* Y's much larger raw numbers. X and Z move
in **exact opposition** (X rises while Z falls), so $r = -1$ and the correlation
distance is $1 - (-1) = 2$: as dissimilar as it gets. So by correlation distance:
**X&Y most similar (0), X&Z least similar (2)** — the opposite of what raw Euclidean
size might suggest. This is exactly the "scale/pattern" distinction from the lesson:
correlation distance ignores overall size and asks only about the shared up-and-down
pattern.

```{r p6-answer}
X <- c(1, 2, 3)
Y <- c(2, 4, 6)
Z <- c(3, 2, 1)
1 - cor(X, Y)    # 0 — perfectly aligned pattern (Y is X scaled up)
1 - cor(X, Z)    # 2 — perfectly opposed pattern
```
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/distance-similarity-and-geometry" pass="4"}

::: {.quiz-q qid="dsg-euclidean" qtype="numeric" answer="10" tolerance="0"}
Two points differ by $6$ units east and $8$ units north. What is the straight-line
(Euclidean) distance between them — the square root of the summed squared gaps?
:::

::: {.quiz-q qid="dsg-manhattan-vs-euclidean" qtype="mc" answer="b"}
Two towns are $3$ units apart east and $4$ north. A student adds the gaps, $3 + 4 = 7$,
and calls that the distance. What is $7$?

a) The Euclidean (straight-line) distance
b) The Manhattan (grid-walking) distance; the straight-line distance is $\sqrt{3^2 + 4^2} = 5$
c) An error — a distance cannot be found from coordinate gaps
:::

::: {.quiz-q qid="dsg-scaling-dominates" qtype="mc" answer="a"}
In a table with income (in thousands) and age (in tens), income dominates every
distance. Why?

a) Distance weighs each column by its raw numeric spread, so the large-scale column dominates until you rescale
b) Income is genuinely the more important variable
c) Age should be dropped from the calculation
:::

::: {.quiz-q qid="dsg-zero-dot-orthogonal" qtype="mc" answer="c"}
Two vectors have a dot product of $0$. What does that mean geometrically?

a) They point in the same direction
b) They have zero length
c) They are orthogonal — at right angles, carrying separate information
:::

::: {.quiz-q qid="dsg-vector-norm" qtype="numeric" answer="13" tolerance="0"}
(From vectors & summation.) A vector's length (norm) is $\sqrt{\sum v_i^2}$. For
$v = (5, 12)$, what is its length?
:::

:::

::::
