---
title: "Practice: Quadratic forms & the chi-square statistic"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Do them on paper or in an R session; only then open
the answer.

## Problem 1 — the toy example, new numbers

Deviations $\mathbf{x} = (6, 2)$, with variances $9$ and $4$ and no co-movement. Compute the
standardised squared distance by hand, then check in R.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
x <- c(6, 2)
v <- c(9, 4)
```

::: {.callout-note collapse="true" title="Answer"}
Term by term: $6^2/9 = 36/9 = 4$ and $2^2/4 = 4/4 = 1$; add to get $4 + 1 = 5$.

```{r p1-ans}
x <- c(6, 2); v <- c(9, 4)
sum(x^2 / v)   # 5
```
:::

## Problem 2 — fill in the missing step

For $\mathbf{x} = (3, 4)$ and the diagonal covariance
$V = \begin{bmatrix} 9 & 0 \\ 0 & 4 \end{bmatrix}$, someone computes $\mathbf{x}^\top V^{-1}\mathbf{x}$:

- Step 1: $V^{-1} = \begin{bmatrix} 1/9 & 0 \\ 0 & 1/4 \end{bmatrix}$ (flip each diagonal entry).
- Step 2: **your step here** — compute $V^{-1}\mathbf{x}$.
- Step 3: dot $\mathbf{x}$ with the step-2 result: $3\cdot\tfrac13 + 4\cdot 1 = 5$.

Fill in step 2.

```{webr}
#| autorun: false
# p2-starter — edit and Run to check the blank
x <- c(3, 4)
V <- matrix(c(9, 0, 0, 4), nrow = 2, byrow = TRUE)
```

::: {.callout-note collapse="true" title="Answer"}
$V^{-1}\mathbf{x} = (3/9,\ 4/4) = (1/3,\ 1)$ — multiply each deviation by the matching
reciprocal variance. That's the "divide each deviation by its variance" move; dotting it back
against $\mathbf{x}$ then squares-and-adds.

```{r p2-ans}
x <- c(3, 4); V <- matrix(c(9,0,0,4), nrow = 2, byrow = TRUE)
solve(V) %*% x           # (1/3, 1)
as.numeric(t(x) %*% solve(V) %*% x)   # 5
```
:::

## Problem 3 — a covariance with co-movement

Now the two measurements co-move. For $\mathbf{x} = (3, 4)$ and
$V = \begin{bmatrix} 9 & 3 \\ 3 & 4 \end{bmatrix}$, compute $\mathbf{x}^\top V^{-1}\mathbf{x}$
in R, and say whether it is bigger or smaller than the no-co-movement answer of $5$.

```{webr}
#| autorun: false
# p3-starter — your job: compute t(x) %*% solve(V) %*% x
x <- c(3, 4)
V <- matrix(c(9, 3, 3, 4), nrow = 2, byrow = TRUE)
```

::: {.callout-note collapse="true" title="Answer"}
```{r p3-ans}
x <- c(3, 4); V <- matrix(c(9,3,3,4), nrow = 2, byrow = TRUE)
as.numeric(t(x) %*% solve(V) %*% x)
```
The result is $4$ — **smaller** than $5$. When two measurements are positively correlated and
both drift in the same (positive) direction, some of that joint drift is "expected together":
the inverse-covariance sandwich discounts the shared movement instead of double-counting it, so
the standardised distance comes out lower than the no-co-movement $5$.
:::

## Problem 4 — spot the error

A colleague wants the p-value for a statistic of $10$ built from a **length-4** deviation
vector. They wrote — it contains **exactly one mistake**. Find it and fix it.

```{webr}
#| autorun: false
# p4-starter — this contains exactly one mistake
stat <- 10
n_observations <- 500     # the study had 500 rows
p <- pchisq(stat, df = n_observations, lower.tail = FALSE)
```

::: {.callout-note collapse="true" title="Answer"}
The mistake is the **degrees of freedom**: `df` must be the *length of the sandwiched vector*
(here $4$), not the number of observations (500). With `df = 500`, a statistic of 10 looks
almost impossibly small (p ≈ 1); with the correct `df = 4` it is genuinely surprising.

```{r p4-ans}
pchisq(10, df = 4, lower.tail = FALSE)   # ~0.040 — the correct p-value
```
:::

## Problem 5 — mixing it with standardization *(uses variance-structure-and-standardization's covariance matrix)*

You are handed two effect estimates with deviations $\mathbf{x} = (2, -1)$, standard errors
$(2, 1)$, and a correlation of $0.5$ between them. First **build the covariance matrix** from
the standard errors and correlation (the standardization move: $V_{ii} = \text{se}_i^2$,
$V_{12} = \text{se}_1\,\text{se}_2\,\rho$), then compute the standardised squared distance.

```{webr}
#| autorun: false
# p5-starter — edit and Run to solve, right here
x  <- c(2, -1)
se <- c(2, 1)
rho <- 0.5
```

::: {.callout-note collapse="true" title="Answer"}
Diagonal: $\text{se}_1^2 = 4$, $\text{se}_2^2 = 1$. Off-diagonal:
$\text{se}_1\,\text{se}_2\,\rho = 2\cdot 1\cdot 0.5 = 1$. So
$V = \begin{bmatrix} 4 & 1 \\ 1 & 1 \end{bmatrix}$.

```{r p5-ans}
x <- c(2, -1); se <- c(2, 1); rho <- 0.5
V <- matrix(c(se[1]^2,          se[1]*se[2]*rho,
              se[1]*se[2]*rho,  se[2]^2), nrow = 2, byrow = TRUE)
V
as.numeric(t(x) %*% solve(V) %*% x)
```
The standardised squared distance is $4$. Building $V$ from standard errors and a correlation
is exactly the covariance-matrix construction you'll meet again as $V = (\sigma\sigma^\top)\odot\Sigma$
in a later course.
:::

## Problem 6 — transfer: read a statistic back as a sentence

A goodness-of-fit test reports a chi-square statistic of $2.1$ on $3$ degrees of freedom.
Without computing anything precisely, is the data surprising under the null, and what does that
say about the model?

::: {.callout-note collapse="true" title="Answer"}
A chi-square with $3$ degrees of freedom has an average of $3$ under pure noise, so a statistic
of $2.1$ is *below* average — completely unremarkable. In one sentence: "the data are no
further from what the model predicts than random noise would put them, so there is no evidence
the model is wrong." (Numerically, `pchisq(2.1, 3, lower.tail = FALSE)` ≈ $0.55$, a large
p-value — no reason to reject.)
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**, once the
material isn't fresh. A pass then is real evidence you've retained it. Five questions, graded on
the page; 4/5 or better means this module is solid. Results are saved in this browser only.

::: {.module-check module="foundations/quadratic-forms-and-the-chi-square-statistic" pass="4"}

::: {.quiz-q qid="qfcs-standardised-distance" qtype="numeric" answer="5" tolerance="0"}
For deviations $\mathbf{x} = (4, 3)$ with variances $8$ and $3$ and no co-movement, what is the
standardised squared distance $\sum x_i^2 / v_i$?
:::

::: {.quiz-q qid="qfcs-df-meaning" qtype="mc" answer="c"}
In $\mathbf{x}^\top V^{-1}\mathbf{x}$, the degrees of freedom of the resulting chi-square is:

a) the number of observations in the study
b) the number of nonzero entries of $\mathbf{x}$
c) the length of the vector $\mathbf{x}$
d) always 1
:::

::: {.quiz-q qid="qfcs-raw-vs-standardised" qtype="mc" answer="b"}
Why divide each squared deviation by its variance instead of just squaring and adding the raw
deviations?

a) It makes the arithmetic easier to carry out by hand
b) So noisy measurements count for less than precise ones
c) It guarantees the result is always a whole number
d) Raw squared deviations can be negative and this fixes it
:::

::: {.quiz-q qid="qfcs-pvalue-direction" qtype="mc" answer="a"}
Holding the degrees of freedom fixed, a **larger** value of the chi-square statistic gives:

a) a smaller p-value (more surprising under pure noise)
b) a larger p-value (less surprising under pure noise)
c) the same p-value
d) an undefined p-value
:::

::: {.quiz-q qid="qfcs-cov-build" qtype="numeric" answer="1" tolerance="0"}
*(uses variance-structure-and-standardization)* Two estimates have standard errors $2$ and $1$
and correlation $0.5$. What is the off-diagonal entry of their covariance matrix,
$\text{se}_1\,\text{se}_2\,\rho$?
:::

:::

::::
