---
title: "Practice: Bagging, random forests & boosting"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c("rpart", "randomForest")
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Do them on paper or in an R session; only then
open the answer.


## Problem 1 — confidence rep with new numbers

Six observations:

| $x$ | 1 | 2 | 3 | 4 | 5 | 6 |
|---|---|---|---|---|---|---|
| $y$ | 10 | 12 | 8 | 35 | 30 | 40 |

(a) Find the best one-split tree (the cutpoint with the lowest total RSS). What are the
left and right means?

(b) Bag $B = 25$ one-split trees and report the bagged prediction at $x = 5$.

```{webr}
#| autorun: false
# p1-starter
x <- c(1, 2, 3, 4, 5, 6)
y <- c(10, 12, 8, 35, 30, 40)
```

::: {.callout-note collapse="true" title="Answer"}
(a) Best split at $x = 3.5$: left = (10, 12, 8), mean = 10; right = (35, 30, 40),
mean = 35. Left RSS = 8, right RSS = 50, total = 58.

```{r}
x <- c(1, 2, 3, 4, 5, 6); y <- c(10, 12, 8, 35, 30, 40)
cuts <- (x[-length(x)] + x[-1]) / 2
rss <- sapply(cuts, function(s) {
  L <- y[x <= s]; R <- y[x > s]
  sum((L - mean(L))^2) + sum((R - mean(R))^2)
})
names(rss) <- cuts; round(rss, 2)  # 3.5 wins with RSS = 58
```

(b) Bagging 25 trees at $x = 4$:

```{r}
stump <- function(x, y, new_x) {
  ux <- sort(unique(x)); if (length(ux) < 2) return(mean(y))
  cuts <- (ux[-length(ux)] + ux[-1]) / 2
  rss <- sapply(cuts, function(s) {
    L <- y[x <= s]; R <- y[x > s]
    sum((L - mean(L))^2) + sum((R - mean(R))^2)
  })
  best <- cuts[which.min(rss)]
  ifelse(new_x <= best, mean(y[x <= best]), mean(y[x > best]))
}

x <- c(1, 2, 3, 4, 5, 6); y <- c(10, 12, 8, 35, 30, 40)
set.seed(42)
bp <- replicate(25, { i <- sample(1:6, 6, TRUE); stump(x[i], y[i], 5) })
cat("Bagged prediction at x=5:", round(mean(bp), 2), "\n")
cat("Single tree at x=5:", stump(x, y, 5), "\n")
```

The single tree predicts 35 (right region). The bagged prediction is close but not
identical — each bootstrap tree sees slightly different data, so the right-region mean
varies across trees. Averaging smooths this out.
:::


## Problem 2 — faded worked example

Below is the procedure for bagging the lesson's toy data (predict at $x = 5$, $B = 25$).
One step is blanked out.

- Step 1: set $B = 25$ and create an empty vector for predictions.
- Step 2: for $b = 1$: draw a bootstrap sample of 6 from the data (with replacement).
  Suppose we draw indices {1, 4, 1, 2, 5, 3}, giving ages (1, 4, 1, 2, 5, 3) and
  spending (5, 20, 5, 8, 25, 6).
- Step 3: **your step here** — given this bootstrap sample, what do you do to get a
  prediction at $x = 5$?
- Step 4: store the prediction. Repeat steps 2–4 for $b = 2, \ldots, 25$.
- Step 5: average all 25 stored predictions → the bagged prediction.

::: {.callout-note collapse="true" title="Answer"}
Step 3: **fit a one-split tree on the bootstrap sample** — try every cutpoint, pick the
one with the lowest total RSS, then predict at $x = 5$ using the left or right mean.

For this bootstrap sample, unique sorted ages are (1, 2, 3, 4, 5), so the cutpoints are
1.5, 2.5, 3.5, 4.5. The best cut is 3.5 (lowest RSS): left = (5, 5, 8, 6), mean = 6;
right = (20, 25), mean = 22.5. Since $x = 5 > 3.5$, the prediction is **22.5**.

This is the decision-trees module's best-split search, applied to a bootstrap sample
instead of the full data. Each of the 25 trees repeats this on its own bootstrap sample.
:::


## Problem 3 — OOB error computation

You bagged 4 trees on the lesson's toy data. Here are the OOB predictions for each
observation (from the trees that excluded it):

| Obs | $y$ | OOB tree predictions | OOB mean |
|---|---|---|---|
| 1 | 5 | 6, 7 | ? |
| 2 | 8 | 6, 5, 7 | ? |
| 3 | 6 | 7 | ? |
| 4 | 20 | 22, 23, 21 | ? |
| 5 | 25 | 23, 22 | ? |
| 6 | 22 | 23, 24, 22 | ? |

(a) Compute each observation's OOB mean prediction.
(b) Compute the OOB MSE.

```{webr}
#| autorun: false
# p3-starter
y       <- c(5, 8, 6, 20, 25, 22)
oob_preds <- list(c(6, 7), c(6, 5, 7), c(7), c(22, 23, 21), c(23, 22), c(23, 24, 22))
```

::: {.callout-note collapse="true" title="Answer"}
(a) OOB means: average each row's predictions.

```{r}
y <- c(5, 8, 6, 20, 25, 22)
oob_preds <- list(c(6, 7), c(6, 5, 7), c(7), c(22, 23, 21), c(23, 22), c(23, 24, 22))
oob_means <- sapply(oob_preds, mean)
round(oob_means, 2)   # 6.5, 6, 7, 22, 22.5, 23
```

(b) OOB MSE = average of $(y_i - \text{OOB mean}_i)^2$:

```{r}
oob_mse <- mean((y - oob_means)^2)
round(oob_mse, 2)     # 3.08
```

$(5 - 6.5)^2 + (8 - 6)^2 + (6 - 7)^2 + (20 - 22)^2 + (25 - 22.5)^2 + (22 - 23)^2$
$= 2.25 + 4 + 1 + 4 + 6.25 + 1 = 18.5$. Divide by 6: **OOB MSE = 3.08**. This
approximates test error without needing a separate test set.
:::


## Problem 4 — spot the error

A colleague says: *"I trained a random forest with 200 trees and got 94% accuracy. Then I
tried 2000 trees and got 93.8%. The model is overfitting — I need to use fewer trees."*
This contains **exactly one mistake** — find it.

::: {.callout-note collapse="true" title="Answer"}
The mistake: claiming the random forest is **overfitting** because more trees gave
slightly lower accuracy. Random forests (and bagging) do NOT overfit as $B$ increases —
each tree is trained independently on a random sample, so adding more trees just
stabilizes the average. The tiny accuracy drop from 94% to 93.8% is random noise in the
OOB or test estimate, not overfitting. The correct response is that 200 and 2000 trees
give essentially the same performance, confirming the error has settled. Overfitting from
too many trees is a boosting concern, not a random forest concern.
:::


## Problem 5 — interleaved *(uses resampling's universal loop pattern)*

Compare the bagging loop to the bootstrap CI loop from the resampling module:

(a) Both use `replicate(B, { ... })`. What does the "resample" step inside the loop do
in each case?
(b) What does the "summarize the pile" step do after the loop in each case?
(c) In one sentence, what is the key difference between the two uses of the same loop?

::: {.callout-note collapse="true" title="Answer"}
(a) Both resample the data **with replacement** (`sample(x, length(x), replace = TRUE)`).
In the bootstrap CI, you resample the data and compute a *statistic* (like the mean). In
bagging, you resample the data and *fit a tree then predict* at a new point.

(b) The bootstrap CI takes `sd()` of the pile for a standard error and
`quantile(pile, c(.025, .975))` for a 95% interval. Bagging takes `mean()` of the pile
for the final prediction — the average of all trees.

(c) The loop shape is identical; the difference is the *goal*: the bootstrap CI quantifies
the *uncertainty* of an estimate, while bagging uses the same resampling engine to produce
a *better* (more stable) prediction.
:::


## Problem 6 — transfer: boosting shrinkage

In the lesson's boosting demo ($\lambda = 0.1$), predictions after 2 iterations were
$(0.76, 0.76, 3.14, 3.14)$ — still far from the actual $y = (3, 5, 15, 18)$.

(a) If you changed $\lambda$ from 0.1 to 0.01, would you need more or fewer iterations
to reach the same accuracy? Why?
(b) Despite being slower, why is a smaller $\lambda$ sometimes preferred?

::: {.callout-note collapse="true" title="Answer"}
(a) **More iterations.** Each tree's contribution is scaled by $\lambda$, so with
$\lambda = 0.01$ each stump contributes ten times less per iteration than with
$\lambda = 0.1$. It takes roughly ten times as many iterations to reach the same
prediction accuracy.

(b) Smaller $\lambda$ (slower learning) often **generalizes better** to new data. Each
tree makes a smaller correction, so the model doesn't latch onto noise in the training
data as quickly. This is the same principle as regularization: constraining how fast the
model fits prevents overfitting. The cost is computational — more trees to train — but
the payoff is lower test error.
:::


## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/bagging-random-forests-and-boosting" pass="4"}

::: {.quiz-q qid="brfb-bag-effect" qtype="mc" answer="c"}
What does bagging do to individual tree predictions?

a) Prunes each tree aggressively before averaging their predictions together
b) Makes each tree deeper by adding more splits at every branching node
c) Averages many bootstrap trees, reducing the variance of predictions
:::

::: {.quiz-q qid="brfb-oob-fraction" qtype="mc" answer="b"}
Roughly what fraction of observations is left out of each bootstrap sample (the OOB observations)?

a) About one-half of the training observations are left out of each bootstrap sample
b) About one-third are left out of each bootstrap sample
c) About one-tenth of the training observations are left out of each bootstrap sample
:::

::: {.quiz-q qid="brfb-rf-vs-bag" qtype="mc" answer="a"}
What is the key difference between random forests and plain bagging?

a) At each split, random forests pick from a random subset of predictors
b) Random forests grow each individual tree to a shallower depth than bagging does
c) Random forests apply a shrinkage penalty to scale down each individual tree's contribution
:::

::: {.quiz-q qid="brfb-tree-variance" qtype="mc" answer="c"}
(From decision trees.) Why are single trees the natural building block for bagging?

a) Trees are the only model type that can work with bootstrap resampling of the data
b) Trees run faster than any other model type on bootstrapped datasets
c) Trees have high variance from instability, which averaging effectively reduces
:::

::: {.quiz-q qid="brfb-boost-target" qtype="mc" answer="a"}
In boosting, each new tree is fit to:

a) The residuals — what the current ensemble has not yet explained
b) A fresh bootstrap sample drawn from the original training data
c) The original response values, ignoring all of the previous trees
:::

:::

::::
