---
title: "Sherman–Morrison: update an inverse without redoing it"
subtitle: "Poke one rank-1 change into a matrix and fix its inverse cheaply — exactly"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

## Where we're going

Inverting a matrix is expensive. If you already have $A^{-1}$ and then make a *small*
change to $A$, redoing the whole inverse from scratch is wasteful. The **Sherman–Morrison
formula** patches $A^{-1}$ into the inverse of the changed matrix directly — and it is
**exact, not an approximation**. This short module builds the one identity, shows it's
exact to machine precision, counts why it's fast, and names the family it belongs to
(including the *bordering* trick that grows a matrix by one row and column — the exact move
a variable-selection search needs when it adds a predictor).

**Builds on:** [matrices-and-linear-transforms](../matrices-and-linear-transforms/lesson.qmd)
(matrix × vector, the transpose, the inverse as "undo").

## Warm-up: do you remember?

**1.** *(matrices)* What does the inverse $A^{-1}$ do, and how do you get it in R?

::: {.callout-note collapse="true" title="Answer"}
$A^{-1}$ is the matrix that undoes $A$: $A^{-1}A = I$ (the identity). In R: `solve(A)`.
See [matrices-and-linear-transforms](../matrices-and-linear-transforms/lesson.qmd).
:::

**2.** *(rank & linear independence)* If $u$ and $v$ are two length-$p$ column vectors, what shape is the
product $u v^\top$ ("u times v-transpose")?

::: {.callout-note collapse="true" title="Answer"}
A $p \times p$ matrix (an "outer product") — a column times a row. It is a **rank-1**
matrix: every row is a multiple of $v^\top$. In R: `u %*% t(v)`.
Shaky? Revisit [rank & linear independence](../rank-and-linear-independence/lesson.qmd),
whose "Rank-1: a column times a row" section teaches exactly this.
:::

## A tiny example first

Start with $A = \begin{pmatrix} 2 & 0 \\ 0 & 2 \end{pmatrix}$, whose inverse is easy:
$A^{-1} = \begin{pmatrix} 0.5 & 0 \\ 0 & 0.5 \end{pmatrix}$.

Now bump the top-left entry by 1, using $u = v = (1, 0)$: the change $u v^\top$ is
$\begin{pmatrix} 1 & 0 \\ 0 & 0 \end{pmatrix}$, so
$A + u v^\top = \begin{pmatrix} 3 & 0 \\ 0 & 2 \end{pmatrix}$, whose inverse is just the
reciprocals down the diagonal, $\begin{pmatrix} 1/3 & 0 \\ 0 & 0.5 \end{pmatrix}$.

Sherman–Morrison gets that *same* answer from $A^{-1}$ without re-inverting.

**PREDICT:** $A + uv^\top$ bumps the (1,1) entry to 3. What top-left value should its
inverse have? Write your guess, then run the cell to check.

```{webr}
#| autorun: false
# PREDICT: what top-left value should the inverse of diag(3,2) have?
# Write your guess above, then run to check.
A <- matrix(c(2, 0, 0, 2), 2); u <- c(1, 0); v <- c(1, 0)
Ai <- solve(A)                                   # the inverse we already have

# Sherman-Morrison: A^{-1} minus a rank-1 correction
correction <- (Ai %*% u %*% t(v) %*% Ai) / as.numeric(1 + t(v) %*% Ai %*% u)
sm <- Ai - correction

list(sherman_morrison = sm, direct = solve(A + u %*% t(v)),
     identical = isTRUE(all.equal(sm, solve(A + u %*% t(v)))))   # TRUE
```

::: {.callout-note collapse="true" title="Answer"}
Plug into the formula: $A^{-1}u = (0.5, 0)$, $v^\top A^{-1} u = 0.5$, so the correction
is the rank-1 matrix $\begin{pmatrix} 0.25 & 0 \\ 0 & 0\end{pmatrix}$ divided by
$1 + 0.5 = 1.5$, i.e. $\begin{pmatrix} 1/6 & 0 \\ 0 & 0 \end{pmatrix}$. Subtract from
$A^{-1}$: $0.5 - 1/6 = 1/3$ in the top-left, everything else unchanged --
$\begin{pmatrix} 1/3 & 0 \\ 0 & 0.5 \end{pmatrix}$. Exactly right.
:::

## Getting a feel for it

**Patch the pothole, don't re-pave the road.** Computing $A^{-1}$ from scratch is
re-paving an entire road. A rank-1 change $u v^\top$ is one pothole. Sherman–Morrison
patches just that spot — it adjusts the inverse with a single cheap correction built from
$A^{-1}$ you already have, instead of rebuilding the whole thing.

**Why it's exact, not approximate.** This is the misconception to kill up front: the
formula is an *algebraic identity*, not a first-order approximation or an iterative guess.
It comes from multiplying $(A + uv^\top)$ by the claimed inverse and watching everything
cancel to $I$. So it matches a fresh inverse to machine precision (`all.equal` returns
`TRUE` above) — there is no accuracy penalty for the speed.

**Why it's fast.** A fresh inverse grows cubically with the matrix size — doubling the
matrix size multiplies the work by roughly eight. The Sherman–Morrison update is just a
few matrix-times-vector products, which grow only quadratically — doubling the size only
multiplies the work by four. The gap widens fast as the matrix gets bigger, and you'll
time it below to see the payoff firsthand.

## What the notation is going to say

The formula will say: *the new inverse is the old inverse, minus a rank-1 correction
made from the old inverse applied to $u$ and to $v$, all divided by a single number*
$1 + v^\top A^{-1} u$. If that number is not zero, the update exists and is exact.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $A$ | "ay" | the matrix we already inverted | $\left(\begin{smallmatrix}2&0\\0&2\end{smallmatrix}\right)$ |
| $A^{-1}$ | "A inverse" | its inverse, already computed (in R: `solve(A)`) | $\left(\begin{smallmatrix}.5&0\\0&.5\end{smallmatrix}\right)$ |
| $u,\ v$ | "you", "vee" | the two vectors defining the rank-1 change | both $(1,0)$ |
| $u v^\top$ | "u v-transpose" | the rank-1 change to $A$ (a column times a row) | bumps entry (1,1) by 1 |
| $v^\top A^{-1} u$ | "v-transpose A-inverse u" | a single scalar; $1$ plus it is the divisor | $0.5$ |
| $I$ | "eye" | the identity matrix ($A^{-1}A = I$; the "do nothing" matrix) | — |
| $O(p^2),\ O(p^3)$ | "order p-squared / p-cubed" | how the work grows with matrix size $p$ | update vs fresh |

## The formal version

For an invertible $A$ and vectors $u, v$, **as long as $1 + v^\top A^{-1} u \neq 0$**:

$$
(A + u v^\top)^{-1} = A^{-1} - \frac{A^{-1} u\, v^\top A^{-1}}{1 + v^\top A^{-1} u}.
$$

Reading it aloud: "the inverse of A plus u-v-transpose equals A-inverse, minus (A-inverse
u, times v-transpose A-inverse) over one plus v-transpose A-inverse u" — or in everyday
words: "take the inverse you already have and subtract one cheap rank-1 patch."

- The numerator $A^{-1}u\,v^\top A^{-1}$ is itself a rank-1 matrix — a column
  ($A^{-1}u$) times a row ($v^\top A^{-1}$) — so the whole correction is built from two
  matrix–vector products, never a full inversion.
- The denominator $1 + v^\top A^{-1} u$ is a single number. If it were zero, $A + uv^\top$
  would be singular (no inverse) and the formula would (correctly) blow up.

**Sanity check (toy numbers):** with $A^{-1}u = (0.5,0)$ and $v^\top A^{-1} u = 0.5$, the
formula gave the top-left entry $0.5 - \tfrac{0.25}{1.5} = 0.5 - \tfrac16 = \tfrac13$,
matching the obvious inverse of $\left(\begin{smallmatrix}3&0\\0&2\end{smallmatrix}\right)$.

::: {.callout-note title="Optional: the family — Woodbury and the bordering update"}
Sherman–Morrison is the rank-1 case of the **Woodbury identity**, which updates an inverse
after a rank-$k$ change. A close relative, the **bordering** (block-inverse) update, grows
a matrix by a whole new row *and* column and produces the enlarged inverse from the old
one — and a matching **partition** formula shrinks it by one. A variable-selection search
uses exactly these: adding a predictor borders the correlation matrix by one row/column,
removing one partitions it away, and a swap is a remove-then-add. The
swap-stepwise-selection course derives and uses them; here we just verify one
numerically:

```{r bordering}
set.seed(1); k <- 4
M <- crossprod(matrix(rnorm(k * k), k)); Mi <- solve(M)   # a k x k matrix + its inverse
b <- rnorm(k); d <- rnorm(1)^2 + 2                        # the new column/row and corner
Mnew <- rbind(cbind(M, b), c(b, d))                       # grown (k+1) x (k+1) matrix

s   <- as.numeric(d - t(b) %*% Mi %*% b)                  # Schur complement (a scalar)
Mib <- Mi %*% b
border <- rbind(cbind(Mi + (Mib %*% t(Mib)) / s, -Mib / s), c(-t(Mib) / s, 1 / s))
max(abs(border - solve(Mnew)))                            # ~1e-15: exact, no re-inversion
```
:::

## Explore it

```{r timing}
# Invert a 200x200 matrix ONCE, then compare: 1000 fresh solve()s of a rank-1-changed
# matrix vs 1000 Sherman-Morrison updates. (Your exact times vary by machine; the gap is
# the point.) This is baked at build time — a heavy timing loop, not a browser cell.
set.seed(2); p <- 200
A <- crossprod(matrix(rnorm(p * p), p)); Ai <- solve(A)   # the one expensive inverse
u <- rnorm(p); v <- rnorm(p)

t_fresh <- system.time(for (i in 1:1000) solve(A + u %*% t(v)))[["elapsed"]]
t_update <- system.time(for (i in 1:1000) {
  Aiu <- Ai %*% u
  Ai - (Aiu %*% (t(v) %*% Ai)) / as.numeric(1 + t(v) %*% Aiu)
})[["elapsed"]]
c(fresh_solve_s = round(t_fresh, 2), sm_update_s = round(t_update, 2),
  speedup = round(t_fresh / t_update, 1))
```

```{webr}
#| autorun: true
# TRY IT: change u and v (any vectors) and re-run — the update always matches a fresh
# solve to ~1e-15. PREDICT: will `identical` ever come back FALSE for a non-singular case?
set.seed(9); p <- 25
A <- crossprod(matrix(rnorm(p * p), p)); Ai <- solve(A)
u <- rnorm(p); v <- rnorm(p)

Aiu <- Ai %*% u
sm  <- Ai - (Aiu %*% (t(v) %*% Ai)) / as.numeric(1 + t(v) %*% Aiu)
c(max_diff = max(abs(sm - solve(A + u %*% t(v)))),
  identical = isTRUE(all.equal(sm, solve(A + u %*% t(v)))))
```

## Common traps

**Trap: "Sherman–Morrison is an approximation, so it drifts."** No — it is an exact
algebraic identity. The `all.equal` checks above return `TRUE` and the max difference is
~$10^{-15}$ (ordinary floating-point rounding, the same you'd get comparing two ways of
computing any inverse). There is no approximation error to accumulate.

**Trap: "it always works."** It fails in exactly one case: when $1 + v^\top A^{-1} u = 0$,
the denominator is zero — and that is precisely when $A + uv^\top$ is **singular** and has
no inverse at all. The formula doesn't break the math; it faithfully reports "no inverse
here." Try $A = I_2$, $u = (1,0)^\top$, $v = (-1,0)^\top$: then $v^\top A^{-1}u = -1$, so
the denominator $1 + v^\top A^{-1}u = 0$ — the formula gives `NaN`. And indeed
`det(A + u %*% t(v))` is zero: the updated matrix is singular.

## Check yourself

**1.** In the update formula, what shape is the correction term $\dfrac{A^{-1}u\,v^\top A^{-1}}{1 + v^\top A^{-1} u}$
— matrix or scalar — and what is the denominator?

::: {.callout-note collapse="true" title="Answer"}
The correction is a **matrix** (rank-1: a column $A^{-1}u$ times a row $v^\top A^{-1}$).
The denominator $1 + v^\top A^{-1} u$ is a **scalar** (a single number). You subtract the
matrix from $A^{-1}$.
:::

**2.** Roughly how does the cost of a fresh $p \times p$ inverse compare to a
Sherman–Morrison update, in big-O terms, and why does the gap matter in a loop?

::: {.callout-note collapse="true" title="Answer"}
Fresh inverse is $O(p^3)$; the update is $O(p^2)$ (a few matrix–vector products). In a
search that changes the matrix thousands of times, replacing each $O(p^3)$ re-inversion
with an $O(p^2)$ update is the difference between minutes and seconds.
:::

**3.** In your own words: explain to a colleague, in two sentences, what Sherman–Morrison
buys you and why it doesn't cost accuracy.

::: {.callout-note collapse="true" title="A model answer"}
If you already have a matrix's inverse and make a small (rank-1) change, Sherman–Morrison
gives you the new inverse with one cheap correction instead of re-inverting from scratch.
It costs no accuracy because it's an exact algebraic identity, not an approximation — it
matches a fresh inverse to floating-point precision.
:::

## What you can now do

- **State the Sherman–Morrison formula** and identify its rank-1 correction and its scalar
  denominator.
- **Explain that it is exact**, and verify it against a fresh `solve()` with `all.equal`.
- **Give the cost argument** ($O(p^3)$ fresh vs $O(p^2)$ update) and why it matters in a
  loop.
- **Name the family** (Woodbury; the bordering/partition updates that grow or shrink a
  matrix by one row+column) that a variable-selection search uses to add/remove predictors.

If any feels shaky, revisit [The formal version](#the-formal-version) — the switching
search reuses this identity on every move.

## Where this goes next

A model search that adds and drops predictors re-inverts a correlation matrix on every
step; the `swap-stepwise-selection` course replaces each re-inversion with a
bordering/Sherman–Morrison update and times the payoff. Ready for reps? Do the
[practice problems](practice.qmd). External readings: [resources](resources.md).
