---
title: "Practice: Sherman–Morrison rank-1 updates"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Do them on paper or in an R session; only then
open the answer.

## Problem 1 — one update by hand, new numbers

Let $A = \begin{pmatrix} 4 & 0 \\ 0 & 1 \end{pmatrix}$ (so
$A^{-1} = \begin{pmatrix} 0.25 & 0 \\ 0 & 1 \end{pmatrix}$), and bump the bottom-right
entry by 1 with $u = v = (0, 1)$. Use Sherman–Morrison to find $(A + uv^\top)^{-1}$, then
check against the obvious inverse.

```{webr}
#| autorun: false
# p1-starter
A <- matrix(c(4, 0, 0, 1), 2); u <- c(0, 1); v <- c(0, 1)
```

::: {.callout-note collapse="true" title="Answer"}
$A^{-1}u = (0, 1)$, $v^\top A^{-1} u = 1$, denominator $= 1 + 1 = 2$. The rank-1
correction is $\begin{pmatrix}0&0\\0&1\end{pmatrix}/2 = \begin{pmatrix}0&0\\0&0.5\end{pmatrix}$.
Subtract from $A^{-1}$: bottom-right becomes $1 - 0.5 = 0.5$, so
$(A+uv^\top)^{-1} = \begin{pmatrix}0.25&0\\0&0.5\end{pmatrix}$ — the inverse of
$\begin{pmatrix}4&0\\0&2\end{pmatrix}$, as expected.
:::

## Problem 2 — fill in the missing step

Applying Sherman–Morrison, someone has computed $A^{-1}u = (2, -1)$, $v^\top A^{-1} = (2, -1)$
(as a row), and $v^\top A^{-1} u = 5$.

- Step 1: numerator (rank-1 matrix) $= (A^{-1}u)(v^\top A^{-1}) = \begin{pmatrix}2\\-1\end{pmatrix}(2, -1) = \begin{pmatrix}4 & -2\\ -2 & 1\end{pmatrix}$.
- Step 2: **your step here** — form the full correction term.
- Step 3: subtract it from $A^{-1}$ to get the updated inverse.

::: {.callout-note collapse="true" title="Answer"}
Step 2: divide the numerator by the scalar $1 + v^\top A^{-1}u = 1 + 5 = 6$:
$\text{correction} = \dfrac{1}{6}\begin{pmatrix}4 & -2\\ -2 & 1\end{pmatrix}
= \begin{pmatrix}2/3 & -1/3\\ -1/3 & 1/6\end{pmatrix}$. The easy-to-forget move is dividing
by $1 + v^\top A^{-1}u$ (not by $v^\top A^{-1}u$ alone).
:::

## Problem 3 — it matches `solve()`, in R

For the random 5×5 matrix below, update its inverse for the rank-1 change $u v^\top$ using
Sherman–Morrison, and confirm it matches a fresh `solve(A + u %*% t(v))`.

```{webr}
#| autorun: false
# p3-starter — build the SM update and compare to a fresh solve
set.seed(4); p <- 5
A <- crossprod(matrix(rnorm(p * p), p)); Ai <- solve(A)
u <- rnorm(p); v <- rnorm(p)
```

::: {.callout-note collapse="true" title="Answer"}
The update reproduces a fresh inverse to ~$10^{-15}$.

```{r}
set.seed(4); p <- 5
A <- crossprod(matrix(rnorm(p * p), p)); Ai <- solve(A); u <- rnorm(p); v <- rnorm(p)
Aiu <- Ai %*% u
sm  <- Ai - (Aiu %*% (t(v) %*% Ai)) / as.numeric(1 + t(v) %*% Aiu)
max(abs(sm - solve(A + u %*% t(v))))      # ~1e-15
```
:::

## Problem 4 — spot the error

A colleague writes: *"Sherman–Morrison is only an approximation, so for anything important
I re-invert from scratch to be safe — the update would drift."* This contains **exactly
one mistake** — find it.

::: {.callout-note collapse="true" title="Answer"}
The mistake: calling it an approximation that "drifts." Sherman–Morrison is an **exact
algebraic identity** — it equals a fresh inverse to floating-point precision (`all.equal`
returns `TRUE`; the difference is ~$10^{-15}$, the same rounding you'd see between any two
computations of an inverse). There is no drift and no accuracy reason to re-invert; the
only reason to re-invert would be a zero denominator (a singular result).
:::

## Problem 5 — mixing it with matrices *(uses matrices-and-linear-transforms' outer product)*

Without any inverse: for $u = (1, 2, 0)$ and $v = (3, 0, 1)$, write out the rank-1 matrix
$u v^\top$, and state its rank.

```{webr}
#| autorun: false
u <- c(1, 2, 0)
v <- c(3, 0, 1)
```

::: {.callout-note collapse="true" title="Answer"}
$u v^\top = \begin{pmatrix}1\\2\\0\end{pmatrix}(3, 0, 1)
= \begin{pmatrix} 3 & 0 & 1 \\ 6 & 0 & 2 \\ 0 & 0 & 0 \end{pmatrix}$. Its **rank is 1** —
every row is a multiple of $(3, 0, 1)$ (and every column a multiple of $u$). That is why
the change is called "rank-1," and why the correction to the inverse is also rank-1.
:::

## Problem 6 — transfer: when the update refuses

Sherman–Morrison has a denominator $1 + v^\top A^{-1} u$. Construct (in words) a situation
where this is zero, say what it means about $A + uv^\top$, and why that's the *right*
behavior. *(In the next course, this is what a ridge $\lambda$ prevents.)*

::: {.callout-note collapse="true" title="Answer"}
If $v^\top A^{-1} u = -1$, the denominator is 0 and the formula blows up. That's not a bug:
it happens exactly when the rank-1 change $u v^\top$ makes $A + u v^\top$ **singular** (no
inverse exists), so there is nothing for the formula to return. The update faithfully
reports "no inverse here." In a variable-selection search, near-singular correlation
matrices make this denominator tiny and the update unstable — which is why the next course
adds a ridge $\lambda$ to the diagonal to keep the matrix comfortably invertible.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/sherman-morrison-rank-one-updates" pass="4"}

::: {.quiz-q qid="smru-exact" qtype="mc" answer="b"}
Is the Sherman–Morrison update an approximation?

a) Yes — small errors accumulate over many updates
b) No — it is an exact algebraic identity; differences are only $\sim 10^{-15}$ floating-point rounding
c) Yes — it is a first-order Taylor approximation
:::

::: {.quiz-q qid="smru-fails-when" qtype="mc" answer="a"}
Sherman–Morrison fails in exactly one case. Which?

a) When $1 + v^\top A^{-1} u = 0$ — precisely when $A + uv^\top$ is singular and has no inverse
b) When the update vector $u$ is longer than $v$
c) When $A$ is symmetric
:::

::: {.quiz-q qid="smru-cost" qtype="mc" answer="c"}
Why use Sherman–Morrison instead of recomputing the inverse from scratch after a rank-1
change?

a) It is more accurate than a fresh solve
b) A fresh inverse cannot handle rank-1 changes
c) It costs $O(p^2)$ per update versus $O(p^3)$ for a fresh inverse — a big saving inside a loop
:::

::: {.quiz-q qid="smru-woodbury" qtype="mc" answer="a"}
Sherman–Morrison generalizes to higher-rank updates under what name?

a) The Woodbury identity
b) The quadratic formula
c) Gaussian elimination
:::

::: {.quiz-q qid="smru-singular" qtype="mc" answer="b"}
(From matrices & linear transforms.) Sherman–Morrison fails exactly when the updated
matrix is singular. What does "singular" mean?

a) The matrix is symmetric
b) The matrix has no inverse (redundant rows or columns)
c) The matrix is very large
:::

:::

::::
