---
title: "The pairwise correction"
subtitle: "One test is weak, two selected variants is biased — the paper's fix runs every pair and corrects for multiplicity"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live cells below re-state their own data. Base R only.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

::: {.callout-tip title="This lesson runs in your browser" collapse="true"}
The code cells below with a **Run** button are live — powered by WebR (R compiled to
WebAssembly), they run right here on your phone or laptop, no install needed. Edit any of them
and re-run. The first run downloads a small R engine, so give it a few seconds. Cells without a
Run button are ordinary listings whose output was computed when the page was built.
:::

## Where we're going

Module 4 gave you a working pairwise test — but a single pairwise test, with only $n - 1 = 1$
degree of freedom, is *weak*: it often can't reject even when it should. The paper describes two
tempting-but-flawed fixes, then its own. Using **all $n$ variants** in one test is still weak
(an $n-1$ df test spreads thin). Using the **two most-associated variants** is powerful but
*biased* — selecting the variants you test on the same data you test with inflates the false-
positive rate. The paper's compromise: run the pairwise test on **all $n(n-1)/2$ pairs**, then
apply a **multiplicity correction** over the total number of possible pairs, and reject
colocalisation if **any** corrected pair is significant. A neat bonus — the test's p-value is
bounded below by the smallest marginal GWAS p-value — lets you test only a subset. You'll build
`prop_run()` and **self-check it live** against the frozen shared (not rejected) and distinct
(rejected) references.

**Builds on:** [Module 4 (the proportional test statistic)](../04-the-proportional-test-statistic/lesson.qmd)
and the foundation
[the false discovery rate](../../../../foundations/false-discovery-rate/lesson.qmd) (multiple
testing, `p.adjust`, FWER vs FDR, Holm vs Benjamini–Hochberg).

## Warm-up: do you remember?

Three quick reps first — produce each from memory, then check.

**1.** *(Module 4)* What does `prop_test_pair()` return for a pair, and how many degrees of
freedom does its chi-square have?

::: {.callout-note collapse="true" title="Answer"}
It returns the chi-square statistic, the estimated slope $\hat\eta$, and the p-value for the null
of proportionality; the chi-square has $n - 1 = 1$ degree of freedom for a pair. See
[Module 4](../04-the-proportional-test-statistic/lesson.qmd).
:::

**2.** *(foundation)* If you run $m$ independent tests each at the 0.05 threshold and nothing is
real, roughly what's the chance of at least one false alarm as $m$ grows?

::: {.callout-note collapse="true" title="Answer"}
It climbs toward 1: $1 - 0.95^m$. For $m = 15$ that's about $0.54$. Running many tests without
correction guarantees false alarms — hence `p.adjust`. See
[the false discovery rate](../../../../foundations/false-discovery-rate/lesson.qmd).
:::

**3.** *(foundation)* What does `p.adjust(p)` do by **default** (no `method =` argument), and is
that a family-wise or a false-discovery-rate correction?

::: {.callout-note collapse="true" title="Answer"}
By default `p.adjust` uses **Holm**, a **family-wise** error correction (controls the chance of
*any* false alarm) — *not* Benjamini–Hochberg (which is `method = "BH"`). See
[the false discovery rate](../../../../foundations/false-discovery-rate/lesson.qmd). This
distinction matters at the end of this lesson.
:::

## A tiny example first

A region with $n = 4$ SNPs. There are $\binom{4}{2} = \frac{4\cdot 3}{2} = 6$ possible pairs:
(1,2), (1,3), (1,4), (2,3), (2,4), (3,4). The recipe:

1. Run `prop_test_pair()` on each of the 6 pairs → 6 raw p-values.
2. Correct all 6 for multiplicity with `p.adjust`, using the total number of possible pairs
   ($\text{ntests} = 6$) as the denominator.
3. **Reject colocalisation if the smallest corrected p-value is below your threshold** (any one
   significant pair is enough).

Why "reject if *any* pair is significant"? Because colocalisation (the null) requires *every*
pair to be proportional — the effects lie on one line through the origin. If even one pair
provably departs from proportionality, the whole region can't colocalise. So a single significant
pair breaks the null. And because we're now doing several tests, we must correct, or we'd
"discover" non-colocalisation by chance.

Concretely: say the 6 raw p-values are $(0.9, 0.5, 0.8, 0.7, 0.8, 0.9)$ — all large, nothing
proportional-breaking. Correct them (Holm, $\text{ntests} = 6$) and the minimum corrected value
stays near 1 → **fail to reject** → colocalised. But if one pair returned a tiny raw p (a real
departure), its corrected value would survive and drive the minimum below 0.05 → **reject** →
not colocalised.

```{r toy-example}
# 4 SNPs -> 6 possible pairs
n <- 4
pairs <- combn(n, 2)                     # 2 x 6: every (i, j) with i < j
pairs
n_possible <- ncol(pairs)                # 6 = n(n-1)/2
n_possible

# Suppose these are the 6 raw pairwise p-values (all large -> nothing breaks proportionality)
raw_p <- c(0.9, 0.5, 0.8, 0.7, 0.8, 0.9)
adjusted <- p.adjust(raw_p, method = "holm", n = n_possible)   # correct over all possible pairs
min(adjusted)                            # near 1 -> fail to reject -> COLOCALISED

# Now imagine one pair really departs (tiny raw p)
raw_p2 <- c(0.9, 0.5, 1e-5, 0.7, 0.8, 0.9)
min(p.adjust(raw_p2, method = "holm", n = n_possible))         # < 0.05 -> reject -> NOT colocalised
```

## Getting a feel for it

**Analogy — a chain is only as strong as its weakest link.** Colocalisation claims the whole
region's effects lie on one line — every pair of SNPs must be consistent with that line, like
every link in a chain holding. To *disprove* the chain, you don't need every link to fail; **one
broken link** is enough. So you inspect each link (each pairwise test) and declare the chain
broken (not colocalised) if any single link snaps. But you're inspecting many links, so some will
*look* cracked by chance — that's why you correct for multiplicity (the false-discovery
foundation): only count a link as truly broken after adjusting for how many you inspected.

Where the analogy breaks: the links aren't independent — pairs share SNPs and the SNPs are in LD,
so the pairwise p-values are correlated. The correction (Holm/Bonferroni over the total possible
pairs) is deliberately conservative to stay safe under that dependence.

**Why not just one big $n$-variant test?** Because an $n$-variant proportional test has $n - 1$
degrees of freedom, and a high-df chi-square needs a *lot* of departure to reach significance —
the evidence from one badly non-proportional pair gets diluted across all $n-1$ dimensions. The
pairwise approach concentrates the evidence: the one non-proportional pair gets its own
low-df, high-power test, and the correction lets you still combine them honestly.

**Why not just pick the two strongest variants?** Because you'd be choosing *which* variants to
test using the same data you then test — a selection bias that inflates the false-positive rate
(the paper cites this being shown to badly break Type I error control). Testing *all* pairs and
correcting sidesteps the selection entirely: you never cherry-pick.

**The min-marginal-p bonus.** The paper shows the proportional test's p-value for a pair can never
be smaller than the smallest single-SNP GWAS p-value among the variants in that pair. So a pair
of two strongly-associated SNPs is the only kind that *can* return a tiny proportional p — pairs
involving a null SNP are bounded away from significance. That means you only need to run the test
on the handful of well-associated variants, as long as you still use the *total* number of
possible pairs as the correction denominator. Fewer tests, same guarantee.

## What the notation is going to say

The formula is bookkeeping over Module 4's pairwise test: enumerate all pairs, run
`prop_test_pair()` on each to get a vector of raw p-values, call `p.adjust(p, n = ntests)` with
$\text{ntests} = n(n-1)/2$ the total possible pairs, and report the *minimum* adjusted p-value —
reject colocalisation if it's below threshold. No new statistic; the new idea is "test all pairs,
correct over all possible pairs, reject on the weakest link."

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $n$ | "en" | number of SNPs in the region | $4$ |
| $\binom{n}{2}$ | "n choose 2" | number of possible pairs, $= n(n-1)/2$ | $6$ |
| $\text{pairs}$ | "pairs" | the $2\times m$ table of index pairs $(i, j)$ to test (in R: `combn(n, 2)`) | 6 columns |
| $\text{ntests}$ | "en-tests" | the correction denominator: total possible pairs | $6$ |
| $p_m$ | "p sub em" | the raw pairwise p-value for the $m$-th pair | one per pair |
| $\tilde p_m$ | "p sub em tilde" | the multiplicity-adjusted p-value (in R: `p.adjust`) | one per pair |
| $\min \tilde p$ | "min p-tilde" | the smallest adjusted p-value; reject colocalisation if it's small | the decision number |

## The formal version

**The decision rule.** For a region of $n$ SNPs, over all $M = n(n-1)/2$ pairs indexed by $m$:

$$
p_m = \text{prop\_test\_pair}(\text{pair } m), \qquad
\tilde p_m = \text{p.adjust}(p_1, \dots; \ n = M)_m, \qquad
\text{reject colocalisation} \iff \min_m \tilde p_m < \alpha .
$$

Reading it aloud: "p-sub-m is the pairwise proportional p-value for pair m; p-tilde-sub-m is those
p-values adjusted for M total tests; reject colocalisation if the smallest adjusted p-value is
below alpha." Everyday words: "test every pair, correct for how many pairs are possible, and call
the region non-colocalised if the strongest single departure survives correction."

**Mapping it onto the toy example.** With $n = 4$, $M = 6$, all-large raw p's gave
$\min \tilde p \approx 1$ (fail to reject → colocalised), while a single tiny raw p ($10^{-5}$)
gave $\min \tilde p < 0.05$ (reject → not colocalised). Now build `prop_run()`, the loop-and-
correct wrapper around Module 4's `prop_test_pair()`:

```{r formal-check}
prop_V <- function(vbeta, rho) { V <- diag(vbeta); V[1,2] <- V[2,1] <- sqrt(vbeta[1]*vbeta[2])*rho; V }
prop_test_pair <- function(beta1, vbeta1, beta2, vbeta2, rho1, rho2 = rho1) {
  V1 <- prop_V(vbeta1, rho1); V2 <- prop_V(vbeta2, rho2)
  chisq <- function(theta) { d <- sin(theta)*beta1 - cos(theta)*beta2
    Vs <- sin(theta)^2*V1 + cos(theta)^2*V2; as.numeric(t(d) %*% solve(Vs, d)) }
  oL <- optimize(chisq, c(0, pi/2)); oR <- optimize(chisq, c(pi/2, pi))
  o <- if (oL$objective < oR$objective) oL else oR
  list(chisquare = o$objective, p = pchisq(o$objective, df = length(beta1) - 1L, lower.tail = FALSE))
}

# Run the pairwise test over a supplied set of index pairs and correct over ntests total pairs
prop_run <- function(beta1, vbeta1, beta2, vbeta2, LD, pairs,
                     ntests = NULL, method = "holm") {
  n <- length(beta1)
  if (is.null(ntests)) ntests <- n * (n - 1) / 2       # default: all possible pairs
  chi <- numeric(ncol(pairs)); p <- numeric(ncol(pairs))
  for (m in seq_len(ncol(pairs))) {
    i <- pairs[1, m]; j <- pairs[2, m]
    r <- prop_test_pair(beta1[c(i, j)], vbeta1[c(i, j)],
                        beta2[c(i, j)], vbeta2[c(i, j)], LD[i, j])   # LD between this pair
    chi[m] <- r$chisquare; p[m] <- r$p
  }
  adjusted <- p.adjust(p, method = method, n = ntests)
  list(chisquare = chi, p = p, adjusted = adjusted, min_adjusted = min(adjusted))
}
# (self-checked against the frozen references below)
```

## Explore it

```{webr}
#| autorun: true
# TRY IT: change n and watch the number of possible pairs grow quadratically. PREDICT: how many
# pairs for n = 6? (15.) For n = 10? (45.)
for (n in c(4, 6, 10, 20)) cat("n =", n, "-> possible pairs =", n*(n-1)/2, "\n")
```

```{webr}
#| autorun: true
#| fig-alt: "A bar chart of six adjusted pairwise p-values. In the shared scenario all bars are near 1 (fail to reject); in the distinct scenario one bar is near zero, driving the minimum below the significance line."
# TRY IT: two sets of adjusted p-values -- shared (all large) vs distinct (one tiny). PREDICT:
# which scenario's MINIMUM crosses below 0.05 (the reject line)?
shared_adj   <- c(1, 1, 1, 1, 1, 1)                    # no pair breaks proportionality
distinct_adj <- c(1, 1, 2e-8, 1, 1, 1)                 # one pair departs strongly
barplot(rbind(shared = shared_adj, distinct = distinct_adj), beside = TRUE,
        ylim = c(0, 1.1), names.arg = paste0("pair", 1:6),
        col = c("steelblue", "tomato"), ylab = "adjusted p-value",
        main = "Reject colocalisation if ANY adjusted p dips below the line")
abline(h = 0.05, lty = 2)
legend("topright", fill = c("steelblue", "tomato"),
       legend = c("shared (fail to reject)", "distinct (one pair rejects)"), bty = "n")
```

### Self-check 1 — the shared region (not rejected) {#self-check-shared}

The frozen `equivalence/fixtures/propcoloc-run-shared.json`: 6 SNPs, all 15 pairs, LD $0.7$,
$\text{ntests} = 15$. The signal is genuinely shared, so no pair should break proportionality —
the reference `min_adjusted` is $1$ (fail to reject → colocalised). Your `prop_run()` should match
to about $10^{-6}$.

```{webr}
#| autorun: true
# LIVE SELF-CHECK (shared): prop_run() vs the FROZEN colocPropTest::run_proptests reference.
prop_V <- function(vbeta, rho) { V <- diag(vbeta); V[1,2] <- V[2,1] <- sqrt(vbeta[1]*vbeta[2])*rho; V }
prop_test_pair <- function(beta1, vbeta1, beta2, vbeta2, rho1, rho2 = rho1) {
  V1 <- prop_V(vbeta1, rho1); V2 <- prop_V(vbeta2, rho2)
  chisq <- function(theta) { d <- sin(theta)*beta1 - cos(theta)*beta2
    Vs <- sin(theta)^2*V1 + cos(theta)^2*V2; as.numeric(t(d) %*% solve(Vs, d)) }
  oL <- optimize(chisq, c(0, pi/2)); oR <- optimize(chisq, c(pi/2, pi))
  o <- if (oL$objective < oR$objective) oL else oR
  list(chisquare = o$objective, p = pchisq(o$objective, df = length(beta1)-1L, lower.tail = FALSE))
}
prop_run <- function(beta1, vbeta1, beta2, vbeta2, LD, pairs, ntests = NULL, method = "holm") {
  n <- length(beta1); if (is.null(ntests)) ntests <- n*(n-1)/2
  chi <- numeric(ncol(pairs)); p <- numeric(ncol(pairs))
  for (m in seq_len(ncol(pairs))) { i <- pairs[1,m]; j <- pairs[2,m]
    r <- prop_test_pair(beta1[c(i,j)], vbeta1[c(i,j)], beta2[c(i,j)], vbeta2[c(i,j)], LD[i,j])
    chi[m] <- r$chisquare; p[m] <- r$p }
  adjusted <- p.adjust(p, method = method, n = ntests)
  list(chisquare = chi, p = p, adjusted = adjusted, min_adjusted = min(adjusted))
}
# frozen inputs, copied from equivalence/fixtures/propcoloc-run-shared.json
beta1 <- c(0.2934792701896923, 0.003326222732736133, -0.01010033789150867,
           -0.001208909373011913, 0.000494075258618496, 0.0114138766483409)
beta2 <- c(0.3123757971125193, 0.007052889898247461, 0.0199359295932678,
           0.008701789835589922, 0.0172386493197919, 0.002554704130661046)
vbeta1 <- rep(0.0009, 6); vbeta2 <- rep(0.001225, 6)
LD <- matrix(0.7, 6, 6); diag(LD) <- 1
pairs <- combn(6, 2)                                     # the 15 pairs (i < j)
ref_min_adjusted <- 1                                    # frozen reference

out <- prop_run(beta1, vbeta1, beta2, vbeta2, LD, pairs, ntests = 15)
cat("min adjusted p:", round(out$min_adjusted, 6), " -> reject colocalisation?",
    out$min_adjusted < 0.05, "\n")
cat("matches reference (", ref_min_adjusted, ") within 1e-6?",
    abs(out$min_adjusted - ref_min_adjusted) < 1e-6, "\n")   # TRUE -> COLOCALISED
```

### Self-check 2 — the distinct region (rejected) {#self-check-distinct}

The mirror: `equivalence/fixtures/propcoloc-run-distinct.json`, same shape but the signals are
distinct. One pair breaks proportionality hard, so the reference `min_adjusted` is
$\approx 2.04\times 10^{-8}$ (reject → not colocalised).

```{webr}
#| autorun: true
# LIVE SELF-CHECK (distinct): prop_run() vs the FROZEN reference.
prop_V <- function(vbeta, rho) { V <- diag(vbeta); V[1,2] <- V[2,1] <- sqrt(vbeta[1]*vbeta[2])*rho; V }
prop_test_pair <- function(beta1, vbeta1, beta2, vbeta2, rho1, rho2 = rho1) {
  V1 <- prop_V(vbeta1, rho1); V2 <- prop_V(vbeta2, rho2)
  chisq <- function(theta) { d <- sin(theta)*beta1 - cos(theta)*beta2
    Vs <- sin(theta)^2*V1 + cos(theta)^2*V2; as.numeric(t(d) %*% solve(Vs, d)) }
  oL <- optimize(chisq, c(0, pi/2)); oR <- optimize(chisq, c(pi/2, pi))
  o <- if (oL$objective < oR$objective) oL else oR
  list(chisquare = o$objective, p = pchisq(o$objective, df = length(beta1)-1L, lower.tail = FALSE))
}
prop_run <- function(beta1, vbeta1, beta2, vbeta2, LD, pairs, ntests = NULL, method = "holm") {
  n <- length(beta1); if (is.null(ntests)) ntests <- n*(n-1)/2
  chi <- numeric(ncol(pairs)); p <- numeric(ncol(pairs))
  for (m in seq_len(ncol(pairs))) { i <- pairs[1,m]; j <- pairs[2,m]
    r <- prop_test_pair(beta1[c(i,j)], vbeta1[c(i,j)], beta2[c(i,j)], vbeta2[c(i,j)], LD[i,j])
    chi[m] <- r$chisquare; p[m] <- r$p }
  adjusted <- p.adjust(p, method = method, n = ntests)
  list(chisquare = chi, p = p, adjusted = adjusted, min_adjusted = min(adjusted))
}
# frozen inputs, copied from equivalence/fixtures/propcoloc-run-distinct.json
beta1 <- c(0.2773643146945664, -0.02213648545494297, -0.0228141260938878,
           -0.03025834543924165, -0.02386502276581572, -0.03947348135824345)
beta2 <- c(-0.01553201036207071, -0.03848563446746702, -0.01816800749239067,
           0.2847805230112141, -0.02141569115469859, -0.005980808079645841)
vbeta1 <- rep(0.0009, 6); vbeta2 <- rep(0.001225, 6)
LD <- matrix(0.7, 6, 6); diag(LD) <- 1
pairs <- combn(6, 2)
ref_min_adjusted <- 2.044213129215881e-08

out <- prop_run(beta1, vbeta1, beta2, vbeta2, LD, pairs, ntests = 15)
cat("min adjusted p:", format(out$min_adjusted, scientific = TRUE),
    " -> reject colocalisation?", out$min_adjusted < 0.05, "\n")
cat("matches reference within 1e-6?", abs(out$min_adjusted - ref_min_adjusted) < 1e-6, "\n")  # TRUE
```

**Validation tier: FULL.** Both matches are against `colocPropTest::run_proptests` 0.9.3's actual
output on these inputs, verified to about $10^{-6}$, and re-verified by `equivalence/check.R` in
CI. Your `prop_run()` *is* colocPropTest's multi-pair procedure.

## The honest FDR-vs-Holm note

The paper's prose calls the correction a "false discovery rate (FDR)." But the released
`colocPropTest` 0.9.3 actually calls `p.adjust(p, n = ntests)` **with its default method**, which
is **Holm** — a family-wise (Bonferroni-type) correction — while naming the output column `fdr`.
So the *concept* the paper describes is FDR (control the fraction of false discoveries, from the
[false-discovery-rate foundation](../../../../foundations/false-discovery-rate/lesson.qmd)), but
the *code* applies Holm (control the chance of *any* false alarm). Our `prop_run()` matches the
package: `method = "holm"`. It matters because Holm is stricter than Benjamini–Hochberg — it will
reject slightly *less* often — so if you read the column labelled `fdr` as a true
Benjamini–Hochberg FDR, you'd mis-describe what the package guarantees. Teach the FDR idea; report
the code honestly as Holm.

## Common traps

**Trap: "reject colocalisation only if the *average* pair is significant."** No — reject if **any**
single pair is significant (the minimum adjusted p-value). Colocalisation needs *every* pair
proportional, so one broken pair disproves it. Our distinct fixture has 14 large adjusted p's and
one at $2\times 10^{-8}$; the average is near 1, but the *minimum* triggers rejection. Averaging
would hide the one real departure.

**Trap: "the column is called `fdr`, so it's a Benjamini–Hochberg false discovery rate."** As
above — no. `colocPropTest` calls `p.adjust`'s default (Holm, family-wise), just naming the column
`fdr`. Holm and BH give different numbers: for raw p's $(0.01, 0.02, 0.04)$ over 3 tests, Holm
gives $(0.03, 0.04, 0.04)$ but BH gives $(0.03, 0.03, 0.04)$. Don't assume the label; check the
method.

## Check yourself

**1.** A region has $n = 6$ SNPs. How many possible pairs, and what `ntests` should the correction
use even if you only actually run the test on 3 of the pairs?

::: {.callout-note collapse="true" title="Answer"}
$\binom{6}{2} = \frac{6\cdot 5}{2} = 15$ possible pairs. Even if you test only 3, the correction
denominator stays $\text{ntests} = 15$ — you must account for all possible tests to keep the
guarantee (the min-marginal-p bound is what lets you skip the other 12).
:::

**2.** `prop_run()` returns `min_adjusted = 0.9`. Colocalised or not, and why?

::: {.callout-note collapse="true" title="Answer"}
**Colocalised** (fail to reject). The smallest adjusted p-value across all pairs is $0.9$, well
above 0.05, so no pair provably breaks proportionality — the region is consistent with one shared
line through the origin. (Remember: for this test, a *large* p is the colocalisation-friendly
outcome.)
:::

**3.** In your own words: explain to a colleague, in two sentences, why the paper tests all pairs
and corrects over the total, instead of testing one big $n$-variant model or the two strongest
SNPs.

::: {.callout-note collapse="true" title="A model answer"}
A single $n$-variant test has $n-1$ degrees of freedom and dilutes the evidence from any one
non-proportional pair, so it's weak; picking the two strongest SNPs is powerful but biased because
you select the variants using the same data you test. Running all pairs concentrates the evidence
into low-df, high-power tests and correcting over the total number of possible pairs removes the
selection bias, so you can reject on the single weakest link while still controlling false alarms.
:::

## What you can now do

- **Enumerate all pairs** of a region ($n(n-1)/2$) and run the pairwise proportional test on each.
- **Correct for multiplicity** with `p.adjust(p, n = ntests)` over the total possible pairs, and
  **reject colocalisation if the minimum adjusted p-value is small** (any one significant pair).
- **Explain the two rejected alternatives** (weak $n$-variant test; biased two-variant selection)
  and the min-marginal-p bound that lets you test a subset.
- **State the FDR-vs-Holm nuance** honestly: the paper says "FDR," but the package applies
  `p.adjust`'s default (Holm), a family-wise correction.
- **Match the frozen `colocPropTest::run_proptests` reference** on both a shared and a distinct
  region to $10^{-6}$.

If any of these feels shaky, revisit the toy example and the two self-checks before the capstone.

## Where this goes next

You've now built the whole method from scratch: `prop_V`, `prop_test`/`prop_test_pair`, and
`prop_run`. The capstone,
[Reading colocPropTest](../06-capstone-reading-colocproptest/lesson.qmd), returns to the source —
decoding the paper's equations and the *real* `colocPropTest` interface (`run_proptests`,
`estprop`, `tester_marg`, `marg_with_V`), backlinking each term to the module that built it, and
assembling the full pipeline on shared and distinct regions.

Ready for more reps? Do the [practice problems](practice.qmd).
