---
title: "Lambda for stability"
subtitle: "When indicators are near-identical, a ridge λ keeps the reconstruction from exploding"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)

sim_collinear <- function(n = 200, seed = 3, rho = 0.999) {
  set.seed(seed); p <- 12; R <- diag(p)
  R[1,2] <- R[2,1] <- 0.9; R[3,4] <- R[4,3] <- rho
  R[8,10] <- R[10,8] <- 0.7; R[11,12] <- R[12,11] <- 0.65
  X <- matrix(rnorm(n*p), n, p) %*% chol(R)
  colnames(X) <- c("rate_10y","rate_2y","cpi_infl","pce_infl","unemp","oil",
                   "usd_index","sentiment","housing","pmi","credit_spread","vix")
  beta <- numeric(p); beta[c(1,3,5)] <- c(0.5,-0.4,-0.3)
  list(X = X, y = drop(X %*% beta + rnorm(n)))
}
```

## Where we're going

The reconstruction $\hat\beta = R^{-1}b$ has a failure mode: when two indicators are *almost
the same* (correlation near 1), the correlation matrix $R$ is nearly impossible to invert,
and $R^{-1}b$ blows up into huge, opposite-signed coefficients that swing wildly with the
tiniest change. The fix is the [ridge](../../../../foundations/ridge-regularization/lesson.qmd)
idea, applied to $R$ itself: replace $R$ with $R + \lambda I$ — add a small $\lambda$ down the
diagonal. That is invertible for *any* $\lambda > 0$, and it trades a little shrinkage for a
lot of stability. This module builds that safeguard and sweeps $\lambda$ to see it work.

**Builds on:** [ridge-regularization](../../../../foundations/ridge-regularization/lesson.qmd)
($(X^\top X + \lambda I)$ inverts for any $\lambda>0$; standardize first),
[Module 1](../01-joint-fit-from-summaries/lesson.qmd) ($\hat\beta = R^{-1}b$),
[Module 3](../03-rank-one-speedups/lesson.qmd) (a tiny Schur complement signals collinearity),
[Multiple Regression, Module 3](../../../multiple-regression/modules/03-marginal-vs-joint/lesson.qmd)
(collinearity blows coefficients up).

## Warm-up: do you remember?

**1.** *(ridge)* What does adding $\lambda I$ to a matrix before inverting guarantee, for any
$\lambda > 0$?

::: {.callout-note collapse="true" title="Answer"}
That the matrix is **invertible** — $R + \lambda I$ is non-singular for any $\lambda > 0$,
even if $R$ alone is singular. See [ridge-regularization](../../../../foundations/ridge-regularization/lesson.qmd).
:::

**2.** *(Module 1)* How do we reconstruct the joint coefficients from the correlation matrix
$R$ and marginal vector $b$?

::: {.callout-note collapse="true" title="Answer"}
$\hat\beta = R^{-1}b$ — solve $R\hat\beta = b$. See
[Module 1](../01-joint-fit-from-summaries/lesson.qmd).
:::

**3.** *(Multiple Regression, Module 3)* What happens to a regression's coefficients when two
predictors are correlated ~0.999?

::: {.callout-note collapse="true" title="Answer"}
They become huge and unstable — often large and opposite in sign — because the fit can't tell
the two nearly-identical predictors apart. See
[Multiple Regression, Module 3](../../../multiple-regression/modules/03-marginal-vs-joint/lesson.qmd).
:::

## A tiny example first

Two indicators correlated at **0.999** — nearly the same variable — with marginal
coefficients $b = (0.5, 0.49)$:

$$
R = \begin{pmatrix} 1 & 0.999 \\ 0.999 & 1 \end{pmatrix}, \qquad b = (0.5,\ 0.49).
$$

**Unregularized.** $\det R = 1 - 0.999^2 = 0.001999$ — a whisker above zero, so $R^{-1}$ has
enormous entries. Reconstructing:

$$
\hat\beta = R^{-1}b = \frac{1}{0.001999}\begin{pmatrix} 1 & -0.999 \\ -0.999 & 1 \end{pmatrix}\begin{pmatrix}0.5 \\ 0.49\end{pmatrix}
= \frac{1}{0.001999}\begin{pmatrix} 0.01049 \\ -0.0095 \end{pmatrix} = \begin{pmatrix} 5.25 \\ -4.75 \end{pmatrix}.
$$

A tiny 0.01 gap between two marginal coefficients got amplified into $+5.25$ and $-4.75$ —
absurd, unstable coefficients from a stable-looking input.

**Regularized.** Add $\lambda = 0.1$ down the diagonal: $R + \lambda I = \begin{pmatrix}1.1 & 0.999 \\ 0.999 & 1.1\end{pmatrix}$,
$\det = 1.21 - 0.998 = 0.212$ — comfortably away from zero. Now

$$
\hat\beta = (R + \lambda I)^{-1} b = \frac{1}{0.212}\begin{pmatrix} 0.0605 \\ 0.0395 \end{pmatrix} = \begin{pmatrix} 0.285 \\ 0.186 \end{pmatrix}.
$$

Sane, similar, positive coefficients — the two near-identical indicators now sensibly *share*
the effect instead of fighting over it.

```{r toy-example}
R <- matrix(c(1, 0.999, 0.999, 1), 2); b <- c(0.5, 0.49)
rbind(unregularized      = solve(R, b),                    # 5.25, -4.75  (explodes)
      ridge_lambda_0.1   = solve(R + 0.1 * diag(2), b))    # 0.285, 0.186 (stable)
```

## Getting a feel for it

**Leaning a wobbly table.** With $R$ near-singular, the least-squares problem is a nearly-flat
valley — a long trough where wildly different coefficient pairs fit almost equally well, so
the solver skids to an extreme. Adding $\lambda I$ tilts the floor of that trough, giving it a
single clear low point near the origin. It's the ridge module's **leash**, now leashing the
*reconstruction*: $\lambda$ pulls the coefficients back toward small, sensible values instead
of letting them fly apart. Slack leash ($\lambda \to 0$): back to the exploding $R^{-1}b$.
Tight leash (large $\lambda$): everything shrinks toward 0.

**Why adding to the diagonal works.** A matrix is hard to invert when it has an eigenvalue
near 0 (a direction it nearly flattens). Adding $\lambda I$ lifts *every* eigenvalue by
$\lambda$, so the smallest becomes $\lambda_{\min} + \lambda$ — safely positive for any
$\lambda > 0$ (VMLS §15.3). The near-zero determinant $0.001999$ became $0.212$; the matrix's
**condition number** (how much it amplifies small changes) drops from enormous to modest.

**What $\lambda$ costs.** Stability isn't free: like any ridge, $\lambda$ shrinks the
recovered coefficients toward zero (bias) in exchange for taming their variance. Too small and
the reconstruction still wobbles; too large and you over-shrink real effects. The useful
$\lambda$ is the smallest one that makes the inversion well-behaved — often a value you sweep
and inspect, as below.

## What the notation is going to say

The formula will say: *before inverting, add a small positive $\lambda$ to each diagonal entry
of $R$; then reconstruct as usual.* $R + \lambda I$ is always invertible, so the exploding
$R^{-1}b$ becomes the stable $(R + \lambda I)^{-1}b$; $\lambda$ dials how much stability (and
how much shrinkage) you buy.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $R$ | "the correlation matrix R" | predictor correlations; near-singular when two correlate ~1 | $\left(\begin{smallmatrix}1&.999\\.999&1\end{smallmatrix}\right)$ |
| $I$ | "the identity matrix" | 1s on the diagonal, 0 elsewhere | $\left(\begin{smallmatrix}1&0\\0&1\end{smallmatrix}\right)$ |
| $\lambda$ | "lambda" | the ridge amount added to $R$'s diagonal ($>0$) | 0.1 |
| $R + \lambda I$ | "R plus lambda I" | the regularized, safely-invertible matrix | $\left(\begin{smallmatrix}1.1&.999\\.999&1.1\end{smallmatrix}\right)$ |
| $\det$ | "determinant" | near 0 ⇒ near-singular; away from 0 ⇒ safe to invert | $0.002 \to 0.212$ |
| $\kappa(R)$ | "kappa of R" (condition number) | how much $R$ amplifies small errors; huge ⇒ unstable | large $\to$ modest |

## The formal version

**The problem.** As two indicators' correlation $\rho \to 1$, $R$ has an eigenvalue $\to 0$,
so $\det R \to 0$ and $R^{-1}$ has entries $\to \infty$. The reconstruction $\hat\beta = R^{-1}b$
then amplifies any tiny wiggle in $b$ into gigantic, opposite-signed coefficients (the toy's
$+5.25 / -4.75$).

**The fix.** Reconstruct with a ridge $\lambda$ on the correlation matrix:

$$
\boxed{\;\hat\beta_{\lambda} = (R + \lambda I)^{-1} b\;}.
$$

- $R + \lambda I$ adds $\lambda$ to every eigenvalue, so its smallest eigenvalue is
  $\lambda_{\min}(R) + \lambda \ge \lambda > 0$ — **invertible for any $\lambda > 0$** (VMLS
  §15.3), even when $R$ is exactly singular.
- This is precisely the ridge move from the foundation, applied to the summary-world Gram
  matrix $R$ instead of $X^\top X$: it minimizes $\lVert$reconstruction error$\rVert^2 + \lambda\lVert\beta\rVert^2$,
  so $\lambda$ shrinks $\hat\beta_\lambda$ toward 0.

Reading it aloud: "beta-hat-lambda equals the inverse of (R plus lambda I), times b" — or in
everyday words: "pad the correlation matrix's diagonal a little so it inverts cleanly, then
reconstruct."

**Sanity check (toy):** $\det(R + 0.1 I) = 0.212 \gg 0.002 = \det R$, and the coefficients went
from $(5.25, -4.75)$ to a stable $(0.285, 0.186)$ — matching the R output.

## Explore it

### Sweep λ on a ρ = 0.999 panel

```{r sweep}
sim <- sim_collinear(n = 200, seed = 3, rho = 0.999)      # cpi_infl / pce_infl at 0.999
R <- cor(sim$X); b <- as.numeric(cor(sim$X, sim$y))

sweep <- t(sapply(c(0, 0.01, 0.05, 0.1, 0.5), function(lam) {
  beta <- solve(R + lam * diag(12), b)
  c(lambda = lam, cpi_infl = beta[3], pce_infl = beta[4], max_abs_coef = max(abs(beta)))
}))
round(sweep, 3)
```

At $\lambda = 0$ the inflation twins explode to `cpi_infl` ≈ −2.02 and `pce_infl` ≈ +1.76
(opposite signs, biggest coefficient ~2.0); by $\lambda = 0.1$ they settle to a sensible,
similar −0.14 / −0.11 and the largest coefficient is ~0.28. The reconstruction is usable
again.

```{webr}
#| autorun: true
#| fig-alt: "The largest reconstructed coefficient plotted against lambda from 0 to 0.5: it starts very large near lambda 0 and drops steeply, flattening to a small stable value as lambda increases."
# TRY IT: change rho (the 0.999) toward 1, or change b's tiny gap, and re-run. Watch how
# large the unregularized (lambda = 0) coefficients get, and how quickly lambda tames them.
# PREDICT: at lambda = 0, roughly how big is the largest |coefficient| — and about where does
# the curve flatten as lambda grows? Write your guess, then read the plot.
rho <- 0.999
R <- matrix(c(1, rho, rho, 1), 2); b <- c(0.5, 0.49)
lams <- seq(0, 0.5, length.out = 60)
maxc <- sapply(lams, function(l) max(abs(solve(R + l * diag(2), b))))
plot(lams, maxc, type = "l", lwd = 2, col = "firebrick",
     xlab = "ridge lambda", ylab = "largest |coefficient|",
     main = "lambda tames the exploding reconstruction")
abline(h = 0, lty = 3)
```

The selected *set*, meanwhile, is robust: running the switching search on this panel returns
`{rate_10y, cpi_infl, unemp}` at every $\lambda$ swept — because the winning subset avoids
keeping *both* twins, so its own small $R_S$ was never ill-conditioned. $\lambda$ rescues the
*coefficients* (and any subset that does contain a near-collinear pair), not the selection
here.

## Common traps

**Trap: "$\lambda$ just adds bias, so avoid it and use plain $R^{-1}b$."** At mild correlation,
sure — $\lambda = 0$ is fine. But at $\rho \to 1$ the plain reconstruction is *unusable*
(coefficients in the thousands, flipping sign on rounding). A small $\lambda$ trades a little
bias for a reconstruction that exists and is stable — no contest when $R$ is near-singular.

**Trap: "bigger $\lambda$ is safer, so crank it up."** Too-large $\lambda$ over-shrinks: every
coefficient collapses toward 0 and you lose the real effects (the ridge module's "$\lambda$
huge ⇒ predict the mean"). The goal is the *smallest* $\lambda$ that makes the inversion
well-conditioned — big enough to stabilize, small enough to preserve signal.

## Check yourself

**1.** $R$ has smallest eigenvalue $0.0004$ (nearly singular). After adding $\lambda = 0.05\,I$,
what is the new smallest eigenvalue, and is the matrix now safely invertible?

::: {.callout-note collapse="true" title="Answer"}
Adding $\lambda I$ lifts every eigenvalue by $\lambda$, so the smallest becomes
$0.0004 + 0.05 = 0.0504$ — comfortably positive, so $R + \lambda I$ is safely invertible.
That's why any $\lambda > 0$ guarantees invertibility.
:::

**2.** Two indicators correlate 0.9999 and the unregularized reconstruction gives coefficients
$(83, -82)$. What does that pattern signal, and what will a small $\lambda$ do to them?

::: {.callout-note collapse="true" title="Answer"}
Huge opposite-signed coefficients are the signature of near-perfect collinearity: the two
almost-identical indicators are fighting over the shared effect, and the near-singular $R$
amplified a tiny difference. A small $\lambda$ will pull them to small, similar values that
sensibly split the shared effect (like the toy's $(0.285, 0.186)$).
:::

**3.** In your own words: explain to a colleague, in two sentences, why near-collinear
indicators break $R^{-1}b$ and how $\lambda$ fixes it.

::: {.callout-note collapse="true" title="A model answer"}
When two indicators are nearly identical, $R$ is almost singular (an eigenvalue near zero), so
$R^{-1}$ has enormous entries that amplify tiny differences into wild, opposite-signed
coefficients. Adding $\lambda$ to $R$'s diagonal lifts that eigenvalue away from zero, making
the inverse well-behaved and pulling the coefficients back to stable, sensible values.
:::

## What you can now do

- **Recognize the collinearity blow-up** (near-zero $\det R$, huge opposite-signed
  coefficients) in a summary-based reconstruction.
- **Apply the ridge fix** $\hat\beta_\lambda = (R + \lambda I)^{-1}b$, and explain why $R+\lambda I$
  is invertible for any $\lambda > 0$ (every eigenvalue lifted by $\lambda$).
- **Read a $\lambda$ sweep**: coefficients settle from exploding to stable, and pick the
  smallest $\lambda$ that stabilizes without over-shrinking.
- **Separate what $\lambda$ fixes** (the coefficients / ill-conditioned subsets) from what can
  stay robust (the selected set).

If any feels shaky, revisit the [tiny example](#a-tiny-example-first) — the capstone assembles
this $\lambda$ into the full procedure.

## Where this goes next

You now have all five pieces — reconstruct from summaries, search by switching, stop with BIC,
speed up with rank-1 updates, stabilize with $\lambda$. Next: [the capstone](../05-capstone-switchstep/lesson.qmd)
assembles them into one `switchstep()` function and runs it end-to-end on a simulated
meta-analysis. Ready for reps? Do the [practice problems](practice.qmd). External readings:
[resources](resources.md).
