---
title: "Practice: Proportionality is colocalisation"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

Work these in order — they get harder. Do them on paper or in an R session; only then open the
answer.

## Problem 1 — the toy example, new numbers

Trait 2 effects $b_2 = (0.2, 0.1, 0.4)$. The traits colocalise with $\eta = 2$. Give $b_1$ and
confirm the three ratios $b_1/b_2$ are equal.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
b2 <- c(0.2, 0.1, 0.4)
eta <- 2
```

::: {.callout-note collapse="true" title="Answer"}
```{r p1-ans}
b2 <- c(0.2, 0.1, 0.4); eta <- 2
b1 <- eta * b2; b1              # 0.4 0.2 0.8
b1 / b2                          # 2 2 2 — constant slope, on the origin-line
```
Because $b_1 = 2 b_2$ exactly, all ratios are $2$; the points $(b_2, b_1)$ sit on the line
through the origin with slope $2$.
:::

## Problem 2 — fill in the missing step

Build $V = (\sigma\sigma^\top)\odot\Sigma$ for a pair with $\text{se} = (0.05, 0.02)$, LD
$\rho = 0.8$:

- Step 1: diagonal is $\sigma_1^2 = 0.0025$, $\sigma_2^2 = 0.0004$.
- Step 2: **your step here** — compute the off-diagonal $\sigma_1\sigma_2\rho$.
- Step 3: assemble $V$.

Fill in step 2.

```{webr}
#| autorun: false
# p2-starter — edit and Run to check the blank
se <- c(0.05, 0.02); rho <- 0.8
```

::: {.callout-note collapse="true" title="Answer"}
Step 2: $\sigma_1\sigma_2\rho = 0.05\times 0.02\times 0.8 = 0.0008$. So
$V = \begin{bmatrix} 0.0025 & 0.0008 \\ 0.0008 & 0.0004 \end{bmatrix}$.

```{r p2-ans}
prop_V <- function(vbeta, rho) { V <- diag(vbeta); V[1,2] <- V[2,1] <- sqrt(vbeta[1]*vbeta[2])*rho; V }
prop_V(c(0.05^2, 0.02^2), 0.8)
```
:::

## Problem 3 — shared or distinct?

Three SNPs, $b_2 = (0.10, 0.20, 0.30)$ and $b_1 = (0.12, 0.24, 0.35)$. Compute the ratios and
decide (ignoring noise) whether this looks shared or distinct.

```{webr}
#| autorun: false
# p3-starter — your job: compute b1/b2 and judge
b2 <- c(0.10, 0.20, 0.30)
b1 <- c(0.12, 0.24, 0.35)
```

::: {.callout-note collapse="true" title="Answer"}
```{r p3-ans}
b2 <- c(0.10, 0.20, 0.30); b1 <- c(0.12, 0.24, 0.35)
round(b1 / b2, 3)              # 1.200 1.200 1.167
```
The first two ratios are $1.2$ but the third is $1.167$ — *nearly* constant. The first two SNPs
look proportional; the third drifts a little off. Whether that drift is "real distinctness" or
just noise is what the test statistic decides (next module) — by eye, it's borderline-shared.
:::

## Problem 4 — spot the error

A colleague ran a proportional colocalisation test — the reasoning contains **exactly one
mistake**. Find it.

> "The proportional test gave p = 0.002. Since that's a small p-value, we have strong evidence
> that the two traits colocalise."

::: {.callout-note collapse="true" title="Answer"}
The mistake is the direction. In the proportional test the **null hypothesis is colocalisation**
($b_1 = \eta b_2$), so a small p-value is evidence **against** the null — i.e. evidence they do
**NOT** colocalise. p = 0.002 means "reject proportionality" = "distinct signals." The correct
statement: "small p → evidence they do *not* colocalise." (This is the opposite of coloc, where a
large PP4 supports colocalisation.)
:::

## Problem 5 — mixing it with standardization *(uses variance-structure-and-standardization's covariance matrix)*

You have a pair of SNPs with variances $\text{vbeta} = (0.0009, 0.0009)$ and LD $\rho = 0.6$.
Build $V$, then extract the **correlation** between the two effect estimates from $V$ (covariance
divided by the product of standard deviations) and confirm it equals $\rho$.

```{webr}
#| autorun: false
# p5-starter — edit and Run to solve, right here
prop_V <- function(vbeta, rho) { V <- diag(vbeta); V[1,2] <- V[2,1] <- sqrt(vbeta[1]*vbeta[2])*rho; V }
vbeta <- c(0.0009, 0.0009); rho <- 0.6
```

::: {.callout-note collapse="true" title="Answer"}
```{r p5-ans}
prop_V <- function(vbeta, rho) { V <- diag(vbeta); V[1,2] <- V[2,1] <- sqrt(vbeta[1]*vbeta[2])*rho; V }
V <- prop_V(c(0.0009, 0.0009), 0.6)
V[1, 2] / sqrt(V[1, 1] * V[2, 2])     # covariance / (sd1 * sd2) = 0.6 = rho
```
The recovered correlation is exactly $\rho = 0.6$ — because $V$ was *built* as
$(\sigma\sigma^\top)\odot\Sigma$, standardising the off-diagonal back out returns the LD you put
in. This is the covariance↔correlation conversion from
variance-structure-and-standardization.
:::

## Problem 6 — transfer: interpret a p-value flip

Two analyses of the same region: `coloc` reports PP4 = 0.92, and the proportional test reports
p = 0.75. Do these two results *agree* or *conflict*, and what would you conclude?

::: {.callout-note collapse="true" title="Answer"}
They **agree** — both support colocalisation. coloc's high PP4 (0.92) says "shared." The
proportional test's *large* p-value (0.75) means "fail to reject the null of proportionality,"
and since the null *is* colocalisation, a large p-value is the colocalisation-friendly outcome.
The trap would be to read p = 0.75 as "weak evidence"; here a large p is exactly what you want.
Two methods with different assumptions agreeing gives extra confidence to design a follow-up.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**, once the
material isn't fresh. A pass then is real evidence you've retained it. Five questions, graded on
the page; 4/5 or better means this module is solid. Results are saved in this browser only.

::: {.module-check module="courses/proportional-colocalisation/modules/03-proportionality-is-colocalisation" pass="4"}

::: {.quiz-q qid="pic-proportional-means" qtype="mc" answer="b"}
"The two traits colocalise" is equivalent to which geometric statement about the points
$(b_{2,j}, b_{1,j})$?

a) they fall on a horizontal line
b) they fall on a straight line through the origin
c) they fall on any straight line
d) they fall on a circle
:::

::: {.quiz-q qid="pic-pvalue-flip" qtype="mc" answer="c"}
In the proportional colocalisation test, a **small** p-value means:

a) strong evidence the traits colocalise
b) the effect estimates are exactly proportional
c) evidence the traits do NOT colocalise (reject the null of proportionality)
d) the sample size was too small
:::

::: {.quiz-q qid="pic-origin" qtype="mc" answer="a"}
Why must the fitted line pass through the origin for colocalisation, rather than having an
intercept?

a) a shared cause means zero effect on one trait exactly where there's zero on the other; an intercept implies a second, independent cause
b) lines through the origin are easier to fit
c) the standard errors force it there
d) it makes the slope equal to 1
:::

::: {.quiz-q qid="pic-V-offdiag" qtype="numeric" answer="0.0008" tolerance="0.00001"}
For a pair with standard errors $0.05$ and $0.02$ and LD $\rho = 0.8$, what is the off-diagonal of
$V = (\sigma\sigma^\top)\odot\Sigma$, i.e. $\sigma_1\sigma_2\rho$?
:::

::: {.quiz-q qid="pic-cov-interleave" qtype="numeric" answer="0.6" tolerance="0.001"}
*(uses variance-structure-and-standardization)* A covariance matrix has diagonal $(0.0009,
0.0009)$ and off-diagonal $0.00054$. What is the correlation between the two estimates,
$\text{cov}/(\sigma_1\sigma_2)$? (Enter to 1 decimal.)
:::

:::

::::
