---
title: "Rank-one speed-ups"
subtitle: "Grow and shrink the correlation matrix's inverse on every move — without re-inverting"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)

# Bordering update: grow a k x k matrix (inverse Mi known) by one row/col [b_new; d].
sw_border <- function(Mi, b_new, d) {
  Mib <- Mi %*% b_new
  s   <- as.numeric(d - crossprod(b_new, Mib))
  TL  <- Mi + (Mib %*% t(Mib)) / s
  TR  <- -Mib / s
  rbind(cbind(TL, TR), c(t(TR), 1 / s))
}
```

## Where we're going

The switching search re-scores dozens of candidate subsets on every step, and each score
needs the inverse of that subset's correlation matrix $R_S$. Re-inverting from scratch every
time is the search's bottleneck. But consecutive candidates differ by just **one** indicator
— add one, remove one, or swap — which changes $R_S$ by a single border (one row and one
column). The Sherman–Morrison / **bordering** update (from the foundation module) produces
the new inverse from the old one cheaply and *exactly*. This module wires that update into
the search and measures the payoff.

**Builds on:** [sherman-morrison-rank-one-updates](../../../../foundations/sherman-morrison-rank-one-updates/lesson.qmd)
(update an inverse without re-inverting; exact; $O(p^2)$ vs $O(p^3)$),
[Module 2](../02-the-switching-search/lesson.qmd) (the search re-solves $R_S^{-1}$ each move).

## Warm-up: do you remember?

**1.** *(Sherman–Morrison)* Is updating an inverse after a small change an approximation or
exact, and roughly how does its cost compare to a fresh inverse?

::: {.callout-note collapse="true" title="Answer"}
**Exact** (an algebraic identity, matches a fresh inverse to machine precision), and cheaper:
about $O(p^2)$ work vs $O(p^3)$ to re-invert. See
[sherman-morrison-rank-one-updates](../../../../foundations/sherman-morrison-rank-one-updates/lesson.qmd).
:::

**2.** *(Module 2)* In the switching search, what has to be computed to score each candidate
subset $S$, and how many candidates are scored per step?

::: {.callout-note collapse="true" title="Answer"}
Each candidate needs $R_S^{-1}$ (to get $b_S^\top R_S^{-1} b_S$ for its BIC), and a step
scores every add/remove/swap neighbour — on the order of $p$ of them. So the inverse is
computed over and over. See [Module 2](../02-the-switching-search/lesson.qmd).
:::

**3.** *(matrices)* If you add one indicator to a subset of size $k$, what happens to the
shape of its correlation matrix $R_S$?

::: {.callout-note collapse="true" title="Answer"}
It grows from $k \times k$ to $(k{+}1) \times (k{+}1)$ — one new row and one new column (the
new indicator's correlations with the others, and a 1 on the diagonal). That's a "border."
:::

## A tiny example first

You already have two indicators in the model, with correlation matrix and inverse

$$
R_S = \begin{pmatrix} 1 & 0.5 \\ 0.5 & 1 \end{pmatrix}, \qquad
R_S^{-1} = \frac{1}{0.75}\begin{pmatrix} 1 & -0.5 \\ -0.5 & 1 \end{pmatrix} = \begin{pmatrix} 1.333 & -0.667 \\ -0.667 & 1.333 \end{pmatrix}.
$$

Now **add** a third indicator whose correlations with the first two are $b_{\text{new}} = (0.3, 0.4)$
(and 1 on its own diagonal). The grown matrix is

$$
R_S^{+} = \begin{pmatrix} 1 & 0.5 & 0.3 \\ 0.5 & 1 & 0.4 \\ 0.3 & 0.4 & 1 \end{pmatrix}.
$$

Bordering builds $\big(R_S^{+}\big)^{-1}$ from the $R_S^{-1}$ you already have. The one new
number you need is the **Schur complement** — a scalar measuring how much *new* information
the third indicator adds beyond the first two:

$$
s = d - b_{\text{new}}^\top R_S^{-1} b_{\text{new}} = 1 - (0.3, 0.4)\,R_S^{-1}\,(0.3, 0.4)^\top = 0.8267.
$$

Everything else in the update is built from $R_S^{-1} b_{\text{new}}$ and that scalar $s$ —
no fresh $3\times3$ inversion. It reproduces `solve()` of $R_S^{+}$ exactly:

```{r toy-example}
RS <- matrix(c(1, 0.5, 0.5, 1), 2); RSi <- solve(RS)   # the inverse we already hold
b_new <- c(0.3, 0.4); d <- 1                            # new indicator's border

s <- as.numeric(d - t(b_new) %*% RSi %*% b_new)         # Schur complement
grown_inv <- sw_border(RSi, b_new, d)                   # bordering update
RS_plus <- rbind(cbind(RS, b_new), c(b_new, d))
c(schur = s, max_diff_vs_solve = max(abs(grown_inv - solve(RS_plus))))   # s = 0.827; diff ~1e-16
```

## Getting a feel for it

**Add a border, patch the inverse.** Growing $R_S$ by one indicator literally adds a border —
a new bottom row and right column — to the matrix. The bordering formula patches the inverse
to match, reusing the $R_S^{-1}$ you already computed. It's the "fix the pothole, don't
re-pave the road" idea from the Sherman–Morrison module, specialized to "the change is a new
row+column." Removing an indicator is the same idea in reverse (a *partition* update peels
the border back off), and a **swap** is just a remove followed by an add.

**Why the search cares.** A search over many indicators makes *thousands* of moves, each
currently paying an $O(k^3)$ re-inversion. Replacing each with an $O(k^2)$ update changes the
cost per move by a factor of about $k$ — the difference between a search that crawls and one
that's instant. And because the update is exact, the search's answers are *identical* to the
re-invert-every-time version; you pay nothing in accuracy for the speed.

**What the Schur complement feels like.** The scalar $s = d - b_{\text{new}}^\top R_S^{-1} b_{\text{new}}$
is the new indicator's *unique* variance — what's left after the indicators already in the
model explain it. When the new indicator is nearly a combination of the existing ones (highly
collinear), $s$ shrinks toward 0, the update divides by a tiny number, and things get
unstable — foreshadowing exactly why Module 4 adds a ridge $\lambda$.

## What the notation is going to say

The update says: *to grow the inverse, compute the new indicator's leftover-variance scalar
$s$; then the new inverse's blocks are the old inverse nudged by a rank-1 term (built from
$R_S^{-1}b_{\text{new}}$ and $s$), a new corner $1/s$, and matching off-diagonal pieces.* No
full inversion — just matrix-times-vector work and one division.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $R_S$ | "R sub S" | correlation matrix of the indicators currently in subset $S$ | $\left(\begin{smallmatrix}1&.5\\.5&1\end{smallmatrix}\right)$ |
| $R_S^{-1}$ | "R sub S inverse" | its inverse, already held from the previous move | $\left(\begin{smallmatrix}1.33&-.67\\-.67&1.33\end{smallmatrix}\right)$ |
| $b_{\text{new}}$ | "b new" | the new indicator's correlations with those already in $S$ | $(0.3, 0.4)$ |
| $d$ | "dee" | the new indicator's own diagonal entry (1 for a correlation matrix) | 1 |
| $s$ | "ess" (Schur complement) | new indicator's leftover variance $d - b_{\text{new}}^\top R_S^{-1} b_{\text{new}}$ | 0.827 |
| $O(k^2),\,O(k^3)$ | "order k-squared / k-cubed" | update cost vs fresh-inverse cost | — |

## The formal version

**Grow by one (bordering).** If $R_S$ (size $k$) has known inverse $R_S^{-1}$, and we append a
border — column/row $b_{\text{new}}$ and corner $d$ — then with the Schur complement
$s = d - b_{\text{new}}^\top R_S^{-1} b_{\text{new}}$ and helper $w = R_S^{-1} b_{\text{new}}$,

$$
\big(R_S^{+}\big)^{-1} =
\begin{pmatrix} R_S^{-1} + \dfrac{w\,w^\top}{s} & -\dfrac{w}{s} \\[1.2em] -\dfrac{w^\top}{s} & \dfrac{1}{s} \end{pmatrix}.
$$

- Top-left: the old inverse plus a rank-1 correction $w w^\top / s$ — the only place the old
  indicators' inverse changes, and it's rank-1 (built from the vector $w$).
- Corner $1/s$ and off-diagonals $-w/s$: the new indicator's own entries.
- Everything is matrix–vector products and one division by $s$ — cost $O(k^2)$, no $O(k^3)$
  inversion. (This is the Woodbury/bordering member of the Sherman–Morrison family named in
  the foundation.)

**Shrink by one (partition).** Removing an indicator reverses the formula: from the grown
inverse's blocks you recover $R_S^{-1}$ as (top-left block) $- w w^\top / s$ — again rank-1,
again $O(k^2)$.

**Swap** an indicator for another = shrink then grow (remove + add), two cheap updates.

Reading it aloud: "the new inverse is the old inverse plus a rank-1 patch, bordered by minus
w-over-s and a corner of one-over-s." Everything hangs on the single scalar $s$.

## Explore it

```{r timing}
# The search's inner loop, timed two ways on a p=12 correlation matrix: re-inverting each
# grown subset from scratch vs bordering the previous inverse. (Times vary by machine; the
# ratio is the point — this heavy loop is baked at build, not a browser cell.)
set.seed(1); p <- 12
Rt <- diag(p); Rt[1,2] <- Rt[2,1] <- 0.9; Rt[3,4] <- Rt[4,3] <- 0.95
X <- matrix(rnorm(400 * p), 400, p) %*% chol(Rt); R <- cor(X)

grow_fresh  <- function(k) solve(R[1:(k+1), 1:(k+1)])                  # re-invert
grow_border <- function(k) sw_border(solve(R[1:k, 1:k]), R[1:k, k+1], R[k+1, k+1])
reps <- 3000
t_fresh  <- system.time(for (i in 1:reps) grow_fresh(8))[["elapsed"]]
t_border <- system.time(for (i in 1:reps) grow_border(8))[["elapsed"]]
c(fresh_solve_s = round(t_fresh, 2), bordering_s = round(t_border, 2),
  max_diff = max(abs(grow_fresh(8) - grow_border(8))))                 # identical to ~1e-15
```

```{webr}
#| autorun: true
# TRY IT: change which indicator you add (b_new) or the base subset, and re-run. The
# bordering update always matches a fresh solve of the grown matrix to ~1e-15.
# PREDICT: the Schur complement s shrinks as the new indicator gets more collinear — watch it.
sw_border <- function(Mi, b_new, d) { Mib <- Mi %*% b_new
  s <- as.numeric(d - crossprod(b_new, Mib))
  rbind(cbind(Mi + (Mib %*% t(Mib))/s, -Mib/s), c(-t(Mib)/s, 1/s)) }

RS <- matrix(c(1, 0.5, 0.5, 1), 2); RSi <- solve(RS)
b_new <- c(0.3, 0.4); d <- 1
grown <- sw_border(RSi, b_new, d)
RS_plus <- rbind(cbind(RS, b_new), c(b_new, d))
c(schur_s = as.numeric(d - t(b_new) %*% RSi %*% b_new),
  max_diff = max(abs(grown - solve(RS_plus))))
```

## Common traps

**Trap: "the bordering update accumulates rounding error over many moves, so re-invert
occasionally to be safe."** Each update is an exact identity, matching a fresh inverse to
~$10^{-15}$; there's no systematic drift to accumulate over a search. (If you ever worry,
compare to a fresh `solve()` at the end — it agrees.)

**Trap: "a swap needs a fresh inverse."** No — a swap is a **remove** (partition update) then
an **add** (bordering update), two $O(k^2)$ steps. You never re-invert for any single move:
add, remove, and swap are all rank-1-style updates.

## Check yourself

**1.** Adding an indicator, you compute Schur complement $s = 0.02$. What does that tiny value
say about the new indicator, and what risk does dividing by it create?

::: {.callout-note collapse="true" title="Answer"}
$s$ is the new indicator's leftover variance after the others explain it; $s = 0.02$ means it
is *almost* a linear combination of indicators already in the model (highly collinear). The
update divides by $s$, so a tiny $s$ makes the new inverse huge and numerically unstable —
the motivation for Module 4's ridge $\lambda$.

:::

**2.** Why is replacing fresh inversion with bordering worth it in the search but pointless for
a single one-off fit?

::: {.callout-note collapse="true" title="Answer"}
A one-off fit inverts once — no savings. The search inverts thousands of times, each on a
matrix that differs from the last by one border, so each $O(k^3)$ re-inversion becomes an
$O(k^2)$ update; multiplied over thousands of moves, that's the difference between fast and
slow. The update pays off precisely because the work is *repeated on nearly-identical
matrices*.
:::

**3.** In your own words: explain to a colleague, in two sentences, how the search avoids
re-inverting $R_S$ on every move.

::: {.callout-note collapse="true" title="A model answer"}
Consecutive candidate subsets differ by only one indicator, so $R_S$ changes by just one
row and column, and the bordering (Sherman–Morrison) update produces the new inverse from
the old one with cheap matrix–vector work instead of a full re-inversion. It's exact, so the
search gives the same answers as re-inverting every time, only much faster.
:::

## What you can now do

- **Grow $R_S^{-1}$ by bordering** and **shrink it by partitioning**, and treat a swap as
  remove-then-add — all without re-inverting.
- **Compute and interpret the Schur complement** $s$ as the new indicator's leftover variance,
  and see why a tiny $s$ threatens stability.
- **Explain the $O(k^2)$-vs-$O(k^3)$ payoff** in the search's repeated-inversion loop, and that
  the update is exact.

If any feels shaky, revisit [The formal version](#the-formal-version) — Module 4 fixes the
tiny-$s$ instability the update exposed.

## Where this goes next

When indicators are near-collinear, the Schur complement (and the whole matrix) gets
dangerously close to singular. Next: [Lambda for stability](../04-lambda-stability/lesson.qmd)
adds a ridge $\lambda$ to keep the correlation matrix comfortably invertible. Ready for reps?
Do the [practice problems](practice.qmd). External readings: [resources](../../resources.md).
