---
title: "Profile likelihood"
subtitle: "When a formula has a parameter you don't care about, tune it away — and why doing so costs one degree of freedom"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live cells below re-state their own data. Base R only.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```


## Where we're going

Sometimes a formula depends on two knobs, but you only care about one of them. The other — a
**nuisance parameter** — is just in the way: you don't want to report it, you only need it out
of your hair so you can judge the knob you *do* care about. **Profiling** is the clean way to do
that: for every setting of the knob you care about, quietly tune the nuisance knob to its best
value, and keep only the resulting "best-case" score. This lesson builds that idea from a tiny
example, shows it in both directions (maximising a likelihood, minimising a fit statistic), and
explains the one accounting rule you must remember: **each nuisance parameter you profile out
costs you one degree of freedom.**

**Builds on:**
[likelihood & log-likelihood](../likelihood-and-log-likelihood/lesson.qmd) (the likelihood as
"how plausible is this parameter given the data," and finding its peak), and
[loss functions & optimization](../loss-functions-and-optimization/lesson.qmd) (a function with
a knob, and letting a computer search for the knob that makes it smallest/largest with
`optimize()`).

## Warm-up: do you remember?

Two quick reps first — produce each from memory, then check.

**1.** In one sentence, what does a **likelihood** measure, and what does the *maximum*
likelihood estimate pick out?

::: {.callout-note collapse="true" title="Answer"}
The likelihood measures how plausible a particular parameter value is given the data you saw
(data fixed, parameter varies). The maximum likelihood estimate is the parameter value that
makes the observed data most plausible — the peak of that curve. See
[likelihood & log-likelihood](../likelihood-and-log-likelihood/lesson.qmd).
:::

**2.** What does `optimize(f, interval = c(0, 10))` do in R, and what two things does it return?

::: {.callout-note collapse="true" title="Answer"}
It searches for the value of the input in $[0, 10]$ that **minimises** `f` (or maximises, with
`maximum = TRUE`), and returns `$minimum` (the best input) and `$objective` (the value of `f`
there). See [loss functions & optimization](../loss-functions-and-optimization/lesson.qmd).
:::

## A tiny example first

Here's a score that depends on two knobs, $a$ (the one you care about) and $b$ (the nuisance):

$$
S(a, b) = (a - 3)^2 + (a + b - 5)^2 .
$$

Smaller is better (think of it as "total miss," like a fit error). You want to know: **for a
given $a$, how good can the score possibly be?** — but you don't care what $b$ ends up being.

Take $a = 2$. The score is $(2-3)^2 + (2 + b - 5)^2 = 1 + (b - 3)^2$. You get to choose $b$ to
make this as small as possible. Setting $b = 3$ kills the second term, leaving $1 + 0 = 1$. So
the **best-case score at $a = 2$ is 1** — and we reached it by tuning the nuisance $b$ to 3.

Now take $a = 4$: score is $(4-3)^2 + (4 + b - 5)^2 = 1 + (b - 1)^2$. Best $b$ is $1$, giving
best-case score $1$ again. And at $a = 3$: score is $0 + (3 + b - 5)^2 = (b-2)^2$, best $b = 2$
gives best-case score **0**. So the *profiled* score — the best-case score as a function of $a$
alone — is $0$ at $a = 3$ and rises as $a$ moves away. We've collapsed a two-knob problem to a
one-knob curve by tuning the nuisance out at every point.

The same thing in R:

```{r toy-example}
# The two-knob score: a is the parameter of interest, b is the nuisance
S <- function(a, b) (a - 3)^2 + (a + b - 5)^2

# "Profile out b": for a fixed a, find the b that minimises S, keep only the best score
profile_at <- function(a) optimize(function(b) S(a, b), interval = c(-10, 10))$objective

# Best-case score at a few values of a
sapply(c(2, 3, 4), profile_at)   # 1, 0, 1 — matches the hand computation
```

## Getting a feel for it

**Analogy — a hiker photographing a mountain ridge from the valley floor.** The full landscape
$S(a, b)$ is a 2-D surface (two knobs, two map directions). You only care about the east–west
direction $a$; the north–south direction $b$ is a nuisance. Profiling is standing at each
east–west position and recording only the **highest point of the ridge** you can see looking
north–south (for a likelihood you keep the peak; for a fit statistic you keep the valley floor).
You throw away the north–south coordinate and keep a single ridge-line curve over $a$. That
skyline is the *profile* — it tells you everything about $a$ once $b$ has been optimally chosen
at every step.

Where the analogy breaks: the "best $b$" isn't fixed — it *slides* as you move along $a$ (in our
toy, best $b$ was 3 at $a=2$ but 1 at $a=4$). Profiling re-optimises the nuisance at every value
of the parameter of interest; it's not a single fixed setting.

**Why re-optimise the nuisance at every $a$ instead of fixing it once?** Because fixing $b$ at
one value would unfairly penalise every $a$ that would have preferred a different $b$. By giving
each $a$ its *best possible* $b$, the profile curve is a fair "this is the best $a$ can do"
comparison — you're judging $a$ on its merits, not blaming it for a bad nuisance setting.

**What does the profiled curve feel like?** It's always at least as good (as low, for a fit
statistic) as the full surface at that $a$, because you optimised the other knob. Its bottom
(here, $0$ at $a = 3$) is the overall best of the whole 2-D surface — profiling never loses the
global optimum, it just hides the nuisance coordinate.

## What the notation is going to say

The profile of a two-knob function will say: "for each value of the parameter you care about,
plug in the *best* value of the nuisance parameter and record the result." For a **likelihood**
you keep the *maximum* over the nuisance (most plausible); for a **fit statistic** like our
$S$ you keep the *minimum* (smallest miss). In our toy example the profiled statistic was
"$\min_b S(a,b)$," which we found by hand to be $0$ at $a = 3$ — the same value the full
two-knob minimisation gives.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $a$ | "a" | the parameter of **interest** — the knob you want to judge | ranges over values; best is $3$ |
| $b$ | "b" | the **nuisance** parameter — needed for the formula but not reported | slides with $a$ (3 at $a=2$, 1 at $a=4$) |
| $S(a, b)$ | "S of a, b" | the two-knob score (here a fit statistic; smaller = better) | $(a-3)^2 + (a+b-5)^2$ |
| $\min_b$ | "the minimum over b" | tune $b$ to make what follows as small as possible | the profiling step |
| $\max_b$ | "the maximum over b" | tune $b$ to make what follows as large as possible (used for likelihoods) | — |
| $\hat b(a)$ | "b-hat of a" | the best nuisance value *at a given $a$* (it depends on $a$) | $3$ at $a=2$, $1$ at $a=4$ |
| $S_{\text{prof}}(a)$ | "S profile of a" | the profiled score: $S$ with $b$ optimally tuned away, a function of $a$ only | $\min_b S(a,b)$ |
| $\chi^2_k$ | "chi-square with k degrees of freedom" | the distribution a profiled statistic follows; each profiled nuisance drops $k$ by 1 | — |

## The formal version

The **profiled** version of a two-knob score, keeping only the parameter of interest, is

$$
S_{\text{prof}}(a) \;=\; \min_{b}\, S(a, b),
$$

read aloud "S-profile of a equals the minimum over b of S of a, b" — everyday words: "the best
score $a$ can achieve, once you've tuned the nuisance $b$ to help it as much as possible." (For a
likelihood you'd write $\max_b$ instead — same idea, keep the *most plausible* nuisance.)

**Mapping it onto the toy example.** At each $a$:

- pick $\hat b(a)$, the $b$ that minimises $S(a, b)$ — the "tune the nuisance" step;
- record $S(a, \hat b(a))$ — the "keep only the best score" step.

We did this by hand: $\hat b(2) = 3 \Rightarrow S_{\text{prof}}(2) = 1$;
$\hat b(3) = 2 \Rightarrow S_{\text{prof}}(3) = 0$; $\hat b(4) = 1 \Rightarrow S_{\text{prof}}(4) = 1$.

Sanity-check that profiling then minimising over $a$ gives the same global minimum as minimising
the full two-knob surface directly:

```{r formal-check}
S <- function(a, b) (a - 3)^2 + (a + b - 5)^2

# Path 1: profile out b, then minimise the profile over a
prof <- function(a) optimize(function(b) S(a, b), interval = c(-20, 20))$objective
best_a <- optimize(prof, interval = c(-20, 20))
c(a_hat = best_a$minimum, profiled_min = best_a$objective)   # a ~ 3, min ~ 0

# Path 2: minimise the full 2-knob surface directly (grid, for a cross-check)
grid <- expand.grid(a = seq(0, 6, 0.01), b = seq(-4, 8, 0.01))
grid$S <- with(grid, S(a, b))
grid[which.min(grid$S), ]     # also a ~ 3, b ~ 2, S ~ 0 — same global optimum
```

### Why the profiled statistic keeps the right distribution — and loses one degree of freedom

Here is the accounting rule. Suppose $S(a, b)$ is a fit statistic that, at the *true*
parameter values, would follow a chi-square distribution. When you **profile out** a nuisance
parameter — replacing $b$ by its best-fit value $\hat b(a)$ — you've *spent* the data on
estimating that nuisance. The profiled statistic still follows a chi-square, but with **one fewer
degree of freedom** for each parameter you profiled out.

The intuition: degrees of freedom count "how many independent pieces of surprise are left after
you've used the data to set the knobs." Every knob you tune to *best-fit* the data soaks up one
piece of that surprise (it can always make the fit look a little better), so it no longer counts
toward the leftover. Start with a statistic that would be $\chi^2_n$ if you knew all the knobs;
profile out one nuisance parameter, and the leftover is $\chi^2_{n-1}$. This is the same
bookkeeping that turns $n$ into $n-1$ when you estimate a mean before computing a variance — you
"used up" one number.

```{webr}
#| autorun: true
# TRY IT: SIMULATE the degrees-of-freedom rule. We build a fit statistic over n=3 noisy numbers
# that would be chi-square_3 if the shared level were known, then PROFILE OUT that level (its
# best-fit is the mean). PREDICT: after profiling one parameter, is the average statistic 3 or 2?
set.seed(1)
n <- 3
n_sim <- 30000
stats <- replicate(n_sim, {
  y <- rnorm(n)                       # 3 standard-normal draws (true shared level is 0)
  lvl <- mean(y)                      # PROFILE OUT the nuisance "level": its best-fit is the mean
  sum((y - lvl)^2)                    # leftover squared miss after tuning the level away
})
cat("mean of profiled statistic:", round(mean(stats), 3),
    " (chi-square df should be n-1 = 2)\n")
# A chi-square with k df has mean k; we get ~2, NOT 3 -> profiling one parameter cost one df
```

## Explore it

```{webr}
#| autorun: true
#| fig-alt: "A U-shaped curve of the profiled score against the parameter of interest a, bottoming out at zero near a = 3 and rising symmetrically on both sides."
# TRY IT: change the '5' in (a+b-5) to another number and re-run — the bottom of the profile
# stays at score 0 but the best a shifts. PREDICT: does the CURVE'S BOTTOM value change?
S <- function(a, b) (a - 3)^2 + (a + b - 5)^2
prof <- function(a) optimize(function(b) S(a, b), interval = c(-20, 20))$objective
a_grid <- seq(0, 6, length.out = 200)
plot(a_grid, sapply(a_grid, prof), type = "l", lwd = 2,
     xlab = "a (parameter of interest)", ylab = "profiled score  min_b S(a, b)",
     main = "The profile curve: nuisance b tuned away at every a")
abline(v = 3, lty = 2, col = "tomato")
legend("top", legend = "best a = 3", lty = 2, col = "tomato", bty = "n")
```

```{webr}
#| autorun: true
# TRY IT: watch the nuisance SLIDE. For several a values, report the best-fit b-hat(a). PREDICT:
# does b-hat stay fixed or move as a changes?
S <- function(a, b) (a - 3)^2 + (a + b - 5)^2
bhat <- function(a) optimize(function(b) S(a, b), interval = c(-20, 20))$minimum
a_vals <- c(1, 2, 3, 4, 5)
data.frame(a = a_vals, best_b = round(sapply(a_vals, bhat), 3))
# b-hat(a) = 5 - a here: 4, 3, 2, 1, 0 — it slides, it is not one fixed value
```

```{webr}
#| autorun: true
# TRY IT: verify that the hand algebra and optimize() give the same profiled score.
# At a = 2 the nuisance best-fit is b-hat = 5 - a = 3, so S(2,3) = (2-3)^2 + (2+3-5)^2 = 1.
S <- function(a, b) (a - 3)^2 + (a + b - 5)^2

# By hand: plug b-hat(a) = 5 - a into S
hand_score <- S(2, 5 - 2)

# By optimize(): let R find the best b at a = 2
opt_score <- optimize(function(b) S(2, b), c(-10, 10))$objective

cat("hand formula:", hand_score, "\n")
cat("optimize():  ", opt_score, "\n")
cat("match:       ", all.equal(hand_score, opt_score), "\n")
```

## Common traps

**Trap: "fix the nuisance at one value and be done."** Tempting, because re-optimising every
time is more work. But a single fixed $b$ mis-scores every $a$ that would have preferred another
$b$. In our toy, fixing $b = 3$ (the best value *at $a=2$*) gives score $(a-3)^2 + (a-2)^2$,
whose minimum over $a$ is $0.5$ at $a = 2.5$ — **not** the true global minimum of $0$ at $a=3$.
Profiling (re-optimising $b$ at every $a$) recovers the real optimum; fixing $b$ does not.

**Trap: "profiling out a parameter doesn't change the degrees of freedom."** No — each nuisance
you profile out costs **one** degree of freedom. Our simulation showed a statistic that would be
$\chi^2_3$ becomes $\chi^2_2$ after profiling one parameter (average dropped from 3 to 2). Using
the un-reduced $df$ would give the wrong p-value: `pchisq(4, df = 3, lower.tail = FALSE)` ≈
$0.26$ versus the correct `pchisq(4, df = 2, ...)` ≈ $0.14$.

## Check yourself

**1.** For the score $S(a, b) = (a - 1)^2 + (a + b - 4)^2$, what is the best-fit $\hat b$ at
$a = 1$, and what is the profiled score there?

```{webr}
#| autorun: false
# check1-starter — edit and Run to check
S <- function(a, b) (a - 1)^2 + (a + b - 4)^2
```

::: {.callout-note collapse="true" title="Answer"}
At $a = 1$: $S = 0 + (1 + b - 4)^2 = (b - 3)^2$, minimised at $\hat b = 3$, giving profiled
score **0**. In R: `optimize(function(b) S(1, b), c(-10, 10))` returns `$minimum` near 3 and
`$objective` near 0.
:::

**2.** You start with a fit statistic that would follow $\chi^2_5$ if all knobs were known, then
profile out **two** nuisance parameters. What distribution does the profiled statistic follow?

::: {.callout-note collapse="true" title="Answer"}
$\chi^2_3$ — each profiled nuisance costs one degree of freedom, so $5 - 2 = 3$. You'd get the
p-value with `pchisq(stat, df = 3, lower.tail = FALSE)`.
:::

**3.** In your own words: explain to a colleague, in two sentences, what profiling out a nuisance
parameter does and why it costs a degree of freedom.

::: {.callout-note collapse="true" title="A model answer"}
Profiling replaces a parameter you don't care about with its best-fit value at every setting of
the parameter you do care about, collapsing a multi-knob score into a fair curve over just the
knob of interest. It costs one degree of freedom because tuning that nuisance to best-fit the
data soaks up one independent piece of the data's surprise, leaving one fewer for the leftover
statistic to carry.
:::

## What you can now do

- **Identify a nuisance parameter** and profile it out by optimising over it at each value of the
  parameter of interest, in R with `optimize` inside a function of the parameter you keep.
- **Explain why the nuisance is re-optimised at every point** (a single fixed value mis-scores
  the rest) and that the profile reaches the same global optimum as the full surface.
- **Apply the degrees-of-freedom rule:** each parameter profiled out drops the chi-square degrees
  of freedom by one.
- **Get the right p-value** by using the *reduced* degrees of freedom in `pchisq`.

If any of these feels shaky, revisit the toy example and the degrees-of-freedom section — a test
statistic that profiles out a parameter assumes all of them.

## Where this goes next

Any test statistic with a nuisance parameter needs profiling to get both a clean curve and the
right distribution. In a later course, you'll see a fit statistic that depends on an
angle (a reparametrised slope) you don't care about reporting — you profile that angle out by
minimising the statistic over it, and the "n minus 1" degrees of freedom you see there is
exactly this "started at n, profiled out one parameter" rule.

Ready for more reps? Do the [practice problems](practice.qmd).
