---
title: "Practice: Profile likelihood"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

Work these in order — they get harder. Do them on paper or in an R session; only then open the
answer.

## Problem 1 — the toy example, new numbers

For $S(a, b) = (a - 2)^2 + (a + b - 6)^2$, profile out $b$ at $a = 2$: find $\hat b$ and the
profiled score. Then check in R.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
S <- function(a, b) (a - 2)^2 + (a + b - 6)^2
```

::: {.callout-note collapse="true" title="Answer"}
At $a = 2$: $S = 0 + (2 + b - 6)^2 = (b - 4)^2$, minimised at $\hat b = 4$, giving profiled
score **0**.

```{r p1-ans}
S <- function(a, b) (a - 2)^2 + (a + b - 6)^2
o <- optimize(function(b) S(2, b), c(-10, 10))
c(b_hat = o$minimum, profiled_score = o$objective)   # 4, 0
```
:::

## Problem 2 — fill in the missing step

For $S(a, b) = (a - 3)^2 + (a + b - 5)^2$, profiling out $b$ at $a = 4$:

- Step 1: plug in $a = 4$: $S = (4-3)^2 + (4 + b - 5)^2 = 1 + (b - 1)^2$.
- Step 2: **your step here** — which $b$ minimises this, and what's the resulting score?
- Step 3: so $S_{\text{prof}}(4) = 1$.

Fill in step 2.

```{webr}
#| autorun: false
# p2-starter — edit and Run to check the blank
S <- function(a, b) (a - 3)^2 + (a + b - 5)^2
```

::: {.callout-note collapse="true" title="Answer"}
Step 2: $(b - 1)^2$ is smallest when $b = 1$, killing the second term and leaving $1 + 0 = 1$.
So $\hat b(4) = 1$ and the profiled score is $1$ — matching step 3. The move is: tune the
nuisance $b$ to zero out the term it appears in.

```{r p2-ans}
S <- function(a, b) (a - 3)^2 + (a + b - 5)^2
optimize(function(b) S(4, b), c(-10, 10))   # $minimum ~ 1, $objective ~ 1
```
:::

## Problem 3 — the whole profile curve

For $S(a, b) = (a - 3)^2 + (a + b - 5)^2$, compute the profiled score at $a = 0, 1, 2, 3, 4, 5,
6$ in R, and say where the curve bottoms out.

```{webr}
#| autorun: false
# p3-starter — your job: build prof(a) and apply it to the grid
S <- function(a, b) (a - 3)^2 + (a + b - 5)^2
a_vals <- c(0, 1, 2, 3, 4, 5, 6)
```

::: {.callout-note collapse="true" title="Answer"}
```{r p3-ans}
S <- function(a, b) (a - 3)^2 + (a + b - 5)^2
prof <- function(a) optimize(function(b) S(a, b), c(-20, 20))$objective
a_vals <- c(0, 1, 2, 3, 4, 5, 6)
round(sapply(a_vals, prof), 3)
```
The profiled scores are $9, 4, 1, 0, 1, 4, 9$ — a U-shape bottoming at **$a = 3$** with score
$0$. Profiling out $b$ leaves a clean one-knob curve over $a$, and its minimum is the whole
surface's global minimum.
:::

## Problem 4 — spot the error

A colleague wants the p-value for a fit statistic that starts as $\chi^2_4$, after profiling out
**one** nuisance parameter. They wrote — it contains **exactly one mistake**. Find it and fix
it.

```{webr}
#| autorun: false
# p4-starter — this contains exactly one mistake
stat <- 6
df <- 4                 # started as chi-square with 4 df
p <- pchisq(stat, df = df, lower.tail = FALSE)
```

::: {.callout-note collapse="true" title="Answer"}
The mistake is the **degrees of freedom**: profiling out one nuisance parameter drops $df$ from
$4$ to $3$. Using $df = 4$ overstates the p-value.

```{r p4-ans}
pchisq(6, df = 3, lower.tail = FALSE)   # ~0.112 — the correct p-value after profiling
```
:::

## Problem 5 — mixing it with optimization *(uses loss-functions-and-optimization's `optimize`)*

You have a fit statistic $S(a, b) = 2(a - 1)^2 + (b - a)^2$. Profile out $b$ **twice** — once by
hand, once with `optimize` — at $a = 1$, and confirm they agree.

```{webr}
#| autorun: false
# p5-starter — edit and Run to solve, right here
S <- function(a, b) 2 * (a - 1)^2 + (b - a)^2
```

::: {.callout-note collapse="true" title="Answer"}
By hand at $a = 1$: $S = 2\cdot 0 + (b - 1)^2 = (b - 1)^2$, minimised at $\hat b = 1$, score
$0$.

```{r p5-ans}
S <- function(a, b) 2 * (a - 1)^2 + (b - a)^2
optimize(function(b) S(1, b), c(-10, 10))   # $minimum ~ 1, $objective ~ 0 — agrees
```
The `optimize` call reproduces the hand answer ($\hat b = 1$, score $0$). Profiling is just
"minimise over the nuisance," which is exactly what `optimize` does — the tool you already know
from loss-functions-and-optimization.
:::

## Problem 6 — transfer: read the degrees-of-freedom rule

A test statistic combines $n = 6$ measurements and would be $\chi^2_6$ if a certain slope were
known. The method instead *estimates* that slope by profiling it out. What are the degrees of
freedom of the reported statistic, and why?

::: {.callout-note collapse="true" title="Answer"}
$\chi^2_5$: profiling out the one slope parameter costs one degree of freedom, so $6 - 1 = 5$.
The estimated slope soaks up one independent piece of the data's surprise, leaving five for the
statistic. (This is exactly the "$n - 1$ degrees of freedom" you'll meet in the proportional
colocalisation test, where the profiled parameter is the effect-size ratio.)
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**, once the
material isn't fresh. A pass then is real evidence you've retained it. Five questions, graded on
the page; 4/5 or better means this module is solid. Results are saved in this browser only.

::: {.module-check module="foundations/profile-likelihood" pass="4"}

::: {.quiz-q qid="pl-profiled-score" qtype="numeric" answer="1" tolerance="0"}
For $S(a, b) = (a - 3)^2 + (a + b - 5)^2$, what is the profiled score $\min_b S(a, b)$ at
$a = 2$?
:::

::: {.quiz-q qid="pl-what-is-nuisance" qtype="mc" answer="b"}
A nuisance parameter is:

a) the parameter you most want to report
b) a parameter the formula needs but that you don't care about reporting
c) a parameter fixed at zero
d) the degrees of freedom of the test
:::

::: {.quiz-q qid="pl-reoptimise" qtype="mc" answer="c"}
When profiling, why re-optimise the nuisance parameter at every value of the parameter of
interest instead of fixing it once?

a) It runs faster
b) The nuisance is always zero anyway
c) A single fixed value mis-scores every setting that would have preferred a different nuisance
d) It changes which parameter is the nuisance
:::

::: {.quiz-q qid="pl-df-rule" qtype="numeric" answer="3" tolerance="0"}
A statistic that would be $\chi^2_5$ with all knobs known has how many degrees of freedom after
you profile out **two** nuisance parameters?
:::

::: {.quiz-q qid="pl-optimize-interleave" qtype="mc" answer="a"}
*(uses loss-functions-and-optimization)* In R, `optimize(function(b) S(a, b), c(-10, 10))`
profiles out $b$ at a fixed $a$ by returning:

a) `$minimum` (the best $b$) and `$objective` (the profiled score there)
b) the derivative of $S$ with respect to $a$
c) the full 2-D surface of $S$
d) the degrees of freedom of the statistic
:::

:::

::::
