---
title: "Matrices, and what it means to multiply them"
subtitle: "How a table of numbers acts on a vector — the machinery papers write as Ax, XᵀX, and A⁻¹"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live cells below re-state their own data. Base R is all
# we need, so nothing to install.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

::: {.callout-tip title="This lesson runs in your browser" collapse="true"}
The code cells below with a **Run** button are live — powered by WebR (R compiled to
WebAssembly), they run right here on your phone or laptop, no install needed. Edit any
of them and re-run. The first run downloads a small R engine, so give it a few seconds.
Cells without a Run button are ordinary listings whose output was computed when the
page was built.
:::

## Where we're going

A **matrix** is a table of numbers — rows and columns of them — and almost every
statistical method eventually packs its data into one. The magic isn't the table
itself; it's what *multiplying* by a matrix does: it takes a whole vector of numbers
and transforms it into a new one, in a single stroke. Once you can read `A %*% x`,
`t(X) %*% X`, and `solve(A)` as ordinary R you already understand, the matrix
notation in papers ($A\mathbf{x}$, $X^\top X$, $A^{-1}$) stops being a wall.

This lesson stays with concrete, hand-checkable tables (recipes, prices, batches) and
base R. We will *not* touch eigenvalues, decompositions, or any regression framing —
just the five moves you need to read matrix notation: building one, multiplying it by
a vector, multiplying two of them, flipping one over (transpose), and undoing one
(inverse).

**Builds on:** [vectors & summation](../vectors-and-summation/lesson.qmd) — matrix
multiplication is just one dot product per row, and a dot product is a sum of products
($\sum_i a_i b_i$); and [reading math notation](../reading-math-notation/lesson.qmd)
(soft — for the symbol conventions like subscripts and bold letters).

## Warm-up: do you remember?

Three quick reps first — produce each answer from memory, then check.

**1.** For `a <- c(2, 1, 1)` and `b <- c(3, 4, 5)`, what single number is the **dot
product** of `a` and `b`? Say the formula in words, then give the value.

```{webr}
#| autorun: false
# warmup1-starter — edit and Run to check
a <- c(2, 1, 1)
b <- c(3, 4, 5)
```

::: {.callout-note collapse="true" title="Answer"}
"Multiply the paired entries and add them up": $2\times3 + 1\times4 + 1\times5 =
6 + 4 + 5 = 15$. In R that's `sum(a * b)`, which is exactly what the dot product means.
This one operation — a **sum of products** — is the entire engine of matrix
multiplication below. Shaky? Revisit
[vectors & summation](../vectors-and-summation/lesson.qmd).
:::

**2.** Read this aloud and give its value for the same `b <- c(3, 4, 5)`:
$\sum_{i=1}^{n} b_i$. What are $\sum_{i=1}^{n} b_i$ and $n$ here?

```{webr}
#| autorun: false
# warmup2-starter — edit and Run to check
b <- c(3, 4, 5)
```

::: {.callout-note collapse="true" title="Answer"}
"The sum from i equals 1 to n of b-sub-i" — add them all: $3 + 4 + 5 = 12$, with
$n = 3$. In R: `sum(b)` and `length(b)`. See
[vectors & summation](../vectors-and-summation/lesson.qmd).
:::

**3.** In R, `x[3]` picks the third entry of a vector. If `A` is a table of numbers,
how do you think you'd pick the number in **row 2, column 3**?

::: {.callout-note collapse="true" title="Answer"}
`A[2, 3]` — row index first, then column index. That two-number address is the whole
idea of a matrix: every entry has a *(row, column)* label. We'll write it $A_{ij}$ in
math ("A-sub-i-j"), which is just `A[i, j]` in R clothing — subscript addressing like
this is [reading math notation](../reading-math-notation/lesson.qmd) territory.
:::

## A tiny example first

You run a tiny bakery with **two recipes**. Each recipe needs some cups of three
ingredients — flour, sugar, butter:

- **Recipe 1** needs 2 flour, 1 sugar, 1 butter.
- **Recipe 2** needs 1 flour, 2 sugar, 0 butter.

Stack those two rows into a grid and you have a **matrix** — call it $A$. It has 2
rows (one per recipe) and 3 columns (one per ingredient), so we say it's a "2 by 3"
matrix:

$$
A = \begin{bmatrix} 2 & 1 & 1 \\ 1 & 2 & 0 \end{bmatrix}
$$

The number in **row 1, column 2** is 1 (recipe 1 uses 1 cup of sugar). The number in
**row 2, column 3** is 0 (recipe 2 uses no butter). That's all a matrix is: a labelled
grid, read *(row, column)*.

**Now the useful part — matrix times vector.** Suppose the three ingredients cost 3, 4,
5 dollars per cup. Put those prices in a vector $\mathbf{x} = (3, 4, 5)$. What does each
recipe *cost*? You'd take each recipe's row and dot it with the prices:

- Recipe 1: $2\times3 + 1\times4 + 1\times5 = 6 + 4 + 5 = \mathbf{15}$.
- Recipe 2: $1\times3 + 2\times4 + 0\times5 = 3 + 8 + 0 = \mathbf{11}$.

So $A$ times the price vector gives a new vector $(15, 11)$ — the cost of each recipe.
**Matrix times vector = one dot product per row.** That single idea is the heart of
everything below.

The same thing in R, so you can check every step:

```{r toy-example}
# The recipe matrix: 2 rows (recipes) x 3 columns (ingredients).
# byrow = TRUE fills row-by-row, so it reads like the grid above.
A <- matrix(c(2, 1, 1,
              1, 2, 0), nrow = 2, byrow = TRUE)
A

dim(A)        # 2 3  -> two rows, three columns
A[1, 2]       # 1  -> row 1, column 2 (recipe 1's sugar)
A[2, 3]       # 0  -> row 2, column 3 (recipe 2's butter)

# Prices per ingredient
x <- c(3, 4, 5)

# Matrix times vector: %*% does one dot product per row automatically
A %*% x       # 15  (recipe 1's cost) and 11 (recipe 2's cost)

# Same thing, one row at a time, to prove %*% is just dot products:
sum(A[1, ] * x)   # 15
sum(A[2, ] * x)   # 11
```

## Getting a feel for it

**A matrix is a machine that eats a vector and spits out a vector.** The picture to
hold: feed the price vector $(3, 4, 5)$ into the box labelled $A$, and out comes the
cost vector $(15, 11)$. Change the prices and different costs come out; the *box*
(the recipes) stays the same. A matrix isn't really a static table — it's a **rule for
transforming one list of numbers into another**. That's why the topic is called
"linear transforms": each matrix is one transformation. (Where the analogy strains: a
matrix can only do *linear* transformations — scale-and-add — never anything with a
square or a curve in it. Doubling every price exactly doubles every cost, always.)

**Why one dot product per row?** Because each row is a self-contained little recipe,
and a dot product is exactly the "how much does this cost" question: pair each
ingredient's amount with its price, multiply, and add. Row 1 asks the cost question
for recipe 1; row 2 asks it for recipe 2. Stacking the answers gives the output
vector. The rows never interact — that's why a matrix with more rows just produces a
longer answer, one entry per row.

**Why the column count has to match.** Recipe 1 lists 3 ingredient amounts, so the
price vector *must* have 3 entries — you can't dot a length-3 row with a length-2
list; there'd be a leftover number with nothing to pair. This is the whole reason
matrix shapes have to "line up," which we'll pin down in a moment. For now: **the
number of columns of the matrix must equal the length of the vector it eats.**

**What the output feels like.** Our costs came out as $(15, 11)$ — recipe 1 is the
pricier one, and that feels right: it uses more total, pricier ingredients. Drop
butter's price to 0 and recipe 1's cost falls to 10 while recipe 2 (which uses no
butter) doesn't move at all. The output vector is just "the same question answered for
every row," so it always has one entry per row of the matrix.

## What the notation is going to say

When a paper writes $A\mathbf{x}$ (a matrix $A$ times a vector $\mathbf{x}$, with no
symbol between them — the invisible "×" is just "apply the matrix"), it means exactly
what we just did by hand: *go down the rows of $A$, and for each row take its dot
product with $\mathbf{x}$; collect those numbers into a new vector.* In our example
$A\mathbf{x}$ is the "$2\times3 + 1\times4 + 1\times5 = 15$, then
$1\times3 + 2\times4 + 0\times5 = 11$" we already computed — the answer is $(15, 11)$.
The transpose $A^\top$ and the inverse $A^{-1}$ we'll meet the same way: a familiar R
call dressed in a symbol.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $A$ | "the matrix A" | a table of numbers, rows × columns (in R: `matrix(...)`) | the 2×3 recipe grid |
| $A_{ij}$ | "A-sub-i-j" | the entry in row $i$, column $j$ (in R: `A[i, j]`) | $A_{12} = 1$, $A_{23} = 0$ |
| $\mathbf{x}$ (bold) | "the vector x" | a whole column of numbers at once (an R vector!) | the prices $(3, 4, 5)$ |
| $x_j$ | "x-sub-j" | the $j$-th entry of the vector $\mathbf{x}$ (in R: `x[j]`) | $x_2 = 4$ |
| $n \times p$ | "n by p" | a matrix's shape: $n$ rows, $p$ columns | $A$ is $2 \times 3$ |
| $A\mathbf{x}$ | "A times x" | matrix times vector: one dot product per row (in R: `A %*% x`) | $(15, 11)$ |
| $AB$ | "A times B" | matrix times matrix: every row of $A$ dotted with every column of $B$ (in R: `A %*% B`) | (below) |
| $B_{jk}$ | "B-sub-j-k" | the entry in row $j$, column $k$ of $B$ (in R: `B[j, k]`) | used in $AB$ below |
| $\sum_{j=1}^{p}$ | "the sum from j equals 1 to p" | add the term to its right once per column $j$ — a dot product ([vectors & summation](../vectors-and-summation/lesson.qmd)) | $\sum_j A_{1j}x_j = 15$ |
| $A^\top$ or $A'$ | "A transpose" | the matrix flipped so rows become columns (in R: `t(A)`) | a 3×2 grid |
| $A^\top A$ or $A'A$ | "A transpose A" | a matrix times its own flip — always square and symmetric | 3×3 (below) |
| $I$ | "the identity matrix" | the matrix that changes nothing, 1s on the diagonal, 0s elsewhere (in R: `diag(n)`) | `diag(3)` |
| $A^{-1}$ | "A inverse" | the matrix that undoes $A$: $A^{-1}A = I$ (in R: `solve(A)`) | (below, for a square $A$) |

Bold lowercase ($\mathbf{x}$) is a vector; a capital letter ($A$, $I$) is a matrix.
When two symbols sit side by side with nothing between them ($A\mathbf{x}$, $AB$),
that juxtaposition *is* the multiply — the same "implied multiplication" as $2x$
meaning $2 \times x$.

## The formal version

**Matrix times vector.** The output's $i$-th entry is the dot product of row $i$ with
the vector:

$$
(A\mathbf{x})_i = \sum_{j=1}^{p} A_{ij}\, x_j
$$

Reading it aloud: "the i-th entry of A-x equals the sum, from j equals 1 to p, of
A-sub-i-j times x-sub-j" — or in everyday words: **"walk along row $i$, multiply each
matrix entry by the matching vector entry, and add them up."** The sum runs over $j$
(the columns) because that's what a dot product does; there's a separate answer for
each row $i$. Mapping onto the toy example, row $i = 1$ gives
$A_{11}x_1 + A_{12}x_2 + A_{13}x_3 = 2\cdot3 + 1\cdot4 + 1\cdot5 = 15$ — the "$15$" we
got by hand.

**Matrix times matrix.** Now multiply $A$ (2×3) by another matrix $B$. Think of $B$ as
holding several *different* vectors side by side, one per column — say two price lists,
"weekday" and "weekend." You just do the matrix-times-vector move once per column of
$B$. Entry $(i, k)$ of the answer is **row $i$ of $A$ dotted with column $k$ of $B$**:

$$
(AB)_{ik} = \sum_{j=1}^{p} A_{ij}\, B_{jk}
$$

Reading it aloud: "the (i, k) entry of A-B equals the sum over j of A-sub-i-j times
B-sub-j-k" — everyday words: **"row $i$ of $A$, dotted with column $k$ of $B$."**

Here's the shape rule that falls straight out of "the row and column must have equal
length": for row $i$ of $A$ (length = $A$'s column count) to dot with column $k$ of
$B$ (length = $B$'s row count), **$A$'s column count must equal $B$'s row count.** In
shorthand, an $n \times p$ matrix can only multiply a $p \times m$ matrix — the two
inner numbers ($p$ and $p$) must match, and the answer is $n \times m$ (the two outer
numbers). This matching-up is called **conformability**; when the inner numbers
disagree, the multiplication is simply undefined and R throws a "non-conformable"
error. A matrix-times-*vector* is just the special case where $B$ has one column
($p \times 1$).

Let's compute an $AB$ by hand, then check it. Take

$$
B = \begin{bmatrix} 1 & 0 \\ 2 & 1 \\ 0 & 3 \end{bmatrix}
\quad (3\times 2), \qquad A\ \text{is}\ 2\times 3,\ \text{so}\ AB\ \text{is}\ 2\times 2.
$$

- $(AB)_{11}$ = row 1 of $A$ · column 1 of $B$ = $2\cdot1 + 1\cdot2 + 1\cdot0 = 4$
- $(AB)_{12}$ = row 1 of $A$ · column 2 of $B$ = $2\cdot0 + 1\cdot1 + 1\cdot3 = 4$
- $(AB)_{21}$ = row 2 of $A$ · column 1 of $B$ = $1\cdot1 + 2\cdot2 + 0\cdot0 = 5$
- $(AB)_{22}$ = row 2 of $A$ · column 2 of $B$ = $1\cdot0 + 2\cdot1 + 0\cdot3 = 2$

So $AB = \begin{bmatrix} 4 & 4 \\ 5 & 2 \end{bmatrix}$. Sanity check in R:

```{r ab-check}
A <- matrix(c(2, 1, 1,
              1, 2, 0), nrow = 2, byrow = TRUE)
B <- matrix(c(1, 0,
              2, 1,
              0, 3), nrow = 3, byrow = TRUE)

A %*% B        # matches our hand answer: 4 4 / 5 2
```

**Transpose — flip rows and columns.** The transpose $A^\top$ turns each row into a
column (and each column into a row): a $2\times3$ becomes a $3\times2$. Entry $(i, j)$
of $A$ lands at entry $(j, i)$ of $A^\top$ — that's the whole rule, "swap the address."

$$
A = \begin{bmatrix} 2 & 1 & 1 \\ 1 & 2 & 0 \end{bmatrix}
\qquad\Longrightarrow\qquad
A^\top = \begin{bmatrix} 2 & 1 \\ 1 & 2 \\ 1 & 0 \end{bmatrix}
$$

Why do papers write $A^\top A$ (or $A'A$ — the prime is an older notation for the same
flip) *everywhere*? Because it's a compact way to say "dot every column of $A$ with
every other column of $A$" — all the pairwise sums-of-products among the columns, laid
out in one square, symmetric grid. Whatever $A$'s shape, $A^\top A$ is always square
($p \times p$) and always conformable, which is exactly why it shows up whenever
someone needs a square matrix built from rectangular data. Let's build ours:

```{r transpose-check}
A <- matrix(c(2, 1, 1,
              1, 2, 0), nrow = 2, byrow = TRUE)

t(A)              # the flip: a 3x2 matrix

t(A) %*% A        # A-transpose-A: a 3x3 SYMMETRIC matrix
#      [,1] [,2] [,3]
# [1,]    5    4    2
# [2,]    4    5    1
# [3,]    2    1    1
# Symmetric: entry (1,2) = entry (2,1) = 4, and so on. Always true for t(A) %*% A.
```

**The identity, and the inverse.** Among all matrices there's one that changes
nothing: the **identity** $I$, with 1s down the diagonal and 0s everywhere else.
Multiplying by $I$ is like multiplying a number by 1 — you get back exactly what you
started with. And just as every nonzero number $c$ has a reciprocal $1/c$ that undoes
multiplication ($c \times \tfrac{1}{c} = 1$), a square matrix $A$ can have an
**inverse** $A^{-1}$ that undoes it: $A^{-1}A = I$. If $A$ transforms $\mathbf{v}$
into $\mathbf{w}$, then $A^{-1}$ transforms $\mathbf{w}$ back into $\mathbf{v}$.

We need a *square* matrix for an inverse to make sense (undoing has to land you back
where you started). Take
$M = \begin{bmatrix} 2 & 1 \\ 1 & 1 \end{bmatrix}$. In R:

```{r inverse-check}
I3 <- diag(3)     # the 3x3 identity: 1s on the diagonal, 0s elsewhere
I3

A <- matrix(c(2, 1, 1,
              1, 2, 0), nrow = 2, byrow = TRUE)
A %*% diag(3)     # multiplying by I changes NOTHING: we get A back

M <- matrix(c(2, 1,
              1, 1), nrow = 2, byrow = TRUE)
solve(M)          # the inverse — clean integers here: 1 -1 / -1 2
M %*% solve(M)    # M times its inverse = the identity (1s on diagonal)

# "Undo" in action: transform v, then invert to recover it
v <- c(3, 5)
w <- M %*% v      # 11, 8  -- M sends v to w
solve(M) %*% w    # 3, 5   -- the inverse brings w back to v
```

`solve()` is R's inverse (its name comes from *solving* $A\mathbf{x} = \mathbf{b}$,
which the inverse does). Reading $A^{-1}$ aloud: "A inverse" — the matrix that undoes
$A$.

## Explore it

```{webr}
#| autorun: true
# explore-1 — LIVE: matrix times vector IS one dot product per row.
# TRY IT: change the prices in x, or edit a recipe row in A, then Run.
# PREDICT: if you set x <- c(1, 1, 1), what will A %*% x give? (Hint: it counts
# total cups per recipe.) Commit to an answer, then run.
A <- matrix(c(2, 1, 1,
              1, 2, 0), nrow = 2, byrow = TRUE)
x <- c(3, 4, 5)

built_in <- A %*% x                              # R's matrix multiply
by_hand  <- c(sum(A[1, ] * x), sum(A[2, ] * x))  # the same, dot product per row

cbind(built_in = built_in, by_hand = by_hand)    # identical columns
```

```{webr}
#| autorun: true
# explore-2 — LIVE: order matters! A %*% B is NOT the same as B %*% A.
# PREDICT: A is 2x3 and B is 3x2. What SHAPE is A %*% B? What shape is B %*% A?
# Write both down, then run and see they differ in shape AND numbers.
A <- matrix(c(2, 1, 1,
              1, 2, 0), nrow = 2, byrow = TRUE)
B <- matrix(c(1, 0,
              2, 1,
              0, 3), nrow = 3, byrow = TRUE)

A %*% B      # a 2x2 result
B %*% A      # a 3x3 result -- completely different: matrix multiply is not commutative
```

```{webr}
#| autorun: false
# explore-3 — LIVE: the inverse undoes the matrix. Then break it.
# PREDICT: solve(M) %*% M -- what matrix should you get back? Write it, then run.
M <- matrix(c(2, 1,
              1, 1), nrow = 2, byrow = TRUE)

solve(M) %*% M          # the identity: undoing then doing = no change

# Now try a SINGULAR matrix (row 2 is exactly 2x row 1) -- uncomment and Run:
# S <- matrix(c(1, 2,
#               2, 4), nrow = 2, byrow = TRUE)
# solve(S)              # ERROR: "system is exactly singular" -- no inverse exists
```

## Common traps

**Trap 1 — "matrix multiplication is commutative, so $AB = BA$."** With ordinary
numbers $3 \times 4 = 4 \times 3$, so it's tempting to assume matrices behave the same.
They do not — order changes both the numbers *and often the shape*. With our $A$
(2×3) and $B$ (3×2), $AB$ is a 2×2 matrix but $BA$ is a 3×3 matrix — they can't even
be compared, let alone equal. Even when both orders are defined and same-shape, the
entries usually differ, because "row $i$ of $A$ dotted with column $k$ of $B$" is a
different computation from "row $i$ of $B$ dotted with column $k$ of $A$." Always keep
the order the paper wrote.

**Trap 2 — "every square matrix has an inverse."** Nonzero numbers always have a
reciprocal, so it feels like square matrices should too. But a matrix whose rows carry
redundant information has *no* inverse — it's called **singular**. Take
$S = \begin{bmatrix} 1 & 2 \\ 2 & 4 \end{bmatrix}$: row 2 is exactly twice row 1, so
$S$ throws away information (it squashes different inputs onto the same output), and
there's no way to undo that. `solve(S)` doesn't return a matrix — it errors with
"system is exactly singular." This is the matrix version of "you can't divide by
zero," and it's precisely why papers fuss over whether a matrix is *invertible*.

## Check yourself

**1.** For the matrix below and the vector $\mathbf{u} = (4, 5)$, compute $C\mathbf{u}$
by hand (two dot products), then check in R.

$$
C = \begin{bmatrix} 1 & 2 \\ 3 & 0 \end{bmatrix}
$$

```{webr}
#| autorun: false
# check1-starter — edit and Run to check your hand answer
C <- matrix(c(1, 2,
              3, 0), nrow = 2, byrow = TRUE)
u <- c(4, 5)
```

::: {.callout-note collapse="true" title="Answer"}
Row 1 · $\mathbf{u}$: $1\times4 + 2\times5 = 4 + 10 = 14$. Row 2 · $\mathbf{u}$:
$3\times4 + 0\times5 = 12 + 0 = 12$. So $C\mathbf{u} = (14, 12)$. In R,
`C %*% u` returns 14 and 12 — one dot product per row, exactly as by hand.
:::

**2.** A matrix $X$ is $5 \times 3$ (5 rows, 3 columns). What shape is $X^\top X$? Is
it guaranteed to be square? Could you form $X X$ (i.e. $X$ times $X$)?

::: {.callout-note collapse="true" title="Answer"}
$X^\top$ is $3 \times 5$, so $X^\top X$ is $(3\times5)\times(5\times3) = 3 \times 3$ —
the inner numbers (5 and 5) match, and the result is the square of the *column* count.
$X^\top X$ is always square. But $X X$ is $(5\times3)\times(5\times3)$: the inner
numbers are 3 and 5, which disagree, so $X X$ is **not conformable** and undefined.
That's why the transpose is there — it makes the shapes line up.
:::

**3.** In your own words: explain to a colleague, in two sentences, what it means to
multiply a matrix by a vector, and why the vector's length has to match the matrix's
number of columns.

::: {.callout-note collapse="true" title="A model answer"}
Multiplying a matrix by a vector runs one dot product per row: each row is paired
up entry-by-entry with the vector, multiplied, and summed, giving one output number
per row. The vector must have one entry for every column, because otherwise a row
entry would have no vector entry to pair with — the dot product wouldn't line up.
:::

## What you can now do

- **Build and read a matrix in R** — create one with `matrix(..., byrow = TRUE)`, read
  its shape with `dim()`, and pull entry $(i, j)$ with `A[i, j]`.
- **Multiply a matrix by a vector by hand**, as one dot product per row, and get the
  same answer as `%*%`.
- **Multiply two matrices** as row-by-column dot products, and check *conformability*
  (inner dimensions must match) before you even start.
- **Take a transpose** with `t()`, and explain why $X^\top X$ is always a square,
  symmetric matrix that papers lean on.
- **Recognize the identity and the inverse** — that $I$ changes nothing, that
  $A^{-1}$ undoes $A$ via `solve()`, and that a *singular* matrix has no inverse.

If any of these feels shaky, revisit [A tiny example first](#a-tiny-example-first) and
[The formal version](#the-formal-version) before continuing — anything that packs data
into a matrix assumes all of these.

## Where this goes next

Matrices are the language every multi-variable method is written in: once several
predictors, several observations, or several equations enter the picture, they get
stacked into a matrix and the method is written as a few matrix operations. The moves
here — $A\mathbf{x}$, $AB$, $A^\top$, $A^{-1}$ — are the alphabet for reading those.
Ready for more reps? Do the [practice problems](practice.qmd).
