---
title: "Practice: Matrices, and what it means to multiply them"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only (a {webr} starter does not see it — each starter re-states its
# own data). Base R is all we need, so nothing to install.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

Work these in order — they get harder. Do them on paper or in an R session; only
then open the answer.

## Problem 1 — the toy example, new numbers

A new recipe grid $C$ and a price vector $\mathbf{u}$:

$$
C = \begin{bmatrix} 1 & 2 \\ 3 & 0 \end{bmatrix}, \qquad \mathbf{u} = (4, 5)
$$

By hand, compute $C\mathbf{u}$ as one dot product per row. Then check with `%*%`.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
C <- matrix(c(1, 2,
              3, 0), nrow = 2, byrow = TRUE)
u <- c(4, 5)
```

::: {.callout-note collapse="true" title="Answer"}
Row 1 dotted with $\mathbf{u}$: $1\times4 + 2\times5 = 4 + 10 = 14$.
Row 2 dotted with $\mathbf{u}$: $3\times4 + 0\times5 = 12 + 0 = 12$.
So $C\mathbf{u} = (14, 12)$.

```{r p1-answer}
C <- matrix(c(1, 2,
              3, 0), nrow = 2, byrow = TRUE)
u <- c(4, 5)
C %*% u                              # 14, 12
c(sum(C[1, ] * u), sum(C[2, ] * u))  # 14, 12 — same, one dot product per row
```
:::

## Problem 2 — fill in the missing step (matrix × matrix)

Here is $PQ$ worked out — but the **core step is blanked**. Fill in ONLY the blanked
line: the four row-by-column dot products.

$$
P = \begin{bmatrix} 1 & 2 \\ 0 & 1 \end{bmatrix}, \qquad
Q = \begin{bmatrix} 3 & 1 \\ 2 & 4 \end{bmatrix}
$$

> Both are 2×2, and $P$'s columns (2) match $Q$'s rows (2), so $PQ$ is 2×2.
> **`<your step here>`** — each entry is row $i$ of $P$ dotted with column $k$ of $Q$:
> $(PQ)_{11} = \underline{\phantom{00}}$, $(PQ)_{12} = \underline{\phantom{00}}$,
> $(PQ)_{21} = \underline{\phantom{00}}$, $(PQ)_{22} = \underline{\phantom{00}}$.
> Collect them into the 2×2 answer.

```{webr}
#| autorun: false
# p2-starter — edit and Run to check your four entries
P <- matrix(c(1, 2,
              0, 1), nrow = 2, byrow = TRUE)
Q <- matrix(c(3, 1,
              2, 4), nrow = 2, byrow = TRUE)
```

::: {.callout-note collapse="true" title="Answer"}
The blanked step is the heart of matrix multiplication — row of $P$, column of $Q$:

- $(PQ)_{11}$ = row 1 of $P$ · column 1 of $Q$ = $1\times3 + 2\times2 = 7$
- $(PQ)_{12}$ = row 1 of $P$ · column 2 of $Q$ = $1\times1 + 2\times4 = 9$
- $(PQ)_{21}$ = row 2 of $P$ · column 1 of $Q$ = $0\times3 + 1\times2 = 2$
- $(PQ)_{22}$ = row 2 of $P$ · column 2 of $Q$ = $0\times1 + 1\times4 = 4$

So $PQ = \begin{bmatrix} 7 & 9 \\ 2 & 4 \end{bmatrix}$.

```{r p2-answer}
P <- matrix(c(1, 2,
              0, 1), nrow = 2, byrow = TRUE)
Q <- matrix(c(3, 1,
              2, 4), nrow = 2, byrow = TRUE)
P %*% Q                    # 7 9 / 2 4 — same as the hand answer
```
:::

## Problem 3 — transpose and $X^\top X$

For the matrix $X$ below, write $X^\top$ (the flip), then compute $X^\top X$ by hand
and confirm it is square and symmetric. Check in R.

$$
X = \begin{bmatrix} 1 & 2 \\ 3 & 4 \\ 5 & 6 \end{bmatrix}
$$

```{webr}
#| autorun: false
# p3-starter — your job: t(X), then t(X) %*% X
X <- matrix(c(1, 2,
              3, 4,
              5, 6), nrow = 3, byrow = TRUE)
```

::: {.callout-note collapse="true" title="Answer"}
The transpose swaps rows and columns, so the 3×2 $X$ becomes a 2×3:
$X^\top = \begin{bmatrix} 1 & 3 & 5 \\ 2 & 4 & 6 \end{bmatrix}$.

$X^\top X$ is $(2\times3)\times(3\times2) = 2\times2$. Each entry is a column of $X$
dotted with a column of $X$:

- $(1,1)$ = column 1 · column 1 = $1^2 + 3^2 + 5^2 = 1 + 9 + 25 = 35$
- $(1,2) = (2,1)$ = column 1 · column 2 = $1\cdot2 + 3\cdot4 + 5\cdot6 = 2 + 12 + 30 = 44$
- $(2,2)$ = column 2 · column 2 = $2^2 + 4^2 + 6^2 = 4 + 16 + 36 = 56$

So $X^\top X = \begin{bmatrix} 35 & 44 \\ 44 & 56 \end{bmatrix}$ — square, and
symmetric because the off-diagonal entries are equal (44 = 44).

```{r p3-answer}
X <- matrix(c(1, 2,
              3, 4,
              5, 6), nrow = 3, byrow = TRUE)
t(X)                # the flip: 2x3
t(X) %*% X          # 35 44 / 44 56 — square and symmetric
```
:::

## Problem 4 — spot the error (conformability)

A colleague wants the product of two matrices and writes the following. The reasoning
is **flawed** — find the core error, explain why it's wrong, and say what (if anything)
would fix it.

> "Let $D$ be 2×3 and $E$ be 2×3. Both are the same shape, so I can multiply them:
> $DE$ is another 2×3 matrix. I'll just dot each row of $D$ with the matching row of
> $E$."

```{webr}
#| autorun: false
# p4-starter — copy this to see R reject the bad multiplication
D <- matrix(1:6, nrow = 2)   # a 2x3 matrix
E <- matrix(1:6, nrow = 2)   # another 2x3 matrix
# D %*% E   # <- uncomment: R errors "non-conformable arguments"
```

::: {.callout-note collapse="true" title="Answer"}
The core error is "same shape means I can multiply them." Conformability is about the
**inner** dimensions: to compute $DE$ you dot each *row* of $D$ (length = $D$'s column
count = 3) with each *column* of $E$ (length = $E$'s row count = 2). Those lengths, 3
and 2, don't match, so $DE$ is undefined — `D %*% E` errors with "non-conformable
arguments." (A second giveaway: matrix multiply is row-of-first *dotted with column-of-
second*, never "row with row.")

A fix: transpose $E$ so its shape becomes 3×2. Then $D\,E^\top$ is
$(2\times3)\times(3\times2) = 2\times2$, which is well-defined.

```{r p4-answer}
D <- matrix(1:6, nrow = 2)          # 2x3
E <- matrix(1:6, nrow = 2)          # 2x3
D %*% t(E)                          # works: 2x2 (inner dims 3 and 3 match)
```
:::

## Problem 5 — mixing it with vectors & summation *(uses [vectors-and-summation](../vectors-and-summation/lesson.qmd)'s dot product)*

Two recipes' ingredient amounts sit in $A$; ingredient prices sit in the vector
$\mathbf{w}$. Compute $A\mathbf{w}$ two ways — (a) directly with `%*%`, and (b) as an
explicit **dot product** per row using `sum(...)`, the way you learned in
vectors & summation — and confirm they agree.

$$
A = \begin{bmatrix} 1 & 0 & 2 \\ 2 & 1 & 1 \end{bmatrix}, \qquad \mathbf{w} = (2, 3, 1)
$$

**PREDICT** first: what will the two entries of $A\mathbf{w}$ be? Write them down,
then run.

```{webr}
#| autorun: false
# p5-starter — edit and Run to solve, right here
A <- matrix(c(1, 0, 2,
              2, 1, 1), nrow = 2, byrow = TRUE)
w <- c(2, 3, 1)
```

::: {.callout-note collapse="true" title="Answer"}
Matrix-times-vector is *literally* one dot product per row — the very sum-of-products
from vectors & summation, $\sum_j A_{ij} w_j$:

- Row 1 · $\mathbf{w}$: $1\times2 + 0\times3 + 2\times1 = 2 + 0 + 2 = 4$
- Row 2 · $\mathbf{w}$: $2\times2 + 1\times3 + 1\times1 = 4 + 3 + 1 = 8$

So $A\mathbf{w} = (4, 8)$.

```{r p5-answer}
A <- matrix(c(1, 0, 2,
              2, 1, 1), nrow = 2, byrow = TRUE)
w <- c(2, 3, 1)

A %*% w                                 # 4, 8
c(sum(A[1, ] * w), sum(A[2, ] * w))     # 4, 8 — the dot product per row, spelled out
```

The `sum(A[i, ] * w)` line is exactly the dot product from
[vectors & summation](../vectors-and-summation/lesson.qmd); `%*%` just does it for
every row at once.
:::

## Problem 6 — a gentle cliffhanger (solving a system)

You have two equations in two unknowns:

$$
2x + \phantom{1}y = 5, \qquad x + 3y = 10.
$$

These can be packed into a matrix equation $A\mathbf{z} = \mathbf{b}$, where
$A = \begin{bmatrix} 2 & 1 \\ 1 & 3 \end{bmatrix}$, $\mathbf{z} = (x, y)$, and
$\mathbf{b} = (5, 10)$. You just met the tool that undoes a matrix — the inverse.
**Predict:** if $A\mathbf{z} = \mathbf{b}$, and $A^{-1}$ undoes $A$, what do you get if
you multiply both sides by $A^{-1}$ on the left? Then run the starter to find $x$ and
$y$.

```{webr}
#| autorun: false
# p6-starter — solve for z = (x, y). Try solve(A) %*% b, and also solve(A, b).
A <- matrix(c(2, 1,
              1, 3), nrow = 2, byrow = TRUE)
b <- c(5, 10)
```

::: {.callout-note collapse="true" title="Answer"}
Multiplying both sides by $A^{-1}$ on the left gives
$A^{-1}A\,\mathbf{z} = A^{-1}\mathbf{b}$. But $A^{-1}A = I$ (the inverse undoes the
matrix), and $I$ changes nothing, so the left side is just $\mathbf{z}$. That leaves
$\mathbf{z} = A^{-1}\mathbf{b}$ — **the inverse solves the system.**

```{r p6-answer}
A <- matrix(c(2, 1,
              1, 3), nrow = 2, byrow = TRUE)
b <- c(5, 10)

solve(A) %*% b     # z = (1, 3): so x = 1, y = 3
solve(A, b)        # 1, 3 — R's direct solver (preferred; no explicit inverse)
```

Check: $2(1) + 3 = 5$ ✓ and $1 + 3(3) = 10$ ✓. This is why `solve()` is *named*
"solve" — its everyday job is solving $A\mathbf{z} = \mathbf{b}$, and computing the
full inverse is really a special case. Turning a whole method into "stack it into a
matrix, then apply an inverse" is exactly what matrix notation buys you.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/matrices-and-linear-transforms" pass="4"}

::: {.quiz-q qid="mlt-product-shape" qtype="mc" answer="a"}
With $A$ of shape $2 \times 3$ and $B$ of shape $3 \times 2$, what shape is $AB$?

a) $2 \times 2$ — the outer dimensions, once the inner $3$s match
b) $3 \times 3$
c) $2 \times 3$
:::

::: {.quiz-q qid="mlt-not-commutative" qtype="mc" answer="b"}
Is matrix multiplication commutative — does $AB = BA$?

a) Yes, just like ordinary numbers
b) No — order changes the entries and often the shape, so keep the order the paper wrote
c) Only for square matrices of the same size
:::

::: {.quiz-q qid="mlt-singular-no-inverse" qtype="mc" answer="c"}
The matrix $\begin{bmatrix} 1 & 2 \\ 2 & 4 \end{bmatrix}$ has row $2$ equal to twice row
$1$. What about its inverse?

a) It is found by swapping the two diagonal entries
b) Every square matrix has an inverse, so it does
c) It has none — the matrix is singular (a redundant row), the matrix version of dividing by zero
:::

::: {.quiz-q qid="mlt-matrix-vector" qtype="numeric" answer="3" tolerance="0"}
For $A = \begin{bmatrix} 1 & 2 \\ 3 & 4 \end{bmatrix}$ and $v = (1, 1)$, the first entry
of $Av$ is row $1$ of $A$ dotted with $v$. What is that first entry?
:::

::: {.quiz-q qid="mlt-dot-product" qtype="numeric" answer="23" tolerance="0"}
(From vectors & summation.) Each entry of a matrix–vector product is a dot product — a
sum of products. For the row $(2, 3)$ and the vector $(4, 5)$, what is that dot product?
:::

:::

::::
