---
title: "The algebra moves a derivation is made of"
subtitle: "The handful of legal steps — expand, factor, collect, substitute, cancel, split a sum — that every 'and therefore…' line is quietly using"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live cells below re-state their own data. Base R is all
# we need, so nothing to install.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```


## Where we're going

When a paper writes a chain of equations — line, then line, then "and therefore" — each
line is not magic. It is *one small, legal algebra move* applied to the line above:
multiply out a bracket, collect the terms that look alike, swap in something you know,
cancel a factor from both sides. There are only a handful of these moves, you met most
of them in school, and every derivation is just them stacked up. This lesson names the
moves so that a multi-line derivation reads as **"one move per line, and here is the name
of each"** instead of a wall of symbols.

We stay with tiny hand-checkable numbers and base R, and — crucially — we check *every*
move by plugging in a number and confirming both sides agree. If the two forms of a line
are really the same algebra, they must give the same number for any input; that is the
whole trust-building trick.

**Builds on:** [vectors & summation](../vectors-and-summation/lesson.qmd) — we *extend*
its three Σ-rules (split a sum, pull a constant out, sum a constant) with two more sum
moves; and [reading math notation](../reading-math-notation/lesson.qmd) — for
implied multiplication and how far a symbol reaches (we lean on those, we don't re-teach
them).

## Warm-up: do you remember?

Before the new stuff, three quick reps — answers from memory first, then check.

**1.** In [vectors & summation](../vectors-and-summation/lesson.qmd) you learned that a
sum can be *split*. For `x <- c(2, 4, 6)` and `y <- c(1, 1, 2)`, what is
$\sum_i (x_i + y_i)$, and does it equal $\sum_i x_i + \sum_i y_i$?

```{webr}
#| autorun: false
# warmup1-starter — edit and Run to check
x <- c(2, 4, 6)
y <- c(1, 1, 2)
```

::: {.callout-note collapse="true" title="Answer"}
$\sum_i (x_i + y_i) = (2+1)+(4+1)+(6+2) = 3+5+8 = 16$, and
$\sum_i x_i + \sum_i y_i = 12 + 4 = 16$ — equal. In R, `sum(x + y)` and
`sum(x) + sum(y)` both give `16`. That's Rule 1, and we build on it below. Shaky?
Revisit [vectors & summation](../vectors-and-summation/lesson.qmd).
:::

**2.** Two symbols written side by side with nothing between them — like $3x$ or
$\beta_1 x$ — mean what operation? And is `2 + 3 * 4` equal to 20 or 14?

::: {.callout-note collapse="true" title="Answer"}
Side-by-side means **multiply** — $3x$ is $3 \times x$ (implied multiplication, from
[reading math notation](../reading-math-notation/lesson.qmd)). And `2 + 3 * 4` is
`14`, because multiplication binds before addition. Both facts matter the moment we
start expanding brackets below.
:::

**3.** For `x <- c(2, 4, 6)`, pull the constant out: what is `sum(2 * x)`, and does it
equal `2 * sum(x)`?

::: {.callout-note collapse="true" title="Answer"}
`sum(2 * x)` = $4 + 8 + 12 = 24$, and `2 * sum(x)` = $2 \times 12 = 24$ — equal.
That's Rule 2 (a constant factors out of a sum), taught in
[vectors & summation](../vectors-and-summation/lesson.qmd); factoring is one of the moves
we name properly today.
:::

## A tiny example first

Here is a two-line "derivation" so small you can hold it in your head. Someone claims:

> $(a + b)^2$ is the same as $a^2 + 2ab + b^2$.

Let's see it as *one move*. Take $a = 3$ and $b = 5$. The left side is
$(3 + 5)^2 = 8^2 = 64$. Now the right side, term by term: $a^2 = 9$, then $2ab = 2\times3\times5 = 30$,
then $b^2 = 25$; add them: $9 + 30 + 25 = 64$. **Same number.**

Why are they equal? Because "squared" means "times itself," so $(a+b)^2 = (a+b)(a+b)$, and
multiplying that out — every term in the first bracket times every term in the second —
gives $aa + ab + ba + bb = a^2 + ab + ab + b^2 = a^2 + 2ab + b^2$. That multiplying-out
step is the move called **expanding**. Read the other direction — going from
$a^2 + 2ab + b^2$ *back* to $(a+b)^2$ — and the same move is called **factoring**. One
move, two directions.

The same thing in R, so you can check every step:

```{r toy-example}
# The toy values
a <- 3
b <- 5

# Left side: square the sum
(a + b)^2                 # (3 + 5)^2 = 64

# Right side: the expanded form, term by term
a^2 + 2 * a * b + b^2     # 9 + 30 + 25 = 64  -> same number, so the move is legal
```

## Getting a feel for it

**Mental picture: algebra moves are like refactoring code.** You already trust this idea
from R. When you rewrite `sum(x) / length(x)` as `mean(x)`, you have not changed *what the
program computes* — same input, same output — you've only changed how it's *written*, to
make the next step easier. A derivation is exactly that: each line is a refactor of the
line above that leaves the value untouched but reshapes the symbols toward the form you
want. "Expand the bracket," "collect like terms," "cancel the common factor" are the
algebra equivalents of extracting a helper or inlining a variable. Nothing about the
underlying quantity changes; only its spelling does.

**Where the analogy breaks:** a code refactor can change *speed* without changing output;
an algebra move never changes speed (there's no runtime) — it only ever changes
readability and what the *next* move can grab onto. And a few "moves" are illegal in a way
refactors usually aren't: you may not divide both sides by something that could be zero,
just as you'd never divide by a variable in code without checking it's nonzero.

**Why these particular moves.** Each move exists to *isolate the thing you care about*. A
derivation almost always has a goal — "get $x$ by itself," "get everything over a common
form," "make a square appear so I can recognize it." Expanding breaks a clump apart so you
can regroup it; collecting like terms gathers the scattered pieces of the thing you want;
substituting swaps a big expression for a smaller name (or vice versa); cancelling strips
away a factor that's just riding along. They are the verbs of "rearrange until the answer
is staring at you."

**What a checked move feels like.** When we confirm $(a+b)^2$ and $a^2+2ab+b^2$ both hit
64, that agreement is not a coincidence to marvel at — it's the *definition* of a legal
move. Two algebraic forms are "equal" precisely when they return the same number for
**every** input, so a single agreeing test is strong evidence and a single *disagreeing*
test is proof the move was wrong. We'll lean on that constantly: derive, then plug in a
number.

## What the notation is going to say

Each move we're about to write is an equation with the *same expression* on both sides,
just spelled differently. "$(a+b)^2 = a^2 + 2ab + b^2$" says: "the square of a sum is the
first squared, plus twice the product, plus the second squared" — and it holds for any $a$
and $b$, which is why our $a=3, b=5$ check landing on 64 both ways is the whole proof in
miniature. The sum moves extend the ones you already read fluently: alongside
$\sum(x_i + y_i) = \sum x_i + \sum y_i$ (splitting), we'll say that *shifting the index* —
relabelling the counter — leaves a sum's value untouched. Every line below is "left form =
right form, and here's a number proving it."

## How to read the symbols

Every symbol used anywhere in this lesson gets a row. (Constants and indices you met in
the modules this one builds on are recalled here, not re-taught.)

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $a, b$ | "ay, bee" | stand-in numbers (any values you like) | $a = 3$, $b = 5$ |
| $x$ | "ex" | an unknown number we're solving for, or a variable input | $x = 4$ in the collect/substitute examples |
| $t$, $u$ | "tee", "you" | other stand-in variables (used like $x$ in the substitute examples) | $t$, $u$ appear in Check-yourself |
| $y$ | "why" | a stand-in variable that equals some expression in $x$ | $y = 2x + 1$ in the substitute example |
| $k$ | "kay" | a constant offset inside a square, $(x+k)^2$ | $k = 3$ in the completing-the-square example |
| $(a+b)^2$ | "the quantity a plus b, squared" | the sum $a+b$ multiplied by itself; the exponent reaches over the whole bracket | $(3+5)^2 = 64$ |
| $a^2$ | "ay squared" | $a \times a$ (superscript = power — [reading math notation](../reading-math-notation/lesson.qmd)) | $3^2 = 9$ |
| $2ab$ | "two ay bee" | $2 \times a \times b$ (side-by-side = multiply) | $2\times3\times5 = 30$ |
| $c$ | "see" | a constant — a fixed number that doesn't change with the index | $6$ in the cancel example |
| $x_i$ | "ex sub eye" | the i-th value of the list x (in R: `x[i]`) | $x_1=2, x_2=4, x_3=6$ |
| $y_i$ | "why sub eye" | the i-th value of a second list y (in R: `y[i]`) | $y_1=1, y_2=1, y_3=2$ |
| $n$ | "en" | how many items in the list (in R: `length(x)`) | 3 |
| $i$, $j$ | "eye", "jay" | the counter that walks a sum; renaming it changes nothing | $i$ runs 1..3 |
| $\sum_{i=1}^{n}$ | "the sum from i equals 1 to n" | add the term to its right, once per $i$ (in R: `sum(...)`) — from [vectors & summation](../vectors-and-summation/lesson.qmd) | $x_1+x_2+x_3$ |
| $\beta_1$ | "beta sub one" | a named parameter (Greek letter; subscript = index) — used here only as an example of implied multiplication; taught in full in [reading math notation](../reading-math-notation/lesson.qmd) | — |
| $\Longrightarrow$ | "double right arrow" or "implies" | the left side is true, so the right side follows — a step-follows-from-step marker | used in Move 4's equation chain |

## The formal version

We take the seven moves in turn. For each: the move as an equation, a one-line reading, and
a numeric check. Treat each check as the proof — if the numbers agreed, the move is legal.

### Move 1 — expand and factor (two directions of one move)

$$
(a + b)^2 = a^2 + 2ab + b^2
$$

Reading it aloud: "the quantity a plus b, all squared, equals a-squared plus two-a-b plus
b-squared" — everyday words: "square of a sum = first squared + twice the product + second
squared." Going **left to right** is *expanding*; going **right to left** is *factoring*
(recognising the pattern and packing it back into a bracket).

- The single move is: $(a+b)^2 = (a+b)(a+b)$, then multiply every term by every term
  ("expand the brackets") and combine the two $ab$ copies ("collect like terms",
  Move 2 below).

Mapping onto the toy numbers with $a=3, b=5$: left is $8^2 = 64$; right is $9 + 30 + 25 = 64$.

```{r move1-check}
a <- 3; b <- 5
c(squared = (a + b)^2,                 # 64
  expanded = a^2 + 2 * a * b + b^2)    # 64 — expand and factor are the same value
```

### Move 2 — collect like terms (regroup to isolate)

"Like terms" are terms with the *same variable part* — $3x$ and $2x$ are alike (both are
"some number of $x$"), but $3x$ and $3x^2$ are not. Collecting means adding the coefficients
of like terms so the scattered pieces become one:

$$
3x + 5 + 2x - 1 = 5x + 4
$$

Reading it aloud: "three-x plus five plus two-x minus one equals five-x plus four" —
everyday words: "gather the x-terms ($3x + 2x = 5x$) and gather the plain numbers
($5 - 1 = 4$)."

- The move is: regroup the sum so like terms sit together, then add their coefficients —
  legal because addition can be reordered and regrouped freely.

Check with $x = 4$: left is $12 + 5 + 8 - 1 = 24$; right is $20 + 4 = 24$.

```{r move2-check}
x <- 4
c(scattered = 3 * x + 5 + 2 * x - 1,   # 24
  collected = 5 * x + 4)               # 24 — regrouping didn't change the value
```

### Move 3 — substitute one expression into another

Substitution swaps a symbol for something you know equals it. Say you know $y = 2x + 1$,
and you meet $3y - 4$. Replace **every** $y$ with the thing it equals — in a bracket, so
the operations still apply to the whole of it:

$$
\begin{aligned}
3y - 4 &= 3(2x + 1) - 4 \\
       &= 6x + 3 - 4 \\
       &= 6x - 1
\end{aligned}
$$

- Line 1: substitute $y = 2x + 1$, wrapped in a bracket so the $3\times$ hits all of it.
- Line 2: expand the bracket — multiply the 3 through (Move 1's expanding, the simple case).
- Line 3: collect like terms — $3 - 4 = -1$ (Move 2).

Reading the result aloud: "three-y minus four equals six-x minus one." Check with $x = 4$:
then $y = 2\times4 + 1 = 9$, so $3y - 4 = 27 - 4 = 23$; and $6x - 1 = 24 - 1 = 23$.

```{r move3-check}
x <- 4
y <- 2 * x + 1                 # the thing we substitute in: y = 9
c(before = 3 * y - 4,          # 23, using y directly
  after  = 6 * x - 1)          # 23 — substitution + expand + collect, same value
```

The clean-substitution habit: **always parenthesise what you drop in.** Writing $3 \cdot 2x + 1$
without the bracket would wrongly leave the $+1$ outside the multiplication.

### Move 4 — divide both sides / cancel a common factor (and when you can't)

If two things are equal, you may do the *same* operation to both and they stay equal. To
solve $6x = 30$, divide both sides by 6:

$$
6x = 30 \quad\Longrightarrow\quad \frac{6x}{6} = \frac{30}{6} \quad\Longrightarrow\quad x = 5
$$

- The move is: divide both sides by the same nonzero number; on the left, the 6 in the
  numerator and denominator **cancel** (they're a common factor), leaving $x$.

Cancelling a common factor from a fraction is the same idea:
$\dfrac{6x}{6} = x$ because the 6 above and below divide out.

::: {.callout-important title="The one rule: never divide by something that could be zero"}
Cancelling a factor is only legal when that factor is **not zero**. You cannot divide both
sides by $x$ unless you know $x \neq 0$ — if $x$ might be 0, dividing by it is undefined
and can silently destroy solutions. This is the algebra face of "you can't divide by
zero," and it is the single most common illegal move in a botched derivation.
:::

Check: $6 \times 5 = 30$ ✓, and the common-factor cancel:

```{r move4-check}
# Solve 6x = 30 by dividing both sides by 6
30 / 6                          # 5  -> x = 5
6 * 5                           # 30 -> plug back in: the solution checks out

# Cancelling a common factor: (6 * x) / 6 is just x, for any nonzero divisor
x <- 5
(6 * x) / 6                     # 5 -> the 6s cancel
```

### Move 5 — sum manipulation beyond the basic three (split, and reindex)

You already have three Σ-rules from
[vectors & summation](../vectors-and-summation/lesson.qmd): split a sum, pull out a
constant, sum a constant. Here are two more moves derivations lean on.

**Splitting extended.** Rule 1 splits $\sum(x_i + y_i)$ into $\sum x_i + \sum y_i$. The
same move handles a term-by-term expression: a sum of a two-part term splits into two sums,
which you can then simplify each of with the constant rules. (We just *use* Rule 1 here, we
don't re-derive it.)

**Reindexing (shifting the index).** The counter's *name and starting number* are
bookkeeping — slide them and the sum's value is unchanged, as long as the terms picked out
stay the same. Adding all of $x_1, x_2, x_3$ is the same whether you call the counter $i$
running 1→3 and take $x_i$, or call it $j$ running 2→4 and take $x_{j-1}$:

$$
\sum_{i=1}^{3} x_i = \sum_{j=2}^{4} x_{j-1}
$$

- The move is: rename the counter and shift its bounds, compensating inside so each term is
  the same one as before — legal because a sum doesn't care what you *call* the counter,
  only which terms it collects.

Reading it aloud: "the sum from i equals 1 to 3 of x-sub-i equals the sum from j equals 2
to 4 of x-sub-(j minus 1)." When $j = 2$ you get $x_1$; when $j = 4$ you get $x_3$ — the
same three terms. Check with $x = (2, 4, 6)$: both are $2 + 4 + 6 = 12$.

```{r move5-check}
x <- c(2, 4, 6)

# Original indexing: i from 1 to 3, term x[i]
sum(x)                                          # 12

# Reindexed: j from 2 to 4, term x[j - 1] -- same three terms, same total
sum(sapply(2:4, function(j) x[j - 1]))          # 12

# Split still works too (Rule 1 from vectors-and-summation):
y <- c(1, 1, 2)
c(split_lhs = sum(x + y), split_rhs = sum(x) + sum(y))   # 16 and 16
```

Reindexing feels pointless on its own — its payoff is *lining two sums up* so their terms
match and can be combined; you'll see it whenever a derivation needs $\sum_{i=1}^{n}$ and
$\sum_{i=0}^{n-1}$ to have the same shape.

### Move 6 — completing the square (recognize it by name) {#completing-the-square}

**Optional to *do*; essential to *recognize*.** This move rewrites $x^2 + (\text{something})x + (\text{something})$
as a perfect square (like Move 1's $(x+k)^2$) plus a leftover number. You don't need to
drill the mechanics here — you need to know the name, so that when a derivation says
"completing the square" you know it means "force a $(x+k)^2$ to appear." (It shows up
wherever a derivation needs to read the minimum off a squared term; naming it now is
enough.)

$$
x^2 + 6x + 5 = (x + 3)^2 - 4
$$

Reading it aloud: "x-squared plus six-x plus five equals the quantity x plus three,
squared, minus four." The name comes from *making a complete square appear*: half of the
6 is 3, and $(x+3)^2$ expands (Move 1) to $x^2 + 6x + 9$ — that's 4 too big, so we subtract
4 to compensate. What buys you: a lone $(x+k)^2$ is easy to reason about (its smallest value
is 0, at $x = -k$), which is why the move shows up when someone wants to find a minimum.

Check with $x = 4$: left is $16 + 24 + 5 = 45$; right is $(4+3)^2 - 4 = 49 - 4 = 45$.

```{r move6-check}
x <- 4
c(before = x^2 + 6 * x + 5,        # 45
  completed = (x + 3)^2 - 4)       # 45 — same value, now a square plus a leftover
```

### Move 7 — reading a multi-line derivation: one move per line, name the move

The payoff. A derivation is a stack of the moves above, **one per line**, each turning the
line above into the line below. The skill is to read it that way: for every line, ask "what
single move got me here from the previous line?" Watch it on a short chain that simplifies
$(x + 2)^2 - 3x$:

$$
\begin{aligned}
(x + 2)^2 - 3x &= (x^2 + 4x + 4) - 3x \\
               &= x^2 + 4x - 3x + 4 \\
               &= x^2 + x + 4
\end{aligned}
$$

- Line 1: **expand** the square (Move 1): $(x+2)^2 = x^2 + 4x + 4$.
- Line 2: **regroup** so like terms are adjacent (Move 2's setup) — just reordering.
- Line 3: **collect like terms** (Move 2): $4x - 3x = x$.

Each line does exactly one nameable thing. Sanity-check the whole chain with $x = 4$: the
start $(4+2)^2 - 12 = 36 - 12 = 24$, and the end $16 + 4 + 4 = 24$ — the derivation
preserved the value at every step, as a legal derivation must.

```{r move7-check}
x <- 4
c(start = (x + 2)^2 - 3 * x,       # 24
  end   = x^2 + x + 4)             # 24 — three moves, value unchanged throughout
```

## Explore it

```{webr}
#| autorun: true
# explore-1 — LIVE: expanding (a+b)^2 is a real algebra identity, not just true for 3 & 5.
# TRY IT: change a and b to ANY numbers (negatives, decimals) and Run.
# PREDICT: will the two forms ever disagree? Commit yes/no, then run.
a <- 3
b <- 5

squared  <- (a + b)^2               # square-the-sum form
expanded <- a^2 + 2 * a * b + b^2   # the expanded form

c(squared = squared, expanded = expanded)   # identical for ANY a, b
```

```{webr}
#| autorun: true
# explore-2 — LIVE: collect like terms, then check the collected form matches the scattered
# one for a whole grid of x values at once (a stronger test than a single number).
# PREDICT: will "all TRUE" print? Commit, then run.
xs <- seq(-5, 5, by = 0.5)          # many values of x

scattered <- 3 * xs + 5 + 2 * xs - 1
collected <- 5 * xs + 4

all(scattered == collected)         # TRUE everywhere -> the collect move is legal
```

```{webr}
#| autorun: false
# explore-3 — LIVE: the cancel trap. Cancelling a factor is illegal if it might be zero.
# PREDICT: solving 6x = 30 gives x = 5. Now: what happens if you try to "cancel x" from
# the equation  x * x = x  ? (Hint: x = 0 is a solution you'd throw away.) Run to see.
30 / 6                               # 5 -> legal: dividing both sides by nonzero 6

# x * x = x has TWO solutions: x = 0 and x = 1.
# Cancelling an x (dividing both sides by x) gives only x = 1 -- and silently loses x = 0,
# because dividing by x is illegal when x could be 0.
candidates <- c(0, 1)
data.frame(x = candidates, holds = candidates * candidates == candidates)  # both TRUE
```

## Common traps

**Trap 1 — "$(a+b)^2 = a^2 + b^2$."** Hugely tempting, because squaring *looks* like it
should distribute across the plus the way multiplication distributes. It does not: you drop
the middle term $2ab$. With $a = 3, b = 4$: the correct $(a+b)^2 = 7^2 = 49$, but the wrong
$a^2 + b^2 = 9 + 16 = 25$. Off by $2ab = 24$ — not a rounding slip, a whole missing term.
The square of a sum always carries that cross term.

```{r trap1}
a <- 3; b <- 4
c(correct = (a + b)^2,     # 49
  wrong   = a^2 + b^2,     # 25 -- missing the 2ab = 24 cross term
  cross   = 2 * a * b)     # 24
```

**Trap 2 — "cancel the $x$ from both sides."** Given $x^2 = x$, it's tempting to divide
both sides by $x$ and conclude $x = 1$. But that quietly assumes $x \neq 0$, and $x = 0$ is
*also* a solution ($0^2 = 0$). Dividing by $x$ threw it away. The safe move is to bring
everything to one side and factor: $x^2 - x = 0$, so $x(x - 1) = 0$, giving $x = 0$ *or*
$x = 1$. **Never cancel a factor that could be zero** — it loses solutions.

## Check yourself

**1.** Expand $(x + 4)^2$ into three terms. Then check your expansion at $x = 2$ by
computing both forms.

```{webr}
#| autorun: false
# check1-starter — edit and Run to check your expansion at x = 2
x <- 2
```

::: {.callout-note collapse="true" title="Answer"}
$(x+4)^2 = x^2 + 2\cdot4\cdot x + 4^2 = x^2 + 8x + 16$ (Move 1, expanding — twice the
product is $2\times4\times x = 8x$). At $x = 2$: the square form is $(2+4)^2 = 36$; the
expanded form is $4 + 16 + 16 = 36$. Same number, so the expansion is right. In R,
`(x + 4)^2` and `x^2 + 8*x + 16` both give `36`.
:::

**2.** You know $u = 3t - 1$. Substitute it into $2u + 5$ and simplify to a single
expression in $t$. What move do you use right after substituting?

::: {.callout-note collapse="true" title="Answer"}
Substitute (parenthesised!): $2u + 5 = 2(3t - 1) + 5$. Then **expand** the bracket
(multiply the 2 through): $6t - 2 + 5$. Then **collect like terms** ($-2 + 5 = 3$):
$6t + 3$. The move right after substituting is *expanding the bracket*. Check at $t = 2$:
$u = 5$, so $2u + 5 = 15$; and $6t + 3 = 15$. Match.
:::

**3.** In your own words: explain to a colleague, in two sentences, what it means for one
line of a derivation to be a "legal move," and how you'd check that a move is legal.

::: {.callout-note collapse="true" title="A model answer"}
A legal move rewrites the line above into a new form that has the *exact same value for
every possible input* — it changes only how the expression is spelled, never what it
computes, the way refactoring `sum(x)/length(x)` into `mean(x)` leaves the answer alone.
To check one, plug a concrete number into both the before and after forms: if they ever
disagree, the move was illegal; if they always agree, it's a genuine rearrangement.
:::

## What you can now do

- **Expand and factor** $(a+b)^2$ — recognize it as one move read in two directions, and
  spot the $2ab$ cross term that the "$a^2 + b^2$" trap drops.
- **Collect like terms and substitute cleanly** — gather same-variable terms into one, and
  drop an expression in for a symbol *in a bracket* before expanding.
- **Divide both sides / cancel a common factor** — and refuse to cancel anything that
  could be zero, because that loses solutions.
- **Reshape a sum** beyond the basic three rules — split a term-by-term sum, and reindex
  (shift the counter) to line two sums up, knowing the value is unchanged.
- **Name "completing the square" on sight** and read a multi-line derivation as one named
  move per line, checking the value is preserved at each step.

If any of these feels shaky, revisit [The formal version](#the-formal-version) and re-run
the checks — every derivation you'll meet is just these moves stacked, and reading it means
naming each one.

## Where this goes next

These moves are the *grammar* of every "and therefore…" line: whenever a lesson later
expands a squared residual, splits a sum, or cancels a factor to isolate a coefficient, it
is using exactly what you just named. Keep this page open as a reference the first few times
a derivation gets long — soon you'll name the moves without looking. Ready for more reps? Do
the [practice problems](practice.qmd).
