---
title: "Practice: The algebra moves a derivation is made of"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only (a {webr} starter does not see it — each starter re-states its
# own data). Base R is all we need, so nothing to install.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Do them on paper or in an R session; only
then open the answer.

## Problem 1 — the toy example, new numbers

Expand $(a + b)^2$ into its three terms for $a = 2$ and $b = 4$, by hand. Then check
both forms in R and confirm they agree.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
a <- 2
b <- 4
```

::: {.callout-note collapse="true" title="Answer"}
The move is *expanding* (Move 1): $(a+b)^2 = a^2 + 2ab + b^2$. Term by term with
$a = 2, b = 4$: $a^2 = 4$, $2ab = 2\times2\times4 = 16$, $b^2 = 16$; total $4 + 16 + 16 = 36$.
And the square form: $(2 + 4)^2 = 6^2 = 36$. Same number, so the expansion is correct.

```{r p1-answer}
a <- 2; b <- 4
c(squared  = (a + b)^2,                 # 36
  expanded = a^2 + 2 * a * b + b^2)     # 36 — expand and factor are the same value
```
:::

## Problem 2 — fill in the missing step (collect like terms)

Here is a two-line simplification of $4x + 7 + 3x - 2$ — but the **core move is blanked**.
Fill in ONLY the blanked line: what has to happen to turn the scattered form into the
collected one?

> Start: $4x + 7 + 3x - 2$.
> **`<your step here>`** — gather the like terms.
> Result: $7x + 5$.

```{webr}
#| autorun: false
# p2-starter — edit and Run to check your collected form at x = 3
x <- 3
```

::: {.callout-note collapse="true" title="Answer"}
The blanked step is **collecting like terms** (Move 2): add the coefficients of the
$x$-terms, $4x + 3x = 7x$, and add the plain numbers, $7 - 2 = 5$, giving $7x + 5$. It's
legal because addition can be reordered and regrouped freely. Check at $x = 3$: the
scattered form is $12 + 7 + 9 - 2 = 26$; the collected form is $21 + 5 = 26$.

```{r p2-answer}
x <- 3
c(scattered = 4 * x + 7 + 3 * x - 2,    # 26
  collected = 7 * x + 5)                # 26 — same value, tidier form
```
:::

## Problem 3 — substitute, then simplify

You know $u = 3t - 2$. Substitute it into $2u + 5$ and simplify to a single expression in
$t$. Do it by hand (naming each move), then check at $t = 5$ in R.

```{webr}
#| autorun: false
# p3-starter — your job: build u from t, then compare 2u+5 to your simplified form
t <- 5
```

::: {.callout-note collapse="true" title="Answer"}
- **Substitute** (in a bracket): $2u + 5 = 2(3t - 2) + 5$.
- **Expand** the bracket (Move 1, simple case — multiply the 2 through): $6t - 4 + 5$.
- **Collect like terms** (Move 2): $-4 + 5 = 1$, giving $6t + 1$.

Check at $t = 5$: $u = 3\times5 - 2 = 13$, so $2u + 5 = 26 + 5 = 31$; and $6t + 1 = 30 + 1 = 31$.

```{r p3-answer}
t <- 5
u <- 3 * t - 2                 # the substituted expression: u = 13
c(before = 2 * u + 5,          # 31
  after  = 6 * t + 1)          # 31 — substitute, expand, collect: value unchanged
```
:::

## Problem 4 — spot the error

A colleague expanded a square like this. It contains **exactly one mistake** — find it,
say why it's wrong with the numbers, and give the corrected expansion.

> "$(a + b)^2 = a^2 + b^2$. So for $a = 3, b = 4$ it's $9 + 16 = 25$."

```{webr}
#| autorun: false
# p4-starter — copy this to see the correct and wrong forms side by side
a <- 3
b <- 4
```

::: {.callout-note collapse="true" title="Answer"}
The mistake is dropping the cross term: $(a+b)^2$ is **not** $a^2 + b^2$; it's
$a^2 + 2ab + b^2$. The missing $2ab$ is the whole error. With $a = 3, b = 4$ the correct
value is $(3+4)^2 = 7^2 = 49$, while the colleague's $a^2 + b^2 = 25$ — short by exactly
$2ab = 2\times3\times4 = 24$. Squaring does not distribute across a plus; the square of a
sum always carries the cross term.

```{r p4-answer}
a <- 3; b <- 4
c(correct = (a + b)^2,     # 49
  wrong   = a^2 + b^2,     # 25 -- missing 2ab
  cross   = 2 * a * b)     # 24 -- the term that was dropped
```
:::

## Problem 5 — mixing it with a sum *(uses [vectors-and-summation](../vectors-and-summation/lesson.qmd)'s sum rules)*

Simplify $\sum_{i=1}^{n} (3x_i + 2)$ into a form using $\sum x_i$ and $n$, then evaluate it
for $x = (1, 2, 3)$. **PREDICT** the number first, then run.

```{webr}
#| autorun: false
# p5-starter — edit and Run to solve, right here
x <- c(1, 2, 3)
```

::: {.callout-note collapse="true" title="Answer"}
Chain the sum moves from [vectors & summation](../vectors-and-summation/lesson.qmd):

- **Split** the sum (Rule 1): $\sum (3x_i + 2) = \sum 3x_i + \sum 2$.
- **Pull the constant out** (Rule 2): $\sum 3x_i = 3\sum x_i$.
- **Sum a constant** (Rule 3): $\sum_{i=1}^{n} 2 = 2n$.

So $\sum (3x_i + 2) = 3\sum x_i + 2n$. For $x = (1,2,3)$: $\sum x_i = 6$ and $n = 3$, giving
$3\times6 + 2\times3 = 18 + 6 = 24$.

```{r p5-answer}
x <- c(1, 2, 3)
n <- length(x)
c(direct    = sum(3 * x + 2),         # 24
  simplified = 3 * sum(x) + 2 * n)    # 24 — split, factor, sum-the-constant
```
:::

## Problem 6 — a gentle cliffhanger (completing the square)

Someone rewrites $x^2 + 4x + 1$ as $(x + 2)^2 - 3$ and calls the move "completing the
square." You met the *name* in the lesson but didn't drill the mechanics. **Predict:**
without doing the full derivation, can you confirm the two forms are equal by plugging in
$x = 5$? Then run the starter and say what "completing the square" bought here.

```{webr}
#| autorun: false
# p6-starter — check the two forms agree at x = 5 (and try other x values too)
x <- 5
```

::: {.callout-note collapse="true" title="Answer"}
At $x = 5$: the original is $25 + 20 + 1 = 46$; the completed-square form is
$(5+2)^2 - 3 = 49 - 3 = 46$ — equal, so the rewrite is legal. You can verify the identity
without deriving it, exactly the checking habit from the lesson. *Why do it:* the completed
form $(x+2)^2 - 3$ makes the smallest possible value obvious — $(x+2)^2$ is never negative,
so the whole thing bottoms out at $-3$, reached when $x = -2$. That "read the minimum off a
square" payoff is why the move appears in derivations that hunt for a minimum — the full
mechanics wait for a lesson that needs them.

```{r p6-answer}
x <- 5
c(original  = x^2 + 4 * x + 1,     # 46
  completed = (x + 2)^2 - 3)       # 46 — same value; the square makes the minimum readable
```
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/algebra-moves-for-derivations" pass="4"}

::: {.quiz-q qid="amd-cross-term" qtype="mc" answer="a"}
A student expands $(a + b)^2$ as $a^2 + b^2$. For $a = 3, b = 5$, what did they drop?

a) The cross term $2ab = 30$; the correct value is $64$, not $34$
b) Nothing — $(a + b)^2 = a^2 + b^2$
c) They should have written $a^2 b^2$
:::

::: {.quiz-q qid="amd-cancel-zero" qtype="mc" answer="b"}
To solve $x^2 = x$, a student divides both sides by $x$ and concludes $x = 1$. What went
wrong?

a) Nothing — $x = 1$ is the only solution
b) Dividing by $x$ assumes $x \neq 0$ and drops the solution $x = 0$; factor $x(x-1) = 0$ instead
c) You can never divide both sides of an equation by a variable
:::

::: {.quiz-q qid="amd-reindex" qtype="numeric" answer="12" tolerance="0"}
For $x = (2, 4, 6)$, evaluate $\sum_{j=2}^{4} x_{j-1}$. (Hint: reindexing shifts the
counter but picks out the same terms.)
:::

::: {.quiz-q qid="amd-two-ab" qtype="numeric" answer="20" tolerance="0"}
Expand and evaluate $(a + b)^2 - a^2 - b^2$ for $a = 2, b = 5$. (It is the cross term the
"$a^2 + b^2$" trap drops.)
:::

::: {.quiz-q qid="amd-complete-square" qtype="mc" answer="c"}
Completing the square rewrites $x^2 + 6x$ as $(x + 3)^2 + c$. What is $c$?

a) $+9$
b) $0$
c) $-9$
:::

::: {.quiz-q qid="amd-argmin" qtype="mc" answer="a"}
(From reading math notation.) What does $\arg\min_{\beta} f(\beta)$ denote?

a) The input $\beta$ that makes $f$ smallest
b) The smallest value that $f$ attains
c) The slope of $f$ at its minimum
:::

:::

::::
