---
title: "Practice: Lambda for stability"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
set.seed(42)
```

Work these in order — they get harder. Do them on paper or in an R session; only then
open the answer.

## Problem 1 — explode, then stabilize, new numbers

Two indicators correlated 0.99, with marginals $b = (0.6, 0.55)$ and
$R = \begin{pmatrix}1 & 0.99\\0.99 & 1\end{pmatrix}$. Compute the unregularized
reconstruction $R^{-1}b$, then the ridge version $(R + 0.1\,I)^{-1}b$.

```{webr}
#| autorun: false
# p1-starter
R <- matrix(c(1, 0.99, 0.99, 1), 2); b <- c(0.6, 0.55)
```

::: {.callout-note collapse="true" title="Answer"}
$\det R = 1 - 0.99^2 = 0.0199$, tiny, so $R^{-1}b \approx (2.79, -2.21)$ — huge, opposite
signs, from marginals only 0.05 apart. With $\lambda = 0.1$, $\det(R + 0.1I) = 1.21 - 0.9801 = 0.2299$,
and $(R + 0.1I)^{-1}b \approx (0.50, 0.05)$ — small, same-signed, sensible.

```{r}
R <- matrix(c(1, 0.99, 0.99, 1), 2); b <- c(0.6, 0.55)
rbind(unregularized = solve(R, b), ridge_0.1 = solve(R + 0.1 * diag(2), b))
```
:::

## Problem 2 — fill in the missing step

$R$ has eigenvalues $\{1.99, 0.01\}$ (the small one makes it nearly singular). You add
$\lambda = 0.05$ down the diagonal.

- Step 1: adding $\lambda I$ adds $\lambda$ to *every* eigenvalue.
- Step 2: **your step here** — what are the new eigenvalues, and is the matrix now safely
  invertible?
- Step 3: since the smallest eigenvalue is now well above 0, the inverse is well-behaved.

::: {.callout-note collapse="true" title="Answer"}
Step 2: the new eigenvalues are $\{1.99 + 0.05,\ 0.01 + 0.05\} = \{2.04,\ 0.06\}$. The smallest
went from 0.01 to 0.06 — comfortably positive — so $R + \lambda I$ is safely invertible. (Any
$\lambda > 0$ lifts the smallest eigenvalue to at least $\lambda$.)
:::

## Problem 3 — a lambda sweep, in R

For the near-collinear pair below, reconstruct at several $\lambda$ values and watch the largest
coefficient shrink from huge to stable.

```{webr}
#| autorun: false
# p3-starter — reconstruct at each lambda and report the largest |coefficient|
# PREDICT: at lambda = 0 it will be large; where does it settle?
R <- matrix(c(1, 0.995, 0.995, 1), 2); b <- c(0.7, 0.6)
lambdas <- c(0, 0.01, 0.05, 0.2)
```

::: {.callout-note collapse="true" title="Answer"}
At $\lambda = 0$ the coefficients are large and opposite-signed; by $\lambda = 0.05$–$0.2$ they
are small and stable.

```{r}
R <- matrix(c(1, 0.995, 0.995, 1), 2); b <- c(0.7, 0.6)
t(sapply(c(0, 0.01, 0.05, 0.2), function(l) {
  beta <- solve(R + l * diag(2), b); c(lambda = l, beta1 = beta[1], beta2 = beta[2])
}))
```
The unregularized row is wild; adding a small $\lambda$ tames it.
:::

## Problem 4 — spot the error

A colleague says: *"To be safe against collinearity I always use a very large $\lambda$, like
100 — bigger $\lambda$ means more stable, so more is better."* This contains **exactly one
mistake** — find it.

::: {.callout-note collapse="true" title="Answer"}
The mistake: "bigger $\lambda$ is always better." A very large $\lambda$ over-shrinks — it drags
every coefficient toward 0, erasing the real effects (at $\lambda = 100$ the fit is essentially
"predict the mean"). Stability and signal trade off; the goal is the *smallest* $\lambda$ that
makes the matrix well-conditioned, not the largest.
:::

## Problem 5 — mixing it with the ridge foundation *(uses $(X^\top X + \lambda I)$ invertibility)*

The ridge foundation showed $(X^\top X + \lambda I)$ is invertible for any $\lambda > 0$. Explain,
in a sentence, why the *same* guarantee applies to $R + \lambda I$ here, even when two indicators
are perfectly correlated ($\rho = 1$, so $R$ is exactly singular).

::: {.callout-note collapse="true" title="Answer"}
$R$ is symmetric with non-negative eigenvalues (it's a correlation matrix), and adding $\lambda I$
lifts every eigenvalue by $\lambda$; so even if $R$ has an eigenvalue of exactly 0 (perfect
collinearity), $R + \lambda I$ has smallest eigenvalue $\lambda > 0$ and is therefore invertible.
It's the identical diagonal-loading argument, applied to $R$ instead of $X^\top X$.
:::

## Problem 6 — transfer: which lambda?

A $\lambda$ sweep shows the selected set and coefficients are wild at $\lambda = 0$, stabilize by
$\lambda \approx 0.05$, and by $\lambda = 1$ every coefficient has shrunk close to 0. You must
pick one $\lambda$. Which region would you choose and why — and what standard technique (named,
not built here) would automate the choice? *(It's outside this course.)*

::: {.callout-note collapse="true" title="Answer"}
Pick a $\lambda$ in the *stable-but-not-over-shrunk* region — around 0.05 here: large enough that
the reconstruction has settled, small enough that real effects aren't crushed. The standard way
to choose it automatically is **cross-validation** — try a grid of $\lambda$ values, measure
out-of-sample prediction error for each, and keep the $\lambda$ that minimizes it. (This course
fixes or sweeps $\lambda$ by hand; CV-based selection is a natural next step.)
:::
