---
title: "Exponentials and logs: undoing each other"
subtitle: "The number e, the natural log, and the handful of rules that make them tame"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live cells below re-state their own data. Base R only.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

::: {.callout-tip title="This lesson runs in your browser" collapse="true"}
The code cells below with a **Run** button are live — powered by WebR (R compiled to
WebAssembly), they run right here on your phone or laptop, no install needed. Edit any
of them and re-run. The first run downloads a small R engine, so give it a few seconds.
Cells without a Run button are ordinary listings whose output was computed when the
page was built.
:::

## Where we're going

Two functions show up in almost every statistical formula that involves growth,
decay, probabilities, or likelihoods: the **exponential** $e^x$ and its exact
opposite, the **natural logarithm** $\log$. They look intimidating written down, but
they are just a matched pair of buttons on a calculator that undo each other — like
"square" and "square root," or "add 3" and "subtract 3." This module makes them
comfortable: what $e$ is, what $\log$ gives you back, and the five little rules that
turn ugly products into tidy sums.

Methods that lean on them: anything with the word *likelihood*, logistic regression
and other models built on probabilities, compound growth and half-lives, and the
information/entropy measures in machine learning.

**Builds on:** just the baseline (you can already raise a number to a power and take
a square root, and you know that some operations undo others).

## Warm-up: do you remember?

This is a foundation module, so there's nothing earlier to recall — just three
baseline reflexes the idea leans on. Produce each answer before opening it.

**1.** What is $2^3$? And what single operation *undoes* "raise 2 to a power" — that
is, if $2^{?} = 8$, how would you describe finding that `?`?

::: {.callout-note collapse="true" title="Answer"}
$2^3 = 2 \times 2 \times 2 = 8$. Undoing it means asking "2 to *what* power gives 8?" —
the answer is 3. That question is exactly what a **logarithm** answers, and it's the
whole idea of this lesson.
:::

**2.** Square roots undo squares: $\sqrt{9} = 3$ because $3^2 = 9$. In the same spirit,
what number, squared, gives 25? What operation did you just perform?

::: {.callout-note collapse="true" title="Answer"}
$5$, because $5^2 = 25$ — you took a square root. The point to hold onto: functions
come in **undo-pairs**. Exponentials and logs are one such pair.
:::

**3.** What is $\tfrac{1}{2^3}$ as a decimal? (No calculator — just $2^3 = 8$.)

::: {.callout-note collapse="true" title="Answer"}
$\tfrac{1}{8} = 0.125$. A *negative* exponent will turn out to mean exactly this — "one
over" — so $2^{-3} = \tfrac{1}{8}$. Worth having fresh, because $e^{-z}$ is everywhere
in the courses that use this module.
:::

## A tiny example first

Meet the number **$e \approx 2.718$**. It's just a fixed number, like $\pi \approx
3.14159$ — nothing mystical, a specific constant that happens to make growth math come
out clean. "Raising $e$ to a power" is the **exponential** function, written $e^x$ or
`exp(x)`.

Let's compute a few by hand-ish reasoning and then check them:

- $e^0 = 1$. (Anything to the power 0 is 1 — no exceptions for $e$.)
- $e^1 = e \approx 2.718$. (One copy of $e$.)
- $e^2 = e \times e \approx 2.718 \times 2.718 \approx 7.389$.
- $e^{-1} = \tfrac{1}{e} \approx \tfrac{1}{2.718} \approx 0.368$. (Negative power = "one
  over," from the warm-up.)

Now the **natural log**, written $\log$ here (base $e$), is the undo button: $\log(y)$
answers "$e$ to *what* power gives $y$?" So:

- $\log(1) = 0$, because $e^0 = 1$.
- $\log(e) = 1$, because $e^1 = e$.
- $\log(7.389) \approx 2$, because $e^2 \approx 7.389$.

Feed a number through $e^x$ and then through $\log$, and you get the original number
straight back. That "there and back again" is the single most useful fact here.

The same thing in R, so you can check every step:

```{r toy-example}
# e is exp(1); "raise e to a power" is exp()
exp(0)          # e^0 = 1
exp(1)          # e^1 = e ~ 2.718
exp(2)          # e^2 ~ 7.389
exp(-1)         # e^-1 = 1/e ~ 0.368

# log() with no base is the NATURAL log (base e) in R -- the undo button
log(1)          # 0, since e^0 = 1
log(exp(2))     # exactly 2: log undoes exp
exp(log(5))     # exactly 5: exp undoes log (the round trip, other direction)
```

## Getting a feel for it

**A mental picture: a stretchy ruler and its rewind button.** Think of $e^x$ as a
machine that takes a distance $x$ and *stretches* it into a scale factor. At $x = 0$
there's no stretch, so you get $1$. Each step to the right multiplies by another factor
of $e$ (about 2.7×), so the output rockets upward — that's *exponential growth*. Step
*left* into negative $x$ and you divide instead of multiply, so the output shrinks
toward zero but never quite reaches it. The **log** is the rewind button: hand it a
scale factor and it tells you how far along the ruler you'd walked to get there.

Where the picture bends: a real ruler is evenly spaced, but the *outputs* of $e^x$ are
not — they get dramatically farther apart as $x$ grows and bunch up toward zero as $x$
goes negative. That uneven spacing is the whole personality of exponential growth.

**Why $e$ and not just $10$?** You *can* take logs in any base (base-10 logs power the
Richter scale, base-2 logs power bits). Base $e$ is the "natural" choice because it's
the one where the growth rate equals the current size — the slope of $e^x$ *is* $e^x$
— which makes the calculus in likelihoods and growth models come out without stray
constants. For reading formulas, the practical rule is simpler: **in stats papers,
"$\log$" means natural log (base $e$) unless it explicitly says otherwise.**

**Why logs turn products into sums.** Here's the property that makes logs beloved:
multiplying two numbers becomes *adding* their logs. Multiplying a long chain of
probabilities (all tiny, prone to underflow) is a nightmare; adding their logs is
stable and easy. That single trick — "products become sums" — is why every likelihood
in statistics gets a log slapped on it before anyone does any work.

**What the numbers feel like.** $e^x$ is always **positive**, no matter what $x$ is —
even $e^{-100}$ is a tiny positive sliver, never zero or negative. And $\log$ only
accepts **positive** inputs (you can't ask "$e$ to what power gives $-3$?" — nothing
works). $\log$ of a number bigger than 1 is positive; $\log(1) = 0$; $\log$ of a number
between 0 and 1 is *negative*. Keep those signs in your pocket — they catch a lot of
mistakes.

## What the notation is going to say

The rules below all say variations of one idea: **$e^x$ and $\log$ undo each other, and
because exponents add when you multiply powers, logs turn multiplication into
addition.** When you see $\log(ab) = \log a + \log b$, read it as "the log of a product
is the sum of the logs" — the same fact as "$e^2 \cdot e^3 = e^{2+3} = e^5$," just
viewed through the rewind button.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $e$ | "ee" (Euler's number) | a fixed constant $\approx 2.718$, the natural base (in R: `exp(1)`) | $2.718\ldots$ |
| $e^x$ | "ee to the ex" | the exponential function: raise $e$ to the power $x$ (in R: `exp(x)`) | $e^2 \approx 7.389$ |
| $\exp(x)$ | "the exponential of ex" | another way to write $e^x$ | same as $e^x$ |
| $\log$ | "log" (natural log) | the natural logarithm, base $e$ — the undo of $e^x$ (in R: `log(x)`) | $\log(e^2) = 2$ |
| $\ln$ | "ell-en" (natural log) | the same natural log; some sources write $\ln$ where we write $\log$ | $\ln(e) = 1$ |
| $x^{-1}$ | "ex to the minus one" | one over: $x^{-1} = 1/x$; a negative exponent means "reciprocal" | $e^{-1} \approx 0.368$ |
| $a, b, y$ | "ay, bee, why" | generic stand-ins for *any* number, used to state the rules | e.g. $a = 4$, $b = 3$ below |

## The formal version

Here is the whole toolkit — five rules, each one just the undo-relationship or the
"exponents add" fact in disguise. We'll verify each numerically right after.

**The undo pair (the two most important):**

$$
\log(e^x) = x \qquad\text{and}\qquad e^{\log y} = y \ \ (y > 0)
$$

Reading it aloud: "log of ee-to-the-ex is ex" and "ee-to-the-log-of-why is why" — in
everyday words: *exp and log cancel each other, in either order.*

**The three algebra rules:**

$$
\begin{aligned}
\log(a \cdot b) &= \log a + \log b \\
\log(a / b)     &= \log a - \log b \\
\log(a^b)       &= b \cdot \log a
\end{aligned}
$$

- Line 1: **a product becomes a sum** — because $e^{p}\cdot e^{q} = e^{p+q}$, so the
  exponents (which are the logs) add.
- Line 2: **a quotient becomes a difference** — same fact with division, so exponents
  subtract. A special case: $\log(1/b) = \log 1 - \log b = -\log b$ (since $\log 1 = 0$).
- Line 3: **a power comes down out front as a multiplier** — because $a^b$ is $a$
  multiplied by itself $b$ times, and by Line 1 each copy contributes one $\log a$.

Reading Line 1 aloud: "log of a-times-b equals log-a plus log-b" — "multiplying inside
the log is the same as adding outside it."

Sanity-check every rule on friendly numbers (using $\log$ base $e$):

```{r formal-check}
a <- 4; b <- 3

# Undo pair
c(log_of_exp = log(exp(2.5)), exp_of_log = exp(log(9)))   # 2.5 and 9 -- unchanged

# Product -> sum
c(lhs = log(a * b), rhs = log(a) + log(b))                # both ~ 2.4849

# Quotient -> difference (and the 1/b special case)
c(lhs = log(a / b), rhs = log(a) - log(b))                # both ~ 0.2877
c(lhs = log(1 / b), rhs = -log(b))                        # both ~ -1.0986

# Power -> multiplier out front
c(lhs = log(a^b), rhs = b * log(a))                       # both ~ 4.1589
```

Every pair matches to the last digit — the rules aren't magic, they're the same
arithmetic written two ways.

## Explore it

```{webr}
#| autorun: true
#| fig-alt: "The curve of e^x rising steeply for positive x and flattening toward (but never touching) zero for negative x, with a dashed horizontal line at y = 0."
# explore-1 — LIVE: the shape of e^x. Positive everywhere, explosive right, hugs 0 left.
# PREDICT: is e^x ever zero or negative for any x? Write your guess, then run and look.
curve(exp(x), from = -3, to = 3, lwd = 2, col = "steelblue",
      ylab = "exp(x)", main = "e^x is always positive")
abline(h = 0, lty = 3)          # the floor it approaches but never reaches
points(0, exp(0), pch = 16)     # e^0 = 1
```

```{webr}
#| autorun: true
# explore-2 — LIVE: log and exp really are inverses. Feed a number through both.
# PREDICT: what will `log(exp(z))` and `exp(log(z))` each equal? Guess, then run.
z <- 7                          # <-- try other POSITIVE numbers
c(round_trip_1 = log(exp(z)),   # exp then log
  round_trip_2 = exp(log(z)))   # log then exp -- both give z back
```

```{webr}
#| autorun: false
# explore-3 — LIVE: the "products become sums" rule is what makes logs useful.
# PREDICT: multiply these three numbers, then take the log. Will it equal the SUM of
# their three individual logs? Write yes/no, then run.
vals <- c(2, 5, 10)
by_product <- log(prod(vals))              # log of the product
by_sum     <- sum(log(vals))               # sum of the logs
c(by_product = by_product, by_sum = by_sum)
```

## Common traps

**Trap: "$\log(a + b)$ is $\log a + \log b$."** Very tempting, because the product rule
looks so close. But the rule turns *products* into sums, not *sums* into sums. Check it:
$\log(4 + 3) = \log(7) \approx 1.946$, while $\log 4 + \log 3 \approx 1.386 + 1.099 =
2.485$. **Not equal.** There is no rule that simplifies $\log$ of a sum — leave it
alone.

**Trap: "a negative exponent makes the answer negative."** No — a negative exponent
means "one over," and the result stays **positive**. $e^{-2} = 1/e^{2} \approx 0.135$,
a small *positive* number, not $-7.389$. Exponentials never go negative; only their
*exponent* can.

## Check yourself

**1.** Without a calculator: what is $\log(e^5)$? And what is $e^{\log(12)}$?

::: {.callout-note collapse="true" title="Answer"}
$\log(e^5) = 5$ and $e^{\log(12)} = 12$. Both are the undo pair in action — exp and log
cancel, leaving the exponent (5) or the input (12) untouched.
:::

**2.** Use the log rules to rewrite $\log\!\left(\dfrac{a}{b}\right)$ as a sum/difference
of simpler logs, then check it in R for $a = 6, b = 2$.

```{webr}
#| autorun: false
# check2-starter — edit and Run to check your rewrite
a <- 6; b <- 2
```

::: {.callout-note collapse="true" title="Answer"}
The quotient rule gives $\log(a/b) = \log a - \log b$.

```{r check2-answer}
a <- 6; b <- 2
c(lhs = log(a / b), rhs = log(a) - log(b))   # both ~ 1.0986
```
Both equal about $1.0986$ ($= \log 3$, since $6/2 = 3$) — the quotient inside the log
became a difference outside it.
:::

**3.** In your own words: explain to a colleague, in two sentences, what the natural log
gives you and why turning products into sums is so useful.

::: {.callout-note collapse="true" title="A model answer"}
The natural log of a number answers "$e$ raised to what power gives this number?" — it's
the rewind button for exponential growth. Its product-to-sum rule lets you replace
multiplying a long chain of numbers (which is fiddly and can underflow to zero) with
adding their logs, which is why every likelihood in statistics is worked on the
log scale.
:::

## What you can now do

- **Evaluate** $e^x$ and $\log(x)$ for simple inputs and know their signs (both $e^x$
  and valid $\log$ inputs are positive; $\log$ of a number below 1 is negative).
- **Use the undo pair** $\log(e^x) = x$ and $e^{\log y} = y$ to cancel exp and log in
  either order.
- **Apply the three log rules** — product→sum, quotient→difference, power→multiplier —
  and know that $\log$ of a *sum* does not simplify.
- **Explain why** likelihoods are taken on the log scale (products become sums).

If any of these feels shaky, revisit [Getting a feel for it](#getting-a-feel-for-it)
and [The formal version](#the-formal-version) — courses that use this module lean on
the undo pair and the product→sum rule constantly.

## Where this goes next

This is a shared building block, so it feeds several courses. You'll use $e^{-z}$ to
build the **sigmoid** function in logistic regression, the natural log to define
**log-odds** and to turn a product of probabilities into a **log-likelihood** you can
actually work with, and these same rules whenever a derivation says "take the log of
both sides." Ready for more reps? Do the [practice problems](practice.qmd).
