---
title: "Practice: Exponentials and logs"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

Work these in order — they get harder. Use paper or the live cells; only then open the
answer. Reminder: `exp(x)` is $e^x$ and `log(x)` (no base) is the natural log.

## Problem 1 — the toy example, new numbers

Compute each of these by reasoning it out, then check in R: $e^0$, $e^{-1}$,
$\log(1)$, and $\log(e^3)$.

```{webr}
#| autorun: false
# p1-starter — edit and Run to check your four answers
c(exp(0), exp(-1), log(1), log(exp(3)))
```

::: {.callout-note collapse="true" title="Answer"}
- $e^0 = 1$ (anything to the 0 is 1).
- $e^{-1} = 1/e \approx 0.368$ (negative power means "one over").
- $\log(1) = 0$ (because $e^0 = 1$).
- $\log(e^3) = 3$ (the undo pair: log cancels exp, leaving the exponent).

```{r p1-answer}
c(exp0 = exp(0), exp_m1 = exp(-1), log1 = log(1), log_e3 = log(exp(3)))
# 1, 0.368, 0, 3
```
:::

## Problem 2 — fill in the missing step

We want to simplify $\log(a^2 b)$ into separate logs. Here's the work with the
**core step blanked**:

> Step 1: $\log(a^2 b) = \log(a^2) + \log(b)$  *(product rule: a product becomes a sum)*
> Step 2: **your step here** — simplify the $\log(a^2)$ term
> Step 3: $= 2\log(a) + \log(b)$

Supply Step 2.

```{webr}
#| autorun: false
# p2-starter — check your simplification for a = 3, b = 5
a <- 3; b <- 5
```

::: {.callout-note collapse="true" title="Answer"}
Step 2 uses the **power rule**: $\log(a^2) = 2\log(a)$ — the exponent 2 comes down out
front as a multiplier. Putting it together gives $\log(a^2 b) = 2\log(a) + \log(b)$.

```{r p2-answer}
a <- 3; b <- 5
c(lhs = log(a^2 * b), rhs = 2 * log(a) + log(b))   # both ~ 3.807
```
Both sides equal about $3.807$ — the rewrite is exact.
:::

## Problem 3 — a round trip on the log scale

A colleague has a probability-like number $p = 0.2$. They compute its **log-odds**,
$\log\!\big(\tfrac{p}{1-p}\big)$, and then want to get $p$ back from that log-odds value.
Compute the log-odds, then recover $p$ using the undo pair, and confirm you land back on
$0.2$.

```{webr}
#| autorun: false
# p3-starter — compute the log-odds, then invert it back to p
p <- 0.2
```

::: {.callout-note collapse="true" title="Answer"}
Forward: the odds are $p/(1-p) = 0.2/0.8 = 0.25$, and the log-odds is $\log(0.25)
\approx -1.386$. To go back, exponentiate (that undoes the log) to recover the odds,
then solve odds $= p/(1-p)$ for $p$:

```{r p3-answer}
p <- 0.2
log_odds <- log(p / (1 - p))          # forward: ~ -1.386
odds_back <- exp(log_odds)            # exp undoes log -> back to the odds, 0.25
p_back    <- odds_back / (1 + odds_back)  # solve odds = p/(1-p) for p
c(log_odds = log_odds, p_back = p_back)   # -1.386, and 0.2 recovered
```
`exp()` cancels the `log()`, handing back the odds; rearranging then returns $p = 0.2$.
This exact round trip is how logistic regression turns a modeled log-odds back into a
probability.
:::

## Problem 4 — spot the error

A colleague is simplifying a log expression and writes the following. It contains
**exactly one mistake** — find it and fix it.

> $\log(a + b) = \log(a) + \log(b)$, so for $a = 5, b = 5$ we get
> $\log(10) = \log(5) + \log(5) \approx 1.609 + 1.609 = 3.219$.

```{webr}
#| autorun: false
# p4-starter — run both sides and see they disagree
a <- 5; b <- 5
c(claimed = log(a) + log(b), actual = log(a + b))
```

::: {.callout-note collapse="true" title="Answer"}
The mistake is the very first equality: **there is no rule** $\log(a+b) = \log a + \log
b$. The product rule turns a *product* into a sum, not a *sum* into a sum. The numbers
expose it: $\log(10) \approx 2.303$, but $\log 5 + \log 5 \approx 3.219$ — not equal.

```{r p4-answer}
a <- 5; b <- 5
c(claimed = log(a) + log(b), actual = log(a + b))   # 3.219 vs 2.303 -- different
```
What the colleague *could* have written is $\log(a \cdot b) = \log a + \log b$: indeed
$\log(25) \approx 3.219$ matches. A sum inside a log doesn't break apart.
:::

## Problem 5 — interpretation

You're reading a paper. In one spot it reports a quantity on the **log scale** as
$-2.3$, and in another spot a different quantity as $+1.6$. Without computing exact
values: which of the two underlying (un-logged) quantities is *bigger than 1*, and which
is *between 0 and 1*? Explain the reasoning in one sentence.

::: {.callout-note collapse="true" title="Answer"}
The quantity whose log is $+1.6$ is **bigger than 1**, and the one whose log is $-2.3$
is **between 0 and 1**. Reason: $\log(1) = 0$, so a positive log means the original
number exceeded 1 (you had to raise $e$ to a positive power), while a negative log means
the original number was a fraction below 1 (a negative power of $e$). Checking:
$e^{1.6} \approx 4.95$ and $e^{-2.3} \approx 0.10$ — exactly as the signs predicted.
:::

## Problem 6 — a gentle cliffhanger

The **sigmoid** function, which you'll meet in the logistic-regression course, is
$\sigma(z) = \dfrac{1}{1 + e^{-z}}$. Using only what you know now, evaluate it at
$z = 0$, and predict (in words) what happens to $\sigma(z)$ as $z$ grows very large
positive, and as $z$ goes very large negative.

```{webr}
#| autorun: false
# p6-starter — evaluate the sigmoid at a few z values
sigma <- function(z) 1 / (1 + exp(-z))
sigma(c(-6, 0, 6))
```

::: {.callout-note collapse="true" title="Answer"}
At $z = 0$: $e^{-0} = 1$, so $\sigma(0) = \tfrac{1}{1+1} = 0.5$.

As $z \to +\infty$: $e^{-z} \to 0$ (a large *negative* exponent), so $\sigma(z) \to
\tfrac{1}{1+0} = 1$.

As $z \to -\infty$: $e^{-z} \to +\infty$ (a large *positive* exponent), so the
denominator blows up and $\sigma(z) \to 0$.

```{r p6-answer}
sigma <- function(z) 1 / (1 + exp(-z))
sigma(c(-6, 0, 6))   # ~0.0025, 0.5, ~0.9975 -- squashed into (0, 1)
```
So the sigmoid takes *any* number and squashes it into the interval $(0, 1)$ — landing
at exactly $0.5$ when its input is $0$. That "squash any number into a probability" move
is the whole reason logistic regression starts here — the next course builds on it.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/exponentials-and-logs" pass="4"}

::: {.quiz-q qid="eal-product-rule" qtype="numeric" answer="2.995" tolerance="0.001"}
Given $\log 4 \approx 1.386$ and $\log 5 \approx 1.609$, use the product rule to find
$\log 20$. Round to three decimals.
:::

::: {.quiz-q qid="eal-log-of-sum" qtype="mc" answer="a"}
A student writes $\log(a + b) = \log a + \log b$. For which positive inputs does this
hold?

a) For no positive $a$ and $b$ — $\log$ of a sum has no simplification rule
b) For all positive $a$ and $b$
c) Only when $a = b$
:::

::: {.quiz-q qid="eal-neg-exponent-sign" qtype="mc" answer="b"}
What is the sign of $e^{-2}$?

a) Negative — the minus sign carries through, giving about $-7.389$
b) Positive — a negative exponent means "one over", so $e^{-2} = 1/e^{2} \approx 0.135$
c) Zero — a negative exponent collapses the exponential to zero
:::

::: {.quiz-q qid="eal-undo-pair" qtype="numeric" answer="8" tolerance="0"}
The undo pair says exp and log cancel. What is $e^{\log(8)}$?
:::

::: {.quiz-q qid="eal-why-log-likelihood" qtype="mc" answer="c"}
Why are likelihoods worked on the log scale?

a) The log makes likelihood values positive, which is required before maximizing
b) The log converts probabilities into log-odds, linking to the logit
c) The log turns a product of many probabilities into a sum, avoiding underflow to zero
:::

:::

::::
