---
title: "Practice: Reading a methods section — the conventions"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only (a {webr} starter does not see it — each starter re-states its
# own data). Base R is all we need, so nothing to install.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Most are reading/translation exercises (this is a
reading module); a couple hand you R to *feel* an asymptotic word. Try each honestly, then open
the answer.

## Problem 1 — the toy example, a new jar

The lesson felt "iid" and "consistent" with a fair coin (jar = {0, 1}, true center 0.5). Redo
the same idea with a **fair six-sided die** (jar = {1, 2, 3, 4, 5, 6}). The true center (average
of 1..6) is 3.5. Draw 5 iid rolls and take their mean; then watch the mean settle toward 3.5 as
$n$ grows. **PREDICT** first: will 5 rolls land close to 3.5, or is 5 too few to count on?

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
set.seed(42)
rolls <- sample(1:6, size = 5, replace = TRUE)   # 5 iid rolls: same die, put it back each time
rolls
mean(rolls)
# now loop n = 5, 50, 500, 5000 and print the mean each time
```

::: {.callout-note collapse="true" title="Answer"}
Five rolls is far too few to count on — the sample mean bounces around. But as $n$ grows the
mean homes in on 3.5, exactly the "consistent as $n \to \infty$" promise, now for a die instead
of a coin. Each roll is iid: the *same* die (identically distributed) rolled *separately*
(independent), with `replace = TRUE` standing in for "the die is unchanged between rolls."

```{r p1-answer}
set.seed(42)
rolls <- sample(1:6, size = 5, replace = TRUE)
rolls           # 1 5 1 1 2
mean(rolls)     # 2.0 — nowhere near 3.5; five rolls is noisy

set.seed(42)
for (n in c(5, 50, 500, 5000)) {
  cat(sprintf("n = %5d   mean roll = %.3f\n", n, mean(sample(1:6, n, replace = TRUE))))
}
# n =     5   mean roll = 2.000
# n =    50   mean roll = 3.340
# n =   500   mean roll = 3.492
# n =  5000   mean roll = 3.457
```

At $n = 5$ the mean was 2.0 (way off); by $n = 500$ it's 3.492, hugging the true 3.5.
:::

## Problem 2 — fill in the missing step (decode an excerpt)

Here is an invented one-line excerpt with a **word-for-word translation**, but the translation
of one phrase is **blanked**. Supply ONLY the blanked translation.

> **Excerpt:** "WLOG assume the errors have mean 0. It suffices to show the estimator is
> unbiased; the rest follows."

> **Translation:** "We simplify by shifting the errors to average 0 — and this costs us nothing,
> because the general case works the same way after undoing the shift.
> **`<your step here>`** — proving *that* smaller claim is enough to prove the whole result.
> Everything else is a direct consequence."

What single ritual phrase does the blanked sentence translate, and what is its plain-English
meaning?

::: {.callout-note collapse="true" title="Answer"}
The blanked line translates **"it suffices to show the estimator is unbiased."** Plain English:
"proving this one smaller claim (that the estimator is unbiased) is *enough* — we don't need to
attack the big result directly; establishing the smaller claim gets us there." The phrase tells
you the proof has just **swapped to an easier target**. (The other two phrases were already
translated: "WLOG" = the cost-free simplification, "the rest follows" = a logic connector, no new
fact.)
:::

## Problem 3 — sort the sentences into the four beats

A methods section contains these five sentences (invented). Label each with the beat it belongs
to — **model**, **estimation**, **inference**, or **validation** — and give a one-line reason.

> **(a)** "Assume $x_1, \dots, x_n$ are iid with finite variance."
> **(b)** "We estimate the center by minimizing the sum of squared deviations."
> **(c)** "Across 1000 simulated datasets, the method recovered the true value 96% of the time."
> **(d)** "The estimator is consistent, with error of order $O(1/\sqrt{n})$."
> **(e)** "Let the true center be denoted $\mu$, the quantity we wish to recover."

::: {.callout-note collapse="true" title="Answer"}
- **(a) Model** — it states the assumptions about the data (iid, finite variance): the fine
  print the guarantees ride on.
- **(b) Estimation** — it gives the *recipe* for computing an answer (minimize squared
  deviations): how you get a number.
- **(c) Validation** — simulation evidence that the method behaves as claimed: "does it actually
  work?"
- **(d) Inference** — a claim about *trust* in the answer: consistency (homes in as $n$ grows)
  plus an error *rate* ($O(1/\sqrt{n})$: 4× the data ≈ half the error).
- **(e) Model** — it defines the object being estimated ($\mu$, the true center); part of the
  setup of *what we're assuming/targeting*.

The skill is spotting *what each sentence is for*: assumptions and the target object = model;
the recipe = estimation; trust/error claims = inference; evidence-it-works = validation.
:::

## Problem 4 — spot the error *(from the lesson's proof-trust trap)*

A colleague explains how they read the paper's main result. Their reasoning contains **exactly
one mistake** — find it, say why it's wrong, and give the correct reading.

> "Theorem 1 states the estimator is consistent, and a full proof follows, ending in a little
> $\square$ box. I couldn't follow the algebra in the proof, so I concluded I'm not allowed to
> rely on the consistency result — if you can't verify the proof yourself, you can't use the
> theorem."

::: {.callout-note collapse="true" title="Answer"}
The mistake is the conclusion **"if you can't verify the proof yourself, you can't use the
theorem."** That inverts the whole point of the boxed-result convention. A **Theorem** is a
result the authors (and reviewers) have *already* verified — the proof is *their* justification,
and the $\square$ box marks where it ends. You are invited to read the **statement**, trust it,
and use it, skipping the proof on a first pass; you trace the proof only when you specifically
want the *why* or intend to audit it. So the correct reading is the reverse: not following the
algebra is *fine* — the result is still yours to rely on, and the $\square$ is precisely your cue
that it's safe to skip ahead. (This is Trap 2 from the lesson: "a proof I can't follow means I
can't trust or use the result" — tempting, but backwards.)
:::

## Problem 5 — mixing it with reading math notation *(uses [reading-math-notation](../reading-math-notation/lesson.qmd)'s glyph decoding)*

Reading a full asymptotic sentence needs *both* skills at once: the glyphs (from
[reading math notation](../reading-math-notation/lesson.qmd)) and the conventions (this lesson).
Decode this invented sentence completely — read every symbol aloud, then say what the whole
sentence *promises*:

> "As $n \to \infty$, $\hat{\theta} \to \theta$, with error $O(1/\sqrt{n})$."

::: {.callout-note collapse="true" title="Answer"}
**Symbol-by-symbol** (glyphs are [reading math notation](../reading-math-notation/lesson.qmd)
territory; the arrow, infinity, the hat accent, big-O are all decoded there or in this lesson):

- "$n \to \infty$" — "as $n$ goes to infinity" = as the sample size grows without bound.
- "$\hat{\theta}$" — "theta-hat": our *estimate* (the hat = "estimated from data") of the
  unknown true value.
- "$\theta$" — "theta": the true, unknown value we're after.
- "$\hat{\theta} \to \theta$" — "theta-hat goes to theta": the estimate approaches the truth.
- "$O(1/\sqrt{n})$" — "big-oh of one-over-root-n": the error shrinks *like* one over the square
  root of $n$.

**What it promises, in plain words:** "The more data you collect, the closer the estimate gets
to the true value (that's **consistency**), and the typical error shrinks like $1/\sqrt{n}$ — so
gathering **4×** the data roughly **halves** the error." You needed reading-math-notation for the
$\hat{}$, $\to$, $\infty$, $\sqrt{}$ glyphs and this lesson for what "$\to \infty$",
"consistency," and big-O *mean*.
:::

## Problem 6 — a gentle transfer (an assumption you must check)

New context: you want to apply a method whose guarantees are stated **"under Assumptions
A1–A2,"** where:

> **(A1)** The observations are iid.
> **(A2)** The variance is finite.

Your actual data are **daily temperature readings from a single weather station, one per day for
a year**. For each assumption, decide whether your data plausibly satisfies it, and say what
follows for the paper's guarantees if one fails. (No code — this is a reading-and-judgment
problem, exactly what the assumption-list convention is *for*.)

::: {.callout-note collapse="true" title="Answer"}
**A1 (iid) — almost certainly violated.** Daily temperatures are *not independent*: a warm day
is very likely followed by another warm day (strong day-to-day carry-over), and they're arguably
*not identically distributed* either, because summer and winter days come from seasonally
different "jars." So the "same jar, separate draws, none nudging another" promise fails on both
counts.

**A2 (finite variance) — plausibly fine.** Temperatures don't swing to unbounded extremes, so a
finite variance is a reasonable assumption here.

**What follows:** because the guarantees are stated *"under A1–A2,"* and **A1 fails**, the
paper's promises (e.g. its consistency or error-rate claims) **may not hold for your
data** — quietly, with nothing in the result itself to warn you. That is exactly why the
assumption list is the *stop-and-check* bucket: reading it against your own situation is how you
learn whether a method's guarantees are actually yours to keep.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/paper-and-proof-conventions" pass="4"}

::: {.quiz-q qid="ppc-iid-two-promises" qtype="mc" answer="c"}
What does "iid" promise about data?

a) Only that the data was collected randomly
b) That the data is normally distributed
c) Two promises: the observations are independent, and identically distributed (same underlying source)
:::

::: {.quiz-q qid="ppc-independence-violation" qtype="mc" answer="b"}
Repeated measurements on the same subject most directly break which part of iid?

a) Identical distribution — the readings come from different sources
b) Independence — today's reading is tied to yesterday's
c) Neither — repeated measurements are always iid
:::

::: {.quiz-q qid="ppc-trust-boxed" qtype="mc" answer="a"}
You reach a Theorem whose proof you cannot follow. What is the right move on a first
pass?

a) Trust the boxed statement — it is already verified — and use it; read the proof only when you need the *why*
b) Stop, because an unread proof means the result cannot be trusted
c) Rewrite the proof yourself before going on
:::

::: {.quiz-q qid="ppc-consistent" qtype="mc" answer="b"}
A paper calls an estimator "consistent." What does that promise?

a) It is exactly correct for every sample size
b) It converges to the true value as the sample size $n \to \infty$
c) It has the smallest possible variance
:::

::: {.quiz-q qid="ppc-hat-vs-true" qtype="mc" answer="a"}
(From reading math notation.) In a paper, how do $\hat\beta$ and $\beta$ differ?

a) $\hat\beta$ is an estimate computed from data; $\beta$ is the true, unobserved value
b) They mean the same thing
c) $\hat\beta$ is always larger than $\beta$
:::

:::

::::
