---
title: "Practice: Vectors, subscripts, and the big Σ"
subtitle: "Answers are hidden — try honestly first, then click to check."
---

```{r setup}
#| message: false
# Packages this practice set uses. Missing ones are installed automatically.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

Work these in order — they get harder. Do them on paper or in an R session; only
then open the answer.

## Problem 1 — the toy example, new numbers

You measured 4 mushrooms: 3 cm, 1 cm, 4 cm, 2 cm. Using the notation
$x = (3, 1, 4, 2)$: what are $n$, $x_2$, and $\sum_{i=1}^{n} x_i$? Do it on paper
first, then check in R.

```{r p1-starter}
# Starter — copy this to begin
x <- c(3, 1, 4, 2)
```

::: {.callout-note collapse="true" title="Answer"}
$n = 4$ (four mushrooms). $x_2 = 1$ (the second value — subscripts point at
positions, not sizes). The sum: $3 + 1 = 4$, $4 + 4 = 8$, $8 + 2 = 10$, so
$\sum_{i=1}^{4} x_i = 10$.

```{r p1-answer}
x <- c(3, 1, 4, 2)
length(x)   # n = 4
x[2]        # x sub 2 = 1
sum(x)      # the big sigma = 10
```
:::

## Problem 2 — read it aloud, then compute

For the same $x = (3, 1, 4, 2)$, compute by hand:
$$\sum_{i=1}^{4} (x_i - 1)$$
Then re-do it the lazy way using Rule 1 (split the sum) and Rule 3 (sum of a
constant), without touching the individual values again.

```{r p2-starter}
# Starter — copy this to begin
x <- c(3, 1, 4, 2)
```

::: {.callout-note collapse="true" title="Answer"}
Read aloud: "the sum, from i equals 1 to 4, of x-sub-i minus one."

By hand, item by item: $(3-1) + (1-1) + (4-1) + (2-1) = 2 + 0 + 3 + 1 = 6$.

The lazy way — one algebraic step per line:

$$
\begin{aligned}
\sum_{i=1}^{4} (x_i - 1) &= \sum_{i=1}^{4} x_i - \sum_{i=1}^{4} 1 \\
&= 10 - 4 \times 1 = 6
\end{aligned}
$$

- Line 1: split the sum (Rule 1, with $y_i = -1$).
- Line 2: we already know $\sum x_i = 10$; summing the constant 1 four times is
  $4 \times 1$ (Rule 3).

Same answer, no re-adding. This "subtract a constant from every value" pattern is
exactly what happens with deviations from the mean in the next module.
:::

## Problem 3 — spot the trap

Without computing anything: which is bigger for $x = (3, 1, 4, 2)$,
$\sum x_i^2$ or $\left(\sum x_i\right)^2$? Then verify in R.

```{r p3-starter}
# Starter — copy this to begin
x <- c(3, 1, 4, 2)
```

::: {.callout-note collapse="true" title="Answer"}
Add-then-square is bigger here: $\left(\sum x_i\right)^2 = 10^2 = 100$, while
square-then-add is $9 + 1 + 16 + 4 = 30$. (With all-positive values, add-then-square
always wins because the cross-terms pile up.)

```{r p3-answer}
x <- c(3, 1, 4, 2)
sum(x^2)    # 30
sum(x)^2    # 100
```

Papers rely on you noticing which one is written. Misreading this single
parenthesis is one of the most common formula-reading errors.
:::

## Problem 4 — the rules on real-sized data (in R)

Starter below. Show numerically that
$\sum_{i=1}^{n} (5 x_i + 2) = 5 \sum_{i=1}^{n} x_i + 2n$ for this data, and say
which rule justifies each move.

```{r p4-starter}
# Starter data — your job: verify the identity above two ways
set.seed(42)
x <- round(rnorm(100, mean = 50, sd = 10))
n <- length(x)
```

::: {.callout-note collapse="true" title="Answer"}
```{r p4-answer}
set.seed(42)
x <- round(rnorm(100, mean = 50, sd = 10))
n <- length(x)

# Left side: transform every value, then add
lhs <- sum(5 * x + 2)

# Right side: use the rules to avoid touching individual values
rhs <- 5 * sum(x) + 2 * n

c(lhs = lhs, rhs = rhs)   # identical
```

Justification, one move per line:

$$
\begin{aligned}
\sum (5x_i + 2) &= \sum 5x_i + \sum 2 \\
&= 5\sum x_i + \sum 2 \\
&= 5\sum x_i + 2n
\end{aligned}
$$

- Line 1: split the sum (Rule 1).
- Line 2: pull the constant multiplier out (Rule 2).
- Line 3: summing the constant 2, n times, is 2n (Rule 3).
:::

## Problem 5 — a gentle cliffhanger

For $x = (3, 1, 4, 2)$ the average is $10/4 = 2.5$. Compute
$\sum_{i=1}^{4} (x_i - 2.5)$ — the total of the "distances from the average". Notice
anything? (You can't fully explain it yet — the next module,
[mean, variance & covariance](../mean-variance-covariance/lesson.qmd), is exactly
about why this happens.)

```{r p5-starter}
# Starter — copy this to begin
x <- c(3, 1, 4, 2)
```

::: {.callout-note collapse="true" title="Answer"}
$(3-2.5) + (1-2.5) + (4-2.5) + (2-2.5) = 0.5 - 1.5 + 1.5 - 0.5 = 0$.

It's exactly zero — and that's no accident: deviations from the mean ALWAYS sum to
zero (the overshoots and undershoots cancel perfectly). That's why variance squares
the deviations before adding them, which is where the next lesson picks up.

```{r p5-answer}
x <- c(3, 1, 4, 2)
sum(x - mean(x))   # 0 (up to tiny floating-point dust)
```
:::
