---
title: "Practice: Vectors, subscripts, and the big Σ"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| message: false
# Packages this practice set uses. Missing ones are installed automatically.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Do them on paper or in an R session; only
then open the answer.

## Problem 1 — the toy example, new numbers

You measured 4 mushrooms: 3 cm, 1 cm, 4 cm, 2 cm. Using the notation
$x = (3, 1, 4, 2)$: what are $n$, $x_2$, and $\sum_{i=1}^{n} x_i$? Do it on paper
first, then check in R.

```{webr}
#| autorun: false
# p1-starter — copy this to begin
x <- c(3, 1, 4, 2)
```

::: {.callout-note collapse="true" title="Answer"}
$n = 4$ (four mushrooms). $x_2 = 1$ (the second value — subscripts point at
positions, not sizes). The sum: $3 + 1 = 4$, $4 + 4 = 8$, $8 + 2 = 10$, so
$\sum_{i=1}^{4} x_i = 10$.

```{r p1-answer}
x <- c(3, 1, 4, 2)
length(x)   # n = 4
x[2]        # x sub 2 = 1
sum(x)      # the big sigma = 10
```
:::

## Problem 2 — faded worked example

For $x = (3, 1, 4, 2)$, simplify $\sum_{i=1}^{4} (x_i - 3)$ using the sum rules.
A solution is partially worked below — one core step is left blank. Fill it in, then
verify in R.

$$
\begin{aligned}
\sum_{i=1}^{4} (x_i - 3)
  &= \sum_{i=1}^{4} x_i - \sum_{i=1}^{4} 3 \\
  &= \langle\text{your step here}\rangle \\
  &= -2
\end{aligned}
$$

*Which rule turns line 2 into line 3? Apply it.*

```{webr}
#| autorun: false
# p2-starter — copy this to begin
x <- c(3, 1, 4, 2)
```

::: {.callout-note collapse="true" title="Answer"}
The blank step applies Rule 3 (summing a constant $n$ times gives $n \cdot c$) and
substitutes the known value of $\sum x_i$ from Problem 1:

$$
\begin{aligned}
\sum_{i=1}^{4} (x_i - 3)
  &= \sum_{i=1}^{4} x_i - \sum_{i=1}^{4} 3 \quad\text{(Rule 1 — split the sum)}\\
  &= 10 - 4 \times 3 \quad\text{(Rule 3 — sum of constant; }\sum x_i = 10\text{)}\\
  &= 10 - 12 = -2
\end{aligned}
$$

```{r p2-answer}
x <- c(3, 1, 4, 2)
sum(x - 3)         # -2 — same answer
```

The pattern "subtract a constant from every value" recurs constantly; Rule 1 + Rule 3
always collapse it to $\sum x_i - nc$.
:::

## Problem 3 — spot the error

A student computed $\sum_{i=1}^{4} x_i^2$ for $x = (3, 1, 4, 2)$ and wrote:

> "I added up all the values first: $3 + 1 + 4 + 2 = 10$, then squared: $10^2 = 100$.
> So $\sum x_i^2 = 100$."

This contains **exactly one mistake**. Find it, name which trap it falls into, and
give the correct value. Then verify both numbers in R.

```{webr}
#| autorun: false
# p3-starter — copy this to begin
x <- c(3, 1, 4, 2)
```

::: {.callout-note collapse="true" title="Answer"}
The mistake is the order of operations: the student computed $\left(\sum x_i\right)^2$
(add first, then square) but called it $\sum x_i^2$ (which means square first, then
add). This is Trap 1 from the lesson — the exponent sits *on* $x_i$, so the Σ must
reach only to $x_i^2$, not to the whole sum.

Correct value: $3^2 + 1^2 + 4^2 + 2^2 = 9 + 1 + 16 + 4 = 30$.

The student's answer of 100 is $\left(\sum x_i\right)^2$, a different expression.

```{r p3-answer}
x <- c(3, 1, 4, 2)
sum(x^2)    # 30  ← correct: square each term, THEN add
sum(x)^2    # 100 ← the student's expression: add first, THEN square
```
:::

## Problem 4 — the rules on real-sized data (in R)

Starter below. Show numerically that
$\sum_{i=1}^{n} (5 x_i + 2) = 5 \sum_{i=1}^{n} x_i + 2n$ for this data, and say
which rule justifies each move.

```{webr}
#| autorun: false
# p4-starter — your job: verify the identity above two ways
set.seed(42)
x <- round(rnorm(100, mean = 50, sd = 10))
n <- length(x)
```

::: {.callout-note collapse="true" title="Answer"}
```{r p4-answer}
set.seed(42)
x <- round(rnorm(100, mean = 50, sd = 10))
n <- length(x)

# Left side: transform every value, then add
lhs <- sum(5 * x + 2)

# Right side: use the rules to avoid touching individual values
rhs <- 5 * sum(x) + 2 * n

c(lhs = lhs, rhs = rhs)   # identical
```

Justification, one move per line:

$$
\begin{aligned}
\sum (5x_i + 2) &= \sum 5x_i + \sum 2 \\
&= 5\sum x_i + \sum 2 \\
&= 5\sum x_i + 2n
\end{aligned}
$$

- Line 1: split the sum (Rule 1).
- Line 2: pull the constant multiplier out (Rule 2).
- Line 3: summing the constant 2, n times, is 2n (Rule 3).
:::

## Problem 5 — interleaved *(uses reading-math-notation)*

Read the following expression aloud using the pronunciation rules from
[reading math notation](../reading-math-notation/lesson.qmd), then compute it for
$x = (5, 2, 8, 1)$:

$$\sum_{i=1}^{4} (x_i - x_2)$$

Do it on paper first: identify what $x_2$ means (it's a subscript, not a power),
substitute the values, and add up the result. Then verify in R.

```{webr}
#| autorun: false
# p5-starter — copy this to begin
x <- c(5, 2, 8, 1)
```

::: {.callout-note collapse="true" title="Answer"}
Read aloud: "the sum, from i equals 1 to 4, of x-sub-i minus x-sub-two."

$x_2 = 2$ (the second position in the list — a subscript marks a position, not a
power). Substituting:

$(5-2) + (2-2) + (8-2) + (1-2) = 3 + 0 + 6 + (-1) = 8$.

Or use Rule 1 + Rule 3 (treating $x_2 = 2$ as the constant $c$):
$\sum x_i - 4 \times x_2 = 16 - 4 \times 2 = 8$.

```{r p5-answer}
x <- c(5, 2, 8, 1)
sum(x - x[2])   # 8
```

The key reading skill: $x_2$ inside the Σ is the *fixed* second value (read once,
not re-evaluated each iteration), while $x_i$ is the *running* value that changes
with each step of the loop.
:::

## Problem 6 — a gentle cliffhanger

For $x = (3, 1, 4, 2)$ the average is $10/4 = 2.5$. Compute
$\sum_{i=1}^{4} (x_i - 2.5)$ — the total of the "distances from the average". Notice
anything? (You can't fully explain it yet — the next module,
[mean, variance & covariance](../mean-variance-covariance/lesson.qmd), is exactly
about why this happens.)

```{webr}
#| autorun: false
# p6-starter — copy this to begin
x <- c(3, 1, 4, 2)
```

::: {.callout-note collapse="true" title="Answer"}
$(3-2.5) + (1-2.5) + (4-2.5) + (2-2.5) = 0.5 - 1.5 + 1.5 - 0.5 = 0$.

It's exactly zero — and that's no accident: deviations from the mean ALWAYS sum to
zero (the overshoots and undershoots cancel perfectly). That's why variance squares
the deviations before adding them, which is where the next lesson picks up.

```{r p6-answer}
x <- c(3, 1, 4, 2)
sum(x - mean(x))   # 0 (up to tiny floating-point dust)
```
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/vectors-and-summation" pass="4"}

::: {.quiz-q qid="vs-square-then-add" qtype="numeric" answer="29" tolerance="0"}
For $x = (2, 3, 4)$, what is $\sum_{i=1}^{3} x_i^2$?
:::

::: {.quiz-q qid="vs-sum-then-square" qtype="mc" answer="b"}
For $x = (1, 3, 5)$, what is $\left(\sum_{i=1}^{3} x_i\right)^2$?

a) $1^2 + 3^2 + 5^2 = 35$ — squaring each term first
b) $(1 + 3 + 5)^2 = 81$ — summing first, then squaring
c) $1 + 3 + 5 = 9$ — forgetting to square at all
:::

::: {.quiz-q qid="vs-sigma-reach-constant" qtype="mc" answer="a"}
In $\sum_{i=1}^{n} x_i + c$, where does the $+c$ land?

a) Added once, outside the sum: $\left(\sum x_i\right) + c$
b) Added to every term: $\sum (x_i + c)$, so $c$ is added $n$ times
c) Multiplied by the sum: $c \sum x_i$
:::

::: {.quiz-q qid="vs-summed-constant" qtype="numeric" answer="15" tolerance="0"}
Using the sum rule for a constant, what is $\sum_{i=1}^{5} 3$?
:::

::: {.quiz-q qid="vs-subscript-index" qtype="numeric" answer="4" tolerance="0"}
(From reading math notation.) For the list $x = (2, 4, 6)$, the subscript in $x_2$ marks
a position, not a power. What is $x_2$?
:::

:::

::::
