---
title: "Fitting a line by descent — and matching a real optimizer"
subtitle: "Roll downhill on the error surface, and land where the professionals land"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

::: {.callout-tip title="This lesson runs in your browser" collapse="true"}
The code cells below with a **Run** button are live — powered by WebR (R compiled to
WebAssembly), they run right here on your phone or laptop, no install needed. Edit any
of them and re-run. The first run downloads a small R engine, so give it a few seconds.
Cells without a Run button are ordinary listings whose output was computed when the
page was built.
:::

## Where we're going

Time to point gradient descent at something real: **fitting a straight line to data.**
The two knobs are the line's intercept and slope; the thing to minimize is the total
squared prediction error as those knobs vary — a bowl over $(b_0, b_1)$, exactly the
kind of surface Module 2 rolled down. We'll build the fitter from scratch, then do the
thing that makes it trustworthy: run a **real, professional optimizer** (`scipy`'s
L-BFGS-B, compiled Fortran) on the very same data and watch our hand-built descent land
on the *same coefficients* — to seven decimal places.

**Builds on:** [the gradient (Module 2)](../02-the-gradient/lesson.qmd) and
[rolling downhill (Module 1)](../01-rolling-downhill/lesson.qmd). (If you've taken the
[simple-linear-regression course](../../../simple-linear-regression/modules/01-least-squares-idea/lesson.qmd),
this minimizes the very same squared-error score — here by descent instead of a formula.)

## Warm-up: do you remember?

Three quick recalls. Produce each before opening it.

**1.** *(Module 2)* What is the gradient of a function, and what does one vector step of
gradient descent look like?

::: {.callout-note collapse="true" title="Answer"}
The gradient is the vector of slopes, one per input. A step is
$\mathbf{v} \leftarrow \mathbf{v} - \alpha \nabla f(\mathbf{v})$ — move against the whole
slope-vector at once. See [Module 2](../02-the-gradient/lesson.qmd).
:::

**2.** *(Module 1)* What goes wrong if the learning rate is too large?

::: {.callout-note collapse="true" title="Answer"}
Each step overshoots the bottom and lands farther up the far side; the value grows
instead of settling — it diverges. See [Module 1](../01-rolling-downhill/lesson.qmd).
:::

**3.** *(baseline / regression idea)* A line is $\hat{y} = b_0 + b_1 x$. For the line
$b_0 = 1, b_1 = 2$, what does it predict at $x = 3$?

::: {.callout-note collapse="true" title="Answer"}
$\hat{y} = 1 + 2 \times 3 = 7$. "Start at the intercept, add the slope once per unit of
$x$." That prediction rule is what we'll score.
:::

## A tiny example first

Three data points that happen to sit on a perfectly straight line — so we can check the
answer by eye. $x = (1, 2, 3)$, $y = (3, 5, 7)$ (that's the line $y = 1 + 2x$).

A candidate line makes predictions $\hat{y}_i = b_0 + b_1 x_i$; its **error** on point
$i$ is the miss $\hat{y}_i - y_i$; we square the misses (so they can't cancel and so big
misses hurt most) and average them. That average is the **mean squared error**, MSE —
the height of our bowl at the knob-setting $(b_0, b_1)$.

Score a bad guess, $b_0 = 0, b_1 = 1$: predictions $1, 2, 3$; misses $1-3, 2-5, 3-7 =
-2, -3, -4$; squared $4, 9, 16$; averaged $= 29/3 \approx 9.67$. Now the true line
$b_0 = 1, b_1 = 2$: predictions $3, 5, 7$; misses all $0$; MSE $= 0$. The good line sits
at the bottom of the bowl (error zero), the bad line up on the wall. Gradient descent's
job is to roll from the bad guess down to the good one — without being *told* the answer.

The same thing in R:

```{r toy-example}
x <- c(1, 2, 3); y <- c(3, 5, 7)     # the line y = 1 + 2x, exactly

# MSE of a candidate line b = c(intercept, slope): average squared miss
mse <- function(b) mean((b[1] + b[2] * x - y)^2)

mse(c(0, 1))   # a bad guess -> ~9.67 (up the wall)
mse(c(1, 2))   # the true line -> 0    (bottom of the bowl)
```

## Getting a feel for it

**The error bowl, and the same ball.** Fix the data and let the two knobs vary: every
$(b_0, b_1)$ gives some MSE, and plotting that is a bowl-shaped surface — low where the
line fits well, high where it fits badly. Fitting the line is nothing new: it's rolling
Module 2's ball to the bottom of *this particular* bowl. The bottom is the best line.
(Where the picture bends: for squared error the bowl is guaranteed to have a single
bottom — no side valleys to get stuck in — which is why every optimizer, ours included,
finds the same answer.)

**Why squared error makes the bowl.** Squaring the misses is what turns "how wrong is
this line?" into a smooth bowl with a clear lowest point. Raw misses would cancel (a line
that's too high on some points and too low on others could fake a good score); absolute
misses have a sharp crease that's awkward to roll down. Squared error is smooth and
bowl-shaped, so the slope is defined everywhere and points cleanly downhill — ideal
terrain for gradient descent.

**What the fit feels like.** Start the knobs anywhere and descent drifts them toward the
setting that threads the line through the cloud of points. With clean collinear data it
reaches MSE $= 0$; with noisy real data (coming up in "Explore it") it stops at the
lowest *achievable* error — above zero, because no straight line hits noisy points
exactly, but lower than any other line manages. And here's the punchline you'll verify:
a from-scratch descent and a state-of-the-art optimizer, rolling down the same bowl, must
arrive at the same bottom — there's only one.

## What the notation is going to say

The formula below says: "the mean squared error of a candidate line $(b_0, b_1)$ is the
average, over the data points, of the squared miss $b_0 + b_1 x_i - y_i$." Then fitting
is just Module 2 applied to this function: measure its gradient (how the error changes as
you nudge the intercept, and as you nudge the slope), and step downhill until both slopes
vanish.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $x_i, y_i$ | "ex sub eye, why sub eye" | the $i$-th data point's input and output | $(1,3), (2,5), (3,7)$ |
| $n$ | "en" | the number of data points | 3 |
| $\sum_{i=1}^{n}$ | "the sum from eye equals one to en" | add up the term to its right, once per data point (in R: `sum(...)`) | sums the squared misses |
| $b_0, b_1$ | "bee zero, bee one" | the candidate line's intercept and slope (the two knobs) | true line: $1, 2$ |
| $\mathbf{b}_t$ | "bee sub tee" | the position: the two knobs $(b_0, b_1)$ together after $t$ steps | starts at `c(0, 0)` |
| $\hat{y}_i$ | "why hat sub eye" | the line's prediction, $b_0 + b_1 x_i$ | $3, 5, 7$ for the true line |
| $\mathrm{MSE}(b_0, b_1)$ | "em-ess-ee of bee-zero, bee-one" | mean squared error — the bowl's height at $(b_0, b_1)$ | $0$ at the true line |
| $\nabla \mathrm{MSE}$ | "grad em-ess-ee" | the gradient of the error: (slope in $b_0$, slope in $b_1$) | $(0,0)$ at the bottom |
| $\alpha$ | "alpha" | the learning rate | $0.1$ below |

## The formal version

The loss — mean squared error as a function of the two knobs:

$$
\mathrm{MSE}(b_0, b_1) = \frac{1}{n} \sum_{i=1}^{n} \big( b_0 + b_1 x_i - y_i \big)^2
$$

Reading it aloud: "em-ess-ee of bee-zero and bee-one equals one over en, times the sum
over the points, of the quantity bee-zero plus bee-one ex-sub-eye minus why-sub-eye,
squared" — plainly: "average the squared misses of the line." For the true line on our
toy data every miss is 0, so MSE $= 0$, matching the hand computation.

Fitting is Module 2's vector update, with the position being the two knobs
$\mathbf{b} = (b_0, b_1)$ and the surface being the MSE:

$$
\mathbf{b}_{t+1} = \mathbf{b}_t - \alpha \, \nabla \mathrm{MSE}(\mathbf{b}_t)
$$

Nothing new — just Module 2 with MSE as the bowl. Measure the gradient numerically
(reusing the exact tool from Module 2), loop, and watch it reach $(1, 2)$:

```{r formal-check}
x <- c(1, 2, 3); y <- c(3, 5, 7)
mse <- function(b) mean((b[1] + b[2] * x - y)^2)

# Module 2's numerical gradient, unchanged
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) {
    up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h
    (f(up) - f(dn)) / (2 * h)
  })

b <- c(0, 0)                                   # start: flat line at 0
for (t in 1:2000) b <- b - 0.1 * gradient(mse, b)
b   # ~ (1, 2): descent recovered the true line
```

## Explore it

Now the real test. Here is a small, **noisy** dataset — a line with a bit of measurement
scatter — that was handed to a professional optimizer, `scipy.optimize.minimize` with the
L-BFGS-B method (compiled Fortran, the kind of routine real software uses). Its answer
was frozen into this course's equivalence fixture. Let's run *our* from-scratch descent
on the identical data and compare.

```{webr}
#| autorun: true
#| fig-alt: "Scatter of eight noisy points with the gradient-descent best-fit line drawn through them, sloping downward."
# explore-1 — LIVE: fit the noisy data with our own descent, and plot the line.
# PREDICT: the points fall from upper-left to lower-right — will the fitted slope be
# positive or negative? Guess, then run.
x <- c(-3.5, -2.5, -1.5, -0.5, 0.5, 1.5, 2.5, 3.5)
y <- c(12.53, 9.40, 6.58, 3.59, 0.30, -2.63, -5.49, -8.53)

mse <- function(b) mean((b[1] + b[2] * x - y)^2)
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) {
    up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
  })

b <- c(0, 0)
for (t in 1:2000) b <- b - 0.1 * gradient(mse, b)   # lr = 0.1, 2000 steps

plot(x, y, pch = 16, cex = 1.4, main = sprintf("fitted line: y-hat = %.3f + %.3f x", b[1], b[2]))
abline(a = b[1], b = b[2], col = "steelblue", lwd = 2)
b
```

```{webr}
#| autorun: true
# explore-2 — LIVE: the moment of truth. Compare our from-scratch coefficients to the
# frozen answer from the REAL optimizer (scipy L-BFGS-B), on the same data.
# PREDICT: how many decimal places will match? Guess, then run.
x <- c(-3.5, -2.5, -1.5, -0.5, 0.5, 1.5, 2.5, 3.5)
y <- c(12.53, 9.40, 6.58, 3.59, 0.30, -2.63, -5.49, -8.53)

mse <- function(b) mean((b[1] + b[2] * x - y)^2)
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) {
    up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
  })
b <- c(0, 0); for (t in 1:2000) b <- b - 0.1 * gradient(mse, b)

# scipy L-BFGS-B's answer, frozen at authoring time in the equivalence fixture
# (equivalence/fixtures/gradient-descent-full-scipy.json — validation tier: FULL)
scipy_frozen <- c(1.968750029034487, -3.009404764462221)

rbind(from_scratch = b, scipy_LBFGSB = scipy_frozen)
cat("largest difference:", max(abs(b - scipy_frozen)),
    " -> match within 1e-6?", max(abs(b - scipy_frozen)) < 1e-6, "\n")
```

Our hand-built descent and a compiled Fortran optimizer agree to seven decimal places
(their largest coefficient difference is about $3 \times 10^{-8}$). They took completely
different routes — ours plodded straight downhill in 2000 small steps,
L-BFGS-B used curvature to leap there in a handful — but the bowl has exactly one bottom,
so they *must* meet at it. That agreement is what "we rebuilt it correctly" looks like.

::: {.callout-note title="Runs on your machine, not here"}
The reference above was produced in **Python with `scipy`**, which compiles Fortran and
therefore **cannot run in this browser** (WebR is R-only, no Fortran compiler — see
[`docs/webr-decision.md`](../../../../docs/webr-decision.md)). So it was run once at
authoring time and its result frozen into the course. The exact code — copy it into a
Python session on your own machine to reproduce the frozen numbers:

```python
import numpy as np
from scipy.optimize import minimize

x = np.array([-3.5, -2.5, -1.5, -0.5, 0.5, 1.5, 2.5, 3.5])
y = np.array([12.53, 9.40, 6.58, 3.59, 0.30, -2.63, -5.49, -8.53])

def mse(b):            # b = [intercept, slope]
    return np.mean((b[0] + b[1] * x - y) ** 2)

res = minimize(mse, np.zeros(2), method="L-BFGS-B")
print(res.x)           # [ 1.96875003 -3.00940476 ]  <- the frozen reference
```
:::

R's own built-in optimizer agrees too — and *it* is base R, so it runs right here:

```{webr}
#| autorun: false
# explore-3 — LIVE: R's built-in optim() is a professional optimizer as well. Run it on
# the same MSE and confirm it lands on the same coefficients as our descent and scipy.
x <- c(-3.5, -2.5, -1.5, -0.5, 0.5, 1.5, 2.5, 3.5)
y <- c(12.53, 9.40, 6.58, 3.59, 0.30, -2.63, -5.49, -8.53)
mse <- function(b) mean((b[1] + b[2] * x - y)^2)
optim(c(0, 0), mse, method = "BFGS")$par    # ~ (1.9688, -3.0094), same bottom
```

## Common traps

**Trap: "a good fit should reach MSE $= 0$."** True only if the points sit *exactly* on
a line. The noisy data above never can — no straight line passes through all eight
scattered points — so the bottom of its error bowl is *above* zero. "Best fit" means the
lowest achievable error, not zero. (Our clean toy hit 0 precisely because those three
points were exactly collinear.)

**Trap: "different optimizers give different answers, so which do I trust?"** For this
squared-error bowl they give the *same* answer — our descent, scipy's L-BFGS-B, and R's
`optim()` agreed to seven decimal places. The bowl has a single bottom, and every correct
optimizer finds it; they differ only in the *path* and *speed*, not the destination. If
two optimizers disagreed on a bowl like this, one of them (or your gradient) has a bug —
which is exactly why an equivalence check like explore-2 is worth running.

## Check yourself

**1.** Fit a line to $x = (0, 1, 2, 3)$, $y = (1, 3, 5, 7)$ (the line $y = 1 + 2x$) with
your from-scratch descent. What coefficients should it find?

```{webr}
#| autorun: false
# check1-starter — edit and Run to solve, right here
x <- c(0, 1, 2, 3); y <- c(1, 3, 5, 7)
mse <- function(b) mean((b[1] + b[2] * x - y)^2)
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) {
    up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
  })
```

::: {.callout-note collapse="true" title="Answer"}
It finds intercept $\approx 1$, slope $\approx 2$ — the exact line, MSE $\approx 0$.

```{r check1-answer}
x <- c(0, 1, 2, 3); y <- c(1, 3, 5, 7)
mse <- function(b) mean((b[1] + b[2] * x - y)^2)
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) {
    up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
  })
b <- c(0, 0); for (t in 1:3000) b <- b - 0.1 * gradient(mse, b)
b   # ~ (1, 2)
```
:::

**2.** Our descent and scipy's L-BFGS-B reached the same coefficients but took very
different numbers of steps. Does that mean one of them is more *accurate*? Explain.

::: {.callout-note collapse="true" title="Answer"}
No — they reached the *same* bottom of the *same* bowl, so they're equally accurate here;
they differ only in efficiency (how many steps to get there). L-BFGS-B uses information
about the surface's curvature to take smart, large steps, so it converges in far fewer
iterations; plain gradient descent takes many small fixed-direction steps. Same
destination, different travel time.
:::

**3.** In your own words: explain to a colleague, in two sentences, how fitting a line
becomes a gradient-descent problem, and why our from-scratch answer matches a
professional optimizer's.

::: {.callout-note collapse="true" title="A model answer"}
Fitting a line means choosing the intercept and slope that minimize the average squared
prediction error, which is a bowl-shaped surface over those two knobs — so we roll down
it with gradient descent. Because that bowl has a single lowest point, any correct
optimizer (ours or a professional one) must land on the same coefficients; only the route
and speed differ.
:::

## What you can now do

- **Write the squared-error loss** of a candidate line as a function of its two knobs,
  and recognize it as a bowl to roll down.
- **Fit a line from scratch** by running Module 2's gradient descent on that loss, in
  base R, live in the browser.
- **Validate a from-scratch method against a real reference**: run a professional
  optimizer on the same data and check the answers match within tolerance.
- **Explain why they match** — one bowl, one bottom — and why that check catches bugs.

If any of these feels shaky, revisit [Explore it](#explore-it) — the capstone leans on
your being able to read what an optimizer is doing.

## Where this goes next

You've rebuilt a fitter and shown it matches the professionals. The capstone closes the
loop: we open the actual documentation of a real optimizer (`optim()` and
`scipy`'s `minimize`) and decode it line by line, mapping every piece of jargon back to
what you built here: [decode the optimizer](../04-capstone-decode-the-optimizer/lesson.qmd).
Ready for more reps first? Do the [practice problems](practice.qmd).
