---
title: "Practice: Fitting a line by descent"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

Work these in order. The numerical gradient tool, to paste into any starter:

```r
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) {
    up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
  })
```

## Problem 1 — the toy example, new numbers

Fit a line to $x = (1, 2, 3, 4)$, $y = (5, 7, 9, 11)$ — the line $y = 3 + 2x$ — with
from-scratch descent ($\alpha = 0.05$, 3000 steps, start $(0,0)$). What coefficients
should it find, and what should the MSE be there?

```{webr}
#| autorun: false
# p1-starter
x <- c(1, 2, 3, 4); y <- c(5, 7, 9, 11)
mse <- function(b) mean((b[1] + b[2] * x - y)^2)
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) {
    up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
  })
```

::: {.callout-note collapse="true" title="Answer"}
The points are exactly on $y = 3 + 2x$, so descent finds intercept $\approx 3$, slope
$\approx 2$, with MSE $\approx 0$.

```{r p1-answer}
x <- c(1, 2, 3, 4); y <- c(5, 7, 9, 11)
mse <- function(b) mean((b[1] + b[2] * x - y)^2)
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) {
    up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
  })
b <- c(0, 0); for (t in 1:3000) b <- b - 0.05 * gradient(mse, b)
c(intercept = b[1], slope = b[2], mse = mse(b))   # ~ (3, 2, 0)
```
:::

## Problem 2 — fill in the missing step

The MSE loss for a candidate line is written below with its **core expression blanked**.
Supply it.

> ```
> mse <- function(b) {
>   predictions <- b[1] + b[2] * x     # the line's predictions
>   your step here                     # turn predictions and y into the mean squared error
> }
> ```

```{webr}
#| autorun: false
# p2-starter — complete mse(), then Run to fit
x <- c(1, 2, 3, 4); y <- c(5, 7, 9, 11)
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) {
    up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
  })
```

::: {.callout-note collapse="true" title="Answer"}
The missing line is the mean of the squared misses: `mean((predictions - y)^2)`. Subtract
the actual values to get the misses, **square** them (so they can't cancel), and average.

```{r p2-answer}
x <- c(1, 2, 3, 4); y <- c(5, 7, 9, 11)
mse <- function(b) {
  predictions <- b[1] + b[2] * x
  mean((predictions - y)^2)
}
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) {
    up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
  })
b <- c(0, 0); for (t in 1:3000) b <- b - 0.05 * gradient(mse, b)
b   # ~ (3, 2)
```
:::

## Problem 3 — noisy data, and reading the leftover error

Fit this **noisy** dataset and report both the coefficients and the final MSE. Is the
MSE zero? Say, in one sentence, why or why not.

```{webr}
#| autorun: false
# p3-starter
x <- c(1, 2, 3, 4, 5); y <- c(2.1, 3.9, 6.2, 7.8, 10.1)   # ~ line 0 + 2x, with scatter
mse <- function(b) mean((b[1] + b[2] * x - y)^2)
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) {
    up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
  })
```

::: {.callout-note collapse="true" title="Answer"}
```{r p3-answer}
x <- c(1, 2, 3, 4, 5); y <- c(2.1, 3.9, 6.2, 7.8, 10.1)
mse <- function(b) mean((b[1] + b[2] * x - y)^2)
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) {
    up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
  })
b <- c(0, 0); for (t in 1:5000) b <- b - 0.02 * gradient(mse, b)
c(intercept = b[1], slope = b[2], mse = mse(b))
```
The MSE is small but **not zero**: the five points don't lie exactly on any straight
line (there's scatter), so even the best line misses each point a little, and those
leftover squared misses can't be driven to zero. Best means lowest achievable, not zero.
:::

## Problem 4 — spot the error

A colleague's line-fitter drifts to nonsense. Their loss has **exactly one mistake** —
find it and fix it.

```{webr}
#| autorun: false
# p4-starter — run it (bad result), then fix the one mistake in mse()
x <- c(1, 2, 3, 4); y <- c(5, 7, 9, 11)
mse <- function(b) mean(b[1] + b[2] * x - y)     # <- one mistake here
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) {
    up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
  })
b <- c(0, 0); for (t in 1:2000) b <- b - 0.05 * gradient(mse, b)
b
```

::: {.callout-note collapse="true" title="Answer"}
The mistake: the misses are **not squared** — `mean(b[1] + b[2]*x - y)` averages the raw
misses, which can cancel (positive and negative misses offset). That "loss" isn't a bowl;
it slopes off to $-\infty$, so descent chases the intercept downward forever. Square the
misses:

```{r p4-answer}
x <- c(1, 2, 3, 4); y <- c(5, 7, 9, 11)
mse <- function(b) mean((b[1] + b[2] * x - y)^2)   # squared: a real bowl
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) {
    up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
  })
b <- c(0, 0); for (t in 1:2000) b <- b - 0.05 * gradient(mse, b)
b   # ~ (3, 2)
```
:::

## Problem 5 — a fit that blows up *(uses Module 1's learning-rate lesson)*

Take the good fitter from Problem 1 but crank the learning rate up to $\alpha = 1.5$.
Predict what happens (recall Module 1), then run it and read the coefficients.

```{webr}
#| autorun: false
# p5-starter — same data, reckless learning rate
x <- c(1, 2, 3, 4); y <- c(5, 7, 9, 11)
mse <- function(b) mean((b[1] + b[2] * x - y)^2)
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) {
    up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
  })
```

::: {.callout-note collapse="true" title="Answer"}
Module 1's warning strikes: too large a learning rate makes each step overshoot the
bottom of the error bowl and land farther up the far wall, so the coefficients explode to
huge (often `NaN`/`Inf`) values instead of settling.

```{r p5-answer}
x <- c(1, 2, 3, 4); y <- c(5, 7, 9, 11)
mse <- function(b) mean((b[1] + b[2] * x - y)^2)
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) {
    up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
  })
b <- c(0, 0); for (t in 1:2000) b <- b - 1.5 * gradient(mse, b)
b   # huge / NaN — diverged
```
The fix is the same as in Module 1: lower the learning rate until the steps settle
instead of overshoot.
:::

## Problem 6 — transfer: match the standard tool

R's `lm()` fits a line by the exact least-squares formula (no descent). Fit the noisy
data below with your from-scratch descent, then compare your coefficients to `lm()`'s.
They should match — another independent confirmation of your rebuild.

```{webr}
#| autorun: false
# p6-starter — compare your descent to R's lm()
x <- c(-2, -1, 0, 1, 2); y <- c(5.1, 3.0, 0.8, -1.1, -3.2)
mse <- function(b) mean((b[1] + b[2] * x - y)^2)
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) {
    up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
  })
```

::: {.callout-note collapse="true" title="Answer"}
```{r p6-answer}
x <- c(-2, -1, 0, 1, 2); y <- c(5.1, 3.0, 0.8, -1.1, -3.2)
mse <- function(b) mean((b[1] + b[2] * x - y)^2)
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) {
    up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
  })
b <- c(0, 0); for (t in 1:5000) b <- b - 0.05 * gradient(mse, b)

rbind(from_scratch = b,
      lm = unname(coef(lm(y ~ x))))    # intercept, slope — should match
```
Your descent and `lm()`'s exact formula agree: that's four routes now (your descent, R's
`optim()`, scipy's L-BFGS-B, and `lm()`'s formula) all reaching the same bottom of the
same squared-error bowl. Different methods, one answer — which is exactly the confidence
an equivalence check buys you.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="courses/gradient-descent/modules/03-fitting-a-line-by-descent" pass="4"}

::: {.quiz-q qid="fld-mse-candidate" qtype="numeric" answer="4.667" tolerance="0.01"}
**On the lesson's toy data $x = (1, 2, 3)$, $y = (3, 5, 7)$, what is the MSE of the
candidate line $b_0 = 1, b_1 = 1$?** (Round to three decimals.)
:::

::: {.quiz-q qid="fld-mse-zero" qtype="mc" answer="b"}
**When can the best-fitting line reach MSE $= 0$?**

a) Whenever gradient descent runs for enough steps
b) Only when the data points sit exactly on one straight line
c) Whenever the learning rate is chosen well
:::

::: {.quiz-q qid="fld-optimizers-agree" qtype="mc" answer="c"}
**Your from-scratch descent and scipy's L-BFGS-B landed on the same coefficients to
seven decimals. Why did that have to happen?**

a) Both methods take the same downhill steps internally
b) It is a coincidence of this dataset; other data would split them
c) The squared-error bowl has a single bottom, and every correct optimizer finds it
:::

::: {.quiz-q qid="fld-why-square" qtype="mc" answer="a"}
**In the MSE, why are the misses squared before averaging?**

a) Raw misses can cancel each other, letting a bad line fake a good score
b) Squared misses are smaller, which speeds up the descent
c) Squaring keeps the error in the same units as $y$
:::

::: {.quiz-q qid="fld-update-interleave" qtype="numeric" answer="1.4" tolerance="0.01"}
**(From Module 1, *rolling downhill*.) You are at $x = 2$ on a curve whose slope there
is $3$. With learning rate $\alpha = 0.2$, where does one gradient-descent step put
you?** (Enter one decimal.)
:::

:::

::::
