---
title: "Practice: Rank-one speed-ups"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
set.seed(42)
```

Work these in order — they get harder. Do them on paper or in an R session; only then
open the answer.

## Problem 1 — the Schur complement, new numbers

You hold two indicators with $R_S = \begin{pmatrix}1 & 0.4\\0.4 & 1\end{pmatrix}$, whose
inverse is $R_S^{-1} = \frac{1}{0.84}\begin{pmatrix}1 & -0.4\\-0.4 & 1\end{pmatrix} =
\begin{pmatrix}1.190 & -0.476\\-0.476 & 1.190\end{pmatrix}$. You add a third indicator with
correlations $b_{\text{new}} = (0.5, 0.2)$ to the first two (and $d = 1$). Compute the Schur
complement $s = d - b_{\text{new}}^\top R_S^{-1} b_{\text{new}}$.

```{webr}
#| autorun: false
# p1-starter
RS <- matrix(c(1, 0.4, 0.4, 1), 2); RSi <- solve(RS); b_new <- c(0.5, 0.2); d <- 1
```

::: {.callout-note collapse="true" title="Answer"}
The helper $w = R_S^{-1} b_{\text{new}} = (1.190\cdot0.5 - 0.476\cdot0.2,\ -0.476\cdot0.5 + 1.190\cdot0.2) = (0.5,\ 0)$,
so $b_{\text{new}}^\top R_S^{-1} b_{\text{new}} = b_{\text{new}}^\top w = 0.5\cdot0.5 + 0.2\cdot0 = 0.25$,
and $s = 1 - 0.25 = 0.75$. The third indicator keeps 75% of its variance as *new* information
after the first two — comfortably far from 0, so the update is stable.

```{r}
RS <- matrix(c(1, 0.4, 0.4, 1), 2); RSi <- solve(RS); b_new <- c(0.5, 0.2)
as.numeric(1 - t(b_new) %*% RSi %*% b_new)      # 0.75
```
:::

## Problem 2 — fill in the missing step

Growing a $k\times k$ inverse by bordering, you have the helper $w = R_S^{-1}b_{\text{new}}$ and
Schur complement $s$. The new corner (bottom-right of the grown inverse) is one simple thing.

- Step 1: compute $w = R_S^{-1}b_{\text{new}}$ and $s = d - b_{\text{new}}^\top w$.
- Step 2: **your step here** — what is the bottom-right corner of $(R_S^{+})^{-1}$?
- Step 3: the off-diagonal block is $-w/s$, and the top-left is $R_S^{-1} + ww^\top/s$.

::: {.callout-note collapse="true" title="Answer"}
Step 2: the corner is $\dfrac{1}{s}$. (Everything in the bordering update is built from $w$
and the single scalar $s$: corner $1/s$, off-diagonals $-w/s$, top-left $R_S^{-1} + ww^\top/s$.)
:::

## Problem 3 — bordering matches `solve()`, in R

Take a random $4\times4$ correlation-like matrix with known inverse, add a border, and confirm
the bordering update equals a fresh `solve()` of the grown $5\times5$ matrix.

```{webr}
#| autorun: false
# p3-starter — build the bordering update and compare to solve()
sw_border <- function(Mi, b_new, d) { Mib <- Mi %*% b_new
  s <- as.numeric(d - crossprod(b_new, Mib))
  rbind(cbind(Mi + (Mib %*% t(Mib))/s, -Mib/s), c(-t(Mib)/s, 1/s)) }
set.seed(3); k <- 4
M <- crossprod(matrix(rnorm(k*k), k)); Mi <- solve(M); b_new <- rnorm(k); d <- rnorm(1)^2 + 3
```

::: {.callout-note collapse="true" title="Answer"}
The update reproduces a fresh inverse to ~$10^{-15}$.

```{r}
sw_border <- function(Mi, b_new, d) { Mib <- Mi %*% b_new
  s <- as.numeric(d - crossprod(b_new, Mib))
  rbind(cbind(Mi + (Mib %*% t(Mib))/s, -Mib/s), c(-t(Mib)/s, 1/s)) }
set.seed(3); k <- 4
M <- crossprod(matrix(rnorm(k*k), k)); Mi <- solve(M); b_new <- rnorm(k); d <- rnorm(1)^2 + 3
M_grown <- rbind(cbind(M, b_new), c(b_new, d))
max(abs(sw_border(Mi, b_new, d) - solve(M_grown)))     # ~1e-15
```
:::

## Problem 4 — spot the error

A colleague says: *"Our search makes 5,000 moves, and each swap changes two indicators, so a
swap needs a fresh matrix inversion — the rank-1 trick only helps for pure adds."* This contains
**exactly one mistake** — find it.

::: {.callout-note collapse="true" title="Answer"}
The mistake: thinking a swap needs a fresh inversion. A swap = **remove** one indicator (a
partition update) then **add** another (a bordering update) — two $O(k^2)$ rank-1-style
updates, no full inversion. Adds, removes, *and* swaps are all handled by the update family;
none re-inverts.
:::

## Problem 5 — mixing it with Sherman–Morrison *(uses the foundation's cost argument)*

A single move updates a $k\times k$ inverse. If re-inverting costs about $k^3$ operations and the
bordering update costs about $k^2$, how many times faster (in operation count) is the update at
$k = 20$, and why does that ratio matter across a 5,000-move search?

::: {.callout-note collapse="true" title="Answer"}
Roughly $k^3 / k^2 = k = 20\times$ fewer operations per move at $k = 20$. Across 5,000 moves that
compounds: the update version does about 5,000 cheap $O(k^2)$ steps instead of 5,000 expensive
$O(k^3)$ ones — the difference between an interactive search and one you wait on. The savings
come from the work being *repeated* on nearly-identical matrices.
:::

## Problem 6 — transfer: when the Schur complement warns

In problem 1 the Schur complement was $s = 0.75$. Suppose adding a different indicator gives
$s = 0.003$. What does that tiny value tell you about the new indicator, and what will happen
numerically when the update divides by it? *(Module 4 is the fix.)*

::: {.callout-note collapse="true" title="Answer"}
$s$ is the new indicator's leftover variance after the ones already in the model explain it;
$s = 0.003$ means it is almost a linear combination of them (near-collinear). Dividing by such a
tiny $s$ makes the updated inverse's entries huge and unstable — the same near-singular blow-up
that plagues $R^{-1}b$. Module 4 adds a ridge $\lambda$ to the diagonal so the matrix (and this
denominator) stays comfortably away from zero.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="courses/swap-stepwise-selection/modules/03-rank-one-speedups" pass="4"}

::: {.quiz-q qid="r1u-schur-meaning" qtype="mc" answer="c"}
When you add an indicator, the bordering update's Schur complement is
$s = d - b_{\text{new}}^\top R_S^{-1} b_{\text{new}}$. A value of $s$ near 0 means the new
indicator:

a) is uncorrelated with the indicators already present
b) adds a large, unique effect on $y$
c) is nearly a combination of the indicators already present
:::

::: {.quiz-q qid="r1u-swap-cost" qtype="mc" answer="b"}
How does the search update $R_S^{-1}$ for a swap move (one indicator out, one in)?

a) A fresh $O(k^3)$ inversion of the new matrix
b) A remove then an add — two $O(k^2)$ updates
c) It cannot update; swaps are skipped
:::

::: {.quiz-q qid="r1u-exact" qtype="mc" answer="c"}
Across thousands of search moves, how do the bordering update's results compare with
re-inverting from scratch each time?

a) Approximate — re-invert occasionally to be safe
b) Faster, but they drift over many moves
c) Exactly equal (an algebraic identity), but faster
:::

::: {.quiz-q qid="r1u-schur-compute" qtype="numeric" answer="0.75" tolerance="0.01"}
The indicators already in the model are uncorrelated, so $R_S^{-1} = I$. A new indicator has
correlations $b_{\text{new}} = (0.3,\ 0.4)$ with them and diagonal $d = 1$. Compute the Schur
complement $s = d - b_{\text{new}}^\top R_S^{-1} b_{\text{new}}$. (Enter 2 decimals.)
:::

::: {.quiz-q qid="r1u-why-updates" qtype="mc" answer="a"}
(From Module 2.) Why do these cheap inverse updates help the switching search specifically,
when they wouldn't help a single one-off regression fit?

a) The search re-scores many subsets differing by one indicator, re-inverting $R_S$ repeatedly
b) The search inverts only one matrix, once
c) A one-off fit inverts more often than the search
:::

:::

::::
