---
title: "Lines and the least-squares idea"
subtitle: "What 'the best-fitting line' actually means"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live cells below re-state their own data. Base R is all
# we need, so nothing to install.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```


## Where we're going

Regression starts with a deceptively simple question: given a cloud of points, which
straight line through it is *best*? This module builds the decision rule — score
every candidate line by its **sum of squared errors**, and call the line with the
lowest score the winner. Everything in the rest of the course (the formulas, the
`lm()` output, the paper tables) is just the consequences of that one rule.

**Builds on:** [vectors & summation](../../../../foundations/vectors-and-summation/lesson.qmd)
and [mean, variance & covariance](../../../../foundations/mean-variance-covariance/lesson.qmd).

## Warm-up: do you remember?

Three quick recalls from the foundations this module stands on. Produce each answer
before opening it.

**1.** *(vectors & summation)* Say $\sum_{i=1}^{n} a_i^2$ aloud, then compute it for
$a = (-2, 0, 2)$.

::: {.callout-note collapse="true" title="Answer"}
"The sum, from i equals 1 to n, of a-sub-eye squared" — square each, then add:
$(-2)^2 + 0^2 + 2^2 = 4 + 0 + 4 = 8$. In R: `sum(a^2)`. See
[vectors & summation](../../../../foundations/vectors-and-summation/lesson.qmd).
:::

**2.** *(mean, variance & covariance)* In one sentence: why do the deviations of any
list of numbers from their own mean always add up to zero?

::: {.callout-note collapse="true" title="Answer"}
The mean is the balance point, so the amounts above it and below it are equal and
cancel exactly — $\sum (x_i - \bar{x}) = 0$ for any list. This is the fact that will
force us to *square* misses in a moment. See
[mean, variance & covariance](../../../../foundations/mean-variance-covariance/lesson.qmd).
:::

**3.** *(mean, variance & covariance)* Compute the mean of $y = (2, 3, 5, 6)$ — the
tomato counts we're about to model.

::: {.callout-note collapse="true" title="Answer"}
$\bar{y} = (2 + 3 + 5 + 6) / 4 = 16 / 4 = 4$. In R: `mean(y)`.
:::

## A tiny example first

You gave 4 tomato plants different doses of fertilizer, $x = (1, 2, 3, 4)$ scoops,
and counted their tomatoes, $y = (2, 3, 5, 6)$. More fertilizer, more tomatoes —
but which *line* summarizes the relationship?

A line is two numbers: where it starts (the **intercept**: its height at x = 0) and
how fast it climbs (the **slope**: extra tomatoes per extra scoop).

Let's *guess* a line: intercept 1, slope 1 — "start at 1, add 1 per scoop". What
does it predict, and how wrong is it?

| plant | scoops $x$ | actual $y$ | predicted $1 + 1\cdot x$ | miss (actual − predicted) | miss² |
|---|---|---|---|---|---|
| 1 | 1 | 2 | 2 | 0 | 0 |
| 2 | 2 | 3 | 3 | 0 | 0 |
| 3 | 3 | 5 | 4 | +1 | 1 |
| 4 | 4 | 6 | 5 | +1 | 1 |

Add up the squared misses: 0 + 0 + 1 + 1 = **2**. That total is the line's score
(lower = better). Why square the misses before adding? The same reason variance did:
positive and negative misses would cancel, and a line that's wildly wrong in both
directions could score a perfect 0.

Try a second guess: intercept 0, slope 1.5. Predictions: 1.5, 3, 4.5, 6. Misses:
+0.5, 0, +0.5, 0. Squared and summed: 0.25 + 0 + 0.25 + 0 = **0.5**. Better! Guess
two beats guess one. Is something even better hiding out there? (Yes — hold that
thought.)

The same thing in R, so you can check every step:

```{r toy-example}
# The toy data: fertilizer scoops and tomato counts for 4 plants
x <- c(1, 2, 3, 4)
y <- c(2, 3, 5, 6)

# Guess 1: intercept 1, slope 1 — "start at 1, add 1 tomato per scoop"
pred1  <- 1 + 1 * x        # the line's predictions at each x
miss1  <- y - pred1        # actual minus predicted: the misses
sum(miss1^2)               # squared, then added: score = 2

# Guess 2: intercept 0, slope 1.5
pred2  <- 0 + 1.5 * x
miss2  <- y - pred2
sum(miss2^2)               # score = 0.5 — a better line
```

## Getting a feel for it

**Picture a spring from each point to the line.** Imagine every data point tied to
the line by a little vertical spring: the further the line sits from the point, the
more that spring is stretched. Squaring the miss is exactly the energy stored in a
stretched spring — stretch it twice as far and it holds *four* times the energy. The
least-squares line is the one where the springs are, all together, as relaxed as
possible. (Where the picture bends: the springs only pull straight up and down, never
sideways — regression measures misses in the $y$ direction only.)

**Why square the misses instead of just adding them?** Two reasons, both visible in
our toy numbers. First, raw misses *cancel*: you recalled in the warm-up that
deviations from a balance point sum to zero, so a wildly wrong line can post a fake
score of 0 just by missing high as much as it misses low (Trap 1 below shows the flat
line at $\bar{y}=4$ doing exactly that). Squaring makes every miss count as a
positive penalty, so cancelling is impossible. Second, squaring *punishes a big miss
far more than two small ones* — a spring stretched by 2 costs 4, but two springs
stretched by 1 cost only 2 — so the line can't win by fitting three points nicely and
abandoning the fourth. It's pushed toward being decent everywhere.

**What the score feels like.** Lower is a tighter fit: guess 1 scored 2, guess 2
scored 0.5, so guess 2 hugs the points more closely. The one move that always lowers
the score is nudging the line toward wherever its springs are stretched most. You
can't drive it all the way to 0, though — our four points don't sit on any single
straight line, so some stretch always remains. In the "Explore it" plots below, that
leftover shows up as a valley whose bottom sits *above* zero.

## What the notation is going to say

The formulas below say: "a candidate line has intercept beta-zero and slope
beta-one; its prediction for plant i is beta-zero plus beta-one times x-sub-i; the
plant's *residual* is actual minus predicted; the line's score is the sum of the
squared residuals; and the best line is the (beta-zero, beta-one) pair that makes
that score as small as possible." Exactly the table game we just played — twice.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $x_i$ | "ex sub eye" | the i-th predictor value — here, scoops of fertilizer (in R: `x[i]`) | 1, 2, 3, 4 |
| $y_i$ | "why sub eye" | the i-th observed outcome — here, tomatoes counted (in R: `y[i]`) | 2, 3, 5, 6 |
| $\bar{x}$ | "ex bar" | the mean of the x values (in R: `mean(x)`) | 2.5 |
| $\bar{y}$ | "why bar" | the mean of the y values (in R: `mean(y)`) | 4 |
| $n$ | "en" | the number of data points (in R: `length(x)`) | 4 |
| $b_0, b_1$ | "bee-zero, bee-one" | any candidate intercept and slope we're testing | guess 1: (1, 1); guess 2: (0, 1.5) |
| $\beta_0$ | "beta zero" (or "beta naught") | the intercept: the line's height at $x = 0$ | guess 1 used 1; guess 2 used 0 |
| $\beta_1$ | "beta one" | the slope: how much $y$ rises per unit of $x$ | guess 1 used 1; guess 2 used 1.5 |
| $\hat{y}_i$ | "why hat sub eye" | the line's prediction for observation i — the hat means "estimated, not observed" | guess 1 predicted 2, 3, 4, 5 |
| $e_i$ | "ee sub eye" / "residual i" | the miss: $y_i - \hat{y}_i$, actual minus predicted | guess 1: 0, 0, +1, +1 |
| $\mathrm{SSE}$ | "sum of squared errors" | the line's badness score: $\sum e_i^2$ | 2 for guess 1; 0.5 for guess 2 |
| $\arg\min_{b_0,\, b_1}$ | "arg min over bee-zero and bee-one" | the VALUES of $b_0, b_1$ that minimize what follows (not the minimum itself) | the best line's intercept & slope |
| $\hat{\beta}_0, \hat{\beta}_1$ | "beta-zero hat, beta-one hat" | the winning intercept and slope, estimated from data | found in the next module! |

(Convention: $\beta$ for the unknown true values, $b$ for "any candidate we're
trying", and hats for "the estimate we computed from data". Papers lean on this
Greek/Latin/hat code constantly — same idea as $\mu$ vs. $\bar{x}$.)

## The formal version

A candidate line and its predictions:

$$
\hat{y}_i = b_0 + b_1 x_i
$$

Reading it aloud: "why-hat sub eye equals bee-zero plus bee-one times ex-sub-eye" —
*the prediction for plant i is: start at the intercept, climb the slope once per
scoop.* For guess 1: $\hat{y}_3 = 1 + 1 \times 3 = 4$ — row 3 of our table.

The score of that line:

$$
\mathrm{SSE}(b_0, b_1) = \sum_{i=1}^{n} \big(y_i - (b_0 + b_1 x_i)\big)^2
$$

Reading it aloud: "the SSE of a candidate line equals the sum, from i equals 1 to n,
of: actual y minus the line's prediction, squared." That's the miss² column, summed —
our 2 and our 0.5. Note SSE is written as a *function of* $(b_0, b_1)$: every
candidate line gets its own score.

And the decision rule — the definition of least squares:

$$
(\hat{\beta}_0, \hat{\beta}_1) = \arg\min_{b_0,\, b_1} \; \sum_{i=1}^{n} \big(y_i - (b_0 + b_1 x_i)\big)^2
$$

Reading it aloud: "beta-zero hat and beta-one hat are the arg min, over all
candidate intercepts and slopes, of the sum of squared errors" — *the best line is
whichever candidate scores lowest.* This single line of notation IS "ordinary least
squares" (OLS). When a paper says "we fit the model by OLS", it means exactly this
game.

## Explore it

```{webr}
#| autorun: true
#| fig-alt: "Scatter plot of the toy data with a candidate line and its residual segments drawn as dashed verticals"
# explore-1 — LIVE: edit b0 and b1 below, press Run, watch the line move, the residual
# segments stretch, and the SSE in the title update. Can you beat 0.5?
# PREDICT: at b0 = 0, will raising the slope from b1 = 1.5 to 1.6 make SSE go up or
# down? (Our line scores 0.5 right now.) Then edit and run.
x <- c(1, 2, 3, 4)
y <- c(2, 3, 5, 6)
b0 <- 0; b1 <- 1.5                      # <-- your candidate line

pred <- b0 + b1 * x                     # y-hat: the line's predictions
sse  <- sum((y - pred)^2)               # the score

plot(x, y, pch = 16, cex = 1.5, xlim = c(0, 5), ylim = c(0, 7),
     main = sprintf("y-hat = %.2f + %.2f x   |   SSE = %.3f", b0, b1, sse))
abline(a = b0, b = b1, col = "steelblue", lwd = 2)     # the candidate line
segments(x, y, x, pred, col = "firebrick", lty = 2)    # the residuals, drawn
```

```{webr}
#| autorun: true
#| fig-alt: "Curve of SSE versus candidate slope, U-shaped with a single lowest point"
# explore-2 — LIVE: instead of guessing one line at a time, score MANY slopes at once
# (intercept fixed at 0.5 for now) and plot the scores. The best slope is the
# bottom of the valley. TRY IT: change the by= step or the fixed intercept and re-run.
# PREDICT: will the valley bottom land left or right of slope = 1.5? Write your guess,
# then run and read it off the plot title.
x <- c(1, 2, 3, 4)
y <- c(2, 3, 5, 6)

slopes <- seq(0, 3, by = 0.01)                    # 301 candidate slopes
sse_of <- function(b1) sum((y - (0.5 + b1 * x))^2)  # score each with b0 = 0.5
scores <- sapply(slopes, sse_of)

plot(slopes, scores, type = "l", lwd = 2,
     xlab = "candidate slope b1", ylab = "SSE",
     main = sprintf("Valley bottom at b1 = %.2f", slopes[which.min(scores)]))
abline(v = slopes[which.min(scores)], col = "firebrick", lty = 2)
```

That smooth valley with a single bottom is the key picture of this course. "Find the
arg min" means "find the bottom of this valley" — and next module, calculus finds it
*exactly*, without trying 301 candidates.

## Common traps

**Trap 1 — scoring a line by $\sum e_i$ instead of $\sum e_i^2$.** Adding up the raw
misses feels simpler than squaring first. But raw misses cancel: the flat line at
$\bar{y} = 4$ has misses $-2, -1, +1, +2$, which sum to **0** — a "perfect" score for
a line that ignores fertilizer entirely. Squaring stops the cancelling, so that same
useless line scores $4 + 1 + 1 + 4 = 10$, honestly bad. This is the same reason
variance squares its deviations. (`sum(y - 4)` gives 0; `sum((y - 4)^2)` gives 10.)

**Trap 2 — expecting the best line to score $\mathrm{SSE} = 0$.** A score of 0 would
mean every point sits exactly on the line. Our four points don't lie on a single
straight line (going 1→2→3→4 in scoops, tomatoes jump 2, 3, 5, 6 — not by a constant
step), so **no** line can hit them all: the valley bottom in the plot above sits
*above* zero, not at it. "Best" means lowest achievable score, not zero — real data
keeps some irreducible scatter no straight line removes.

## Check yourself

**1.** Read this aloud and say what it equals for guess 1 (intercept 1, slope 1):
$\;\hat{y}_4 = b_0 + b_1 x_4$

::: {.callout-note collapse="true" title="Answer"}
"Why-hat sub four equals bee-zero plus bee-one times ex-sub-four." With $b_0 = 1$,
$b_1 = 1$, $x_4 = 4$: $\hat{y}_4 = 1 + 1 \times 4 = 5$. The actual $y_4$ is 6, so
the residual is $e_4 = 6 - 5 = +1$ — the line under-predicts plant 4 by one tomato.
:::

**2.** A friend proposes scoring lines with $\sum e_i$ (no squaring). Using guess
"intercept 4, slope 0" (a flat line at the mean of y), show why that's a broken
score.

```{webr}
#| autorun: false
# check2-starter — LIVE: edit and Run to compare sum(y - 4) with sum((y - 4)^2).
x <- c(1, 2, 3, 4); y <- c(2, 3, 5, 6)
```

::: {.callout-note collapse="true" title="Answer"}
The flat line predicts 4 for everyone. Misses: $2-4 = -2$, $3-4 = -1$, $5-4 = +1$,
$6-4 = +2$. Un-squared sum: $-2 - 1 + 1 + 2 = 0$ — a "perfect" score for a line
that ignores x entirely! (Deviations from a mean always cancel — the same fact we
met in the foundations module.) Squaring stops the cancelling: the squared score is
$4 + 1 + 1 + 4 = 10$, honestly bad. `sum(y - 4)` vs `sum((y - 4)^2)` confirms.
:::

**3.** In two sentences, explain to a colleague what SSE measures and why the misses
are squared instead of just added.

::: {.callout-note collapse="true" title="Answer"}
SSE adds up the squared gaps between each data point and the line's prediction, so a
larger SSE means a worse-fitting line. We square the gaps because raw gaps can be
positive or negative and would cancel each other out, giving a misleadingly low score
to a line that misses badly in both directions.
:::

## What you can now do

- **Score any candidate line** on a dataset by summing its squared residuals (SSE),
  by hand and with `sum((y - pred)^2)` in R.
- **Explain why the misses are squared** — that raw misses cancel, and that squaring
  punishes big misses hardest — instead of just added.
- **State what "least squares" means**: the best line is the intercept/slope pair
  with the lowest SSE (the arg-min), i.e. the bottom of the SSE valley.
- **Recognize the phrase "fit by OLS"** in a paper as exactly this scoring game.

If any of these feels shaky, revisit [Getting a feel for it](#getting-a-feel-for-it)
and [The formal version](#the-formal-version) before continuing — the next module
finds that valley bottom with calculus and assumes all of the above.

## Where this goes next

We can *score* any line, and we saw the valley has one bottom. Next module finds the
bottom exactly: [deriving the OLS formulas](../02-deriving-the-ols-formulas/lesson.qmd) —
with every algebra and calculus trick taught on the way. Ready for more reps first?
Do the [practice problems](practice.qmd).
