---
title: "Practice: Lines and the least-squares idea"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — each starter below re-states its own data. Base R throughout.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

Work these in order — they get harder. Each **Starter** cell is live: edit it and press
Run to solve right here in the page (WebR boots on the first run — give it a moment).
Try honestly first; the worked answer's output stays baked in, so it's there the moment
you open the callout.

## Problem 1 — the toy example, new numbers

Three plants: $x = (1, 2, 3)$ scoops, $y = (3, 4, 6)$ tomatoes. Score the candidate
line "intercept 2, slope 1" by hand: predictions, residuals, SSE. Check in R.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
x <- c(1, 2, 3); y <- c(3, 4, 6)
```

::: {.callout-note collapse="true" title="Answer"}
Predictions $\hat{y}_i = 2 + 1 \cdot x_i$: 3, 4, 5. Residuals $y_i - \hat{y}_i$:
$3-3 = 0$, $4-4 = 0$, $6-5 = +1$. Squared and summed: $0 + 0 + 1 = 1$. SSE = **1**.

```{r p1-answer}
x <- c(1, 2, 3); y <- c(3, 4, 6)
pred <- 2 + 1 * x        # the line's predictions
sum((y - pred)^2)        # SSE = 1
```
:::

## Problem 2 — faded worked example (score a line)

New data: $x = (0, 1, 2)$ scoops, $y = (1, 4, 4)$ tomatoes; candidate line "intercept
1, slope 2". The scoring is worked out below, but the **core step is blanked** — fill
in ONLY that line.

> Predictions $\hat{y}_i = 1 + 2 x_i$: at $x = 0,1,2$ → $1, 3, 5$.
> Residuals $y_i - \hat{y}_i$: $1-1 = 0$, $\;4-3 = +1$, $\;4-5 = -1$.
> **`<your step here>`** — square each residual and add them up:
> $\mathrm{SSE} = \underline{\phantom{00}}$.

```{webr}
#| autorun: false
# p2-starter — edit and Run to check the blank
x <- c(0, 1, 2); y <- c(1, 4, 4)
```

::: {.callout-note collapse="true" title="Answer"}
The blanked step is the move that defines the score — square the residuals so they
can't cancel, then sum: $0^2 + 1^2 + (-1)^2 = 0 + 1 + 1 = 2$. So $\mathrm{SSE} = 2$.

```{r p2-answer}
x <- c(0, 1, 2); y <- c(1, 4, 4)
pred <- 1 + 2 * x        # 1 3 5
sum((y - pred)^2)        # 0 + 1 + 1 = 2
```
:::

## Problem 3 — spot the error

A colleague scores the candidate line "intercept 4, slope 0" (a flat line) on the
tomato data $x = (1,2,3,4)$, $y = (2,3,5,6)$. Their work contains **exactly one
mistake** — find it, and give the correct score.

> "The flat line predicts 4 for every plant. Misses: $2-4=-2$, $3-4=-1$, $5-4=+1$,
> $6-4=+2$. Add the misses: $-2-1+1+2 = 0$. Score = 0 — a perfect fit!"

```{webr}
#| autorun: false
# p3-starter — edit and Run to check the corrected score
x <- c(1, 2, 3, 4); y <- c(2, 3, 5, 6)
```

::: {.callout-note collapse="true" title="Answer"}
The mistake is summing the **raw** misses instead of the **squared** misses. Raw
residuals always cancel around a mean-height line (overshoots balance undershoots),
so $\sum e_i = 0$ would rate this x-ignoring flat line as "perfect" — nonsense. SSE
squares first: $(-2)^2 + (-1)^2 + 1^2 + 2^2 = 4 + 1 + 1 + 4 = 10$ — honestly bad, as
it should be.

```{r p3-answer}
x <- c(1, 2, 3, 4); y <- c(2, 3, 5, 6)
pred <- rep(4, 4)
sum(y - pred)        # 0  — the broken "score"
sum((y - pred)^2)    # 10 — the real SSE
```
:::

## Problem 4 — read the paper's sentence

A paper writes: "Parameters were estimated by minimizing
$\sum_{i=1}^{n}(y_i - b_0 - b_1 x_i)^2$." In one or two plain-English sentences:
what did the authors do? (Hint: it's this module in disguise — note that
$y_i - b_0 - b_1 x_i$ is just $y_i - (b_0 + b_1 x_i)$ with the parentheses dropped.)

::: {.callout-note collapse="true" title="Answer"}
They played exactly our scoring game: for every candidate line (every intercept
$b_0$ and slope $b_1$), they computed each observation's miss — actual $y_i$ minus
the line's prediction $b_0 + b_1 x_i$ — squared the misses, summed them, and chose
the line with the smallest total. In other words: ordinary least squares. "Minimize
the sum of squared errors" is the formal way of saying "pick the line with the best
badness score."
:::

## Problem 5 — interleaved *(uses mean-variance-covariance)*

The flat line at the mean, $\hat{y}_i = \bar{y}$, has a special SSE. For
$y = (2, 3, 5, 6)$ (so $\bar{y} = 4$): compute this flat line's SSE by hand, then
compute the sample variance $s_y^2$ from the
[foundations module](../../../../foundations/mean-variance-covariance/lesson.qmd).
How are the two numbers related? (Reach back: $s_y^2 = \frac{1}{n-1}\sum(y_i-\bar{y})^2$.)

```{webr}
#| autorun: false
# p5-starter — edit and Run to solve, right here
y <- c(2, 3, 5, 6)
```

::: {.callout-note collapse="true" title="Answer"}
Flat-line residuals are exactly the deviations from the mean: $-2, -1, +1, +2$. So
$\mathrm{SSE} = 4 + 1 + 1 + 4 = 10$. The variance divides that same sum of squared
deviations by $n - 1 = 3$: $s_y^2 = 10/3 \approx 3.33$. So
$\mathrm{SSE}_{\text{flat}} = (n-1)\, s_y^2$ — the flat line's badness *is* the
spread of $y$ (times $n-1$). That is why a good regression line is judged by how far
below this number it can push the SSE: it's the "no predictor" baseline.

```{r p5-answer}
y <- c(2, 3, 5, 6); n <- length(y)
sse_flat <- sum((y - mean(y))^2)     # 10
c(sse_flat = sse_flat, var_y = var(y), n_minus_1_times_var = (n - 1) * var(y))
```
:::

## Problem 6 — the SSE surface (in R)

Module 1's valley plot fixed the intercept and varied only the slope. Now vary BOTH
with the starter below: fill in the `sse` computation, then find the grid's best
$(b_0, b_1)$ pair. Compare it with `coef(lm(y ~ x))`.

```{webr}
#| autorun: false
# p6-starter — your job: compute the SSE for every (b0, b1) pair in the grid
x <- c(1, 2, 3, 4)
y <- c(2, 3, 5, 6)
grid <- expand.grid(b0 = seq(-1, 2, by = 0.05),
                    b1 = seq(0.5, 2.5, by = 0.05))
```

::: {.callout-note collapse="true" title="Answer"}
```{r p6-answer}
x <- c(1, 2, 3, 4)
y <- c(2, 3, 5, 6)
grid <- expand.grid(b0 = seq(-1, 2, by = 0.05),
                    b1 = seq(0.5, 2.5, by = 0.05))

# For each candidate pair: predictions, misses, squared, summed
grid$sse <- apply(grid, 1, function(b) sum((y - (b["b0"] + b["b1"] * x))^2))

grid[which.min(grid$sse), ]     # best pair in the grid: b0 = 0.5, b1 = 1.4
coef(lm(y ~ x))                 # lm agrees: intercept 0.5, slope 1.4
```

The grid search and `lm()` land on the same line — because `lm()` is solving the
same arg-min, just exactly (with Module 2's formulas) instead of by brute force.
:::

## Problem 7 — a gentle cliffhanger

The valley plot had ONE bottom, and grid search found it — but only to the grid's
resolution, and it needed over 2,500 candidate evaluations. From AP Calculus, hazily:
what tool finds the exact bottom of a smooth valley in one move? (Next lesson
re-teaches it from scratch before using it.)

::: {.callout-note collapse="true" title="Answer"}
The derivative. At the bottom of a smooth valley the slope of the curve is exactly
zero — so instead of trying thousands of candidates, you take the derivative of
SSE with respect to the line's parameters, set it equal to zero, and solve. That
"take the derivative, set it to zero, solve" move is the entire engine of the next
module — and of a remarkable fraction of every methods section you'll ever read.
:::
