---
title: "Capstone — assembling Swap-Stepwise"
subtitle: "Return to the source: decode the method, build it, run it end-to-end"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)

sim_study <- function(n, seed) {
  set.seed(seed); p <- 12; R <- diag(p)
  R[1,2] <- R[2,1] <- 0.9; R[3,4] <- R[4,3] <- 0.95
  R[8,10] <- R[10,8] <- 0.7; R[11,12] <- R[12,11] <- 0.65
  X <- matrix(rnorm(n*p), n, p) %*% chol(R)
  colnames(X) <- c("rate_10y","rate_2y","cpi_infl","pce_infl","unemp","oil",
                   "usd_index","sentiment","housing","pmi","credit_spread","vix")
  beta <- numeric(p); beta[c(1,3,5)] <- c(0.5,-0.4,-0.3)
  list(X = X, y = drop(X %*% beta + rnorm(n)))
}
```

## Where we're going

No new concepts — this is the proof you reached the goal. We return to the method's own
description, decode every claim in it and point each to the module that built it, then
assemble the five pieces into one small base-R function, `swap_stepwise()`, and run it
end-to-end on a simulated multi-study meta-analysis — comparing its answer to a full-data
oracle. If you can read the method spec and rebuild it, you've reached the course goal.

**Builds on:** every module — the summary reconstruction
([M1](../01-joint-fit-from-summaries/lesson.qmd)), the switching search
([M2](../02-the-switching-search/lesson.qmd)), the rank-one updates
([M3](../03-rank-one-speedups/lesson.qmd)), and the ridge $\lambda$
([M4](../04-lambda-stability/lesson.qmd)).

## Warm-up: do you remember?

**1.** *(M1)* From a correlation matrix $R$ and marginal coefficients $b$, how do you get the
joint fit and its $R^2$?

::: {.callout-note collapse="true" title="Answer"}
$\hat\beta = R^{-1}b$ and $R^2 = b^\top R^{-1}b$. See [M1](../01-joint-fit-from-summaries/lesson.qmd).
:::

**2.** *(M2)* Name the three moves the switching search tries at each step, and its stopping
rule.

::: {.callout-note collapse="true" title="Answer"}
Add, remove, and swap; stop when no move lowers **BIC**. See
[M2](../02-the-switching-search/lesson.qmd).
:::

**3.** *(M3, M4)* What makes each search move cheap, and what keeps the reconstruction stable
when two indicators are near-identical?

::: {.callout-note collapse="true" title="Answer"}
Rank-1 (Sherman–Morrison / bordering) **updates** of $R_S^{-1}$ make each move $O(k^2)$
instead of re-inverting; a ridge **$\lambda$** on $R$'s diagonal keeps it invertible and the
coefficients stable. See [M3](../03-rank-one-speedups/lesson.qmd),
[M4](../04-lambda-stability/lesson.qmd).
:::

## Return to the source

Here is the method, quoted from the course's own specification. We decode it claim by claim,
linking each to the module that rebuilt it.

> **THE METHOD:** A frequentist, likelihood-based procedure for variable selection in linear
> regression. Two settings matter: the ordinary setting where you have the full data table,
> and a META-ANALYSIS setting where you must combine results across many published studies
> and only have each indicator's reported one-at-a-time (marginal) regression coefficient
> plus a known correlation matrix among the indicators — never the raw rows. A key skill is
> reconstructing the JOINT multiple-regression fit from those two ingredients alone.

*Decode.* "Frequentist, likelihood-based" — it scores models by the Gaussian log-likelihood
(no priors), through BIC. "Marginal coefficient + correlation matrix → joint fit" is the
identity $\hat\beta = R^{-1}b$ from **[M1](../01-joint-fit-from-summaries/lesson.qmd)**: in the
standardized world $R$ is $X^\top X/n$ and $b$ is $X^\top y/n$, so $R^{-1}b$ *is* the joint
least-squares fit — computable from summaries, "never the raw rows."

> - Because predictors are correlated, marginal effects != joint effects; the method recovers
>   the joint effects.

*Decode.* The whole reason a lone coefficient can mislead — **Multiple Regression Module 3's**
marginal-vs-joint story — and $R^{-1}$ is what converts one into the other
([M1](../01-joint-fit-from-summaries/lesson.qmd)).

> - It searches models by iteratively SWITCHING predictors in and out (not just forward
>   stepwise), evaluating the full joint likelihood at each step.

*Decode.* The **[M2](../02-the-switching-search/lesson.qmd)** switching search: add / remove /
swap, each candidate scored by the joint likelihood (via BIC) computed from summaries — the
"switch" move is what beats plain forward stepwise.

> - It chooses model size with an information criterion (BIC) rather than a p-value threshold.

*Decode.* BIC from the [model-selection foundation](../../../../foundations/model-selection-and-information-criteria/lesson.qmd),
used as the stopping rule; **[M2](../02-the-switching-search/lesson.qmd)** explained why a
post-selection p-value threshold would be dishonest.

> - It stays fast by using rank-1 (Sherman–Morrison) updates to avoid re-inverting the
>   correlation matrix each time a predictor is added or removed.

*Decode.* **[M3](../03-rank-one-speedups/lesson.qmd)**: the bordering/partition updates grow
and shrink $R_S^{-1}$ in $O(k^2)$, exactly, so the search never re-inverts.

> - It regularizes the correlation matrix (ridge-style, a lambda parameter) so it inverts
>   stably when predictors are near-collinear.

*Decode.* **[M4](../04-lambda-stability/lesson.qmd)**: replace $R$ with $R + \lambda I$,
invertible for any $\lambda > 0$, taming the near-collinear blow-up.

Every claim maps to a piece you built. Now assemble them.

## The assembled procedure

The whole method is a short base-R suite. Each function is one module's idea; `swap_stepwise()`
wires them together.

```{r suite}
# --- Swap-Stepwise: likelihood-based variable selection from summaries (R, b, n, lambda) ---

# M1 + M4: subset BIC from summaries, with a ridge lambda on the correlation sub-matrix.
ss_bic <- function(R, b, n, cols, lambda = 0) {
  if (length(cols) == 0) return(0)
  RS <- R[cols, cols, drop = FALSE] + lambda * diag(length(cols))
  bS <- b[cols]
  r2 <- min(as.numeric(crossprod(bS, solve(RS, bS))), 1 - 1e-9)   # M1: b' R^{-1} b
  n * log(1 - r2) + length(cols) * log(n)                          # BIC (up to a constant)
}

# M2: the switching search — add / remove / swap, stop when no move lowers BIC.
ss_search <- function(R, b, n, lambda = 0) {
  p <- length(b); cur <- integer(0); cur_bic <- ss_bic(R, b, n, cur, lambda)
  repeat {
    mv <- list()
    for (j in setdiff(1:p, cur)) mv <- c(mv, list(sort(c(cur, j))))                 # add
    for (j in cur)               mv <- c(mv, list(setdiff(cur, j)))                 # remove
    for (j in cur) for (k in setdiff(1:p, cur)) mv <- c(mv, list(sort(c(setdiff(cur, j), k))))  # swap
    if (!length(mv)) break
    bx <- sapply(mv, function(m) ss_bic(R, b, n, m, lambda)); i <- which.min(bx)
    if (bx[i] < cur_bic - 1e-9) { cur <- mv[[i]]; cur_bic <- bx[i] } else break
  }
  sort(cur)
}

# The whole method: select a subset, then reconstruct its ridge-stabilised joint coefficients.
swap_stepwise <- function(b, R, n, lambda = 0) {
  S <- ss_search(R, b, n, lambda)
  beta <- setNames(numeric(length(b)), names(b))
  if (length(S)) beta[S] <- solve(R[S, S, drop = FALSE] + lambda * diag(length(S)), b[S])  # M1 + M4
  list(selected = S, beta = beta)
}
```

(The $O(k^2)$ rank-1 updates from [M3](../03-rank-one-speedups/lesson.qmd) would slot into
`ss_search` to avoid the `solve()` inside `ss_bic` — omitted here for readability, since for
$p = 12$ the plain solves are already instant; they matter when $p$ is large.)

## Explore it — end-to-end on a simulated meta-analysis

Five "published studies" each report marginal coefficients for the same twelve indicators
(different sample sizes, no shared raw data). We inverse-variance-average them into one $b$,
take $R$ from a reference panel, and run `swap_stepwise()` — then compare to a **full-data
oracle** that pools every study's raw rows and does exhaustive best-subset selection.

```{r meta-end-to-end}
ns <- c(120, 200, 90, 150, 180); seeds <- 101:105       # five studies
B <- SE <- matrix(NA, 5, 12); allX <- NULL; allY <- NULL
for (i in 1:5) {
  s <- sim_study(ns[i], seeds[i])
  B[i, ]  <- as.numeric(cor(s$X, s$y))                  # study i's marginal std coefs
  SE[i, ] <- (1 - B[i, ]^2) / sqrt(ns[i])               # approx SE of each correlation
  allX <- rbind(allX, s$X); allY <- c(allY, s$y)        # (oracle only) pool raw rows
}
w <- 1 / SE^2
b_combined <- setNames(colSums(w * B) / colSums(w), colnames(allX))   # meta: precision-weighted b
R_ref <- cor(sim_study(1000, 999)$X)                                  # reference-panel correlations
N <- sum(ns)

fit <- swap_stepwise(b_combined, R_ref, N, lambda = 0.05)   # <- summaries only, never allX/allY
names(fit$beta)[fit$selected]                            # selected indicators
round(fit$beta[fit$selected], 3)                         # their joint coefficients
```

```{r meta-oracle}
# Full-data oracle: pool the raw rows, exhaustive best-subset over BIC(lm).
exhaustive <- function(X, y) {
  p <- ncol(X); best <- integer(0); best_bic <- BIC(lm(y ~ 1))
  for (k in 1:p) { cb <- combn(p, k)
    for (i in seq_len(ncol(cb))) { v <- BIC(lm(y ~ X[, cb[, i], drop = FALSE]))
      if (v < best_bic) { best_bic <- v; best <- cb[, i] } } }
  sort(best)
}
oracle_set <- exhaustive(allX, allY)
list(meta_selected   = colnames(allX)[fit$selected],
     oracle_selected = colnames(allX)[oracle_set],
     sets_match      = identical(fit$selected, oracle_set))
```

`swap_stepwise()`, working only from per-study summaries and a reference correlation matrix,
selects the same indicators the pooled full-data oracle does — the true
`{rate_10y, cpi_infl, unemp}`. The joint coefficients are close (not identical): the small gap
is the price of combining summaries instead of raw rows, plus the $\lambda$ shrinkage — an
honest reconstruction, not magic.

```{webr}
#| autorun: true
# TRY IT: change lambda (0, 0.05, 0.5) or the studies' seeds and re-run swap-stepwise.
# PREDICT: does the selected set stay {rate_10y, cpi_infl, unemp} as lambda changes?
ss_bic <- function(R,b,n,cols,lambda=0){ if(!length(cols)) return(0)
  RS <- R[cols,cols,drop=FALSE]+lambda*diag(length(cols))
  r2 <- min(as.numeric(crossprod(b[cols], solve(RS, b[cols]))), 1-1e-9)
  n*log(1-r2)+length(cols)*log(n) }
ss_search <- function(R,b,n,lambda=0){ p<-length(b); cur<-integer(0); cb<-ss_bic(R,b,n,cur,lambda)
  repeat{ mv<-list(); for(j in setdiff(1:p,cur)) mv<-c(mv,list(sort(c(cur,j))))
    for(j in cur) mv<-c(mv,list(setdiff(cur,j)))
    for(j in cur) for(k in setdiff(1:p,cur)) mv<-c(mv,list(sort(c(setdiff(cur,j),k))))
    if(!length(mv))break; bx<-sapply(mv,function(m)ss_bic(R,b,n,m,lambda)); i<-which.min(bx)
    if(bx[i]<cb-1e-9){cur<-mv[[i]];cb<-bx[i]}else break }; sort(cur) }

# One study's summaries as a quick stand-in (b = marginal std coefs, R = its correlations)
set.seed(101); p<-12; Rt<-diag(p); Rt[1,2]<-Rt[2,1]<-.9; Rt[3,4]<-Rt[4,3]<-.95
Rt[8,10]<-Rt[10,8]<-.7; Rt[11,12]<-Rt[12,11]<-.65
X<-matrix(rnorm(200*p),200,p)%*%chol(Rt); bt<-numeric(p); bt[c(1,3,5)]<-c(.5,-.4,-.3)
y<-drop(X%*%bt+rnorm(200)); R<-cor(X); b<-as.numeric(cor(X,y))
ss_search(R, b, 200, lambda = 0.05)      # the selected indicator indices
```

## Common traps

**Trap: "the meta-analysis reconstruction should exactly equal the pooled-data fit."** The
*selected set* matches the oracle here, but the coefficients are only close: averaging
published marginals (with a reference-panel $R$) and adding a stabilizing $\lambda$ is an
approximation to having every raw row. It recovers the right model and approximately the right
effects — honest, and often all the raw data would let you conclude anyway.

**Trap: "Swap-Stepwise found a set, so those indicators are *the* true causal drivers."** It found
the subset that best trades fit against complexity under BIC, given correlated indicators. Under
heavy collinearity (Module 4's twins), a different seed can select the *other* twin — as the
equivalence harness's collinear fixture shows, matching the oracle both times. Selection tells
you which indicators best *predict* jointly, not which are causal.

## What you can now do

- **Decode the method spec** claim by claim and name the module (and formula) behind each.
- **Reconstruct a joint fit from summaries** ($R^{-1}b$), **search** subsets by switching under
  BIC, **speed** it with rank-1 updates, and **stabilize** it with a ridge $\lambda$ — the five
  pieces of Swap-Stepwise.
- **Assemble `swap_stepwise(b, R, n, lambda)`** and run it end-to-end on a multi-study
  meta-analysis, validating the selected set against a full-data oracle.
- **State the honest limits**: it's a greedy (heuristic) search, coefficients from summaries are
  approximate, and selection under collinearity is about prediction, not causation.

If any bullet feels shaky, revisit the module it points to — these can-dos *are* the course goal.

## Where this goes next

**What you still can't do (yet).** Swap-Stepwise selects and reconstructs, but it doesn't give you
*honest* post-selection confidence intervals for the chosen coefficients — that needs
post-selection inference methods this course only named. Other natural next steps: the **lasso**
(selection by an $\ell_1$ penalty, from the [ridge foundation](../../../../foundations/ridge-regularization/lesson.qmd)),
cross-validation to *choose* $\lambda$, and time-series structure in the indicators (our
simulator is deliberately iid — real economic series are autocorrelated). Ready for a final
workout? Do the cumulative [practice problems](practice.qmd).
