---
title: "Practice: Capstone — assembling Swap-Stepwise"
subtitle: "Cumulative. No labels saying which tool — recalling that is part of the test."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
set.seed(42)
```

These draw on the whole course, in mixed order — deciding *which* idea each problem needs is
part of the workout. Try each honestly before opening the answer.

## Problem 1 — warm-up reconstruction

From $R = \begin{pmatrix}1 & 0.3\\0.3 & 1\end{pmatrix}$ and marginal coefficients $b = (0.5, 0.4)$,
reconstruct the joint coefficients.

```{webr}
#| autorun: false
# p1-starter
R <- matrix(c(1, 0.3, 0.3, 1), 2); b <- c(0.5, 0.4)
```

::: {.callout-note collapse="true" title="Answer"}
$\hat\beta = R^{-1}b \approx (0.418, 0.275)$. With only mild correlation (0.3), the joint
coefficients are close to the marginals but discounted a little for the overlap.

```{r}
solve(matrix(c(1, 0.3, 0.3, 1), 2), c(0.5, 0.4))     # 0.418, 0.275
```
:::

## Problem 2 — score a subset

Using $\mathrm{BIC}(S) = n\log(1 - b_S^\top R_S^{-1}b_S) + |S|\log n$, compute the BIC of the
single-indicator model with $b_1 = 0.5$ (so $R_S = 1$) at $n = 80$.

::: {.callout-note collapse="true" title="Answer"}
$b_S^\top R_S^{-1} b_S = 0.5^2 = 0.25$, so $\mathrm{BIC} = 80\log(0.75) + \log(80) = -23.01 + 4.38 = -18.63$
(up to the additive constant common to all models on this data).
:::

## Problem 3 — spot the error

A report reads: *"Swap-Stepwise selected `unemp`, `cpi_infl`, and `rate_10y`; each has $p < 0.01$ in
the final fit, and the model's high $R^2$ confirms these are the true causal drivers of GDP."*
This contains **more than one** questionable claim — name the two biggest.

::: {.callout-note collapse="true" title="Answer"}
(1) **Post-selection p-values**: the $p < 0.01$ figures come from a model chosen *by searching*,
so they're optimistically small and not honest evidence. (2) **Selection ≠ causation**: Swap-Stepwise
finds the subset that best predicts jointly under BIC; a high $R^2$ and being selected say nothing
about causality — under collinearity it could even have picked a correlated stand-in (a noise twin)
for a true driver. (Bonus: $R^2$ measures fit, not the trustworthiness of individual coefficients.)
:::

## Problem 4 — the Schur complement warns

While growing the selected set, the bordering update reports a Schur complement of $s = 0.004$ for
the next indicator. What does that mean, and what should the method do?

::: {.callout-note collapse="true" title="Answer"}
$s$ is the candidate indicator's leftover variance after those already selected explain it;
$s = 0.004$ means it's almost a linear combination of them (near-collinear), so dividing by $s$ in
the update would produce a huge, unstable inverse. The method should add (or increase) the ridge
$\lambda$ on the correlation matrix, which lifts the small eigenvalue and keeps the inversion — and
the reconstructed coefficients — stable.
:::

## Problem 5 — stabilize a blow-up

Two indicators correlate 0.99 with marginals $b = (0.6, 0.5)$ and
$R = \begin{pmatrix}1 & 0.99\\0.99 & 1\end{pmatrix}$. Show the unregularized reconstruction blows
up, and that $\lambda = 0.1$ tames it.

```{webr}
#| autorun: false
# p5-starter
R <- matrix(c(1, 0.99, 0.99, 1), 2); b <- c(0.6, 0.5)
```

::: {.callout-note collapse="true" title="Answer"}
Unregularized: $R^{-1}b \approx (5.28, -4.72)$ — enormous, opposite signs. With $\lambda = 0.1$:
$(R + 0.1I)^{-1}b \approx (0.72, -0.19)$ — brought back to a sane range.

```{r}
R <- matrix(c(1, 0.99, 0.99, 1), 2); b <- c(0.6, 0.5)
rbind(unregularized = solve(R, b), ridge_0.1 = solve(R + 0.1 * diag(2), b))
```
:::

## Problem 6 — interpret the meta-analysis result

Running `swap_stepwise()` on five studies' summaries selected the *same* indicators as a full-data
oracle that pooled every study's raw rows, but the joint coefficients differed by up to ~0.05.
Is that a failure? Explain in two or three sentences.

::: {.callout-note collapse="true" title="Answer"}
No — it's the expected, honest outcome. Combining published marginal coefficients (with a
reference-panel correlation matrix) plus a stabilizing $\lambda$ is an *approximation* to having
every raw row, so the selected set can match exactly while the coefficients are only close. The
method recovered the right model and approximately the right effects without ever seeing the raw
data — which is the whole point of the meta-analysis setting.
:::

## Problem 7 — decode a fresh claim

Here is a sentence from a different method's description you haven't seen:

> "To keep the search tractable on hundreds of predictors, we avoid recomputing the fit from
> scratch after each single-variable change, and we penalize model size with a criterion that
> grows its per-parameter cost with the sample size."

Name the two Swap-Stepwise ideas this describes, and give the formula or move behind each.

::: {.callout-note collapse="true" title="Answer"}
(1) "Avoid recomputing the fit after each single-variable change" = **rank-1 / bordering updates**
(Sherman–Morrison): grow/shrink $R_S^{-1}$ with an $O(k^2)$ update instead of re-inverting. (2)
"A criterion whose per-parameter cost grows with the sample size" = **BIC**, $-2\log L + k\log n$
(the penalty $\log n$ grows with $n$, unlike AIC's fixed $2k$). Both are pieces you assembled into
`swap_stepwise()`.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="courses/swap-stepwise-selection/modules/05-capstone-swap-stepwise" pass="4"}

::: {.quiz-q qid="cap-reconstruct-piece" qtype="mc" answer="a"}
In `swap_stepwise`, which piece turns the marginal coefficients $b$ and correlation matrix $R$
into joint coefficients?

a) $\hat\beta = R^{-1}b$, the summary reconstruction
b) The add / remove / swap search
c) The ridge $\lambda$ on the diagonal
:::

::: {.quiz-q qid="cap-swap-vs-forward" qtype="mc" answer="c"}
`swap_stepwise` beats plain forward stepwise mainly because its search can also:

a) stop using a $p < 0.05$ threshold
b) invert the correlation matrix exactly
c) remove and swap predictors, not only add
:::

::: {.quiz-q qid="cap-meta-approx" qtype="mc" answer="b"}
Run on combined study summaries, `swap_stepwise` selects the same set as a full-data oracle,
but its coefficients differ slightly. Why?

a) The search selected the wrong subset entirely
b) Combining marginals only approximates having the raw data
c) The full-data oracle made a rounding error
:::

::: {.quiz-q qid="cap-not-causal" qtype="mc" answer="a"}
`swap_stepwise` selects three indicators. What does that support?

a) That they jointly predict $y$ well — not that they cause it
b) That they are the true causal drivers of $y$
c) That no other subset predicts $y$ comparably
:::

::: {.quiz-q qid="cap-ridge-eigenvalue" qtype="numeric" answer="0.021" tolerance="0.001"}
`swap_stepwise` stabilises a near-singular $R$ by adding $\lambda$ down the diagonal. If $R$'s
smallest eigenvalue is $0.001$ and $\lambda = 0.02$, what is the smallest eigenvalue of
$R + \lambda I$? (Enter 3 decimals.)
:::

:::

::::
