---
title: "Distributions and densities"
subtitle: "Which values are common vs rare — the bell curve, why AREA is probability, and R's d/p/q/r question-askers"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — a live `{webr}` cell does NOT see anything defined here, so
# every teaching cell below re-states its own data. Base R is all we need (dnorm/pnorm/
# qnorm/rnorm and hist all ship with R), so nothing to install.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```


## Where we're going

A **distribution** is the answer to one plain question: *across all the values a
quantity can take, which are common and which are rare?* Measure a hundred commute
times, a thousand test scores, ten thousand coin-flip counts — pile them up and a
shape emerges. That shape *is* the distribution. This module turns that pile-up into
something you can compute with: a smooth curve, the famous **bell curve** (the
*normal* distribution), and the four everyday questions R lets you ask any
distribution with its `dnorm`/`pnorm`/`qnorm`/`rnorm` family.

The one idea that trips everyone up — and the one this module drills — is that for a
smooth curve **the AREA under the curve is a probability, but the HEIGHT of the curve
is not.** Get that straight and the rest is bookkeeping.

**Builds on:** [mean, variance & covariance](../mean-variance-covariance/lesson.qmd)
— the normal curve's two knobs are exactly the mean (center) and the standard
deviation (spread) you already know; [probability, odds & log-odds](../probability-and-odds/lesson.qmd)
— probability as a number in $[0,1]$; and [the derivative as a slope](../derivative-as-slope/lesson.qmd)
— soft, for the companion "area under a curve" picture of an integral we lean on when
we say area = probability. This module does **not** cover likelihood or
sampling-distributions/standard-error — those are separate modules; we only point at
them.

## Warm-up: do you remember?

Before the new stuff, three quick reps — answers from memory first, then check.

**1.** For `x <- c(28, 30, 32, 29, 31, 30)`, what is the **mean** $\bar{x}$? Say the
recipe in words, then give the number.

```{webr}
#| autorun: false
# warmup1-starter — edit and Run to check
x <- c(28, 30, 32, 29, 31, 30)
```

::: {.callout-note collapse="true" title="Answer"}
"Add them all up and divide by how many": $28+30+32+29+31+30 = 180$, and $180 / 6 =
\mathbf{30}$. In R that's `mean(x)`. Shaky? Revisit
[mean, variance & covariance](../mean-variance-covariance/lesson.qmd).
:::

**2.** A probability is a number that must live between two bounds — which two? And in
plain words, what does a probability of $0.68$ mean?

::: {.callout-note collapse="true" title="Answer"}
Every probability sits in $[0, 1]$ (0 = impossible, 1 = certain). A probability of
$0.68$ means "about 68 times out of 100" — roughly two-thirds of the time. Revisit
[probability, odds & log-odds](../probability-and-odds/lesson.qmd) if that felt slow.
:::

**3.** You have the *idea* of an integral from calculus: if you draw a curve and shade
the region between it and the horizontal axis, what does that shaded region's **area**
represent, loosely? (One word will do.)

::: {.callout-note collapse="true" title="Answer"}
**Area.** The integral of a curve between two points is just "the area under it,
between those two points." We use nothing more mechanical than that all lesson — the
companion to the slope picture in
[the derivative as a slope](../derivative-as-slope/lesson.qmd). If the word "integral"
scares you, mentally replace it with "area under the curve" every time.
:::

## A tiny example first

You clock your morning commute on **six** days. In minutes:

$$
28,\quad 30,\quad 32,\quad 29,\quad 31,\quad 30.
$$

Which values are common, which are rare? Just *count how many times each shows up* —
that count is called a **frequency**:

- 28 minutes: happened **once**
- 29 minutes: **once**
- 30 minutes: **twice** (the most common — the peak)
- 31 minutes: **once**
- 32 minutes: **once**

Draw those counts as bars over a number line and you have a **histogram** — the crudest
possible picture of a distribution. The tallest bar sits at 30 (the common value); the
edges (28 and 32) are shorter (rarer). That "tall in the middle, shorter at the edges"
shape is the seed of the bell curve.

Turn the counts into **proportions** and you have probabilities: 30 minutes happened 2
times out of 6, a proportion of $2/6 \approx 0.33$; each of the others happened 1 in 6,
$\approx 0.17$. Those six proportions add to exactly 1 — the commute *had* to be one of
these values. **That "everything adds to 1" is the rule every distribution obeys.**

The same thing in R, so you can check every count:

```{r toy-example}
# Six commute times, in minutes
x <- c(28, 30, 32, 29, 31, 30)

table(x)          # frequency of each value: 30 appears twice, the rest once
mean(x)           # 30 — the balance point / center
sd(x)             # ~1.41 — the spread (standard deviation)

table(x) / length(x)   # proportions: 0.1667 ... 0.3333 ... — they sum to 1
sum(table(x) / length(x))   # 1 — a distribution's proportions always total 1
```

## Getting a feel for it

**(a) The mental picture — a crowd finding its seats.** Picture a thousand people
filing into a long, narrow stadium whose seats are labelled by value (commute time,
say). Everyone sits in the row matching their own value. Popular values fill up into
tall stacks of people; rare values stay nearly empty. Look down from above and the
skyline of stacks *is* the distribution — a rough shape you'd sketch as a curve:
high where the crowd piled up, low where it thinned out. A **histogram** is that
skyline drawn with a handful of wide bars; a **density curve** is the same skyline
after you let the crowd grow to infinity and shrink the bars to zero width, so the
jagged tops smooth into one continuous line.

Where the analogy strains: real people are countable, so a histogram bar is an honest
"how many." A smooth density curve pretends the value can be *any* real number (30.0000…
minutes), so no single exact value has a headcount anymore — which is exactly why we
have to switch from "height = how many" to "area = how much probability." That switch
is the whole point of the next beat.

**(b) Why AREA and not height.** Once the curve is smooth, ask "what's the probability
the commute is *exactly* 30.000000 minutes?" The honest answer is essentially zero —
there are infinitely many nearby values (30.0001, 29.9998, …) and no single razor-thin
one carries any real weight. So a smooth curve can't put probability *at a point*.
What it can do is put probability *over a stretch*: "between 29 and 31 minutes." And
the natural way to measure "how much curve sits over the stretch from 29 to 31" is the
**area** of that slice — base times height, summed up. Height alone tells you nothing
until you multiply it by a width to get an area. That is why the curve's height is
called a **density** (probability *per unit width*, like people-per-row), never a
probability itself — and why the total area under the whole curve is exactly 1 (all the
probability there is).

**(c) What the numbers feel like.** For our commute curve (centered at 30), the height
right at the peak is about $0.13$ — and $0.13$ is emphatically **not** "a 13% chance of
commuting exactly 30 minutes." It's a density: to turn it into a probability you'd
multiply by a width. Take a 1-minute-wide sliver around 30 (from 29.5 to 30.5): its
area is roughly height × width $\approx 0.13 \times 1 = 0.13$, and *that* 0.13 is a
genuine probability (about a 13% chance the commute lands in that one-minute window).
Widen the window and the probability grows; the whole area totals 1. Densities can even
be bigger than 1 (a very sharp, narrow peak can tower above height 1) — impossible for
a probability, and the surest proof that height is not probability. You'll see all of
this move in **Explore it**.

## What the notation is going to say

The curve we'll write down is the **normal** (bell) curve. Its formula takes a value
$x$ and returns the curve's **height** there — the density. The formula has exactly two
tuning knobs: $\mu$ (the center, where the peak sits) and $\sigma$ (the spread, how wide
the bell is). For our commute story we'll set the center to 30 and the spread to 3, and
the formula will say, in plain English: *"you're highest right at the center 30, and
you fall off — fast at first, then gently — the farther $x$ gets from 30, and how fast
you fall off is set by the spread."* Feeding in $x = 30$ (the center) gives the tallest
height; feeding in $x = 36$ (two spreads out) gives a much smaller one. We will never
read a single height as a probability — only areas between two $x$ values are that.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $x$ | "ex" | a value the quantity can take (here: a commute time) | e.g. $x = 30$ minutes |
| $\bar{x}$ | "ex bar" | the mean (average) of the data (in R: `mean(x)`) | 30 |
| $\mu$ | "mew" | the distribution's true center — the peak of the bell (in R: the `mean` argument) | 30 |
| $\sigma$ | "sigma (lowercase)" | the distribution's spread — the standard deviation, how wide the bell is (in R: the `sd` argument) | 3 |
| $\sigma^2$ | "sigma squared" | the variance — spread squared | 9 |
| $f(x)$ | "eff of ex" | the **density**: the curve's HEIGHT at $x$ — probability per unit width, NOT a probability (in R: `dnorm(x, mu, sigma)`) | $f(30) \approx 0.133$ |
| $P(a < X < b)$ | "the probability that X is between a and b" | the AREA under the curve from $a$ to $b$ — a genuine probability in $[0,1]$ | $P(27<X<33)\approx 0.68$ |
| $\int_a^b f(x)\,dx$ | "the integral from a to b of f of x dee-ex" | the area under the curve $f$ between $a$ and $b$ — exactly that probability | the shaded slice |
| $X \sim N(\mu, \sigma^2)$ | "X is distributed normal with mean mew and variance sigma squared" | $X$ follows a bell curve centered at $\mu$ with spread $\sigma$ | commute $\sim N(30, 9)$ |
| $e$ | "ee" | the constant $\approx 2.718$ (see [exponentials & logs](../exponentials-and-logs/lesson.qmd)) | inside the formula |
| $\pi$ | "pie" | the constant $\approx 3.14159$ | inside the formula |
| $q$ | "cue" | a value you ask about (a quantile / cutoff) | $q = 33$ in `pnorm(33, ...)` |
| $p$ | "pee" | a probability, a number in $[0,1]$ (here, an area to the left) | $p = 0.5$ in `qnorm(0.5, ...)` |
| $n$ | "en" | how many random draws you ask for | $n = 5$ in `rnorm(5, ...)` |

The height $f(x)$ and the area $P(a<X<b)$ are **different kinds of thing**: a height is
a density (can exceed 1), an area is a probability (never exceeds 1). Keeping those two
rows apart in your head is 90% of this module.

## The formal version

The normal (bell) curve's height at a value $x$, for center $\mu$ and spread $\sigma$,
is:

$$
f(x) = \frac{1}{\sigma\sqrt{2\pi}}\; e^{-\frac{1}{2}\left(\frac{x-\mu}{\sigma}\right)^2}
$$

You do **not** need to memorize or hand-compute this — `dnorm()` does it for you. But
it's worth reading the two pieces, because each maps onto something you already feel:

- The exponent $-\tfrac{1}{2}\left(\tfrac{x-\mu}{\sigma}\right)^2$ is the whole shape.
  $x-\mu$ is "how far $x$ is from the center"; dividing by $\sigma$ measures that
  distance *in spreads*; squaring makes it symmetric (left and right of center fall off
  the same) and the minus sign flips it upside-down so the curve is **highest at the
  center** ($x=\mu$ makes the exponent 0, and $e^0 = 1$, the peak) and drops toward 0
  as $x$ runs away from $\mu$. That squared-then-negated bump is the bell.
- The front factor $\tfrac{1}{\sigma\sqrt{2\pi}}$ is just a scaler chosen so the
  **total area under the whole curve comes out to exactly 1** — the "all probability
  adds to 1" rule, enforced. Bigger $\sigma$ shrinks the peak height (a wider bell must
  be shorter to keep the same area of 1).

Reading it aloud: "eff of ex equals one over sigma-root-two-pi, times e to the minus
one-half times (x minus mew over sigma) squared" — or in everyday words: **"start
tallest at the center, and fall off symmetrically the farther you get from it, scaled
so the whole area is 1."**

Mapping it onto the toy example ($\mu = 30$, $\sigma = 3$):

- At the center, $x = 30$: the exponent's $x-\mu$ is 0, so $e^0 = 1$ and the height is
  just the front factor — the tallest point. `dnorm(30, 30, 3)` returns $\approx
  0.133$.
- Six minutes out, $x = 36$ (two spreads away): the exponent is
  $-\tfrac{1}{2}(6/3)^2 = -2$, so the height is far smaller. `dnorm(36, 30, 3)`
  returns $\approx 0.018$ — rarer, exactly as the picture promised.

Sanity check that the height is really "per unit width," not a probability: a 1-minute
sliver around the center (29.5 to 30.5) has area $\approx$ height × width $= 0.133
\times 1 = 0.133$, and the exact area `pnorm(30.5, 30, 3) - pnorm(29.5, 30, 3)` is
$0.1324$ — essentially the same. Height times a width gives a probability; height alone
does not.

## Explore it

```{webr}
#| autorun: true
#| fig-alt: "A bell-shaped normal density curve centered at 30, with the region from 27 to 33 shaded; the shaded area is labelled 0.68, the probability."
# explore-1 — LIVE: the curve, and AREA = probability.
# TRY IT: change lo and hi (the shaded window), then Run — watch the area change.
# PREDICT: the window 27..33 is one spread (3) either side of center 30. What
# fraction of the area do you expect it to hold? Commit to a number, then run.
mu <- 30; sigma <- 3
xs <- seq(20, 40, length.out = 400)
plot(xs, dnorm(xs, mu, sigma), type = "l", lwd = 2,
     xlab = "commute time (min)", ylab = "density = height f(x)",
     main = "Normal curve: area under a slice IS the probability")

lo <- 27; hi <- 33                       # the window we want the probability for
gx <- seq(lo, hi, length.out = 200)
polygon(c(lo, gx, hi), c(0, dnorm(gx, mu, sigma), 0), col = "grey80", border = NA)
lines(xs, dnorm(xs, mu, sigma), lwd = 2) # redraw the curve on top of the shading

area <- pnorm(hi, mu, sigma) - pnorm(lo, mu, sigma)  # the shaded area = P(lo < X < hi)
text(mu, 0.05, paste0("area = ", round(area, 3)))
```

```{webr}
#| autorun: true
# explore-2 — LIVE: HEIGHT is a density, NOT a probability.
# The height at the center is > 0.13 but shrinks as you widen the spread — because a
# wider bell must be shorter to keep its total area at 1.
# PREDICT: which is a bigger HEIGHT at the center: spread 1 or spread 6? Then run.
c(height_at_center_spread1 = dnorm(30, 30, 1),   # ~0.399 — tall, narrow bell
  height_at_center_spread6 = dnorm(30, 30, 6))   # ~0.066 — short, wide bell

# A height can even EXCEED 1 (impossible for a probability) when the bell is very sharp:
dnorm(0, 0, 0.3)   # ~1.33 — a density above 1, the clearest proof height != probability
```

```{webr}
#| autorun: true
# explore-3 — LIVE: the formula from scratch vs. the built-in.
# The formal section gave the normal pdf: f(x) = 1/(sigma*sqrt(2*pi)) * exp(-0.5*((x-mu)/sigma)^2).
# Does computing that by hand match what dnorm() returns? Run and see.
x <- 30; mu <- 30; sigma <- 3
f_manual <- (1 / (sigma * sqrt(2 * pi))) * exp(-0.5 * ((x - mu) / sigma)^2)
f_builtin <- dnorm(x, mu, sigma)
c(from_formula = f_manual, from_dnorm = f_builtin)
all.equal(f_manual, f_builtin)   # TRUE — they are the same computation
```

```{webr}
#| autorun: false
# explore-4 — LIVE: the four questions you can ask a distribution (d / p / q / r).
# PREDICT before running: qnorm(0.5, 30, 3) asks "which value splits the area in half?"
# For a symmetric bell centered at 30, what must that value be? Write it, then run.
dnorm(30, 30, 3)          # d = DENSITY (height) at 30
pnorm(33, 30, 3)          # p = PROBABILITY area to the LEFT of 33 (i.e. P(X < 33))
qnorm(0.5, 30, 3)         # q = QUANTILE: the value with 0.5 of the area to its left
set.seed(42)
rnorm(5, 30, 3)           # r = RANDOM draws from the distribution
```

**The four questions, in one breath.** Every distribution in R ships as a family of
four functions sharing one root name (`norm` for normal). They are the four things you
can ask:

- **`dnorm(x, ...)`** — **d**ensity: "how *tall* is the curve at $x$?" (the height
  $f(x)$; not a probability).
- **`pnorm(q, ...)`** — **p**robability: "how much area lies to the *left* of $q$?"
  i.e. $P(X < q)$. To get the area of a slice $a$ to $b$, subtract:
  `pnorm(b) - pnorm(a)`.
- **`qnorm(p, ...)`** — **q**uantile: the inverse of `pnorm` — "which value has
  probability $p$ to its left?" (`qnorm(0.5)` is the median/center; `qnorm(0.975)` is
  the value with 97.5% below it).
- **`rnorm(n, ...)`** — **r**andom: "give me $n$ fresh draws from this distribution."

## A whole zoo, same grammar (name-drop only)

The normal is one curve among many. There's the **t** (a heavier-tailed cousin of the
normal), **chi-squared** (spread of squared things), **binomial** (counts of
successes in a fixed number of yes/no trials), **Poisson** (counts of rare events in a
window), and more. You don't need any of their details here. The one thing to carry
away: **they all speak the same d/p/q/r grammar** — `dbinom`/`pbinom`/`qbinom`/`rbinom`,
`dt`/`pt`/`qt`/`rt`, and so on. Learn the four questions once and you can interrogate
any distribution R knows.

## Common traps

**Trap 1 — "the curve's height at a point is the probability of that point."** It is
tempting because for a *histogram* the bar height really is a count/proportion. But a
smooth curve's height is a **density** — probability *per unit width* — and a single
exact point has zero width, hence zero probability. Numbers: `dnorm(30, 30, 3)` is
$0.133$, which is **not** "a 13.3% chance of exactly 30 minutes"; the actual
probability of any single exact value is 0. Only an *area* over a stretch is a
probability: `pnorm(30.5, 30, 3) - pnorm(29.5, 30, 3)` $= 0.132$ is the real
probability of "between 29.5 and 30.5." Height needs a width multiplied in before it
becomes a probability. (And a density can exceed 1 — `dnorm(0, 0, 0.3)` $= 1.33$ — which
no probability ever can.)

**Trap 2 — "`pnorm` gives the probability *at* the value."** `pnorm(33, 30, 3)` is
$0.841$ — that is the area to the **left** of 33 (i.e. $P(X < 33)$), the whole
cumulative pile up to 33, not the chance of landing on 33. To get the probability of a
*slice* you must subtract two `pnorm`s: $P(27 < X < 33)$ is
`pnorm(33, 30, 3) - pnorm(27, 30, 3)` $= 0.841 - 0.159 = 0.682$ (the exact R result
is $0.6827 \approx 0.683$; the tiny gap is rounding in the intermediate values).
Reading a single `pnorm` as a slice probability double-counts everything below the
lower edge.

## Check yourself

**1.** For a normal curve with center $\mu = 30$ and spread $\sigma = 3$, what does
`dnorm(30, 30, 3)` return — and is that number a probability? What single R expression
would give the *probability* that a value falls between 29.5 and 30.5?

```{webr}
#| autorun: false
# check1-starter — edit and Run to check your answer
mu <- 30; sigma <- 3
```

::: {.callout-note collapse="true" title="Answer"}
`dnorm(30, 30, 3)` returns $\approx 0.133$ — the **height** (density) at the center. It
is **not** a probability (densities can even exceed 1). The probability of the slice
29.5 to 30.5 is an **area**: `pnorm(30.5, 30, 3) - pnorm(29.5, 30, 3)` $\approx 0.132$.
The density and that slice-probability are numerically close here only because the slice
is exactly 1 unit wide (area $\approx$ height × width).
:::

**2.** `pnorm(33, 30, 3)` returns $0.841$. In one sentence, what does that number mean —
and what does it *not* mean?

::: {.callout-note collapse="true" title="Answer"}
It means "about 84.1% of the area lies to the left of 33," i.e. $P(X < 33) = 0.841$ —
the chance a value is *below* 33. It does **not** mean the probability of landing
*exactly on* 33 (that's 0), nor the probability of a slice around 33.
:::

**3.** In your own words: explain to a colleague, in two sentences, why the AREA under
a density curve is a probability but the HEIGHT is not.

::: {.callout-note collapse="true" title="A model answer"}
The curve's height is a *density* — a per-unit-width rate, like people-per-row — so it
only becomes a probability once you multiply it by a width, i.e. take the area of a
slice. A single exact value has zero width and therefore zero probability; probability
lives in the area over a *range* of values, and the whole area under the curve adds to 1.
:::

## What you can now do

- **Read a pile of numbers as a distribution** — count frequencies into a histogram,
  and picture the smooth density curve it approaches.
- **Keep density and probability straight** — say why the curve's HEIGHT is a density
  (per unit width, can exceed 1) and only the AREA under a stretch is a probability.
- **Describe a normal curve by its two knobs** — the center $\mu$ (the mean) and the
  spread $\sigma$ (the standard deviation), and predict how the curve moves when each
  changes.
- **Ask a distribution the four questions in R** — `dnorm` (height), `pnorm` (area to
  the left), `qnorm` (the value at a given left-area), `rnorm` (random draws) — and
  build a slice probability as `pnorm(b) - pnorm(a)`.
- **Recognize the wider zoo** — t, chi-squared, binomial, Poisson — as other curves
  spoken with the same d/p/q/r grammar.

If any of these feels shaky, revisit [Getting a feel for it](#getting-a-feel-for-it)
and [The formal version](#the-formal-version) before continuing — anything that reasons
about "how likely is a value" leans on all of these.

## Where this goes next

Distributions are the vocabulary of every statistical claim: a p-value, a confidence
interval, a model's assumed noise — all are statements about areas under some curve.
Two ideas built directly on this one, each its own module:
**[likelihood](../likelihood-and-log-likelihood/lesson.qmd)** (turning the
density around to ask "how well does this curve explain the data I saw?") and the
**[sampling distribution / standard error](../randomness-samples-and-sampling-distributions/lesson.qmd)**
(the distribution of a *summary* like the mean, not of raw values). We deliberately did not touch those here. Ready for more
reps? Do the [practice problems](practice.qmd).
