---
title: "Practice: Distributions and densities"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only (a {webr} starter does not see it — each starter re-states its
# own data). Base R is all we need, so nothing to install.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Do them on paper or in an R session; only
then open the answer.

## Problem 1 — the toy example, new numbers

Six new commute times (minutes): $19, 21, 20, 22, 20, 18$. As in the lesson: count how
many times each value shows up (the frequencies), turn those into proportions, and
confirm the proportions sum to 1. What is the mean, and which value is the most common?

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
y <- c(19, 21, 20, 22, 20, 18)
```

::: {.callout-note collapse="true" title="Answer"}
Frequencies: 18 once, 19 once, **20 twice** (the peak), 21 once, 22 once. As
proportions that's $1/6 \approx 0.167$ for each single, and $2/6 \approx 0.333$ for 20;
they add to $0.167\times4 + 0.333 \approx 1$. The mean is $(19+21+20+22+20+18)/6 = 120/6 =
\mathbf{20}$, and 20 is the most common value.

```{r p1-answer}
y <- c(19, 21, 20, 22, 20, 18)
table(y)                     # 20 appears twice; the rest once
table(y) / length(y)         # proportions
sum(table(y) / length(y))    # 1 — always
mean(y)                      # 20
```
:::

## Problem 2 — fill in the missing step (slice probability)

Here is a worked computation of $P(45 < X < 55)$ for a normal curve with center
$\mu = 50$ and spread $\sigma = 5$ — but the **core step is blanked**. Fill in ONLY the
blanked line.

> To get the probability of a *slice*, you take the area to the left of the upper edge
> and subtract the area to the left of the lower edge.
> Area left of 55 is `pnorm(55, 50, 5)` $= 0.841$; area left of 45 is
> `pnorm(45, 50, 5)` $= 0.159$.
> **`<your step here>`** — combine those two into the slice probability.
> So $P(45 < X < 55) = \underline{\phantom{00000}}$.

```{webr}
#| autorun: false
# p2-starter — edit and Run to check your step
mu <- 50; sigma <- 5
```

::: {.callout-note collapse="true" title="Answer"}
The blanked step is the subtraction — a slice's area is (area left of the top) minus
(area left of the bottom):

$$
P(45 < X < 55) = \texttt{pnorm(55, 50, 5)} - \texttt{pnorm(45, 50, 5)} = 0.841 - 0.159
= 0.682.
$$

(The exact R result is $0.6827 \approx 0.683$; the tiny gap comes from rounding the
intermediate values.)

That is why you never read one `pnorm` as a slice probability — a single `pnorm` is
everything to the *left*, so you must subtract off the part below the lower edge.

```{r p2-answer}
mu <- 50; sigma <- 5
pnorm(55, mu, sigma) - pnorm(45, mu, sigma)   # 0.6827 — the slice probability
```
(This is the 45-to-55 window: one spread either side of the center, so ~68% — the
familiar "about two-thirds within one standard deviation.")
:::

## Problem 3 — the four questions on one curve

A neutral measurement follows a normal curve with center $\mu = 100$ and spread
$\sigma = 15$. Answer each with the *right* member of the d/p/q/r family, then check in
R. (a) What is the curve's HEIGHT at 100? (b) What is $P(85 < X < 115)$? (c) What value
has 90% of the area to its left?

```{webr}
#| autorun: false
# p3-starter — your job: pick d / p / q for each part
mu <- 100; sigma <- 15
```

::: {.callout-note collapse="true" title="Answer"}
- **(a) Height** → `dnorm`: `dnorm(100, 100, 15)` $\approx 0.0266$. This is a density
  (per unit width), not a probability.
- **(b) Slice area** → two `pnorm`s subtracted:
  `pnorm(115, 100, 15) - pnorm(85, 100, 15)` $\approx 0.683$ (one spread either side of
  center — the ~68% again).
- **(c) The value at a given left-area** → `qnorm`: `qnorm(0.9, 100, 15)` $\approx
  119.2$.

```{r p3-answer}
mu <- 100; sigma <- 15
dnorm(100, mu, sigma)                              # (a) 0.0266 — a HEIGHT
pnorm(115, mu, sigma) - pnorm(85, mu, sigma)       # (b) 0.6827 — an AREA
qnorm(0.9, mu, sigma)                              # (c) 119.2 — a VALUE
```
:::

## Problem 4 — spot the error

A colleague wants the probability that the measurement in Problem 3 is *exactly* 100.
Their reasoning contains **exactly one mistake** — find it, explain why it's wrong, and
say what the right answer is.

> "The height of the curve at 100 is `dnorm(100, 100, 15)` = 0.0266. Height is
> probability, so the probability of getting exactly 100 is about 0.0266 — roughly a
> 2.7% chance."

```{webr}
#| autorun: false
# p4-starter — copy this to see the height vs. an actual slice probability
mu <- 100; sigma <- 15
```

::: {.callout-note collapse="true" title="Answer"}
The mistake is **"height is probability."** For a smooth density curve the height
`dnorm(100, 100, 15)` = 0.0266 is a **density** (probability per unit width), not a
probability. The probability of *exactly* 100 is 0 — a single point has zero width, so
zero area. Probability only exists over a *range*: e.g. the chance of landing within a
1-unit slice around 100 is `pnorm(100.5, 100, 15) - pnorm(99.5, 100, 15)` (an area),
not the bare height.

```{r p4-answer}
mu <- 100; sigma <- 15
dnorm(100, mu, sigma)                                  # 0.0266 — a density, NOT a prob
pnorm(100.5, mu, sigma) - pnorm(99.5, mu, sigma)       # ~0.0266 — a real slice prob
```
The two numbers happen to be close because the slice is ~1 unit wide (area ≈ height ×
width), but they are different *kinds* of thing: the first is a height, the second an
area.
:::

## Problem 5 — mixing it with the mean & spread *(uses [mean-variance-covariance](../mean-variance-covariance/lesson.qmd)'s mean and sd)*

Draw 500 random values from the Problem 3 curve ($\mu = 100$, $\sigma = 15$) with
`rnorm`, then compute the sample `mean` and `sd` of those draws.

**PREDICT** first: the curve was *built* with center 100 and spread 15 — what do you
expect the draws' mean and sd to be, roughly? Write your guess, then run.

```{webr}
#| autorun: false
# p5-starter — set the seed so your numbers match the answer, then draw
set.seed(42)
d <- rnorm(500, 100, 15)
```

::: {.callout-note collapse="true" title="Answer"}
`rnorm` draws from the curve, so the sample `mean` and `sd` — the very quantities from
[mean-variance-covariance](../mean-variance-covariance/lesson.qmd) — should land *near*
the curve's two knobs (100 and 15), but not exactly, because 500 random draws wobble
around the true values.

```{r p5-answer}
set.seed(42)
d <- rnorm(500, 100, 15)
mean(d)     # ~99.5 — close to the curve's center mu = 100
sd(d)       # ~14.6 — close to the curve's spread sigma = 15
```
The knobs $\mu$ and $\sigma$ that *define* the curve are exactly what the data's `mean`
and `sd` *estimate*. (How far off a 500-draw estimate tends to be is the
sampling-distribution / standard-error story — a separate module.)
:::

## Problem 6 — a gentle cliffhanger (turning the curve around)

You've been asking "given the curve, how likely is this data?" — feeding a value to
`dnorm`/`pnorm`. Now flip the question. Suppose you saw a single measurement of 108 but
you're **unsure of the curve's center**: was it centered at $\mu = 100$ or at $\mu =
110$? A natural move is to ask which center makes the observed 108 *least surprising* —
i.e. gives it the higher density. Compute `dnorm(108, 100, 15)` and `dnorm(108, 110,
15)` and see which is larger. (You can't fully name what you're doing yet — that's the
next idea.)

```{webr}
#| autorun: false
# p6-starter — same value 108, two candidate centers
value <- 108
```

::: {.callout-note collapse="true" title="Answer"}
```{r p6-answer}
dnorm(108, 100, 15)   # density of 108 if the curve is centered at 100
dnorm(108, 110, 15)   # density of 108 if the curve is centered at 110
```
The center 110 gives 108 the higher density (108 is nearer that peak), so 110 makes the
observation *less surprising*. You just used the density to **score how well a candidate
curve explains an observed value** — that's the seed of **likelihood**, which the
likelihood module builds out fully. Here it's only a cliffhanger: we flipped `dnorm`
from "curve → height of a value" into "which curve best fits this value?"
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/distributions-and-densities" pass="4"}

::: {.quiz-q qid="dd-height-is-density" qtype="mc" answer="b"}
For a smooth density curve, `dnorm(30, 30, 3)` is $0.133$. Is that "a 13.3% chance of
exactly $30$"?

a) Yes — the curve's height is the probability of that value
b) No — the height is a *density* (probability per unit width); an exact point has zero width, so its probability is $0$
c) No — the height must first be divided by the standard deviation
:::

::: {.quiz-q qid="dd-pnorm-left" qtype="mc" answer="a"}
`pnorm(33, 30, 3)` is $0.841$. What does $0.841$ represent?

a) The area to the *left* of $33$, i.e. $P(X < 33)$
b) The probability of landing exactly on $33$
c) The height of the curve at $33$
:::

::: {.quiz-q qid="dd-slice-prob" qtype="numeric" answer="0.682" tolerance="0.0015"}
For a normal, $P(X < 33) = 0.841$ and $P(X < 27) = 0.159$. What is the slice probability
$P(27 < X < 33)$? Round to three decimals.
:::

::: {.quiz-q qid="dd-normal-knobs" qtype="mc" answer="c"}
A normal curve is fully described by which two numbers?

a) Its minimum and its maximum
b) Its total area and its peak height
c) Its center $\mu$ (the mean) and its spread $\sigma$ (the standard deviation)
:::

::: {.quiz-q qid="dd-sd-from-variance" qtype="numeric" answer="3" tolerance="0"}
(From mean, variance & covariance.) The normal's spread knob is the standard deviation.
If a sample has variance $9$, what is its standard deviation?
:::

:::

::::
