---
title: "Capstone practice — the whole course, mixed"
subtitle: "No labels saying which tool to use — recalling that is part of the test. Answers hidden; try first."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| message: false
# Packages this practice set uses. Missing ones are installed automatically.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

These draw on every module, in mixed order — figuring out *which* idea each problem
needs is the point, so there are no "uses Module N" hints. Do each honestly before
opening the answer.

## Problem 1

Three cars: $x = (2, 4, 6)$ mph over the limit, $y = (3, 5, 8)$ tickets that year.
Score the candidate line "intercept 1, slope 1" by hand — predictions, residuals,
and its total score. Check in R.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
x <- c(2, 4, 6); y <- c(3, 5, 8)
```

::: {.callout-note collapse="true" title="Answer"}
Predictions $1 + 1\cdot x$: 3, 5, 7. Residuals $y - \hat{y}$: 0, 0, +1. Square and
sum (never just add — raw misses would cancel): $0 + 0 + 1 = 1$. The score is
SSE $= 1$.

```{r p1-answer}
x <- c(2, 4, 6); y <- c(3, 5, 8)
sum((y - (1 + 1 * x))^2)   # 1
```
:::

## Problem 2

New data: $x = (1, 2, 3)$, $y = (2, 1, 6)$. Fit the *best* line by hand: compute
$\bar{x}, \bar{y}, s_{xy}, s_x^2$, then the slope and intercept. Confirm with `lm()`.

```{webr}
#| autorun: false
# p2-starter — edit and Run to solve, right here
x <- c(1, 2, 3); y <- c(2, 1, 6)
```

::: {.callout-note collapse="true" title="Answer"}
$\bar{x} = 2$, $\bar{y} = 3$. Deviations of x: $(-1, 0, 1)$; of y: $(-1, -2, 3)$.
Products: $1, 0, 3$; sum 4, so $s_{xy} = 4/2 = 2$. Squared x-deviations: $1, 0, 1$;
sum 2, so $s_x^2 = 2/2 = 1$.

Slope $\hat{\beta}_1 = s_{xy}/s_x^2 = 2/1 = 2$; intercept
$\hat{\beta}_0 = \bar{y} - \hat{\beta}_1\bar{x} = 3 - 2\times 2 = -1$.

```{r p2-answer}
x <- c(1, 2, 3); y <- c(2, 1, 6)
c(slope = cov(x, y) / var(x),
  intercept = mean(y) - (cov(x, y) / var(x)) * mean(x))   # 2 and -1
coef(lm(y ~ x))                                            # matches
```
:::

## Problem 3

A journal table reports a predictor as `2.61 (0.87)` with the note "standard errors
in parentheses." Reconstruct its t value, and say in one plain sentence what that t
value tells you.

::: {.callout-note collapse="true" title="Answer"}
$t = \text{estimate} / \text{SE} = 2.61 / 0.87 = 3.0$. The estimate sits 3 bounce
("standard error") lengths from zero — comfortably past the ±2 rule of thumb, so it's
unlikely that luck alone (a true coefficient of zero) produced an effect this large;
the paper would flag it as statistically significant.
:::

## Problem 4 — spot the error

A colleague computes the regression slope for $x = (1, 2, 3)$, $y = (2, 1, 6)$ like
this. It has **exactly one mistake** — find it and give the correct slope.

> "Covariance is 2, variance of x is 1. The slope is variance over covariance,
> $s_x^2 / s_{xy} = 1/2 = 0.5$."

```{webr}
#| autorun: false
# p4-starter — edit and Run to check the corrected slope
x <- c(1, 2, 3); y <- c(2, 1, 6)
```

::: {.callout-note collapse="true" title="Answer"}
The ratio is upside down. The slope is **covariance over variance**,
$\hat{\beta}_1 = s_{xy}/s_x^2 = 2/1 = 2$ — not variance over covariance. (A quick
sanity check catches it: this data trends upward overall, so the slope should
be a solid positive number like 2, not a shallow 0.5.)

```{r p4-answer}
x <- c(1, 2, 3); y <- c(2, 1, 6)
cov(x, y) / var(x)   # 2 — covariance over variance
```
:::

## Problem 5

Fit `mpg ~ hp` on the built-in `mtcars` (fuel efficiency vs horsepower). The
horsepower coefficient comes out strongly significant ($p < 0.001$), and $R^2 \approx
0.60$. A friend concludes: "significant means horsepower predicts mpg almost
perfectly." Are they right? Explain.

```{webr}
#| autorun: false
# p5-starter — edit and Run to solve, right here
fit <- lm(mpg ~ hp, data = mtcars)
```

::: {.callout-note collapse="true" title="Answer"}
No — they're conflating two different questions. The tiny p-value says the *slope is
almost certainly not zero*: more horsepower really is associated with lower mpg. But
$R^2 = 0.60$ says the line still leaves **40%** of mpg's variation unexplained — a
lot of scatter around the line. A relationship can be rock-solid (highly significant)
and still be a loose predictor. Significance answers "is there an effect?"; $R^2$
answers "how tightly does the line fit?"

```{r p5-answer}
summary(lm(mpg ~ hp, data = mtcars))$coefficients   # hp: p ~ 1e-7, very significant
summary(lm(mpg ~ hp, data = mtcars))$r.squared       # ~0.60 — 40% unexplained
```
:::

## Problem 6 — decode a fresh table

Here is a regression you haven't seen, from R's built-in `women` data (average weight
in pounds regressed on height in inches for 15 age groups), in journal format:

> | | Weight (lbs) |
> |---|---|
> | Height (in) | 3.450*** |
> | | (0.091) |
> | Constant | −87.517*** |
> | | (5.937) |
> | Observations | 15 |
> | R² | 0.991 |
>
> *Notes:* SEs in parentheses. \*\*\*p < 0.001.

(a) In plain English, what does the Height coefficient mean? (b) Reconstruct the
Height t value from the table. (c) The intercept is −87.5 lbs — is a 0-inch person
predicted to weigh negative? What's really going on? (d) Is this a tight fit?

```{webr}
#| autorun: false
# p6-starter — edit and Run to check your reading against the real fit
fit <- lm(weight ~ height, data = women)
```

::: {.callout-note collapse="true" title="Answer"}
(a) Each additional inch of height is associated with about **3.45 more pounds** of
weight (that's $\hat{\beta}_1 = s_{xy}/s_x^2$).

(b) $t = 3.450 / 0.091 \approx 37.9$ — an enormous distance from zero; the `***`
(p < 0.001) is no surprise.

(c) No real 0-inch person exists — the intercept is just the line's height at
$x = 0$, far outside the data (the shortest woman here is 58 inches). Extrapolating
the line that far gives a meaningless negative; interpret the intercept only as the
anchor $\bar{y} - \hat{\beta}_1\bar{x}$, not a physical weight.

(d) Yes — $R^2 = 0.991$ means height explains 99% of the weight variation; these are
smoothed group averages, so the line fits almost exactly.

```{r p6-answer}
fit <- lm(weight ~ height, data = women)
summary(fit)$coefficients   # slope 3.450 (0.091), intercept -87.517 (5.937)
summary(fit)$r.squared       # 0.991
3.450 / 0.091                # ~37.9, the reconstructed t
```
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="courses/simple-linear-regression/modules/04-capstone-reading-a-real-regression" pass="4"}

::: {.quiz-q qid="cap-prediction" qtype="numeric" answer="41.4" tolerance="0.05"}
The `cars` regression line is $\hat{y} = -17.5791 + 3.9324\,x$. At $x = 15$ mph,
what stopping distance does it predict? Round to one decimal.
:::

::: {.quiz-q qid="cap-r2-interpret" qtype="mc" answer="c"}
The `cars` regression has $R^2 = 0.65$ and $p \approx 10^{-12}$ for speed. What do
these two numbers together tell you?

a) Speed's effect is rock-solid and explains nearly all of the variation in stopping distance
b) The p-value and R² contradict each other, so only one of them can be meaningful at once
c) Speed's effect is reliable, but about 35% of stopping variation remains unexplained
:::

::: {.quiz-q qid="cap-slope" qtype="numeric" answer="1.4" tolerance="0"}
For the tomato data, $s_{xy} = 7/3$ and $s_x^2 = 5/3$. What is the OLS slope
$\hat{\beta}_1 = s_{xy}/s_x^2$? Enter one decimal.
:::

::: {.quiz-q qid="cap-t-recon" qtype="numeric" answer="9.5" tolerance="0.05"}
From the `cars` regression: slope $= 3.9324$ and $\mathrm{SE} = 0.4155$. Compute
the t value (estimate ÷ SE). Round to one decimal.
:::

::: {.quiz-q qid="cap-intercept" qtype="mc" answer="b"}
The `cars` intercept is $-17.58$. A car at 0 mph is "predicted" to stop in
$-17.58$ feet. What is the right conclusion?

a) The model is wrong — a negative prediction proves the linear form breaks down in this data
b) The intercept is just the line's height at $x=0$, far outside the `cars` data range
c) Stopping distance at 0 mph must be estimated separately from different data
:::

:::

::::
