---
title: "Clustering: k-means and hierarchical"
subtitle: "Group unlabeled points into clusters — two algorithms, one goal"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

## Where we're going

You already know how to measure distance between data points and how to compute a
mean. Now the question is: **given a cloud of points with no labels, can you find
the groups?** That is *clustering* — discovering structure that nobody told you was
there. This lesson teaches two ways to do it. **k-means** says "pick k centers,
assign each point to its nearest center, move each center to the middle of its group,
repeat." **Hierarchical clustering** says "start with every point alone, then keep
merging the two closest groups until one big group remains — and read the merge
history to choose your clusters." Both rest on the distance and mean machinery you
already own.

**Builds on:**
[distance, similarity & geometry](../distance-similarity-and-geometry/lesson.qmd) —
Euclidean distance, why column scale matters;
[mean, variance & covariance](../mean-variance-covariance/lesson.qmd) — the mean as
a center point.

## Warm-up: do you remember?

Three quick reps from earlier modules — answers from memory first, then check.

**1.** Compute the Euclidean distance between the points $(1, 2)$ and $(4, 6)$.

```{webr}
#| autorun: false
# warmup1-starter — edit and Run to check
a <- c(1, 2)
b <- c(4, 6)
```

::: {.callout-note collapse="true" title="Answer"}
Square the coordinate gaps: $(4-1)^2 = 9$, $(6-2)^2 = 16$. Add: $9 + 16 = 25$.
Square root: $\sqrt{25} = 5$. In R: `sqrt(sum((a - b)^2))` gives 5. Taught in
[distance, similarity & geometry](../distance-similarity-and-geometry/lesson.qmd);
revisit if shaky.
:::

**2.** The values are 3, 7, 5. What is their mean?

::: {.callout-note collapse="true" title="Answer"}
$(3 + 7 + 5) / 3 = 15 / 3 = 5$. The mean is **5** — the balance point. Taught in
[mean, variance & covariance](../mean-variance-covariance/lesson.qmd); revisit if
shaky.
:::

**3.** Why does column scale matter when computing distances between data points?

::: {.callout-note collapse="true" title="Answer"}
A column measured in large numbers (like income in dollars) dominates the distance
calculation, drowning out columns measured in small numbers (like age in years). The
data's raw scale, not its importance, decides what matters. That is why you
standardize columns before computing distances. Taught in
[distance, similarity & geometry](../distance-similarity-and-geometry/lesson.qmd);
revisit if shaky.
:::

---

## The toy data (shared by both algorithms)

Six points in 2D. They naturally fall into two groups — three points in the
lower-left corner and three in the upper-right.

```{r toy-data}
#| fig-alt: "Six points plotted on a 2D grid, with three points clustered in the lower-left near (1.5,1.5) and three in the upper-right near (5.5,5.5), labeled 1 through 6."
# Six 2D points — three near (1,1) and three near (5.5, 5.5)
x <- c(1, 2, 1.5, 5, 6, 5.5)
y <- c(1, 1.5, 2, 5, 5.5, 6)
dat <- data.frame(x = x, y = y)
dat

# Quick look: you can see the two clusters by eye
plot(dat$x, dat$y, pch = 16, cex = 2, col = "steelblue",
     xlab = "x", ylab = "y", main = "Six points — two obvious groups",
     xlim = c(0, 7), ylim = c(0, 7))
text(dat$x, dat$y, labels = 1:6, pos = 3)
```

Points 1, 2, 3 sit near $(1.5, 1.5)$. Points 4, 5, 6 sit near $(5.5, 5.5)$. Both
algorithms should discover exactly these two groups — let's see how.

---

# Part 1: k-means clustering

## A tiny example first

**k-means by hand, step by step.** We set $k = 2$ (we want two clusters) and pick
two initial centroids — starting guesses for the cluster centers. For predictability,
start with centroid 1 at $(1, 1)$ and centroid 2 at $(6, 6)$.

**Iteration 1 — assign each point to its nearest centroid.**

For each of the 6 points, compute the distance to both centroids and assign the point
to whichever centroid is closer.

- Point 1 $(1, 1)$: distance to centroid 1 $(1,1)$ = $\sqrt{0+0} = 0$; distance to
  centroid 2 $(6,6)$ = $\sqrt{25+25} = \sqrt{50} \approx 7.07$. Assign to cluster 1.
- Point 2 $(2, 1.5)$: distance to C1 = $\sqrt{1+0.25} = \sqrt{1.25} \approx 1.12$;
  distance to C2 = $\sqrt{16+20.25} = \sqrt{36.25} \approx 6.02$. Assign to cluster 1.
- Point 3 $(1.5, 2)$: distance to C1 = $\sqrt{0.25+1} = \sqrt{1.25} \approx 1.12$;
  distance to C2 = $\sqrt{20.25+16} = \sqrt{36.25} \approx 6.02$. Assign to cluster 1.
- Point 4 $(5, 5)$: distance to C1 = $\sqrt{16+16} = \sqrt{32} \approx 5.66$;
  distance to C2 = $\sqrt{1+1} = \sqrt{2} \approx 1.41$. Assign to cluster 2.
- Point 5 $(6, 5.5)$: distance to C1 = $\sqrt{25+20.25} = \sqrt{45.25} \approx 6.73$;
  distance to C2 = $\sqrt{0+0.25} = \sqrt{0.25} = 0.5$. Assign to cluster 2.
- Point 6 $(5.5, 6)$: distance to C1 = $\sqrt{20.25+25} = \sqrt{45.25} \approx 6.73$;
  distance to C2 = $\sqrt{0.25+0} = \sqrt{0.25} = 0.5$. Assign to cluster 2.

Cluster 1 gets points {1, 2, 3}. Cluster 2 gets points {4, 5, 6}.

**Now update the centroids** — each centroid moves to the mean of its assigned points.

- Centroid 1: x-mean = $(1 + 2 + 1.5) / 3 = 4.5 / 3 = 1.5$,
  y-mean = $(1 + 1.5 + 2) / 3 = 4.5 / 3 = 1.5$. New centroid 1: $(1.5, 1.5)$.
- Centroid 2: x-mean = $(5 + 6 + 5.5) / 3 = 16.5 / 3 = 5.5$,
  y-mean = $(5 + 5.5 + 6) / 3 = 16.5 / 3 = 5.5$. New centroid 2: $(5.5, 5.5)$.

**Iteration 2 — reassign.**

Recompute distances with the new centroids $(1.5, 1.5)$ and $(5.5, 5.5)$:

- Point 1 $(1, 1)$: distance to C1 = $\sqrt{0.25+0.25} \approx 0.71$;
  distance to C2 = $\sqrt{20.25+20.25} \approx 6.36$. Still cluster 1.
- Point 2 $(2, 1.5)$: distance to C1 = $\sqrt{0.25+0} = 0.5$;
  distance to C2 = $\sqrt{12.25+16} \approx 5.31$. Still cluster 1.
- Point 3 $(1.5, 2)$: distance to C1 = $\sqrt{0+0.25} = 0.5$;
  distance to C2 = $\sqrt{16+12.25} \approx 5.31$. Still cluster 1.
- Points 4, 5, 6 are all much closer to C2 — assignments unchanged.

No point changed clusters, so the centroids stay at $(1.5, 1.5)$ and $(5.5, 5.5)$.
**Converged.** The algorithm stops.

The same computation in R:

```{r kmeans-by-hand}
# The 6 points as a matrix (one row per point)
x <- c(1, 2, 1.5, 5, 6, 5.5)
y <- c(1, 1.5, 2, 5, 5.5, 6)
dat <- data.frame(x = x, y = y)

# Starting centroids: (1,1) and (6,6)
centroids <- matrix(c(1, 1, 6, 6), nrow = 2, byrow = TRUE)
colnames(centroids) <- c("x", "y")

# --- Iteration 1 ---
# Distance from each point to each centroid (6 x 2 matrix)
dist_to_c1 <- sqrt((dat$x - centroids[1, 1])^2 + (dat$y - centroids[1, 2])^2)
dist_to_c2 <- sqrt((dat$x - centroids[2, 1])^2 + (dat$y - centroids[2, 2])^2)

# Assign each point to the nearest centroid
cluster <- ifelse(dist_to_c1 <= dist_to_c2, 1, 2)
cluster  # 1 1 1 2 2 2

# Update centroids: mean of assigned points
centroids[1, ] <- colMeans(dat[cluster == 1, ])  # (1.5, 1.5)
centroids[2, ] <- colMeans(dat[cluster == 2, ])  # (5.5, 5.5)
centroids

# --- Iteration 2 ---
dist_to_c1 <- sqrt((dat$x - centroids[1, 1])^2 + (dat$y - centroids[1, 2])^2)
dist_to_c2 <- sqrt((dat$x - centroids[2, 1])^2 + (dat$y - centroids[2, 2])^2)
cluster_new <- ifelse(dist_to_c1 <= dist_to_c2, 1, 2)
cluster_new  # 1 1 1 2 2 2 — same as before, converged!

all(cluster == cluster_new)  # TRUE: nothing changed, algorithm stops
```

And with R's built-in `kmeans()`:

```{r kmeans-builtin}
# kmeans() on the same data, k = 2, one random start
set.seed(42)
km <- kmeans(dat, centers = 2, nstart = 1)
km$cluster   # cluster assignments (1 or 2 — the label numbers may be swapped)
km$centers   # the final centroids
```

The centroids from `kmeans()` land at the same positions (up to label swapping —
"cluster 1" and "cluster 2" are just names, and `kmeans()` may number them
differently from our hand computation).

## Getting a feel for it (k-means)

**The magnet analogy.** Picture two magnets on a table scattered with iron filings.
Each filing snaps to its nearest magnet. Then each magnet slides to the center of its
pile of filings. After it slides, some filings near the boundary might now be closer
to the other magnet, so they switch. The magnets slide again. Repeat until nothing
moves. That is k-means: the magnets are the centroids, the filings are the data
points. (Where the analogy strains: real magnets exert force that varies with
distance; k-means only cares about "nearest", with no force weighting.)

**Why iterate — why not just pick good centers and be done?** Because the initial
guess is arbitrary. The loop alternates between two questions: "given these centers,
who belongs where?" (the assign step) and "given these members, where should the
center be?" (the update step). Each step can only improve the total squared distance
or leave it unchanged — never worsen it — so the algorithm always converges. But it
finds a *local* minimum, not necessarily the global best, which is why the starting
positions matter and why you run it multiple times (more on that in Common traps).

**What the grouping feels like.** Our two clusters have centroids at $(1.5, 1.5)$ and
$(5.5, 5.5)$ — they are 4 units apart in each coordinate, a clear separation. Each
point sits less than 1 unit from its own centroid. If the two groups were closer
together, or if the points were more spread out, the clustering would be harder — the
boundary between clusters gets fuzzy when the gap between centroids shrinks relative
to the scatter within each cluster.

## WCSS: how good is this clustering?

The **within-cluster sum of squares** (WCSS) measures clustering quality. For each
point, compute the squared distance from that point to its own cluster's centroid.
Add them all up. A smaller WCSS means tighter clusters.

For our toy data:

- Cluster 1, centroid $(1.5, 1.5)$:
  - Point 1: $(1-1.5)^2 + (1-1.5)^2 = 0.25 + 0.25 = 0.5$
  - Point 2: $(2-1.5)^2 + (1.5-1.5)^2 = 0.25 + 0 = 0.25$
  - Point 3: $(1.5-1.5)^2 + (2-1.5)^2 = 0 + 0.25 = 0.25$
  - Subtotal: $0.5 + 0.25 + 0.25 = 1.0$
- Cluster 2, centroid $(5.5, 5.5)$:
  - Point 4: $(5-5.5)^2 + (5-5.5)^2 = 0.25 + 0.25 = 0.5$
  - Point 5: $(6-5.5)^2 + (5.5-5.5)^2 = 0.25 + 0 = 0.25$
  - Point 6: $(5.5-5.5)^2 + (6-5.5)^2 = 0 + 0.25 = 0.25$
  - Subtotal: $0.5 + 0.25 + 0.25 = 1.0$
- **Total WCSS = $1.0 + 1.0 = 2.0$.**

This is what k-means tries to minimize. `kmeans()` reports it as `tot.withinss`:

```{r wcss-check}
x <- c(1, 2, 1.5, 5, 6, 5.5)
y <- c(1, 1.5, 2, 5, 5.5, 6)
dat <- data.frame(x = x, y = y)
set.seed(42)
km <- kmeans(dat, centers = 2, nstart = 1)
km$tot.withinss  # 2.0 — matches our hand computation
```

## What the notation is going to say (k-means)

The formula will say: for each cluster $k$, take every point $\mathbf{x}_i$ in that
cluster, compute the squared distance from $\mathbf{x}_i$ to its centroid
$\boldsymbol{\mu}_k$, and add them all up across every cluster. The centroid update
will say: the new centroid is the mean of all the points assigned to that cluster. In
our toy example that is exactly the "compute squared gaps, add up to get 2.0" we just
did, and the "average the x-coordinates and y-coordinates separately" we did in the
update step.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $K$ | "big K" | the number of clusters you choose | $K = 2$ |
| $k$ | "little k" | the index of one particular cluster ($k = 1, 2, \ldots, K$) | cluster 1 or cluster 2 |
| $\boldsymbol{\mu}_k$ | "mew sub k" | the centroid of cluster $k$ — the mean position of its members (in R: `colMeans(dat[cluster == k, ])`) | $\boldsymbol{\mu}_1 = (1.5, 1.5)$, $\boldsymbol{\mu}_2 = (5.5, 5.5)$ |
| $C_k$ | "see sub k" | the set of points assigned to cluster $k$ | $C_1 = \{1, 2, 3\}$, $C_2 = \{4, 5, 6\}$ |
| $\mathbf{x}_i$ | "ex sub eye" (bold) | the $i$-th data point, as a vector of coordinates (in R: `dat[i, ]`) | $\mathbf{x}_1 = (1, 1)$, $\mathbf{x}_4 = (5, 5)$ |
| $n_k$ or $\lvert C_k \rvert$ | "en sub k" or "the size of see sub k" | how many points are in cluster $k$ | $n_1 = 3$, $n_2 = 3$ |
| $\lVert \mathbf{x}_i - \boldsymbol{\mu}_k \rVert^2$ | "the squared norm of ex-sub-i minus mew-sub-k" | the squared Euclidean distance from point $i$ to its centroid (in R: `sum((dat[i,] - centroid_k)^2)`) | point 1 to $\boldsymbol{\mu}_1$: $0.5$ |
| WCSS | "double-you see ess ess" | within-cluster sum of squares — the total of all those squared distances | $2.0$ |
| $d(\mathbf{x}_i, \boldsymbol{\mu}_k)$ | "d of ex-sub-i and mew-sub-k" | the distance from point $i$ to centroid $k$ (in R: `sqrt(sum((dat[i,] - centroid_k)^2))`) | point 1 to $\boldsymbol{\mu}_1$: $\sqrt{0.5} \approx 0.71$ |

## The formal version (k-means)

**The k-means objective — what the algorithm tries to minimize:**
[@boyd2018appliedlinearalgebra, ch. 4; @james2021introductionstatisticallearning, pp. 515-517]

$$
\text{WCSS} = \sum_{k=1}^{K} \sum_{i \in C_k} \lVert \mathbf{x}_i - \boldsymbol{\mu}_k \rVert^2
$$

Reading it aloud: "the WCSS equals the sum, from k equals 1 to K, of the sum over
all i in C-sub-k, of the squared norm of x-sub-i minus mu-sub-k" — or in everyday
words: **"for every point, square its distance to its own cluster center; add all of
those up across all clusters."**

Mapping it onto the toy example:

- The outer sum $\sum_{k=1}^{K}$ says "do this for cluster 1 and cluster 2" — our
  $K = 2$.
- The inner sum $\sum_{i \in C_k}$ says "for every point in this cluster" — three
  points each.
- $\lVert \mathbf{x}_i - \boldsymbol{\mu}_k \rVert^2$ is the squared Euclidean
  distance from point $i$ to its centroid — e.g. $\lVert (1,1) - (1.5,1.5) \rVert^2
  = 0.25 + 0.25 = 0.5$.
- Adding it all up: $0.5 + 0.25 + 0.25 + 0.5 + 0.25 + 0.25 = 2.0$.

**The centroid update rule:** [@boyd2018appliedlinearalgebra, p. 74]

$$
\boldsymbol{\mu}_k = \frac{1}{\lvert C_k \rvert} \sum_{i \in C_k} \mathbf{x}_i
$$

Reading it aloud: "mu-sub-k equals one over the size of C-sub-k, times the sum of
x-sub-i for all i in C-sub-k" — or in everyday words: **"the centroid is the average
position of all the points in the cluster."** For cluster 1:
$\boldsymbol{\mu}_1 = \frac{1}{3}\bigl((1,1) + (2,1.5) + (1.5,2)\bigr) = (4.5/3,\;4.5/3) = (1.5,\;1.5)$.

---

# Part 2: hierarchical clustering

## A tiny example first

**Hierarchical agglomerative clustering by hand.** We use the same 6 points. Instead
of picking $k$ up front, we start with each point as its own cluster (6 clusters)
and repeatedly merge the two closest clusters until only one remains.
[@james2021introductionstatisticallearning, pp. 519-524]

**Step 1 — compute all pairwise distances.** We need the Euclidean distance between
every pair:

```{r pairwise-distances}
# The same 6 points
x <- c(1, 2, 1.5, 5, 6, 5.5)
y <- c(1, 1.5, 2, 5, 5.5, 6)
dat <- data.frame(x = x, y = y)

# The full distance matrix (we only need the upper/lower triangle)
d <- dist(dat)
round(as.matrix(d), 2)
```

The key distances: points 1-2 are about 1.12 apart, points 1-3 are about 1.12 apart,
points 2-3 are about 0.71 apart. On the other side, points 4-5 are about 1.12 apart,
points 4-6 are about 1.12 apart, points 5-6 are about 0.71 apart. The cross-group
distances (e.g. 1-4) are all around 5 or more.

**Step 2 — merge the closest pair.** The smallest distance is 0.71, between points
2 and 3 (and also between points 5 and 6). Pick one — say 2 and 3. Merge them into a
new cluster {2, 3}. We now have 5 clusters: {1}, {2,3}, {4}, {5}, {6}.

**Step 3 — update distances.** We need the distance from the merged cluster {2, 3}
to every other cluster. Using **single linkage** (minimum distance between any member
of one cluster and any member of the other):

- $d(\{2,3\}, \{1\})$ = $\min(d(2,1),\;d(3,1))$ = $\min(1.12,\;1.12)$ = $1.12$
- $d(\{2,3\}, \{4\})$ = $\min(d(2,4),\;d(3,4))$ = $\min(4.61,\;4.61)$ = $4.61$
- $d(\{2,3\}, \{5\})$ = $\min(d(2,5),\;d(3,5))$ = $\min(5.66,\;5.70)$ = $5.66$
- $d(\{2,3\}, \{6\})$ = $\min(d(2,6),\;d(3,6))$ = $\min(5.70,\;5.66)$ = $5.66$

**Step 4 — find the next closest pair and merge.** Scanning all current inter-cluster
distances, the smallest is 0.71 between {5} and {6}. Merge into {5, 6}. Now 4
clusters: {1}, {2,3}, {4}, {5,6}.

**Step 5 — keep going.** The next smallest distance is 1.12 between {1} and {2,3}.
Merge into {1, 2, 3}. Now 3 clusters: {1,2,3}, {4}, {5,6}.

**Step 6.** The next merge is {4} into {5,6}, giving {4,5,6}, at a single-linkage
distance of about 1.12. Now 2 clusters: {1,2,3} and {4,5,6}.

**Step 7.** The last merge combines {1,2,3} and {4,5,6} into one big cluster, at a
distance of about 4.61 (the nearest member of one group to the nearest member of the
other).

The whole merge history — which pairs merged and at what distance — is the
**dendrogram**: a tree diagram that records the clustering at every level.

The same thing in R with `hclust()`:

```{r hclust-example}
#| fig-alt: "Dendrogram from single-linkage hierarchical clustering of 6 points, showing within-group merges at low heights and the final between-group merge at a much greater height."
x <- c(1, 2, 1.5, 5, 6, 5.5)
y <- c(1, 1.5, 2, 5, 5.5, 6)
dat <- data.frame(x = x, y = y)

# hclust() needs a dist object, and we specify single linkage
d <- dist(dat)
hc <- hclust(d, method = "single")

# The dendrogram: the tree of merges
plot(hc, main = "Dendrogram (single linkage)", xlab = "", sub = "",
     labels = paste0("pt", 1:6))
```

## Getting a feel for it (hierarchical)

**The party analogy.** Imagine six people at a party, each standing alone. The two
standing closest start chatting and form a pair. Then the next-closest person or pair
joins a conversation. Groups grow by absorbing whoever is nearest, and eventually the
whole room is one big group. The dendrogram is the *guest list* that records who
joined whom and when. (Where the analogy breaks: at a real party, someone might
join based on who they know, not pure physical distance — hierarchical clustering uses
only the numbers.)

**Why bottom-up?** Starting with every point separate and merging upward means you
never have to pick the number of clusters in advance. The dendrogram records *every*
possible grouping — from 6 clusters down to 1 — and you choose after the fact by
cutting the tree at a height that makes sense. That flexibility is the key advantage
over k-means, which demands you commit to $K$ before running.

**What the dendrogram tells you.** The height at which two branches merge is the
distance between those clusters at the moment they joined. A tall vertical gap in the
dendrogram before a merge means those groups were far apart — a natural break. In our
toy data, the biggest gap is between the merges at height ~1.12 (the within-group
merges) and the final merge at height ~4.61 (the between-group merge). That gap
screams "two clusters."

## Cutting the dendrogram

The dendrogram stores the full merge history. To get a specific number of clusters,
you **cut** it — draw a horizontal line across the tree at some height, and each
branch below the line is one cluster.

```{r cutree-example}
x <- c(1, 2, 1.5, 5, 6, 5.5)
y <- c(1, 1.5, 2, 5, 5.5, 6)
dat <- data.frame(x = x, y = y)
d <- dist(dat)
hc <- hclust(d, method = "single")

# Cut to get exactly 2 clusters
clusters_2 <- cutree(hc, k = 2)
clusters_2  # 1 1 1 2 2 2 — same two groups as k-means found

# Cut to get 3 clusters
clusters_3 <- cutree(hc, k = 3)
clusters_3  # one of the tight groups splits further
```

`cutree(hc, k = 2)` gives us {1,2,3} and {4,5,6} — the same two groups k-means
found.

## Linkage types

The linkage rule decides how to measure the distance between two clusters (not just
two individual points). This choice matters a lot.
[@james2021introductionstatisticallearning, pp. 523-524]

**Single linkage** = the distance between the **closest** members of the two
clusters: $d(A, B) = \min_{i \in A,\;j \in B} d(\mathbf{x}_i, \mathbf{x}_j)$.
It can chain: if cluster A has a point near cluster B, which has a point near cluster
C, single linkage may merge A-B-C into one long straggling chain even when A and C
are far apart. Good for detecting elongated shapes; bad if you want compact round
clusters.

**Complete linkage** = the distance between the **farthest** members:
$d(A, B) = \max_{i \in A,\;j \in B} d(\mathbf{x}_i, \mathbf{x}_j)$. A merge only
happens if even the most remote members are close. This gives compact, roughly
spherical clusters — the conservative choice.

**Average linkage** = the average of all pairwise distances between members of the
two clusters. A middle ground between single and complete: less prone to chaining
than single, less extreme than complete.

**Ward's linkage** = merge the pair of clusters that increases the WCSS the least.
It directly optimizes compactness and is the most common choice in practice. (The
`method = "ward.D2"` argument in `hclust()` implements this using squared distances.)

```{r linkage-comparison}
#| fig-alt: "Two dendrograms side by side: single linkage (left) and Ward's linkage (right) for the same 6 points, both showing the same two-cluster structure."
x <- c(1, 2, 1.5, 5, 6, 5.5)
y <- c(1, 1.5, 2, 5, 5.5, 6)
dat <- data.frame(x = x, y = y)
d <- dist(dat)

# Compare dendrograms: set up a 1-row, 2-column layout
par(mfrow = c(1, 2))
plot(hclust(d, method = "single"), main = "Single linkage",
     xlab = "", sub = "", labels = paste0("pt", 1:6))
plot(hclust(d, method = "ward.D2"), main = "Ward's linkage",
     xlab = "", sub = "", labels = paste0("pt", 1:6))
par(mfrow = c(1, 1))
```

For our toy data both linkage methods give the same two-cluster solution — the groups
are so well separated that the linkage choice doesn't matter. On messier real data,
the choice matters a great deal.

---

## What the notation is going to say (hierarchical)

The single-linkage formula will say: the distance between two clusters $A$ and $B$ is
the *smallest* distance between any one member of $A$ and any one member of $B$. In
our toy example, that is exactly the "which two points from different groups are
closest?" question.

## How to read the symbols (hierarchical)

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $A$, $B$ | "cluster A, cluster B" | two clusters being considered for a merge | {1} and {2,3} during step 5 |
| $d(A, B)$ | "d of A and B" | the inter-cluster distance (depends on linkage) | single linkage: the smallest point-to-point distance across the two clusters |
| $\min_{i \in A,\;j \in B}$ | "the minimum over i in A and j in B" | the smallest value you get when checking all pairs with one member from A and one from B (in R: `min(dist_matrix[A, B])`) | $\min(1.12, 1.12) = 1.12$ for $d(\{1\}, \{2,3\})$ |
| $\max_{i \in A,\;j \in B}$ | "the maximum over i in A and j in B" | the largest such pairwise distance | used in complete linkage |

## The formal version (hierarchical)
[@james2021introductionstatisticallearning, pp. 523-524, Table 12.3]

**Single linkage:**

$$
d(A, B) = \min_{i \in A,\;j \in B} d(\mathbf{x}_i, \mathbf{x}_j)
$$

Reading it aloud: "d of A and B equals the minimum, over all i in A and j in B, of
d of x-sub-i and x-sub-j" — or in everyday words: **"the distance between two
clusters is the distance between their two closest members."**

For our toy example, the distance between cluster {1,2,3} and cluster {4,5,6} under
single linkage is the shortest point-to-point distance across the two groups. The
closest cross-group pair is points 3 and 4: $d((1.5, 2), (5, 5)) = \sqrt{3.5^2 + 3^2}
= \sqrt{12.25 + 9} = \sqrt{21.25} \approx 4.61$. That is the merge height for the
final merge in the dendrogram.

**Complete linkage** swaps min for max:

$$
d(A, B) = \max_{i \in A,\;j \in B} d(\mathbf{x}_i, \mathbf{x}_j)
$$

"The distance between two clusters is the distance between their two **farthest**
members."

---

## Choosing k: the elbow method

Neither algorithm tells you the "right" number of clusters. For k-means, a common
trick is the **elbow plot**: run k-means for $k = 1, 2, 3, \ldots$ and plot the
total WCSS at each $k$. WCSS always decreases as you add more clusters (at the
extreme, every point would be its own cluster and WCSS = 0), but at some point the
improvement slows sharply — the "elbow." The $k$ at the elbow is a reasonable choice.
[@boyd2018appliedlinearalgebra, p. 76]

```{r elbow-plot}
#| fig-alt: "Elbow plot showing WCSS versus number of clusters k from 1 to 5, with a sharp drop from k=1 to k=2 and nearly flat after k=2, indicating the elbow at k=2."
x <- c(1, 2, 1.5, 5, 6, 5.5)
y <- c(1, 1.5, 2, 5, 5.5, 6)
dat <- data.frame(x = x, y = y)

# Run kmeans for k = 1 to 5, collect WCSS (k = 6 = n would be trivial: every
# point its own cluster, WCSS = 0 — and R's kmeans doesn't allow k = n)
wcss <- numeric(5)
for (k in 1:5) {
  set.seed(42)
  wcss[k] <- kmeans(dat, centers = k, nstart = 25)$tot.withinss
}
wcss

# The elbow plot
plot(1:5, wcss, type = "b", pch = 16, col = "steelblue",
     xlab = "Number of clusters (k)", ylab = "Total WCSS",
     main = "Elbow plot — WCSS vs k")
```

The big drop is from $k = 1$ to $k = 2$. After $k = 2$ the line is nearly flat — the
elbow says $k = 2$, which matches what we see by eye.

---

## Explore it

```{webr}
#| autorun: true
#| fig-alt: "Side-by-side comparison of k-means clustering results with nstart=1 versus nstart=25 on the toy data, showing cluster assignments and centroids."
# explore-1 — LIVE: why nstart matters.
# PREDICT: will running kmeans() twice with nstart=1 give the same clusters?
# Try it — run this cell, note the results, then run it AGAIN.
x <- c(1, 2, 1.5, 5, 6, 5.5)
y <- c(1, 1.5, 2, 5, 5.5, 6)
dat <- data.frame(x = x, y = y)

# nstart=1: one random starting position
km1 <- kmeans(dat, centers = 2, nstart = 1)

# nstart=25: pick the best of 25 random starts
km25 <- kmeans(dat, centers = 2, nstart = 25)

par(mfrow = c(1, 2))
plot(dat$x, dat$y, col = km1$cluster, pch = 16, cex = 2,
     main = paste("nstart=1, WCSS =", round(km1$tot.withinss, 2)),
     xlab = "x", ylab = "y", xlim = c(0, 7), ylim = c(0, 7))
points(km1$centers, pch = 4, cex = 3, lwd = 3)

plot(dat$x, dat$y, col = km25$cluster, pch = 16, cex = 2,
     main = paste("nstart=25, WCSS =", round(km25$tot.withinss, 2)),
     xlab = "x", ylab = "y", xlim = c(0, 7), ylim = c(0, 7))
points(km25$centers, pch = 4, cex = 3, lwd = 3)
par(mfrow = c(1, 1))
```

```{webr}
#| autorun: true
#| fig-alt: "Two dendrograms side by side showing single linkage (left) and complete linkage (right) for the same 6-point dataset."
# explore-2 — LIVE: linkage comparison.
# TRY IT: change method to "average" or "ward.D2" and re-run.
# PREDICT: will single linkage and complete linkage give different 2-cluster cuts
# for this well-separated data? What about data where clusters are closer together?
x <- c(1, 2, 1.5, 5, 6, 5.5)
y <- c(1, 1.5, 2, 5, 5.5, 6)
dat <- data.frame(x = x, y = y)
d <- dist(dat)

par(mfrow = c(1, 2))
hc_s <- hclust(d, method = "single")
hc_c <- hclust(d, method = "complete")
plot(hc_s, main = "Single linkage", xlab = "", sub = "")
plot(hc_c, main = "Complete linkage", xlab = "", sub = "")
par(mfrow = c(1, 1))

# The 2-cluster assignments — same for both on this clean data
cat("Single linkage, k=2:", cutree(hc_s, k = 2), "\n")
cat("Complete linkage, k=2:", cutree(hc_c, k = 2), "\n")
```

```{webr}
#| autorun: true
#| fig-alt: "Elbow plot showing WCSS versus number of clusters k from 1 to 5, with the elbow at k=2."
# explore-3 — LIVE: the elbow plot for choosing k.
# TRY IT: what if you added a 7th point at (3.5, 3.5) — right between the groups?
# Would the elbow still be at k=2?
x <- c(1, 2, 1.5, 5, 6, 5.5)
y <- c(1, 1.5, 2, 5, 5.5, 6)
dat <- data.frame(x = x, y = y)

wcss <- numeric(5)
for (k in 1:5) {
  set.seed(42)
  wcss[k] <- kmeans(dat, centers = k, nstart = 25)$tot.withinss
}

plot(1:5, wcss, type = "b", pch = 16, cex = 1.5, col = "steelblue",
     xlab = "Number of clusters (k)", ylab = "Total WCSS",
     main = "Elbow plot — where does the curve bend?")
# The sharp drop from k=1 to k=2 and flat afterward — elbow at k=2
```

---

## Common traps

**Trap 1 — "k-means always finds the best clustering."** k-means finds a *local*
minimum of WCSS, not the global one. With unlucky starting centroids, it can get stuck
in a worse solution. On our toy data the two clusters are so obvious that any start
works, but here is a 1D example where starting positions matter.
[@boyd2018appliedlinearalgebra, p. 76; @james2021introductionstatisticallearning, pp. 517-518]

```{r trap-nstart}
# 9 points in 1D: three tight groups at 0, 5, and 10
pts <- data.frame(val = c(0, 0.1, -0.1, 5, 5.1, 4.9, 10, 10.1, 9.9))
set.seed(1)
bad <- kmeans(pts, centers = 3, nstart = 1)
set.seed(42)
good <- kmeans(pts, centers = 3, nstart = 25)
cat("nstart=1 WCSS:", bad$tot.withinss, "\n")
cat("nstart=25 WCSS:", good$tot.withinss, "\n")
# nstart=25 is more likely to find the best (lowest WCSS) solution
```

The fix is `nstart = 25` (or more) — run the algorithm 25 times from random starts
and keep the best. Always use `nstart` > 1 in practice.

**Trap 2 — "k-means works for any shape of cluster."** k-means assumes roughly
spherical (round) clusters because it assigns points to the *nearest centroid*. If the
true clusters are crescent-shaped or elongated, k-means slices them with straight
borders and gets it wrong:

```{r trap-shape}
#| fig-alt: "Two crescent-shaped clusters that k-means splits incorrectly because it assumes round clusters."
# Two crescents (half-moons) — k-means can't handle these
set.seed(42)
n <- 100
theta1 <- seq(0, pi, length.out = n)
theta2 <- seq(0, pi, length.out = n)
moon1 <- data.frame(x = cos(theta1), y = sin(theta1))
moon2 <- data.frame(x = cos(theta2) + 0.5, y = -sin(theta2) + 0.5)
moons <- rbind(moon1, moon2)
true_label <- rep(c(1, 2), each = n)

km_moons <- kmeans(moons, centers = 2, nstart = 25)

plot(moons$x, moons$y, col = km_moons$cluster, pch = 16,
     xlab = "x", ylab = "y",
     main = "k-means on crescents — wrong split")
```

k-means draws a straight boundary between the two crescents instead of following
their curved shapes. Hierarchical clustering with single linkage handles elongated
shapes better — it follows the chain of nearby points.

**Trap 3 — "Leaves that sit close together in a dendrogram are similar."** Look
again at the single-linkage dendrogram from *A tiny example first* in Part 2.
Points 3 and 4 are drawn right next to each other on the x-axis — the picture
invites you to read that as "similar." Check the height at which they actually
first end up in the same cluster instead:

```{r trap-dendrogram-order}
x <- c(1, 2, 1.5, 5, 6, 5.5)
y <- c(1, 1.5, 2, 5, 5.5, 6)
dat <- data.frame(x = x, y = y)
hc <- hclust(dist(dat), method = "single")

hc$order                       # leaf order left-to-right in the plot: 1 2 3 4 5 6
hc$height[length(hc$height)]   # height of the very last (highest) merge
```

Points 3 and 4 don't join the same cluster until the **last, highest merge in the
whole tree**, at height $\approx 4.61$ — they are the *least* similar adjacent pair
here, not the most. Similarity in a dendrogram is read from the **vertical axis**
(the height where two branches first fuse), never from horizontal position. Horizontal
order is largely a drawing convenience: a dendrogram with $n$ leaves has $2^{n-1}$
different left-to-right layouts, all showing the exact same clustering, because you
can always flip either side of any fork without changing which points end up
together. [@james2021introductionstatisticallearning, pp. 521-522]

**Trap 4 — "The clusters my run found are THE truth."** Clustering results are not
as robust as they look — remove a few observations and re-cluster, and the grouping
can shift. Force $k = 3$ on the toy data (splitting one of the two obvious groups,
since 3 isn't the natural number of clusters here) and compare the full data against
the same data with point 4 dropped:

```{r trap-perturbation}
x <- c(1, 2, 1.5, 5, 6, 5.5)
y <- c(1, 1.5, 2, 5, 5.5, 6)
dat <- data.frame(x = x, y = y, id = 1:6)

# All 6 points, k = 3
full <- cutree(hclust(dist(dat[, c("x", "y")]), method = "single"), k = 3)
names(full) <- dat$id
full   # point 4 ends up alone in its own group

# Drop point 4 (about 1/6, close to a 20% perturbation) and re-cluster
sub <- dat[dat$id != 4, ]
no4 <- cutree(hclust(dist(sub[, c("x", "y")]), method = "single"), k = 3)
names(no4) <- sub$id
no4    # point 1 is alone now instead
```

With all 6 points, point 4 is the odd one out (its own singleton group). Drop that
one point and re-cluster the remaining 5, and point 1 becomes the new odd one out —
a genuinely different grouping, not just a relabeling. If a single removed point can
change *which* observation looks like the outlier on data this clean, imagine what
removing a chunk of a messier, real dataset can do. Treat a clustering result as a
hypothesis to sanity-check — re-run it on a perturbed subset before treating the
grouping as settled fact. [@james2021introductionstatisticallearning, pp. 529-530]

**Trap 5 — "It doesn't matter which options I pick."** The dissimilarity metric, linkage method, scaling
decision, and choice of $k$ are not minor implementation details — every one of them
can materially change which points end up grouped together. This lesson already
showed it: the warm-up recalled how raw column scale silently decides what counts as
"close", *Linkage types* explained why single and complete linkage can disagree on
messier data, and Trap 4 showed the grouping shifting when a single point was
removed. There is no universally "correct" combination, so always report your choices —
state the dissimilarity measure, the linkage rule (or `nstart` for k-means), whether
and how you scaled, and the $k$ you chose, so a reader can assess and reproduce the
result.
[@james2021introductionstatisticallearning, pp. 528--530]

## Check yourself

**1.** You run `kmeans()` with `nstart = 1` and get a WCSS of 15. You run it again
with `nstart = 1` and get a WCSS of 12. Which run found the better clustering?

::: {.callout-note collapse="true" title="Answer"}
The run with WCSS = **12** found the better clustering. Lower WCSS means points are
closer to their own centroids, i.e., tighter clusters. The fact that two runs gave
different answers is why you should use `nstart = 25` or more — it keeps the best of
many tries.
:::

**2.** In a dendrogram, what does a large vertical gap between two merge heights tell
you?

::: {.callout-note collapse="true" title="Answer"}
A large gap means the clusters on either side of that gap were far apart — they
resisted merging for a long time. This is a natural break in the data and suggests
a good place to cut the dendrogram. In our toy data, the big gap between the
within-group merges (~1.12) and the final between-group merge (~4.61) signals two
well-separated clusters.
:::

**3.** In your own words: explain to a colleague when you would prefer k-means over
hierarchical clustering, and when you might choose the other way.

::: {.callout-note collapse="true" title="A model answer"}
k-means is fast and works well when you have a rough idea of how many clusters you
want and the clusters are roughly round — you set k, run it, and get assignments.
Hierarchical clustering is better when you want to explore different numbers of
clusters without re-running anything — the dendrogram shows the full merge history,
so you cut it wherever makes sense. Hierarchical also handles elongated cluster shapes
better (with single linkage), but it is slower on large datasets because it computes
all pairwise distances.
:::

## What you can now do

- **Run k-means by hand**: assign points to nearest centroids, recompute centroids as
  means, iterate until convergence — and reproduce it with `kmeans()` in R.
- **Compute WCSS** for a clustering — the sum of squared distances from each point to
  its centroid — and read `kmeans()$tot.withinss`.
- **Run hierarchical clustering by hand**: compute pairwise distances, merge the
  closest pair, update distances, and read the resulting dendrogram — and reproduce it
  with `hclust()` and `plot()`.
- **Cut a dendrogram** at a chosen height or number of clusters using `cutree()`.
- **Choose k** using the elbow plot (WCSS vs k) and explain why `nstart` > 1 matters
  for k-means.

If any of these feels shaky, revisit
[A tiny example first (k-means)](#a-tiny-example-first) or
[A tiny example first (hierarchical)](#a-tiny-example-first-1) before continuing —
any method that groups data assumes all of these.

## Where this goes next

Clustering groups similar points; **[PCA & SVD](../pca-and-svd/lesson.qmd)** (principal
component analysis) finds the directions along which points *vary most* — both use the
covariance matrix's geometry, and they often work in tandem (reduce dimensions with
PCA, then cluster in that reduced space). The distance and mean machinery you used
here will come right back.

Ready for more reps? Do the [practice problems](practice.qmd).
