---
title: "Practice: Clustering (k-means & hierarchical)"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Do them on paper or in an R session; only
then open the answer.

## Problem 1 — the toy example, new numbers (k-means by hand)

Six new 2D points:

```{webr}
#| autorun: false
# p1-starter — edit and Run to check your hand work
x <- c(2, 3, 2.5, 8, 9, 8.5)
y <- c(2, 2.5, 3, 8, 8.5, 9)
dat <- data.frame(x = x, y = y)
```

With $k = 2$ and initial centroids at $(2, 2)$ and $(9, 9)$:

a. Assign each point to its nearest centroid.
b. Compute the new centroids.
c. Do a second iteration. Does anything change?
d. Compute the WCSS.

::: {.callout-note collapse="true" title="Answer"}
**a. Assign (iteration 1):**

- Point 1 $(2, 2)$: dist to C1 $(2,2)$ = 0; dist to C2 $(9,9)$ = $\sqrt{49+49} \approx 9.90$. Cluster 1.
- Point 2 $(3, 2.5)$: dist to C1 = $\sqrt{1+0.25} \approx 1.12$; dist to C2 = $\sqrt{36+42.25} \approx 8.85$. Cluster 1.
- Point 3 $(2.5, 3)$: dist to C1 = $\sqrt{0.25+1} \approx 1.12$; dist to C2 = $\sqrt{42.25+36} \approx 8.85$. Cluster 1.
- Point 4 $(8, 8)$: dist to C1 = $\sqrt{36+36} \approx 8.49$; dist to C2 = $\sqrt{1+1} \approx 1.41$. Cluster 2.
- Point 5 $(9, 8.5)$: dist to C1 = $\sqrt{49+42.25} \approx 9.56$; dist to C2 = $\sqrt{0+0.25} = 0.5$. Cluster 2.
- Point 6 $(8.5, 9)$: dist to C1 = $\sqrt{42.25+49} \approx 9.56$; dist to C2 = $\sqrt{0.25+0} = 0.5$. Cluster 2.

Cluster 1: {1, 2, 3}. Cluster 2: {4, 5, 6}.

**b. New centroids:**

- C1: $((2+3+2.5)/3,\;(2+2.5+3)/3) = (7.5/3,\;7.5/3) = (2.5, 2.5)$.
- C2: $((8+9+8.5)/3,\;(8+8.5+9)/3) = (25.5/3,\;25.5/3) = (8.5, 8.5)$.

**c. Iteration 2:** All points are still much closer to their own centroid. No
reassignments. Converged.

**d. WCSS:**

- Cluster 1: $(2-2.5)^2+(2-2.5)^2 + (3-2.5)^2+(2.5-2.5)^2 + (2.5-2.5)^2+(3-2.5)^2 = 0.5 + 0.25 + 0.25 = 1.0$
- Cluster 2: $(8-8.5)^2+(8-8.5)^2 + (9-8.5)^2+(8.5-8.5)^2 + (8.5-8.5)^2+(9-8.5)^2 = 0.5 + 0.25 + 0.25 = 1.0$
- Total WCSS = $1.0 + 1.0 = 2.0$.

```{r p1-answer}
x <- c(2, 3, 2.5, 8, 9, 8.5)
y <- c(2, 2.5, 3, 8, 8.5, 9)
dat <- data.frame(x = x, y = y)
set.seed(42)
km <- kmeans(dat, centers = 2, nstart = 25)
km$cluster    # the two clusters
km$centers    # centroids at (2.5, 2.5) and (8.5, 8.5)
km$tot.withinss  # 2.0
```
:::

## Problem 2 — fill in the missing step (hierarchical merge)

Here is the start of hierarchical clustering (single linkage) on these 4 points:

```{webr}
#| autorun: false
# p2-starter — use this data to check the blank
pts <- data.frame(x = c(0, 1, 5, 6), y = c(0, 1, 5, 6))
```

Step 1: Compute all pairwise distances.

- $d(1,2) = \sqrt{1+1} \approx 1.41$
- $d(1,3) = \sqrt{25+25} \approx 7.07$
- $d(1,4) = \sqrt{36+36} \approx 8.49$
- $d(2,3) = \sqrt{16+16} \approx 5.66$
- $d(2,4) = \sqrt{25+25} \approx 7.07$
- $d(3,4) = \sqrt{1+1} \approx 1.41$

Step 2: The two closest pairs are {1,2} and {3,4}, both at distance 1.41. Merge {1,2}.

Step 3: **Your step here** — compute the single-linkage distance from the merged
cluster {1,2} to each remaining cluster ({3} and {4}), and decide which merge happens
next.

Step 4: After all merges, cut at $k = 2$. The two clusters are {1,2} and {3,4}.

Fill in step 3.

::: {.callout-note collapse="true" title="Answer"}
**Step 3:** Using single linkage (minimum distance between any member of one cluster
and any member of the other):

- $d(\{1,2\}, \{3\}) = \min(d(1,3), d(2,3)) = \min(7.07, 5.66) = 5.66$
- $d(\{1,2\}, \{4\}) = \min(d(1,4), d(2,4)) = \min(8.49, 7.07) = 7.07$
- $d(\{3\}, \{4\}) = 1.41$

The smallest inter-cluster distance is $d(\{3\}, \{4\}) = 1.41$, so {3} and {4}
merge next. Now we have two clusters: {1,2} and {3,4}.

The final merge at distance $d(\{1,2\}, \{3,4\}) = \min(7.07, 5.66, 8.49, 7.07)
= 5.66$ joins everything. The dendrogram would show merges at heights 1.41, 1.41,
and 5.66.

```{r p2-answer}
#| fig-alt: "Dendrogram of 4 points under single linkage, showing two merges at height 1.41 and a final merge at height 5.66."
pts <- data.frame(x = c(0, 1, 5, 6), y = c(0, 1, 5, 6))
d <- dist(pts)
hc <- hclust(d, method = "single")
plot(hc, main = "Dendrogram (single linkage)", xlab = "", sub = "",
     labels = paste0("pt", 1:4))
cutree(hc, k = 2)  # 1 1 2 2
```
:::

## Problem 3 — implement k-means from scratch

Write your own k-means in R: a `while` loop that alternates between assigning points
to the nearest centroid and recomputing centroids, stopping when assignments stop
changing. Use the data below with $k = 2$. Compare your result with `kmeans()`.

```{webr}
#| autorun: false
# p3-starter — your job: implement k-means from scratch
x <- c(1, 2, 1.5, 5, 6, 5.5)
y <- c(1, 1.5, 2, 5, 5.5, 6)
dat <- data.frame(x = x, y = y)

# Start with these centroids
centroids <- matrix(c(1, 1, 6, 6), nrow = 2, byrow = TRUE)

# Your code here: a while loop that:
# 1. Computes distance from each point to each centroid
# 2. Assigns each point to the nearest centroid
# 3. Recomputes centroids as means of assigned points
# 4. Stops when assignments don't change
```

::: {.callout-note collapse="true" title="Answer"}
```{r p3-answer}
x <- c(1, 2, 1.5, 5, 6, 5.5)
y <- c(1, 1.5, 2, 5, 5.5, 6)
dat <- data.frame(x = x, y = y)

# Starting centroids
centroids <- matrix(c(1, 1, 6, 6), nrow = 2, byrow = TRUE)
colnames(centroids) <- c("x", "y")

cluster <- rep(0, nrow(dat))  # initial (will be overwritten)
converged <- FALSE
iter <- 0

while (!converged) {
  iter <- iter + 1

  # Step 1: distance from each point to each centroid
  dist_c1 <- sqrt((dat$x - centroids[1, 1])^2 + (dat$y - centroids[1, 2])^2)
  dist_c2 <- sqrt((dat$x - centroids[2, 1])^2 + (dat$y - centroids[2, 2])^2)

  # Step 2: assign to nearest centroid
  new_cluster <- ifelse(dist_c1 <= dist_c2, 1, 2)

  # Step 3: check convergence (did assignments change?)
  converged <- all(new_cluster == cluster)
  cluster <- new_cluster

  # Step 4: update centroids (mean of assigned points)
  centroids[1, ] <- colMeans(dat[cluster == 1, ])
  centroids[2, ] <- colMeans(dat[cluster == 2, ])

  cat("Iteration", iter, "- clusters:", cluster, "\n")
}

cat("Converged after", iter, "iterations\n")
cat("Final centroids:\n")
print(centroids)   # (1.5, 1.5) and (5.5, 5.5)

# Compare with kmeans()
set.seed(42)
km <- kmeans(dat, centers = 2, nstart = 25)
cat("kmeans() clusters:", km$cluster, "\n")
cat("kmeans() WCSS:", km$tot.withinss, "\n")  # 2.0

# WCSS from our implementation
wcss <- sum((dat$x[cluster==1] - centroids[1,1])^2 +
            (dat$y[cluster==1] - centroids[1,2])^2) +
        sum((dat$x[cluster==2] - centroids[2,1])^2 +
            (dat$y[cluster==2] - centroids[2,2])^2)
cat("Our WCSS:", wcss, "\n")  # 2.0 — matches
```
:::

## Problem 4 — spot the error

A colleague clusters 9 one-dimensional measurements that fall into three tight,
well-separated groups (around 0, around 5, and around 10 — the same data as the
lesson's Trap 1). They run this code:

```{webr}
#| autorun: false
# p4-starter — the colleague's exact analysis
pts <- data.frame(val = c(0, 0.1, -0.1, 5, 5.1, 4.9, 10, 10.1, 9.9))
set.seed(1)
km <- kmeans(pts, centers = 3, nstart = 1)
km$cluster       # the clustering they reported
km$tot.withinss  # the quality they reported
```

They report: "`kmeans()` converged, so this is the best 3-cluster solution for this
data — WCSS is about 37.5."

This analysis contains **exactly one mistake**. Find it and explain the fix.

::: {.callout-note collapse="true" title="Answer"}
**The mistake:** trusting a **single random start** (`nstart = 1`). "Converged" only
means k-means reached a *local* minimum of WCSS — it says nothing about whether a
better clustering exists. With this unlucky start, the colleague's run lumps the
groups at 0 and 5 into one big cluster and splits the group at 10 into two
(assignments `2 2 2 2 2 2 1 3 1`), giving WCSS $\approx 37.5$.
[@boyd2018appliedlinearalgebra, p. 76; @james2021introductionstatisticallearning, pp. 517-518]

**The fix:** run from many random starts with `nstart = 25` (or more) and let
`kmeans()` keep the run with the lowest `tot.withinss`:

```{r p4-answer}
pts <- data.frame(val = c(0, 0.1, -0.1, 5, 5.1, 4.9, 10, 10.1, 9.9))

# The colleague's run: one random start
set.seed(1)
bad <- kmeans(pts, centers = 3, nstart = 1)
bad$cluster        # 2 2 2 2 2 2 1 3 1 — two groups merged, one split
bad$tot.withinss   # 37.545 — a poor local minimum

# The fix: 25 random starts, keep the best
set.seed(42)
good <- kmeans(pts, centers = 3, nstart = 25)
good$cluster       # 1 1 1 2 2 2 3 3 3 — the three real groups
good$tot.withinss  # 0.06 — about 600 times smaller
good$centers       # 0, 5, 10 — right on the group centers
```

The correct clustering finds the three real groups with WCSS $0.06$ — about 600
times smaller than the reported $37.545$. Same function, same data, same $k$; the
only thing that changed is not betting everything on one starting position. Whenever
a clustering matters, run it with `nstart = 25` or more and compare `tot.withinss`.
:::

## Problem 5 — mixing it with distance *(uses distance-similarity-and-geometry's Manhattan distance)*

Recluster the toy data using **Manhattan distance** instead of Euclidean distance for
hierarchical clustering. Do the 2-cluster assignments change? Why or why not?

```{webr}
#| autorun: false
# p5-starter — edit and Run to compare
x <- c(1, 2, 1.5, 5, 6, 5.5)
y <- c(1, 1.5, 2, 5, 5.5, 6)
dat <- data.frame(x = x, y = y)

# Euclidean (default)
hc_euc <- hclust(dist(dat), method = "single")

# Manhattan distance
hc_man <- hclust(dist(dat, method = "manhattan"), method = "single")

# Compare 2-cluster cuts
# Your prediction: same or different?
```

::: {.callout-note collapse="true" title="Answer"}
```{r p5-answer}
#| fig-alt: "Two dendrograms side by side comparing Euclidean and Manhattan distance hierarchical clustering, both giving the same two-cluster solution."
x <- c(1, 2, 1.5, 5, 6, 5.5)
y <- c(1, 1.5, 2, 5, 5.5, 6)
dat <- data.frame(x = x, y = y)

# Euclidean hierarchical clustering
hc_euc <- hclust(dist(dat), method = "single")
cat("Euclidean, k=2:", cutree(hc_euc, k = 2), "\n")

# Manhattan hierarchical clustering
hc_man <- hclust(dist(dat, method = "manhattan"), method = "single")
cat("Manhattan, k=2:", cutree(hc_man, k = 2), "\n")

# The merge heights differ but the cluster assignments are the same
par(mfrow = c(1, 2))
plot(hc_euc, main = "Euclidean", xlab = "", sub = "")
plot(hc_man, main = "Manhattan", xlab = "", sub = "")
par(mfrow = c(1, 1))
```

The 2-cluster assignments are **the same**: {1,2,3} and {4,5,6}. The two groups are
so well separated that both distance metrics agree on which points belong together.
The merge heights change — Manhattan distance between $(1,1)$ and $(2,1.5)$ is
$|1| + |0.5| = 1.5$ instead of Euclidean $\sqrt{1.25} \approx 1.12$ — but the
*ordering* of which pairs are closest stays the same. Manhattan adds the absolute gaps
instead of squaring, adding, and rooting (from
[distance, similarity & geometry](../distance-similarity-and-geometry/lesson.qmd)),
so the numbers differ but the ranking is preserved when clusters are well separated.
With more ambiguous data the choice of distance metric could change the clustering.
:::

## Problem 6 — reading a dendrogram (transfer)

Here is the dendrogram from the toy data (complete linkage). Without running any code,
answer:

a. If you cut at $k = 3$, which points form the three clusters?
b. If you cut at $k = 2$, which points form the two clusters?
c. Which cut looks more natural — $k = 2$ or $k = 3$? Why?

```{webr}
#| autorun: false
#| fig-alt: "Dendrogram from complete-linkage hierarchical clustering of six points, showing merge heights."
# p6-starter — run this to see the dendrogram, then answer from the picture
x <- c(1, 2, 1.5, 5, 6, 5.5)
y <- c(1, 1.5, 2, 5, 5.5, 6)
dat <- data.frame(x = x, y = y)
d <- dist(dat)
hc <- hclust(d, method = "complete")
plot(hc, main = "Dendrogram (complete linkage)", xlab = "", sub = "",
     labels = paste0("pt", 1:6))
# Add horizontal lines at the cut heights to help visualize
abline(h = 3, col = "red", lty = 2)   # a cut around here gives 2 clusters
abline(h = 1, col = "blue", lty = 2)  # a cut around here gives more clusters
```

::: {.callout-note collapse="true" title="Answer"}
**a. k = 3:** Cutting at a height that splits one of the two main branches further.
With complete linkage: points 2 and 3 merge first (they are closest), then point 1
joins them. On the other side, points 5 and 6 merge, then point 4 joins. At $k = 3$,
one of the branches stays intact while the other splits, giving something like
{1}, {2,3}, {4,5,6} or {1,2,3}, {4}, {5,6} depending on which side splits at a lower
height.

```{r p6-answer}
x <- c(1, 2, 1.5, 5, 6, 5.5)
y <- c(1, 1.5, 2, 5, 5.5, 6)
dat <- data.frame(x = x, y = y)
d <- dist(dat)
hc <- hclust(d, method = "complete")
cat("k=3:", cutree(hc, k = 3), "\n")
cat("k=2:", cutree(hc, k = 2), "\n")
```

**b. k = 2:** {1,2,3} and {4,5,6} — the two obvious groups.

**c. k = 2 looks more natural.** The dendrogram shows a large vertical gap between
the within-group merges (at heights below ~1.5) and the final between-group merge (at
a height above 6). That big gap indicates a strong, natural separation into two
groups. Going to $k = 3$ splits one of those tight groups that has no reason to be
split — there is no large gap within either branch to justify it.
:::

## Problem 7 — which scaling is "right"? (transfer)

An online retailer sells two things: socks and computers. For six customers, you
have last year's purchase counts. [@james2021introductionstatisticallearning, pp. 527-528]

```{webr}
#| autorun: false
# p7-starter — copy this to begin
socks     <- c(2, 3, 1, 11, 12, 10)   # pairs of socks each customer bought
computers <- c(0, 1, 1, 0, 1, 0)      # computers each customer bought
shop <- data.frame(socks = socks, computers = computers)
```

Cluster the customers into $k = 2$ groups (use `nstart = 25`) three different ways,
and **predict each result before you run it**:

a. **Raw counts**, as given. Which customers group together — and which column
   decided that?
b. **Scaled to unit SD** (`scale(shop)`). What changes about each column's
   contribution to the distance? Does the 2-cluster split change?
c. **Dollars spent**: socks cost \$2 a pair, computers cost \$2,000 each
   (`shop$socks * 2` and `shop$computers * 2000`). Now which column decides?
d. **Interpretation:** one clustering changed and two didn't — which, and why? Is
   any of the three versions the "correct" one?

::: {.callout-note collapse="true" title="Answer"}
```{r p7-answer}
socks     <- c(2, 3, 1, 11, 12, 10)
computers <- c(0, 1, 1, 0, 1, 0)
shop <- data.frame(socks = socks, computers = computers)

# a. Raw counts — the sock column has all the spread
apply(shop, 2, var)   # socks 25.1, computers 0.3 — socks carry ~99% of the variance
set.seed(42)
kmeans(shop, centers = 2, nstart = 25)$cluster        # 1 1 1 2 2 2

# b. Scaled to unit SD — both columns now run about -0.9 to +1.1
shop_scaled <- scale(shop)
round(shop_scaled, 2)
set.seed(42)
kmeans(shop_scaled, centers = 2, nstart = 25)$cluster  # 1 1 1 2 2 2 — same split

# c. Dollars — the computer column now has almost all the spread
shop_dollars <- data.frame(socks = socks * 2, computers = computers * 2000)
apply(shop_dollars, 2, var)  # socks 100.4, computers 1,200,000
set.seed(42)
kmeans(shop_dollars, centers = 2, nstart = 25)$cluster # 1 2 2 1 2 1 — different!
```

**a. Raw counts:** {1, 2, 3} vs {4, 5, 6} — the light sock buyers vs the heavy sock
buyers. The sock column decided alone: its variance (25.1) dwarfs the computer
column's (0.3), so sock differences dominate every Euclidean distance and computer
purchases barely register.

**b. Scaled:** each column now has SD 1, so a "one computer" difference finally
counts as much as a proportional sock difference — both columns genuinely contribute
to the distances. On this data the best 2-cluster split happens to stay
{1, 2, 3} vs {4, 5, 6}: even with both columns weighted equally, the sock gap between
the two customer groups is the strongest pattern. Same partition, but for a better
reason — it now survives with both variables heard, instead of winning by drowning
the other out.

**c. Dollars:** {1, 4, 6} vs {2, 3, 5} — a *different* clustering: the
computer-buyers vs the non-buyers. A single computer is a \$2,000 gap while the
entire sock range spans \$2 to \$24, so computer ownership dominates every distance.

**d.** The dollar version changed the grouping; raw and scaled agreed here (though
they weighted the columns very differently). And none of the three is "the correct
one" — each answers a different question. Raw counts cluster by *items handled*
(inventory/shipping view), unit-SD scaling gives *each product an equal vote*
(product-mix view), and dollars cluster by *revenue* (finance view). Scaling is not
a technicality you get right or wrong; it encodes what you, the analyst, mean by
"similar customers" — which is exactly why a clustering report must state the choice
it made. [@james2021introductionstatisticallearning, pp. 527-528]
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/clustering-k-means-and-hierarchical" pass="4"}

::: {.quiz-q qid="clust-centroid-update" qtype="mc" answer="c"}
After the assign step of k-means, what does the update step compute for each cluster?

a) The point in the cluster farthest from all other clusters
b) The median of each coordinate across the cluster's points
c) The mean of each coordinate across the cluster's points
d) The point in the cluster closest to all other points in the dataset
:::

::: {.quiz-q qid="clust-wcss-toy" qtype="numeric" answer="2" tolerance="0"}
For the lesson's toy data (6 points, $k = 2$, centroids at $(1.5, 1.5)$ and $(5.5, 5.5)$), what is the total WCSS?
:::

::: {.quiz-q qid="clust-linkage-choice" qtype="mc" answer="b"}
Which linkage method measures the distance between two clusters as the distance between their two closest members?

a) Complete linkage
b) Single linkage
c) Ward's linkage
d) Average linkage
:::

::: {.quiz-q qid="clust-nstart" qtype="mc" answer="d"}
Why should you use `nstart = 25` (or more) when running `kmeans()` in R?

a) It searches over 25 different values of k and picks the best cluster count
b) It speeds up convergence by spreading the work across 25 parallel runs
c) It guarantees the algorithm reaches the global minimum WCSS on every run
d) It runs 25 random starts and keeps the lowest-WCSS result
:::

::: {.quiz-q qid="clust-dist-scale" qtype="mc" answer="a"}
*(Interleaved from distance-similarity-and-geometry.)* Before running k-means on a dataset where column A ranges from 0 to 1 and column B ranges from 0 to 100,000, what must you do first?

a) Scale both columns so neither dominates the distance
b) Remove column B because its range is too large for clustering
c) Use Manhattan distance, which handles different scales automatically
d) Set nstart to a very large number to compensate for the scale difference
:::

:::

::::
