---
title: "Deriving the OLS formulas"
subtitle: "The 'take the derivative, set it to zero' trick — taught, then used for real"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| message: false
# Packages this lesson uses. Missing ones are installed automatically.
needed <- c()   # base R is all we need
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

## Where we're going

Last module ended at a valley: SSE plotted against candidate slopes, with the best
line at the bottom. This module finds the bottom *exactly*. The tool is the single
most reused move in all of applied math — **take the derivative, set it to zero,
solve** — and by the end, that move will hand us the famous formulas
$\hat{\beta}_1 = s_{xy}/s_x^2$ and $\hat{\beta}_0 = \bar{y} - \hat{\beta}_1 \bar{x}$.
Every algebra step on the way uses a trick you've already been taught (mostly the
sum rules), and each step says which.

**Builds on:** [module 01](../01-least-squares-idea/lesson.qmd) (SSE, argmin),
[vectors & summation](../../../../foundations/vectors-and-summation/lesson.qmd)
(sum Rules 1–3),
[derivative as slope](../../../../foundations/derivative-as-slope/lesson.qmd),
[mean, variance & covariance](../../../../foundations/mean-variance-covariance/lesson.qmd),
[loss functions & optimization](../../../../foundations/loss-functions-and-optimization/lesson.qmd)
(the "differentiate, set to zero, solve" recipe),
[differentiation rules](../../../../foundations/differentiation-rules/lesson.qmd)
(power rule, chain rule), and
[partial derivatives](../../../../foundations/partial-derivatives-and-the-gradient/lesson.qmd).

## Warm-up: do you remember?

Everything this derivation leans on, retrieved first. Produce each answer before you
open it.

**1.** *(module 1)* In one sentence: what does $\mathrm{SSE}$ measure, and what does
"the best line" mean in least squares?

::: {.callout-note collapse="true" title="Answer"}
SSE (sum of squared errors) is a line's *badness score* — the sum of its squared
misses, $\sum (y_i - \hat{y}_i)^2$. The best line is the $(b_0, b_1)$ pair that makes
that score smallest (the arg-min — the bottom of the valley). See
[module 1](../01-least-squares-idea/lesson.qmd).
:::

**2.** *(vectors & summation)* A constant $b_0$ is added inside a sum over $n$ terms:
what is $\sum_{i=1}^{n} b_0$?

::: {.callout-note collapse="true" title="Answer"}
$n\,b_0$ — the constant appears once per term, $n$ times over (sum-Rule 3). We'll use
this exact move in Step 1. See
[vectors & summation](../../../../foundations/vectors-and-summation/lesson.qmd).
:::

**3.** *(mean, variance & covariance)* Fill in the blank: for any list, the deviations
from the mean sum to ______ , i.e. $\sum_{i=1}^{n}(x_i - \bar{x}) = $ ______ .

::: {.callout-note collapse="true" title="Answer"}
Zero: $\sum (x_i - \bar{x}) = 0$. This is the "centering trick" that turns the raw
slope formula into covariance-over-variance in Step 3. See
[mean, variance & covariance](../../../../foundations/mean-variance-covariance/lesson.qmd).
:::

## A tiny example first

Forget regression for a moment. Here's the smallest possible valley problem:

> $f(b) = (b - 3)^2 + 2$. Which $b$ makes $f$ smallest?

Try a few values by hand: $f(1) = (-2)^2 + 2 = 6$; $f(2) = 1 + 2 = 3$;
$f(3) = 0 + 2 = 2$; $f(4) = 1 + 2 = 3$; $f(5) = 6$. The scores fall, bottom out at
$b = 3$, and rise again — a valley with its bottom at 3.

Now the calculus way, no guessing. From the
[loss-functions-and-optimization](../../../../foundations/loss-functions-and-optimization/lesson.qmd)
foundation: the derivative is the slope of the curve, and at the exact bottom of a
valley the slope is zero — so differentiate, set to zero, solve.

Two rules from
[differentiation-rules](../../../../foundations/differentiation-rules/lesson.qmd)
do all the work:

- **Power rule:** the derivative of $b^2$ is $2b$.
- **Chain rule:** the derivative of $(\text{stuff})^2$ is
  $2 \cdot (\text{stuff}) \cdot (\text{derivative of the stuff})$.

Apply to our $f$: the derivative of $(b-3)^2$ is $2(b-3) \cdot 1$ (the inside
$b - 3$ has derivative 1), and the $+2$ is a constant — flat, derivative 0. So
$f'(b) = 2(b - 3)$. Set it to zero: $2(b - 3) = 0$, so $b = 3$. The bottom, exactly,
in two lines — no grid search.

The same thing in R, so you can check every step:

```{r toy-example}
#| fig-alt: "A parabola with its minimum at b = 3 where the slope is zero"
# The tiny valley: f(b) = (b - 3)^2 + 2
f <- function(b) (b - 3)^2 + 2

# Guess-and-check, like we did by hand
f(c(1, 2, 3, 4, 5))          # 6 3 2 3 6 — bottom at b = 3

# The derivative we computed: f'(b) = 2(b - 3). Zero exactly at b = 3.
fprime <- function(b) 2 * (b - 3)
fprime(c(1, 2, 3, 4, 5))     # -4 -2 0 2 4 — negative going in, zero at bottom

b <- seq(0, 6, 0.01)
plot(b, f(b), type = "l", lwd = 2, main = "The slope is 0 exactly at the bottom")
abline(v = 3, col = "firebrick", lty = 2)
```

One more idea and we're armed. SSE depends on TWO knobs, $b_0$ and $b_1$. From
[partial derivatives](../../../../foundations/partial-derivatives-and-the-gradient/lesson.qmd):
a partial derivative takes the slope in one direction while holding the other knobs
still. A 2-knob valley's bottom is where BOTH partial slopes are zero — two
equations, two unknowns.

## Getting a feel for it

**A ball in a bowl.** Picture the SSE valley from module 1 as a smooth bowl and drop
a marble in. It rolls downhill, overshoots, rolls back, and finally settles at one
spot — and that resting spot is the *only* place where the ground under it is
perfectly **level**. Anywhere the ground tilts, the marble keeps rolling. That's the
whole idea in one picture: the bottom of a valley is the place where the ground is
flat.

**Why "derivative equals zero" is exactly that.** The derivative is nothing but the
steepness of the ground — how tilted the curve is at a point. So "level ground"
*is* "derivative equals zero." Instead of the module-1 grid search (score 301
candidate slopes and eyeball the lowest), we can ask algebra directly: *where is the
ground level?* Set the slope-of-the-curve to zero and solve. In our toy valley
$f(b) = (b-3)^2 + 2$, the steepness is $f'(b) = 2(b-3)$: it reads $-4$ at $b=1$ and
$-2$ at $b=2$ (ground tilting down, marble rolls right), shrinks to $0$ exactly at
$b=3$, then goes positive past it. The steepness *changing sign through zero* is the
marble arriving at the bottom and stopping.

**Why the bowl shape is what lets this work.** Squaring the misses (module 1) is what
makes the SSE surface a bowl — curving *up* on every side — so it has exactly one
level spot, and that spot is a bottom, not a peak or a saddle. That's the payoff of
squaring showing up again: because SSE is a bowl, "find where the slope is zero" is
guaranteed to find *the* minimum, and with two knobs we just do it in both
directions at once (both partial slopes zero). The rest of this lesson is that single
move, carried out carefully with the sum rules doing the bookkeeping.

## What the notation is going to say

The derivation below says: "write down the SSE score from module 1; take its
partial derivative with respect to the intercept, set it to zero, and simplify with
the sum rules — that gives one equation. Do the same for the slope — a second
equation. Solve the pair: the intercept equation says *the best line passes through
the point* $(\bar{x}, \bar{y})$, and the slope equation then simplifies to
*covariance over variance*."

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $x_i$ | "ex sub eye" | the i-th predictor value (in R: `x[i]`) | 1, 2, 3, 4 |
| $y_i$ | "why sub eye" | the i-th observed outcome (in R: `y[i]`) | 2, 3, 5, 6 |
| $n$ | "en" | the number of data points (in R: `length(x)`) | 4 |
| $b_0, b_1$ | "bee-zero, bee-one" | the intercept and slope *knobs* we're minimizing over (any candidate line) | swept during the search |
| $f'(b)$ | "f prime of b" | the derivative: the slope of $f$ at $b$ | $f'(b) = 2(b-3)$ |
| $\frac{\partial}{\partial b_0}$ | "the partial derivative with respect to bee-zero" | slope in the $b_0$ direction, holding $b_1$ still | see derivation |
| $\mathrm{SSE}(b_0, b_1)$ | "the SSE at bee-zero, bee-one" | the score as a function of both knobs | $\sum (y_i - b_0 - b_1 x_i)^2$ |
| $\hat{\beta}_0, \hat{\beta}_1$ | "beta-zero hat, beta-one hat" | the solution: the arg-min pair | 0.5 and 1.4 (verified below) |
| $\bar{x}, \bar{y}$ | "ex bar, why bar" | means of x and y | 2.5 and 4 |
| $s_{xy}$ | "the covariance of x and y" | $\frac{1}{n-1}\sum(x_i-\bar{x})(y_i-\bar{y})$ | $7/3$ |
| $s_x^2$ | "the variance of x" | $\frac{1}{n-1}\sum(x_i-\bar{x})^2$ | $5/3$ |
| $\Rightarrow$ | "which implies" / "therefore" | the next line follows from this one | between derivation lines |

## The formal version

The problem, from module 1 (tomato data $x = (1,2,3,4)$, $y = (2,3,5,6)$ as our
running check):

$$
(\hat{\beta}_0, \hat{\beta}_1) = \arg\min_{b_0,\, b_1} \;\underbrace{\sum_{i=1}^{n} (y_i - b_0 - b_1 x_i)^2}_{\mathrm{SSE}(b_0,\, b_1)}
$$

### Step 1 — partial derivative with respect to $b_0$, set to zero

$$
\begin{aligned}
\frac{\partial\, \mathrm{SSE}}{\partial b_0}
&= \sum_{i=1}^{n} 2\,(y_i - b_0 - b_1 x_i)\cdot(-1) \\
&= -2 \sum_{i=1}^{n} (y_i - b_0 - b_1 x_i)
\end{aligned}
$$

- Line 1: chain rule, lite — derivative of (stuff)² is 2·(stuff)·(derivative of
  stuff). Treating $b_1$ as a constant, the "stuff" $y_i - b_0 - b_1 x_i$ has
  derivative $-1$ with respect to $b_0$. Also: the derivative of a sum is the sum
  of the derivatives (differentiate each term of the Σ separately).
- Line 2: pull the constant $-2$ out of the sum (sum Rule 2).

Set it to zero and simplify (dividing both sides by $-2$ first):

$$
\begin{aligned}
\sum_{i=1}^{n} (y_i - b_0 - b_1 x_i) &= 0 \\
\Rightarrow\quad \sum y_i \;-\; n\,b_0 \;-\; b_1 \sum x_i &= 0 \\
\Rightarrow\quad \bar{y} - b_0 - b_1 \bar{x} &= 0 \\
\Rightarrow\quad \boxed{\;\hat{\beta}_0 = \bar{y} - \hat{\beta}_1\, \bar{x}\;}
\end{aligned}
$$

- Line 2: split the sum into three sums (Rule 1); $b_0$ summed n times is $n b_0$
  (Rule 3); pull $b_1$ out (Rule 2).
- Line 3: divide the whole equation by $n$; $\frac{1}{n}\sum y_i$ is $\bar{y}$ and
  $\frac{1}{n}\sum x_i$ is $\bar{x}$ — the mean's definition.
- Line 4: solve for $b_0$ (move terms across the equals sign).

Plain English: **the best line always passes through the point of means**
$(\bar{x}, \bar{y})$ — pick any slope, and the best intercept for it just slides
the line through that anchor point.

### Step 2 — partial derivative with respect to $b_1$, set to zero

$$
\begin{aligned}
\frac{\partial\, \mathrm{SSE}}{\partial b_1}
&= \sum_{i=1}^{n} 2\,(y_i - b_0 - b_1 x_i)\cdot(-x_i)
\;=\; -2 \sum_{i=1}^{n} x_i\,(y_i - b_0 - b_1 x_i)
\end{aligned}
$$

- Chain rule again — this time the inside's derivative with respect to $b_1$ is
  $-x_i$; then pull out the constant $-2$ (Rule 2).

Set to zero, then substitute Step 1's result $b_0 = \bar{y} - b_1 \bar{x}$:

$$
\begin{aligned}
\sum x_i \big(y_i - (\bar{y} - b_1 \bar{x}) - b_1 x_i\big) &= 0 \\
\Rightarrow\quad \sum x_i \big( (y_i - \bar{y}) - b_1 (x_i - \bar{x}) \big) &= 0 \\
\Rightarrow\quad \sum x_i (y_i - \bar{y}) \;-\; b_1 \sum x_i (x_i - \bar{x}) &= 0 \\
\Rightarrow\quad \hat{\beta}_1 = \frac{\sum x_i (y_i - \bar{y})}{\sum x_i (x_i - \bar{x})}
\end{aligned}
$$

- Line 1: substitution (replace $b_0$ with what Step 1 proved it must be).
- Line 2: regroup the inside: $y_i - \bar{y} + b_1\bar{x} - b_1 x_i$ becomes
  $(y_i - \bar{y}) - b_1(x_i - \bar{x})$ — just collecting the $b_1$ terms (factoring).
- Line 3: split the sum (Rule 1) and pull $b_1$ out (Rule 2).
- Line 4: solve for $b_1$.

### Step 3 — the centering trick makes it recognizable

The last expression doesn't look like covariance/variance yet. One small lemma
fixes both top and bottom. Remember from the foundations module that deviations sum
to zero: $\sum (y_i - \bar{y}) = 0$. Multiply that zero by the constant $\bar{x}$:

$$
\begin{aligned}
\sum (x_i - \bar{x})(y_i - \bar{y})
&= \sum x_i (y_i - \bar{y}) - \bar{x} \underbrace{\sum (y_i - \bar{y})}_{=\,0} \\
&= \sum x_i (y_i - \bar{y})
\end{aligned}
$$

- Line 1: expand the first factor and split the sum (Rule 1), pulling the constant
  $\bar{x}$ out (Rule 2).
- Line 2: the second term is $\bar{x} \times 0 = 0$ and vanishes.

So the numerator can be rewritten with $(x_i - \bar{x})$ in it for free; the same
argument with $y \to x$ rewrites the denominator. Therefore:

$$
\hat{\beta}_1
= \frac{\sum (x_i - \bar{x})(y_i - \bar{y})}{\sum (x_i - \bar{x})^2}
= \frac{s_{xy}}{s_x^2}
\qquad\text{and}\qquad
\hat{\beta}_0 = \bar{y} - \hat{\beta}_1\,\bar{x}
$$

(For the last equality: divide top and bottom by the same $n - 1$ — it cancels,
which is why the covariance/variance form and the sum form are the same number.)

Reading it aloud: "beta-one hat equals the covariance of x and y, divided by the
variance of x" — *the slope is: how much x and y move together, per unit of how
much x moves on its own.*

Mapping back to the toy numbers: $\bar{x} = 2.5$, $\bar{y} = 4$,
$s_{xy} = 7/3$, $s_x^2 = 5/3$, so
$\hat{\beta}_1 = (7/3)/(5/3) = 7/5 = 1.4$ and
$\hat{\beta}_0 = 4 - 1.4 \times 2.5 = 0.5$. Module 1's grid search found exactly
this pair — now you know why.

## Explore it

```{webr}
#| autorun: true
# explore-1 — LIVE: the formulas, straight from the derivation — then lm() as the referee.
# PREDICT: will the by-hand slope (cov/var) match lm()'s slope exactly, or only
# roughly? Write your guess, then read the two lines of output below.
x <- c(1, 2, 3, 4)
y <- c(2, 3, 5, 6)

b1_hat <- cov(x, y) / var(x)          # slope: covariance over variance
b0_hat <- mean(y) - b1_hat * mean(x)  # intercept: through the point of means

c(b0 = b0_hat, b1 = b1_hat)           # 0.5 and 1.4
coef(lm(y ~ x))                       # lm() agrees exactly
```

```{webr}
#| autorun: true
# explore-2 — LIVE: swap in ANY data — the agreement is not a coincidence of the toy
# numbers. mtcars: predict fuel efficiency (mpg) from weight (wt).
x <- mtcars$wt
y <- mtcars$mpg

c(by_hand = cov(x, y) / var(x),
  lm      = unname(coef(lm(y ~ x))[2]))     # identical slopes

# And the anchor-point fact from Step 1: the fitted line passes through
# (mean of x, mean of y). Predict at x-bar and compare with y-bar:
fit <- lm(y ~ x)
c(prediction_at_xbar = unname(predict(fit, data.frame(x = mean(x)))),
  ybar               = mean(y))              # identical
```

```{webr}
#| autorun: true
#| fig-alt: "SSE valley over slopes with the derivative-derived optimum marked"
# explore-3 — LIVE: module 1's valley, revisited — but now the red line is placed by the
# FORMULA, not by search. Change y's noise (sd below) and re-run: the formula
# tracks the valley bottom perfectly every time.
# PREDICT: the true slope used to build y is 2. Will the formula's red line land
# exactly on 2, or a little off? Guess, then read the value in the plot title.
set.seed(42)
x <- 1:20
y <- 3 + 2 * x + rnorm(20, sd = 4)

b1_hat <- cov(x, y) / var(x)
b0_best_for <- function(b1) mean(y) - b1 * mean(x)   # Step 1, reused
sse_of <- function(b1) sum((y - (b0_best_for(b1) + b1 * x))^2)

slopes <- seq(0, 4, 0.01)
plot(slopes, sapply(slopes, sse_of), type = "l", lwd = 2,
     xlab = "candidate slope b1", ylab = "SSE (with best b0 for each slope)",
     main = sprintf("Formula says bottom at b1 = %.3f", b1_hat))
abline(v = b1_hat, col = "firebrick", lty = 2)
```

## Common traps

**Trap 1 — dropping the inside-derivative factor of the chain rule.** It's easy to
differentiate $(y_i - b_0 - b_1 x_i)^2$ as just $2(y_i - b_0 - b_1 x_i)$ and forget
the $\cdot(-x_i)$ that the chain rule tacks on for the $b_1$ direction. That missing
factor is not cosmetic: without the $x_i$, the slope equation collapses to
$\sum (y_i - b_0 - b_1 x_i) = 0$ — which is *identical* to the intercept equation
from Step 1. Two identical equations can't pin down two unknowns, so you'd never get
a unique slope. The $-x_i$ is exactly what makes the slope equation carry
information about $x$. (Practice turns this into a spot-the-error.)

**Trap 2 — "derivative = 0 means minimum."** Setting the slope to zero finds *level*
ground, which by itself could be a valley bottom, a hilltop, or a flat saddle. We're
entitled to call it the minimum only because SSE is a **bowl** — squaring the misses
curves the surface up on every side (Trap-2's payoff from module 1), so its one level
spot must be a bottom. On a curve that bends the other way, "derivative zero" would
hand you the *worst* point, not the best.

## Check yourself

**1.** Read this aloud, then answer: what is $\frac{\partial}{\partial b_0}\,(y_3 - b_0 - b_1 x_3)^2$
for the toy data (where $x_3 = 3$, $y_3 = 5$), evaluated at $b_0 = 0.5, b_1 = 1.4$?

::: {.callout-note collapse="true" title="Answer"}
"The partial derivative, with respect to bee-zero, of: why-three minus bee-zero
minus bee-one times ex-three, squared."

Chain rule lite: $2\,(y_3 - b_0 - b_1 x_3)\cdot(-1)$. Plug in:
$y_3 - b_0 - b_1 x_3 = 5 - 0.5 - 4.2 = 0.3$, so the answer is
$2 \times 0.3 \times (-1) = -0.6$. (Each data point contributes a term like this;
at the optimum they all sum to zero, even though individual terms — like this
−0.6 — don't have to be zero themselves.)
:::

**2.** Using only Step 1's boxed result (no computation): a fitted regression line
has slope 2, and the data have $\bar{x} = 10$, $\bar{y} = 30$. What is the
intercept?

::: {.callout-note collapse="true" title="Answer"}
$\hat{\beta}_0 = \bar{y} - \hat{\beta}_1 \bar{x} = 30 - 2 \times 10 = 10$. The best
line must pass through the point of means $(10, 30)$; with slope 2, walking back
from $x = 10$ to $x = 0$ descends $2 \times 10 = 20$, landing at 10.
:::

**3.** Explain in two sentences, without formulas, why setting the derivative to zero
finds the best line.

::: {.callout-note collapse="true" title="Answer"}
The derivative tells you whether nudging the line's parameters would make the fit
better or worse: a positive derivative means "this direction makes it worse," a
negative one means "keep going." At the bottom of the bowl, nudging in any direction
makes the fit worse, so the derivative is exactly zero there, and that is the one
spot where no further improvement is possible.
:::

## What you can now do

- **Find the bottom of a one-variable valley** by taking the derivative, setting it
  to zero, and solving — e.g. minimize $f(b) = (b-3)^2 + 2$ in two lines.
- **Explain "derivative equals zero"** as "level ground at the bottom of the bowl,"
  and say why the bowl shape (from squaring) guarantees it's a minimum.
- **Walk the OLS derivation**, naming the trick behind each step (chain rule, the
  three sum rules, the centering trick), from $\mathrm{SSE}$ down to
  $\hat{\beta}_1 = s_{xy}/s_x^2$ and $\hat{\beta}_0 = \bar{y} - \hat{\beta}_1\bar{x}$.
- **Compute a fit by hand** with `cov(x,y)/var(x)` and `mean(y) - b1*mean(x)`, and
  use the point-of-means fact to get an intercept from a slope.

If any of these feels shaky, revisit [Getting a feel for it](#getting-a-feel-for-it)
and [The formal version](#the-formal-version) before continuing — the next module
reads the output of exactly this fit.

## Where this goes next

You now know what `lm()` computes and why. The last module decodes what it *prints* —
standard errors, t values, p-values, $R^2$ — and how the same numbers appear in
papers: [reading lm() output and papers](../03-reading-lm-output-and-papers/lesson.qmd).
Ready for more reps first? Do the [practice problems](practice.qmd).
