---
title: "Practice: Deriving the OLS formulas"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| message: false
# Packages this practice set uses. Missing ones are installed automatically.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Do them on paper or in an R session; only
then open the answer.

## Problem 1 — the toy example, new numbers

Find the bottom of $f(b) = (b - 5)^2 + 1$ two ways: (a) guess-and-check at
$b = 3, 4, 5, 6, 7$; (b) take the derivative, set it to zero, solve.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
f <- function(b) (b - 5)^2 + 1
```

::: {.callout-note collapse="true" title="Answer"}
(a) $f(3) = 5$, $f(4) = 2$, $f(5) = 1$, $f(6) = 2$, $f(7) = 5$ — bottom at $b = 5$.

(b) Chain rule lite: $f'(b) = 2(b - 5) \cdot 1$ (the $+1$ is constant, derivative
0). Set to zero: $2(b - 5) = 0 \Rightarrow b = 5$. Same answer, no guessing.

```{r p1-answer}
f <- function(b) (b - 5)^2 + 1
optimize(f, c(0, 10))$minimum   # R's numeric optimizer agrees: 5
```
:::

## Problem 2 — one derivation line, on your own

In Step 1 we went from $\sum (y_i - b_0 - b_1 x_i) = 0$ to
$\sum y_i - n\,b_0 - b_1 \sum x_i = 0$. Justify that move yourself, naming the sum
rule used for each of the three pieces.

::: {.callout-note collapse="true" title="Answer"}
Split the sum into three sums (Rule 1):
$\sum y_i - \sum b_0 - \sum b_1 x_i = 0$. Then:

- $\sum y_i$ — stays as is.
- $\sum b_0$: summing the constant $b_0$, n times, gives $n\,b_0$ (Rule 3).
- $\sum b_1 x_i$: pull the constant multiplier $b_1$ out, giving $b_1 \sum x_i$
  (Rule 2).

Those three sum rules from the foundations module carry the whole derivation —
that's why they were taught first.
:::

## Problem 3 — spot the error

To find the slope, a colleague differentiates SSE with respect to $b_1$ and writes:

> "$\dfrac{\partial\, \mathrm{SSE}}{\partial b_1} = \sum 2\,(y_i - b_0 - b_1 x_i)$.
> Set it to zero: $\sum (y_i - b_0 - b_1 x_i) = 0$."

Their derivative contains **exactly one mistake**. Name it, and explain why the
mistake would make the slope impossible to solve for.

::: {.callout-note collapse="true" title="Answer"}
They dropped the chain rule's *inside-derivative* factor. The "stuff"
$y_i - b_0 - b_1 x_i$ has derivative $-x_i$ with respect to $b_1$ (not 1), so the
correct line is
$$\frac{\partial\, \mathrm{SSE}}{\partial b_1} = \sum 2\,(y_i - b_0 - b_1 x_i)\,(-x_i).$$

Why it matters: without the $-x_i$, setting the derivative to zero gives
$\sum (y_i - b_0 - b_1 x_i) = 0$ — which is **exactly** the intercept equation from
Step 1. Two identical equations can't pin down two unknowns ($b_0$ and $b_1$), so the
slope would be unsolvable. The $x_i$ factor is precisely what makes the slope
equation carry information about $x$, and it's what turns the final answer into
$s_{xy}/s_x^2$.
:::

## Problem 4 — compute a fit fully by hand *(uses Module 1)*

Three plants: $x = (1, 2, 3)$, $y = (3, 4, 6)$ — the same data you scored a *guessed*
line on in module 01's practice (Problem 1). There you rated one line; now use the
formulas to find the **best** one: compute $\bar{x}$, $\bar{y}$, $s_{xy}$, $s_x^2$,
then $\hat{\beta}_1$ and $\hat{\beta}_0$. Check with `lm()`, and note how much lower
the best line's SSE is than the guessed line's (which scored 1 back in module 01).

```{webr}
#| autorun: false
# p4-starter — edit and Run to solve, right here
x <- c(1, 2, 3); y <- c(3, 4, 6)
```

::: {.callout-note collapse="true" title="Answer"}
$\bar{x} = 2$, $\bar{y} = 13/3$. Deviations of x: $(-1, 0, 1)$; of y:
$(3 - \tfrac{13}{3}, 4 - \tfrac{13}{3}, 6 - \tfrac{13}{3}) = (-\tfrac{4}{3}, -\tfrac{1}{3}, \tfrac{5}{3})$.

Products: $\tfrac{4}{3}, 0, \tfrac{5}{3}$; sum = 3; so $s_{xy} = 3/2$.
Squared x-deviations: $1, 0, 1$; sum = 2; so $s_x^2 = 1$.

$\hat{\beta}_1 = \frac{3/2}{1} = 1.5$ and
$\hat{\beta}_0 = \frac{13}{3} - 1.5 \times 2 = \frac{13}{3} - 3 = \frac{4}{3}$.

So the best line is intercept $4/3$, slope $1.5$ — and its SSE is $1/6 \approx 0.167$,
far below the guessed line's score of 1 in module 01. The formulas found the exact
valley bottom that guessing could only approach.

```{r p4-answer}
x <- c(1, 2, 3); y <- c(3, 4, 6)
c(b1 = cov(x, y) / var(x), b0 = mean(y) - (cov(x, y) / var(x)) * mean(x))
coef(lm(y ~ x))                       # intercept 1.333, slope 1.5 — matches
sum(resid(lm(y ~ x))^2)               # SSE = 0.1667 = 1/6, the exact minimum
```
:::

## Problem 5 — break the formula (in R)

The slope formula divides by $s_x^2$. Predict what happens when all x values are
identical (say every plant got exactly 2 scoops) — then confirm with the starter,
and explain in plain English why no slope *could* be estimated.

```{webr}
#| autorun: false
# p5-starter — edit and Run to solve, right here
x <- c(2, 2, 2, 2)
y <- c(2, 3, 5, 6)
```

::: {.callout-note collapse="true" title="Answer"}
$s_x^2 = 0$ (no spread — every deviation is zero), so the formula is $s_{xy}/0$:
division by zero, undefined.

```{r p5-answer}
x <- c(2, 2, 2, 2)
y <- c(2, 3, 5, 6)
cov(x, y) / var(x)    # NaN — 0/0, the formula collapses
coef(lm(y ~ x))       # lm returns NA for the slope: it refuses too
```

Plain English: a slope answers "how much does y change *per unit change in x*?" —
but x never changes in this data, so the question is unanswerable. Every vertical
line of candidate slopes through the point of means scores the same SSE. When a
paper says a predictor was dropped for "no variation" or warns about
"(multi)collinearity", this is the failure mode it's talking about.
:::

## Problem 6 — a gentle cliffhanger

Fit `lm(mpg ~ wt, data = mtcars)` and run `summary()` on it. You can now explain
the two numbers in the `Estimate` column. Look at the neighboring column,
`Std. Error` — from its name and the Greek/Latin letter code, take a guess at what
question it answers. (The next module makes it precise.)

::: {.callout-note collapse="true" title="Answer"}
```{r p6-answer}
summary(lm(mpg ~ wt, data = mtcars))$coefficients
```

`Estimate` is $\hat{\beta}_0 = 37.29$ and $\hat{\beta}_1 = -5.34$ — covariance over
variance, and the anchor through the means, exactly as derived.

`Std. Error` answers: *if we could rerun the study on a fresh sample of 32 cars,
how much would the estimate typically bounce around?* A hat marks a quantity
computed from one sample ($\hat{\beta}$ estimates the true, hatless $\beta$), so a
different sample gives different values.
The standard error is the estimated standard deviation of that bouncing. Making
this precise (and turning it into t values and p-values) is the next module.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="courses/simple-linear-regression/modules/02-deriving-the-ols-formulas" pass="4"}

::: {.quiz-q qid="ols-slope" qtype="numeric" answer="1.4" tolerance="0"}
For the tomato data ($x = (1,2,3,4)$, $y = (2,3,5,6)$), the derivation gives
$s_{xy} = 7/3$ and $s_x^2 = 5/3$. Compute $\hat{\beta}_1 = s_{xy}/s_x^2$.
Enter one decimal.
:::

::: {.quiz-q qid="ols-intercept" qtype="numeric" answer="0.5" tolerance="0"}
With $\hat{\beta}_1 = 1.4$, $\bar{y} = 4$, and $\bar{x} = 2.5$, compute
$\hat{\beta}_0 = \bar{y} - \hat{\beta}_1\bar{x}$. Enter one decimal.
:::

::: {.quiz-q qid="ols-through-means" qtype="mc" answer="c"}
What does the Step-1 result $\hat{\beta}_0 = \bar{y} - \hat{\beta}_1\bar{x}$ imply
geometrically?

a) It tells you the slope must be positive for any real dataset with an upward trend
b) It proves the best line has the smallest possible slope among all candidate lines
c) The fitted line always passes through the point of means $(\bar{x},\bar{y})$
:::

::: {.quiz-q qid="ols-chain-rule" qtype="mc" answer="b"}
When differentiating SSE with respect to $b_1$, which extra factor does the chain
rule add compared with the $b_0$ step?

a) $-b_0$, the intercept term that also enters through the chain rule for $b_0$
b) $-x_i$, the predictor that multiplies $b_1$ inside each residual
c) $-2$, the constant pulled out by the power rule
:::

::: {.quiz-q qid="ols-sse-from-fit" qtype="numeric" answer="0.20" tolerance="0"}
(From Module 1.) Using $\hat{y}_i = 0.5 + 1.4\,x_i$ on $x = (1,2,3,4)$,
$y = (2,3,5,6)$: the residuals are $(0.1,\,-0.3,\,0.3,\,-0.1)$. What is the SSE?
Round to two decimals.
:::

:::

::::
