---
title: "Practice: The summary of an association"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

Work these in order — they get harder. Do them on paper or in an R session; only then
open the answer.

## Problem 1 — the toy example, new numbers

An association reports $\hat\beta = 0.8$ and $\text{se} = 0.2$. Compute the variance $V$,
the z-statistic $Z$, and the two-sided p-value.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
beta_hat <- 0.8
se       <- 0.2
```

::: {.callout-note collapse="true" title="Answer"}
$V = 0.2^2 = 0.04$. $Z = 0.8/0.2 = 4$. $p = 2\,\texttt{pnorm}(-4) \approx 6.3 \times
10^{-5}$ — an even stronger signal than the lesson's $Z = 3$.

```{r p1-answer}
beta_hat <- 0.8; se <- 0.2
c(V = se^2, Z = beta_hat / se, p = 2 * pnorm(-abs(beta_hat / se)))   # 0.04, 4, 6.3e-05
```
:::

## Problem 2 — fill in the missing step

An association has $\hat\beta = 0.5$ and $\text{se} = 0.25$. Here is the worked solution
with the core step blanked — supply only that step.

- Step 1 (shown): the variance is $V = 0.25^2 = 0.0625$.
- Step 2 (**your step here**): compute the z-statistic $Z$ from $\hat\beta$ and
  $\text{se}$. Write the exact arithmetic.
- Step 3 (shown): the two-sided p-value is $2\,\texttt{pnorm}(-Z) \approx 0.0455$.

```{webr}
#| autorun: false
# p2-starter — edit and Run to check the blank
beta_hat <- 0.5
se       <- 0.25
```

::: {.callout-note collapse="true" title="Answer"}
The missing move is the **z-statistic**: $Z = \hat\beta/\text{se} = 0.5/0.25 = 2$. That's
"the estimate is 2 standard errors from zero," which the p-value then reads off the normal
curve as $2\,\texttt{pnorm}(-2) \approx 0.0455$ — right at the edge of significance.

```{r p2-answer}
beta_hat <- 0.5; se <- 0.25
Z <- beta_hat / se               # 2
c(Z = Z, p = 2 * pnorm(-abs(Z)))  # 2, 0.0455
```
:::

## Problem 3 — a protective allele

A variant reports $\hat\beta = -0.30$ (a log odds ratio) with $\text{se} = 0.10$. (a)
Compute $Z$ and the two-sided p-value. (b) Convert $\hat\beta$ to an odds ratio and say
whether the allele raises or lowers the odds of the trait.

```{webr}
#| autorun: false
# p3-starter — your job: compute Z, the p-value, and the odds ratio
beta_hat <- -0.30
se       <- 0.10
```

::: {.callout-note collapse="true" title="Answer"}
(a) $Z = -0.30/0.10 = -3$; the p-value uses $|Z|$, so $p = 2\,\texttt{pnorm}(-3) \approx
0.0027$ — significant. (b) Odds ratio $= e^{-0.30} \approx 0.741$, which is **below 1**, so
the allele *lowers* the odds of the trait (a protective effect): each copy multiplies the
odds by about $0.74$.

```{r p3-answer}
beta_hat <- -0.30; se <- 0.10
Z <- beta_hat / se
c(Z = Z, p = 2 * pnorm(-abs(Z)), odds_ratio = exp(beta_hat))    # -3, 0.0027, 0.741
```
:::

## Problem 4 — spot the error

A colleague summarizes a variant — it contains **exactly one mistake** — find it and fix
it.

> "The fitted effect is $\hat\beta = 0.6$, so the odds ratio is $0.6$ — the allele barely
> changes the odds. Its standard error is $0.2$, giving $Z = 0.6/0.2 = 3$ and $p \approx
> 0.0027$, so it *is* statistically significant."

::: {.callout-note collapse="true" title="Answer"}
The mistake is **reading the log odds ratio as the odds ratio**. $\hat\beta = 0.6$ is on
the log-odds scale; the odds ratio is $e^{0.6} \approx 1.82$, not $0.6$. So the allele
raises the odds by about $82\%$ — the opposite of "barely changes the odds" (an odds ratio
near $0.6$ would actually mean a *drop*). The $Z$, p-value, and significance conclusion are
all correct; only the odds-ratio reading is wrong.

```{r p4-answer}
beta_hat <- 0.6
c(wrong_OR = beta_hat, correct_OR = exp(beta_hat))    # 0.6 vs 1.822
```
:::

## Problem 5 — mixing it with logistic regression *(uses logistic regression's odds ratio)*

A paper reports a variant's association as an **odds ratio of $1.5$** with a $95\%$
confidence interval on the odds-ratio scale of $[1.13, 1.99]$. To feed this into a
summary-statistic method you need $(\hat\beta, \text{se})$ on the log-odds scale. (a) What
is $\hat\beta$? (b) Recover the standard error from the CI, using that a $95\%$ CI for
$\hat\beta$ is $\hat\beta \pm 1.96\,\text{se}$ (so the log-OR CI is $[\log 1.13, \log
1.99]$).

```{webr}
#| autorun: false
# p5-starter — edit and Run to solve, right here
OR    <- 1.5
CI_OR <- c(1.13, 1.99)      # 95% CI on the odds-ratio scale
```

::: {.callout-note collapse="true" title="Answer"}
(a) $\hat\beta = \log(\text{OR}) = \log(1.5) \approx 0.405$ (converting an odds ratio back
to the log-odds scale — the inverse of $e^{\hat\beta}$ from
[logistic regression](../../../logistic-regression/modules/02-odds-log-odds-and-the-logit/lesson.qmd)).
(b) On the log scale the CI is $[\log 1.13, \log 1.99] \approx [0.122, 0.688]$; its width is
$0.688 - 0.122 = 0.566$, which equals $2 \times 1.96\,\text{se} = 3.92\,\text{se}$, so
$\text{se} = 0.566/3.92 \approx 0.144$.

```{r p5-answer}
OR <- 1.5; CI_OR <- c(1.13, 1.99)
beta_hat <- log(OR)                                # 0.405 -- log odds ratio
CI_log   <- log(CI_OR)                              # CI on the log-odds scale
se       <- (CI_log[2] - CI_log[1]) / (2 * 1.96)    # recover se from the CI width
c(beta_hat = round(beta_hat, 3), se = round(se, 3)) # 0.405, 0.144
```
This is the everyday move for reading a published GWAS/epidemiology result into a
summary-statistic pipeline: log the odds ratio, and back out the SE from its confidence
interval.
:::

## Problem 6 — transfer (interpretation)

A GWAS scans a million variants. Variant P has $Z = 5$; variant Q has $Z = 5$ as well, but
Q came from a study with **ten times** as many people. Do the two variants have the same
two-sided p-value? And — a cliffhanger the next modules will sharpen — should you be
*equally* impressed by both? Answer the p-value question fully; give your best guess on the
"equally impressed" question.

::: {.callout-note collapse="true" title="Answer"}
**The p-values are identical.** The two-sided p-value is a transform of $Z$ alone
($2\,\texttt{pnorm}(-5) \approx 5.7\times10^{-7}$), and both have $Z = 5$, so sample size
does not enter the p-value at all once $Z$ is fixed.

On "equally impressed": your instinct that something differs is right, and it's the
headline of this course. With ten times the data, Q's standard error is smaller, so the
*same* $Z = 5$ corresponds to a *smaller* estimated effect $\hat\beta$ (since $Z =
\hat\beta/\text{se}$). A tiny true effect detected only because $n$ is huge is arguably
weaker evidence of an *interesting* association than a solid effect in a smaller study —
yet the p-value can't see the difference. The Approximate Bayes Factor (Module 3) and its
posterior probability (Module 4) *can*, and that is exactly why Wakefield argued the
p-value threshold should shrink as $n$ grows. You can't fully resolve this yet — the tools
are three modules away.

```{r p6-answer}
c(p_at_Z5 = 2 * pnorm(-5))          # ~5.7e-07 -- same for both, sample size irrelevant here
```
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="courses/approximate-bayes-factors/modules/01-the-summary-of-an-association" pass="4"}

::: {.quiz-q qid="soa-variance-numeric" qtype="numeric" answer="0.0025" tolerance="0.0001"}
An association reports a standard error $\text{se} = 0.05$. What is the variance $V$ of the
estimate?
:::

::: {.quiz-q qid="soa-zstat-numeric" qtype="numeric" answer="3" tolerance="0"}
An association reports $\hat\beta = 0.36$ and $\text{se} = 0.12$. Compute the z-statistic
$Z$.
:::

::: {.quiz-q qid="soa-pvalue-transform" qtype="mc" answer="b"}
Why can a method use the pair $(\hat\beta, V)$ alone and lose nothing the p-value carried?

a) The p-value is always larger than $Z$, so $Z$ is the more informative number
b) The two-sided p-value is a deterministic transform of $Z = \hat\beta/\sqrt{V}$, so it adds no information beyond the pair
c) The p-value depends on the sample size, which the pair also records
:::

::: {.quiz-q qid="soa-logor-reading" qtype="numeric" answer="1.822" tolerance="0.005"}
A variant's effect estimate is $\hat\beta = 0.6$ on the log-odds scale. What odds ratio
does it correspond to? Round to three decimals.
:::

::: {.quiz-q qid="soa-evidence-strength" qtype="mc" answer="c"}
*(Interleaved — uses hypothesis tests & standard errors.)* Two variants have the same
effect estimate $\hat\beta = 0.4$. Variant A has $\text{se} = 0.1$; variant B has
$\text{se} = 0.4$. Which has the stronger evidence of association?

a) Variant B, because a larger standard error means a more thorough study
b) Neither — they have the same $\hat\beta$, so the evidence is identical
c) Variant A, because its smaller standard error gives a larger $Z$ ($4$ vs $1$)
:::

:::

::::
