---
title: "<Module title — plain words, e.g. 'The mean, and distance from it'>"
subtitle: "<one line: what this module unlocks>"
engine: knitr
---

<!-- INTERACTIVE LESSONS: if this lesson has any runnable in-browser `{webr}` cells,
     keep the include line below (it wires WebR into this page). If the lesson is
     entirely static `{r}` chunks, delete both the line below and `engine: knitr`
     above. Full rules: .claude/course-authoring/interactive-webr.md. Do NOT add a
     per-file `format:` — the project renders every page as live-html already. -->
{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

<!--
TEMPLATE INSTRUCTIONS (delete all HTML comments when filling in):
- This template implements root TEACHING.md's golden order. Keep every section
  below, in this exact order. When in doubt, TEACHING.md wins.
- Replace <placeholders>. Keep headings' wording-style: plain, non-scary.
- Every symbol used anywhere in this lesson MUST have a row in the pronunciation
  table (§ "How to read the symbols"). Copy rows from .claude/course-authoring/notation-style.md.
- The formal formula may appear ONLY in "The formal version" or later.
- INTERACTIVE vs STATIC (full contract: .claude/course-authoring/interactive-webr.md):
  a chunk the learner TWEAKS or SOLVES is a live `{webr}` cell; everything else —
  the hidden setup chunk, listings the learner only reads, worked answers, and any
  code needing a package WebAssembly can't run — stays a baked `{r}` chunk. Don't
  make a chunk live by reflex.
- R chunks: echo everything, comment the MATH each line computes; hidden setup uses
  set.seed(42).
- Anything beyond the course's applied need: mark it **Optional** with one line on
  what the extra depth buys.
-->

```{r setup}
#| include: false
# Build-time setup only — a live `{webr}` cell does NOT see anything defined here,
# so every teaching cell below re-states its own data. Packages this lesson uses are
# installed automatically (into renv; run renv::snapshot() afterwards if you add one).
needed <- c()   # e.g. c("ggplot2") — leave empty if base R suffices
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

## Where we're going

<!-- 2–3 sentences, no formulas: what this concept IS and why the learner needs it
     (for course modules: connect to the course goal; for foundation modules: name
     the kinds of methods that use it). -->

<What we'll learn, in plain words, and why it matters.>

**Builds on:** <links to prerequisite lessons, or "just the baseline">

## Warm-up: do you remember?

<!-- 2–3 RETRIEVAL questions on the prerequisites THIS lesson stands on. Rules:
     - Retrieval, not re-reading: ask the learner to PRODUCE something — a value
       from small numbers, a pronunciation, a one-line explanation. Never "did you
       read X?".
     - Reach back to ≥2 different earlier points when they exist (an earlier module
       AND a foundation module); for entry modules, quiz the baseline (arithmetic,
       reading a scatter plot, the idea of an average).
     - Each answer: collapsed callout, brief worked answer, plus a link back to the
       lesson/section that taught it ("shaky? revisit <link>").
     - If a question hands the learner data to compute with, give a copy-able
       live starter cell, same rule as practice problems (no knitr label — a
       leading comment, `# warmupN-starter`). -->

Before the new stuff, a few quick reps — answers from memory first, then check.

**1.** <recall question from a prerequisite module>

::: {.callout-note collapse="true" title="Answer"}
<brief worked answer> — taught in [<module name>](<relative link>); revisit if shaky.
:::

**2.** <recall question from a different earlier point>

::: {.callout-note collapse="true" title="Answer"}
<brief worked answer + backlink>
:::

## A tiny example first

<!-- THE most important section. Rules:
     - n ≤ 6 data points, small friendly numbers (integers if possible).
     - Work the computation in PROSE, digit by digit ("add 2 + 4 + 6 to get 12,
       divide by 3 to get 4").
     - Then reproduce the exact same numbers in a commented R chunk.
     - NO formal notation yet. Words only. -->

<Set up the toy scenario in one or two sentences — concrete and friendly, e.g.
"You measured 3 plants: 2 cm, 4 cm, 6 cm.">

<Work the whole computation in prose.>

The same thing in R, so you can check every step:

```{r toy-example}
# The toy data: <what it is>
x <- c(2, 4, 6)

# <what this computes, in math words — e.g. "add up all the values">
total <- sum(x)
total

# <next step of the same hand computation>
```

## Getting a feel for it

<!-- Build the INTUITION before any notation. The learner reads R fluently but does
     not yet FEEL what this concept measures. Still NO formulas here. Three beats:
     (a) ANALOGY / MENTAL PICTURE — tie the concept to something they already have a
         feel for (a balance point, two dancers moving together, a typical bet).
         Blurry-to-sharp: a rough picture first, sharpened later by the formula.
         REUSE the repo's established anchors where one fits (grep earlier lessons)
         so anchors compound across courses; and say where the analogy breaks.
     (b) WHY THIS CONSTRUCTION — why square / why multiply / why divide by THIS, in
         plain words, tied to the toy numbers ("raw deviations cancel to zero, so we
         square to stop the cancelling"). This is the design reason, not "it follows".
     (c) WHAT THE NUMBER FEELS LIKE — is our toy result big or small? What single
         change would make it bigger or smaller? Point at the picture in "Explore it
         in R" if the concept has one (or sketch a number line here).
     ANTI-FILLER RULE: rewording the formula in prose is NOT intuition. This section
     is a defect unless it contains a genuine analogy AND a why-it's-built-this-way
     reason. If the concept is purely mechanical and truly has no intuition to build,
     say so in one line rather than padding. -->

<Analogy / mental picture, why it's built this way, and what the toy number feels like.>

## What the notation is going to say

<!-- Bridge section, still NO formula. In 1 short paragraph, narrate what the
     upcoming formula says, in English, referring to the toy numbers:
     "The formula will say: take each measurement, ... . In our toy example
     that's exactly the 'add 2+4+6, divide by 3' we just did." -->

<Narrate the upcoming formula in plain English.>

## How to read the symbols

<!-- One row per symbol appearing ANYWHERE in this lesson (including practice-
     relevant ones). Copy pronunciations from .claude/course-authoring/notation-style.md so
     they stay consistent across all modules. The last column anchors each
     symbol to the toy example. -->

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $x$ | "ex" | <meaning> | <value(s)> |
| $n$ | "en" | how many data points | 3 |

## The formal version

<!-- ONLY NOW show the formula. Rules:
     - Multi-term equations: assemble ONE TERM AT A TIME — why the term exists,
       what breaks without it — then show the finished equation.
     - Map every part back to the toy numbers ("the Σ part is the 2+4+6 step; the
       1/n part is the 'divide by 3' step"), and re-do the toy computation once
       more, narrated THROUGH the formula.
     - Give the reading-aloud line: symbol-by-symbol AND everyday words.
     - For derivations: one algebraic step per line (aligned $$ ... $$ block), a
       bullet after EVERY line naming the trick used ("expand the square", "a sum
       can be split into two sums", ...); each trick taught already — in a
       foundation module or earlier in this lesson. Then SANITY-CHECK the result
       numerically: plug the toy numbers in and show they match the earlier hand
       computation. -->

$$
<formula>
$$

Reading it aloud: "<the whole formula spoken in words, using the pronunciation
table>" — or in everyday words: "<plain-English sentence>".

Mapping it onto the toy example:

- <formula part> is the "<prose step>" we did above → <number>
- ...

## Explore it

<!-- 2–4 chunks the learner should RUN AND TWEAK, live in the page. These are the
     lesson's INTERACTIVE cells: write them as `{webr}` (see
     .claude/course-authoring/interactive-webr.md). Each chunk:
     - a one-line "TRY IT: change ..." instruction in a comment at the top,
     - a PREDICT prompt for anything with a checkable outcome: "PREDICT: write down
       what you expect <X> to be — then run." (2–3 predict-then-run moments per
       lesson; committing to a prediction is what makes the result teach),
     - computes the concept BOTH from the raw formula (by hand, sum()/length()
       etc.) AND with the built-in R function when one exists, showing they
       match — that's what makes notation trustworthy,
     - a plot whenever the concept has a picture (with fig-alt). A concept-carrying
       plot uses `#| autorun: true` so the picture is there on page load.
     If a cell needs a package WebAssembly can't run, it is NOT a `{webr}` cell —
     keep it a baked `{r}` chunk under the static-fallback callout (pattern in
     interactive-webr.md), clearly labelled "runs on your machine, not here". -->

```{webr}
#| autorun: true
#| fig-alt: "<describe the picture for a screen reader>"
# TRY IT: change the values below and re-run — watch the result move.
# PREDICT: write down what you expect <X> to be — then run and check.
x <- c(2, 4, 6)

# By hand, straight from the formula: <formula in words>
by_hand <- sum(x) / length(x)

# R's built-in shortcut for the same thing
built_in <- mean(x)

c(by_hand = by_hand, built_in = built_in)  # identical — the formula IS the function
```

## Common traps

<!-- 1–2 predictable misconceptions, refuted WITH NUMBERS (refutation beats
     omission: naming the trap inoculates; hiding it leaves it live). Pattern per
     trap: state the tempting wrong belief plainly → say why it's tempting → show
     it fail on this lesson's numbers (prose or a tiny chunk). Pick traps a real
     learner would hit (order-of-operations, n vs n−1, correlation≠slope, ...).
     One of these becomes the practice set's spot-the-error problem. -->

**Trap: "<the tempting wrong belief>".** <Why it's tempting, and the numeric
refutation using this lesson's toy numbers.>

## Check yourself

<!-- 2–3 quick questions, answerable in <1 minute each, straight from this lesson.
     The LAST one is always the own-words prompt: "Explain to a colleague, in two
     sentences, what <concept> measures and why it's built the way it is" — with a
     collapsed MODEL answer (a genuinely good two-sentence explanation, not a
     formula reading). Answers in collapsed callouts with the worked step(s), not
     just the result. If a question gives the learner a vector / data frame to
     compute with, add a small live ```{webr} starter cell (autorun: false) defining
     that data right after the question (same copy-able-data rule as the practice
     set; no knitr label — name it in a leading comment, `# checkN-starter`). Skip
     it for scalar-only plug-ins already stated in prose. -->

**1.** <question>

```{webr}
#| autorun: false
# check1-starter — edit and Run to solve, right here (omit if scalar-only)
x <- c(...)
```

::: {.callout-note collapse="true" title="Answer"}
<worked answer>
:::

**2.** <question>

::: {.callout-note collapse="true" title="Answer"}
<worked answer>
:::

**3.** In your own words: explain to a colleague, in two sentences, what <concept>
measures and why it's built the way it is.

::: {.callout-note collapse="true" title="A model answer"}
<two crisp sentences a non-mathematician could say — analogy welcome, no formula
recitation>
:::

## What you can now do

<!-- 3–5 CAN-DO bullets — abilities, not topics ("compute a covariance by hand from
     paired deviations", not "covariance"). Then the mastery rule, naming the exact
     place to revisit. -->

- <can-do 1>
- <can-do 2>
- <can-do 3>

If any of these feels shaky, revisit <section / module link> before continuing —
the next module assumes all of them.

## Where this goes next

<!-- 2–3 sentences: which module comes next and what THIS concept will be used
     for there. Link with a relative path. Then point to practice.qmd. If the
     course's COURSE-REQUEST gave applied-practice pointers for this module, end
     with the Applied-practice line (format in course-structure.md). -->

<Next step and link. Then: "Ready for more reps? Do the
[practice problems](practice.qmd).">

<!-- Optional, only when the course request provides it:
**Applied practice (<external repo>):** open `<path/to/notebook-or-doc>` — <one
line on what to do there>. -->
