---
title: "Resampling: the bootstrap & permutation tests"
subtitle: "Get a standard error or a null when no formula exists — by redrawing and reshuffling your own data"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live cells below re-state their own data. This whole
# module is ABOUT randomness, so set.seed() is what makes every histogram, CI, and
# p-value reproducible. Base R only, nothing to install.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```


## Where we're going

Sometimes you can look up a formula for the wobble of a statistic — for the sample
*mean*, the standard error is $s/\sqrt{n}$ (where $s$ (the sample standard deviation)
and $n$ (the sample size) are decoded in the table below), and you're done. But for a **median**, a
**correlation**, a **90th percentile**, or the gap between two groups, there is often
no clean formula at all. **Resampling** is the escape hatch: instead of deriving the
wobble on paper, you *manufacture* it from the one sample you already have — by
redrawing it (the **bootstrap**) or reshuffling it (a **permutation test**). It's the
same simulate-and-summarize engine you already met, turned loose on real data when the
population is a mystery.

This lesson stays with neutral, hand-checkable examples (a few measurements, two small
groups) and base R. Two building blocks are **borrowed, not re-taught**: what a
*standard error* is (the wobble of a statistic across redraws) and what a *p-value*
is (how surprising your result would be if nothing were going on). Here you only learn
the two resampling *recipes* that produce them, and the one loop pattern behind both.

**Builds on:**
[randomness, samples & sampling distributions](../randomness-samples-and-sampling-distributions/lesson.qmd)
— we reuse the **standard error** (the spread of a statistic across redraws) and the
`replicate()` + histogram engine, and do not re-derive them; and
[hypothesis tests, p-values & confidence intervals](../hypothesis-tests-p-values-and-confidence-intervals/lesson.qmd)
— we reuse what a **p-value** and a **confidence interval** *mean*, and only supply the
resampling recipe that produces them.

## Warm-up: do you remember?

Before the new stuff, three quick reps — produce each answer from memory, then check.

**1.** *(sampling distributions)* In one sentence, what is the **standard error** of a
statistic, and how did you *simulate* one — the `replicate()` recipe?

::: {.callout-note collapse="true" title="Answer"}
The **standard error** is the *wobble of the statistic itself* — how much a number you
compute from data (a mean, a slope) would bounce around if you redrew the sample many
times. You simulated it by drawing many fresh samples, recording the statistic each time,
and taking the `sd()` of that list:
`sd(replicate(2000, mean(rnorm(25, 100, 15))))`. This lesson keeps that exact engine but
feeds it *your own data* instead of a known population. Shaky? Revisit
[randomness, samples & sampling distributions](../randomness-samples-and-sampling-distributions/lesson.qmd).
:::

**2.** *(hypothesis tests)* In one sentence, what is a **p-value** — and what is the
"null" it's measured against?

::: {.callout-note collapse="true" title="Answer"}
A **p-value** is the probability of seeing a result *at least as extreme* as the one you
got, **if the null hypothesis were true** — the null being the boring "nothing is going
on" world (no effect, no difference). Small p-value → your result would be surprising in
that boring world. Here you'll *build* that boring world by hand, by shuffling labels.
Shaky? Revisit
[hypothesis tests, p-values & confidence intervals](../hypothesis-tests-p-values-and-confidence-intervals/lesson.qmd).
:::

**3.** In R, what is the difference between `sample(x, 4, replace = TRUE)` and
`sample(x, 4, replace = FALSE)`?

```{webr}
#| autorun: false
# warmup3-starter — edit and Run to see the difference
x <- c(3, 5, 8, 9)
```

::: {.callout-note collapse="true" title="Answer"}
`replace = TRUE` draws **with replacement** — each pick is put back before the next, so
the same value can appear more than once (and some can be missed). `replace = FALSE`
draws **without replacement** — each value can be used at most once, so drawing all 4
just *reshuffles* the four values into a new order. Those two switches are the whole
lesson: the bootstrap uses `replace = TRUE`; the permutation test uses `replace = FALSE`.
:::

## A tiny example first

You measured **five** things — call them a sample: **4, 7, 6, 9, 5**. Their average is
$(4+7+6+9+5)/5 = 31/5 = 6.2$. That 6.2 is your one estimate. But you know from the last
module that if you could redraw a *fresh* sample of five, you'd get a slightly different
average — the estimate has a wobble (a standard error). The problem: **you can't redraw.**
You don't have the population; you have these five numbers and nothing else.

**The bootstrap's one idea: pretend your sample *is* the population.** To fake a redraw,
reach into your own five numbers and pull out five again — *with replacement*, so some
repeat and some go missing. That faked sample is called a **bootstrap resample**. Compute
its average, and you have one faked redraw of the statistic. Do it thousands of times and
you've faked the whole sampling distribution — no formula, no real population needed.

Let's make one resample by hand-in-R (a fixed seed so you see these exact draws):

```{r toy-example}
mysample <- c(4, 7, 6, 9, 5)
mean(mysample)                 # 6.2 — our one estimate

set.seed(1)
# ONE bootstrap resample: draw 5 from our own 5, WITH replacement
b1 <- sample(mysample, size = 5, replace = TRUE)
b1                             # 4 9 4 7 5  -- the 4 repeats, the 6 went missing
mean(b1)                       # 5.8 -- a faked "redraw" of the average

set.seed(2)
b2 <- sample(mysample, size = 5, replace = TRUE)
b2                             # 5 4 5 4 9  -- different again
mean(b2)                       # 5.4
```

Two faked redraws gave **5.8** and **5.4** — both near 6.2, both a little off. Nothing
was measured twice; the *resample* changed, so the faked average changed. Repeat that a
few thousand times and the spread of those faked averages *is* the standard error — the
wobble you couldn't compute by formula.

## Getting a feel for it

**The bootstrap mental picture: photocopy your sample into a giant bag, then redraw.**
Imagine your five numbers, each photocopied a million times and dumped into a huge bag.
That bag is your best guess at the population — it has the same shape as your data because
it *is* your data, just repeated. Now draw five from the bag: because the bag is enormous,
drawing "with replacement" from your five numbers gives the same thing as drawing five
from that bag. Each draw is a plausible alternate sample you *might* have collected. The
histogram of the statistic across those draws is your faked sampling distribution. Where
the analogy strains: the real population might contain values your five numbers never
saw (a rare huge outlier), and the bag can never invent one — so the bootstrap can only
be as honest as the sample it's copying. That's the root of its small-$n$ limitation.

**Why redraw *with* replacement, and why that's the whole trick.** If you drew your five
numbers *without* replacement, you'd just get the same five in a new order — the average
would be 6.2 every single time, wobble zero, useless. Replacement is what lets some values
repeat and others drop out, which is exactly the kind of luck-of-the-draw variation you'd
have gotten from real redraws. The bootstrap steals the *shape* of sampling variation
from the only sample you have.

**The permutation mental picture: if the label doesn't matter, you can swap it.** Now a
different question — a *comparison*. Group A came out to 7 on average, group B to 4; the
gap is 3. Is a gap of 3 a real effect, or just which numbers happened to land in which
group? Here's the "no effect" world made concrete: if the group label truly did nothing,
then a value's label is arbitrary — you could **shuffle** all the labels onto the pooled
values and the gap should look about the same. So reshuffle the labels many times, compute
the gap each time, and you've built the "no effect" distribution *by hand*. If your real
gap of 3 sits way out in the tail of that shuffled pile, the label mattered. (This is
where we hand off to the [p-value module](../hypothesis-tests-p-values-and-confidence-intervals/lesson.qmd):
"how far out in the tail" *is* the p-value — we just build the pile here.)

**What the numbers will feel like.** Our bootstrap of the mean of five numbers gives a
standard error around 0.76 and a 95% range of roughly 4.8 to 7.8 — a wide net, because
five numbers don't pin down much. Our permutation gap of 3 turns out to be the *biggest*
gap any reshuffle can produce, so it will look genuinely extreme. Both feel intuitive once
you watch the histograms below.

## What the notation is going to say

There is almost no new notation here — resampling is a *procedure*, not a formula. The one
new symbol is a star: $\bar{x}^{*}$ ("x-bar-star") is the statistic computed on **one
resample** rather than on the original data, and the star is the universal flag for "this
came from a resample, not the real sample." The bootstrap standard error is then the same
sentence as last module — the standard deviation of the statistic across redraws — only
the redraws are faked: **take your resampled statistics $\bar{x}^{*}_1, \dots, \bar{x}^{*}_R$
and compute their `sd()`.** The 95% confidence interval is just the middle 95% of that same
pile: chop off the lowest 2.5% and the highest 2.5% with `quantile(., c(.025, .975))`.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $\bar{x}$ | "ex bar" | the statistic on the **original** sample (in R: `mean(x)`) | 6.2 |
| $\bar{x}^{*}$ | "ex bar star" | the same statistic on **one resample**; the star means "from a resample" | 5.8 for the first resample |
| $\bar{x}^{*}_1, \dots, \bar{x}^{*}_R$ | "ex bar star sub one … up to sub R" | the resampled statistic from resample 1, 2, …, $R$ — the pile we histogram | 5.8, 5.4, … |
| $\mathrm{SE}$ | "the standard error" | the wobble of the statistic — here the `sd()` of the resampled pile | ≈ 0.76 |
| $\mathrm{SE}_{\text{boot}}$ | "S-E sub boot" | the bootstrap estimate of the standard error — the `sd()` of the resampled statistics | ≈ 0.76 |
| $s$ | "ess" | the sample standard deviation — the spread of your data (in R: `sd(x)`) | ≈ 1.92 |
| $n$ | "en" | the sample size — how many numbers we resample each time (in R: `length(x)`) | 5 |
| $R$ | "big R" | how many resamples we take (in R: `replicate(R, ...)`) | 2000 |
| $d$ | "dee" | the observed statistic in a comparison — here the gap between two group means | 3 |
| $d^{*}$ | "dee star" | the gap computed after **shuffling the labels** — one draw from the "no effect" pile | between $-3$ and $3$ |
| $q_{2.5\%},\, q_{97.5\%}$ | "the 2.5th and 97.5th percentiles" | the cutoffs below which 2.5% / 97.5% of the pile sits — the CI endpoints (in R: `quantile(pile, c(.025, .975))`) | ≈ 4.8 and 7.8 |
| $\binom{6}{3}$ | "six choose three" | the number of ways to pick 3 of 6 things (here, which 3 values get label A) | 20 |

The star ($^{*}$) is the only convention to memorize: **starred = computed on resampled
data.** Everything else — $\bar{x}$, $\mathrm{SE}$, $n$, $R$ — you already read fluently
from the last module.

## The formal version

There is no equation to derive — resampling replaces the derivation. Instead, here are the
two recipes as procedures, each mapped onto the toy numbers and sanity-checked in R.

### Recipe 1 — the bootstrap (a standard error and a CI, from your data alone)

1. **Resample:** draw $n$ values from your sample *with replacement*.
2. **Recompute:** compute your statistic on that resample → one $\bar{x}^{*}$.
3. **Repeat** $R$ times (a few thousand) → the pile $\bar{x}^{*}_1, \dots, \bar{x}^{*}_R$.
4. **Summarize:** the **standard error** is the `sd()` of the pile; the **95% CI** is its
   middle 95%, `quantile(pile, c(.025, .975))`.

$$
\mathrm{SE}_{\text{boot}} \;=\; \mathrm{sd}\big(\bar{x}^{*}_1,\; \dots,\; \bar{x}^{*}_R\big)
\qquad
\text{95\% CI} \;=\; \big[\,q_{2.5\%},\; q_{97.5\%}\,\big]
$$

Reading it aloud: "the bootstrap standard error is the standard deviation of the starred
x-bars; the 95% interval runs from the 2.5th percentile to the 97.5th percentile of that
same pile" — everyday words: **"redraw your own data thousands of times, and read the
spread and the middle 95% straight off the resulting histogram."**

```{r boot-recipe}
mysample <- c(4, 7, 6, 9, 5)

set.seed(42)
# Steps 1-3: resample-with-replacement, recompute the mean, repeat 2000 times.
# replicate() is the "do this whole thing R times" tool from the last module.
boot_means <- replicate(2000, mean(sample(mysample, size = 5, replace = TRUE)))

# Step 4: read the answers off the pile
sd(boot_means)                          # ~0.758  -- the bootstrap STANDARD ERROR
quantile(boot_means, c(0.025, 0.975))   # ~4.8 to 7.8 -- the 95% percentile CI
```

**Sanity check against the one formula we do have.** For the mean, the last module's
shortcut $s/\sqrt{n}$ should land near the bootstrap SE:

```{r boot-sanity}
mysample <- c(4, 7, 6, 9, 5)
sd(mysample) / sqrt(length(mysample))   # ~0.86 -- the formula SE for the mean
# vs the bootstrap SE ~0.758 above: close. The bootstrap agrees with the formula
# WHERE a formula exists -- which is why we trust it where one does NOT.
```

### Recipe 2 — the permutation test (build the "no effect" null by shuffling)

1. **Observe:** compute the statistic on the real groups → the observed $d$.
2. **Pool & shuffle:** put all the values in one pot and randomly reassign the labels
   (`sample()` without replacement — a shuffle).
3. **Recompute:** compute the same statistic under the shuffle → one $d^{*}$.
4. **Repeat** $R$ times → the "no effect" pile $d^{*}_1, \dots, d^{*}_R$.
5. **Locate:** see where the observed $d$ falls in that pile. (Turning "how far out" into
   a p-value is the [hypothesis-tests module's](../hypothesis-tests-p-values-and-confidence-intervals/lesson.qmd)
   job; we only build the pile and point at $d$.)

```{r perm-recipe}
groupA <- c(6, 8, 7)   # e.g. a treated group
groupB <- c(3, 5, 4)   # e.g. a control group

obs_diff <- mean(groupA) - mean(groupB)   # step 1: observed gap
obs_diff                                   # 3

all_vals <- c(groupA, groupB)              # pool the 6 values
labels   <- rep(c("A", "B"), each = 3)     # their real labels: A A A B B B

set.seed(1)
one_shuffle <- sample(labels)              # step 2: shuffle the labels (no replacement)
one_shuffle                                # "A" "B" "A" "B" "A" "B"
mean(all_vals[one_shuffle == "A"]) - mean(all_vals[one_shuffle == "B"])   # step 3: 1
```

Under that one shuffle the gap was only **1** — smaller than the real **3**. Now do it
thousands of times to build the whole "no effect" pile:

```{r perm-null}
groupA <- c(6, 8, 7); groupB <- c(3, 5, 4)
obs_diff <- mean(groupA) - mean(groupB)
all_vals <- c(groupA, groupB)
labels   <- rep(c("A", "B"), each = 3)

set.seed(42)
# steps 2-4: shuffle labels, recompute the gap, 2000 times -> the "no effect" pile
perm_diffs <- replicate(2000, {
  s <- sample(labels)                                  # reshuffle the labels
  mean(all_vals[s == "A"]) - mean(all_vals[s == "B"])  # the gap under "no effect"
})

mean(perm_diffs)                       # ~0: the no-effect pile is centered on zero
range(perm_diffs)                      # -3 to 3: the biggest gap ANY shuffle can make
mean(abs(perm_diffs) >= abs(obs_diff)) # ~0.103: fraction of shuffles as extreme as 3
```

Mapping it onto the numbers: with only six values split 3-and-3, there are exactly
$\binom{6}{3} = 20$ ways to hand out the labels. The observed split (A gets the three
largest) is the *single* most extreme one, giving the biggest possible gap of 3. So the
observed $d = 3$ sits at the very edge of the shuffled pile — the label really did seem to
matter. (**Exactly** 2 of the 20 splits reach a gap of $\pm 3$, so the exact two-sided
tail is $2/20 = 0.10$ — the simulation's 0.103 nails it.)

::: {.callout-tip collapse="true" title="The universal loop pattern (concept 4) — memorize this shape"}
Every resampling method — bootstrap, permutation, and dozens you'll meet later — is the
**same three-line loop**:

```r
pile <- replicate(R, {          # 1. REPEAT R times:
  resampled <- <redraw or reshuffle the data>   # 2. make a fake dataset
  <compute a statistic on it>                    # 3. return one number
})
<summarize the pile>            # 4. sd(), quantile(), or "where does the observed fall?"
```

Only line 2 changes: **`sample(x, replace = TRUE)` for the bootstrap**, **`sample(labels)`
(a shuffle) for a permutation test.** Learn the shape once and every resampling method is a
fill-in-the-blank.
:::

## Explore it

```{webr}
#| autorun: true
#| fig-alt: "Histogram of 2000 bootstrap sample means in a bell around 6.2, with dashed lines marking the 2.5% and 97.5% percentiles near 4.8 and 7.8"
# explore-1 — LIVE the bootstrap: watch the faked sampling distribution and its CI.
# TRY IT: change mysample, or swap mean for median, and re-run.
# PREDICT: will the CI get WIDER or NARROWER if you make the sample bigger? Guess, then run.
set.seed(42)
mysample   <- c(4, 7, 6, 9, 5)
boot_means <- replicate(2000, mean(sample(mysample, length(mysample), replace = TRUE)))
ci <- quantile(boot_means, c(0.025, 0.975))

hist(boot_means, breaks = 30, col = "steelblue",
     main = sprintf("bootstrap SE = %.2f;  95%% CI = [%.1f, %.1f]",
                    sd(boot_means), ci[1], ci[2]),
     xlab = "resampled mean (x-bar-star) over 2000 resamples")
abline(v = ci, col = "red", lwd = 2, lty = 2)   # the 2.5% and 97.5% cutoffs
```

```{webr}
#| autorun: true
# explore-2 — LIVE: the bootstrap works for statistics with NO clean SE formula.
# The MEDIAN has no tidy s/sqrt(n). The recipe does not care -- swap mean -> median.
# PREDICT: which pile is spread wider, the mean's or the median's? Write it, then run.
set.seed(42)
mysample <- c(4, 7, 6, 9, 5)
boot_mean   <- replicate(2000, mean(  sample(mysample, length(mysample), replace = TRUE)))
boot_median <- replicate(2000, median(sample(mysample, length(mysample), replace = TRUE)))

c(SE_of_mean   = sd(boot_mean),      # ~0.76 -- also gettable from a formula
  SE_of_median = sd(boot_median))    # ~1.14 -- NO clean formula; only resampling gives it
```

```{webr}
#| autorun: true
#| fig-alt: "Histogram of 2000 permutation gaps centered on zero, spanning -3 to 3, with a red line at the observed gap of 3 sitting at the far right edge"
# explore-3 — LIVE the permutation test: build the "no effect" pile, then locate d.
# TRY IT: edit groupA/groupB and re-run -- watch where the red line lands.
# PREDICT: if you make the two groups nearly identical, will the red line be in the
# MIDDLE of the pile or at the EDGE? Guess, then run.
set.seed(42)
groupA <- c(6, 8, 7); groupB <- c(3, 5, 4)
obs_diff <- mean(groupA) - mean(groupB)
all_vals <- c(groupA, groupB); labels <- rep(c("A", "B"), each = 3)

perm_diffs <- replicate(2000, {
  s <- sample(labels)
  mean(all_vals[s == "A"]) - mean(all_vals[s == "B"])
})

hist(perm_diffs, breaks = 20, col = "seagreen",
     main = "the 'no effect' pile (shuffled labels)",
     xlab = "gap between group means under a random shuffle")
abline(v = obs_diff, col = "red", lwd = 3)   # the observed gap: far out at the edge
```

The observed red line sits at the extreme right edge of the shuffled pile: a gap that big
almost never happens when the labels are meaningless, so the label looks like it matters.
**How extreme "extreme enough" is** — turning the picture into a p-value and a decision —
is the [hypothesis-tests module's](../hypothesis-tests-p-values-and-confidence-intervals/lesson.qmd)
job; resampling's job is just to build the pile.

```{webr}
#| autorun: true
# explore-4 — LIVE sanity check: bootstrap SE ≈ formula SE for the mean.
# For the mean we DO have a formula: s/sqrt(n). Run this and confirm the two agree.
# PREDICT: will they match exactly, or just approximately? Why?
set.seed(42)
mysample <- c(4, 7, 6, 9, 5)
boot_means <- replicate(2000, mean(sample(mysample, length(mysample), replace = TRUE)))

boot_se    <- sd(boot_means)                          # bootstrap SE
formula_se <- sd(mysample) / sqrt(length(mysample))   # the s/sqrt(n) shortcut

c(bootstrap_SE = boot_se,    # ~0.76
  formula_SE   = formula_se) # ~0.86
# They're close — the bootstrap agrees with the formula where a formula exists,
# which is exactly why we trust it where one does NOT.
```

## When resampling beats a formula — and when it fails

**When it wins (concept 5, the payoff).** Reach for resampling whenever your statistic has
**no clean standard-error formula or closed form**: a median, a trimmed mean, a
correlation, a ratio, the difference of two medians, a 90th percentile, the output of a
messy multi-step pipeline. The bootstrap gives you a standard error and a CI for *any* of
them with the same three-line loop; the permutation test gives you a null for *any*
comparison by shuffling the labels. You trade a page of derivation (which may not even
exist) for a few thousand cheap redraws.

**When it fails.** Resampling is not magic — two limits to respect:

- **Small $n$.** The bootstrap can only reshuffle the values you *have*; it can never
  invent a value your sample missed. With very few data points the faked distribution is
  coarse and can badly understate the true wobble. Watch how few *distinct* answers a
  bootstrap of just three numbers can even produce:

  ```{r limit-small-n}
  tiny <- c(4, 7, 6)                # only 3 data points
  set.seed(42)
  bt <- replicate(2000, mean(sample(tiny, 3, replace = TRUE)))
  length(unique(bt))               # only 9 distinct means possible -- a coarse, chunky pile
  ```

- **Dependence.** Both recipes assume the data points are *interchangeable* — that
  reshuffling or redrawing them is fair. If the observations are **linked** (measurements
  over time, repeated readings on the same subject, spatial neighbors), a naive shuffle
  breaks a real structure and the answer is wrong. Plain resampling wants roughly
  independent, exchangeable data; dependent data needs specialized variants (block
  bootstrap, restricted permutations) beyond this module.

## Common traps

**Trap 1 — "the bootstrap draws *without* replacement, like an ordinary sample."**
Tempting, because that's what a normal `sample()` call defaults to. But without replacement
you'd draw all $n$ of your values every time — just in a new order — so every resampled
mean would be **identical** and the standard error would come out **exactly zero**. Watch
the bug on our five numbers:

```{r trap1-numbers}
mysample <- c(4, 7, 6, 9, 5)
set.seed(42)
buggy <- replicate(2000, mean(sample(mysample, 5, replace = FALSE)))  # WRONG: no replacement
sd(buggy)          # 0 -- every resample is just a reordering, so every mean is 6.2
# Correct version uses replace = TRUE, giving sd ~0.758 as we saw.
```

Zero wobble is the tell-tale sign you forgot `replace = TRUE`. **Bootstrap = with
replacement, always.**

**Trap 2 — "a permutation test just shuffles the numbers around within each group."**
People reach for resampling and shuffle the wrong thing. Reordering the values *inside*
each group changes nothing — a group's mean doesn't care what order its numbers sit in —
so every gap stays exactly the observed 3 and you learn nothing about the null. Watch it
fail:

```{r trap2-numbers}
A <- c(6, 8, 7); B <- c(3, 5, 4)      # the two groups
set.seed(42)
within <- replicate(2000, mean(sample(A)) - mean(sample(B)))  # WRONG: shuffle within groups
sd(within)                            # 0 -- every gap is still 3; nothing was learned
```

The permutation test instead keeps the six values fixed and **reassigns which group each
belongs to** — shuffling the A/B *labels* across the pooled values — because the hypothesis
being tested is precisely "the *label* doesn't matter." Shuffle the labels, not the numbers
within a group.

## Check yourself

**1.** You have the sample `x <- c(10, 12, 15, 11, 13)`. Write the R to bootstrap the
**standard error of the mean** with 2000 resamples, and to get a 95% percentile CI.

```{webr}
#| autorun: false
# check1-starter — edit and Run to solve, right here
set.seed(42)
x <- c(10, 12, 15, 11, 13)
# your job: replicate( ... mean(sample(x, ..., replace = TRUE)) ... ), then sd() and quantile()
```

::: {.callout-note collapse="true" title="Answer"}
The three-line loop, with `replace = TRUE` and the pooled summaries:

```{r check1-answer}
set.seed(42)
x <- c(10, 12, 15, 11, 13)
boot <- replicate(2000, mean(sample(x, length(x), replace = TRUE)))
sd(boot)                           # ~0.75 -- the bootstrap SE of the mean
quantile(boot, c(0.025, 0.975))    # ~10.8 to 13.8 -- the 95% percentile CI
```

The whole method is "resample with replacement, recompute the mean, repeat, summarize."
:::

**2.** In a permutation test, you shuffle the labels **without** replacement (a plain
`sample(labels)`). But the bootstrap draws **with** replacement. In one sentence each, why
does each recipe need the setting it uses?

::: {.callout-note collapse="true" title="Answer"}
The **permutation** test asks "does the label matter?", so it keeps the exact same set of
values and only *reassigns* them — a shuffle, which is sampling the labels **without**
replacement (every value gets exactly one label). The **bootstrap** asks "how much would
my estimate wobble across fresh samples?", so it must let values repeat and drop out to
mimic real sampling variation — which requires drawing **with** replacement (without it,
every resample is identical and the wobble is a fake zero).
:::

**3.** In your own words: explain to a colleague, in two sentences, what the bootstrap is
and why you'd use it instead of a formula.

::: {.callout-note collapse="true" title="A model answer"}
The bootstrap treats your one sample as a stand-in for the whole population and fakes many
fresh samples by redrawing from it *with replacement*, so the spread of the statistic
across those fake redraws estimates its standard error and confidence interval. You reach
for it when your statistic — a median, a correlation, some messy pipeline — has no clean
formula for its wobble, because the same three-line resampling loop works for any statistic
at all.
:::

## What you can now do

- **Sample with vs without replacement** on purpose — knowing `replace = TRUE` lets values
  repeat and `replace = FALSE` merely reshuffles, and knowing which recipe needs which.
- **Bootstrap a standard error and a percentile CI** for *any* statistic: resample with
  replacement, recompute, repeat with `replicate()`, then `sd()` and
  `quantile(., c(.025, .975))`.
- **Run a permutation test by hand** — pool the values, shuffle the labels to build the
  "no effect" pile, and locate where the observed statistic falls in it.
- **Recognize the one universal loop** (`replicate` → resample → compute a statistic →
  summarize) behind every resampling method, and swap only the "resample" line.
- **Judge when resampling is the right tool** — when no clean formula exists — and when to
  distrust it (very small $n$, dependent/non-exchangeable data).

If any of these feels shaky, revisit [The formal version](#the-formal-version) and
[Common traps](#common-traps) before continuing — anything that quantifies uncertainty
without a formula assumes all of these.

## Where this goes next

Resampling is the quiet workhorse under a huge range of modern methods: any time a paper
reports a standard error or confidence interval for something with no textbook formula, or
a p-value from "we permuted the labels 10,000 times," this is the machinery. The bootstrap
reuses the last module's simulate-and-take-`sd()` engine when you can't name the population;
the permutation test reuses the p-value idea when you can't write down a null by formula.
Ready for more reps? Do the [practice problems](practice.qmd).
