---
title: "Practice: Resampling — the bootstrap & permutation tests"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only (a {webr} starter does not see it — each starter re-states its
# own data). This module is about randomness, so every drawing chunk sets its own seed.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Do them on paper or in an R session; only
then open the answer.

## Problem 1 — the toy example, new numbers

A fresh sample of five: `x <- c(10, 12, 15, 11, 13)`. Bootstrap the **standard error of
the mean** with 2000 resamples, and report a 95% percentile confidence interval. Use
`set.seed(42)` so your numbers match.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
set.seed(42)
x <- c(10, 12, 15, 11, 13)
# your job: bootstrap the mean, then sd() and quantile(., c(.025, .975))
```

::: {.callout-note collapse="true" title="Answer"}
The three-line loop: resample **with replacement**, recompute the mean, repeat; then read
the spread and the middle 95% off the pile.

```{r p1-answer}
set.seed(42)
x <- c(10, 12, 15, 11, 13)
boot <- replicate(2000, mean(sample(x, length(x), replace = TRUE)))
sd(boot)                           # ~0.75 -- the bootstrap SE of the mean
quantile(boot, c(0.025, 0.975))    # ~10.8 to 13.8 -- the 95% percentile CI
```

The sample mean is $(10+12+15+11+13)/5 = 12.2$, and the CI sits symmetrically around it —
the resampling just measures how far the mean could wobble.
:::

## Problem 2 — fill in the missing step

Here is a bootstrap of the mean of `s <- c(4, 8, 6, 10)`, worked out — but the **core
move is blanked**. Fill in ONLY the blanked line.

> The sample mean is $(4+8+6+10)/4 = 7$. To fake many redraws:
> **`<your step here>`** — build a pile of 2000 resampled means, each from a
> resample-with-replacement of `s`.
> Then `sd(pile)` is the bootstrap SE (~1.1) and `quantile(pile, c(.025, .975))` is the CI.

```{webr}
#| autorun: false
# p2-starter — edit and Run to check the blanked line
set.seed(42)
s <- c(4, 8, 6, 10)
# pile <- replicate( ... )   # <- fill this in
```

::: {.callout-note collapse="true" title="Answer"}
The blanked line is the whole bootstrap engine — `replicate()` wrapped around a
resample-with-replacement and a `mean()`:

```{r p2-answer}
set.seed(42)
s <- c(4, 8, 6, 10)
pile <- replicate(2000, mean(sample(s, length(s), replace = TRUE)))   # the missing step
sd(pile)                           # ~1.08 -- bootstrap SE
quantile(pile, c(0.025, 0.975))    # 5 to 9 -- 95% CI
```

The key detail is `replace = TRUE` inside `sample()` — that is what lets values repeat
and produces a non-zero wobble.
:::

## Problem 3 — bootstrap a statistic with no formula

You have six measurements: `d <- c(2, 5, 3, 8, 4, 6)`. The **median** has no clean
standard-error formula. Bootstrap its standard error and a 95% CI (2000 resamples,
`set.seed(42)`), and say in one line why you *had* to resample here.

```{webr}
#| autorun: false
# p3-starter — your job: bootstrap the MEDIAN
set.seed(42)
d <- c(2, 5, 3, 8, 4, 6)
```

::: {.callout-note collapse="true" title="Answer"}
Same loop as the mean — only the statistic changes from `mean` to `median`, which is
exactly the point: the recipe doesn't care which statistic you feed it.

```{r p3-answer}
set.seed(42)
d <- c(2, 5, 3, 8, 4, 6)
boot_med <- replicate(2000, median(sample(d, length(d), replace = TRUE)))
sd(boot_med)                           # ~1.08 -- SE of the median
quantile(boot_med, c(0.025, 0.975))    # 2.5 to 7 -- 95% CI for the median
```

The observed median is 4.5. You *had* to resample because there is no tidy $s/\sqrt{n}$
for the median — the bootstrap is the only easy way to get its wobble.
:::

## Problem 4 — spot the error

A colleague wants a bootstrap standard error for the mean of `y <- c(4, 7, 6, 9, 5)` and
writes the following. It contains **exactly one mistake**. Find it, explain why it's
wrong, and fix it.

> "I'll redraw my sample 2000 times and take the sd of the means:
> `sd(replicate(2000, mean(sample(y, 5, replace = FALSE))))`.
> That gives a standard error of 0 — great, my estimate is perfectly precise!"

```{webr}
#| autorun: false
# p4-starter — run the buggy line, then the fix, and compare
set.seed(42)
y <- c(4, 7, 6, 9, 5)
```

::: {.callout-note collapse="true" title="Answer"}
The mistake is **`replace = FALSE`**. Drawing all 5 values without replacement just
*reorders* them, so every resample has the same five numbers and therefore the identical
mean 6.2 — the sd of a constant is 0. That zero is a red flag, not a triumph: it means the
bootstrap never varied. Fix it with `replace = TRUE`, which lets values repeat and produces
real sampling variation.

```{r p4-answer}
set.seed(42)
y <- c(4, 7, 6, 9, 5)
sd(replicate(2000, mean(sample(y, 5, replace = FALSE))))   # 0 -- the bug: no variation
sd(replicate(2000, mean(sample(y, 5, replace = TRUE))))    # ~0.77 -- the correct SE
```

Bootstrap = **with replacement, always.**
:::

## Problem 5 — mixing it with hypothesis tests *(uses [hypothesis-tests-p-values-and-confidence-intervals](../hypothesis-tests-p-values-and-confidence-intervals/lesson.qmd)'s p-value)*

Two small groups: `gA <- c(9, 11, 10)` and `gB <- c(6, 5, 7)`. Build the "no effect" pile
by shuffling the labels (3000 shuffles, `set.seed(42)`), then compute a two-sided p-value
as the fraction of shuffles whose gap is at least as extreme as the observed gap. The
p-value *definition* is the hypothesis-tests module's; your job is to feed it the shuffled
pile.

**PREDICT** first: the observed gap is $10 - 6 = 4$. With three-and-three values, do you
expect the observed gap to land in the middle of the shuffled pile or at its edge?

```{webr}
#| autorun: false
# p5-starter — edit and Run to solve, right here
set.seed(42)
gA <- c(9, 11, 10); gB <- c(6, 5, 7)
```

::: {.callout-note collapse="true" title="Answer"}
Pool the six values, shuffle the labels (no replacement), recompute the gap, repeat; then
the p-value is just "what fraction of the shuffled pile is as extreme as the real gap" —
the definition from the [hypothesis-tests module](../hypothesis-tests-p-values-and-confidence-intervals/lesson.qmd).

```{r p5-answer}
set.seed(42)
gA <- c(9, 11, 10); gB <- c(6, 5, 7)
obs_diff <- mean(gA) - mean(gB)          # 4
all_vals <- c(gA, gB); labels <- rep(c("A", "B"), each = 3)

perm_diffs <- replicate(3000, {
  s <- sample(labels)                                  # shuffle the labels
  mean(all_vals[s == "A"]) - mean(all_vals[s == "B"])
})
mean(abs(perm_diffs) >= abs(obs_diff))   # ~0.10 -- the two-sided permutation p-value
```

The observed gap of 4 is the *biggest* gap the six values can produce, so it sits at the
edge of the pile. Exactly 2 of the $\binom{6}{3} = 20$ possible splits reach $\pm 4$, so
the exact two-sided p is $2/20 = 0.10$ — the simulation lands right on it. The resampling
built the null; the p-value idea came from the hypothesis-tests module.
:::

## Problem 6 — transfer: a confidence interval for a difference of medians

New context. You have two groups and want a CI for the **difference of their medians** —
a quantity with no textbook standard-error formula:

```{webr}
#| autorun: false
# p6-starter — bootstrap the DIFFERENCE OF MEDIANS and give a 95% CI
set.seed(42)
gX <- c(3, 7, 5, 9, 6)
gY <- c(2, 4, 3, 5, 6)
# hint: resample EACH group with replacement, take median(bx) - median(by), repeat
```

::: {.callout-note collapse="true" title="Answer"}
The bootstrap generalizes with no new ideas: resample **each** group with replacement,
compute the difference of medians on the resamples, repeat, and read the CI off the pile.

```{r p6-answer}
set.seed(42)
gX <- c(3, 7, 5, 9, 6)
gY <- c(2, 4, 3, 5, 6)
median(gX) - median(gY)                  # observed: 6 - 4 = 2

boot_diff <- replicate(2000, {
  bx <- sample(gX, length(gX), replace = TRUE)
  by <- sample(gY, length(gY), replace = TRUE)
  median(bx) - median(by)
})
sd(boot_diff)                            # ~1.6 -- SE of the difference of medians
quantile(boot_diff, c(0.025, 0.975))     # ~ -1 to 5 -- 95% CI
```

The CI includes 0, so with only five values per group the difference of medians isn't
pinned down away from zero. Notice you never needed a formula — the same resample-recompute-
summarize loop handled a two-group difference of medians exactly as it handled one mean.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/resampling-bootstrap-and-permutation" pass="4"}

::: {.quiz-q qid="rbp-with-replacement" qtype="mc" answer="b"}
A bootstrap resample of $n$ values is drawn:

a) Without replacement — a reshuffle of the same $n$ values
b) With replacement — values can repeat, so resamples genuinely differ
c) By dropping one value each time
:::

::: {.quiz-q qid="rbp-zero-se" qtype="mc" answer="a"}
You bootstrap a standard error and get exactly $0$. What went wrong?

a) You forgot `replace = TRUE`, so every resample is just a reordering with the same mean
b) The statistic has no variance by nature
c) You used too many replications
:::

::: {.quiz-q qid="rbp-shuffle-labels" qtype="mc" answer="c"}
A permutation test for a group difference shuffles what?

a) The values within each group
b) The order of the rows, leaving the groups intact
c) The group *labels* across the pooled values — reassigning which group each value belongs to
:::

::: {.quiz-q qid="rbp-when-right" qtype="mc" answer="a"}
When is resampling the right tool?

a) When no clean formula exists for the statistic's uncertainty — though it struggles with very small $n$ or dependent data
b) Always — it replaces every formula
c) Only when the data are normally distributed
:::

::: {.quiz-q qid="rbp-bootstrap-se" qtype="mc" answer="b"}
(From randomness & sampling distributions.) A bootstrap estimates a standard error as the
spread of the resampled statistics. Which R function gives that spread?

a) `mean()`
b) `sd()` — the standard deviation of the resampled statistics
c) `sum()`
:::

:::

::::
