---
title: "Marginal vs joint: the vanishing coefficient"
subtitle: "Why a predictor that matters alone can vanish — or flip sign — in the full model"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)

sim_indicators <- function(n = 200, seed = 1, rho_max = 0.95) {
  set.seed(seed)
  nm <- c("rate_10y", "rate_2y", "cpi_infl", "pce_infl", "unemp",
          "oil", "usd_index", "sentiment", "housing", "pmi",
          "credit_spread", "vix")
  p <- length(nm)
  R <- diag(p); dimnames(R) <- list(nm, nm)
  R["rate_10y", "rate_2y"]  <- R["rate_2y", "rate_10y"]  <- 0.90
  R["cpi_infl", "pce_infl"] <- R["pce_infl", "cpi_infl"] <- rho_max
  R["sentiment", "pmi"]     <- R["pmi", "sentiment"]     <- 0.70
  R["credit_spread", "vix"] <- R["vix", "credit_spread"] <- 0.65
  X <- matrix(rnorm(n * p), n, p) %*% chol(R)
  colnames(X) <- nm
  beta <- setNames(numeric(p), nm)
  beta[c("rate_10y", "cpi_infl", "unemp")] <- c(0.5, -0.4, -0.3)
  y <- drop(X %*% beta + rnorm(n, sd = 1))
  list(X = X, y = y, beta_true = beta)
}
```

## Where we're going

This is the module the course was built for. You can now fit a multiple regression and
read its coefficients "holding the others fixed." But what happens when the predictors
*refuse to be held fixed independently* — when they travel together, like two inflation
measures that rise and fall as a pair? Then a predictor can look strongly related to the
outcome **on its own** and yet contribute almost nothing — or the *opposite* of what you
expect — **in the joint fit**. That is not a bug. It is exactly what "holding the others
fixed" costs when there is barely any independent variation left to hold fixed. By the
end you'll see it happen, measure it, and know not to trust a lone coefficient.

**Builds on:** [Module 2](../02-least-squares-and-the-normal-equations/lesson.qmd) (the
joint fit via the normal equations),
[Module 1](../01-many-knobs-one-model/lesson.qmd) ("holding the others fixed"),
[variance-structure-and-standardization](../../../../foundations/variance-structure-and-standardization/lesson.qmd)
(correlation between predictors — "two knobs wired to one wheel").

## Warm-up: do you remember?

**1.** *(Module 1)* State the meaning of the coefficient $\beta_2$ in a two-predictor
model, including the qualifier.

::: {.callout-note collapse="true" title="Answer"}
$\beta_2$ is the change in predicted $y$ for a one-unit increase in $x_2$ **holding
$x_1$ fixed**. The "holding $x_1$ fixed" part is what today's whole story turns on. See
[Module 1](../01-many-knobs-one-model/lesson.qmd).
:::

**2.** *(variance structure)* Two predictors have correlation 0.95. In one plain
sentence, what does that say about how they move?

::: {.callout-note collapse="true" title="Answer"}
They move almost in lockstep — when one is above its average, the other almost always is
too. There is very little variation in one that isn't shared with the other. See
[variance-structure-and-standardization](../../../../foundations/variance-structure-and-standardization/lesson.qmd).
:::

**3.** *(Module 2)* A "marginal" (one-at-a-time) regression fits $y$ on a single
predictor. How would you get the marginal slope of $y$ on $x_2$ in R?

::: {.callout-note collapse="true" title="Answer"}
`coef(lm(y ~ x2))[2]` — just $y$ against $x_2$ alone, ignoring every other predictor.
The **joint** slope, by contrast, comes from `lm(y ~ x1 + x2)`. Today is about how
different those two can be.
:::

## A tiny example first

Four observations, two predictors, and — the crucial ingredient — the two predictors
are strongly correlated. To keep arithmetic clean, everything is centred (means are 0),
so we can fit through the origin:

$$
x_1 = (-2, -1, 1, 2), \qquad x_2 = (-1, -1, 1, 1), \qquad y = (-2, -1, 1, 2).
$$

Two things to notice before any fitting. First, **$y$ is a perfect copy of $x_1$** —
$y$ is driven *entirely* by $x_1$, and $x_2$ has nothing to do with generating it.
Second, **$x_1$ and $x_2$ are strongly correlated** ($r \approx 0.95$): whenever $x_1$
is negative $x_2$ tends to be negative, and likewise positive.

**The marginal fit — $y$ on $x_2$ alone.** The one-at-a-time slope (through the origin)
is $\sum x_{i2} y_i / \sum x_{i2}^2$:

$$
\sum x_{i2} y_i = (-1)(-2) + (-1)(-1) + (1)(1) + (1)(2) = 2 + 1 + 1 + 2 = 6, \qquad
\sum x_{i2}^2 = 1 + 1 + 1 + 1 = 4,
$$

so the marginal slope of $x_2$ is $6/4 = 1.5$. Judged **alone**, $x_2$ looks like a
strong, positive predictor of $y$.

**The joint fit — $y$ on $x_1$ and $x_2$ together.** Use Module 2's normal equations.
The dot-product grid and right-hand side are

$$
X^\top X = \begin{pmatrix} 10 & 6 \\ 6 & 4 \end{pmatrix}, \qquad X^\top y = \begin{pmatrix} 10 \\ 6 \end{pmatrix},
$$

with determinant $10\cdot4 - 6\cdot6 = 40 - 36 = 4$. Solving,

$$
b_1 = \frac{4\cdot10 - 6\cdot6}{4} = \frac{4}{4} = 1, \qquad
b_2 = \frac{10\cdot6 - 6\cdot10}{4} = \frac{0}{4} = 0 .
$$

**The joint coefficient on $x_2$ is exactly zero.** The very same $x_2$ that looked like
a strong predictor on its own contributes *nothing* once $x_1$ is in the room. The same
thing in R:

```{r toy-example}
x1 <- c(-2, -1, 1, 2); x2 <- c(-1, -1, 1, 1); y <- c(-2, -1, 1, 2)

marginal_x2 <- coef(lm(y ~ x2 - 1))[1]         # x2 alone, through origin: slope 1.5
joint       <- coef(lm(y ~ x1 + x2 - 1))       # x1 and x2 together: (1, 0)
c(marginal_x2 = marginal_x2, joint_x1 = joint[1], joint_x2 = joint[2])
```

## Getting a feel for it

**Two knobs wired to one wheel.** Imagine a machine whose output really depends on one
hidden wheel. You don't get to touch the wheel; you get two knobs, $x_1$ and $x_2$, and
both are connected to that same wheel by nearly-rigid linkages — turn one and the other
almost has to turn with it. If someone asks "what does knob $x_2$ do *while $x_1$ is held
perfectly still*?", the honest answer is "almost nothing I can measure" — because holding
$x_1$ still barely lets $x_2$ move at all, and the tiny wiggle of $x_2$ that *is*
independent had nothing to do with the output. That is the joint coefficient. The
marginal fit, by contrast, never held $x_1$ still — so when it wiggled $x_2$, the wheel
(and $x_1$) came along for the ride, and $x_2$ got *credit for the wheel's work*.

**Why the credit gets reassigned.** In the toy, $x_2$'s marginal slope of 1.5 is
borrowed glory: $x_2$ correlates with $x_1$, $x_1$ drives $y$, so $x_2$ *appears* to
predict $y$. The joint fit refuses to double-count: it asks what $x_2$ adds **beyond**
$x_1$, finds nothing, and sets its coefficient to 0. With real, noisy data the joint fit
can go further and hand back a *negative* coefficient to a truly-positive predictor — a
**sign flip** — when noise makes one correlated twin look slightly better than the
other. You'll watch exactly that happen on the economic panel below.

**What it feels like as correlation grows.** The more correlated the two predictors, the
less independent variation is left, and the more violently the joint fit swings the
coefficients around trying to split a shared effect. At correlation 0 the marginal and
joint coefficients agree; as correlation climbs toward 1 they diverge and the joint
estimates get wild and unstable. The **variance inflation factor** puts a number on this
(below): for our toy it is exactly 10.

## What the notation is going to say

We'll write down two things. First, a name for "how badly is predictor $j$ tangled up
with the others": regress $x_j$ on all the *other* predictors, read off that fit's
$R^2_j$, and the **variance inflation factor** is $1/(1 - R^2_j)$ — it says how many
times wider $x_j$'s coefficient uncertainty is, compared to if $x_j$ were uncorrelated
with the rest. Second, nothing new for the fit itself: "marginal" is Module 2's fit with
one predictor, "joint" is Module 2's fit with all of them.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $r$ | "little r" | correlation between two predictors, in $[-1, 1]$ (in R: `cor`) | $r(x_1, x_2) \approx 0.95$ |
| $\beta_j^{\text{marg}}$ | "beta jay marginal" | slope of $y$ on $x_j$ **alone** (one-at-a-time fit) | $\beta_2^{\text{marg}} = 1.5$ |
| $\beta_j^{\text{joint}}$ | "beta jay joint" | coefficient on $x_j$ in the **full** model, others held fixed | $\beta_2^{\text{joint}} = 0$ |
| $R^2_j$ | "R squared sub jay" | fraction of $x_j$ explained by the *other* predictors | $0.90$ (regress $x_2$ on $x_1$) |
| $\mathrm{VIF}_j$ | "vif sub jay" | variance inflation factor $= 1/(1 - R^2_j)$ | $1/(1-0.9) = 10$ |
| $\hat\beta$ | "beta hat" | the estimated joint coefficient vector | $(1, 0)$ |

**Carried from [Module 2](../02-least-squares-and-the-normal-equations/lesson.qmd)** (re-used in the formulas below):

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $x_1$, $x_2$ | "ex one", "ex two" | the predictor columns | $(-2,-1,1,2)$, $(-1,-1,1,1)$ |
| $y$ | "why" | the response column | $(-2,-1,1,2)$ |
| $b_1$, $b_2$ | "bee one", "bee two" | the fitted coefficients we solve for | $1$, $0$ |
| $X^\top X$ | "ex transpose ex" | grid of dot products among predictor columns | $\left(\begin{smallmatrix}10&6\\6&4\end{smallmatrix}\right)$ |
| $X^\top y$ | "ex transpose why" | each column's dot product with $y$ | $(10, 6)$ |
| $\sum$ | "the sum" | add one term per observation | $\sum x_{i2} y_i = 6$ |

## The formal version

There is no new fitting formula here — "marginal" and "joint" are both just Module 2's
least-squares fit, run on one predictor or on all of them. The new object is the
**variance inflation factor**, the standard one-number measure of how collinear a
predictor is with the rest:

$$
\mathrm{VIF}_j = \frac{1}{1 - R^2_j},
$$

where $R^2_j$ is the $R^2$ from regressing predictor $x_j$ on **all the other
predictors**. Reading it aloud: "vif-sub-jay equals one over, one minus R-squared-sub-jay"
— or in everyday words: "if the other predictors already explain most of $x_j$ (high
$R^2_j$), then $x_j$ adds little independent information, and its coefficient becomes
that many times harder to pin down."

Mapping onto the toy: regressing $x_2$ on $x_1$ gives $R^2_2 = 0.90$ (since
$r = 0.9487$ and $0.9487^2 = 0.90$), so $\mathrm{VIF}_2 = 1/(1 - 0.90) = 10$. A common
rule of thumb flags $\mathrm{VIF} > 10$ as serious collinearity — our toy sits right at
the line, which is why $x_2$'s joint coefficient was so easily driven to zero.

```{webr}
#| autorun: true
# TRY IT: change x2 to be LESS correlated with x1 (e.g. x2 <- c(-1, 1, -1, 1)) and re-run.
# PREDICT: if x2 moves independently of x1, will R^2_2 be near 0 or near 1? What VIF?
x1 <- c(-2, -1, 1, 2); x2 <- c(-1, -1, 1, 1)
R2_2 <- summary(lm(x2 ~ x1 - 1))$r.squared
c(R2_2 = round(R2_2, 2), VIF_2 = round(1 / (1 - R2_2), 1))
```

```{r vif-toy}
# VIF for the toy, from the definition: regress x2 on x1, read R^2, invert (1 - R^2)
x1 <- c(-2, -1, 1, 2); x2 <- c(-1, -1, 1, 1)
R2_2 <- summary(lm(x2 ~ x1))$r.squared
c(R2_2 = R2_2, VIF_2 = 1 / (1 - R2_2))          # 0.90 and 10
```

## Explore it

### The economic panel: marginal vs joint, side by side

Here is the real thing on the course's simulated panel ($n = 200$). We built it so the
**true** driver in each correlated pair is only *one* of the twins: `rate_10y` (not
`rate_2y`) and `cpi_infl` (not `pce_infl`). Watch what the marginal and joint fits say:

```{r panel-marg-vs-joint}
sim   <- sim_indicators(n = 200, seed = 1)
panel <- data.frame(sim$X, gdp = sim$y)

joint <- coef(lm(gdp ~ ., data = panel))[-1]                       # all together
marg  <- sapply(colnames(sim$X),                                    # one at a time
                function(v) coef(lm(sim$y ~ sim$X[, v]))[2])

cmp <- data.frame(true = sim$beta_true, marginal = round(marg, 3), joint = round(joint, 3))
cmp[c("rate_10y", "rate_2y", "cpi_infl", "pce_infl"), ]
```

Read the two twin pairs:

- **Yield twins** (`rate_10y`, `rate_2y`, correlated 0.90). The true effect lives
  entirely on `rate_10y` (0.5); `rate_2y`'s true effect is 0. Yet **marginally the two
  look identical** (both ≈ 0.49) — because `rate_2y` rides along with `rate_10y`. The
  joint fit splits the single real effect between them (≈ 0.27 and ≈ 0.30), crediting
  neither with the full 0.5.
- **Inflation twins** (`cpi_infl`, `pce_infl`, correlated 0.95). The true driver is
  `cpi_infl` (−0.4); `pce_infl`'s true effect is 0. Both look moderately negative
  marginally (≈ −0.29 and ≈ −0.31). But in the joint fit the near-collinear pair is so
  hard to separate that the model hands the big negative coefficient to the **wrong**
  twin, `pce_infl` (≈ −0.43), and **flips `cpi_infl` positive** (≈ +0.11). A predictor
  with a genuine negative effect shows up with the opposite sign — purely because its
  correlated partner absorbed the role.

The lesson: **a single coefficient in a collinear model is not a verdict on that
predictor.** The fit is still excellent at *prediction* (the twins together capture the
inflation signal); it just can't reliably attribute the effect to one twin.

### Turn the correlation knob yourself

```{webr}
#| autorun: true
#| fig-alt: "Two curves plotted against the predictor correlation rho from 0 to about 1: the marginal slope of x2 (blue) rises steadily from near 0 toward about 1, while its joint coefficient (red) stays near 0 but grows jagged and swings negative as rho approaches 1."
# TRY IT: change the rho grid (e.g. go all the way to 0.999) or n, and re-run.
# PREDICT: as rho -> 1, which curve heads toward 1, and which stays near 0 but gets jumpy?
# Throughout, y depends ONLY on x1 — x2's TRUE effect is 0.
set.seed(1); n <- 400
rhos <- seq(0, 0.99, by = 0.03)
res <- sapply(rhos, function(rho) {
  x1 <- rnorm(n)
  x2 <- rho * x1 + sqrt(1 - rho^2) * rnorm(n)   # correlated with x1 at level rho
  y  <- x1 + rnorm(n)                            # only x1 drives y
  c(marginal = unname(coef(lm(y ~ x2))[2]),      # x2 alone: borrows x1's glory
    joint    = unname(coef(lm(y ~ x1 + x2))[3])) # x2 with x1 held fixed: true role is 0
})
plot(rhos, res["marginal", ], type = "l", lwd = 2, col = "steelblue", ylim = range(res),
     xlab = "correlation between x1 and x2  (rho)", ylab = "x2's estimated slope",
     main = "x2 has NO true effect — only the joint fit knows")
lines(rhos, res["joint", ], lwd = 2, col = "firebrick")
abline(h = 0, lty = 3)
legend("topleft", c("marginal (x2 alone)", "joint (x1 held fixed)"),
       col = c("steelblue", "firebrick"), lwd = 2, bty = "n")
```

The blue marginal curve climbs steadily toward 1 as `rho` grows — `x2` looks ever more
important on its own. The red joint curve tells the truth (`x2`'s real effect is 0),
staying near 0 — but notice it grows jagged and even swings negative near `rho = 1`: with
almost no independent variation left in `x2`, the joint fit can barely tell the twins
apart, so its estimate gets unstable even while hovering around the correct 0.

## Common traps

**Trap: "a big (or 'significant') coefficient means the predictor matters."** In the
panel, `pce_infl` gets the *largest* inflation coefficient (≈ −0.43) in the joint fit —
yet its true effect is exactly 0. It won the coefficient only by being 0.95-correlated
with the real driver. Size alone, in a collinear model, tells you nothing reliable about
which predictor is truly responsible. (This is also why a coefficient's p-value from a
model you *searched* for is not an honest p-value — a thread the next course picks up.)

**Trap: "the marginal and joint coefficients should roughly agree; if they don't,
something is broken."** They agree only when predictors are uncorrelated. When
predictors are correlated they *should* differ — that difference is the whole point of a
multiple regression. In the toy, marginal $x_2 = 1.5$ and joint $x_2 = 0$ are both
correct answers to *different questions*: "how does $x_2$ track $y$ ignoring everything
else?" versus "what does $x_2$ add once $x_1$ is accounted for?"

## Check yourself

**1.** A predictor has $R^2_j = 0.75$ when regressed on the others. What is its VIF, and
is that cause for concern by the rule of thumb?

```{webr}
#| autorun: false
# check1-starter (scalar plug-in — or just do it by hand)
# PREDICT: R^2_j = 0.75 means 75% of x_j is explained by the others. Will VIF be above or below 10?
R2j <- 0.75
```

::: {.callout-note collapse="true" title="Answer"}
$\mathrm{VIF} = 1/(1 - 0.75) = 1/0.25 = 4$. Below the usual "> 10" flag, so moderate
collinearity — its coefficient's uncertainty is 4× inflated, worth noting but not alarming.
:::

**2.** In a fitted model, $x_A$ has a large positive marginal slope but a near-zero joint
coefficient. Give a one-sentence explanation a colleague would accept.

::: {.callout-note collapse="true" title="Answer"}
$x_A$ is probably correlated with another predictor that carries the real effect; on its
own $x_A$ borrows that predictor's association with $y$ (large marginal slope), but once
the real driver is held fixed, $x_A$ adds nothing (near-zero joint coefficient).
:::

**3.** In your own words: explain to a colleague, in two sentences, why a predictor's
coefficient can flip sign between the marginal and joint fits, and what that does *not*
imply about the model's predictions.

::: {.callout-note collapse="true" title="A model answer"}
When two predictors are strongly correlated, the joint fit struggles to split their
shared effect, so noise can push one twin's coefficient negative while the other goes
positive — even reversing the sign of a truly positive predictor. It does *not* mean the
model predicts badly: the twins together still capture the signal; the instability is
only in attributing the effect to one predictor rather than the other.
:::

## What you can now do

- **Distinguish a marginal fit from a joint fit** and explain why correlated predictors
  make them disagree — using "holding the others fixed" as the reason.
- **Diagnose the vanishing/sign-flipping coefficient**: a predictor correlated with the
  true driver borrows glory marginally and can lose it (or reverse) jointly.
- **Compute and interpret a VIF** as $1/(1 - R^2_j)$, and apply the "> 10" rule of thumb.
- **State the honest limit**: in a collinear model, an individual coefficient is not a
  reliable verdict on that predictor, even though prediction can stay excellent.

If any of these feels shaky, revisit [Getting a feel for it](#getting-a-feel-for-it)
and the panel comparison before continuing — the capstone reads all of this off a real
`summary(lm())`.

## Where this goes next

The capstone puts every piece together: [reading a multiple regression](../04-capstone-reading-a-multiple-regression/lesson.qmd)
decodes a full `summary(lm())` on a fresh panel line by line, and connects the
marginal-vs-joint story to how a statistics text frames it. Beyond this course,
a natural next question is *which* subset of these tangled predictors the model
should even keep — that's the world of variable selection. Ready for more reps first? Do the
[practice problems](practice.qmd).
