---
title: "Practice: Marginal vs joint"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Do them on paper or in an R session; only then
open the answer.

## Problem 1 — the vanishing coefficient, new numbers

Centred data (fit through the origin): $x_1 = (-3, -1, 1, 3)$, $x_2 = (-1, -1, 1, 1)$,
and $y = (-3, -1, 1, 3)$ — so $y$ is a perfect copy of $x_1$, and $x_2$ is correlated
with $x_1$. Compute (a) the **marginal** slope of $y$ on $x_2$ alone, and (b) the
**joint** coefficient on $x_2$ in the fit on both predictors. Are they the same?

```{webr}
#| autorun: false
# p1-starter — marginal (x2 alone) vs joint (both), through the origin
x1 <- c(-3, -1, 1, 3); x2 <- c(-1, -1, 1, 1); y <- c(-3, -1, 1, 3)
```

::: {.callout-note collapse="true" title="Answer"}
Marginal: $\sum x_2 y / \sum x_2^2 = (3+1+1+3)/4 = 8/4 = 2$. Joint: the normal equations
give $X^\top X = \left(\begin{smallmatrix}20&8\\8&4\end{smallmatrix}\right)$,
$X^\top y = (20, 8)$, determinant 16, so $b_2 = (20\cdot8 - 8\cdot20)/16 = 0$. Marginally
$x_2$ looks like a slope-2 predictor; jointly it contributes **nothing** — its apparent
effect was entirely borrowed from the correlated real driver $x_1$.

```{r}
x1 <- c(-3, -1, 1, 3); x2 <- c(-1, -1, 1, 1); y <- c(-3, -1, 1, 3); X <- cbind(x1, x2)
c(marginal_x2 = sum(x2 * y) / sum(x2^2),
  joint = solve(t(X) %*% X, t(X) %*% y)[, 1])      # marginal 2; joint (1, 0)
```
:::

## Problem 2 — fill in the missing step

A predictor is regressed on all the others and gets $R^2_j = 0.80$. Compute its variance
inflation factor.

- Step 1: the VIF formula is $\mathrm{VIF}_j = \dfrac{1}{1 - R^2_j}$.
- Step 2: **your step here** — plug in $R^2_j = 0.80$ and simplify.
- Step 3: interpret the number in one phrase.

::: {.callout-note collapse="true" title="Answer"}
Step 2: $\mathrm{VIF}_j = 1/(1 - 0.80) = 1/0.20 = 5$. Step 3: this predictor's coefficient
has 5× the sampling variance it would have if the predictor were uncorrelated with the
others — moderate collinearity, below the usual "> 10" alarm but worth noting.
:::

## Problem 3 — marginal vs joint, in R

Here is a small correlated dataset. Fit $y$ on $x_2$ **alone**, then on $x_1$ and $x_2$
**together**, and compare $x_2$'s slope in the two fits. Explain the direction of the
change.

```{webr}
#| autorun: false
# p3-starter — compare marginal and joint slopes of x2
# PREDICT: x1 and x2 are strongly correlated; will x2's joint slope be bigger or smaller?
x1 <- c(-2, -1, 0, 1, 2); x2 <- c(-1, -1, 0, 1, 1); y <- c(-1.5, -1, 0.5, 1, 1.5)
```

::: {.callout-note collapse="true" title="Answer"}
`x1` and `x2` are correlated ≈ 0.95. The marginal slope of `x2` (≈ 1.25) overstates its
role because `x2` tracks the real driver `x1`; in the joint fit, with `x1` held fixed,
`x2`'s coefficient shrinks (to ≈ 0.5). The joint fit strips out the part of `x2`'s
apparent effect that was really `x1`'s.

```{r}
x1 <- c(-2, -1, 0, 1, 2); x2 <- c(-1, -1, 0, 1, 1); y <- c(-1.5, -1, 0.5, 1, 1.5)
c(cor_x1x2   = round(cor(x1, x2), 3),
  marginal_x2 = round(coef(lm(y ~ x2))[2], 3),
  joint_x2    = round(coef(lm(y ~ x1 + x2))[3], 3))
```
:::

## Problem 4 — spot the error

Reading the economic panel's joint fit, a colleague says: *"`pce_infl` has the largest
(most negative) inflation coefficient in the joint model, so it must be the most
important inflation driver of GDP."* We simulated the data with `pce_infl`'s true effect
equal to **zero**. This reasoning contains **exactly one mistake** — name it.

::: {.callout-note collapse="true" title="Answer"}
The mistake is treating coefficient **size** as importance in a **collinear** model.
`pce_infl` is 0.95-correlated with `cpi_infl` (the real driver), so the joint fit can't
reliably attribute the shared inflation effect — it happened to load onto `pce_infl` (and
even flipped `cpi_infl` positive). A large coefficient on a highly collinear predictor is
not evidence it's the true driver; you'd check the VIFs (both ≈ 10 on this panel) and
recognize the attribution is unreliable.
:::

## Problem 5 — mixing it with Module 2 *(uses Module 2's normal equations)*

For a centred two-predictor problem you're told
$X^\top X = \left(\begin{smallmatrix}8&4\\4&4\end{smallmatrix}\right)$ and
$X^\top y = (8, 2)$. (a) Solve the normal equations for the joint $(b_1, b_2)$. (b) The
marginal slope of $y$ on $x_2$ alone is $(X^\top y)_2 / (X^\top X)_{22}$. Compute it and
compare its **sign** to the joint $b_2$.

```{webr}
#| autorun: false
# p5-starter — solve the joint system, then the marginal x2 slope
XtX <- matrix(c(8, 4, 4, 4), 2); Xty <- c(8, 2)
```

::: {.callout-note collapse="true" title="Answer"}
(a) Determinant $8\cdot4 - 4\cdot4 = 16$; $b_1 = (4\cdot8 - 4\cdot2)/16 = 24/16 = 1.5$,
$b_2 = (8\cdot2 - 4\cdot8)/16 = -16/16 = -1$. So joint $(b_1, b_2) = (1.5, -1)$. (b)
Marginal $x_2$ slope $= 2/4 = 0.5$ — **positive**, while the joint $b_2 = -1$ is
**negative**: a full sign flip. On its own $x_2$ trends up with $y$; once $x_1$ is
accounted for, $x_2$'s residual role is negative.

```{r}
XtX <- matrix(c(8, 4, 4, 4), 2); Xty <- c(8, 2)
c(joint = solve(XtX, Xty), marginal_x2 = Xty[2] / XtX[2, 2])   # (1.5, -1) vs 0.5
```
:::

## Problem 6 — interpretation

You fit a model and report these VIFs: `age` 1.2, `income` 8.5, `savings` 9.1,
`net_worth` 22.0. A reviewer asks which coefficient to trust least and what you might do.
Answer in two or three sentences (no computation needed).

::: {.callout-note collapse="true" title="Answer"}
Trust `net_worth`'s coefficient least: its VIF of 22 is well over the "> 10" flag, meaning
it's largely explained by the other predictors (probably `income` and `savings`), so its
individual coefficient is unstable and hard to interpret. Reasonable responses: drop or
combine the redundant predictors (e.g. keep `net_worth` and drop the pieces that make it
up, or vice versa), or note explicitly that the collinear group's coefficients shouldn't
be read individually. `age` (VIF 1.2) is essentially independent and its coefficient is
trustworthy.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="courses/multiple-regression/modules/03-marginal-vs-joint" pass="4"}

::: {.quiz-q qid="mvj-vif" qtype="numeric" answer="5" tolerance="0"}
A predictor regressed on all the others gives $R^2_j = 0.8$. Its VIF is $1/(1 - R^2_j)$.
Compute it.
:::

::: {.quiz-q qid="mvj-marginal-joint-disagree" qtype="mc" answer="b"}
A predictor's marginal and joint coefficients disagree sharply. Is something broken?

a) Yes — they should always roughly agree
b) No — they agree only when predictors are uncorrelated; disagreement under correlation is the whole point of a joint fit
c) Yes — the data must be re-collected
:::

::: {.quiz-q qid="mvj-borrowed-glory" qtype="mc" answer="a"}
In a collinear model, a predictor has the largest coefficient yet a true effect of zero.
How did that happen?

a) It is correlated with the real driver and borrowed its glory in the joint fit
b) Large coefficients always identify the true driver regardless of correlation
c) Its units are smaller than the others', inflating the coefficient artificially
:::

::: {.quiz-q qid="mvj-vif-rule" qtype="mc" answer="c"}
A VIF above the rule-of-thumb value of $10$ tells you:

a) The predictor's coefficient is definitely zero and can be dropped safely
b) The model as a whole must be discarded and rebuilt from scratch
c) That predictor is badly collinear, so its coefficient is unreliable
:::

::: {.quiz-q qid="mvj-collinear-singular" qtype="mc" answer="a"}
(From module 2.) The collinearity that inflates VIF is the same problem that makes what
fail in the normal equations?

a) The inversion of $X^\top X$ — a collinear column makes it singular
b) The computation of the residuals
c) The prediction step
:::

:::

::::
