---
title: "Maximum a posteriori estimation"
subtitle: "MLE plus a prior — and why ridge regression is secretly Bayesian"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```


## Where we're going

You already know two estimation ideas. **Maximum likelihood** (MLE) asks: "which
parameter makes my observed data most probable?" — it maximises the likelihood.
**Bayesian updating** says: "multiply the likelihood by a prior, and the result (the
posterior) is your updated belief about the parameter."
[@chan2021probabilitydatascience, pp. 503--504]

**MAP estimation** combines both: it finds the parameter value at the **peak** (mode) of
the posterior. Because the posterior is likelihood × prior, the MAP estimate is pulled
between the data (likelihood) and your prior belief. With enough data, the data win and
MAP converges to MLE. With a flat (uniform) prior, MAP IS MLE — the prior drops out.
[@chan2021probabilitydatascience, pp. 506--507]

The payoff: MAP explains *why* ridge regression and the lasso exist. Ridge is MAP with a
Gaussian prior; lasso is MAP with a Laplace prior. The shape of the prior controls
whether coefficients shrink smoothly toward zero (ridge) or snap exactly to zero (lasso).
[@chan2021probabilitydatascience, pp. 518--520]

**Builds on:**
[likelihood & log-likelihood](../likelihood-and-log-likelihood/lesson.qmd) —
the MLE idea and the product-to-sum log rule;
[priors, posteriors & updating](../priors-posteriors-and-updating/lesson.qmd) —
posterior ∝ likelihood × prior;
[ridge regularization](../ridge-regularization/lesson.qmd) —
the ridge and lasso penalties.


## Warm-up: do you remember?

Three quick reps — produce each answer from memory first, then check.

**1.** *(likelihood)* You flip a coin 5 times and see 3 heads, 2 tails. What is the
likelihood $L(p)$ of a particular value $p$? Which $p$ maximises it?

::: {.callout-note collapse="true" title="Answer"}
$L(p) = p^3 (1-p)^2$ — three heads contribute $p$ each, two tails contribute $(1-p)$
each, multiplied by independence. The MLE is $\hat{p} = 3/5 = 0.6$. Taught in
[likelihood & log-likelihood](../likelihood-and-log-likelihood/lesson.qmd); revisit if
shaky.
:::

**2.** *(priors)* Write Bayes' theorem in one line using "posterior ∝ likelihood × prior".
What does the ∝ mean?

::: {.callout-note collapse="true" title="Answer"}
$P(\theta \mid \text{data}) \propto P(\text{data} \mid \theta) \times P(\theta)$. The
∝ ("proportional to") means we drop the denominator $P(\text{data})$ because it doesn't
depend on $\theta$ — it's just a normalising constant. Taught in
[priors, posteriors & updating](../priors-posteriors-and-updating/lesson.qmd); revisit if
shaky.
:::

**3.** *(priors)* If your prior says $\theta = 0.5$ and your data strongly suggest
$\theta = 0.8$, does the posterior sit closer to 0.5 or to 0.8? What happens as you
collect more data?

::: {.callout-note collapse="true" title="Answer"}
With little data the posterior sits closer to the prior (0.5). With more data the
likelihood sharpens and the posterior shifts toward 0.8 (the data's answer). With
enough data the posterior concentrates on the MLE — the prior's influence fades.
Taught in
[priors, posteriors & updating](../priors-posteriors-and-updating/lesson.qmd);
revisit if shaky.
:::


## A tiny example first

Same coin: 5 flips, 3 heads. The MLE says $\hat{p} = 0.6$ — just the fraction of
heads. But suppose you believe, before seeing any data, that the coin is probably
close to fair. Encode that belief as a **Beta(3, 3)** prior — a hill centred at 0.5
that gives low weight to extreme values like 0.1 or 0.9.

**What Beta(3, 3) means in plain language:** think of it as "phantom observations." A
Beta($\alpha$, $\beta$) prior acts like you have already seen $\alpha - 1$ phantom heads
and $\beta - 1$ phantom tails before touching the real data. Beta(3, 3) means 2 phantom
heads and 2 phantom tails — 4 phantom flips that split evenly, pulling toward fair. The
MAP formula adds the phantom counts to the real counts.

The MAP estimate — the peak of the posterior — has a closed form for the Beta-Bernoulli
case. For Beta($\alpha$, $\beta$) prior with $\alpha + \text{heads} > 1$ and
$\beta + \text{tails} > 1$ (so the mode exists in the interior):
[@chan2021probabilitydatascience, pp. 514--515]

MAP = (heads + a − 1) / (n + a + b − 2)  = (3 + 3 − 1) / (5 + 3 + 3 − 2) = 5/9 ≈ 0.556

Read it as: "(real heads + phantom heads) / (real flips + phantom flips)" =
(3 + 2) / (5 + 4) = 5/9.

The MLE was 0.6 (pulled entirely by the data). The MAP is 0.556 — pulled partway
back toward the prior's mode at 0.5. The prior "tugged" the estimate away from the
data, toward fair.

Now try a **uniform** prior — Beta(1, 1), which says "every value of p is equally likely
before I see data":

MAP = (3 + 1 − 1) / (5 + 1 + 1 − 2) = 3/5 = 0.6 = MLE

With no prior opinion, MAP gives the same answer as MLE. The prior is what makes the
difference — and the shape of the prior determines *how* the estimate is pulled.

```{r toy-example}
# 5 flips, 3 heads — compute MAP under two priors
n <- 5; heads <- 3

mle <- heads / n                                         # 0.6
map_beta33 <- (heads + 3 - 1) / (n + 3 + 3 - 2)        # 5/9 ≈ 0.556
map_uniform <- (heads + 1 - 1) / (n + 1 + 1 - 2)       # 3/5 = 0.6

cat("MLE:", mle, "\n")
cat("MAP (Beta(3,3) prior):", round(map_beta33, 3), "\n")
cat("MAP (Uniform prior):", map_uniform, "— same as MLE\n")
```


## Getting a feel for it

**The mental picture: a tug-of-war.** The likelihood is a team of horses, one per data
point, pulling the estimate toward the data's answer. The prior is a bull anchored at
its favourite value. With few data points the bull wins — MAP stays near the prior. Add
more horses and they overpower the bull — MAP drifts toward the MLE. (Where the
analogy strains: the tug-of-war implies one equilibrium point between two pulls. That
works for the Beta-Bernoulli case and for ridge/lasso, but with a multimodal prior or a
non-concave posterior the MAP can jump between multiple peaks — more like a landscape
than a rope.)
[@chan2021probabilitydatascience, pp. 509--511]

In the toy, the bull (Beta(3, 3)) is anchored at 0.5. The 5 horses pull toward 0.6.
Result: 0.556, closer to the horses than the bull but not all the way. With 50 flips
and 30 heads the MAP would be $(30 + 2)/(50 + 4) \approx 0.593$ — almost at the MLE
of 0.6 — because 50 horses overwhelm one bull.

**Why it matters beyond coins.** In regression, "the prior" becomes a belief about how
big the coefficients should be. A Gaussian prior centred at zero says "I expect small
coefficients" — the bull pulls each coefficient toward zero. That is exactly what the
[ridge penalty](../ridge-regularization/lesson.qmd) does: it shrinks coefficients toward
zero. MAP with a Gaussian prior IS ridge. That's not a metaphor — it is the same
formula, derived below.


## What the notation is going to say

The upcoming formula says: "instead of just maximising the log-likelihood (as MLE
does), add the log of the prior and maximise the sum." The extra log-prior term acts as
a **penalty** — it rewards parameter values the prior likes and penalises values it
dislikes. In the coin example, the Beta(3, 3) log-prior adds a hill at 0.5, pushing the
peak from 0.6 down to 0.556.


## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $\hat{\theta}_{\text{MAP}}$ | "theta-hat MAP" | the MAP estimate: the mode (peak) of the posterior | 5/9 ≈ 0.556 |
| $\hat{\theta}_{\text{ML}}$ | "theta-hat ML" | the MLE: the mode of the likelihood | 3/5 = 0.6 |
| $f(\theta)$ | "the prior" | the prior density — your belief about $\theta$ before seeing data | Beta(3, 3), centred at 0.5 |
| $L(\theta)$ | "the likelihood" | how probable the data are under parameter $\theta$ | $p^3(1-p)^2$ |
| $\ell(\theta)$ | "the log-likelihood" | $\log L(\theta)$ — same peak, easier arithmetic | $3\log p + 2\log(1-p)$ |
| $\log f(\theta)$ | "log prior" | the log of the prior density — becomes the penalty term in MAP | $(3-1)\log p + (3-1)\log(1-p)$ |
| $\lambda$ | "lambda" (the regularisation weight) | in ridge/lasso: how hard to penalise large coefficients; in MAP: $\sigma^2/\sigma_0^2$ | strength of the prior's pull |
| $\arg\max$ | "arg max" | the value of the input that makes the expression largest | find the $p$ that peaks the posterior |
| $\beta$ | "beta" (coefficient vector) | the vector of regression coefficients being estimated | slope in the toy regression |
| $\sigma^2$ | "sigma squared" (noise variance) | variance of the observation noise $\varepsilon$ in the regression model | assumed known; controls how much the data pull |
| $\sigma_0^2$ | "sigma-zero squared" (prior variance) | variance of the Gaussian prior on each coefficient; small = tight prior | $\sigma^2/\lambda$ — determines the ridge penalty strength |


## The formal version

### MAP = argmax (log-likelihood + log-prior)

The posterior is $f(\theta \mid \text{data}) \propto L(\theta) \times f(\theta)$.
Taking the log: [@chan2021probabilitydatascience, pp. 506--507]

$$
\hat{\theta}_{\text{MAP}} = \arg\max_\theta \;
\bigl[\,\ell(\theta) + \log f(\theta)\,\bigr]
$$

Reading it aloud: "theta-hat MAP is the argmax over theta of the log-likelihood plus the
log-prior." In everyday words: "find the $\theta$ that maximises the posterior by
maximising the sum of two terms — one from the data, one from the prior."

Map it onto the coin toy with a Beta(3, 3) prior. The log-prior of Beta($\alpha, \beta$)
is (up to a constant) $(\alpha - 1)\log p + (\beta - 1)\log(1-p) = 2\log p + 2\log(1-p)$.
The log-likelihood is $3\log p + 2\log(1-p)$. Sum:

$$
3\log p + 2\log(1-p) + 2\log p + 2\log(1-p) = 5\log p + 4\log(1-p)
$$

This is the log of $p^5(1-p)^4$ — a Beta(6, 5) shape whose mode is
$5/(5+4) = 5/9 \approx 0.556$, exactly what we computed.

With a **uniform** prior ($\log f(\theta)$ = constant), the log-prior adds nothing
to the optimisation and MAP reduces to MLE. The prior is the *only* difference between
MAP and MLE.

```{r map-formula}
# MAP on the coin: maximise log-likelihood + log-prior
log_posterior <- function(p, heads, tails, alpha, beta) {
  heads * log(p) + tails * log(1 - p) +   # log-likelihood
  (alpha - 1) * log(p) + (beta - 1) * log(1 - p)  # log-prior
}

p_grid <- seq(0.01, 0.99, by = 0.001)
lp <- log_posterior(p_grid, heads = 3, tails = 2, alpha = 3, beta = 3)
cat("MAP (grid search):", p_grid[which.max(lp)], "\n")  # ≈ 0.556
cat("MAP (formula):    ", 5/9, "\n")
```


### Gaussian prior → ridge

Now apply MAP to regression. You have data $(x_i, y_i)$ and a linear model
$y = X\beta + \varepsilon$ with Gaussian noise $\varepsilon \sim N(0, \sigma^2)$.
The log-likelihood is (up to a constant):
[@chan2021probabilitydatascience, pp. 518--519]

$$
\ell(\beta) = -\frac{1}{2\sigma^2} \sum_i (y_i - x_i^\top \beta)^2
$$

Put a **Gaussian** prior on $\beta$: each $\beta_j \sim N(0, \sigma_0^2)$. Its log
(up to a constant) is:

$$
\log f(\beta) = -\frac{1}{2\sigma_0^2} \sum_j \beta_j^2
$$

Add them and maximise:

$$
\hat{\beta}_{\text{MAP}} = \arg\max_\beta \left[
  -\frac{1}{2\sigma^2}\sum_i(y_i - x_i^\top\beta)^2
  - \frac{1}{2\sigma_0^2}\sum_j \beta_j^2 \right]
$$

Maximising a negative sum is the same as *minimising* the positive version. Multiply
through by $2\sigma^2$:

$$
\hat{\beta}_{\text{MAP}} = \arg\min_\beta \left[
  \sum_i(y_i - x_i^\top\beta)^2 + \frac{\sigma^2}{\sigma_0^2}\sum_j \beta_j^2 \right]
$$

Set $\lambda = \sigma^2 / \sigma_0^2$ and this is **ridge regression**:

$$
\text{RSS} + \lambda \sum_j \beta_j^2
$$

The $\lambda$ that ridge treats as a "tuning knob" is the ratio of noise variance to
prior variance. A tight prior ($\sigma_0^2$ small) makes $\lambda$ large — strong
shrinkage. A wide prior ($\sigma_0^2$ large) makes $\lambda$ small — almost no
shrinkage, nearly MLE.

```{r ridge-map}
# Simple regression: y ~ bx (centered, no intercept)
x <- c(-1, 0, 1); y <- c(-1, 0.5, 2)

b_ols   <- sum(x * y) / sum(x^2)          # 1.5 (MLE)
b_ridge1 <- sum(x * y) / (sum(x^2) + 1)   # 1.0 (ridge, lambda=1)
b_ridge3 <- sum(x * y) / (sum(x^2) + 3)   # 0.6 (ridge, lambda=3)

cat("OLS (= MLE = MAP with flat prior):", b_ols, "\n")
cat("Ridge lambda=1 (= MAP, sigma0^2 = sigma^2):", b_ridge1, "\n")
cat("Ridge lambda=3 (= MAP, tight prior):", b_ridge3, "\n")
```


### Laplace prior → lasso

Replace the Gaussian prior with a **Laplace** (double-exponential) prior:
each $\beta_j$ has density $f(\beta_j) \propto \exp(-|\beta_j|/\alpha)$.
Its log is: [@chan2021probabilitydatascience, pp. 519--520]

$$
\log f(\beta) = -\frac{1}{\alpha}\sum_j |\beta_j|
$$

The MAP problem becomes (multiply through by $2\sigma^2$ as before — but the Laplace
has no $\tfrac{1}{2}$, so a factor of 2 survives):

$$
\hat{\beta}_{\text{MAP}} = \arg\min_\beta \left[
  \sum_i(y_i - x_i^\top\beta)^2 + \frac{2\sigma^2}{\alpha}\sum_j |\beta_j| \right]
$$

This is the **lasso** — the absolute-value penalty from the
[ridge-regularization module](../ridge-regularization/lesson.qmd). The Laplace
distribution has a sharp peak at zero — it concentrates more mass near zero than the
Gaussian does. Importantly, a continuous prior like the Laplace does not assign nonzero
probability to any single value (including zero); the sparsity comes from the
**non-differentiability** of the absolute-value penalty at zero. The crease in
$|\beta_j|$ at zero means the MAP optimisation can land exactly at zero — the penalty's
corner acts as a "catch" that traps small coefficients.

The difference in one picture: the Gaussian prior is a smooth bell (the squared penalty
is differentiable everywhere, so the MAP solution glides toward zero but never reaches
it), while the Laplace prior is a tent with a crease at zero (the absolute-value
penalty's corner is where the MAP solution snaps coefficients to exactly zero).


## Explore it

```{webr}
#| autorun: true
#| fig-alt: "Two-panel plot showing MAP coin estimate vs data size: with few flips MAP stays near the prior, with many flips it converges to MLE"
# explore-1 — LIVE: watch MAP move from the prior toward MLE as data accumulates.
# TRY IT: change alpha/beta to c(10,10) (strong prior) — how many flips before MAP
# reaches 0.59?
# PREDICT: with alpha=1, beta=1 (uniform prior), will MAP ever differ from MLE?
alpha <- 3; beta <- 3   # prior: Beta(alpha, beta)
true_p <- 0.6            # the "true" coin bias

flips <- c(5, 10, 20, 50, 100, 200)
results <- sapply(flips, function(n) {
  heads <- round(n * true_p)
  mle <- heads / n
  map <- (heads + alpha - 1) / (n + alpha + beta - 2)
  c(MLE = mle, MAP = map)
})

matplot(flips, t(results), type = "b", pch = c(1, 19), lty = 1, lwd = 2,
        col = c("grey50", "steelblue"),
        xlab = "number of flips", ylab = "estimate",
        main = sprintf("MAP (Beta(%d,%d) prior) converges to MLE", alpha, beta))
abline(h = 0.5, lty = 3, col = "tomato")
abline(h = true_p, lty = 2, col = "grey50")
legend("right", legend = c("MLE", "MAP", "prior centre", "true p"),
       col = c("grey50", "steelblue", "tomato", "grey50"),
       lty = c(1, 1, 3, 2), pch = c(1, 19, NA, NA), bty = "n")
```

```{webr}
#| autorun: true
#| fig-alt: "Plot comparing Gaussian and Laplace prior densities: Gaussian is a smooth bell, Laplace has a sharp peak at zero"
# explore-2 — LIVE: the Gaussian vs Laplace prior — why lasso zeros out.
# TRY IT: change sigma0 to 0.5 (tighter Gaussian) — does it look more like Laplace?
# PREDICT: which prior puts more weight near beta = 0?
sigma0 <- 1    # Gaussian prior std dev
alpha  <- 1    # Laplace prior scale

b_grid <- seq(-4, 4, length.out = 300)
gauss  <- dnorm(b_grid, 0, sigma0)
laplace <- (1 / (2 * alpha)) * exp(-abs(b_grid) / alpha)

plot(b_grid, gauss, type = "l", col = "steelblue", lwd = 2,
     xlab = expression(beta), ylab = "prior density",
     main = "Gaussian prior (ridge) vs Laplace prior (lasso)", ylim = c(0, max(laplace) * 1.1))
lines(b_grid, laplace, col = "tomato", lwd = 2)
abline(v = 0, lty = 3)
legend("topright", legend = c("Gaussian → ridge", "Laplace → lasso"),
       col = c("steelblue", "tomato"), lwd = 2, bty = "n")
```

```{webr}
#| autorun: false
# explore-3 — LIVE: MAP by grid search vs closed-form formula — they agree.
# TRY IT: change to alpha=10, beta=10 (stronger prior) — does MAP move?
n <- 5; heads <- 3; tails <- 2
alpha <- 3; beta <- 3

p_grid <- seq(0.01, 0.99, by = 0.001)
log_post <- heads * log(p_grid) + tails * log(1 - p_grid) +
            (alpha - 1) * log(p_grid) + (beta - 1) * log(1 - p_grid)
map_grid <- p_grid[which.max(log_post)]
map_formula <- (heads + alpha - 1) / (n + alpha + beta - 2)
cat("MAP (grid search):", map_grid, "\n")
cat("MAP (formula):    ", round(map_formula, 4), "\n")
cat("They agree:", abs(map_grid - map_formula) < 0.002, "\n")
```

```{webr}
#| autorun: false
# explore-4 — LIVE: ridge MAP in action on the toy regression.
# TRY IT: change lambda to 0 — does the estimate match OLS?
# PREDICT: what lambda makes the MAP slope exactly 0.5?
x <- c(-1, 0, 1); y <- c(-1, 0.5, 2)
lambda <- 1

b_ols   <- sum(x * y) / sum(x^2)
b_ridge <- sum(x * y) / (sum(x^2) + lambda)

cat("OLS slope:", b_ols, "\n")
cat("Ridge MAP slope (lambda =", lambda, "):", b_ridge, "\n")
cat("Shrinkage:", round((1 - b_ridge / b_ols) * 100, 1), "% toward zero\n")
```


## Common traps

**Trap: "MAP is always better than MLE."** MAP is optimal for maximising the posterior;
MLE is optimal for maximising the likelihood. Neither is universally superior. A strong
prior that is *wrong* (e.g. centred at the wrong value) makes the MAP estimate worse
than MLE, especially with little data. MAP's advantage depends entirely on the prior
being reasonable. In the toy, a Beta(50, 2) prior "knows" the coin is very biased —
if the coin is actually fair, that prior actively hurts.
[@chan2021probabilitydatascience, p. 520]

**Trap: "Ridge and lasso are just ad hoc penalties with no justification."** They *are*
justified — each is the MAP estimate under a specific prior. Ridge penalises
$\sum \beta_j^2$ because the Gaussian prior says "large coefficients are unlikely."
Lasso penalises $\sum |\beta_j|$ because the Laplace prior concentrates mass near zero,
and the non-differentiability of $|\beta_j|$ at zero lets the MAP solution snap
coefficients to exactly zero. The penalty isn't a hack — it is a *belief*, written in
math.
[@chan2021probabilitydatascience, pp. 518--520]


## Check yourself

**1.** If you use a uniform prior (every parameter value equally likely before seeing
data), what does the MAP estimate equal?

::: {.callout-note collapse="true" title="Answer"}
The MLE. A uniform prior's log is a constant, which drops out of the optimisation
$\arg\max[\ell(\theta) + \text{const}]$, leaving just $\arg\max \ell(\theta)$.
:::

**2.** A colleague fits a ridge regression with $\lambda = 5$. In MAP language, is
their Gaussian prior on the coefficients tight (small $\sigma_0^2$) or wide (large
$\sigma_0^2$)? Why?

::: {.callout-note collapse="true" title="Answer"}
Tight (small $\sigma_0^2$). Since $\lambda = \sigma^2 / \sigma_0^2$, a large $\lambda$
means $\sigma_0^2$ is small relative to $\sigma^2$. A small $\sigma_0^2$ means the
prior is concentrated near zero — a strong belief that coefficients should be small —
so the estimate shrinks aggressively.
:::

**3.** In your own words: explain to a colleague, in two sentences, what MAP does and
how it relates to ridge regression.

::: {.callout-note collapse="true" title="A model answer"}
MAP finds the most probable parameter value by maximising the log-likelihood plus the
log-prior — it is MLE with a penalty for parameter values the prior dislikes. Ridge
regression is MAP with a Gaussian prior centred at zero: the penalty
$\lambda\sum\beta_j^2$ is exactly the negative log of that prior.
:::


## What you can now do

- **State the MAP formula** — argmax of log-likelihood + log-prior — and explain how
  it differs from MLE by exactly one term.
- **Compute a MAP estimate** for the coin-flip toy under a Beta prior, by hand and in R.
- **Derive the ridge penalty** from a Gaussian prior on the coefficients, and name the
  relationship $\lambda = \sigma^2 / \sigma_0^2$.
- **Explain why lasso zeros out coefficients and ridge does not** — the Laplace prior
  has a sharp peak at zero, the Gaussian prior does not.

If any of these feels shaky, revisit [A tiny example first](#a-tiny-example-first)
before continuing — the MAP idea is one formula, and every regularisation method is a
disguised version of it.


## Where this goes next

The MAP-ridge and MAP-lasso equivalences reappear whenever a paper says "regularisation."
Knowing that the penalty is a prior lets you choose one intelligently: if you believe most coefficients are small but nonzero, use ridge (Gaussian prior); if
you believe most are near zero and want sparsity, use lasso (Laplace prior — the
non-differentiable corner, not a point mass at zero, is what drives coefficients to
exactly zero). More complex priors lead to more complex
penalties — elastic net, horseshoe, spike-and-slab — but the MAP recipe is always the
same: log-likelihood plus log-prior.
[@chan2021probabilitydatascience, pp. 517--520]

Ready for more reps? Do the [practice problems](practice.qmd).
