---
title: "Practice: From a line to a probability"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

Work these in order — they get harder. Use paper or the live cells; only then open the
answer. Paste this into any starter as needed:
`sigmoid <- function(z) 1 / (1 + exp(-z))`.

## Problem 1 — the toy example, new numbers

A logistic model has intercept $\beta_0 = 2$ and slope $\beta_1 = -0.5$. Compute the
linear score $\eta = \beta_0 + \beta_1 x$ for $x = 0$, $x = 4$, and $x = 8$, then squash
each with the sigmoid. Do the $x = 4$ case **by hand** first.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
sigmoid <- function(z) 1 / (1 + exp(-z))
b0 <- 2; b1 <- -0.5
x <- c(0, 4, 8)
```

::: {.callout-note collapse="true" title="Answer"}
By hand at $x = 4$: $\eta = 2 + (-0.5)(4) = 0$, so $\sigma(0) = 0.5$.

```{r p1-answer}
sigmoid <- function(z) 1 / (1 + exp(-z))
b0 <- 2; b1 <- -0.5
x   <- c(0, 4, 8)
eta <- b0 + b1 * x
rbind(eta = eta, prob = sigmoid(eta))   # eta = 2, 0, -2 ; prob = 0.881, 0.5, 0.119
```
The score falls as $x$ rises (negative slope), so the probability slides from $0.881$
down through $0.5$ to $0.119$ — all legal probabilities.
:::

## Problem 2 — fill in the missing step

We're evaluating the sigmoid at a score of $z = 3$. Here's the work with the **core
step blanked**:

> Step 1: $e^{-z} = e^{-3} \approx 0.0498$
> Step 2: **your step here** — build the denominator and take the reciprocal
> Step 3: $\sigma(3) \approx 0.953$

Supply Step 2.

```{webr}
#| autorun: false
# p2-starter — check sigma(3)
sigmoid <- function(z) 1 / (1 + exp(-z))
z <- 3
```

::: {.callout-note collapse="true" title="Answer"}
Step 2: add 1 to get the denominator, then take one over it: $1 + e^{-3} \approx 1.0498$,
and $\sigma(3) = 1/1.0498 \approx 0.953$.

```{r p2-answer}
z <- 3
denom <- 1 + exp(-z)     # ~1.0498
1 / denom                # ~0.953
```
The "add 1, then reciprocal" is the whole squashing move — it's what forces the answer
below 1 (denominator over 1) and above 0 (one over a positive number).
:::

## Problem 3 — reading probabilities off scores

Below are the linear scores for five customers from a logistic model. Turn them into
predicted probabilities, and say how many are predicted "more likely than not" (probability
above $0.5$).

```{webr}
#| autorun: false
# p3-starter — your job: convert scores to probabilities and count those above 0.5
sigmoid <- function(z) 1 / (1 + exp(-z))
scores <- c(-2.5, -0.3, 0.0, 1.2, 4.0)
```

::: {.callout-note collapse="true" title="Answer"}
```{r p3-answer}
sigmoid <- function(z) 1 / (1 + exp(-z))
scores <- c(-2.5, -0.3, 0.0, 1.2, 4.0)
probs  <- sigmoid(scores)
round(probs, 3)          # 0.076, 0.426, 0.500, 0.769, 0.982
sum(probs > 0.5)         # 2
```
Probabilities: $0.076, 0.426, 0.500, 0.769, 0.982$. Exactly **2** are above $0.5$ (the
scores $1.2$ and $4.0$). The score of $0$ lands exactly on $0.5$ (not above), and the two
negative scores fall below — the sign of the score decides which side of $0.5$ you're on.
:::

## Problem 4 — spot the error

A colleague wants the failure probability for a part with linear score $z = -3$ and
writes the following. It contains **exactly one mistake** — find it and fix it.

```{webr}
#| autorun: false
# p4-starter — run it; the answer comes out above 0.5, which can't be right for z = -3
bad_sigmoid <- function(z) 1 / (1 + exp(z))   # note the exponent
bad_sigmoid(-3)
```

::: {.callout-note collapse="true" title="Answer"}
The mistake is the **sign in the exponent**: they wrote `exp(z)` instead of `exp(-z)`.
That computes $1/(1 + e^{z})$, which is $\sigma(-z)$, not $\sigma(z)$ — so a negative
score wrongly comes out *above* $0.5$. For $z = -3$ the correct probability is small:

```{r p4-answer}
sigmoid <- function(z) 1 / (1 + exp(-z))
c(wrong = 1 / (1 + exp(-3)),   # what the buggy code returns for z = -3: ~0.953
  right = sigmoid(-3))          # correct: ~0.047
```
A part with a *negative* score should be *unlikely* to fail, so its probability must be
**below** $0.5$ — the correct $\sigma(-3) \approx 0.047$, not $0.953$. The minus sign in
$e^{-z}$ is doing essential work; drop it and the S-curve runs backwards.
:::

## Problem 5 — mixing it with the line *(uses simple-linear-regression's $\beta_0 + \beta_1 x$)*

A logistic model for whether a seed germinates uses temperature $x$ (in °C) with
$\beta_0 = -6$ and $\beta_1 = 0.3$. First, using the **linear-regression skill**, find
the temperature $x$ where the linear score $\eta$ equals $0$. Then explain, using this
module, what the germination probability is at that temperature.

```{webr}
#| autorun: false
# p5-starter — solve for x where eta = 0, then interpret with the sigmoid
sigmoid <- function(z) 1 / (1 + exp(-z))
b0 <- -6; b1 <- 0.3
```

::: {.callout-note collapse="true" title="Answer"}
Setting the line to zero (the linear-regression move — solve $\beta_0 + \beta_1 x = 0$
for $x$): $-6 + 0.3x = 0 \Rightarrow x = 6/0.3 = 20$°C.

```{r p5-answer}
sigmoid <- function(z) 1 / (1 + exp(-z))
b0 <- -6; b1 <- 0.3
x_star <- -b0 / b1            # solve the line = 0: x = 20
c(x_at_score_0 = x_star, prob_there = sigmoid(b0 + b1 * x_star))   # 20, and 0.5
```
At $x = 20$°C the score is $0$, so the germination probability is exactly $\sigma(0) =
0.5$ — a 50/50 seed. This temperature is the model's *tipping point*: below it germination
is predicted unlikely, above it likely. (The score-of-zero location is where the S-curve
crosses $0.5$ — the same crossing you saw in the lesson's explore plot.)
:::

## Problem 6 — a gentle cliffhanger

We keep saying a straight line is fit to something inside the sigmoid — but to *what*,
exactly? Here's a hint you can check now. Take a probability $p$, and compute
$\log\!\big(\tfrac{p}{1-p}\big)$ (its log-odds, from the
[probability-and-odds](../../../../foundations/probability-and-odds/lesson.qmd) module).
Do this for the three probabilities the sigmoid produced from scores $-1, 0, 1$. What do
you notice about the log-odds you get back?

```{webr}
#| autorun: false
# p6-starter — run the sigmoid forward, then take log-odds to go back
sigmoid <- function(z) 1 / (1 + exp(-z))
scores <- c(-1, 0, 1)
p <- sigmoid(scores)
```

::: {.callout-note collapse="true" title="Answer"}
```{r p6-answer}
sigmoid <- function(z) 1 / (1 + exp(-z))
scores <- c(-1, 0, 1)
p       <- sigmoid(scores)
log_odds <- log(p / (1 - p))
rbind(score = scores, prob = round(p, 3), log_odds = round(log_odds, 3))
```
The log-odds you recover are exactly the original scores: $-1, 0, 1$. That's the punchline
of the whole course sneaking up on you — **the sigmoid turns a score into a probability,
and the log-odds turns that probability right back into the score.** So the "line inside
the sigmoid" is really a line for the *log-odds*. You can't fully unpack why that's the
natural choice yet — [the next module](../02-odds-log-odds-and-the-logit/lesson.qmd) makes
it precise.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="courses/logistic-regression/modules/01-from-line-to-probability" pass="4"}

::: {.quiz-q qid="ltp-sigmoid-zero" qtype="mc" answer="b"}
The part at $x = 2$ (from the lesson's model with $\beta_0 = -1$, $\beta_1 = 0.5$)
has linear score $\eta = 0$. What is $\sigma(0)$?

a) $\sigma(0) = 0$: a score of zero means the outcome has no predicted probability
b) $\sigma(0) = 0.5$: $e^{-0} = 1$ so the denominator is 2 and the reciprocal is $0.5$
c) $\sigma(0) = 1$: zero is where the S-curve transitions from below 1 to at 1
:::

::: {.quiz-q qid="ltp-sigmoid-neg4" qtype="numeric" answer="0.018" tolerance="0.0005"}
For the lesson's part model ($\beta_0 = -1$, $\beta_1 = 0.5$), the part at $x = -6$
gets linear score $\eta = -4$. Compute $\sigma(-4)$. Round to three decimals.
:::

::: {.quiz-q qid="ltp-exact-bounds" qtype="mc" answer="a"}
For any finite score $z$, what keeps $\sigma(z) = 1/(1 + e^{-z})$ from ever equaling
exactly 1?

a) $e^{-z}$ is always positive, so the denominator $1 + e^{-z}$ is always greater than 1
b) The sigmoid is bounded by the probability axiom, which forbids values of exactly 1
c) For very large $z$, the computer rounds the output up to display exactly 1
:::

::: {.quiz-q qid="ltp-sigmoid-job" qtype="mc" answer="c"}
The lesson's part model scores the part at $x = 10$ as $\eta = 4$. What is the
sigmoid's job here?

a) Re-center the score by subtracting the mean so it lands near zero
b) Take the natural log of the score to compress it into $[0, 1]$
c) Squash the runaway score to a probability strictly inside $(0, 1)$
:::

::: {.quiz-q qid="ltp-linear-score" qtype="numeric" answer="4" tolerance="0"}
(From simple linear regression's line $\beta_0 + \beta_1 x$.) For the lesson's part
model ($\beta_0 = -1$, $\beta_1 = 0.5$), what is the linear score $\eta$ at $x = 10$?
Enter a whole number.
:::

:::

::::
