---
title: "Partial derivatives, and the gradient"
subtitle: "When a function has more than one knob: the slope along each direction, collected into the vector papers write as ∇f"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live cells below re-state their own data. Base R is all
# we need, so nothing to install.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

::: {.callout-tip title="This lesson runs in your browser" collapse="true"}
The code cells below with a **Run** button are live — powered by WebR (R compiled to
WebAssembly), they run right here on your phone or laptop, no install needed. Edit any
of them and re-run. The first run downloads a small R engine, so give it a few seconds.
Cells without a Run button are ordinary listings whose output was computed when the
page was built.
:::

## Where we're going

You already know how to measure the **slope** of a function that has *one* input — one
knob to turn — and you know that the bottom of a valley is where that slope is zero. But
the functions in real methods almost always have *many* inputs at once: several settings,
several parameters, several knobs. This lesson answers the natural next question: **what
does "slope" even mean when there is more than one knob?**

The answer has two parts. First, the **partial derivative**: freeze every knob but one,
and measure the slope along that single direction — a plain one-input slope in disguise.
Second, the **gradient**: do that once for each knob and collect all the slopes into a
vector. That vector points straight uphill, and its negative points straight downhill. We
will build both from the numeric slope tool you already have, and read the symbol papers
use for the gradient — $\nabla f$ — as ordinary R you can run.

**Builds on:** [the derivative as a slope](../derivative-as-slope/lesson.qmd) — we reuse
its numeric `slope_at()` central-difference tool and the fact that *slope zero means
momentarily flat*; [vectors & summation](../vectors-and-summation/lesson.qmd) — the
gradient IS a vector, so we lean on `sum`, indexing, and vector length; and
[loss functions & optimization](../loss-functions-and-optimization/lesson.qmd) — for the
"why we want slopes" framing: a slope is what tells you which way is downhill toward a
minimum.

## Warm-up: do you remember?

Before the new material, three quick reps — produce each answer from memory, then check.

**1.** From [the derivative as a slope](../derivative-as-slope/lesson.qmd): the numeric
slope tool is `slope_at <- function(g, x, h = 1e-5) (g(x + h) - g(x - h)) / (2 * h)`. For
$g(x) = x^2$, what does `slope_at(g, 3)` give, and what is the exact slope of $x^2$ at
$x = 3$?

```{webr}
#| autorun: false
# warmup1-starter — edit and Run to check
slope_at <- function(g, x, h = 1e-5) (g(x + h) - g(x - h)) / (2 * h)
g <- function(x) x^2
```

::: {.callout-note collapse="true" title="Answer"}
`slope_at(g, 3)` returns essentially **6** — a tiny rise ($g(3+h)$) minus a tiny fall
($g(3-h)$) divided by the total run $2h$. The exact slope of $x^2$ is $2x$, which is
$2\times3 = 6$. Shaky on why this measures the slope? Revisit
[the derivative as a slope](../derivative-as-slope/lesson.qmd).
:::

**2.** From [loss functions & optimization](../loss-functions-and-optimization/lesson.qmd):
for a U-shaped curve like $f(x) = x^2$, where along it is the slope exactly **zero**, and
what is special about that spot?

::: {.callout-note collapse="true" title="Answer"}
At the very **bottom** of the U — here $x = 0$. The slope is zero because the curve is
momentarily flat there, neither uphill nor downhill. That "flat at the bottom" fact is
what lets a slope *point you toward* a minimum. See
[loss functions & optimization](../loss-functions-and-optimization/lesson.qmd).
:::

**3.** From [vectors & summation](../vectors-and-summation/lesson.qmd): for the vector
`v <- c(6, 8)`, what is its **length** (also called its norm) — the everyday "how far from
the origin" distance? Say the formula in words, then give the value.

```{webr}
#| autorun: false
# warmup3-starter — edit and Run to check
v <- c(6, 8)
```

::: {.callout-note collapse="true" title="Answer"}
"Square each entry, add them, take the square root": $\sqrt{6^2 + 8^2} = \sqrt{36 + 64} =
\sqrt{100} = 10$. In R: `sqrt(sum(v^2))`. This is the same Pythagorean distance you know;
we'll use it at the end to ask *how steep* the gradient is. See
[vectors & summation](../vectors-and-summation/lesson.qmd).
:::

## A tiny example first

Picture a small machine with **two dials**, $a$ and $b$. However you set them, the machine
reports back a single number — think of it as a "cost". The rule is:

$$
f(a, b) = a^2 + b^2 .
$$

(For now just treat that as the machine's readout; we'll unpack the notation soon.) Set
both dials to zero and the cost is $0^2 + 0^2 = 0$ — the cheapest the machine gets. Set
$a = 3$, $b = 2$ and the cost is $3^2 + 2^2 = 9 + 4 = 13$.

**One dial at a time — the partial derivative.** Suppose you're standing at $a = 3$,
$b = 2$, and you ask: *if I nudge only dial $a$ and leave $b$ frozen at 2, how fast does
the cost change?* That's a plain one-input slope question — the very thing `slope_at()`
answers — except the "function of one input" is "cost as $a$ varies, with $b$ nailed to 2":

- Nudge $a$ up a hair to $3 + h$: cost is $(3+h)^2 + 2^2$.
- Nudge $a$ down a hair to $3 - h$: cost is $(3-h)^2 + 2^2$.
- Rise over run: the $2^2$ cancels, leaving the slope of $a^2$ at $a = 3$, which is 6.

So the slope *along the $a$ direction* is **6**. Do the same for dial $b$ (freeze $a = 3$,
wiggle $b$): the slope along the $b$ direction is the slope of $b^2$ at $b = 2$, which is
**4**. These two numbers — 6 and 4 — are the **partial derivatives** at $(3, 2)$.

**Collect them — the gradient.** Stack those two direction-slopes into a vector:
$(6, 4)$. That vector is the **gradient** at $(3, 2)$. It has one entry per dial, and each
entry is "how fast cost climbs if I push *that* dial (holding the others still)."

The same computations in R, reusing the exact slope tool from the derivative lesson:

```{r toy-example}
# The two-dial machine: cost = a^2 + b^2. Written to take a length-2 vector p = c(a, b).
f <- function(p) p[1]^2 + p[2]^2

# The numeric slope tool from the derivative-as-slope lesson (one input):
slope_at <- function(g, x, h = 1e-5) (g(x + h) - g(x - h)) / (2 * h)

# Partial in the 'a' direction at (3, 2): freeze b = 2, treat cost as a function of a.
df_da <- slope_at(function(a) f(c(a, 2)), 3)   # slope of a^2 at a = 3
df_da                                          # 6

# Partial in the 'b' direction at (3, 2): freeze a = 3, treat cost as a function of b.
df_db <- slope_at(function(b) f(c(3, b)), 2)   # slope of b^2 at b = 2
df_db                                          # 4

# The gradient = the two partials collected into one vector:
c(df_da, df_db)                                # 6 4
```

## Getting a feel for it

**A surface with one knob per input — the landscape picture.** A one-input function is a
*curve*: one knob, and the readout traces a line you can walk left and right along. Add a
second knob and it becomes a *surface* — a landscape. Our $f(a, b) = a^2 + b^2$ is a round
bowl: the floor is at $(0, 0)$ and the ground rises as you move away in any direction, like
standing in the bottom of a valley. Dial $a$ is your east-west position and dial $b$ is your
north-south position; the cost is your **altitude**. (Where the analogy strains: a real
landscape has exactly two ground directions, but a function can have ten knobs — a
ten-dimensional landscape you can't draw. The *math* works identically no matter how many
knobs; the picture is just the 2-knob case you can actually see.)

**Why "hold the others still"?** On a hillside, "what's the slope here?" has no single
answer — it depends which way you face. Walk due east and you might be climbing; walk due
north and you might be on the level. So to get *one* number you must commit to *one*
direction. The partial derivative $\partial f/\partial a$ commits to the pure-$a$ direction:
face straight along the $a$-axis, freeze every other dial, and read the ordinary
one-input slope. Freezing the others is not a trick — it's the only way the question
"the slope" has a definite answer. That's why each partial is *literally* a `slope_at()`
call on a one-input slice of the surface.

**Why collect them into a vector — and what it points at.** One partial tells you the slope
along one axis; but you can face *any* direction, not just along an axis. Here's the payoff:
the gradient vector $(6, 4)$ — the axis-slopes bundled together — points in the direction of
**steepest ascent**, the single compass bearing that climbs fastest from where you stand.
Turn it around, $-(6, 4) = (-6, -4)$, and you have **steepest descent** — the fastest way
*downhill*. That's the whole reason the gradient matters: it's the arrow that says "downhill
is *that* way," which is exactly what the loss-function story
([loss functions & optimization](../loss-functions-and-optimization/lesson.qmd)) says you
want when hunting for a minimum.

**What the numbers feel like.** At $(3, 2)$ the gradient is $(6, 4)$: both entries positive,
so from here pushing *either* dial up raises the cost — you're on the outside wall of the
bowl, and every axis climbs. The $a$-entry (6) is bigger than the $b$-entry (4), meaning the
wall is steeper in the $a$ direction than the $b$ direction — sensible, since you're farther
from the floor in $a$ (3) than in $b$ (2). Now walk to the very bottom, $(0, 0)$: both
partials become **zero**, so the gradient is $(0, 0)$ — flat in *every* direction at once.
That all-zero gradient is the multi-knob version of "slope zero at the bottom," and it's how
you recognize you've arrived at a valley floor (or a hilltop).

## What the notation is going to say

Two new symbols are coming. The first, $\partial f / \partial a$, is just "the slope of $f$
when only $a$ moves and everything else is held still" — the **6** we measured by freezing
$b$. The curly $\partial$ ("partial") is there only to remind you *other knobs exist and are
frozen*; if the function had a single input it would be the ordinary derivative $df/dx$ you
already know. The second, $\nabla f$, is a shorthand for "collect every partial into one
vector" — the $(6, 4)$ we assembled. So the whole story reads: *measure a one-knob slope for
each knob, then stack them.* Nothing beyond what we just did by hand.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $f(a, b)$ | "eff of a, b" | a function of two inputs — a surface, one knob per input (in R: `f <- function(p) ...`) | $f(a,b) = a^2 + b^2$ |
| $a, b$ | "a, b" | the two inputs (dials); together the point $(a, b)$ where you stand | $a = 3$, $b = 2$ |
| $h$ | "aitch" | the tiny step used to measure a slope (in R: `h = 1e-5`) | $10^{-5}$ |
| $\frac{\partial f}{\partial a}$ | "the partial derivative of f with respect to a" | slope of $f$ when only $a$ moves, all other inputs held still (in R: `slope_at()` on an $a$-only slice) | 6 |
| $\frac{\partial f}{\partial b}$ | "the partial derivative of f with respect to b" | slope of $f$ when only $b$ moves, all else held still | 4 |
| $\partial$ | "partial" (the curly-dee) | the "hold the other inputs still" version of the derivative-$d$ | used in $\partial f/\partial a$ |
| $\nabla f$ | "grad f" / "del f" | the **gradient**: the vector of every partial derivative, one per input (in R: our `grad()` below) | $(6, 4)$ |
| $-\nabla f$ | "minus grad f" | the gradient flipped — the steepest-**downhill** direction | $(-6, -4)$ |
| $\nabla f = \mathbf{0}$ | "grad f equals the zero vector" | every partial is zero — flat in every direction: a bottom or top | at $(0, 0)$ |
| $\Longleftrightarrow$ | "if and only if" | the two sides mean the same thing — each holds exactly when the other does | $\nabla f = \mathbf{0} \Longleftrightarrow$ flat everywhere |
| $\|\nabla f\|$ | "the norm of grad f" | the length of the gradient vector — *how steep* the steepest climb is (in R: `sqrt(sum(g^2))`) | $\sqrt{6^2+4^2} \approx 7.21$ |
| $k$ | "kay" | the number of inputs (dials); the gradient has one entry per input | $k = 2$ |

The curly $\partial$ and the ordinary straight $d$ mean the *same operation* — measure a
slope — the curly one just flags "there are other inputs, and they're frozen." The upside-down
triangle $\nabla$ ("nabla") is not an operation you compute entry-by-entry in your head;
read it as a *packaging instruction*: "gather the partials into a vector."

## The formal version

**One partial derivative.** The slope in the $a$ direction is the ordinary one-input slope
of $f$ measured while $b$ stays put — exactly the difference quotient from the derivative
lesson, applied to the $a$-slice:

$$
\frac{\partial f}{\partial a} \;\approx\; \frac{f(a + h,\, b) - f(a - h,\, b)}{2h}
$$

Reading it aloud: "the partial of f with respect to a is approximately: f at (a plus h, b)
minus f at (a minus h, b), all over two h" — everyday words: **"nudge only $a$ by a hair
each way, keep $b$ fixed, and take rise over run."** Notice $b$ is identical in both terms —
that's "held still." Plugging in our numbers at $(3, 2)$ with $f = a^2 + b^2$: the $b^2$
appears in both the plus and minus terms and cancels, leaving the slope of $a^2$ at 3, which
is **6** — the hand answer.

**The gradient.** The gradient just lists the partials, one per input:

$$
\nabla f(a, b) \;=\; \left( \frac{\partial f}{\partial a},\; \frac{\partial f}{\partial b} \right)
$$

Reading it aloud: "grad f equals the vector: partial of f with respect to a, partial of f
with respect to b" — everyday words: **"the gradient is the list of every direction's
slope."** For a function with $k$ inputs it has $k$ entries, one per input. At $(3, 2)$ it is
$(6, 4)$ — the vector we stacked. And the "you're at a bottom" condition is simply every
entry being zero:

$$
\nabla f = \mathbf{0} \quad\Longleftrightarrow\quad \text{flat in every direction (a bottom or top).}
$$

At $(0, 0)$ both partials are 0, so $\nabla f = (0, 0)$ — the multi-knob echo of "slope zero
at the valley floor" from the loss-function lesson. (This lesson *reads* that condition; the
recipe of setting it to zero and solving for the bottom by algebra belongs to
[loss functions & optimization](../loss-functions-and-optimization/lesson.qmd), and actually
*walking* downhill step by step is the gradient-descent course's job — neither is done here.)

**The numeric gradient — one central difference per coordinate.** The single-input
`slope_at()` measures one direction. To get every direction, we call the same central
difference once per coordinate, nudging *only that coordinate* each time. An `sapply` loop
does exactly that — it's `slope_at()` extended from one knob to many:

```{r grad-tool}
f <- function(p) p[1]^2 + p[2]^2       # cost = a^2 + b^2, taking p = c(a, b)

# grad(): one central difference per coordinate. This EXTENDS the 1-D slope_at()
# tool — for each coordinate j, nudge only p[j] by +/- h and take rise over run.
grad <- function(f, p, h = 1e-5) {
  sapply(seq_along(p), function(j) {
    p_plus  <- p; p_plus[j]  <- p_plus[j]  + h   # step ONLY coordinate j up
    p_minus <- p; p_minus[j] <- p_minus[j] - h   # step ONLY coordinate j down
    (f(p_plus) - f(p_minus)) / (2 * h)           # rise over run for that coordinate
  })
}

grad(f, c(3, 2))     # 6 4  — the gradient, matching our hand answer
grad(f, c(0, 0))     # 0 0  — flat in every direction: the bottom of the bowl
-grad(f, c(3, 2))    # -6 -4 — the steepest-DOWNHILL direction from (3, 2)
```

**Sanity-check against the by-hand partials.** Each entry of `grad()` should equal the
matching `slope_at()` call on that coordinate's slice — because that's exactly what it is:

```{r grad-sanity}
f <- function(p) p[1]^2 + p[2]^2
slope_at <- function(g, x, h = 1e-5) (g(x + h) - g(x - h)) / (2 * h)
grad <- function(f, p, h = 1e-5) {
  sapply(seq_along(p), function(j) {
    p_plus <- p; p_plus[j] <- p_plus[j] + h
    p_minus <- p; p_minus[j] <- p_minus[j] - h
    (f(p_plus) - f(p_minus)) / (2 * h)
  })
}

# The gradient, and the two partials computed the 1-D way, side by side:
rbind(
  grad     = grad(f, c(3, 2)),
  slice_by = c(slope_at(function(a) f(c(a, 2)), 3),   # a-slice slope
               slope_at(function(b) f(c(3, b)), 2))   # b-slice slope
)                                                     # both rows: 6 4
```

## Explore it

```{webr}
#| autorun: true
# explore-1 — LIVE: the gradient IS one 1-D slope per coordinate.
# TRY IT: move to a different point, e.g. change p to c(1, -2), then Run.
# PREDICT: at c(1, -2) for f = a^2 + b^2, the exact partials are 2a and 2b.
# Write down what grad should be, then run and check.
f <- function(p) p[1]^2 + p[2]^2
slope_at <- function(g, x, h = 1e-5) (g(x + h) - g(x - h)) / (2 * h)
grad <- function(f, p, h = 1e-5) {
  sapply(seq_along(p), function(j) {
    p_plus <- p; p_plus[j] <- p_plus[j] + h
    p_minus <- p; p_minus[j] <- p_minus[j] - h
    (f(p_plus) - f(p_minus)) / (2 * h)
  })
}
p <- c(3, 2)

by_hand <- c(slope_at(function(a) f(c(a, p[2])), p[1]),   # a-slice
             slope_at(function(b) f(c(p[1], b)), p[2]))   # b-slice
built    <- grad(f, p)

rbind(by_hand = by_hand, grad_tool = built)   # identical rows
```

```{webr}
#| autorun: true
#| fig-alt: "A contour map of the bowl f(a,b)=a^2+b^2 with concentric circles; a red arrow from the point (3,2) points outward away from the center (the uphill gradient), and a blue arrow points inward toward the center (downhill, minus the gradient)."
# explore-2 — LIVE: the gradient points UPHILL, minus-gradient points DOWNHILL.
# PREDICT: from (3, 2) on the bowl, will the uphill arrow point toward the center
# (0,0) or away from it? Commit, then run and look.
f <- function(p) p[1]^2 + p[2]^2
grad <- function(f, p, h = 1e-5) {
  sapply(seq_along(p), function(j) {
    p_plus <- p; p_plus[j] <- p_plus[j] + h
    p_minus <- p; p_minus[j] <- p_minus[j] - h
    (f(p_plus) - f(p_minus)) / (2 * h)
  })
}
p <- c(3, 2)
g <- grad(f, p)

# Draw the bowl's contour lines (rings of equal cost), then the two arrows.
as <- seq(-4, 4, length.out = 60)
z  <- outer(as, as, function(a, b) a^2 + b^2)
contour(as, as, z, asp = 1, xlab = "dial a", ylab = "dial b",
        main = "Bowl contours; red = uphill (grad), blue = downhill (-grad)")
points(p[1], p[2], pch = 16)
s <- 0.15                                   # shrink arrows so they fit the plot
arrows(p[1], p[2], p[1] + s*g[1],  p[2] + s*g[2],  col = "red",  lwd = 2)   # uphill
arrows(p[1], p[2], p[1] - s*g[1],  p[2] - s*g[2],  col = "blue", lwd = 2)   # downhill
```

```{webr}
#| autorun: false
# explore-3 — LIVE: grad f = 0 means flat in EVERY direction.
# PREDICT: at the bottom (0, 0), what will grad(f, c(0, 0)) be? And how steep is the
# gradient there — what is its length? Write both down, then run.
f <- function(p) p[1]^2 + p[2]^2
grad <- function(f, p, h = 1e-5) {
  sapply(seq_along(p), function(j) {
    p_plus <- p; p_plus[j] <- p_plus[j] + h
    p_minus <- p; p_minus[j] <- p_minus[j] - h
    (f(p_plus) - f(p_minus)) / (2 * h)
  })
}

g_bottom <- grad(f, c(0, 0))
g_wall   <- grad(f, c(3, 2))
c(bottom_grad = g_bottom, bottom_steepness = sqrt(sum(g_bottom^2)))  # 0 0 and 0
c(wall_grad   = g_wall,   wall_steepness   = sqrt(sum(g_wall^2)))    # 6 4 and ~7.21
```

## Common traps

**Trap 1 — "the partial derivative moves all the inputs at once."** It's tempting to
nudge $a$ *and* $b$ together, since both are inputs. But a partial derivative freezes every
input except the one named. Watch what wrongly moving both does at $(3, 2)$: nudge $a$ and
$b$ together by $h$ and take rise over run, and you get
$\big(f(3+h, 2+h) - f(3-h, 2-h)\big)/(2h) = \mathbf{10}$ — which is just $6 + 4$, the two
partials *added*. That's a slope along the diagonal direction, **not** the partial in the
$a$ direction (6) or the $b$ direction (4). The whole point of $\partial f/\partial a$ is
"only $a$ moves"; move both and you've measured a different thing entirely.

```{r trap1-numbers}
f <- function(a, b) a^2 + b^2
h <- 1e-5
wrong_both_move <- (f(3 + h, 2 + h) - f(3 - h, 2 - h)) / (2 * h)
wrong_both_move          # 10 — this is 6 + 4, a diagonal slope, NOT a partial derivative
```

**Trap 2 — "the gradient is a single slope number."** With one input the derivative is one
number, so it feels like the gradient should be too. But the gradient is a **vector** — one
slope *per input*. Our gradient at $(3, 2)$ is $(6, 4)$, two numbers, because there are two
dials; a machine with five dials has a five-entry gradient. Collapsing it to a single number
throws away *which direction* is steep. (If you want one summary number, that's the gradient's
*length* $\|\nabla f\| = \sqrt{6^2 + 4^2} \approx 7.21$ — "how steep," not "which way." They
answer different questions.)

## Check yourself

**1.** For $f(a, b) = a^2 + b^2$ at the point $(4, 1)$, give the gradient by hand (the exact
partials of $a^2 + b^2$ are $2a$ and $2b$), then check with the `grad()` tool.

```{webr}
#| autorun: false
# check1-starter — edit and Run to check your hand answer
f <- function(p) p[1]^2 + p[2]^2
grad <- function(f, p, h = 1e-5) {
  sapply(seq_along(p), function(j) {
    p_plus <- p; p_plus[j] <- p_plus[j] + h
    p_minus <- p; p_minus[j] <- p_minus[j] - h
    (f(p_plus) - f(p_minus)) / (2 * h)
  })
}
```

::: {.callout-note collapse="true" title="Answer"}
Partial in $a$: $2a = 2\times4 = 8$. Partial in $b$: $2b = 2\times1 = 2$. So the gradient is
$(8, 2)$. `grad(f, c(4, 1))` returns essentially `8 2`. Both entries are positive, so from
$(4, 1)$ every dial climbs — you're on the outer wall of the bowl, steeper in $a$ than $b$.
:::

**2.** A function has gradient $(0, 0)$ at some point. What does that tell you about the
surface there, in one sentence? Is that point definitely the bottom?

::: {.callout-note collapse="true" title="Answer"}
It tells you the surface is **flat in every direction** at that point — no direction goes up
or down, so you can't decrease $f$ by nudging any single dial. It's a bottom *or* a top (or a
saddle): $\nabla f = \mathbf{0}$ says "flat," not "lowest." For our bowl the only such point is
the true bottom $(0, 0)$, but "grad = 0" on its own only means flat. (Actually *sorting out*
bottom-vs-top-vs-saddle is beyond this lesson.)
:::

**3.** In your own words: explain to a colleague, in two sentences, what the gradient is and
why it's a vector rather than a single number.

::: {.callout-note collapse="true" title="A model answer"}
The gradient is the collection of a function's slopes — one slope for each input, measured by
wiggling that input alone while holding the others still. It has to be a vector because a
surface can be steep in one direction and flat in another, so a single number couldn't say
*which way* is uphill; the vector both points uphill and, when you flip its sign, points the
fastest way down.
:::

## What you can now do

- **Compute a partial derivative** by hand or with `slope_at()` — freeze every input but one,
  and read the ordinary one-input slope along that direction.
- **Assemble the gradient** as the vector of all the partials, using a `grad()` wrapper that
  runs one central difference per coordinate (the `slope_at()` tool extended to many knobs).
- **Read the gradient's meaning** — it points straight uphill, its negative $-\nabla f$ points
  straight downhill, and its length says *how steep* the steepest climb is.
- **Recognize $\nabla f = \mathbf{0}$** as "flat in every direction" — the multi-knob version
  of slope-zero-at-the-bottom, marking a valley floor or hilltop.

If any of these feels shaky, revisit [A tiny example first](#a-tiny-example-first) and
[The formal version](#the-formal-version) before continuing — anything that optimizes a
function of several inputs assumes all of these.

## Where this goes next

The gradient is the compass every multi-input optimizer steers by: once you can point to
"downhill is *that* way" from any spot on a surface, you have what it takes to *walk* toward a
minimum — repeatedly stepping in the $-\nabla f$ direction, which is the gradient-descent
idea. This lesson stops at reading the compass; taking the steps (and choosing how big) is the
next course's job. Ready for more reps? Do the [practice problems](practice.qmd).
