---
title: "Practice: Partial derivatives, and the gradient"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only (a {webr} starter does not see it — each starter re-states its
# own data). Base R is all we need, so nothing to install.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

Work these in order — they get harder. Do them on paper or in an R session; only then
open the answer. The gradient tool from the lesson is re-supplied in each starter so you
can copy it straight in.

## Problem 1 — the toy example, new numbers

A two-dial machine with cost $g(a, b) = 3a^2 + b^2$. At the point $(1, 2)$, find the two
partial derivatives by hand (the exact partials of $3a^2 + b^2$ are $6a$ and $2b$), then
assemble the gradient. Check with the `grad()` tool.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
g <- function(p) 3*p[1]^2 + p[2]^2
grad <- function(f, p, h = 1e-5) {
  sapply(seq_along(p), function(j) {
    p_plus <- p; p_plus[j] <- p_plus[j] + h
    p_minus <- p; p_minus[j] <- p_minus[j] - h
    (f(p_plus) - f(p_minus)) / (2 * h)
  })
}
```

::: {.callout-note collapse="true" title="Answer"}
Partial in $a$ (freeze $b = 2$): $6a = 6\times1 = 6$.
Partial in $b$ (freeze $a = 1$): $2b = 2\times2 = 4$.
So $\nabla g(1, 2) = (6, 4)$.

```{r p1-answer}
g <- function(p) 3*p[1]^2 + p[2]^2
grad <- function(f, p, h = 1e-5) {
  sapply(seq_along(p), function(j) {
    p_plus <- p; p_plus[j] <- p_plus[j] + h
    p_minus <- p; p_minus[j] <- p_minus[j] - h
    (f(p_plus) - f(p_minus)) / (2 * h)
  })
}
grad(g, c(1, 2))     # 6 4 — matches the hand answer
```
:::

## Problem 2 — fill in the missing step

Here is the gradient of $f(a, b) = a^2 + b^2$ at $(5, -1)$ worked out — but the **core
step is blanked**. Fill in ONLY the blanked line: the two partial derivatives.

> The gradient collects one partial per input. The exact partials of $a^2 + b^2$ are $2a$
> and $2b$.
> **`<your step here>`** — freeze the other input and read each slope:
> $\dfrac{\partial f}{\partial a} = \underline{\phantom{00}}$ (at $a = 5$),
> $\dfrac{\partial f}{\partial b} = \underline{\phantom{00}}$ (at $b = -1$).
> Collect them into $\nabla f(5, -1) = (\underline{\phantom{00}},\ \underline{\phantom{00}})$.

```{webr}
#| autorun: false
# p2-starter — edit and Run to check your two partials
f <- function(p) p[1]^2 + p[2]^2
grad <- function(f, p, h = 1e-5) {
  sapply(seq_along(p), function(j) {
    p_plus <- p; p_plus[j] <- p_plus[j] + h
    p_minus <- p; p_minus[j] <- p_minus[j] - h
    (f(p_plus) - f(p_minus)) / (2 * h)
  })
}
```

::: {.callout-note collapse="true" title="Answer"}
The blanked step is the heart of the gradient — one held-still slope per input:

- $\dfrac{\partial f}{\partial a} = 2a = 2\times5 = 10$ (freeze $b = -1$).
- $\dfrac{\partial f}{\partial b} = 2b = 2\times(-1) = -2$ (freeze $a = 5$).

So $\nabla f(5, -1) = (10, -2)$. The $b$-entry is negative: from here, pushing $b$ up
(toward 0) actually *lowers* the cost — you're on the $b < 0$ side of the bowl.

```{r p2-answer}
f <- function(p) p[1]^2 + p[2]^2
grad <- function(f, p, h = 1e-5) {
  sapply(seq_along(p), function(j) {
    p_plus <- p; p_plus[j] <- p_plus[j] + h
    p_minus <- p; p_minus[j] <- p_minus[j] - h
    (f(p_plus) - f(p_minus)) / (2 * h)
  })
}
grad(f, c(5, -1))     # 10 -2 — same as the hand answer
```
:::

## Problem 3 — a full problem with a mixed term

A cost surface where the two dials interact: $h(a, b) = a^2 + a\,b + b^2$. Now the partial
in $a$ *depends on $b$* (and vice versa). The exact partials are
$\partial h/\partial a = 2a + b$ and $\partial h/\partial b = a + 2b$.

(a) Find $\nabla h$ at $(1, 1)$ by hand, then check.
(b) Find $\nabla h$ at $(0, 0)$. What does the result say about that point?

```{webr}
#| autorun: false
# p3-starter — edit and Run to solve, right here
h_fun <- function(p) p[1]^2 + p[1]*p[2] + p[2]^2
grad <- function(f, p, delta = 1e-5) {
  sapply(seq_along(p), function(j) {
    p_plus <- p; p_plus[j] <- p_plus[j] + delta
    p_minus <- p; p_minus[j] <- p_minus[j] - delta
    (f(p_plus) - f(p_minus)) / (2 * delta)
  })
}
```

::: {.callout-note collapse="true" title="Answer"}
(a) At $(1, 1)$: $\partial h/\partial a = 2a + b = 2 + 1 = 3$; $\partial h/\partial b =
a + 2b = 1 + 2 = 3$. So $\nabla h(1, 1) = (3, 3)$.

(b) At $(0, 0)$: both partials are $0$, so $\nabla h(0, 0) = (0, 0)$ — **flat in every
direction**. That marks a bottom (or top). For this bowl-like surface it's the bottom.

```{r p3-answer}
h_fun <- function(p) p[1]^2 + p[1]*p[2] + p[2]^2
grad <- function(f, p, delta = 1e-5) {
  sapply(seq_along(p), function(j) {
    p_plus <- p; p_plus[j] <- p_plus[j] + delta
    p_minus <- p; p_minus[j] <- p_minus[j] - delta
    (f(p_plus) - f(p_minus)) / (2 * delta)
  })
}
grad(h_fun, c(1, 1))     # 3 3
grad(h_fun, c(0, 0))     # 0 0 — flat in every direction: a bottom
```
:::

## Problem 4 — spot the error

A colleague wants the partial derivative of $f(a, b) = a^2 + b^2$ with respect to $a$ at
the point $(2, 3)$, and writes the following. It contains **exactly one mistake** — find
it, explain why it's wrong, and give the correct value.

> "The partial in $a$ is just rise over run for a tiny step. I'll step the whole point by
> $h$ in *both* coordinates and divide by $2h$:
> $\dfrac{f(2+h,\ 3+h) - f(2-h,\ 3-h)}{2h}$. That gives **10**, so
> $\partial f/\partial a = 10$."

```{webr}
#| autorun: false
# p4-starter — copy this to see the two computations side by side
f <- function(a, b) a^2 + b^2
h <- 1e-5
```

::: {.callout-note collapse="true" title="Answer"}
The mistake is **moving both inputs at once**. A partial derivative in $a$ must hold $b$
*still* and nudge only $a$: the correct step is $f(2+h, 3)$ minus $f(2-h, 3)$, with $b$
identical in both terms. By stepping $b$ too, the colleague measured a *diagonal* slope,
not the $a$-partial. Their **10** is actually $\partial f/\partial a + \partial f/\partial b
= 4 + 6$ — the two partials added, which is a different quantity entirely.

The correct partial: $\partial f/\partial a = 2a = 2\times2 = 4$.

```{r p4-answer}
f <- function(a, b) a^2 + b^2
h <- 1e-5

# The colleague's (wrong) both-move step:
(f(2 + h, 3 + h) - f(2 - h, 3 - h)) / (2 * h)   # 10 — a diagonal slope, NOT the a-partial

# The correct a-partial: hold b = 3 still, nudge only a:
(f(2 + h, 3) - f(2 - h, 3)) / (2 * h)           # 4 — this is partial f / partial a
```
:::

## Problem 5 — how steep is the climb? *(uses [vectors & summation](../vectors-and-summation/lesson.qmd)'s vector length)*

The gradient is a vector, so it has a **length** — and that length is exactly *how steep*
the steepest climb is. For $f(a, b) = a^2 + b^2$ at the point $(3, 4)$:

(a) Find $\nabla f(3, 4)$ (the exact partials are $2a$ and $2b$).
(b) Using the vector-length formula $\|\mathbf{v}\| = \sqrt{\sum v_i^2}$ from
    [vectors & summation](../vectors-and-summation/lesson.qmd), compute $\|\nabla f\|$ — the
    steepness.

**PREDICT** first: write down the two gradient entries and guess the length, then run.

```{webr}
#| autorun: false
# p5-starter — edit and Run to solve, right here
f <- function(p) p[1]^2 + p[2]^2
grad <- function(f, p, delta = 1e-5) {
  sapply(seq_along(p), function(j) {
    p_plus <- p; p_plus[j] <- p_plus[j] + delta
    p_minus <- p; p_minus[j] <- p_minus[j] - delta
    (f(p_plus) - f(p_minus)) / (2 * delta)
  })
}
```

::: {.callout-note collapse="true" title="Answer"}
(a) $\partial f/\partial a = 2a = 6$; $\partial f/\partial b = 2b = 8$. So
$\nabla f(3, 4) = (6, 8)$.

(b) Its length is the same Pythagorean distance from
[vectors & summation](../vectors-and-summation/lesson.qmd):
$\|\nabla f\| = \sqrt{6^2 + 8^2} = \sqrt{36 + 64} = \sqrt{100} = 10$. So the steepest climb
from $(3, 4)$ rises at a rate of 10 — steeper than at $(3, 2)$, where it was only
$\sqrt{6^2 + 4^2} \approx 7.21$, because you're farther out on the wall of the bowl.

```{r p5-answer}
f <- function(p) p[1]^2 + p[2]^2
grad <- function(f, p, delta = 1e-5) {
  sapply(seq_along(p), function(j) {
    p_plus <- p; p_plus[j] <- p_plus[j] + delta
    p_minus <- p; p_minus[j] <- p_minus[j] - delta
    (f(p_plus) - f(p_minus)) / (2 * delta)
  })
}
g <- grad(f, c(3, 4))
g                    # 6 8
sqrt(sum(g^2))       # 10 — the vector length = the steepness of the climb
```
:::

## Problem 6 — a gentle cliffhanger (three knobs, then walking downhill)

Everything so far had two dials so we could draw the surface. But the `grad()` tool never
cared how many there were — it runs one central difference *per coordinate*, however many
you give it. Consider a three-dial cost $f(a, b, c) = a^2 + b^2 + c^2$ at the point
$(1, 2, 3)$.

**Predict:** the exact partials are $2a$, $2b$, $2c$. What will $\nabla f(1, 2, 3)$ be —
and which direction, $+\nabla f$ or $-\nabla f$, would you step in to make the cost
*smaller*? Then run the starter.

```{webr}
#| autorun: false
# p6-starter — grad() handles any number of coordinates. Run it on a length-3 point.
f <- function(p) p[1]^2 + p[2]^2 + p[3]^2
grad <- function(f, p, delta = 1e-5) {
  sapply(seq_along(p), function(j) {
    p_plus <- p; p_plus[j] <- p_plus[j] + delta
    p_minus <- p; p_minus[j] <- p_minus[j] - delta
    (f(p_plus) - f(p_minus)) / (2 * delta)
  })
}
```

::: {.callout-note collapse="true" title="Answer"}
$\nabla f(1, 2, 3) = (2a, 2b, 2c) = (2, 4, 6)$ — three entries, one per dial, and the tool
produces them with no change at all. All three are positive, so **stepping in the
$-\nabla f$ direction**, $(-2, -4, -6)$, is the way to make the cost smaller: downhill is
the negative of the gradient.

```{r p6-answer}
f <- function(p) p[1]^2 + p[2]^2 + p[3]^2
grad <- function(f, p, delta = 1e-5) {
  sapply(seq_along(p), function(j) {
    p_plus <- p; p_plus[j] <- p_plus[j] + delta
    p_minus <- p; p_minus[j] <- p_minus[j] - delta
    (f(p_plus) - f(p_minus)) / (2 * delta)
  })
}
grad(f, c(1, 2, 3))      # 2 4 6 — the gradient in three dimensions
-grad(f, c(1, 2, 3))     # -2 -4 -6 — the downhill direction
```

**You can't fully solve this next part yet:** *how big* a step to take in the $-\nabla f$
direction, and how to repeat it until you reach the bottom, is the **gradient-descent**
recipe — a whole course of its own. This lesson gives you the compass ($-\nabla f$ points
downhill); the next one teaches you how to walk.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/partial-derivatives-and-the-gradient" pass="4"}

::: {.quiz-q qid="pdg-partial-value" qtype="numeric" answer="6" tolerance="0"}
For $f(a, b) = a^2 + b^2$, the partial $\partial f / \partial a$ at $(3, 2)$ freezes $b$
and measures the slope in the $a$ direction. What is it?
:::

::: {.quiz-q qid="pdg-freeze-others" qtype="mc" answer="b"}
To compute the partial $\partial f / \partial a$, how do you nudge the inputs?

a) Nudge $a$ and $b$ together
b) Nudge only $a$, holding $b$ fixed
c) Nudge $b$, holding $a$ fixed
:::

::: {.quiz-q qid="pdg-gradient-is-vector" qtype="mc" answer="c"}
For a function of two inputs, the gradient is:

a) A single slope number
b) The larger of the two partial derivatives
c) A vector — one partial derivative per input
:::

::: {.quiz-q qid="pdg-gradient-length" qtype="numeric" answer="10" tolerance="0"}
The gradient at a point is $(6, 8)$. Its length $\lVert \nabla f \rVert$ — how steep the
steepest climb is — equals what?
:::

::: {.quiz-q qid="pdg-zero-gradient" qtype="mc" answer="a"}
(From loss functions & optimization.) With one input, the bottom of a bowl has slope
zero. What is the gradient at the floor of a multi-input valley?

a) The zero vector — flat in every direction
b) A vector pointing straight down
c) Undefined
:::

:::

::::
