---
title: "The switching search"
subtitle: "Add, remove, AND swap predictors — escaping the traps greedy stepwise walks into"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)

# Summary-stat BIC and the two searches (mirrored from equivalence/reimplementations).
sw_bic <- function(R, b, n, cols) {
  if (length(cols) == 0) return(0)
  RS <- R[cols, cols, drop = FALSE]; bS <- b[cols]
  r2 <- min(as.numeric(crossprod(bS, solve(RS, bS))), 1 - 1e-9)  # cap: summary R2 can exceed 1
  n * log(1 - r2) + length(cols) * log(n)
}
forward_only <- function(R, b, n) {
  p <- length(b); cur <- integer(0)
  repeat {
    cand <- setdiff(1:p, cur); if (!length(cand)) break
    bics <- sapply(cand, function(j) sw_bic(R, b, n, sort(c(cur, j))))
    if (min(bics) < sw_bic(R, b, n, cur) - 1e-9) cur <- sort(c(cur, cand[which.min(bics)])) else break
  }
  sort(cur)
}
switching <- function(R, b, n, start = integer(0)) {
  p <- length(b); cur <- sort(start); cb <- sw_bic(R, b, n, cur)
  repeat {
    mv <- list()
    for (j in setdiff(1:p, cur)) mv[[length(mv)+1]] <- sort(c(cur, j))
    for (j in cur)              mv[[length(mv)+1]] <- setdiff(cur, j)
    for (j in cur) for (k in setdiff(1:p, cur)) mv[[length(mv)+1]] <- sort(c(setdiff(cur, j), k))
    if (!length(mv)) break
    bics <- sapply(mv, function(m) sw_bic(R, b, n, m)); i <- which.min(bics)
    if (bics[i] < cb - 1e-9) { cur <- mv[[i]]; cb <- bics[i] } else break
  }
  sort(cur)
}
sim6 <- function(seed = 15, n = 200) {
  set.seed(seed); p <- 6
  Rt <- diag(p); Rt[1,2] <- Rt[2,1] <- 0.85; Rt[3,4] <- Rt[4,3] <- 0.6
  X <- matrix(rnorm(n*p), n, p) %*% chol(Rt)
  y <- drop(X %*% c(0.3, 0.45, 0.5, 0, 0, 0) + rnorm(n, sd = 1.2))
  list(X = X, y = y)
}
```

## Where we're going

We can now score any subset of indicators by its BIC, using only summaries (Module 1). Now
we *search*: which subset is best? Checking all $2^p$ subsets is hopeless for many
indicators, and plain forward stepwise — add the best predictor each step, never remove —
can walk into a trap it can't climb out of. The **switching search** adds one move: at each
step it also tries **removing** a predictor and **swapping** one for another, and takes
whichever move most improves BIC. That extra freedom lets it escape greedy's traps and,
on our tests, land on the same subset an exhaustive search would.

**Builds on:** [Module 1](../01-joint-fit-from-summaries/lesson.qmd) (summary-stat BIC),
[model-selection-and-information-criteria](../../../../foundations/model-selection-and-information-criteria/lesson.qmd)
(BIC, best-subset, why greedy stepwise gets stuck),
[hypothesis-tests-p-values-and-confidence-intervals](../../../../foundations/hypothesis-tests-p-values-and-confidence-intervals/lesson.qmd)
(for the p-value contrast).

## Warm-up: do you remember?

**1.** *(model selection)* Why can't we just keep the subset with the smallest RSS, and what
do we use instead?

::: {.callout-note collapse="true" title="Answer"}
RSS only falls as predictors are added, so it always picks the biggest model. We use a
penalized score — here **BIC** $= -2\log L + k\log n$ — that charges for complexity. See
[model-selection](../../../../foundations/model-selection-and-information-criteria/lesson.qmd).
:::

**2.** *(model selection)* What is the documented weakness of *forward* stepwise selection?

::: {.callout-note collapse="true" title="Answer"}
It's greedy and can't undo an early pick: once it adds a predictor, that predictor stays,
so an early commitment can trap it on a subset a fuller search would beat (ISLR Table 6.1).
:::

**3.** *(Module 1)* How do we get a subset $S$'s BIC from summaries $R$ and $b$?

::: {.callout-note collapse="true" title="Answer"}
$\mathrm{BIC}(S) = n\log(1 - b_S^\top R_S^{-1} b_S) + |S|\log n$ (up to a constant common to
all models), where $R_S, b_S$ are the sub-matrix and sub-vector for $S$. See
[Module 1](../01-joint-fit-from-summaries/lesson.qmd).
:::

## A tiny example first

Suppose you've scored four candidate subsets of indicators by their summary-stat BIC (lower
is better):

| Subset | BIC |
|---|---:|
| $\{x_2\}$ | −31.6 |
| $\{x_2, x_3\}$ | −58.0 |
| $\{x_1, x_2, x_3\}$ | −58.3 |
| $\{x_1, x_3\}$ | **−61.5** |

A **forward-only** search, starting empty, adds whichever single predictor helps most:
it adds $x_2$ (−31.6), then $x_3$ (−58.0), then $x_1$ (−58.3) — and now every *add* makes
BIC worse, so it **stops at $\{x_1, x_2, x_3\}$**. But look: $\{x_1, x_3\}$ scores −61.5,
*better* than where forward stopped. Forward can never reach it, because getting there means
**removing** $x_2$ — a move forward doesn't have. The switching search does have it: from
$\{x_1, x_2, x_3\}$ it tries removing each predictor, sees that dropping $x_2$ improves BIC
to −61.5, and takes it. These are real numbers from the panel below.

```{r toy-example}
sim <- sim6(seed = 15)                              # 6 indicators; x1,x2 correlated 0.86
R <- cor(sim$X); b <- as.numeric(cor(sim$X, sim$y)); n <- nrow(sim$X)

c(fwd_stops_at = paste(forward_only(R, b, n), collapse = ","),
  its_BIC      = round(sw_bic(R, b, n, forward_only(R, b, n)), 2),
  switch_finds = paste(switching(R, b, n), collapse = ","),
  its_BIC      = round(sw_bic(R, b, n, switching(R, b, n)), 2))
```

Forward stops at `1,2,3` (BIC −58.3); switching removes the redundant `2` and finishes at
`1,3` (BIC −61.5). Why is `x2` redundant? It's 0.86-correlated with `x1`, so once `x1` is in,
`x2` adds little — but forward had already committed to `x2` first (it was marginally the
strongest) and couldn't let go.

## Getting a feel for it

**Hiring vs hiring-and-firing.** Forward stepwise is a manager who can only *hire*: each
round they add the most impressive-looking candidate and can never let anyone go. If they
hire someone early who later turns out redundant once a better colleague joins, they're
stuck with the bloated team. The switching search can **hire, fire, and swap** — so when
`x1` joins and makes `x2` redundant, it fires `x2`. That one extra freedom is what lets it
reach subsets forward can't.

**Why swapping escapes traps.** Forward's problem is that its path is one-directional: the
only neighbours it considers are "current set plus one." A subset like $\{x_1, x_3\}$ isn't
a neighbour of $\{x_1, x_2, x_3\}$ under "add," but it *is* under "remove," and $\{x_2, x_4\}$
is a neighbour of $\{x_1, x_4\}$ under "swap." By considering add **and** remove **and** swap,
the search's neighbourhood is rich enough to walk downhill past the ridges that trap forward
— much like the multiple-regression course's point that a predictor's value depends on which
others are present.

**What "best" means here, honestly.** The switching search is still *greedy* on BIC — it
takes the best single move each step and stops at a **local** optimum (no single move helps).
That's not guaranteed to equal the global best-subset optimum. But it's dramatically closer
than forward-only, and on every seed we tested (see the validation below) it matched the
exhaustive best-subset answer exactly — for far less work than checking all $2^p$ subsets.

## What the notation is going to say

The search says: *hold a current subset $S$; look at every subset one move away — add any
one predictor, remove any one, or swap an in for an out; if the best of those has lower BIC
than $S$, move there; otherwise stop.* The stopping subset is the answer. "One move away" is
the whole difference from forward, which only ever looks at "add one."

## How to read the symbols

| Symbol | How to say it | What it means | In our example |
|---|---|---|---|
| $S$ | "ess" | the current subset of selected indicators | e.g. $\{1,2,3\}$ |
| $|S|$ | "size of S" | how many indicators are in $S$ | 3 |
| $\mathrm{BIC}(S)$ | "the BIC of S" | the summary-stat BIC of subset $S$ (Module 1) | −58.3 |
| add / remove / **swap** | — | the three moves: $S\cup\{j\}$, $S\setminus\{j\}$, $(S\setminus\{j\})\cup\{k\}$ | remove 2 |
| $\arg\min$ | "arg min" | the move giving the smallest BIC | pick best move |

## The formal version

The switching search, step by step:

1. **Start** from a subset $S$ (often empty, or forward's answer) and compute
   $\mathrm{BIC}(S)$ from summaries.
2. **Enumerate the neighbours** — every subset reachable in one move:
   - *add*: $S \cup \{j\}$ for each indicator $j \notin S$,
   - *remove*: $S \setminus \{j\}$ for each $j \in S$,
   - *swap*: $(S \setminus \{j\}) \cup \{k\}$ for each in-out pair.
3. **Score** each neighbour's BIC (each is a small $R^{-1}$ solve on summaries).
4. **Move** to the neighbour with the lowest BIC **if** it beats $\mathrm{BIC}(S)$;
   otherwise **stop** — $S$ is a local BIC optimum and is returned.

Reading it aloud: "score the current set; try every add, remove, and swap; jump to the best
if it lowers BIC, else stop." The stopping rule *is* BIC — no p-value threshold anywhere
(the next point explains why that matters).

**Validation.** For $p \le 12$ we can still afford the exhaustive best-subset search
(all $2^p$ subsets), so we can check the switching search against the true optimum. This
repo's equivalence harness freezes exactly that comparison on two panels (**tier: full**,
oracle = exhaustive best-subset over `BIC(lm)`): on a clean seed the switching search
returns the true `{rate_10y, cpi_infl, unemp}`; on a collinear seed it returns
`{rate_2y, pce_infl, unemp}` — and **both match the exhaustive oracle exactly**. (The second
picks the *noise* twins, honestly reproducing what full-data BIC does under collinearity —
the search is faithful, not clairvoyant.)

**BIC, not a p-value threshold.** A tempting alternative is "keep every predictor with
$p < 0.05$." But once you *search* over models, the usual p-values are no longer honest:
you deliberately hunted for the predictors that look most significant, so their p-values are
optimistically small (the [hypothesis-tests](../../../../foundations/hypothesis-tests-p-values-and-confidence-intervals/lesson.qmd)
module's caution, amplified by searching). This is **post-selection inference**, and naïve
p-values understate the true uncertainty. BIC sidesteps the threshold game: it scores whole
models by fit-plus-complexity, and comparing model scores is not the same as testing each
coefficient. (Honest post-selection p-values exist but need special methods — beyond this
course.)

## Explore it

```{webr}
#| autorun: true
# TRY IT: change the seed (or the true coefficients / correlations) and re-run. Watch how
# often forward-only STOPS at a bigger, worse-BIC subset than the switching search.
# PREDICT: will the two ever agree? (Yes — when forward's greedy path happens to be optimal.)
sw_bic <- function(R, b, n, cols) { if (!length(cols)) return(0)
  r2 <- min(as.numeric(crossprod(b[cols], solve(R[cols,cols,drop=FALSE], b[cols]))), 1 - 1e-9)
  n*log(1-r2) + length(cols)*log(n) }
forward_only <- function(R,b,n){p<-length(b);cur<-integer(0)
  repeat{cand<-setdiff(1:p,cur);if(!length(cand))break
    bx<-sapply(cand,function(j)sw_bic(R,b,n,sort(c(cur,j))))
    if(min(bx)<sw_bic(R,b,n,cur)-1e-9)cur<-sort(c(cur,cand[which.min(bx)]))else break};sort(cur)}
switching<-function(R,b,n){p<-length(b);cur<-integer(0);cb<-sw_bic(R,b,n,cur)
  repeat{mv<-list();for(j in setdiff(1:p,cur))mv<-c(mv,list(sort(c(cur,j))))
    for(j in cur)mv<-c(mv,list(setdiff(cur,j)))
    for(j in cur)for(k in setdiff(1:p,cur))mv<-c(mv,list(sort(c(setdiff(cur,j),k))))
    if(!length(mv))break;bx<-sapply(mv,function(m)sw_bic(R,b,n,m));i<-which.min(bx)
    if(bx[i]<cb-1e-9){cur<-mv[[i]];cb<-bx[i]}else break};sort(cur)}

set.seed(15); n <- 200; p <- 6
Rt <- diag(p); Rt[1,2]<-Rt[2,1]<-0.85; Rt[3,4]<-Rt[4,3]<-0.6
X <- matrix(rnorm(n*p),n,p) %*% chol(Rt); y <- drop(X %*% c(.3,.45,.5,0,0,0) + rnorm(n,sd=1.2))
R <- cor(X); b <- as.numeric(cor(X,y))
list(forward = forward_only(R,b,n), forward_BIC = round(sw_bic(R,b,n,forward_only(R,b,n)),2),
     switching = switching(R,b,n),  switch_BIC  = round(sw_bic(R,b,n,switching(R,b,n)),2))
```

## Common traps

**Trap: "the switching search is guaranteed to find the best subset."** It is a *greedy*
search that stops at a local BIC optimum, so in general it can miss the global best — it's a
heuristic, just a much stronger one than forward-only. It happened to match the exhaustive
optimum on every tested seed here, but that's an empirical result for these problems, not a
guarantee. (For small $p$ you can still run exhaustive best-subset to be sure.)

**Trap: "the selected predictors' p-values tell me they're significant."** After a search,
no. You picked the models that look best, so the reported p-values are optimistically small
— they don't account for the searching. Selection is done by BIC precisely to avoid leaning
on those dishonest thresholds; treating a post-selection p-value as a normal one overstates
your confidence.

## Check yourself

**1.** Forward-only stops at $\{x_1, x_4, x_6\}$ with BIC 102. The switching search reports
$\{x_1, x_6\}$ with BIC 99. Which move did switching make that forward could not, and is the
switching model better?

::: {.callout-note collapse="true" title="Answer"}
A **remove** (drop $x_4$) — forward can only add, so it could never shrink $\{x_1,x_4,x_6\}$.
Lower BIC (99 < 102) means the switching model is preferred: $x_4$ wasn't earning its
penalty once $x_1$ and $x_6$ were present.
:::

**2.** Why does the search use BIC as its stopping rule rather than "add predictors while
each is significant at $p < 0.05$"?

::: {.callout-note collapse="true" title="Answer"}
Because searching over many models invalidates the usual p-values — you cherry-picked the
significant-looking predictors, so their p-values are optimistically small (post-selection
inference). BIC scores whole models by fit-plus-complexity and doesn't rely on a per-
coefficient significance threshold, so it isn't fooled by the search the same way.
:::

**3.** In your own words: explain to a colleague, in two sentences, what the "switch" move
adds over forward stepwise and why it helps.

::: {.callout-note collapse="true" title="A model answer"}
Forward stepwise can only add predictors and never remove them, so an early pick that later
becomes redundant is stuck in the model. The switching search can remove and swap as well as
add, so it can undo early mistakes and reach lower-BIC subsets forward can never get to.
:::

## What you can now do

- **Run the switching search**: from a current subset, try every add/remove/swap, move to the
  best if it lowers BIC, stop otherwise.
- **Explain why it beats forward stepwise** (remove/swap escape early greedy commitments) and
  why it's still a heuristic (local, not guaranteed global).
- **Validate it** against exhaustive best-subset BIC for small $p$, and read the honest
  result (it can select noise twins under collinearity, matching the oracle).
- **Say why BIC, not a p-value threshold**, is the stopping rule (post-selection inference).

If any feels shaky, revisit [Getting a feel for it](#getting-a-feel-for-it) — Module 3 makes
this search fast enough to run on many indicators.

## Where this goes next

Each move re-solves an $R_S^{-1}$; on many indicators that's the bottleneck. Next:
[Rank-one speed-ups](../03-rank-one-speedups/lesson.qmd) replaces each re-inversion with a
cheap Sherman–Morrison update. Ready for reps? Do the [practice problems](practice.qmd).
External readings: [resources](../../resources.md).
