---
title: "Combining ABFs with priors"
subtitle: "The lH0..lH4 formulas, the sum-over-i-not-j trick, and building coloc_abf() from scratch"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live {webr} cells below re-state their own data. Base R only.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

::: {.callout-tip title="This lesson runs in your browser" collapse="true"}
The R cells below with a **Run** button are live (WebR) — you build `coloc_abf()` and check it
against the real `coloc` reference numbers right here. All of it is base R.
:::

## Where we're going

This is the method. You have the two per-SNP lABF vectors $\ell_1$, $\ell_2$ (Module 1) and
the five hypotheses with their priors (Module 2). Now we turn them into the five posterior
probabilities. The engine is three moves: (1) each hypothesis gets a **log support**
$\ell_{H_k}$ = log(prior) + log(evidence); (2) the tricky "distinct" hypothesis $H_3$ uses a
neat algebraic identity — the $\sum_{i\neq j}$ trick — to avoid a double loop over SNP pairs;
(3) everything is done in **log space** with two small helpers, `logsum` and `logdiff`, then
normalised to $\text{PP0}\dots\text{PP4}$. By the end you'll have built `coloc_abf()` and
checked it against `coloc`'s own output.

**Builds on:** [Module 1](../01-the-colocalisation-question/lesson.qmd) ($\ell_1, \ell_2$),
[Module 2](../02-five-hypotheses/lesson.qmd) (the five hypotheses and priors), and the
foundations [likelihood-and-log-likelihood](../../../../foundations/likelihood-and-log-likelihood/lesson.qmd),
[marginal-likelihood-and-model-evidence](../../../../foundations/marginal-likelihood-and-model-evidence/lesson.qmd),
and [bayes-factors-and-posterior-odds](../../../../foundations/bayes-factors-and-posterior-odds/lesson.qmd).

## Warm-up: do you remember?

Before the new stuff, three quick reps — answers from memory first, then check.

**1.** *(Module 2)* Which hypothesis means "one shared causal SNP for both traits," and which
means "two distinct causal SNPs"?

::: {.callout-note collapse="true" title="Answer"}
$H_4$ = shared (report PP4); $H_3$ = distinct (report PP3). See
[Module 2](../02-five-hypotheses/lesson.qmd).
:::

**2.** *(foundation)* Why do we add logs instead of multiplying probabilities directly?

::: {.callout-note collapse="true" title="Answer"}
Products of many tiny probabilities underflow to zero and lose precision; adding their logs
is numerically safe and turns "multiply the evidence" into "add the log-evidence." See
[likelihood-and-log-likelihood](../../../../foundations/likelihood-and-log-likelihood/lesson.qmd).
:::

**3.** *(foundation)* To turn a set of unnormalised scores into probabilities that sum to 1,
what do you do?

::: {.callout-note collapse="true" title="Answer"}
Divide each score by the total (normalise). In log space: subtract the log of the total,
$\log\!\sum e^{\text{score}}$, then exponentiate. See
[bayes-factors-and-posterior-odds](../../../../foundations/bayes-factors-and-posterior-odds/lesson.qmd).
:::

## A tiny example first

To see every step by hand, let's skip the ABF machine for a moment and work with **clean
per-SNP Bayes factors** directly — the actual $\ell$s are just logs of these. Take a
three-SNP region with

$$\text{BF}_1 = (1, 2, 5), \qquad \text{BF}_2 = (2, 1, 4).$$

(So $\ell_1 = \log \text{BF}_1$ and $\ell_2 = \log \text{BF}_2$.) The evidence pieces we need
are all sums:

- Trait 1's total evidence: $\sum \text{BF}_1 = 1 + 2 + 5 = 8$.
- Trait 2's total evidence: $\sum \text{BF}_2 = 2 + 1 + 4 = 7$.
- The **shared** ($H_4$) evidence, one SNP causal for both:
  $\sum_i \text{BF}_{1,i}\text{BF}_{2,i} = 1\cdot2 + 2\cdot1 + 5\cdot4 = 2 + 2 + 20 = 24$.
- The **distinct** ($H_3$) evidence, trait 1 at SNP $i$ and trait 2 at a *different* SNP $j$:
  $\sum_{i\neq j}\text{BF}_{1,i}\text{BF}_{2,j}$. Instead of listing all 6 off-diagonal pairs,
  use the identity (derived below): $(\sum \text{BF}_1)(\sum \text{BF}_2) - \sum_i
  \text{BF}_{1,i}\text{BF}_{2,i} = 8 \times 7 - 24 = 56 - 24 = 32$.

Now attach priors. Use toy priors $p_1 = p_2 = 0.1$, $p_{12} = 0.05$ (bigger than coloc's
real defaults so the toy numbers stay readable). The five (unnormalised) supports are
prior $\times$ evidence:

$$
\begin{aligned}
S_0 &= 1 \\
S_1 &= p_1 \cdot 8 = 0.8 \\
S_2 &= p_2 \cdot 7 = 0.7 \\
S_3 &= p_1 p_2 \cdot 32 = 0.01 \times 32 = 0.32 \\
S_4 &= p_{12} \cdot 24 = 0.05 \times 24 = 1.2
\end{aligned}
$$

Normalise by the total $S_0+\dots+S_4 = 1 + 0.8 + 0.7 + 0.32 + 1.2 = 4.02$:
$\text{PP4} = 1.2 / 4.02 \approx 0.299$ — the biggest, so the toy region leans (weakly)
toward a shared variant.

The same thing in R, done directly (we'll switch to log space right after):

```{r toy-example}
BF1 <- c(1, 2, 5); BF2 <- c(2, 1, 4)
p1 <- 0.1; p2 <- 0.1; p12 <- 0.05

sum1 <- sum(BF1)                 # 8   total evidence, trait 1
sum2 <- sum(BF2)                 # 7   total evidence, trait 2
diag <- sum(BF1 * BF2)           # 24  shared (H4): same SNP for both
offdiag <- sum1 * sum2 - diag    # 32  distinct (H3): the i != j trick

S <- c(H0 = 1,
       H1 = p1 * sum1,
       H2 = p2 * sum2,
       H3 = p1 * p2 * offdiag,
       H4 = p12 * diag)
round(S / sum(S), 4)             # PP0..PP4, summing to 1
```

## Getting a feel for it

**Analogy — a multiplication table with the diagonal removed.** Lay trait 1's Bayes factors
down the rows and trait 2's across the columns of a $3\times 3$ grid; each cell $(i,j)$ holds
$\text{BF}_{1,i}\text{BF}_{2,j}$ — the evidence that trait 1 is causal at SNP $i$ **and**
trait 2 at SNP $j$. The **whole grid's total** is $(\sum \text{BF}_1)(\sum \text{BF}_2) = 56$.
The **diagonal** cells ($i = j$) are the *shared* configurations (same SNP for both) — that's
$H_4$, summing to 24. Everything **off** the diagonal ($i \neq j$) is the *distinct*
configurations — that's $H_3$. So the distinct evidence is just "the whole grid minus its
diagonal": $56 - 24 = 32$. That is the entire $\sum_{i\neq j}$ trick, seen at a glance.

Where the analogy breaks: for a real region of thousands of SNPs the grid has millions of
cells, so you would never build it. The identity is precious exactly *because* it computes
"grid minus diagonal" from three cheap sums, never touching the grid.

**Why prior $\times$ evidence, then normalise?** Each hypothesis's support is "how plausible
before data" (its prior) times "how well it explains the region's evidence" (a sum of Bayes
factors over the configurations it allows). That product is an unnormalised posterior weight —
the numerator of Bayes' rule. Dividing by the total is Bayes' rule's denominator, and it
forces the five to sum to 1. It's the exact machinery from
[bayes-factors-and-posterior-odds](../../../../foundations/bayes-factors-and-posterior-odds/lesson.qmd),
just with five models instead of two.

**What the numbers feel like.** In the toy, $S_4 = 1.2$ edged out everything because SNP 3
was loud in *both* traits ($5$ and $4$), pumping the diagonal. If trait 2's loud SNP had been
a *different* one, the diagonal would shrink and the off-diagonal (distinct) would swell,
pushing weight from PP4 to PP3. With coloc's real priors ($p_{12} = 10^{-5}$, ten times
rarer than $p_1$), sharing needs *much* louder aligned evidence to win — which is why real
PP4 near 1 is a strong statement.

## What the notation is going to say

Everything above, but in **log space**, because real BFs are astronomically large or tiny.
Two helpers do the heavy lifting:

- $\text{logsum}(x) = \log\!\big(\sum_i e^{x_i}\big)$ — "add things that are stored as logs,"
  computed stably. Applied to $\ell_1$ it gives $\log \sum \text{BF}_1$.
- $\text{logdiff}(x, y) = \log\!\big(e^x - e^y\big)$ for $x > y$ — "subtract two things stored
  as logs." It's how we do "grid minus diagonal" without leaving log space.

Then each hypothesis's log support is $\log(\text{prior}) + \log(\text{evidence})$, and the
distinct case's log evidence is exactly $\text{logdiff}$ of "whole grid (in logs)" and
"diagonal (in logs)."

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $\ell_{1,i}, \ell_{2,i}$ | "ell one/two sub eye" | per-SNP log Bayes factors (Module 1) | $\ell_1 = \log(1,2,5)$ |
| $\text{BF}_{1,i}$ | "bee-eff one sub eye" | trait 1's Bayes factor at SNP $i$ ($=e^{\ell_{1,i}}$) | $1, 2, 5$ |
| $\sum_i$ | "the sum over i" | add over all SNPs (in R: `sum(...)`) | $\sum \text{BF}_1 = 8$ |
| $\sum_{i\neq j}$ | "the sum over i not equal to j" | add over all SNP *pairs with different indices* | the off-diagonal, $32$ |
| $\text{logsum}(x)$ | "log-sum of ex" | $\log\sum e^{x_i}$, stably (log of a sum) | $\text{logsum}(\ell_1)=\log 8$ |
| $\text{logdiff}(x,y)$ | "log-diff of ex and why" | $\log(e^x - e^y)$, $x>y$ (log of a difference) | $\log(56-24)=\log 32$ |
| $s_1, s_2$ | "ess one, ess two" | log total evidence per trait, $\text{logsum}(\ell_1)$ and $\text{logsum}(\ell_2)$ | $s_1=\log 8,\ s_2=\log 7$ |
| $s_{12}$ | "ess one-two" | log diagonal (shared) evidence, $\text{logsum}(\ell_1+\ell_2)$ | $s_{12}=\log 24$ |
| $\ell_{H_k}$ | "ell aitch sub kay" | log support of hypothesis $H_k$ = log prior + log evidence | $\ell_{H_4}=\log(0.05\cdot24)$ |
| $p_1, p_2, p_{12}$ | "pee one, pee two, pee one-two" | the per-SNP priors (Module 2) | toy: $0.1, 0.1, 0.05$ |
| $\text{PP}_k$ | "pee-pee sub kay" | posterior prob. of $H_k$ (sum to 1) | PP4 $\approx 0.299$ |

## The formal version

First the identity that powers $H_3$. We want the distinct evidence $\sum_{i\neq j}
\text{BF}_{1,i}\text{BF}_{2,j}$. Build it up:

$$
\begin{aligned}
\Big(\sum_i \text{BF}_{1,i}\Big)\Big(\sum_j \text{BF}_{2,j}\Big)
  &= \sum_i \sum_j \text{BF}_{1,i}\,\text{BF}_{2,j} \\
  &= \sum_{i = j} \text{BF}_{1,i}\,\text{BF}_{2,j} \;+\; \sum_{i \neq j} \text{BF}_{1,i}\,\text{BF}_{2,j} \\
  &= \sum_i \text{BF}_{1,i}\,\text{BF}_{2,i} \;+\; \sum_{i \neq j} \text{BF}_{1,i}\,\text{BF}_{2,j}.
\end{aligned}
$$

- Line 1: a product of two sums is the sum of all pairwise products (expand the product —
  the distributive law, one term per cell of the grid).
- Line 2: split the double sum into the diagonal ($i = j$) plus the off-diagonal ($i \neq j$) —
  a sum can be split into two sums over disjoint pieces (taught in
  [vectors-and-summation](../../../../foundations/vectors-and-summation/lesson.qmd)).
- Line 3: on the diagonal $j$ equals $i$, so $\text{BF}_{2,j} = \text{BF}_{2,i}$ — just
  relabel.

Rearranging the last line gives the identity we use:

$$
\sum_{i \neq j} \text{BF}_{1,i}\,\text{BF}_{2,j}
= \Big(\sum_i \text{BF}_{1,i}\Big)\Big(\sum_j \text{BF}_{2,j}\Big) - \sum_i \text{BF}_{1,i}\,\text{BF}_{2,i}.
$$

Reading it aloud: "the distinct-pair sum equals the product of the two totals, minus the
matched (same-SNP) sum" — grid total minus diagonal.

Sanity check on the toy: $8 \times 7 - 24 = 56 - 24 = 32$. ✓ (matches the hand count above).

Now the five log supports. With $s_1 = \text{logsum}(\ell_1)$, $s_2 = \text{logsum}(\ell_2)$,
and $s_{12} = \text{logsum}(\ell_1 + \ell_2)$ (note: $\ell_{1,i} + \ell_{2,i} = \log(\text{BF}_{1,i}\text{BF}_{2,i})$, so $e^{s_{12}}$ is the diagonal sum):

$$
\begin{aligned}
\ell_{H_0} &= 0 \\
\ell_{H_1} &= \log p_1 + s_1 \\
\ell_{H_2} &= \log p_2 + s_2 \\
\ell_{H_3} &= \log p_1 + \log p_2 + \text{logdiff}(s_1 + s_2,\; s_{12}) \\
\ell_{H_4} &= \log p_{12} + s_{12}
\end{aligned}
$$

- $\ell_{H_0} = 0$: the null's Bayes factor against itself is 1, and $\log 1 = 0$ — it's the
  reference everything else is measured against.
- $\ell_{H_1}, \ell_{H_2}$: log prior + log of that trait's total evidence ($s_1 = \log\sum\text{BF}_1$).
- $\ell_{H_3}$: log priors + $\text{logdiff}$ of "log grid total" ($s_1 + s_2$, since
  $\log[(\sum\text{BF}_1)(\sum\text{BF}_2)] = s_1 + s_2$) minus "log diagonal" ($s_{12}$) —
  the identity, in log space.
- $\ell_{H_4}$: log prior + log diagonal ($s_{12}$).

Finally normalise: $\text{PP}_k = \exp\!\big(\ell_{H_k} - \text{logsum}(\ell_{H_0},\dots,\ell_{H_4})\big)$.

Sanity check that the log-space route reproduces the direct toy numbers:

```{r formal-sanity}
logsum  <- function(x) { m <- max(x); m + log(sum(exp(x - m))) }
logdiff <- function(x, y) { m <- max(x, y); m + log(exp(x - m) - exp(y - m)) }

l1 <- log(c(1, 2, 5)); l2 <- log(c(2, 1, 4))   # the toy BFs, as logs
p1 <- 0.1; p2 <- 0.1; p12 <- 0.05

s1  <- logsum(l1)        # = log 8
s2  <- logsum(l2)        # = log 7
s12 <- logsum(l1 + l2)   # = log 24 (the diagonal)

lH <- c(H0 = 0,
        H1 = log(p1)  + s1,
        H2 = log(p2)  + s2,
        H3 = log(p1)  + log(p2)  + logdiff(s1 + s2, s12),  # logdiff = log(56 - 24) = log 32
        H4 = log(p12) + s12)
pp <- exp(lH - logsum(lH))
round(pp, 4)             # identical to the direct S/sum(S) above: PP4 ~ 0.2985
```

## Explore it

```{webr}
#| autorun: true
# TRY IT: move trait 2's loud SNP. Right now BF2's biggest (4) is at SNP 3, same as BF1's (5)
#         -> shared -> PP4 wins. Change BF2 to c(4, 1, 2) so its loud SNP is #1 instead.
# PREDICT: with the loud SNPs on DIFFERENT SNPs, which grows -- PP3 or PP4?
logsum  <- function(x) { m <- max(x); m + log(sum(exp(x - m))) }
logdiff <- function(x, y) { m <- max(x, y); m + log(exp(x - m) - exp(y - m)) }

BF1 <- c(1, 2, 5); BF2 <- c(2, 1, 4)      # <- edit BF2
l1 <- log(BF1); l2 <- log(BF2)
p1 <- 0.1; p2 <- 0.1; p12 <- 0.05

s1 <- logsum(l1); s2 <- logsum(l2); s12 <- logsum(l1 + l2)
lH <- c(0, log(p1)+s1, log(p2)+s2, log(p1)+log(p2)+logdiff(s1+s2, s12), log(p12)+s12)
pp <- exp(lH - logsum(lH))
setNames(round(pp, 4), c("PP0","PP1","PP2","PP3","PP4"))
```

Now assemble the real thing — `coloc_abf()` — starting from raw summary statistics, exactly
as `coloc.abf` does. It stacks the ABF machine from Module 1 in front of the combination
above.

```{webr}
#| autorun: true
# LIVE: the complete from-scratch method (base R). This IS what coloc.abf computes.
# PREDICT: SNP 2 is the loud one in both traits here -- which PP will come out largest?
wakefield_abf <- function(beta, se, W) { z <- beta/se; V <- se^2; r <- W/(W+V); 0.5*(log(1-r)+r*z^2) }
.logsum  <- function(x) { m <- max(x); m + log(sum(exp(x - m))) }
.logdiff <- function(x, y) { m <- max(x, y); m + log(exp(x - m) - exp(y - m)) }

coloc_abf <- function(beta1, se1, beta2, se2, W1, W2,
                      p1 = 1e-4, p2 = 1e-4, p12 = 1e-5) {
  l1 <- wakefield_abf(beta1, se1, W1)   # per-SNP lABF, trait 1  (Module 1)
  l2 <- wakefield_abf(beta2, se2, W2)   # per-SNP lABF, trait 2
  s1  <- .logsum(l1)                    # log total evidence, trait 1
  s2  <- .logsum(l2)                    # log total evidence, trait 2
  s12 <- .logsum(l1 + l2)               # log diagonal (shared) evidence
  lH0 <- 0
  lH1 <- log(p1)  + s1
  lH2 <- log(p2)  + s2
  lH3 <- log(p1)  + log(p2)  + .logdiff(s1 + s2, s12)   # distinct: grid - diagonal
  lH4 <- log(p12) + s12                                 # shared
  all <- c(lH0, lH1, lH2, lH3, lH4)
  list(pp = exp(all - .logsum(all)))
}

# a quick shared-looking region (loud SNP #2 in both traits)
out <- coloc_abf(beta1 = c(0.01, 0.20, -0.02), se1 = rep(0.03, 3),
                 beta2 = c(-0.01, 0.19, 0.02), se2 = rep(0.035, 3),
                 W1 = 0.04, W2 = 0.04)
setNames(round(out$pp, 4), c("PP0","PP1","PP2","PP3","PP4"))
```

### The self-check against `coloc` {#self-check}

Prove your `coloc_abf()` matches the real package. The cell embeds the **frozen reference**
from `equivalence/fixtures/coloc-abf-shared.json` — the exact inputs and the
$\text{PP0}\dots\text{PP4}$ vector that `coloc::coloc.abf` 5.2.3 produced — and compares.

```{webr}
#| autorun: true
# LIVE SELF-CHECK: your coloc_abf() vs the FROZEN coloc reference (scenario: shared).
wakefield_abf <- function(beta, se, W) { z <- beta/se; V <- se^2; r <- W/(W+V); 0.5*(log(1-r)+r*z^2) }
.logsum  <- function(x) { m <- max(x); m + log(sum(exp(x - m))) }
.logdiff <- function(x, y) { m <- max(x, y); m + log(exp(x - m) - exp(y - m)) }
coloc_abf <- function(beta1, se1, beta2, se2, W1, W2, p1 = 1e-4, p2 = 1e-4, p12 = 1e-5) {
  l1 <- wakefield_abf(beta1, se1, W1); l2 <- wakefield_abf(beta2, se2, W2)
  s1 <- .logsum(l1); s2 <- .logsum(l2); s12 <- .logsum(l1 + l2)
  all <- c(0, log(p1)+s1, log(p2)+s2, log(p1)+log(p2)+.logdiff(s1+s2, s12), log(p12)+s12)
  list(pp = exp(all - .logsum(all)))
}

# frozen inputs + reference, copied verbatim from equivalence/fixtures/coloc-abf-shared.json
beta1 <- c(-0.009781094715461567, 0.01657385566257415, -0.02024831531867491, 0.16,
           0.00932307651940807, 0.03521898862688097, 0.01856369566877904,
           -0.003382029442626436, 0.02751084868538136, -0.006697780938817877)
se1 <- rep(0.03, 10)
beta2 <- c(0.01579344296620927, -0.02384533306245165, 0.04283266634049109, 0.15,
           -0.007100501358086871, -0.005800138949925011, -0.02549264221001584,
           0.00175396493548505, -0.02453011067627874, -0.06150923446918894)
se2 <- rep(0.035, 10)
W1 <- 0.04; W2 <- 0.04; p1 <- 1e-4; p2 <- 1e-4; p12 <- 1e-5
reference_pp <- c(0.0004764450235039068, 0.007762352170233302, 6.09890757517626e-05,
                  1.949819640442383e-06, 0.9916982639108702)

mine <- coloc_abf(beta1, se1, beta2, se2, W1, W2, p1, p2, p12)$pp
cat("PP4 (shared):", round(mine[5], 4), "-> colocalisation\n")
cat("max |mine - coloc reference| =", format(max(abs(mine - reference_pp)), scientific = TRUE), "\n")
cat("match within 1e-6?", max(abs(mine - reference_pp)) < 1e-6, "\n")   # TRUE
```

**PP4 $\approx 0.992$** — a clean shared-variant call, matching `coloc` to about $10^{-6}$.
You just rebuilt `coloc.abf`'s core in base R.

::: {.callout-note title="Runs on your machine, not here"}
The reference numbers above came from the **real `coloc` package** (`coloc::coloc.abf`
5.2.3), which cannot run in the browser (WebR has no `coloc`). On your own machine the same
check is what `equivalence/check.R` re-verifies against `equivalence/fixtures/`. The call
looks like this — shown for reading, not running here:

```r
# On your machine (not in the browser): coloc's own routine, GPL-3, shown for reference only
library(coloc)
d1 <- list(beta = beta1, varbeta = se1^2, snp = paste0("rs", 1:10),
           type = "cc", N = 5000, s = 0.5, MAF = maf)      # trait 1 (e.g. GWAS)
d2 <- list(beta = beta2, varbeta = se2^2, snp = paste0("rs", 1:10),
           type = "cc", N = 5000, s = 0.5, MAF = maf)      # trait 2 (e.g. eQTL)
coloc.abf(d1, d2, p1 = 1e-4, p2 = 1e-4, p12 = 1e-5)$summary   # PP.H0.abf .. PP.H4.abf
```
:::

**Validation tier: FULL.** The match is against `coloc::coloc.abf` 5.2.3's actual output on
these inputs (not a reported number or a closed form re-derived by us), and
`equivalence/check.R` re-verifies it in CI. Your `coloc_abf()` *is* coloc's colocalisation core.

## Common traps

**Trap: "the distinct evidence is just the product of the two totals."** No — that product,
$(\sum\text{BF}_1)(\sum\text{BF}_2) = 56$, is the *whole grid*, which includes the shared
(diagonal) configurations. The distinct evidence must **subtract the diagonal**: $56 - 24 =
32$. Forgetting the subtraction double-counts the shared cases inside $H_3$ and inflates PP3.
That's precisely why $\ell_{H_3}$ uses `logdiff(s1 + s2, s12)`, not just $s_1 + s_2$.

**Trap: "you can compute PP4 by exponentiating $\ell_{H_4}$ directly."** With real BFs,
$s_{12}$ can be $+50$ or $-50$; $e^{50}$ overflows and $e^{-50}$ underflows. You must
normalise on the log scale first — subtract $\text{logsum}$ of all five, *then* exponentiate.
`.logsum` subtracts the max before exponentiating for exactly this reason. Skipping it gives
`NaN` or `Inf` on realistic data even though the toy numbers "work."

## Check yourself

**1.** For $\text{BF}_1 = (2, 3)$ and $\text{BF}_2 = (1, 4)$, compute the distinct ($H_3$)
evidence with the trick.

```{webr}
#| autorun: false
# check1-starter — edit and Run
BF1 <- c(2, 3); BF2 <- c(1, 4)
```

::: {.callout-note collapse="true" title="Answer"}
Totals: $\sum\text{BF}_1 = 5$, $\sum\text{BF}_2 = 5$, grid $= 25$. Diagonal:
$2\cdot1 + 3\cdot4 = 2 + 12 = 14$. Distinct $= 25 - 14 = 11$. (Check by listing off-diagonal
pairs: $2\cdot4 + 3\cdot1 = 8 + 3 = 11$. ✓)
:::

**2.** In the shared self-check, PP4 came out $\approx 0.992$ and PP3 $\approx 0.000002$. What
would happen to PP3 if trait 2's loud SNP (SNP 4) had instead been a *different* SNP from
trait 1's?

::: {.callout-note collapse="true" title="Answer"}
PP3 would shoot up and PP4 would collapse. Aligned loud SNPs pump the diagonal ($s_{12}$),
which feeds $\ell_{H_4}$; misaligned loud SNPs leave the diagonal small but keep the grid
total large, so `logdiff(s1 + s2, s12)` — the distinct evidence — dominates and PP3 wins.
That's the distinct fixture (PP3 $\approx 1$) you'll meet in the capstone.
:::

**3.** In your own words: explain to a colleague, in two sentences, what the $\sum_{i\neq j}$
trick does and why it matters.

::: {.callout-note collapse="true" title="A model answer"}
It computes the "distinct causal variants" evidence — every pair of *different* SNPs, one per
trait — as the product of the two evidence totals minus the same-SNP (diagonal) sum, instead
of looping over all pairs. It matters because a real region has thousands of SNPs, so the
identity replaces a million-cell double sum with three cheap sums.
:::

## What you can now do

- **Write the five log supports** $\ell_{H_0}\dots\ell_{H_4}$ from $\ell_1$, $\ell_2$, and
  the priors, and normalise them to $\text{PP0}\dots\text{PP4}$.
- **Derive and apply the $\sum_{i\neq j}$ identity** (grid total minus diagonal), and its
  log-space form via `logdiff`.
- **Implement `logsum` and `logdiff` stably in base R** and explain why log-space normalising
  is necessary on real data.
- **Build `coloc_abf()` from scratch** and confirm it matches `coloc::coloc.abf` to $\sim
  10^{-6}$.

If any of these feels shaky, revisit the "multiplication table minus the diagonal" picture or
the derivation before continuing — the capstone assembles exactly this and runs it on both a
shared and a distinct region.

## Where this goes next

You've built the method. [The capstone](../04-capstone-reading-coloc-abf/lesson.qmd) returns
to the source: it decodes the paper's own $H_0$–$H_4$ equations and the `coloc::coloc.abf`
interface line by line, then runs your `coloc_abf()` on both a shared and a distinct region,
interprets PP4 vs PP3, adds the paper's sample-size caveat, and states the validation tier.
Ready for reps? Do the [practice problems](practice.qmd).
