---
title: "Decision trees"
subtitle: "Split the data into boxes — each box makes one prediction"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c("rpart")
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```


## Where we're going

Every method we have seen so far fits a **single** formula to the whole dataset — a
line, a curve, a discriminant score. A decision tree does something different: it
**splits** the data into regions and makes a separate prediction in each region. The
result is a flowchart of yes/no questions: "Is age < 30? Yes → go left. Is income >
50k? Yes → predict class A." That flowchart IS the model — you can draw it and read it
like a recipe. [@james2021introductionstatisticallearning, pp. 327--329]

The trade-off: trees are easy to interpret but they are **unstable** — small changes in
the data can produce a completely different tree. The next module solves that problem by
averaging many trees together.

**Builds on:**
[model fit, error & validation](../model-fit-error-and-validation/lesson.qmd) — RSS as
the measure of fit, and train/test splitting to avoid overfitting;
[classification evaluation](../classification-evaluation/lesson.qmd) — confusion
matrices and accuracy for grading classification trees


## Warm-up: do you remember?

Before the new stuff, a few quick reps — answers from memory first, then check.

**1.** You fit a model to a training set and get RSS = 20. On the test set, RSS = 80.
In one sentence, what is happening?

::: {.callout-note collapse="true" title="Answer"}
The model is **overfitting** — it fits the training data well (low RSS) but fails to
generalize (high test RSS). Taught in
[model fit, error & validation](../model-fit-error-and-validation/lesson.qmd); revisit
if shaky.
:::

**2.** A confusion matrix has 40 true positives and 10 false negatives. What is the
sensitivity?

::: {.callout-note collapse="true" title="Answer"}
Sensitivity = TP / (TP + FN) = 40 / 50 = 0.80. Taught in
[classification evaluation](../classification-evaluation/lesson.qmd); revisit if shaky.
:::


## A tiny example first

Twelve customers, each described by one predictor (account age in years) and one
outcome (dollars spent last month):

| Customer | Age (yrs) | Spent ($) |
|----------|-----------|-----------|
| 1        | 1         | 10        |
| 2        | 2         | 15        |
| 3        | 3         | 12        |
| 4        | 4         | 18        |
| 5        | 5         | 40        |
| 6        | 6         | 45        |
| 7        | 7         | 50        |
| 8        | 8         | 42        |
| 9        | 9         | 80        |
| 10       | 10        | 85        |
| 11       | 11        | 78        |
| 12       | 12        | 90        |

Suppose you want to predict spending from age. A linear regression would draw one
line through all 12 points. A tree asks: **what single cutpoint on age best splits
these customers into two groups with the smallest total RSS?**

Try splitting at age = 4.5 (customers 1--4 go left, 5--12 go right):

- Left group: (10, 15, 12, 18). Mean = 13.75. RSS = $(10-13.75)^2 + (15-13.75)^2 + (12-13.75)^2 + (18-13.75)^2 = 14.06 + 1.56 + 3.06 + 18.06 = 36.75$.
- Right group: (40, 45, 50, 42, 80, 85, 78, 90). Mean = 63.75. RSS = sum of squared deviations = 3185.5.
- Total RSS = 36.75 + 3185.5 = 3222.25.

Now try splitting at age = 8.5 (customers 1--8 go left, 9--12 go right):

- Left group: (10, 15, 12, 18, 40, 45, 50, 42). Mean = 29. RSS = 1954.
- Right group: (80, 85, 78, 90). Mean = 83.25. RSS = 86.75.

- Total RSS = 1954 + 86.75 = 2040.75.

The age = 8.5 split has lower total RSS. The tree algorithm tries **every** possible
cutpoint and picks the one with the smallest total RSS.

```{r toy-example}
# The toy data
age   <- 1:12
spent <- c(10, 15, 12, 18, 40, 45, 50, 42, 80, 85, 78, 90)

# Try every possible split and compute total RSS
try_split <- function(cutpoint) {
  left  <- spent[age <= cutpoint]
  right <- spent[age > cutpoint]
  rss_left  <- sum((left  - mean(left))^2)
  rss_right <- sum((right - mean(right))^2)
  rss_left + rss_right
}

# Candidate cutpoints: midpoints between consecutive ages
cuts <- 1:11 + 0.5
rss  <- sapply(cuts, try_split)
names(rss) <- cuts
round(rss, 1)

# Best first split
cat("Best split at age =", cuts[which.min(rss)],
    "with total RSS =", round(min(rss), 1), "\n")
```


## Getting a feel for it

**The mental picture: a tree carves the number line (or the predictor space) into
boxes, and each box gets its own flat prediction.** One split on age creates two
boxes; a second split on one of those boxes creates three; and so on. The prediction
inside each box is just the average of the training $y$-values that landed there.

**Why this construction?** Because some patterns are not lines. If spending jumps
sharply at certain age thresholds (new customers spend little, established ones spend
a lot, long-tenured ones spend even more), a tree discovers those thresholds
automatically. No one needs to specify the breakpoints in advance — the algorithm
finds the split that minimizes RSS at each step.

**Where the analogy breaks:** a tree can only make axis-aligned cuts (vertical or
horizontal lines in 2D). It cannot tilt a boundary diagonally the way LDA or logistic
regression can. For smooth, diagonal relationships, a line often beats a tree.

**What the numbers feel like.** In our toy example, the best first split separates
low spenders from high spenders. The RSS inside each group is much smaller than the
total RSS before the split — the tree "explains" the jump by finding where it happens.


## What the notation is going to say

The formula will say: partition the predictor space into regions $R_1, \ldots, R_J$,
and for each region predict the mean of the training $y$-values in that region. The
algorithm picks the regions by greedily choosing the split (which predictor, which
cutpoint) that reduces RSS the most at each step. In our toy, the first region might
be "age $\leq$ 4.5" (predicted spending ≈ 14) and the second "age > 4.5" (predicted
spending ≈ 64).


## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $x_i$, $y_i$ | "x-i", "y-i" | the predictor value and response for observation $i$ | $x_3 = 3$ (age), $y_3 = 12$ (spending) |
| $n_j$ | "n-j" | number of training observations in region $R_j$ | 4 customers in the left group after splitting at 4.5 |
| $R_j$ | "R-j" | the $j$-th region (leaf) of the tree | $R_1$: customers with age $\leq$ the cutpoint; $R_2$: the rest |
| $J$ | "J" | number of terminal nodes (leaves) | 2 after the first split |
| $\hat{y}_{R_j}$ | "y-hat R-j" | the predicted value in region $R_j$: the mean of training $y$-values there | mean spending for customers in that box |
| $\in$ | "in" | "is a member of" — $x_i \in R_j$ means observation $i$ falls in region $j$ | customer 3 $\in R_1$ (age 3 is in the left group) |
| $s$ | "s" | the cutpoint value for a split | 4.5 (the first cutpoint we tried) |
| $X_j$ | "X-j" | the predictor being split on (when there are multiple predictors) | age (our only predictor here) |
| $\lvert T \rvert$ | "size of T" | number of terminal nodes in tree $T$ | 2 for a single split, 3 for two splits, etc. |
| $\alpha$ | "alpha" | the pruning penalty on tree size — larger $\alpha$ favors smaller trees | (used in pruning below) |
| $G$ | "Gini index" | a measure of node impurity for classification: small when a node is pure (mostly one class) | (used for classification trees below) |


## The formal version

### The prediction rule

A decision tree partitions the predictor space into $J$ non-overlapping regions
$R_1, \ldots, R_J$, and for each observation falling in region $R_j$, predicts
[@james2021introductionstatisticallearning, pp. 330--331]:

$$
\hat{y} = \hat{y}_{R_j} = \frac{1}{n_j} \sum_{i:\, x_i \in R_j} y_i
$$

Reading it aloud: "y-hat equals the mean of the training y-values in region R-j" —
or in everyday words: "look up which box the new point falls in, and predict the
average of the training data in that box."

Mapping it onto the toy: if the tree splits at age = 4.5, then a customer with age 3
falls in $R_1$ and is predicted 13.75 (the mean of 10, 15, 12, 18).

### How splits are chosen (greedy, top-down)

At each step, the algorithm considers every predictor $X_j$ and every cutpoint $s$,
and picks the pair that minimizes the total RSS across the two resulting regions
[@james2021introductionstatisticallearning, pp. 330--331]:

$$
\sum_{i:\, x_i \in R_1(j,s)} (y_i - \hat{y}_{R_1})^2 \;+\; \sum_{i:\, x_i \in R_2(j,s)} (y_i - \hat{y}_{R_2})^2
$$

Reading it aloud: "the RSS in the left region plus the RSS in the right region" —
this is exactly the two-group computation we did by hand above.

The process is **greedy**: at each step it picks the locally best split without
looking ahead. This is fast (try all cutpoints, pick the winner) but means the
globally optimal tree is not guaranteed.

### For classification: Gini index

When the response is a class label instead of a number, RSS does not apply. Instead,
trees need a measure of how **mixed** (impure) a node is. Imagine a node with 3
class-A observations and 1 class-B. The proportion of A is 3/4 = 0.75, B is 1/4 =
0.25. How impure is that? Multiply each proportion by its complement and add:
$0.75 \times 0.25 + 0.25 \times 0.75 = 0.375$. That is the **Gini index** — a
score from 0 (pure) to 0.5 (maximally mixed for 2 classes).
[@james2021introductionstatisticallearning, pp. 335--336]

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $\hat{p}_{mk}$ | "p-hat m-k" | proportion of observations in node $m$ from class $k$ | if 3 of 4 customers in a node are "low spenders," $\hat{p}_{m,\text{low}} = 0.75$ |
| $K$ | "K" | total number of classes | 2 (low spenders, high spenders) |

The formal version:

$$
G = \sum_{k=1}^{K} \hat{p}_{mk}(1 - \hat{p}_{mk})
$$

Reading it aloud: "the Gini index equals the sum over all classes of p-hat-m-k times
one minus p-hat-m-k" — in everyday words: "how mixed is this node?" If a node is
pure (all one class), every $\hat{p}_{mk}$ is 0 or 1, so $G = 0$. If the node is
evenly split between two classes, $G = 2 \times 0.5 \times 0.5 = 0.5$ — maximum
impurity. The tree picks the split that reduces the weighted Gini index the most.

### Pruning: grow big, then cut back

A tree that keeps splitting until every leaf has one observation will overfit badly —
it memorizes the training data. The fix is **cost-complexity pruning**: grow the
largest tree $T_0$, then find the subtree $T$ that minimizes
[@james2021introductionstatisticallearning, pp. 331--333]:

$$
\sum_{m=1}^{|T|} \sum_{i:\, x_i \in R_m} (y_i - \hat{y}_{R_m})^2 + \alpha\,|T|
$$

Reading it aloud: "the total RSS of the tree plus alpha times the number of leaves"
— or in everyday words: "fit plus a penalty for complexity." When $\alpha = 0$, the
full tree wins (lowest RSS). As $\alpha$ grows, the penalty for extra leaves
increases, and the best tree shrinks. Select $\alpha$ by cross-validation — same
idea as choosing $\lambda$ in ridge or lasso.

PREDICT: how many leaves will the unpruned tree have for 12 data points? Write your
guess, then check the plot below.

```{r pruning-demo}
#| fig-alt: "Two regression trees for the toy data: the left tree is unpruned (many leaves, low training RSS), the right tree is pruned to 3 leaves."
library(rpart)
df <- data.frame(age = 1:12,
                 spent = c(10, 15, 12, 18, 40, 45, 50, 42, 80, 85, 78, 90))

# Grow a big tree (no pruning constraints)
big_tree <- rpart(spent ~ age, data = df, control = rpart.control(minsplit = 2, cp = 0))

# Prune to 3 leaves using a complexity parameter
pruned <- prune(big_tree, cp = 0.05)

par(mfrow = c(1, 2), mar = c(2, 2, 3, 1))
plot(big_tree, uniform = TRUE, main = "Unpruned"); text(big_tree, cex = 0.8)
plot(pruned, uniform = TRUE, main = "Pruned (3 leaves)"); text(pruned, cex = 0.8)
```


## Explore it

### Find the best split by hand

```{webr}
#| autorun: true
#| fig-alt: "Bar chart showing total RSS for each possible split point on age (1.5 through 11.5). The minimum RSS is marked, showing the best first split."
# The toy data
age   <- 1:12
spent <- c(10, 15, 12, 18, 40, 45, 50, 42, 80, 85, 78, 90)

# TRY IT: change the spending values and re-run — does the best split move?
# PREDICT: which cutpoint will have the lowest RSS? Write your guess first.

# Try every cutpoint
cuts <- 1:11 + 0.5
rss <- numeric(length(cuts))
for (i in seq_along(cuts)) {
  left  <- spent[age <= cuts[i]]
  right <- spent[age >  cuts[i]]
  rss[i] <- sum((left - mean(left))^2) + sum((right - mean(right))^2)
}

barplot(rss, names.arg = cuts, xlab = "Cutpoint", ylab = "Total RSS",
        main = "RSS for each possible split", col = ifelse(rss == min(rss), "coral", "steelblue"))
cat("Best split: age =", cuts[which.min(rss)], " RSS =", round(min(rss), 1), "\n")
```

### `rpart` agrees with our hand computation

::: {.callout-note collapse="true" title="Static R — runs at build time, not in the browser"}
The `rpart` package fits decision trees. This cell runs when the page is built; you
can run it on your own machine with `Rscript`.
:::

```{r rpart-toy}
library(rpart)
df <- data.frame(age = 1:12,
                 spent = c(10, 15, 12, 18, 40, 45, 50, 42, 80, 85, 78, 90))
fit <- rpart(spent ~ age, data = df, control = rpart.control(minsplit = 2, cp = 0.01))

# The first split rpart chose
cat("rpart's splits:\n")
print(fit)

# Compare: our hand-computed best split
cuts <- 1:11 + 0.5
rss <- sapply(cuts, function(s) {
  L <- df$spent[df$age <= s]; R <- df$spent[df$age > s]
  sum((L - mean(L))^2) + sum((R - mean(R))^2)
})
cat("\nHand search best split: age =", cuts[which.min(rss)], "\n")
cat("rpart first split agrees:", grepl(as.character(cuts[which.min(rss)]),
    paste(capture.output(print(fit)), collapse = " ")), "\n")
```

### Classification tree on iris

PREDICT: iris has 4 predictors — will the tree use all 4, or will a few suffice?

```{r iris-tree}
#| fig-alt: "A classification tree for the iris dataset, splitting on Petal.Length and Petal.Width to classify three species."
library(rpart)
fit_iris <- rpart(Species ~ ., data = iris, method = "class")
plot(fit_iris, uniform = TRUE, margin = 0.1); text(fit_iris, cex = 0.8)

preds <- predict(fit_iris, type = "class")
cat("Training accuracy:", mean(preds == iris$Species), "\n")
```


## Common traps

**Trap: "A split that barely reduces RSS is useless — just stop early."** Stopping
early is risky because a weak split now can enable a strong split later. For example,
in our toy data, suppose the first split at age = 8.5 looks mediocre. If we stop, we
miss the chance to split the left group at age = 4.5, which would cleanly separate
the low and mid spenders. The standard approach is: grow the full tree, then prune
back. Let the pruning penalty decide what to keep, not an early-stopping rule.
[@james2021introductionstatisticallearning, p. 331]

**Trap: "Trees are as accurate as linear models."** On our toy data, a single tree
with 3 leaves might predict 14, 44, and 83 for the three groups. A well-fit line
through the same data could have lower test error if the true relationship is smooth.
Trees shine when the true pattern has sharp jumps or interactions — but for smooth
trends, a line or generalized linear model usually wins. The real power of trees comes
from ensembles (next module).
[@james2021introductionstatisticallearning, pp. 338--340]


## Check yourself

**1.** In our toy data, the best first split is at age = 8.5 (from the R output). What
is the predicted spending for a customer with age = 3?

```{webr}
#| autorun: false
# check1-starter
age   <- 1:12
spent <- c(10, 15, 12, 18, 40, 45, 50, 42, 80, 85, 78, 90)
# What is the mean of spent for customers with age <= 8.5?
```

::: {.callout-note collapse="true" title="Answer"}
Age 3 falls in the left region (age $\leq$ 8.5). The training values there are 10,
15, 12, 18, 40, 45, 50, 42. Mean = 232/8 = 29. So the prediction is \$29.
:::

**2.** A node has 60 observations: 45 from class A and 15 from class B. Compute the
Gini index.

::: {.callout-note collapse="true" title="Answer"}
$\hat{p}_A = 45/60 = 0.75$, $\hat{p}_B = 15/60 = 0.25$.
$G = 0.75 \times 0.25 + 0.25 \times 0.75 = 0.375$.
(A pure node would have $G = 0$; this node is moderately impure.)
:::

**3.** In your own words: explain to a colleague, in two sentences, what a decision
tree does and why it needs pruning.

::: {.callout-note collapse="true" title="A model answer"}
A decision tree splits the data into boxes using yes/no questions about the
predictors, and predicts the average (or majority class) in each box. Without
pruning it keeps splitting until each box has one observation — memorizing the
training data — so you grow the full tree and then cut back branches that don't
improve test-set performance.
:::


## What you can now do

- Find the best binary split on a predictor by trying every cutpoint and picking the
  one with the lowest RSS (or Gini index for classification)
- Read a tree diagram: follow the branches to get a prediction for a new observation
- Explain why pruning (grow big, then cut back with a penalty) is safer than
  stopping early
- State the key weakness of single trees: instability (small data changes, very
  different tree)

If any of these feels shaky, revisit the relevant section above before continuing —
the next module builds on trees as the base learner for ensemble methods.


## Where this goes next

A single tree is interpretable but fragile — small changes in the data can produce a
completely different tree. The next module — **bagging, random forests & boosting** —
fixes this by growing **many** trees and averaging their predictions, trading
interpretability for dramatically better accuracy and stability.

Ready for more reps? Do the [practice problems](practice.qmd).
