---
title: "Practice: Decision trees"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
library(rpart)
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Do them on paper or in an R session; only then
open the answer.


## Problem 1 — confidence rep with new numbers

Eight observations, one predictor ($x$), one response ($y$):

| $x$ | 1  | 2  | 3  | 4  | 5  | 6  | 7  | 8  |
|-----|----|----|----|----|----|----|----|----|
| $y$ | 5  | 8  | 6  | 7  | 20 | 25 | 22 | 23 |

Try splitting at $x = 4.5$. Compute the mean $y$ and RSS in each group. What is the
total RSS?

```{webr}
#| autorun: false
# p1-starter
x <- 1:8
y <- c(5, 8, 6, 7, 20, 25, 22, 23)
```

::: {.callout-note collapse="true" title="Answer"}
Left ($x \leq 4.5$): $y$ = (5, 8, 6, 7). Mean = 6.5.
RSS = $(5-6.5)^2 + (8-6.5)^2 + (6-6.5)^2 + (7-6.5)^2 = 2.25 + 2.25 + 0.25 + 0.25 = 5$.
Right ($x > 4.5$): $y$ = (20, 25, 22, 23). Mean = 22.5.
RSS = $(20-22.5)^2 + (25-22.5)^2 + (22-22.5)^2 + (23-22.5)^2 = 6.25 + 6.25 + 0.25 + 0.25 = 13$.
Total RSS = 5 + 13 = 18.
:::


## Problem 2 — faded worked example

Below is a best-split search for the lesson's toy data (age 1--12, spending as
given). One step is blanked out.

- Step 1: list all candidate cutpoints: 1.5, 2.5, ..., 11.5 (midpoints between
  consecutive ages).
- Step 2: for cutpoint = 4.5, the left group is (10, 15, 12, 18) with mean 13.75,
  and the right group is (40, 45, 50, 42, 80, 85, 78, 90) with mean 63.75.
- Step 3: **your step here** — compute the RSS in the left group and the RSS in the
  right group for the cutpoint 4.5.
- Step 4: add left RSS + right RSS to get the total RSS for this cutpoint.
- Step 5: repeat for all cutpoints, pick the one with the smallest total.

::: {.callout-note collapse="true" title="Answer"}
Step 3: Left RSS = $(10-13.75)^2 + (15-13.75)^2 + (12-13.75)^2 + (18-13.75)^2$
$= 14.06 + 1.56 + 3.06 + 18.06 = 36.75$.
Right RSS = $(40-63.75)^2 + (45-63.75)^2 + (50-63.75)^2 + (42-63.75)^2 + (80-63.75)^2 + (85-63.75)^2 + (78-63.75)^2 + (90-63.75)^2$
$= 564.06 + 351.56 + 189.06 + 473.06 + 264.06 + 451.56 + 203.06 + 689.06 = 3185.5$.
Total = 36.75 + 3185.5 = 3222.25. (Not the best split — the R code in the lesson
finds a cutpoint with lower total RSS.)
:::


## Problem 3 — Gini index computation

A classification node has 80 observations: 60 from class A and 20 from class B.

(a) Compute the Gini index for this node.
(b) After splitting, the left child has 55 A and 5 B; the right child has 5 A and
15 B. Compute the Gini index of each child. Is the split an improvement?

```{webr}
#| autorun: false
# p3-starter
# PREDICT: will the split improve (lower) the weighted Gini?
```

::: {.callout-note collapse="true" title="Answer"}
(a) $\hat{p}_A = 60/80 = 0.75$, $\hat{p}_B = 0.25$.
$G_\text{parent} = 0.75 \times 0.25 + 0.25 \times 0.75 = 0.375$.

(b) Left (60 obs): $\hat{p}_A = 55/60 = 0.917$, $\hat{p}_B = 5/60 = 0.083$.
$G_\text{left} = 0.917 \times 0.083 + 0.083 \times 0.917 = 0.153$.
Right (20 obs): $\hat{p}_A = 5/20 = 0.25$, $\hat{p}_B = 15/20 = 0.75$.
$G_\text{right} = 0.25 \times 0.75 + 0.75 \times 0.25 = 0.375$.
Weighted Gini = $(60/80) \times 0.153 + (20/80) \times 0.375 = 0.115 + 0.094 = 0.209$.
Yes, the split improves impurity: parent Gini = 0.375, weighted child Gini = 0.209.
The left child is much purer (mostly A).
:::


## Problem 4 — spot the error

A colleague says: *"My unpruned tree has 0% training error — it must be the best
model, so I'll deploy it as-is."* This contains **exactly one mistake** — find it.

::: {.callout-note collapse="true" title="Answer"}
The mistake: equating 0% training error with "best model." An unpruned tree can
achieve 0% training error by creating one leaf per observation, but that tree is
**overfitting** — it memorizes the training data and will perform poorly on new data.
A pruned tree (fewer leaves, higher training error) almost always has lower **test**
error. This is the "grow big then cut back" lesson: training error alone does not
measure quality.
:::


## Problem 5 — interleaved *(uses model fit, error & validation's train/test idea)*

Using the lesson's toy data (age 1--12), suppose you hold out customers 11 and 12 as
a test set and build a tree on the remaining 10.

(a) Training on customers 1--10 with a split at age = 8.5: the right group (ages 9,
10; spending 80, 85) has mean 82.5. What is the test RSS for customers 11 and 12
(actual spending: 78 and 90)?
(b) Is this test RSS higher or lower than the training RSS for the same tree? What
does that tell you?

```{webr}
#| autorun: false
# p5-starter
# Test customers: age 11 (spent 78), age 12 (spent 90)
# Both fall in the right region (age > 8.5), predicted = 82.5 (from training data only)
test_actual <- c(78, 90)
test_pred   <- 82.5
```

::: {.callout-note collapse="true" title="Answer"}
(a) Both test customers have age > 8.5, so both are predicted 82.5 (the mean of the
training right group: ages 9--10).
Test RSS = $(78 - 82.5)^2 + (90 - 82.5)^2 = 20.25 + 56.25 = 76.5$.

(b) The test RSS (76.5 for 2 observations) is proportionally higher than the
training RSS per observation — expected, because the tree was fit to the training
data, not these new points. Moderate test error on unseen data is normal; the
concern would be if test RSS were dramatically worse than training, signaling
overfitting.
:::


## Problem 6 — transfer: tree instability

Remove customer 12 (age = 12, spent = 90) from the toy dataset and re-find the best
first split on the remaining 11 customers. Does the best cutpoint change? What does
this say about tree stability?

```{webr}
#| autorun: false
# p6-starter
# PREDICT: will removing one customer change the best split? Write your guess.
age   <- 1:11
spent <- c(10, 15, 12, 18, 40, 45, 50, 42, 80, 85, 78)
cuts  <- 1:10 + 0.5
```

::: {.callout-note collapse="true" title="Answer"}
```{r}
age <- 1:11; spent <- c(10, 15, 12, 18, 40, 45, 50, 42, 80, 85, 78)
rss <- sapply(1:10 + 0.5, function(s) {
  L <- spent[age <= s]; R <- spent[age > s]
  sum((L - mean(L))^2) + sum((R - mean(R))^2)
})
names(rss) <- 1:10 + 0.5; round(rss, 1)
cat("Best split:", names(which.min(rss)), "\n")
```

Removing one customer can change which cutpoint wins (or at least change the RSS
values substantially). This is tree **instability**: small data changes can produce
different trees. It is the main weakness of single decision trees and the motivation
for ensemble methods (bagging, random forests) in the next module.
:::


## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/decision-trees" pass="4"}

::: {.quiz-q qid="dt-split-criterion" qtype="mc" answer="c"}
A regression tree picks each split by choosing the cutpoint that:

a) Maximizes the distance between the two group means after splitting
b) Puts an equal number of observations in each resulting child node
c) Minimizes the total RSS (sum of squared deviations) in both resulting groups
:::

::: {.quiz-q qid="dt-gini-pure" qtype="mc" answer="a"}
A classification node contains 30 observations, all from class A. What is its Gini index?

a) 0 — the node is perfectly pure, so every term in the Gini sum is zero
b) 0.5 — because there is only one class present in this particular node
c) 1 — because the node has maximum certainty about its predicted class
:::

::: {.quiz-q qid="dt-pruning-why" qtype="mc" answer="b"}
Why is "grow big, then prune" preferred over "stop splitting when RSS gain is small"?

a) Pruning always produces a tree with exactly two leaves, which is simplest
b) A weak early split can enable a strong later split that stopping would miss
c) Pruning uses the training RSS directly, avoiding any need for cross-validation
:::

::: {.quiz-q qid="dt-instability" qtype="mc" answer="a"}
What is the main weakness of a single decision tree compared to linear regression?

a) Small changes in the data can produce a very different tree (high variance)
b) Trees cannot handle classification problems, only regression problems
c) Trees require the predictors to be normally distributed within each region
:::

::: {.quiz-q qid="dt-train-rss" qtype="mc" answer="c"}
(From model fit, error & validation.) An unpruned tree achieves 0% training error. This means:

a) The tree will also achieve 0% error on new data from the same population
b) The tree has found the true underlying relationship between predictors and response
c) The tree has likely memorized the training data and will overfit on new observations
:::

:::

::::
