---
title: "Rolling downhill: finding a minimum one step at a time"
subtitle: "Use the slope as a compass — step against it, and you walk to the bottom"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live cells below re-state their own data. Base R only.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```


## Where we're going

Loads of applied math boils down to one question: *what input makes this quantity as
small as possible?* The smallest prediction error, the lowest cost, the best fit. This
module builds the workhorse answer — **gradient descent** — in its simplest form: one
input, one curve, find the bottom. The recipe is almost embarrassingly simple: stand
somewhere on the curve, measure the slope, take a step *downhill*, repeat. The rest of
the course scales this exact move up to real problems.

**Builds on:** [the derivative as a slope you can measure](../../../../foundations/derivative-as-slope/lesson.qmd).

## Warm-up: do you remember?

Three quick recalls from the foundation this module stands on. Produce each answer
before opening it.

**1.** *(derivative-as-slope)* How do you *measure* the slope of a function $f$ at a
point $x$ numerically, in one line of R?

::: {.callout-note collapse="true" title="Answer"}
The central difference: `(f(x + h) - f(x - h)) / (2 * h)` for a small `h` (say
`1e-5`) — rise over the run, looking a small step each way. See
[derivative-as-slope](../../../../foundations/derivative-as-slope/lesson.qmd).
:::

**2.** *(derivative-as-slope)* At the very bottom of a U-shaped curve, what is the
slope? And just to the *left* of the bottom, is the slope positive or negative?

::: {.callout-note collapse="true" title="Answer"}
At the bottom the slope is **zero**. Just to the left it's **negative** (the curve is
still heading downhill as $x$ increases). This sign pattern is the compass this whole
module rides on.
:::

**3.** *(baseline)* If you're standing on a hillside and the ground slopes *up* to your
right (increasing $x$), which way do you step to go *down* — left or right?

::: {.callout-note collapse="true" title="Answer"}
Left — *opposite* the uphill direction. "Downhill is the opposite of the slope" is the
entire idea, in plain words.
:::

## A tiny example first

Here's a bowl-shaped curve: $f(x) = (x - 4)^2 + 1$. Its lowest point is at $x = 4$
(where the squared term is zero), and we want to *find* that bottom without knowing the
answer in advance — just by feeling our way downhill.

Start at $x = 1$. Measure the slope there. The slope of this curve at $x = 1$ is $-6$
(steeply downhill — you can measure it with last module's tool). Negative slope means
"downhill is to the *right*", so we should increase $x$. Here's the move: **step
against the slope**, by an amount scaled down by a small factor we'll call the
*learning rate*, $\alpha = 0.1$:

$$
x_{\text{new}} = x_{\text{old}} - \alpha \times (\text{slope}) = 1 - 0.1 \times (-6) = 1 + 0.6 = 1.6
$$

Subtracting a negative moved us *right*, toward the bottom — exactly what we wanted.
Now repeat from $x = 1.6$: the slope there is $2(1.6 - 4) = -4.8$, so
$x_{\text{new}} = 1.6 - 0.1 \times (-4.8) = 2.08$. Again: from $2.08$ the slope is
$-3.84$, giving $2.46$. The steps are $1 \to 1.6 \to 2.08 \to 2.46 \to \dots$ — each
one closer to $4$, and each one *smaller* than the last, because the curve flattens as
we near the bottom (smaller slope → smaller step). We coast to a stop right at the
minimum.

The same thing in R, so you can watch every step:

```{r toy-example}
# The bowl and a way to measure its slope (from the foundation)
f        <- function(x) (x - 4)^2 + 1
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)

x  <- 1        # start here
lr <- 0.1      # learning rate: how much of the slope to step by
for (step in 1:4) {
  slope <- slope_at(f, x)
  x     <- x - lr * slope    # step downhill (against the slope)
  cat("step", step, ": x =", round(x, 3), " (slope was", round(slope, 2), ")\n")
}
# 1.6, 2.08, 2.464, ... marching toward the bottom at x = 4
```

## Getting a feel for it

**A ball rolling into a valley.** Picture the curve as the cross-section of a valley
and set a ball down on the slope. Gravity pulls it downhill, and the *steeper* the
ground, the harder it accelerates. Gradient descent is that ball, but taking discrete
steps: at each spot it checks how steep the ground is and in which direction, then
hops downhill accordingly. Where the valley is steep (far from the bottom) it takes big
hops; where it flattens out (near the bottom) the hops shrink to nothing and it settles.
(Where the picture bends: a real ball has momentum and can roll *through* the bottom and
up the other side; plain gradient descent has no momentum — each step is decided fresh
from the slope right where it stands.)

**Why step *against* the slope, and why scale it down?** The slope points in the
*uphill* direction (positive slope = uphill to the right), and we want *down* — so we
move the opposite way, which is the minus sign in the update. And why multiply by a
small $\alpha$ instead of stepping by the whole slope? Because the slope only tells you
the direction and steepness *right where you're standing* — it doesn't know there's a
bottom ahead. Trust it a little (a small step), re-measure, trust it a little more. Take
the whole slope in one leap and you can overshoot the bottom entirely and end up higher
than you started.

**What the steps feel like.** With a sensible learning rate the moves start big and
automatically taper — $0.6$, then $0.48$, then $0.38$ in our example — because the slope
itself shrinks as the ground flattens near the bottom. You don't have to *tell* it to
slow down; the geometry does it for you. That self-tapering is why "just keep stepping"
actually converges instead of rattling around forever.

## What the notation is going to say

The formula below says: "to get your next position, take your current position and
subtract the learning rate times the slope at your current position." The minus is
"go downhill, opposite the slope"; the learning rate is "take a modest step, don't
lurch." Do it over and over and you're doing the $1 \to 1.6 \to 2.08 \to \dots$ walk we
just did by hand.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $f(x)$ | "eff of ex" | the curve we're minimizing (the thing we want small) | $(x-4)^2 + 1$ |
| $x_t$ | "ex sub tee" | our position after $t$ steps ($t$ is the step counter) | $x_0 = 1$, $x_1 = 1.6$, … |
| $f'(x_t)$ | "eff prime of ex sub tee" | the slope at our current position (measured numerically) | $f'(1) = -6$ |
| $\alpha$ | "alpha" | the *learning rate*: the fraction of the slope we step by | $0.1$ |
| $x_{t+1}$ | "ex sub tee plus one" | our position after the next step | $x_1 = 1.6$ |
| $\leftarrow$ | "gets updated to" | assignment — replace the old value with the new one | $x \leftarrow 1.6$ |

## The formal version

One step of gradient descent, in one dimension:

$$
x_{t+1} = x_t - \alpha \, f'(x_t)
$$

Reading it aloud: "ex sub tee-plus-one equals ex sub tee, minus alpha times eff-prime
of ex sub tee" — or in everyday words: "your next spot is your current spot, moved
against the slope by a fraction alpha of it."

Mapping it onto the toy example, first step ($t = 0$, starting at $x_0 = 1$):

- $f'(x_0)$ is the slope at $x = 1$ → $-6$
- $\alpha \, f'(x_0)$ is $0.1 \times (-6)$ → $-0.6$
- $x_1 = x_0 - (-0.6) = 1 + 0.6$ → $1.6$, matching our hand step exactly.

Run that update in a loop and you have the whole algorithm:

```{r formal-check}
f        <- function(x) (x - 4)^2 + 1
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)

gradient_descent_1d <- function(f, x, lr, n_steps) {
  for (t in 1:n_steps) x <- x - lr * slope_at(f, x)   # the update rule, repeated
  x
}
gradient_descent_1d(f, x = 1, lr = 0.1, n_steps = 100)   # ~4, the true minimum
```

## Explore it

```{webr}
#| autorun: true
#| fig-alt: "The bowl-shaped curve (x-4)^2+1 with the gradient-descent path drawn as points hopping down the curve toward the minimum at x=4."
# explore-1 — LIVE: watch gradient descent walk down the bowl. Each dot is one step.
# PREDICT: starting at x = 1 with lr = 0.1, will 15 steps be enough to basically reach
# the bottom at x = 4? Write your guess, then run.
f        <- function(x) (x - 4)^2 + 1
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)

x <- 1; lr <- 0.1; n <- 15          # <-- try changing lr and n
path <- numeric(n + 1); path[1] <- x
for (t in 1:n) { x <- x - lr * slope_at(f, x); path[t + 1] <- x }

curve(f, from = -1, to = 9, lwd = 2, ylab = "f(x)",
      main = sprintf("lr = %.2f : after %d steps, x = %.3f", lr, n, x))
points(path, f(path), pch = 16, col = "firebrick")
lines(path, f(path), col = "firebrick", lty = 3)
```

```{webr}
#| autorun: true
#| fig-alt: "Three gradient-descent paths on the same bowl for small, good, and too-large learning rates, showing crawling, converging, and overshooting."
# explore-2 — LIVE: the learning rate is everything. Run three rates side by side.
# PREDICT: which rate below never settles down? Guess, then run and read the paths.
f        <- function(x) (x - 4)^2 + 1
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
run <- function(lr, n = 20, x = 1) {
  p <- numeric(n + 1); p[1] <- x
  for (t in 1:n) { x <- x - lr * slope_at(f, x); p[t + 1] <- x }
  p
}

curve(f, from = -2, to = 10, lwd = 2, ylab = "f(x)", main = "same start, three learning rates")
for (lr in c(0.01, 0.1, 1.02)) {              # crawl / good / overshoots & grows
  p <- run(lr); lines(p, f(p), lwd = 2, lty = 2)
  points(p, f(p), pch = 16)
}
legend("top", legend = c("0.01 crawls", "0.1 converges", "1.02 diverges"), bty = "n")
```

That second plot is the whole drama of the learning rate: too small and you barely
move; just right and you glide to the bottom; too big and each step overshoots *farther*
than the last, climbing the walls instead of settling.

## Common traps

**Trap: "a bigger learning rate always gets there faster."** Tempting — bigger steps,
fewer of them, right? But past a threshold the step overshoots the bottom and lands
*higher* up the opposite wall; the next step overshoots even more, and the value
explodes. Try it: from $x = 1$ on our bowl, $\alpha = 1.02$ sends $x$ to
$1 - 1.02(-6) = 7.12$, then the slope there ($+6.24$) throws it to
$7.12 - 1.02(6.24) = 0.76$, then further out again — diverging, not converging.
`gradient_descent_1d(f, 1, lr = 1.02, n_steps = 20)` runs away to a huge number. The
right rate is a Goldilocks choice, not "as big as possible."

**Trap: "it doesn't matter whether you add or subtract the slope."** It absolutely
does. The update is $x \leftarrow x - \alpha f'(x)$: the minus means "step *against*
the slope," i.e. downhill. Flip it to a plus and you step *with* the slope — uphill —
and walk away from the minimum forever (gradient *ascent*). A sign slip in code is
the fastest way to turn a minimizer into a maximizer.

**Trap: "gradient descent always finds the lowest point of the whole curve."** On a
single bowl, yes. But on a curve with several dips, the ball settles into whichever
valley it happened to start above — a *local* minimum, not necessarily the deepest one.
Our $(x-4)^2 + 1$ has just one valley so it's safe here; real problems can have many,
which is why where you *start* can matter. *(Optional: this is why people run descent
from several starting points — worth knowing, not needed to follow this course.)*

## Check yourself

**1.** Minimize $f(x) = (x - 2)^2$ (bottom at $x = 2$) with gradient descent starting
from $x = 5$, learning rate $0.1$, for 50 steps. Where does it end up?

```{webr}
#| autorun: false
# check1-starter — edit and Run to solve, right here
f        <- function(x) (x - 2)^2
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
```

::: {.callout-note collapse="true" title="Answer"}
It converges to about **2**, the bottom of the bowl.

```{r check1-answer}
f        <- function(x) (x - 2)^2
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
x <- 5; lr <- 0.1
for (t in 1:50) x <- x - lr * slope_at(f, x)
x   # ~2
```
Starting to the right of the bottom, the slope is positive (uphill to the right), so
subtracting it moves $x$ *left*, back toward 2. The compass works from either side.
:::

**2.** In the update $x_{t+1} = x_t - \alpha f'(x_t)$, why is there a *minus* sign and
not a plus?

::: {.callout-note collapse="true" title="Answer"}
Because the slope $f'(x_t)$ points *uphill*, and we want to go *downhill* — the opposite
direction. Subtracting the slope moves us against it, toward smaller $f$. (A plus sign
gives *gradient ascent*, which climbs to a maximum instead.)
:::

**3.** In your own words: explain to a colleague, in two sentences, what gradient
descent does and why the learning rate has to be neither too big nor too small.

::: {.callout-note collapse="true" title="A model answer"}
Gradient descent finds the bottom of a curve by repeatedly measuring the slope where it
stands and stepping in the downhill (opposite) direction. The learning rate sets the
step size: too small and it barely moves and takes forever; too big and it overshoots
the bottom and can spiral outward instead of settling.
:::

## What you can now do

- **Run gradient descent in one dimension**: from a starting point, repeatedly measure
  the slope and step against it by a learning rate, in R with a short loop.
- **Explain the update rule** $x \leftarrow x - \alpha f'(x)$ — why it's a minus, and
  what $\alpha$ controls.
- **Diagnose the learning rate**: recognize crawling (too small), converging (good), and
  diverging/oscillating (too big) from the path.
- **State the one caveat**: descent settles into whatever valley it starts above.

If any of these feels shaky, revisit [Getting a feel for it](#getting-a-feel-for-it)
and [Explore it](#explore-it) — the next module puts a *second* knob on the problem and
everything here has to still hold.

## Where this goes next

Real problems have more than one number to tune. Next we minimize a curve that lives
over *two* inputs at once, where "the slope" becomes a little bundle of slopes called the
**gradient**: [the gradient — downhill in more than one direction](../02-the-gradient/lesson.qmd).
Ready for more reps first? Do the [practice problems](practice.qmd).
