---
title: "Practice: Rolling downhill"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

Work these in order — they get harder. Use paper or the live cells; only then open the
answer. A slope-measuring helper you can paste into any starter:
`slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)`.

## Problem 1 — the toy example, new numbers

Minimize the bowl $f(x) = (x - 6)^2 + 2$ (bottom at $x = 6$). Start at $x = 2$, learning
rate $0.1$. Do the **first two steps by hand** (the slope of this curve at $x$ is
$2(x-6)$), then run it in R for 100 steps and confirm where it lands.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
f        <- function(x) (x - 6)^2 + 2
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
x <- 2; lr <- 0.1
```

::: {.callout-note collapse="true" title="Answer"}
By hand: at $x = 2$, slope $= 2(2 - 6) = -8$, so $x \leftarrow 2 - 0.1(-8) = 2.8$. At
$x = 2.8$, slope $= 2(2.8 - 6) = -6.4$, so $x \leftarrow 2.8 - 0.1(-6.4) = 3.44$. Both
steps move right, toward 6, and shrink as the slope eases.

```{r p1-answer}
f        <- function(x) (x - 6)^2 + 2
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
x <- 2; lr <- 0.1
for (t in 1:100) x <- x - lr * slope_at(f, x)
x   # ~6
```
:::

## Problem 2 — fill in the missing step

A gradient-descent loop for $f(x) = (x - 1)^2$ starting at $x = 4$, $lr = 0.2$, is
written below with its **core line blanked**. Supply just that line.

> ```
> x <- 4
> for (t in 1:50) {
>   slope <- slope_at(f, x)
>   your step here          # update x using slope, lr, and the descent rule
> }
> ```

```{webr}
#| autorun: false
# p2-starter — fill in the update line, then Run
f        <- function(x) (x - 1)^2
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
x <- 4; lr <- 0.2
```

::: {.callout-note collapse="true" title="Answer"}
The blanked line is the update rule — step *against* the slope by `lr`:
`x <- x - lr * slope`. The move is: multiply the slope by the learning rate and
*subtract* it (downhill, not up).

```{r p2-answer}
f        <- function(x) (x - 1)^2
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
x <- 4; lr <- 0.2
for (t in 1:50) { slope <- slope_at(f, x); x <- x - lr * slope }
x   # ~1
```
:::

## Problem 3 — start on the other side

Minimize $f(x) = (x - 3)^2$ starting from $x = 10$ (to the *right* of the bottom this
time), $lr = 0.1$, 100 steps. Where does it end up, and — in one sentence — why does it
move *left* now when in the lesson it moved right?

```{webr}
#| autorun: false
# p3-starter — your job: run descent from x = 10 and explain the direction
f        <- function(x) (x - 3)^2
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
x <- 10; lr <- 0.1
```

::: {.callout-note collapse="true" title="Answer"}
```{r p3-answer}
f        <- function(x) (x - 3)^2
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
x <- 10; lr <- 0.1
for (t in 1:100) x <- x - lr * slope_at(f, x)
x   # ~3
```
It lands at about **3**. To the right of the bottom the slope is *positive* (uphill to
the right), so subtracting it moves $x$ *left*, back toward the bottom. The minus sign
always points you downhill, whichever side you start on.
:::

## Problem 4 — spot the error

A colleague wants to minimize $f(x) = (x-4)^2$ but their loop drifts *away* from the
bottom. Their code has **exactly one mistake** — find it and fix it.

```{webr}
#| autorun: false
# p4-starter — run it, watch x run off, then fix the one mistake
f        <- function(x) (x - 4)^2
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
x <- 1; lr <- 0.1
for (t in 1:20) x <- x + lr * slope_at(f, x)   # x heads the wrong way
x
```

::: {.callout-note collapse="true" title="Answer"}
The mistake is the **sign**: they wrote `x + lr * slope`, which steps *up* the slope
(gradient *ascent*) — so from $x = 1$ it runs off toward $-\infty$, away from the
minimum. Descent subtracts:

```{r p4-answer}
f        <- function(x) (x - 4)^2
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
x <- 1; lr <- 0.1
for (t in 1:20) x <- x - lr * slope_at(f, x)   # minus: downhill
x   # ~4
```
:::

## Problem 5 — did it really reach the bottom? *(uses derivative-as-slope)*

Run descent on $f(x) = (x - 5)^2 + 3$ from $x = 0$, $lr = 0.1$, 200 steps. Then use the
foundation's slope-measuring tool to check the slope *at the point where descent
stopped*. What should that slope be if descent truly found the bottom?

```{webr}
#| autorun: false
# p5-starter — descend, then measure the slope where you landed
f        <- function(x) (x - 5)^2 + 3
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
x <- 0; lr <- 0.1
```

::: {.callout-note collapse="true" title="Answer"}
```{r p5-answer}
f        <- function(x) (x - 5)^2 + 3
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
x <- 0; lr <- 0.1
for (t in 1:200) x <- x - lr * slope_at(f, x)
c(landed_at = x, slope_there = slope_at(f, x))   # ~5, and slope ~0
```
Descent lands at about $x = 5$, and the measured slope there is essentially **zero** —
the signature of a minimum from the [derivative module](../../../../foundations/derivative-as-slope/lesson.qmd).
Measuring the slope where you stop is a cheap, honest way to confirm you've actually
arrived at the bottom rather than merely run out of steps.
:::

## Problem 6 — a gentle cliffhanger

Suppose the thing you want to minimize depends on **two** numbers, not one — say a cost
$C(a, b)$ that changes as you tune both a dial $a$ and a dial $b$. You can still imagine
a "downhill", but now there are two directions to worry about at once. How would you
extend the one-dimensional rule "step against the slope" to this two-dial situation?

::: {.callout-note collapse="true" title="Answer"}
You can't fully solve this yet — but here's the honest sketch, and it's exactly the next
lesson. You measure a slope *for each dial separately* (how does the cost change if I
nudge only $a$? only $b$?), bundle those two slopes into a little vector called the
**gradient**, and step against *both* at once: $a \leftarrow a - \alpha(\text{slope in }
a)$ and $b \leftarrow b - \alpha(\text{slope in } b)$. Same downhill idea, one step per
direction. [The next module](../02-the-gradient/lesson.qmd) makes this precise.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="courses/gradient-descent/modules/01-rolling-downhill" pass="4"}

::: {.quiz-q qid="rd-update-step" qtype="numeric" answer="5.6" tolerance="0.01"}
**On the lesson's bowl $f(x) = (x-4)^2 + 1$, you are standing at $x = 6$, where the
slope is $4$. With learning rate $\alpha = 0.1$, where does one update step put you?**
(Enter one decimal.)
:::

::: {.quiz-q qid="rd-minus-sign" qtype="mc" answer="a"}
**In the update $x_{t+1} = x_t - \alpha f'(x_t)$, why is the slope *subtracted*?**

a) The slope points uphill, so moving opposite to it goes downhill
b) Slopes near a minimum are negative, and subtracting a negative moves you forward
c) Subtraction is what makes each step smaller than the last
:::

::: {.quiz-q qid="rd-lr-diagnosis" qtype="mc" answer="b"}
**Starting from $x = 1$ on the lesson's bowl, your iterates are $1$, then $7.12$, then
$0.75$, each landing farther from $x = 4$ than the last. What is going on?**

a) The learning rate is too small, so the walk barely crawls forward
b) The learning rate is too big, so each step overshoots the bottom
c) The walk has slipped into a different valley of the curve
d) This is normal — the steps grow for a while before they shrink
:::

::: {.quiz-q qid="rd-step-taper" qtype="mc" answer="c"}
**With a fixed $\alpha = 0.1$, the steps in the lesson shrank on their own: $0.6$, then
$0.48$, then $0.38$. What makes them shrink?**

a) The algorithm senses the minimum approaching and gently brakes
b) The learning rate quietly decays a little on every iteration
c) The slope flattens near the bottom, so $\alpha \times$ slope shrinks too
:::

::: {.quiz-q qid="rd-slope-at-bottom" qtype="mc" answer="b"}
**(From the foundation *derivative-as-slope*.) At the exact bottom of a smooth U-shaped
curve, and just to the *left* of that bottom, the slopes are:**

a) zero at the bottom; positive just left of it
b) zero at the bottom; negative just left of it
c) negative at the bottom; zero just left of it
:::

:::

::::
