---
title: "Practice: Marginal likelihood and model evidence"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

Work these in order — they get harder. Do them on paper or in an R session; only then
open the answer.

## Problem 1 — the toy example, new numbers

Same coin data (heads, heads, tails), so $L(p) = p^2(1-p)$. This time your prior puts
**equal weight $1/3$** on $p = 0.3$, $p = 0.5$, and $p = 0.7$. Compute the marginal
likelihood of the data by hand, then check in R.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
lik   <- function(p) p^2 * (1 - p)
p     <- c(0.3, 0.5, 0.7)
prior <- c(1/3, 1/3, 1/3)
```

::: {.callout-note collapse="true" title="Answer"}
Likelihood at each value: $L(0.3) = 0.3^2 \times 0.7 = 0.09 \times 0.7 = 0.063$;
$L(0.5) = 0.5^2 \times 0.5 = 0.125$; $L(0.7) = 0.7^2 \times 0.3 = 0.49 \times 0.3 =
0.147$. Weighted average (equal $1/3$ weights): $\tfrac13(0.063 + 0.125 + 0.147) =
\tfrac13(0.335) = 0.1117$.

```{r p1-answer}
lik   <- function(p) p^2 * (1 - p)
p     <- c(0.3, 0.5, 0.7)
prior <- c(1/3, 1/3, 1/3)
sum(prior * lik(p))          # 0.11167 -- the marginal likelihood
```
:::

## Problem 2 — fill in the missing step

Same data and $L(p) = p^2(1-p)$, but now an **unequal** prior: weight $0.2$ on $p = 0.4$,
$0.5$ on $p = 0.6$, and $0.3$ on $p = 0.8$. Here is the worked solution with the core
step blanked out — supply only that step.

- Step 1 (shown): likelihoods are $L(0.4) = 0.096$, $L(0.6) = 0.144$, $L(0.8) = 0.128$.
- Step 2 (**your step here**): combine these into the marginal likelihood using the
  prior weights. Write the exact arithmetic.
- Step 3 (shown): the marginal likelihood is $0.1296$.

```{webr}
#| autorun: false
# p2-starter — edit and Run to check the blank
lik   <- function(p) p^2 * (1 - p)
p     <- c(0.4, 0.6, 0.8)
prior <- c(0.2, 0.5, 0.3)
```

::: {.callout-note collapse="true" title="Answer"}
The missing move is the **prior-weighted sum**: multiply each likelihood by its prior
weight and add.

$$
0.2(0.096) + 0.5(0.144) + 0.3(0.128) = 0.0192 + 0.072 + 0.0384 = 0.1296.
$$

That's the move that turns a curve of likelihoods into one number for the model — each
value counts in proportion to how much you believed it beforehand.

```{r p2-answer}
lik   <- function(p) p^2 * (1 - p)
p     <- c(0.4, 0.6, 0.8)
prior <- c(0.2, 0.5, 0.3)
sum(prior * lik(p))          # 0.1296
```
:::

## Problem 3 — the normal closed form

An estimate $\hat\theta = 0.10$ has variance $V = 0.04$. You put a normal prior $N(0, W)$
on the true value with $W = 0.05$. (a) What is the variance of the marginal
(prior-predictive) distribution of $\hat\theta$? (b) Using the closed form, what is the
marginal density at $\hat\theta = 0.10$? Confirm it matches a brute-force integral in R.

```{webr}
#| autorun: false
# p3-starter — your job: compute the marginal variance, then the closed-form density
theta_hat <- 0.10
V <- 0.04
W <- 0.05
```

::: {.callout-note collapse="true" title="Answer"}
(a) Variances add: the marginal variance is $V + W = 0.04 + 0.05 = 0.09$ (so the marginal
is $N(0, 0.09)$, standard deviation $\sqrt{0.09} = 0.3$).

(b) The marginal density is the height of that $N(0, 0.09)$ curve at $0.10$:
`dnorm(0.10, 0, sqrt(0.09))` $\approx 1.258$.

```{r p3-answer}
theta_hat <- 0.10; V <- 0.04; W <- 0.05

closed <- dnorm(theta_hat, 0, sqrt(V + W))         # closed form: N(0, V+W)

integrand <- function(theta) dnorm(theta_hat, theta, sqrt(V)) * dnorm(theta, 0, sqrt(W))
brute <- integrate(integrand, -Inf, Inf)$value      # brute-force average over the prior

c(closed_form = closed, brute_force = brute)         # equal -- variances add
```
:::

## Problem 4 — spot the error

A colleague computes the marginal spread when combining an estimate with variance
$V = 0.0025$ and a prior with variance $W = 0.04$. They write — it contains **exactly one
mistake** — find it and fix it.

> "The estimate's standard deviation is $\sqrt{0.0025} = 0.05$ and the prior's is
> $\sqrt{0.04} = 0.2$. So the marginal standard deviation is $0.05 + 0.2 = 0.25$, and the
> marginal is $N(0, 0.25^2)$."

::: {.callout-note collapse="true" title="Answer"}
The mistake is **adding the standard deviations**. It is the *variances* that add, not the
standard deviations. The correct marginal variance is $V + W = 0.0025 + 0.04 = 0.0425$, so
the marginal standard deviation is $\sqrt{0.0425} \approx 0.206$ — not $0.25$. The
colleague's $0.25$ overstates the spread (it corresponds to a variance of $0.0625$, which
is bigger than the true $0.0425$).

```{r p4-answer}
V <- 0.0025; W <- 0.04
c(correct_var = V + W, correct_sd = sqrt(V + W),   # 0.0425, 0.206
  wrong_sd = sqrt(V) + sqrt(W))                     # 0.25 -- the error
```
:::

## Problem 5 — mixing it with priors & posteriors *(uses priors-posteriors-and-updating's normalization)*

For the coin data ($L(p) = p^2(1-p)$) with equal prior $1/3$ on $p \in \{0.4, 0.6, 0.8\}$,
the marginal likelihood is $0.1227$ (from the lesson). Now go one step further and compute
the **posterior** weight on each $p$ — that is, prior $\times$ likelihood, then normalized
so the three posterior weights sum to $1$. Notice the marginal likelihood is exactly the
normalizing denominator.

```{webr}
#| autorun: false
# p5-starter — edit and Run to solve, right here
lik   <- function(p) p^2 * (1 - p)
p     <- c(0.4, 0.6, 0.8)
prior <- c(1/3, 1/3, 1/3)
```

::: {.callout-note collapse="true" title="Answer"}
Posterior $\propto$ prior $\times$ likelihood (from
[priors, posteriors & updating](../priors-posteriors-and-updating/lesson.qmd)). The
unnormalized weights are $\tfrac13(0.096), \tfrac13(0.144), \tfrac13(0.128) = 0.032,
0.048, 0.0427$; their sum is $0.1227$ — the marginal likelihood. Dividing each by that
sum gives the posterior weights $0.261, 0.391, 0.348$.

```{r p5-answer}
lik   <- function(p) p^2 * (1 - p)
p     <- c(0.4, 0.6, 0.8); prior <- c(1/3, 1/3, 1/3)

unnorm    <- prior * lik(p)              # prior x likelihood
marginal  <- sum(unnorm)                 # 0.1227 -- the normalizing denominator
posterior <- unnorm / marginal           # normalized to sum to 1
round(posterior, 4)                       # 0.2609, 0.3913, 0.3478
```
The marginal likelihood is not just a model score — it is precisely the denominator that
turns "prior $\times$ likelihood" into a proper posterior distribution.
:::

## Problem 6 — transfer

You are comparing **two models** for the same coin data (heads, heads, tails). Model A
believes only $p = 0.6$ (all prior weight there). Model B spreads its belief equally over
$p \in \{0.2, 0.6, 0.9\}$. Compute each model's marginal likelihood, and say which model
the data favors — and give a one-sentence intuition for why.

```{webr}
#| autorun: false
# p6-starter — edit and Run to solve, right here
lik <- function(p) p^2 * (1 - p)
```

::: {.callout-note collapse="true" title="Answer"}
Model A's marginal is just $L(0.6) = 0.144$ (all weight on one value). Model B's is
$\tfrac13\big(L(0.2) + L(0.6) + L(0.9)\big) = \tfrac13(0.032 + 0.144 + 0.081) =
\tfrac13(0.257) = 0.0857$. **Model A wins** ($0.144 > 0.0857$).

```{r p6-answer}
lik <- function(p) p^2 * (1 - p)
A <- lik(0.6)                                  # all weight on p = 0.6
B <- mean(lik(c(0.2, 0.6, 0.9)))               # equal weight on three values
c(model_A = A, model_B = round(B, 4))          # 0.144 vs 0.0857
```
Intuition: Model A bet its whole prior on a value that fits the data well, so its average
is that good fit. Model B diluted its belief with values that fit poorly ($p = 0.2$ makes
two heads unlikely), so averaging drags its evidence down — the honesty penalty for a
vaguer model.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/marginal-likelihood-and-model-evidence" pass="4"}

::: {.quiz-q qid="mlme-what-marginal" qtype="mc" answer="c"}
What does the marginal likelihood (model evidence) $P(D)$ measure?

a) The single parameter value that makes the data most probable
b) The probability that the parameter equals its best value
c) How probable the data is, averaged over the parameter values the prior weights
d) The largest likelihood any parameter value can achieve
:::

::: {.quiz-q qid="mlme-discrete-numeric" qtype="numeric" answer="0.1345" tolerance="0.001"}
For coin data with $L(p) = p^2(1-p)$ and a prior putting equal weight $0.5$ on $p = 0.5$
and $p = 0.6$, compute the marginal likelihood. Round to four decimals.
:::

::: {.quiz-q qid="mlme-variances-add" qtype="numeric" answer="0.09" tolerance="0.0001"}
An estimate has variance $V = 0.02$; the prior on the true value is $N(0, W)$ with
$W = 0.07$. What is the variance of the marginal (prior-predictive) distribution of the
estimate?
:::

::: {.quiz-q qid="mlme-why-average" qtype="mc" answer="b"}
Why average the likelihood over the prior instead of just taking the maximum likelihood?

a) Averaging always gives a larger number, which favors the true model
b) Maximizing always flatters a more complex model, so averaging is the fairer judge
c) The maximum likelihood cannot be computed for continuous parameters
:::

::: {.quiz-q qid="mlme-normal-closedform" qtype="mc" answer="a"}
*(Interleaved — uses distributions & densities.)* An estimate is normal with variance $V$
around the true value, and the prior on the true value is $N(0, W)$. What is the marginal
(prior-predictive) distribution of the estimate?

a) Normal, $N(0, V + W)$ — the variances add
b) Normal, $N(0, V \times W)$ — the variances multiply
c) Normal, $N(0, \sqrt{V} + \sqrt{W})$ — the standard deviations add
:::

:::

::::
