---
title: "Practice: Capstone — the whole descent, mixed"
subtitle: "No hints about which idea each problem needs — recalling that is the test."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

These draw on every module, in mixed order — figuring out *which* idea each one needs is
part of the test, so there are no "uses Module N" labels. Two helpers to paste as needed:

```r
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) {
    up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
  })
```

## Problem 1

Measure the slope of $f(x) = x^3 - 2x$ at $x = 2$, numerically. Is the function heading
uphill or downhill there?

```{webr}
#| autorun: false
# p1-starter
f <- function(x) x^3 - 2 * x
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
```

::: {.callout-note collapse="true" title="Answer"}
```{r p1-answer}
f <- function(x) x^3 - 2 * x
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
slope_at(f, 2)   # 10
```
The slope is about $+10$ — positive, so the curve is heading **uphill** at $x = 2$. (The
exact slope is $3x^2 - 2 = 10$, but you found it by measuring.)
:::

## Problem 2

Minimize $f(x) = (x + 3)^2 + 2$ (a bowl with its bottom at $x = -3$) with gradient
descent from $x = 4$, learning rate $0.1$, 100 steps. Where does it land?

```{webr}
#| autorun: false
# p2-starter
f <- function(x) (x + 3)^2 + 2
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
```

::: {.callout-note collapse="true" title="Answer"}
```{r p2-answer}
f <- function(x) (x + 3)^2 + 2
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
x <- 4; for (t in 1:100) x <- x - 0.1 * slope_at(f, x)
x   # ~ -3
```
It lands at about $x = -3$. Starting to the right of the bottom, the slope is positive,
so subtracting it walks $x$ left toward $-3$.
:::

## Problem 3

For $g(a, b) = (a - 5)^2 + (b + 1)^2$, compute the gradient at $(0, 0)$ and take one
descent step with $\alpha = 0.1$. Which direction does each knob move, and why?

```{webr}
#| autorun: false
# p3-starter
g <- function(v) (v[1] - 5)^2 + (v[2] + 1)^2
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) {
    up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
  })
```

::: {.callout-note collapse="true" title="Answer"}
```{r p3-answer}
g <- function(v) (v[1] - 5)^2 + (v[2] + 1)^2
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) {
    up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
  })
gr <- gradient(g, c(0, 0)); gr        # (-10, 2)
c(0, 0) - 0.1 * gr                    # (1, -0.2)
```
Gradient $(-10, +2)$. Subtracting $0.1$ times it moves $a$ **up** (toward its target 5,
since the $a$-slope is negative → downhill is toward larger $a$) and $b$ **down** (toward
$-1$). Each knob heads toward its own target.
:::

## Problem 4

Fit a line to this noisy data by descent, then report the coefficients **and** the final
MSE. Should the MSE be zero? Explain in a sentence.

```{webr}
#| autorun: false
# p4-starter
x <- c(0, 1, 2, 3, 4); y <- c(1.1, 2.8, 5.2, 6.9, 9.1)
mse <- function(b) mean((b[1] + b[2] * x - y)^2)
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) {
    up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
  })
```

::: {.callout-note collapse="true" title="Answer"}
```{r p4-answer}
x <- c(0, 1, 2, 3, 4); y <- c(1.1, 2.8, 5.2, 6.9, 9.1)
mse <- function(b) mean((b[1] + b[2] * x - y)^2)
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) {
    up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
  })
b <- c(0, 0); for (t in 1:5000) b <- b - 0.02 * gradient(mse, b)
c(intercept = b[1], slope = b[2], mse = mse(b))
```
Coefficients near intercept $\approx 1$, slope $\approx 2$. The MSE is **not** zero: the
points carry scatter, so no straight line hits them all — the best line just minimizes the
leftover squared error.
:::

## Problem 5 — spot the error

This code is meant to fit a line, but the gradient is computed wrongly — **exactly one
mistake**. Find it and fix it.

```{webr}
#| autorun: false
# p5-starter — the gradient nudges both knobs together instead of one at a time. Fix it.
x <- c(1, 2, 3, 4); y <- c(3, 5, 7, 9)     # line y = 1 + 2x
mse <- function(b) mean((b[1] + b[2] * x - y)^2)
bad_gradient <- function(f, v, h = 1e-5) {
  (f(v + h) - f(v - h)) / (2 * h)          # nudges the whole vector at once
}
b <- c(0, 0); for (t in 1:2000) b <- b - 0.05 * bad_gradient(mse, b)
b
```

::: {.callout-note collapse="true" title="Answer"}
The gradient must nudge **one coordinate at a time** (a partial slope), holding the other
fixed. `f(v + h)` adds $h$ to both knobs simultaneously, measuring a diagonal slope, not
the per-knob slopes — so the fit is wrong. Fix it with the proper per-coordinate gradient:

```{r p5-answer}
x <- c(1, 2, 3, 4); y <- c(3, 5, 7, 9)
mse <- function(b) mean((b[1] + b[2] * x - y)^2)
gradient <- function(f, v, h = 1e-5)
  sapply(seq_along(v), function(i) {
    up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
  })
b <- c(0, 0); for (t in 1:2000) b <- b - 0.05 * gradient(mse, b)
b   # ~ (1, 2)
```
:::

## Problem 6 — interpretation

You fit the same data three ways — your from-scratch descent, R's `optim()`, and `lm()` —
and all three return the same intercept and slope, but your descent took 3000 iterations
while `optim()` reports only 12. A colleague says "so `optim()` is more accurate." Are
they right? Answer in two or three sentences.

::: {.callout-note collapse="true" title="Answer"}
No. All three found the *same* bottom of the *same* single-bottomed error bowl, so they're
equally accurate on this problem — the answer is identical. They differ only in
*efficiency*: `optim()` (a BFGS-type method) uses the surface's curvature to take large,
well-chosen steps and so converges in far fewer iterations, while plain gradient descent
takes many small fixed steps. Fewer iterations means faster, not more correct.
:::

## Problem 7 — decode a fresh artifact

Here is the output of a real optimizer you haven't seen — `optim()` minimizing some
objective. Read each field and say, in plain words, what it tells you (no need to know
the objective).

```{r p7-artifact}
# A fresh optimizer run to decode (you don't need to know the objective)
set.seed(1)
obj <- function(v) (v[1] - 7)^2 + 3 * (v[2] + 2)^2 + 5
optim(par = c(0, 0), fn = obj, method = "BFGS")[c("par", "value", "counts", "convergence")]
```

::: {.callout-note collapse="true" title="Answer"}
- **`par`** ≈ $(7, -2)$ — the parameters at the answer: the position of the bottom of this
  objective's bowl (the values that minimize it).
- **`value`** ≈ $5$ — the objective *at* that bottom: the smallest value it reached (here
  the "+5" floor of the bowl).
- **`counts`** — how many times it evaluated the function / its gradient: a proxy for the
  number of downhill steps it took.
- **`convergence`** = `0` — the success flag: it reached a flat spot (gradient ≈ 0) and
  stopped cleanly rather than running out of iterations.

Reading those four fields off any optimizer's output — that's the whole point of the
course, and you just did it on something new.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only. Like the capstone itself, these questions draw on the whole course.

::: {.module-check module="courses/gradient-descent/modules/04-capstone-decode-the-optimizer" pass="4"}

::: {.quiz-q qid="cdo-par-field" qtype="mc" answer="a"}
**In `optim()`'s output, what is `$par`?**

a) The fitted parameters — the position of the bowl's bottom
b) The leftover error the best fit could not remove
c) The number of downhill steps the optimizer took
:::

::: {.quiz-q qid="cdo-convergence-flag" qtype="mc" answer="b"}
**`convergence = 0` in R (or `success = True` in scipy) tells you that:**

a) The optimizer found the best possible answer for any surface shape
b) The optimizer reached a spot where the gradient is near zero and stopped cleanly
c) The optimizer ran out of iterations before settling
:::

::: {.quiz-q qid="cdo-lbfgs-adds" qtype="mc" answer="c"}
**What does a method like L-BFGS-B add on top of the plain gradient descent you built?**

a) A fundamentally different search that does not use gradients
b) A guarantee of finding the deepest valley on any surface
c) A smarter step-size rule that reads the bowl's curvature, so no hand-set learning rate
:::

::: {.quiz-q qid="cdo-predict-at-answer" qtype="numeric" answer="-1.04" tolerance="0.01"}
**The capstone's fitted line is $\hat{y} = 1.969 + (-3.009)x$. What does it predict at
$x = 1$?** (Round to two decimals.)
:::

::: {.quiz-q qid="cdo-objective-field" qtype="mc" answer="a"}
**In the call `optim(par = c(0, 0), fn = mse, method = "BFGS")`, what is `fn`?**

a) The objective — the function being minimized, the bowl whose height is the error
b) The formula for the fitted line's predictions
c) The function that computes the gradient
:::

:::

::::
