---
title: "How to read a formula out loud"
subtitle: "The unwritten grammar every equation assumes you already know"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live cells below re-state their own data. Base R is all
# we need, so nothing to install.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```


## Where we're going

A formula is a **compressed language**. Someone who reads them fluently is running a
dozen silent conventions in their head — which marks mean "multiply," which letters
are quantities you'll never observe, how far a symbol's reach extends, what a little
hat *does* to the letter under it. Nobody writes those rules down, because to the
people writing papers they're as invisible as spaces between words. To everyone else
they're why an equation looks like a wall.

This lesson hands you that grammar. It won't teach any one method — it teaches the
*reading habits* every method assumes, so that when you meet $\hat{\beta}$,
$\sum_i$, $\mathbb{E}[X]$, or $p \in [0,1]$ later, you already know how to say them out
loud and what they're asking for. Every formula in every later module leans on this.

**Builds on:** just the baseline. You already read R fluently — and R is the perfect
bridge, because R forces you to *write out* almost everything math leaves invisible.

## Warm-up: do you remember?

This is the entry point, so there's nothing earlier to recall — just three R reflexes
the notation leans on. Produce each from memory before opening the answer.

**1.** For `x <- c(2, 4, 6)`, what do `x[2]` and `length(x)` return?

::: {.callout-note collapse="true" title="Answer"}
`x[2]` is `4` (the 2nd item) and `length(x)` is `3` (how many items). That indexing
bracket is exactly the subscript $x_i$ you're about to meet, and `length(x)` is the
$n$ that sits on top of a sum.
:::

**2.** Is `2 + 3 * 4` equal to `20` or `14`? Why?

::: {.callout-note collapse="true" title="Answer"}
`14` — multiplication binds before addition ($3 \times 4 = 12$, then $+2$). "What
binds first?" is a reading habit, and math relies on it even harder than R because it
often hides the $\times$ sign entirely.
:::

**3.** What does `log(exp(1))` return in R, and what base is R's `log()`?

::: {.callout-note collapse="true" title="Answer"}
`1`. R's `log()` is the **natural** log (base $e$) — `log10()` is the base-10 one.
Papers write $\log$ and almost always mean natural log too; that unspoken default is
exactly the kind of convention this lesson is about.
:::

## A tiny example first

Here is a formula you already know cold, dressed in full notation:

$$
\bar{x} = \frac{1}{n}\sum_{i=1}^{n} x_i
$$

It looks like four moving parts. It says one thing you've typed a thousand times:

> add up the numbers and divide by how many there are.

In R that whole wall is `mean(x)`. Nothing in the formula is new to you — the bar on
top means "the average of," the big $\sum$ is a `for` loop that adds, the $x_i$ is
`x[i]`, the $n$ is `length(x)`. **Reading a formula is exactly this move**: take the
compressed marks, name each one out loud, and watch a plain idea fall out. The rest of
this lesson is the vocabulary that lets you do it for formulas you *don't* already
know.

```{r toy-example}
# The formula, made literal — build the mean from its pieces...
x <- c(2, 4, 6)
n <- length(x)          # the n on top of the sum
sum(x) / n              # (1/n) * sum of x_i  ->  4

# ...and the one-word version the notation is compressing
mean(x)                 # 4 — identical
```

## Getting a feel for it

**The mental picture: notation is your R code with the operators turned invisible.**
Imagine taking `sum(2 * x[i])`, erasing the `*`, erasing the function names, and
shrinking every variable to a single letter with a little mark on top. That's
essentially what happened to produce $\sum 2x_i$. Reading notation is running that
compression *backwards*: put the invisible operators back, expand the letters into
quantities, and you're holding runnable code again.

**Why it's built this compressed way.** Two reasons, both practical. A formula has to
fit on a page and be read aloud in a lecture, so anything that can be dropped — the
multiplication sign, the loop boilerplate, the word "average" — gets dropped. And it
has to describe a computation for a list of *any* size at once, so concrete numbers
become indexed letters ($x_i$) and counts become symbols ($n$). Compression plus
generality: that's the whole design.

**Where the analogy breaks — and this is the hard part.** A programming language has
*one* meaning per symbol; math notation does not. The same bars $|\cdot|$ mean four
different things depending on what's inside them; a superscript is usually a power but
sometimes a label; one author's $r$ is a correlation and another's is a residual. So
fluent reading isn't memorizing a fixed dictionary — it's reading *around* each symbol,
using context to pick which meaning is live. This lesson gives you the common meanings
and the context cues that disambiguate them.

## What the notation is going to say

We'll decode two formulas you'll genuinely meet. The first,
$\hat{\beta} = \arg\min_{\beta} \sum_i (y_i - \beta_0 - \beta_1 x_i)^2$, says in
English: *"the slope and intercept I'll report are the pair that make the total
squared prediction error as small as possible."* The second is a cluster of
probability shorthand — $\mathbb{E}[X]$, $P(A \mid B)$, $p \in [0,1]$,
$P(\theta \mid \text{data}) \propto \dots$ — saying *"the long-run average of X,"*
*"the chance of A once you know B,"* *"p is a number between 0 and 1,"* and
*"proportional to, up to a constant."* By the end you'll say each of those aloud
without flinching.

## How to read the symbols

This is the reference heart of the lesson. Every symbol you'll meet in the two
formulas — and most you'll meet anywhere in these courses — is below, grouped by the
*job* the mark does. Skim it now; come back to it whenever a formula stumps you.

### Marks that sit *on* a letter (accents = operators)

An accent is a verb applied to the letter under it — it changes what the letter means.

| Symbol | How to say it | What it means | In R |
|---|---|---|---|
| $\bar{x}$ | "ex bar" | the mean (average) of the x values | `mean(x)` |
| $\hat{\beta}$ | "beta hat" | an **estimate from data** — the hat always says "computed from a sample" | `coef(fit)` |
| $\hat{y}_i$ | "why hat sub eye" | the model's predicted value for observation i | `predict(fit)[i]` |
| $\tilde{x}$ | "ex tilde" | an adjusted/modified version of x (context says how) | — |
| $\dot{x}$ | "ex dot" | a rate of change over time (physics-flavored) | — |

### Which alphabet a letter is in (Greek vs Latin)

| Symbol | How to say it | What it means | In R |
|---|---|---|---|
| $\mu, \sigma, \beta, \theta$ | "mew, sigma, beta, THAY-ta" | **Greek lowercase = an unknown truth** you never observe (a *parameter*) | — (you estimate it) |
| $\hat{\beta}, \bar{x}$ | "beta hat, ex bar" | **Hatted = estimated from data** — the hat turns any letter (Greek or Latin) into a statistic | `coef(fit)`, `mean(x)` |
| $s, n$ | "ess, en" | **Bare Latin = statistic or count from your data** (no hat needed) | `sd(x)`, `length(x)` |
| $\Sigma$ vs $\sigma$ | "big sigma / little sigma" | **case matters:** $\Sigma$ is often a sum *or* a covariance matrix; $\sigma$ is a spread | `sum(...)` / — |

: The rule is really *parameter vs statistic*; "Greek = truth" is a strong guess with
real exceptions ($\pi$, $\alpha$, the operator $\Sigma$). See the deep-dive in this
module's [resources](resources.md).

### Little numbers beside a letter (sub- vs superscript)

| Symbol | How to say it | What it means | In R |
|---|---|---|---|
| $x_i$ | "ex sub eye" | **subscript = an index/label:** the i-th item, or a named slot | `x[i]` |
| $y_i$ | "why sub eye" | the i-th observed value of a second list y | `y[i]` |
| $\beta_0, \beta_1$ | "beta zero, beta one" | subscript as a **name**, not a count: intercept, slope | two separate numbers |
| $x^2$ | "ex squared" | **superscript = a power** (the usual case) | `x^2` |
| $x^{(i)}$ | "ex super eye" | superscript in **parentheses = a label**, not a power (the i-th one) | `x[[i]]` |
| $X^\top$ | "ex transpose" | superscript $\top$ = flip a matrix's rows and columns | `t(X)` |
| $x^{*}$ | "ex star" | superscript star = a special/optimal value | — |

### Relations (how two sides compare)

| Symbol | How to say it | What it means | In R |
|---|---|---|---|
| $=$ | "equals" | the two sides are the same number | `==` (test) / `<-` (define) |
| $:=$ or $\equiv$ | "is defined as" | this is a *definition*, not a fact to prove | `<-` |
| $\approx$ | "is approximately" | nearly equal | `isTRUE(all.equal(a, b))` |
| $\propto$ | "is proportional to" | equal **up to some constant multiplier** you're ignoring | `a <- k * b` |
| $\in$ | "is in / is an element of" | belongs to a set or interval | `%in%` |

### Operators that look like functions

| Symbol | How to say it | What it means | In R |
|---|---|---|---|
| $n$ | "en" | the number of data points (the count the loop runs to) | `length(x)` |
| $\sum_{i=1}^{n}$ | "the sum from i equals 1 to n" | add the expression to its right, once per i | `sum(...)` |
| $\prod_{i=1}^{n}$ | "the product from 1 to n" | multiply instead of add | `prod(...)` |
| $\mathbb{E}[X]$ | "the expected value of X" | the long-run average of the random quantity X | `mean(x)` (empirically) |
| $\mathrm{Var}(X)$ | "the variance of X" | the long-run spread of X | `var(x)` |
| $P(A)$ | "the probability of A" | how likely event A is (a number in $[0,1]$) | — |
| $P(A \mid B)$ | "the probability of A **given** B" | chance of A once you know B happened | — |
| $\arg\min_{\beta} f(\beta)$ | "arg min over beta of f" | the **value of $\beta$** that makes $f$ smallest — *not* the smallest $f$ | `beta[which.min(f)]` |
| $\arg\max_{\beta} f(\beta)$ | "arg max over beta of f" | the **value of $\beta$** that makes $f$ largest | `beta[which.max(f)]` |

### Grouping, and marks that mean different things by context

| Symbol | How to say it | What it means | In R |
|---|---|---|---|
| $\beta_1 x$ | "beta one ex" | **juxtaposition = multiply:** two things written together are multiplied | `b1 * x` |
| $(\,\cdot\,)$ | "parentheses" | group first — the operator reaches over everything inside | `( )` |
| $\lvert x \rvert$ | "absolute value of x" | distance from zero (one number) | `abs(x)` |
| $\lVert \mathbf{x} \rVert$ | "the norm of x" | the length of a vector (double bars) | `sqrt(sum(x^2))` |
| $\lvert A \rvert$ | "size of A" | the number of elements in a set | `length(A)` |
| $\mathbf{x}, \mathbf{X}$ | "the vector x / matrix X" | **bold = a whole list/table** at once | an R vector / matrix |
| $A_{ij}$ | "A sub i j" | the entry in **row i, column j** | `A[i, j]` |
| $[0, 1]$ vs $(0, 1)$ | "closed / open interval" | square bracket **includes** the endpoint; round **excludes** it | `x >= 0 & x <= 1` / `x > 0 & x < 1` |

## The formal version

### Formula 1 — the least-squares line

$$
\hat{\beta} = \arg\min_{\beta} \sum_{i=1}^{n} \left(y_i - \beta_0 - \beta_1 x_i\right)^2
$$

**Reading it aloud:** "beta-hat equals arg-min over beta of the sum, from i equals 1
to n, of the quantity y-sub-i minus beta-zero minus beta-one times x-sub-i, all
squared." **In everyday words:** "the slope and intercept I'll report are the pair
that make the total squared prediction error smallest."

Now watch every convention from the table above fire at once:

- $\hat{\beta}$ — the **hat** says these are estimates *from the data*, as opposed to
  the true $\beta$ we can never see.
- $\arg\min_{\beta}$ — we want the **value of $\beta$** that minimizes, not the
  minimum error itself. (The single most common misreading; see the traps.)
- $\beta_0, \beta_1$ — subscripts here are **names** (intercept, slope), not the i-th
  item of anything.
- $\sum_{i=1}^{n}$ — a loop adding one term per data point; it **reaches over the whole
  squared parenthesis** to its right.
- $y_i, x_i$ — subscripts here *are* indices: the i-th observation, `y[i]` and `x[i]`.
- $\beta_1 x_i$ — written together, so **multiply**: `b1 * x[i]`.
- $(\dots)^2$ — the superscript is a genuine **power**; the parentheses tell it to
  square the *whole* residual.

**Sanity-check it in R.** "Arg min" is just "try many $\beta$s and keep the best one."
With a 3-point dataset, scan a grid of slopes and confirm the winner matches `lm()`:

```{r argmin-check}
x <- c(1, 2, 3)
y <- c(1, 3, 2)                 # deliberately not a perfect line

# For a fixed intercept, total squared error as a function of the slope b1:
b0     <- mean(y) - mean(x) * cov(x, y) / var(x)   # least-squares intercept
slopes <- seq(-2, 3, by = 0.001)
sse    <- sapply(slopes, function(b1) sum((y - b0 - b1 * x)^2))

c(argmin_slope = slopes[which.min(sse)],   # the VALUE of b1 that wins  (arg min)
  min_sse      = min(sse),                 # the smallest error itself  (the min)
  lm_slope     = unname(coef(lm(y ~ x))[2]))  # what lm() reports — should match
```

The `argmin_slope` and `lm_slope` agree, and notice they are *different* from
`min_sse`: the arg-min is the slope, the min is the leftover error. Two different
questions, one symbol apart.

### Formula 2 — the probability shorthand

These four short lines carry most of the probability notation you'll meet:

$$
p \in [0, 1], \qquad
\mathbb{E}[X] = \mu, \qquad
P(A \mid B), \qquad
P(\theta \mid \text{data}) \propto P(\text{data} \mid \theta)\, P(\theta)
$$

**Reading them aloud:** "p is in the closed interval zero to one" · "the expected
value of X equals mew" · "the probability of A given B" · "the probability of theta
given the data is proportional to the probability of the data given theta, times the
probability of theta."

- $p \in [0,1]$ — **membership** plus a **closed interval**: $p$ is a number, and it
  can be exactly 0 or 1 (square brackets include the ends). In R: `p >= 0 & p <= 1`.
- $\mathbb{E}[X] = \mu$ — $\mathbb{E}[\cdot]$ is an **operator** (long-run average),
  and the answer is named with a **Greek** letter because it's the *true* mean, not a
  sample estimate.
- $P(A \mid B)$ — the **bar means "given":** we've restricted to the world where B
  already happened.
- $\propto$ — "**proportional to**": the two sides differ by a constant we're not
  writing. It is *not* an equals sign, and forgetting that is its own trap below.

## Explore it

R is unusually good at teaching notation, because it refuses to hide the operators
math leaves out. Run and tweak these.

```{webr}
#| autorun: false
# explore-1 — LIVE: juxtaposition means MULTIPLY. Math writes "beta1 x"; R makes you
# type the star. PREDICT: what does R do with `beta1 * x`? And what would bare
# `x(2)` do — multiply, or something else? Commit, then run.
beta1 <- 3
x     <- 4
beta1 * x          # 12 — "beta1 x" in a formula is this

# The gotcha: in R, `name(...)` is a FUNCTION CALL, not multiplication.
# Math's  x(y + z)  means multiply; R's  x(y + z)  tries to CALL x. Uncomment to see:
# x(1 + 1)
```

```{webr}
#| autorun: false
# explore-2 — LIVE: the SAME bars |.| mean four different things. Predict each, then
# run. PREDICT: which of these equals 5?
abs(-3)                       # |x|  absolute value: distance from 0
sqrt(sum(c(3, 4)^2))          # ||x|| vector norm (double bars): length of (3,4)
length(c(10, 20, 30))         # |A|  size of a set
det(matrix(c(2, 0, 0, 2), 2)) # |A|  determinant of a matrix
```

```{webr}
#| autorun: false
# explore-3 — LIVE: "arg min" is the WINNING INPUT, not the winning value.
# PREDICT: which.min returns the position; what slope wins? Edit y and re-run.
x <- c(1, 2, 3)
y <- c(1, 3, 2)
slopes <- seq(-2, 3, by = 0.01)
sse    <- sapply(slopes, function(b1) sum((y - mean(y) - b1 * (x - mean(x)))^2))

c(arg_min = slopes[which.min(sse)],  # the slope that minimizes  <- arg min
  the_min = min(sse))                # the minimum error itself   <- NOT arg min
```

```{webr}
#| autorun: false
# explore-4 — LIVE: set/interval membership. PREDICT each TRUE/FALSE before running.
p <- 0.3
p >= 0 & p <= 1     # is p in the CLOSED interval [0, 1]?
2 %in% c(1, 2, 3)   # is 2 an element of {1, 2, 3}?  (the ∈ symbol)
0 %in% c(1, 2, 3)   # and 0?
```

## Common traps

**Trap 1 — reading a subscript as a superscript (or vice-versa).** $x_2$ and $x^2$
look almost the same at a glance and mean completely different things. $x_2$ is the
**second item** of a list (`x[2]`); for `x <- c(2, 4, 6)` that's `4`. $x^2$ is
**x squared** (`x^2`); for a single $x = 4$ that's `16`. The little number's *position*
— below or above — is the whole message. And a superscript *in parentheses*, $x^{(i)}$,
flips back to being a label ("the i-th one"), not a power at all.

**Trap 2 — thinking $\arg\min$ is the minimum.** $\arg\min_{\beta} f(\beta)$ is the
**input** $\beta$ that makes $f$ smallest, not the smallest value of $f$. In
`explore-3` the arg-min was a *slope* (around 0.5) while the min was a *sum of squared
errors* (a different number entirely). If a formula defines an estimate as an
$\arg\min$ or $\arg\max$ — as nearly every fitting method does — it's telling you *which
setting to choose*, and the leftover value is a separate quantity. Say "the beta that
minimizes," never "the minimum," and you won't slip.

## Check yourself

**1.** Read this aloud, then translate it to one line of R:
$\;\hat{y}_i = \hat{\beta}_0 + \hat{\beta}_1 x_i$.

```{webr}
#| autorun: false
# check1-starter — set the pieces, then write the prediction line and Run.
b0 <- 2      # beta-hat-zero (intercept)
b1 <- 3      # beta-hat-one  (slope)
x  <- c(0, 1, 2)
# your line here: yhat <- ...
```

::: {.callout-note collapse="true" title="Answer"}
"why-hat sub eye equals beta-hat-zero plus beta-hat-one times x-sub-eye." Every hat
means "estimated from data"; the subscripts on $\beta$ are names (intercept, slope);
$\hat{\beta}_1 x_i$ is written together, so multiply. In R:

```{r check1-answer}
b0 <- 2; b1 <- 3; x <- c(0, 1, 2)
yhat <- b0 + b1 * x     # the invisible × becomes an explicit *
yhat                    # 2 5 8
```
:::

**2.** No computing — just interpretation. (a) What's the difference between $\sigma$
and $s$? (b) Between $x_2$ and $x^2$?

::: {.callout-note collapse="true" title="Answer"}
(a) $\sigma$ (Greek) is the **true, unknown** standard deviation of the population; $s$
(Latin) is the estimate you **compute from your sample** (`sd(x)`). Same quantity, one
is truth and one is your data's guess at it. (b) $x_2$ is the **second element** of a
list (`x[2]`); $x^2$ is **x squared** (`x^2`). Subscript = index, superscript = power.
:::

**3.** In your own words: explain to a colleague, in two sentences, what "reading a
formula" actually involves.

::: {.callout-note collapse="true" title="A model answer"}
Reading a formula is decompressing it: you put back the operators it hides (a $\times$
between two letters written together, a `for` loop inside a $\sum$) and you check each
letter's costume — a hat or a Greek letter tells you whether it's data or an unknown
truth. Then you notice how far each operator reaches, and a plain-English sentence
falls out.
:::

## What you can now do

- **Say a formula out loud** — name each accent, subscript, operator, and relation —
  and turn it into runnable R by making the hidden operators explicit.
- **Tell data from unknowns at a glance:** a hat on any letter — Greek or Latin — signals
  "estimated from data"; a bare Latin letter ($s$, $n$) is also a statistic; a bare Greek
  letter ($\beta$, $\sigma$) usually means "true but unobserved."
- **Read a subscript as an index or a name, a superscript as a power or a label,** using
  position and parentheses as the cue.
- **Not confuse $\arg\min$ with the minimum,** $\propto$ with $=$, or the four meanings
  of $\lvert \cdot \rvert$ — reading each symbol *in context*.

If any of these feels shaky, revisit [How to read the symbols](#how-to-read-the-symbols)
and keep it open as a reference — every later formula is built from these marks.

## Where this goes next

The single most common symbol you just met — the big $\sum$ — gets a whole module of
its own next: [vectors, subscripts & summation](../vectors-and-summation/lesson.qmd),
where you'll learn the three algebra moves for reshaping sums that power almost every
derivation. From there, [mean, variance & covariance](../mean-variance-covariance/lesson.qmd)
puts these symbols to work on real formulas.

Ready for more reps? Do the [practice problems](practice.qmd).
