---
title: "Practice: How to read a formula out loud"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Say each formula out loud first, then check.

## Problem 1 — decode a known formula, new numbers

The lesson decoded $\bar{x} = \frac{1}{n}\sum_{i=1}^{n} x_i$ into `mean(x)`. Here's the
same shape for a different letter and fresh data:

$$
\bar{w} = \frac{1}{n}\sum_{i=1}^{n} w_i, \qquad w = (5, 10, 15, 10)
$$

Read it aloud, then give its value and the one-word R call that computes it.

```{webr}
#| autorun: false
# p1-starter — edit and Run to solve, right here
w <- c(5, 10, 15, 10)
```

::: {.callout-note collapse="true" title="Answer"}
"w-bar equals one over n, times the sum from i equals 1 to n, of w-sub-i" — the mean
of the w values. Add $5 + 10 + 15 + 10 = 40$, divide by $n = 4$, get **10**.

```{r p1-answer}
w <- c(5, 10, 15, 10)
sum(w) / length(w)   # 10 — the formula, literal
mean(w)              # 10 — the one word it compresses to
```
:::

## Problem 2 — fill in the missing step

Translating $\hat{y}_i = \hat{\beta}_0 + \hat{\beta}_1 x_i$ to R goes in three moves.
Two are shown; supply the middle one.

- **Step 1 (read the marks):** every hat means "estimated from data"; $\hat\beta_0$ and
  $\hat\beta_1$ are the intercept and slope (subscripts as *names*); $x_i$ is `x[i]`.
- **Step 2:** **_your step here_** — what does the notation $\hat{\beta}_1 x_i$, with the
  two symbols written side by side, tell you to put between them in R?
- **Step 3 (write it):** `yhat <- b0 + b1 * x`.

```{webr}
#| autorun: false
# p2-starter — set the pieces, write yhat, and Run to check the blank
b0 <- 1; b1 <- 2; x <- c(0, 1, 2)
```

::: {.callout-note collapse="true" title="Answer"}
The missing move: **juxtaposition means multiply.** $\hat{\beta}_1 x_i$ — two things
written together with no symbol — is a product, so it becomes `b1 * x` in R (the
explicit `*` is exactly what math leaves invisible).

```{r p2-answer}
b0 <- 1; b1 <- 2; x <- c(0, 1, 2)
yhat <- b0 + b1 * x
yhat                 # 1 3 5
```
:::

## Problem 3 — match the symbol to its meaning (interpretation, no computing)

For each, say what the mark is doing. No arithmetic — this is pure reading.

(a) In $\hat{\sigma}$, what is the hat telling you, given that $\sigma$ is a Greek
letter?
(b) $x_3$ versus $x^3$ — which is "the third item" and which is "x cubed"?
(c) In $\arg\max_{\theta} L(\theta)$, is the answer a value of $\theta$ or the largest
value of $L$?
(d) $P(\text{rain} \mid \text{clouds})$ — read the bar aloud and say what it restricts.

::: {.callout-note collapse="true" title="Answer"}
(a) The hat says "**estimated from data**," and it **overrides the alphabet** — $\sigma$
alone is the unknown true spread, but $\hat{\sigma}$ is the sample estimate of it.
(b) $x_3$ (subscript) is the **third item**, `x[3]`; $x^3$ (superscript) is **x cubed**,
`x^3`.
(c) A **value of $\theta$** — the setting that maximizes $L$, not the maximum of $L$
itself.
(d) "The probability of rain **given** clouds" — it restricts attention to the world
where it's already cloudy.
:::

## Problem 4 — spot the error

A colleague is reading the least-squares formula
$\hat{\beta} = \arg\min_{\beta} \sum_i (y_i - \beta_0 - \beta_1 x_i)^2$ aloud and writes
down its meaning. It contains **exactly one mistake**. Find and fix it.

> "beta-hat is the **smallest total squared error** you can get by choosing the
> intercept and slope."

::: {.callout-note collapse="true" title="Answer"}
The mistake is confusing $\arg\min$ with the minimum. $\hat{\beta}$ is **not** the
smallest error — it's the **pair (intercept, slope) that achieves** the smallest error.
The $\arg\min$ hands you the *inputs*; the minimum error itself is a different, leftover
number.

Corrected: "beta-hat is the intercept and slope that **make** the total squared error
as small as possible." (In `explore-3` of the lesson, the arg-min was a slope ≈ 0.5
while the min was a sum-of-squares — two different quantities.)
:::

## Problem 5 — transfer to a formula you haven't seen

Here's a real one from the next module, dressed up. You can't *compute* it yet, but you
can already **read every mark**:

$$
s_{xy} = \frac{1}{n-1}\sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y})
$$

Without computing, answer: (a) is $s_{xy}$ data or an unknown truth? (b) how many
$\times$ signs are hidden in $(x_i - \bar{x})(y_i - \bar{y})$? (c) how far right does the
$\sum$ reach?

::: {.callout-note collapse="true" title="Answer"}
(a) **Data** — it's a Latin $s$ (with subscripts as names), a *statistic* computed from
your sample, not a Greek parameter. (b) **One** hidden $\times$: the two parenthesized
factors are written side by side, so they're multiplied — `(x - mean(x)) * (y - mean(y))`
in R. (The $\bar{x}, \bar{y}$ are means, not multiplications.) (c) The $\sum$ reaches
over the **whole product of both parentheses**, once per i; the $\frac{1}{n-1}$ sits
*outside* it.

You've just read a covariance formula cold — that's the payoff. The next module,
[vectors, subscripts & summation](../vectors-and-summation/lesson.qmd), teaches the sum
machinery to actually evaluate it.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh in your head. A pass *then* is real evidence you've
retained it, and the effort of recalling it cold is itself what makes it stick. Five
questions, graded on the page; 4/5 or better means this module is solid. Results are
saved in this browser only.

::: {.module-check module="foundations/reading-math-notation" pass="4"}

::: {.quiz-q qid="rmn-hat-meaning" qtype="mc" answer="b"}
**In $\hat{\beta}$, what does the hat mean?**

a) The true parameter value we can never observe
b) An estimate computed from a sample
c) The largest value $\beta$ can take
d) A vector rather than a single number
:::

::: {.quiz-q qid="rmn-sub-vs-super" qtype="mc" answer="a"}
**For `x <- c(2, 4, 6)`, what is $x_2$?**

a) 4
b) 16
c) 36
d) 2
:::

::: {.quiz-q qid="rmn-sum-toy" qtype="numeric" answer="14" tolerance="0"}
**For $x = (3, 5, 6)$, what is $\sum_{i=1}^{n} x_i$?**
:::

::: {.quiz-q qid="rmn-argmin" qtype="mc" answer="b"}
**What is $\arg\min_{\beta} f(\beta)$?**

a) The smallest value $f$ reaches
b) The value of $\beta$ that makes $f$ smallest
c) The average of $f$
d) The number of minima
:::

::: {.quiz-q qid="rmn-propto" qtype="mc" answer="c"}
**What does $\propto$ mean?**

a) Approximately equal
b) Defined as
c) Equal up to a constant multiplier
d) An element of
:::

:::

::::
