---
title: "Orthogonality, projection & least squares"
subtitle: "Why the best-fit line works: the residual is perpendicular to every predictor column"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```


## Where we're going

When three data points don't fall perfectly on a line, no choice of intercept and
slope makes the errors zero — the system of equations is **inconsistent**. The best
you can do is find the line whose errors are as small as possible. That "best
approximate answer" turns out to be a **projection**: you drop a perpendicular from the
data onto the space of all possible fitted values, and the foot of that perpendicular
is the fitted line. The normal equations are nothing but the algebraic statement of
that perpendicularity — and `lm()` finds the same least-squares solution (via QR
decomposition, which is numerically more stable than solving the normal equations
directly).
[@austin2022understandinglinearalgebra, pp. 397--398]

This lesson extends the "shadow" you already know from the distance module into a full
projection machine, then shows that ordinary least squares IS that machine.

**Builds on:**
[distance, similarity & geometry](../distance-similarity-and-geometry/lesson.qmd) —
the shadow of one vector onto another (dot product divided by norm);
[matrices & linear transforms](../matrices-and-linear-transforms/lesson.qmd) —
$A^\top A$ and matrix multiplication;
[rank & linear independence](../rank-and-linear-independence/lesson.qmd) — why
independent columns matter (the normal equations need $A^\top A$ invertible).

## Warm-up: do you remember?

Three quick reps — produce each answer from memory first, then check.

**1.** *(distance)* For $\mathbf{a} = (4, 2)$ and $\mathbf{b} = (3, 4)$, what is the
shadow length of $\mathbf{a}$ along $\mathbf{b}$?

::: {.callout-note collapse="true" title="Answer"}
Shadow length = $\mathbf{a} \cdot \mathbf{b} \;/\; \lVert \mathbf{b} \rVert$
$= (4 \times 3 + 2 \times 4) / \sqrt{9 + 16} = 20 / 5 = \mathbf{4}$. Taught in
[distance, similarity & geometry §Formal version](../distance-similarity-and-geometry/lesson.qmd#the-formal-version);
revisit if shaky.
:::

**2.** *(matrices)* For $A = \begin{bmatrix} 1 & 1 \\ 1 & 2 \\ 1 & 3 \end{bmatrix}$,
what is $A^\top A$? (It's a $2 \times 2$ matrix — compute each entry.)

::: {.callout-note collapse="true" title="Answer"}
$A^\top A = \begin{bmatrix} 1\!\cdot\!1 + 1\!\cdot\!1 + 1\!\cdot\!1 &
1\!\cdot\!1 + 1\!\cdot\!2 + 1\!\cdot\!3 \\ 1\!\cdot\!1 + 2\!\cdot\!1 + 3\!\cdot\!1 &
1\!\cdot\!1 + 2\!\cdot\!2 + 3\!\cdot\!3 \end{bmatrix} =
\begin{bmatrix} 3 & 6 \\ 6 & 14 \end{bmatrix}$. Taught in
[matrices & linear transforms §Transpose](../matrices-and-linear-transforms/lesson.qmd);
revisit if shaky.
:::

**3.** *(rank)* If a matrix $A$ has 3 rows and 2 columns, and both columns point in
genuinely different directions (neither is a multiple of the other), what is the rank
of $A$? Can you invert $A^\top A$?

::: {.callout-note collapse="true" title="Answer"}
Rank = 2 (two independent columns). Since the columns are independent, $A^\top A$ is
a $2 \times 2$ invertible matrix — `solve(t(A) %*% A)` works. Taught in
[rank & linear independence](../rank-and-linear-independence/lesson.qmd); revisit if
shaky.
:::


## A tiny example first

Three students: hours studied = 1, 2, 3; exam scores = 1, 3, 2. You want the
best-fit line $y = a + bx$, but the points don't line up:

| Student | Hours ($x$) | Score ($y$) |
|---|---|---|
| A | 1 | 1 |
| B | 2 | 3 |
| C | 3 | 2 |

To fit the line $y = a + bx$, you need $a$ and $b$ such that:

$$
a + 1b = 1, \quad a + 2b = 3, \quad a + 3b = 2
$$

Three equations, two unknowns — no exact solution exists. Write this as a matrix
equation $A\mathbf{x} = \mathbf{y}$ where the design matrix $A$ stacks the
coefficients:

$$
\underbrace{\begin{bmatrix} 1 & 1 \\ 1 & 2 \\ 1 & 3 \end{bmatrix}}_{A}
\underbrace{\begin{bmatrix} a \\ b \end{bmatrix}}_{\mathbf{x}} =
\underbrace{\begin{bmatrix} 1 \\ 3 \\ 2 \end{bmatrix}}_{\mathbf{y}}
$$

No $\mathbf{x}$ makes this exact. So which $\mathbf{x}$ comes closest?

The one that makes the **residual** $\mathbf{e} = \mathbf{y} - A\hat{\mathbf{x}}$ as
short as possible — that is, minimises $\lVert \mathbf{e} \rVert^2 = e_1^2 + e_2^2 +
e_3^2$. The answer turns out to be $a = 1$, $b = 0.5$:

- Fitted values: $\hat{\mathbf{y}} = (1.5,\; 2,\; 2.5)$
- Residuals: $\mathbf{e} = (1 - 1.5,\; 3 - 2,\; 2 - 2.5) = (-0.5,\; 1,\; -0.5)$

Here is the key fact — check it with arithmetic:

- Column 1 of $A$ is $(1, 1, 1)$. Dot it with $\mathbf{e} = (-0.5, 1, -0.5)$:
  $1(-0.5) + 1(1) + 1(-0.5) = \mathbf{0}$.
- Column 2 of $A$ is $(1, 2, 3)$. Dot it with $\mathbf{e}$:
  $1(-0.5) + 2(1) + 3(-0.5) = -0.5 + 2 - 1.5 = \mathbf{0}$.

The residual is **orthogonal to both columns of $A$**. That is not a coincidence — it
is the whole reason least squares works. The rest of this lesson is about *why*.

```{r toy-example}
x <- c(1, 2, 3); y <- c(1, 3, 2)
A <- cbind(1, x)
xhat <- solve(t(A) %*% A, t(A) %*% y)   # solve A^T A xhat = A^T y
cat("intercept:", xhat[1], " slope:", xhat[2], "\n")

yhat <- A %*% xhat                        # fitted values
e    <- y - yhat                           # residuals
cat("fitted:", yhat, "\n")
cat("residuals:", e, "\n")

cat("col1 dot e:", sum(A[,1] * e), "\n")   # 0 — orthogonal
cat("col2 dot e:", sum(A[,2] * e), "\n")   # 0 — orthogonal
```


## Getting a feel for it

In the distance module you learned the **shadow**: the shadow of $\mathbf{a}$ onto
$\mathbf{b}$ tells you how much of $\mathbf{a}$ lies along $\mathbf{b}$'s direction.
The leftover — $\mathbf{a}$ minus its shadow — is perpendicular to $\mathbf{b}$.

Least squares is the same move, scaled up. Instead of projecting one vector onto a
*line* (the direction of $\mathbf{b}$), you project the response vector $\mathbf{y}$
onto the **flat surface** spanned by the columns of $A$. The surface contains every
vector you could build by mixing those columns — every possible set of fitted values
$a \cdot (1,1,1) + b \cdot (1,2,3)$. The projection $\hat{\mathbf{y}}$ is the point on
that surface closest to $\mathbf{y}$, and the residual $\mathbf{e}$ is the perpendicular
dropped from $\mathbf{y}$ down to the surface.
[@austin2022understandinglinearalgebra, pp. 371--372, 397--398]

Why perpendicular? Because a perpendicular is the shortest path from a point to a flat
surface — the same reason a plumb line hangs straight down. If the residual tilted even
slightly toward the surface, you could slide the foot of the perpendicular a little and
get a shorter residual. Perpendicular means you have already slid as far as possible —
the errors are as small as they can be.

In the toy example, $\mathbf{y} = (1, 3, 2)$ lives in 3D space (one dimension per
student). The two columns of $A$ span a 2D plane inside that 3D space. The fitted
values $(1.5, 2, 2.5)$ are the point on that plane closest to $(1, 3, 2)$, and the
residual $(-0.5, 1, -0.5)$ is the perpendicular from $\mathbf{y}$ down to the plane.
That perpendicular has squared length $0.25 + 1 + 0.25 = 1.5$ — any other line would
give a longer residual.


## What the notation is going to say

The formula will say: "find $\hat{\mathbf{x}}$ so that $A\hat{\mathbf{x}}$ is the
closest point in the span of $A$'s columns to $\mathbf{y}$." The condition for
"closest" is that the residual $\mathbf{y} - A\hat{\mathbf{x}}$ is orthogonal to every
column. Writing "orthogonal to every column" in matrix language gives
$A^\top (\mathbf{y} - A\hat{\mathbf{x}}) = \mathbf{0}$, which rearranges to the
**normal equations**: $A^\top A\,\hat{\mathbf{x}} = A^\top \mathbf{y}$. In our toy,
that's $\begin{bmatrix}3 & 6 \\ 6 & 14\end{bmatrix}\hat{\mathbf{x}} =
\begin{bmatrix}6 \\ 13\end{bmatrix}$, exactly the system `solve(t(A) %*% A, t(A) %*% y)`
solves.


## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $A$ | "the design matrix" | each row is one observation; columns are predictor variables (first column often all 1s for the intercept) | $\begin{bmatrix}1&1\\1&2\\1&3\end{bmatrix}$ |
| $\mathbf{y}$ | "why" (the response vector) | the column of observed outcomes, one entry per observation | $(1, 3, 2)$ |
| $\hat{\mathbf{x}}$ | "ex hat" | the vector of coefficients that minimises the residual — the least-squares solution; the hat means "estimated from data" | $(1, 0.5)$ = intercept 1, slope 0.5 |
| $\hat{\mathbf{y}}$ | "why hat" (the fitted values) | $A\hat{\mathbf{x}}$ — the projection of $\mathbf{y}$ onto the column space of $A$ | $(1.5, 2, 2.5)$ |
| $\mathbf{e}$ | "the residual vector" | $\mathbf{y} - \hat{\mathbf{y}}$ — what the model missed; perpendicular to every column of $A$ | $(-0.5, 1, -0.5)$ |
| $\text{Col}(A)$ | "the column space of A" | the set of all vectors you can build as $A\mathbf{x}$ — every possible set of fitted values; this is the **span** of $A$'s columns | a 2D plane inside $\mathbb{R}^3$ |
| $A^\top A$ | "A transpose A" | the Gram matrix — always square, always symmetric; invertible when columns are independent | $\begin{bmatrix}3&6\\6&14\end{bmatrix}$ |
| $A^\top \mathbf{y}$ | "A transpose why" | the right-hand side of the normal equations — each entry is a column of $A$ dotted with $\mathbf{y}$ | $(6, 13)$ |
| $\mathbf{a}$ | "a" (the vector being projected) | the vector whose shadow we want | $(4, 2)$ in the projection example |
| $\mathbf{b}$ | "b" (the direction projected onto) | the vector defining the line we project onto | $(3, 4)$ in the projection example |
| $\hat{\mathbf{a}}$ | "a-hat" (the projection of a onto b) | the shadow as a vector — the point on the line through $\mathbf{b}$ closest to $\mathbf{a}$ | $(2.4, 3.2)$ |
| $\perp$ | "perpendicular to" / "orthogonal to" | at right angles — dot product is 0 | $\mathbf{e} \perp \text{Col}(A)$ |


## The formal version

### Projection vector (one vector onto another)

The distance module gave you the shadow **length**:
$\mathbf{a} \cdot \mathbf{b} / \lVert \mathbf{b} \rVert$. Now we need the shadow as a
**vector** — the actual point on the line through $\mathbf{b}$ closest to $\mathbf{a}$.
Scale $\mathbf{b}$ to the right length:
[@austin2022understandinglinearalgebra, pp. 371--374]

$$
\hat{\mathbf{a}} \;=\;
\frac{\mathbf{a} \cdot \mathbf{b}}{\mathbf{b} \cdot \mathbf{b}} \;\mathbf{b}
$$

Reading it aloud: "a-hat equals a-dot-b over b-dot-b, times b." In everyday words:
"scale b by how much of a points along it."

Take $\mathbf{a} = (4, 2)$, $\mathbf{b} = (3, 4)$:
$\hat{\mathbf{a}} = (20/25) \times (3, 4) = 0.8 \times (3, 4) = (2.4, 3.2)$.
The leftover $\mathbf{a} - \hat{\mathbf{a}} = (1.6, -1.2)$ is orthogonal to
$\mathbf{b}$: $(1.6)(3) + (-1.2)(4) = 4.8 - 4.8 = 0$. The perpendicular from
$\mathbf{a}$ to the line through $\mathbf{b}$ has zero dot product with $\mathbf{b}$
— exactly the orthogonality that makes this the closest point.

```{r proj-vector}
a <- c(4, 2); b <- c(3, 4)
ahat <- (sum(a * b) / sum(b * b)) * b     # projection vector
leftover <- a - ahat                       # the perpendicular part
cat("projection:", ahat, "\n")             # 2.4  3.2
cat("leftover:", leftover, "\n")           # 1.6 -1.2
cat("check orthogonal:", sum(leftover * b), "\n")  # 0
```

### Projection onto a column space (least squares)

Now instead of projecting onto one vector, project $\mathbf{y}$ onto the **span** of
several columns — the column space $\text{Col}(A)$. The closest point
$\hat{\mathbf{y}} = A\hat{\mathbf{x}}$ satisfies: the residual
$\mathbf{e} = \mathbf{y} - A\hat{\mathbf{x}}$ is orthogonal to **every** column of $A$.
[@austin2022understandinglinearalgebra, pp. 397--398]

"Orthogonal to every column" means each column's dot product with $\mathbf{e}$ is 0.
Stack those dot products into a matrix multiplication and you get:

$$
A^\top \mathbf{e} = \mathbf{0}
\qquad\Longleftrightarrow\qquad
A^\top (\mathbf{y} - A\hat{\mathbf{x}}) = \mathbf{0}
$$

Distribute $A^\top$ and rearrange:

$$
A^\top A\,\hat{\mathbf{x}} = A^\top \mathbf{y}
\qquad\text{(the normal equations)}
$$

Reading it aloud: "A-transpose-A times x-hat equals A-transpose y." In everyday words:
"the coefficients that make the residual perpendicular to every column."
[@boyd2018appliedlinearalgebra, pp. 228--229]

When the columns of $A$ are linearly independent (full rank), $A^\top A$ is invertible
and the unique solution is:

$$
\hat{\mathbf{x}} = (A^\top A)^{-1}\,A^\top \mathbf{y}
$$

Mapping it onto the toy example:

- $A^\top A = \begin{bmatrix}3 & 6 \\ 6 & 14\end{bmatrix}$ — the "how do the columns
  relate to each other" matrix (warm-up Q2).
- $A^\top \mathbf{y} = \begin{bmatrix}1(1)+1(3)+1(2) \\ 1(1)+2(3)+3(2)\end{bmatrix} =
  \begin{bmatrix}6 \\ 13\end{bmatrix}$ — each column dotted with $\mathbf{y}$.
- Solve $\begin{bmatrix}3&6\\6&14\end{bmatrix}\hat{\mathbf{x}} =
  \begin{bmatrix}6\\13\end{bmatrix}$: multiply row 1 by 2 and subtract from row 2 →
  $2b = 1$, so $b = 0.5$; back-substitute → $3a = 6 - 3 = 3$, so $a = 1$.
- $\hat{\mathbf{x}} = (1, 0.5)$ — intercept 1, slope 0.5, exactly what `lm()` returns.

```{r normal-equations}
x <- c(1, 2, 3); y <- c(1, 3, 2)
A <- cbind(1, x)

AtA <- t(A) %*% A        # [3 6; 6 14]
Aty <- t(A) %*% y        # [6; 13]
xhat <- solve(AtA, Aty)  # solve the normal equations
cat("xhat:", xhat, "\n") # 1.0  0.5

yhat <- A %*% xhat
e <- y - yhat
cat("A^T e:", t(A) %*% e, "\n")  # 0  0  — orthogonal, as promised

cat("\nlm() agrees:\n")
print(coef(lm(y ~ x)))   # intercept = 1, slope = 0.5
```


## Explore it

```{webr}
#| autorun: true
#| fig-alt: "Scatter plot of 3 points with fitted line and vertical residual segments; residuals are perpendicular to the column space"
# explore-1 — LIVE: plot the data, fitted line, and residuals.
# TRY IT: change y to c(1, 2, 4) — how does the line move?
# PREDICT: will the residuals still sum to zero? Guess, then check.
x <- c(1, 2, 3); y <- c(1, 3, 2)
A <- cbind(1, x)
xhat <- solve(t(A) %*% A, t(A) %*% y)
yhat <- A %*% xhat
e <- y - yhat

plot(x, y, pch = 19, cex = 1.5, xlim = c(0.5, 3.5), ylim = c(0, 4),
     xlab = "hours", ylab = "score",
     main = sprintf("y = %.1f + %.1fx   residuals sum to %.1f",
                    xhat[1], xhat[2], sum(e)))
abline(xhat[1], xhat[2], col = "steelblue", lwd = 2)
segments(x, y, x, yhat, col = "tomato", lwd = 2, lty = 2)
points(x, yhat, pch = 4, col = "steelblue", cex = 1.5, lwd = 2)
legend("topleft", legend = c("data", "fitted", "residual"),
       pch = c(19, 4, NA), lty = c(NA, NA, 2),
       col = c("black", "steelblue", "tomato"), lwd = 2, bty = "n")
```

```{webr}
#| autorun: true
#| fig-alt: "Bar chart showing the dot product of each column of A with the residual vector is zero, confirming orthogonality"
# explore-2 — LIVE: verify the residual is orthogonal to EVERY column of A.
# TRY IT: add a third predictor column (say x^2) and check again.
# PREDICT: will column 3's dot product with e also be zero?
x <- c(1, 2, 3); y <- c(1, 3, 2)
A <- cbind(intercept = 1, hours = x)
xhat <- solve(t(A) %*% A, t(A) %*% y)
e <- y - A %*% xhat

dots <- as.numeric(t(A) %*% e)
barplot(dots, names.arg = colnames(A), col = "steelblue",
        main = "dot product of each column with residual",
        ylab = "column · e", ylim = c(-1, 1))
abline(h = 0, lty = 2)
cat("A^T e =", dots, "— both zero: residual is orthogonal to Col(A)\n")
cat("\nlm() agrees:", coef(lm(y ~ x)), "\n")
```

```{webr}
#| autorun: false
# explore-3 — LIVE: projection vector recap. The shadow as a VECTOR.
# TRY IT: change a to (0, 5) — what direction is the projection?
# PREDICT: for a = (3, -4) (orthogonal to b), what is the projection vector?
a <- c(4, 2); b <- c(3, 4)

proj <- (sum(a * b) / sum(b * b)) * b   # projection vector
leftover <- a - proj                     # perpendicular part

cat("projection of a onto b:", proj, "\n")
cat("leftover (perpendicular):", leftover, "\n")
cat("check: leftover dot b =", sum(leftover * b), "\n")
```


## Common traps

**Trap: "The projection drops a perpendicular to the fitted LINE."** The
perpendicular is not to the line *in the scatter plot* — those are the vertical
residuals, and they are not generally at right angles to the fitted line. The
orthogonality is in $n$-dimensional **observation space**: the residual vector
$\mathbf{e}$ (one entry per data point) is perpendicular to the columns of $A$ (also
one entry per data point). In our toy, $\mathbf{e} = (-0.5, 1, -0.5)$ lives in
$\mathbb{R}^3$; the perpendicularity is there, not in the 2D scatter plot. The
vertical segments in the plot show the *sizes* of the residuals, not their geometric
direction.
[@austin2022understandinglinearalgebra, pp. 397--398]

**Trap: "The normal equations are just algebra to memorise."** They are the
*orthogonality condition* $A^\top \mathbf{e} = \mathbf{0}$, rearranged. Memorising the
formula hides the geometry; remembering "the residual must be perpendicular to every
column" recovers the formula in one line: $A^\top(\mathbf{y} - A\hat{\mathbf{x}}) = 0$
→ distribute → $A^\top A\hat{\mathbf{x}} = A^\top \mathbf{y}$.
[@boyd2018appliedlinearalgebra, p. 231]


## Check yourself

**1.** For $\mathbf{a} = (6, 0)$ and $\mathbf{b} = (3, 4)$, compute the projection
*vector* of $\mathbf{a}$ onto $\mathbf{b}$. (Not the shadow length — the actual vector.)

```{webr}
#| autorun: false
# check1-starter
a <- c(6, 0); b <- c(3, 4)
```

::: {.callout-note collapse="true" title="Answer"}
$\hat{\mathbf{a}} = \frac{\mathbf{a}\cdot\mathbf{b}}{\mathbf{b}\cdot\mathbf{b}}
\mathbf{b} = \frac{18}{25}(3, 4) = (2.16, 2.88)$.
Check: leftover = $(3.84, -2.88)$; dot with $\mathbf{b}$:
$3.84 \times 3 + (-2.88) \times 4 = 11.52 - 11.52 = 0$. Orthogonal.

```{r}
a <- c(6, 0); b <- c(3, 4)
proj <- (sum(a * b) / sum(b * b)) * b
cat("projection:", proj, "\n")
cat("check:", sum((a - proj) * b), "\n")
```
:::

**2.** Using the toy data ($x = (1,2,3)$, $y = (1,3,2)$), what is the entry
$(A^\top \mathbf{y})_2$ — the second entry of $A^\top \mathbf{y}$? What does it
represent geometrically?

```{webr}
#| autorun: false
# check2-starter
x <- c(1, 2, 3); y <- c(1, 3, 2)
A <- cbind(1, x)
```

::: {.callout-note collapse="true" title="Answer"}
$(A^\top \mathbf{y})_2 = 1(1) + 2(3) + 3(2) = 1 + 6 + 6 = 13$. Geometrically, it is
the **dot product of column 2 of $A$** (the hours column $(1,2,3)$) with the response
$\mathbf{y}$. It measures how much $\mathbf{y}$ points along that column's direction —
the same "how much lies along" idea as the shadow from the distance module.
:::

**3.** In your own words: explain to a colleague, in two sentences, what the normal
equations say and why they work.

::: {.callout-note collapse="true" title="A model answer"}
The normal equations say: find the coefficients that make the residual perpendicular to
every predictor column. That works because perpendicular is the shortest path — any
other coefficients would leave a longer residual.
:::


## What you can now do

- **Compute a projection vector** — the shadow as a vector, not just a length — and
  verify it is orthogonal to the leftover.
- **Write down the normal equations** $A^\top A\hat{\mathbf{x}} = A^\top \mathbf{y}$
  from the orthogonality condition, and solve them by hand for a $2 \times 2$ system.
- **Explain why least squares IS projection** — the fitted values are the closest
  point in the column space, and the residual is the perpendicular.
- **Verify orthogonality in R** — `t(A) %*% e` should be a vector of zeros.

If any of these feels shaky, revisit [A tiny example first](#a-tiny-example-first) and
[The formal version](#the-formal-version) before continuing — anything that uses `lm()`
finds the same least-squares solution under the hood (via QR decomposition rather than
forming $A^\top A$ directly, for numerical stability).


## Where this goes next

The [multiple-regression course](../../courses/multiple-regression/syllabus.md) derives
and solves these normal equations for real data with many predictors — this module gives
you the geometric reason *why* those equations are the right ones to solve. The
[ridge-regularization module](../ridge-regularization/lesson.qmd) adds a penalty that
nudges the solution away from the plain least-squares answer, trading a small increase
in bias for a large decrease in variance. Ready for more reps? Do the
[practice problems](practice.qmd).
