---
title: "Practice: Orthogonality, projection & least squares"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Do them on paper or in an R session; only then
open the answer.


## Problem 1 — confidence rep with new numbers

Four observations:

| $x$ | 1 | 2 | 3 | 4 |
|---|---|---|---|---|
| $y$ | 2 | 5 | 4 | 7 |

(a) Write the $4 \times 2$ design matrix $A$ for fitting $y = a + bx$.
(b) Compute $A^\top A$ and $A^\top \mathbf{y}$.
(c) Solve the normal equations for $\hat{\mathbf{x}} = (a, b)$.
(d) Compute the residuals and verify $A^\top \mathbf{e} = \mathbf{0}$.

```{webr}
#| autorun: false
# p1-starter
x <- c(1, 2, 3, 4); y <- c(2, 5, 4, 7)
A <- cbind(1, x)
```

::: {.callout-note collapse="true" title="Answer"}
(a) $A = \begin{bmatrix}1&1\\1&2\\1&3\\1&4\end{bmatrix}$

(b)
```{r}
x <- c(1, 2, 3, 4); y <- c(2, 5, 4, 7)
A <- cbind(1, x)
AtA <- t(A) %*% A; Aty <- t(A) %*% y
cat("A^T A:\n"); print(AtA)   # [4 10; 10 30]
cat("A^T y:\n"); print(Aty)   # [18; 52]
```

(c) Solve $\begin{bmatrix}4&10\\10&30\end{bmatrix}\hat{\mathbf{x}} =
\begin{bmatrix}18\\52\end{bmatrix}$: multiply row 1 by 2.5 and subtract from row 2 →
$5b = 52 - 45 = 7$, so $b = 1.4$; back-substitute → $4a = 18 - 14 = 4$, so $a = 1$.

```{r}
xhat <- solve(AtA, Aty)
cat("intercept:", xhat[1], " slope:", xhat[2], "\n")  # 1  1.4
```

(d) Fitted values: $(2.4, 3.8, 5.2, 6.6)$. Residuals: $(-0.4, 1.2, -1.2, 0.4)$.

```{r}
yhat <- A %*% xhat; e <- y - yhat
cat("fitted:", yhat, "\n")
cat("residuals:", e, "\n")
cat("A^T e:", t(A) %*% e, "\n")  # 0  0
```

Both entries of $A^\top \mathbf{e}$ are zero — the residual is orthogonal to both
columns, confirming this is the least-squares solution.
:::


## Problem 2 — fill in the missing step

Here is the procedure for finding the least-squares line through
$(x, y) = (1,1), (2,3), (3,2)$:

- Step 1: Write the design matrix $A = \begin{bmatrix}1&1\\1&2\\1&3\end{bmatrix}$ and
  the response $\mathbf{y} = (1, 3, 2)$.
- Step 2: Compute $A^\top A = \begin{bmatrix}3&6\\6&14\end{bmatrix}$ and
  $A^\top \mathbf{y} = (6, 13)$.
- Step 3: **your step here** — what equation do you solve, and why does it give the
  best-fit coefficients?
- Step 4: Compute $\hat{\mathbf{y}} = A\hat{\mathbf{x}} = (1.5, 2, 2.5)$ and residuals
  $\mathbf{e} = (-0.5, 1, -0.5)$.
- Step 5: Verify $A^\top \mathbf{e} = \mathbf{0}$.

::: {.callout-note collapse="true" title="Answer"}
Step 3: **Solve the normal equations**
$A^\top A \hat{\mathbf{x}} = A^\top \mathbf{y}$ — that is,
$\begin{bmatrix}3&6\\6&14\end{bmatrix}\hat{\mathbf{x}} =
\begin{bmatrix}6\\13\end{bmatrix}$.

This gives $\hat{\mathbf{x}} = (1, 0.5)$: intercept $a = 1$, slope $b = 0.5$.

The normal equations are the orthogonality condition
$A^\top(\mathbf{y} - A\hat{\mathbf{x}}) = \mathbf{0}$ rearranged: they find the
coefficients that make the residual perpendicular to every column of $A$, which is
what makes $\hat{\mathbf{y}}$ the closest point in the column space to $\mathbf{y}$.
:::


## Problem 3 — projection vector

Compute the projection of $\mathbf{a} = (5, 1)$ onto $\mathbf{b} = (2, 6)$. Then
verify the leftover is orthogonal to $\mathbf{b}$.

```{webr}
#| autorun: false
# p3-starter
a <- c(5, 1); b <- c(2, 6)
```

::: {.callout-note collapse="true" title="Answer"}
$\hat{\mathbf{a}} = \frac{\mathbf{a} \cdot \mathbf{b}}{\mathbf{b} \cdot \mathbf{b}}
\mathbf{b} = \frac{10 + 6}{4 + 36}(2, 6) = \frac{16}{40}(2, 6) = 0.4 \times (2, 6)
= (0.8, 2.4)$.

Leftover: $(5 - 0.8, 1 - 2.4) = (4.2, -1.4)$.
Check: $4.2 \times 2 + (-1.4) \times 6 = 8.4 - 8.4 = 0$. Orthogonal.

```{r}
a <- c(5, 1); b <- c(2, 6)
proj <- (sum(a * b) / sum(b * b)) * b
leftover <- a - proj
cat("projection:", proj, "\n")
cat("leftover:", leftover, "\n")
cat("check:", sum(leftover * b), "\n")  # 0
```
:::


## Problem 4 — spot the error

A colleague says: *"To check whether my least-squares fit is correct, I plotted the
data and the fitted line, then measured the angle between each residual segment and
the fitted line. They're not 90°, so something went wrong."*

This contains **exactly one mistake** — find it.

::: {.callout-note collapse="true" title="Answer"}
The mistake: checking for **right angles in the scatter plot**. The orthogonality of
least squares is not between the vertical residual segments and the fitted line in the
2D plot. It is between the residual *vector* $\mathbf{e}$ (one entry per data point,
living in $n$-dimensional observation space) and each column of the design matrix $A$.
The correct check is $A^\top \mathbf{e} = \mathbf{0}$ — dot the residual vector with
each column and verify the result is zero. The vertical segments in a scatter plot show
residual *magnitudes*, not geometric directions.
:::


## Problem 5 — mixing it with rank *(uses rank's "independent columns" idea)*

You have two predictor columns: `hours` $= (1, 2, 3)$ and `double_hours` $= (2, 4, 6)$.

(a) What is the rank of $A = \begin{bmatrix}1&1&2\\1&2&4\\1&3&6\end{bmatrix}$
(intercept + hours + double_hours)?
(b) Can you solve the normal equations $A^\top A\hat{\mathbf{x}} = A^\top \mathbf{y}$?
Why or why not?
(c) What does `lm(y ~ hours + double_hours)` do in this situation?

```{webr}
#| autorun: false
# p5-starter
hours <- c(1, 2, 3); double_hours <- 2 * hours
A <- cbind(1, hours, double_hours)
```

::: {.callout-note collapse="true" title="Answer"}
(a) Rank = 2. Column 3 is exactly $2 \times$ column 2, so it is not independent — the
third column adds nothing new.

(b) No. $A^\top A$ is a $3 \times 3$ matrix but only rank 2, so it is singular —
`solve(t(A) %*% A)` fails. The normal equations need $A^\top A$ invertible, which
requires the columns of $A$ to be linearly independent.

(c) R's `lm()` detects the dependence and reports `NA` for the redundant coefficient:

```{r}
y <- c(1, 3, 2); hours <- c(1, 2, 3); double_hours <- 2 * hours
coef(lm(y ~ hours + double_hours))
```

It drops `double_hours` and fits the same line as `lm(y ~ hours)`. The rank module
warned that a dependent column does not enlarge the span — the column space of $A$ is
the same plane with or without `double_hours`.
:::


## Problem 6 — transfer: what R^2 means geometrically

The toy data has $\mathbf{y} = (1, 3, 2)$ and fitted $\hat{\mathbf{y}} = (1.5, 2, 2.5)$.

(a) Compute $\lVert \mathbf{e} \rVert^2$ (total squared residual).
(b) Compute $\lVert \mathbf{y} - \bar{y} \rVert^2$ (total variation of $\mathbf{y}$
around its mean $\bar{y} = 2$).
(c) Compute $R^2 = 1 - \lVert \mathbf{e} \rVert^2 / \lVert \mathbf{y} - \bar{y} \rVert^2$.
What fraction of the variation does the line explain?
[@austin2022understandinglinearalgebra, pp. 400, 412--413]

```{webr}
#| autorun: false
# p6-starter
y <- c(1, 3, 2); yhat <- c(1.5, 2, 2.5)
ybar <- mean(y)
```

::: {.callout-note collapse="true" title="Answer"}
(a) $\lVert \mathbf{e} \rVert^2 = (-0.5)^2 + 1^2 + (-0.5)^2 = 0.25 + 1 + 0.25 = 1.5$.

(b) $\mathbf{y} - \bar{y} = (1-2, 3-2, 2-2) = (-1, 1, 0)$.
$\lVert \mathbf{y} - \bar{y} \rVert^2 = 1 + 1 + 0 = 2$.

(c) $R^2 = 1 - 1.5/2 = 1 - 0.75 = 0.25$.

The line explains 25% of the variation in $\mathbf{y}$. The other 75% is in the
residual — the perpendicular part. Geometrically, $R^2$ is the fraction of
$\lVert \mathbf{y} - \bar{y} \rVert^2$ that lies in the column space direction.

```{r}
y <- c(1, 3, 2); yhat <- c(1.5, 2, 2.5); ybar <- mean(y)
SSE <- sum((y - yhat)^2)            # 1.5
SST <- sum((y - ybar)^2)            # 2
R2  <- 1 - SSE / SST                # 0.25
cat("SSE:", SSE, " SST:", SST, " R^2:", R2, "\n")
```
:::


## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/orthogonality-projection-and-least-squares" pass="4"}

::: {.quiz-q qid="opls-resid-ortho" qtype="mc" answer="b"}
What geometric property characterises the least-squares residual vector?

a) It points in the same direction as the longest column of the design matrix
b) It is orthogonal to every column of the design matrix
c) It has the same length as the response vector
:::

::: {.quiz-q qid="opls-normal-eqn" qtype="mc" answer="a"}
The normal equations $A^\top A \hat{\mathbf{x}} = A^\top \mathbf{y}$ are equivalent to which condition?

a) The residual is perpendicular to the column space of $A$
b) The fitted values equal the response values exactly for every observation
c) The columns of $A$ are mutually orthogonal to one another
:::

::: {.quiz-q qid="opls-proj-vec" qtype="mc" answer="c"}
The projection of $\mathbf{a}$ onto $\mathbf{b}$ is $(\mathbf{a} \cdot \mathbf{b} / \mathbf{b} \cdot \mathbf{b})\,\mathbf{b}$. What happens to the projection when $\mathbf{a}$ is orthogonal to $\mathbf{b}$?

a) The projection equals $\mathbf{a}$ itself because the vectors do not interact with each other
b) The projection equals $\mathbf{b}$ because the formula always returns a multiple of $\mathbf{b}$
c) The projection is the zero vector because the dot product in the numerator is zero
:::

::: {.quiz-q qid="opls-ata-toy" qtype="numeric" answer="14" tolerance="0"}
For the toy design matrix $A = \begin{bmatrix}1&1\\1&2\\1&3\end{bmatrix}$, what is the bottom-right entry of $A^\top A$?
:::

::: {.quiz-q qid="opls-rank-needed" qtype="mc" answer="b"}
(From rank.) Why does solving the normal equations require the columns of $A$ to be linearly independent?

a) Dependent columns make the residual vector longer, which increases the total error of fit
b) Dependent columns make $A^\top A$ singular, so `solve()` fails on the normal equations
c) Dependent columns make the response vector lie outside the column space of $A$
:::

:::

::::
