---
title: "Probability, odds, and log-odds"
subtitle: "Three ways to say the same thing about 'how likely' — and how to move between them"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
# Build-time setup only — the live cells below re-state their own data. Base R only.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

::: {.callout-tip title="This lesson runs in your browser" collapse="true"}
The code cells below with a **Run** button are live — powered by WebR (R compiled to
WebAssembly), they run right here on your phone or laptop, no install needed. Edit any
of them and re-run. The first run downloads a small R engine, so give it a few seconds.
Cells without a Run button are ordinary listings whose output was computed when the
page was built.
:::

## Where we're going

"How likely is it?" has three standard answers, and they're all the same information
in different clothes: a **probability** (a number from 0 to 1), the **odds** (a ratio
like "3 to 1"), and the **log-odds** (the natural log of the odds). Gamblers speak in
odds, forecasters speak in probabilities, and — as you'll see in courses that model
yes/no outcomes — statisticians love the log-odds because it's the version that ranges
over *all* numbers, which makes it play nicely with straight-line models. This module
makes all three comfortable and teaches you to convert between them in your sleep.

Methods that lean on this: logistic regression, risk and epidemiology (odds ratios,
relative risk), betting markets, and any model whose outcome is a yes/no event.

**Builds on:** the [natural log and $e^x$](../exponentials-and-logs/lesson.qmd) (log-odds
is *the log of* the odds), plus the baseline idea of probability.

## Warm-up: do you remember?

Two of these reach back to [exponentials-and-logs](../exponentials-and-logs/lesson.qmd);
one is a baseline reflex. Produce each answer before opening it.

**1.** *(baseline)* A fair coin has probability $0.5$ of heads. If you flip a bent coin
that lands heads $70\%$ of the time, what is its probability of heads as a decimal, and
what is its probability of *tails*?

::: {.callout-note collapse="true" title="Answer"}
Heads: $0.7$. Tails: $1 - 0.7 = 0.3$. The two must add to $1$, because the coin lands
*some* way every time — "probability of not-A is one minus probability of A" is the fact
this whole lesson leans on.
:::

**2.** *(exponentials-and-logs)* What is $\log(1)$, and is $\log(0.25)$ positive or
negative?

::: {.callout-note collapse="true" title="Answer"}
$\log(1) = 0$, and $\log(0.25)$ is **negative** (about $-1.386$) because $0.25$ is below
1. Shaky? Revisit [exponentials-and-logs](../exponentials-and-logs/lesson.qmd) — the sign
of a log will tell us which way a bet is leaning.
:::

**3.** *(exponentials-and-logs)* What does `exp()` do to `log(x)` — that is, what is
$e^{\log x}$?

::: {.callout-note collapse="true" title="Answer"}
It gives back $x$: exp and log are the undo pair, so $e^{\log x} = x$. We'll use this to
turn a log-odds back into ordinary odds. See
[exponentials-and-logs](../exponentials-and-logs/lesson.qmd).
:::

## A tiny example first

Suppose that, out of a long history, some event happens on $3$ days and doesn't happen
on $1$ day, for every 4 days — a probability of $p = 3/4 = 0.75$.

**Odds** ask a slightly different question than probability. Probability compares the
event to *all* outcomes ("3 out of 4 days"); **odds** compare the event to its
*opposite* ("3 days it happens for every 1 day it doesn't") — a ratio of $3$ to $1$,
which we write as the single number $3$. In symbols, odds $= p / (1-p)$:

$$
\text{odds} = \frac{p}{1-p} = \frac{0.75}{0.25} = 3.
$$

So "probability $0.75$" and "odds of 3" (or "3 to 1") are the same fact. Now take the
**log-odds** — literally the natural log of the odds:

$$
\text{log-odds} = \log(\text{odds}) = \log(3) \approx 1.099.
$$

A positive log-odds means the event is *more* likely than not (odds above 1); a
log-odds of exactly $0$ means a 50/50 event (odds of 1, probability $0.5$); a negative
log-odds means it's *less* likely than not.

And you can walk back the other way. Given odds, probability is $p = \text{odds} /
(1 + \text{odds}) = 3/4 = 0.75$ again. Given a log-odds, exponentiate first to recover
the odds ($e^{1.099} \approx 3$), then convert.

The same thing in R, so you can check every step:

```{r toy-example}
p <- 0.75

odds     <- p / (1 - p)          # 3 -- "3 to 1"
log_odds <- log(odds)            # ~1.099 -- the log of the odds
c(p = p, odds = odds, log_odds = log_odds)

# Walk back: odds -> probability, and log-odds -> odds -> probability
p_from_odds     <- odds / (1 + odds)             # 0.75 again
p_from_logodds  <- exp(log_odds) / (1 + exp(log_odds))  # exp() undoes the log first
c(p_from_odds = p_from_odds, p_from_logodds = p_from_logodds)
```

## Getting a feel for it

**A mental picture: a tug-of-war rope.** Put "it happens" on one side of a rope and "it
doesn't" on the other. **Probability** is the fraction of the *whole rope* your side
holds ($0.75$ = three-quarters of the rope). **Odds** is how your side compares to
*their* side — 3 units of rope to their 1, a ratio. When the sides are even (50/50),
probability is $0.5$ and odds are exactly $1$. As your side wins more rope, probability
crawls toward 1 (but can never pass it) while odds shoot up with no ceiling; as your
side loses, probability sinks toward 0 while odds shrink toward 0.

Where the picture bends: probability is bounded and symmetric around $0.5$; odds are
*not* symmetric — they run from 0 up to infinity, with the "even" point sitting at 1,
not in the middle. That lopsidedness is the reason we reach for the log next.

**Why bother with odds and log-odds at all?** Because of their *ranges*. A probability
is trapped in $[0, 1]$ — a cramped little box. Odds open the top of the box: they run
from $0$ to $\infty$. And taking the log opens the *bottom* too: **log-odds run over
every number, $-\infty$ to $+\infty$.** That matters enormously the moment you want to
model "how likely" with a straight line, because a straight line also runs over every
number — you can't cram a line into $[0,1]$ without it poking out, but it fits the
log-odds scale perfectly. Odds stretch the box open; the log makes it symmetric around
zero.

**What the numbers feel like.** Watch the same probabilities in all three scales:

| Probability $p$ | Odds $p/(1-p)$ | Log-odds $\log$ of odds |
|---|---|---|
| $0.5$ | $1$ | $0$ |
| $0.75$ | $3$ | $\approx 1.10$ |
| $0.9$ | $9$ | $\approx 2.20$ |
| $0.1$ | $0.111$ | $\approx -2.20$ |

Notice the pleasing symmetry the log buys you: $p = 0.9$ and $p = 0.1$ are mirror
images ($+2.20$ and $-2.20$), which they are *not* on the probability scale ($0.9$ and
$0.1$ aren't mirror images around anything convenient) nor on the odds scale ($9$ vs
$0.111$). A probability and "one minus it" always give log-odds that are negatives of
each other — a fact worth trusting.

## What the notation is going to say

The formulas below just name the three moves you did with the toy example: **odds
is probability over one-minus-probability; log-odds is the log of that; and you invert
by exponentiating and dividing.** When you read $\text{logit}(p) = \log\!\big(\tfrac{p}
{1-p}\big)$, hear "the logit of $p$ is the log of the odds of $p$" — it's a name for the
whole odds-then-log pipeline.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $p$ | "pee" | a probability: a number from 0 to 1 | $0.75$ |
| $1 - p$ | "one minus pee" | the probability of the *opposite* event | $0.25$ |
| $\dfrac{p}{1-p}$ | "pee over one-minus-pee" | the **odds**: event vs. its opposite | $3$ |
| $\log$ | "log" (natural log) | natural log, base $e$ (in R: `log`); undoes $e^x$ | $\log(3) \approx 1.099$ |
| $\operatorname{logit}(p)$ | "the LOH-jit of pee" | the **log-odds**: $\log\!\big(\tfrac{p}{1-p}\big)$ | $\approx 1.099$ |
| $\ell$ | "ell" | a log-odds *value* (a plain number on the log-odds scale), used when inverting back to $p$ | $1.099$ |
| $e^{x}$ | "ee to the ex" | the exponential (in R: `exp`); turns a log-odds back into odds | $e^{1.099} \approx 3$ |

## The formal version

The three quantities and the round trip between them:

$$
\text{odds} = \frac{p}{1-p}, \qquad
\operatorname{logit}(p) = \log\!\left(\frac{p}{1-p}\right).
$$

Reading it aloud: "odds equals pee over one-minus-pee" and "logit of pee equals log of
(pee over one-minus-pee)" — in everyday words: *odds is the event weighed against its
opposite, and the logit is the log of that.*

And the inverses — how to climb back to a probability:

$$
p = \frac{\text{odds}}{1 + \text{odds}}, \qquad
p = \frac{e^{\,\ell}}{1 + e^{\,\ell}} \quad (\ell = \text{log-odds}).
$$

- Left: solve $\text{odds} = \tfrac{p}{1-p}$ for $p$ (multiply out, collect the $p$'s) —
  ordinary algebra, the "solve a linear equation for the unknown" move.
- Right: since $\ell = \log(\text{odds})$, apply the undo pair $e^{\log(\cdot)} = \cdot$
  (taught in [exponentials-and-logs](../exponentials-and-logs/lesson.qmd)) to get
  $\text{odds} = e^{\ell}$, then reuse the left formula.

Mapping onto the toy example ($p = 0.75$):

- $\tfrac{p}{1-p}$ is "0.75 over 0.25" → odds $= 3$.
- $\log(3)$ → log-odds $\approx 1.099$.
- Back: $e^{1.099} \approx 3$, then $3/(1+3) = 0.75$. Full circle.

Sanity-check the whole round trip numerically:

```{r formal-check}
p <- 0.75

odds     <- p / (1 - p)          # 3
logit_p  <- log(odds)            # ~1.099

# Invert both ways -- must return 0.75
back_from_odds   <- odds / (1 + odds)
back_from_logit  <- exp(logit_p) / (1 + exp(logit_p))
c(odds = odds, logit = logit_p,
  back_from_odds = back_from_odds, back_from_logit = back_from_logit)
# 3, 1.099, 0.75, 0.75 -- the conversions undo each other exactly
```

## Explore it

```{webr}
#| autorun: true
#| fig-alt: "The logit curve: log-odds plotted against probability p, rising from large negative values near p = 0, crossing zero at p = 0.5, and shooting up toward large positive values near p = 1."
# explore-1 — LIVE: the logit stretches the cramped [0,1] probability scale onto the
# whole number line. PREDICT: at p = 0.5, what is the log-odds? And what happens as p
# approaches 0 or 1? Write your guesses, then run.
p <- seq(0.01, 0.99, by = 0.01)
logit <- log(p / (1 - p))
plot(p, logit, type = "l", lwd = 2, col = "steelblue",
     xlab = "probability p", ylab = "log-odds  log(p/(1-p))",
     main = "the logit: [0,1] stretched onto all real numbers")
abline(h = 0, v = 0.5, lty = 3)      # 50/50 sits at log-odds 0
```

```{webr}
#| autorun: false
# explore-2 — LIVE: probability, odds, and log-odds are the same fact three ways.
# PREDICT: for p = 0.8, will the odds be 4 and the log-odds be positive? Guess, then run.
p <- 0.8                              # <-- try 0.5, 0.1, 0.99
odds     <- p / (1 - p)
log_odds <- log(odds)
c(p = p, odds = odds, log_odds = log_odds)
```

```{webr}
#| autorun: false
# explore-3 — LIVE: the mirror-image fact. Compare a probability to "one minus it".
# PREDICT: will the log-odds of 0.9 and the log-odds of 0.1 be negatives of each other?
lo <- function(p) log(p / (1 - p))
c(logodds_0.9 = lo(0.9), logodds_0.1 = lo(0.1))   # equal and opposite?
```

## Common traps

**Trap: "odds and probability are the same number."** Tempting, because casual speech
blurs them ("the odds of rain are 75%"). But probability $0.75$ is **odds of 3**, not
odds of $0.75$. Check: a 50/50 event has probability $0.5$ but odds of $1$; a $0.9$
event has probability $0.9$ but odds of $9$. They only coincide near very small
probabilities. Always ask "compared to *all* outcomes (probability) or to the *opposite*
outcome (odds)?"

**Trap: "log-odds is stuck between 0 and 1 like a probability."** No — that's the whole
point of taking the log. Probability lives in $[0,1]$, odds live in $[0, \infty)$, and
log-odds live in **all** of $(-\infty, +\infty)$. A log-odds of $-2.20$ (that's $p =
0.1$) or $+2.20$ ($p = 0.9$) is perfectly ordinary. If you ever see a "log-odds" clamped
to $[0,1]$, something has gone wrong.

## Check yourself

**1.** An event has probability $p = 0.2$. What are its odds, and its log-odds? Is the
log-odds positive or negative, and why does that make sense?

```{webr}
#| autorun: false
# check1-starter — edit and Run to solve, right here
p <- 0.2
```

::: {.callout-note collapse="true" title="Answer"}
Odds $= 0.2 / 0.8 = 0.25$ (or "1 to 4"). Log-odds $= \log(0.25) \approx -1.386$.

```{r check1-answer}
p <- 0.2
c(odds = p / (1 - p), log_odds = log(p / (1 - p)))   # 0.25, -1.386
```
It's **negative** because the event is *less* likely than not ($p < 0.5$), so its odds
are below 1, and the log of a number below 1 is negative — exactly the sign fact from
[exponentials-and-logs](../exponentials-and-logs/lesson.qmd).
:::

**2.** A forecaster gives an event a log-odds of $0$. What is the probability? (No
calculator needed.)

::: {.callout-note collapse="true" title="Answer"}
$p = 0.5$. Log-odds $0$ means odds $= e^0 = 1$ (an even, "1 to 1" bet), and odds of $1$
means $p/(1-p) = 1$, i.e. $p = 1-p$, so $p = 0.5$. Zero log-odds is always dead-even.
:::

**3.** In your own words: explain to a colleague, in two sentences, what the odds of an
event are (versus its probability) and why statisticians reach for the log-odds.

::: {.callout-note collapse="true" title="A model answer"}
The odds of an event are how often it happens compared to how often it *doesn't* (a
ratio like "3 to 1"), whereas the probability compares it to *all* outcomes (a fraction
of 1). Statisticians use the log-odds because, unlike probability (stuck in $[0,1]$) or
odds (stuck above 0), it ranges over every number — so a straight-line model can predict
it without ever poking outside a valid range.
:::

## What you can now do

- **Convert a probability to odds** with $p/(1-p)$, and read odds back as a
  probability with $\text{odds}/(1+\text{odds})$.
- **Compute the log-odds** (logit) of a probability, and invert it (exponentiate, then
  convert) to recover $p$.
- **Read the sign of a log-odds**: positive → more likely than not, zero → 50/50,
  negative → less likely than not.
- **Explain the ranges** — why probability lives in $[0,1]$, odds in $[0,\infty)$, and
  log-odds over all real numbers — and why that last range is the useful one for
  line-based models.

If any of these feels shaky, revisit [Getting a feel for it](#getting-a-feel-for-it) —
the "three scales, one fact" table is the thing to have solid before any course uses it.

## Where this goes next

This is a shared building block. In the logistic-regression course you'll see that a
straight line is fit to the **log-odds** (never directly to the probability), precisely
because the log-odds is the scale a line can live on — and the sigmoid function is
exactly the "log-odds → probability" inversion you practised here, run in reverse. Ready
for more reps? Do the [practice problems](practice.qmd).
