---
title: "Practice: Probability, odds, and log-odds"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

Work these in order — they get harder. Use paper or the live cells; only then open the
answer. Reminders: odds $= p/(1-p)$, log-odds $= \log(\text{odds})$, and to invert,
$p = \text{odds}/(1+\text{odds})$.

## Problem 1 — the toy example, new numbers

An event has probability $p = 0.8$. Compute its odds and its log-odds by hand, then
check in R. State the "X to Y" reading of the odds.

```{webr}
#| autorun: false
# p1-starter — edit and Run to check your two answers
p <- 0.8
```

::: {.callout-note collapse="true" title="Answer"}
Odds $= 0.8 / 0.2 = 4$ — that's "**4 to 1**." Log-odds $= \log(4) \approx 1.386$
(positive, because the event is more likely than not).

```{r p1-answer}
p <- 0.8
c(odds = p / (1 - p), log_odds = log(p / (1 - p)))   # 4, 1.386
```
:::

## Problem 2 — fill in the missing step

We're converting **odds back to a probability**. The odds are $2$ (a "2 to 1" event).
Here's the work with the **core step blanked**:

> Step 1: odds $= \dfrac{p}{1 - p} = 2$
> Step 2: **your step here** — solve this equation for $p$
> Step 3: $p = \dfrac{2}{3} \approx 0.667$

Supply Step 2 (the algebra that gets from Step 1 to Step 3).

```{webr}
#| autorun: false
# p2-starter — check that odds = 2 gives p = 2/3
odds <- 2
```

::: {.callout-note collapse="true" title="Answer"}
Step 2: multiply both sides by $(1-p)$ to clear the fraction, then collect the $p$'s:
$2(1 - p) = p \Rightarrow 2 - 2p = p \Rightarrow 2 = 3p \Rightarrow p = 2/3$. In general
this rearrangement is the formula $p = \text{odds}/(1 + \text{odds})$.

```{r p2-answer}
odds <- 2
odds / (1 + odds)   # 0.6667
```
Odds of $2$ mean the event is twice as likely to happen as not — a probability of about
$0.667$, not $0.5$.
:::

## Problem 3 — the full round trip

Start from a log-odds value of $\ell = -1.5$. Recover the odds (undo the log), then the
probability. Then, as a check, feed that probability *forward* through odds and log to
confirm you get $-1.5$ back.

```{webr}
#| autorun: false
# p3-starter — invert the log-odds, then re-derive it as a check
ell <- -1.5
```

::: {.callout-note collapse="true" title="Answer"}
```{r p3-answer}
ell  <- -1.5
odds <- exp(ell)                 # exp undoes log -> odds ~ 0.223
p    <- odds / (1 + odds)        # odds -> probability ~ 0.182
c(odds = odds, p = p)

# Forward check: p -> odds -> log-odds should return -1.5
log(p / (1 - p))                 # -1.5, full circle
```
The log-odds $-1.5$ corresponds to odds $\approx 0.223$ and probability $\approx 0.182$
(a fairly unlikely event, as the negative log-odds warned). Running it forward returns
exactly $-1.5$ — the conversions are genuine inverses.
:::

## Problem 4 — spot the error

A colleague wants the odds of a probability $p = 0.6$ and writes the following. It
contains **exactly one mistake** — find it and fix it.

> "The odds are just the probability, so odds $= 0.6$. Then the log-odds is $\log(0.6)
> \approx -0.511$, which is negative — so the event is *less* likely than not."

```{webr}
#| autorun: false
# p4-starter — compute the CORRECT odds and log-odds for p = 0.6
p <- 0.6
```

::: {.callout-note collapse="true" title="Answer"}
The mistake is "**the odds are just the probability**." Odds are $p/(1-p)$, not $p$. For
$p = 0.6$ the odds are $0.6/0.4 = 1.5$, and the log-odds is $\log(1.5) \approx +0.405$ —
**positive**, correctly signalling the event is *more* likely than not (as it should be,
since $p > 0.5$).

```{r p4-answer}
p <- 0.6
c(wrong_odds = p, right_odds = p / (1 - p),
  right_log_odds = log(p / (1 - p)))   # 0.6 (wrong) vs 1.5, +0.405
```
The colleague's error even flipped the *conclusion*: they got a negative log-odds for an
event that's actually more likely than not. Confusing odds with probability is the
classic slip.
:::

## Problem 5 — mixing it with logs *(uses exponentials-and-logs' log rules)*

Two events have odds $o_1 = 6$ and $o_2 = 2$. Their **log-odds difference** is
$\log(o_1) - \log(o_2)$. Using a log rule (not a calculator), rewrite that difference as
a *single* log of one number, then confirm in R. What single number's log is it?

```{webr}
#| autorun: false
# p5-starter — check that log(6) - log(2) collapses to one log
o1 <- 6; o2 <- 2
```

::: {.callout-note collapse="true" title="Answer"}
By the quotient rule (from
[exponentials-and-logs](../exponentials-and-logs/lesson.qmd)), $\log(o_1) - \log(o_2) =
\log(o_1 / o_2) = \log(6/2) = \log(3) \approx 1.099$. So the difference of two log-odds is
the log of the **ratio of the odds** — that ratio ($6/2 = 3$) is exactly what gets called
an *odds ratio* in applied work.

```{r p5-answer}
o1 <- 6; o2 <- 2
c(difference = log(o1) - log(o2), single_log = log(o1 / o2))   # both ~ 1.099
```
This "a difference of log-odds is the log of an odds ratio" fact is the seed of how
logistic-regression coefficients get read — you'll meet it in that course.
:::

## Problem 6 — interpretation

Two medical tests are described. Test A: the odds of a true positive are $9$. Test B:
the odds are $1.5$. Without converting to probabilities, which test is more likely to
give a true positive, and roughly how do their probabilities compare (both above $0.5$,
both below, one each)? Explain in a sentence or two.

::: {.callout-note collapse="true" title="Answer"}
Test A is much more likely to give a true positive — odds of $9$ (nine hits per miss)
dwarf odds of $1.5$ (three hits per two misses). Both odds exceed $1$, so **both**
probabilities are above $0.5$: Test A's is $9/10 = 0.9$ and Test B's is $1.5/2.5 = 0.6$.
The lesson: any odds above 1 means "more likely than not," and the bigger the odds, the
closer the probability creeps toward 1 (without ever reaching it).
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/probability-and-odds" pass="4"}

::: {.quiz-q qid="po-prob-to-odds" qtype="numeric" answer="4" tolerance="0"}
A probability of $0.8$ corresponds to what odds? Use $p/(1-p)$.
:::

::: {.quiz-q qid="po-odds-to-prob" qtype="numeric" answer="0.667" tolerance="0.001"}
An event has odds of $2$. What is its probability? Use $\text{odds}/(1+\text{odds})$.
Round to three decimals.
:::

::: {.quiz-q qid="po-odds-not-prob" qtype="mc" answer="a"}
Someone says "the odds of rain are 75%." What is the confusion?

a) A probability of $0.75$ is odds of $3$ (that is, $0.75/0.25$), not odds of $0.75$
b) Nothing — a probability of $0.75$ is the same number as odds of $0.75$
c) Odds expressed as "75%" means a probability of $0.43$
:::

::: {.quiz-q qid="po-logodds-range" qtype="mc" answer="b"}
A log-odds (logit) value of $-2.20$ is reported. Is it valid, and what does the sign say?

a) Invalid — log-odds must lie in $[0, 1]$ like a probability
b) Valid — log-odds range over all real numbers; negative means less likely than not
c) Valid, but a negative log-odds means the probability itself is negative
:::

::: {.quiz-q qid="po-invert-logit" qtype="mc" answer="c"}
(From exponentials & logs.) The undo pair says $e^{\log y} = y$. To recover the odds from
a log-odds value $\ell$, which operation do you apply?

a) Take the log again: $\text{odds} = \log \ell$
b) Negate it: $\text{odds} = -\ell$
c) Exponentiate: $\text{odds} = e^{\ell}$
:::

:::

::::
