---
title: "The gradient: downhill in more than one direction"
subtitle: "When a curve has two knobs, the slope becomes a little vector of slopes"
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```


## Where we're going

Last module we rolled downhill on a curve with a single knob $x$. Real problems have
*several* knobs — a line to fit has an intercept *and* a slope, a model can have dozens
of parameters. You already know from the foundation what the gradient is (the vector of
partial slopes) and how to measure it numerically. This module puts it to work: we take
the gradient you built in the foundation and turn it into a **descent step** — the same
"step against the slope" idea from Module 1, now applied to every knob at once.

**Builds on:** [rolling downhill (Module 1)](../01-rolling-downhill/lesson.qmd),
[partial derivatives & the gradient](../../../../foundations/partial-derivatives-and-the-gradient/lesson.qmd),
and [vectors & summation](../../../../foundations/vectors-and-summation/lesson.qmd).

## Warm-up: do you remember?

Three quick recalls. Produce each before opening it.

**1.** *(Module 1)* Write the one-dimensional gradient-descent update rule, and say why
it has a minus sign.

::: {.callout-note collapse="true" title="Answer"}
$x_{t+1} = x_t - \alpha f'(x_t)$. The minus is there because the slope points *uphill*,
and we want to go *downhill* — the opposite way. See
[Module 1](../01-rolling-downhill/lesson.qmd).
:::

**2.** *(partial-derivatives-and-the-gradient)* For $f(a,b) = a^2 + b^2$ at the point
$(3, 2)$, what is the gradient $\nabla f$, and which direction does it point?

::: {.callout-note collapse="true" title="Answer"}
$\nabla f(3,2) = (6, 4)$ — the vector of partial slopes. It points in the
steepest-*uphill* direction; its negative $(-6, -4)$ points steepest-downhill. See
[partial derivatives & the gradient](../../../../foundations/partial-derivatives-and-the-gradient/lesson.qmd).
:::

**3.** *(partial-derivatives-and-the-gradient)* How do you measure the gradient
numerically in R? What does the tool do, in one sentence?

::: {.callout-note collapse="true" title="Answer"}
Run one central-difference `slope_at()` per coordinate (nudge only that coordinate,
hold the rest fixed), and collect the results into a vector with `sapply`. See
[partial derivatives & the gradient](../../../../foundations/partial-derivatives-and-the-gradient/lesson.qmd).
:::

## A tiny example first

Here's a bowl over *two* inputs — think of a real valley you could walk around in, with
an east–west direction $a$ and a north–south direction $b$:

$$
f(a, b) = (a - 2)^2 + (b - 5)^2
$$

Its lowest point is at $a = 2, b = 5$ (both squared terms zero). Stand at the point
$a = 0, b = 0$ and ask: *which way is downhill?* Now there are two questions, one per
direction:

- **Slope in the $a$ direction** (nudge $a$, hold $b = 0$): the curve $(a-2)^2 + 25$ has
  slope $2(a - 2) = 2(0 - 2) = -4$ at $a = 0$. Negative → downhill is toward larger $a$.
- **Slope in the $b$ direction** (nudge $b$, hold $a = 0$): the curve $4 + (b-5)^2$ has
  slope $2(b - 5) = 2(0 - 5) = -10$ at $b = 0$. Negative → downhill is toward larger $b$.

Bundle those two slopes into a vector — that's the **gradient** at $(0,0)$: $(-4, -10)$.
It says "to go uphill, head in direction $(-4,-10)$", so to go *downhill* we step the
opposite way. One gradient-descent step with $\alpha = 0.1$ nudges *both* knobs at once:

$$
a \leftarrow 0 - 0.1(-4) = 0.4, \qquad b \leftarrow 0 - 0.1(-10) = 1.0
$$

We moved from $(0,0)$ toward $(0.4, 1.0)$ — heading for $(2, 5)$ on both axes at once.

The same thing in R, measuring each slope numerically so no calculus is needed:

```{r toy-example}
# The two-input bowl; f takes a vector v = c(a, b)
f <- function(v) (v[1] - 2)^2 + (v[2] - 5)^2

# Slope in direction i: nudge ONLY v[i] up and down, rise over run
partial <- function(f, v, i, h = 1e-5) {
  up <- v; dn <- v
  up[i] <- up[i] + h; dn[i] <- dn[i] - h
  (f(up) - f(dn)) / (2 * h)
}
# The gradient: one partial slope per input, bundled into a vector
gradient <- function(f, v) sapply(seq_along(v), function(i) partial(f, v, i))

v <- c(0, 0)
gradient(f, v)            # about (-4, -10) — the two slopes at (0,0)
v - 0.1 * gradient(f, v)  # one downhill step -> about (0.4, 1.0)
```

## Getting a feel for it

**The gradient as your compass, from Module 1 to many knobs.** You already know from the
[foundation](../../../../foundations/partial-derivatives-and-the-gradient/lesson.qmd)
that the gradient is an arrow pointing steepest-uphill, and its negative points steepest-
downhill. Module 1 said "step against the slope" with one knob. Now the slope is a
*vector* of slopes (the gradient), and "step against it" means stepping against the
whole vector at once, nudging every knob simultaneously. Each descent step reads the
gradient arrow and takes a scaled hop in the opposite direction. (The gradient is only
accurate *right where you stand*; a step away the surface has turned, so you must
re-measure — which is why it's a loop, not one leap.)

**What the steps feel like.** At $(0,0)$ our gradient was $(-4, -10)$: the $b$
direction is steeper (magnitude 10) than the $a$ direction (magnitude 4), so the first
steps move $b$ faster than $a$. As each knob nears its target the matching slope shrinks
toward zero, so that knob's steps taper — the same self-slowing you saw in one dimension,
now happening independently on each axis. At the very bottom $(2,5)$ **both** slopes are
zero: the gradient is the zero vector, and descent stops.

## What the notation is going to say

The formulas below say: "the gradient of $f$, written $\nabla f$, is the vector whose
entries are the slope in the $a$ direction and the slope in the $b$ direction; and one
gradient-descent step subtracts the learning rate times that whole vector from your
current position." It's Module 1's rule with vectors in place of single numbers — the
$-4$ and the $-10$ handled together.

## How to read the symbols

| Symbol | How to say it | What it means | In our toy example |
|---|---|---|---|
| $f(a, b)$ | "eff of ay, bee" | a function of two inputs — a surface with a knob for each | $(a-2)^2 + (b-5)^2$ |
| $\mathbf{v}$ | "vee" | the position, held as a vector of the inputs (in R: `c(a, b)`) | starts at `c(0, 0)` |
| $\dfrac{\partial f}{\partial a}$ | "the partial of eff with respect to ay" | the slope in the $a$ direction, holding $b$ fixed | $-4$ at $(0,0)$ |
| $\dfrac{\partial f}{\partial b}$ | "the partial of eff with respect to bee" | the slope in the $b$ direction, holding $a$ fixed | $-10$ at $(0,0)$ |
| $\nabla f$ | "grad eff" (or "del eff") | the **gradient**: the vector of all the partial slopes | $(-4, -10)$ at $(0,0)$ |
| $\alpha$ | "alpha" | the learning rate (same as Module 1) | $0.1$ |
| $\mathbf{v}_t$ | "vee sub tee" | the current position after $t$ steps | starts at `c(0, 0)` |
| $\mathbf{v}_{t+1}$ | "vee sub tee plus one" | the position after the next step | $(0.4, 1.0)$ |

(The $\partial$ ("partial") is just a $d$ for "a slope, but only along this one direction
while the others hold still". The $\nabla$ ("grad") gathers those slopes into one arrow.)

## The formal version

The gradient is the vector of partial slopes:

$$
\nabla f(a, b) = \left( \frac{\partial f}{\partial a}, \; \frac{\partial f}{\partial b} \right)
$$

Reading it aloud: "grad eff equals the vector of: the partial of eff with respect to ay,
and the partial of eff with respect to bee" — or plainly: "list the slope in each
direction." At $(0,0)$ that's $(-4, -10)$, the two numbers we measured.

And one step of gradient descent, now in vector form:

$$
\mathbf{v}_{t+1} = \mathbf{v}_t - \alpha \, \nabla f(\mathbf{v}_t)
$$

Reading it aloud: "vee sub tee-plus-one equals vee sub tee minus alpha times grad eff at
vee sub tee" — *your next position is your current position, stepped against the gradient
arrow by a fraction alpha.* It's identical to Module 1's $x_{t+1} = x_t - \alpha f'(x_t)$
— only now $\mathbf{v}$ and $\nabla f$ are vectors, so the single subtraction updates
*both* knobs at once. Check the first step numerically:

```{r formal-check}
f <- function(v) (v[1] - 2)^2 + (v[2] - 5)^2
partial  <- function(f, v, i, h = 1e-5) {
  up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h
  (f(up) - f(dn)) / (2 * h)
}
gradient <- function(f, v) sapply(seq_along(v), function(i) partial(f, v, i))

descend <- function(f, v, lr, n_steps) {
  for (t in 1:n_steps) v <- v - lr * gradient(f, v)   # the vector update, repeated
  v
}
descend(f, v = c(0, 0), lr = 0.1, n_steps = 100)   # ~ (2, 5), the bottom
```

## Explore it

```{webr}
#| autorun: true
#| fig-alt: "A contour map of the two-input bowl with the gradient-descent path drawn as connected points spiraling in toward the minimum at (2,5)."
# explore-1 — LIVE: a contour map (like a topographic map) of the bowl, with the descent
# path drawn on top. Each ring is a height; the path walks across rings toward the center.
# PREDICT: from (0,0) with lr = 0.1, roughly how many steps until the path reaches the
# center at (2,5)? Guess, then run. Try changing the start or lr.
f <- function(v) (v[1] - 2)^2 + (v[2] - 5)^2
partial  <- function(f, v, i, h = 1e-5) {
  up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
}
gradient <- function(f, v) sapply(seq_along(v), function(i) partial(f, v, i))

v <- c(0, 0); lr <- 0.1; n <- 25
path <- matrix(NA, n + 1, 2); path[1, ] <- v
for (t in 1:n) { v <- v - lr * gradient(f, v); path[t + 1, ] <- v }

a <- seq(-1, 5, length.out = 60); b <- seq(-1, 8, length.out = 60)
z <- outer(a, b, function(A, B) (A - 2)^2 + (B - 5)^2)
contour(a, b, z, xlab = "a", ylab = "b", main = sprintf("after %d steps: (%.2f, %.2f)", n, v[1], v[2]))
lines(path[, 1], path[, 2], col = "firebrick", lwd = 2)
points(path[, 1], path[, 2], pch = 16, col = "firebrick", cex = 0.7)
points(2, 5, pch = 3, cex = 2, lwd = 2)   # the true minimum
```

```{webr}
#| autorun: false
#| fig-alt: "Contour map of a stretched bowl where the descent path zig-zags toward the minimum."
# explore-2 — LIVE: a STRETCHED bowl, 10x steeper in b than in a. Watch the path
# zig-zag instead of heading straight for the center. PREDICT: will the negative
# gradient point straight AT the minimum here? Run and see why not (Trap 2 below).
f <- function(v) (v[1] - 2)^2 + 10 * (v[2] - 5)^2
partial  <- function(f, v, i, h = 1e-5) {
  up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
}
gradient <- function(f, v) sapply(seq_along(v), function(i) partial(f, v, i))

v <- c(-1, 8); lr <- 0.08; n <- 40
path <- matrix(NA, n + 1, 2); path[1, ] <- v
for (t in 1:n) { v <- v - lr * gradient(f, v); path[t + 1, ] <- v }
a <- seq(-2, 5, length.out = 60); b <- seq(2, 8, length.out = 60)
z <- outer(a, b, function(A, B) (A - 2)^2 + 10 * (B - 5)^2)
contour(a, b, z, xlab = "a", ylab = "b", main = "stretched bowl: the path zig-zags")
lines(path[, 1], path[, 2], col = "firebrick", lwd = 2); points(2, 5, pch = 3, cex = 2, lwd = 2)
```

## Common traps

**Trap: "tune one knob all the way, then tune the other."** It's tempting to minimize
over $a$ first (holding $b$ fixed), then over $b$. Gradient descent does *not* do this —
it nudges **both at once**, a little each step, using the current slopes. Tuning one
fully then the other can take many more passes and, on tilted surfaces, get stuck
ping-ponging. The gradient's whole point is to move in all directions *simultaneously*
in the right proportions.

**Trap: "to measure the slope in the $a$ direction, nudge both $a$ and $b$."** A
partial slope holds *every other input fixed* and nudges only the one you're asking
about. Nudging all inputs at once measures a diagonal slope — a mix of all the
partials added together — not the slope in any single direction. In code: `f(v + h)`
adds $h$ to every coordinate; the correct version changes only `v[i]`.

**Trap: "the negative gradient points straight at the minimum."** On our first, round
bowl it very nearly does — but that's luck of symmetry. On the *stretched* bowl in
explore-2, the steepest-downhill direction is not aimed at the center, so the path
zig-zags in toward it rather than making a beeline. The gradient only promises the
steepest direction *locally*, right where you stand; it makes no promise about pointing
at the far-off bottom. (This is why step size and surface shape matter so much in
practice.)

## Check yourself

**1.** For $f(a,b) = (a-1)^2 + (b+3)^2$ (bottom at $a=1, b=-3$), compute the gradient at
$(0, 0)$ numerically, then take one descent step with $\alpha = 0.1$.

```{webr}
#| autorun: false
# check1-starter — edit and Run to solve, right here
f <- function(v) (v[1] - 1)^2 + (v[2] + 3)^2
partial  <- function(f, v, i, h = 1e-5) {
  up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
}
gradient <- function(f, v) sapply(seq_along(v), function(i) partial(f, v, i))
```

::: {.callout-note collapse="true" title="Answer"}
The slopes at $(0,0)$: in $a$, $2(0-1) = -2$; in $b$, $2(0+3) = +6$. So
$\nabla f(0,0) = (-2, 6)$. One step: $(0,0) - 0.1(-2, 6) = (0.2, -0.6)$ — moving toward
$a=1$ (right) and toward $b=-3$ (down), as it should.

```{r check1-answer}
f <- function(v) (v[1] - 1)^2 + (v[2] + 3)^2
partial  <- function(f, v, i, h = 1e-5) {
  up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
}
gradient <- function(f, v) sapply(seq_along(v), function(i) partial(f, v, i))
g <- gradient(f, c(0, 0)); g            # (-2, 6)
c(0, 0) - 0.1 * g                       # (0.2, -0.6)
```
:::

**2.** At the bottom of any bowl, what is the gradient vector, and how does descent
"know" to stop there?

::: {.callout-note collapse="true" title="Answer"}
At the bottom every partial slope is zero, so the gradient is the **zero vector**
$(0, 0)$. The update $\mathbf{v} - \alpha \cdot \mathbf{0} = \mathbf{v}$ leaves the
position unchanged — descent stops moving because there's no downhill left in any
direction.
:::

**3.** In your own words: explain to a colleague, in two sentences, what the gradient is
and how gradient descent uses it when there's more than one input.

::: {.callout-note collapse="true" title="A model answer"}
The gradient is the bundle of slopes — one per input — that together point in the
steepest-uphill direction on the surface. Gradient descent measures it at the current
spot and steps the opposite way (downhill) in all directions at once, scaled by the
learning rate, repeating until every slope is zero.
:::

## What you can now do

- **Measure a partial slope**: the steepness in one input's direction, by nudging only
  that input.
- **Build the gradient** of a multi-input function numerically — the vector of partial
  slopes — and read it as a steepest-uphill arrow.
- **Run gradient descent in vector form**, updating all inputs at once with
  $\mathbf{v} \leftarrow \mathbf{v} - \alpha \nabla f(\mathbf{v})$.
- **Explain why descent stops at the bottom** (the gradient is zero) and why paths can
  zig-zag on stretched surfaces.

If any of these feels shaky, revisit [Getting a feel for it](#getting-a-feel-for-it) and
[Explore it](#explore-it) — the next module puts this exact machinery to work on real
data.

## Where this goes next

Now we point this at something real: fitting a straight line to data by gradient
descent — the loss becomes an error-over-two-knobs (intercept and slope), and we watch
our from-scratch descent land on the *same* answer as a professional optimizer:
[fitting a line by descent](../03-fitting-a-line-by-descent/lesson.qmd). Ready for more
reps first? Do the [practice problems](practice.qmd).
