---
title: "Practice: The gradient"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order. Handy helpers to paste into any starter:

```r
partial  <- function(f, v, i, h = 1e-5) {
  up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
}
gradient <- function(f, v) sapply(seq_along(v), function(i) partial(f, v, i))
```

## Problem 1 — the toy example, new numbers

For $f(a,b) = (a - 3)^2 + (b - 1)^2$ (bottom at $a=3, b=1$), compute the gradient at
$(0, 0)$ by hand (the slope in each direction is $2(\text{coord} - \text{target})$), then
take one descent step with $\alpha = 0.1$. Check in R.

```{webr}
#| autorun: false
# p1-starter
f <- function(v) (v[1] - 3)^2 + (v[2] - 1)^2
partial  <- function(f, v, i, h = 1e-5) {
  up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
}
gradient <- function(f, v) sapply(seq_along(v), function(i) partial(f, v, i))
```

::: {.callout-note collapse="true" title="Answer"}
Slopes at $(0,0)$: in $a$, $2(0-3) = -6$; in $b$, $2(0-1) = -2$. So gradient $=(-6,-2)$.
One step: $(0,0) - 0.1(-6,-2) = (0.6, 0.2)$ — moving toward $(3,1)$ on both axes.

```{r p1-answer}
f <- function(v) (v[1] - 3)^2 + (v[2] - 1)^2
partial  <- function(f, v, i, h = 1e-5) {
  up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
}
gradient <- function(f, v) sapply(seq_along(v), function(i) partial(f, v, i))
g <- gradient(f, c(0, 0)); g       # (-6, -2)
c(0, 0) - 0.1 * g                  # (0.6, 0.2)
```
:::

## Problem 2 — fill in the missing step

A vector gradient-descent loop is written with its **update line blanked**. Supply it.

> ```
> v <- c(0, 0)
> for (t in 1:100) {
>   g <- gradient(f, v)
>   your step here          # update the whole vector v using g and lr
> }
> ```

```{webr}
#| autorun: false
# p2-starter — fill in the update line, then Run
f <- function(v) (v[1] - 3)^2 + (v[2] - 1)^2
partial  <- function(f, v, i, h = 1e-5) {
  up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
}
gradient <- function(f, v) sapply(seq_along(v), function(i) partial(f, v, i))
v <- c(0, 0); lr <- 0.1
```

::: {.callout-note collapse="true" title="Answer"}
The blanked line is the vector update — subtract the learning rate times the whole
gradient: `v <- v - lr * g`. Because `v` and `g` are both length-2 vectors, this one
line updates *both* knobs at once.

```{r p2-answer}
f <- function(v) (v[1] - 3)^2 + (v[2] - 1)^2
partial  <- function(f, v, i, h = 1e-5) {
  up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
}
gradient <- function(f, v) sapply(seq_along(v), function(i) partial(f, v, i))
v <- c(0, 0); lr <- 0.1
for (t in 1:100) { g <- gradient(f, v); v <- v - lr * g }
v   # ~ (3, 1)
```
:::

## Problem 3 — descend to the bottom

Minimize $f(a,b) = (a + 2)^2 + (b - 4)^2$ (bottom at $a=-2, b=4$) from the start
$(5, 5)$, $\alpha = 0.1$, 150 steps. Report the final $(a, b)$.

```{webr}
#| autorun: false
# p3-starter
f <- function(v) (v[1] + 2)^2 + (v[2] - 4)^2
partial  <- function(f, v, i, h = 1e-5) {
  up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
}
gradient <- function(f, v) sapply(seq_along(v), function(i) partial(f, v, i))
```

::: {.callout-note collapse="true" title="Answer"}
```{r p3-answer}
f <- function(v) (v[1] + 2)^2 + (v[2] - 4)^2
partial  <- function(f, v, i, h = 1e-5) {
  up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
}
gradient <- function(f, v) sapply(seq_along(v), function(i) partial(f, v, i))
v <- c(5, 5); lr <- 0.1
for (t in 1:150) v <- v - lr * gradient(f, v)
v   # ~ (-2, 4)
```
It converges to about $(-2, 4)$. Note $a$ started far away (at 5, target $-2$) and $b$
started close (at 5, target 4), so $a$ took bigger early steps — each knob moves at a
pace set by its own slope.
:::

## Problem 4 — spot the error

A colleague's "partial slope in direction $i$" nudges the wrong thing. Their code has
**exactly one mistake** — find it and fix it.

```{webr}
#| autorun: false
# p4-starter — this partial() nudges every coordinate, not just coordinate i. Fix it.
f <- function(v) (v[1] - 3)^2 + (v[2] - 1)^2
partial <- function(f, v, i, h = 1e-5) {
  (f(v + h) - f(v - h)) / (2 * h)     # nudges ALL of v, not only v[i]
}
partial(f, c(0, 0), 1)   # should be -6 (slope in a only); gives something else
```

::: {.callout-note collapse="true" title="Answer"}
The mistake: `f(v + h)` adds $h$ to **every** coordinate at once, so it measures a
diagonal slope, not the slope in direction $i$ alone. A partial must nudge *only* the
$i$-th coordinate, holding the rest fixed:

```{r p4-answer}
f <- function(v) (v[1] - 3)^2 + (v[2] - 1)^2
partial <- function(f, v, i, h = 1e-5) {
  up <- v; dn <- v
  up[i] <- up[i] + h; dn[i] <- dn[i] - h   # change ONLY coordinate i
  (f(up) - f(dn)) / (2 * h)
}
partial(f, c(0, 0), 1)   # -6, the slope in the a-direction, as it should be
```
:::

## Problem 5 — two knobs that don't talk *(uses Module 1's 1-D descent)*

The bowl $f(a,b) = (a - 3)^2 + (b - 1)^2$ splits cleanly: the $a$ part doesn't involve
$b$ and vice versa. *Predict*, using Module 1 thinking, what 2-D descent will find — then
confirm the two coordinates behave exactly like two independent 1-D descents.

```{webr}
#| autorun: false
# p5-starter — compare 2-D descent to two separate 1-D descents
f     <- function(v) (v[1] - 3)^2 + (v[2] - 1)^2
fa    <- function(x) (x - 3)^2      # the a-slice alone (a 1-D bowl, Module 1)
fb    <- function(x) (x - 1)^2      # the b-slice alone
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
partial  <- function(f, v, i, h = 1e-5) {
  up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
}
gradient <- function(f, v) sapply(seq_along(v), function(i) partial(f, v, i))
```

::: {.callout-note collapse="true" title="Answer"}
Prediction: since the knobs don't interact, the $a$-coordinate should behave like a
Module-1 descent on $(a-3)^2$ (landing at 3) and the $b$-coordinate like a descent on
$(b-1)^2$ (landing at 1) — completely independently. Confirm:

```{r p5-answer}
f     <- function(v) (v[1] - 3)^2 + (v[2] - 1)^2
fa    <- function(x) (x - 3)^2
fb    <- function(x) (x - 1)^2
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
partial  <- function(f, v, i, h = 1e-5) {
  up <- v; dn <- v; up[i] <- up[i] + h; dn[i] <- dn[i] - h; (f(up) - f(dn)) / (2 * h)
}
gradient <- function(f, v) sapply(seq_along(v), function(i) partial(f, v, i))

# 1-D descents (Module 1) on each slice
xa <- 0; xb <- 0
for (t in 1:100) { xa <- xa - 0.1 * slope_at(fa, xa); xb <- xb - 0.1 * slope_at(fb, xb) }

# 2-D descent, same start and rate
v <- c(0, 0)
for (t in 1:100) v <- v - 0.1 * gradient(f, v)

rbind(one_d = c(xa, xb), two_d = v)   # identical: (3, 1)
```
When the inputs don't interact, the gradient's coordinates are just separate 1-D slopes,
so 2-D descent *is* two Module-1 descents running side by side.
:::

## Problem 6 — a gentle cliffhanger

To fit a straight line $\hat y = b_0 + b_1 x$ to data, the two knobs are the intercept
$b_0$ and the slope $b_1$, and the thing to minimize is the total squared prediction
error as those knobs vary — a bowl over $(b_0, b_1)$. Given everything in this module,
what would gradient descent *do* to fit the line, in one sentence?

::: {.callout-note collapse="true" title="Answer"}
It would treat $(b_0, b_1)$ as the position $\mathbf v$, measure the gradient of the
squared-error surface (how the error changes as you nudge the intercept, and as you nudge
the slope), and step against it repeatedly until both slopes are zero — i.e. roll to the
bottom of the error bowl. That bottom is the best-fitting line, and
[the next module](../03-fitting-a-line-by-descent/lesson.qmd) runs it — then checks the
answer against a professional optimizer.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="courses/gradient-descent/modules/02-the-gradient" pass="4"}

::: {.quiz-q qid="tg-partial-at-point" qtype="numeric" answer="-2" tolerance="0"}
**On the lesson's bowl $f(a, b) = (a-2)^2 + (b-5)^2$, what is the slope in the $a$
direction at the point $(a, b) = (1, 2)$?** (The lesson's slope rule for this bowl:
the $a$-direction slope is $2(a-2)$.)
:::

::: {.quiz-q qid="tg-stop-condition" qtype="mc" answer="b"}
**At the bottom of any smooth bowl, why does gradient descent stop moving?**

a) The function value reaches zero there, which ends the loop
b) The gradient is zero there, so the step changes nothing
c) The learning rate has decayed to zero by the time it arrives
:::

::: {.quiz-q qid="tg-both-knobs" qtype="mc" answer="a"}
**A function has two knobs, $a$ and $b$. How does one gradient-descent step move them?**

a) Both at once, each nudged against its own partial slope
b) It tunes $a$ all the way to its best value first, then does the same for $b$
c) It moves only the knob whose slope is steeper this step
:::

::: {.quiz-q qid="tg-zigzag" qtype="mc" answer="c"}
**On the stretched bowl of explore-2, does the negative gradient $-\nabla f$ point
straight at the minimum?**

a) Yes — the steepest-downhill direction always aims at the lowest point
b) No — the gradient points uphill, so the path drifts away from the minimum
c) No — it is the steepest direction only *right where you stand*, so the path zig-zags inward
:::

::: {.quiz-q qid="tg-sum-recall" qtype="numeric" answer="20" tolerance="0"}
**(From the foundation *vectors & summation*.) For $v = (4, 7, 9)$, what is
$\sum_{i=1}^{3} v_i$?**
:::

:::

::::
