---
title: "Practice: Differentiation rules"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Do them on paper or in the live cells; only
then open the answer.

## Problem 1 — the toy example, new numbers

In the lesson we found the derivative of $f(x) = 3x^2 + 5x - 7$ was $f'(x) = 6x + 5$.
Now differentiate $g(x) = 4x^3 - 2x^2 + x - 10$ by hand (use constant-multiple, power,
sum/difference, and constant rules). Then check your answer numerically at $x = 2$.

```{webr}
#| autorun: false
# p1-starter — edit and Run to check your hand calculation
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
g <- function(x) 4*x^3 - 2*x^2 + x - 10
x0 <- 2
```

::: {.callout-note collapse="true" title="Answer"}
Apply the rules term by term:

- $4x^3$: constant-multiple × power rule → $4 \cdot 3x^2 = 12x^2$
- $-2x^2$: constant-multiple × power rule → $-2 \cdot 2x = -4x$
- $x$: power rule ($x = x^1$, exponent 1 → $1 \cdot x^0 = 1$) → $1$
- $-10$: constant rule → $0$

So $g'(x) = 12x^2 - 4x + 1$. At $x = 2$: $12(4) - 4(2) + 1 = 48 - 8 + 1 = 41$.

```{r p1-answer}
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
g <- function(x) 4*x^3 - 2*x^2 + x - 10
slope_at(g, 2)   # 41 — matches 12*(4) - 4*(2) + 1
```
:::

---

## Problem 2 — fill in the missing step

Here is a worked derivation of the slope of $f(x) = (3x - 2)^2$ at $x = 3$, using the
chain rule. One step is blanked out — supply it.

> **Step 1 — identify outer and inner:**
> Outer function: $(\cdot)^2$; inner function: $g(x) = 3x - 2$.
>
> **Step 2 — outer derivative:**
> Derivative of $u^2$ is $2u$; evaluated at inner: $2(3x - 2)$.
>
> **Step 3 — inner derivative:**
> `<your step here>`: compute $g'(x)$ = derivative of $3x - 2$.
>
> **Step 4 — chain rule: outer × inner:**
> $f'(x) = 2(3x - 2) \cdot g'(x)$.
>
> **Step 5 — evaluate at $x = 3$:**
> Inner value: $3(3) - 2 = 7$. So $f'(3) = 2(7) \cdot g'(x)$.

```{webr}
#| autorun: false
# p2-starter — edit and Run to check the blank
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
f <- function(x) (3*x - 2)^2
x0 <- 3
```

::: {.callout-note collapse="true" title="Answer"}
The missing step is the **inner derivative**: $g(x) = 3x - 2$, so $g'(x) = 3$
(constant-multiple × power rule: $3 \cdot 1 \cdot x^0 = 3$).

Completing the chain rule: $f'(x) = 2(3x - 2) \cdot 3 = 6(3x - 2)$.
At $x = 3$: $6(3 \cdot 3 - 2) = 6 \cdot 7 = 42$.

```{r p2-answer}
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
f <- function(x) (3*x - 2)^2
slope_at(f, 3)   # 42 — matches 6*(3*3 - 2) = 6*7 = 42
```
:::

---

## Problem 3 — full problem: applying all the rules

The function $h(x) = 2(x - 1)^3 + 4x - 5$ looks complex but it falls apart cleanly.
Differentiate it by hand, then evaluate $h'(x)$ at $x = 2$.

- The $2(x - 1)^3$ term needs chain rule (outer: $(\cdot)^3$; inner: $x - 1$) plus
  constant-multiple.
- The remaining terms need power and constant rules.

Check numerically.

```{webr}
#| autorun: false
# p3-starter — edit and Run to check your answer
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
h_fn <- function(x) 2*(x - 1)^3 + 4*x - 5
x0 <- 2
```

::: {.callout-note collapse="true" title="Answer"}
Break into terms:

**Term $2(x-1)^3$:**
- Chain rule: outer $(\cdot)^3$ → outer derivative $3(\cdot)^2$; evaluated at inner
  $x-1$: $3(x-1)^2$
- Inner $x - 1$ → inner derivative $1$
- Chain rule product: $3(x-1)^2 \cdot 1 = 3(x-1)^2$
- Constant-multiple $2$: $2 \cdot 3(x-1)^2 = 6(x-1)^2$

**Term $4x$:** → $4$

**Term $-5$:** → $0$ (constant rule)

So $h'(x) = 6(x-1)^2 + 4$. At $x = 2$: $6(1)^2 + 4 = 6 + 4 = 10$.

```{r p3-answer}
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
h_fn <- function(x) 2*(x - 1)^3 + 4*x - 5
slope_at(h_fn, 2)   # 10 — matches 6*(2-1)^2 + 4 = 6 + 4
```
:::

---

## Problem 4 — spot the error

A student differentiates $f(x) = (5x + 2)^2$ and gets $f'(x) = 2(5x + 2)$. Their work
is shown below — it contains **exactly one mistake**. Find it and write the correct
answer. Then check numerically at $x = 1$.

> "Outer function is $(\cdot)^2$, so the outer derivative is $2(\cdot)^1 = 2(5x + 2)$.
> Inner function is $5x + 2$, but its derivative is just 1 because the coefficient
> doesn't matter. So $f'(x) = 2(5x + 2) \cdot 1 = 2(5x + 2)$."

```{webr}
#| autorun: false
# p4-starter — run to see the true slope, then identify the error above
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
f <- function(x) (5*x + 2)^2
x0 <- 1
```

::: {.callout-note collapse="true" title="Answer"}
The mistake is in the **inner derivative**. The inner function is $g(x) = 5x + 2$.
Its derivative is $g'(x) = 5$ (not 1) — the constant-multiple rule gives $5 \cdot 1
= 5$ for the $5x$ term. The coefficient of $x$ in the inner function absolutely does
matter.

Correct chain rule: $f'(x) = 2(5x + 2) \cdot 5 = 10(5x + 2)$.
At $x = 1$: $10(5 + 2) = 10 \cdot 7 = 70$.

```{r p4-answer}
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
f <- function(x) (5*x + 2)^2

wrong   <- 2 * (5*1 + 2) * 1   # student's answer: 14
correct <- 2 * (5*1 + 2) * 5   # true: 70
measured <- slope_at(f, 1)      # 70 — confirms the correct version

c(wrong = wrong, correct = correct, measured = measured)
```

This is exactly the trap from Common traps: forgetting the inner derivative makes the
answer off by a factor of $g'(x)$ (here, off by 5).
:::

---

## Problem 5 — interleaved *(uses derivative-as-slope's numerical tool)*

*(Uses [derivative-as-slope](../derivative-as-slope/lesson.qmd))*

You have the function $f(x) = x^4 - 8x^2 + 7$.

**Part A:** Use the rules from this module to find $f'(x)$ algebraically.

**Part B:** Use `slope_at()` from the derivative-as-slope module to measure the slope
at $x = -2$, $x = 0$, and $x = 2$. Do the numerical measurements agree with your
algebraic derivative? What do you notice about the slopes at these three points — and
what does that tell you about the shape of the function?

```{webr}
#| autorun: false
# p5-starter — Part A: work out f'(x) on paper first, then check here
# Part B: measure the slopes and compare
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
f <- function(x) x^4 - 8*x^2 + 7
xs <- c(-2, 0, 2)
```

::: {.callout-note collapse="true" title="Answer"}
**Part A:** Apply sum rule + constant-multiple + power rule term by term:
- $x^4$ → $4x^3$
- $-8x^2$ → $-8 \cdot 2x = -16x$
- $7$ → $0$

So $f'(x) = 4x^3 - 16x$.

Evaluate by hand: $f'(-2) = 4(-8) - 16(-2) = -32 + 32 = 0$; $f'(0) = 0$;
$f'(2) = 4(8) - 16(2) = 32 - 32 = 0$.

```{r p5-answer}
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
f <- function(x) x^4 - 8*x^2 + 7

xs      <- c(-2, 0, 2)
numeric <- sapply(xs, function(x) slope_at(f, x))   # all ~0
rule    <- 4*xs^3 - 16*xs                            # all exactly 0

rbind(by_rule = rule, numerical = round(numeric, 6))
```

**Part B interpretation:** All three slopes are (essentially) zero, which means $x =
-2$, $x = 0$, and $x = 2$ are all flat spots — local minima or maxima. The function has
a "W" shape with two valleys and a peak. The derivative tells you *where* those turning
points are; you'd need to check second derivatives (or just plot it) to distinguish
the valleys from the peak.
:::

---

## Problem 6 — transfer: the OLS pattern

*(Preview — a tool from the next derivation you'll encounter)*

The ordinary least-squares loss for fitting a line $y = a + bx$ to data involves
minimising a sum of squared gaps. For each data point, the squared residual has the
form $(y_i - a - b x_i)^2$. With $y_i$, $a$, and $x_i$ treated as fixed constants,
this is a function of the single knob $b$.

**Your task:** write down $\frac{d}{db}(y_i - a - b x_i)^2$ using the chain rule.

Treat $y_i - a - b x_i$ as the inner function (a linear function of $b$), and
$(\cdot)^2$ as the outer. The derivative will involve $x_i$ (the $x_i$ in the inner
function).

You don't have enough machinery yet to finish the whole OLS derivation — that's the
next course's job. But the derivative of this one squared term is the key move.

```{webr}
#| autorun: false
# p6-starter — verify your chain-rule answer numerically for one concrete case:
# set y_i = 5, a = 1, x_i = 2, then measure d/db of (y_i - a - b*x_i)^2 at b = 1
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
y_i <- 5; a <- 1; x_i <- 2
f_of_b <- function(b) (y_i - a - b * x_i)^2
b0 <- 1
```

::: {.callout-note collapse="true" title="Answer"}
**By chain rule:**
- Outer: $(\cdot)^2$; outer derivative: $2(\cdot)$; evaluated at inner: $2(y_i - a - b x_i)$
- Inner: $y_i - a - b x_i$; derivative with respect to $b$: $-x_i$ (the constant $-x_i$
  multiplies $b^1$, so the derivative is $-x_i$; the constant terms $y_i - a$ vanish)
- Chain rule product: $2(y_i - a - b x_i) \cdot (-x_i) = -2 x_i (y_i - a - b x_i)$

At the concrete values ($y_i = 5$, $a = 1$, $x_i = 2$, $b = 1$):
inner $= 5 - 1 - 1 \cdot 2 = 2$; derivative $= -2 \cdot 2 \cdot 2 = -8$.

```{r p6-answer}
slope_at <- function(f, x, h = 1e-5) (f(x + h) - f(x - h)) / (2 * h)
y_i <- 5; a <- 1; x_i <- 2
f_of_b <- function(b) (y_i - a - b * x_i)^2

by_rule    <- -2 * x_i * (y_i - a - 1 * x_i)   # at b = 1: -2*2*2 = -8
by_numeric <- slope_at(f_of_b, 1)               # should also be -8

c(rule = by_rule, numeric = by_numeric)
```

When you sum these $-2 x_i (y_i - a - b x_i)$ terms over all data points and set the
sum to zero, that's the equation whose solution gives the OLS slope estimate.
The chain rule just made the hardest step of that derivation.
:::

---

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/differentiation-rules" pass="4"}

::: {.quiz-q qid="dr-power-rule-apply" qtype="numeric" answer="12" tolerance="0"}
By the power rule, $\frac{d}{dx}(x^3) = 3x^2$. What is $3x^2$ evaluated at $x = 2$?
:::

::: {.quiz-q qid="dr-constant-rule" qtype="mc" answer="c"}
You differentiate $f(x) = 8$ and get $f'(x) = ?$

a) $8$
b) $8x$
c) $0$
:::

::: {.quiz-q qid="dr-sum-rule-eval" qtype="numeric" answer="17" tolerance="0"}
Using the sum, constant-multiple, and power rules, what is the derivative of
$3x^2 + 5x - 7$ evaluated at $x = 2$?
:::

::: {.quiz-q qid="dr-chain-inner" qtype="mc" answer="b"}
When applying the chain rule to $f(x) = (4x - 1)^3$, what is the inner derivative
$g'(x)$ where $g(x) = 4x - 1$?

a) $1$
b) $4$
c) $3$
:::

::: {.quiz-q qid="dr-das-central-slope" qtype="numeric" answer="6" tolerance="0"}
*(Uses [derivative-as-slope](../derivative-as-slope/lesson.qmd))* Using the power rule,
the derivative of $x^2$ is $2x$. What does the central-difference formula
$(f(x+h) - f(x-h))/(2h)$ give for $f(x) = x^2$ at $x = 3$ when you measure it
numerically? (Recall this lesson used this fact to verify the power rule.)
:::

:::

::::
