---
title: "Practice: The mean, variance & covariance"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| message: false
# Packages this practice set uses. Missing ones are installed automatically.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

{{< include /includes/practice-timing.qmd >}}

Work these in order — they get harder. Do them on paper or in an R session; only
then open the answer.

## Problem 1 — the toy example, new numbers

Four mushrooms weigh $x = (3, 1, 4, 2)$ grams. By hand: find $\bar{x}$, the four
deviations, and $s_x^2$. Check with `mean()` and `var()`.

```{webr}
#| autorun: false
# p1-starter — Starter: copy this to begin
x <- c(3, 1, 4, 2)
```

::: {.callout-note collapse="true" title="Answer"}
Mean: $3 + 1 + 4 + 2 = 10$, divided by 4 → $\bar{x} = 2.5$.

Deviations: $3 - 2.5 = 0.5$; $1 - 2.5 = -1.5$; $4 - 2.5 = 1.5$; $2 - 2.5 = -0.5$.
(Sanity check: $0.5 - 1.5 + 1.5 - 0.5 = 0$ ✓.)

Squared: $0.25, 2.25, 2.25, 0.25$. Sum: $5$. Divide by $n - 1 = 3$:
$s_x^2 = 5/3 \approx 1.67$.

```{r p1-answer}
x <- c(3, 1, 4, 2)
mean(x)          # 2.5
x - mean(x)      # 0.5 -1.5  1.5 -0.5
var(x)           # 1.6667 = 5/3
```
:::

## Problem 2 — faded worked example (covariance by hand)

Three plots got $x = (1, 3, 5)$ units of fertilizer and yielded $y = (2, 2, 8)$
crates. Below is the covariance worked out — but the **core step is blanked**. Fill
in ONLY the blanked line (the paired-deviation products and their sum), then finish
the last line.

> $\bar{x} = 3$, $\bar{y} = 4$.
> Deviations: $x - \bar{x} = (-2,\, 0,\, 2)$ and $y - \bar{y} = (-2,\, -2,\, 4)$.
> **`<your step here>`** — multiply each *pair* of deviations, then add the products:
> $\sum (x_i - \bar{x})(y_i - \bar{y}) = \underline{\phantom{00}}$
> Finally divide by $n - 1 = 2$: $\;s_{xy} = \underline{\phantom{00}}$.

```{webr}
#| autorun: false
# p2-starter — Starter: copy this to check your two blanks
x <- c(1, 3, 5)
y <- c(2, 2, 8)
```

::: {.callout-note collapse="true" title="Answer"}
The blanked step is the heart of covariance — pair up each plot's two deviations and
multiply: $(-2)(-2) = 4$, $(0)(-2) = 0$, $(2)(4) = 8$. Sum: $4 + 0 + 8 = 12$.

Divide by $n - 1 = 2$: $s_{xy} = 12 / 2 = 6$. Positive — more-fertilized plots tend
to be the higher-yielding ones.

```{r p2-answer}
x <- c(1, 3, 5); y <- c(2, 2, 8)
n <- length(x)
sum((x - mean(x)) * (y - mean(y))) / (n - 1)   # 6
cov(x, y)                                       # 6 — same
```
:::

## Problem 3 — greek vs. latin

A paper writes: "heights are distributed with mean $\mu = 170$ and variance
$\sigma^2 = 81$; in our sample of $n = 40$ we observed $\bar{x} = 172.1$ and
$s^2 = 76.4$." In plain English: what is the difference between $\mu$ and $\bar{x}$
here, and between $\sigma^2$ and $s^2$? Which pair would change if they re-ran the
study on 40 new people?

::: {.callout-note collapse="true" title="Answer"}
$\mu$ and $\sigma^2$ (Greek) are the *true* mean and variance of the whole
population — fixed but unknown quantities the paper is assuming or estimating.
$\bar{x}$ and $s^2$ (Latin) are *computed from the 40 sampled people* — estimates of
$\mu$ and $\sigma^2$.

A new sample of 40 people would give new values of $\bar{x}$ and $s^2$ (they bounce
around from sample to sample), while $\mu$ and $\sigma^2$ would stay exactly what
they were. The Greek-vs-Latin letter code is how papers silently tell you which kind
of quantity each symbol is.
:::

## Problem 4 — sign of the covariance, no arithmetic

For each pair, say whether the covariance should be positive, near zero, or
negative — *before* computing anything:

a. hours studied & exam score
b. car weight & fuel efficiency (miles per gallon)
c. shoe size & R programming skill

Then verify (b) in R with the built-in `mtcars` data (`wt` and `mpg`).

```{webr}
#| autorun: false
# p4-starter — check your prediction for (b)
```

::: {.callout-note collapse="true" title="Answer"}
a. Positive — above-average studiers tend to score above average (deviations share
   a sign, products positive).
b. Negative — heavier-than-average cars tend to have below-average mpg.
c. Near zero — no reason for the deviations to align.

```{r p4-answer}
cov(mtcars$wt, mtcars$mpg)   # about -5.1: negative, as predicted
cor(mtcars$wt, mtcars$mpg)   # about -0.87: and strongly so
```

Note how the correlation (unit-free, in [−1, 1]) is much easier to interpret than
the raw covariance (whose units are tons·mpg!).
:::

## Problem 5 — spot the error

A colleague computes the variance of $x = (4, 8, 12)$ like this. It contains
**exactly one mistake** — find it, then give the correct $s_x^2$.

> "Mean is $(4 + 8 + 12)/3 = 8$. Deviations $-4, 0, 4$; squared $16, 0, 16$; they
> sum to $32$. There are $n = 3$ values, so the variance is $32 / 3 \approx 10.67$."

```{webr}
#| autorun: false
# p5-starter — Starter: copy this to check the corrected answer
x <- c(4, 8, 12)
```

::: {.callout-note collapse="true" title="Answer"}
Everything is right up to the last step. The mistake is dividing by $n = 3$ instead
of $n - 1 = 2$: **sample** variance uses $n - 1$. The correct value is
$32 / 2 = 16$, not $10.67$. (Dividing by $n$ would be the *population* variance,
which assumes you have the entire population, not a sample.)

```{r p5-answer}
x <- c(4, 8, 12)
n <- length(x)
sum((x - mean(x))^2) / (n - 1)   # 16 — the correct sample variance
var(x)                            # 16 — R agrees (it uses n - 1)
```
:::

## Problem 6 — interleaved *(uses vectors & summation's sum rules)*

For $x = (5, 10, 15)$: expand $\sum_{i=1}^{3}(x_i - \bar{x})^2$ step by step
using sum rules from [vectors & summation](../vectors-and-summation/lesson.qmd).
First find $\bar{x}$, then write out each $(x_i - \bar{x})^2$ term, sum them, and
finally divide by $n - 1$ to get $s_x^2$. Check with `var()`.

```{webr}
#| autorun: false
# p6-interleaved — Starter
x <- c(5, 10, 15)
```

::: {.callout-note collapse="true" title="Answer"}
$\bar{x} = (5 + 10 + 15)/3 = 10$.

Deviations: $5 - 10 = -5$, $10 - 10 = 0$, $15 - 10 = 5$.

Expanding the sum term by term (writing out $\sum_{i=1}^{3}$ as three addends):
$$\sum_{i=1}^{3}(x_i - \bar{x})^2 = (-5)^2 + (0)^2 + (5)^2 = 25 + 0 + 25 = 50$$

Divide by $n - 1 = 2$: $s_x^2 = 50 / 2 = 25$.

```{r p6-interleaved-answer}
x <- c(5, 10, 15)
sum((x - mean(x))^2) / (length(x) - 1)   # 25
var(x)                                    # 25 — same
```
:::

## Problem 7 — the formula IS the function (in R)

Starter below. Compute the covariance of `x` and `y` **without** `cov()` — build it
straight from the formula $s_{xy} = \frac{1}{n-1}\sum (x_i - \bar{x})(y_i - \bar{y})$
— then confirm against `cov(x, y)`.

```{webr}
#| autorun: false
# p7-starter — Starter data: your job is covariance by hand, then check with cov()
set.seed(42)
x <- rnorm(500, mean = 10, sd = 2)
y <- 3 * x + rnorm(500, sd = 5)
```

::: {.callout-note collapse="true" title="Answer"}
```{r p7-answer}
set.seed(42)
x <- rnorm(500, mean = 10, sd = 2)
y <- 3 * x + rnorm(500, sd = 5)
n <- length(x)

# Straight from the formula:
# each x-deviation times its paired y-deviation, summed, divided by n - 1
by_hand <- sum((x - mean(x)) * (y - mean(y))) / (n - 1)

c(by_hand = by_hand, built_in = cov(x, y))   # identical
```

If you can write this line, you can *read* the covariance formula in any paper —
they are the same statement in two languages.
:::

## Problem 8 — name that formula

Each line below describes a computation. For each one, name the summary it computes:
**mean**, **variance**, **standard deviation**, **covariance**, or **correlation**.
No calculation needed — just pattern-match the recipe to the right name.

a. "Add all the values, divide by $n$."
b. "For each observation, subtract the mean, square the result, add those squared
   numbers up, divide by $n - 1$."
c. "For each observation, multiply that observation's x-deviation by its y-deviation,
   add those products, divide by $n - 1$."
d. "Compute (c), then divide the result by the standard deviation of x times the
   standard deviation of y."
e. "Compute (b), then take the square root."

::: {.callout-note collapse="true" title="Answer"}
a. **Mean** ($\bar{x} = \frac{1}{n}\sum x_i$) — add and divide by count.

b. **Variance** ($s_x^2 = \frac{1}{n-1}\sum (x_i - \bar{x})^2$) — squared
   deviations, averaged with $n - 1$.

c. **Covariance** ($s_{xy} = \frac{1}{n-1}\sum (x_i - \bar{x})(y_i - \bar{y})$) —
   the key difference from variance is that you multiply *paired* deviations from
   two variables instead of squaring one variable's deviations.

d. **Correlation** ($r = s_{xy} / (s_x \cdot s_y)$) — covariance divided by both
   standard deviations, which cancels the units and forces the result into [−1, 1].

e. **Standard deviation** ($s_x = \sqrt{s_x^2}$) — the square root of variance,
   back in the original units so you can feel it as a distance.

The two easiest to mix up are variance and covariance: both divide by $n - 1$, but
variance *squares* one variable's deviations while covariance *multiplies* two
variables' deviations. If you see $(x_i - \bar{x})^2$, it's variance; if you see
$(x_i - \bar{x})(y_i - \bar{y})$, it's covariance.
:::

## Problem 9 — all five summaries, one dataset

Three days of temperature ($x$, in Celsius) and ice-cream sales ($y$, in dozens):
$x = (2, 5, 8)$, $y = (6, 4, 5)$. Compute all five summaries by hand: $\bar{x}$,
$\bar{y}$, $s_x^2$, $s_{xy}$, and $r$. (You will need $s_x$ and $s_y$ along the
way.) Then check every one in R.

This is the exercise that proves you can tell the formulas apart when they sit next
to each other — same data, five different recipes.

```{webr}
#| autorun: false
# p9-starter — copy this to begin
x <- c(2, 5, 8)
y <- c(6, 4, 5)
```

::: {.callout-note collapse="true" title="Answer"}
**Means.** $\bar{x} = (2 + 5 + 8)/3 = 15/3 = 5$.
$\bar{y} = (6 + 4 + 5)/3 = 15/3 = 5$.

**Deviations.** $x - \bar{x}$: $-3, 0, 3$. $\;y - \bar{y}$: $1, -1, 0$.

**Variance of x.** Square the x-deviations: $9, 0, 9$. Sum: $18$. Divide by
$n - 1 = 2$: $s_x^2 = 9$. So $s_x = \sqrt{9} = 3$.

**Variance of y.** Square the y-deviations: $1, 1, 0$. Sum: $2$. Divide by $2$:
$s_y^2 = 1$. So $s_y = 1$.

**Covariance.** Multiply *paired* deviations (this is where it differs from
variance — paired products, not squares): $(-3)(1) = -3$, $(0)(-1) = 0$,
$(3)(0) = 0$. Sum: $-3$. Divide by $2$: $s_{xy} = -1.5$. Negative — warmer days
do not consistently bring more sales in this tiny dataset.

**Correlation.** $r = s_{xy}/(s_x \cdot s_y) = -1.5/(3 \times 1) = -0.5$.

```{r p9-answer}
x <- c(2, 5, 8); y <- c(6, 4, 5)

mean(x)          # 5
mean(y)          # 5
var(x)           # 9
var(y)           # 1
sd(x)            # 3
sd(y)            # 1
cov(x, y)        # -1.5
cor(x, y)        # -0.5
```

The discrimination test: variance squared one variable's deviations ($9, 0, 9$);
covariance multiplied two variables' deviations ($-3, 0, 0$). Same data, different
operation, different answer.
:::

## Problem 10 — read the R, name the summary

Each R expression below computes one of the five summaries (mean, variance, SD,
covariance, correlation). Name which summary each line computes, and say which
single built-in R function would give the same answer.

a. `sum(x) / length(x)`
b. `sum((x - mean(x))^2) / (length(x) - 1)`
c. `sum((x - mean(x)) * (y - mean(y))) / (length(x) - 1)`
d. `sqrt(sum((x - mean(x))^2) / (length(x) - 1))`
e. `cov(x, y) / (sd(x) * sd(y))`

::: {.callout-note collapse="true" title="Answer"}
a. **Mean** — adds all x values, divides by how many. Built-in: `mean(x)`.

b. **Variance** — squared deviations, summed, divided by $n - 1$. Built-in:
   `var(x)`. The giveaway is `^2` applied to one variable's deviations.

c. **Covariance** — paired deviations multiplied (`*` between two variables'
   deviations), summed, divided by $n - 1$. Built-in: `cov(x, y)`. The giveaway
   is `(x - mean(x)) * (y - mean(y))` — two different variables, multiplied, not
   squared.

d. **Standard deviation** — the same thing as (b), wrapped in `sqrt()`. Built-in:
   `sd(x)`. Whenever you see `sqrt(variance)`, it's the SD.

e. **Correlation** — covariance divided by both standard deviations. Built-in:
   `cor(x, y)`. The division by `sd(x) * sd(y)` cancels the units and squeezes the
   result into [−1, 1].

The pair to watch: (b) vs (c). The *only* difference is `^2` (squaring one
variable's deviations = variance) vs `* (y - mean(y))` (multiplying two
variables' deviations = covariance). That one substitution is the entire difference
between "how spread out is x" and "do x and y move together."
:::

## Problem 11 — spot the wrong formula

A student is asked to compute the **covariance** of $x = (3, 6, 9)$ and
$y = (2, 5, 2)$. Their work contains **exactly one mistake** — find it and give the
correct answer.

> "Means: $\bar{x} = 6$, $\bar{y} = 3$. Deviations: $x - \bar{x} = (-3, 0, 3)$
> and $y - \bar{y} = (-1, 2, -1)$.
> Now I square the x-deviations: $(-3)^2 + 0^2 + 3^2 = 9 + 0 + 9 = 18$.
> Divide by $n - 1 = 2$: covariance $= 18 / 2 = 9$."

```{webr}
#| autorun: false
# p11-starter — copy this to check the corrected answer
x <- c(3, 6, 9)
y <- c(2, 5, 2)
```

::: {.callout-note collapse="true" title="Answer"}
The student *squared the x-deviations* instead of *multiplying the paired
deviations from x and y*. They computed the **variance** of x (which is indeed 9),
not the **covariance** of x and y.

The covariance formula uses $(x_i - \bar{x})(y_i - \bar{y})$, not
$(x_i - \bar{x})^2$. The correct paired products are:

$(-3)(-1) = 3$, $\;(0)(2) = 0$, $\;(3)(-1) = -3$.

Sum: $3 + 0 + (-3) = 0$. Divide by $n - 1 = 2$: $s_{xy} = 0$.

The covariance is zero — there is no linear pattern between x and y here. The
student's answer of 9 was the variance of x, a completely different question.

```{r p11-answer}
x <- c(3, 6, 9); y <- c(2, 5, 2)
var(x)              # 9 — what the student actually computed (variance of x)
cov(x, y)           # 0 — what they were asked for (covariance of x and y)
```

This is the most common formula mix-up: variance squares one variable's deviations,
covariance multiplies two variables' deviations. Watch for $(x_i - \bar{x})^2$ vs
$(x_i - \bar{x})(y_i - \bar{y})$ — they look similar, but they answer
different questions.
:::

## Problem 12 — a gentle cliffhanger

Using the `mtcars` variables again: compute `cov(mtcars$wt, mtcars$mpg) /
var(mtcars$wt)`. Keep that number. Then run `coef(lm(mpg ~ wt, data = mtcars))` and
compare. (Explaining WHY those match is the whole point of the
[simple-linear-regression course](../../courses/simple-linear-regression/syllabus.md).)

```{webr}
#| autorun: false
# p12-starter — compute the ratio, then the regression slope, and compare
```

::: {.callout-note collapse="true" title="Answer"}
```{r p12-answer}
cov(mtcars$wt, mtcars$mpg) / var(mtcars$wt)   # -5.344
coef(lm(mpg ~ wt, data = mtcars))             # slope: -5.344
```

They're identical: the regression slope IS covariance divided by variance,
$\hat{\beta}_1 = s_{xy}/s_x^2$. You just computed your first regression coefficient
from raw ingredients — the course shows where that formula comes from and why it's
the "best" line.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/mean-variance-covariance" pass="4"}

::: {.quiz-q qid="mvc-mean" qtype="numeric" answer="5" tolerance="0"}
For $x = (2, 4, 9)$, what is the mean $\bar{x}$?
:::

::: {.quiz-q qid="mvc-variance-n-minus-1" qtype="numeric" answer="4" tolerance="0"}
For $x = (1, 3, 5)$ the mean is $3$ and the squared deviations are $4, 0, 4$, summing to
$8$. What is the sample variance $s^2$ (divide by $n - 1$)?
:::

::: {.quiz-q qid="mvc-divide-by-n-trap" qtype="mc" answer="b"}
A student divides the sum of squared deviations by $n$ instead of $n - 1$. What is the
consequence?

a) Nothing — $n$ and $n - 1$ give the same variance
b) The variance comes out too small; sample variance uses $n - 1$ (as R's `var()` does)
c) The variance comes out negative
:::

::: {.quiz-q qid="mvc-covariance-units" qtype="mc" answer="a"}
Rescaling $y$ from liters to milliliters multiplies the covariance by $1000$, though the
relationship is unchanged. What does this say about judging strength by covariance size?

a) Covariance size is unreliable; the sign is the trustworthy part
b) A bigger covariance always means a stronger relationship
c) The relationship genuinely became $1000$ times stronger
:::

::: {.quiz-q qid="mvc-sum-of-squares" qtype="numeric" answer="14" tolerance="0"}
(From vectors & summation.) The variance formula sums squared deviations. As a warm-up,
for $x = (1, 2, 3)$ what is $\sum_{i=1}^{3} x_i^2$?
:::

:::

::::
