---
title: "Practice: The mean, variance & covariance"
subtitle: "Answers are hidden — try honestly first, then click to check."
---

```{r setup}
#| message: false
# Packages this practice set uses. Missing ones are installed automatically.
needed <- c()
to_install <- setdiff(needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(needed, library, character.only = TRUE))
set.seed(42)
```

Work these in order — they get harder. Do them on paper or in an R session; only
then open the answer.

## Problem 1 — the toy example, new numbers

Four mushrooms weigh $x = (3, 1, 4, 2)$ grams. By hand: find $\bar{x}$, the four
deviations, and $s_x^2$. Check with `mean()` and `var()`.

```{r p1-starter}
# Starter — copy this to begin
x <- c(3, 1, 4, 2)
```

::: {.callout-note collapse="true" title="Answer"}
Mean: $3 + 1 + 4 + 2 = 10$, divided by 4 → $\bar{x} = 2.5$.

Deviations: $3 - 2.5 = 0.5$; $1 - 2.5 = -1.5$; $4 - 2.5 = 1.5$; $2 - 2.5 = -0.5$.
(Sanity check: $0.5 - 1.5 + 1.5 - 0.5 = 0$ ✓.)

Squared: $0.25, 2.25, 2.25, 0.25$. Sum: $5$. Divide by $n - 1 = 3$:
$s_x^2 = 5/3 \approx 1.67$.

```{r p1-answer}
x <- c(3, 1, 4, 2)
mean(x)          # 2.5
x - mean(x)      # 0.5 -1.5  1.5 -0.5
var(x)           # 1.6667 = 5/3
```
:::

## Problem 2 — faded worked example (covariance by hand)

Three plots got $x = (1, 3, 5)$ units of fertilizer and yielded $y = (2, 2, 8)$
crates. Below is the covariance worked out — but the **core step is blanked**. Fill
in ONLY the blanked line (the paired-deviation products and their sum), then finish
the last line.

> $\bar{x} = 3$, $\bar{y} = 4$.
> Deviations: $x - \bar{x} = (-2,\, 0,\, 2)$ and $y - \bar{y} = (-2,\, -2,\, 4)$.
> **`<your step here>`** — multiply each *pair* of deviations, then add the products:
> $\sum (x_i - \bar{x})(y_i - \bar{y}) = \underline{\phantom{00}}$
> Finally divide by $n - 1 = 2$: $\;s_{xy} = \underline{\phantom{00}}$.

```{r p2-starter}
# Starter — copy this to check your two blanks
x <- c(1, 3, 5)
y <- c(2, 2, 8)
```

::: {.callout-note collapse="true" title="Answer"}
The blanked step is the heart of covariance — pair up each plot's two deviations and
multiply: $(-2)(-2) = 4$, $(0)(-2) = 0$, $(2)(4) = 8$. Sum: $4 + 0 + 8 = 12$.

Divide by $n - 1 = 2$: $s_{xy} = 12 / 2 = 6$. Positive — more-fertilized plots tend
to be the higher-yielding ones.

```{r p2-answer}
x <- c(1, 3, 5); y <- c(2, 2, 8)
n <- length(x)
sum((x - mean(x)) * (y - mean(y))) / (n - 1)   # 6
cov(x, y)                                       # 6 — same
```
:::

## Problem 3 — greek vs. latin

A paper writes: "heights are distributed with mean $\mu = 170$ and variance
$\sigma^2 = 81$; in our sample of $n = 40$ we observed $\bar{x} = 172.1$ and
$s^2 = 76.4$." In plain English: what is the difference between $\mu$ and $\bar{x}$
here, and between $\sigma^2$ and $s^2$? Which pair would change if they re-ran the
study on 40 new people?

::: {.callout-note collapse="true" title="Answer"}
$\mu$ and $\sigma^2$ (Greek) are the *true* mean and variance of the whole
population — fixed but unknown quantities the paper is assuming or estimating.
$\bar{x}$ and $s^2$ (Latin) are *computed from the 40 sampled people* — estimates of
$\mu$ and $\sigma^2$.

A new sample of 40 people would give new values of $\bar{x}$ and $s^2$ (they bounce
around from sample to sample), while $\mu$ and $\sigma^2$ would stay exactly what
they were. The Greek-vs-Latin letter code is how papers silently tell you which kind
of quantity each symbol is.
:::

## Problem 4 — sign of the covariance, no arithmetic

For each pair, say whether the covariance should be positive, near zero, or
negative — *before* computing anything:

a. hours studied & exam score
b. car weight & fuel efficiency (miles per gallon)
c. shoe size & R programming skill

Then verify (b) in R with the built-in `mtcars` data (`wt` and `mpg`).

::: {.callout-note collapse="true" title="Answer"}
a. Positive — above-average studiers tend to score above average (deviations share
   a sign, products positive).
b. Negative — heavier-than-average cars tend to have below-average mpg.
c. Near zero — no reason for the deviations to align.

```{r p4-answer}
cov(mtcars$wt, mtcars$mpg)   # about -5.1: negative, as predicted
cor(mtcars$wt, mtcars$mpg)   # about -0.87: and strongly so
```

Note how the correlation (unit-free, in [−1, 1]) is much easier to interpret than
the raw covariance (whose units are tons·mpg!).
:::

## Problem 5 — spot the error

A colleague computes the variance of $x = (4, 8, 12)$ like this. It contains
**exactly one mistake** — find it, then give the correct $s_x^2$.

> "Mean is $(4 + 8 + 12)/3 = 8$. Deviations $-4, 0, 4$; squared $16, 0, 16$; they
> sum to $32$. There are $n = 3$ values, so the variance is $32 / 3 \approx 10.67$."

```{r p5-starter}
# Starter — copy this to check the corrected answer
x <- c(4, 8, 12)
```

::: {.callout-note collapse="true" title="Answer"}
Everything is right up to the last step. The mistake is dividing by $n = 3$ instead
of $n - 1 = 2$: **sample** variance uses $n - 1$. The correct value is
$32 / 2 = 16$, not $10.67$. (Dividing by $n$ would be the *population* variance,
which assumes you have the entire population, not a sample.)

```{r p5-answer}
x <- c(4, 8, 12)
n <- length(x)
sum((x - mean(x))^2) / (n - 1)   # 16 — the correct sample variance
var(x)                            # 16 — R agrees (it uses n - 1)
```
:::

## Problem 6 — the formula IS the function (in R)

Starter below. Compute the covariance of `x` and `y` **without** `cov()` — build it
straight from the formula $s_{xy} = \frac{1}{n-1}\sum (x_i - \bar{x})(y_i - \bar{y})$
— then confirm against `cov(x, y)`.

```{r p6-starter}
# Starter data — your job: covariance by hand, then check with cov()
set.seed(42)
x <- rnorm(500, mean = 10, sd = 2)
y <- 3 * x + rnorm(500, sd = 5)
```

::: {.callout-note collapse="true" title="Answer"}
```{r p6-answer}
set.seed(42)
x <- rnorm(500, mean = 10, sd = 2)
y <- 3 * x + rnorm(500, sd = 5)
n <- length(x)

# Straight from the formula:
# each x-deviation times its paired y-deviation, summed, divided by n - 1
by_hand <- sum((x - mean(x)) * (y - mean(y))) / (n - 1)

c(by_hand = by_hand, built_in = cov(x, y))   # identical
```

If you can write this line, you can *read* the covariance formula in any paper —
they are the same statement in two languages.
:::

## Problem 7 — a gentle cliffhanger

Using the `mtcars` variables again: compute `cov(mtcars$wt, mtcars$mpg) /
var(mtcars$wt)`. Keep that number. Then run `coef(lm(mpg ~ wt, data = mtcars))` and
compare. (Explaining WHY those match is the whole point of the
[simple-linear-regression course](../../courses/simple-linear-regression/syllabus.md).)

::: {.callout-note collapse="true" title="Answer"}
```{r p7-answer}
cov(mtcars$wt, mtcars$mpg) / var(mtcars$wt)   # -5.344
coef(lm(mpg ~ wt, data = mtcars))             # slope: -5.344
```

They're identical: the regression slope IS covariance divided by variance,
$\hat{\beta}_1 = s_{xy}/s_x^2$. You just computed your first regression coefficient
from raw ingredients — the course shows where that formula comes from and why it's
the "best" line.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="foundations/mean-variance-covariance" pass="4"}

::: {.quiz-q qid="mvc-mean" qtype="numeric" answer="5" tolerance="0"}
For $x = (2, 4, 9)$, what is the mean $\bar{x}$?
:::

::: {.quiz-q qid="mvc-variance-n-minus-1" qtype="numeric" answer="4" tolerance="0"}
For $x = (1, 3, 5)$ the mean is $3$ and the squared deviations are $4, 0, 4$, summing to
$8$. What is the sample variance $s^2$ (divide by $n - 1$)?
:::

::: {.quiz-q qid="mvc-divide-by-n-trap" qtype="mc" answer="b"}
A student divides the sum of squared deviations by $n$ instead of $n - 1$. What is the
consequence?

a) Nothing — $n$ and $n - 1$ give the same variance
b) The variance comes out too small; sample variance uses $n - 1$ (as R's `var()` does)
c) The variance comes out negative
:::

::: {.quiz-q qid="mvc-covariance-units" qtype="mc" answer="a"}
Rescaling $y$ from liters to milliliters multiplies the covariance by $1000$, though the
relationship is unchanged. What does this say about judging strength by covariance size?

a) Size is unreliable — covariance carries the variables' units; the *sign* is the trustworthy part, and correlation gives a comparable size
b) A bigger covariance always means a stronger relationship
c) The relationship genuinely became $1000$ times stronger
:::

::: {.quiz-q qid="mvc-sum-of-squares" qtype="numeric" answer="14" tolerance="0"}
(From vectors & summation.) The variance formula sums squared deviations. As a warm-up,
for $x = (1, 2, 3)$ what is $\sum_{i=1}^{3} x_i^2$?
:::

:::

::::
