---
title: "Practice: The switching search"
subtitle: "Answers are hidden — try honestly first, then click to check."
engine: knitr
---

{{< include /_extensions/r-wasm/live/_knitr.qmd >}}

```{r setup}
#| include: false
set.seed(42)
```

Work these in order — they get harder. Do them on paper or in an R session; only then
open the answer.

## Problem 1 — pick the move, new numbers

The search is currently at $S = \{x_1, x_4\}$ with BIC 210. Its candidate moves score:

| Move | New subset | BIC |
|---|---|---:|
| add $x_2$ | $\{x_1, x_2, x_4\}$ | 208 |
| remove $x_4$ | $\{x_1\}$ | 213 |
| swap $x_4\to x_5$ | $\{x_1, x_5\}$ | 205 |

Which move does the search take, and does it then stop?

::: {.callout-note collapse="true" title="Answer"}
It takes the **lowest-BIC** move that beats the current 210: the swap $x_4\to x_5$ (BIC 205).
It does **not** stop — 205 < 210 is an improvement, so it moves to $\{x_1, x_5\}$ and looks for
another improving move from there. It stops only when *no* move beats the current BIC.
:::

## Problem 2 — fill in the missing step

Forward stepwise, starting empty, has BICs: add $x_2 \to -30$, then from $\{x_2\}$ add
$x_3 \to -58$, then from $\{x_2, x_3\}$ every add makes BIC worse, so it stops at
$\{x_2, x_3\}$. But $\{x_1, x_3\}$ scores $-61$.

- Step 1: forward stops at $\{x_2, x_3\}$ (BIC $-58$) because it can only *add*, and no add helps.
- Step 2: **your step here** — what single move, unavailable to forward, reaches $\{x_1, x_3\}$ from a set forward could hold?
- Step 3: the switching search takes it, ending at $\{x_1, x_3\}$ (BIC $-61$).

::: {.callout-note collapse="true" title="Answer"}
Step 2: a **swap** — from $\{x_2, x_3\}$, swap $x_2 \to x_1$ to get $\{x_1, x_3\}$ (or,
if the search had first added $x_1$ to reach $\{x_1, x_2, x_3\}$, a **remove** of $x_2$).
Either way it's a move forward stepwise doesn't have, which is why forward is stuck at the
worse-BIC $\{x_2, x_3\}$.
:::

## Problem 3 — forward vs switching, in R

On the 6-indicator panel below (seed 8), run forward-only and the switching search and compare
their selected subsets and BICs.

```{webr}
#| autorun: false
# p3-starter — the searches are provided; just run them and compare
# PREDICT: will forward-only stop at a different (worse-BIC) subset than switching?
sw_bic <- function(R,b,n,cols){ if(!length(cols)) return(0)
  r2 <- min(as.numeric(crossprod(b[cols], solve(R[cols,cols,drop=FALSE], b[cols]))), 1 - 1e-9)
  n*log(1-r2)+length(cols)*log(n) }
forward_only <- function(R,b,n){p<-length(b);cur<-integer(0)
  repeat{cd<-setdiff(1:p,cur);if(!length(cd))break
    bx<-sapply(cd,function(j)sw_bic(R,b,n,sort(c(cur,j))))
    if(min(bx)<sw_bic(R,b,n,cur)-1e-9)cur<-sort(c(cur,cd[which.min(bx)]))else break};sort(cur)}
switching<-function(R,b,n){p<-length(b);cur<-integer(0);cb<-sw_bic(R,b,n,cur)
  repeat{mv<-list();for(j in setdiff(1:p,cur))mv<-c(mv,list(sort(c(cur,j))))
    for(j in cur)mv<-c(mv,list(setdiff(cur,j)))
    for(j in cur)for(k in setdiff(1:p,cur))mv<-c(mv,list(sort(c(setdiff(cur,j),k))))
    if(!length(mv))break;bx<-sapply(mv,function(m)sw_bic(R,b,n,m));i<-which.min(bx)
    if(bx[i]<cb-1e-9){cur<-mv[[i]];cb<-bx[i]}else break};sort(cur)}
set.seed(8); n<-200; p<-6; Rt<-diag(p); Rt[1,2]<-Rt[2,1]<-.85; Rt[3,4]<-Rt[4,3]<-.6
X<-matrix(rnorm(n*p),n,p)%*%chol(Rt); y<-drop(X%*%c(.3,.45,.5,0,0,0)+rnorm(n,sd=1.2))
R<-cor(X); b<-as.numeric(cor(X,y))
```

::: {.callout-note collapse="true" title="Answer"}
Forward-only stops at `{2,3}`; the switching search reaches `{1,3}` with lower BIC — `x1` and
`x2` are 0.85-correlated, so once the search can swap the marginally-strong-but-redundant `x2`
for `x1`, BIC improves.

```{r}
sw_bic <- function(R,b,n,cols){ if(!length(cols)) return(0)
  r2 <- min(as.numeric(crossprod(b[cols], solve(R[cols,cols,drop=FALSE], b[cols]))), 1 - 1e-9)
  n*log(1-r2)+length(cols)*log(n) }
forward_only <- function(R,b,n){p<-length(b);cur<-integer(0)
  repeat{cd<-setdiff(1:p,cur);if(!length(cd))break
    bx<-sapply(cd,function(j)sw_bic(R,b,n,sort(c(cur,j))))
    if(min(bx)<sw_bic(R,b,n,cur)-1e-9)cur<-sort(c(cur,cd[which.min(bx)]))else break};sort(cur)}
switching<-function(R,b,n){p<-length(b);cur<-integer(0);cb<-sw_bic(R,b,n,cur)
  repeat{mv<-list();for(j in setdiff(1:p,cur))mv<-c(mv,list(sort(c(cur,j))))
    for(j in cur)mv<-c(mv,list(setdiff(cur,j)))
    for(j in cur)for(k in setdiff(1:p,cur))mv<-c(mv,list(sort(c(setdiff(cur,j),k))))
    if(!length(mv))break;bx<-sapply(mv,function(m)sw_bic(R,b,n,m));i<-which.min(bx)
    if(bx[i]<cb-1e-9){cur<-mv[[i]];cb<-bx[i]}else break};sort(cur)}
set.seed(8); n<-200; p<-6; Rt<-diag(p); Rt[1,2]<-Rt[2,1]<-.85; Rt[3,4]<-Rt[4,3]<-.6
X<-matrix(rnorm(n*p),n,p)%*%chol(Rt); y<-drop(X%*%c(.3,.45,.5,0,0,0)+rnorm(n,sd=1.2))
R<-cor(X); b<-as.numeric(cor(X,y))
list(forward = forward_only(R,b,n), switching = switching(R,b,n),
     bic_fwd = round(sw_bic(R,b,n,forward_only(R,b,n)),2),
     bic_sw  = round(sw_bic(R,b,n,switching(R,b,n)),2))
```
:::

## Problem 4 — spot the error

After running the switching search, a colleague reports: *"`cpi_infl` was selected and its
coefficient's p-value is 0.001, so it's very significantly related to GDP."* This contains
**exactly one mistake** — find it.

::: {.callout-note collapse="true" title="Answer"}
The mistake: quoting a naïve p-value **for a coefficient in a model you selected by
searching**. Because the search cherry-picked the best-looking indicators, the usual p-values
are optimistically small (post-selection inference) — 0.001 overstates the evidence. That's
exactly why the method stops with BIC rather than a p-value threshold; honest post-selection
p-values need special methods.
:::

## Problem 5 — mixing it with Module 1 *(uses the summary-stat BIC)*

Using Module 1's formula $\mathrm{BIC}(S) = n\log(1 - b_S^\top R_S^{-1}b_S) + |S|\log n$,
compute the BIC of the single-indicator model $S = \{1\}$ when $b_1 = 0.6$ (so $R_S = 1$) and
$n = 100$.

```{webr}
#| autorun: false
# p5-starter
b1 <- 0.6; n <- 100
```

::: {.callout-note collapse="true" title="Answer"}
For a single indicator, $R_S = 1$, so $b_S^\top R_S^{-1} b_S = b_1^2 = 0.36$. Then
$\mathrm{BIC} = 100\log(1 - 0.36) + 1\cdot\log(100) = 100\log(0.64) + 4.605 = 100(-0.446) + 4.605
= -44.63 + 4.605 = -40.03$. (Up to the additive constant common to all models on this data.)

```{r}
b1 <- 0.6; n <- 100
n * log(1 - b1^2) + 1 * log(n)      # about -40.03
```
:::

## Problem 6 — transfer: is the switch enough?

For $p = 12$ indicators there are $2^{12} = 4096$ subsets — small enough to check *all* of them
(exhaustive best-subset). For $p = 40$ there are over $10^{12}$. In one or two sentences: why do
we validate the switching search against exhaustive best-subset only for small $p$, and what do
we rely on for large $p$?

::: {.callout-note collapse="true" title="Answer"}
For small $p$ we can afford to enumerate every subset, so we can *check* that the switching
search lands on the true global-best BIC subset (it did, on every tested seed). For large $p$
exhaustive search is computationally impossible, so we rely on the switching search as a
heuristic — trusting the small-$p$ validation as evidence it behaves well, while accepting it's
no longer guaranteed to hit the global optimum.
:::

## Module check

:::: {.callout-tip collapse="true" title="⏳ Wait at least a day before taking this quiz"}

This is a checkpoint, not more practice — come back to it **at least a day from now**,
once the material isn't fresh. A pass then is real evidence you've retained it. Five
questions, graded on the page; 4/5 or better means this module is solid. Score under
4? Revisit what the missed questions point at, then retake in a **later session** —
passing after another gap is the evidence that counts. Results are saved in this
browser only.

::: {.module-check module="courses/swap-stepwise-selection/modules/02-the-switching-search" pass="4"}

::: {.quiz-q qid="sws-which-move" qtype="mc" answer="b"}
Forward-only selection stops at $\{x_1, x_2, x_3\}$, but $\{x_1, x_3\}$ scores a lower (better)
BIC. Which move — one forward-only lacks — lets the switching search reach it?

a) Add — include another predictor
b) Remove — drop $x_2$
c) None; forward-only could reach it too
:::

::: {.quiz-q qid="sws-why-bic" qtype="mc" answer="a"}
Why does the switching search stop using BIC rather than "keep every predictor with
$p < 0.05$"?

a) Searching over many models makes the usual p-values dishonestly small
b) BIC is faster to compute than a p-value
c) p-values cannot be computed from summary statistics
:::

::: {.quiz-q qid="sws-best-bic" qtype="numeric" answer="-61.5" tolerance="0.1"}
Four candidate subsets score BIC $-31.6$, $-58.0$, $-58.3$, and $-61.5$ (lower is better). What
BIC value should the search finish at?
:::

::: {.quiz-q qid="sws-not-guaranteed" qtype="mc" answer="b"}
The switching search matched the exhaustive best-subset answer on every seed tested here. Does
that guarantee it always finds the global-best subset?

a) Yes — matching the oracle on those tests proves it always will
b) No — it's a greedy search that can stop at a local optimum
c) Yes — the swap move guarantees the global optimum
:::

::: {.quiz-q qid="sws-r2-term" qtype="mc" answer="c"}
(From Module 1.) A subset's summary-stat BIC is $n\log(1 - b_S^\top R_S^{-1}b_S) + |S|\log n$.
What does the quantity $b_S^\top R_S^{-1}b_S$ represent?

a) The number of predictors in the subset
b) The subset's residual (unexplained) variance
c) The subset's $R^2$ (variance explained)
:::

:::

::::
